### 2022-11-08 proposal.tex
Linting Linear Core proposal draft

\documentclass{article}

\title{Linting Linearity in Core/System FC}
\author{Rodrigo Mesquita}

\begin{document}

\maketitle

\section{Introduction}

Since Linear Haskell's\ref{}
publication and implementation release in GHC 9.2, Haskell's type system
supports linearity annotations in functions -- bringing linear types into a
mainstream pure and lazy functional language.

System FC is the formal system in which the implementation of GHC's intermediate
representation language \emph{Core} is based on.

There are at least two distinct typecheckers in core. The first is run on the frontend
language, i.e. the Haskell we write, and is a big and complex typechecker. The
second is run on the intermediate language \emph{Core} that we obtain from
desugaring Haskell.

\emph{Core} is a much smaller and more principled language than the whole of
Haskell (even though we can compile the whole of Haskell to it), and the
typechecker for it is small and fast due to \emph{Core} being explicitly typed
and having a very small abstract syntax tree. This typechecker is called
\emph{Lint} and gives us guarantees of correctness (i.e. sanity) in face of
the complexity of all the transformations a Haskell program undergoes, such as
type inference, desugaring and optimising transformations (and other Core
related passes?).

The addition of linear types comes at two levels: frontend and Core.
What's the first and foremost reason for permeating Core with linearity? I can't
phrase it correctly, something wrt to the foundational formal design and system FC.
In the frontend, we can now annotate with linearity type declarations and
typecheck programs that use them accordingly (a linear fuction will consume its
argument exactly once if it is consumed exactly once, for some definition of
\emph{consume} that I should revise here). In Core, we still have linearity
annotations and ideally \emph{Lint} would check that all the GHC transformations
to our linear program preserved its linearity. However, \textbf{it can't}!

Despite the strong formal foundations of linear types driving the
implementation, their interaction with the whole of GHC is still far from
trivial. Diverse problems with linearity spring up when we get past the
desugarer. In particular, optimizing transformations, coercions from GADTs and
type families, recursive lets, and empty case expressions don't currently fit in
with linearity.

We believe that GHC's transformations are correct, and it is the linear type
system that can't accommodate the resulting programs. We aim to formalise a type
system that is able to type check \emph{Core}/\emph{System FC} at all points in
the core pipeline and implement it into GHC to be able to preserve linearity
accross the stages and to enable \emph{Lint} to preserve our sanity regarding
linearity.

\section{Motivation}

\begin{itemize}
    \item Are we really preserving linearity?
    \item Inform/unlock other optimisations that take into account linearity
\end{itemize}

\section{Typing Usage Environments}

The first set of problems appears in the core-to-core optimisation passes. GHC
applies many optimising transformations to \emph{Core} and we believe those
transformations preserve linearity. However, our linear type system cannot check
that they indeed preserve linearity.

The simpler examples come from straightforward and common optimising
transformations. Then we have recursive let definitions that don't accommodate
linearity even though it might converge to only use the value once. Finally, we
have the empty case expression introduced with the \emph{EmptyCase} language
extension that we currently can't typecheck either.

\begin{itemize}
    \item Transformations, occurrences?
    \item Recursive lets
    \item Empty case expression
\end{itemize}

\section{Multiplicity Coercions}

The second set of problems arises from our inability to coerce a multiplicity
into another (or say that one is submultiple of another?).

When we pattern match on a GADT we ...

Taking this example (copy example over) we can see that we don't know how to say
that x is indeed linear in one case and unrestricted in the other, even though
it is according to its type. We'd need some sort of coercion to coerce through
the multiplicity to the new one we uncover when we pattern match on the GADT
evidence (...)

\begin{itemize}
    \item Are we really preserving linearity?
    \item Inform/unlock other optimisations that take into account linearity
\end{itemize}

\end{document}

### 2022-11-08 proposal.tex
Draft usage environment section

@@
 \documentclass{article}
+
+\usepackage{fancyvrb}
+\DefineVerbatimEnvironment{code}{Verbatim}{fontsize=\small}
+\DefineVerbatimEnvironment{example}{Verbatim}{fontsize=\small}
 
@@
 
+\begin{itemize}
+    \item In GHC, the type Type is shared between the surface language and Core.
+        So modifying the types implies that Core is modified as well.
+
+    \item More importantly, because Core is a good check that our implementation
+        works as intended. That is, a linearly typed Core will ensure that
+        linearly-typed programs are indeed desugared to linearly-typed programs
+        in Core. And that optimisations do not destroy linearity.
+\end{itemize}
+
 The addition of linear types comes at two levels: frontend and Core.
-What's the first and foremost reason for permeating Core with linearity? I can't
-phrase it correctly, something wrt to the foundational formal design and system FC.
 In the frontend, we can now annotate with linearity type declarations and
@@
 
+% Our key innovation is that, by recognising join points as a language construct,
+% we both preserve join points through subsequent transformations and exploit them
+% to make those transformations more effective. Next, we formalize this ap-
+% proach; subsequent sections develop the consequences.
+
 \section{Typing Usage Environments}
@@
 
+A usage environment is a mapping from variables to multiplicities.
+
+The key idea is to annotate every variable with either its multiplicity, if it's
+lambda bound, or with its usage environment, if it's let bound.
+
+For example, if y and z are linear, in the following code it might not look as
+if y and z were both being consumed linearly, but indeed they are since in the
+first branch we use x which means using y and z linearly, and we use y and z
+directly on the second branch. Note that let binding x doesn't consume y and z,
+only using x itself does, because of laziness.
+\begin{code}
+let x = (y, z) in
+case e of
+  Pat1 -> … x …
+  Pat2 -> … y … z …
+\end{code}
+
+If we annotate the x bound by let with a usage environment $\delta$ mapping all (free?)
+variables in its binder to a multiplicity ($\delta = [y := 1, z := 1]$), we
+could, upon finding x, simply emit that y and z are consumed once. When typing
+the second branch we'd also see that y and z are used exactly once, hence the
+linearity in the two branches match.
+
+Currently, in GHC, we don't annotate let-bound variables with a usage
+environment, but we already calculate a usage environment and use it to check
+some things (which things?)
+
 \begin{itemize}
@@
 \end{itemize}
+
+\subsection{Join Points}
+
+When duplicating a case (in the case-of-case transformation), to avoid code
+explosion, the branches of the case are first made into join points
+
+\begin{code}
+case e of
+  Pat1 -> u
+  Pat2 -> v
+~~>
+let j1 = u in
+let j2 = v in
+case e of
+  Pat1 -> j1
+  Pat2 -> j2
+\end{code}
+
+If there is any linear variable in u and v, then the standard
+let rule above will fail (since j1 occurs only in one branch, and
+so does j2).
+
+However, if j1 and j2 were annotated with their usage environment,
 

### 2022-11-08 proposal.tex
Add more examples

@@
-\documentclass{article}
+\documentclass[10pt, a4paper, draft]{article}
 
@@
 \begin{document}
-
 \maketitle
@@
 
+\subsection{Recursive Lets}
+
+Optimisations can create a letrec which uses a variable linearly. The following
+example uses 'x' linearly, but this is not seen by the linter.
+\begin{code}
+
+letrec f True = f False
+       f False = x
+in f True
+\end{code}
+
+Plan: make let-bound variables remember the usage environment.
+See ticket #18694.
+
+\subsection{Empty Case}
+
+\begin{itemize}
+\item https://gitlab.haskell.org/ghc/ghc/-/issues/20058
+\item https://gitlab.haskell.org/ghc/ghc/-/issues/18768
+\end{itemize}
+
+\begin{code}
+{-# LANGUAGE LinearTypes, EmptyCase #-}
+module M where
+
+{-# NOINLINE f #-}
+f :: a %1-> ()
+f x = case () of {}
+\end{code}
+
 \section{Multiplicity Coercions}
@@
 
+\begin{code}
+data SBool :: Bool -> Type where
+  STrue :: SBool True
+  SFalse :: SBool False
+
+type family If b t f where
+  If True t _ = t
+  If False _ f = f
+
+dep :: SBool b -> Int %(If b One Many) -> Int
+dep STrue x = x
+dep SFalse _ = 0
+\end{code}
+
 \end{document}

### 2022-11-08 proposal.tex
Initial idea for let-recs

@@
 
-letrec f True = f False
-       f False = x
+letrec f = \case
+        True -> f False
+        False -> x
 in f True
@@
 
-Plan: make let-bound variables remember the usage environment.
-See ticket #18694.
+Following the idea of making let-bound variables remember the usage environment
+here's an informal description that that example is well typed. We want to
+anotate the let-bound variable $f$ with a usage environment delta $\delta$, and
+use the binding body to compute it.
+
+Now, how to compute the usage environment of a case expression? It's described
+here if I'm not mistaken
+https://gitlab.haskell.org/ghc/ghc/-/wikis/uploads/355cd9a03291a852a518b0cb42f960b4/minicore.pdf.
+
+However, not understanding it, my take would be that we can compute the usage
+environment of every branch of the case expression and make sure that they all
+unify (wrt to submultiplicities). The special case is when the branch happens to
+call $f$ itself while computing the usage environment of $f$. If it were another
+let bound variable we'd add its own usage environment to the one we're
+computing; if it were a lambda bound variable we'd add [itself $:=$ its
+multiplicity].
+
+My idea is that we can emit a special usage [$rec := p$], which, when unified
+against the other case branches $\delta$ will always succeed with a unification
+mapping from $rec \rightarrow p\delta$ scaled by the multiplicity of $rec$
+
+So taking the example, to compute the usage environment of $f$, we'd compute for
+the second branch $[x := 1]$ and for the first branch $[rec := 1]$. Then, we'd
+unify them with $rec \rightarrow [x := 1*1]$, and somehow result in $[x := 1]$
+
+An example which should break linearity because $x$ is not linear in the first
+branch:
+\begin{code}
+letrec f = \case
+         True -> f False + f False
+         False -> x
+    in f True
+\end{code}
+
+To compute the usage environment of $f$ we take the second branch usage
+environment $[x := 1]$ and the first branch usage $[rec := 1+1]$ and unify them,
+somehow resulting in $[x := 2]$. Now, to lint the linearity in the whole let
+expression we must ensure that the body of the let uses $x$ linearly. The body
+is $f True$, which is a let-bound variable (how to distinguish functions that
+must be applied vs variables) so we take its usage environment $[x := 2]$ which
+does not use $x$ linearly and thus breaks linearity.
 

### 2022-11-11 proposal.tex
Add some comments
Expand letrec solution

@@
-\documentclass[10pt, a4paper, draft]{article}
+\documentclass[a4paper, draft]{article}
 
@@
 \section{Introduction}
+
+\section{Background}
+
+\section{Technical Details}
 
@@
 
+What do we currently do?
+
+When we lint a core expression, we get both its type and its usage environment.
+That means that to lint linearity in an expression, whenever we come across a
+free variable we compute its usage environment and take it into account
+
 \subsection{Join Points}
@@
 \begin{code}
-
 letrec f = \case
@@
 
+\textbf{Re-explained:} to compute the usage environment of the recursive
+let-bound function $f$ when applied, we compute $Z$ such that for all branch
+alternatives $U,V,...$, $U \subset Z$, $V \subset Z$ and so on by tracking the
+multiplicities and usage environments of variables that show up in the body and
+by emitting a special keyword $rec$ everytime we find a saturated call of $f$
+(how to handle unsaturated calls?) (how to have a more general solution that
+doesn't require a keyword, perhaps something to do with fixed points); Then we
+scale $Z \setminus \{(rec,_)\}$ by $Z[rec]$ and get $T$ which is the usage
+environment of $f$ when saturated.
+
+\textbf{Example 1 revisited:}
+\begin{enumerate}
+    \item Take $U = \{rec := 1\}$
+    \item Take $V = \{x := 1\}$
+    \item Take $Z = \{x := 1, rec := 1\}$
+    \item Take $\pi = Z[rec] = 1$
+    \item Take $W = Z \setminus \{rec\} = \{x := 1\}$
+    \item Take $T = \pi W = \{x := \pi * 1\} = \{x := 1\}$
+    \item Linearity OK
+\end{enumerate}
+
+\textbf{Example 2 revisited:}
+\begin{enumerate}
+    \item Take $U = \{rec := 2\}$
+    \item Take $V = \{x := 1\}$
+    \item Take $Z = \{x := 1, rec := 1\}$
+    \item Take $\pi = Z[rec] = 2$
+    \item Take $W = Z \setminus \{rec\} = \{x := 1\}$
+    \item Take $T = \pi W = \{x := \pi * 1\} = \{x := 2\}$
+    \item Linearity not OK
+\end{enumerate}
+
 \subsection{Empty Case}
+
+For case expressions, the usage environment is computed by checking all branches
+and taking sup. However, this trick doesn't work when there are no branches.
 
@@
 \item https://gitlab.haskell.org/ghc/ghc/-/issues/18768
+
+\item (1) Just like a case expression remembers its type (Note [Why does Case have a
+'Type' field?] in Core.hs), it should remember its usage environment. This data
+should be verified by Lint.
+
+\item (2) Once this is done, we can remove the Bottom usage and the second field of
+UsageEnv. In this step, we have to infer the correct usage environment for empty
+case in the typechecker.
 \end{itemize}
@@
 \end{code}
+
+This example is well typed: a function is linear if it consumes its argument
+exactly once when it's consumed exactly once. It seems like the function isn't
+linear since it won't consume x because of the empty case, however, that also
+means f won't be consumed due to the same empty case, thus linearity is
+preserved.
+
+\begin{code}
+* In the case of empty types (see Note [Bottoming expressions]), say
+  data T
+we do NOT want to replace
+  case (x::T) of Bool {}   -->   error Bool "Inaccessible case"
+because x might raise an exception, and *that*'s what we want to see!
+(#6067 is an example.) To preserve semantics we'd have to say
+   x `seq` error Bool "Inaccessible case"
+but the 'seq' is just such a case, so we are back to square 1.
+\end{code}
+
+There are three different problems:
+
+\begin{itemize}
+\item castBottomExpr converts (case x :: T of {}) :: T to x.
+\item Worker/wrapper moves the empty case to a separate binding
+\item CorePrep eliminates empty case, just like point 1 (See -- Eliminate empty
+    case in GHC.CoreToStg.Prep
+\end{itemize}
+
+With castBottomExpr, we get the example above to
+\begin{code}
+    f = \ @a (x (%1) :: a) -> ()
+\end{code}
+And if we don't, 
+\begin{code}
+    f = \ @a (x (%1) :: a) -> case () of {}
+\end{code}
+And that supposedly if we had a usage environment in the case expression we
+could avoid the error. How is it typed without the transformation in face
+of the bottom? (Even knowing that theoretically it's because of divergence?)
 
@@
 evidence (...)
-
-\begin{itemize}
-    \item Are we really preserving linearity?
-    \item Inform/unlock other optimisations that take into account linearity
-\end{itemize}
 

### 2022-11-11 laboratory/EmptyCase.hs

@@ -1,0 +1,7 @@
+{-# LANGUAGE LinearTypes, EmptyCase #-}
+module M where
+
+{-# NOINLINE f #-}
+f :: a %1-> ()
+f x = case () of {}
+

### 2022-11-15 proposal.tex
Tweaks
Expand on transformations going wrong

@@
 
+\section{Motivation}
+
+\begin{itemize}
+    \item Are we really preserving linearity?
+    \item Inform/unlock other optimisations that take into account linearity
+\end{itemize}
+
 \section{Background}
+
+\begin{itemize}
+    \item Linear types
+    \item Linear Haskell
+    \item Core
+    \item Sistemas de inferência
+    \item GADTs e Coercions
+\end{itemize}
+
+\section{Related Work}
 
@@
 
-\section{Motivation}
-
-\begin{itemize}
-    \item Are we really preserving linearity?
-    \item Inform/unlock other optimisations that take into account linearity
-\end{itemize}
-
 % Our key innovation is that, by recognising join points as a language construct,
@@
 \begin{itemize}
-    \item Transformations, occurrences?
-    \item Recursive lets
+    \item occurrences? store usage environment in Ids (vars)
+    \item Recursive lets (can it be rewritten using fix)
     \item Empty case expression
@@
 
-\subsection{Join Points}
+\subsection{Inlining}
 
-When duplicating a case (in the case-of-case transformation), to avoid code
-explosion, the branches of the case are first made into join points
-
-\begin{code}
-case e of
-  Pat1 -> u
-  Pat2 -> v
-~~>
-let j1 = u in
-let j2 = v in
-case e of
-  Pat1 -> j1
-  Pat2 -> j2
-\end{code}
-
-If there is any linear variable in u and v, then the standard
-let rule above will fail (since j1 occurs only in one branch, and
-so does j2).
-
-However, if j1 and j2 were annotated with their usage environment,
+If we annotate the let bound variables with their usage and emit that usage when
+we come across those variables, we can solve the linearity issues with inlining.
 
@@
 
+\section{Typing After Transformations}
+
+\subsection{Result of desugar (before optimization)}
+
+This resulting program fails lint with \emph{Linearity failure in lambda: x
+'Many $\not\subseteq$ p}
+\begin{code}
+($!) :: forall {rep} a (b :: TYPE rep) p q. (a %p -> b) %q -> a %p -> b
+($!) f !x = f x
+~~>
+($!)
+  :: forall {rep :: RuntimeRep} a (b :: TYPE rep) (p :: Multiplicity)
+            (q :: Multiplicity).
+     (a %p -> b) %q -> a %p -> b
+($!)
+  = \ (@(rep :: RuntimeRep))
+      (@a)
+      (@(b :: TYPE rep))
+      (@(p :: Multiplicity))
+      (@(q :: Multiplicity))
+      (f :: a %p -> b)
+      (x :: a) ->
+      case x of x { __DEFAULT -> f x }
+\end{code}
+
+How are strictness annotations typed in the frontend? The issue is this program
+consumes to value once to force it, and then again to determine the return
+value.
+
+Work better the meaning of consumption (does it mean for a value to be reduced
+to WHNF?). Is consuming = forcing? Why is the above program linear?
+
+The resulting program from -ddump-simpl is
+\begin{code}
+($!)
+  :: forall {rep :: GHC.Types.RuntimeRep} a (b :: TYPE rep)
+            (p :: GHC.Types.Multiplicity) (q :: GHC.Types.Multiplicity).
+     (a %p -> b) %q -> a %p -> b
+($!)
+  = \ (@(rep :: GHC.Types.RuntimeRep))
+      (@a)
+      (@(b :: TYPE rep_aSJ))
+      (@(p :: GHC.Types.Multiplicity))
+      (@(q :: GHC.Types.Multiplicity))
+      (f :: a %p -> b)
+      (x :: a) ->
+      case x of y { __DEFAULT -> f_aDV y }
+\end{code}
+So in this program we use another name for the case binder, but it still stands
+for the resulting of evaluating to WHNF; how is it technically different?
+
+\subsection{In Result of TcGblEnv axioms}
+
+I don't understand this one yet. I would guess it has to do with multiplicity
+coercions. Incorrect incompatible branch: CoAxBranch (src/Prelude/Linear/GenericUtil.hs:112:3-51): [S1
+\begin{code}
+type family Fixup (f :: Type -> Type) (g :: Type -> Type) :: Type -> Type where
+  Fixup (D1 c f) (D1 _c g) = D1 c (Fixup f g)
+  Fixup (C1 c f) (C1 _c g) = C1 c (Fixup f g)
+  Fixup (S1 c f) (S1 _c (MP1 m f)) = S1 c (MP1 m f) -- error in this constructor
+  Fixup (S1 c f) (S1 _c f) = S1 c f
+  Fixup (f :*: g) (f' :*: g') = Fixup f f' :*: Fixup g g'
+  Fixup (f :+: g) (f' :+: g') = Fixup f f' :+: Fixup g g'
+  Fixup V1 V1 = V1
+  Fixup _ _ = TypeError ('Text "FixupMetaData: representations do not match.")
+\end{code}
+
+\subsection{Common Sub-expression Elimination}
+
+Currently, the CSE seems to transform a linear program that pattern matches on
+constant and returns the same constant into a program that breaks linearity that
+pattern matches on the argument and returns the argument (where in the frontend
+a constant equal to the argument would be returned)
+
+Example
+\begin{code}
+(&&) :: Bool %1 -> Bool %1 -> Bool
+False && False = False
+False && True = False
+True && x = x
+~~>
+(&&) :: Bool %1 -> Bool %1 -> Bool
+(&&)
+  = \ (x :: Bool) (y :: Bool) ->
+      case x of w1 {
+        False -> case y of w2 { __DEFAULT -> x };
+        True -> y
+      }
+\end{code}
+In the first branch case, we pattern match on y to force it and then return $x$
+rather than the constant $False$.
+
+At first look, the resulting program is impossible to typecheck.
+
+Idea: if $x$ was anotated with some information regarding the CSE then perhaps
+it could be typechecked (in a system that considered said anotations)
+
+Confusingly: the resulting optimised program from running the complete simplifier
+and outputting with -ddump-simpl actually does what we expected it to regarding
+the constants and linearity. So is the issue from running linear lint after a
+particular CSE but it would be fine in the end?
+\begin{code}
+(&&) :: Bool %1 -> Bool %1 -> Bool
+(&&)
+  = \ (x :: Bool) (y :: Bool) ->
+      case x of {
+        False -> case y of { __DEFAULT -> GHC.Types.False };
+        True -> y
+      }
+\end{code}
+
+\subsection{Join Points}
+
+When duplicating a case (in the case-of-case transformation), to avoid code
+explosion, the branches of the case are first made into join points
+
+\begin{code}
+case e of
+  Pat1 -> u
+  Pat2 -> v
+~~>
+let j1 = u in
+let j2 = v in
+case e of
+  Pat1 -> j1
+  Pat2 -> j2
+\end{code}
+
+If there is any linear variable in u and v, then the standard
+let rule above will fail (since j1 occurs only in one branch, and
+so does j2).
+
+However, if j1 and j2 were annotated with their usage environment,
+
 \end{document}

### 2022-11-16 proposal.tex
case __DEFAULT error when building GHC

@@
 
+\usepackage{cmll}
+\usepackage{amssymb}
+\usepackage{amsmath}
+\usepackage{mathpartir}
+\usepackage{hyperref}
 \usepackage{fancyvrb}
@@
 
+Draft typing rule:
+
+\begin{mathparpagebreakable}
+    \infer*[right=(letrec)]
+    {\Gamma ; \Delta/ \Delta' ; \Omega \vdash M : A \Uparrow \and \Gamma ;
+    \Delta/ \Delta'' ; \Omega \vdash N : B \Uparrow \and \Delta' = \Delta''}
+    {\Gamma ; \Delta/\Delta' ; \Omega \vdash  (M \with N) : A
+    \with B \Uparrow}
+\end{mathparpagebreakable}
+
 \subsection{Empty Case}
@@
 
+\subsection{Compiling ghc with -dlinear-core-lint}
+
+From the definition of groupBy in GHC.Data.List.Infinite
+\begin{code}
+groupBy :: (a -> a -> Bool) -> Infinite a -> Infinite (NonEmpty a)
+groupBy eq = go
+  where
+    go (Inf a as) = Inf (a:|bs) (go cs)
+      where (bs, cs) = span (eq a) as
+\end{code}
+we get the following core which violates linearity.
+\begin{code}
+groupBy = \ (@a) (eq :: a -> a -> Bool) (eta :: Infinite a) ->
+  letrec {
+    go :: Infinite a -> Infinite (NonEmpty a)
+    go = \ (inf :: Infinite a) -> case inf of {
+      Inf x xs -> let {
+        ds :: ([a], Infinite a)
+        ds = let {
+            parteq :: a -> Bool
+            parteq = eq x
+        } in
+          letrec {
+            goZ :: Infinite a -> ([a], Infinite a)
+            goZ = \ (inf' :: Infinite a) -> case wgo inf' of { (# wA, wB #) -> (wA, wB) };
+
+            wgo :: Infinite a -> (# [a], Infinite a #)
+            wgo = \ (inf' :: Infinite a) -> case inf' of wildX2 {
+              Inf y ys ->
+                join {
+                    jp :: ([a], Infinite a) \%1 -> (# [a], Infinite a #)
+                    jp (wwj :: ([a], Infinite a)) =
+                        case wwj of wwj {
+                            DEFAULT -> case wwj of { (wA, wB) -> (# wA, wB #) } }
+                    } in
+                      case parteq y of {
+                        False ->
+                            let {
+                                ww :: ([a], Infinite a)
+                                ww = ([] @a, wildX2)
+                            } in jump jp ww;
+                        True ->
+                            let {
+                                dy :: ([a], Infinite a)
+                                dy = case wgo ys of { (# wA, wB #) -> (wA, wB)
+                                }
+                            } in
+                                let {
+                                    wwB :: [a]
+                                    wwB = case dy of { (bs, cs) -> bs }
+                                } in
+                                    let {
+                                        wwA :: [a]
+                                        wwA = : @a y wwB
+                                    } in let {
+                                        wwC :: Infinite a
+                                        wwC = case dy of { (bs, cs) -> cs }
+                                    } in let {
+                                        ww :: ([a], Infinite a)
+                                        ww = (wwA, wwC)
+                                    } in jump jp ww
+                                }
+              };
+          } in
+              case wgo xs of { (# wA, wB #) -> (wA, wB) }
+      } in
+          Inf @(NonEmpty a) (:| @a x (case ds of { (bs, cs) -> bs })) (case ds of { (bs, cs) -> go cs })
+    };
+  } in go eta
+\end{code}
+The issue is in the linear function that shows up in the core output (how does
+the linearity end up there?). The wwj variable is used once in the case
+expression, bound as the case binder, and used in the case body once again. Why
+do we do that instead of pattern matching right away? Seems a bit redundant.
+
+Observation: If the branch is DEFAULT, then the case binder binds the case
+scrutinee which was just forced?, but hasn't been actually consumed, because we
+haven't consumed its components. As long as the same branch doesn't consume the
+pattern matching result and the case binder at the same time it should be fine?
+
+In this example, we force wwj with the case binder, but we don't really
+consume it (more precise definition of consume...), so we can use the case
+binder meaning we're simply using x for the first time. Needs to be
+formalised....
+
+\begin{code}
+jp :: ([a], Infinite a) \%1 -> (# [a], Infinite a #)
+jp (wwj :: ([a], Infinite a)) =
+    case wwj of wwj {
+        DEFAULT -> case wwj of { (wA, wB) -> (# wA, wB #) }
+    }
+\end{code}
+
 \end{document}

### 2022-11-17 proposal.tex
Case binder unifying solution

@@
 Idea: if $x$ was anotated with some information regarding the CSE then perhaps
-it could be typechecked (in a system that considered said anotations)
+it could be typechecked (in a system that considered said anotations) -- This is
+too specific and the case binder environment solution is much cleaner. It just
+requires that the example is changed into using the case binder $w1$ rather than
+$x$.
+
+Solution with the (unifying) case binder idea:
+Annotate the case binder with its usage environment. For enums without arguments
+this usage environment will always be empty, meaning we can always use the case
+binder instead of the actual enum without issues. This solves this problem as
+well as a handful of other.
+
+In this exact example, the binder has usage environment empty ($w1 := []$),
+meaning that in the False branch, when we use $x$, if we instead used $w1$ which
+is equivalent and seems more correct (since it doesn't need the case scrutinee
+expression to be a variable), then the usage of $w1$ would imply the usage of
+$[]$ which is nothing and therefore we would preserve linearity
 
@@
 
+Solution with the case binder idea that unifies a different problems:
+The case binder is annotated with a usage environment $U_wwj = []$, and its
+usage is equal to using the usage environment
+
+Here it would mean that second \textbf{case of} $wwj$ doesn't actually use any
+variables and hence it isn't used twice.
+
+\subsection{Artificial examples}
+
+Following the difficulties in consuming or not consuming the case binder and its
+associated bound variables linearly we constructed some additional examples that
+bring out the problem quite clearly
+\begin{code}
+case x of w1
+    (x1,x2) -> case y:Bool of
+                DEFAULT -> (x1,x2)
+\end{code}
+Is equal to (note that the case binder is being used rather than the $x$ case
+scrutinee which could be an arbitrary expression and hence really cannot be
+consumed multiple times?)
+\begin{code}
+case x of w1
+    (x1,x2) -> case y:Bool of
+                DEFAULT -> w1
+\end{code}
+
+We initially wondered whether x was being consumed if the pattern match ignored
+some of its variables. We put that if it does have wildcards then it isn't
+consuming x fully
+\begin{code}
+case x of w1
+    (_,x2) -> Is x being consumed? no because not all of its components are
+    being consumed?
+\end{code}
+
+Can we have a solution that even handles weird cases in which we pattern match
+twice on the same expression but only on one constructor argument per time?
+\begin{code}
+case exp of w1
+    (x1,_) -> case w1 of w2
+                (_, x2) -> (x1, x2)
+\end{code}
+
 \end{document}

### 2022-11-18 proposal.tex
Fix another example using the case binder unifying solution
Rewrite CSE and Desugarer examples

@@
 
-\section{Motivation}
+\subsection{Motivation}
 
@@
 
-\subsection{Result of desugar (before optimization)}
-
-This resulting program fails lint with \emph{Linearity failure in lambda: x
-'Many $\not\subseteq$ p}
-\begin{code}
-($!) :: forall {rep} a (b :: TYPE rep) p q. (a %p -> b) %q -> a %p -> b
-($!) f !x = f x
-~~>
-($!)
-  :: forall {rep :: RuntimeRep} a (b :: TYPE rep) (p :: Multiplicity)
-            (q :: Multiplicity).
-     (a %p -> b) %q -> a %p -> b
-($!)
-  = \ (@(rep :: RuntimeRep))
-      (@a)
-      (@(b :: TYPE rep))
-      (@(p :: Multiplicity))
-      (@(q :: Multiplicity))
-      (f :: a %p -> b)
-      (x :: a) ->
-      case x of x { __DEFAULT -> f x }
-\end{code}
-
-How are strictness annotations typed in the frontend? The issue is this program
-consumes to value once to force it, and then again to determine the return
-value.
-
-Work better the meaning of consumption (does it mean for a value to be reduced
-to WHNF?). Is consuming = forcing? Why is the above program linear?
-
-The resulting program from -ddump-simpl is
-\begin{code}
-($!)
-  :: forall {rep :: GHC.Types.RuntimeRep} a (b :: TYPE rep)
-            (p :: GHC.Types.Multiplicity) (q :: GHC.Types.Multiplicity).
-     (a %p -> b) %q -> a %p -> b
-($!)
-  = \ (@(rep :: GHC.Types.RuntimeRep))
-      (@a)
-      (@(b :: TYPE rep_aSJ))
-      (@(p :: GHC.Types.Multiplicity))
-      (@(q :: GHC.Types.Multiplicity))
-      (f :: a %p -> b)
-      (x :: a) ->
-      case x of y { __DEFAULT -> f_aDV y }
-\end{code}
-So in this program we use another name for the case binder, but it still stands
-for the resulting of evaluating to WHNF; how is it technically different?
-
-\subsection{In Result of TcGblEnv axioms}
-
-I don't understand this one yet. I would guess it has to do with multiplicity
-coercions. Incorrect incompatible branch: CoAxBranch (src/Prelude/Linear/GenericUtil.hs:112:3-51): [S1
-\begin{code}
-type family Fixup (f :: Type -> Type) (g :: Type -> Type) :: Type -> Type where
-  Fixup (D1 c f) (D1 _c g) = D1 c (Fixup f g)
-  Fixup (C1 c f) (C1 _c g) = C1 c (Fixup f g)
-  Fixup (S1 c f) (S1 _c (MP1 m f)) = S1 c (MP1 m f) -- error in this constructor
-  Fixup (S1 c f) (S1 _c f) = S1 c f
-  Fixup (f :*: g) (f' :*: g') = Fixup f f' :*: Fixup g g'
-  Fixup (f :+: g) (f' :+: g') = Fixup f f' :+: Fixup g g'
-  Fixup V1 V1 = V1
-  Fixup _ _ = TypeError ('Text "FixupMetaData: representations do not match.")
-\end{code}
-
-\subsection{Common Sub-expression Elimination}
-
-Currently, the CSE seems to transform a linear program that pattern matches on
-constant and returns the same constant into a program that breaks linearity that
-pattern matches on the argument and returns the argument (where in the frontend
-a constant equal to the argument would be returned)
-
-Example
-\begin{code}
-(&&) :: Bool %1 -> Bool %1 -> Bool
-False && False = False
-False && True = False
-True && x = x
-~~>
-(&&) :: Bool %1 -> Bool %1 -> Bool
-(&&)
-  = \ (x :: Bool) (y :: Bool) ->
-      case x of w1 {
-        False -> case y of w2 { __DEFAULT -> x };
-        True -> y
-      }
-\end{code}
-In the first branch case, we pattern match on y to force it and then return $x$
-rather than the constant $False$.
-
-At first look, the resulting program is impossible to typecheck.
-
-Idea: if $x$ was anotated with some information regarding the CSE then perhaps
-it could be typechecked (in a system that considered said anotations) -- This is
-too specific and the case binder environment solution is much cleaner. It just
-requires that the example is changed into using the case binder $w1$ rather than
-$x$.
-
-Solution with the (unifying) case binder idea:
-Annotate the case binder with its usage environment. For enums without arguments
-this usage environment will always be empty, meaning we can always use the case
-binder instead of the actual enum without issues. This solves this problem as
-well as a handful of other.
-
-In this exact example, the binder has usage environment empty ($w1 := []$),
-meaning that in the False branch, when we use $x$, if we instead used $w1$ which
-is equivalent and seems more correct (since it doesn't need the case scrutinee
-expression to be a variable), then the usage of $w1$ would imply the usage of
-$[]$ which is nothing and therefore we would preserve linearity
-
-Confusingly: the resulting optimised program from running the complete simplifier
-and outputting with -ddump-simpl actually does what we expected it to regarding
-the constants and linearity. So is the issue from running linear lint after a
-particular CSE but it would be fine in the end?
-\begin{code}
-(&&) :: Bool %1 -> Bool %1 -> Bool
-(&&)
-  = \ (x :: Bool) (y :: Bool) ->
-      case x of {
-        False -> case y of { __DEFAULT -> GHC.Types.False };
-        True -> y
-      }
-\end{code}
-
 \subsection{Join Points}
@@
 However, if j1 and j2 were annotated with their usage environment,
+
+\section{Novel rules}
+
+\subsection{The usage environment in a Recursive Let}
+
+\subsection{The usage environment of a Case Binder\label{caseBinderKey}}
+
+\begin{mathparpagebreakable}
+    \infer*[right=(letrec)]
+    {\Gamma ; \Delta/ \Delta' ; \Omega \vdash M : A \Uparrow \and \Gamma ;
+    \Delta/ \Delta'' ; \Omega \vdash N : B \Uparrow \and \Delta' = \Delta''}
+    {\Gamma ; \Delta/\Delta' ; \Omega \vdash  (M \with N) : A \leadsto U}
+\end{mathparpagebreakable}
+
+\section{Examples}
+
+In this section we present worked examples of programs that currently fail to
+compile with \textbf{-dlinear-core-lint} but would succeed according to the
+novel typing rules we introduced above. Each example belongs to a subsection
+that indicates after which transformation the linting failed.
+
+\subsection{After the desugarer (before optimizations)}
+
+The definition for $\$!$ in \textbf{linear-base}\cite{} fails to lint after the
+desugarer (before any optimisation takes place) with \emph{Linearity failure in
+lambda: x 'Many $\not\subseteq$ p}
+\begin{code}
+($!) :: forall {rep} a (b :: TYPE rep) p q. (a %p -> b) %q -> a %p -> b
+($!) f !x = f x
+\end{code}
+
+The desugared version of that function follows below. It violates (naive?)
+linearity by using $x$ twice, once to force the value and a second time to call
+$f$. However, the $x$ passed as an argument is actually the case binder and it
+must be handled in its own way. The case binder is the key (as seen in ~\ref{caseBinderKey}) to solving this
+and many other examples.
+\begin{code}
+($!)
+  :: forall {rep :: RuntimeRep} a (b :: TYPE rep) (p :: Multiplicity)
+            (q :: Multiplicity).
+     (a %p -> b) %q -> a %p -> b
+($!)
+  = \ (@(rep :: RuntimeRep))
+      (@a)
+      (@(b :: TYPE rep))
+      (@(p :: Multiplicity))
+      (@(q :: Multiplicity))
+      (f :: a %p -> b)
+      (x :: a) ->
+      case x of x { __DEFAULT -> f x }
+\end{code}
+
+% Intuitively, this program is linear because forcing the polymorphic value to
+% WHNF ...? Need better semantics of consuming
+%
+The case binder usage rule typechecks this program because $x$ is consumed once,
+the usage environment of the case binder is empty ($x :_\emptyset a$) and,
+therefore, when $x$ is used in $f(x)$, we use its usage environment which is
+empty, so we don't use anything new.
+
+\emph{How are strictness annotations typed in the frontend? The issue is this program
+consumes to value once to force it, and then again to determine the return
+value.}
+
+% Work better the meaning of consumption (does it mean for a value to be reduced
+% to WHNF?). Is consuming = forcing? Why is the above program linear?
+
+As a finishing note on this example, we show the resulting program from
+\textbf{-ddump-simpl} that uses a different name for the case binder.
+\begin{code}
+($!)
+  :: forall {rep :: RuntimeRep} a (b :: TYPE rep)
+            (p :: Multiplicity) (q :: Multiplicity).
+     (a %p -> b) %q -> a %p -> b
+($!)
+  = \ (@(rep :: RuntimeRep))
+      (@a)
+      (@(b :: TYPE rep_aSJ))
+      (@(p :: Multiplicity))
+      (@(q :: Multiplicity))
+      (f :: a %p -> b)
+      (x :: a) ->
+      case x of y { __DEFAULT -> f_aDV y }
+\end{code}
+% So in this program we use another name for the case binder, but it still stands
+% for the resulting of evaluating to WHNF; how is it technically different?
+
+\subsection{Common Sub-expression Elimination}
+
+Currently, the CSE seems to transform a linear program that pattern matches on
+constant and returns the same constant into a program that breaks linearity that
+pattern matches on the argument and returns the argument (where in the frontend
+a constant equal to the argument would be returned)
+
+The definition for $\&\&$ in \textbf{linear-base} fails to lint after the common
+sub-expression elimination (CSE) pass transforms the program.
+\begin{code}
+(&&) :: Bool %1 -> Bool %1 -> Bool
+False && False = False
+False && True = False
+True && x = x
+\end{code}
+The issue with the program resulting from the transformation is $x$ being used
+twice in the first branch of the first case expression. We pattern match on y to
+force it (since it's a type without constructor arguments, forcing is consuming)
+and then return $x$ rather than the constant $False$
+\begin{code}
+(&&) :: Bool %1 -> Bool %1 -> Bool
+(&&) = \ (x :: Bool) (y :: Bool) ->
+  case x of w1 {
+    False -> case y of w2 { __DEFAULT -> x };
+    True -> y
+  }
+\end{code}
+At a first glance, the resulting program is impossible to typecheck linearly.
+
+The key observation is that after $x$ is forced into either $True$ or $False$,
+we know $x$ to be a constructor without arguments (which can be created
+without restrictions) and know that where we see $x$ we can as well have
+$True$ or $False$ depending on the branch. However, using $x$ here is very
+unsatisfactory (and linearly unsound?) because $x$ might be an expression, and
+we can't really associate $x$ to the value we pattern matched on (be it $True$
+or $False$). What we really want to have instead of $x$ is the $w1$ case binder --
+it relates directly to the value we pattern matched on, it's a variable rather
+than an expression, and, most importantly, our case-binder-usage idea could be
+applied here as well
+
+% Idea: if $x$ was anotated with some information regarding the CSE then perhaps
+% it could be typechecked (in a system that considered said anotations) -- This is
+% too specific and the case binder environment solution is much cleaner. It just
+% requires that the example is changed into using the case binder $w1$ rather than
+% $x$.
+
+The solution with the unifying case binder usage idea means annotating the case
+binder with its usage environment. For $True$ and $False$ (and other data
+constructors without arguments) the usage environment is always empty (using
+them doesn't entail using any other variable), meaning we can always use the
+case binder instead of the actual constant constructor without issues.
+%
+In concrete, if we had $w1$ instead of the second occurrence of $x$, we'd have
+an empty usage environment for $w1$ in the $False$ branch ($w1 :_\emptyset
+Bool$) and upon using $w1$ we wouldn't use any extra resources
+
+To make this work, we'd have to look at the current transformations and see
+whether it would be possible to ensure that CSEing the case scrutinee would
+entail using the case binder instead of the scrutinee. I don't know of a
+situation in which we'd really want the scrutinee rather than the case binder,
+so I hypothetize the modified transformation would work out in practice and
+solve this linearity issue.
+
+% In this exact example, the binder has usage environment empty ($w1 := []$),
+% meaning that in the False branch, when we use $x$, if we instead used $w1$ which
+% is equivalent and seems more correct (since it doesn't need the case scrutinee
+% expression to be a variable), then the usage of $w1$ would imply the usage of
+% $[]$ which is nothing and therefore we would preserve linearity
+
+Curiously, the optimised program resulting from running all transformations
+actually does what we expected it with regard to using the constant constructors
+and preserving linearity. So is the issue from running linear lint after a
+particular CSE but it would be fine in the end?
+\begin{code}
+(&&) :: Bool %1 -> Bool %1 -> Bool
+(&&) = \ (x :: Bool) (y :: Bool) ->
+  case x of {
+    False -> case y of { __DEFAULT -> GHC.Types.False };
+    True -> y
+  }
+\end{code}
 
@@
 
+\subsection{In Result of TcGblEnv axioms}
+
+I don't understand this one yet. I would guess it has to do with multiplicity
+coercions. Incorrect incompatible branch: CoAxBranch (src/Prelude/Linear/GenericUtil.hs:112:3-51):
+\begin{code}
+type family Fixup (f :: Type -> Type) (g :: Type -> Type) :: Type -> Type where
+  Fixup (D1 c f) (D1 _c g) = D1 c (Fixup f g)
+  Fixup (C1 c f) (C1 _c g) = C1 c (Fixup f g)
+  Fixup (S1 c f) (S1 _c (MP1 m f)) = S1 c (MP1 m f) -- error in this constructor
+  Fixup (S1 c f) (S1 _c f) = S1 c f
+  Fixup (f :*: g) (f' :*: g') = Fixup f f' :*: Fixup g g'
+  Fixup (f :+: g) (f' :+: g') = Fixup f f' :+: Fixup g g'
+  Fixup V1 V1 = V1
+  Fixup _ _ = TypeError ('Text "FixupMetaData: representations do not match.")
+\end{code}
+
 \subsection{Artificial examples}

### 2022-11-19 proposal.tex
Algorithm for usage environment of bindings in a letrec

@@
 \usepackage{mathpartir}
+\usepackage[ruled,vlined]{algorithm2e}
 \usepackage{hyperref}
@@
 
+The difficulty of calculating a usage environment for the recursive let recs
+increases with the number of recursive definitions in the same let.
+
+We devise an algorithm for computing the usage environment of a set of recursive
+definitions. The proof that the algorithm works follows...
+
+\begin{itemize}
+    \item Given a list of functions and their naive environment computed from
+        the body and including the recursive function names ($(f, F), (g, G),
+        (h, H)$ in which there might exist multiple occurrences of $f, g, h$ in $F, G, H$
+    \item For each bound rec var and its environment, update all bindings and
+        their usage as described in the algorithm
+    \item After iterating through all bound rec vars, all usage environments
+        will be free of recursive bind usages, and hence "final"
+\end{itemize}
+
+\begin{algorithm}
+$usageEnvs \gets naiveUsageEnvs.map(fst)$\;
+\For{$(bind, U) \in naiveUsageEnvs$}{
+    \For{$V \in usageEnvs$}{
+        $V \gets sup(V[bind]*U\setminus\{bind\}, V\setminus\{bind\})$
+    }
+}
+\caption{computeRecUsageEnvs}
+\end{algorithm}
+
+In haskell that would look like this;
+\begin{code}
+computeRecUsageEnvs :: [(Var, UsageEnv)] -- Recursive usage environments associated to their recursive call
+                    -> [(Var, UsageEnv)] -- Non-recursive usage environments
+computeRecUsageEnvs l =
+  foldl (flip $ \(v,vEnv) -> map $ \(n, nEnv) ->
+      (n, ((fromMaybe 0 $ v `M.lookup` nEnv) `scale` (M.delete v vEnv)) `sup` (M.delete v nEnv))) l l
+
+sup :: UsageEnv -> UsageEnv -> UsageEnv
+sup = M.merge M.preserveMissing M.preserveMissing (M.zipWithMatched $ \_ x y -> max x y)
+
+scale :: Mult -> UsageEnv -> UsageEnv
+scale m = M.map (*m)
+\end{code}
+
 \subsection{Empty Case}
@@
 
-\section{Typing After Transformations}
-
-\subsection{Join Points}
-
-When duplicating a case (in the case-of-case transformation), to avoid code
-explosion, the branches of the case are first made into join points
-
-\begin{code}
-case e of
-  Pat1 -> u
-  Pat2 -> v
-~~>
-let j1 = u in
-let j2 = v in
-case e of
-  Pat1 -> j1
-  Pat2 -> j2
-\end{code}
-
-If there is any linear variable in u and v, then the standard
-let rule above will fail (since j1 occurs only in one branch, and
-so does j2).
-
-However, if j1 and j2 were annotated with their usage environment,
-
 \section{Novel rules}
@@
 As a finishing note on this example, we show the resulting program from
-\textbf{-ddump-simpl} that uses a different name for the case binder.
+\textbf{-ddump-simpl} that simply uses a different name for the case binder.
 \begin{code}
@@
 
+A double let rec in which both use a linear bound variable $y$
+\begin{code}
+letrec f = \case
+        True  -> y
+        False -> g True
+       g = \case
+        True -> f True
+        False -> y
+    in f False
+\end{code}
+
+We have to compute the usage environment of f and g.
+For f, we have the first branch with usage environment $y$ and the second with
+usage $g$, meaning we have $F = \{g := 1, y := 1\}$
+For g, we have $G = \{f := 1, y := 1\}$. To calculate the usage environment of
+$f$ to emit upon $f False$, we calculate ...
+
+Refer to the algorithm for computing recursive environment
+
+\subsection{Join Points}
+
+When duplicating a case (in the case-of-case transformation), to avoid code
+explosion, the branches of the case are first made into join points
+
+\begin{code}
+case e of
+  Pat1 -> u
+  Pat2 -> v
+~~>
+let j1 = u in
+let j2 = v in
+case e of
+  Pat1 -> j1
+  Pat2 -> j2
+\end{code}
+
+If there is any linear variable in u and v, then the standard
+let rule above will fail (since j1 occurs only in one branch, and
+so does j2).
+
+However, if j1 and j2 were annotated with their usage environment,
+
 \end{document}

### 2022-11-19 laboratory/algorithms/UsageEnv.hs

@@ -1,0 +1,23 @@
+{-# LANGUAGE LambdaCase #-}
+module UsageEnv where
+
+import Data.Maybe
+import qualified Data.Map.Internal as M
+
+type Var  = String
+type Mult = Int
+type UsageEnv = M.Map Var Mult
+
+
+computeRecUsageEnvs :: [(Var, UsageEnv)] -- ^ Recursive usage environments associated to their recursive call
+                    -> [(Var, UsageEnv)] -- ^ Non-recursive usage environments
+computeRecUsageEnvs l =
+  foldl (flip $ \(v,vEnv) -> map $ \(n, nEnv) ->
+      (n, ((fromMaybe 0 $ v `M.lookup` nEnv) `scale` (M.delete v vEnv)) `sup` (M.delete v nEnv))) l l
+
+sup :: UsageEnv -> UsageEnv -> UsageEnv
+sup = M.merge M.preserveMissing M.preserveMissing (M.zipWithMatched $ \_ x y -> max x y)
+  
+scale :: Mult -> UsageEnv -> UsageEnv
+scale m = M.map (*m)
+

### 2022-11-20 proposal.tex
Letrec draft rule

@@
     \infer*[right=(letrec)]
-    {\Gamma ; \Delta/ \Delta' ; \Omega \vdash M : A \Uparrow \and \Gamma ;
-    \Delta/ \Delta'' ; \Omega \vdash N : B \Uparrow \and \Delta' = \Delta''}
-    {\Gamma ; \Delta/\Delta' ; \Omega \vdash  (M \with N) : A
-    \with B \Uparrow}
+    {\Gamma ; x_1 : A_1 \dots x_n : A_n \vdash t_i : A_i \leadsto \{U_{i_\text{naive}}\} \\
+     (U_1 \dots U_n) = computeRecUsages(U_{1_\text{naive}} \dots U_{n_\text{naive}}) \\
+     \Gamma ; x_1 :_{U_1} A_1 \dots x_n :_{U_n} A_n \vdash u : C \leadsto \{V\}}
+    {\Gamma \vdash \text{let } x_1 :_{U_1} A_1 = t_1 \dots x_n :_{U_n} A_n = t_n \text{ in } u : C \leadsto \{V\}}
 \end{mathparpagebreakable}
@@
 
+-- I should probably use the re-computed usageEnvs instead of the
+naiveUsageEnvs. I'm pretty sure it might fail in some inputs if I keep using the
+naiveUsageEnvs.
 \begin{algorithm}
@@
 }
-\caption{computeRecUsageEnvs}
+\caption{computeRecUsages}
 \end{algorithm}

### 2022-11-20 proposal.tex
Case binder usage based typing

@@
 computeRecUsageEnvs l =
-  foldl (flip $ \(v,vEnv) -> map $ \(n, nEnv) ->
-      (n, ((fromMaybe 0 $ v `M.lookup` nEnv) `scale` (M.delete v vEnv)) `sup` (M.delete v nEnv))) l l
+  foldl (flip \\(v,vEnv) -> map \\(n, nEnv) -> (n, ((fromMaybe 0 \$ v `M.lookup` nEnv) `scale` (M.delete v vEnv)) `sup` (M.delete v nEnv))) l l
 
 sup :: UsageEnv -> UsageEnv -> UsageEnv
-sup = M.merge M.preserveMissing M.preserveMissing (M.zipWithMatched $ \_ x y -> max x y)
+sup = M.merge M.preserveMissing M.preserveMissing (M.zipWithMatched \$ \_ x y -> max x y)
 
@@
 \end{code}
+
+\subsection{Handling the case binder}
+
+The current typing rule for case expressions describes that using the case
+binder is as using the case scrutinee multiple times (though it seems like it
+actually just starts counting after the second usage which would make the rule
+wrong for a case expression that doesn't use the binder -- the scrutinee should
+also be consumed), so we scale the usage of
+the case scrutinee by the superior multiplicity of using the case binder across
+all branches. However, we hypothetize that this prevents us from typing many
+valid programs and ultimately doesn't capture correctly the usage of variable,
+and present a seemingly viable alternative.
+
+Firstly, we remember the definition of \emph{consuming a value} from Linear
+Haskell as
+\begin{itemize}
+    \item consuming an atomic value is forcing it to Normal Form (NF)
+    \item consuming a value that is constructed with more than 1 argument is
+        consuming all of its arguments
+\end{itemize}
+and further note that when an expression is scrutinized by a case expression it
+is evaluated to Weak Head Normal Form (WHNF), which in the case of a nullary data
+constructor is equal to NF.
+
+Now, with the following example, my interpretation of the current rule says that
+the following case expression consumes the scrutinee twice because the case
+binder is used. However, we know for sure that in the branch where z is used, z
+is equal to False, and using False doesn't violate linearity since we can
+unrestrictedly create nullary data constructors. A situation similar to this one
+below happens after the CSE transformation.
+\begin{code}
+    case <complex expression> $\leadsto U$ of z {
+        False -> case y of z' { DEFAULT -> z }
+        True  -> y
+    }
+
+A different example is using the case binder $z$ instead of the arguments we
+pattern matched on $x,y$. This currently violates linearity because both $x$ and
+$y$ aren't used and because $z$ is used twice. This doesn't typecheck in the
+frontend either. However, we know that in this branch, using $z$ is effectively
+the same as using $(x,y)$!
+\begin{code}
+    case <complex expression> $\leadsto U$ of z {
+        (x,y) -> z
+    }
+\end{code}
+
+The good news is both these two programs and a handful of others listed in
+Section~\ref{examples} are accepted with a new rule we've devised for the
+linear linter. We still need to prove we don't accept any invalid programs as
+valid.
+
+The key idea of the case-binder-usage solution is annotating the case binder
+with independent usage environments for each pattern match. That is, for
+each of the possible branches, using the case binder $z$ will equate to
+using its usage environment in that branch. That makes it so that using the
+case binder instead of a nullary data constructor is the same, using the case
+binder instead of reconstructing the value with the bound pattern variables too,
+and other situations in which we previously couldn't typecheck linearity are
+now accepted.
+
+The typing rule for the case expression using the case binder solution:
+
+\begin{mathparpagebreakable}
+    \infer*[right=(case)]
+    {\Gamma \vdash t : D_{\pi_1 \dots \pi_n} \leadsto \{U\} \\
+     \Gamma ; z :_{U_k} D_{\pi_1 \dots \pi_n} \vdash b_k : C \leadsto \{V_k\}
+     \and V_k \leq V}
+    {\Gamma \vdash \text{case } t \text{ of } z :_{(U_1\dots U_n)} D_{\pi_1 \dots \pi_n} \{b_k\}_1^m : C \leadsto \{U + V\}}
+\end{mathparpagebreakable}
 
@@
 
-\section{Novel rules}
+% \section{Novel rules}
 
-\subsection{The usage environment in a Recursive Let}
+% \subsection{The usage environment in a Recursive Let}
 
-\subsection{The usage environment of a Case Binder\label{caseBinderKey}}
+% \subsection{The usage environment of a Case Binder\label{caseBinderKey}}
 
-\begin{mathparpagebreakable}
-    \infer*[right=(letrec)]
-    {\Gamma ; \Delta/ \Delta' ; \Omega \vdash M : A \Uparrow \and \Gamma ;
-    \Delta/ \Delta'' ; \Omega \vdash N : B \Uparrow \and \Delta' = \Delta''}
-    {\Gamma ; \Delta/\Delta' ; \Omega \vdash  (M \with N) : A \leadsto U}
-\end{mathparpagebreakable}
+% \begin{mathparpagebreakable}
+%     \infer*[right=(letrec)]
+%     {\Gamma ; \Delta/ \Delta' ; \Omega \vdash M : A \Uparrow \and \Gamma ;
+%     \Delta/ \Delta'' ; \Omega \vdash N : B \Uparrow \and \Delta' = \Delta''}
+%     {\Gamma ; \Delta/\Delta' ; \Omega \vdash  (M \with N) : A \leadsto U}
+% \end{mathparpagebreakable}
 
-\section{Examples}
+\section{Examples\label{examples}}
 

### 2022-11-20 laboratory/UsingPair.hs

@@ -1,0 +1,7 @@
+{-# LANGUAGE LinearTypes #-}
+
+p :: (Int, Int) %1 -> (Int, Int)
+p x = case x of
+        w@(y,z) -> w
+
+main = pure ()

### 2022-11-22 proposal.tex
Proposal introduction and foreword

@@
 
+\usepackage{todonotes}
 \usepackage{cmll}
@@
 
+% \section{Introduction}
+
+% \subsection{Motivation}
+
+% \begin{itemize}
+%     \item Are we really preserving linearity?
+%     \item Inform/unlock other optimisations that take into account linearity
+% \end{itemize}
+
+% \section{Background}
+
+% \begin{itemize}
+%     \item Linear types
+%     \item Linear Haskell
+%     \item Core
+%     \item Sistemas de inferência
+%     \item GADTs e Coercions
+% \end{itemize}
+
+% \section{Related Work}
+
+% \section{Technical Details}
+
+\section{Foreword}
+
+This document is a work in progress proposal that I will deliver at the end of
+the semester as part of my master thesis preparation. The document needs a
+proper introduction and lengthy background and related work section besides the
+content (to be extended) that follows, such that an outsider might understand
+what I propose to do. However, this initial iteration is instead targeted at
+those already familiar with the problem and serves as a (less bureaucratic)
+research proposal on linting linearity in Core and to showcase the progress I
+have made so far.
+
 \section{Introduction}
 
-\subsection{Motivation}
+Since Linear Haskell's\cite{} publication and implementation release in GHC 9.0,
+Haskell's type system supports linearity annotations in functions -- bringing
+linear types into a mainstream pure and lazy functional language. In concrete,
+function arrows can be annotated with a multiplicity. If a function is linear it
+must consume its argument exactly once if it is consumed exactly once, for some
+definition of \emph{consume} that I should revise here.
 
-\begin{itemize}
-    \item Are we really preserving linearity?
-    \item Inform/unlock other optimisations that take into account linearity
-\end{itemize}
+Linearly typed programs are proven/checked to be linear by the type system, so
+the addition of linear types evidently implied changing the type checker to
+support linearity. There exist, however, two distinct type checkers in Haskell.
+The first is run on the surface language, i.e. the Haskell we write, and is a
+big and complex type checker that now also supports typing linearity. The second
+is run on the intermediate language \emph{Core} that we obtain from desugaring
+Haskell.
 
-\section{Background}
+\emph{Core} is a much smaller language than the whole of Haskell (even though we
+can compile the whole of Haskell to it!) and the type checker for it is small
+and fast due to \emph{Core} being explicitly typed and having a small abstract syntax
+tree. This type checker (called \emph{Lint}) gives us guarantees of correctness
+(i.e. sanity) in face of all the complex transformations a Haskell program
+undergoes such as desugaring and optimising transformations because the linter
+is always run on the resulting code before being compiled to untyped machine
+code.
 
-\begin{itemize}
-    \item Linear types
-    \item Linear Haskell
-    \item Core
-    \item Sistemas de inferência
-    \item GADTs e Coercions
-\end{itemize}
+% System FC is the formal system in which the implementation of GHC's intermediate
+% representation language \emph{Core} is based on.
 
-\section{Related Work}
+We want Core to give us guarantees about our desugaring and transformations with
+regard to linearity too -- a linearly typed Core will ensure that linearly-typed
+programs remain correct (i.e.  linearity is preserved) after desugaring and all
+GHC transformations (optimisations shouldn't destroy linearity).
 
-\section{Technical Details}
+Core is already annotated with linearity, but \textbf{we currently ignore it}!
+In spite of the strong formal foundations of linear types driving the
+implementation, their interaction with the whole of GHC is still far from
+trivial, and the implemented formal system cannot accomodate various
+optimising transformations and from them resulting programs, that seemingly
+violate linearity, but which do preserve it semantically. Therefore, we reject
+various valid programs with regard to linearity after desugaring: the current
+solution, because disabling optimisations can inccur great performance costs, is
+to disable the linear linter. We believe that GHC's transformations are correct,
+and it is the linear type system that can't accommodate the resulting programs.
+% diverse problems with linearity spring up when we get past the desugarer and the
+% linter rejects valid programs which is quite undesirable.
 
-Since Linear Haskell's\ref{}
-publication and implementation release in GHC 9.2, Haskell's type system
-supports linearity annotations in functions -- bringing linear types into a
-mainstream pure and lazy functional language.
+% For example, optimizing transformations, coercions and type families, recursive
+% lets, and empty case expressions don't currently fit in with linearity.
 
-System FC is the formal system in which the implementation of GHC's intermediate
-representation language \emph{Core} is based on.
+We propose to formalise a type system that is able to type check linear
+\emph{Core} at all points in the core pipeline and then to implement it into
+GHC. Proposal... \todo{proposal paragraph}
 
-There are at least two distinct typecheckers in core. The first is run on the frontend
-language, i.e. the Haskell we write, and is a big and complex typechecker. The
-second is run on the intermediate language \emph{Core} that we obtain from
-desugaring Haskell.
+% to be able to preserve linearity accross the stages and to enable \emph{Lint} to
+% preserve our sanity regarding linearity and eventually inform with linearity new
+% transformations.
 
-\emph{Core} is a much smaller and more principled language than the whole of
-Haskell (even though we can compile the whole of Haskell to it), and the
-typechecker for it is small and fast due to \emph{Core} being explicitly typed
-and having a very small abstract syntax tree. This typechecker is called
-\emph{Lint} and gives us guarantees of correctness (i.e. sanity) in face of
-the complexity of all the transformations a Haskell program undergoes, such as
-type inference, desugaring and optimising transformations (and other Core
-related passes?).
-
-\begin{itemize}
-    \item In GHC, the type Type is shared between the surface language and Core.
-        So modifying the types implies that Core is modified as well.
-
-    \item More importantly, because Core is a good check that our implementation
-        works as intended. That is, a linearly typed Core will ensure that
-        linearly-typed programs are indeed desugared to linearly-typed programs
-        in Core. And that optimisations do not destroy linearity.
-\end{itemize}
-
-The addition of linear types comes at two levels: frontend and Core.
-In the frontend, we can now annotate with linearity type declarations and
-typecheck programs that use them accordingly (a linear fuction will consume its
-argument exactly once if it is consumed exactly once, for some definition of
-\emph{consume} that I should revise here). In Core, we still have linearity
-annotations and ideally \emph{Lint} would check that all the GHC transformations
-to our linear program preserved its linearity. However, \textbf{it can't}!
-
-Despite the strong formal foundations of linear types driving the
-implementation, their interaction with the whole of GHC is still far from
-trivial. Diverse problems with linearity spring up when we get past the
-desugarer. In particular, optimizing transformations, coercions from GADTs and
-type families, recursive lets, and empty case expressions don't currently fit in
-with linearity.
-
-We believe that GHC's transformations are correct, and it is the linear type
-system that can't accommodate the resulting programs. We aim to formalise a type
-system that is able to type check \emph{Core}/\emph{System FC} at all points in
-the core pipeline and implement it into GHC to be able to preserve linearity
-accross the stages and to enable \emph{Lint} to preserve our sanity regarding
-linearity.
+% A value allocated to be passed to a linear function and then never again used could
+% bypass garbage collection.
 
@@
     }
+\end{code}
 
@@
 
-Taking this example (copy example over) we can see that we don't know how to say
+Taking the following example we can see that we don't know how to say
 that x is indeed linear in one case and unrestricted in the other, even though

### 2022-11-22 proposal.tex
Finish first iteration of proposal introduction

@@
 
-We propose to formalise a type system that is able to type check linear
-\emph{Core} at all points in the core pipeline and then to implement it into
-GHC. Proposal... \todo{proposal paragraph}
+Finally, we propose to work and write a master thesis on a formal type system
+that is able to accommodate linear programs as seen in Core after optimising
+transformations, prove its soundness, and ultimately implement it into GHC's
+linter. In concrete, the type system will build on the current linear core
+system and add rules based on usage annotations for let, letrec and case
+binder bound variables; possibly a new kind of coercion related to
+multiplicities, and other rules that might come along.
+%
+The ultimate measure of success is the -dlinear-core-lint flag that right now,
+when enabled, rejects many linearly valid programs in the linter. Ideally, by
+the end of our research and implementation, this flag could be enabled by
+default and accommodate all existing transformations. Realistically, we want to
+formally (in the type system) accept as many diverse transformations as possible
+while still preserving linearity, even if we are unable to fix them all.
 

### 2022-11-24 proposal.tex
Apply text patches

--- 2022-11-22 proposal.tex
@@
 Haskell's type system supports linearity annotations in functions -- bringing
-linear types into a mainstream pure and lazy functional language. In concrete,
-function arrows can be annotated with a multiplicity. If a function is linear it
-must consume its argument exactly once if it is consumed exactly once, for some
+linear types into a mainstream pure and lazy functional language. Concretely,
+function types can be annotated with a multiplicity (a multiplicity of One
+requires the argument to be consumed exactly once, a multiplicity of Many allows
+the argument to be consumed unrestrictedly, i.e., zero or more times). A
+function is linear if it consumes its arguments exactly once when it itself is
+consumed exactly once (its arguments have multiplicity of one), for some
 definition of \emph{consume} that I should revise here.
@@
 The first is run on the surface language, i.e. the Haskell we write, and is a
-big and complex type checker that now also supports typing linearity. The second
-is run on the intermediate language \emph{Core} that we obtain from desugaring
+very complex type checker that now also supports typing linearity. The second is
+run on the intermediate language \emph{Core} that we obtain from desugaring
 Haskell.
@@
 can compile the whole of Haskell to it!) and the type checker for it is small
-and fast due to \emph{Core} being explicitly typed and having a small abstract syntax
-tree. This type checker (called \emph{Lint}) gives us guarantees of correctness
-(i.e. sanity) in face of all the complex transformations a Haskell program
-undergoes such as desugaring and optimising transformations because the linter
-is always run on the resulting code before being compiled to untyped machine
-code.
+and fast due to \emph{Core} being explicitly typed and much closer to a
+polymorphic lambda calculus than its surface-level counterpart. This type
+checker (called \emph{Lint}) gives us guarantees of correctness in face of all
+the complex transformations a Haskell program undergoes, such as desugaring and
+core-to-core passes, because the linter is always run on the resulting code
+before being compiled to untyped machine code.
 
@@
 regard to linearity too -- a linearly typed Core will ensure that linearly-typed
-programs remain correct (i.e.  linearity is preserved) after desugaring and all
-GHC transformations (optimisations shouldn't destroy linearity).
+programs remain correct (linearity is preserved) after desugaring and all GHC
+transformations (optimisations shouldn't destroy linearity).
 
-Core is already annotated with linearity, but \textbf{we currently ignore it}!
-In spite of the strong formal foundations of linear types driving the
-implementation, their interaction with the whole of GHC is still far from
-trivial, and the implemented formal system cannot accomodate various
-optimising transformations and from them resulting programs, that seemingly
-violate linearity, but which do preserve it semantically. Therefore, we reject
-various valid programs with regard to linearity after desugaring: the current
-solution, because disabling optimisations can inccur great performance costs, is
-to disable the linear linter. We believe that GHC's transformations are correct,
-and it is the linear type system that can't accommodate the resulting programs.
-% diverse problems with linearity spring up when we get past the desugarer and the
+Core is already annotated with linearity, but the \textbf{linter currently
+ignores it}! In spite of the strong formal foundations of linear types driving
+the implementation, their interaction with the whole of GHC is still far from
+trivial, and the implemented type system cannot accomodate various optimising
+transformations and, from them resulting, programs that seemingly violate
+linearity but which do preserve it semantically; for example, laziness makes it
+harder to reason about linearity. Therefore, we reject various valid programs
+with regard to linearity after desugaring: the current solution, because
+disabling optimisations can inccur great performance costs, is to disable the
+linear linter. However, we believe that GHC's transformations are correct, and
+it is the linear type system that can't accommodate the resulting programs.
+% diverse problems with linearity spring up when we get past the desugarer and
+% the
 % linter rejects valid programs which is quite undesirable.
@@
 
-We propose to formalise a type system that is able to type check linear
-\emph{Core} at all points in the core pipeline and then to implement it into
-GHC. Proposal... \todo{proposal paragraph}
+We propose to extend Core's formal type system with more linearity information
+to be able to accommodate linear programs as seen in Core after optimising
+transformations, prove its soundness/type safety, validate core-to-core
+optimisation passes, and ultimately implement it into GHC's linter. Concretely,
+we'll extend the linear Core system with rules based on usage annotations for
+let, letrec and case binder bound variables, possibly a new kind of coercion for
+multiplicities, and eventually other rules.
+%
+The ultimate measure of success is the -dlinear-core-lint flag that, right now,
+when enabled, rejects many linearly valid programs in the linter. Ideally, by
+the end of our research and implementation, this flag could be enabled by
+default and accommodate all existing transformations. Realistically, we want to
+formally (in the type system) accept as many diverse transformations as possible
+while still preserving linearity, even if we are unable to fix them all.
 
@@
 applies many optimising transformations to \emph{Core} and we believe those
-transformations preserve linearity. However, our linear type system cannot check
+transformations preserve linearity. However, Core's linear type system cannot check
 that they indeed preserve linearity.
@@
 have the empty case expression introduced with the \emph{EmptyCase} language
-extension that we currently can't typecheck either.
-
-A usage environment is a mapping from variables to multiplicities.
+extension that Core currently can't typecheck either.
 
@@
 lambda bound, or with its usage environment, if it's let bound.
+
+A usage environment is a mapping from variables to multiplicities.
 
@@
 %
-In concrete, if we had $w1$ instead of the second occurrence of $x$, we'd have
+Concretely, if we had $w1$ instead of the second occurrence of $x$, we'd have
 an empty usage environment for $w1$ in the $False$ branch ($w1 :_\emptyset

### 2022-11-25 proposal.tex
Typing Usage Envs and Lets sections

@@
 
-We propose to extend Core's formal type system with more linearity information
-to be able to accommodate linear programs as seen in Core after optimising
-transformations, prove its soundness/type safety, validate core-to-core
-optimisation passes, and ultimately implement it into GHC's linter. Concretely,
-we'll extend the linear Core system with rules based on usage annotations for
-let, letrec and case binder bound variables, possibly a new kind of coercion for
-multiplicities, and eventually other rules.
+We propose to extend Core's type system with additional linearity information to
+be able to accommodate linear programs as seen in Core after optimising
+transformations. We will prove the system's soundness/type safety and validate
+some core-to-core optimisation passes. Ultimately, we will implement the
+extension into GHC's linter. Concretely, we'll extend Core's linear type system
+with rules based on usage annotations for let, letrec and case binder bound
+variables. Additionally, we will explore a new kind of coercion for
+multiplicities to validate programs that combine GADTs with multiplicities.
 %
@@
 default and accommodate all existing transformations. Realistically, we want to
-formally (in the type system) accept as many diverse transformations as possible
-while still preserving linearity, even if we are unable to fix them all.
+accept as many diverse transformations as possible while still preserving
+linearity, even if we are unable to fix them all.
+
+In section~\ref{typingUsageEnvs} we describe usage environments and how they
+solve three distinct problems. In section~\ref{multiplicityCoercions} we discuss
+the beginning of multiplicity coercions. In section~\ref{examples} we take
+programs that are currently rejected and show how the type system with our
+extensions can accept them.
 
@@
 
-\section{Typing Usage Environments}
+\section{Typing Usage Environments\label{typingUsageEnvs}}
 
-The first set of problems appears in the core-to-core optimisation passes. GHC
-applies many optimising transformations to \emph{Core} and we believe those
-transformations preserve linearity. However, Core's linear type system cannot check
-that they indeed preserve linearity.
+Haskell has call-by-need semantics that entail that an expression isn't
+evaluated when it is let bound but rather when the binding variable is used.
+This makes it hard to reason about linearity for let bound variables. The
+following example doesn't typecheck in Haskell, but semantically it is indeed
+linear because the let binding is lazy.
 
-The simpler examples come from straightforward and common optimising
-transformations. Then we have recursive let definitions that don't accommodate
-linearity even though it might converge to only use the value once. Finally, we
-have the empty case expression introduced with the \emph{EmptyCase} language
-extension that Core currently can't typecheck either.
+\begin{code}
+f :: a %1 -> a
+f x = let y = x in y
+\end{code}
 
-The key idea is to annotate every variable with either its multiplicity, if it's
-lambda bound, or with its usage environment, if it's let bound.
+Despite not being accepted by the surface-level language, linear programs using
+lets can occur in Core due to optimising transformations that create let
+bindings. In a similar fashion, programs which violate syntatic linearity
+for other reasons other than let bindings are formed from Core transformations.
 
-A usage environment is a mapping from variables to multiplicities.
+The solution we found to a handful of problems regarding binders is to have a
+\emph{usage environment}. A usage environment is a mapping from variables to
+multiplicities. The main idea is to annotate those binders with a usage
+annotation rather than a multiplicity (in contrast, a lambda-bound variable has
+exactly one multiplicity). When we find a bound variable with a usage
+environment, we type linearity as if we were using all variables with
+corresponding multiplicities from that usage environment.
 
-For example, if y and z are linear, in the following code it might not look as
-if y and z were both being consumed linearly, but indeed they are since in the
-first branch we use x which means using y and z linearly, and we use y and z
-directly on the second branch. Note that let binding x doesn't consume y and z,
-only using x itself does, because of laziness.
+In the above example, this would amount to annotating $y$ with a usage
+environemnt $x := 1$ (because the expression bound by $y$ uses $x$ one time).
+Upon using $y$, we are actually using $x$ one time, and therefore linearity is
+preserved.
+
+% The first set of problems appears in the core-to-core optimisation passes. GHC
+% applies many optimising transformations to \emph{Core} and we believe those
+% transformations preserve linearity. However, Core's linear type system cannot check
+% that they indeed preserve linearity.
+
+% The simpler examples come from straightforward and common optimising
+% transformations. Then we have recursive let definitions that don't accommodate
+% linearity even though it might converge to only use the value once. Finally, we
+% have the empty case expression introduced with the \emph{EmptyCase} language
+% extension that Core currently can't typecheck either.
+
+% The key idea is to annotate every variable with either its multiplicity, if it's
+% lambda bound, or with its usage environment, if it's let bound.
+
+% A usage environment is a mapping from variables to multiplicities.
+
+% When we lint a core expression, we get both its type and its usage environment.
+% That means that to lint linearity in an expression, whenever we come across a
+% free variable we compute its usage environment and take it into account
+
+% TODO: When we type an expression we get both the type and usage environment
+
+\subsection{Let}
+
+Let bindings in Core are the first family of problems we tackle with usage
+environment annotations. Multiple transformations can introduce let-bindings,
+such as CSE and join points. By extending the type system to allow lets in core
+we start paving our way to a linear linter.
+
+Right now, Core annotates every variable with a multiplicity at its binding
+site. The multiplicity for let-bound variables must be ignored thorought the
+transformation pipeline or otherwise too many programs valid programs would be
+rejected for violating linearity in its transformed type.
+
+However, programs with let bindings can be correctly typed by associating a
+usage environment to the bound variables. Instead of associating a multiplicity
+to every binder, we want to associate a multiplicity if the variable is lambda
+bound and a usage environment when it is let bound. We then instance the usage
+environment solution to lets in particular -- a let bound variable is annotated
+with the usage environment computed from the binder expression; in the let body
+expression, when we find an occurrence of the let bound variable we emit its
+usage environment. The typing rule is the following:
+
+\begin{mathparpagebreakable}
+    \infer*[right=(let)]
+    {\Gamma \vdash t : A \leadsto \{U\} \\
+     \Gamma ; x :_{U} A \vdash u : C \leadsto \{V\}}
+    {\Gamma \vdash \text{let } x :_{U} A = t \text{ in } u : C \leadsto \{V\}}
+\end{mathparpagebreakable}
+
+Take for example an expression in which $y$ and $z$ are lambda bound with a
+multiplicity of exactly one. In the following code it might not look as if $y$
+and $z$ are both being consumed linearly, but indeed they are since in the first
+branch we use $x$ -- which means using $y$ and $z$ linearly -- and we use $y$
+and $z$ directly on the second branch. Note again that let binding $x$ doesn't
+consume $y$ and $z$ because of laziness. Only \emph{using} $x$ consumes $y$ and
+$z$.
+
 \begin{code}
@@
 
-If we annotate the x bound by let with a usage environment $\delta$ mapping all (free?)
-variables in its binder to a multiplicity ($\delta = [y := 1, z := 1]$), we
-could, upon finding x, simply emit that y and z are consumed once. When typing
-the second branch we'd also see that y and z are used exactly once, hence the
-linearity in the two branches match.
+If we annotate the $x$ bound by the let with a usage environment $\delta$
+mapping all free variables in its binder to a multiplicity ($\delta = [y := 1, z
+:= 1]$), we could, upon finding $x$, simply emit that $y$ and $z$ are consumed
+once.  When typing the second branch we'd also see that $y$ and $z$ are used
+exactly once. Because both branches use $y$ and $z$ linearly, the whole case
+expression uses $y$ and $z$ linearly.
 
-Currently, in GHC, we don't annotate let-bound variables with a usage
-environment, but we already calculate a usage environment and use it to check
-some things (which things?)
+\subsection{Letrec}
 
-\begin{itemize}
-    \item occurrences? store usage environment in Ids (vars)
-    \item Recursive lets (can it be rewritten using fix)
-    \item Empty case expression
-\end{itemize}
+% Currently, in GHC, we don't annotate let-bound variables with a usage
+% environment, but we already calculate a usage environment and use it to check
+% some things (which things?)
 
-What do we currently do?
+% \begin{itemize}
+%     \item occurrences? store usage environment in Ids (vars)
+%     \item Recursive lets (can it be rewritten using fix)
+%     \item Empty case expression
+% \end{itemize}
 
-When we lint a core expression, we get both its type and its usage environment.
-That means that to lint linearity in an expression, whenever we come across a
-free variable we compute its usage environment and take it into account
+% \subsection{Inlining}
 
-\subsection{Inlining}
-
-If we annotate the let bound variables with their usage and emit that usage when
-we come across those variables, we can solve the linearity issues with inlining.
+% If we annotate the let bound variables with their usage and emit that usage when
+% we come across those variables, we can solve the linearity issues with inlining.
 
@@
 
-Now, how to compute the usage environment of a case expression? It's described
-here if I'm not mistaken
-https://gitlab.haskell.org/ghc/ghc/-/wikis/uploads/355cd9a03291a852a518b0cb42f960b4/minicore.pdf.
+% Description of computing usage environment of a case expression:
+% https://gitlab.haskell.org/ghc/ghc/-/wikis/uploads/355cd9a03291a852a518b0cb42f960b4/minicore.pdf.
 
-However, not understanding it, my take would be that we can compute the usage
-environment of every branch of the case expression and make sure that they all
-unify (wrt to submultiplicities). The special case is when the branch happens to
-call $f$ itself while computing the usage environment of $f$. If it were another
-let bound variable we'd add its own usage environment to the one we're
-computing; if it were a lambda bound variable we'd add [itself $:=$ its
-multiplicity].
+% However, not understanding it, my take would be that we can compute the usage
+% environment of every branch of the case expression and make sure that they all
+% unify (wrt to submultiplicities). The special case is when the branch happens to
+% call $f$ itself while computing the usage environment of $f$. If it were another
+% let bound variable we'd add its own usage environment to the one we're
+% computing; if it were a lambda bound variable we'd add [itself $:=$ its
+% multiplicity].
 
@@
     {\Gamma ; x_1 : A_1 \dots x_n : A_n \vdash t_i : A_i \leadsto \{U_{i_\text{naive}}\} \\
-     (U_1 \dots U_n) = computeRecUsages(U_{1_\text{naive}} \dots U_{n_\text{naive}}) \\
+     (U_1 \dots U_n) = \mathit{computeRecUsages}(U_{1_\text{naive}} \dots U_{n_\text{naive}}) \\
      \Gamma ; x_1 :_{U_1} A_1 \dots x_n :_{U_n} A_n \vdash u : C \leadsto \{V\}}
@@
 
-\section{Multiplicity Coercions}
+\section{Multiplicity Coercions\label{multiplicityCoercions}}
 
@@
 associated bound variables linearly we constructed some additional examples that
-bring out the problem quite clearly
+bring out the problem quite clearly.
 \begin{code}
@@
 We initially wondered whether x was being consumed if the pattern match ignored
-some of its variables. We put that if it does have wildcards then it isn't
+some of its variables. We pose that if it does have wildcards then it isn't
 consuming x fully
@@
 Can we have a solution that even handles weird cases in which we pattern match
-twice on the same expression but only on one constructor argument per time?
+twice on the same expression but only on one constructor argument per time? I
+don't think so.
 \begin{code}
@@
 
+\section{Core2Core Validdations}
+
+\subsection{INlining}
+\subsection{CSE}
+
 \end{document}

### 2022-11-25 laboratory/LetBound.hs

@@ -1,0 +1,8 @@
+{-# LANGUAGE LinearTypes #-}
+
+
+f :: a %1 -> a
+f x = let y = x in y
+
+main = pure ()
+

### 2022-11-25 proposal.tex
Recursive lets section

@@
 
-\subsection{Letrec}
-
 % Currently, in GHC, we don't annotate let-bound variables with a usage
@@
 
-Optimisations can create a letrec which uses a variable linearly. The following
-example uses 'x' linearly, but this is not seen by the linter.
+A recursive let binds a variable to an expression that might use the bound
+variable in its body. Certain optimisations can create a recursive let that uses
+linearly bound variables in its binder body. We want to treat recursive lets
+similarly to lets: attribute to bound variables a usage environment and emit it
+when the variable is used.
+% TODO: Replace "certain" with concrete examples
+
+The tricky part about determining the usage environment of the variables bound
+by a recursive let is knowing what usage to emit inside their own body when
+computing said usage environment. The example below uses $x$ linearly % \ref{example_letrec}
+but how can we prove it? If we are able to somehow compute $f$'s usage
+environment to $x := 1$, we know $x$ is used once when $f$ is applied to
+$\mathit{True}$.
+
 \begin{code}
@@
 
-Following the idea of making let-bound variables remember the usage environment
-here's an informal description that that example is well typed. We want to
-anotate the let-bound variable $f$ with a usage environment delta $\delta$, and
-use the binding body to compute it.
+The key idea to computing the usage environment of multiple variables that use
+themselves in their body is to do the computation in two separate passes. First,
+calculate a naive environment by emiting free variables multiplicities as usual
+while treating the recursive-let bound variables as free ones. Second, pass the
+variables names and their corresponding naive environment as input to
+algorithm~\ref{computeRecUsages} to get a final usage environment. Intuitively,
+the algorithm, for each recursive binder, goes over all (initially naive) usage
+environment and substitutes the recursive binder by the corresponding usage
+environment, scaled up by the amount of times that recursive binder is used in
+the environment being updated.
+
+We present ahead an algorithm for computing the usage environment of a set of
+recursive definitions, and its textual description. An (at least informal) proof
+that the algorithm works should eventually follow this.
+
+\begin{itemize}
+    \item Given a list of functions and their naive environment computed from
+        the body and including the recursive function names ($(f, F), (g, G),
+        (h, H)$ in which there might exist multiple occurrences of $f, g, h$ in $F, G, H$
+    \item For each bound rec var and its environment, update all bindings and
+        their usage as described in the algorithm
+    \item After iterating through all bound rec vars, all usage environments
+        will be free of recursive bind usages, and hence "final"
+\end{itemize}
+
+Note that the difficulty of calculating a usage environment for $n$
+recursive-let bound variables increases quadratically, i.e. the algorithm has
+complexity $O(n^2)$, but this is not a problem since it's rare to have more than
+a handful of binders in the same recursive let block.
+
+% TODO: I should probably use the re-computed usageEnvs instead of the naiveUsageEnvs.
+% I'm pretty sure it might fail in some inputs if I keep using the naiveUsageEnvs.
+\begin{algorithm}
+$usageEnvs \gets naiveUsageEnvs.map(fst)$\;
+\For{$(bind, U) \in naiveUsageEnvs$}{
+    \For{$V \in usageEnvs$}{
+        $V \gets sup(V[bind]*U\setminus\{bind\}, V\setminus\{bind\})$
+    }
+}
+\caption{computeRecUsages\label{computeRecUsages}}
+\end{algorithm}
+
+Putting the usage environment idea together with the algorithm we get the following typing rule:
+
+\begin{mathparpagebreakable}
+    \infer*[right=(letrec)]
+    {\Gamma ; x_1 : A_1 \dots x_n : A_n \vdash t_i : A_i \leadsto \{U_{i_\text{naive}}\} \\
+     (U_1 \dots U_n) = \mathit{computeRecUsages}(U_{1_\text{naive}} \dots U_{n_\text{naive}}) \\
+     \Gamma ; x_1 :_{U_1} A_1 \dots x_n :_{U_n} A_n \vdash u : C \leadsto \{V\}}
+    {\Gamma \vdash \text{let } x_1 :_{U_1} A_1 = t_1 \dots x_n :_{U_n} A_n = t_n \text{ in } u : C \leadsto \{V\}}
+\end{mathparpagebreakable}
+
+Instancing the usage environment solution according to the above description,
+here's an informal proof that the previous example is well typed: We %\ref{example_letrec}
+compute the naive usage environment of $f$ to be $x := 1, f := 1$. We compute
+its actual usage environment by scaling the naive environment of $f$ without $f$
+by the amount of times $f$ appears in the naive environment of $f$ ($1*[x :=
+1]$) and get $x := 1$. Finally, we emit $x := 1$ from applying $f$ in the let's
+body.
+
+% Following the idea of making let-bound variables remember the usage environment
+% here's an informal description that that example is well typed. We want to
+% anotate the let-bound variable $f$ with a usage environment delta $\delta$, and
+% use the binding body to compute it.
 
@@
 
-My idea is that we can emit a special usage [$rec := p$], which, when unified
-against the other case branches $\delta$ will always succeed with a unification
-mapping from $rec \rightarrow p\delta$ scaled by the multiplicity of $rec$
+% My idea is that we can emit a special usage [$rec := p$], which, when unified
+% against the other case branches $\delta$ will always succeed with a unification
+% mapping from $rec \rightarrow p\delta$ scaled by the multiplicity of $rec$
 
-So taking the example, to compute the usage environment of $f$, we'd compute for
-the second branch $[x := 1]$ and for the first branch $[rec := 1]$. Then, we'd
-unify them with $rec \rightarrow [x := 1*1]$, and somehow result in $[x := 1]$
+% So taking the example, to compute the usage environment of $f$, we'd compute for
+% the second branch $[x := 1]$ and for the first branch $[rec := 1]$. Then, we'd
+% unify them with $rec \rightarrow [x := 1*1]$, and somehow result in $[x := 1]$
 
-An example which should break linearity because $x$ is not linear in the first
-branch:
+The next example is a linearly invalid program because $x$ is a %~\ref{example_letrec_2}
+linearly bound variable that is used more than once, and shows how this method
+correctly rejects it.
+
 \begin{code}
@@
 
-To compute the usage environment of $f$ we take the second branch usage
-environment $[x := 1]$ and the first branch usage $[rec := 1+1]$ and unify them,
-somehow resulting in $[x := 2]$. Now, to lint the linearity in the whole let
-expression we must ensure that the body of the let uses $x$ linearly. The body
-is $f True$, which is a let-bound variable (how to distinguish functions that
-must be applied vs variables) so we take its usage environment $[x := 2]$ which
-does not use $x$ linearly and thus breaks linearity.
+To compute the usage environment of $f$ we take its naive environment to be $x
+:= 1, f := 2$. Then, we compute the final environment to be $x := 1$ scaled by
+the usage of $f$, $2*[x := 1]$, resulting in $x := 2$. Now, to lint the
+linearity of the whole let we must ensure the body of the let uses $x$ linearly.
+We emit from the body the usage environment of $f$ ($x := 2$) which uses $x$
+more than once, i.e. not linearly.
 
-\textbf{Re-explained:} to compute the usage environment of the recursive
-let-bound function $f$ when applied, we compute $Z$ such that for all branch
-alternatives $U,V,...$, $U \subset Z$, $V \subset Z$ and so on by tracking the
-multiplicities and usage environments of variables that show up in the body and
-by emitting a special keyword $rec$ everytime we find a saturated call of $f$
-(how to handle unsaturated calls?) (how to have a more general solution that
-doesn't require a keyword, perhaps something to do with fixed points); Then we
-scale $Z \setminus \{(rec,_)\}$ by $Z[rec]$ and get $T$ which is the usage
-environment of $f$ when saturated.
+% To compute the usage environment of $f$ we take the second branch usage
+% environment $[x := 1]$ and the first branch usage $[rec := 1+1]$ and unify them,
+% somehow resulting in $[x := 2]$. Now, to lint the linearity in the whole let
+% expression we must ensure that the body of the let uses $x$ linearly. The body
+% is $f True$, which is a let-bound variable (how to distinguish functions that
+% must be applied vs variables) so we take its usage environment $[x := 2]$ which
+% does not use $x$ linearly and thus breaks linearity.
 
-\textbf{Example 1 revisited:}
-\begin{enumerate}
-    \item Take $U = \{rec := 1\}$
-    \item Take $V = \{x := 1\}$
-    \item Take $Z = \{x := 1, rec := 1\}$
-    \item Take $\pi = Z[rec] = 1$
-    \item Take $W = Z \setminus \{rec\} = \{x := 1\}$
-    \item Take $T = \pi W = \{x := \pi * 1\} = \{x := 1\}$
-    \item Linearity OK
-\end{enumerate}
+% \textbf{Re-explained:} to compute the usage environment of the recursive
+% let-bound function $f$ when applied, we compute $Z$ such that for all branch
+% alternatives $U,V,...$, $U \subset Z$, $V \subset Z$ and so on by tracking the
+% multiplicities and usage environments of variables that show up in the body and
+% by emitting a special keyword $rec$ everytime we find a saturated call of $f$
+% (how to handle unsaturated calls?) (how to have a more general solution that
+% doesn't require a keyword, perhaps something to do with fixed points); Then we
+% scale $Z \setminus \{(rec,_)\}$ by $Z[rec]$ and get $T$ which is the usage
+% environment of $f$ when saturated.
 
-\textbf{Example 2 revisited:}
-\begin{enumerate}
-    \item Take $U = \{rec := 2\}$
-    \item Take $V = \{x := 1\}$
-    \item Take $Z = \{x := 1, rec := 1\}$
-    \item Take $\pi = Z[rec] = 2$
-    \item Take $W = Z \setminus \{rec\} = \{x := 1\}$
-    \item Take $T = \pi W = \{x := \pi * 1\} = \{x := 2\}$
-    \item Linearity not OK
-\end{enumerate}
+% \textbf{Example 1 revisited:}
+% \begin{enumerate}
+%     \item Take $U = \{rec := 1\}$
+%     \item Take $V = \{x := 1\}$
+%     \item Take $Z = \{x := 1, rec := 1\}$
+%     \item Take $\pi = Z[rec] = 1$
+%     \item Take $W = Z \setminus \{rec\} = \{x := 1\}$
+%     \item Take $T = \pi W = \{x := \pi * 1\} = \{x := 1\}$
+%     \item Linearity OK
+% \end{enumerate}
 
-Draft typing rule:
+% \textbf{Example 2 revisited:}
+% \begin{enumerate}
+%     \item Take $U = \{rec := 2\}$
+%     \item Take $V = \{x := 1\}$
+%     \item Take $Z = \{x := 1, rec := 1\}$
+%     \item Take $\pi = Z[rec] = 2$
+%     \item Take $W = Z \setminus \{rec\} = \{x := 1\}$
+%     \item Take $T = \pi W = \{x := \pi * 1\} = \{x := 2\}$
+%     \item Linearity not OK
+% \end{enumerate}
 
-\begin{mathparpagebreakable}
-    \infer*[right=(letrec)]
-    {\Gamma ; x_1 : A_1 \dots x_n : A_n \vdash t_i : A_i \leadsto \{U_{i_\text{naive}}\} \\
-     (U_1 \dots U_n) = \mathit{computeRecUsages}(U_{1_\text{naive}} \dots U_{n_\text{naive}}) \\
-     \Gamma ; x_1 :_{U_1} A_1 \dots x_n :_{U_n} A_n \vdash u : C \leadsto \{V\}}
-    {\Gamma \vdash \text{let } x_1 :_{U_1} A_1 = t_1 \dots x_n :_{U_n} A_n = t_n \text{ in } u : C \leadsto \{V\}}
-\end{mathparpagebreakable}
+% Draft typing rule:
 
-The difficulty of calculating a usage environment for the recursive let recs
-increases with the number of recursive definitions in the same let.
+% In haskell that would look like this;
+% \begin{code}
+% computeRecUsageEnvs :: [(Var, UsageEnv)] -- Recursive usage environments associated to their recursive call
+%                     -> [(Var, UsageEnv)] -- Non-recursive usage environments
+% computeRecUsageEnvs l =
+%   foldl (flip \\(v,vEnv) -> map \\(n, nEnv) -> (n, ((fromMaybe 0 \$ v `M.lookup` nEnv) `scale` (M.delete v vEnv)) `sup` (M.delete v nEnv))) l l
 
-We devise an algorithm for computing the usage environment of a set of recursive
-definitions. The proof that the algorithm works follows...
+% sup :: UsageEnv -> UsageEnv -> UsageEnv
+% sup = M.merge M.preserveMissing M.preserveMissing (M.zipWithMatched \$ \_ x y -> max x y)
 
-\begin{itemize}
-    \item Given a list of functions and their naive environment computed from
-        the body and including the recursive function names ($(f, F), (g, G),
-        (h, H)$ in which there might exist multiple occurrences of $f, g, h$ in $F, G, H$
-    \item For each bound rec var and its environment, update all bindings and
-        their usage as described in the algorithm
-    \item After iterating through all bound rec vars, all usage environments
-        will be free of recursive bind usages, and hence "final"
-\end{itemize}
-
--- I should probably use the re-computed usageEnvs instead of the
-naiveUsageEnvs. I'm pretty sure it might fail in some inputs if I keep using the
-naiveUsageEnvs.
-\begin{algorithm}
-$usageEnvs \gets naiveUsageEnvs.map(fst)$\;
-\For{$(bind, U) \in naiveUsageEnvs$}{
-    \For{$V \in usageEnvs$}{
-        $V \gets sup(V[bind]*U\setminus\{bind\}, V\setminus\{bind\})$
-    }
-}
-\caption{computeRecUsages}
-\end{algorithm}
-
-In haskell that would look like this;
-\begin{code}
-computeRecUsageEnvs :: [(Var, UsageEnv)] -- Recursive usage environments associated to their recursive call
-                    -> [(Var, UsageEnv)] -- Non-recursive usage environments
-computeRecUsageEnvs l =
-  foldl (flip \\(v,vEnv) -> map \\(n, nEnv) -> (n, ((fromMaybe 0 \$ v `M.lookup` nEnv) `scale` (M.delete v vEnv)) `sup` (M.delete v nEnv))) l l
-
-sup :: UsageEnv -> UsageEnv -> UsageEnv
-sup = M.merge M.preserveMissing M.preserveMissing (M.zipWithMatched \$ \_ x y -> max x y)
-
-scale :: Mult -> UsageEnv -> UsageEnv
-scale m = M.map (*m)
-\end{code}
+% scale :: Mult -> UsageEnv -> UsageEnv
+% scale m = M.map (*m)
+% \end{code}
 

### 2022-11-25 proposal.tex
Cleaning up the work in progress parts

@@
 \begin{itemize}
-    \item consuming an atomic value is forcing it to Normal Form (NF)
-    \item consuming a value that is constructed with more than 1 argument is
+    \item Consuming an atomic value is forcing it to Normal Form (NF)
+    \item Consuming a value that is constructed with more than 1 argument is
         consuming all of its arguments
@@
 \begin{code}
-    case <complex expression> $\leadsto U$ of z {
+    case <complex expression> of z {
         False -> case y of z' { DEFAULT -> z }
@@
 \end{code}
+% $\leadsto U$
 
@@
 \begin{code}
-    case <complex expression> $\leadsto U$ of z {
+    case <complex expression> of z {
         (x,y) -> z
@@
 
-\subsection{Empty Case}
-
-For case expressions, the usage environment is computed by checking all branches
-and taking sup. However, this trick doesn't work when there are no branches.
-
-\begin{itemize}
-\item https://gitlab.haskell.org/ghc/ghc/-/issues/20058
-\item https://gitlab.haskell.org/ghc/ghc/-/issues/18768
-
-\item (1) Just like a case expression remembers its type (Note [Why does Case have a
-'Type' field?] in Core.hs), it should remember its usage environment. This data
-should be verified by Lint.
-
-\item (2) Once this is done, we can remove the Bottom usage and the second field of
-UsageEnv. In this step, we have to infer the correct usage environment for empty
-case in the typechecker.
-\end{itemize}
-
-\begin{code}
-{-# LANGUAGE LinearTypes, EmptyCase #-}
-module M where
-
-{-# NOINLINE f #-}
-f :: a %1-> ()
-f x = case () of {}
-\end{code}
-
-This example is well typed: a function is linear if it consumes its argument
-exactly once when it's consumed exactly once. It seems like the function isn't
-linear since it won't consume x because of the empty case, however, that also
-means f won't be consumed due to the same empty case, thus linearity is
-preserved.
-
-\begin{code}
-* In the case of empty types (see Note [Bottoming expressions]), say
-  data T
-we do NOT want to replace
-  case (x::T) of Bool {}   -->   error Bool "Inaccessible case"
-because x might raise an exception, and *that*'s what we want to see!
-(#6067 is an example.) To preserve semantics we'd have to say
-   x `seq` error Bool "Inaccessible case"
-but the 'seq' is just such a case, so we are back to square 1.
-\end{code}
-
-There are three different problems:
-
-\begin{itemize}
-\item castBottomExpr converts (case x :: T of {}) :: T to x.
-\item Worker/wrapper moves the empty case to a separate binding
-\item CorePrep eliminates empty case, just like point 1 (See -- Eliminate empty
-    case in GHC.CoreToStg.Prep
-\end{itemize}
-
-With castBottomExpr, we get the example above to
-\begin{code}
-    f = \ @a (x (%1) :: a) -> ()
-\end{code}
-And if we don't, 
-\begin{code}
-    f = \ @a (x (%1) :: a) -> case () of {}
-\end{code}
-And that supposedly if we had a usage environment in the case expression we
-could avoid the error. How is it typed without the transformation in face
-of the bottom? (Even knowing that theoretically it's because of divergence?)
-
 \section{Multiplicity Coercions\label{multiplicityCoercions}}
@@
 
-When we pattern match on a GADT we ...
+% When we pattern match on a GADT we ...
 
@@
 
-From the definition of groupBy in GHC.Data.List.Infinite
+From the definition of $\mathit{groupBy}$ in $\mathit{GHC.Data.List.Infinite}$:
+
 \begin{code}
@@
 \end{code}
-we get the following core which violates linearity.
-\begin{code}
-groupBy = \ (@a) (eq :: a -> a -> Bool) (eta :: Infinite a) ->
-  letrec {
-    go :: Infinite a -> Infinite (NonEmpty a)
-    go = \ (inf :: Infinite a) -> case inf of {
-      Inf x xs -> let {
-        ds :: ([a], Infinite a)
-        ds = let {
-            parteq :: a -> Bool
-            parteq = eq x
-        } in
-          letrec {
-            goZ :: Infinite a -> ([a], Infinite a)
-            goZ = \ (inf' :: Infinite a) -> case wgo inf' of { (# wA, wB #) -> (wA, wB) };
 
-            wgo :: Infinite a -> (# [a], Infinite a #)
-            wgo = \ (inf' :: Infinite a) -> case inf' of wildX2 {
-              Inf y ys ->
-                join {
-                    jp :: ([a], Infinite a) \%1 -> (# [a], Infinite a #)
-                    jp (wwj :: ([a], Infinite a)) =
-                        case wwj of wwj {
-                            DEFAULT -> case wwj of { (wA, wB) -> (# wA, wB #) } }
-                    } in
-                      case parteq y of {
-                        False ->
-                            let {
-                                ww :: ([a], Infinite a)
-                                ww = ([] @a, wildX2)
-                            } in jump jp ww;
-                        True ->
-                            let {
-                                dy :: ([a], Infinite a)
-                                dy = case wgo ys of { (# wA, wB #) -> (wA, wB)
-                                }
-                            } in
-                                let {
-                                    wwB :: [a]
-                                    wwB = case dy of { (bs, cs) -> bs }
-                                } in
-                                    let {
-                                        wwA :: [a]
-                                        wwA = : @a y wwB
-                                    } in let {
-                                        wwC :: Infinite a
-                                        wwC = case dy of { (bs, cs) -> cs }
-                                    } in let {
-                                        ww :: ([a], Infinite a)
-                                        ww = (wwA, wwC)
-                                    } in jump jp ww
-                                }
-              };
-          } in
-              case wgo xs of { (# wA, wB #) -> (wA, wB) }
-      } in
-          Inf @(NonEmpty a) (:| @a x (case ds of { (bs, cs) -> bs })) (case ds of { (bs, cs) -> go cs })
-    };
-  } in go eta
-\end{code}
-The issue is in the linear function that shows up in the core output (how does
-the linearity end up there?). The wwj variable is used once in the case
-expression, bound as the case binder, and used in the case body once again. Why
-do we do that instead of pattern matching right away? Seems a bit redundant.
-
-Observation: If the branch is DEFAULT, then the case binder binds the case
-scrutinee which was just forced?, but hasn't been actually consumed, because we
-haven't consumed its components. As long as the same branch doesn't consume the
-pattern matching result and the case binder at the same time it should be fine?
-
-In this example, we force wwj with the case binder, but we don't really
-consume it (more precise definition of consume...), so we can use the case
-binder meaning we're simply using x for the first time. Needs to be
-formalised....
+We get a somewhat large resulting core expression, in the middle of which the
+following expression with a linear type -- which currently syntatically violates
+linearity.
 
@@
 
-Solution with the case binder idea that unifies a different problems:
-The case binder is annotated with a usage environment $U_wwj = []$, and its
-usage is equal to using the usage environment
+Using the case binder usage solution, the case binder is annotated with a usage
+environment for the only branch ($U_wwj = \emptyset$) and using $wwj$ is equal
+to using the $\emptyset$. Here it would mean that second \textbf{case of} $wwj$
+doesn't actually use any variables and hence the first $wwj$ isn't used twice.
 
-Here it would mean that second \textbf{case of} $wwj$ doesn't actually use any
-variables and hence it isn't used twice.
+% \begin{code}
+% groupBy = \ (@a) (eq :: a -> a -> Bool) (eta :: Infinite a) ->
+%   letrec {
+%     go :: Infinite a -> Infinite (NonEmpty a)
+%     go = \ (inf :: Infinite a) -> case inf of {
+%       Inf x xs -> let {
+%         ds :: ([a], Infinite a)
+%         ds = let {
+%             parteq :: a -> Bool
+%             parteq = eq x
+%         } in
+%           letrec {
+%             goZ :: Infinite a -> ([a], Infinite a)
+%             goZ = \ (inf' :: Infinite a) -> case wgo inf' of { (# wA, wB #) -> (wA, wB) };
 
-\subsection{In Result of TcGblEnv axioms}
+%             wgo :: Infinite a -> (# [a], Infinite a #)
+%             wgo = \ (inf' :: Infinite a) -> case inf' of wildX2 {
+%               Inf y ys ->
+%                 join {
+%                     jp :: ([a], Infinite a) \%1 -> (# [a], Infinite a #)
+%                     jp (wwj :: ([a], Infinite a)) =
+%                         case wwj of wwj {
+%                             DEFAULT -> case wwj of { (wA, wB) -> (# wA, wB #) } }
+%                     } in
+%                       case parteq y of {
+%                         False ->
+%                             let {
+%                                 ww :: ([a], Infinite a)
+%                                 ww = ([] @a, wildX2)
+%                             } in jump jp ww;
+%                         True ->
+%                             let {
+%                                 dy :: ([a], Infinite a)
+%                                 dy = case wgo ys of { (# wA, wB #) -> (wA, wB)
+%                                 }
+%                             } in
+%                                 let {
+%                                     wwB :: [a]
+%                                     wwB = case dy of { (bs, cs) -> bs }
+%                                 } in
+%                                     let {
+%                                         wwA :: [a]
+%                                         wwA = : @a y wwB
+%                                     } in let {
+%                                         wwC :: Infinite a
+%                                         wwC = case dy of { (bs, cs) -> cs }
+%                                     } in let {
+%                                         ww :: ([a], Infinite a)
+%                                         ww = (wwA, wwC)
+%                                     } in jump jp ww
+%                                 }
+%               };
+%           } in
+%               case wgo xs of { (# wA, wB #) -> (wA, wB) }
+%       } in
+%           Inf @(NonEmpty a) (:| @a x (case ds of { (bs, cs) -> bs })) (case ds of { (bs, cs) -> go cs })
+%     };
+%   } in go eta
+% \end{code}
+% The issue is in the linear function that shows up in the core output (how does
+% the linearity end up there?). The wwj variable is used once in the case
+% expression, bound as the case binder, and used in the case body once again. Why
+% do we do that instead of pattern matching right away? Seems a bit redundant.
 
-I don't understand this one yet. I would guess it has to do with multiplicity
-coercions. Incorrect incompatible branch: CoAxBranch (src/Prelude/Linear/GenericUtil.hs:112:3-51):
-\begin{code}
-type family Fixup (f :: Type -> Type) (g :: Type -> Type) :: Type -> Type where
-  Fixup (D1 c f) (D1 _c g) = D1 c (Fixup f g)
-  Fixup (C1 c f) (C1 _c g) = C1 c (Fixup f g)
-  Fixup (S1 c f) (S1 _c (MP1 m f)) = S1 c (MP1 m f) -- error in this constructor
-  Fixup (S1 c f) (S1 _c f) = S1 c f
-  Fixup (f :*: g) (f' :*: g') = Fixup f f' :*: Fixup g g'
-  Fixup (f :+: g) (f' :+: g') = Fixup f f' :+: Fixup g g'
-  Fixup V1 V1 = V1
-  Fixup _ _ = TypeError ('Text "FixupMetaData: representations do not match.")
-\end{code}
+% Observation: If the branch is DEFAULT, then the case binder binds the case
+% scrutinee which was just forced?, but hasn't been actually consumed, because we
+% haven't consumed its components. As long as the same branch doesn't consume the
+% pattern matching result and the case binder at the same time it should be fine?
+
+% In this example, we force wwj with the case binder, but we don't really
+% consume it (more precise definition of consume...), so we can use the case
+% binder meaning we're simply using x for the first time. Needs to be
+% formalised....
 
@@
 
+\section{Work in progress}
+
+\subsection{In Result of TcGblEnv axioms}
+
+I don't understand this one yet. I would guess it has to do with multiplicity
+coercions. Incorrect incompatible branch: CoAxBranch (src/Prelude/Linear/GenericUtil.hs:112:3-51):
+\begin{code}
+type family Fixup (f :: Type -> Type) (g :: Type -> Type) :: Type -> Type where
+  Fixup (D1 c f) (D1 _c g) = D1 c (Fixup f g)
+  Fixup (C1 c f) (C1 _c g) = C1 c (Fixup f g)
+  Fixup (S1 c f) (S1 _c (MP1 m f)) = S1 c (MP1 m f) -- error in this constructor
+  Fixup (S1 c f) (S1 _c f) = S1 c f
+  Fixup (f :*: g) (f' :*: g') = Fixup f f' :*: Fixup g g'
+  Fixup (f :+: g) (f' :+: g') = Fixup f f' :+: Fixup g g'
+  Fixup V1 V1 = V1
+  Fixup _ _ = TypeError ('Text "FixupMetaData: representations do not match.")
+\end{code}
+
+% \section{Core to Core Validation}
+\subsection{Inlining}
+
+\subsection{CSE}
+
 \subsection{Join Points}
@@
 
-\section{Core2Core Validdations}
+\subsection{Empty Case}
 
-\subsection{INlining}
-\subsection{CSE}
+For case expressions, the usage environment is computed by checking all branches
+and taking sup. However, this trick doesn't work when there are no branches.
+
+\begin{itemize}
+\item https://gitlab.haskell.org/ghc/ghc/-/issues/20058
+\item https://gitlab.haskell.org/ghc/ghc/-/issues/18768
+
+\item (1) Just like a case expression remembers its type (Note [Why does Case have a
+'Type' field?] in Core.hs), it should remember its usage environment. This data
+should be verified by Lint.
+
+\item (2) Once this is done, we can remove the Bottom usage and the second field of
+UsageEnv. In this step, we have to infer the correct usage environment for empty
+case in the typechecker.
+\end{itemize}
+
+\begin{code}
+{-# LANGUAGE LinearTypes, EmptyCase #-}
+module M where
+
+{-# NOINLINE f #-}
+f :: a %1-> ()
+f x = case () of {}
+\end{code}
+
+This example is well typed: a function is linear if it consumes its argument
+exactly once when it's consumed exactly once. It seems like the function isn't
+linear since it won't consume x because of the empty case, however, that also
+means f won't be consumed due to the same empty case, thus linearity is
+preserved.
+
+\begin{code}
+* In the case of empty types (see Note [Bottoming expressions]), say
+  data T
+we do NOT want to replace
+  case (x::T) of Bool {}   -->   error Bool "Inaccessible case"
+because x might raise an exception, and *that*'s what we want to see!
+(#6067 is an example.) To preserve semantics we'd have to say
+   x `seq` error Bool "Inaccessible case"
+but the 'seq' is just such a case, so we are back to square 1.
+\end{code}
+
+There are three different problems:
+
+\begin{itemize}
+\item castBottomExpr converts (case x :: T of {}) :: T to x.
+\item Worker/wrapper moves the empty case to a separate binding
+\item CorePrep eliminates empty case, just like point 1 (See -- Eliminate empty
+    case in GHC.CoreToStg.Prep
+\end{itemize}
+
+With castBottomExpr, we get the example above to
+\begin{code}
+    f = \ @a (x (%1) :: a) -> ()
+\end{code}
+And if we don't, 
+\begin{code}
+    f = \ @a (x (%1) :: a) -> case () of {}
+\end{code}
+And that supposedly if we had a usage environment in the case expression we
+could avoid the error. How is it typed without the transformation in face
+of the bottom? (Even knowing that theoretically it's because of divergence?)
 

### 2022-11-25 proposal.tex
Add references

@@
 
-This document is a work in progress proposal that I will deliver at the end of
+This document is a work in progress proposal that I will deliver (in
+an extended form) at the end of
 the semester as part of my master thesis preparation. The document needs a
@@
 those already familiar with the problem and serves as a (less bureaucratic)
-research proposal on linting linearity in Core and to showcase the progress I
+proposal on linting linearity in Core and to showcase the preliminary progress I
 have made so far.
@@
 
-Since Linear Haskell's\cite{} publication and implementation release in GHC 9.0,
+Since the publication of Linear Haskell~\cite{linearhaskell} and its release in GHC 9.0,
 Haskell's type system supports linearity annotations in functions -- bringing
-linear types into a mainstream pure and lazy functional language. Concretely,
+linear types to a mainstream, pure, and lazy functional language. Concretely,
 function types can be annotated with a multiplicity (a multiplicity of One
@@
 Linearly typed programs are proven/checked to be linear by the type system, so
-the addition of linear types evidently implied changing the type checker to
-support linearity. There exist, however, two distinct type checkers in Haskell.
+the addition of linear types evidently required changing the type checker to
+support linearity. There exist, however, two distinct type checkers in GHC.
 The first is run on the surface language, i.e. the Haskell we write, and is a
-very complex type checker that now also supports typing linearity. The second is
-run on the intermediate language \emph{Core} that we obtain from desugaring
-Haskell.
+complex type checker that now also supports typing linearity. The second is
+run on programs written in the intermediate language \emph{Core}, that
+are obtained from \emph{desugaring} Haskell.
 
-\emph{Core} is a much smaller language than the whole of Haskell (even though we
-can compile the whole of Haskell to it!) and the type checker for it is small
-and fast due to \emph{Core} being explicitly typed and much closer to a
-polymorphic lambda calculus than its surface-level counterpart. This type
+Core is a much smaller language than the whole of Haskell (even though we
+can compile the whole of Haskell to it!) and its type checker is
+simple and fast due to Core being explicitly typed. In essence, Core
+is close to a higher-order polymorphic lambda calculus. This type
 checker (called \emph{Lint}) gives us guarantees of correctness in face of all
 the complex transformations a Haskell program undergoes, such as desugaring and
-core-to-core passes, because the linter is always run on the resulting code
-before being compiled to untyped machine code.
+core-to-core optimization passes, because the linter is always run on the resulting code
+before being ultimately compiled to (untyped) machine code.
 
@@
 
-We want Core to give us guarantees about our desugaring and transformations with
-regard to linearity too -- a linearly typed Core will ensure that linearly-typed
-programs remain correct (linearity is preserved) after desugaring and all GHC
-transformations (optimisations shouldn't destroy linearity).
+We want Core and its type system to give us guarantees about
+desugaring and transformations with regard to linearity -- a linearly
+typed Core ensures that linearly-typed programs remain correct
+(i.e.,~linearity is preserved) after desugaring and all GHC
+transformations (i.e.,~optimisations should not destroy linearity).
 
-Core is already annotated with linearity, but the \textbf{linter currently
-ignores it}! In spite of the strong formal foundations of linear types driving
-the implementation, their interaction with the whole of GHC is still far from
-trivial, and the implemented type system cannot accomodate various optimising
-transformations and, from them resulting, programs that seemingly violate
-linearity but which do preserve it semantically; for example, laziness makes it
-harder to reason about linearity. Therefore, we reject various valid programs
-with regard to linearity after desugaring: the current solution, because
-disabling optimisations can inccur great performance costs, is to disable the
-linear linter. However, we believe that GHC's transformations are correct, and
-it is the linear type system that can't accommodate the resulting programs.
+In this sense, Core is already annotated with linearity, but the \textbf{linter currently
+  ignores linearity annotations}.
+%
+In spite of the strong formal foundations of linear types driving the
+implementation, their interaction with the whole of GHC is still far
+from trivial, and the implemented type system cannot accomodate
+several optimising transformations that produce programs that violate
+linearity in a syntactically (i.e.,~multiple occurrences of linear
+variables in the program text), but ultimately preserve it in a
+semantic sense, where a linear term is still consumed exactly once --
+this is compounded by lazy evaluation driving a non-trivial mismatch
+between syntactic and semantic linearity. Therefore, linear linting
+rejects various valid programs (with regard to linearity) after
+desugaring. The current solution to this is to effectively disable the
+linear linter, since disabling optimisations can incur significant
+performance costs. However, we believe that GHC's transformations are
+correct, and it is the linear linter and its underlying type system
+that that cannot sufficiently accommodate the
+resulting programs.
 % diverse problems with linearity spring up when we get past the desugarer and
@@
 
-We propose to extend Core's type system with additional linearity information to
-be able to accommodate linear programs as seen in Core after optimising
-transformations. We will prove the system's soundness/type safety and validate
-some core-to-core optimisation passes. Ultimately, we will implement the
-extension into GHC's linter. Concretely, we'll extend Core's linear type system
-with rules based on usage annotations for let, letrec and case binder bound
+We propose to extend Core's type system with additional linearity
+information in order to to accommodate linear program in Core that
+result from optimising transformations. We will prove the system's
+soundness/type safety and show how it validates some core-to-core optimisation
+passes. We also plan to implement the extension into GHC's
+linter. Concretely, we will extend Core's linear type system with
+\emph{usage annotations} for let, letrec and case binder bound
 variables. Additionally, we will explore a new kind of coercion for
-multiplicities to validate programs that combine GADTs with multiplicities.
+multiplicities to validate programs that combine GADTs with
+multiplicities.
 %
-The ultimate measure of success is the -dlinear-core-lint flag that, right now,
-when enabled, rejects many linearly valid programs in the linter. Ideally, by
+The ultimate measure of success is the \verb=-dlinear-core-lint= flag,
+which activates the linear linter. In its current implementation,
+enabling this flag rejects many linearly valid programs. Ideally, by
 the end of our research and implementation, this flag could be enabled by
@@
 accept as many diverse transformations as possible while still preserving
-linearity, even if we are unable to fix them all.
+linearity, even if we are unable to account for all of them.
 
-In section~\ref{typingUsageEnvs} we describe usage environments and how they
-solve three distinct problems. In section~\ref{multiplicityCoercions} we discuss
-the beginning of multiplicity coercions. In section~\ref{examples} we take
+In Section~\ref{typingUsageEnvs} we describe usage environments and how they
+solve three distinct problems. In Section~\ref{multiplicityCoercions} we discuss
+the beginning of multiplicity coercions. In Section~\ref{examples} we take
 programs that are currently rejected and show how the type system with our
@@
 
-Haskell has call-by-need semantics that entail that an expression isn't
-evaluated when it is let bound but rather when the binding variable is used.
-This makes it hard to reason about linearity for let bound variables. The
-following example doesn't typecheck in Haskell, but semantically it is indeed
-linear because the let binding is lazy.
+Haskell has call-by-need semantics that entail that an expression is not
+evaluated when it is let-bound but rather when the binding variable is used.
+This makes it challenging to reason about linearity for let-bound variables. The
+following example fails to typecheck in Haskell, but semantically it is indeed
+linear due to lazy evaluation:
 
@@
 Despite not being accepted by the surface-level language, linear programs using
-lets can occur in Core due to optimising transformations that create let
+lets occur in Core due to optimising transformations that create let
 bindings. In a similar fashion, programs which violate syntatic linearity
-for other reasons other than let bindings are formed from Core transformations.
+for other reasons other than let bindings are produced by Core transformations.
 
@@
 In the above example, this would amount to annotating $y$ with a usage
-environemnt $x := 1$ (because the expression bound by $y$ uses $x$ one time).
+environment $x := 1$ (because the expression bound by $y$ uses $x$ one time).
 Upon using $y$, we are actually using $x$ one time, and therefore linearity is
@@
 environment annotations. Multiple transformations can introduce let-bindings,
-such as CSE and join points. By extending the type system to allow lets in core
-we start paving our way to a linear linter.
+such as CSE and join points. By extending the type system to allow let-bindings in Core
+we start paving the way to a linear linter.
 
-Right now, Core annotates every variable with a multiplicity at its binding
-site. The multiplicity for let-bound variables must be ignored thorought the
+Currently, every variable in Core is annotated with a multiplicity at its binding
+site. The multiplicity for let-bound variables must be ignored throughout the
 transformation pipeline or otherwise too many programs valid programs would be
@@
 to every binder, we want to associate a multiplicity if the variable is lambda
-bound and a usage environment when it is let bound. We then instance the usage
+bound and a usage environment when it is let-bound. We then instantiate the usage
 environment solution to lets in particular -- a let bound variable is annotated
@@
 
-Take for example an expression in which $y$ and $z$ are lambda bound with a
-multiplicity of exactly one. In the following code it might not look as if $y$
+Take for example an expression in which $y$ and $z$ are lambda-bound with a
+multiplicity of one. In the following code it might not appear as if $y$
 and $z$ are both being consumed linearly, but indeed they are since in the first
@@
 and $z$ directly on the second branch. Note again that let binding $x$ doesn't
-consume $y$ and $z$ because of laziness. Only \emph{using} $x$ consumes $y$ and
+consume $y$ and $z$ because of lazy evaluation. Only \emph{using} $x$ consumes $y$ and
 $z$.
@@
 
-The tricky part about determining the usage environment of the variables bound
-by a recursive let is knowing what usage to emit inside their own body when
+The challenging part about determining the usage environment of the variables bound
+by a recursive let is knowing what usage to emit inside their own body, when
 computing said usage environment. The example below uses $x$ linearly % \ref{example_letrec}
@@
 The key idea to computing the usage environment of multiple variables that use
-themselves in their body is to do the computation in two separate passes. First,
+themselves in their definition body is to perform the computation in two separate passes. First,
 calculate a naive environment by emiting free variables multiplicities as usual
@@
 algorithm~\ref{computeRecUsages} to get a final usage environment. Intuitively,
-the algorithm, for each recursive binder, goes over all (initially naive) usage
-environment and substitutes the recursive binder by the corresponding usage
+the algorithm, for each recursive binder, iterates over all (initially naive) usage
+environments and substitutes the recursive binder by the corresponding usage
 environment, scaled up by the amount of times that recursive binder is used in
@@
 
-We present ahead an algorithm for computing the usage environment of a set of
+We describe below an algorithm for computing the usage environment of a set of
 recursive definitions, and its textual description. An (at least informal) proof
-that the algorithm works should eventually follow this.
+that the algorithm is correct should eventually follow this.
 
@@
 recursive-let bound variables increases quadratically, i.e. the algorithm has
-complexity $O(n^2)$, but this is not a problem since it's rare to have more than
+$O(n^2)$ complexity, but this is not a problem since it's rare to have more than
 a handful of binders in the same recursive let block.
@@
 
-Putting the usage environment idea together with the algorithm we get the following typing rule:
+Putting the usage environment idea together with the algorithm we obtain the following typing rule:
 
@@
 
-Instancing the usage environment solution according to the above description,
+Instantiating the usage environment solution according to the above description,
 here's an informal proof that the previous example is well typed: We %\ref{example_letrec}
@@
 
-\subsection{Handling the case binder}
+\subsection{Handling the case binder\label{casebinder}}
 
@@
 all branches. However, we hypothetize that this prevents us from typing many
-valid programs and ultimately doesn't capture correctly the usage of variable,
+valid programs and ultimately does not correctly capture the usage of a variable,
 and present a seemingly viable alternative.
 
-Firstly, we remember the definition of \emph{consuming a value} from Linear
+Firstly, we recall the definition of \emph{consuming a value} from Linear
 Haskell as
@@
 the following case expression consumes the scrutinee twice because the case
-binder is used. However, we know for sure that in the branch where z is used, z
-is equal to False, and using False doesn't violate linearity since we can
+binder is used. However, we know that in the branch where z is used, z
+is equal to False, and using False does not violate linearity since we can
 unrestrictedly create nullary data constructors. A situation similar to this one
@@
 
-The good news is both these two programs and a handful of others listed in
-Section~\ref{examples} are accepted with a new rule we've devised for the
+The good news is that both these programs (and a handful of others listed in
+Section~\ref{examples}) are accepted with a new rule we have devised for the
 linear linter. We still need to prove we don't accept any invalid programs as
@@
 
-The definition for $\$!$ in \textbf{linear-base}\cite{} fails to lint after the
+The definition for $\$!$ in \textbf{linear-base}\cite{linearbase} fails to lint after the
 desugarer (before any optimisation takes place) with \emph{Linearity failure in
@@
 $f$. However, the $x$ passed as an argument is actually the case binder and it
-must be handled in its own way. The case binder is the key (as seen in ~\ref{caseBinderKey}) to solving this
+must be handled in its own way. The case binder is the key (as seen
+in~\ref{casebinder}) to solving this
 and many other examples.
@@
 
+\bibliographystyle{plain}
+\bibliography{references}
+
 \end{document}

### 2022-11-25 proposal.tex
tweaks

@@
 
-Since the publication of Linear Haskell~\cite{linearhaskell} and its release in GHC 9.0,
+Since the publication of Linear Haskell~\cite{} and its release in GHC 9.0,
 Haskell's type system supports linearity annotations in functions -- bringing
@@
 
-\subsection{Handling the case binder\label{casebinder}}
+\subsection{Handling the case binder}
 
@@
 
-The definition for $\$!$ in \textbf{linear-base}\cite{linearbase} fails to lint after the
+The definition for $\$!$ in \textbf{linear-base}\cite{} fails to lint after the
 desugarer (before any optimisation takes place) with \emph{Linearity failure in
@@
 $f$. However, the $x$ passed as an argument is actually the case binder and it
-must be handled in its own way. The case binder is the key (as seen
-in~\ref{casebinder}) to solving this
+must be handled in its own way. The case binder is the key (as seen in ~\ref{caseBinderKey}) to solving this
 and many other examples.
@@
 
-\bibliographystyle{plain}
-\bibliography{references}
-
 \end{document}

### 2022-11-26 proposal.tex
Tweaks and small extensions

--- 2022-11-25 proposal.tex
@@
 
-This document is a work in progress proposal that I will deliver (in
-an extended form) at the end of
-the semester as part of my master thesis preparation. The document needs a
-proper introduction and lengthy background and related work section besides the
-content (to be extended) that follows, such that an outsider might understand
-what I propose to do. However, this initial iteration is instead targeted at
-those already familiar with the problem and serves as a (less bureaucratic)
-proposal on linting linearity in Core and to showcase the preliminary progress I
-have made so far.
+This document is a work in progress proposal that I will deliver (in an extended
+form) at the end of the semester as part of my master thesis preparation. The
+document still needs a proper introduction, lengthy background and related work
+section  such that an outsider might understand what I propose to do. However,
+this initial iteration is instead targeted at those already familiar with the
+problem and serves as a (less bureaucratic) proposal on linting linearity in
+Core and to showcase the preliminary progress I have made so far.
 
@@
 function is linear if it consumes its arguments exactly once when it itself is
-consumed exactly once (its arguments have multiplicity of one), for some
-definition of \emph{consume} that I should revise here.
+consumed exactly once.
+% TODO: for some definition of \emph{consume} that I should revise here.
 
@@
 core-to-core optimization passes, because the linter is always run on the resulting code
-before being ultimately compiled to (untyped) machine code.
+before ultimately being compiled to (untyped) machine code.
 
@@
 implementation, their interaction with the whole of GHC is still far
-from trivial, and the implemented type system cannot accomodate
+from trivial. The implemented type system cannot accomodate
 several optimising transformations that produce programs that violate
-linearity in a syntactically (i.e.,~multiple occurrences of linear
+linearity syntactically (i.e.,~multiple occurrences of linear
 variables in the program text), but ultimately preserve it in a
@@
 desugaring. The current solution to this is to effectively disable the
-linear linter, since disabling optimisations can incur significant
+linear linter, since otherwise disabling optimisations can incur significant
 performance costs. However, we believe that GHC's transformations are
 correct, and it is the linear linter and its underlying type system
-that that cannot sufficiently accommodate the
+that cannot sufficiently accommodate the
 resulting programs.
@@
 We propose to extend Core's type system with additional linearity
-information in order to to accommodate linear program in Core that
+information in order to accommodate linear programs in Core that
 result from optimising transformations. We will prove the system's
@@
 Despite not being accepted by the surface-level language, linear programs using
-lets occur in Core due to optimising transformations that create let
+lets occur naturally in Core due to optimising transformations that create let
 bindings. In a similar fashion, programs which violate syntatic linearity
@@
 site. The multiplicity for let-bound variables must be ignored throughout the
-transformation pipeline or otherwise too many programs valid programs would be
+transformation pipeline or otherwise too many valid programs would be
 rejected for violating linearity in its transformed type.
@@
 
-We describe below an algorithm for computing the usage environment of a set of
-recursive definitions, and its textual description. An (at least informal) proof
+The algorithm for computing the usage environment of a set of
+recursive definitions works as follows. An at least informal proof
 that the algorithm is correct should eventually follow this.
@@
 
+% (though in the rule seems to only start counting after the second usage which
+% would make the rule wrong for a case expression that doesn't use the binder --
+% the scrutinee should also be consumed)
 The current typing rule for case expressions describes that using the case
-binder is as using the case scrutinee multiple times (though it seems like it
-actually just starts counting after the second usage which would make the rule
-wrong for a case expression that doesn't use the binder -- the scrutinee should
-also be consumed), so we scale the usage of
+binder is as using the case scrutinee multiple times, so we scale the usage of
 the case scrutinee by the superior multiplicity of using the case binder across
 all branches. However, we hypothetize that this prevents us from typing many
-valid programs and ultimately does not correctly capture the usage of a variable,
-and present a seemingly viable alternative.
+valid programs and ultimately does not correctly capture the usage of a
+variable and present a seemingly viable alternative.
 
@@
 
-Now, with the following example, my interpretation of the current rule says that
-the following case expression consumes the scrutinee twice because the case
-binder is used. However, we know that in the branch where z is used, z
-is equal to False, and using False does not violate linearity since we can
-unrestrictedly create nullary data constructors. A situation similar to this one
-below happens after the CSE transformation.
+Now, with the following example, my interpretation of the current rule dictates
+that the case expression consumes the scrutinee twice because the case binder is
+used. However, we know that in the branch where $z$ is used, $z$ is equal to
+False, and using False does not violate linearity since we can unrestrictedly
+create nullary data constructors. A situation similar to this one below happens
+after the CSE transformation.
 \begin{code}
@@
 
-The key idea of the case-binder-usage solution is annotating the case binder
-with independent usage environments for each pattern match. That is, for
+The key idea of the case-binder-usage solution is \textbf{annotating the case binder
+with independent usage environments for each pattern match}. That is, for
 each of the possible branches, using the case binder $z$ will equate to
@@
 \end{mathparpagebreakable}
+
+Taking the first example in this section, we annotate both branches (True and
+False) with $U_z = \emptyset$ because the pattern match doesn't introduce any
+new variables. We consume \emph{complex expression} once and $y$ once, and then use
+$z$ once. However, using $z$ once equates to not using anything (semantically
+this relates to being able to use False or True unrestrictedly) and therefore
+linearity is preserved.
+
+In the second example, we know annotate the only branch with usage environment
+$U_z = [x := 1, y := 1]$. Upon the usage of $z$ we emit $x$ and $y$ --
+effectively using both -- thus linearity is preserved.
 

### 2023-01-02 report.tex
Chapters and sections in report document

--- 2022-11-26 proposal.tex
@@
-\documentclass[a4paper, draft]{article}
+\documentclass[a4paper, draft]{report}
 
@@
 \maketitle
-
-% \section{Introduction}
-
-% \subsection{Motivation}
-
-% \begin{itemize}
-%     \item Are we really preserving linearity?
-%     \item Inform/unlock other optimisations that take into account linearity
-% \end{itemize}
-
-% \section{Background}
+\tableofcontents
+
+\chapter{Introduction}
+
+\subsection{Motivation}
+
+\begin{itemize}
+    \item Are we really preserving linearity?
+    \item Inform/unlock other optimisations that take into account linearity
+\end{itemize}
+
+\chapter{Background}
+
+\section{Linear Types}
+
+\section{Haskell}
+
+\section{Linear Haskell}
+
+\section{Core Haskell}
+
+\section{GHC Pipeline}
 
@@
 
-% \section{Related Work}
-
-% \section{Technical Details}
-
-\section{Foreword}
-
-This document is a work in progress proposal that I will deliver (in an extended
-form) at the end of the semester as part of my master thesis preparation. The
-document still needs a proper introduction, lengthy background and related work
-section  such that an outsider might understand what I propose to do. However,
-this initial iteration is instead targeted at those already familiar with the
-problem and serves as a (less bureaucratic) proposal on linting linearity in
-Core and to showcase the preliminary progress I have made so far.
+\chapter{Related Work}
+
+\section{System FC}
+
+\section{Linear Haskell}
+
+\chapter{Technical Details}
+
+% \section{Foreword}
+
+% This document is a work in progress proposal that I will deliver (in an extended
+% form) at the end of the semester as part of my master thesis preparation. The
+% document still needs a proper introduction, lengthy background and related work
+% section  such that an outsider might understand what I propose to do. However,
+% this initial iteration is instead targeted at those already familiar with the
+% problem and serves as a (less bureaucratic) proposal on linting linearity in
+% Core and to showcase the preliminary progress I have made so far.
 

### 2023-01-02 report.tex
Items on background sections

@@
 
+\newcommand{\mypara}[1]{\paragraph{\textbf{#1}.}}
+\newcommand{\lolli}{\multimap}
+\newcommand{\tensor}{\otimes}
+\newcommand{\one}{\mathbf{1}}
+\newcommand{\bang}{{!}}
+
+\newcommand{\llet}[2]{\mathsf{let}\ #1\ \mathsf{in}\ #2}
+
 \title{Linting Linearity in Core/System FC}
@@
 
+\mypara{Type Systems} A type system can be formally described through
+a set of inference rules that inductively define a judgment of the
+form $\Gamma \vdash M : A$, stating that program expression $M$ has
+type $A$ according to the \emph{typing assumptions} for variables
+tracked in $\Gamma$. For instance,
+$x{:}\mathsf{Int}, y{:}\mathsf{Int} \vdash x+y : \mathsf{Int}$ states
+that $x+y$ has type $\mathsf{Int}$ under the assumption that $x$ and
+$y$ have type $\mathsf{Int}$.  An expression $M$ is deemed well-typed
+with a given type $A$ if one can construct a typing derivation with $M :
+A$ as its conclusion, by repeated application of the inference rules.
+
+The simply-typed $\lambda$-calculus is a typed core functional
+language~\cite{10.5555/509043} that captures the essence of a type system in a simple and familiar environment. Its syntax consists of
+functional abstraction, written $\lambda x{:}A.M$, denoting
+an (anonymous) function that takes an argument of type $A$, bound to
+$x$ in $M$; and application $M\,N$, with the standard meaning, and
+variables $x$. For instance, the term $\lambda x{:}A.x$, denoting the identity function, is a functional abstraction
+taking an argument of type $A$ and returning it back.
+
+\mypara{Linear Logic}
+
+Linear logic \cite{DBLP:journals/tcs/Girard87} can be seen as a
+resource-aware logic, where propositions are interpreted as resources
+that are consumed during the inference process.  Where in standard
+propositional logic we are able to use an assumption as many times as
+we want, in linear logic every resource (i.e., every assumption) must
+be used \emph{exactly once}, or \emph{linearly}. This usage
+restriction gives rise to new logical connectives, based on the way
+the ambient resources are used. For instance, conjunction, usually
+written as $A\wedge B$, appears in two forms in linear logic:
+multiplicative or simultaneous conjunction (written $A\tensor B$); and
+additive or alternative conjunction (written $A\with
+B$). Multiplicative conjunction denotes the simultaneous availability
+of resources $A$ and $B$, requiring both of them to be
+used. Alternative conjunction denotes the availability of $A$ and $B$,
+but where only one of the two resources may be used. Similarly,
+implication becomes linear implication, written $A\lolli B$,
+denoting a resource that will consume (exactly one) resource $A$ to
+produce a resource $B$.
+
+To present the formalization of this logic, besides the new
+connectives, we need to introduce the \emph{resource-aware context}
+$\Delta$.  In contrast to the previously seen $\Gamma$, $\Delta$ is
+also a list of variables and their types, but where each and every
+variable must be used exactly once during inference.  So, to introduce
+the connective $\tensor$ which defines a multiplicative pair of
+propositions, we must use exactly all the resources
+($\Delta_1, \Delta_2$) needed to realize the ($\Delta_1$)
+proposition $A$, and ($\Delta_2$) proposition $B$:
+\[
+    \infer*[right=($\tensor I$)]
+    {\Delta_1 \vdash M : A \and \Delta_2 \vdash N : B}
+    {\Delta_1 , \Delta_2 \vdash (M \tensor N) : A \tensor B}
+\]
+Out of the logical connectives, we need to mention one more, since it
+augments the form of the judgment and it's the one that ensures
+logical strength i.e. we're able to translate intuitionistic logic
+into linear logic.  The proposition $\bang A$ (read \emph{of course}
+$A$) is used (under certain conditions) to make a resource
+``infinite'' i.e. to make it useable an arbitrary number of times. To
+distinguish the ``infinite'' variables, a separate, unrestricted,
+context is used -- $\Gamma$. So $\Gamma$ holds the ``infinite''
+resources, and $\Delta$ the resources that can only be used once.  The
+linear typing judgment for the introduction of the exponential $\bang A$
+takes the form:
+\[
+    \infer*[right=($\bang I$)]
+    {\Gamma ; \emptyset \vdash M : A}
+    {\Gamma ; \emptyset \vdash \bang M : \bang A}
+\]
+Logically, a proof of $\bang A$ cannot use linear resources since
+$\bang A$ denotes an unbounded (potentially $0$) number of copies of
+$A$. Proofs of $\bang A$ may use other unrestricted or exponential
+resources, tracked by context $\Gamma$.
+From a computational perspective, the type $\bang A$
+internalizes the simply-typed $\lambda$-calculus in the linear
+$\lambda$-calculus.
+
+The elimination form for the exponential, written $\llet{ !u
+= M}{ N}$, warrants the use of resource $A$ an unbounded
+number of times in $N$ via the variable $u$:
+\[
+    \infer*[right=($\bang E$)]
+    {\Gamma ; \Delta_1 \vdash M : \bang A \and \Gamma, u{:}A ; \Delta_2 \vdash N : C}
+    {\Gamma ; \Delta_1, \Delta_2 \vdash \llet{ !u = M}{ N} : C}
+\]
+
+% TODO: Revise the story above instead of dumping so much information
+
+\mypara{Linear Types}
+
+Girard's \emph{linear logic} suggests a new \emph{type system} for functional
+languages through the Curry-Howard isomorphism -- every linear logic proposition
+maps to a different linear type.
+
+Values belonging to a linear type must be used exactly once. For example, the
+type corresponding to the linear implication ($\lolli$), the linear function, is
+a function that consumes its argument exactly once. A linear function that
+receives an integer as an input and consumes it exactly once to
+produce a character would have type $\textbf{int} \lolli
+\textbf{char}$.
+
+\begin{itemize}
+    \item Ideia dos linear types
+    \item Calculo lambda linear (como forma de falar da linearidade)
+    \item Gestão de recursos
+    \item Rust (core is linear lambda calculus?)
+\end{itemize}
+
 \section{Haskell}
 
+\begin{itemize}
+    \item Features fundamentais
+    \item Exemplos, introduzir syntax
+    \item (Ao contrário de OCaml) tem um sistema de tipos muito sufesticado
+    \item Eventualmente explicação de GADTs? c/ exemplo do Vec
+    \item Uma das features mais avaçadas/experimentais é o Linear Haskell
+\end{itemize}
+
 \section{Linear Haskell}
 
+\begin{itemize}
+    \item Linear Haskell definition
+    \item Consuming values precisely
+    \item Multiplicities
+    \item Multiplicity polymorphism
+    \item Que relaciona o systemFC com o calculo lambda linear
+\end{itemize}
+
 \section{Core Haskell}
 
+\begin{itemize}
+    \item What is Core (IR) e as particularidades
+    \item Em cima do SystemFC
+    \item Optimizações feitas no Core
+    \item Base SystemFC tem de ser "merged" com o linear lambda calculus
+    \item Ou porque é que é importante o Core ter linearidade também/a que nível
+        foi introduzido
+    \item Falar das coercions em especifico
+    \item E das local equalities que são implicitas no front end
+\end{itemize}
+
 \section{GHC Pipeline}
 
+\begin{itemize}
+    \item Parser -> Rename -> Typechecker -> Desugar
+    \item Core2Core transformations
+    \item GHC unique em haver tantas transformações sempre sobre a mesma intermediate
+        representation como input e output
+    \item 
+\end{itemize}
+
+% Previous itemize. Talvez ainda tenha de ter secções sobre os últimos dois
+% pontos
 % \begin{itemize}
@@
 \section{Linear Haskell}
+
+\section{OutsideIn(X)}
+
+Se for modificar o typechecker com as multiplicity coercions vou ter de falar
+disto.
 

### 2023-01-02 report.tex
Expand a bit on the Haskell section

@@
 
-\mypara{Linear Types}
+% \mypara{Linear Types}
+\subsection{Linear Types}
 
@@
 \section{Haskell}
+
+Haskell is a functional programming language defined by the "Haskell
+Report"\cite{jones1999haskell,marlow2010haskell} and whose \emph{de-facto}
+implementation is GHC, the Glasgow Haskell Compiler\cite{GHC}. Haskell is a
+\emph{non-strict}, \emph{purely functional} language, i.e., functions cannot
+have side effects or mutate data, and contrary to many programming languages,
+arguments are \emph{not} evaluated when passed to functions, but rather are only
+evaluated when the value is needed.
+
+A function to compute the length of a list in Haskell is defined by pattern
+matching on the list and returning $0$ if the list is empty and adding one to
+the (computed recursively) length of the rest of the list otherwise. The
+following snippet defines the said \textbf{length} function.
+\begin{code}
+    length [] = 0
+    length (x:xs) = 1 + length xs
+\end{code}
+The function \textbf{length} receives a list and returns an integer, and since
+it treats all list's elements the same, it is said to be polymorphic over the
+list's elements type. In Haskell, function signatures typically appear above the
+definition, and for this particular function we'd have
+\begin{code}
+    length :: [a] -> Int
+\end{code}
+where \textbf{a} is a universally quantified type variable, that is,
+\textbf{length} can be called on lists of ints, lists of characters, lists of
+lists of strings, lists of anything.
+
+The laziness evaluation strategy entails expressions are only evaluated when
+needed. If we were to call just the above function on a list of computationally
+expensive elements, such as
+\begin{code}
+    length [5^1234, 6^4567, 8^(9^42)]
+\end{code}
+We would get the result 3 instantaneously: the elements of the list would never
+be evaluated because the \textbf{length} function does not require the value of
+each element to compute a result.
+
+% TODO:
+Is it outside the scope of this report to discuss the merits and disadvantages
+of laziness and purely functional properties? Não estou nada contente com esta
+secção em geral. Acho que preciso de escrever o resto para conseguir alguma
+coerência.
+
+Another defining feature of Haskell is its powerful type system....
 
@@
 disto.
+
+\section{Rust}
+
+They have a linearly typed something?
 

### 2023-01-09 report.tex
Introduction on linear types

@@
 
+Linear type systems\cite{} add expressiveness to existing type systems by
+defining that linear resources must be used exactly once (linearly). In
+programming languages with a linear type system, not using certain resources or
+using them twice are type errors. Linear types can, thus, be used, e.g., to
+guarantee file handles, socket descriptors and allocated memory are freed
+exactly once (leaks and double-frees become type errors)\cite{}, implement safe
+high-performance language interop\cite{}, guarantee quantum entangled states are
+not duplicated~\cite{}, handle mutable state\cite{}, a bit on session
+types guarantees\cite{}, and much more.
+
+% TODO: Also include bit about how it makes the basis of other type systems such
+% as rusts owneship. How ?
+
+As an example, consider the following program which allocates some memory and
+frees it twice. The variable $p$ is a linear resource created by the
+\textbf{malloc} keyword.
+
+\begin{code}
+let p = malloc 4
+ in free p
+    free p
+\end{code}
+
+Because $p$ is a linear variable, it must be used exactly once. However, we are
+using it in both calls to \textbf{free}. The example program doesn't typecheck!
+In section~\ref{linear-types} we describe the fundamentals of linear types in
+more detail with a formal definition and many examples.
+
+Despite their usefulness, most mainstream programming languages don't have
+linear types. In practice, it can be hard to balance linear and non-linear types
+and avoid the boilerplate threading of linear resources that we'll see ahead.
+
+% The main contributions of this thesis are:
+% \begin{itemize}
+%     \item Core Type system which accepts optimized linear programs
+% \end{itemize}
+
+% stores a value to the allocated memory, reads from it and finally frees
+
+% Linear type systems add expressiven at the type level regarding values that
+% must be used exactly once thorought their lifetime.
+
+% about the usage of resources in a
+% programming language.
+
+\section {Introduction}
+
+A história que eu quero contar:
+
 \subsection{Motivation}
@@
 
-\section{Linear Types}
+\section{Linear Types\label{linear-types}}
+
+Programming languages with linear type systems might differ in their definition of linear
+resources, but usually \emph{variables} are considered linear resources
+% That is, we can rely on the type system to prove properties regarding the usage of certain resources.
+The \emph{Hello World} of linearly typed programs is a linearly typed pointer.
+In a non-memory managed language, an allocated pointer must be freed
+\emph{exactly once}.
+
+\begin{code}
+let p = malloc(5)
+    p' = store(p, "x")
+    (p'', v) = load(p)
+    free p''
+\end{code}
+
+% We'd like to hide the boilerplate involved \cite{Linear constraints}
+
+In order to give well defined semantics to a linear type system, we present a
+linearly typed lambda calculus which is a very simple language with linear
+types.
 
@@
 
-\section{Linear Haskell}
+\subsection{Linear Haskell}
 
@@
 
-\section{Core Haskell}
+\subsection{Core Haskell}
 

### 2023-01-09 report.tex
Extend linear types introduction story

@@
 Linear type systems\cite{} add expressiveness to existing type systems by
-defining that linear resources must be used exactly once (linearly). In
+stating that linear resources must be used exactly once (linearly). In
 programming languages with a linear type system, not using certain resources or
-using them twice are type errors. Linear types can, thus, be used, e.g., to
-guarantee file handles, socket descriptors and allocated memory are freed
+using them twice are type errors. Linear types can, thus, be used to, e.g.,
+guarantee that file handles, socket descriptors, and allocated memory are freed
 exactly once (leaks and double-frees become type errors)\cite{}, implement safe
-high-performance language interop\cite{}, guarantee quantum entangled states are
-not duplicated~\cite{}, handle mutable state\cite{}, a bit on session
-types guarantees\cite{}, and much more.
+high-performance language interoperability\cite{}, guarantee that quantum
+entangled states are not duplicated~\cite{}, handle mutable state\cite{},
+something on session types guarantees\cite{}, and much more.
 
@@
 
-As an example, consider the following program which allocates some memory and
-frees it twice. The variable $p$ is a linear resource created by the
-\textbf{malloc} keyword.
+As an example, consider the following linearly typed program which allocates
+some memory and frees it twice. The variable $p$ is a linear resource created by
+the \textbf{malloc} keyword.
 
@@
 Because $p$ is a linear variable, it must be used exactly once. However, we are
-using it in both calls to \textbf{free}. The example program doesn't typecheck!
-In section~\ref{linear-types} we describe the fundamentals of linear types in
-more detail with a formal definition and many examples.
-
-Despite their usefulness, most mainstream programming languages don't have
-linear types. In practice, it can be hard to balance linear and non-linear types
-and avoid the boilerplate threading of linear resources that we'll see ahead.
+using it in two different calls to \textbf{free}. The example program
+\emph{doesn't typecheck}! In section~\ref{linear-types} we describe the
+fundamentals of linear types in more detail with a formal definition and many
+examples.
+
+Despite their great promise and research literature, few programming languages
+have linear types --
+% , with Rust\cite{} (with its ownership types), Idris\{}, and recently
+% Haskell\cite{} being the most popular.
+it is hard to balance linear and non-linear types in a pratical language. Linear
+Haskell\cite{linearhaskell} tackles this difficulty and describes a type system
+to retrofit linear types into Haskell, a purely-functional lazy language, while
+retaining backwards-compatibility and code reuse across linear and non-linear
+users of libraries.
+% , and, e.g., avoid the required boilerplate threading of linear resources that we will get to know ahead.
+
+% In an attempt to bridge the gap between the theory and practicality of linear
+% types, Linear Haskell\cite{} describes , the 9.0 version series of Haskell's
+% \emph{de facto} compiler, GHC, supports linear types. However, retrofitting
+% linear types to a purely-functional lazy language
 

### 2023-01-10 report.tex
Nova thesis lightweight template
Improve thesis theme
Extend introduction on Haskell's linear types

@@
-\documentclass[a4paper, draft]{report}
+\documentclass[]{lwnovathesis}
 
@@
 \begin{document}
+
+\frontmatter
+
 \maketitle
@@
 
+\mainmatter
+
 \chapter{Introduction}
 
+\section{Introduction}
+
 Linear type systems\cite{} add expressiveness to existing type systems by
-stating that linear resources must be used exactly once (linearly). In
-programming languages with a linear type system, not using certain resources or
-using them twice are type errors. Linear types can, thus, be used to, e.g.,
-guarantee that file handles, socket descriptors, and allocated memory are freed
-exactly once (leaks and double-frees become type errors)\cite{}, implement safe
+stating that linear resources must be used \emph{exactly once}. In programming
+languages with a linear type system, not using certain resources or using them
+twice are type errors. Linear types can thus be used to, for example, guarantee
+that file handles, socket descriptors, and allocated memory are freed exactly
+once (leaks and double-frees become type errors)\cite{}, implement safe
 high-performance language interoperability\cite{}, guarantee that quantum
-entangled states are not duplicated~\cite{}, handle mutable state\cite{},
+entangled states are not duplicated~\cite{}, handle mutable state safely\cite{},
 something on session types guarantees\cite{}, and much more.
@@
 
-Despite their great promise and research literature, few programming languages
-have linear types --
-% , with Rust\cite{} (with its ownership types), Idris\{}, and recently
-% Haskell\cite{} being the most popular.
-it is hard to balance linear and non-linear types in a pratical language. Linear
-Haskell\cite{linearhaskell} tackles this difficulty and describes a type system
-to retrofit linear types into Haskell, a purely-functional lazy language, while
-retaining backwards-compatibility and code reuse across linear and non-linear
-users of libraries.
-% , and, e.g., avoid the required boilerplate threading of linear resources that we will get to know ahead.
+Despite their great promise and research literature, it is hard to balance
+linear and non-linear types in practice and, consequently, few programming
+languages have linear type systems. Among them are Idris2\cite{}, a linearly and
+dependently typed language based on Quantitative Type Theory, Rust\cite{}, a
+language whose ownership types build on linear ones to guarantee memory safety
+without garbage collection or reference counting, and, more recently,
+Haskell\cite{}, a \emph{purely functional} and \emph{lazy} language.
+
+Linearity in Haskell stands out from linearity in Rust and Idris2 for two
+distinct reasons:
+
+\begin{itemize}
+    \item Linear types were only added to the language roughly \emph{31 years
+        after} Haskell's inception. It is an especially difficult endeavour to
+        add linear types to a well-established language with a large library
+        ecosystem and many active users, rather than to develop the language
+        from the ground up with linear types in mind, and the successful
+        approach as implemented in GHC 9.0, the leading Haskell compiler, was
+        based on Linear Haskell\cite{linearhaskell}, where a linear type system
+        designed with retaining backwards-compatibility and allowing code reuse
+        across linear and non-linear users of the same libraries in mind was
+        described.
+    \item 
+\end{itemize}
+
+% Besides Haskell's supporting linear
+% types according to the said paper, Idris2\cite{} supports linear types in a
+% dependently typed setting, Clean\cite{} has uniqueness types which are closely
+% related to linear types, and Rust\cite{} has ownership types which build from
+% linear types. 
+% released in the Glasgow Haskell Compiler (GHC) version 9.0.
+% , and, e.g., avoid the required boilerplate threading of linear resources that
+% we will get to know ahead.
+
+% Linear types were introduced in GHC 9.0
 
@@
 \end{document}
+
+% TODO:Use glossary

### 2023-01-10 report.tex
Continue describing Haskell linearity differences

@@
 
-Linearity in Haskell stands out from linearity in Rust and Idris2 for two
-distinct reasons:
+Linearity in Haskell stands out from linearity in Rust and Idris2 for, at least,
+the two following distinct reasons:
 
@@
         described.
-    \item 
+    \item Linear types permeate Haskell down to (its) \textbf{Core}, the intermediate
+        language Haskell is desugared to. \textbf{Core} is a very small
+        explicitly typed language to which multiple Core-to-Core optimizing
+        transformations are applied in succession. The final Core program can be
+        typechecked very fast (because of the explicit type annotations) and
+        doing so serves as a sanity check to the correction of the
+        source transformations. If the resulting optimized Core program doesn't
+        typecheck, something \emph{very wrong} has happened in a transformation!
 \end{itemize}
+
+Even though Linear Haskell was successful in integrating linear types in an
+existing language 
+% In spite of the smooth integration of linear types in an existing language
+with regard to backwards compatibility and (re)usability,
+
+The advent of linear types in Haskell bring forth
 
@@
 
-\section {Introduction}
-
-A história que eu quero contar:
-
-\subsection{Motivation}
+\section{Motivation}
 

### 2023-01-11 report.tex
Stuck on linear type background cohesion

@@
 
+% TODO: linear functions to allow safe/controlled use of reified language
+% implementation objects like *single-use continuations* (single use
+% continuations? can we make them faster?)
+
 \chapter{Introduction}
@@
         across linear and non-linear users of the same libraries in mind was
-        described.
+        described. We describe Linear Haskell in detail in
+        section~\ref{linear-haskell-section}.
     \item Linear types permeate Haskell down to (its) \textbf{Core}, the intermediate
@@
         typecheck, something \emph{very wrong} has happened in a transformation!
+        We discuss Core in detail in section~\ref{core-section}.
 \end{itemize}
 
-Even though Linear Haskell was successful in integrating linear types in an
-existing language 
+In theory, the linearity annotations in \textbf{Core} should allow for the same
+typechecking verification we had with the existing type annotations prior to the
+addition of linear types, that is, our Core program with linearity annotations
+should be typechecked after the optimizing transformations...
+
+% Even though Linear Haskell was successful in integrating linear types in an
+% existing language 
 % In spite of the smooth integration of linear types in an existing language
-with regard to backwards compatibility and (re)usability,
-
-The advent of linear types in Haskell bring forth
+% with regard to backwards compatibility and (re)usability,
+
+% The advent of linear types in Haskell bring forth
 
@@
 
+A linear type system is a type system in which ...!
+
+% A linear type system is a type system in which statements such as "this function
+% will use its argument exactly once" can be formally expressed and checked.
+
+Girard's \emph{linear logic} suggests a new \emph{type system} for functional
+languages through the Curry-Howard isomorphism -- every linear logic proposition
+maps to a different linear type.
+
+Values belonging to a linear type must be used exactly once. For example, the
+type corresponding to the linear implication ($\lolli$), the linear function, is
+a function that consumes its argument exactly once. A linear function that
+receives an integer as an input and consumes it exactly once to
+produce a character would have type $\textbf{int} \lolli
+\textbf{char}$.
+
 Programming languages with linear type systems might differ in their definition of linear
-resources, but usually \emph{variables} are considered linear resources
+resources, but usually \emph{variables} are considered linear resources?
 % That is, we can rely on the type system to prove properties regarding the usage of certain resources.
-The \emph{Hello World} of linearly typed programs is a linearly typed pointer.
-In a non-memory managed language, an allocated pointer must be freed
-\emph{exactly once}.
-
-\begin{code}
-let p = malloc(5)
-    p' = store(p, "x")
-    (p'', v) = load(p)
-    free p''
-\end{code}
+% The \emph{Hello World} of linearly typed programs is a linearly typed pointer.
+% In a non-memory managed language, an allocated pointer must be freed
+% \emph{exactly once}.
+
+% \begin{code}
+% let p = malloc(5)
+%     p' = store(p, "x")
+%     (p'', v) = load(p)
+%     free p''
+% \end{code}
 
@@
 implementation is GHC, the Glasgow Haskell Compiler\cite{GHC}. Haskell is a
-\emph{non-strict}, \emph{purely functional} language, i.e., functions cannot
+\emph{lazy}, \emph{purely functional} language, i.e., functions cannot
 have side effects or mutate data, and contrary to many programming languages,
@@
 evaluated when the value is needed.
+
+Another defining feature of Haskell is its powerful type system....
+
+% TODO: File example
 
@@
 % TODO:
-Is it outside the scope of this report to discuss the merits and disadvantages
-of laziness and purely functional properties? Não estou nada contente com esta
-secção em geral. Acho que preciso de escrever o resto para conseguir alguma
-coerência.
-
-Another defining feature of Haskell is its powerful type system....
+% Is it outside the scope of this report to discuss the merits and disadvantages
+% of laziness and purely functional properties?
 
@@
 
-\subsection{Linear Haskell}
+\subsection{Linear Haskell\label{linear-haskell-section}}
 
@@
 
-\subsection{Core Haskell}
+\subsection{Core Haskell\label{core-section}}
 

### 2023-01-11 report.tex
introduction: linearity in core

@@
 
+% Linear types are cool
+% Among the few, Haskell has linear types
+% But Haskell differs from other languages with linear types
+%   * added 31 years after inception
+%   * added to its core
+% Core has linear types for good reasons
+% But they are broken -- why?
+%   * Optimizations changes source -- e.g. if recursive lets are introduced, how to type linearity?
+%
+
 \section{Introduction}
@@
 the \textbf{malloc} keyword.
-
 \begin{code}
@@
         We discuss Core in detail in section~\ref{core-section}.
+    % TODO: \item values in rust are linear by default while non-linear is the
+        % haskell default?
 \end{itemize}
 
-In theory, the linearity annotations in \textbf{Core} should allow for the same
-typechecking verification we had with the existing type annotations prior to the
-addition of linear types, that is, our Core program with linearity annotations
-should be typechecked after the optimizing transformations...
+Core was imbued with linear types because an implementation detail required it,
+and, more importantly, to guarantee linear programs remain linear after being
+desugared and optimized: much the same way a \emph{typed} Core ensures that
+desugaring, and optimizing transformations, are correctly implemented and result
+in a \emph{well-typed} program -- a design decision that has well served the
+compiler developers with a safety net for a long time. It is crucial that a
+program behaviour enforced by linear types, regarding the usage of certain
+resources, is \emph{not} changed in the desugaring or optimization process
+(optimizations should not destroy linearity!) and a linearity aware Core
+typechecker can guarantee that.
+
+Additionally, while not yet a reality, linearity in Core could be used to inform
+certain program optimizations, i.e. having linear types in Core could be used to
+further optimize certain programs and, therefore, benefit the runtime
+performance characteristics of our programs. For example, Linear Haskell\cite{}
+describes as future work an improvement to the inlining optimization: Inlining
+is a centerpiece program optimization primarily because of the subsequent
+optimizing opportunities unlocked by inlining. However, it relies on a heuristic
+process known as \emph{cardinality analysis} to discover safe inlining
+opportunities. Unfortunately, heuristics can be volatile and fail in identifying
+crucial inlining opportunities resulting in programs that fall short of their
+performance potential. On the contrary, the linearity information could be
+integrated in the analysis and used as a (non-heuristic) cardinality
+\emph{declaration}.
+
+However, currently, typechecking linearity in Core is broken in the face of
+optimizations! The reason is not evidently clear: if we can typecheck linearity
+in the surface level Haskell why can't we in Core? In fact, if we were to
+typecheck the programs in Core as they were at the surface level, we would
+succeed, but programs are \emph{transformed} (by semantic-preserving optimizing
+transformations) and hence end up different from how they were originally
+written. For example, the following recursive let definition, where $x$ is a
+linear variable that must be used exactly once, would not typecheck in a source
+linear Haskell program since the typechecker cannot tell that $x$ is used
+linearly, but this program might occur naturally in Core from
+transformations on a program that did succesfully typecheck:
+\begin{code}
+letrec f = \case
+        True -> f False
+        False -> x
+in f True
+\end{code}
+
+Is this program really still linear? If one looks at it
+
+% Alternatively, one might imagine Haskell being desugared into an intermediate
+% representation to which multiple different optimizing transformations are
+% applied but on which no integrity checks are done
+
+Despite \emph{want} a linearly-typed core
+because a linearly-typed Core ensures that desugaring
+Haskell and optimizing transformations don't destroy linearity.
+
+% In theory, because the Core language must also know about linearity, we should
+
+% A remedy is to use the multiplicity annotations of λq → as cardinality declarations. Formalising
+% and implementing the integration of multiplicities in the cardinality analysis is left as future work.
+
+% In theory, we should typecheck \emph{linearity} in \textbf{Core} just the same
+% as the typechecking verification we had with the existing type annotations prior
+% to the addition of linear types, that is, our Core program with linearity
+% annotations should be typechecked after the optimizing transformations...
 

### 2023-01-11 report.tex
Start linear lambda calculus grammar

@@
 \usepackage{fancyvrb}
+
+% BNF grammars
+\usepackage[center]{formal-grammar}
+
 \DefineVerbatimEnvironment{code}{Verbatim}{fontsize=\small}
@@
 linear variable that must be used exactly once, would not typecheck in a source
-linear Haskell program since the typechecker cannot tell that $x$ is used
-linearly, but this program might occur naturally in Core from
-transformations on a program that did succesfully typecheck:
+Haskell program since the typechecker cannot tell that $x$ is used linearly, but
+this program might occur naturally in Core from transformations on a program
+that did succesfully typecheck:
 \begin{code}
@@
 
-Is this program really still linear? If one looks at it
+Is this program really still linear? Yes, but ...
+
+% If one looks at it
 
@@
 
-Despite \emph{want} a linearly-typed core
-because a linearly-typed Core ensures that desugaring
-Haskell and optimizing transformations don't destroy linearity.
+% Despite \emph{want} a linearly-typed core
+% because a linearly-typed Core ensures that desugaring
+% Haskell and optimizing transformations don't destroy linearity.
 
@@
 
-% Linear type systems add expressiven at the type level regarding values that
-% must be used exactly once thorought their lifetime.
-
 % about the usage of resources in a
@@
 
+% TODO? Another defining feature of Haskell is its powerful type system. The
+% interaction between the new linearity polymorphism and the existing
+% type-level facilities is quite interesting
+
 \chapter{Background}
@@
 
-A linear type system is a type system in which ...!
+A linear type system is a type system in which linear resources must be used
+exactly once...!
 
@@
 % We'd like to hide the boilerplate involved \cite{Linear constraints}
-
-In order to give well defined semantics to a linear type system, we present a
-linearly typed lambda calculus which is a very simple language with linear
-types.
 
@@
 % TODO: Revise the story above instead of dumping so much information
+
+In order to give well defined semantics to a linear type system, we present a
+linearly typed lambda calculus which is a very simple language with linear
+types.
+
+\begin{grammar}
+    \firstcase{\t}{\star}{Unit type}
+    \otherform{t \tensor t'}{Multiplicative conjunction type}
+    % t ::= a $\lolli$ b
+\end{grammar}
 

### 2023-01-12 report.tex
minor changes

@@
 
-% BNF grammars
-\usepackage[center]{formal-grammar}
-
 \DefineVerbatimEnvironment{code}{Verbatim}{fontsize=\small}
@@
 
+% Statically typed languages 
 % Linear types are cool
@@
 %   * Optimizations changes source -- e.g. if recursive lets are introduced, how to type linearity?
-%
-
-\section{Introduction}
-
-Linear type systems\cite{} add expressiveness to existing type systems by
+% Goals - O que eu vou fazer
+% Document Structure (pode-se eventualmente combinar com a seccao dos goals)
+
+%\section{Introduction}
+
+Linear type systems~\cite{} add expressiveness to existing type systems by
 stating that linear resources must be used \emph{exactly once}. In programming
@@
 that file handles, socket descriptors, and allocated memory are freed exactly
-once (leaks and double-frees become type errors)\cite{}, implement safe
-high-performance language interoperability\cite{}, guarantee that quantum
-entangled states are not duplicated~\cite{}, handle mutable state safely\cite{},
-something on session types guarantees\cite{}, and much more.
+once (leaks and double-frees become type errors)~\cite{}, implement safe
+high-performance language interoperability~\cite{}, guarantee that quantum
+entangled states are not duplicated~\cite{}, handle mutable state safely~\cite{},
+something on session types guarantees~\cite{}, and much more.
 
@@
 
-Core was imbued with linear types because an implementation detail required it,
-and, more importantly, to guarantee linear programs remain linear after being
+Core was imbued with linear types to guarantee linear programs remain linear after being
 desugared and optimized: much the same way a \emph{typed} Core ensures that
@@
 
-\begin{grammar}
-    \firstcase{\t}{\star}{Unit type}
-    \otherform{t \tensor t'}{Multiplicative conjunction type}
-    % t ::= a $\lolli$ b
-\end{grammar}
+\[
+  \begin{array}{lcl}
+    s,t & ::= & \star\\
+        & \mid & t \lolli s\\
+       \dots
+   \end{array}
+\]
 

### 2023-01-12 report.tex
Introduction on linear types

@@
 
+Static type systems have brought programmers ...
+Programming languages rely on type systems to statically eliminate classes of
+errors ...
+
 Linear type systems~\cite{} add expressiveness to existing type systems by
@@
 
-As an example, consider the following linearly typed program which allocates
-some memory and frees it twice. The variable $p$ is a linear resource created by
-the \textbf{malloc} keyword.
-\begin{code}
-let p = malloc 4
- in free p
-    free p
-\end{code}
-
-Because $p$ is a linear variable, it must be used exactly once. However, we are
-using it in two different calls to \textbf{free}. The example program
-\emph{doesn't typecheck}! In section~\ref{linear-types} we describe the
-fundamentals of linear types in more detail with a formal definition and many
-examples.
+As an example, consider the following C-like program in which allocated memory
+is freed twice. We know this to be the dreaded double-free error which will
+crash our programs and crush our dreams. Regardless, a C-like type system will
+accept this program without any issue.
+\begin{code}
+let p = malloc 4;
+ in free p;
+    free p;
+\end{code}
+
+Under the lens of a linear type system, consider the variable $p$ to be a linear
+resource created by the \textbf{malloc} keyword. Because $p$ is a linear
+variable, it must be used exactly once. However, we are using it in two
+different calls to \textbf{free}. With our new linearity assumptions, the
+% TODO:
+example program \emph{would not typecheck}! Some conclusion on linearity
+goodness. In section~\ref{linear-types} we describe the fundamentals of linear
+types in more detail with a formal definition and many examples.
 
@@
 
-Linearity in Haskell stands out from linearity in Rust and Idris2 for, at least,
-the two following distinct reasons:
+Linearity in Haskell stands out % from linearity in Rust and Idris2
+for the two following distinct reasons:
 
@@
     \item Linear types were only added to the language roughly \emph{31 years
-        after} Haskell's inception. It is an especially difficult endeavour to
-        add linear types to a well-established language with a large library
-        ecosystem and many active users, rather than to develop the language
-        from the ground up with linear types in mind, and the successful
-        approach as implemented in GHC 9.0, the leading Haskell compiler, was
-        based on Linear Haskell\cite{linearhaskell}, where a linear type system
-        designed with retaining backwards-compatibility and allowing code reuse
-        across linear and non-linear users of the same libraries in mind was
-        described. We describe Linear Haskell in detail in
+        after} Haskell's inception, unlike Rust and Idris2 which were
+        designed with linearity from the start. It is an especially difficult
+        endeavour to add linear types to a well-established language with a
+        large library ecosystem and many active users, rather than to develop
+        the language from the ground up with linear types in mind, and the
+        successful approach as implemented in GHC 9.0, the leading Haskell
+        compiler, was based on Linear Haskell\cite{linearhaskell}, where a
+        linear type system designed with retaining backwards-compatibility and
+        allowing code reuse across linear and non-linear users of the same
+        libraries in mind was described. We describe Linear Haskell in detail in
         section~\ref{linear-haskell-section}.
-    \item Linear types permeate Haskell down to (its) \textbf{Core}, the intermediate
-        language Haskell is desugared to. \textbf{Core} is a very small
-        explicitly typed language to which multiple Core-to-Core optimizing
-        transformations are applied in succession. The final Core program can be
-        typechecked very fast (because of the explicit type annotations) and
-        doing so serves as a sanity check to the correction of the
-        source transformations. If the resulting optimized Core program doesn't
-        typecheck, something \emph{very wrong} has happened in a transformation!
-        We discuss Core in detail in section~\ref{core-section}.
-    % TODO: \item values in rust are linear by default while non-linear is the
-        % haskell default?
+
+    \item Linear types permeate Haskell down to (its) \textbf{Core}, the
+        intermediate language Haskell is desugared to. \textbf{Core} is a very
+        small explicitly typed functional language to which multiple
+        Core-to-Core optimizing transformations are applied in succession. The
+        final Core program can be typechecked very fast (because of the explicit
+        type annotations) and doing so serves as a sanity check to the
+        correction of the source transformations. If the resulting optimized
+        Core program doesn't typecheck, something \emph{very wrong} has happened
+        in a transformation!  We discuss Core in detail in
+        section~\ref{core-section}.
+        % TODO: \item values in rust are linear by default while non-linear is
+        % the haskell default?
 \end{itemize}
@@
 
-However, currently, typechecking linearity in Core is broken in the face of
-optimizations! The reason is not evidently clear: if we can typecheck linearity
-in the surface level Haskell why can't we in Core? In fact, if we were to
-typecheck the programs in Core as they were at the surface level, we would
-succeed, but programs are \emph{transformed} (by semantic-preserving optimizing
-transformations) and hence end up different from how they were originally
-written. For example, the following recursive let definition, where $x$ is a
-linear variable that must be used exactly once, would not typecheck in a source
-Haskell program since the typechecker cannot tell that $x$ is used linearly, but
-this program might occur naturally in Core from transformations on a program
-that did succesfully typecheck:
+However, currently, programs transformed by optimizations can fail to typecheck
+in Core because of the newly added linearity! The reason is not evidently clear:
+if we can typecheck linearity in the surface level Haskell why can't we in Core?
+In fact, if we were to typecheck the programs in Core as they were at the
+surface level, we would succeed, but programs are \emph{transformed} (by
+semantic-preserving optimizing transformations) and hence end up different from
+how they were originally written. For example, the following recursive let
+definition, where $x$ is a linear variable that must be used exactly once, would
+not typecheck in a source Haskell program since the typechecker cannot tell that
+$x$ is used linearly, but this program might occur naturally in Core from
+transformations on a program that did succesfully typecheck:
 \begin{code}
@@
 
-A linear type system is a type system in which linear resources must be used
-exactly once...!
-
-% A linear type system is a type system in which statements such as "this function
-% will use its argument exactly once" can be formally expressed and checked.
-
-Girard's \emph{linear logic} suggests a new \emph{type system} for functional
-languages through the Curry-Howard isomorphism -- every linear logic proposition
-maps to a different linear type.
-
-Values belonging to a linear type must be used exactly once. For example, the
-type corresponding to the linear implication ($\lolli$), the linear function, is
-a function that consumes its argument exactly once. A linear function that
-receives an integer as an input and consumes it exactly once to
-produce a character would have type $\textbf{int} \lolli
-\textbf{char}$.
-
-Programming languages with linear type systems might differ in their definition of linear
-resources, but usually \emph{variables} are considered linear resources?
-% That is, we can rely on the type system to prove properties regarding the usage of certain resources.
-% The \emph{Hello World} of linearly typed programs is a linearly typed pointer.
-% In a non-memory managed language, an allocated pointer must be freed
-% \emph{exactly once}.
-
-% \begin{code}
-% let p = malloc(5)
-%     p' = store(p, "x")
-%     (p'', v) = load(p)
-%     free p''
-% \end{code}
-
-% We'd like to hide the boilerplate involved \cite{Linear constraints}
-
-\mypara{Type Systems} A type system can be formally described through
-a set of inference rules that inductively define a judgment of the
-form $\Gamma \vdash M : A$, stating that program expression $M$ has
-type $A$ according to the \emph{typing assumptions} for variables
-tracked in $\Gamma$. For instance,
-$x{:}\mathsf{Int}, y{:}\mathsf{Int} \vdash x+y : \mathsf{Int}$ states
-that $x+y$ has type $\mathsf{Int}$ under the assumption that $x$ and
-$y$ have type $\mathsf{Int}$.  An expression $M$ is deemed well-typed
-with a given type $A$ if one can construct a typing derivation with $M :
-A$ as its conclusion, by repeated application of the inference rules.
-
-The simply-typed $\lambda$-calculus is a typed core functional
-language~\cite{10.5555/509043} that captures the essence of a type system in a simple and familiar environment. Its syntax consists of
-functional abstraction, written $\lambda x{:}A.M$, denoting
-an (anonymous) function that takes an argument of type $A$, bound to
-$x$ in $M$; and application $M\,N$, with the standard meaning, and
-variables $x$. For instance, the term $\lambda x{:}A.x$, denoting the identity function, is a functional abstraction
-taking an argument of type $A$ and returning it back.
-
-\mypara{Linear Logic}
-
-Linear logic \cite{DBLP:journals/tcs/Girard87} can be seen as a
-resource-aware logic, where propositions are interpreted as resources
-that are consumed during the inference process.  Where in standard
-propositional logic we are able to use an assumption as many times as
-we want, in linear logic every resource (i.e., every assumption) must
-be used \emph{exactly once}, or \emph{linearly}. This usage
-restriction gives rise to new logical connectives, based on the way
-the ambient resources are used. For instance, conjunction, usually
-written as $A\wedge B$, appears in two forms in linear logic:
-multiplicative or simultaneous conjunction (written $A\tensor B$); and
-additive or alternative conjunction (written $A\with
-B$). Multiplicative conjunction denotes the simultaneous availability
-of resources $A$ and $B$, requiring both of them to be
-used. Alternative conjunction denotes the availability of $A$ and $B$,
-but where only one of the two resources may be used. Similarly,
-implication becomes linear implication, written $A\lolli B$,
-denoting a resource that will consume (exactly one) resource $A$ to
-produce a resource $B$.
-
-To present the formalization of this logic, besides the new
-connectives, we need to introduce the \emph{resource-aware context}
-$\Delta$.  In contrast to the previously seen $\Gamma$, $\Delta$ is
-also a list of variables and their types, but where each and every
-variable must be used exactly once during inference.  So, to introduce
-the connective $\tensor$ which defines a multiplicative pair of
-propositions, we must use exactly all the resources
-($\Delta_1, \Delta_2$) needed to realize the ($\Delta_1$)
-proposition $A$, and ($\Delta_2$) proposition $B$:
+Much the same way type systems can statically eliminate various kinds of
+programs that would fail at runtime, such as a program that derreferences an
+integer value rather than a pointer, linear type systems can guarantee that
+certain errors (regarding resource usage) are forbidden.
+
+In linear type systems, so called linear resources must be used \emph{exactly
+once}. Not using a linear resource at all or using said resource multiple times
+will result in a type error. We can model many real-world resources such as file
+handles, socket descriptors, and allocated memory, as linear resources. This
+way, because a file handle must be used exactly once, forgetting to close the
+file handle is a type error, and closing the handle twice is also a type error.
+With linear types, avoiding leaks and double frees is no longer a programmer's
+worry because the compiler can guarantee the resource is used exactly once, or
+\emph{linearly}.
+
+To understand how linear types are defined and used in practice, we present two
+examples of anonymous functions, try to understand how they could be disregarded
+as incorrect, and work our way up to linear types from them. The first receives a
+file handle as an argument, ignores it, and returns $\star$ (read unit), which
+is equivalent to C's
+\textbf{void}.
+% TODO: Better code formatting
 \[
-    \infer*[right=($\tensor I$)]
-    {\Delta_1 \vdash M : A \and \Delta_2 \vdash N : B}
-    {\Delta_1 , \Delta_2 \vdash (M \tensor N) : A \tensor B}
+    \lambda h.~\textrm{return}~\star;
 \]
-Out of the logical connectives, we need to mention one more, since it
-augments the form of the judgment and it's the one that ensures
-logical strength i.e. we're able to translate intuitionistic logic
-into linear logic.  The proposition $\bang A$ (read \emph{of course}
-$A$) is used (under certain conditions) to make a resource
-``infinite'' i.e. to make it useable an arbitrary number of times. To
-distinguish the ``infinite'' variables, a separate, unrestricted,
-context is used -- $\Gamma$. So $\Gamma$ holds the ``infinite''
-resources, and $\Delta$ the resources that can only be used once.  The
-linear typing judgment for the introduction of the exponential $\bang A$
-takes the form:
+
+And, similarly, the second receives a file handle and closes it
+twice. Note that \textbf{close} has type $\textbf{Handle} \to \star$, that is, it's a
+function that receives a handle and returns unit.
 \[
-    \infer*[right=($\bang I$)]
-    {\Gamma ; \emptyset \vdash M : A}
-    {\Gamma ; \emptyset \vdash \bang M : \bang A}
+    \lambda h.~\textrm{close}~h;~\textrm{close}~h;
 \]
-Logically, a proof of $\bang A$ cannot use linear resources since
-$\bang A$ denotes an unbounded (potentially $0$) number of copies of
-$A$. Proofs of $\bang A$ may use other unrestricted or exponential
-resources, tracked by context $\Gamma$.
-From a computational perspective, the type $\bang A$
-internalizes the simply-typed $\lambda$-calculus in the linear
-$\lambda$-calculus.
-
-The elimination form for the exponential, written $\llet{ !u
-= M}{ N}$, warrants the use of resource $A$ an unbounded
-number of times in $N$ via the variable $u$:
-\[
-    \infer*[right=($\bang E$)]
-    {\Gamma ; \Delta_1 \vdash M : \bang A \and \Gamma, u{:}A ; \Delta_2 \vdash N : C}
-    {\Gamma ; \Delta_1, \Delta_2 \vdash \llet{ !u = M}{ N} : C}
-\]
-
-% TODO: Revise the story above instead of dumping so much information
+
+Both the above functions share the same type ($\textbf{Handle} \to \star$),
+under a simple type system such as C's they are type correct, and both have
+erroneous behaviours. The first forgets to close the handle and the second
+closes it twice. Our goal is now to reach a type system that rejects these
+two programs.
+
+The key observation is that the argument of the function must be linear, and we
+can augment the common function type to account for the linearity of its
+argument. We take the function type $A \to B$ and replace the arrow with the
+$\lolli$
+%
+\footnote{Since linear types are born from a
+correspondence\cite{curry-howard} (the Curry-Howard isomrphism) with linear
+logic\cite{linear-logic}, we borrow the $\lolli$ symbol and other linear logic
+connectives to describe linear types.}
+%
+operator to denote a function that uses its argument exactly once: $A \lolli B$.
+% Linear types are a powerful idea because they allow us to statically reason
+% about resources in our program. A so called linear resource
+
+% For example, to define a ...
+
+% Linear types are a powerful idea
 
@@
 linearly typed lambda calculus which is a very simple language with linear
-types.
+types...
 
@@
         & \mid & t \lolli s\\
-       \dots
+        & \mid & t \oplus s\\
+        & \mid & t \tensor s\\
+        & \mid & t \with s\\
+        & \mid & \bang t\\
    \end{array}
@@
 
+% A linear type system is a type system in which linear resources must be used
+% exactly once...!
+
+% % A linear type system is a type system in which statements such as "this function
+% % will use its argument exactly once" can be formally expressed and checked.
+
+% Girard's \emph{linear logic} suggests a new \emph{type system} for functional
+% languages through the Curry-Howard isomorphism -- every linear logic proposition
+% maps to a different linear type.
+
+% Values belonging to a linear type must be used exactly once. For example, the
+% type corresponding to the linear implication ($\lolli$), the linear function, is
+% a function that consumes its argument exactly once. A linear function that
+% receives an integer as an input and consumes it exactly once to
+% produce a character would have type $\textbf{int} \lolli
+% \textbf{char}$.
+
+% Programming languages with linear type systems might differ in their definition of linear
+% resources, but usually \emph{variables} are considered linear resources?
+% % That is, we can rely on the type system to prove properties regarding the usage of certain resources.
+% % The \emph{Hello World} of linearly typed programs is a linearly typed pointer.
+% % In a non-memory managed language, an allocated pointer must be freed
+% % \emph{exactly once}.
+
+% % \begin{code}
+% % let p = malloc(5)
+% %     p' = store(p, "x")
+% %     (p'', v) = load(p)
+% %     free p''
+% % \end{code}
+
+% % We'd like to hide the boilerplate involved \cite{Linear constraints}
+
+% \mypara{Type Systems} A type system can be formally described through
+% a set of inference rules that inductively define a judgment of the
+% form $\Gamma \vdash M : A$, stating that program expression $M$ has
+% type $A$ according to the \emph{typing assumptions} for variables
+% tracked in $\Gamma$. For instance,
+% $x{:}\mathsf{Int}, y{:}\mathsf{Int} \vdash x+y : \mathsf{Int}$ states
+% that $x+y$ has type $\mathsf{Int}$ under the assumption that $x$ and
+% $y$ have type $\mathsf{Int}$.  An expression $M$ is deemed well-typed
+% with a given type $A$ if one can construct a typing derivation with $M :
+% A$ as its conclusion, by repeated application of the inference rules.
+
+% The simply-typed $\lambda$-calculus is a typed core functional
+% language~\cite{10.5555/509043} that captures the essence of a type system in a simple and familiar environment. Its syntax consists of
+% functional abstraction, written $\lambda x{:}A.M$, denoting
+% an (anonymous) function that takes an argument of type $A$, bound to
+% $x$ in $M$; and application $M\,N$, with the standard meaning, and
+% variables $x$. For instance, the term $\lambda x{:}A.x$, denoting the identity function, is a functional abstraction
+% taking an argument of type $A$ and returning it back.
+
+% \mypara{Linear Logic}
+
+% Linear logic \cite{DBLP:journals/tcs/Girard87} can be seen as a
+% resource-aware logic, where propositions are interpreted as resources
+% that are consumed during the inference process.  Where in standard
+% propositional logic we are able to use an assumption as many times as
+% we want, in linear logic every resource (i.e., every assumption) must
+% be used \emph{exactly once}, or \emph{linearly}. This usage
+% restriction gives rise to new logical connectives, based on the way
+% the ambient resources are used. For instance, conjunction, usually
+% written as $A\wedge B$, appears in two forms in linear logic:
+% multiplicative or simultaneous conjunction (written $A\tensor B$); and
+% additive or alternative conjunction (written $A\with
+% B$). Multiplicative conjunction denotes the simultaneous availability
+% of resources $A$ and $B$, requiring both of them to be
+% used. Alternative conjunction denotes the availability of $A$ and $B$,
+% but where only one of the two resources may be used. Similarly,
+% implication becomes linear implication, written $A\lolli B$,
+% denoting a resource that will consume (exactly one) resource $A$ to
+% produce a resource $B$.
+
+% To present the formalization of this logic, besides the new
+% connectives, we need to introduce the \emph{resource-aware context}
+% $\Delta$.  In contrast to the previously seen $\Gamma$, $\Delta$ is
+% also a list of variables and their types, but where each and every
+% variable must be used exactly once during inference.  So, to introduce
+% the connective $\tensor$ which defines a multiplicative pair of
+% propositions, we must use exactly all the resources
+% ($\Delta_1, \Delta_2$) needed to realize the ($\Delta_1$)
+% proposition $A$, and ($\Delta_2$) proposition $B$:
+% \[
+%     \infer*[right=($\tensor I$)]
+%     {\Delta_1 \vdash M : A \and \Delta_2 \vdash N : B}
+%     {\Delta_1 , \Delta_2 \vdash (M \tensor N) : A \tensor B}
+% \]
+% Out of the logical connectives, we need to mention one more, since it
+% augments the form of the judgment and it's the one that ensures
+% logical strength i.e. we're able to translate intuitionistic logic
+% into linear logic.  The proposition $\bang A$ (read \emph{of course}
+% $A$) is used (under certain conditions) to make a resource
+% ``infinite'' i.e. to make it useable an arbitrary number of times. To
+% distinguish the ``infinite'' variables, a separate, unrestricted,
+% context is used -- $\Gamma$. So $\Gamma$ holds the ``infinite''
+% resources, and $\Delta$ the resources that can only be used once.  The
+% linear typing judgment for the introduction of the exponential $\bang A$
+% takes the form:
+% \[
+%     \infer*[right=($\bang I$)]
+%     {\Gamma ; \emptyset \vdash M : A}
+%     {\Gamma ; \emptyset \vdash \bang M : \bang A}
+% \]
+% Logically, a proof of $\bang A$ cannot use linear resources since
+% $\bang A$ denotes an unbounded (potentially $0$) number of copies of
+% $A$. Proofs of $\bang A$ may use other unrestricted or exponential
+% resources, tracked by context $\Gamma$.
+% From a computational perspective, the type $\bang A$
+% internalizes the simply-typed $\lambda$-calculus in the linear
+% $\lambda$-calculus.
+
+% The elimination form for the exponential, written $\llet{ !u
+% = M}{ N}$, warrants the use of resource $A$ an unbounded
+% number of times in $N$ via the variable $u$:
+% \[
+%     \infer*[right=($\bang E$)]
+%     {\Gamma ; \Delta_1 \vdash M : \bang A \and \Gamma, u{:}A ; \Delta_2 \vdash N : C}
+%     {\Gamma ; \Delta_1, \Delta_2 \vdash \llet{ !u = M}{ N} : C}
+% \]
+
+% % TODO: Revise the story above instead of dumping so much information
+
 % \mypara{Linear Types}
-\subsection{Linear Types}
-
-Girard's \emph{linear logic} suggests a new \emph{type system} for functional
-languages through the Curry-Howard isomorphism -- every linear logic proposition
-maps to a different linear type.
-
-Values belonging to a linear type must be used exactly once. For example, the
-type corresponding to the linear implication ($\lolli$), the linear function, is
-a function that consumes its argument exactly once. A linear function that
-receives an integer as an input and consumes it exactly once to
-produce a character would have type $\textbf{int} \lolli
-\textbf{char}$.
+% \subsection{Linear Types}
+
+% Girard's \emph{linear logic} suggests a new \emph{type system} for functional
+% languages through the Curry-Howard isomorphism -- every linear logic proposition
+% maps to a different linear type.
+
+% Values belonging to a linear type must be used exactly once. For example, the
+% type corresponding to the linear implication ($\lolli$), the linear function, is
+% a function that consumes its argument exactly once. A linear function that
+% receives an integer as an input and consumes it exactly once to
+% produce a character would have type $\textbf{int} \lolli
+% \textbf{char}$.
 
@@
 
-\section{Work in progress}
-
-\subsection{In Result of TcGblEnv axioms}
-
-I don't understand this one yet. I would guess it has to do with multiplicity
-coercions. Incorrect incompatible branch: CoAxBranch (src/Prelude/Linear/GenericUtil.hs:112:3-51):
-\begin{code}
-type family Fixup (f :: Type -> Type) (g :: Type -> Type) :: Type -> Type where
-  Fixup (D1 c f) (D1 _c g) = D1 c (Fixup f g)
-  Fixup (C1 c f) (C1 _c g) = C1 c (Fixup f g)
-  Fixup (S1 c f) (S1 _c (MP1 m f)) = S1 c (MP1 m f) -- error in this constructor
-  Fixup (S1 c f) (S1 _c f) = S1 c f
-  Fixup (f :*: g) (f' :*: g') = Fixup f f' :*: Fixup g g'
-  Fixup (f :+: g) (f' :+: g') = Fixup f f' :+: Fixup g g'
-  Fixup V1 V1 = V1
-  Fixup _ _ = TypeError ('Text "FixupMetaData: representations do not match.")
-\end{code}
-
-% \section{Core to Core Validation}
-\subsection{Inlining}
-
-\subsection{CSE}
-
-\subsection{Join Points}
-
-When duplicating a case (in the case-of-case transformation), to avoid code
-explosion, the branches of the case are first made into join points
-
-\begin{code}
-case e of
-  Pat1 -> u
-  Pat2 -> v
-~~>
-let j1 = u in
-let j2 = v in
-case e of
-  Pat1 -> j1
-  Pat2 -> j2
-\end{code}
-
-If there is any linear variable in u and v, then the standard
-let rule above will fail (since j1 occurs only in one branch, and
-so does j2).
-
-However, if j1 and j2 were annotated with their usage environment,
-
-\subsection{Empty Case}
-
-For case expressions, the usage environment is computed by checking all branches
-and taking sup. However, this trick doesn't work when there are no branches.
-
-\begin{itemize}
-\item https://gitlab.haskell.org/ghc/ghc/-/issues/20058
-\item https://gitlab.haskell.org/ghc/ghc/-/issues/18768
-
-\item (1) Just like a case expression remembers its type (Note [Why does Case have a
-'Type' field?] in Core.hs), it should remember its usage environment. This data
-should be verified by Lint.
-
-\item (2) Once this is done, we can remove the Bottom usage and the second field of
-UsageEnv. In this step, we have to infer the correct usage environment for empty
-case in the typechecker.
-\end{itemize}
-
-\begin{code}
-{-# LANGUAGE LinearTypes, EmptyCase #-}
-module M where
-
-{-# NOINLINE f #-}
-f :: a %1-> ()
-f x = case () of {}
-\end{code}
-
-This example is well typed: a function is linear if it consumes its argument
-exactly once when it's consumed exactly once. It seems like the function isn't
-linear since it won't consume x because of the empty case, however, that also
-means f won't be consumed due to the same empty case, thus linearity is
-preserved.
-
-\begin{code}
-* In the case of empty types (see Note [Bottoming expressions]), say
-  data T
-we do NOT want to replace
-  case (x::T) of Bool {}   -->   error Bool "Inaccessible case"
-because x might raise an exception, and *that*'s what we want to see!
-(#6067 is an example.) To preserve semantics we'd have to say
-   x `seq` error Bool "Inaccessible case"
-but the 'seq' is just such a case, so we are back to square 1.
-\end{code}
-
-There are three different problems:
-
-\begin{itemize}
-\item castBottomExpr converts (case x :: T of {}) :: T to x.
-\item Worker/wrapper moves the empty case to a separate binding
-\item CorePrep eliminates empty case, just like point 1 (See -- Eliminate empty
-    case in GHC.CoreToStg.Prep
-\end{itemize}
-
-With castBottomExpr, we get the example above to
-\begin{code}
-    f = \ @a (x (%1) :: a) -> ()
-\end{code}
-And if we don't, 
-\begin{code}
-    f = \ @a (x (%1) :: a) -> case () of {}
-\end{code}
-And that supposedly if we had a usage environment in the case expression we
-could avoid the error. How is it typed without the transformation in face
-of the bottom? (Even knowing that theoretically it's because of divergence?)
+% \section{Work in progress}
+
+% % \section{Core to Core Validation}
+% \subsection{Inlining}
+
+% \subsection{CSE}
+
+% \subsection{Join Points}
+
+% When duplicating a case (in the case-of-case transformation), to avoid code
+% explosion, the branches of the case are first made into join points
+
+% \begin{code}
+% case e of
+%   Pat1 -> u
+%   Pat2 -> v
+% ~~>
+% let j1 = u in
+% let j2 = v in
+% case e of
+%   Pat1 -> j1
+%   Pat2 -> j2
+% \end{code}
+
+% If there is any linear variable in u and v, then the standard
+% let rule above will fail (since j1 occurs only in one branch, and
+% so does j2).
+
+% However, if j1 and j2 were annotated with their usage environment,
+
+% \subsection{Empty Case}
+
+% For case expressions, the usage environment is computed by checking all branches
+% and taking sup. However, this trick doesn't work when there are no branches.
+
+% \begin{itemize}
+% \item https://gitlab.haskell.org/ghc/ghc/-/issues/20058
+% \item https://gitlab.haskell.org/ghc/ghc/-/issues/18768
+
+% \item (1) Just like a case expression remembers its type (Note [Why does Case have a
+% 'Type' field?] in Core.hs), it should remember its usage environment. This data
+% should be verified by Lint.
+
+% \item (2) Once this is done, we can remove the Bottom usage and the second field of
+% UsageEnv. In this step, we have to infer the correct usage environment for empty
+% case in the typechecker.
+% \end{itemize}
+
+% \begin{code}
+% {-# LANGUAGE LinearTypes, EmptyCase #-}
+% module M where
+
+% {-# NOINLINE f #-}
+% f :: a %1-> ()
+% f x = case () of {}
+% \end{code}
+
+% This example is well typed: a function is linear if it consumes its argument
+% exactly once when it's consumed exactly once. It seems like the function isn't
+% linear since it won't consume x because of the empty case, however, that also
+% means f won't be consumed due to the same empty case, thus linearity is
+% preserved.
+
+% \begin{code}
+% * In the case of empty types (see Note [Bottoming expressions]), say
+%   data T
+% we do NOT want to replace
+%   case (x::T) of Bool {}   -->   error Bool "Inaccessible case"
+% because x might raise an exception, and *that*'s what we want to see!
+% (#6067 is an example.) To preserve semantics we'd have to say
+%    x `seq` error Bool "Inaccessible case"
+% but the 'seq' is just such a case, so we are back to square 1.
+% \end{code}
+
+% There are three different problems:
+
+% \begin{itemize}
+% \item castBottomExpr converts (case x :: T of {}) :: T to x.
+% \item Worker/wrapper moves the empty case to a separate binding
+% \item CorePrep eliminates empty case, just like point 1 (See -- Eliminate empty
+%     case in GHC.CoreToStg.Prep
+% \end{itemize}
+
+% With castBottomExpr, we get the example above to
+% \begin{code}
+%     f = \ @a (x (%1) :: a) -> ()
+% \end{code}
+% And if we don't, 
+% \begin{code}
+%     f = \ @a (x (%1) :: a) -> case () of {}
+% \end{code}
+% And that supposedly if we had a usage environment in the case expression we
+% could avoid the error. How is it typed without the transformation in face
+% of the bottom? (Even knowing that theoretically it's because of divergence?)
 

### 2023-01-16 report.tex
Typeset linearly typed lambda calculus

@@
 
-\newcommand{\llet}[2]{\mathsf{let}\ #1\ \mathsf{in}\ #2}
+\newcommand{\llet}[2]{\mathsf{let}~#1~\mathsf{in}~#2}
+\newcommand{\ccase}[2]{\mathsf{case}~#1~\mathsf{of}~#2}
 
@@
 To understand how linear types are defined and used in practice, we present two
-examples of anonymous functions, try to understand how they could be disregarded
-as incorrect, and work our way up to linear types from them. The first receives a
-file handle as an argument, ignores it, and returns $\star$ (read unit), which
-is equivalent to C's
-\textbf{void}.
+examples of anonymous functions that receive as an argument a handle to work
+with and are responsible for closing it, we try to understand how the examples
+could be disregarded as incorrect, and work our way up to linear types from
+them. The first function ignores the received file handle and returns $\star$
+(read unit), which is equivalent to C's \textbf{void}.
 % TODO: Better code formatting
-\[
+\begin{mathpar}
     \lambda h.~\textrm{return}~\star;
-\]
-
-And, similarly, the second receives a file handle and closes it
-twice. Note that \textbf{close} has type $\textbf{Handle} \to \star$, that is, it's a
-function that receives a handle and returns unit.
-\[
+    \and
     \lambda h.~\textrm{close}~h;~\textrm{close}~h;
-\]
-
-Both the above functions share the same type ($\textbf{Handle} \to \star$),
-under a simple type system such as C's they are type correct, and both have
-erroneous behaviours. The first forgets to close the handle and the second
-closes it twice. Our goal is now to reach a type system that rejects these
-two programs.
-
-The key observation is that the argument of the function must be linear, and we
-can augment the common function type to account for the linearity of its
-argument. We take the function type $A \to B$ and replace the arrow with the
-$\lolli$
+\end{mathpar}
+% TODO: Equation references?
+
+Ignoring the file handle which should have been closed by the function makes the
+first function incorrect. Similarly, the second function receives the file handle and
+closes it twice, which is incorrect not because it falls short of the
+specification, but rather because the program will crash while executing it.
 %
-\footnote{Since linear types are born from a
-correspondence\cite{curry-howard} (the Curry-Howard isomrphism) with linear
-logic\cite{linear-logic}, we borrow the $\lolli$ symbol and other linear logic
-connectives to describe linear types.}
+Additionally, both functions share the same type, $\textbf{Handle} \to \star$, i.e.
+a function that takes a \textbf{Handle} and returns $\star$. The second function also
+shares this type because \textbf{close} has type $\textbf{Handle} \to \star$.
+%
+Under a simple type system such as C's both functions are type correct (the
+compiler will happily succeed), but both have erroneous behaviours. The first
+forgets to close the handle and the second closes it twice. Our goal is now to
+reach a type system that rejects these two programs.
+
+The key observation to invalidating these programs is to focus on the function
+type going between \textbf{Handle} and $\star$ and augment it to indicate that
+\emph{the argument must be used exactly once}, or, in other words, that the
+argument of the function must be linear. We take the function type $A \to B$
+and replace the function arrow ($\to$) with the linear function arrow ($\lolli$)
+%
+\footnote{Since linear types are born from a correspondence with linear
+logic\cite{linear-logic} (the Curry-Howard isomorphism\cite{curry-howard}), we
+borrow the $\lolli$ symbol and other linear logic connectives to describe linear
+types.}
 %
 operator to denote a function that uses its argument exactly once: $A \lolli B$.
+%
+Providing the more restrictive linear function signature $\textbf{Handle} \lolli
+\star$ to the example programs would make both of them fail to typecheck because
+they do not satisfy the linearity specification that the function argument
+should only be used exactly once.
+
 % Linear types are a powerful idea because they allow us to statically reason
-% about resources in our program. A so called linear resource
-
-% For example, to define a ...
-
-% Linear types are a powerful idea
-
-In order to give well defined semantics to a linear type system, we present a
-linearly typed lambda calculus which is a very simple language with linear
-types...
-
-\[
+% about resources in our program. A so called linear resource ?
+
+In order to further give well defined semantics to a linear type system, we
+present a linearly typed lambda calculus, a very simple language with linear
+types, by defining what are syntatically valid programs through the grammar and
+what programs are well typed through the typing rules.
+
+An expression in the linearly typed lambda calculus can be $\star$...
+A type in the linearly typed lambda calculus can be $\star$...
+
+\begin{mathpar}
   \begin{array}{lcl}
@@
    \end{array}
-\]
+
+\and
+
+  \begin{array}{lcl}
+      M,N & ::= & \star \mid \llet{\star = M}{N} \\
+        & \mid & u \\
+        & \mid & \lambda u. M \mid M~N\\
+        & \mid & \textrm{inl}~M \mid \textrm{inr}~M \mid
+        \ccase{M}{\textrm{inl}~u_1 \to N_1 \mid \textrm{inr}~u_2 \to N_2}\\
+        & \mid & M \tensor N \mid \llet{u_1 \tensor u_2 = M}{N}\\
+        & \mid & M \with N \mid \textrm{fst}~M \mid \textrm{snd}~M \\
+        & \mid & \bang M \mid \llet{!u = M}{N} \\
+   \end{array}
+\end{mathpar}
+
+\begin{mathpar}
+    \infer*[right=($\tensor I$)]
+    {\Delta \vdash M : A \and \Delta' \vdash N : B}
+    {\Delta , \Delta' \vdash (M \tensor N) : A \tensor B}
+
+    \and
+\end{mathpar}
+
+\begin{mathpar}
+    \infer*[right=($\tensor E$)]
+    {\Delta \vdash M : A \tensor B \and \Delta', u{:}A, v{:}B \vdash N : C}
+    {\Delta , \Delta' \vdash \llet{u \tensor v}{N} : C}
+\end{mathpar}
+
+\begin{mathpar}
+    \infer*[right=($\lolli I$)]
+    {\Delta , u{:}A \vdash M : B}
+    {\Delta \vdash \lambda u. M : A \lolli B}
+
+    \and
+
+    \infer*[right=($\lolli E$)]
+    {\Delta \vdash M : A \lolli B \and \Delta' \vdash N : A}
+    {\Delta, \Delta' \vdash M~N : B}
+\end{mathpar}
+
+\begin{mathpar}
+    \infer*[right=($\with I$)]
+    {\Delta \vdash M : A \and \Delta \vdash N : B}
+    {\Delta \vdash M \with N : A \with B}
+
+    \and
+
+    \infer*[right=($\with E_L$)]
+    {\Delta \vdash M : A \with B}
+    {\Delta \vdash \textrm{fst}~M : A}
+
+    \and
+
+    \infer*[right=($\with E_R$)]
+    {\Delta \vdash M : A \with B}
+    {\Delta \vdash \textrm{snd}~M : B}
+\end{mathpar}
+
+\begin{mathpar}
+    \infer*[right=($\oplus I_L$)]
+    {\Delta \vdash M : A}
+    {\Delta \vdash \textrm{inl}~M : A \oplus B}
+
+    \and
+
+    \infer*[right=($\oplus I_R$)]
+    {\Delta \vdash M : B}
+    {\Delta \vdash \textrm{inr}~M : A \oplus B}
+
+    \and
+
+    \infer*[right=($\oplus E$)]
+    {\Delta \vdash M : A \oplus B \and \Delta', w_1{:}A \vdash N_1 : C \and
+    \Delta', w_2{:}B \vdash N_2 : C}
+    {\Delta, \Delta' \vdash \ccase{M}{\textrm{inl}~w_1 \to N_1 \mid \textrm{inr}~w_2 \to N_2} : C}
+\end{mathpar}
+
+\begin{mathpar}
+    \infer*[right=($\star I$)]
+    { }
+    {\Delta \vdash \star : \star}
+
+    \and
+
+    \infer*[right=($\star E$)]
+    {\Delta \vdash M : \star \and \Delta' \vdash N : B}
+    {\Delta, \Delta' \vdash \llet{\star = M}{N} : B}
+
+\end{mathpar}
+
+\begin{mathpar}
+    \infer*[right=($\bang I$)]
+    {\Gamma ; \cdot \vdash M : A}
+    {\Gamma ; \cdot \vdash !M : !A}
+
+    \and
+
+    \infer*[right=($\bang E$)]
+    {\Gamma ; \Delta \vdash M : \bang A \and \Gamma, u{:}A ; \Delta' \vdash N : C}
+    {\Gamma ; \Delta, \Delta' \vdash \llet{!u = M}{N} : C}
+\end{mathpar}
 

### 2023-01-16 report.tex
Expand Haskell section introduction

@@
 
-\begin{itemize}
-    \item Ideia dos linear types
-    \item Calculo lambda linear (como forma de falar da linearidade)
-    \item Gestão de recursos
-    \item Rust (core is linear lambda calculus?)
-\end{itemize}
+% \begin{itemize}
+%     \item Ideia dos linear types
+%     \item Calculo lambda linear (como forma de falar da linearidade)
+%     \item Gestão de recursos
+%     \item Rust (core is linear lambda calculus?)
+% \end{itemize}
 
@@
 arguments are \emph{not} evaluated when passed to functions, but rather are only
-evaluated when the value is needed.
+evaluated when the value is needed. The combination of purety and laziness is
+unique to Haskell among mainstream programming languages...
+
+Haskell is a large language but its small core is based on a typed lambda
+calculus. As such, there exist no statements and computation is done simply
+through the evaluation of functions. Besides functions, one can define types and
+their constructors and pattern match on said constructors. Function application
+is denoted by an empty space (\textbf{f~a} means \textbf{f} applied to
+\textbf{a}) and pattern matching is done with the \textbf{case} keyword followed
+by the enumerated alternatives. All variable names start with lower case and
+types start with upper case (excluding type variables). To make explicit the
+type of an expression, the \textbf{::} operator is used (e.g.
+$\textbf{f~::~Int}\to \textbf{Bool}$ is read \textbf{f} \emph{has type} function
+from \textbf{Int} to \textbf{Bool}).
+
+Because Haskell is a pure programming language, input/ouptut side-effects are
+"made pure" by modelling them on the type level through the
+higher-kinded\footnote{\textbf{IO} has kind $\textbf{Type}\to\textbf{Type}$,
+that is, it is only a type after another type is passed as a parameter (e.g.
+\textbf{IO~Int}, \textbf{IO~Bool}); \textbf{IO} by itself is a \emph{type
+constructor}} type \textbf{IO}. Some of the example programs will look though as
+if they had statements, but, in reality, the sequential appearence is just
+syntatic sugar to an expression using Monadic operators. The main take away is
+that computations that do I/O may be sequenced together with other operations
+that do I/O while retaining the lack of statements and the language purity
+guarantees.
+
+% TODO: Perhaps elaborate a bit more on the constructors and pattern matching?
+
+As an example, consider these functions that do I/O and their types. The first
+opens a file by path and returns its handle, the second gets the size of a file
+from its handle, and the third closes the handle. It is important that the
+handle be closed exactly once, but currently nothing enforces that usage policy.
+
+\begin{code}
+openFile :: FilePath -> IOMode -> IO Handle
+hFileSize :: Handle -> IO Integer
+hClose   :: Handle -> IO ()
+\end{code}
+
+The following function makes use of the above definitions to return the size of
+a file given its path.
+\begin{code}
+countWords :: FilePath -> IO Integer
+countWords path = do
+    handle <- openFile path ReadMode
+    size   <- hFileSize handle
+    return size
+\end{code}
+% hClose handle
 

### 2023-01-18 report.tex
Start writing about GADTs

@@
 if they had statements, but, in reality, the sequential appearence is just
-syntatic sugar to an expression using Monadic operators. The main take away is
+syntatic sugar to an expression using monadic operators. The main take away is
 that computations that do I/O may be sequenced together with other operations
@@
 The following function makes use of the above definitions to return the size of
-a file given its path.
+a file given its path. Note that the function silently leaks the handle to the
+file, despite compiling successfully.
+
 \begin{code}
@@
 
-Another defining feature of Haskell is its powerful type system....
-
-% TODO: File example
-
-A function to compute the length of a list in Haskell is defined by pattern
-matching on the list and returning $0$ if the list is empty and adding one to
-the (computed recursively) length of the rest of the list otherwise. The
-following snippet defines the said \textbf{length} function.
-\begin{code}
-    length [] = 0
-    length (x:xs) = 1 + length xs
-\end{code}
-The function \textbf{length} receives a list and returns an integer, and since
-it treats all list's elements the same, it is said to be polymorphic over the
-list's elements type. In Haskell, function signatures typically appear above the
-definition, and for this particular function we'd have
-\begin{code}
-    length :: [a] -> Int
-\end{code}
-where \textbf{a} is a universally quantified type variable, that is,
-\textbf{length} can be called on lists of ints, lists of characters, lists of
-lists of strings, lists of anything.
-
-The laziness evaluation strategy entails expressions are only evaluated when
-needed. If we were to call just the above function on a list of computationally
-expensive elements, such as
-\begin{code}
-    length [5^1234, 6^4567, 8^(9^42)]
-\end{code}
-We would get the result 3 instantaneously: the elements of the list would never
-be evaluated because the \textbf{length} function does not require the value of
-each element to compute a result.
+Another defining feature of Haskell is its powerful type system. Features such
+as type classes, arbitrary-rank types, functional dependencies, GADTs, type
+families and (now) even impredicative polymorphism\cite{10.1145/3408971} have become
+commonplace in Haskell code enforcing application level invariants and program
+correctness through the types. For example, consider the definition of
+\textbf{head} in the standard library, a function which takes the first element
+of a list by pattern matching on the list constructors.
+
+\begin{code}
+    head :: [a] -> a
+    head [] = error "List is empty!"
+    head (x:xs) = x
+\end{code}
+
+If the list is empty, we terminate the program with an error, otherwise, we take
+the first element. This function is unsafe -- our program might crash if we use
+it on an invalid input. A common technique in Haskell is to use more expressive
+types to assert properties about the values and get rid of the impossible cases
+(e.g. a \textbf{NonEmpty} is a list that can not be empty). A well liked motto is
+"make invalid states unrepresentable". In this light, we will introduce
+Generalized Algebraic Data Types (GADTs) and create a list type indexed by size
+for which we can write a completely safe \textbf{head} function by expressing
+that the size of the list must be at least one, at the type level.
+
+\textbf{GADTs}~\cite{10.1145/1159803.1159811,,,} are a wildly popular type level
+feature that allow the user to define data types by enumerating its constructors
+with explicit type signatures, where the result type may differ in the type
+parameters across branches; pattern matching against GADTs might introduce type
+refinements locally, i.e. in individual case alternatives. We develop the
+length-indexed lists example without discussing the type system and type
+inference details of GADTs which we explore later in~\ref{related-work-gadts}.
+First, we define the natural numbers inductively: a natural number is either
+zero (\textbf{Z}) or a successor (\textbf{S}) of another natural number (e.g.
+the successor of the successor of zero is a natural number). The following
+definition creates the \textbf{Nat} type, the \textbf{Z} and \textbf{S} data
+constructors which construct values, \emph{and}, the \textbf{Z} and \textbf{S}
+\emph{type} constructors which construct types. The latter two exist due to the
+\emph{DataKinds} language feature which promotes term-level constructors to
+type-level constructors.
+%
+\begin{code}
+    data Nat = Z | S Nat
+\end{code}
+% Before introducing the data type of
+
+Besides the aforementioned type-level features of Haskell, \emph{linear types},
+which allow for a finer control over the usage of certain resources at the type
+level, are another example of an extension to Haskell's type system that makes
+it yet more expressive.
+
+% A function to compute the length of a list in Haskell is defined by pattern
+% matching on the list and returning $0$ if the list is empty and adding one to
+% the (computed recursively) length of the rest of the list otherwise. The
+% following snippet defines the said \textbf{length} function.
+% \begin{code}
+%     length [] = 0
+%     length (x:xs) = 1 + length xs
+% \end{code}
+% The function \textbf{length} receives a list and returns an integer, and since
+% it treats all list's elements the same, it is said to be polymorphic over the
+% list's elements type. In Haskell, function signatures typically appear above the
+% definition, and for this particular function we'd have
+% \begin{code}
+%     length :: [a] -> Int
+% \end{code}
+% where \textbf{a} is a universally quantified type variable, that is,
+% \textbf{length} can be called on lists of ints, lists of characters, lists of
+% lists of strings, lists of anything.
+
+% The laziness evaluation strategy entails expressions are only evaluated when
+% needed. If we were to call just the above function on a list of computationally
+% expensive elements, such as
+% \begin{code}
+%     length [5^1234, 6^4567, 8^(9^42)]
+% \end{code}
+% We would get the result 3 instantaneously: the elements of the list would never
+% be evaluated because the \textbf{length} function does not require the value of
+% each element to compute a result.
 
@@
 \begin{itemize}
-    \item Features fundamentais
-    \item Exemplos, introduzir syntax
-    \item (Ao contrário de OCaml) tem um sistema de tipos muito sufesticado
+    % \item Features fundamentais
+    % \item Exemplos, introduzir syntax
+    \item (Ao contrário de OCaml) tem um sistema de tipos muito sofisticado
     \item Eventualmente explicação de GADTs? c/ exemplo do Vec
@@
 
-\section{OutsideIn(X)}
+\section{OutsideIn(X)\label{related-work-gadts}}
 

### 2023-01-18 report.tex
Write more on the length-indexed list example

@@
 "made pure" by modelling them on the type level through the
-higher-kinded\footnote{\textbf{IO} has kind $\textbf{Type}\to\textbf{Type}$,
+non-nullary\footnote{\textbf{IO} has kind $\textbf{Type}\to\textbf{Type}$,
 that is, it is only a type after another type is passed as a parameter (e.g.
 \textbf{IO~Int}, \textbf{IO~Bool}); \textbf{IO} by itself is a \emph{type
-constructor}} type \textbf{IO}. Some of the example programs will look though as
+constructor}} type constructor \textbf{IO}. Some of the example programs will look though as
 if they had statements, but, in reality, the sequential appearence is just
@@
 
-\textbf{GADTs}~\cite{10.1145/1159803.1159811,,,} are a wildly popular type level
-feature that allow the user to define data types by enumerating its constructors
-with explicit type signatures, where the result type may differ in the type
-parameters across branches; pattern matching against GADTs might introduce type
-refinements locally, i.e. in individual case alternatives. We develop the
-length-indexed lists example without discussing the type system and type
-inference details of GADTs which we explore later in~\ref{related-work-gadts}.
+% TODO: But what are ADTs? Perhaps expand Haskell introduction section to
+% include recursive ADTs
+\textbf{GADTs}~\cite{10.1145/1159803.1159811,,,} are a very popular feature with
+which users can define data types as they would common recursive algebraic data
+types, but that allows explicit type signatures for the data constructors where
+the result type may differ in the type parameters across branches. Pattern
+matching against GADTs can introduce local type refinements, that is, pattern
+matching against certain data constructor will refine the type information known
+for typechecking in individual case alternatives. We develop the length-indexed
+lists example without discussing the type system and type inference details of
+GADTs which we explore later in~\ref{related-work-gadts}.
+
+% TODO: Explain Type in Type?
+
 First, we define the natural numbers inductively: a natural number is either
@@
 definition creates the \textbf{Nat} type, the \textbf{Z} and \textbf{S} data
-constructors which construct values, \emph{and}, the \textbf{Z} and \textbf{S}
-\emph{type} constructors which construct types. The latter two exist due to the
+constructors which construct values, \emph{and} the \textbf{Z} and \textbf{S}
+\emph{type} constructors which construct types. The latter two exist because of
 \emph{DataKinds} language feature which promotes term-level constructors to
 type-level constructors.
-%
+
 \begin{code}
@@
 \end{code}
-% Before introducing the data type of
-
-Besides the aforementioned type-level features of Haskell, \emph{linear types},
-which allow for a finer control over the usage of certain resources at the type
-level, are another example of an extension to Haskell's type system that makes
-it yet more expressive.
+
+Secondly, we define the data type in GADT syntax for length-index lists which
+takes two type parameters. The first type parameter is the length of the list
+and the type of the type parameter (i.e. the kind of the first type parameter)
+is \textbf{Nat}. To construct a type of kind \textbf{Nat} we can only use the
+type constructors \textbf{Z} and \textbf{S}. The second type parameter is the
+type of the values contained in the list, and any type is valid, hence the
+\textbf{Type} kind.
+
+\begin{code}
+    data Vec (n :: Nat) (a :: Type) where
+        Nil :: Vec Z a
+        Cons :: a -> Vec m a -> Vec (S m) a
+\end{code}
+
+The length-indexed list is defined inductively as either an empty list \emph{of
+size zero}, or the construction of a list by appending a new element to an
+existing list \emph{of size $m$} whose final size is $m+1$ (\textbf{S m}). The
+list \textbf{Cons~2~(Cons~1~Nil)} has type \textbf{Vec~(S~(S~Z))~Int} because
+\textbf{Nil~::~Vec~Z~Int} and \textbf{Cons~1~Nil~::~Vec~(S~Z)~Int}. GADTs make
+possible different data type constructors being parametrized over different type
+parameters as we do with \textbf{Vec}'s size parameter being different depending
+on the constructor that constructs the list.
+    To define the safe \textbf{head} function, we must specify the type of the
+input list taking into account that the size must not be zero. To that effect,
+the function takes as input a \textbf{Vec (S n) a}, that is, a vector with size
+(n+1) for all possible n's, which makes it impossible to pass a list of type
+\textbf{Vec Z a} to \textbf{head}.
+
+\begin{code}
+    head :: Vec (S n) a -> a
+    head (Cons x xs) = x
+\end{code}
+
+Pattern matching on the \textbf{Nil} constructor would result in an innacessible
+pattern warning because it is impossible to pass a list that would match
+\textbf{Nil} -- it would entail the list has size zero; however, it's impossible
+to call \textbf{head} on a list of size zero!
+
+Linear types, which allow for a finer control over the usage of certain resources at the type
+level, are another example of an extension to Haskell's type system, besides the aforementioned type-level
+features, that makes it even more expressive.
 

### 2023-01-18 laboratory/Vec.hs

@@ -1,0 +1,21 @@
+{-# LANGUAGE DataKinds #-}
+{-# LANGUAGE GADTs #-}
+{-# LANGUAGE GHC2021 #-}
+
+import Data.Kind
+import Prelude hiding (head)
+
+data Nat = Z | S Nat
+
+data Vec (n :: Nat) (a :: Type) where
+  Nil :: Vec Z a
+  Cons :: a -> Vec m a -> Vec (S m) a
+
+head :: Vec (S n) a -> a
+head (Cons x _) = x
+head Nil = undefined
+
+main = do
+  pure $ head $ Cons '1' Nil
+  -- pure $ head $ Nil
+  pure ()

### 2023-01-18 report.tex
Some ideas on empty sections

@@
 
-\begin{itemize}
+% \begin{itemize}
     % \item Features fundamentais
     % \item Exemplos, introduzir syntax
-    \item (Ao contrário de OCaml) tem um sistema de tipos muito sofisticado
-    \item Eventualmente explicação de GADTs? c/ exemplo do Vec
-    \item Uma das features mais avaçadas/experimentais é o Linear Haskell
-\end{itemize}
-
-\subsection{Linear Haskell\label{linear-haskell-section}}
+    % \item (Ao contrário de OCaml) tem um sistema de tipos muito sofisticado
+    % \item Eventualmente explicação de GADTs? c/ exemplo do Vec
+    % \item Uma das features mais avaçadas/experimentais é o Linear Haskell
+% \end{itemize}
+
+\section{Linear Haskell\label{linear-haskell-section}}
+
+Introduction of linear types to Haskell. How do they work, what's the syntax?
+Multiplicities, examples. Monads, multiplicity polymorphism, multiplicity
+polymoprhism applied to monadic bind to define resource-aware sequencing of
+operations. Linear haskell relates systemFC with the linear lambda calculus?
 
@@
 
-\subsection{Core Haskell\label{core-section}}
+\section{Core Haskell\label{core-section}}
+
+Haskel's intermediate representation that is statically typed
 
@@
 
+The architecture of GHC
+Both typecheckers, the backends Core is transformed into, and \textbf{\emph{all
+core transformations}}.
+
 \begin{itemize}
@@
 
+A brief introduction to the related work section?
+
 \section{System FC}
 
+Haskell Core's foundational language was imbued with linear types, but it does
+not account for linearity with the whole of the type system
+
+Multiplicity annotations in SystemFC?
+
+Rules?
+
 \section{Linear Haskell}
 
+The paper is concerned with retrofitted lienar types but couldn't account for
+optimizations in the Core. It originally did not set out to modify Core but had
+to, and it's better this way bc we get optimizations typechecking safety
+guarantees.
+
 \section{OutsideIn(X)\label{related-work-gadts}}
+
+Defines constraint-based type system parametrized over X which does not account
+for local type refinements regarding linearity.
 
@@
 
-They have a linearly typed something?
+Rust has a core based on linear types. Describe Rust's architecture?
+How do they handle linearity plus optimizations
+They probabluy don't typecheck linearity in Core
 

### 2023-01-19 report.tex
rules
tweaks

@@
    \end{array}
-
 \and
-
   \begin{array}{lcl}
@@
 
-\begin{mathpar}
+\[
+  \begin{array}{c}
     \infer*[right=($\tensor I$)]
@@
     {\Delta , \Delta' \vdash (M \tensor N) : A \tensor B}
-
-    \and
-\end{mathpar}
-
-\begin{mathpar}
+\qquad
     \infer*[right=($\tensor E$)]
@@
     {\Delta , \Delta' \vdash \llet{u \tensor v}{N} : C}
-\end{mathpar}
-
-\begin{mathpar}
+\\[1em]
     \infer*[right=($\lolli I$)]
@@
     {\Delta \vdash \lambda u. M : A \lolli B}
-
-    \and
-
+\qquad
     \infer*[right=($\lolli E$)]
@@
     {\Delta, \Delta' \vdash M~N : B}
-\end{mathpar}
-
-\begin{mathpar}
+\\[1em]
     \infer*[right=($\with I$)]
@@
     {\Delta \vdash M \with N : A \with B}
-
-    \and
-
+    \qquad
     \infer*[right=($\with E_L$)]
@@
     {\Delta \vdash \textrm{fst}~M : A}
-
-    \and
-
+\qquad
     \infer*[right=($\with E_R$)]
@@
     {\Delta \vdash \textrm{snd}~M : B}
-\end{mathpar}
-
-\begin{mathpar}
+\\[1em]
     \infer*[right=($\oplus I_L$)]
@@
 
-    \and
-
+\qquad
     \infer*[right=($\oplus I_R$)]
@@
     {\Delta \vdash \textrm{inr}~M : A \oplus B}
-
-    \and
+\\[1em]
 
@@
     {\Delta, \Delta' \vdash \ccase{M}{\textrm{inl}~w_1 \to N_1 \mid \textrm{inr}~w_2 \to N_2} : C}
-\end{mathpar}
-
-\begin{mathpar}
+    \\[1em]
     \infer*[right=($\star I$)]
@@
     {\Delta \vdash \star : \star}
-
-    \and
-
-    \infer*[right=($\star E$)]
+\quad    \infer*[right=($\star E$)]
     {\Delta \vdash M : \star \and \Delta' \vdash N : B}
     {\Delta, \Delta' \vdash \llet{\star = M}{N} : B}
-
-\end{mathpar}
-
-\begin{mathpar}
+\\[1em]
     \infer*[right=($\bang I$)]
@@
     {\Gamma ; \cdot \vdash !M : !A}
-
-    \and
-
-    \infer*[right=($\bang E$)]
+\qquad    \infer*[right=($\bang E$)]
     {\Gamma ; \Delta \vdash M : \bang A \and \Gamma, u{:}A ; \Delta' \vdash N : C}
     {\Gamma ; \Delta, \Delta' \vdash \llet{!u = M}{N} : C}
-\end{mathpar}
+  \end{array}
+  \]
 
@@
 as type classes, arbitrary-rank types, functional dependencies, GADTs, type
-families and (now) even impredicative polymorphism\cite{10.1145/3408971} have become
-commonplace in Haskell code enforcing application level invariants and program
+families and (more recently) even impredicative polymorphism\cite{10.1145/3408971} have become
+commonplace in Haskell code, enforcing application level invariants and program
 correctness through the types. For example, consider the definition of
@@
 
-If the list is empty, we terminate the program with an error, otherwise, we take
-the first element. This function is unsafe -- our program might crash if we use
-it on an invalid input. A common technique in Haskell is to use more expressive
-types to assert properties about the values and get rid of the impossible cases
-(e.g. a \textbf{NonEmpty} is a list that can not be empty). A well liked motto is
-"make invalid states unrepresentable". In this light, we will introduce
-Generalized Algebraic Data Types (GADTs) and create a list type indexed by size
-for which we can write a completely safe \textbf{head} function by expressing
-that the size of the list must be at least one, at the type level.
+Note that if the list is empty, we terminate the program with an error. This
+function is unsafe -- our program might crash if we use it on an invalid input.
+A common technique in Haskell is to use more expressive types to assert
+properties about the values and get rid of the unsafe cases (e.g. a
+\textbf{NonEmpty} is a list that can not be empty). A well liked motto is "make
+invalid states unrepresentable". In this light, we will introduce Generalized
+Algebraic Data Types (GADTs) and create a list type indexed by size for which we
+can write a completely safe \textbf{head} function. This is done by expressing that the size
+of the list must be at least one at the type level.
 
@@
 which users can define data types as they would common recursive algebraic data
-types, but that allows explicit type signatures for the data constructors where
-the result type may differ in the type parameters across branches. Pattern
-matching against GADTs can introduce local type refinements, that is, pattern
-matching against certain data constructor will refine the type information known
-for typechecking in individual case alternatives. We develop the length-indexed
-lists example without discussing the type system and type inference details of
-GADTs which we explore later in~\ref{related-work-gadts}.
+types, with the added abilitiy to give explicit type signatures for the data
+constructors where the result type may differ in the type parameters across
+branches. Pattern matching against GADTs can introduce local type refinements,
+that is, pattern matching against certain data constructors will refine the type
+information used for typechecking individual case alternatives. We develop
+the length-indexed lists example without discussing the type system and type
+inference details of GADTs which we later explore in~\ref{related-work-gadts}.
 
@@
 \emph{type} constructors which construct types. The latter two exist because of
-\emph{DataKinds} language feature which promotes term-level constructors to
+the \emph{DataKinds} language feature which promotes term-level constructors to
 type-level constructors.
@@
 existing list \emph{of size $m$} whose final size is $m+1$ (\textbf{S m}). The
-list \textbf{Cons~2~(Cons~1~Nil)} has type \textbf{Vec~(S~(S~Z))~Int} because
-\textbf{Nil~::~Vec~Z~Int} and \textbf{Cons~1~Nil~::~Vec~(S~Z)~Int}. GADTs make
-possible different data type constructors being parametrized over different type
-parameters as we do with \textbf{Vec}'s size parameter being different depending
-on the constructor that constructs the list.
-    To define the safe \textbf{head} function, we must specify the type of the
-input list taking into account that the size must not be zero. To that effect,
-the function takes as input a \textbf{Vec (S n) a}, that is, a vector with size
+list \textbf{Cons~'a'~(Cons~'b'~Nil)} has type \textbf{Vec~(S~(S~Z))~Char}
+because \textbf{Nil~::~Vec~Z~Char} and \textbf{Cons~'a'~Nil~::~Vec~(S~Z)~Char}.
+GADTs make possible different data constructors being parametrized over
+different type parameters as we do with \textbf{Vec}'s size parameter being
+different depending on the constructor that constructs the list.
+
+To define the safe \textbf{head} function, we must specify the type of the input
+list taking into account that the size must not be zero. To that effect, the
+function takes as input a \textbf{Vec (S n) a}, that is, a vector with size
 (n+1) for all possible n's, which makes it impossible to pass a list of type
@@
 pattern warning because it is impossible to pass a list that would match
-\textbf{Nil} -- it would entail the list has size zero; however, it's impossible
-to call \textbf{head} on a list of size zero!
+\textbf{Nil} -- it would entail the argument list has size zero; however, it's
+impossible to call \textbf{head} on a list of size zero!
 
 Linear types, which allow for a finer control over the usage of certain resources at the type
-level, are another example of an extension to Haskell's type system, besides the aforementioned type-level
-features, that makes it even more expressive.
+level, are another example of an extension to Haskell's type system
+%, besides the aforementioned type-level features
+that makes it even more expressive.
 

### 2023-01-19 report.tex
typeset rules and code

@@
 Under the lens of a linear type system, consider the variable $p$ to be a linear
-resource created by the \textbf{malloc} keyword. Because $p$ is a linear
+resource created by the \texttt{malloc} keyword. Because $p$ is a linear
 variable, it must be used exactly once. However, we are using it in two
-different calls to \textbf{free}. With our new linearity assumptions, the
+different calls to \texttt{free}. With our new linearity assumptions, the
 % TODO:
@@
 them. The first function ignores the received file handle and returns $\star$
-(read unit), which is equivalent to C's \textbf{void}.
+(read unit), which is equivalent to C's \texttt{void}.
 % TODO: Better code formatting
@@
 %
-Additionally, both functions share the same type, $\textbf{Handle} \to \star$, i.e.
-a function that takes a \textbf{Handle} and returns $\star$. The second function also
-shares this type because \textbf{close} has type $\textbf{Handle} \to \star$.
+Additionally, both functions share the same type, $\texttt{Handle} \to \star$, i.e.
+a function that takes a \texttt{Handle} and returns $\star$. The second function also
+shares this type because \texttt{close} has type $\texttt{Handle} \to \star$.
 %
@@
 The key observation to invalidating these programs is to focus on the function
-type going between \textbf{Handle} and $\star$ and augment it to indicate that
+type going between \texttt{Handle} and $\star$ and augment it to indicate that
 \emph{the argument must be used exactly once}, or, in other words, that the
@@
 %
-Providing the more restrictive linear function signature $\textbf{Handle} \lolli
+Providing the more restrictive linear function signature $\texttt{Handle} \lolli
 \star$ to the example programs would make both of them fail to typecheck because
@@
 
-\begin{mathpar}
+\begin{figure}[h]
+\[
+\begin{array}{c}
   \begin{array}{lcl}
-    s,t & ::= & \star\\
-        & \mid & t \lolli s\\
-        & \mid & t \oplus s\\
-        & \mid & t \tensor s\\
-        & \mid & t \with s\\
-        & \mid & \bang t\\
+    A,B & ::= & \star\\
+        & \mid & A \lolli B\\
+        & \mid & A \oplus B\\
+        & \mid & A \tensor B\\
+        & \mid & A \with B\\
+        & \mid & \bang A\\
    \end{array}
-\and
+\qquad
   \begin{array}{lcl}
@@
         & \mid & \textrm{inl}~M \mid \textrm{inr}~M \mid
-        \ccase{M}{\textrm{inl}~u_1 \to N_1 \mid \textrm{inr}~u_2 \to N_2}\\
+        \ccase{M}{\textrm{inl}~u_1 \to N_1;\textrm{inr}~u_2 \to N_2}\\
         & \mid & M \tensor N \mid \llet{u_1 \tensor u_2 = M}{N}\\
@@
    \end{array}
-\end{mathpar}
-
+\end{array}
+\]
+\caption{Grammar for a linearly-typed lambda calculus}
+\end{figure}
+
+\begin{figure}[h]
 \[
@@
     \infer*[right=($\tensor I$)]
-    {\Delta \vdash M : A \and \Delta' \vdash N : B}
-    {\Delta , \Delta' \vdash (M \tensor N) : A \tensor B}
+    {\Gamma ; \Delta \vdash M : A \and \Gamma ; \Delta' \vdash N : B}
+    {\Gamma ; \Delta , \Delta' \vdash (M \tensor N) : A \tensor B}
 \qquad
     \infer*[right=($\tensor E$)]
-    {\Delta \vdash M : A \tensor B \and \Delta', u{:}A, v{:}B \vdash N : C}
-    {\Delta , \Delta' \vdash \llet{u \tensor v}{N} : C}
+    {\Gamma ; \Delta \vdash M : A \tensor B \and \Gamma ; \Delta', u{:}A, v{:}B \vdash N : C}
+    {\Gamma ; \Delta , \Delta' \vdash \llet{u \tensor v}{N} : C}
 \\[1em]
     \infer*[right=($\lolli I$)]
-    {\Delta , u{:}A \vdash M : B}
-    {\Delta \vdash \lambda u. M : A \lolli B}
+    {\Gamma ; \Delta , u{:}A \vdash M : B}
+    {\Gamma ; \Delta \vdash \lambda u. M : A \lolli B}
 \qquad
     \infer*[right=($\lolli E$)]
-    {\Delta \vdash M : A \lolli B \and \Delta' \vdash N : A}
-    {\Delta, \Delta' \vdash M~N : B}
+    {\Gamma ; \Delta \vdash M : A \lolli B \and \Gamma ; \Delta' \vdash N : A}
+    {\Gamma ; \Delta, \Delta' \vdash M~N : B}
 \\[1em]
     \infer*[right=($\with I$)]
-    {\Delta \vdash M : A \and \Delta \vdash N : B}
-    {\Delta \vdash M \with N : A \with B}
+    {\Gamma ; \Delta \vdash M : A \and \Gamma ; \Delta \vdash N : B}
+    {\Gamma ; \Delta \vdash M \with N : A \with B}
     \qquad
     \infer*[right=($\with E_L$)]
-    {\Delta \vdash M : A \with B}
-    {\Delta \vdash \textrm{fst}~M : A}
+    {\Gamma ; \Delta \vdash M : A \with B}
+    {\Gamma ; \Delta \vdash \textrm{fst}~M : A}
 \qquad
     \infer*[right=($\with E_R$)]
-    {\Delta \vdash M : A \with B}
-    {\Delta \vdash \textrm{snd}~M : B}
+    {\Gamma ; \Delta \vdash M : A \with B}
+    {\Gamma ; \Delta \vdash \textrm{snd}~M : B}
 \\[1em]
     \infer*[right=($\oplus I_L$)]
-    {\Delta \vdash M : A}
-    {\Delta \vdash \textrm{inl}~M : A \oplus B}
-
+    {\Gamma ; \Delta \vdash M : A}
+    {\Gamma ; \Delta \vdash \textrm{inl}~M : A \oplus B}
 \qquad
     \infer*[right=($\oplus I_R$)]
-    {\Delta \vdash M : B}
-    {\Delta \vdash \textrm{inr}~M : A \oplus B}
+    {\Gamma ; \Delta \vdash M : B}
+    {\Gamma ; \Delta \vdash \textrm{inr}~M : A \oplus B}
 \\[1em]
-
     \infer*[right=($\oplus E$)]
-    {\Delta \vdash M : A \oplus B \and \Delta', w_1{:}A \vdash N_1 : C \and
-    \Delta', w_2{:}B \vdash N_2 : C}
-    {\Delta, \Delta' \vdash \ccase{M}{\textrm{inl}~w_1 \to N_1 \mid \textrm{inr}~w_2 \to N_2} : C}
+    {\Gamma ; \Delta \vdash M : A \oplus B \and \Gamma ; \Delta', w_1{:}A \vdash N_1 : C \and
+    \Gamma ; \Delta', w_2{:}B \vdash N_2 : C}
+    {\Gamma ; \Delta, \Delta' \vdash \ccase{M}{\textrm{inl}~w_1 \to N_1 \mid \textrm{inr}~w_2 \to N_2} : C}
     \\[1em]
@@
     { }
-    {\Delta \vdash \star : \star}
+    {\Gamma ; \Delta \vdash \star : \star}
 \quad    \infer*[right=($\star E$)]
-    {\Delta \vdash M : \star \and \Delta' \vdash N : B}
-    {\Delta, \Delta' \vdash \llet{\star = M}{N} : B}
+    {\Gamma ; \Delta \vdash M : \star \and \Gamma ; \Delta' \vdash N : B}
+    {\Gamma ; \Delta, \Delta' \vdash \llet{\star = M}{N} : B}
 \\[1em]
@@
 \]
+\caption{Typing rules for a linearly-typed lambda calculus}
+\end{figure}
 
@@
 their constructors and pattern match on said constructors. Function application
-is denoted by an empty space (\textbf{f~a} means \textbf{f} applied to
-\textbf{a}) and pattern matching is done with the \textbf{case} keyword followed
+is denoted by an empty space (\texttt{f~a} means \texttt{f} applied to
+\texttt{a}) and pattern matching is done with the \texttt{case} keyword followed
 by the enumerated alternatives. All variable names start with lower case and
 types start with upper case (excluding type variables). To make explicit the
-type of an expression, the \textbf{::} operator is used (e.g.
-$\textbf{f~::~Int}\to \textbf{Bool}$ is read \textbf{f} \emph{has type} function
-from \textbf{Int} to \textbf{Bool}).
+type of an expression, the \texttt{::} operator is used (e.g.
+$\texttt{f~::~Int}\to \texttt{Bool}$ is read \texttt{f} \emph{has type} function
+from \texttt{Int} to \texttt{Bool}).
 
@@
 "made pure" by modelling them on the type level through the
-non-nullary\footnote{\textbf{IO} has kind $\textbf{Type}\to\textbf{Type}$,
+non-nullary\footnote{\texttt{IO} has kind $\texttt{Type}\to\texttt{Type}$,
 that is, it is only a type after another type is passed as a parameter (e.g.
-\textbf{IO~Int}, \textbf{IO~Bool}); \textbf{IO} by itself is a \emph{type
-constructor}} type constructor \textbf{IO}. Some of the example programs will look though as
+\texttt{IO~Int}, \texttt{IO~Bool}); \texttt{IO} by itself is a \emph{type
+constructor}} type constructor \texttt{IO}. Some of the example programs will look though as
 if they had statements, but, in reality, the sequential appearence is just
@@
 correctness through the types. For example, consider the definition of
-\textbf{head} in the standard library, a function which takes the first element
+\texttt{head} in the standard library, a function which takes the first element
 of a list by pattern matching on the list constructors.
@@
 properties about the values and get rid of the unsafe cases (e.g. a
-\textbf{NonEmpty} is a list that can not be empty). A well liked motto is "make
+\texttt{NonEmpty} is a list that can not be empty). A well liked motto is "make
 invalid states unrepresentable". In this light, we will introduce Generalized
 Algebraic Data Types (GADTs) and create a list type indexed by size for which we
-can write a completely safe \textbf{head} function. This is done by expressing that the size
+can write a completely safe \texttt{head} function. This is done by expressing that the size
 of the list must be at least one at the type level.
@@
 First, we define the natural numbers inductively: a natural number is either
-zero (\textbf{Z}) or a successor (\textbf{S}) of another natural number (e.g.
+zero (\texttt{Z}) or a successor (\texttt{S}) of another natural number (e.g.
 the successor of the successor of zero is a natural number). The following
-definition creates the \textbf{Nat} type, the \textbf{Z} and \textbf{S} data
-constructors which construct values, \emph{and} the \textbf{Z} and \textbf{S}
-\emph{type} constructors which construct types. The latter two exist because of
-the \emph{DataKinds} language feature which promotes term-level constructors to
-type-level constructors.
+definition creates the \texttt{Nat} type, the \texttt{Z} and \texttt{S} data
+constructors which construct values, \emph{and} the \texttt{Z} and \texttt{S}
+\emph{type} constructors which construct types.
+%
+% The latter two exist because of the \emph{DataKinds} language feature which
+% promotes term-level constructors to type-level constructors.
 
@@
 and the type of the type parameter (i.e. the kind of the first type parameter)
-is \textbf{Nat}. To construct a type of kind \textbf{Nat} we can only use the
-type constructors \textbf{Z} and \textbf{S}. The second type parameter is the
+is \texttt{Nat}. To construct a type of kind \texttt{Nat} we can only use the
+type constructors \texttt{Z} and \texttt{S}. The second type parameter is the
 type of the values contained in the list, and any type is valid, hence the
-\textbf{Type} kind.
+\texttt{Type} kind.
 
@@
 size zero}, or the construction of a list by appending a new element to an
-existing list \emph{of size $m$} whose final size is $m+1$ (\textbf{S m}). The
-list \textbf{Cons~'a'~(Cons~'b'~Nil)} has type \textbf{Vec~(S~(S~Z))~Char}
-because \textbf{Nil~::~Vec~Z~Char} and \textbf{Cons~'a'~Nil~::~Vec~(S~Z)~Char}.
+existing list \emph{of size $m$} whose final size is $m+1$ (\texttt{S m}). The
+list \texttt{Cons~'a'~(Cons~'b'~Nil)} has type \texttt{Vec~(S~(S~Z))~Char}
+because \texttt{Nil~::~Vec~Z~Char} and \texttt{Cons~'a'~Nil~::~Vec~(S~Z)~Char}.
 GADTs make possible different data constructors being parametrized over
-different type parameters as we do with \textbf{Vec}'s size parameter being
+different type parameters as we do with \texttt{Vec}'s size parameter being
 different depending on the constructor that constructs the list.
 
-To define the safe \textbf{head} function, we must specify the type of the input
+To define the safe \texttt{head} function, we must specify the type of the input
 list taking into account that the size must not be zero. To that effect, the
-function takes as input a \textbf{Vec (S n) a}, that is, a vector with size
+function takes as input a \texttt{Vec (S n) a}, that is, a vector with size
 (n+1) for all possible n's, which makes it impossible to pass a list of type
-\textbf{Vec Z a} to \textbf{head}.
+\texttt{Vec Z a} to \texttt{head}.
 
@@
 
-Pattern matching on the \textbf{Nil} constructor would result in an innacessible
+Pattern matching on the \texttt{Nil} constructor would result in an innacessible
 pattern warning because it is impossible to pass a list that would match
-\textbf{Nil} -- it would entail the argument list has size zero; however, it's
-impossible to call \textbf{head} on a list of size zero!
+\texttt{Nil} -- it would entail the argument list has size zero; however, it's
+impossible to call \texttt{head} on a list of size zero!
 
@@
 % the (computed recursively) length of the rest of the list otherwise. The
-% following snippet defines the said \textbf{length} function.
+% following snippet defines the said \texttt{length} function.
 % \begin{code}
@@
 % \end{code}
-% The function \textbf{length} receives a list and returns an integer, and since
+% The function \texttt{length} receives a list and returns an integer, and since
 % it treats all list's elements the same, it is said to be polymorphic over the
@@
 % \end{code}
-% where \textbf{a} is a universally quantified type variable, that is,
-% \textbf{length} can be called on lists of ints, lists of characters, lists of
+% where \texttt{a} is a universally quantified type variable, that is,
+% \texttt{length} can be called on lists of ints, lists of characters, lists of
 % lists of strings, lists of anything.
@@
 % We would get the result 3 instantaneously: the elements of the list would never
-% be evaluated because the \textbf{length} function does not require the value of
+% be evaluated because the \texttt{length} function does not require the value of
 % each element to compute a result.

### 2023-01-19 report.tex
texing help

--- 2023-01-18 report.tex
@@
     {\Delta, \Delta' \vdash \ccase{M}{\textrm{inl}~w_1 \to N_1 \mid \textrm{inr}~w_2 \to N_2} : C}
-\end{mathpar}
-
-\begin{mathpar}
+    \\[1em]
     \infer*[right=($\star I$)]
     { }
-    {\Delta \vdash \star : \star}
-
-    \and
-
-    \infer*[right=($\star E$)]
+    {\cdot \vdash \star : \star}
+\quad    \infer*[right=($\star E$)]
     {\Delta \vdash M : \star \and \Delta' \vdash N : B}
     {\Delta, \Delta' \vdash \llet{\star = M}{N} : B}
-
-\end{mathpar}
-
-\begin{mathpar}
+\\[1em]
     \infer*[right=($\bang I$)]
@@
 the first element. This function is unsafe -- our program might crash if we use
-it on an invalid input. A common technique in Haskell is to use more expressive
+it on an invalid input.
+\todo[inline]{When applied to the empty list, ...}
+
+A common technique in Haskell is to use more expressive
 types to assert properties about the values and get rid of the impossible cases
-(e.g. a \textbf{NonEmpty} is a list that can not be empty). A well liked motto is
+(e.g. a \textbf{NonEmpty} is a list that can not be empty).
+
+\todo[inline]{Leveraging Haskell's more advanced features, we can
+  .... }
+
+A well liked motto is
 "make invalid states unrepresentable". In this light, we will introduce
@@
 
-% TODO: But what are ADTs? Perhaps expand Haskell introduction section to
-% include recursive ADTs
-\textbf{GADTs}~\cite{10.1145/1159803.1159811,,,} are a very popular feature with
+\textbf{GADTs}~\cite{10.1145/1159803.1159811,,,} are a very popular\todo{nope} feature with
 which users can define data types as they would common recursive algebraic data
@@
 existing list \emph{of size $m$} whose final size is $m+1$ (\textbf{S m}). The
-list \textbf{Cons~2~(Cons~1~Nil)} has type \textbf{Vec~(S~(S~Z))~Int} because
+list \texttt{Cons~2~(Cons~1~Nil)} has type \textbf{Vec~(S~(S~Z))~Int} because
 \textbf{Nil~::~Vec~Z~Int} and \textbf{Cons~1~Nil~::~Vec~(S~Z)~Int}. GADTs make

### 2023-01-19 report.tex

--- 2023-01-19 report.tex
@@
     { }
-    {\Gamma ; \Delta \vdash \star : \star}
+    {\Gamma ; \cdot \vdash \star : \star}
 \quad    \infer*[right=($\star E$)]
@@
 
-Note that if the list is empty, we terminate the program with an error. This
-function is unsafe -- our program might crash if we use it on an invalid input.
-A common technique in Haskell is to use more expressive types to assert
-properties about the values and get rid of the unsafe cases (e.g. a
-\texttt{NonEmpty} is a list that can not be empty). A well liked motto is "make
-invalid states unrepresentable". In this light, we will introduce Generalized
-Algebraic Data Types (GADTs) and create a list type indexed by size for which we
-can write a completely safe \texttt{head} function. This is done by expressing that the size
-of the list must be at least one at the type level.
-
-% TODO: But what are ADTs? Perhaps expand Haskell introduction section to
-% include recursive ADTs
-\textbf{GADTs}~\cite{10.1145/1159803.1159811,,,} are a very popular feature with
+If the list is empty, we terminate the program with an error, otherwise, we take
+the first element. This function is unsafe -- our program might crash if we use
+it on an invalid input.
+\todo[inline]{When applied to the empty list, ...}
+
+A common technique in Haskell is to use more expressive
+types to assert properties about the values and get rid of the impossible cases
+(e.g. a \textt{NonEmpty} is a list that can not be empty).
+
+\todo[inline]{Leveraging Haskell's more advanced features, we can
+  .... }
+
+A well liked motto is
+"make invalid states unrepresentable". In this light, we will introduce
+Generalized Algebraic Data Types (GADTs) and create a list type indexed by size
+for which we can write a completely safe \texttt{head} function by expressing
+that the size of the list must be at least one, at the type level.
+
+\textbf{GADTs}~\cite{10.1145/1159803.1159811,,,} are a very popular\todo{nope} feature with
 which users can define data types as they would common recursive algebraic data

### 2023-01-19 report.tex
Linear types section

@@
         libraries in mind was described. We describe Linear Haskell in detail in
-        section~\ref{linear-haskell-section}.
+        section~\ref{sec:linear-haskell}.
 
@@
 present a linearly typed lambda calculus, a very simple language with linear
-types, by defining what are syntatically valid programs through the grammar and
-what programs are well typed through the typing rules.
-
-An expression in the linearly typed lambda calculus can be $\star$...
-A type in the linearly typed lambda calculus can be $\star$...
+types, by defining what are syntatically valid programs through the grammar
+in Fig.~\ref{fig:llcgrammar} and what programs are well typed through the typing
+rules in Fig.~\ref{fig:llcrules}. The language features functions and function application ($\lolli$), two
+flavours of pairs, additive ($\with$) and multiplicative ($\tensor$), a
+disjunction operator ($\oplus$) and the $\bang$ modality operator. A typing
+judgement for the linearly typed lambda calculus has the form
+%
+\[ \Gamma; \Delta \vdash M : A \]
+%
+where $\Gamma$ is the context of resources that may be used unrestrictedly, that
+is, any number of times, $\Delta$ is the context of resources that must be used
+linearly (\emph{exactly once}), $M$ is the program to type and $A$ is its type.
+When resources from the linear context are used, they are removed from the
+context and no longer available, and all resources in the linear context must be
+used exactly once.
+
+% Explicar o que é modality operator?
 
@@
 \caption{Grammar for a linearly-typed lambda calculus}
+\label{fig:llcgrammar}
 \end{figure}
+
+The function abstraction is typed according to the linear function introduction
+rule ($\lolli I$). The rule states that a a function abstraction, written
+$\lambda u. M$, is a linear function (i.e. has type $A \lolli B$) given the
+unrestricted context $\Gamma$ and the linear context $\Delta$, if the program $M$
+has type $B$ given the same unrestricted context $\Gamma$ and the linear context
+$\Delta,u{:}A$. That is, if $M$ has type $B$ using $u$ of type $A$ exactly once
+besides the other resources in $\Delta$, then the lambda abstraction has the
+linear function type.
+%
+\[
+    \infer*[right=($\lolli I$)]
+    {\Gamma ; \Delta , u{:}A \vdash M : B}
+    {\Gamma ; \Delta \vdash \lambda u. M : A \lolli B}
+\qquad
+    \infer*[right=($\lolli E$)]
+    {\Gamma ; \Delta \vdash M : A \lolli B \and \Gamma ; \Delta' \vdash N : A}
+    {\Gamma ; \Delta, \Delta' \vdash M~N : B}
+\]
+%
+Function application is typed according to the elimination
+rule for the same type ($\lolli E$). To type an application $M~N$ as $B$, $M$
+must have type $A \lolli B$ and $N$ must have type $A$. To account for the
+linear resources that might be used while proving both that $M{:}A \lolli B$ and
+$N{:}A$, the linear context must be split in two such that both typing judgments
+succeed using exactly once every resource in their linear context (while the
+resources in $\Gamma$ might be used unrestrictedly), hence the separation of the
+linear context in $\Delta$ and $\Delta'$.
+
+The multiplicative pair ($M \tensor N$) is constructed from two linearly typed expressions that
+can each be typed with a division of the given linear context, as we see in its
+introduction rule ($\tensor I$). Upon deconstruction, the multiplicative pair
+elimination rule ($\tensor E$) requires that both of the pair constituents be
+consumed exactly once.
+%
+\[
+    \infer*[right=($\tensor I$)]
+    {\Gamma ; \Delta \vdash M : A \and \Gamma ; \Delta' \vdash N : B}
+    {\Gamma ; \Delta , \Delta' \vdash (M \tensor N) : A \tensor B}
+\qquad
+    \infer*[right=($\tensor E$)]
+    {\Gamma ; \Delta \vdash M : A \tensor B \and \Gamma ; \Delta', u{:}A, v{:}B \vdash N : C}
+    {\Gamma ; \Delta , \Delta' \vdash \llet{u \tensor v}{N} : C}
+\]
+%
+On the other hand, the additive pair requires that both elements of the pair can
+be proved with the same linear context, and upon deconstruction only one of the
+pair elements might be used, rather than both simultaneously.
+
+Finally, the modality operator ($\bang$) constructs a resource that can be used
+unrestrictedly. Its introduction rule ($\bang I$) states that to construct this
+resource means to add a resource to the unrestricted context, which can then be
+used freely. To construct an unrestricted value, however, the linear context
+must be empty -- an unrestricted value can only be constructed if it does not
+depend on any linear resource.
+%
+\[
+    \infer*[right=($\bang I$)]
+    {\Gamma ; \cdot \vdash M : A}
+    {\Gamma ; \cdot \vdash !M : !A}
+\qquad    \infer*[right=($\bang E$)]
+    {\Gamma ; \Delta \vdash M : \bang A \and \Gamma, u{:}A ; \Delta' \vdash N : C}
+    {\Gamma ; \Delta, \Delta' \vdash \llet{!u = M}{N} : C}
+\]
+%
+To utilize an unrestricted value $M$, we must
+bind it to $u$ with $\llet{\bang u = M}{N}$ and can then use it in $N$
+unrestrictedly, because $u$ extends the unrestricted context rather than the
+linear context as we have seen thus far.
+
+In section~\ref{sec:linear-haskell} we describe how linear types are defined in
+Haskell, a programming language more featureful than the linearly typed lambda
+calculus. We will see that, the theoretical principles underlying the linear
+lambda calculus and linear Haskell are the same, and by studying them in this
+minimal setting we can understand them at large.
+
+% TODO: Variable rules
 
@@
 \caption{Typing rules for a linearly-typed lambda calculus}
+\label{fig:llcrules}
 \end{figure}
@@
 
-\todo[inline]{Leveraging Haskell's more advanced features, we can
-  .... }
+\todo[inline]{Leveraging Haskell's more advanced features, we can .... }
 
@@
 
-\section{Linear Haskell\label{linear-haskell-section}}
+\section{Linear Haskell\label{sec:linear-haskell}}
 

### 2023-01-22 report.tex
notas

@@
 In order to further give well defined semantics to a linear type system, we
-present a linearly typed lambda calculus, a very simple language with linear
+present a linearly typed lambda calculus\todo{citar o girard
+  provavelmente ou o Barber (\url{https://www.lfcs.inf.ed.ac.uk/reports/96/ECS-LFCS-96-347/})}, a very simple language with linear
 types, by defining what are syntatically valid programs through the grammar
@@
 flavours of pairs, additive ($\with$) and multiplicative ($\tensor$), a
-disjunction operator ($\oplus$) and the $\bang$ modality operator. A typing
+disjunction operator ($\oplus$)\todo{sum types} and the $\bang$
+modality operator\todo{que serve para}. A typing
 judgement for the linearly typed lambda calculus has the form
@@
     {\Gamma ; \Delta , \Delta' \vdash (M \tensor N) : A \tensor B}
-\qquad
+\quad
     \infer*[right=($\tensor E$)]
@@
     {\Gamma ; \Delta \vdash M \with N : A \with B}
-    \qquad
+    \quad
     \infer*[right=($\with E_L$)]
@@
     {\Gamma ; \Delta \vdash \textrm{fst}~M : A}
-\qquad
+\quad
     \infer*[right=($\with E_R$)]
@@
 
-Haskell is a functional programming language defined by the "Haskell
-Report"\cite{jones1999haskell,marlow2010haskell} and whose \emph{de-facto}
+Haskell is a functional programming language defined by the Haskell
+Report~\cite{jones1999haskell,marlow2010haskell} and whose \emph{de-facto}
 implementation is GHC, the Glasgow Haskell Compiler\cite{GHC}. Haskell is a
@@
 arguments are \emph{not} evaluated when passed to functions, but rather are only
-evaluated when the value is needed. The combination of purety and laziness is
-unique to Haskell among mainstream programming languages...
-
-Haskell is a large language but its small core is based on a typed lambda
+evaluated when its value is needed. The combination of purity and laziness is
+unique to Haskell among mainstream programming languages.
+
+Haskell is a large feature-rich language but its relatively small core is based on a typed lambda
 calculus. As such, there exist no statements and computation is done simply
@@
 their constructors and pattern match on said constructors. Function application
-is denoted by an empty space (\texttt{f~a} means \texttt{f} applied to
+is denoted by an empty space\todo{juxtaposition} (\texttt{f~a} means \texttt{f} applied to
 \texttt{a}) and pattern matching is done with the \texttt{case} keyword followed
@@
 Because Haskell is a pure programming language, input/ouptut side-effects are
-"made pure" by modelling them on the type level through the
+``made pure'' by modelling them on the type level through the
 non-nullary\footnote{\texttt{IO} has kind $\texttt{Type}\to\texttt{Type}$,
@@
 \texttt{IO~Int}, \texttt{IO~Bool}); \texttt{IO} by itself is a \emph{type
-constructor}} type constructor \texttt{IO}. Some of the example programs will look though as
+constructor}} type constructor \texttt{IO}.\todo{IO monad; made pure
+nao quer dizer nada. A ideia é que valores do tipo IO t capturam
+\emph{computacoes} que quando executadas produzem efeitos e produzem
+valores do tipo $t$, os operadores monadicos permitem compor
+computacoes, mas nao executam nada, por isso e que e tudo puro.}
+Some of the example programs will look though as
 if they had statements, but, in reality, the sequential appearence is just
@@
 types to assert properties about the values and get rid of the impossible cases
-(e.g. a \textt{NonEmpty} is a list that can not be empty).
+(e.g. a \texttt{NonEmpty} is a list that can not be empty).
 
@@
 A well liked motto is
-"make invalid states unrepresentable". In this light, we will introduce
+"make invalid states unrepresentable". In this light, we will\todo{nao
+  uses o futuro} introduce
 Generalized Algebraic Data Types (GADTs) and create a list type indexed by size
@@
 \textbf{GADTs}~\cite{10.1145/1159803.1159811,,,} are a very popular\todo{nope} feature with
-which users can define data types as they would common recursive algebraic data
-types, with the added abilitiy to give explicit type signatures for the data
+which users can define data types as they would common algebraic data
+types, with the added ability to give explicit type signatures for the data
 constructors where the result type may differ in the type parameters across
-branches. Pattern matching against GADTs can introduce local type refinements,
+branches.
+\todo{nao ha branches, é nos varios construtores. This allows for
+  additional type invariants to be represented with GADTs through
+  their type parameters, which restricts the use of specific
+  constructors and their subsequent deconstruction through pattern
+  matching}
+
+Pattern matching against GADTs can introduce local type refinements,
 that is, pattern matching against certain data constructors will refine the type
-information used for typechecking individual case alternatives. We develop
+information used for typechecking individual case alternatives.
+\todo{acho que a outra frase que escrevi e melhor que esta / complementa}
+We develop
 the length-indexed lists example without discussing the type system and type
@@
   \end{code}
+
+  \todo[inline]{Acho que a parte dos naturais nao e necessaria}
 

### 2023-01-23 proposal.tex
More on linear haskell

@@
 implementation, their interaction with the whole of GHC is still far
-from trivial. The implemented type system cannot accomodate
+from trivial. The implemented type system cannot accommodate 
 several optimising transformations that produce programs that violate
@@
 lets occur naturally in Core due to optimising transformations that create let
-bindings. In a similar fashion, programs which violate syntatic linearity
+bindings. In a similar fashion, programs which violate syntactic linearity
 for other reasons other than let bindings are produced by Core transformations.
@@
 themselves in their definition body is to perform the computation in two separate passes. First,
-calculate a naive environment by emiting free variables multiplicities as usual
+calculate a naive environment by emitting free variables multiplicities as usual
 while treating the recursive-let bound variables as free ones. Second, pass the

### 2023-01-23 report.tex
More on linear haskell

@@
 \begin{code}
-let p = malloc 4;
- in free p;
-    free p;
+let p = malloc(4);
+ in free(p);
+    free(p);
 \end{code}
@@
         successful approach as implemented in GHC 9.0, the leading Haskell
-        compiler, was based on Linear Haskell\cite{linearhaskell}, where a
+        compiler, was based on Linear Haskell\cite{cite:linearhaskell}, where a
         linear type system designed with retaining backwards-compatibility and
@@
 
+% TODO: Rewrite this whole second part of the introduction.
+% It needs to be much clearer
+
 Additionally, while not yet a reality, linearity in Core could be used to inform
@@
 
-\chapter{Background}
-
-\section{Linear Types\label{linear-types}}
+\chapter{Background and Related Work}
+
+\section{Linear Types\label{sec:linear-types}}
 
@@
 
-Finally, the modality operator ($\bang$) constructs a resource that can be used
-unrestrictedly. Its introduction rule ($\bang I$) states that to construct this
-resource means to add a resource to the unrestricted context, which can then be
-used freely. To construct an unrestricted value, however, the linear context
-must be empty -- an unrestricted value can only be constructed if it does not
-depend on any linear resource.
+Finally, the "of-course" operator ($\bang$) can be used to construct a resource that can be
+used unrestrictedly ($\bang M$). Its introduction rule ($\bang I$) states that
+to construct this resource means to add a resource to the unrestricted context,
+which can then be used freely. To construct an unrestricted value, however, the
+linear context \emph{must be empty} -- an unrestricted value can only be constructed if
+it does not depend on any linear resource.
 %
@@
 
-% TODO: Variable rules
-
 \begin{figure}[h]
@@
   \begin{array}{c}
+    \infer*[right=($u$)]
+    { }
+    {\Gamma ; u{:}A \vdash u : A}
+\qquad
+    \infer*[right=($u$)]
+    { }
+    {\Gamma , u{:}A ; \cdot \vdash u : A}
+\\[1em]
     \infer*[right=($\tensor I$)]
@@
 \emph{lazy}, \emph{purely functional} language, i.e., functions cannot
-have side effects or mutate data, and contrary to many programming languages,
+have side effects or mutate data, and, contrary to many programming languages,
 arguments are \emph{not} evaluated when passed to functions, but rather are only
@@
 
-Another defining feature of Haskell is its powerful type system. Features such
-as type classes, arbitrary-rank types, functional dependencies, GADTs, type
-families and (more recently) even impredicative polymorphism\cite{10.1145/3408971} have become
-commonplace in Haskell code, enforcing application level invariants and program
-correctness through the types. For example, consider the definition of
-\texttt{head} in the standard library, a function which takes the first element
-of a list by pattern matching on the list constructors.
+Another defining feature of Haskell is its powerful type system. In contrast to
+most mainstream programming languages, such as OCaml and Java, Haskell supports a myriad of
+advanced type level features, such as:
+\begin{itemize}
+
+    \item Multiple forms of advanced polymorphism: where languages with whole
+        program type inference usually stick to Damas–Hindley–Milner type
+        inference~\cite{DBLP:conf/popl/DamasM82}, Haskell goes much further with, e.g.,
+        arbitrary-rank types~\cite{10.1017/S0956796806006034}, type-class
+        polymorphism~\cite{10.1145/227699.227700}, levity polymorphism~\cite{10.1145/3062341.3062357},
+        multiplicity polymorphism~\cite{cite:linearhaskell}, and, more recently, impredicative
+        polymorphism~\cite{10.1145/3408971}.
+
+    \item Type level computation by means of type classes and Haskell's type
+        families~\cite{,,}, which permit a direct encoding of type-level
+        functions resembling rewrite rules.
+
+    \item Local equality constraints and existential types by using GADTs, we
+        explain ahead in more detail. A design for first class existential types
+        with bi-directional type inference in Haskell has been published
+        in~\cite{10.1145/3473569}, despite not being yet implemented in GHC.
+
+\end{itemize}
+% Haskell's type families, supports abstraction through type classes, and GADTs
+% which we explain in more detail ahead.  as type classes, arbitrary-rank types,
+% functional dependencies, GADTs, type families and (more recently) even
+% impredicative polymorphism\cite{}
+These advanced features have become commonplace in Haskell code, enforcing
+application level invariants and program correctness through the types. As an
+example to work through this section while we introduce compile-time invariants
+with GADTs, consider the definition of \texttt{head} in the standard library, a
+function which takes the first element of a list by pattern matching on the list
+constructors.
 
@@
 
-If the list is empty, we terminate the program with an error, otherwise, we take
-the first element. This function is unsafe -- our program might crash if we use
-it on an invalid input.
-\todo[inline]{When applied to the empty list, ...}
-
-A common technique in Haskell is to use more expressive
-types to assert properties about the values and get rid of the impossible cases
-(e.g. a \texttt{NonEmpty} is a list that can not be empty).
-
-\todo[inline]{Leveraging Haskell's more advanced features, we can .... }
-
-A well liked motto is
-"make invalid states unrepresentable". In this light, we will\todo{nao
-  uses o futuro} introduce
-Generalized Algebraic Data Types (GADTs) and create a list type indexed by size
-for which we can write a completely safe \texttt{head} function by expressing
-that the size of the list must be at least one, at the type level.
-
-\textbf{GADTs}~\cite{10.1145/1159803.1159811,,,} are a very popular\todo{nope} feature with
-which users can define data types as they would common algebraic data
-types, with the added ability to give explicit type signatures for the data
-constructors where the result type may differ in the type parameters across
+When applied to the empty list, \texttt{head} terminates the program with an
+error. This function is unsafe -- our program might crash if we use it on an
+invalid input.
+%
+Leveraging Haskell's more advanced features, we can use more expressive types to
+assert properties about the values and get rid of the invalid cases (e.g. we
+could define a \texttt{NonEmpty} type to model a list that can not be empty).
+%
+A well liked motto is "make invalid states unrepresentable". In this light, we
+introduce Generalized Algebraic Data Types (GADTs) and create a list type
+indexed by size for which we can write a completely safe \texttt{head} function
+by expressing that the size of the list must be at least one, at the type level.
+
+\textbf{GADTs}~\cite{10.1145/1159803.1159811,,,} are an advanced Haskell feature
+which allows users to define data types as they would common algebraic
+data types, with the added abilitiy to give explicit type signatures for the
+data constructors where the result type may differ in the type parameters across
 branches.
@@
   matching}
-
-Pattern matching against GADTs can introduce local type refinements,
-that is, pattern matching against certain data constructors will refine the type
-information used for typechecking individual case alternatives.
-\todo{acho que a outra frase que escrevi e melhor que esta / complementa}
-We develop
-the length-indexed lists example without discussing the type system and type
+% TODO: Rewrite this bit below
+Pattern matching against GADTs can introduce local type refinements, that is,
+pattern matching against certain data constructors will refine the type
+information used for typechecking individual case alternatives. We develop the
+length-indexed lists example without discussing the type system and type
 inference details of GADTs which we later explore in~\ref{related-work-gadts}.
@@
 list \texttt{Cons~'a'~(Cons~'b'~Nil)} has type \texttt{Vec~(S~(S~Z))~Char}
-because \texttt{Nil~::~Vec~Z~Char} and \texttt{Cons~'a'~Nil~::~Vec~(S~Z)~Char}.
+because \texttt{Nil} has type \texttt{Vec~Z~Char} and \texttt{Cons~'a'~Nil} has
+type \texttt{Vec~(S~Z)~Char}.
 GADTs make possible different data constructors being parametrized over
@@
 function takes as input a \texttt{Vec (S n) a}, that is, a vector with size
-(n+1) for all possible n's, which makes it impossible to pass a list of type
-\texttt{Vec Z a} to \texttt{head}.
+(n+1) for all possible n's, making a call to \texttt{head} on a list of type
+\texttt{Vec Z a} (an empty list) is a compile-time type error.
 
@@
 
-Pattern matching on the \texttt{Nil} constructor would result in an innacessible
-pattern warning because it is impossible to pass a list that would match
-\texttt{Nil} -- it would entail the argument list has size zero; however, it's
-impossible to call \texttt{head} on a list of size zero!
-
-Linear types, which allow for a finer control over the usage of certain resources at the type
-level, are another example of an extension to Haskell's type system
+Pattern matching on the \texttt{Nil} constructor is not needed, despite it being
+indeed a constructor of \texttt{Vec}. The input type just doesn't match the type
+of \texttt{Nil} ($\texttt{S~n} \neq \texttt{Z}$), so the corresponding pattern
+case alternative would be innacessible because the typechecker will not allow
+calling \texttt{head} on \texttt{Nil} (once again, its type, \texttt{Vec~Z~a},
+does not match the input type of \texttt{head}, \texttt{Vec~(S~n)~a}).
+
+In practice, the idea of using more expressive types to enforce invariants at
+compile time, that is illustrated by this simple example, can be taken much
+further, e.g., to implement type-safe artificial neural
+networks\cite{https://blog.jle.im/entry/practical-dependent-types-in-haskell-1.html},
+enforce size compatibility in operations between matrices and vectors\cite{}, to
+implement red-black trees guaranteeing its invariants at compile-time, or to
+implement a material system in a game engine\cite{}.
+% TODO: cite: https://blog.jle.im/entry/practical-dependent-types-in-haskell-1.html
+% TODO: haskell-servant?
+
+Linear types are, similarly, an extension to Haskell's type system that makes it
+even more expressive, by providing a finer control over the usage of certain
+resources at the type level.
 %, besides the aforementioned type-level features
-that makes it even more expressive.
 
@@
 
-Introduction of linear types to Haskell. How do they work, what's the syntax?
-Multiplicities, examples. Monads, multiplicity polymorphism, multiplicity
-polymoprhism applied to monadic bind to define resource-aware sequencing of
-operations. Linear haskell relates systemFC with the linear lambda calculus?
+% Introduction of linear types to Haskell. How do they work, what's the syntax?
+% Multiplicities, examples. Monads, multiplicity polymorphism, multiplicity
+% polymoprhism applied to monadic bind to define resource-aware sequencing of
+% operations. Linear haskell relates systemFC with the linear lambda calculus?
+
+The introduction of linear types to Haskell's type system is originally
+described in Linear Haskell\cite{cite:linearhaskell}. While
+in~\ref{sec:related-work-linear-haskell} we discuss the reasoning and design
+choices behind retrofitting linear types to Haskell, here we focus on
+linear types solely as they exist in the language, and re-work the file
+handle example seen in the previous section along the way.
+% TODO: useless information but the term "Linear Haskell" is sometimes also used
+% to refer to the existence of linear types in Haskell in general.
+
+A linear function ($f :: A \lolli B$) guarantees that if ($f~x$) is consumed
+exactly once, then the argument $x$ is consumed exactly once. The precise
+definition of \emph{consuming a value} depends on the value as follows,
+paraphrasing Linear Haskell~\cite{cite:linearhaskell}:
+
+\begin{itemize}
+    \item To consume a value of atomic base type (like \texttt{Int} or
+        \texttt{Ptr}) exactly once, just evaluate it
+    \item To consume a function value exactly once, apply it to one argument,
+        and consume its result exactly once.
+    \item To consume a value of an algebraic datatype exactly once,
+        pattern-match on it, and consume all its linear components exactly once.
+        For example, a linear pair (equivalent to $\tensor$) is consumed exactly
+        once if pattern-matched on \emph{and} both the first and second element are
+        consumed once.
+\end{itemize}
+
+In Haskell, linear types are introduced through \emph{linearity on the function
+arrow}, rather than through \emph{linearity in the kinds}. In practice, this
+means function types are annotated with a linearity that defines whether a
+function argument must be consumed \emph{exactly once} or whether it can be
+consumed \emph{unrestrictedly} (many times). Additionally, algebraic data types
+constructor consume its arguments linearly by default, but the constructor can
+be specified to receive unrestricted arguments which, in contrast, do not need
+to be consumed exactly once when pattern matched on.
+%
+As an example, consider from the following functions the first, which doesn't
+typecheck because it is a linear function (annotated with the linearity of
+\texttt{1}) that consumes its argument more than once, and the second, which is
+an unrestricted function (annoted with \texttt{Many}) that typechecks because
+its type allows the argument to be consumed unrestrictedly.
+
+\begin{code}
+    f :: a %1 -> (a, a)
+    f x = (x, x)
+\end{code}
+\begin{code}
+    g :: a %Many -> (a, a)
+    g x = (x, x)
+\end{code}
+
+The function annotated with the linearity annotation of \texttt{1} is equivalent
+to the linear function type ($\lolli$) seen in the linear lambda
+calculs~\ref{sec:linear-types}. To encode the multiplicative linear pair
+($\tensor$) we must create a pair data type with two linear components. To
+consume an algebraic data type is to consume all its linear components once, so
+to consume said pair data type we need to consume both its linear components,
+which successfully encodes the multiplicative pair elimination rule ($\tensor
+E$). To construct said pair data type we must provide two linear elements, each
+consuming some required resources to be constructed, which encodes the
+multiplicative pair introduction rule ($\tensor I$).
+
+\begin{code}
+data MultPair a b where
+    MkPair :: a %1 -> b %1 -> MultPair a b
+\end{code}
+
+The linearity annotations are a specialization of the more general so-called
+\emph{multiplicity annotations}. A multiplicity of \texttt{1} entails that the
+function argument must be consumed once, and a function annotated with it
+($\to_1$) is called a linear function (also known as $\lolli$). A function with
+multiplicity of \texttt{Many} ($\to_\omega$) allows the function argument to
+be consumed unrestrictedly. Unrestricted functions are equivalent to the common
+function type and, in fact, the usual function arrow ($\to$) implicitly has
+multiplicity \texttt{Many}.  Multiplicities also smoothly allow for
+\emph{multiplicity polymorphism}, which we describe below.
 
@@
 
-\chapter{Related Work}
+\section{Related Work}
 
@@
 
-\section{System FC}
+\subsection{System FC}
 
@@
 
-\section{Linear Haskell}
+\subsection{Linear Haskell\label{sec:related-work-linear-haskell}}
 
@@
 
-\section{OutsideIn(X)\label{related-work-gadts}}
+Haskell, contrary to most programming languages with linear types or
+linearity-based types (such as ownership types), has existed for 31 years of its
+life \emph{without} linear types. As such, the introduction of linear types to
+Haskell comes with added challenges that can't exist in linearly-typed languages
+that were designed with linear types from the start:
+
+\begin{itemize}
+    \item Backwards compatibility. The addition of linear types shouldn't break
+        all existing Haskell code.
+    \item Code re-usability. The linearly-typed part of Haskell's ecosystem and
+        its non-linearly-typed counterpart should fit in together and it must be
+        possible to define functions readily usable by both sides
+        simultaneously.
+\end{itemize}
+
+\begin{itemize}
+    \item Forwards compatibility -- Haskell, despite being an
+        industrial-strength language, is also a petri-dish for experimentation
+        and innovation in the field of programming languages. Therefore, Linear
+        Haskell takes care to accomodate possible future features, in
+        particular, its design is forward compatible with affine and dependent
+        types.
+\end{itemize}
+
+\subsection{OutsideIn(X)\label{related-work-gadts}}
 
@@
 
-\section{Rust}
+\subsection{Rust}
 
@@
 
-\section{Introduction}
-
-Since the publication of Linear Haskell~\cite{linearhaskell} and its release in GHC 9.0,
+\section{Summary}
+
+Since the publication of Linear Haskell~\cite{cite:linearhaskell} and its release in GHC 9.0,
 Haskell's type system supports linearity annotations in functions -- bringing

### 2023-01-23 laboratory/Files.hs

@@ -1,0 +1,19 @@
+{-# LANGUAGE LinearTypes #-}
+{-# LANGUAGE NoImplicitPrelude #-}
+module Files where
+
+-- import Prelude hiding (Functor, Applicative, Monad)
+
+
+-- class Functor f where
+--   fmap :: (a %1 -> b) -> f a %1 -> f b
+
+-- class Functor f => Applicative f where
+--   (<*>) :: f (a %1 -> b) %1 -> f a %1 -> f b
+--   pure :: a -> f a
+
+-- class Applicative m => Monad m where
+--   (>>=) :: m a %1 -> (a %1 -> m b) %1 -> m b
+
+import Prelude.Linear
+

### 2023-01-23 laboratory/app/Files.hs

@@ -1,0 +1,20 @@
+{-# LANGUAGE LinearTypes #-}
+{-# LANGUAGE NoImplicitPrelude #-}
+module Files where
+
+import Prelude.Linear
+
+-- import Prelude hiding (Functor, Applicative, Monad)
+
+
+-- class Functor f where
+--   fmap :: (a %1 -> b) -> f a %1 -> f b
+
+-- class Functor f => Applicative f where
+--   (<*>) :: f (a %1 -> b) %1 -> f a %1 -> f b
+--   pure :: a -> f a
+
+-- class Applicative m => Monad m where
+--   (>>=) :: m a %1 -> (a %1 -> m b) %1 -> m b
+
+

### 2023-01-23 laboratory/app/Main.hs

@@ -1,0 +1,23 @@
+{-# LANGUAGE LinearTypes #-}
+{-# LANGUAGE QualifiedDo #-}
+{-# LANGUAGE DataKinds #-}
+{-# LANGUAGE BlockArguments #-}
+{-# LANGUAGE NoImplicitPrelude #-}
+{-# LANGUAGE RebindableSyntax #-}
+module Main where
+
+import GHC.Exts (Multiplicity(..))
+
+import Prelude.Linear
+import qualified System.IO.Linear as Linear
+import System.IO
+import Control.Functor.Linear as Linear
+
+-- f1 :: a %1 -> (a, a)
+-- f1 x = (x,x)
+
+
+
+f2 :: a %Many -> (a, a)
+f2 x = (x,x)
+

### 2023-01-24 report.tex
Small improvements

@@
 high-performance language interoperability~\cite{}, guarantee that quantum
-entangled states are not duplicated~\cite{}, handle mutable state safely~\cite{},
-something on session types guarantees~\cite{}, and much more.
-
-% TODO: Also include bit about how it makes the basis of other type systems such
-% as rusts owneship. How ?
+entangled states are not duplicated~\cite{}, something on session types
+guarantees~\cite{}, and much more.
+% handle mutable state safely~\cite{}
 
@@
 % TODO:
-example program \emph{would not typecheck}! Some conclusion on linearity
-goodness. In section~\ref{linear-types} we describe the fundamentals of linear
+example program \emph{would not typecheck}! \todo{Some conclusion on linearity
+goodness.} In section~\ref{linear-types} we describe the fundamentals of linear
 types in more detail with a formal definition and many examples.
@@
 Linearity in Haskell stands out % from linearity in Rust and Idris2
-for the two following distinct reasons:
+two distinct reasons:
 
@@
         in a transformation!  We discuss Core in detail in
-        section~\ref{core-section}.
+        section~\ref{core-section}.\todo{Mention systemFC}
         % TODO: \item values in rust are linear by default while non-linear is
@@
 To understand how linear types are defined and used in practice, we present two
-examples of anonymous functions that receive as an argument a handle to work
-with and are responsible for closing it, we try to understand how the examples
-could be disregarded as incorrect, and work our way up to linear types from
-them. The first function ignores the received file handle and returns $\star$
-(read unit), which is equivalent to C's \texttt{void}.
-% TODO: Better code formatting
+examples of anonymous functions that receive a handle to work with (that must be
+closed before returning), we explore how the examples could be disregarded as
+incorrect, and work our way up to linear types from them. The first function
+ignores the received file handle and returns $\star$ (read unit), which is
+equivalent to C's \texttt{void}.
+%
 \begin{mathpar}
-    \lambda h.~\textrm{return}~\star;
+    \lambda h.~\mathsf{return}~\star;
     \and
-    \lambda h.~\textrm{close}~h;~\textrm{close}~h;
+    \lambda h.~\mathsf{close}~h;~\mathsf{close}~h;
 \end{mathpar}
@@
 Ignoring the file handle which should have been closed by the function makes the
-first function incorrect. Similarly, the second function receives the file handle and
-closes it twice, which is incorrect not because it falls short of the
+first function incorrect. Similarly, the second function receives the file
+handle and closes it twice, which is incorrect not because it falls short of the
 specification, but rather because the program will crash while executing it.
 %
-Additionally, both functions share the same type, $\texttt{Handle} \to \star$, i.e.
-a function that takes a \texttt{Handle} and returns $\star$. The second function also
-shares this type because \texttt{close} has type $\texttt{Handle} \to \star$.
-%
-Under a simple type system such as C's both functions are type correct (the
+Additionally, both functions share the same type, $\texttt{Handle} \to \star$,
+i.e.  a function that takes a \texttt{Handle} and returns $\star$. The second
+function also shares this type because \texttt{close} has type $\texttt{Handle}
+\to \star$.
+%
+Under a simple type system such as C's, both functions are type correct (the
 compiler will happily succeed), but both have erroneous behaviours. The first
-forgets to close the handle and the second closes it twice. Our goal is now to
-reach a type system that rejects these two programs.
+forgets to close the handle and the second closes it twice. Our goal is to reach
+a type system that rejects these two programs.
 
@@
 In order to further give well defined semantics to a linear type system, we
-present a linearly typed lambda calculus\todo{citar o girard
-  provavelmente ou o Barber (\url{https://www.lfcs.inf.ed.ac.uk/reports/96/ECS-LFCS-96-347/})}, a very simple language with linear
-types, by defining what are syntatically valid programs through the grammar
-in Fig.~\ref{fig:llcgrammar} and what programs are well typed through the typing
-rules in Fig.~\ref{fig:llcrules}. The language features functions and function application ($\lolli$), two
-flavours of pairs, additive ($\with$) and multiplicative ($\tensor$), a
-disjunction operator ($\oplus$)\todo{sum types} and the $\bang$
-modality operator\todo{que serve para}. A typing
-judgement for the linearly typed lambda calculus has the form
+present a linearly typed lambda calculus~\cite{}\todo{citar o girard
+provavelmente ou o Barber
+(\url{https://www.lfcs.inf.ed.ac.uk/reports/96/ECS-LFCS-96-347/})}, a very
+simple language with linear types, by defining what are syntatically valid
+programs through the grammar in Fig.~\ref{fig:llcgrammar} and what programs are
+well typed through the typing rules in Fig.~\ref{fig:llcrules}. The language
+features functions and function application ($\lolli$), two flavours of pairs,
+additive ($\with$) and multiplicative ($\tensor$), a disjunction operator
+($\oplus$) to construct sum types, and the $\bang$ modality operator which
+constructs an unrestricted type from a linear one, allowing values inhabiting
+$\bang A$ to be consumed unrestrictedly. A typing judgement for the linearly typed
+lambda calculus has the form
 %
@@
 %
-To utilize an unrestricted value $M$, we must
-bind it to $u$ with $\llet{\bang u = M}{N}$ and can then use it in $N$
-unrestrictedly, because $u$ extends the unrestricted context rather than the
-linear context as we have seen thus far.
+To utilize an unrestricted value $M$, we must bind it to $u$ with $\llet{\bang u
+= M}{N}$ which can then be used in $N$ unrestrictedly, because $u$ extends the
+unrestricted context rather than the linear context as we have seen thus far.
 
@@
 Haskell, a programming language more featureful than the linearly typed lambda
-calculus. We will see that, the theoretical principles underlying the linear
+calculus. We will see that the theoretical principles underlying the linear
 lambda calculus and linear Haskell are the same, and by studying them in this
@@
 
-Haskell is a large feature-rich language but its relatively small core is based on a typed lambda
-calculus. As such, there exist no statements and computation is done simply
-through the evaluation of functions. Besides functions, one can define types and
-their constructors and pattern match on said constructors. Function application
-is denoted by an empty space\todo{juxtaposition} (\texttt{f~a} means \texttt{f} applied to
-\texttt{a}) and pattern matching is done with the \texttt{case} keyword followed
-by the enumerated alternatives. All variable names start with lower case and
-types start with upper case (excluding type variables). To make explicit the
-type of an expression, the \texttt{::} operator is used (e.g.
-$\texttt{f~::~Int}\to \texttt{Bool}$ is read \texttt{f} \emph{has type} function
-from \texttt{Int} to \texttt{Bool}).
+Haskell is a large feature-rich language but its relatively small core is based
+on a typed lambda calculus. As such, there exist no statements and computation
+is done simply through the evaluation of functions. Besides functions, one can
+define types and their constructors and pattern match on said constructors.
+Function application is denoted by the juxtaposition of the function expression
+and its arguments, which often means empty space between terms (\texttt{f~a}
+means \texttt{f} applied to \texttt{a}). Pattern matching is done with the
+\texttt{case} keyword followed by the enumerated alternatives. All variable
+names start with lower case and types start with upper case (excluding type
+variables). To make explicit the type of an expression, the \texttt{::} operator
+is used (e.g.  $\texttt{f~::~Int}\to \texttt{Bool}$ is read \texttt{f} \emph{has
+type} function from \texttt{Int} to \texttt{Bool}).
 
@@
 
-    \item Local equality constraints and existential types by using GADTs, we
-        explain ahead in more detail. A design for first class existential types
-        with bi-directional type inference in Haskell has been published
+    \item Local equality constraints and existential types by using GADTs, which
+        we explain ahead in more detail. A design for first class existential
+        types with bi-directional type inference in Haskell has been published
         in~\cite{10.1145/3473569}, despite not being yet implemented in GHC.
@@
 \textbf{GADTs}~\cite{10.1145/1159803.1159811,,,} are an advanced Haskell feature
-which allows users to define data types as they would common algebraic
-data types, with the added abilitiy to give explicit type signatures for the
-data constructors where the result type may differ in the type parameters across
-branches.
-\todo{nao ha branches, é nos varios construtores. This allows for
-  additional type invariants to be represented with GADTs through
-  their type parameters, which restricts the use of specific
-  constructors and their subsequent deconstruction through pattern
-  matching}
+that allows users to define data types as they would common algebraic data
+types, with the added abilitiy to give explicit type signatures for the data
+constructors where the result type may differ in the type parameters (e.g., we
+might have two constructors of the same data type \texttt{T~a} return values of
+type \texttt{T~Bool} and \texttt{T~Int}).
+%
+This allows for additional type invariants to be represented with GADTs through
+their type parameters, which restricts the use of specific constructors and
+their subsequent deconstruction through pattern matching.
+%
 % TODO: Rewrite this bit below
@@
 (n+1) for all possible n's, making a call to \texttt{head} on a list of type
-\texttt{Vec Z a} (an empty list) is a compile-time type error.
+\texttt{Vec Z a} (an empty list) a compile-time type error.
 
@@
 linear types solely as they exist in the language, and re-work the file
-handle example seen in the previous section along the way.
+handle example seen in the previous section to make sure it doesn't typecheck.
 % TODO: useless information but the term "Linear Haskell" is sometimes also used
@@
 function argument must be consumed \emph{exactly once} or whether it can be
-consumed \emph{unrestrictedly} (many times). Additionally, algebraic data types
-constructor consume its arguments linearly by default, but the constructor can
-be specified to receive unrestricted arguments which, in contrast, do not need
-to be consumed exactly once when pattern matched on.
+consumed \emph{unrestrictedly} (many times).
 %
 As an example, consider from the following functions the first, which doesn't
-typecheck because it is a linear function (annotated with the linearity of
+typecheck because it is a linear function (annotated with the linearity
 \texttt{1}) that consumes its argument more than once, and the second, which is
@@
 its type allows the argument to be consumed unrestrictedly.
-
+%
 \begin{code}
@@
 \end{code}
-
+%
 The function annotated with the linearity annotation of \texttt{1} is equivalent
 to the linear function type ($\lolli$) seen in the linear lambda
-calculs~\ref{sec:linear-types}. To encode the multiplicative linear pair
-($\tensor$) we must create a pair data type with two linear components. To
-consume an algebraic data type is to consume all its linear components once, so
-to consume said pair data type we need to consume both its linear components,
-which successfully encodes the multiplicative pair elimination rule ($\tensor
-E$). To construct said pair data type we must provide two linear elements, each
-consuming some required resources to be constructed, which encodes the
-multiplicative pair introduction rule ($\tensor I$).
+calculs~\ref{sec:linear-types}. 
+%
+Additionally, algebraic data type constructors can specify whether their
+arguments are linear or unrestricted, requiring that, when pattern matched on,
+linear arguments be consumed once while unrestricted arguments need not be
+consumed exactly once.
+%Similarly, when constructing such a constructor, using
+% a linear value as a linear argument of the constructor consumes it once, while
+% using it as an unrestricted argument consumes it unrestrictedly.
+%
+To encode the multiplicative linear pair ($\tensor$) we must create a pair data
+type with two linear components. To consume an algebraic data type is to consume
+all its linear components once, so to consume said pair data type we need to
+consume both its linear components -- successfully encoding the
+multiplicative pair elimination rule ($\tensor E$). To construct said pair data
+type we must provide two linear elements, each consuming some required resources
+to be constructed, thus encoding the multiplicative pair introduction rule
+($\tensor I$). As such, we define \texttt{MultPair} as an algebraic data type
+whose constructor uses a linear arrow for each of the arguments\footnote{By
+default, constructors defined without GADT syntax have linear arguments. We
+could have written \texttt{data MultPair a b = MkPair a b} to the same effect.}
 
@@
 
-The linearity annotations are a specialization of the more general so-called
+The linearity annotations are just a specialization of the more general so-called
 \emph{multiplicity annotations}. A multiplicity of \texttt{1} entails that the
@@
 
+% TODO: Is this enough about multiplicities?
+
 \begin{itemize}
-    \item Linear Haskell definition
-    \item Consuming values precisely
-    \item Multiplicities
+    % \item Linear Haskell definition
+    % \item Consuming values precisely
+    % \item Multiplicities
     \item Multiplicity polymorphism

### 2023-01-24 report.tex
Core Haskell

@@
 
-Haskel's intermediate representation that is statically typed
+Haskell is a big and expressive language with a lot of syntatic constructs and
+many, many features. However, the whole of Haskell can be desugared down to a
+very simple and elegant, explicitly typed, intermediate representation called
+\textbf{Core}. To illustrate the difference in complexity, in GHC, the source
+language (the big and complex Haskell) is defined through dozens of datatypes
+and hundreds of constructors, while Core is defined in 3 types (expressions,
+types, and coercions) and 15 constructors~\cite{}.
+%
+Core is unique as a strongly-typed intermediate language, with GHC as the only
+production-ready compiler having such a typed intermediate representation.
+
+Core allows us to reason about the entirety of Haskell in a much smaller and
+elegant language -- analysis, optimizating transformations, and code generation
+are all done on Core, not Haskell.
+%
+Of equal importance is Core being explicitly typed, which makes Core an internal
+consistency tool that validates desugaring and optimization passes. Desugaring
+and optimizing transformations must produce well-typed Core: having the Core
+typechecker guarantee that Core programs are well-typed provides a verification
+layer of the correctness of the optimizing transformations, of the desugarer,
+and of their implementations.
+%
+Finally, the expressive but simple type-system in Core serves as a sanity-check
+for all the extensions to the source language type system -- if we can desugar a
+type system feature into Core then the feature must be sound by reducibility.
+
+Core is typed very differently from Haskell because Core is explicitly typed and
+its type system is based on the $System_{FC}$~\cite{systemfc} type system, while
+Haskell is implicitly typed and its type system is based on the constraint-based
+type inference system $OutsideIn(X)$~\cite{outsideinx}. Therefore, typechecking
+Core is simple and fast and no inference must be done and, on the contrary,
+Haskell's typechecker is much more complicated and requires ad-hoc restrictions
+to make type inference feasible~\cite{}.
+%
+Haskell is typechecked in addition to Core to do type inference and to provide
+meaningful type errors. If Haskell wasn't typechecked as it is, everything (e.g.
+all binders) would have to be explicitly typed and all error messages would
+refer to the intermediate language rather than the written program.
+
+\todo[inline]{Any feature added to Haskell is only syntatic sugar}
 

### 2023-01-25 report.tex
optimizations subsections list
Tweaks

--- 2023-01-24 report.tex
@@
 
-Linearity in Haskell stands out % from linearity in Rust and Idris2
+Linearity in Haskell stands out from linearity in Rust and Idris2
 two distinct reasons:
@@
 
-% % TODO: Revise the story above instead of dumping so much information
-
 % \mypara{Linear Types}
@@
 
-Haskel's intermediate representation that is statically typed
+Haskell is a big and expressive language with a lot of syntatic constructs and
+many, many features. However, the whole of Haskell can be desugared down to a
+very simple and elegant, explicitly typed, intermediate representation called
+\textbf{Core}. To illustrate the difference in complexity, in GHC, the source
+language (the big and complex Haskell) is defined through dozens of datatypes
+and hundreds of constructors, while Core is defined in 3 types (expressions,
+types, and coercions) and 15 constructors~\cite{}.
+%
+Core is unique as a strongly-typed, lazy, purely functional intermediate
+language akin to a polymorphic lambda calculus, with GHC as the only
+production-ready compiler having such a typed intermediate representation.
+
+Core allows us to reason about the entirety of Haskell in a much smaller and
+elegant functional language -- analysis, optimizating transformations, and code
+generation are all done on Core, not Haskell.
+%
+Of equal importance is Core being explicitly typed, which makes Core an internal
+consistency tool that validates desugaring and optimization passes. Desugaring
+and optimizing transformations must produce well-typed Core, and the Core
+typechecker can guarantee that Core programs are well-typed, thus providing a
+verification layer of the correctness of the optimizing transformations, of the
+desugarer, and of their implementations.
+%
+Finally, the expressive but simple type-system in Core serves as a sanity-check
+for all the extensions to the source language type system -- if we can desugar a
+type system feature into Core then the feature must be sound by reducibility.
+Effectively, any feature added to Haskell is only syntatic sugar if it can be
+desugared to Core.
+
+Core is typed very differently from Haskell because Core is explicitly typed and
+its type system is based on the $System~F_C$~\cite{systemfc} type system, while
+Haskell is implicitly typed and its type system is based on the constraint-based
+type inference system $OutsideIn(X)$~\cite{outsideinx}. Therefore, typechecking
+Core is simple and fast and no inference must be done and, on the contrary,
+Haskell's typechecker is much more complicated and requires ad-hoc restrictions
+to make type inference feasible~\cite{}.
+%
+Haskell is typechecked in addition to Core to (i) do type inference and (ii)
+provide meaningful type errors. If Haskell wasn't typechecked as it is and
+instead we only typechecked Core, everything (e.g.  all binders) would have to
+be explicitly typed and all error messages would refer to the intermediate
+language rather than the written program, which is known to be undesirable.
 
 \begin{itemize}
-    \item What is Core (IR) e as particularidades
+    % \item What is Core (IR) e as particularidades
     \item Em cima do SystemFC
@@
 % \end{itemize}
+
+\section{Core-To-Core Optimizations}
+
+% like inlining, its inverse (CSE), beta-reduction, lambda-lifting, eta-reduction/eta-expansion, binder-swap, case-of-case, case-of-know-constructor, float-out, float-in, worker/wrapper split (this one is big, in comparison), etc…)
+
+\subsection{Inlining}
+
+\subsection{Common Sub-expression elimination}
+
+\subsection{Beta-reduction}
+\subsection{Lambda lifting}
+\subsection{Eta-reduction/Eta-expansion}
+
+\subsection{Binder-swap}
+
+\subsection{Case-of-case}
+\subsection{Case-of-known-constructor}
+\subsection{Float-out and Float-in}
+\subsection{Worker/wrapper split}
 
@@
 
+% TODO: - In the letrec case: congratulations on finding an inference algorithm
+% (though I will want to see the proof some day, I don't see why it works yet).
+% You should clarify that, during linting, you will have a usage environment
+% annotation and won't need to run the inference algorithm. This algorithm is
+% only needed when you first create a letrec.
+
 % TODO:Use glossary

### 2023-02-16 presentation/preparation/slides.tex
slides(preparation): Initial context

\documentclass[14pt,aspectratio=169]{beamer}
\usetheme{Copenhagen}
\usecolortheme{beaver}
\setbeamertemplate{navigation symbols}{}
\setbeamercovered{transparent}

\usepackage[utf8]{inputenc}

\title{Type-checking Linearity in Haskell's\\ Core Intermediate Language}
\author{Rodrigo Mesquita and Bernardo Toninho}
\institute{\small NOVA School of Science and Technology}
\date{\small February 2023}

\begin{document}
\begin{frame}
    \titlepage
    \begin{figure}[htpb]
        \begin{center}
            % \vspace*{-0.5cm}
            \includegraphics[width=0.4\linewidth]{../logo_nova.png}
        \end{center}
    \end{figure}
\end{frame}

\begin{frame}{Context}

\begin{itemize}
    \item<1-> Statically typed programs eliminate errors at compile time
    \item<2-> Linear types are expressive types that can enforce invariants wrt
        resource usage
    \item<3-> Haskell is a lazy, purely functional programming language
    \item<4-> Linear types were recently introduced in GHC Haskell, \emph{31 years} after Haskell's inception.
                % Which came with new challenges such as backwards compatibility, future
                % proofing and code re-use
\end{itemize}

\end{frame}

\begin{frame}{Motivation}
\begin{itemize}
    \item<1-> However, linear types were also introduced in \textbf<1>{Core}, a
        explicitly typed language that serves as Haskell's main intermediate
        representation
        % Core is explicitly typed
        % Core is very important because it serves as an internal consistency
        % tool and validation layer of the implementation
    \item<2-> Core programs undergo a myriad of (Core-to-Core)
        optimizing transformations
        % Which might introduce lazy let bindings (through e.g. beta reduction),
        % inline definitions, move bindings around (let-floating
        % transformations), 
    \item<3-> \textbf<3>{Core's linear type-checker doesn't accept linear
        programs} after optimizing transformations are applied
        % Optimizing transformations destroy linearity as the type-checker sees
        % it. However, we believe optimizing transformations *don't* destroy
        % linearity, it's just a limitation of the type system
    \item<4-> \textbf<4>{Core's linear type-checker is disabled} because
        otherwise disabling optimizations is far worse
        % it doesn't accept transformed Core programs
        % GHC's Core type-checker is disabled because it doesn't accept Core
        % programs that undergo optimizing transformations in Core
        % and otherwise disabling optimizing transformations is not feasible.
\end{itemize}
\end{frame}

\end{document}

### 2023-02-16 presentation/preparation/slides.tex
Simplify slides

@@
 \setbeamertemplate{navigation symbols}{}
-\setbeamercovered{transparent}
+% \setbeamercovered{transparent}
 
@@
 
+\section{Introduction}
+
 \begin{frame}{Context}
@@
     \item<1-> Statically typed programs eliminate errors at compile time
-    \item<2-> Linear types are expressive types that can enforce invariants wrt
-        resource usage
+    \item<2-> Linear types can enforce resource usage invariants
+        % Are expressive types
     \item<3-> Haskell is a lazy, purely functional programming language
-    \item<4-> Linear types were recently introduced in GHC Haskell, \emph{31 years} after Haskell's inception.
+    \item<4-> Linear types were introduced in Haskell, 31 \emph<4>{years} after Haskell's inception.
+                % recently introduced
                 % Which came with new challenges such as backwards compatibility, future
@@
 \begin{itemize}
-    \item<1-> However, linear types were also introduced in \textbf<1>{Core}, a
-        explicitly typed language that serves as Haskell's main intermediate
-        representation
+    \item<1-> However, linear types were also introduced in \textbf<1>{Core}
+        % a typed language that serves as Haskell's main intermediate representation
         % Core is explicitly typed
@@
         % tool and validation layer of the implementation
-    \item<2-> Core programs undergo a myriad of (Core-to-Core)
-        optimizing transformations
+    \item<2-> Core programs undergo a myriad of optimizing transformations
+        % These (core-to-core) transformations follow a modular approach, where
+        % each transformation produces Core that is further used as input to
+        % other transformations
         % Which might introduce lazy let bindings (through e.g. beta reduction),
@@
 
+\begin{frame}{Objectives}
+\begin{itemize}
+    \item<1-> Extend Core's type-system to accommodate more linear Core programs
+        % Big distinction between syntatically valid linearity and semantically
+        % valid. Transformations result in many semantically valid but
+        % syntatically invalid programs, and we need to extend Core to
+        % account for this semantic linearity.
+    \item<2-> Validate optimizing transformations with the improved Core linear type-system
+    \item<3-> Implement the extensions to Core in GHC, the leading Haskell
+        compiler
+\end{itemize}
+\end{frame}
+
+\section{Background}
+
+\begin{frame}{}
+\end{frame}
+
 \end{document}

### 2023-05-24 laboratory/LetRecs.hs
Advance preservation proof of Case

@@ -1,0 +1,20 @@
+
+module LetRecs where
+
+f y =
+  let g x = case x of
+                True -> h x
+                False -> True
+      h x = case x of
+              True -> False
+              False -> g x
+      l x = z x
+      {-# NOINLINE g #-}
+      {-# NOINLINE h #-}
+      {-# NOINLINE l #-}
+    in g y || l y
+
+
+z :: Bool -> Bool
+z = undefined
+{-# NOINLINE z #-}

### 2023-05-24 laboratory/join_points_reclet_1/NameCache.hs

@@ -1,0 +1,12 @@
+module NameCache where
+
+data Name = Name
+data NameCache = NameCache !Int !Name
+
+extendOrigNameCache :: Name -> Name -> Name
+extendOrigNameCache _ _ = Name
+
+initNameCache :: Int -> [Name] -> NameCache
+initNameCache us names
+  = NameCache us (foldl extendOrigNameCache Name names)
+

### 2023-05-24 laboratory/reuniao/A.hs

@@ -1,0 +1,47 @@
+{-# LANGUAGE LinearTypes, UnicodeSyntax, MagicHash #-}
+module A where
+
+import B qualified as Unsafe
+import GHC.Exts
+
+data Alg a = K1 a
+           | K2 a a
+           | K3
+
+f :: FUN 'One Int (IO ())
+f x = case x of
+        -- I# y -> case consumeInt y of
+        --           () -> pure ()
+        _ -> case consumeInt' x of
+               () -> pure ()
+
+-- Não é igual aos outros?
+h :: FUN 'One Int Int
+h x = case x of
+        _ -> x
+-- VS
+g :: FUN 'One Int Int
+g x = case x of
+        y -> y
+-- VS
+-- u :: FUN 'One Int Int
+-- u x = case x of z
+--         _ -> z
+
+v :: Alg Int ⊸ Alg Int
+v x = case x of
+        y -> y
+
+kk :: Alg Int ⊸ Alg Int
+kk x = case v x of
+         y -> y
+
+-- Usage env do case binder acaba por ser o contexto linear de <expr>
+-- Duplicado de forma parecida ao resto das coisas...
+
+consumeInt :: Int# ⊸ ()
+consumeInt = Unsafe.toLinear (\x -> ())
+consumeInt' :: Int ⊸ ()
+consumeInt' = Unsafe.toLinear (\x -> ())
+
+

### 2023-05-24 laboratory/reuniao/B.hs

@@ -1,0 +1,197 @@
+{-# LANGUAGE AllowAmbiguousTypes #-}
+{-# LANGUAGE DataKinds #-}
+{-# LANGUAGE FlexibleInstances #-}
+{-# LANGUAGE GADTs #-}
+{-# LANGUAGE LinearTypes #-}
+{-# LANGUAGE MultiParamTypeClasses #-}
+{-# LANGUAGE PolyKinds #-}
+{-# LANGUAGE ScopedTypeVariables #-}
+{-# LANGUAGE StandaloneKindSignatures #-}
+{-# LANGUAGE TypeApplications #-}
+{-# LANGUAGE TypeFamilies #-}
+{-# LANGUAGE TypeOperators #-}
+{-# LANGUAGE UndecidableInstances #-}
+
+-- | Unsafe coercions for linearly typed code.
+--
+-- Use this module to coerce non-linear functions to be linear or values
+-- bound linearly to be another type. /All/ functions in this module are
+-- unsafe.
+--
+-- Hence:
+--
+-- * Import this module qualifed as Unsafe.
+-- * Do not use this unless you have to. Specifically, if you can write a
+-- linear function @f :: A %1-> B@, do not write a non-linear version and coerce
+-- it.
+module B
+  ( -- * Unsafe Coercions
+    coerce,
+    toLinear,
+    toLinear2,
+    toLinear3,
+    toLinearN,
+    ToLinearN (..),
+  )
+where
+
+import Data.Kind (Constraint)
+import Data.Type.Equality (type (~~))
+import GHC.Exts (RuntimeRep (..), TYPE)
+import GHC.TypeNats
+import Unsafe.Coerce (UnsafeEquality (..), unsafeEqualityProof)
+
+-- | Linearly typed @unsafeCoerce@
+coerce :: forall a b. a %1 -> b
+coerce a =
+  case unsafeEqualityProof @a @b of
+    UnsafeRefl -> a
+{-# INLINE coerce #-}
+
+-- | Converts an unrestricted function into a linear function
+toLinear ::
+  forall
+    (r1 :: RuntimeRep)
+    (r2 :: RuntimeRep)
+    (a :: TYPE r1)
+    (b :: TYPE r2)
+    p
+    x.
+  (a %p -> b) %1 ->
+  (a %x -> b)
+toLinear f = case unsafeEqualityProof @p @x of
+  UnsafeRefl -> f
+
+-- | Like 'toLinear' but for two-argument functions
+toLinear2 ::
+  forall
[…]

### 2023-05-24 laboratory/reuniao/X.hs

@@ -1,0 +1,30 @@
+-- P :: a ⊸ b ⊸ c ⊸ P a b c
+case x of z:p
+  P a b c -> ___ z:(a,b,c) && a:1 b:1 c:1
+  P _ b _ -> ___ z:p && b:1 -- Invalid... does it happen in Core w/ optimisations?...
+  _       -> ___ z:p
+
+P2 :: a -> b ⊸ c -> P a b c
+case x of z:p
+  P2 a b c -> ___ z:(b) && a:Many b:1 c:Many
+  P2 _ b _ -> ___ z:(b) && b:1 -- Invalid... does it happen in Core w/ optimisations?...
+  _        -> ___ z:p
+
+
+-- Our bet
+P3 :: a ⊸ b ⊸ c ⊸ P3 a b c
+case <expr> of z:{...}
+  P3 a b c -> ___ z:(a,b,c) && a:1 b:1 c:1
+  P3 _ b _ -> ___ -- won't happen in practice
+
+
+
+---------------
+., x:1 |- x : σ
+
+\x:1 ->
+  case x of z
+    _  -> ___ z:should be 1
+
+
+

### 2023-07-12 prototype/app/Syntax.hs
Add let and letrec typechecking

@@ -7,8 +7,8 @@ module Syntax where
 import Data.Functor.Foldable.TH (makeBaseFunctor)
 import Data.Map (Map)
 import qualified Data.Map as M
--- import Data.IntMap (IntMap)
--- import qualified Data.IntMap as IM
+import Data.Set (Set)
+import qualified Data.Set as S
 
 {-
 Notes
@@ -69,7 +69,8 @@ data Mult = Many       -- ω
 pattern MV :: Name -> Mult
 pattern MV name = MV' (MkMultVar name)
 
-type UsageEnv = Map Name Mult
+-- | A UsageEnv (Δ) is a set of variables that need to be used linearly
+type UsageEnv = Set Name
 
 data IdBinding = LambdaBound Mult
                | DeltaBound UsageEnv -- ^ Let and case bound variables. Case variables do have many usage environments, but in practice (when they occur in a context), they have just one usage environment
@@ -166,3 +167,11 @@ idBad :: CoreExpr
 idBad =
  Lam (Id (Datatype "A" [MV "p"]) (LambdaBound (MV "p")) "x") $
    Var (Id (Datatype "A" [MV "p"]) (LambdaBound (MV "p")) "y")
+
+varUE :: Var -> Maybe UsageEnv
+varUE (Id _ (DeltaBound ue) _) = Just ue
+varUE _ = Nothing
+
+varId :: Var -> Id
+varId (Id' i) = i
+varId _ = error "varId: Not an Id"

### 2023-07-12 prototype/app/Typechecker.hs

@@ -1,3 +1,4 @@
+{-# LANGUAGE ViewPatterns #-}
 {-# LANGUAGE BlockArguments #-}
 {-# LANGUAGE OverloadedRecordDot #-}
 {-# LANGUAGE DuplicateRecordFields #-}
@@ -7,17 +8,20 @@ module Typechecker where
 
 import Data.Maybe
 import qualified Data.Map as M
+import Data.List.Unicode ((∖))
 import Data.Functor.Foldable
 import Control.Monad.State
 import Control.Monad.Except
 import Control.Monad
 import Syntax
 import Control.Applicative (empty)
+import Data.Set (Set)
+import qualified Data.Set as S
 
 {-
 Notes:
 
-* For now we don't do good error reporting
+* For now we don't do good error reporting (perhaps we could use diagnose?!)
 * We use Input/Output context to track linear variables without context splitting
 * I'm missing the rule for Var_p
 
@@ -30,11 +34,12 @@ type Check = StateT Ctx Maybe
 runClosedCheck :: Check a -> Maybe a
 runClosedCheck check = do
   (a,ctx) <- runStateT check mempty
-  when (not $ null $ M.elems $ M.filter (\case Id _ (LambdaBound One) _ -> True; _ -> False) ctx) (fail "Linear variables escaped")
+  unless (null $ projectCtxLinear ctx) (fail "Linear variables escaped")
   return a
 
-projectLinear :: UsageEnv -> [Name]
-projectLinear = M.keys . M.filter (== One)
+-- TODO: Should project linear also return x{:}_p ?
+projectCtxLinear :: Ctx -> [Name]
+projectCtxLinear = M.keys . M.filter (\case Id _ (LambdaBound One) _ -> True; _ -> False)
 
 -- | Use a variable from the environment.
 -- If it is a linearly-bound variable it will be consumed (and no longer available in the context)
@@ -52,8 +57,7 @@ use name = do
           -- they were linear, bc they might be instantiated to 'One' later on.
           modify (M.delete name)
         DeltaBound ue -> do
-          _ <- mapM use (projectLinear ue) -- all elements should already have mult == One.
-          pure ()
+          mapM_ use (S.toList ue) -- all elements should already have mult == One.
     MultVar' _ ->
       pure ()
   return var
@@ -71,12 +75,22 @@ extend var check = do
       LambdaBound (MV' _) -> do
         True <- wasConsumed name
         return a
[…]

### 2023-07-14 prototype/src/Linear/Core/Check.hs
Structure, testsuite

@@ -4,7 +4,7 @@
 {-# LANGUAGE DuplicateRecordFields #-}
 {-# LANGUAGE LambdaCase #-}
 {-# OPTIONS_GHC -Wno-name-shadowing #-}
-module Typechecker where
+module Linear.Core.Check where
 
 import Data.Maybe
 import qualified Data.Map as M
@@ -13,10 +13,11 @@ import Data.Functor.Foldable
 import Control.Monad.State
 import Control.Monad.Except
 import Control.Monad
-import Syntax
 import Control.Applicative (empty)
 import Data.Set (Set)
 import qualified Data.Set as S
+
+import Linear.Core.Syntax
 
 {-
 Notes:

### 2023-07-14 prototype/src/Linear/Core/Syntax.hs

@@ -2,7 +2,7 @@
 {-# LANGUAGE PatternSynonyms #-}
 {-# LANGUAGE DeriveFunctor, DeriveFoldable,
    DeriveTraversable, TemplateHaskell, TypeFamilies #-}
-module Syntax where
+module Linear.Core.Syntax where
 
 import Data.Functor.Foldable.TH (makeBaseFunctor)
 import Data.Map (Map)
@@ -144,29 +144,6 @@ type CoreExpr = Expr Var
 
 erase :: CoreTerm -> CoreExpr
 erase (Term _ e _) = e
-
--- Let's try writing a simple program like id @A
--- id :: ∀ p. (A p) ->p (A p)
--- id = /\p. \x:_pA -> x:A
-idProg :: CoreTerm
-idProg
-  = Term
-    (M.singleton "MkA" (Id (Scheme "kp" (Datatype "A" [MV "p"])) (DeltaBound mempty) "MkA"))
-    (Lam (MultVar "p") $
-       Lam (Id (Datatype "A" [MV "p"]) (LambdaBound (MV "p")) "x") $
-         Var (Id (Datatype "A" [MV "p"]) (LambdaBound (MV "p")) "x")
-    )
-    (Scheme "p" (FunTy (Datatype "A" [MV "p"]) (MV "p") (Datatype "A" [MV "p"])))
-
-id2 :: CoreExpr
-id2 =
- Lam (Id (Datatype "A" [MV "p"]) (LambdaBound (MV "p")) "x") $
-   Var (Id (Datatype "A" [MV "p"]) (LambdaBound (MV "p")) "x")
-
-idBad :: CoreExpr
-idBad =
- Lam (Id (Datatype "A" [MV "p"]) (LambdaBound (MV "p")) "x") $
-   Var (Id (Datatype "A" [MV "p"]) (LambdaBound (MV "p")) "y")
 
 varUE :: Var -> Maybe UsageEnv
 varUE (Id _ (DeltaBound ue) _) = Just ue

### 2023-07-14 prototype/test/Test.hs

@@ -1,0 +1,47 @@
+import Data.Maybe
+import qualified Data.Map as M
+
+import Test.Tasty
+import Test.Tasty.HUnit
+
+import Linear.Core.Syntax
+import Linear.Core.Check
+
+-- Let's try writing a simple program like id @A
+-- id :: ∀ p. (A p) ->p (A p)
+-- id = /\p. \x:_pA -> x:A
+idProg :: CoreTerm
+idProg
+  = Term
+    (M.singleton "MkA" (Id (Scheme "kp" (Datatype "A" [MV "p"])) (DeltaBound mempty) "MkA"))
+    (Lam (MultVar "p") $
+       Lam (Id (Datatype "A" [MV "p"]) (LambdaBound (MV "p")) "x") $
+         Var (Id (Datatype "A" [MV "p"]) (LambdaBound (MV "p")) "x")
+    )
+    (Scheme "p" (FunTy (Datatype "A" [MV "p"]) (MV "p") (Datatype "A" [MV "p"])))
+
+id2 :: CoreExpr
+id2 =
+ Lam (Id (Datatype "A" [MV "p"]) (LambdaBound (MV "p")) "x") $
+   Var (Id (Datatype "A" [MV "p"]) (LambdaBound (MV "p")) "x")
+
+idBad :: CoreExpr
+idBad =
+ Lam (Id (Datatype "A" [MV "p"]) (LambdaBound (MV "p")) "x") $
+   Var (Id (Datatype "A" [MV "p"]) (LambdaBound (MV "p")) "y")
+
+main :: IO ()
+main = defaultMain $
+  testCase "Typecheck some things" $ do
+
+    -- 2 + 2 @?= 4
+
+    assertBool "Typechecking idProg" $ typechecks (erase idProg)
+
+    assertBool "Typechecking id2" $ typechecks id2
+
+    assertBool "Shouldn't typecheck idBad" $ (typechecks idBad)
+
+  where
+    typechecks = isJust . runClosedCheck . typecheck
+

### 2023-07-14 prototype/src/Linear/Core/Check.hs
Add CI for prototype testsuite
Fix some dependencies

@@ -6,9 +6,9 @@
 {-# OPTIONS_GHC -Wno-name-shadowing #-}
 module Linear.Core.Check where
 
+import Data.List ((\\))
 import Data.Maybe
 import qualified Data.Map as M
-import Data.List.Unicode ((∖))
 import Data.Functor.Foldable
 import Control.Monad.State
 import Control.Monad.Except
@@ -135,7 +135,7 @@ dryRunUsed check = do
     _Δ' <- gets projectCtxLinear
 
     -- The linear variables used to run the 'Check' action
-    let usedΔ = S.fromList (_Δ ∖ _Δ')
+    let usedΔ = S.fromList (_Δ \\ _Δ')
 
     -- Restore the context
     put _Γ_Δ

### 2023-07-14 prototype/src/Linear/Core/Check.hs
Typecheck Case

@@ -42,6 +42,9 @@ runClosedCheck check = do
 projectCtxLinear :: Ctx -> [Name]
 projectCtxLinear = M.keys . M.filter (\case Id _ (LambdaBound One) _ -> True; _ -> False)
 
+projectLinear :: [Var] -> [Var]
+projectLinear = filter (\case Id _ (LambdaBound One) _ -> True; _ -> False)
+
 -- | Use a variable from the environment.
 -- If it is a linearly-bound variable it will be consumed (and no longer available in the context)
 use :: Name -> Check Var
@@ -201,7 +204,7 @@ typecheck = cata \case
       _ -> empty
 
   -- we don't consider type-lets in our system (yet?)
-  LetF (NonRec (MultVar n) _) e' -> fail "We don't yet support type-level lets"
+  LetF (NonRec (MultVar _) _) _ -> fail "We don't yet support type-level lets"
 
   --- * (Let)
   LetF (NonRec id@(Id var_ty var_bndr _) e) e' -> do
@@ -280,7 +283,60 @@ typecheck = cata \case
 
     return (Let (Rec (zip bndr_vars es)) e', t')
 
-  CaseF _ var alts -> undefined
+  -- TODO: Should the var have just one DeltaEnv, like (1) Δ_s, and then when
+  -- passed to the alternatives is changed, (2) or all of them (costly?), or
+  -- (3) None, so that it is more principled in the sense "the u.e. for
+  -- case-vars are computed during the case? For now, we do opt 1, seems good since the env. of $e$ is the hardest to compute, the others are trivial?
+  CaseF scrut var alts -> do
+    {-
+
+
+          Γ;Δ/Δ'⊦e:σ   {Γ;Δ',z/Δ''⊦ρ -> e':σ => φ} 
+          ----------------------------------------
+            Γ;Δ/Δ''⊦case e of z { {ρ -> e'} } :φ
+
+
+    -}
+
+    _Δ  <- gets projectCtxLinear
+    (scrut, t) <- scrut
+    _Γ_Δ' <- get
+
+    let _Δ'  = projectCtxLinear _Γ_Δ'
+        _Δ_s = S.fromList (_Δ \\ _Δ')
+
+    unless (t == varType (varId var)) $
+      fail "The type of the case var doesn't match the type of the scrutinee"
+
+    unless (Just _Δ_s == varUE var) $
[…]

### 2023-07-14 prototype/src/Linear/Core/Syntax.hs

@@ -4,6 +4,7 @@
    DeriveTraversable, TemplateHaskell, TypeFamilies #-}
 module Linear.Core.Syntax where
 
+import Debug.Trace
 import Data.Functor.Foldable.TH (makeBaseFunctor)
 import Data.Map (Map)
 import qualified Data.Map as M
@@ -148,6 +149,10 @@ erase (Term _ e _) = e
 varUE :: Var -> Maybe UsageEnv
 varUE (Id _ (DeltaBound ue) _) = Just ue
 varUE _ = Nothing
+
+setUE :: Var -> UsageEnv -> Var
+setUE (Id t (DeltaBound _ue_old) n) ue = Id t (DeltaBound ue) n
+setUE x _ = trace "Setting the UE of a non DeltaBound var" x
 
 varId :: Var -> Id
 varId (Id' i) = i

### 2023-07-14 prototype/test/Test.hs

@@ -40,7 +40,7 @@ main = defaultMain $
 
     assertBool "Typechecking id2" $ typechecks id2
 
-    assertBool "Shouldn't typecheck idBad" $ (typechecks idBad)
+    assertBool "Shouldn't typecheck idBad" $ not (typechecks idBad)
 
   where
     typechecks = isJust . runClosedCheck . typecheck

### 2023-07-14 prototype/src/Linear/Core/Parser.hs
Fragmented usage environments
* Much Simpler
* And uniform
* More general
* Accepts reverse-binder-swap as linearity preserving

@@ -1,0 +1,14 @@
+module Linear.Core.Parser where
+
+import Data.Text (Text)
+-- import qualified Data.Text as T
+
+import Data.Functor.Identity
+import Data.Void
+
+import Text.Megaparsec
+
+type Parser = ParsecT Void Text Identity
+
+Control
+

### 2023-07-22 call-notes/17-07-2023.hs
A giant overhaul of the system, do things better.
* Lots of call notes
* In the implementation, convert semi-successfully from Core to Linear
  Core (to be dropped in favour of a re-implementation directly on Core
  using the new system)
* Typeset the new rules.

@@ -1,0 +1,279 @@
+{-
+
+
+* Escrever os 3 passos da transformação do Simon
+* Escrever o programa do double free que não pode estar well-typed
+
+
+-- para realmente fechar a handle, resultado tem de ser consumido
+close :: Handle -o RW# -o (# String, RW# #)
+
+map :: (a -o b) -> [a] -o [b]
+
+--
+
+Em call-by-need, os recursos passados a uma função só são consumidos quando o resultado da função é consumido. (Linear Haskell agrees)
+
+Definition 2.1 (Consume exactly once).
+• To consume a value of atomic base type (like Int or Ptr) exactly once, just evaluate it. (NF)
+• To consume a function value exactly once, apply it to one argument, and consume its result
+exactly once. (β)
+• To consume a pair exactly once, pattern-match on it, and consume each component exactly
+once. (WHNF)
+• In general, to consume a value of an algebraic datatype exactly once, pattern-match on it,
+and consume all its linear components exactly once (Sec. 2.5).
+
+--
+
+[suspended computation]
+==force==>
+[resource (string)] -- tem de ser consumido linearmente, para fazer sentido termos consumido os suspended resources
+
+
+    Δ,xΔσ ⊦ x:σ   Γ,x1!σ ⊦ e : τ
+    ----------------------------- (unsuspend)
+        Γ,Δ,xΔσ ⊦ !x; e : τ
+
+
+    Δ,xΔσ ⊦ x:σ    Γ, x1!σ, (a,b,c):x, z:x ⊦ e : τ
+    -----------------------------------------------(unsuspend)
+        Γ,Δ,xΔσ ⊦ case !x of z { (...) -> e } : τ
+
+
+let x = close handle rw -- thunk -- suspended computation -- suspended resources (handle, rw)
+
+!x -- result of suspended computation
+
+
+
+x -> WHNF consumimos algumas resources
+x -> NF consumimos todas as resources
+vs
+x -> WHNF consumimos todas resources
+
+
+
+-}
+
+
+f x = case x of z { K a b -> expensive z }
+
+===>
+data K a b = K a b
+x :: K a b
+g ... = let y = ... in f y
+
+f x = case x of z { K a b -> expensive x }
+-- x:1 is
+-- * evaluated (case x of -- noop)
+-- * not evaluated (case x of -- evaluates, move from Δ -> 1)
+
+
+{--
+ Ao passar WHNF ->
+  * os recursos que estavam suspensos já 
[…]

### 2023-07-22 call-notes/18-07-2023.hs

@@ -1,0 +1,288 @@
+{-
+Bom dia professor, há uma ideia que ainda não consegui formular de coisas
+"construtivas" ou "positivas" vs coisas "negativas"
+
+A ideia é que quando forçamos uma coisa positiva no Case, os recursos ainda
+estão disponíveis no interior do case, porque forçar uma computação "positiva"
+(e.g. vars, construtores) não consome recursos lineares nem os põe no que eu
+ando a chamar "ghost resources".  No entanto, forçar uma computação "negativa"
+(e.g.  application, cases) vai consumir os recursos lineares do scrutinee e
+passá-los para o dito ghost context.
+
+A qualquer momento, um recurso linear `x` ou `y` apenas pode aparecer out no
+/linear context/ ou no /ghost linear context/. Desta forma, um `z` com usage
+environment `{x,y}` pode ser usado desde que `{x,y}` estejam ou no linear
+context, ou no ghost context. Para além disso, recursos lineares no contexto
+linear podem ser usados, mas recursos lineares no ghost context não podem ser
+usados - têm de ser consumidos através do usage environment de outros recursos.
+
+Os usage environments do case binder e dos patterns seriam iguais
+para todos os scrutinees, quer estes fossem positivos ou negativos.
+
+O que difere da regra em que o scrutinee é positivo para a regra em que o
+scrutinee é negativo é que forçar um scrutinee negativo "vai por recursos a serem consumidos" (vai passar os recursos do contexto linear para o ghost context).
+
+Pelo contrário, forçar um scrutinee positivo não consome nenhum recurso, por isso os recursos são mantidos.
+
+* Por isso, para positivos, manter os recursos do scrutinee nas alternativas
+
+* Para negativos, mover os recursos do scrutinee para o ghost context
+
+* Em ambos, as linearly bound variables no pattern têm um "fragmented usage
+  environment" do scrutinee, e o case binder o "usage environment" do scrutinee
+
+* Em ambos, se não houver linearly bound variables, os recursos do scrutinee são consumidos na totalidade (quer desapareçam do ghost ou do linear environment)
[…]

### 2023-07-22 call-notes/19-07-2023.hs

@@ -1,0 +1,41 @@
+{-
+Let's have three case rules:
+* Scrutinee is in WHNF
+* Scrutinee is not in WHNF and is not a var (are vars in WHNF? kind-of)
+* Scrutinee is a var
+
+And the alternative rules are the same for all of them, we simply split the
+resources amongst the linearly bound variables, or consume the environment
+fully if there are no more linear variables bound in the pattern (i.e. we've reached NF up to linearity, i.e. all linear resources are in NF)
+ -}
+
+{-
+Provavelmente seria bom olhar para a Permission com fractions da separation logic
+Mas a minha ideia agora é que se pode dividir os recursos:
+
+   G, x:1 |- e : T
+-------------------------(Split)
+G, x:1/2, x:1/2 |- e : T
+
+e pronto, é só isso quase.
+com uma regra para tirar recursos do ghost context,
+
+--------------------(Var Delta)
+x:D:T ; {D} |- x : T
+
+e outra para tirar do sitio normal
+
+--------------------(Var Delta)
+D, x:D:T ; . |- x : T
+
+para usar o a, b dos patterns, é preciso dividir os recursos em fragmentos dos recursos
+e como não temos o contract, não dá para usar a várias vezes, por isso as frações são suficientes
+(não sei é quanto àquela optimização em que aproveitávamos  a contract como regra admissível)
+
+Ao mantermos dois contextos separados podemos falar das substituições por vazio
+(.) sobre as duas ao mesmo tempo, porque têm o mesmo nome, estão é em sítios
+diferentes.
+
+Depois também é preciso que na regra da linear variable a linear variable não
+possa ser tirada do ghost context
+-}

### 2023-07-22 call-notes/20-07-2023.hs

@@ -1,0 +1,210 @@
+{-
+On context management...
+
+* Proof irrelevant
+
+
+Todos os contextos:
+* ou são distintos ( Γ,Γ'; Δ,Δ'; δ,δ' ; Ω,Ω' ) (unrestricted, linear, delta-bound, proof-irrelevant linear)
+* ou estão todos juntos (muito mais precisos na semântica de coisas como "has no linear variables"...)
+
+explicit better than implicit
+
+"Γ has no linear variables -- ..."
+
+Γ' has no linear variables
+Γ',Δ ⊦ e : σ
+e is in WHNF
+{Γ, Δ, z{Δ} ⊦ p -> e :_Δ σ => φ}
+----------------------------------
+Γ,Γ',Δ ⊦ case e of z {p -> e} : φ
+
+"o que é a que é a tralha toda"
+Prof: é delta-bound, linear e proof irrelevant
+
+---------------- (Var Δ)
+Δ, x:Δ ⊦ x : σ
+
+
+Δ,a:Δ ...---
+--------------------------------(K)
+Δ, a:Δ, Δ', b:Δ' ⊦ K a b : σ
+
+Δ linear variables used in the context
+a e b nao são linear, são Δbound, mas os seus Δ e Δ' associados são...
+
+
+-}
+
+f x y = case g x y of z -- is NOT WHNF
+          K a b -> -- {x:1}, {y:1}, a{x:1/2,y:1/2}, b{x:1/2,y:1/2}, z{x:1,y:1}
+            case h a b of
+              K n p -> ...
+            -- or
+            case h a b of
+              K -> ...
+            -- or
+            -- Δ_s = {x:1/2, y:1/2, x:1/2, y:1/2} ==> {x:1, y:1}
+            case K a b of
+              K n p -> -- {x}, {y}, z{x,y} n{x:1/2, y:1/2}, p{x:1/2, y:1/2}, porque não também a{x:1/2, y:1/2}, b{x:1/2, y:1/2}?
+
+{-
+
+Ser explicito sobre a divisão de todos os (4 contextos) seria muito mais
+uniforme, explícito, e de certa forma não demasiado difícil de modificar as
+provas existentes, com os benefícios de conseguir usar as novas ideias!
+
+
+~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+Vamos separar os 4 contextos
+~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+
+Prof: Primeiro arranjar as regras
+
[…]

### 2023-07-22 call-notes/21-07-2023.hs

@@ -1,0 +1,9 @@
+{-
+On using usage environments instead of unrestricted and linear resources (empty UE vs exactly one UE vs complex UE):
+* This isn't useful for linear resources, since they would have to appear once in the linear context and another in the delta environment, we'd have double the bookkeeping there
+* However, for unrestricted resources, this seems like a great idea:
+  * It would uniformise/get rid completely of the distinction between unrestricted and empty-usage-environment variables
+  * And means we could get rid of the unrestricted environment completely!
+  * However, the delta context doesn't have Contract. Should be simple to fix a way for empty-usage-env/unrestricted variables (talvez uma regra explícita de contract para quando os environments estão vazios, porque fzemos o split na mesma)
+  * Se bem que neste momento estou a por type variables e constructors no Unrestricted. Mas os Constructors parecem-me tmb bem no Delta. As type variables é que já não sei. Se calhar a distinção dá jeito, tendo em conta que no Core há mais coisas unrestricted (coercions, ...)
+-}

### 2023-07-22 prototype/src/Linear/Core.hs

@@ -1,0 +1,13 @@
+{-|
+In this module we attempt to implement the linear type system defined in the
+thesis.
+- Please finish it before implementing it, it'll save me some time.
+
+This implementation works directly on top of Core as defined in GHC, instead of
+using a different reduced syntax
+
+-}
+module Linear.Core where
+
+
+

### 2023-07-22 prototype/src/Linear/Core/Check.hs

@@ -5,6 +5,7 @@
 {-# OPTIONS_GHC -Wno-name-shadowing #-}
 module Linear.Core.Check where
 
+-- import Debug.Trace
 import Data.List ((\\))
 import Data.Maybe
 import qualified Data.Map as M
@@ -15,11 +16,15 @@ import Control.Monad.Except
 import Control.Applicative
 -- import Data.Set (Set)
 import qualified Data.Set as S
-import Data.Text (Text)
+-- import Data.Text (Text)
 import qualified Data.Text as T
+import Prettyprinter
+import Data.Void
 -- import Error.Diagnose
 
 import Linear.Core.Syntax
+import GHC.Driver.Ppr (showPprUnsafe)
+import GHC.Plugins (literalType)
 
 {-
 Notes:
@@ -30,18 +35,25 @@ Notes:
 
 * We don't allow multiplicity variables to appear anywhere besides function
 types, so we don't need a type-level multiplicity variable
+
 -}
 
-newtype Check a = Check {unCheck :: StateT Ctx (Except Text) a} -- romes:Todo: Except (Diagnostic Text)
-  deriving (Functor, Applicative, Monad, MonadState Ctx, MonadError Text, Alternative)
+newtype Check a = Check {unCheck :: StateT Ctx (Except (Doc Void)) a} -- romes:Todo: Except (Diagnostic Text)
+  deriving (Functor, Applicative, Monad, MonadState Ctx, MonadError (Doc Void), Alternative)
 
 instance MonadFail Check where
-  fail txt = throwError (T.pack txt)
+  fail txt = throwError (pretty txt)
 
-runClosedCheck :: Check a -> Either Text a
+runClosedCheck :: Check a -> Either (Doc Void) a
 runClosedCheck check = runExcept do
   (a,ctx) <- runStateT (unCheck check) mempty
   unless (null $ projectCtxLinear ctx) (throwError "Linear variables escaped")
+  return a
+
+runCheck :: Ctx -> Check a -> Either (Doc Void) a
+runCheck ctx check = runExcept do
+  (a,ctx') <- runStateT (unCheck check) ctx
+  unless (projectCtxLinear ctx' == projectCtxLinear ctx) (throwError "Linear variables escaped")
   return a
 
 -- TODO: Should project linear also return x{:}_p ?
@@ -53,24 +65,41 @@ projectLinear = filter (\case Id _ (LambdaBound One) _ -> True; _ -> False)
 
 -- | Use a variable from the environment.
[…]

### 2023-07-22 prototype/src/Linear/Core/Monad.hs

@@ -1,0 +1,78 @@
+{-# LANGUAGE LambdaCase #-}
+{-# LANGUAGE BlockArguments #-}
+{-# LANGUAGE OverloadedStrings #-}
+-- | A monad transformer for computations that consume and thread through resources from a linear
+-- environment
+module Linear.Core.Monad
+  ( LinearCtxT
+  , use
+  , extend
+  , unrestricted
+  )
+  where
+
+import Data.String
+import Control.Monad.State
+import Control.Monad.Except
+import Data.Map  as M
+import Data.List as L
+import Data.Maybe
+
+-- | Computations that thread through a linear context from which resources
+-- must be used exactly once.
+newtype LinearCtxT k v m a = LinearCtxT { unLC :: StateT (M.Map k v) m a }
+  deriving (Functor, Applicative, Monad, MonadTrans)
+
+
+-- | Uses a resource from the linear environment, making it unavailable in subsequent computations.
+-- If the key doesn't exist, no resources are consumed
+use :: (Ord k, Monad m) => k -> LinearCtxT k v m (Maybe v)
+use key = LinearCtxT do
+  mv <- gets (M.lookup key)
+  case mv of
+    Nothing -> return Nothing
+    Just v -> do
+      modify (M.delete key)
+      return (Just v)
+  
+
+-- | Extend a computation that threads linear resources with a resource that
+-- must definitely not escape its scope (i.e. the given computation must
+-- consume the resource it was extended with).
+--
+-- If the resource escapes the given computation, the resulting computation fails.
+--
+-- If the resource was already in the context it also fails.
+extend :: (Ord k, MonadError e m, IsString e, Show k)
+       => k -> v -> LinearCtxT k v m a -> LinearCtxT k v m a
+extend key value computation = LinearCtxT do
+  gets (isNothing . M.lookup key) >>= \case
+    False -> throwError $ fromString $ "Tried to extend a computation with a resource that was already in the environment: " ++ show key ++ "."
+    True -> do
+      modify (M.insert key value)
+      result <- unLC computation
+      wasConsumed key >>= \case
+        True -> return result
+        False -> throwError $ fromString $
+          "The linear resource " ++ show key ++ " wasn't consumed in the computation extended with it"
+        where
+          wasConsumed x = gets $ isNothing . M.lookup x
+
+
+-- | Runs a computation that threads linear resources and fails if the
+-- computation consumed any resource at all (i.e. fails if the input and output
[…]

### 2023-07-22 prototype/src/Linear/Core/Plugin.hs

@@ -1,0 +1,279 @@
+{-# LANGUAGE ViewPatterns #-}
+{-# LANGUAGE LambdaCase #-}
+{-# LANGUAGE OverloadedStrings #-}
+{-# OPTIONS_GHC -Wno-orphans #-}
+module Linear.Core.Plugin where
+
+{-
+Implementation Notes:
+
+fromCore and fromBind will take a Core expression and convert it into a LinearCore expression.
+This conversion isn't complete: the usage environments are only inferred in a
+subsequent pass 'inferUsageEnvironments'.
+Only after the usage environments are inferred can the expressions be typechecked.
+
+* (We could likely compute the usage environment of non-rec lets while converting the expressions, but for let recs it's harder)
+
+* This module is looking way too confusing:
+Perhaps re-do the implementation directly on Core instead of Linear.Core.
+
+--------
+Using usage environments instead of unrestricted and linear resources (empty UE vs exactly one UE vs complex UE):
+* this isn't useful for linear resources, since they would be duplicated
+ -}
+
+import Data.Void
+import Data.Text (Text)
+import qualified Data.Text as T
+import Prettyprinter
+
+import GHC.Core.TyCo.Rep
+import GHC.Types.Var
+import GHC.Plugins
+import Control.Monad
+import Control.Monad.Reader
+import Control.Monad.Except
+
+import qualified Linear.Core.Syntax as LC
+import qualified Linear.Core.Check as LC
+import GHC.Core.Map.Type
+import Data.Maybe
+import Data.Functor
+
+type CoreCheck = ReaderT CoreProgram (ExceptT (Doc Void) CoreM)
+type CoreConv  = ReaderT (VarEnv LC.Var) CoreCheck
+
+plugin :: Plugin
+plugin = defaultPlugin { installCoreToDos = install }
+
+install :: [CommandLineOption] -> [CoreToDo] -> CoreM [CoreToDo]
+install _opts todo =
+  return (concatMap (:[CoreDoPluginPass "Validate Linear Core" validateLinearGuts]) todo)
+
+
+--------------------------------------------------------------------------------
+----- Attempt 1 - Core to Linear Core ------------------------------------------
+--------------------------------------------------------------------------------
+
+validateLinearGuts :: ModGuts -> CoreM ModGuts
+validateLinearGuts guts = do
[…]

### 2023-07-22 prototype/src/Linear/Core/Syntax.hs

@@ -14,6 +14,8 @@ module Linear.Core.Syntax
   )
   where
 
+import GHC.Utils.Outputable (showPprUnsafe, ppr)
+import GHC.Types.Literal
 -- import Data.Kind
 -- import GHC.TypeLits
 import Debug.Trace
@@ -73,9 +75,7 @@ varName :: Var -> Name
 varName (Id _ _ n)  = n
 varName (MultVar n) = n
 
-data Var = Id' Id
-         | MultVar' MultVar
-         deriving (Eq, Show)
+data Var = Id' Id | MultVar' MultVar deriving (Eq, Show)
 {-# COMPLETE Id, MultVar #-}
 
 pattern Id :: Ty -> IdBinding -> Name -> Var
@@ -104,7 +104,10 @@ data IdBinding = LambdaBound Mult
                | DeltaBound UsageEnv -- ^ Let and case bound variables. Case variables do have many usage environments, but in practice (when they occur in a context), they have just one usage environment
                deriving (Eq, Show)
 
+data BindingSite = LambdaBindSite | LetBindSite deriving (Eq, Show)
 
+-- I suppose we need TyMultVar to type mult applications, despite no term with
+-- that type existing.
 data Ty
   -- = TyMultVar Name          -- p -- no, multiplicities can't exist on their own, only attached to functions (or datatype univ. vars)
   = Datatype Name [Mult]    -- K π_1 ... π_n
@@ -114,6 +117,7 @@ data Ty
 
 data Expr b where
   Var  :: b -> Expr b -- Only term variables and constructors. Type variables at the term level would be (Ty (TyMultVar "p"))
+  Lit  :: Literal_ -> Expr b
   Lam  :: b -> Expr b -> Expr b
   App  :: Expr b -> Expr b -> Expr b
   Let  :: Bind b -> Expr b -> Expr b
@@ -146,6 +150,9 @@ data AltCon b
 deriving instance Eq b => Eq (AltCon b)
 deriving instance Show b => Show (AltCon b)
 
+newtype Literal_ = L Literal deriving Eq
+instance Show Literal_ where
+  show (L l) = showPprUnsafe l
 
 -- | All constructors are of the form K~\overline{x{:}_\pi\sigma}
 -- We record the information about the multiplicity and type of each argument
@@ -182,7 +189,7 @@ makeBaseFunctor ''Ty
 
[…]

### 2023-07-26 call-notes/22-07-2023.hs
Updates

@@ -1,0 +1,35 @@
+{-
+(Also good to include this in the thesis as an alternative solution and the one we ended up taking)
+A simpler alternative to annotating every pattern-bound variable with the constructor and the corresponding pattern position,
+we could NOT allow Joining (x:1/2, x:1/2 -> x:1), and to simply split the variables in subsequent constructors further, even if we end up having more of them
+
+  * This means we can no longer type
+    
+    f x y = case (x,y) of
+      (a,b) -> case (x,y) of
+        (n,p) -> (a,p)
+
+    Oh, no. Wait. The `x` and `y` in the second case are still whole because we didn't split them up. Meaning we could return (a,n)... which isn't linear semantically.
+
+Still, I was going to say that
+
+  * We get rid of the Join clauses (We only ever split when we need to instantiate variables with fragmented usage environments, we never rebuild the split variables into one)
+  * Restricts that special case (actually it doesn't, see above)
+  * We don't need complicated positional+constructor arguments
+  * We are back to reasonable arithmetic
+
+Turns out it's not quite right.
+
+Nonetheless, I don't think we will ever need the join rule - just don't split things if you don't intend on using them separately.
+
+
+----------------------------------------
+
+Eu até posso apresentar a prova de soundness sobre a linguagem tal e qual como
+ela está, mesmo com a deficência que permite o double free, e depois digo que
+remover o que permite o double free seria simples, e apenas implicaria mudar o
+outro caso da alternativa, mas que perseguimos outro caminho completamente
+diferente muito melhor.
+
+
+-}

### 2023-07-26 call-notes/23-07-2023.hs

@@ -1,0 +1,114 @@
+{-
+
+In essence, we have one unrestricted context, two linear contexts, and one
+affine context. This matches the relationship between affine types and
+call-by-need explored in~\cite{}.
+
+What do we know about affine type systems that we can use in our system for the \deltas?
+
+Notes from linear haskell:
+Variables in the environments are annotated by a multiplicity (1 or ω), ω-variables are ordinary
+variables. When forced, an ω-variable is replaced by its value (to model lazy sharing), but 1-
+variables must be consumed exactly once: when forced, they are removed from the environment.
+Reduction gets stuck if a 1-variable is used more than once.
+
+
+On NF values:
+f rw = case rw of
+        RealWorld# -> ...?? rw has been consumed, what now? I suppose we can't make up RealWorld#s, only thread them?
+
+
+On contexts:
+Perhaps we can merge the proof irrelevant and linear contexts?
+Dropping Γ no longer seems like that good of an idea, because it would conflate unrestricted with affine variables
+Then again, delta vars with empty usage environments are really not affine... (either we move them, or give explicit contract)
+
+On optimisations:
+Don't forget to re-read the future-work section of Linear Haskell!
+
+On the implementation:
+How do we infer the proof-irrelevant linear context? How do we know what is in
+the proof irrelevant context vs what isn't without typechecking?
+Perhaps in the inference pass we carry a reader context mapping linear
+resources to either the linear or proof irrelevant context, and update it in
+cases expressions where the scrutinee is not in whnf.
+
+Three relevant notes:
+Note [Zero as a usage]
+~~~~~~~~~~~~~~~~~~~~~~
+In the current presentation usages are not exactly multiplicities, because they
+can contain 0, and multiplicities can't.
+
+Why do we need a 0 usage? A function which doesn't use its argument will be
+required to annotate it with `Many`:
+
+    \(x % Many) -> 0
+
+However, we cannot replace absence with Many when computing usages
+compositionally: in
+
+    (x, True)
+
+We expect x to have usage 1. But when computing the usage of x in True we would
+find that x is absent, hence has multiplicity Many. The final multiplicity would
+be One+Many = Many. Oops!
+
+Hence there is a usage Zero for absent variables. Zero is characterised by being
+the neutral element to usage addition.
+
+We may decide to add Zero as a multiplicity in the future. In which case, this
+distinction will go away.
+
+Note [Joining usages]
+~~~~~~~~~~~~~~~~~~~~~
[…]

### 2023-07-26 call-notes/24-07-2023.hs

@@ -1,0 +1,185 @@
+
+{-
+
+G;D;d;[D] |- ...
+
+G ::= . | p,G | x:w,G | K, G
+d ::= x:D,[D],D
+
+K => K:w
+x => x:w
+x:w => x:.,[.]
+x:.,[.] <=> x:w
+
+
+G; D; d; [D]
+
+--------------
+
+* Regra que permite empurrar coisas de `d` para `G`
+ou
+* Ter um contract explícito para coisas com delta env vazio no d
+
+
+G ; D ; d ; [D]
+U ; L ; A ; L
+
+------(var)
+x |- x
+
+não!
+------(var)
+[x] |- x
+
+
+---------(varD)
+x,[y],z:{x,y} |- z
+
+---------(varD)
+x,z:{x,[]} |- z
+
+
+
+-}
+
+f x y = let w = use x y (Δ={x,y})
+         in case h x y of
+              K a b -> Δ={},[Δ]={x,y}
+                return w{x,y}
+
+f x y = let w = use x y (Δ={x,y})
+         in case h x of
+              K a b -> {[x],y}
+                return w
+
+f x y = let w = use x (Δ={x})
+         in case K x of
+              K2 a b -> return w
+
+f x y = let w = use x (Δ={x})
+         in case h x of
+              K2 a b -> return w -- bad!
+
+f x y =
+  case x of
+    _ ->
+      let w = use x y (Δ={x,y})
+      ---ou--
+      case h y of
+        _ -> return w
+      ----- mas isto está bem
+      return w
+      ----
+      case h w of
+        _ -> (Δ={[x,y]})
+
+-- este exemplo mostra porque é que é preciso ter os 2 contextos nas delta
+-- annotations, pq temos de saber as coisas que já usamos na mesma
+f x y =
+  case h x of
+    K a{x/2-1K} b{x/2-2K} ->
+      let w = use a b y (Δ={[x/2,x/2],y})
+      return w
+
+
+-- Como usar os positional binders por constructor
+
+K : \ov{σ ->} φ
+
+Γ;Δ;δ,a{Δ1 -> 1K},b{Δ1 -> 2K},c{Δ1 -> 3K};[Δ] ⊦ e : φ
+----------------------------------------------------
+Γ;Δ;δ;[Δ] ⊦ K a b c -> e : σ => φ
+
+f x y = case (x,y) of
+          (a{xy/2},b{xy/2}) -> case (x,y) of
[…]

### 2023-07-26 prototype/src/Linear/Core.hs

@@ -9,5 +9,58 @@ using a different reduced syntax
 -}
 module Linear.Core where
 
+import Control.Monad.Reader
+import GHC.Core.UsageEnv
+import GHC.Plugins
+import GHC.Core.Lint
+import GHC.Types.Var.Env
+import Data.Map as M
+
+type LCProgram = [LCBind]
+type LCBind = Bind LCVar
+type LCExpr = Expr LCVar
+type LCAlt  = Alt LCVar
+
+-- | Whenever we recurse into the body of a case expression (whose scrutinee is
+-- not in WHNF) to determine the delta annotations of the Delta-bound variables,
+-- we need to move the linear variables from the scrutinee to the
+-- proof-irrelevant context. This context is needed for 'makeDeltaAnnFrom',
+-- which creates a delta-env based on the usage environment of an expression,
+-- but requires splitting the resources between proof-irrelevant and relevant.
+type InferDeltas a = ReaderT (VarEnv LCVar) CoreM
+
+-- Usage environments annotated to delta variables:
+-- The linear variables and proof irrelevant linear variables that suspended on
+-- a computation bound to the annotated delta var
+data DeltaAnn = DeltaAnn
+  { linear :: UsageEnv
+  , proof_irrelevant :: UsageEnv
+  }
+
+data IdBinding = LambdaBound Mult    -- lambda
+               | DeltaBound DeltaAnn -- both let and case binders
+
+data LCVar = LCVar
+  { id :: Var
+  , binding :: Maybe IdBinding -- we only have Id bindings for Ids (not for TyVars)
+  }
 
 
+runLinearCore :: CoreProgram -> CoreM [SDoc]
+runLinearCore = undefined
+
+inferDeltaAnns :: CoreProgram -> LintM LCProgram
+inferDeltaAnns = traverse inferDeltaAnnsBind
+
+inferDeltaAnnsBind :: CoreBind -> LintM LCBind
+inferDeltaAnnsBind (NonRec b e) = do
+  (ty, ue) <- lintCoreExpr e
+  deltaAnn <- makeDeltaAnnFrom ue
+  _
+
+inferDeltaAnnsExpr :: CoreExpr -> LintM LCExpr
+inferDeltaAnnsExpr expr = do
+
+
+makeDeltaAnnFrom :: UsageEnv -> DeltaAnn
+makeDeltaAnnFrom ue

### 2023-07-26 prototype/src/Linear/Core/Monad.hs

@@ -11,6 +11,7 @@ module Linear.Core.Monad
   )
   where
 
+import Control.Monad
 import Data.String
 import Control.Monad.State
 import Control.Monad.Except
@@ -20,6 +21,7 @@ import Data.Maybe
 
 -- | Computations that thread through a linear context from which resources
 -- must be used exactly once.
+-- (i.e. models computations that are well-defined under a linear context)
 newtype LinearCtxT k v m a = LinearCtxT { unLC :: StateT (M.Map k v) m a }
   deriving (Functor, Applicative, Monad, MonadTrans)
 

### 2023-07-26 prototype/src/Linear/Core/Plugin.hs

@@ -34,11 +34,13 @@ import Control.Monad
 import Control.Monad.Reader
 import Control.Monad.Except
 
-import qualified Linear.Core.Syntax as LC
-import qualified Linear.Core.Check as LC
+-- import qualified Linear.Core.Syntax as LC
+-- import qualified Linear.Core.Check as LC
 import GHC.Core.Map.Type
 import Data.Maybe
 import Data.Functor
+
+import Linear.Core
 
 type CoreCheck = ReaderT CoreProgram (ExceptT (Doc Void) CoreM)
 type CoreConv  = ReaderT (VarEnv LC.Var) CoreCheck
@@ -48,232 +50,246 @@ plugin = defaultPlugin { installCoreToDos = install }
 
 install :: [CommandLineOption] -> [CoreToDo] -> CoreM [CoreToDo]
 install _opts todo =
-  return (concatMap (:[CoreDoPluginPass "Validate Linear Core" validateLinearGuts]) todo)
-
+  return (concatMap (:[CoreDoPluginPass "Linear Core Pass" linearCorePass]) todo)
 
 --------------------------------------------------------------------------------
------ Attempt 1 - Core to Linear Core ------------------------------------------
+----- Attempt 2 - Typecheck Linearity in Core directly--------------------------
 --------------------------------------------------------------------------------
 
-validateLinearGuts :: ModGuts -> CoreM ModGuts
-validateLinearGuts guts = do
+linearCorePass :: ModGuts -> CoreM ModGuts
+linearCorePass guts = do
   let prog = mg_binds guts
 
-  msgs <- runExceptT (runReaderT (forM_ prog validateBind) prog)
+  msgs <- Linear.Core.runLinearCore prog
   case msgs of
     Right () -> pure ()
-    Left e -> fatalErrorMsgS (show e)
+    Left e -> fatalErrorMsg e
 
   return guts -- unchanged guts, after validating them.
 
--- LC.Binder and the type of the bound expression
---
--- Perhaps we should guarantee the bindings are already strongly connected
-validateBind :: CoreBind -> CoreCheck () 
-validateBind b = do
-  printP b
-  lcBind <- runConv (fromBind b)
-  printPretty "Validate Bind" lcBind
-  lift . liftEither $
-    case lcBind of
-      LC.NonRec _ e -> void $ LC.runCheck mempty (LC.typecheck e)
[…]

### 2023-07-27 language-v3/Proofs.hs
Typeset some more proofs

@@ -1,0 +1,65 @@
+#!/usr/bin/env cabal
+{- cabal:
+build-depends: base, HaTeX
+-}
+{-# LANGUAGE OverloadedStrings #-}
+{-# LANGUAGE BlockArguments #-}
+
+-- Welcome to Proofs.hs!
+-- This is not used by the real proof... was just an experiment
+
+import Data.String
+import Control.Monad
+import Text.LaTeX
+import Text.LaTeX.Packages.AMSThm hiding (proof)
+import qualified Text.LaTeX.Packages.AMSThm as AMSThm
+
+type Latex = LaTeXM ()
+
+main :: IO ()
+main = renderFile "all_proofs.tex" (execLaTeXM all_proofs)
+
+all_proofs :: Latex
+all_proofs = do
+  linear_subst
+  linear_subst_alt
+
+lemma :: Latex -> Latex
+lemma = theorem "lemma"
+
+proofCase :: Latex -> [Latex] -> Latex
+proofCase titl steps = do
+  item (Just "Case:")
+  titl
+  -- tabbing do
+  --   forM_ (zip steps [1..]) \(step,i) -> do
+  --     ("\"" <> show i <> "\"")
+  --     math step
+
+linear_subst :: Latex
+linear_subst = do
+  lemma "(Substitution of linear variables preserves typing)"
+  AMSThm.proof Nothing do
+    "By structural induction on the first derivation."
+    description do
+      proofCase "CaseWHNF" caseWHNF
+
+linear_subst_alt :: Latex
+linear_subst_alt = do
+  lemma "(Substitution of linear variables in case alternatives preserves typing)"
+  AMSThm.proof Nothing do
+    "By structural induction on the alt judgment."
+    description do
+      proofCase "AltN" altZero
+
+caseWHNF :: [Latex]
+caseWHNF = []
+
+
+altZero :: [Latex]
+altZero =
+  [ j g' bd' ld' e' s'
+  , "Subcase x does not occur"
+  , j 
+  ]
+

### 2023-08-03 prototype/src/Linear/Core.hs
The K-Vars Arrangement Dilemma

@@ -63,4 +63,4 @@ inferDeltaAnnsExpr expr = do
 
 
 makeDeltaAnnFrom :: UsageEnv -> DeltaAnn
-makeDeltaAnnFrom ue
+makeDeltaAnnFrom ue = _

### 2023-08-05 prototype/src/Linear/Core.hs
Update to Typechecking Linearity in Core directly with patched ghc plugin

@@ -1,3 +1,4 @@
+{-# LANGUAGE GHC2021, ViewPatterns, DerivingVia, GeneralizedNewtypeDeriving, OverloadedRecordDot #-}
 {-|
 In this module we attempt to implement the linear type system defined in the
 thesis.
@@ -11,10 +12,14 @@ module Linear.Core where
 
 import Control.Monad.Reader
 import GHC.Core.UsageEnv
+import GHC.Driver.Config.Diagnostic
+import GHC.Driver.Config.Core.Lint
 import GHC.Plugins
+import GHC.Utils.Trace
 import GHC.Core.Lint
 import GHC.Types.Var.Env
 import Data.Map as M
+import Unsafe.Coerce
 
 type LCProgram = [LCBind]
 type LCBind = Bind LCVar
@@ -32,35 +37,76 @@ type InferDeltas a = ReaderT (VarEnv LCVar) CoreM
 -- Usage environments annotated to delta variables:
 -- The linear variables and proof irrelevant linear variables that suspended on
 -- a computation bound to the annotated delta var
-data DeltaAnn = DeltaAnn
+newtype DeltaAnn = DeltaAnn
   { linear :: UsageEnv
-  , proof_irrelevant :: UsageEnv
   }
+  deriving newtype Outputable
 
-data IdBinding = LambdaBound Mult    -- lambda
+data IdBinding = LambdaBound LCMult  -- lambda
                | DeltaBound DeltaAnn -- both let and case binders
+instance Outputable IdBinding where
+  ppr (LambdaBound m) = text "LambdaBound" <+> ppr m
+  ppr (DeltaBound an) = text "DeltaBound" <+> ppr an
+
+data LCMult = Relevant Mult
+            | Irrelevant Mult
+
+instance Outputable LCMult where
+  ppr (Relevant m) = text "Relevant" <+> ppr m
+  ppr (Irrelevant m) = text "Irrelevant" <+> ppr m
 
 data LCVar = LCVar
   { id :: Var
   , binding :: Maybe IdBinding -- we only have Id bindings for Ids (not for TyVars)
   }
 
+instance Outputable LCVar where
+  ppr v = ppr v.id <+> ppr v.binding
+instance OutputableBndr LCVar where
+  pprPrefixOcc v = ppr v
[…]

### 2023-08-05 prototype/src/Linear/Core/Plugin.hs

@@ -43,7 +43,7 @@ import Data.Functor
 import Linear.Core
 
 type CoreCheck = ReaderT CoreProgram (ExceptT (Doc Void) CoreM)
-type CoreConv  = ReaderT (VarEnv LC.Var) CoreCheck
+type CoreConv  = ReaderT (VarEnv Var) CoreCheck
 
 plugin :: Plugin
 plugin = defaultPlugin { installCoreToDos = install }
@@ -62,8 +62,8 @@ linearCorePass guts = do
 
   msgs <- Linear.Core.runLinearCore prog
   case msgs of
-    Right () -> pure ()
-    Left e -> fatalErrorMsg e
+    [] -> pure ()
+    errs -> fatalErrorMsg (ppr errs)
 
   return guts -- unchanged guts, after validating them.
 

### 2023-08-05 prototype/src/Linear/Core/Check.hs
Trying translation again

@@ -12,7 +12,7 @@ import qualified Data.Map as M
 import Data.Functor.Foldable
 import Control.Monad.State
 import Control.Monad.Except
--- import Control.Monad
+import Control.Monad
 import Control.Applicative
 -- import Data.Set (Set)
 import qualified Data.Set as S

### 2023-08-05 prototype/src/Linear/Core/Syntax.hs

@@ -53,6 +53,9 @@ types, instead of by type synonyms
 
 * We must also implement the algorithm for inferring usage environments, which
 we still need to figure out when exactly it should be run
+
+* Nevermind, we simply support type abstractions, variables, and type
+constructor applications, because removing them is too awful.
 -}
 
 type Name = Text
@@ -112,7 +115,7 @@ data Ty
   -- = TyMultVar Name          -- p -- no, multiplicities can't exist on their own, only attached to functions (or datatype univ. vars)
   = Datatype Name [Mult]    -- K π_1 ... π_n
   | FunTy   Ty   Mult Ty    -- φ ->π σ
-  | Scheme  Name Ty         -- ∀p. φ
+  | Scheme  Name Ty         -- ∀p. φ (both multiplicity and type abstractions...)
   deriving (Eq,Show)
 
 data Expr b where
@@ -275,8 +278,8 @@ instance Pretty Ty where
   pretty = cata go where
     go (DatatypeF name [])   = pretty name
     go (DatatypeF name mult) = pretty name <+> hsep (map pretty mult)
-    go (FunTyF t1 One t2)    = t1 <+> "-o" <+> t2
-    go (FunTyF t1 Many t2)   = t1 <+> "->" <+> t2
-    go (FunTyF t1 m t2)      = t1 <+> "%" <> pretty m <+> "->" <+> t2
+    go (FunTyF t1 One t2)    = parens t1 <+> "-o" <+> parens t2
+    go (FunTyF t1 Many t2)   = parens t1 <+> "->" <+> parens t2
+    go (FunTyF t1 m t2)      = parens t1 <+> "%" <> pretty m <+> "->" <+> parens t2
     go (SchemeF n ty)        = "∀" <> pretty n <> "." <+> ty
 

### 2023-08-06 prototype/app/F.hs
Finalize translation pass iter1

@@ -1,0 +1,21 @@
+{-# LANGUAGE LinearTypes, UnicodeSyntax #-}
+module F where
+
+my_f :: Int -> Int
+my_f x = x
+
+my_f1 :: Int %1 -> Int
+my_f1 x = x
+
+s :: (a -> Maybe b) %1 -> (a -> Maybe b)
+s = undefined
+
+
+my_t :: Maybe Int -> Maybe Int
+my_t x = x
+
+my_g :: (a %1 -> b) -> Maybe a %1 -> Maybe b
+my_g f x = case x of
+             Nothing -> Nothing
+             Just y -> Just (f y)
+

### 2023-08-06 prototype/src/Linear/Core/Plugin.hs

@@ -153,15 +153,36 @@ fromCore = \case
   Lam b e
     | Just b' <- fromLamVar b
     -> do
-      e' <- fromJust <$> extend b b' (fromCore e) -- the body of the lambda can't just be a type in a well-typed expression, can it?
+      Just e' <- extend b b' (fromCore e) -- the body of the lambda can't just be a type in a well-typed expression, can it?
       return $ Just $ LC.Lam b' e'
     | otherwise
     -- type abstractions are removed
     -> fromCore e
 
-  Let b e -> undefined
+  Let (NonRec b e1) e2
+    | Just b' <- fromDeltaVar b
+    -> do
+      Just e1' <- fromCore e2
+      Just e2' <- extend b b' (fromCore e1)
+      return $ Just $ LC.Let (LC.NonRec b' e1') e2'
+    | otherwise -> error "hwat"
 
-  Case e b t alts -> undefined
+  Let (Rec (unzip . filter (isId . fst) -> (bs,es))) e'
+    | Just bs' <- traverse fromDeltaVar bs
+    -> do
+      Just es' <- sequence <$> traverse (extends (zip bs bs') . fromCore) es
+      Just e'' <- extends (zip bs bs') $ fromCore e'
+      return $ Just $ LC.Let (LC.Rec (zip bs' es')) e''
+    | otherwise -> error "hwat"
+
+  Case e b _t alts
+    | Just b' <- fromDeltaVar b
+    -> do
+      Just e' <- fromCore e
+      alts' <- traverse (extend b b' . fromCoreAlt) alts
+      return $ Just $ LC.Case e' b' alts'
+
+    | otherwise -> error "what"
 
   Cast e _co -> fromCore e -- ignore coercions
   Tick _tick e -> fromCore e -- and ticks
@@ -171,6 +192,25 @@ fromCore = \case
     pure $ LC.Mult m
 
   Coercion _co -> error "Coercion" -- how to handle moving from coercion to expr? where can coercions occur?
+
+fromCoreAlt :: CoreAlt -> CoreConv LC.CoreAlt
+fromCoreAlt (Alt con bs e)
+  | Just bs' <- traverse fromLamVar bs
+  = do
+    let con' = fromCoreAltCon con
+    Just e' <- extends (zip bs bs') (fromCore e)
+    return $ LC.Alt con' bs' e'
+
[…]

### 2023-08-06 prototype/src/Linear/Core/Syntax.hs

@@ -163,7 +163,7 @@ data DataCon b where
   DataCon :: Name        -- ^ dcName
           -> [Mult]      -- ^ dcUnivMultVars
           -> [Scaled Ty] -- ^ dcArgTys, with corresponding multiplicity
-          -> DataCon Id  -- ^ Elaborated DataCon
+          -> DataCon Var -- ^ Elaborated DataCon
   DataConName :: Name -> DataCon Name -- ^ Parsed DataCon
   
 deriving instance Show b => Show (DataCon b)

### 2023-08-06 prototype/src/Linear/Core/Check.hs
Stop checking types in typechecker

@@ -36,6 +36,16 @@ Notes:
 * We don't allow multiplicity variables to appear anywhere besides function
 types, so we don't need a type-level multiplicity variable
 
+------
+
+* Checking types after translation from Core is anything but trivial, because
+the conversion is very lossy - we lose casts, coercions, type constructors,
+type variables, synonyms, etc...
+
+* Therefore, we no longer check types. Type checking is already done by the Core typechecker
+
+* Instead, we only check that linearity is maintained and ignore the types.
+
 -}
 
 newtype Check a = Check {unCheck :: StateT Ctx (Except (Doc Void)) a} -- romes:Todo: Except (Diagnostic Text)
@@ -92,8 +102,8 @@ use evar = do
             LambdaBound One -> do
               modify (M.delete name)
             LambdaBound (MV' _) -> do
-              -- ROMES:TODO: I suppose we must treat polymorphic variables as if
-              -- they were linear, bc they might be instantiated to 'One' later on.
+              -- We must treat polymorphic variables as if they were linear, bc
+              -- they might be instantiated to 'One' later on.
               modify (M.delete name)
             DeltaBound ue -> do
               mapM_ (use . Right) (S.toList ue) -- all elements should already have mult == One.
@@ -396,9 +406,12 @@ typecheck = cata \case
 -- exprTy :: CoreExpr -> Ty
 -- exprTy = cata undefined
 
+-- Note! We don't check types because of the elaborate types in Core in comparison to Linear (mini) Core.
+-- This is a noop now.
 checkEqTy :: Ty -> Ty -> Check ()
-checkEqTy t1 t2 = if t1 == t2 then pure ()
-                              else throwError $ "Expecting" <+> pretty t1 <+> "and" <+> pretty t2 <+> "to match."
+checkEqTy _ _ = pure ()
+-- checkEqTy t1 t2 = if t1 == t2 then pure ()
+--                               else throwError $ "Expecting" <+> pretty t1 <+> "and" <+> pretty t2 <+> "to match."
 
 allEq :: Eq a => [a] -> Bool
 allEq = allEq' Nothing where

### 2023-08-06 prototype/src/Linear/Core/Check.hs
Add structure to the document with todos for most bits

@@ -418,3 +418,10 @@ allEq = allEq' Nothing where
   allEq' _        []     = True
   allEq' Nothing  (x:xs) = allEq' (Just x) xs
   allEq' (Just y) (x:xs) = x == y && allEq' (Just y) xs
+
+--------------------------------------------------------------------------------
+-- Inferring usage environments ------------------------------------------------
+--------------------------------------------------------------------------------
+
+
+

### 2023-08-09 prototype/app/F.hs
Compute recursive usage environments of recursive binders
Add computeRecUsageEnvs to inferDeltaAnnsBind

@@ -1,6 +1,10 @@
 {-# LANGUAGE LinearTypes, UnicodeSyntax #-}
 module F where
 
+data Y = Y Y Y
+
+f y = let z = Y z y in z
+
 r,t :: Bool -> Bool
 r x = case x of
         True -> t x

### 2023-08-09 prototype/src/Linear/Core.hs

@@ -23,6 +23,7 @@ import GHC.Core.Lint
 import GHC.Types.Var.Env
 import Data.Map as M
 import Data.List as L
+import Data.Maybe
 
 type LCProgram = [LCBind]
 type LCBind = Bind LCVar
@@ -95,11 +96,12 @@ inferDeltaAnnsBind topflag (NonRec b e)
   = do e'        <- inferDeltaAnnsExpr e
        return (NonRec (LCVar b Nothing) e')
 
-inferDeltaAnnsBind topflag (Rec (unzip -> (bs,es))) = do
-  (unzip -> (_tys, naiveUes)) <- mapM lintCoreExpr es
-  pprTraceM "rec things" $ ppr (Rec (zip bs es))
-  pprTraceM "naive ues" $ ppr naiveUes
-  error "OK"
+inferDeltaAnnsBind topflag (Rec bs) = do
+  let (ids,rhss) = unzip bs
+  (rhss', naiveUes) <- lintRecBindings topflag bs (\_ -> traverse inferDeltaAnnsExpr rhss)
+  let ids' = L.map (\(id',ue') -> LCVar id' (deltaBinding ue'))
+              $ computeRecUsageEnvs (zip ids naiveUes)
+  return (Rec (zip ids' rhss'))
 
 
 inferDeltaAnnsExpr :: CoreExpr -> LintM LCExpr
@@ -132,7 +134,7 @@ inferDeltaAnnsAlt :: UsageEnv -- ^ The usage environment of the scrutinee
 inferDeltaAnnsAlt ue (Alt con args rhs) = do
   rhs' <- extends CasePatBind args $ inferDeltaAnnsExpr rhs
   -- ROMES:TODO: This is wrong for now, we need to update this with the right
-  -- way of inferring usage envs. for case pattern variables, but we first
+  -- way of inferring usage envs for case pattern variables, but we first
   -- should decide how they typecheck altogether
   let args' = L.map (\a -> LCVar a (deltaBinding ue)) args
   return (Alt con args' rhs')
@@ -159,9 +161,14 @@ multBinding v
   | isId v    = Just $ LambdaBound $ Relevant (idMult v)
   | otherwise = Nothing
 
--- computeRecUsageEnvs :: [(Var, UsageEnv)] -- ^ Recursive usage environments associated to their recursive call
[…]

### 2023-08-09 prototype/app/F.hs
Realize using UsageEnvs computed by current Core is not good enough

@@ -3,6 +3,7 @@ module F where
 
 data Y = Y Y Y
 
+f :: Y -> Y
 f y = let z = Y z y in z
 
 r,t :: Bool -> Bool
@@ -27,7 +28,7 @@ my_t :: Maybe Int -> Maybe Int
 my_t x = x
 
 my_g :: (a %1 -> b) -> Maybe a %1 -> Maybe b
-my_g f x = case x of
+my_g f' x = case x of
              Nothing -> Nothing
-             Just y -> Just (f y)
+             Just y -> Just (f' y)
 

### 2023-08-09 prototype/src/Linear/Core.hs

@@ -30,6 +30,15 @@ type LCBind = Bind LCVar
 type LCExpr = Expr LCVar
 type LCAlt  = Alt LCVar
 
+{-
+I realize now this whole approach of leveraging `lintCoreExpr` to get the usage
+environments on a pass before typechecking linear core is likely wrong.
+The way Core determines a usage environment is not the same way we do, because
+we see linearity and usage of resources in a different way.
+
+Meaning this pass is not fit for our purposes.
+ -}
+
 -- | Whenever we recurse into the body of a case expression (whose scrutinee is
 -- not in WHNF) to determine the delta annotations of the Delta-bound variables,
 -- we need to move the linear variables from the scrutinee to the
@@ -49,8 +58,8 @@ newtype DeltaAnn = DeltaAnn
 data IdBinding = LambdaBound LCMult  -- lambda
                | DeltaBound DeltaAnn -- both let and case binders
 instance Outputable IdBinding where
-  ppr (LambdaBound m) = text "LambdaBound" <+> ppr m
-  ppr (DeltaBound an) = text "DeltaBound" <+> ppr an
+  ppr (LambdaBound m) = text "λ=" GHC.Plugins.<> ppr m
+  ppr (DeltaBound an) = text "Δ=" GHC.Plugins.<> ppr an
 
 data LCMult = Relevant Mult
             | Irrelevant Mult
@@ -65,7 +74,8 @@ data LCVar = LCVar
   }
 
 instance Outputable LCVar where
-  ppr v = ppr v.id <+> ppr v.binding
+  ppr (LCVar id Nothing) = ppr id
+  ppr (LCVar id (Just b)) = ppr id <+> ppr b
 instance OutputableBndr LCVar where
   pprPrefixOcc v = ppr v
   pprInfixOcc v = text "`" GHC.Plugins.<> ppr v GHC.Plugins.<> text "`"

### 2023-08-09 prototype/src/Linear/Core.hs
Clean up for the typechecking directly on Core LCVar

@@ -14,16 +14,16 @@ module Linear.Core
 
 import GHC.Utils.Outputable
 import Control.Monad.Reader
-import GHC.Core.UsageEnv
-import GHC.Driver.Config.Diagnostic
 import GHC.Driver.Config.Core.Lint
-import GHC.Plugins
+import GHC.Plugins hiding (Mult)
 import GHC.Utils.Trace
-import GHC.Core.Lint
 import GHC.Types.Var.Env
 import Data.Map as M
 import Data.List as L
 import Data.Maybe
+
+import Linear.Core.Multiplicities
+import Data.Functor.Identity
 
 type LCProgram = [LCBind]
 type LCBind = Bind LCVar
@@ -47,38 +47,13 @@ Meaning this pass is not fit for our purposes.
 -- but requires splitting the resources between proof-irrelevant and relevant.
 type InferDeltas a = ReaderT (VarEnv LCVar) CoreM
 
--- Usage environments annotated to delta variables:
--- The linear variables and proof irrelevant linear variables that suspended on
--- a computation bound to the annotated delta var
-newtype DeltaAnn = DeltaAnn
-  { linear :: UsageEnv
-  }
-  deriving newtype Outputable
-
-data IdBinding = LambdaBound LCMult  -- lambda
-               | DeltaBound DeltaAnn -- both let and case binders
-instance Outputable IdBinding where
-  ppr (LambdaBound m) = text "λ=" GHC.Plugins.<> ppr m
-  ppr (DeltaBound an) = text "Δ=" GHC.Plugins.<> ppr an
-
-data LCMult = Relevant Mult
-            | Irrelevant Mult
-
-instance Outputable LCMult where
-  ppr (Relevant m) = text "Relevant" <+> ppr m
-  ppr (Irrelevant m) = text "Irrelevant" <+> ppr m
+data IdBinding = LambdaBound Mult    -- lambdas
+               | DeltaBound UsageEnv -- both let and case binders (including pattern variables)
 
 data LCVar = LCVar
   { id :: Var
   , binding :: Maybe IdBinding -- we only have Id bindings for Ids (not for TyVars)
   }
-
-instance Outputable LCVar where
-  ppr (LCVar id Nothing) = ppr id
-  ppr (LCVar id (Just b)) = ppr id <+> ppr b
-instance OutputableBndr LCVar where
-  pprPrefixOcc v = ppr v
[…]

### 2023-08-09 prototype/src/Linear/Core/Monad.hs

@@ -77,4 +77,3 @@ unrestricted computation = LinearCtxT do
   return result
 
 
-

### 2023-08-09 prototype/src/Linear/Core/Multiplicities.hs

@@ -1,0 +1,45 @@
+{-# LANGUAGE GHC2021, ViewPatterns, DerivingVia, GeneralizedNewtypeDeriving, OverloadedRecordDot #-}
+module Linear.Core.Multiplicities where
+
+import GHC.Utils.Outputable
+import qualified GHC.Core.Multiplicity as C
+import GHC.Types.Name.Env
+import GHC.Types.Var
+
+data Mult = Relevant C.Mult
+          | Irrelevant C.Mult
+
+-- Usage environments annotated to delta variables:
+-- The linear variables and proof irrelevant linear variables that suspended on
+-- a computation bound to the annotated delta var
+newtype UsageEnv = UsageEnv (NameEnv Mult)
+
+data Usage = Zero | LCM Mult
+
+--------------------------------------------------------------------------------
+----- Usage Environments -------------------------------------------------------
+--------------------------------------------------------------------------------
+
+lookupUE :: Var -> UsageEnv -> Usage
+lookupUE = undefined
+
+deleteUE :: Var -> UsageEnv -> UsageEnv
+deleteUE = undefined
+
+supUE :: UsageEnv -> UsageEnv -> UsageEnv
+supUE = undefined
+
+scaleUE :: Usage -> UsageEnv -> UsageEnv
+scaleUE Zero ue = ue
+scaleUE (LCM m) ue = m `undefined` ue
+
+--------------------------------------------------------------------------------
+----- Outputable Instances -----------------------------------------------------
+--------------------------------------------------------------------------------
+
+instance Outputable Mult where
+  ppr (Relevant m) = text "Relevant" <+> ppr m
+  ppr (Irrelevant m) = text "Irrelevant" <+> ppr m
+
+deriving newtype instance Outputable UsageEnv
+

### 2023-08-09 prototype/src/Linear/Core/Plugin.hs

@@ -35,8 +35,8 @@ import Control.Monad
 import Control.Monad.Reader
 import Control.Monad.Except
 
-import qualified Linear.Core.Syntax as LC
-import qualified Linear.Core.Check as LC
+import qualified Linear.Core.Translate.Syntax as LC
+import qualified Linear.Core.Translate.Check as LC
 import GHC.Core.Map.Type
 import Data.Maybe
 import Data.Functor hiding (unzip)
@@ -70,7 +70,9 @@ linearCorePass guts = do
 
 --------------------------------------------------------------------------------
 ----- Attempt 1 - From Core to Linear Core first -------------------------------
+----- See also Linear.Core.Translate.* modules ---------------------------------
 --------------------------------------------------------------------------------
+-- Everything below this line is unused, but still typechecks
 
 validateLinearGuts :: ModGuts -> CoreM ModGuts
 validateLinearGuts guts = do
@@ -364,7 +366,8 @@ printPretty str a = liftIO $ do
 instance MonadFail CoreM where
   fail = error
 
--------------------
--- Tomorrow:
--- Re-do type-checker implementation directly on Core using the new contexts.
--- Nevermind, we're going back and forth between translation and direct typechecking...
+-----------------
+ -- Tomorrow:
+ -- Re-do type-checker implementation directly on Core using the new contexts.
+ -- Nevermind, we're going back and forth between translation and direct typechecking...
+ -- No, do go back to direct typechecking

### 2023-08-09 prototype/src/Linear/Core/Translate/Check.hs

@@ -3,7 +3,7 @@
 {-# LANGUAGE DuplicateRecordFields #-}
 {-# LANGUAGE LambdaCase #-}
 {-# OPTIONS_GHC -Wno-name-shadowing #-}
-module Linear.Core.Check where
+module Linear.Core.Translate.Check where
 
 -- import Debug.Trace
 import Data.List ((\\))
@@ -22,7 +22,7 @@ import Prettyprinter
 import Data.Void
 -- import Error.Diagnose
 
-import Linear.Core.Syntax
+import Linear.Core.Translate.Syntax
 import GHC.Driver.Ppr (showPprUnsafe)
 import GHC.Plugins (literalType)
 

### 2023-08-09 prototype/src/Linear/Core/Translate/Infer.hs

@@ -10,13 +10,13 @@ https://dl.acm.org/doi/pdf/10.1145/3450952
 Perhaps this would all be easier accounting for linearity using the rules in Sec. 6?
 
 -}
-module Linear.Core.Infer where
+module Linear.Core.Translate.Infer where
 
 import Data.Text (Text)
 import qualified Data.Text as T
 import qualified Data.Map as M
 import Prelude hiding (lookup)
-import Linear.Core.Syntax
+import Linear.Core.Translate.Syntax
 import Data.Functor.Foldable
 import Control.Monad
 import Control.Monad.Reader
@@ -27,7 +27,7 @@ import Data.Void
 import Prettyprinter
 
 -- We run 'typecheck' after the naive inference, to check linearity and validation.
-import qualified Linear.Core.Check as Core
+import qualified Linear.Core.Translate.Check as Core
 
 -- We don't have SrcSpans yet, for reporting error messages.
 

### 2023-08-09 prototype/src/Linear/Core/Translate/Parser.hs

@@ -1,7 +1,7 @@
 {-# LANGUAGE OverloadedStrings #-}
 {-# LANGUAGE BlockArguments #-}
 {-# OPTIONS -Wno-orphans #-} -- HasHints Void
-module Linear.Core.Parser where
+module Linear.Core.Translate.Parser where
 
 {-
 It could be more interesting to also transform Core into this language, than to
@@ -24,7 +24,7 @@ import Control.Monad.Combinators.Expr
 import Control.Monad.IO.Class
 import qualified Text.Megaparsec.Char.Lexer as L
 
-import Linear.Core.Syntax
+import Linear.Core.Translate.Syntax
 import Error.Diagnose
 import Error.Diagnose.Compat.Megaparsec
 import System.Exit

### 2023-08-09 prototype/src/Linear/Core/Translate/Syntax.hs

@@ -8,8 +8,8 @@
 {-# LANGUAGE PatternSynonyms #-}
 {-# LANGUAGE MonoLocalBinds, DataKinds #-}
 {-# LANGUAGE DeriveTraversable, TemplateHaskell #-}
-module Linear.Core.Syntax
-  ( module Linear.Core.Syntax
+module Linear.Core.Translate.Syntax
+  ( module Linear.Core.Translate.Syntax
 
   -- * Re-exports
   , pretty

### 2023-08-09 prototype/src/Linear/Core.hs
Start checking Core

@@ -77,6 +77,60 @@ computeRecUsageEnvs l =
     ) l l
 
 --------------------------------------------------------------------------------
+----- Typechecking Linear Core (mostly ignoring types) -------------------------
+--------------------------------------------------------------------------------
+
+checkProgram :: Monad m => CoreProgram -> m LCProgram
+checkProgram = traverse checkBind
+
+checkBind :: Monad m => CoreBind -> m LCBind
+checkBind (NonRec b e)
+  | isId b
+  = do e' <- checkExpr e
+       return (NonRec (LCVar b Nothing) e')
+  | otherwise
+  = do e' <- checkExpr e
+       -- This is really instanced to Nothing, since TyVars are not accounted for as linear resources.
+       return (NonRec (LCVar b Nothing) e')
+
+checkBind (Rec bs) = do
+  let (ids,rhss) = unzip bs
+  rhss' <- traverse checkExpr rhss
+  let ids' = L.map (`LCVar` Nothing) ids
+  return (Rec (zip ids' rhss'))
+
+checkExpr :: Monad m => CoreExpr -> m LCExpr
+checkExpr expr = case expr of
+  Var v       -> return (Var v)
+  Lit l       -> return (Lit l)
+  Type ty     -> return (Type ty)
+  Coercion co -> return (Coercion co)
+  App e1 e2   -> App <$> checkExpr e1 <*> checkExpr e2
+  Lam b e     -> Lam (LCVar b (multBinding b)) <$> checkExpr e
+  Let bind@(NonRec _b _) body
+    -> Let <$> checkBind bind
+           <*> checkExpr body
+  Let bind@(Rec _bs) body
+    -> Let <$> checkBind bind
+           <*> checkExpr body
+  Case e b ty alts -> do
+    Case <$> checkExpr e
+         <*> pure (LCVar b Nothing)
+         <*> pure ty
+         <*> mapM (checkAlt) alts
+
+  Tick t e  -> Tick t <$> checkExpr e
+  Cast e co -> Cast <$> checkExpr e <*> pure co
+
+checkAlt :: Monad m
+           => CoreAlt
+           -> m LCAlt
+checkAlt (Alt con args rhs) = do
+  rhs' <- checkExpr rhs
+  let args' = L.map (\a -> LCVar a Nothing) args
+  return (Alt con args' rhs')
+
+--------------------------------------------------------------------------------
 ----- Initial conversion to operate on LCVar binders ---------------------------
[…]

### 2023-08-09 prototype/src/Linear/Core/Monad.hs

@@ -77,3 +77,7 @@ unrestricted computation = LinearCtxT do
   return result
 
 
+-- | Run a computation and record the linear resources used in that computation
+record :: Monad m => LinearCtxT k v m a -> LinearCtxT k v m (a, [k])
+record computation = Linear
+

### 2023-08-14 prototype/src/Linear/Core.hs
Tweaks to implementation

@@ -138,10 +138,10 @@ checkExpr expr = case expr of
            <*> pure (LCVar b.id (deltaBinding ue))
            <*> pure ty
            <*> extend (b.id) (DeltaBound ue) (mapM (checkAlt ue) alts)
-    -- Expression is definitely not in WHNF (right? or do we mean HNF)
     | otherwise
+    -- Expression is definitely not in WHNF (right? or do we mean HNF)
     -> do
-      (e', ue) <- record $ checkExpr e
+      (e', makeIrrelevant -> ue) <- record $ checkExpr e
       Case <$> pure e'
            <*> pure (LCVar b.id (deltaBinding ue))
            <*> pure ty
@@ -171,14 +171,17 @@ checkAlt _ (Alt (LitAlt _) _ _) = error "impossible"
 checkAlt ue (Alt (DataAlt con) args rhs)
   | L.null $ L.filter (not . isManyTy . scaledMult) (dataConOrigArgTys con)
   = do
-  rhs' <- extends (L.map (\arg -> (arg.id, LambdaBound (Relevant ManyTy))) args) $ Linear.Core.Monad.drop ue $ checkExpr rhs
+  rhs' <- extends (L.map (\arg -> (arg.id, LambdaBound (Relevant ManyTy))) args)
+          $ Linear.Core.Monad.drop ue
+          $ checkExpr rhs
   return (Alt (DataAlt con) args rhs')
 
 --- * ALTN
 checkAlt ue (Alt (DataAlt con) args rhs) = do
-  -- TODO: Partition unrestricted and linear variables
-  rhs' <- checkExpr rhs
+  let (unrestricted_args, linear_args) = L.partition (isManyTy . scaledMult)
   -- TODO: We need to figure out how to typecheck alternatives (in the syntax directed form too) before we do this right.
+  rhs' <- extends (L.map (\arg -> (arg.id, LambdaBound (Relevant ManyTy))) unrestricted_args)
+          $ checkExpr rhs
   let args' = L.map (\a -> LCVar a.id (deltaBinding ue)) args
   return (Alt (DataAlt con) args' rhs')
 

### 2023-08-14 prototype/src/Linear/Core/Multiplicities.hs

@@ -55,6 +55,9 @@ newtype UsageEnv = UsageEnv [(Var,Mult)]
 -- INVARIANT: Two ocurrences of the same variable must have the same mult
   deriving Eq
 
+makeIrrelevant :: UsageEnv -> UsageEnv
+makeIrrelevant (UsageEnv ue) = UsageEnv $ M.map (\case Relevant m -> Irrelevant m; Irrelevant m -> Irrelevant m) ue
+
 lookupUE :: Var -> UsageEnv -> Usage
 lookupUE v (UsageEnv m) = case lookup v m of
                             Nothing   -> Zero

### 2023-08-22 laboratory/A.hs
Prototype tweaks

@@ -1,0 +1,7 @@
+{-# LANGUAGE LinearTypes #-}
+module A
+where
+
+f :: (a %1 -> a) -> a %1 -> a
+f use x = let y = use x in y
+

### 2023-08-22 prototype/src/Linear/Core.hs

@@ -27,6 +27,7 @@ import GHC.Core.TyCo.Rep (Type(..))
 import Linear.Core.Multiplicities
 import Linear.Core.Monad
 import GHC.Core.Multiplicity (scaledMult)
+import GHC.Core.Multiplicity (scaledThing)
 
 type LCProgram = [LCBind]
 type LCBind = Bind LCVar
@@ -68,7 +69,7 @@ runLinearCore pgr = do
 
     lcprg = runIdentity (convertProgram pgr)
 
-  case runExcept $ runLinearCoreT bindingsMap (checkProgram lcprg) of
+  case runIdentity $ runLinearCoreT bindingsMap (checkProgram lcprg) of
     Left e -> pprTraceM "Failed to typecheck linearity!" (ppr lcprg) >> return [text e]
     Right x -> pprTraceM "Safe prog!" (ppr x) >> return []
 
@@ -76,7 +77,7 @@ runLinearCore pgr = do
 -- {{{ Typechecking Linear Core (mostly ignoring types) ------------------------
 --------------------------------------------------------------------------------
 
-type LinearCoreM = LinearCoreT (Except String)
+type LinearCoreM = LinearCoreT Identity
 
 checkProgram :: LCProgram -> LinearCoreM LCProgram
 checkProgram = traverse checkBind
@@ -108,7 +109,7 @@ checkBind (Rec bs) = do
 
 checkExpr :: LCExpr -> LinearCoreM LCExpr
 checkExpr expr = case expr of
-  Var v       -> use v Nothing >> return (Var v)
+  Var v       -> use v >> return (Var v)
   Lit l       -> return (Lit l)
   Type ty     -> return (Type ty)
   Coercion co -> return (Coercion co)
@@ -188,7 +189,7 @@ checkAlt ue (Alt (DataAlt con) args rhs)
 checkAlt ue (Alt (DataAlt con) args rhs) = do
   let (unrestricted_args, linear_args) = L.partition (isManyTy . scaledMult) (dataConOrigArgTys con)
   -- TODO: We need to figure out how to typecheck alternatives (in the syntax directed form too) before we do this right.
-  rhs' <- extends (L.map (\arg -> (arg.id, LambdaBound (Relevant ManyTy))) unrestricted_args)
+  rhs' <- extends (L.map (\arg -> (scaledThing arg, LambdaBound (Relevant ManyTy))) unrestricted_args)
           $ checkExpr rhs
[…]

### 2023-08-22 prototype/src/Linear/Core/Monad.hs

@@ -63,11 +63,14 @@ import qualified Data.Map as M
 -- * The Reader, Bool, determines whether to record all variables used in the writer env.
 --    * This bool will only be true when recordNonLinear sets in order to record variables.
 --    * This ensures we only collect variables within recordNonLinear, after which we wipe the writer
-newtype LinearCoreT m a = LinearCoreT { unLC :: RWST Bool [Var] (M.Map Var (NonEmpty IdBinding)) m a }
-  deriving (Functor, Applicative, Monad, MonadTrans, MonadFail, MonadState (M.Map Var (NonEmpty IdBinding)))
+newtype LinearCoreT m a = LinearCoreT { unLC :: RWST Bool [Var] (M.Map Var (NonEmpty IdBinding)) (ExceptT String m) a }
+  deriving (Functor, Applicative, Monad, MonadState (M.Map Var (NonEmpty IdBinding)), MonadError String)
 
-runLinearCoreT :: (MonadError e m, IsString e) => M.Map Var IdBinding -> LinearCoreT m a -> m a
-runLinearCoreT s comp = do
+instance Monad m => MonadFail (LinearCoreT m) where
+  fail = throwError
+
+runLinearCoreT :: Monad m => M.Map Var IdBinding -> LinearCoreT m a -> m (Either String a)
+runLinearCoreT s comp = runExceptT $ do
   (a, renv, _unrestrictedUsed) <- runRWST (unLC comp) False (M.map (:| []) s)
   if M.null $ M.filter (all isBindingLinear) renv
     then return a
@@ -76,46 +79,47 @@ runLinearCoreT s comp = do
 
 -- | Uses a resource from the environment, making it unavailable in subsequent computations if it was/consumed a linear resource.
 -- If the key doesn't exist, no resources are consumed
---
+use :: Monad m => Var -> LinearCoreT m ()
+use = flip use' Nothing where
 -- The variable being used might be associated to a tag, in which case we must ensure the resources are split before consumed
 -- * The variable will be paired with a tag only in the usage environment of case pattern variables (we recursively invoke this function for usage envs, and that's when the tag exists)
 -- * To model this, we receive an explicit multiplicity that the variable should have as an argument
 -- * And try to instance a var with the requested multiplicity if it exists, or split a linear variable into multiple tagged vars otherwise
-use :: (MonadError e m, IsString e) => Var -> Maybe Mult -> LinearCoreT m ()
[…]

### 2023-08-22 prototype/src/Linear/Core/Translate/Syntax.hs

@@ -215,7 +215,6 @@ varId :: Var -> Id
 varId (Id' i) = i
 varId _ = error "varId: Not an Id"
 
-
 --------------------------------------------------------------------------------
 ----- Pretty-printing ----------------------------------------------------------
 --------------------------------------------------------------------------------

### 2023-08-31 prototype/src/Linear/Core.hs
Structure Linear Core Section!
And advertise it in the introduction to the section: foundations -> usage environments (lets) -> case expressions

@@ -14,7 +14,6 @@ module Linear.Core
 
 import GHC.Utils.Outputable
 import Control.Monad.State
-import Control.Monad.Except
 import GHC.Driver.Config.Core.Lint
 import GHC.Plugins hiding (Mult, count, unrestricted)
 import GHC.Utils.Trace
@@ -65,7 +64,7 @@ runLinearCore pgr = do
     -- The local top-level bindings have empty usage environments, and GlobalIds are treated as constants so we don't need to include them here
     -- See also Note [GlobalId/LocalId]
     bindingsMap = M.fromList $ L.map (,DeltaBound emptyUE) localTopBindings
-    defcfg = initLintConfig dflags localTopBindings
+    -- defcfg = initLintConfig dflags localTopBindings
 
     lcprg = runIdentity (convertProgram pgr)
 

### 2023-08-31 prototype/src/Linear/Core/Monad.hs

@@ -84,6 +84,7 @@ runLinearCoreT s comp = runExceptT $ do
     then return a
     else throwError "Not all linear resources in a computation were consumed"
 
+-- All this state management is a bit complex, i wonder if it would be simpler if defined in terms of e.g. finite automata
 
 -- | Uses a resource from the environment, making it unavailable in subsequent computations if it was/consumed a linear resource.
 -- If the key doesn't exist, no resources are consumed

### 2023-09-11 prototype/src/Linear/Core.hs
Major updates to writing and prototype i think

@@ -188,8 +188,7 @@ checkAlt _ (Alt (LitAlt _) _ _) = error "impossible"
 
 --- * ALT0
 checkAlt ue a@(Alt (DataAlt con) args rhs)
--- ROMES:Isto é fixe para mostrar o HLS
-  | L.null $ L.filter (not . isManyTy . scaledMult) (dataConOrigArgTys con)
+  | all (isManyTy . scaledMult) (dataConOrigArgTys con)
   = pprTrace "ALT0 con" (ppr a Ppr.<> Ppr.text ". (UE):" Ppr.<> ppr ue) $ do
           -- Add unrestricted binders
   rhs' <- extends (L.map (\arg -> (arg.id, LambdaBound (Relevant ManyTy))) args)

### 2023-09-11 prototype/test/Test.hs

@@ -12,6 +12,8 @@ import GHC.Plugins
 import Linear.Core.Plugin
 
 
+test :: a -> a
+test x = x
 
 main :: IO ()
 main = pure ()

### 2023-09-11 prototype/test/Translate/Test.hs

@@ -1,0 +1,139 @@
+{-# LANGUAGE OverloadedStrings #-}
+{-# OPTIONS_GHC -Wno-orphans #-}
+module Translate.Test where
+import Data.Either
+import qualified Data.Map as M
+import Error.Diagnose
+import qualified Data.Text as T
+
+import Data.Text (Text)
+import Test.Tasty
+import Test.Tasty.HUnit
+
+import Prettyprinter
+
+import Linear.Core.Translate.Syntax
+import Linear.Core.Translate.Parser
+import Linear.Core.Translate.Check
+import Linear.Core.Translate.Infer
+
+-- Let's try writing a simple program like id @A
+-- id :: ∀ p. (A p) ->p (A p)
+-- id = /\p. \x:_pA -> x:A
+idProg :: CoreTerm
+idProg
+  = Term
+    (M.singleton "MkA" (Id (Scheme "kp" (Datatype "A" [MV "p"])) (DeltaBound mempty) "MkA"))
+    (Lam (MultVar "p") $
+       Lam (Id (Datatype "A" [MV "p"]) (LambdaBound (MV "p")) "x") $
+         Var (Id (Datatype "A" [MV "p"]) (LambdaBound (MV "p")) "x")
+    )
+    (Scheme "p" (FunTy (Datatype "A" [MV "p"]) (MV "p") (Datatype "A" [MV "p"])))
+
+id2 :: CoreExpr
+id2 =
+ Lam (Id (Datatype "A" [MV "p"]) (LambdaBound (MV "p")) "x") $
+   Var (Id (Datatype "A" [MV "p"]) (LambdaBound (MV "p")) "x")
+
+idBad :: CoreExpr
+idBad =
+ Lam (Id (Datatype "A" [MV "p"]) (LambdaBound (MV "p")) "x") $
+   Var (Id (Datatype "A" [MV "p"]) (LambdaBound (MV "p")) "y")
+
+
+parsingTests :: TestTree
+parsingTests = testCase "Parsing some things" $ do
+
+  assertBool "Parse K @1" $ parses "K @1"
+  assertBool "Parse K @ω" $ parses "K @ω"
+  assertBool "Parse λp -> K @p" $ parses "λp -> K @p"
+  assertBool "Parse T4" $ parses "\\x -> case x of z {Nothing -> True; Just y -> y}"
+  assertBool "Parse T5" $ parses "\\p -> \\x -> x"
[…]

### 2023-09-12 language-v4/proofs/WHNFConvSoundness.tex
Organize optimisation proofs in the document

\WHNFConvSoundness

\begin{proof}
By structural induction on the typing derivation.

\begin{description}

\item[Case:] $\Lambda I$
\begin{tabbing}
    (1)
\end{tabbing}

\end{description}
\end{proof}

### 2023-09-12 language-v4/proofs/optimizations/BetaReduction.tex
Organize optimisation proofs in the document

\BetaReductionTheorem

\begin{proof}~

\begin{tabbing}
    Subcase $\pi = 1,p$\\
    (1) $\G,\G' \vdash (\lambda \x[1,p][\s].~e)~e' : \vp$\\
    (2) $\G \vdash (\lambda \x[1,p][\s].~e) : \s \to_{1,p} \vp$\` by inv. on $\lambda E_{1,p}$\\
    (3) $\G' \vdash e' : \s$ \` by inv. on $\lambda E_{1,p}$\\
    (4) $\G,\x[1,p][\s] \vdash e : \vp $\`by inv. on $\lambda I$\\
    (5) $\x[1,p][\s] \notin \G$\`by inv. on $\lambda I$\\
    (6) $\G,\G' \vdash e[e'/x] : \vp$\` by lin. subst. lemma (4,3) and (5)\\
    Subcase $\pi = \omega$\\
    (1) $\G,\G' \vdash (\lambda \xo.~e)~e' : \vp$\\
    (2) $\G \vdash (\lambda \xo.~e) : \s \to_\omega \vp$\` by inv. on $\lambda E_\omega$\\
    (3) $\G' \vdash e' : \s$ \` by inv. on $\lambda E_\omega$\\
    (4) $\hasnolinearvars{\G'}$ \` by inv. on $\lambda E_\omega$\\
    (5) $\G,\xo \vdash e : \vp$\` by inv. on $\lambda I$\\
    (6) $\xo \notin \G$\` by inv. on $\lambda I$\\
    (7) $\G,\G' \vdash e[e'/x] : \vp$\` by unr. subst. lemma (5,3,4)\\
\end{tabbing}
\end{proof}

\BetaReductionSharingTheorem

\begin{proof}~

\begin{tabbing}
    (1) $\G,\G' \vdash (\lambda \xo.~e)~e' : \vp$\\
    (2) $\G \vdash (\lambda \xo.~e) : \s \to_\omega \vp$\` by inv. on $\lambda E_\omega$\\
    (3) $\G', \cdot \vdash e' : \s$ \` by inv. on $\lambda E_\omega$\\
    (4) $\hasnolinearvars{\G'}$ \` by inv. on $\lambda E_\omega$\\
    (5) $\G,\xo \vdash e : \vp$\` by inv. on $\lambda I$\\
    (6) $\G, \cdot, \x[\cdot][\s] \vdash e : \vp$\` by $\xo = \x[\cdot][\s]$ lemma (TODO)\\
    (7) $\G,\G' \vdash \llet{x = e'}{e} : \vp$\` by let (6,3,4)\\
\end{tabbing}
\end{proof}

\BetaReductionMultTheorem

\begin{proof}
    Trivial through the progress theorem
\end{proof}

### 2023-09-12 language-v4/proofs/optimizations/CaseOfKnownConstructor.tex
Organize optimisation proofs in the document

\CaseOfKnownConstructorTheorem

\begin{proof}
    Trivial through the progress theorem.
\end{proof}

### 2023-09-12 language-v4/proofs/optimizations/Inlining.tex
Organize optimisation proofs in the document

\InliningTheorem

\begin{proof}~

\begin{tabbing}
    (1) $\G,\G',\D \vdash \llet{\xD = e}{e'} : \vp$\\
    (2) $\G,\D,\xD \vdash e' : \vp$\` by inv. on (let)\\
    (3) $\G',\D \vdash e : \s$\` by inv. on (let)\\
    (4) $\hasnolinearvars{\G'}$\\
    (5) $\G,\G',\D\vdash e'[e/x] : \vp$\` by $\D$-subst. lemma (2,3)\\
    (6) $\G,\G',\D,\xD \vdash e'[e/x] : \vp$ \` by $Weaken_\Delta$\\
    (7) $\G,\G',\G',\D \vdash \llet{\xD = e}{e'[e/x]} : \vp$\` by (let) (7,3,4)\\
    (8) $\G,\G',\D \vdash \llet{\xD = e}{e'[e/x]} : \vp$\` by (4)\\
\end{tabbing}
\end{proof}

### 2023-09-12 language-v4/proofs/WHNFConvSoundness.tex
Mre typesetting proofs

@@
 \begin{proof}
-By structural induction on the typing derivation.
+By structural induction on the case alternative typing derivation.
 
@@
 
-\item[Case:] $\Lambda I$
+\item[Case:] $Alt\_$
 \begin{tabbing}
-    (1)
+    (1) $\G,\z[\irr{\D}];\irr{\D}, \D' \vdash_{alt} \_ \to e :^z_{\irr{\D}} \s \Rightarrow \vp$\\
+    (2) $\G, \z[\irr{\D}];\irr{\D},\D' \vdash e : \vp$\\
+    (3) $\irr{\D}$ is used through $z$\`since $\irr{\D}$ can't otherwise be used\\\`and is introduced in this alternative uniquely \\\`(since we have multi-tier proof irrelevance, i.e.
+$\irr{\irr{\D}}\neq\irr{\D}$)\\
+    (4) $\G; \z[1], \D' \vdash e : \vp$\` by $\D \Rightarrow 1$ lemma (2,3)\\
+    (5) $\G, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash e : \vp$\` by $1 \Rightarrow \D$ lemma (4)\\
+    (6) $\G, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash_{alt} \_ \to e :^z_{\ov{\D_i}} \s \Rightarrow \vp$\` by $Alt\_$\\
+\end{tabbing}
+
+\item[Case:] $Alt0$
+\begin{tabbing}
+    (1) $\G, \z[\irr{\D}]; \irr{\D}, \D' \vdash_{alt} K~\ov{\xo}\to e :^z_{\irr{\D}} \s \Rightarrow \vp$\\
+    (2) $\G, \z[\cdot], \ov{\xo}; \D' \vdash e : \vp$\` by inv. on $Alt0$ and def. of empty subst.\\
+    (3) $\G, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash_{alt} K~\ov{\xo}\to e :^z_{\ov{\D_i}} \s \Rightarrow \vp$\` by $Alt0$ (2)\\
+\end{tabbing}
+
+\item[Case:] $AltN_{\textrm{WHNF}}$
+\begin{tabbing}
+    Not applicable since $\Rrightarrow$ is only generalized by $\Rightarrow$, not $\Mapsto$.
+\end{tabbing}
+
+\item[Case:] $AltN_{\textrm{Not WHNF}}$
+\begin{tabbing}
+    (1) $\G, \z[\irr{\D}]; \irr{\D}, \D' \vdash_{alt} K~\ov{\xo},\ov{y_i{:}_{1}\s_i}^n\to e :^z_{\irr{\D}} \s \Rightarrow \vp$\\
+    (2) $\G,\z[\irr{\D}],\ov{\xo},\ov{y_j{:}_{\D_j}\s_j}^n; \irr{\D}, \D' \vdash e : \vp$\`by inv. on $AltN_{\textrm{Not WHNF}}$\\
+    (3) $\ov{\D_j} = \ov{\lctag{\irr{\D}}{K_j}}$\`by inv. on $AltN_{\textrm{Not WHNF}}$\\
+    Subcase $\irr{\D}$ is consumed through $z$\\
+    (4) $\G,\ov{\xo},\ov{y_j{:}_{\D_j}\s_j}^n; \z[1], \D' \vdash e : \vp$\`by $\D \Rightarrow 1$ lemma (2,subcase)\\
+    (5) $\G,\ov{\xo},\ov{y_j{:}_{\D_i}\s_j}^n; \z[1], \D' \vdash e : \vp$\`by $\ov{y}$ does not consume resources,\\\` and Weaken? (HELP move $\D_j$ to $\D_i$)\\
+    (6) $\G,\ov{\xo},\ov{y_j{:}_{\D_i}\s_j}^n, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash e : \vp$\`by $1 \Rightarrow \D$ lemma (5)\\
+    (7) $\G, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash_{alt} K~\ov{\xo},\ov{y_i{:}_{1}\s_i}^n \to e :^z_{\ov{\D_i}} \s \Mapsto \vp$\`by $AltN_{\textrm{WHNF}}$ (6)\\
+    Subcase $\irr{\D}$ is (fully) consumed by $\ov{y}$ (after splitting)\\
+    (4) $\G,\z[\irr{\D}],\ov{\xo};\ov{y_j{:}_{1}\s_j}^n, \D' \vdash e : \vp$\`by $\D \Rightarrow 1$ lemma (2, subcase)\\
+    (5) $\G,\z[\ov{\D_i}],\ov{\xo};\ov{y_j{:}_{1}\s_j}^n, \D' \vdash e : \vp$\`by Weaken and $z$ does not consume resources (HELP SAME KIND)\\
+    (6) $\G,\z[\ov{\D_i}],\ov{\xo},\ov{y_j{:}_{\D_i}\s_j}^n; \ov{\D_i}, \D' \vdash e : \vp$\`by $1 \Rightarrow \D$ lemma (5)\\
+    (7) $\G,\z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash K~\ov{\xo},\ov{y_i{:}_{1}\s_i}^n\to e :^z_{\ov{\D_i}} \s \Mapsto \vp$\`by $AltN_{\textrm{WHNF}}$\\
 \end{tabbing}

### 2023-09-12 language-v4/proofs/optimizations/BetaReduction.tex
Mre typesetting proofs
Improve assumptions, finish beta reduction

@@
 \begin{tabbing}
-    Subcase $\pi = 1,p$\\
-    (1) $\G,\G' \vdash (\lambda \x[1,p][\s].~e)~e' : \vp$\\
-    (2) $\G \vdash (\lambda \x[1,p][\s].~e) : \s \to_{1,p} \vp$\` by inv. on $\lambda E_{1,p}$\\
-    (3) $\G' \vdash e' : \s$ \` by inv. on $\lambda E_{1,p}$\\
-    (4) $\G,\x[1,p][\s] \vdash e : \vp $\`by inv. on $\lambda I$\\
-    (5) $\x[1,p][\s] \notin \G$\`by inv. on $\lambda I$\\
-    (6) $\G,\G' \vdash e[e'/x] : \vp$\` by lin. subst. lemma (4,3) and (5)\\
-    Subcase $\pi = \omega$\\
-    (1) $\G,\G' \vdash (\lambda \xo.~e)~e' : \vp$\\
-    (2) $\G \vdash (\lambda \xo.~e) : \s \to_\omega \vp$\` by inv. on $\lambda E_\omega$\\
-    (3) $\G' \vdash e' : \s$ \` by inv. on $\lambda E_\omega$\\
-    (4) $\hasnolinearvars{\G'}$ \` by inv. on $\lambda E_\omega$\\
-    (5) $\G,\xo \vdash e : \vp$\` by inv. on $\lambda I$\\
-    (6) $\xo \notin \G$\` by inv. on $\lambda I$\\
-    (7) $\G,\G' \vdash e[e'/x] : \vp$\` by unr. subst. lemma (5,3,4)\\
+% Using preservation
+    (1) $\G; \D \vdash (\lambda \xl.~e)~e' : \vp$\\
+    (2) $(\lambda \xl.~e) e' \longrightarrow e[e'/x]$\\
+    (3) $\G; \D \vdash e[e'/x] : \vp$ \` by type preservation theorem (1,2)\\
+
+% Manually
+%    Subcase $\pi = 1,p$\\
+%    (1) $\G,\G' \vdash (\lambda \x[1,p][\s].~e)~e' : \vp$\\
+%    (2) $\G \vdash (\lambda \x[1,p][\s].~e) : \s \to_{1,p} \vp$\` by inv. on $\lambda E_{1,p}$\\
+%    (3) $\G' \vdash e' : \s$ \` by inv. on $\lambda E_{1,p}$\\
+%    (4) $\G,\x[1,p][\s] \vdash e : \vp $\`by inv. on $\lambda I$\\
+%    (5) $\x[1,p][\s] \notin \G$\`by inv. on $\lambda I$\\
+%    (6) $\G,\G' \vdash e[e'/x] : \vp$\` by lin. subst. lemma (4,3) and (5)\\
+%    Subcase $\pi = \omega$\\
+%    (1) $\G,\G' \vdash (\lambda \xo.~e)~e' : \vp$\\
+%    (2) $\G \vdash (\lambda \xo.~e) : \s \to_\omega \vp$\` by inv. on $\lambda E_\omega$\\
+%    (3) $\G' \vdash e' : \s$ \` by inv. on $\lambda E_\omega$\\
+%    (4) $\hasnolinearvars{\G'}$ \` by inv. on $\lambda E_\omega$\\
+%    (5) $\G,\xo \vdash e : \vp$\` by inv. on $\lambda I$\\
+%    (6) $\xo \notin \G$\` by inv. on $\lambda I$\\
+%    (7) $\G,\G' \vdash e[e'/x] : \vp$\` by unr. subst. lemma (5,3,4)\\
 \end{tabbing}
@@
 \begin{tabbing}
-    (1) $\G,\G' \vdash (\lambda \xo.~e)~e' : \vp$\\
-    (2) $\G \vdash (\lambda \xo.~e) : \s \to_\omega \vp$\` by inv. on $\lambda E_\omega$\\
-    (3) $\G', \cdot \vdash e' : \s$ \` by inv. on $\lambda E_\omega$\\
-    (4) $\hasnolinearvars{\G'}$ \` by inv. on $\lambda E_\omega$\\
-    (5) $\G,\xo \vdash e : \vp$\` by inv. on $\lambda I$\\
-    (6) $\G, \cdot, \x[\cdot][\s] \vdash e : \vp$\` by $\xo = \x[\cdot][\s]$ lemma (TODO)\\
-    (7) $\G,\G' \vdash \llet{x = e'}{e} : \vp$\` by let (6,3,4)\\
+    (1) $\G;\D \vdash (\lambda \xo.~e)~e' : \vp$\\
+    (2) $\G; \D \vdash (\lambda \xo.~e) : \s \to_\omega \vp$\` by inv. on $\lambda E_\omega$\\
+    (3) $\G; \cdot \vdash e' : \s$ \` by inv. on $\lambda E_\omega$\\
+    (4) $\G,\xo; \D \vdash e : \vp$\` by inv. on $\lambda I$\\
+    (5) $\G, \x[\cdot][\s]; \D \vdash e : \vp$\` by $\xo = \x[\cdot][\s]$ lemma\\
+    (6) $\G,\G' \vdash \llet{x = e'}{e} : \vp$\` by let (5,3)\\
 \end{tabbing}
@@
 \begin{proof}
-    Trivial through the progress theorem
+    Trivial by invoking preservation using the $\Lambda$ application reduction and the assumption
 \end{proof}

### 2023-09-13 language-v4/proofs/optimizations/EtaConvs.tex
Write and typeset multiple optimisation proofs

\EtaExpansionTheorem

\begin{proof}~
\begin{tabbing}
    Subcase $f$ is linear\\
    (1) $\G; \D \vdash f : \s \lolli \vp$\\
    (2) $\G; \xl \vdash x : \s$\\
    (3) $\G; \D, \xl \vdash f~x : \vp$\`by $\lambda E$\\
    (4) $\G; \D \vdash (\lambda \xl.~f~x) : \s \lolli \vp$\`by $\lambda I$\\
    Subcase $f$ is unrestricted\\
    As above but $x$ is introduced in $\G$ and functions are unrestricted
\end{tabbing}
\end{proof}

\EtaReductionTheorem

\begin{proof}~
\begin{tabbing}
    (1) $\G; \D \vdash (\lambda \x[\pi].~f~x) : \s \to_\pi \vp$\\
    Subcase $\pi = 1$\\
    (2) $\G; \D, \xl \vdash f~x : \vp$\`by inv. on $\lambda I$\\
    (3) $\G; \D \vdash f : \s \to_1 \vp$\`by inv. on $\lambda E$\\
    Subcase $\pi = \omega$\\
    As above but $x$ is introduced in $\G$
\end{tabbing}
\end{proof}

### 2023-09-14 language-v4/proofs/WHNFConvSoundness.tex
More work, improvements, corrections

--- 2023-09-12 language-v4/proofs/WHNFConvSoundness.tex
@@
 
-\item[Case:] $\Lambda I$
+\item[Case:] $Alt\_$
 \begin{tabbing}
-    (1)
+    (1) $\G,\z[\irr{\D}];\irr{\D}, \D' \vdash_{alt} \_ \to e :^z_{\irr{\D}} \s \Rightarrow \vp$\\
+    (2) $\G, \z[\irr{\D}];\irr{\D},\D' \vdash e : \vp$\\
+    (3) $\irr{\D}$ is used through $z$\`since $\irr{\D}$ can't otherwise be used\\\`and is introduced in this alternative uniquely \\\`(since we have multi-tier proof irrelevance, i.e.
+$\irr{\irr{\D}}\neq\irr{\D}$)\\
+    (4) $\G; \z[1], \D' \vdash e : \vp$\` by $\D \Rightarrow 1$ lemma (2,3)\\
+    (5) $\G, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash e : \vp$\` by $1 \Rightarrow \D$ lemma (4)\\
+    (6) $\G, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash_{alt} \_ \to e :^z_{\ov{\D_i}} \s \Rightarrow \vp$\` by $Alt\_$\\
+\end{tabbing}
+
+\item[Case:] $Alt0$
+\begin{tabbing}
+    (1) $\G, \z[\irr{\D}]; \irr{\D}, \D' \vdash_{alt} K~\ov{\xo}\to e :^z_{\irr{\D}} \s \Rightarrow \vp$\\
+    (2) $\G, \z[\cdot], \ov{\xo}; \D' \vdash e : \vp$\` by inv. on $Alt0$ and def. of empty subst.\\
+    (3) $\G, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash_{alt} K~\ov{\xo}\to e :^z_{\ov{\D_i}} \s \Rightarrow \vp$\` by $Alt0$ (2)\\
+\end{tabbing}
+
+\item[Case:] $AltN_{\textrm{WHNF}}$
+\begin{tabbing}
+    Not applicable since $\Rrightarrow$ is only generalized by $\Rightarrow$, not $\Mapsto$.
+\end{tabbing}
+
+\item[Case:] $AltN_{\textrm{Not WHNF}}$
+    We prove the theorem for constructing both an $AltN_{\textrm{Not WHNF}}$
+        and a $AltN_{\textrm{WHNF}}$ from a proof-irrelevant $AltN_{\textrm{Not
+        WHNF}}$, to prove the statement holds for any $\Rightarrow$ kind for
+        $AltN$, rather than just $\Rrightarrow$ or $\Mapsto$.
+\begin{tabbing}
+    (1) $\G, \z[\irr{\D}]; \irr{\D}, \D' \vdash_{alt} K~\ov{\xo},\ov{y_i{:}_{1}\s_i}^n\to e :^z_{\irr{\D}} \s \Rrightarrow \vp$\\
+    (2) $\G,\z[\irr{\D}],\ov{\xo},\ov{y_j{:}_{\D_j}\s_j}^n; \irr{\D}, \D' \vdash e : \vp$\`by inv. on $AltN_{\textrm{Not WHNF}}$\\
+    (3) $\ov{\D_j} = \ov{\lctag{\irr{\D}}{K_j}}$\`by inv. on $AltN_{\textrm{Not WHNF}}$\\
+    Subcase $\irr{\D}$ is consumed through $z$\\
+    (4) $\G,\ov{\xo},\ov{y_j{:}_{\D_j}\s_j}^n; \z[1], \D' \vdash e : \vp$\`by $\D \Rightarrow 1$ lemma (2,subcase)\\
+    Subcase constructing $AltN_{\textrm{WHNF}}$ ($\Mapsto$)\\
+    (5) $\G,\ov{\xo},\ov{y_j{:}_{\D_i}\s_j}^n; \z[1], \D' \vdash e : \vp$\`by $\ov{y}$ does not consume resources, and $Weaken$\\
+    (6) $\G,\ov{\xo},\ov{y_j{:}_{\D_i}\s_j}^n, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash e : \vp$\`by $1 \Rightarrow \D$ lemma (5)\\
+    (7) $\G, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash_{alt} K~\ov{\xo},\ov{y_i{:}_{1}\s_i}^n \to e :^z_{\ov{\D_i}} \s \Mapsto \vp$\`by $AltN_{\textrm{WHNF}}$ (6)\\
+    Subcase constructing $AltN_{\textrm{Not WHNF}}$ ($\Rrightarrow$)\\
+    (5) $\G,\ov{\xo},\ov{y_j{:}_{\lctag{\ov{\D_i}}{K_j}}\s_j}^n; \z[1], \D' \vdash e : \vp$\\\`by $\ov{y}$ does not consume resources, and $Weaken$\\
+    (6) $\G,\ov{\xo},\ov{y_j{:}_{\lctag{\ov{\D_i}}{K_j}}\s_j}^n, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash e : \vp$\`by $1 \Rightarrow \D$ lemma (5)\\
+    (7) $\G, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash_{alt} K~\ov{\xo},\ov{y_i{:}_{1}\s_i}^n \to e :^z_{\ov{\D_i}} \s \Rrightarrow \vp$\`by $AltN_{\textrm{Not WHNF}}$ (6)\\
+    Subcase $\irr{\D}$ is (fully) consumed by $\ov{y}$ (after splitting)\\
+    (4) $\G,\z[\irr{\D}],\ov{\xo};\ov{y_j{:}_{1}\s_j}^n, \D' \vdash e : \vp$\`by $\D \Rightarrow 1$ lemma (2, subcase)\\
+    (5) $\G,\z[\ov{\D_i}],\ov{\xo};\ov{y_j{:}_{1}\s_j}^n, \D' \vdash e : \vp$\`by $Weaken$ and $z$ does not consume resources\\
+    Subcase constructing $AltN_{\textrm{WHNF}}$ ($\Mapsto$)\\
+    (6) $\G,\z[\ov{\D_i}],\ov{\xo},\ov{y_j{:}_{\D_i}\s_j}^n; \ov{\D_i}, \D' \vdash e : \vp$\`by $1 \Rightarrow \D$ lemma (5)\\
+    (7) $\G,\z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash K~\ov{\xo},\ov{y_i{:}_{1}\s_i}^n\to e :^z_{\ov{\D_i}} \s \Mapsto \vp$\`by $AltN_{\textrm{WHNF}}$\\
+    Subcase constructing $AltN_{\textrm{Not WHNF}}$ ($\Rrightarrow$)\\
+    (6) $\G,\z[\ov{\D_i}],\ov{\xo},\ov{y_j{:}_{\lctag{\ov{\D_i}}{K_j}}\s_j}^n; \ov{\D_i}, \D' \vdash e : \vp$\`by $1 \Rightarrow \D$ lemma (5)\\
+    (7) $\G,\z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash K~\ov{\xo},\ov{y_i{:}_{1}\s_i}^n\to e :^z_{\ov{\D_i}} \s \Rrightarrow \vp$\`by $AltN_{\textrm{Not WHNF}}$\\
 \end{tabbing}

### 2023-09-20 prototype/app/Main.hs
Updates; comment translate modules

@@ -1,13 +1,17 @@
-{-# LANGUAGE LinearTypes, UnicodeSyntax, BlockArguments #-}
+{-# LANGUAGE LinearTypes, UnicodeSyntax, BlockArguments, OverloadedStrings #-}
 module Main where
-
+import GHC.Core.Opt.Monad
+import GHC.Driver.Main (initHscEnv)
+import GHC.Plugins
+import Linear.Core
+import Data.String
 
 {- Examples from the thesis -}
 
 {- Lets -}
 
-f1 :: (a ⊸ b) -> a ⊸ b
-f1 use x = let y = use x in y
+-- f1 :: (a ⊸ b) -> a ⊸ b
+-- f1 use x = let y = use x in y
 
 {- Recursive lets -}
 
@@ -19,4 +23,35 @@ f1 use x = let y = use x in y
 -- possible, and `-fdefer-type-errors` makes the program useless, only having
 -- runtime type errors...
 main :: IO ()
-main = putStrLn "Hello, Haskell!"
+main = do
+  putStrLn "Hello, Haskell!"
+
+  hscEnv <- initHscEnv Nothing
+  let
+    modu
+      = mkModule (stringToUnit "linear-core-tests-unit") (mkModuleName "Linear.Core.Tests")
+                 
+
+  print . map showPprUnsafe . fst =<<
+    runCoreM hscEnv
+             mempty
+             's'
+             modu
+             neverQualify -- or alwaysQualify
+             noSrcSpan
+             mainLinearCoreTest
+
+mainLinearCoreTest :: CoreM [SDoc]
+mainLinearCoreTest
+  = runLinearCore
+    [ f1
+    -- , f2
+    ]
+
+f1 :: CoreBind
+f1 = NonRec 
+
+
+mkId :: String -> Mult -> Type -> Id
+mkId s = mkLocalId (mkInternalName (mkVarOcc s) noSrcSpan)
+

### 2023-09-20 prototype/src/Linear/Core.hs

@@ -9,8 +9,9 @@ using a different reduced syntax
 
 -}
 module Linear.Core
-  ( runLinearCore
-  ) where
+  -- ( runLinearCore
+  -- )
+  where
 
 -- import GHC.Driver.Config.Core.Lint
 import Control.Monad.State
@@ -26,9 +27,7 @@ import GHC.Utils.Trace
 import Linear.Core.Multiplicities
 import Linear.Core.Monad
 import GHC.Core.Multiplicity (scaledMult)
-import GHC.Core.Multiplicity (scaledThing)
 import Data.Bifunctor
-import qualified Data.List.NonEmpty as NE
 
 type LCProgram = [LCBind]
 type LCBind = Bind LCVar
@@ -74,7 +73,7 @@ runLinearCore pgr = do
 
   case runIdentity $ runLinearCoreT bindingsMap (checkProgram lcprg) of
     Left e -> do
-      pprTraceM "Failed to typecheck linearity!" (ppr lcprg)
+      pprTraceM "Failed to typecheck linearity!" (ppr pgr)
       return [text e]
     Right x -> do
       pprTraceM "Safe prog!" (ppr x)
@@ -309,7 +308,7 @@ convertBind (NonRec b e)
 convertBind (Rec bs) = do
   let (ids,rhss) = unzip bs
   rhss' <- traverse convertExpr rhss
-  let ids' = L.map (`LCVar` (deltaBinding emptyUE)) ids
+  let ids' = L.map (`LCVar` deltaBinding emptyUE) ids
   return (Rec (zip ids' rhss'))
 
 
@@ -376,7 +375,7 @@ unconvertBind (Rec bs) =
   let (ids,rhss) = unzip bs
       rhss' = L.map unconvertExpr rhss
       ids' = L.map (.id) ids
-   in (Rec (zip ids' rhss'))
+   in Rec (zip ids' rhss')
 
 
 unconvertExpr :: LCExpr -> CoreExpr
@@ -392,7 +391,7 @@ unconvertExpr expr = case expr of
            (unconvertExpr body)
   Case e b ty alts -> do
     Case (unconvertExpr e)
-         (b.id)
+         b.id
          ty
          (L.map unconvertAlt alts)
 
@@ -403,7 +402,7 @@ unconvertAlt :: LCAlt -> CoreAlt
 unconvertAlt (Alt con args rhs) =
   let rhs' = unconvertExpr rhs
[…]

### 2023-09-20 prototype/src/Linear/Core/Multiplicities.hs

@@ -167,8 +167,8 @@ instance Outputable IdBinding where
   ppr (DeltaBound an) = text "Δ=" GHC.Plugins.<> ppr an
 
 instance Outputable Mult where
-  ppr (Relevant m) = text "Relevant" <+> ppr m
-  ppr (Irrelevant m) = text "Irrelevant" <+> ppr m
+  ppr (Relevant m) = text "@" GHC.Utils.Outputable.<> ppr m
+  ppr (Irrelevant m) = text "[@" GHC.Utils.Outputable.<> ppr m GHC.Utils.Outputable.<> text "]"
   ppr (Tagged t m) = ppr m GHC.Utils.Outputable.<> text "#" GHC.Utils.Outputable.<> ppr t
 
 instance Outputable Tag where

### 2023-09-20 prototype/src/Linear/Core/Plugin.hs

@@ -26,7 +26,7 @@ import Data.Functor hiding (unzip)
 import Data.Void
 import Data.Text (Text)
 import qualified Data.Text as T
-import Prettyprinter
+-- import Prettyprinter
 
 import GHC.Core.TyCon
 import GHC.Core.TyCo.Rep
@@ -36,16 +36,16 @@ import Control.Monad
 import Control.Monad.Reader
 import Control.Monad.Except
 
-import qualified Linear.Core.Translate.Syntax as LC
-import qualified Linear.Core.Translate.Check as LC
+-- import qualified Linear.Core.Translate.Syntax as LC
+-- import qualified Linear.Core.Translate.Check as LC
 import GHC.Core.Map.Type
 import Data.Maybe
 
 import Linear.Core
 import qualified Data.List as L
 
-type CoreCheck = ReaderT CoreProgram (ExceptT (Doc Void) CoreM)
-type CoreConv  = ReaderT (VarEnv LC.Var) CoreCheck
+-- type CoreCheck = ReaderT CoreProgram (ExceptT (Doc Void) CoreM)
+-- type CoreConv  = ReaderT (VarEnv LC.Var) CoreCheck
 
 plugin :: Plugin
 plugin = defaultPlugin { installCoreToDos = install }
@@ -78,300 +78,294 @@ linearCorePass guts = do
 --------------------------------------------------------------------------------
 -- Everything below this line is unused, but still typechecks
 
-validateLinearGuts :: ModGuts -> CoreM ModGuts
-validateLinearGuts guts = do
-  let prog = mg_binds guts
+--validateLinearGuts :: ModGuts -> CoreM ModGuts
+--validateLinearGuts guts = do
+--  let prog = mg_binds guts
 
-  msgs <- runExceptT (runReaderT (forM_ prog validateBind) prog)
-  case msgs of
-    Right () -> pure ()
-    Left e -> fatalErrorMsgS (show e)
+--  msgs <- runExceptT (runReaderT (forM_ prog validateBind) prog)
+--  case msgs of
+--    Right () -> pure ()
+--    Left e -> fatalErrorMsgS (show e)
 
-  return guts -- unchanged guts, after validating them.
+--  return guts -- unchanged guts, after validating them.
 
--- LC.Binder and the type of the bound expression
---
--- Perhaps we should guarantee the bindings are already strongly connected
-validateBind :: CoreBind -> CoreCheck () 
-validateBind b = do
-  printP b
[…]

### 2023-09-20 prototype/app/Main.hs
Fix one implementation bug, some writing

@@ -49,9 +49,9 @@ mainLinearCoreTest
     ]
 
 f1 :: CoreBind
-f1 = NonRec 
+f1 = NonRec _ _
 
 
 mkId :: String -> Mult -> Type -> Id
-mkId s = mkLocalId (mkInternalName (mkVarOcc s) noSrcSpan)
+mkId s = mkLocalId (mkInternalName 0 (mkVarOcc s) noSrcSpan)
 

### 2023-09-20 prototype/src/Linear/Core.hs

@@ -170,7 +170,7 @@ checkExpr expr = case expr of
       Case e'
            (LCVar b.id (deltaBinding ue))
            ty
-       <$> withSameEnvMap (extend b.id (DeltaBound ue) . checkAlt ue) alts
+       <$> withSameEnvMap (extend b.id (DeltaBound ue) . checkAlt (makeIrrelevant ue)) alts
 
   Tick t e  -> Tick t <$> checkExpr e
   Cast e co -> Cast <$> checkExpr e <*> pure co
@@ -215,12 +215,11 @@ checkAlt ue alt@(Alt (DataAlt con) args rhs) = pprTrace "ALTN con" (ppr alt) $ d
 
   let (unrestricted_args, linear_args) = bimap (L.map snd) (L.map snd) $
                                           L.partition (isManyTy . scaledMult . fst) (zip (dataConOrigArgTys con) args)
-  -- TODO: We need to figure out how to typecheck alternatives (in the syntax directed form too) before we do this right.
-
 
   -- Add the tag the usage environment with the linear resources with this constructor and an index for each
   -- It will ensure that when we consume the resources by using this environment, we'll just split the resource according to the tag.
   let linear_args' = L.zipWith (\a i -> (a.id, deltaBindingTagged con i ue)) linear_args [1..]
+  pprTraceM "Linear args in ALTN con:" (ppr linear_args')
 
           -- First, extend computation with unrestricted resources
   rhs' <- extends (L.map ((, LambdaBound (Relevant ManyTy)) . (.id)) unrestricted_args)
@@ -411,7 +410,7 @@ unconvertAlt (Alt con args rhs) =
 
 instance Outputable LCVar where
   ppr (LCVar id' Nothing)  = ppr id'
-  ppr (LCVar id' (Just b)) = ppr id' -- <+> ppr b
+  ppr (LCVar id' (Just b)) = ppr id' <+> ppr b
 
 instance OutputableBndr LCVar where
   pprPrefixOcc = ppr

### 2023-09-20 prototype/src/Linear/Core/Monad.hs

@@ -123,7 +123,6 @@ use = flip use' [] where
   use' :: Monad m => Var -> [Tag] -> LinearCoreT m ()
   use' key _    | isGlobalId key = pure () -- See Note [GlobalId/LocalId]; Global ids are top-level imported Ids, which are unrestricted
   use' key mtags = LinearCoreT do
-
     -- When recording, we count when linear resources are used, and allow them
     -- to be consumed more than once (despite that not being possible if we were
     -- typechecking).
@@ -131,6 +130,9 @@ use = flip use' [] where
 
     -- Lookup the variable in the environment
     mv <- gets (M.lookup key)
+
+    pprTraceM "Using var" (Ppr.ppr key Ppr.<+> Ppr.text "with mult" Ppr.<+> Ppr.ppr mv Ppr.<+> Ppr.text "and tag environment" Ppr.<+> Ppr.ppr mtags)
+
     case mv of
 
       -- Variable has already been consumed (it should be guaranteed to be in scope since Core programs are well-typed)
@@ -168,6 +170,7 @@ use = flip use' [] where
             -- given tag (in the form of a tagstack)
             tags -> do
               splits <- splitAsNeededThenConsume allowsIrrelevant tags mult
+              pprTraceM "Splitting key at tags" (Ppr.ppr key Ppr.<+> Ppr.ppr tags Ppr.<+> Ppr.ppr splits)
               case NE.nonEmpty splits of
                     Nothing
                       | isDryRun  -> tell [key]
@@ -209,6 +212,7 @@ use = flip use' [] where
           tags -> do
 
             splits <- join <$> mapM (splitAsNeededThenConsume allowsIrrelevant tags) (NE.toList mults)
+            pprTraceM "Split key at tags" (Ppr.ppr key Ppr.<+> Ppr.ppr tags Ppr.<+> Ppr.ppr splits)
             case NE.nonEmpty splits of
                 Nothing
                   | isDryRun  -> tell [key] -- Again, we don't consume things when dry run

### 2023-09-20 prototype/src/Linear/Core/Multiplicities.hs

@@ -93,6 +93,13 @@ removeTag m = m
 -- handled outside of this function, here they are split just like relevant ones
 splitAsNeededThenConsume :: MonadError String m
                          => AllowIrrelevant -> [Tag] -> Mult -> m [Mult]
+splitAsNeededThenConsume allowsIrrelevant tagstack m
+  | DisallowIrrelevant <- allowsIrrelevant
+  , isIrrelevant m = throwError "Tried to use an irrelevant result when not allowed"
+
+  | extractTags m == tagstack
+  = pure [] -- consume successfully since the mult tags matches exactly the tagstack
+
 splitAsNeededThenConsume allowsIrrelevant (Tag con i:ts) m
   = join <$> traverse (\case
                  m'@(Tagged (Tag _ i') _)

### 2023-09-22 language-v4/proofs/optimizations/BetaReduction.tex
Write on more optimisations

@@
+$\beta$-reduction evaluates a $\lambda$-abstraction application by substituting
+the $\lambda$-bound variable $x$ with the argument $e'$ in the body of the
+$\lambda$-abstraction $e$.
+%
+We consider two definitions of $\beta$-reductions, one that substitutes all
+occurrences of a variable by an expression, as in call-by-name, and other which
+creates a lazy let binding to share the result of computing the argument
+expression amongst uses of the variable, as in call-by-need.
+
+The first kind of $\beta$-reduction on term abstractions can be seen to
+preserve linearity by case analysis. When the function is linear, the binding
+$x$ is used exactly once in the body of the lambda, thus can be substituted by
+an expression typed with linear resources, since the expression is guaranteed
+to be used exactly once in place of $x$. The proof is direct by type preservation.
+
 \BetaReductionTheorem
@@
 % Using preservation
-    (1) $\G; \D \vdash (\lambda \xl.~e)~e' : \vp$\\
-    (2) $(\lambda \xl.~e) e' \longrightarrow e[e'/x]$\\
+    (1) $\G; \D \vdash (\lambda \x[\pi].~e)~e' : \vp$\\
+    (2) $(\lambda \x[\pi].~e) e' \longrightarrow e[e'/x]$\\
     (3) $\G; \D \vdash e[e'/x] : \vp$ \` by type preservation theorem (1,2)\\
@@
 
+\noindent We assume the $\beta$-reduction with sharing (i.e. the one that creates a let
+binding) is only applicable when the $\lambda$-abstraction has an unrestricted
+function type. Otherwise, the call-by-name $\beta$-reduction is always
+favourable, as we know the resource to be used exactly once and hence sharing
+would be counterproductive, and result in an unnecessary heap allocation.
+%
+Consequently, the argument to the function must be unrestricted (hence use no
+linear resources) for the term to be well-typed, and so it vacuously follows
+that linearity is preserved by this transformation.
+
 \BetaReductionSharingTheorem
@@
 
+\noindent Finally, $\beta$-reduction on multiplicity abstractions is also type
+preserving. The argument of the application is a multiplicity rather than an
+expression, so no resources are needed to type it, and since the body of the
+lambda must treat the multiplicity $p$ as though it were a linear, the body
+uses the argument linearly regardless of the instantiation of $p$ at $\pi$.
+
 \BetaReductionMultTheorem

### 2023-09-22 language-v4/proofs/optimizations/CaseOfKnownConstructor.tex
Write and typeset multiple optimisation proofs
Write on more optimisations

@@
+Case-of-known constructor is a transformation that essentially evaluates a case
+expression of a known constructor at compile time, by substituting in the
+matching alternative the case binder by the scrutinee and pattern variables by
+constructor arguments.
+%
+Intuitively, either the case binder is used exactly once to consume the
+resources of the scrutinee, or the pattern components whose matching
+constructor argument uses linear resources are used exactly once, so one of the
+substitutions is a no-op.  If the pattern variables are substituted by the
+matching scrutinee expressions, the expressions are still only used once, and
+if the case binder is substituted by the scrutinee, it is still used exactly
+once.
+%
+The proof follows trivially from the preservation theorem.
+
 \CaseOfKnownConstructorTheorem
 
-\begin{proof}
-    Trivial through the progress theorem.
+\begin{proof}~
+\begin{tabbing}
+    (1) $\G; \D, \D' \vdash \ccase{K~\ov{e}}{\z[\D][\s]~\{..., K~\ov{x} \to e_i\}} : \vp$\\
+    (2) $\ccase{K~\ov{e}}{\z[\D][\s]~\{..., K~\ov{x} \to e_i\}} \longrightarrow e_i\ov{[e/x]}[K~\ov{e}/z]$\\
+    (3) $\G; \D,\D' \vdash e_i\ov{[e/x]}[K~\ov{e}/z] : \vp$\` by preservation theorem (1,2)\\
+\end{tabbing}
 \end{proof}

### 2023-09-22 language-v4/proofs/optimizations/EtaConvs.tex
Write on more optimisations

@@
+The $\eta$-conversion transformations are $\eta$-expansion and
+$\eta$-reduction. In both transformations, linearity is preserved since the
+resources used to type the function $f$ do not change neither when the lambda
+and its argument are removed, nor when we add a lambda and apply $f$ to the
+bound argument.
+
 \EtaExpansionTheorem

### 2023-09-22 language-v4/proofs/optimizations/Inlining.tex
Mre typesetting proofs
Write on more optimisations

@@
+Inlining substitutes occurrences of a let-bound $\D$-variable $x$
+with the expression $e$ it is bound to, which determines its usage environment
+$\D$. Intuitively, in the let body $e'$, $x$ can occur once or not at all: if
+$x$ occurs, then the linear resources $\D$ used indirectly through $x$ are used
+via $e$ instead; if $x$ does not occur, then the resources $\D$ are already
+used linearly in $e'$ and the substitution is a no-op.
+
 \InliningTheorem
@@
 \begin{tabbing}
-    (1) $\G,\G',\D \vdash \llet{\xD = e}{e'} : \vp$\\
-    (2) $\G,\D,\xD \vdash e' : \vp$\` by inv. on (let)\\
-    (3) $\G',\D \vdash e : \s$\` by inv. on (let)\\
-    (4) $\hasnolinearvars{\G'}$\\
-    (5) $\G,\G',\D\vdash e'[e/x] : \vp$\` by $\D$-subst. lemma (2,3)\\
-    (6) $\G,\G',\D,\xD \vdash e'[e/x] : \vp$ \` by $Weaken_\Delta$\\
-    (7) $\G,\G',\G',\D \vdash \llet{\xD = e}{e'[e/x]} : \vp$\` by (let) (7,3,4)\\
-    (8) $\G,\G',\D \vdash \llet{\xD = e}{e'[e/x]} : \vp$\` by (4)\\
+    (1) $\G;\D,\D' \vdash \llet{\xD = e}{e'} : \vp$\\
+    (2) $\G,\D \vdash e : \s$\` by inv. on (let)\\
+    (3) $\G,\xD; \D, \D' \vdash e' : \vp$\` by inv. on (let)\\
+    (4) $\G;\D,\D' \vdash e'[e/x] : \vp$\` by $\D$-subst. lemma (2,3)\\
+    (5) $\G,\xD; \D,\D' \vdash e'[e/x] : \vp$ \` by (admissible) $Weaken_\Delta$\\
+    (6) $\G;\D,\D' \vdash \llet{\xD = e}{e'[e/x]} : \vp$\` by (let) (2,5)\\
 \end{tabbing}

### 2023-09-24 prototype/src/Linear/Core.hs
Evaluation strategies in background

@@ -73,10 +73,10 @@ runLinearCore pgr = do
 
   case runIdentity $ runLinearCoreT bindingsMap (checkProgram lcprg) of
     Left e -> do
-      pprTraceM "Failed to typecheck linearity!" (ppr pgr)
+      -- pprTraceM "Failed to typecheck linearity!" (ppr pgr)
       return [text e]
     Right x -> do
-      pprTraceM "Safe prog!" (ppr x)
+      -- pprTraceM "Safe prog!" (ppr x)
       return []
 
 --------------------------------------------------------------------------------
@@ -87,7 +87,7 @@ type LinearCoreM = LinearCoreT Identity
 
 checkProgram :: LCProgram -> LinearCoreM LCProgram
 checkProgram prog = do
-  pprTraceM "checkProgram" (ppr prog)
+  -- pprTraceM "checkProgram" (ppr prog)
   traverse checkBind prog
 
 -- ROMES:TODO: use isGlobalId and setTopLevelBindingName to set the binding
@@ -118,7 +118,7 @@ checkBind (Rec bs) = do
       -- environments with a linear multiplicity.
       -- In practice, when recursive binders are used, we'll try to use the linear variables more than once, if they exist more than once
   recUes <- mapM (\i -> reconstructUe i.id recUsages inScope) ids
-  pprTraceM "Has checked naiveUsages" (ppr recUsages $$ ppr recUes)
+  -- pprTraceM "Has checked naiveUsages" (ppr recUsages $$ ppr recUes)
   let ids' = L.zipWith (\i b -> LCVar i.id (deltaBinding b)) ids recUes
 
   -- Must typecheck rhss' again with the correct recursive usage environments
@@ -163,10 +163,9 @@ checkExpr expr = case expr of
     -> pprTrace "Case expression" (ppr casee) $ do
 
       (e', ue) <- restoringState $ record $ checkExpr e
-      pprTraceM "Making irrelevant usage environment" (ppr ue)
-      -- ROMES:TODO: Make the resources irrelevant in the actual context, not only in the usage environment
+      -- pprTraceM "Making irrelevant usage environment" (ppr ue)
       makeEnvResourcesIrrelevant ue
-      pprTraceM "Done irr" Ppr.empty
+      -- pprTraceM "Done irr" Ppr.empty
       Case e'
            (LCVar b.id (deltaBinding ue))
            ty
[…]

### 2023-09-24 prototype/src/Linear/Core/Monad.hs

@@ -131,7 +131,7 @@ use = flip use' [] where
     -- Lookup the variable in the environment
     mv <- gets (M.lookup key)
 
-    pprTraceM "Using var" (Ppr.ppr key Ppr.<+> Ppr.text "with mult" Ppr.<+> Ppr.ppr mv Ppr.<+> Ppr.text "and tag environment" Ppr.<+> Ppr.ppr mtags)
+    -- pprTraceM "Using var" (Ppr.ppr key Ppr.<+> Ppr.text "with mult" Ppr.<+> Ppr.ppr mv Ppr.<+> Ppr.text "and tag environment" Ppr.<+> Ppr.ppr mtags)
 
     case mv of
 
@@ -162,7 +162,7 @@ use = flip use' [] where
               if isDryRun
                  then tell [key]
                  else do
-                   pprTraceM "Using lambda bound" (Ppr.ppr key Ppr.<+> Ppr.ppr mult)
+                   -- pprTraceM "Using lambda bound" (Ppr.ppr key Ppr.<+> Ppr.ppr mult)
                    modify (M.delete key)
 
             -- We are trying to use a fragment of this resource, so we split it
@@ -170,12 +170,12 @@ use = flip use' [] where
             -- given tag (in the form of a tagstack)
             tags -> do
               splits <- splitAsNeededThenConsume allowsIrrelevant tags mult
-              pprTraceM "Splitting key at tags" (Ppr.ppr key Ppr.<+> Ppr.ppr tags Ppr.<+> Ppr.ppr splits)
+              -- pprTraceM "Splitting key at tags" (Ppr.ppr key Ppr.<+> Ppr.ppr tags Ppr.<+> Ppr.ppr splits)
               case NE.nonEmpty splits of
                     Nothing
                       | isDryRun  -> tell [key]
                       | otherwise -> do
-                        pprTraceM "Using lambda bound split" (Ppr.ppr key)
+                        -- pprTraceM "Using lambda bound split" (Ppr.ppr key)
                         modify (M.delete key) -- we split the fragment and consumed the only one
                     Just splits'  -> modify (M.insert key (Right splits')) -- we keep the remaining fragments
 
@@ -212,12 +212,12 @@ use = flip use' [] where
           tags -> do
 
             splits <- join <$> mapM (splitAsNeededThenConsume allowsIrrelevant tags) (NE.toList mults)
-            pprTraceM "Split key at tags" (Ppr.ppr key Ppr.<+> Ppr.ppr tags Ppr.<+> Ppr.ppr splits)
[…]

### 2023-09-24 prototype/src/Linear/Core/Plugin.hs

@@ -35,6 +35,7 @@ import GHC.Plugins
 import Control.Monad
 import Control.Monad.Reader
 import Control.Monad.Except
+import qualified GHC.Utils.Trace as Trace
 
 -- import qualified Linear.Core.Translate.Syntax as LC
 -- import qualified Linear.Core.Translate.Check as LC
@@ -67,8 +68,10 @@ linearCorePass guts = do
     [] -> pure ()
     errs ->
       if all (L.isPrefixOf "fail_" . showSDocUnsafe) errs
+       -- the fail_ thing was an attempt to allow the plugin to continue on
+       -- functions marked for failing, but that feature isn't finished
        then fatalErrorMsg (ppr errs)
-       else pprPanic "Linear typechecking failed when it shouldn't!" (ppr errs)
+       else Trace.pprTraceM "[FAILED] Linear typechecking failed when it shouldn't!" (ppr errs)
 
   return guts -- unchanged guts, after validating them.
 

### 2023-09-29 proposal.tex
Almost closed implemenetation section

@@
 variables names and their corresponding naive environment as input to
-algorithm~\ref{computeRecUsages} to get a final usage environment. Intuitively,
+Algorithm~\ref{computeRecUsages} to get a final usage environment. Intuitively,
 the algorithm, for each recursive binder, iterates over all (initially naive) usage

### 2023-09-29 language-v4/proofs/WHNFConvSoundness.tex
Irrelevance using updated assumptions

--- 2023-09-12 language-v4/proofs/WHNFConvSoundness.tex
@@
 
-\item[Case:] $\Lambda I$
+\item[Case:] $Alt\_$
 \begin{tabbing}
-    (1)
+    (1) $\G,\z[\irr{\D}];\irr{\D}, \D' \vdash_{alt} \_ \to e :^z_{\irr{\D}} \s \Rightarrow \vp$\\
+    (2) $\G, \z[\irr{\D}];\irr{\D},\D' \vdash e : \vp$\\
+    (3) $\irr{\D}$ is used through $z$\`since $\irr{\D}$ can't otherwise be used\\\`and is introduced in this alternative uniquely \\\`(since we have multi-tier proof irrelevance, i.e.
+$\irr{\irr{\D}}\neq\irr{\D}$)\\
+    (4) $\G[z/\irr{\D}]; \z[1], \D' \vdash e : \vp$\` by $\D \Rightarrow 1$ lemma (2,3)\\
+    (5) $\G[\ov{\D_i}/\irr{\D}], \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash e : \vp$\` by $1 \Rightarrow \D$ lemma (4)\\
+    (6) $\G, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash e : \vp$\` by (3), resources in $\G$ with $\irr{\D}$ (parts) in\\\` the usage environment cannot be used in $e$\\
+    (7) $\G, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash_{alt} \_ \to e :^z_{\ov{\D_i}} \s \Rightarrow \vp$\` by $Alt\_$\\
+\end{tabbing}
+
+\item[Case:] $Alt0$
+\begin{tabbing}
+    (1) $\G, \z[\irr{\D}]; \irr{\D}, \D' \vdash_{alt} K~\ov{\xo}\to e :^z_{\irr{\D}} \s \Rightarrow \vp$\\
+    (2) $\G, \z[\cdot], \ov{\xo}; \D' \vdash e : \vp$\` by inv. on $Alt0$ and def. of empty subst.\\
+    (3) $\G, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash_{alt} K~\ov{\xo}\to e :^z_{\ov{\D_i}} \s \Rightarrow \vp$\` by $Alt0$ (2)\\
+\end{tabbing}
+
+\item[Case:] $AltN_{\textrm{WHNF}}$
+\begin{tabbing}
+    Not applicable since $\Rrightarrow$ is only generalized by $\Rightarrow$, not $\Mapsto$.
+\end{tabbing}
+
+\item[Case:] $AltN_{\textrm{Not WHNF}}$
+    We prove the theorem for constructing both an $AltN_{\textrm{Not WHNF}}$
+        and an $AltN_{\textrm{WHNF}}$ from a proof-irrelevant $AltN_{\textrm{Not
+        WHNF}}$, to prove the statement holds for any $\Rightarrow$ kind for
+        $AltN$, rather than just $\Rrightarrow$ or $\Mapsto$.
+\begin{tabbing}
+    (1) $\G, \z[\irr{\D}]; \irr{\D}, \D' \vdash_{alt} K~\ov{\xo},\ov{y_i{:}_{1}\s_i}^n\to e :^z_{\irr{\D}} \s \Rrightarrow \vp$\\
+    (2) $\G,\z[\irr{\D}],\ov{\xo},\ov{y_j{:}_{\D_j}\s_j}^n; \irr{\D}, \D' \vdash e : \vp$\`by inv. on $AltN_{\textrm{Not WHNF}}$\\
+    (3) $\ov{\D_j} = \ov{\lctag{\irr{\D}}{K_j}}$\`by inv. on $AltN_{\textrm{Not WHNF}}$\\
+    Subcase $\irr{\D}$ is consumed through $z$\\
+    (4) $\G,\ov{\xo},\ov{y_j{:}_{\D_j[z/\irr{\D}]}\s_j}^n; \z[1], \D' \vdash e : \vp$\`by $\D \Rightarrow 1$ lemma (2,subcase)\\\`and vars in $\G$ with $\irr{\D}$ cannot be used by subcase\\
+    Subcase constructing $AltN_{\textrm{WHNF}}$ ($\Mapsto$)\\
+    (5) $\G,\ov{\xo},\ov{y_j{:}_{\D_i}\s_j}^n; \z[1], \D' \vdash e : \vp$\`by $\ov{y}$ is not used, and $Weaken$\\
+    (6) $\G,\ov{\xo},\ov{y_j{:}_{\D_i}\s_j}^n, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash e : \vp$\`by $1 \Rightarrow \D$ lemma (5)\\\` and $\G$ vars do not mention $z$ by (4) \\
+    (7) $\G, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash_{alt} K~\ov{\xo},\ov{y_i{:}_{1}\s_i}^n \to e :^z_{\ov{\D_i}} \s \Mapsto \vp$\`by $AltN_{\textrm{WHNF}}$ (6)\\
+    Subcase constructing $AltN_{\textrm{Not WHNF}}$ ($\Rrightarrow$)\\
+    (5) $\G,\ov{\xo},\ov{y_j{:}_{\lctag{\ov{\D_i}}{K_j}}\s_j}^n; \z[1], \D' \vdash e : \vp$\`by $\ov{y}$ does not consume resources\\
+    (6) $\G,\ov{\xo},\ov{y_j{:}_{\lctag{\ov{\D_i}}{K_j}}\s_j}^n, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash e : \vp$\`by $1 \Rightarrow \D$ lemma (5)\\\` and $\G$ vars do not mention $z$ by (4) \\
+    (7) $\G, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash_{alt} K~\ov{\xo},\ov{y_i{:}_{1}\s_i}^n \to e :^z_{\ov{\D_i}} \s \Rrightarrow \vp$\`by $AltN_{\textrm{Not WHNF}}$ (6)\\
+    Subcase $\irr{\D}$ is (fully) consumed by $\ov{y}$ (after splitting)\\
+    (4) $\G,\z[\ov{y}],\ov{\xo};\ov{y_j{:}_{1}\s_j}^n, \D' \vdash e : \vp$\`by $\D \Rightarrow 1$ lemma (2, subcase)\\\` and vars in $\G$ with $\irr{\D}$ cannot be used by subcase\\
+    (5) $\G,\z[\ov{\D_i}],\ov{\xo};\ov{y_j{:}_{1}\s_j}^n, \D' \vdash e : \vp$\`by $Weaken$ and $z$ does not consume resources\\
+    Subcase constructing $AltN_{\textrm{WHNF}}$ ($\Mapsto$)\\
+    (6) $\G,\z[\ov{\D_i}],\ov{\xo},\ov{y_j{:}_{\D_i}\s_j}^n; \ov{\D_i}, \D' \vdash e : \vp$\`by $1 \Rightarrow \D$ lemma (5)\\\`and vars in $\G$ do not mention $\ov{y}$ (4,5)\\
+    (7) $\G,\z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash K~\ov{\xo},\ov{y_i{:}_{1}\s_i}^n\to e :^z_{\ov{\D_i}} \s \Mapsto \vp$\`by $AltN_{\textrm{WHNF}}$\\
+    Subcase constructing $AltN_{\textrm{Not WHNF}}$ ($\Rrightarrow$)\\
+    (6) $\G,\z[\ov{\D_i}],\ov{\xo},\ov{y_j{:}_{\lctag{\ov{\D_i}}{K_j}}\s_j}^n; \ov{\D_i}, \D' \vdash e : \vp$\`by $1 \Rightarrow \D$ lemma (5)\\\`and vars in $\G$ do not mention $\ov{y}$ (4,5)\\
+    (7) $\G,\z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash K~\ov{\xo},\ov{y_i{:}_{1}\s_i}^n\to e :^z_{\ov{\D_i}} \s \Rrightarrow \vp$\`by $AltN_{\textrm{Not WHNF}}$\\
 \end{tabbing}

### 2023-10-02 language-v4/proofs/optimizations/BetaReduction.tex
Apply Jacinta's corrections
+ optimiz -> optimis

@@
 preserving. The argument of the application is a multiplicity rather than an
-expression, so no resources are needed to type it, and since the body of the
-lambda must treat the multiplicity $p$ as though it were a linear, the body
-uses the argument linearly regardless of the instantiation of $p$ at $\pi$.
+expression, so no resources are needed to type it, and, since the body of the
+lambda must treat the multiplicity $p$ as though it were linear, the body uses
+the argument linearly regardless of the instantiation of $p$ at $\pi$.
 

### 2023-10-02 prototype/core-plugin-results.tex
Make EXPERIMENT 1 about Case Var work
When run on linear-smc, with EXPERIMENT we get 0 errors, without EXPERIMENT 1 we get 4 errors that the experimente would lift. Great.

\begin{figure}[h]
\centering{
\begin{tabular}{c|c|c|c|c|c}
    \thead{Library} & \thead{Total\\Accepted} & \thead{Total\\Rejected} & \thead{Unique\\Rejected} & \thead{Case-Var\\Rejected} & \thead{Unknown\\Rejected}\\
    linear-smc & 19438 & 4 & 1 & 1 & 0\\
    linear-base & 92308 & - & - & - & -\\
\end{tabular}
}
\caption{Linear Core Plugin on Linear Libraries}
\label{fig:core-plugin-res}
\end{figure}

### 2023-10-02 prototype/src/Linear/Core.hs

@@ -169,10 +169,10 @@ checkExpr expr = case expr of
   casee@(Case e b ty alts)
     -- Expression is in WHNF (See Note [exprIsHNF] and #23771, function is really "exprIsWHNF")
     | exprIsHNF (unconvertExpr e)
-      -- EXPERIENCE 1: Experiment in ignoring the preservation issue and simply
+      -- part 1/2 EXPERIENCE 1: Experiment in ignoring the preservation issue and simply
       -- accepting occurrences of linear variables in scrutinees and in their
       -- bodies
-       || (if | Var _ <- e -> True | otherwise -> False)
+       -- || (if | Var _ <- e -> True | otherwise -> False)
     -> do
       lcs0 <- get
       (e', ue) <- record $ checkExpr e
@@ -181,8 +181,12 @@ checkExpr expr = case expr of
            ty
        <$> withSameEnvMap (\alt -> do
             put lcs0 -- we restore the state for each alternative, not before (otherwise resources aren't consumed in the EmptyCase)
+                     -- in the NOT WHNF case we also restore the state, but then make it irrelevant
             extend CaseBinder b.id (DeltaBound ue) $
-              checkAlt ue b.id (exprType (unconvertExpr e)) alt) alts
+              checkAlt ue (
+                          (if | Var x_s <- e -> (x_s:) | otherwise -> Prelude.id) -- part 2/2 of EXPERIENCE 1
+                          [b.id]
+                          ) (exprType (unconvertExpr e)) alt) alts
     | otherwise
     -- Expression is definitely not in WHNF (or do we really mean HNF?)
     -> do -- pprTrace "Case expression" (ppr casee) $ do
@@ -197,20 +201,20 @@ checkExpr expr = case expr of
            ty
        <$> withSameEnvMap (\alt -> do
             put lcs0 -- we restore the state for each alternative, not before (otherwise resources aren't consumed in the EmptyCase)
-            makeEnvResourcesIrrelevant ue
+            makeEnvResourcesIrrelevant ue -- and then make it irrelevant
             extend CaseBinder b.id (DeltaBound irrUe) $
[…]

### 2023-10-02 prototype/src/Linear/Core/Monad.hs

@@ -12,7 +12,7 @@ module Linear.Core.Monad
   , use
   , extend
   , extends
-  , drop, dropEnvOf
+  , drop, moveToUnrEnv
   , unrestricted
   , record
   , dryRun
@@ -139,9 +139,9 @@ use = flip use' [] where
     -- Lookup the variable in the environment
     mv <- gets (M.lookup key)
 
-    -- pprTraceM "Using var" (Ppr.ppr key Ppr.<+> Ppr.text "with mult" Ppr.<+>
-    --   Ppr.ppr mv Ppr.<+> Ppr.text "and tag environment" Ppr.<+> Ppr.ppr mtags
-    --   Ppr.<+> Ppr.text "; allowIrr: " Ppr.<+> Ppr.text (show allowsIrrelevant))
+    pprTraceM "Using var" (Ppr.ppr key Ppr.<+> Ppr.text "with mult" Ppr.<+>
+      Ppr.ppr mv Ppr.<+> Ppr.text "and tag environment" Ppr.<+> Ppr.ppr mtags
+      Ppr.<+> Ppr.text "; allowIrr: " Ppr.<+> Ppr.text (show allowsIrrelevant))
 
 
     case mv of
@@ -232,8 +232,9 @@ use = flip use' [] where
                     modify (M.delete key) -- OK, we only had one fragment left and consumed it
                 Just splits' -> modify (M.insert key (Right splits'))   -- we keep the remaining fragments
 
-    -- key_val <- gets (M.lookup key)
-    -- pprTraceM "Used var, remaining var in env is:" (Ppr.ppr key Ppr.<+> Ppr.ppr key_val)
+    -- debug
+    key_val <- gets (M.lookup key)
+    pprTraceM "Used var, remaining var in env is:" (Ppr.ppr key Ppr.<+> Ppr.ppr key_val)
 
 
 -- | Extend a computation that threads linear resources with a resource that
@@ -267,8 +268,9 @@ extend bindsite key value computation = LinearCoreT do
       | either isBindingLinear (const True) ms
       , not isDryRun -- When doing a dry run, don't fail, even if the resource isn't consumed
       -> do
+        lcs <- get
         throwError $ fromString $
-           "The linear resource " ++ showPprUnsafe key ++ " wasn't fully consumed in the computation extended with it. [" ++ showPprUnsafe ms ++ "]; Binder: " ++ show bindsite
[…]

### 2023-10-02 prototype/src/Linear/Core/Multiplicities.hs

@@ -24,7 +24,7 @@ import GHC.Core.Type (isMultiplicityVar, isMultiplicityTy)
 ----- Id Binding ---------------------------------------------------------------
 --------------------------------------------------------------------------------
 
-data BindSite = LetBinder | LetRecBinder | LetRecBinderDry | LambdaBinder | CaseBinder | PatternBinder
+data BindSite = LetBinder | LetRecBinder | LetRecBinderDry | LambdaBinder | CaseBinder | PatternBinder | Unknown
   deriving (Show, Eq)
 
 data IdBinding = LambdaBound Mult    -- lambdas

### 2023-10-02 prototype/src/Linear/Core.hs
Revert crazy hack that allowed a case scrut var to become unrestricted if the pattern were unrestricted

@@ -181,12 +181,8 @@ checkExpr expr = case expr of
            ty
        <$> withSameEnvMap (\alt -> do
             put lcs0 -- we restore the state for each alternative, not before (otherwise resources aren't consumed in the EmptyCase)
-                     -- in the NOT WHNF case we also restore the state, but then make it irrelevant
             extend CaseBinder b.id (DeltaBound ue) $
-              checkAlt ue (
-                          (if | Var x_s <- e -> (x_s:) | otherwise -> Prelude.id) -- part 2/2 of EXPERIENCE 1
-                          [b.id]
-                          ) (exprType (unconvertExpr e)) alt) alts
+              checkAlt ue b.id (exprType (unconvertExpr e)) alt) alts
     | otherwise
     -- Expression is definitely not in WHNF (or do we really mean HNF?)
     -> do -- pprTrace "Case expression" (ppr casee) $ do
@@ -201,20 +197,20 @@ checkExpr expr = case expr of
            ty
        <$> withSameEnvMap (\alt -> do
             put lcs0 -- we restore the state for each alternative, not before (otherwise resources aren't consumed in the EmptyCase)
-            makeEnvResourcesIrrelevant ue -- and then make it irrelevant
+            makeEnvResourcesIrrelevant ue
             extend CaseBinder b.id (DeltaBound irrUe) $
-              pprTrace "checking alt" Ppr.empty $ checkAlt irrUe [b.id] (exprType (unconvertExpr e)) alt) alts
+              pprTrace "checking alt" Ppr.empty $ checkAlt irrUe b.id (exprType (unconvertExpr e)) alt) alts
 
   Tick t e  -> Tick t <$> checkExpr e
   Cast e co -> Cast <$> checkExpr e <*> pure co
 
 checkAlt :: UsageEnv -- ^ The scrutinee's usage environment
-         -> [Var]    -- ^ Case binder name and possibly scrutinee-var name, to make unrestricted in ALT0
+         -> Var      -- ^ Case binder name
          -> Type     -- ^ Scrutinee type
          -> LCAlt -> LinearCoreM LCAlt
 
 --- * ALT_
-checkAlt ue zbinds s_ty (Alt DEFAULT [] rhs) = do
+checkAlt ue zbind s_ty (Alt DEFAULT [] rhs) = do
[…]

### 2023-10-02 prototype/src/Linear/Core/Monad.hs

@@ -12,7 +12,7 @@ module Linear.Core.Monad
   , use
   , extend
   , extends
-  , drop, moveToUnrEnv
+  , drop, dropEnvOf
   , unrestricted
   , record
   , dryRun
@@ -139,9 +139,9 @@ use = flip use' [] where
     -- Lookup the variable in the environment
     mv <- gets (M.lookup key)
 
-    pprTraceM "Using var" (Ppr.ppr key Ppr.<+> Ppr.text "with mult" Ppr.<+>
-      Ppr.ppr mv Ppr.<+> Ppr.text "and tag environment" Ppr.<+> Ppr.ppr mtags
-      Ppr.<+> Ppr.text "; allowIrr: " Ppr.<+> Ppr.text (show allowsIrrelevant))
+    -- pprTraceM "Using var" (Ppr.ppr key Ppr.<+> Ppr.text "with mult" Ppr.<+>
+    --   Ppr.ppr mv Ppr.<+> Ppr.text "and tag environment" Ppr.<+> Ppr.ppr mtags
+    --   Ppr.<+> Ppr.text "; allowIrr: " Ppr.<+> Ppr.text (show allowsIrrelevant))
 
 
     case mv of
@@ -232,9 +232,8 @@ use = flip use' [] where
                     modify (M.delete key) -- OK, we only had one fragment left and consumed it
                 Just splits' -> modify (M.insert key (Right splits'))   -- we keep the remaining fragments
 
-    -- debug
-    key_val <- gets (M.lookup key)
-    pprTraceM "Used var, remaining var in env is:" (Ppr.ppr key Ppr.<+> Ppr.ppr key_val)
+    -- key_val <- gets (M.lookup key)
+    -- pprTraceM "Used var, remaining var in env is:" (Ppr.ppr key Ppr.<+> Ppr.ppr key_val)
 
 
 -- | Extend a computation that threads linear resources with a resource that
@@ -332,10 +331,10 @@ drop (UsageEnv env) = do
 --              -> [m]
 --             ) (NE.toList mults)
 
-moveToUnrEnv :: Monad m
+dropEnvOf :: Monad m
           => Var -- ^ The var to make unrestricted by "dropping all resources from it"
           -> LinearCoreT m ()
-moveToUnrEnv v = do
+dropEnvOf v = do
   modify (M.insert v (Left (LambdaBound (Relevant GHC.Core.Type.ManyTy))))
 
 
@@ -351,11 +350,9 @@ unrestricted computation = LinearCoreT do
[…]

### 2023-10-02 prototype/src/Linear/Core/Multiplicities.hs

@@ -24,7 +24,7 @@ import GHC.Core.Type (isMultiplicityVar, isMultiplicityTy)
 ----- Id Binding ---------------------------------------------------------------
 --------------------------------------------------------------------------------
 
-data BindSite = LetBinder | LetRecBinder | LetRecBinderDry | LambdaBinder | CaseBinder | PatternBinder | Unknown
+data BindSite = LetBinder | LetRecBinder | LetRecBinderDry | LambdaBinder | CaseBinder | PatternBinder
   deriving (Show, Eq)
 
 data IdBinding = LambdaBound Mult    -- lambdas

### 2023-10-02 prototype/core-plugin-results.tex
First final version finished

@@
 \centering{
-\begin{tabular}{c|c|c|c|c|c}
-    \thead{Library} & \thead{Total\\Accepted} & \thead{Total\\Rejected} & \thead{Unique\\Rejected} & \thead{Case-Var\\Rejected} & \thead{Unknown\\Rejected}\\
-    linear-smc & 19438 & 4 & 1 & 1 & 0\\
-    linear-base & 92308 & - & - & - & -\\
+\begin{tabular}{c|c|c|c|c|c|c|c}
+    \thead{Library} & \thead{Total\\Accepted} & \thead{Total\\Rejected} & \thead{Unique\\Rejected} & \thead{Linear modulo\\Call-by-name} & \thead{Linear\\Rejected} & \thead{$\lnot$ Linear\\Rejected} & \thead{Unknown\\Rejected}\\
+    linear-smc & 19438 & 4 & 1 & 1 & 0 & 0 & 0\\
+    linear-base & 91516 & 499 & 72 & 10 & 8 & 2 & 52\\
 \end{tabular}

### 2023-10-02 prototype/src/Linear/Core.hs

@@ -172,7 +172,7 @@ checkExpr expr = case expr of
       -- part 1/2 EXPERIENCE 1: Experiment in ignoring the preservation issue and simply
       -- accepting occurrences of linear variables in scrutinees and in their
       -- bodies
-       -- || (if | Var _ <- e -> True | otherwise -> False)
+       || (if | Var _ <- e -> True | otherwise -> False)
     -> do
       lcs0 <- get
       (e', ue) <- record $ checkExpr e

### 2023-10-02 prototype/src/Linear/Core/Monad.hs

@@ -307,30 +307,6 @@ drop (UsageEnv env) = do
   -- Remove resources ocurring in the given usage env from the available resources, then run the computation
   modify (`diffResources` (M.fromListWith (\cases (Right ne1) (Right ne2) -> Right (ne1 <> ne2) ;  _ _ -> error "No left left, left right, or right left") $ L.map (\(v,m) -> if null (extractTags m) then (v, Left (LambdaBound m)) else (v, Right (NE.fromList [m]))) env))
 
---     where
---       -- For keys being dropped that still exist in the environment, 
---       diffgo :: Either IdBinding (NonEmpty Mult)
---              -> Mult
---              -> Maybe (Either IdBinding (NonEmpty Mult))
---       diffgo (Left (LambdaBound _)) (extractTags -> []) = Nothing
---       diffgo (Left (LambdaBound m)) (extractTags -> tags)
---         = Just (case splitAsNeededThenConsume @(Either String) AllowIrrelevant tags m of
---                  Left s -> error s
---                  Right (ms,_) -> Right (NE.fromList ms)
---                )
---       diffgo (Left (DeltaBound _)) _ = error "We shouldn't be dropping a delta var?"
---       diffgo (Right _) (extractTags -> []) = error "unexpected extract tags = []"
---       diffgo (Right mults) (extractTags -> tags) = fmap Right $ NE.nonEmpty $
---         L.concatMap (\m ->
---           if | extractTags m == tags
---              -> []
---              | extractTags m `L.isPrefixOf` tags
---              -> fst . fromRight undefined $
---                splitAsNeededThenConsume AllowIrrelevant tags m
---              | otherwise
---              -> [m]
---             ) (NE.toList mults)
-
 dropEnvOf :: Monad m
           => Var -- ^ The var to make unrestricted by "dropping all resources from it"
           -> LinearCoreT m ()

### 2023-10-02 prototype/core-plugin-results.tex
Corrections

--- 2023-10-02 prototype/core-plugin-results.tex
@@
 \centering{
-\begin{tabular}{c|c|c|c|c|c}
-    \thead{Library} & \thead{Total\\Accepted} & \thead{Total\\Rejected} & \thead{Unique\\Rejected} & \thead{Case-Var\\Rejected} & \thead{Unknown\\Rejected}\\
-    linear-smc & 19438 & 4 & 1 & 1 & 0\\
-    linear-base & 92308 & - & - & - & -\\
+\begin{tabular}{|c|c|c|c|c|c|c|c|}
+    \toprule
+    \thead{Library} & \thead{Total\\Accepted} & \thead{Total\\Rejected} & \thead{Unique\\Rejected} & \thead{Linear modulo\\Call-by-name} & \thead{Linear\\Rejected} & \thead{$\lnot$ Linear\\Rejected} & \thead{Unknown\\Rejected}\\
+    \midrule
+    linear-smc & 19438 & 4 & 1 & 1 & 0 & 0 & 0\\
+    linear-base & 91516 & 499 & 72 & 10 & 8 & 2 & 52\\
+    \makecell{priority-\\sesh} & 83 & 69 & 6 & - & - & - & 6\\
+    \bottomrule
 \end{tabular}

### 2023-10-02 prototype/src/Linear/Core.hs

@@ -172,7 +172,7 @@ checkExpr expr = case expr of
       -- part 1/2 EXPERIENCE 1: Experiment in ignoring the preservation issue and simply
       -- accepting occurrences of linear variables in scrutinees and in their
       -- bodies
-       || (if | Var _ <- e -> True | otherwise -> False)
+       -- || (if | Var _ <- e -> True | otherwise -> False)
     -> do
       lcs0 <- get
       (e', ue) <- record $ checkExpr e

### 2023-10-02 prototype/src/Linear/Core/Monad.hs
Fix major bug in splitting-by-need

@@ -47,6 +47,8 @@ import qualified Data.Map as M
 import qualified GHC.Utils.Panic as Ppr
 import qualified Data.List as L
 import Data.Either (fromRight)
+import Data.Bifunctor (Bifunctor(..))
+import Data.Maybe (catMaybes)
 
 -- | Computations that thread through a linear context from which resources must
 -- be used exactly once, and resources of other multiplicities that can be
@@ -222,7 +224,11 @@ use = flip use' [] where
           -- non-matching tagged fragments untouched.
           tags -> do
 
-            (join -> splits, consumed_ms) <- mapAndUnzipM (splitAsNeededThenConsume allowsIrrelevant tags) (NE.toList mults)
+            (join -> splits, catMaybes -> consumed_ms)
+              <- mapAndUnzipM (\m -> if extractTags m `L.isPrefixOf` tags
+                                        -- if the mult doesn't match the prefix then we don't split, otherwsie we split things that are definitely not supposed to split!
+                                        then second Just <$> splitAsNeededThenConsume allowsIrrelevant tags m
+                                        else return ([m], Nothing)) (NE.toList mults)
             -- pprTraceM "Split key at tags" (Ppr.ppr key Ppr.<+> Ppr.ppr tags Ppr.<+> Ppr.ppr splits)
             if isDryRun
                then tell (map (key,) consumed_ms) -- Again, we don't consume things when dry run

### 2023-10-02 prototype/src/Linear/Core.hs
Drop traces

@@ -117,9 +117,9 @@ checkBind (Rec bs) = do
     (extends LetRecBinderDry
          (L.map (\(LCVar b (Just _d)) -> (b, LambdaBound (Relevant OneTy))) ids0)
          (traverse (dryRun . checkExpr) rhss))
-  pprTraceM "naiveUsages" (ppr naiveUsages)
+  -- pprTraceM "naiveUsages" (ppr naiveUsages)
   let recUsages = computeRecUsageEnvs (zip (L.map (.id) ids0) naiveUsages)
-  pprTraceM "recUsages" (ppr recUsages)
+  -- pprTraceM "recUsages" (ppr recUsages)
 
       -- Repeated ocurrences of linear variables will be represented as many
       -- times as they occur in the recursive bindings in the usage
@@ -128,7 +128,7 @@ checkBind (Rec bs) = do
   recUes <- mapM (\i -> extends LetRecBinder (L.map (\id' -> (id'.id, DeltaBound emptyUE)) ids0) $ reconstructUe i.id recUsages inScope) ids0
 
   let ids1 = L.zipWith (\i b -> LCVar i.id (deltaBinding b)) ids0 recUes
-  pprTraceM "Recursive bindings:" (ppr ids1)
+  -- pprTraceM "Recursive bindings:" (ppr ids1)
 
   -- Must typecheck rhss' again with the correct recursive usage environments
   rhss' <- extends LetRecBinder (L.map (\(LCVar i (Just b)) -> (i,b)) ids1) (traverse checkExpr rhss)
@@ -191,7 +191,7 @@ checkExpr expr = case expr of
       lcs0 <- get
       (e', ue) <- record $ checkExpr e
       let irrUe = makeIrrelevant ue
-      pprTraceM "Make irrelevant usage environment:" (ppr irrUe) 
+      -- pprTraceM "Make irrelevant usage environment:" (ppr irrUe) 
       Case e'
            (LCVar b.id (deltaBinding irrUe))
            ty
@@ -199,7 +199,7 @@ checkExpr expr = case expr of
             put lcs0 -- we restore the state for each alternative, not before (otherwise resources aren't consumed in the EmptyCase)
             makeEnvResourcesIrrelevant ue
             extend CaseBinder b.id (DeltaBound irrUe) $
-              pprTrace "checking alt" Ppr.empty $ checkAlt irrUe b.id (exprType (unconvertExpr e)) alt) alts
[…]

### 2023-10-02 prototype/src/Linear/Core/Monad.hs

@@ -418,7 +418,7 @@ dryRun comp = LinearCoreT do
 
 makeEnvResourcesIrrelevant :: Monad m => UsageEnv -> LinearCoreT m ()
 makeEnvResourcesIrrelevant (UsageEnv vs) = do
-  pprTraceM "Making resource irrelevant in case alternative" $! Ppr.empty
+  -- pprTraceM "Making resource irrelevant in case alternative" $! Ppr.empty
   lcstate0 <- get
   lcstate1 <- forM vs $ \(var,mult) -> do
     case M.lookup var lcstate0 of
@@ -436,7 +436,7 @@ makeEnvResourcesIrrelevant (UsageEnv vs) = do
   -- pprTraceM "mkEnvResIrr lcstate0" (Ppr.ppr lcstate0)
   -- pprTraceM "mkEnvResIrr lcstate1" (Ppr.ppr lcstate1)
   put (M.fromList lcstate1 <> lcstate0)
-  pprTraceM "Made resources irrelevant in case alternative" $! Ppr.empty
+  -- pprTraceM "Made resources irrelevant in case alternative" $! Ppr.empty
 
 -- | Run a list of monadic computation ala 'mapM' but restoring the typing environment
 -- for each individual action

### 2023-10-02 prototype/src/Linear/Core/Plugin.hs
Don't print program on failure

@@ -75,7 +75,7 @@ linearCorePass guts = do
        -- the fail_ thing was an attempt to allow the plugin to continue on
        -- functions marked for failing, but that feature isn't finished
        -- then fatalErrorMsg (ppr errs)
-       Trace.pprTraceM "[FAILED]" (ppr errs Ppr.$$ ppr prog)
+       Trace.pprTraceM "[FAILED]" (ppr errs) -- Ppr.$$ ppr prog)
 
   -- We use a simple bash script to count SUCCESS and FAILED
   return guts -- unchanged guts, after validating them.

### 2023-10-02 prototype/src/Linear/Core/Plugin.hs
Revert "Don't print program on failure"
This reverts commit aa3826224db6402b175166de4065e9495582a124.

@@ -75,7 +75,7 @@ linearCorePass guts = do
        -- the fail_ thing was an attempt to allow the plugin to continue on
        -- functions marked for failing, but that feature isn't finished
        -- then fatalErrorMsg (ppr errs)
-       Trace.pprTraceM "[FAILED]" (ppr errs) -- Ppr.$$ ppr prog)
+       Trace.pprTraceM "[FAILED]" (ppr errs Ppr.$$ ppr prog)
 
   -- We use a simple bash script to count SUCCESS and FAILED
   return guts -- unchanged guts, after validating them.

### 2023-10-02 haskell-symposium/paper.tex
Add thesis_capa
And random files

\documentclass[sigplan,screen,review]{acmart}

\begin{document}

\title{Typechecking Linearity in Haskell's Core Intermediate Language}

\author{Rodrigo Mesquita}
\affiliation{
  \institution{Universidade Nova de Lisboa}
  \city{Lisbon}
  \country{Portugal}
}

\author{Bernardo Toninho}
\affiliation{
  \institution{Universidade Nova de Lisboa}
  \city{Lisbon}
  \country{Portugal}
}

\begin{abstract}
Linear types were introduced to Haskell ...
%
However, the current Core type-checker rejects many linearly valid programs
that originate from Core-to-Core optimizing transformations. As such, linearity
typing is effectively disabled, for otherwise disabling optimizations would be
far more devastating.
%
% This dissertation presents an extension to Core's type system that accepts a
The goal of our proposed dissertation is to develop an extension to Core's type
system that accepts a larger amount of programs and verifies that optimizing
transformations applied to well-typed linear Core produce well-typed linear
Core.
%
Our extension will be based on attaching variable \emph{usage environments} to
binders, which augment the type system with more  fine-grained contextual
linearity information, allowing the system to accept programs which seem to
syntactically violate linearity but preserve linear resource usage. We will
also develop a usage environment inference procedure and integrate the
procedure with the type checker.  We will validate our proposal by showing a
range of Core-to-Core transformations can be typed by our system.
\end{abstract}

\maketitle

\section{Introduction}

\end{document}

### 2024-01-22 icfp.tex
added tex source file

\documentclass[acmsmall, screen, review]{acmart}
\PassOptionsToPackage{dvipsnames}{xcolor}
\usepackage{boldline}
\usepackage{xcolor}
\usepackage{xargs}
\usepackage{framed}
\usepackage[framemethod=default]{mdframed}
\usepackage{mathpartir}
\usepackage{mathtools}
\usepackage{soul}
\usepackage{cmll} % ROMES:TODO: not allowed?
\usepackage{amsthm}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{cleveref}
\usepackage{makecell}
\usepackage{fancyvrb}
\usepackage{thmtools, thm-restate}
\usepackage{listings}

\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
% \newtheorem{lemma}[theorem]{Lemma}%[theorem]
\newtheorem{lemma}{Lemma}%[theorem]
\newtheorem{sublemma}{Lemma}[lemma]
\newtheorem{assumption}{Assumption}

%%%%%%%%%%%%%%  Color-related things   %%%%%%%%%%%%%%

% Work around lhs2tex#82
\let\Bbbk\undefined

%include polycode.fmt
%format ⊸ = "\lolli"

% don't, for now
%%%%subst keyword a = "\textcolor{BlueViolet}{\textbf{" a "}}"

\newcommand{\myFor}[2]{\For{$#1$}{$#2$}}
\newcommand{\id}[1]{\textsf{\textsl{#1}}}

% \renewcommand{\Varid}[1]{\textcolor{Sepia}{\id{#1}}}
% \renewcommand{\Conid}[1]{\textcolor{OliveGreen}{\id{#1}}}

%%%%%%%%%%%%  End of Color-related things   %%%%%%%%%%%%

% It might make sense to add pretty formating of individual things
% like "forall", cf.
% https://github.com/goldfirere/thesis/blob/master/tex/rae.fmt

% colorboxes, from rae's thesis as well
\definecolor{working}{rgb}{0.9,1,0.9}
\newmdenv[hidealllines=true,backgroundcolor=working,innerleftmargin=0pt,innerrightmargin=0pt,innertopmargin=-3pt,innerbottommargin=-3pt,skipabove=3pt,skipbelow=3pt]{working}
\newcommand{\workingcolorname}{light green}

\definecolor{notyet}{rgb}{1,1,0.85}
\newmdenv[hidealllines=true,backgroundcolor=notyet,innerleftmargin=0pt,innerrightmargin=0pt,innertopmargin=-3pt,innerbottommargin=-3pt,skipabove=3pt,skipbelow=3pt]{notyet}
\newcommand{\notyetcolorname}{light yellow}

\definecolor{noway}{rgb}{1,0.9,0.9}
\newmdenv[hidealllines=true,backgroundcolor=noway,innerleftmargin=0pt,innerrightmargin=0pt,innertopmargin=-3pt,innerbottommargin=-3pt,skipabove=3pt,skipbelow=3pt]{noway}
\newcommand{\nowaycolorname}{light red}

\definecolor{limitation}{rgb}{1.0, 0.875, 0.75}
\newmdenv[hidealllines=true,backgroundcolor=limitation,innerleftmargin=0pt,innerrightmargin=0pt,innertopmargin=-3pt,innerbottommargin=-3pt,skipabove=3pt,skipbelow=3pt]{limitation}
\newcommand{\limitationcolorname}{light orange}

\DefineVerbatimEnvironment{code}{Verbatim}{fontsize=\small}
\DefineVerbatimEnvironment{example}{Verbatim}{fontsize=\small}

\newcommand{\parawith}[1]{\paragraph{\emph{#1}}}
\newcommand{\lolli}{\multimap}
\newcommand{\tensor}{\otimes}
\newcommand{\one}{\mathbf{1}}
\newcommand{\bang}{{!}}

\newcommand{\llet}[2]{\mathsf{let}~#1~\mathsf{in}~#2}
\newcommand{\lletrec}[2]{\mathsf{letrec}~#1~\mathsf{in}~#2}
\newcommand{\ccase}[2]{\mathsf{case}~#1~\mathsf{of}~#2}

\lstset{
  frame=none,
  xleftmargin=2pt,
  stepnumber=1,
  numbers=left,
  numbersep=5pt,
  numberstyle=\ttfamily\tiny\color[gray]{0.3},
  belowcaptionskip=\bigskipamount,
  captionpos=b,
  escapeinside={*'}{'*},
  language=haskell,
  tabsize=2,
  emphstyle={\bf},
  commentstyle=\it,
  stringstyle=\mdseries\rmfamily,
  showspaces=false,
  keywordstyle=\bfseries\rmfamily,
  columns=flexible,
  basicstyle=\small\sffamily,
  showstringspaces=false,
  morecomment=[l]\%,
}

\input{proof}
\input{language-v2/proof}
\input{language-v3/proof}
\input{language-v4/proof}
\input{language-v4/Syntax}
\input{language-v4/TypingRules}
\input{language-v4/TheoremsAndLemmas}

\title{Type-checking Linearity in GHC Core}
\subtitle{Semantic Linearity in a Lazy Optimising Compiler}

\author{Rodrigo Mesquita}
\affiliation{
  \department{Departamento de Inform\'{a}tica}
  \institution{FCT-NOVA, Universidade Nova de Lisboa}
  %\city{City}
  \country{Portugal}}
\affiliation{
   \institution{Well-Typed}
  %\city{City}
   \country{Something}}
 \email{xxx\@yyy.com}
\author{Bernardo Toninho}
\orcid{0000-0002-0746-7514}
\affiliation{
  \department{Departamento de Inform\'{a}tica}
  \institution{NOVA-LINCS, FCT-NOVA, Universidade Nova de Lisboa}
  %\city{City}
  \country{Portugal}}
\email{btoninho\@fct.unl.pt}
%% TODO: AFFILIATION, EMAIL
%% TODO: CCS CONCEPTS and KEYWORDS
%% TODO: ACM RIGHTS notice...
% CHECK: https://www.acm.org/binaries/content/assets/publications/taps/latex-best_practices-06-may-2020.pdf

% \date{ }

\begin{document}

\begin{abstract}
Linear type systems guarantee linear resources are used \emph{exactly once}.
Traditionally, using a resource is synonymous with its \emph{syntactic}
occurrence in the program, however, under the lens of \emph{lazy} evaluation,
linearity can be further understood \emph{semantically}, where a
syntactic occurrence of a resource does not necessarily entail
\emph{using} that resource when the program is evaluated.
%
Semantic linearity is especially necessary in optimising compilers for
languages combining linearity and laziness: optimisations leverage laziness to
heavily rewrite the source program, pushing the interaction of linearity and
laziness to its limit, regardless of the original program typing linearity
conservatively.
%
We present Linear Core, the first type system that understands semantic
linearity in the presence of laziness, suitable for the Core intermediate
language of the Glasgow Haskell Compiler. We prove Linear Core is both type
safe and that multiple optimising transformations preserve linearity in Linear
Core while failing to do so in Core. We have implemented Linear Core as a
compiler plugin to validate the system against linearity-heavy libraries,
including \texttt{linear-base}, in the heart of the compiler.
\end{abstract}

\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% {{{ Introduction
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

% Statically safe programming languages provide compile time correctness
% guarantees by having the compiler rule out certain classes of errors
% or invalid programs. Moreover, static typing
% allows programmers to state and enforce (compile-time) invariants
% relevant to their problem domain.
% In this sense, type safety entails that all
% possible executions of a type-correct program cannot exhibit behaviors
% deemed ``wrong'' by the type system design. This idea is captured in
% the motto ``well-typed programs do not go wrong''.%~\cite{}.

Linear type systems~\cite{cite:linear-logic,cite:barberdill} add expressiveness
to existing type systems by enforcing that certain \emph{resources} (e.g.~a
file handle) must be used \emph{exactly once}.
%
In programming languages with a linear type system, not using certain resources
or using them twice is flagged as a type error. Linear types can thus be used
to, for instance, statically guarantee that socket descriptors are closed or
heap-allocated memory is freed, exactly once (leaks and double-frees become
type errors), or guarantee channel-based communication protocols are
deadlock-free~\cite{10.1007/978-3-642-15375-4_16},
% implement safe
% high-performance language interoperability~\cite{}, 
%guarantee that quantum entangled states are not duplicated~\cite{}
among other correctness properties~\cite{10.1145/3373718.3394765,10.1145/3527313,cite:linearhaskell}.
% handle mutable state safely~\cite{}

Consider the following program in which allocated memory is freed twice.
% We know this to be the dreaded double-free error which will crash the program at runtime.
Regardless of the \emph{double-free} error, a C-like type system will
accept such a program:
% \centering
\begin{code}
let p = malloc(4); in free(p); free(p);
\end{code}
Under the lens of a linear type system, if $p$ is deemed to be a
linear resource created by the call to \texttt{malloc}, it
must be used \emph{exactly once}.  However, the program above uses $p$ twice,
in the two calls to \texttt{free} and so fails to type-check.
%
In this sense, linear typing
effectively ensures the program does not compile with a double-free error.

Despite their promise and extensive presence in research
literature~\cite{Wadler1990LinearTC,CERVESATO2000133,10.1093/logcom/2.3.297},
an effective design combining linear and non-linear typing is both
challenging and necessary to bring the advantages of linear typing to
mainstream languages.
%
Consequently, few general purpose programming languages have linear
type systems. Among them are Idris~2~\cite{brady:LIPIcs.ECOOP.2021.9},
% a linearly and dependently typed language based on Quantitative Type Theory,
Rust~\cite{10.1145/2692956.2663188}, a language whose
ownership types build on linear types to guarantee memory safety
without garbage collection or reference counting, and, more recently,
Haskell~\cite{cite:linearhaskell}, a pure, functional, and
\emph{lazy} general purpose programming language.
%
% Besides Haskell's supporting linear
% types according to the said paper, Idris 2\cite{} supports linear types in a
% dependently typed setting, Clean\cite{} has uniqueness types which are closely
% related to linear types, and Rust\cite{} has ownership types which build from
% linear types. 
%
Linearity in Haskell stands out from linearity in other languages for the
following reasons, essential to our work:

\begin{itemize}
    % \item Linear types were only added to the language roughly \emph{31 years
    %     after} Haskell's inception, unlike Rust and Idris 2 which were designed
    %     with linearity from the start. It is an especially difficult endeavour
    %     to add linear types to a well-established language with a large library
    %     ecosystem and many active users, rather than to develop the language
    %     from the ground up with linear types in mind. The successful
    %     approach as implemented in the Glasgow Haskell Compiler (GHC), the leading
    %     Haskell compiler, is based on Linear Haskell~\cite{cite:linearhaskell}, which describes a linear type system designed
    %     with retaining backwards-compatibility and allowing code reuse across
    %     linear and non-linear users of the same libraries.
    %     We describe Linear Haskell in detail in Section~\ref{sec:linear-haskell}.

    \item Linear types permeate Haskell down to (its) Core, the
        intermediate language into which Haskell is
        translated. Core is a minimal, explicitly typed,
        functional language, to which multiple
        Core-to-Core optimising transformations are
        applied during compilation. Due to Core's minimal design, typechecking
        Core programs is very efficient and doing so serves as a sanity check to the
        correction of the source transformations. If the resulting optimised
        Core program fails to typecheck, the optimising
        transformations are very likely to have introduced an error
        in the resulting program. We present Core and its formal
        basis, System~$F_C$~\cite{cite:systemfc}, in Section~\ref{sec:core}.
        % TODO: \item values in rust are linear by default while non-linear is
        % the haskell default?

    \item Both Haskell and its intermediate language Core are \emph{lazily}
        evaluated, i.e., expressions in Haskell are only evaluated when needed,
        unlike C or Rust, which are \emph{eagerly} evaluated.
        %
        Laziness allows an optimising compiler to aggressively transform the source
        program without changing its semantics and, indeed, the Glasgow Haskell
        Compiler (GHC) heavily transforms Core by leveraging its laziness.
        %
        However, lazy evaluation interacts non-trivially with linearity.
        Intuitively, since expressions are not necessarily evaluated, an
        occurrence of a linear resource in an expression does not necessarily
        entail consuming that resource (i.e., if the expression is not evaluated,
        the resource is not used).

        In eagerly evaluated languages, the distinction between syntactic uses of a
        resource and the actual use of linear resources at runtime does not exist --
        an occurrence of a variable in the program always entails consuming it.
        %
        This interaction is unique to Haskell since, as far as we know, it is the
        only language featuring both laziness and linearity.
        %
        We review lazy and eager evaluation strategies in Section~\ref{sec:background:evaluation-strategies}.

\end{itemize}
%

% Linear core good Aligned with the philosophy of having a \emph{typed}
% intermediate language, the integration of linear types in Haskell required
% extending Core with linear types.
%
Much like a typed Core ensures that the translation from Haskell (dubbed
\emph{desugaring}) and the subsequent optimising transformations applied to
Core are correctly implemented, a \emph{linearly typed} Core guarantees that
linear resource usage in the source language is not violated by the translation
process and the compiler optimisation passes.
%
It is crucial that a program's behaviour enforced by linear types is \emph{not}
changed by the compiler in the desugaring or optimisation stages (optimisations
should not destroy linearity!) and a linearity aware Core type-checker is key in
providing such guarantees -- it would be disastrous if the compiler, e.g.,
duplicated a pointer to heap-allocated memory that was previously just used
once and then freed in the original program.
%TODO: linearidade pode informar otimizacoes
%
Even more, linearity in Core can inform Core-to-Core optimising
transformations~\cite{cite:let-floating,peytonjones1997a,cite:linearhaskell},
including inlining and $\beta$-reduction, to produce more performant programs.

% Linear core actually not so good
% Additionally, while not yet a reality, linearity in Core could be used to inform
% certain program optimisations, i.e. having linear types in Core could be used to
% further optimise certain programs and, therefore, benefit the runtime
% performance characteristics of our programs. For example, Linear Haskell\cite{}
% describes as future work an improvement to the inlining optimisation: Inlining
% is a centerpiece program optimisation primarily because of the subsequent
% optimising opportunities unlocked by inlining. However, it relies on a heuristic
% process known as \emph{cardinality analysis} to discover safe inlining
% opportunities. Unfortunately, heuristics can be volatile and fail in identifying
% crucial inlining opportunities resulting in programs that fall short of their
% performance potential. On the contrary, the linearity information could be
% integrated in the analysis and used as a (non-heuristic) cardinality
% \emph{declaration}.

% While the current version of Core is linearity-aware (i.e.,~Core has so-called
% multiplicity annotations in variable binders), its type system does not (fully)
% validate the linearity constraints in the desugared program and essentially
% fails to type-check programs resulting from several optimising transformations
% that are necessary to produce efficient object code. The reason for this latter
% point is not evidently clear:
% %
% if we can typecheck linearity in the surface level Haskell why do we fail to do
% so in Core?
%

In spite of a linearly typed Core ultimately being the desired intermediate
language for the Glasgow Haskell Compiler,
%ever since Linear Haskell was implemented,
both in its expressiveness to completely represent a Haskell with
linear types and in enabling more extensive compiler validations plus improved optimisations,
% it is currently impossible to enable linearity checking in Core. The reason is
linearity is effectively ignored in Core. The reason is not evidently clear:
if we can typecheck linearity in the surface level Haskell, it should also be
possible, and natural, to do so in Core.
% why do we fail to do so in Core?

The desugaring process from surface level Haskell to Core, and the subsequent
Core-to-Core optimising transformations, eliminate, and rearrange, most of the
syntactic constructs through which linearity checking is performed -- often
resulting in programs completely different from the original, especially due to
the flexibility laziness provides a compiler in the optimisations it may
perform. % without changing the program semantics.
%
Crucially, since optimisations do not destroy linearity, the resulting program
is still linear \emph{semantically}, however, the current linear type system
fails to recognize it as linear.
%
% The optimisations transform programs that are linear into ones that don't
% \emph{look} linear, but remain linear \emph{semantically}.
%
For instance, let $x$ be a linear resource in the two following programs, where
the latter results from inlining $y$ in the let body of the former. Despite the
result no longer \emph{looking} linear (as there are now two syntactic
occurrences of the linear resource $x$), the program \emph{is} indeed linear
because the let-bound expression freeing $x$ is never evaluated, so $x$ is
consumed exactly once when it is freed in the let body:
% executed, $x$ will be freed exactly once:
%
% , \emph{is} linear, as the let bound expression freeing $x$
% is never evaluated (because is not needed) -- thus $x$ is only consumed once
% when freed in the let body:
\[
\begin{array}{ccc}
\llet{y = \textsf{free}~x}{y} & \Longrightarrow_{Inlining} & \llet{y = \textsf{free}~x}{\textsf{free}~x}
\end{array}
\]

% \begin{boxedminipage}
% \begin{code}
% let y = free x
% in free x
% \end{code}
% \end{boxedminipage}

The Core optimising transformations expose a fundamental limitation of Core's
linear type system: it does not account for the call-by-need evaluation model
of Core, and thus a whole class of programs that are linear under the lens of
lazy evaluation are rejected by Core's current linear type system.

In this work, we address this limitation (and its implications on validating
and influencing optimising transformations) by designing a type system which
understands semantic linearity in the presence of laziness and is suitable for
the intermediate language of an optimising compiler.
%
In detail, our contributions are:
%
\begin{itemize}

\item We explain and provide insights into \emph{semantic} linearity in
contrast to \emph{syntactic} linearity, in Haskell, by example
($\S$~\ref{sec:linearity-semantically}).

\item We introduce Linear Core, a type system for a linear lazy language with
all key features from Core (except for type equality coercions), which,
crucially, understands semantic linearity in the presence of laziness. To the
best of our knowledge, this is the first type system to understand linearity
semantically in the context of lazy evaluation
($\S$~\ref{sec:main:linear-core}).

\item We prove Linear Core to be sound (well-typed Linear Core programs do not
get \emph{stuck}) and prove that multiple optimising transformations (which
currently violate linearity in Core) preserve types and linearity in Linear
Core ($\S$~\ref{sec:main:metatheory}).

\item We implement Linear Core as a GHC plugin which typechecks linearity in
all intermediate Core programs produced during the compilation process, showing
it accepts the intermediate programs resulting from (laziness-abusing)
transformations in linearity-heavy Haskell libraries, such as
\texttt{linear-base} ($\S$~\ref{sec:discuss:implementation}).

\end{itemize}
%
We review background concepts fundamental to our work in
Chapter~\ref{sec:background}, including linear type systems, linear types in
Haskell, evaluation strategies, Core's type system, and multiple optimising
transformations applied by GHC in its compilation process.
%
We compare our contributions to related work and discuss possible
avenues for further research (highlighting so-called \emph{multiplicity
coercions}) in Chapter~\ref{sec:discussion}, which concludes the document.

% In fact, we are not aware of any linear type system which
% understands linearity in the presence of laziness.

% \todo[inline]{In reality, the Core optimising transformations only expose a
% more fundamental issue in the existing linear types in Haskell -- its mismatch
% with the evaluation model. In call-by-need languages a mention of a variable
% might not entail using that variable - it depends on it being required or not.
% We're the first to do so as far as we know}

% Concluding, by extending Core / System $F_C$ with linearity and multiplicity
% annotations such that we can desugar all of Linear Haskell and validate it
% accross transformations taking into consideration Core's call-by-need
% semantics, we can validate the surface level linear type's implementation, we
% can guarantee optimising transformations do not destroy linearity, and we might
% be able to inform optimising transformations with linearity.

% Consider the following recursive let
% definition, where $x$ is a linear variable that must be used exactly once, would
% not typecheck in a source Haskell program since the typechecker cannot tell that
% $x$ is used linearly, but this program might occur naturally in Core from
% transformations on a program that did succesfully typecheck:
% \begin{code}
% letrec f = \case
%         True -> f False
%         False -> x
% in f True
% \end{code}

% Alternatively, one might imagine Haskell being desugared into an intermediate
% representation to which multiple different optimising transformations are
% applied but on which no integrity checks are done

% Despite \emph{want} a linearly-typed core
% because a linearly-typed Core ensures that desugaring
% Haskell and optimising transformations don't destroy linearity.

% In theory, because the Core language must also know about linearity, we should

% A remedy is to use the multiplicity annotations of λq → as cardinality declarations. Formalising
% and implementing the integration of multiplicities in the cardinality analysis is left as future work.

% In theory, we should typecheck \emph{linearity} in \textbf{Core} just the same
% as the typechecking verification we had with the existing type annotations prior
% to the addition of linear types, that is, our Core program with linearity
% annotations should be typechecked after the optimising transformations...

% }}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% {{{ Linearity, Semantically
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Linearity, Semantically\label{sec:linearity-semantically}}

A linear type system statically guarantees that linear resources are
\emph{consumed} exactly once. Consequently, whether a program is well-typed
under a linear type system intrinsically depends on the precise definition of
\emph{consuming} a resource. Even though \emph{consuming} a resource is
commonly regarded as synonymous with the \emph{syntactic occurrence} of the resource
in the program, that is not always the case.
%
In fact, this chapter highlights the distinction between using resources
\emph{syntactically} and \emph{semantically} as a principal limitation of
linear type systems for non-strict languages, with examples motivated by the
constructs available in GHC Core and how they are evaluated.

% and we show how a linear type system found on
% \emph{consuming resources semantically} rather than \emph{using resources
% syntactically} can accept

Consider the following program in a functional Haskell-like language,
where a computation that closes the given |handle| is bound to |x| before the
|handle| is returned:
%
\begin{code}
f : Handle \lolli Handle
f handle = let x = close handle in handle
\end{code}
%
In this seemingly innocent example, the |handle| appears to be closed before
returned, whereas in fact the handle will only be closed if the let bound
computation is effectively run (i.e. evaluated).
%
The example illustrates that \emph{consuming} a resource is not necessarily
synonymous with using it syntactically, as depending on the evaluation strategy
of the language, the computation that closes the handle might or not be
evaluated, and if it isn't, the |handle| in that unused computation
is not consumed.
%
Expanding on this, consider the above example program under distinct
evaluation strategies:
%
\begin{description}

\item[Call-by-value] With \emph{eager} evaluation semantics, the let bound expression |close
handle| is eagerly evaluated, and the |handle| will be closed before being
returned. It is clear that a linear type system should not accept such a
program since the linear resource |handle| is duplicated -- it is used in a
computation that closes it, while still being made available to the caller of
the function.

\item[Call-by-need] On the other hand, with \emph{lazy} evaluation semantics, the let
bound expression will only be evaluated when the binding |x| is needed. We return
the |handle| right away, and the let binding is forgotten as it cannot be
used outside the scope of the function, so the handle is not closed by |f|.
Under the lens of \emph{call-by-need} evaluation, \emph{using} a resource in a
let binding only results in the resource being \emph{consumed} if the binding
itself is \emph{consumed}. We argue that a linear type system under
\emph{call-by-need} evaluation should accept the above program, unlike a linear
type system for the same program evaluated \emph{call-by-value}.

\end{description}
%
%Rather, \emph{consuming} a resource is intrinsically defined by the evaluation model of the language.
%
% Under distinct evaluation strategies, the above example (and programs in general) have different semantics:
%
%
Intuitively, a computation that depends on a linear resource to produce a
result consumes that resource iff the result is effectively computed; in
contrast, a computation that depends on a linear resource, but is never run,
will not consume that resource.

From this observation, and exploring the connection between computation and evaluation,
% \todo{Alguma dificuldade em dizer exatamente como é que evaluation drives/is computation}
it becomes clear that \emph{linearity} and \emph{consuming resources}, in the
above example and for programs in general, should be defined in function of the
language's evaluation strategy.
%
We turn our focus to \emph{linearity} under \emph{call-by-need}, not only
because GHC Core is \emph{call-by-need}, but also because the distinction
between semantically and syntactically consuming a resource is only exposed
under \emph{non-strict} semantics.
%
Indeed, under \emph{call-by-value}, syntactic occurrences of a linear resource
directly correspond to semantically using that resource\footnote{With the minor exception
of trivial aliases, which don't entail any computation even in
\emph{call-by-value}. In theory, we could use in mutual exclusion any of the
aliases to refer to a resource without loss of linearity} because \emph{all}
expressions are eagerly evaluated -- if all computations are eagerly run, all
linear resources required by computations are \emph{eagerly consumed}.

% ROMES:IMPORTANT:TODO:
% \subsection{Reductions / Function applications}
% 
% \todo[inline]{unrestricted call-by-name with resources can duplicate the resources, as if it were unsound?}
% We reduce function applications in two distinct ways, call by name (for linear
% functions) and call by need (we've now introduced linear lets, so we can look
% at this now)
% 
% Foreshadow to issues with opt (reverse binder swap)?
% 
% Or maybe just drop this section altogether

\subsection{Semantic Linearity by Example\label{sec:semantic-linearity-examples}}

Aligned with our original motivation of typechecking linearity in GHC Core such
that optimising transformations preserve linearity, and with the goal
of understanding linearity in a non-strict context, this section helps the
reader build an intuition for semantic linearity through examples of Core
programs that are semantically linear but rejected by Core's linear type
system.
% BETTER?:
%Aligned with our motivation of typechecking linearity in GHC Core such
%that optimising transformations preserve linearity, and in order to explain
%linearity in a non-strict context, we illustrate semantic linearity
%through examples of Core programs that are semantically linear but rejected by
%Core's linear type system -- building intuition towards semantic linearity in a
%lazy language.
%
In the examples, a \colorbox{working}{\workingcolorname} background highlights programs that are
syntactically linear and are accepted by Core's naive linear type system. A
\colorbox{notyet}{\notyetcolorname} or \colorbox{limitation}{\limitationcolorname} background mark programs that are
semantically linear, but are not seen as linear by Core's (naive wrt laziness) linear type
system. Notably, the linear type system we develop in this work accepts all
\colorbox{notyet}{\notyetcolorname} programs.
% (while the few \colorbox{limitation}{\limitationcolorname} programs are not accepted).
A \colorbox{noway}{\nowaycolorname} background indicates that the program
simply isn't linear, not even semantically, i.e. the program effectively
discards or duplicates linear resources.

\subsubsection{Let bindings}
We start our discussion with non-strict (non-recursive) let bindings, i.e. let bindings whose
body is evaluated only when the binding is needed, rather than when declared.
In Core, a let binding entails the creation of a \emph{thunk} that suspends the
evaluation of the let body (for background, see
Section~\ref{sec:background:evaluation-strategies}). When the binding itself is evaluated, the \emph{thunk} is
\emph{forced} and the evaluation is carried out. The result overwrites the
\emph{thunk} -- the let binding now points to the result of the evaluation.
% A \emph{thunk} is \emph{forced} (and the suspended computation is evaluated) when
% the binding itself is evaluated.

In a linear type system, a non-strict let binding that depends on a linear
resource |x| doesn't consume the resource as long as the binding isn't
evaluated -- the suspended computation only uses the resource if it is run. For
this reason, we can't naively tell whether |x| is consumed just by looking at
the let binding body. In the following example, we assign a computation that
depends on the resource |x| to a binder, which is then returned:
%
\begin{notyet}
\begin{code}
f1 :: (a -o b) -> a -o b
f1 use x =
  let y = use x
   in y
\end{code}
\end{notyet}
%
The linear resource |x| is used exactly once, since it is used exactly once in
the body of the binding and the binding is used exactly once in the let body.
%
According to Linear Haskell's core calculus~$\lambda_{\to}^{q}$~\cite{cite:linearhaskell}, let
bound variables are annotated with a multiplicity which is multiplied (as per
the multiplicity semiring) by the multiplicity of all variables that are free
in the binder's body.
%
In short, if a let binder is linear (has
multiplicity $1$) then the linear variables free in its body are only used
once; if the let binder is unrestricted (has multiplicity $\omega$) then the
resources in its body are consumed many times, meaning no linear variables can
occur in that let binder's body.
%
Unfortunately, GHC's implementation of Linear
Haskell doesn't seem to infer multiplicities for lets yet, so while the above
program should typecheck in Linear Haskell, it is rejected by GHC.

The next example exposes the case in which the let binder is ignored in the let
body. Here, the linear resource |x| is used in |y|'s body and in the let
body, however, the resource is still used semantically linearly because |y| isn't used at all, thus |x| is consumed just once in the let body:
%
\begin{notyet}
\begin{code}
f2 :: (a -o a) -> a -o a
f2 use x =
  let y = use x
  in use x
\end{code}
\end{notyet}
%
Programmers don't often write bindings that are completely unused, yet, an
optimising compiler will produce intermediate programs with unused bindings\footnote{Unused bindings are then
also dropped by the optimising compiler} 
from transformations such as inlining, which can substitute out occurrences
of the binder (e.g. |y| is inlined in the let body).

Let bindings can also go unused if they are defined before branching on case
alternatives. At runtime, depending on the branch taken, the let binding will
be evaluated only if it occurs in that branch.
Both optimising transformations (float-out), and programmers used to non-strict evaluation, can produce
programs with bindings that are selectively used in the case alternatives, for
instance:
%
\begin{notyet}
\begin{code}
f3 :: (a -o a) -> Bool -> a -o a
f3 use bool x =
  let y = use x
  in case bool of
      True -> x
      False -> y
\end{code}
\end{notyet}
%
This example essentially merges |f1| with |f2|, using |x| directly in one
branch and using |y| in the other. Semantically, this program is linear because
the linear resource |x| ends up being used exactly once in both case
alternatives, directly or indirectly.

Shifting our focus from not using a let binding to using it (more than once),
we reiterate that a let binding creates a \emph{thunk} which is only evaluated
once, and re-used subsequently. Despite the binder body only being evaluated
once, and thus its resources only used once to compute a result, we can still
only consume said result of the computation once -- perhaps surprisingly, as the
perception so far is that ``resources are consumed during computation'' and
multiple uses of the same let binder share the result that was computed only
once. Illustratively, the following program must \emph{not} typecheck:
%
\begin{noway}
\begin{code}
f4 :: (a -o b) -> a -o (b, b)
f4 use x =
  let y = use x
  in (y, y)
\end{code}
\end{noway}
%
Intuitively, the result of the computation must also be used exactly once,
despite being effectively computed just once, because said result may still
contain (parts of) the linear resource. The trivial example is |f4| applied to
|id| -- the result of computing |id x| is |x|, and |x| must definitely not be
shared! Indeed, if the result of the computation involving the linear resource
was, e.g., an unrestricted integer, then sharing the result would not involve consuming the
resource more than once.
% ROMES:TODO:!!!!
%\todo{this kind of hints that maybe somehow in the
%alternative scrutinizing a "trivial" atom we could make the let binding
%unrestricted after we were sure it was evaluated once, but that isn't easy
%without erasing too many things before, we kind of tried this once}
%
Concretely, the result of evaluating a let binder body using linear resources, if computed, must be
consumed exactly once, or, otherwise, we risk discarding or duplicating said resources.

Lastly, consider a program which defines two let bindings |z| and |y|, where
|z| uses |y| which in turn uses the linear resource |x|:
%
\begin{noway}
\begin{code}
f5 :: (a -o a) -> a -o ()
f5 use x =
  let y = use x
  let z = use y
  in ()
\end{code}
\end{noway}
%
Even though the binding |y| is used in |z|, |x| is still never consumed because
|z| isn't evaluated in the let body, and consequently |y| isn't evaluated
either -- never consuming |x|. We use this example to highlight that even for
let bound variables, the syntactic occurrence of a variable isn't enough to
determine whether it is used. Instead, we ought to think of uses of |y| as
implying using |x|, and therefore uses of |z| imply using |x|, however, if
neither is used, then |x| isn't used. Since |x| is effectively discarded, this
example also violates linearity.

% \parawith{Summary}
The examples so far build an intuition for semantic linearity in the presence
of lazy let bindings. In essence, an unused let binding doesn't consume any
resources, and a let binding used exactly once consumes its resources exactly
once. Let binders that depend on linear resources must be used \emph{at most
once} -- let bound variables are \emph{affine} in the let body.
%
Moreover, if the let binding ($y$) isn't used in the let body, then the
resources it depends on ($\ov{x}$) must still be used -- the binding $y$ is
mutually exclusive with the resources $\ov{x}$ (for the resources to be used
linearly, either the binder occurs exactly once $y$, or the resources $\ov{x}$
do). We'll later see how we can encode this principle of mutual exclusivity
between let bindings and their dependencies using so called \emph{usage
environments}, in Section~\ref{sec:usage-environments}.

\subsubsection{Recursive let bindings\label{sec:semantic-linearity-examples:recursive-lets}}
Second, we look into recursive let bindings. For the most part,
recursive let bindings behave as non-recursive let bindings, i.e. we must use them \emph{at
most once} because, when evaluated, the linear resources used in the binders
bodies are consumed. The defining property of a group of mutually recursive let
bindings is that the binders bound in that group can occur, without restrictions, in
the bodies of those same binders. The same way that, in a let body, evaluating
a binding that uses some resource exactly once consumes that resource once,
using the binding in its own definition also entails using that resource once.
Consider the following program, that calls a recursive let-bound function
defined in terms of the linear resource |x| and itself:
%
\begin{notyet}
\begin{code}
f6 :: Bool -> a -o a
f6 bool x =
  let go b
        = case b of
           True -> x
           False -> go (not b)
  in go bool
\end{code}
\end{notyet}
%
Function |f6| is semantically linear because, iff it is consumed exactly once,
then |x| is consumed exactly once. We can see this by case analysis on |go|'s argument
\begin{itemize}
\item When |bool| is |True|, we'll use the resource |x|
\item When |bool| is |False|, we recurse by calling |go| on |True|, which in turn will use the resource |x|.
\end{itemize}
In |go|'s body, |x| is used directly in one branch and indirectly in the
other, by recursively calling |go| (which we know will result in using |x|
linearly).
% ROMES:TODO:!!
%\todo{this bit is quite hard to explain. It is some sort of cyclic
%argument -- we kind of assume go uses x linearly s.t. when go itself is used
%then we're using x linearly. recursion...}
%
It so happens that |go| will terminate on any input, and will always consume
|x|. However, termination is not a requirement for a binding to use |x| linearly,
and we could have a similar example in which |go| might never terminate but still
uses |x| linearly if evaluated:
%
\begin{notyet}
\begin{code}
f7 :: Bool -> a -o a
f7 bool x =
  let go b
        = case b of
           True -> x
           False -> go b
  in go bool
\end{code}
\end{notyet}
%
The key to linearity in the presence of non-termination is Linear Haskell's
definition of a linear function: \emph{if a linear function application (|f u|) is
consumed exactly once, then the argument (|u|) is consumed exactly once}.
If |f u| doesn't terminate, it is never consumed, thus the claim holds
vacuously; that's why |f8| typechecks:
\begin{working}
\begin{code}
f8 :: a -o b
f8 x = f8 x
\end{code}
\end{working}
%
If |go| doesn't terminate, we aren't able to compute (nor consume) the result
of |f7|, so we do not promise anything about |x| being consumed (|f7|'s
linearity holds trivially). If it did terminate, it would consume |x| exactly
once (e.g. if |go| was applied to |True|).

Determining the linear resources used in a recursive binding might feel
peculiar since we need to know the linear resources used by the binder to determine the linear resources it uses.
%
The paradoxical definition is difficult to grasp, just how learning that a
function can be defined in terms of itself is perplexing when one is first
introduced to general recursion.
%
Informally, we \emph{assume} the binding will consume some linear resources
exactly once, and use that assumption when reasoning about recursive calls such
that those linear resources are used exactly once.

%This high-level reasoning isn't amenable for a computer that must check
%whether the program is linear.

Generalizing, we need to find a set of linear resources ($\Delta$) that satisfies the recursive equation
\footnote{This set of resources will basically be the least upper bound of the sets of resources used in each
mutually recursive binding scaled by the times each binding was used}
arising from given binding $x$, such that:
\begin{enumerate}
\item Occurrences of $x$ in its own body are synonymous with using all resources in $\Delta$ exactly once,
\item And if the binding $x$ is fully evaluated, then all resources in $\Delta$ are consumed exactly once.
\end{enumerate}
Finding a solution to this equation is akin to finding a (principle) type for a
recursive binding: the binding needs to be given a type such that occurrences of
that binding in its own body typecheck using that type.
%
Foreshadowing, the core system we developed assumes recursive let bindings to
be annotated with a set of resources satisfying the equations; but we also
present an algorithm to determine this solution, and distinguish between an
\emph{inference} and a \emph{checking} phase, where we first determine the
linear resources used by a group of recursive bindings and only then check
whether the binding is linear, in our implementation of checking of recursive
lets.
% ROMES:TODO:!!!
%\todo{We might not need the inferrence phase, somehow. Anyhow it seems like
%our inferrence is not really about determining a solution but more about
%determining how many times each thing gets used}

There might not be a solution to the set of equations. In this case, the
binding undoubtedly entails using a linear resource more than once. For
example, if we use a linear resource |x| in one case alternative, and invoke
the recursive call more than once, we might eventually consume |x| more than
once:
%
\begin{noway}
\begin{code}
f9 :: Bool -> Bool -o Bool
f9 bool x =
  let go b
        = case b of
           True -> x
           False -> go (not b) && go True
  in go bool
\end{code}
\end{noway}
Note that if returned |x| instead of |go bool| in the let body, then, despite
the binding using |x| more than once, we would still consume |x| exactly once,
since recursive bindings are still lazy.

Lastly, we extend our single-binding running example to use two mutually recursive bindings that
depend a linear resource:
\begin{notyet}
\begin{code}
f10 :: Bool -> a -o a
f10 bool x =
  let go1 b
        = case b of
           True -> go2 b
           False -> go1 (not b)
      go2 b
        = case b of
           True -> x
           False -> go1 b
  in go1 bool
\end{code}
\end{notyet}
As before, we must find a solution to the set of equations defined by the
mutually recursive bindings to determine which resources will be consumed.
In this case, |go1| and |go2| both consume |x| exactly once if evaluated.
We additionally note that a strongly connected group of recursive bindings
(i.e. all bindings are transitively reachable from any of them) will always
consume the same set of resources -- if all bindings are potentially reachable,
then all linear resources are too.

% \parawith{Summary}
Summarising, recursive let bindings behave like non-recursive let bindings in
that if they aren't consumed, the resources they depend on aren't consumed
either.  However, recursive let bindings are defined in terms of themselves, so
the set of linear resources that will be consumed when the binder is evaluated
is also defined in terms of itself (we need it to determine what resources are
used when we recurse). We can intuitively think of this set of linear resources
that will be consumed as a solution to a set of equations defined by a group of
mutually recursive bindings, which we are able to reason about without an
algorithm for simpler programs. In our work, the core type system isn't
concerned with deriving said solution, but we present a simple algorithm for
inferring it with our implementation.

\subsubsection{Case expressions}

Finally, we discuss semantic linearity for case expressions, which have been
purposefully left for last as the key ingredient that brings together the
semantic linearity insights developed thus far, because, essentially,
\emph{case expressions drive evaluation} and semantic linearity can only be
understood in function of how expressions are evaluated.

Up until now, the example functions have always linearly transformed linear
resources, taking into consideration how expressions will be evaluated (and thus
consumed) to determine if resources are being used linearly. However, there
have been no examples in which linear resources are \emph{fully consumed} in
the bodies of linear functions. In other words, all example functions so far
return a value that has to be itself consumed exactly once to ensure the linear
argument is, in turn, consumed exactly once -- as opposed to functions whose
application simply needs to be evaluated to guarantee its linear argument is
consumed (functions that return an unrestricted value).
%
%The latter are of particular relevance because linearly-typed abstractions
%usually require such a function to provide newly-created linear resources to.
%
For example, the entry point to the linear array API presented in Linear
Haskell takes such a function as its second argument:
\begin{working}
\begin{code}
newMArray :: Int -> (MArray a -o Unrestricted b) -o b
\end{code}
\end{working}
%(The second argument is a function that consumes |MArray a| linearly
%and returns an unrestricted result -- we don't need to consume said result
%exactly once to guarantee that |MArray a| is used linearly in the function
%body.)

% ROMES:TODO: Há qualquer coisa que não gosto nada! na coerência deste paragrafo

In short, case expressions enable us to consume resources and thus write
functions that fully consume their linear arguments. To understand exactly how,
we turn to the definition of \emph{consuming} a resource from Linear
Haskell~\cite{cite:linearhaskell}:
\begin{itemize}
    \item To consume a value of atomic base type (such as~\texttt{Int} or
        \texttt{Ptr}) exactly once, just evaluate~it.
    \item To consume a function value exactly once, apply it to one argument,
        and consume its result exactly once.
    \item To consume a value of an algebraic datatype exactly once,
        pattern-match on it, and consume all its linear components exactly once.
        % For example, a linear pair (equivalent to $\tensor$) is consumed exactly
        % once if pattern-matched on \emph{and} both the first and second element are
        % consumed once.
\end{itemize}
That is, we can consume a linear resource by fully evaluating it, through case
expressions.

%%% In Section~\ref{sec:generalizing-evaluation-consuming} we generalize the idea
%%% that consuming a resource is deeply tied to evaluation.  Here, we continue
%%% building intuition for semantic linearity, first reviewing how case expressions
%%% evaluate expressions, and then exploring how they consume resources, by way of example.

In Core, case expressions are of the form $\ccase{e_s}{z~\{\ov{\rho_i \to e_i}\}}$,
where $e_s$ is the case \emph{scrutinee}, $z$ is the case \emph{binder}, and
$\ov{\rho_i \to e_i}$ are the case \emph{alternatives}, composed of a pattern
$\rho_i$ and of the corresponding expression $e_i$. Critically:
\begin{enumerate}
\item The case scrutinee is always evaluated to Weak Head Normal Form (WHNF).

\item Evaluating to WHNF an expression that is already in WHNF is a no-op, that is,
no computation whatsoever occurs, unlike evaluating a e.g. function application.

\item The case binder is an alias to the result of evaluating the scrutinee to WHNF.

\item The alternative patterns are always exhaustive, i.e. there always exists
a pattern that matches the WHNF of a value resulting from evaluating the
scrutinee, where a pattern is either a wildcard that matches all expressions
($\_$), or a constructor and its linear and non-linear component binders
($K~\ov{x}\ov{y}$, with $\ov{x}$ as linearly-bound variables and $\ov{y}$ as
unrestricted ones).

\end{enumerate}

% ROMES:TODO: We should talk about WHNF in the background, not here.
% \parawith{WHNF} An expression is in Weak Head Normal Form when

To explore these case properties in the presence of linearity, we start with an
example of a program that constructs a tuple from linear resources then pattern
matches on it, then uses both linearly-bound variables from the tuple pattern
match. This is well-typed in Linear Haskell:
\begin{working}
\begin{code}
f11 :: a -o b -o (a -o b -o c) -> c
f11 x y use = case (x,y) of { (a,b) -> use a b }
\end{code}
\end{working}
What might be more surprising is that a similar program which discards the
pattern variables and instead uses the resources in the scrutinee is also
semantically linear, despite not being accepted by Linear Haskell:
\begin{notyet}
\begin{code}
f12 :: a -o b -o (a -o b -o c) -> c
f12 x y use = case (x,y) of z { (a,b) -> use x y }
\end{code}
\end{notyet}
We justify that |f12| is linear by appealing to property 2 -- since the
expression (the tuple) being scrutinized is already in WHNF, evaluating it will
not consume neither $x$ nor $y$. Even if the tuple was constructed with
two expressions using $x$ and $y$ respectively, no computation would happen
since we aren't using neither $a$ nor $b$ (thereby never forcing the arguments
of the tuple). However, if we did use $a$ in the case body, then $x$ would be unavailable:
\begin{noway}
\begin{code}
f13 :: a -o a -o (a -o a -o c) -> c
f13 x y use = case (x,y) of z { (a,b) -> use a x }
\end{code}
\end{noway}
This idea that $x$ and $a$ are mutually exclusive is the same behind let
bindings being mutually exclusive to the resources that define them.
By forcing the pattern variable (or the let binding), we run the computations
defined in terms of the linear variables used for that constructor argument (or
let binder body), but otherwise, if we don't use those binders, then we don't
run the computation thus no resources are consumed.

A third option for this example is to use the case binder $z$ instead of $a,b$ or $x,y$:
\begin{notyet}
\begin{code}
f14 :: a -o b -o (a -o b -o c) -> c
f14 x y use = case (x,y) of z { (a,b) -> uncurry use z }
\end{code}
\end{notyet}
Again, $z$ is mutually exclusive with $a,b$ and with $x,y$, but at least one of
the three must occur to ensure the linear resources are consumed. In this
example, we can think that using $a$ entails using the resource $x$, $b$ the
resource $y$, and the case binder $z$ entails using both $a$ and $b$.

Dually, consider the scrutinee to be an expression that's not in WHNF, s.t.
evaluating it to WHNF will require doing computation and thus consume linear
resources that are used in it:
\begin{notyet}
\begin{code}
f15 :: a -o b -o (a -o b -o (c,d)) -> (c,d)
f15 x y use = case use x y of z { (a,b) -> z }
\end{code}
\end{notyet}
Unlike when the scrutinee was in WHNF, we can no longer use $x,y$ in the case
alternatives, but we \emph{must} still use either the case binder $z$ or the linear
pattern variables $a,b$, e.g. it would be quite disastrous if any of the following typechecked:
\begin{noway}
\begin{code}
doubleFree :: Ptr -o (Ptr -o Result) -> Result
doubleFree x free = case free x of z { Result v -> free x }
\end{code}
\end{noway}
\begin{noway}
\begin{code}
leakPointer :: Ptr -> ()
leakPointer x = case id x of z { _ -> () }
\end{code}
\end{noway}
%
%whereas this is fine:
%\begin{notyet}
%\begin{code}
%f16 :: Ptr -o (Ptr -o (Value,Int)) -> (Value,Int)
%f16 x free = case K (free x) of z { K y -> free x }
%\end{code}
%\end{notyet}
%
The result of evaluating the scrutinee must be consumed exactly to guarantee
that the resources used in the scrutinee are fully consumed, or risk them being only ``almost'' consumed. Take for example
|use| in |f15| to simply be |(,)|: it is not sufficient for |use x y| to be
evaluated to WHNF to consume |x| and |y|. Otherwise, if all the resources were
considered to be fully consumed after the scrutinee were evaluated in a case
expression, we could simply ignore the pattern variables, effectively
discarding linear resources (for cases such as the |use = (,)| example). In
short, if the scrutinee is not in WHNF we must either consume the case binder
or the linear components of the pattern.

However, we must also consider pattern matches on constructors without
any linear components. If the constructor has no linear fields, it means the
result can be consumed unrestrictedly and, therefore, all linear resources used
in the computation have been fully consumed.
%
Consequently, in a branch of a constructor without linear fields we know the
result of evaluating the scrutinee to be unrestricted, so we can use the case
binder unrestrictedly and refer to it zero or more times. For example, this
program is semantically linear:
\begin{notyet}
\begin{code}
f16 :: () -o ()
f16 x = case x of z { () -> z <> z }
\end{code}
\end{notyet}
A second example of an unrestricted pattern, where |K2| has no fields and |K1|
has one linear field, seems as though it shouldn't typecheck since the resource
$x$ must have been fully consumed to take the |K2| branch, but because the
scrutinee is known to be |K1|, the |K2| branch is \emph{absurd}, so, in reality
any resource could be freely used in that branch, and the example is
\emph{semantically} linear (despite not being seen as so by our system):
%
\begin{limitation}
\begin{code}
f :: a -o a
f x = case K1 x of z { K2 -> x; K1 a -> x }
\end{code}
\end{limitation}
%
This particular example has a known constructor being scrutinized which might
seem like an unrealistic example, but we recall that during the transformations
programs undergo in an optimising compiler, many programs such as this
naturally occur (e.g. if the definition of a function is inlined in the
scrutinee).
%
% However, the scrutinee before evaluation likely isn't in WHNF, thus scrutinee
% resources cannot directly occur in the case alternatives, since they might be
% potentially consumed when evaluating the scrutinee, the program leading up to
% this one wouldn't have been linear (even semantically).

Further exploring that each linear field must be consumed exactly once, and
that resources in WHNF scrutinees aren't consumed, we are able to construct
more contrived examples, the following two of which the first doesn't typecheck
because the same linear field is used twice, but the second one does since it
uses each linear field exactly once (despite pattern matching on the same
components twice)
\begin{noway}
\begin{code}
f w = case w of z
        (a,b) ->
          case (a,b) of z'
            (c,d) ->
               (a,c)
\end{code}
\end{noway}
\begin{notyet}
\begin{code}
f w = case w of z
        (a,b) ->
          case (a,b) of z'
            (c,d) ->
               (a,d)
\end{code}
\end{notyet}

Before last, we consider the default case alternatives, also known as wildcards
(written $\_$), in the presence of linearity: matching against the wildcard
doesn't provide new information, so linearity is seen as before but without
fully consuming the scrutinee linear resources (in non-linear patterns) nor
binding new linear resources (in linear patterns). In short, if the scrutinee
is in WHNF, we can either use the resources from the scrutinee or the case
binder in that alternative, if the scrutinee is not in WHNF, we \emph{must} use
the case binder, as it's the only way to linearly consume the result of
evaluating the scrutinee.

Finally, we discuss the special case of a case expression scrutinizing a
variable $x$:
\[
\lambda x.~\ccase{x}{\_ \to x}
\]
It might seem as though the program is linear:
\begin{itemize}
\item If $x$ is in WHNF, then scrutinizing it is a no-op, and returning $x$
just returns the resource intact.
\item If $x$ is not in WHNF, then scrutinizing it evaluates it to WHNF, and
returning $x$ returns the result of evaluating $x$ that still had to be
consumed exactly once.
\end{itemize}
However, in practice, it depends on the evaluation strategy. If linear function
applications are $\beta$-reduced \emph{call-by-name} (a common practice, as
linear functions use their argument exactly once), and the above function is
considered linear, then an application might duplicate linear resources during
evaluation. For example:
\[
\begin{array}{l}
(\lambda x.~\ccase{x}{\_ \to x})~(free~x)\\
\Longrightarrow_{CBN~\beta-reduction}\\
\ccase{free~x}{\_ \to free~x}
\end{array}
\]
Therefore, the type system we present in the next section, which evaluates
linear applications call-by-name, does not accept the above program as linear.
Foreshadowing, this particular interaction between evaluation and linearity
comes up in the type preservation proof of our program, and is again explored
with the reverse binder swap transformation in
Section~\ref{sec:optimisations-preserve-types-meta}.

% \parawith{Summary}
In summary, case expressions evaluate their scrutinees to WHNF,
introduce a case binder, and bind pattern variables. If the scrutinee is
already in WHNF, all resources occurring in it are still available in the case
alternative, alongside the case binder and the pattern-bound variables. In the
case alternative, either the resources of the scrutinee, the case binder, or
the linearly bound pattern variables must be used exactly once, but mutually
exclusively. For scrutinees not in WHNF, in the case alternative, either the
case binder or the linear pattern variables need to be used, in mutual exclusion.
If the pattern doesn't bind any linear resources, then it may be consumed
unrestrictedly, and therefore the case binder may also be used unrestrictedly.

% This is fine, because |x| in the case alternative is known to be K...
% Variables really are weird case...
% ROMES:TODO: Do something about this?
% \begin{noway}
% \begin{code}
% f :: a -o (a -o K) -o a
% f x use = case x of z { K -> x }
% \end{code}
% \end{noway}

%We used to call NOT IN WHNF "negative" hah
%\begin{code}
%-- This is NOT OK since (g x y) is negative (eliminates g)
%f x y = case g x y of z
%          K1 a b -> (x,y)
%          K2 w   -> (x,y)
%\end{code}

%% The harder harder things

%\todo[inline]{Should we discuss this? It would be fine, but we're not able to see this because of call-by-name substitution}
%\begin{code}
%f x = case x of z { _ -> x }
%\end{code}

%\todo[inline]{the harder ones regarding reverse binder swap. these give some
%intuition, but this is an unsound optimisation in some contexts}
%\begin{code}
%f y = let x = use y
%       in case x of z { K a b -> expensive x; K2 w -> w }
%\end{code}
%
%\begin{code}
%f y = let x = use y
%      let t = expensive x
%       in case x of z { K a b -> t; K2 w -> w }
%\end{code}
%
%\begin{code}
%f x = case x of z { K a b -> expensive x; K2 w -> w }
%
%
%f y = let x = use y
%       in case x of z { K a b -> expensive x; K2 w -> w }
%
%
%f y = let x = use y
%      let t = expensive x
%       in case x of z { K a b -> t; K2 w -> w }
%\end{code}

% ROMES: We don't have time to do this I think
%%%\subsection{Generalizing linearity in function of evaluation\label{sec:generalizing-evaluation-consuming}}
%%%
%%%\todo[inline]{Deixar isto para último, dificil generalizar, se necessario cortar}
%%%
%%%Indeed, as hinted towards in the previous section, there's a deep connection
%%%between \emph{evaluation} and \emph{consuming resources}.
%%%
%%%Definition X.Y: A linear resource is consumed when it is either fully evaluated
%%%(NF) to a value, or when it is returned s.t. an application of that function
%%%being fully evaluated would fully evaluate the resource to a value. Or something like that.
%%%%
%%%Note how this generalizes Linear Haskell's definition of consuming a resource: $\dots$
%%%
%%%\begin{itemize}
%%%\item We could almost say that eventually everything all linear
%%%resources must be evaluated to NF to be consumed, or returned by a function
%%%s.t. a continuation of that function has to evaluate the result to NF., or something.
%%%
%%%\item Discuss our own generalized (call-by-value, call-by-name, etc)
%%%definition of consuming resources by evaluation. Something like, if an
%%%expression is fully evaluated, all linear resources that expression depends on
%%%to compute a result are consumed, or something...
%%%
%%%\item How does this relate to strictness? Reference the section of Linear
%%%Haskell about linearity and strictness, and basically revisit what they say.
%%%
%%%\end{itemize}

% }}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% {{{ Linear Core
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{A Type System for Semantic Linearity in Core\label{sec:main:linear-core}}

In this chapter, we develop a linear calculus $\lambda_\Delta^\pi$, dubbed \emph{Linear Core}, that 
combines the linearity-in-the-arrow and multiplicity polymorphism introduced by
Linear Haskell~\cite{cite:linearhaskell} with all the key features from GHC's Core
language, except for type equality coercions\footnote{We explain a main avenue of
future work, multiplicity coercions, in Section~\ref{sec:future-work}}.
%
Specifically, our core calculus is a linear lambda calculus with algebraic
datatypes, case expressions, recursive let bindings, and multiplicity
polymorphism.

Linear Core makes much more precise the various insights discussed in the
previous chapter by crystallizing them together in a linear type system for which we
prove soundness via the usual preservation and progress theorems. Crucially,
the Linear Core type system accepts all the \emph{semantically linear} example
programs (highlighted with \colorbox{notyet}{\notyetcolorname})
from Section~\ref{sec:semantic-linearity-examples}, which Core currently
rejects.
%
Besides type safety, we prove that multiple optimising Core-to-Core
transformations preserve linearity in Linear Core. These same transformations
don't preserve linearity under Core's current type system. As far as we know,
we are the first to prove optimisations preserve types in a non-strict linear
language.

The first key idea for typing linearity semantically is to delay \emph{consuming a
resource} to when a computation that depends on that resource is effectively
evaluated or returned, by annotating relevant binders with \emph{usage
environments} (\S~\ref{sec:usage-environments}).
%
The second key idea is to have two distinct rules for
case expressions, branching on whether the scrutinee is in Weak Head Normal
Form, and using ``proof irrelevance'' to track resources that are no
longer available but haven't yet been fully consumed (\S~\ref{sec:lc-case-exps}). Additionally, we introduce
tagged resources to split the resources between the pattern-bound variables as
usage environments, encoding that pattern variables jointly ``finish
consuming'' the scrutinee resources.
% 
We also note that despite the focus on GHC Core, the fundamental ideas for
understanding linearity in a call-by-need calculus can be readily applied to
other non-strict languages.

We present Linear Core's syntax and type system iteratively, starting with the
judgements and base linear calculi rules for multiplicity and term lambdas plus
the variable rules ($\S~\ref{sec:base-calculi}$).
%
Then, usage environments, the rule for $\Delta$-bound variables, and rules for
(recursive) let bindings ($\S~\ref{sec:usage-environments}$).
%
Finally, we introduce the rules to typecheck case expressions and alternatives,
along with the key insights to do so, namely branching on WHNF-ness of
scrutinee, proof irrelevant resources, and tagged variables
($\S~\ref{sec:lc-case-exps}$).

% We start by introducing the Core-like language, $\dots$ usage environments as a
% way to encode choice between the way a resource is used $\dots$\todo{$\dots$}
%

%\todo[inline]{Explicar algumas das ideias fundamentais, e apresentar as regras
%iterativamente. Podemos começar com as triviais e avançar para os dois pontos
%mais difíceis : Lets e Cases}

\subsection{Language Syntax and Operational Semantics}

The complete syntax of Linear Core is given by Figure~\ref{fig:full-linear-core-syntax}.
% Figure~\ref{fig:linear-core-types} and Figure~\ref{fig:linear-core-terms}.
%
The types of Linear Core are algebraic datatypes, function types, and
multiplicity schemes to support multiplicity polymorphism: datatypes
($T~\ov{p}$) are parametrised by multiplicities, function types
($\vp\to_\pi\s$) are also annotated with a multiplicity, which can be $1$,
$\omega$ (read \emph{many}), or a multiplicity variable $p$ introduced by a
multiplicity universal scheme ($\forall p.~\vp$).
%
\[
\SyntaxTypes
\]
%
The terms are variables $x,y,z$, data constructors $K$, multiplicity
abstractions $\Lambda p.~e$ and applications $e~\pi$, term abstractions
$\lambda \x[\pi].~e$ and applications $e~e'$, where lambda binders are
annotated with a multiplicity $\pi$ and a type $\s$. Then, there are
non-recursive let bindings $\llet{\xD = e}{e'}$, recursive let bindings
$\lletrec{\ov{\xD = e}}{e'}$, where the overline denotes a set of distinct
bindings $x_1{:}_{\D_1}\s_1\dots x_n{:}_{\D_n}\s_n$ and associated expressions
$e_1\dots e_n$, and case expressions $\ccase{e}{\zD~\{\ov{\rho \to e'}\}}$,
where $z$ is the case binder and the overline denotes a set of distinct
patterns $\rho_1\dots \rho_n$ and corresponding right hand sides $e'_1\dots
e'_n$. Notably, (recursive) let-bound
binders and case-bound binders are annotated
with a so-called \emph{usage environment} $\Delta$ -- a fundamental construct
for type-checking semantic linearity in the presence of laziness we present in
Section~\ref{sec:usage-environments}.
%
Case patterns $\rho$ can be either the \emph{default} or \emph{wildcard}
pattern $\_$, which matches any expression, or a constructor $K$ and a set of
variables that bind its arguments, where each field of the constructor has an
associated multiplicity denoting whether the pattern-bound variables must
consumed linearly (ultimately, in order to consume the scrutinee linearly).
Additionally, the set of patterns in a case expression is guaranteed to be exhaustive,
i.e. there is always at least one pattern which matches the scrutinized expression.
\[
\SyntaxTerms
\]
Linear Core takes the idea of annotating lets with usage environments from the
unpublished Linear Mini-Core document by Arnaud Spiwack et.
all~\cite{cite:minicore}, which first tentatively tackled Core's
linearity issues. We discuss this work in more detail in
Section~\ref{sec:linear-mini-core}.

Datatype declarations $\datatype{T~\overline{p}}{\overline{K:\overline{\sigma
\to_\pi}~T~\overline{p} }}$ involve the name of the type being declared $T$
parametrized over multiplicity variables $\ov{p}$, and a set of the data
constructors $K$ with signatures indicating the type and multiplicity of the
constructor arguments. Note that a linear resource is used many times when a
constructor with an unrestricted field is applied to it, since, dually, pattern
matching on the same constructor with an unrestricted field allows it to be
used unrestrictedly. Programs are a set of declarations and a top-level
expression.

\SyntaxFull

%Linear Core is similar to Core without coercions, and with multiplicity
%abstractions instead of type abstractions.
%The key differences are that lets, recursive lets, case binders and
%pattern-bound variables
%the product of combining Linear Haskell's arrow linearity and
%multiplicity polymorphism with Core's key features, except for type equality
%coercions --
%a linear lambda calculus with algebraic datatypes, case
%expressions, recursive let bindings, and multiplicity abstractions. The

The (small-step) operational semantics of Linear Core are given by
Figure~\ref{fig:linear-core-operational-semantics}. We use call-by-name
evaluation for Linear Core as it captures the non-strict semantics in which
our type system understands linearity, while being simpler to reason about than
call-by-need operational semantics.
% which is traditionally modelled with a
% mutable heap to store \emph{thunks} and the values they are overwritten with.
Furthermore, linear function applications, even in a \emph{call-by-need} system, are
usually reduced \emph{call-by-name} as the function argument is guaranteed to
be used exactly once (thus avoiding unnecessarily allocating memory on the heap
for a redundant \emph{thunk}).
%
Specifically, function applications are reduced by the standard
\emph{call-by-name} $\beta$-reduction, substituting the argument whole by
occurrences of the lambda binder in its body, case expressions evaluate their
scrutinee to WHNF, substituting the result by the case binder and constructor
arguments for pattern-bound bound variables matching on that same constructor.

\input{language-v4/OperationalSemantics}

\subsection{Typing Foundations\label{sec:base-calculi}}

Linear Core ($\lambda^\pi_\Delta$) is a linear lambda calculus akin to Linear
Haskell's $\lambda^q_\to$ in that both have multiplicity polymorphism,
(recursive) let bindings, case expressions, and algebraic data types.
$\lambda^\pi_\Delta$ diverges from $\lambda^q_\to$ primarily when typing lets,
case expressions, and alternatives (in its purpose to type semantic linearity).
%, and secondarily in only treating multiplicity polymorphism superficially.
%
% \todo{And our treatment of multiplicity polymorphism completely ignores
% algebraic treatment of multiplicities with semiring operations!!! Would it be
% sufficient to add a rule for application of variable multiplicity functions?}
%
Otherwise, the base rules of the calculus for, multiplicity and term,
abstraction and application are quite similar. In this section, we present
the linear calculi's typing rules that share much in common with
$\lambda^q_\to$, and, in the subsequent sections, the rules encoding the
novel insights from Linear Core in typing semantic linearity, that were
explored by example in Section~\ref{sec:linearity-semantically}.
%
We note, however, that we handle multiplicity polymorphism differently from
Linear Haskell in ignoring the multiplicity semiring and, instead,
conservatively treating all multiplicity polymorphic functions as linear.
%
The full type system is given by Figure~\ref{fig:linear-core-typing-rules},
with auxiliary judgements given by
Figure~\ref{fig:linear-core-other-judgements}.

\TypingRules
\TypingRulesOther

We start with the main typing judgement. As is customary for linear type
systems, we use two typing environments, an \emph{unrestricted} $\G$ and
a \emph{linear} $\D$ environment.
%
Variables in $\G$ can be freely discarded (\emph{weakened}) and duplicated
(\emph{contracted}), while resources in $\D$ must be used exactly once (hence
can't be discarded nor duplicated). Despite not having explicit weakening and
contraction rules in our system, they are available as admissible rules for
$\G$ (but not for $\D$), since, equivalently (via
\cite{91621fae-5e53-3497-8291-32b2fab5a743}), resources from $\G$ are
duplicated for sub-derivations and may unrestrictedly exist in the variable
rules.
% TODO: Likely rewrite that
%
The main typing judgement reads ``expression $e$ has type $\s \to \vp$ under
the unrestricted environment $\G$ and linear environment $\D$'':
\[
\G;\D \vdash e : \s \to \vp
\]
Occurrences of unrestricted variables from $\G$ are well-typed as long as the linear
environment is empty, while occurrences of linear variables are only well-typed
when the variable being typed is the only resource available in the linear context.
\[
\begin{array}{cc}
\TypeVarOmega & \TypeLinearVar
\end{array}
\]
In both cases, the linear context must contain exactly what is required to
prove the proposition, whereas the unrestricted context may contain arbitrary
variables.
%
Variables in contexts are annotated with their type and multiplicity, so
$\x[\pi]$ is a variable named $x$ of type $\s$ and multiplicity $\pi$.

Linear functions are introduced via the function type ($\s \to_\pi \vp$) with
$\pi = 1$, i.e. a function of type $\s \to_1 \vp$ (or, equivalently, $\s \lolli \vp$)
introduces a linear resource of type $\s$ in the linear environment $\D$ to then type an expression of type $\varphi$.
%
Unrestricted functions are introduced via the function type ($\s \to_\pi \vp$) with $\pi =
\omega$, and the $\lambda$-bound variable is introduced in $\G$:
\[
\begin{array}{cc}
\TypeLamIntroL & \TypeLamIntroW
\end{array}
\]
A linear function application is well-typed if there exists a disjoint split of the
linear resources into $\D,\D'$ s.t. the function and argument, each under a
distinct split, are both well-typed and the argument type matches the
function's expected argument type. Conversely, unrestricted resources are
duplicated and available whole to both sub-derivations.
\[
\TypeLamElimL
\]
An unrestricted function, unlike a linear one, consumes its argument
unrestrictedly (zero or more times). Therefore, in an unrestricted function
application, allowing any linear resources to occur in the argument expression
entails consuming those resources unrestrictedly, since the
variable binding the argument expression could be discarded or used more than
once in the function body. Thus, argument expressions to unrestricted functions must also
be unrestricted, i.e. no linear variables can be used to type them.
\[
\TypeLamElimW
\]
Typing linear and unrestricted function applications separately is less general
than typing applications of functions of any multiplicity $\pi$ by scaling (per the multiplicity
semiring) the multiplicities of the resources used to type the argument by $\pi$, however, our objective of typing semantic
linearity does not benefit much from doing so, and by keeping the simple approach we can have
the linear and unrestricted environments be separate.
% \todo{We might only need multiplicities for the case-of-case definition as it
% exists in ghc?, even then couldn't we do semiring scaling without variables?
% Also, dislike sentence}

Multiplicity abstractions ($\Lambda p.~e$) introduce a multiplicity variable
$p$ (in the unrestricted context), and construct expressions of type
$\forall p.~\dots$, i.e. a type universally quantified over a multiplicity
variable $p$. We note that, in the body of the abstraction, function types annotated
with a $p$ variable and datatype fields with multiplicity $p$ are typed as
thought they are linear functions and linear fields, because $p$ can be
instantiated at both $\omega$ and $1$ (so types using $p$ must be well-typed at both instantiations).
\[
\TypeMultLamIntro
\]
A multiplicity application instantiates a multiplicity-polymorphic type
$\forall p.~\s$ at a particular (argument) multiplicity $\pi$, resulting in an
expression of type $\s$ where occurrences of $p$ are substituted by $\pi$, i.e.
$\s[\pi/p]$.
\[
\TypeMultLamElim
\]
The rule additionally requires that $\pi$ be \emph{well-formed} in order
for the expression to be well-typed, using the judgement $\G \vdash_{mult}
\pi$, where well-formedness is given by $\pi$ either being $1$, $\omega$, or an
in-scope  multiplicity variable in $\G$.
\[
\TypeWellFormedMult
\]

These rules conclude the foundations of our linear calculi. In subsequent
sections we type (recursive) let bindings and case expressions,
accounting for semantic linearity as per the insights construed in
Section~\ref{sec:linearity-semantically}, effectively distilling them into the
key ideas of our work -- encoded as rules.

\subsection{Usage environments\label{sec:usage-environments}}

A \emph{usage environment} $\Delta$ is the means to encode the idea that lazy
bindings don't consume the resources required by the bound expression when
defined, but rather when the bindings themselves are fully consumed.
Specifically, we annotate so-called $\Delta$-bound variables with a \emph{usage
environment} to denote that consuming these variables equates to consuming the
resources in the usage environment $\D$ they are annotated with, where a usage
environment is essentially a multiset of linear resources. $\Delta$-bound
variables are introduced by a handful of constructs, namely, (recursive) let
binders, case binders, and case pattern variables. In the following example, as
per the insights into semantic linearity developed in
Section~\ref{sec:semantic-linearity-examples}, the resources required to
typecheck the body of the binder $u$, $x$ and $y$, are only used if the let-var
$u$ is consumed in the let-body $e$.  Accordingly, the usage environment of
the let-bound $u$ is $\{x,y\}$:
\[
f = \lambda \xl.~\lambda \y[1].~\llet{u = (x,y)}{e}
\]
Furthermore, usage environments guarantee that using a $\Delta$-bound variable
is mutually exclusive with directly using the resources it is annotated with --
using the $\Delta$-bound variable consumes all linear resources listed in its
usage environment, meaning they are no longer available for direct usage.
Dually, using the linear resources directly means they are no longer available
to consume through the usage environment of the $\Delta$-bound variable.

Finally, we note that usage environments bear a strong resemblance to the linear
typing environments to the right of the semicolon in the main typing judgement,
i.e. the environment with the linear resources required to type an expression.
%
In fact, usage environments and linear typing contexts differ only in that the
former are used to annotate variables, while the latter used to type
expressions. Yet, this distinction is slightly blurred after introducing how
typing environments can be moved to usage environments, or otherwise occurs in
rules relating the two.

\subsubsection{\texorpdfstring{$\D$}{Delta}-bound variables}

A $\Delta$-bound variable $u$ is a variable annotated with a usage environment $\Delta$. Crucially, for any $\Delta$-bound variable $u$:
%
\begin{enumerate}
\item Using $u$ is equivalent to using all the linear resources in $\Delta$
%\item $u$ can only be used once, since the $\D$ resources will not be available afterwards
\item Using $u$ is mutually exclusive with using the $\Delta$ resources it depends on elsehow
\item $u$ can be safely discarded as long as the resources in $\Delta$ are consumed elsehow
\end{enumerate}
%
Fortunately, since linear resources must be linearly split across
sub-derivations, (2) follows from (1): consuming the linear resources
in $\Delta$ to type $u$ makes them unavailable in the context of any other
derivation. Therefore, expressions using these resources a second time, directly, or indirectly through the same
(or other) usage environment, is ill-typed, as the resources are
already allocated to the derivation of $u$. Similarly, (3) also follows from
(1), because if the linear resources aren't consumed in the $\Delta$-var
derivation, they must be consumed in an alternative derivation (or otherwise
the expression is ill-typed).

These observations all boil down to one typing rule for $\Delta$-bound
variables, which fundamentally encodes (1), implying the other two bullets:
\[
\TypeVarDelta
\]
The rule reads that an occurrence of a $\Delta$-bound variable is well-typed if
the linear environment is exactly the resources in the usage environment of
that variable.
% ROMES:TODO: Dizer alguma coisa tipo (blurring the distinction between usage envs and typing envs)?

$\D$-variables are always introduced in $\Gamma$ since they can be discarded
and duplicated, despite multiple occurrences of the same $\Delta$-variable not
possibly being well-typed as, ultimately, it would imply non-linear usage of
linear resources.
% multiple occurrences of the same
% $\D$-variable imply non-linear usage of resources that must be used linearly.

\subsubsection{Lazy let bindings}

In Section~\ref{sec:semantic-linearity-examples}, we discussed how linear
resources used in let-bound expressions are only consumed when the same let-bound
expressions are fully evaluated, i.e. linear resources required by let-bound
expressions are consumed lazily.
%
Moreover, resources from a let-bound expression cannot be used \emph{together}
with the variable binding, since said resources would end up being consumed
more than once, violating (semantic) linearity -- the binder has to be used in
mutual exclusion with the linear resources required to type the expression it
binds, and either \emph{must} be used, or we'd be discarding resources.

Indeed, usage environments allow us to encode mutual exclusivity between
alternative ways of consuming linear resources (between $\D$-vars and direct
resource usage). Let-bound variables are the canonical example of a
$\Delta$-bound variable, that is, let-variables bind expressions in which
the resources required to type them are consumed lazily rather than eagerly.
%
Effectively, annotating let-bound variables with a usage environment $\D$
\emph{delays} the consumption of resources to when the variables themselves are
used.

Summarily, let-bindings introduce $\Delta$-variables whose usage environments
are the linear typing environments of the bindings' bodies:
\[
\TypeLet
\]
The rule for non-recursive let bindings splits the linear environment in $\D$
and $\D'$. $\D$ is used to type the body $e$ of the let binding $x$. Perhaps
surprisingly, the resources $\D$ used to type $e$ are still available in the
environment to type the let body $e'$, alongside the unrestricted $x$ binding
annotated with the usage environment $\D$. Ultimately, the resources being
available in $e'$ reflects the fact that typing a lazily bound expression
doesn't consume resources, and the binding $x$ being $\D$-bound reflects that
its usage entails consuming the resources $\D$ the expression $e$ it binds depends on.
% \todo[inline]{Let bindings are hard, if they are used then we use resources. If
% they don't get used then we use no resources! In practice, resources that show
% up in the body of the let must be used, be it by using the let binder, or by
% using them directly. This makes the let binder and the resources in its body
% mutual exclusive.}

% \todo[inline]{Explain the idea of suspended computation, and how resources will
% be consumed to some extent when we force the computation -- also foreshadowing
% that evaluation to WHNF doesn't necessarily consume all resources}

\subsubsection{Recursive let bindings}

Recursive let bindings are very similar to non-recursive ones, the main
exception being that the recursive bindings may be defined in terms of
themselves, and we may have more than one binding. In our system, groups of
recursive let bindings are always assumed to be strongly connected, that is,
all the bindings in a recursive let group are mutually recursive in the sense
that they all (transitively) depend on one another.

As before, recursive let bindings bind expressions \emph{lazily}, so they similarly
introduce a $\D$-variable for each binding, and resources required to type the
let-bindings are still available in the body of the let, to later be consumed
via $\D$-variables or directly, if the let-bindings are unused.
%
However, as shown by example in Section
\ref{sec:semantic-linearity-examples:recursive-lets}, we must consider how
recursive uses of a binder in its own definition entails consuming all resources otherwise
required to type the binder's body.
%, i.e. a least upper bound.
%
Extrapolating to a strongly-connected group of recursive bindings, (mutually)
recursive uses of binders entail consuming all resources
required to type those binders. By definition, those binders in turn
recursively use binders that used them, and thus all the resources otherwise
required to type them.
%
Ultimately, all binders in a strongly-connected group of mutually recursive let
bindings have to be typed with the same linear resources (which are the least
upper bound of resources needed to type the bodies of all binders, accounting
for uses of mutually-recursive binders).
%
% Therefore, we conclude the least upper bound of resources required to type a
% mutually recursive group of let bindings to be the same for any such group that
% is strongly-connected.

The typing rule for recursive groups of bindings leverages our assumption that
all recursive let bindings are strongly connected and exactly the observation
that every binder in a strongly connected group of recursive bindings is typed
with the same linear context. Consequently, all bindings of the recursive group
are introduced as $\D$-vars with the same $\D$ environment -- using any one of
the bindings in a recursive group entails consuming all resources required to
type that same group, which is also why we can use the same linear resources to
type each binder:
\[
\TypeLetRec
\]
Unfortunately, this formulation is ill-suited for a syntax-directed system
(from which an implementation is direct) because determining a particular $\D$
to type and annotate all binders is difficult. We present our system and
metatheory agnostically to the challenge of inferring this linear typing
environment by assuming recursive let expressions are annotated with the
correct typing environment.

In practice, determining this typing environment $\D$ amounts to finding a
least upper bound of the resources needed to type each mutually-recursive
binding that (transitively) uses all binders in the recursive group.
%
We propose a naive algorithm for inferring usage environments of recursive
bindings in Section~\ref{sec:discuss:implementation} orthogonally to the theory
developed in this section.
%
The algorithm is a $O(n^2)$ traversal over the so-called \emph{naive usage
environments} used to type each binding.
%
Inference of usage environments for recursive binding groups bears some
resemblance to the inference of principle types for recursive bindings
traditionally achieved through the Hindley–Milner inference
algorithm~\cite{DBLP:conf/popl/DamasM82}, there might be an opportunity to
develop a better algorithm leveraging existing inference techniques.
%
Despite being a seemingly useful observation, we leave exploring a potential
connection between inference of usage environments and type inference
algorithms as future work.

% We present a naive algorithm for inferring usage environments of recursive bindings in
% Section~\ref{sec:impl:recursive-alg} and leave exploring this potential
% connection as future work.

\subsection{Case Expressions\label{sec:lc-case-exps}}

Case expressions \emph{drive evaluation} --
% thus they are key to realize a type system that understands linearity in the presence of lazy evaluation.
%
%However, the evaluation of case expressions is considerably nuanced,
%
%In lazy let bindings, computations  can be a case
%expression can effectively consume resources rather than just 
%
a case expression \emph{evaluates its scrutinee} to Weak Head Normal Form
(WHNF)~\cite{10.5555/1096899}, \emph{then} selects the case alternative corresponding to the pattern
matching the Weak Head Normal Form of the scrutinee\footnote{In our calculus,
the alternatives are always exhaustive, i.e. there always exists at least one
pattern which matches the scrutinee in its WHNF, so we're guaranteed to have an
expression to progress evaluation.}. An expression in Weak Head Normal Form can
either be:
\begin{itemize}
\item A lambda expression $\lambda x.~e$,
\item or a datatype constructor application $K~\ov{e}$
\end{itemize}
In both cases, the sub-expressions $e$ or $\ov{e}$ occurring in the lambda body
or as constructor arguments needn't be evaluated for the lambda or constructor
application to be in weak head normal form (if otherwise all sub-expressions
were fully evaluated the whole expression would also be in \emph{normal form}).
% (that is why it is called \emph{weak} head normal form).
%
Accordingly, these sub-expressions might still depend on linear resources to be
well-typed (these resources will be consumed when the expression is evaluated).
%
As will be made clear in later sections, we need to devise a specialized typing
judgement for scrutinees that is able to distinguish between terms
in WHNF and terms that are not in WHNF.
%
Following the discussion on expressions in weak head normal form, we present a
typing judgement $\G;\D \Vdash e : \s \gtrdot \ov{\D_i}$, and a rule for each
of the forms given above:
\[
    \TypeWHNFCons
\qquad
    \TypeWHNFLam
\]
This judgement differs from the main typing judgement in that (1) it only
applies to expressions in weak head normal form, and (2) it ``outputs'' (to the right of $\gtrdot$) a
disjoint set of linear environments ($\ov{\D_i}$), where each environment corresponds to the
linear resources used by a sub-expression of the WHNF expression.
%
To type a constructor application $K~\ov{e_\omega e_i}$, where $e_\omega$
are unrestricted arguments and $e_i$ the linear arguments of the
constructor, we split the resources $\D$ into a disjoint set of resources
$\ov{\D_i}$ required to type each linear argument individually and return exactly
that split of the resources; the unrestricted $e_\omega$ expressions must be
typed on an empty linear environment. A lambda expression is typed with
the main typing judgement and trivially ``outputs'' the whole $\D$ environment,
as there is always only a single sub-expression in lambdas (the lambda body).

Recall that the operational semantics encode the evaluation of case expressions as:
%
\begin{tabbing}
$\ccase{e}{z~\{\ov{\rho_i \to e_i}\}} \longrightarrow^* \ccase{e'}{z~\{\ov{\rho_i \to e_i}\}}$\`where $e'$ is in WHNF and $e$ is not\\
$\ccase{K~\ov{e}}{z~\{K~\ov{x} \to e_i\}} \longrightarrow e_i\ov{[e/x]}[K~\ov{e}/z]$\\
$\ccase{e}{z~\{\_ \to e_i\}} \longrightarrow e_i[e/z]$\`where $e$ is in WHNF\\
\end{tabbing}
%
When a scrutinee $K~\ov{e}$ matches a constructor pattern $K~\ov{\x[\pi]}$,
evaluation proceeds in the case alternative corresponding to the matching
pattern, with occurrences of $\ov{x}$ being substituted by $\ov{e}$, and
occurrences of the case binder $z$ substituted by the whole scrutinee
$K~\ov{e}$. Constructors and lambda expressions otherwise match the wildcard
pattern whose alternative body is evaluated only substituting the case binder by the
scrutinee.

We highlight that, when evaluating a case expression, computation only
effectively happens when a scrutinee not in WHNF is evaluated to WHNF. When the
scrutinee is already in WHNF, evaluation continues in the alternative by
substituting in the appropriate scrutinee expressions, but without having
performed any computation.
% (the scrutinee was already in weak head normal form).
%In short, no computation happens if the scrutinee is already in WHNF.
%
In terms of linearity, resources are consumed only if evaluation happens.
Therefore, resources used to type a scrutinee not in
WHNF will be consumed when the case is evaluated, making said resources unavailable in the case
alternatives. Conversely, when the scrutinee is already in WHNF, linear
resources required to type the scrutinee are still available in the alternatives.
The linear resources used by an expression in WHNF are exactly those which
occur to the right of $\gtrdot$ in the WHNF judgement shown above
(corresponding to the resources required to typecheck the lambda body or the
constructor arguments).

%Recalling that patterns are either a wildcard that matches anything or data
%constructors parametrised by variables to bind the constructor arguments in the
%case alternative
%
%Let's consider how resources are consumed when the case exp ...
%
%Patterns data constructors with variables to bind the constructor arguments parametrised
%
%\begin{itemize}
%\item The wildcard pattern matches against anything, and is the only pattern that matches against a lambda function
%\item We can match against the data constructor and 
%\end{itemize}
%
%Considering patterns are either data constructors and variables bounding
%A function (lambda expression) can only be matched against the wildcard pattern $\_$, but 
%
%Evaluation continues in the selected branch by
%substituting, first, the pattern variables by the (possibly unevaluated)
%expressions used as arguments to the

% \todo[inline]{Case expressions are the means by which we do evaluation and
% pattern matching -- when things are scrutinized, we evaluate them (if they
% aren't evaluated -- tag is 0), and then compare the result against multiple
% alternatives}

% \todo[inline]{Item 2.
%     Pattern matching an expression that is evaluated will not consume all
%     the resources that define that computation -- because of laziness, we only
%     evaluate things to WHNF. To fully consume a value, we need to consume all
%     the linear components of that pattern.
% }

\subsubsection{Branching on WHNF-ness}

The dichotomy between evaluation (hence resource usage) of a case expression
whose scrutinee is in weak head normal form, or otherwise, leads to one of our
key insights: we must \emph{branch on weak head normal formed-ness} to
type case expressions.
%
When the scrutinee is already in weak head normal form, the resources are
unused upon evaluation and thus available in the alternatives.
%
When it is not, resources will be consumed and cannot be used in the
alternative.
%
To illustrate, consider a case expression with a scrutinee in weak head normal
form and another whose scrutinee is not:
\[
\begin{array}{ccc}
(1)~\lambda x.~\ccase{K~x}{\_ \to x} &  & (2)~\lambda x.~\ccase{free~x}{\_ \to x}
\end{array}
\]
The first function uses $x$ linearly, while the second does not.
%
Alternatives may also use the case binder or pattern variables, referring to, respectively,
the whole scrutinee (and all resources used to type the scrutinee) or
constructor arguments (and the resources to type each argument).
% Using the case binder entails using all the resources required by the scrutinee, and using a pattern
% variable implies using the resources of the corresponding constructor argument.

Linear resources must be used exactly once, but there are three \emph{competing} ways
to use the resources from a scrutinee in WHNF in a case alternative: directly, via
the case binder, or by using \emph{all} the pattern-bound variables.
%
Recall how $\D$-variables can encode mutual exclusivity between alternative
ways of consuming resources -- it follows that case binders and pattern-bound
variables are another instance of $\D$-bound variables. Intuitively, resources
in a scrutinee that is already in WHNF are only properly consumed when all
(linear) fields of the pattern are used, satisfying the definition of
consuming resources given in Linear Haskell.
%
% We type a case expression whose scrutinee is in weak head normal form with
This suggests the following rule:
\[
\TypeCaseWHNFIntermediate
\]
First, we assert this rule is only applicable to expressions in weak head
normal form. Second, we use the typing judgement for expressions in WHNF
previously introduced to determine the split of resources amongst the scrutinee
sub-expressions. Finally, we type all case alternatives with the same context, using the
$\vdash_{alt}$ judgement. 
%
Specifically:
\begin{itemize}
\item We introduce the case binder $z$ in the environment as a $\D$-bound
variable whose usage environment is the linear resources used to type the
scrutinee.
\item We make all the resources $\ov{\D_i}$ used to type the scrutinee available in the linear typing environment.
\item We annotate the \emph{alt} judgement with the disjoint set of linear resources $\ov{\D_i}$ used
to typecheck the scrutinee sub-expressions.
\item We annotate the judgement with the name of the case binder $z$ and use
the $\Mapsto$ arrow in the judgement -- this is of most importance when typing
the alternative itself, and will be motivated together with the alternative
judgement below.
\end{itemize}
%
Despite the key intuitions of typing a case expressions whose scrutinee is in
WHNF being conveyed by this rule, our treatment of such a case expression is
slightly more involved, as will be presented after discussing cases of
scrutinees not in WHNF.

The alternative judgement $\G;\D \vdash_{alt} \rho \to e :^z_\D \s \Rightarrow
\vp$ is used to type case alternatives, but it encompasses three
``sub-judgements``, distinguished by the arrow that is used:
for alternatives of case expressions whose scrutinee is in WHNF ($\Mapsto$),
for case expressions in which the scrutinee is not in WHNF ($\Rrightarrow$),
and for alternatives agnostic to the WHNF-ness of the scrutinee
($\Rightarrow$), with $\Rightarrow$ also generalizing the other two.
%
Following the $Case_\textrm{WHNF}$ rule in which we use the $\Mapsto$ alternative
judgement, the rule for typing a case alternative whose pattern is a
constructor with $n > 0$ linear components is:
\[
\TypeAltNWHNF
\]
The rule states that, for such a pattern matching a scrutinee already in
WHNF, we introduce the linear components of the pattern as $\D$-bound variables
whose usage environment matches the linear resources required to type the
corresponding constructor argument in the scrutinee, which comes annotated in
the judgement ($\ov{\D_i}$). Unrestricted fields of the constructor are
introduced as unrestricted variables. We note that the typing environment $\D$
always contains the resources $\ov{\D_i}$ in uses of the alternative
judgement.

Secondly, the rule for alternatives that match on the wildcard pattern:
\[
\TypeAltWild
\]
To type a wildcard alternative we simply type the expression with the main
judgement, ignoring all annotations on the judgement; recalling that the case
binder was already introduced in the environment with the appropriate usage
environment by the case expression, rather than in the case alternative rule.

Finally, consider an alternative matching on a case constructor without any
linear components. According to the definition of consuming a resource from
Linear Haskell, the linear resources of a scrutinee matching such a pattern are
fully consumed in the body of the corresponding alternative, since the
scrutinee must have been evaluated to a form that does not have any linear
components.
%
This definition agrees with the intuition we have developed by example in the
previous section, and with the typing rule we devised for alternatives matching
constructors without linear components.

% Indeed, matching on a constructor without
% linear arguments entails that all the scrutinee resources have been fully
% consumed, and thus are no longer available.

Taking into account that case expressions introduce the linear resources of the
scrutinee in the typing environment of all alternatives, and in the usage
environment of the case binder, we must reactively update the typing
environments after matching on such a pattern.
%
The $Alt0$ rule essentially encodes this insight, and is applicable regardless
of the WHNF-ness of the scrutinee (hence the $\Rightarrow$ arrow), as long as
the constructor pattern has no linear fields:
%
\[
\TypeAltZero
\]
The rule deletes the annotated scrutinee environment $\D_s$ from two select environments:
\begin{itemize}

\item The linear typing environment, effectively deleting the resources from
the scrutinee made available here by the case expressions (written
$\D[\cdot/\D_s]$, a substitution of the scrutinee typing environment by the
empty linear environment $\cdot$).

\item The usage environment of the case binder $z$, written $\G[\cdot/\D_s]_z$
to denote replacing the usage environment of the variable $z$ in $\G$, which
is necessarily $\D_s$ (since we always annotate the judgement with the
environment of the scrutinee), by the empty environment.

\end{itemize}
%
The rule faithfully encodes the notion that an expression matching such a
pattern is unrestricted when evaluated to WHNF, implying that all linear
resources have been consumed to produce it, and the result is something that
can be freely discarded or duplicated.
%
It ensures that when we match on an unrestricted pattern we no longer need to
consume the scrutinee resources. Otherwise, for example,
$\ccase{K_1~x}{\{K_2 \to K_2,K_1~y \to K_1~y\}}$ would not be well-typed since
the resource $x$ is not consumed in the first branch.
%
Furthermore, since the case binder in such an alternative refers to the
unrestricted expression, the case binder too may be used unrestrictedly, which
we allow by making its usage environment empty.

It might seem as though deleting the resources from the environment in this
rule is necessary to guarantee a resource is not used after it is consumed.
%
However, let us consider two discrete situations -- pattern matches in a case
expression whose scrutinee is in WHNF, and matches on a case expression whose
scrutinee is \emph{not} in WHNF:
%
\begin{enumerate}
\item When the scrutinee is in WHNF, it is either an unrestricted expression
against which any match will only introduce unrestricted variables, or an
expression that depends on linear resources. The first case trivially allows
any resource from the scrutinee in the alternatives as well. The second is
further divided:
\begin{enumerate}

\item The pattern is unrestricted while the
scrutinee is not, so entering this branch is impossible as long as the case
expression is well-typed; by contradiction, the linear resources from the
scrutinee could occur unrestrictedly in that branch, since from falsity
anything follows (\emph{ex falso quodlibet}).
%
For uniformity, we type such alternatives as those for scrutinees that are not in WHNF.

\item The pattern is linear and matches the scrutinee, in which case the
$AltN_{\textrm{WHNF}}$ is applicable instead of $Alt0$.

\item The pattern is linear but does not match the scrutinee and so, the same
reasoning as (1) above applies: any resource could theoretically be used in
such alternatives, however, for uniformity, it is also typed as though the
scrutinee were not in WHNF.

\end{enumerate}

\item However, if the scrutinee is not in WHNF, the resources occurring in the
scrutinee will be consumed when evaluation occurs. Therefore, the resources
used in the scrutinee cannot occur in the alternative body
(e.g. $x$ cannot occur in the alternative in $\ccase{close~x}{\{K_1 \to x\}}$)
-- regardless of the pattern.
%
% In fact, the resources from a scrutinee that is not in weak head normal form
% cannot occur in any of the alternatives, even ones matching on constructors
% with linear components, as the resources may have been consumed when evaluating
% the expression to weak head normal form.
%
We guarantee resources from a scrutinee that is not in weak head normal form
cannot occur/directly be used in any case alternative, in our rule for typing
cases not in WHNF, which we introduce below.

\end{enumerate}

% \todo[inline]{The trick here is to separate the case rules into two separate
% rules, one that fires when the scrutinee is in WHNF, the other when it isn't.}

\subsubsection{Proof irrelevant resources}

% \todo[inline]{O que é que eu vou fazer aqui? O que falta é conseguir tratar uma
% case expression de forma flexivel o suficiente. Explorar o scrutinee estar em
% WHNF ou não, agora é preciso usar essas ideias em typing rules.  Só preciso de
% dizer que para tipificar o caso em que não está em WHNF de forma rigorosa num
% sistema de tipos: Manter resources in scope pq é preciso consumir; Mas não
% podem ser usadas diretamente}

Resources used in a scrutinee that is not in weak head normal form must
definitely not be used in the case alternatives
% since they have been used in the evaluation of the scrutinee, as shown in
% the example above.
%
However, it is not sufficient to evaluate the scrutinee to weak head normal
form to \emph{fully} consume all resources used in the scrutinee, since
sub-expressions such as constructor arguments will be left unevaluated. To
\emph{fully} consume all resources occurring in the scrutinee, the scrutinee
must be evaluated either to \emph{normal form} or s.t. all linear components of
the scrutinee are fully evaluated, as witnessed by the $Alt0$ rule. In short,
for a case expression whose scrutinee is not in WHNF:
% We tackle this in due time, in the proof irrelevance section.
\begin{itemize}

\item The scrutinee resources must \emph{not} be used directly in the case alternatives;
\item But the result of evaluating the scrutinee to WHNF must still be
consumed, as all sub-expressions of the scrutinee remain unevaluated and must
be consumed.
\item Since the scrutinee resources cannot be consumed directly, they must be
consumed indirectly through $\D$-variables, namely, either the case binder, or
the linear pattern-bound variables introduced in the alternative.

\end{itemize}

% In alternatives of a case where the scrutinee is not in
% WHNF, we must also consume the result of evaluating the scrutinee to WHNF, but the
% scrutinee resources must definitely not be available for consumption. In
% practice, the result of evaluating the scrutinee must be consumed by using
% either the case binder or all the linear components of a constructor pattern,
% except for patterns matching an unrestricted pattern, which are handled with
% the $Alt0$ rule. For WHNF scrutinees, we encode mutual exclusivity between
% consuming resources directly, with the case binder, or through linear pattern
% variables, by introducing the latter two as $\D$-bound variables.  In essence,
% for the counterpart not-WHNF scrutinees, either the case binder or linear
% pattern-bound variables \emph{must} still be used to guarantee the evaluation
% result is consumed (thus their usage environment cannot be empty), but the
% scrutinee resources cannot be used directly.

We introduce \emph{proof irrelevant} resources, denoted as linear resources
within square brackets $[\D]$, to encode linear resources that cannot be
directly used (the $Var$ rule is not applicable). Proof irrelevant resources
are linear resources in all other senses, meaning they must be used
\emph{exactly once}. However, since proof irrelevant resources cannot be
forgotten neither used directly, they have to be consumed \emph{indirectly} --
by $\D$-bound variables.

To type a case expression whose scrutinee is in weak head normal form, we
type the scrutinee with linear resources $\D$ and type the case
alternatives by introducing the case binder with a usage environment $[\D]$,
having the same proof irrelevant linear context $[\D]$ in the typing
environment, annotating the judgement with the proof irrelevant resources,
and using the $\Rrightarrow$ judgement:
\[
\TypeCaseNotWHNF
\]
Note how the rule is quite similar to the one for scrutinees in WHNF, only
diverging in that the resources in the case binder, typing environment, and
judgement annotation, are made irrelevant.

% In practice, we can't know which resources are consumed by evaluating a given
% expression. The resources become in a limbo state -- they cannot be used
% directly because they might have been consumed, but they mustn't be considered
% as consumed, because they might not have been.  We say these resources enter a
% proof irrelevant state. They must still be tracked as though they weren't
% consumed, but they cannot be used directly to construct the program. How can we
% ensure these proof irrelevant resource variables are fully consumed? With usage
% environments -- for the case binder and for the pattern variables, and
% otherwise propagate

% \todo[inline]{The case binder and pattern variables will consume the scrutinee
% resources, be those irrelevant or relevant resources}

Finally, we recall the tentative $Case_\textrm{WHNF}$ rule presented before and
highlight its flaw: the $\G;\D \vdash_{alt} \rho \to e :^z_{\D_s} \s \Mapsto
\vp$ judgement is only well-defined for patterns $\rho$ matching the WHNF form
of the scrutinee, as the distribution of resources per constructor components
only makes sense for the constructor pattern matching the scrutinee.
Essentially, if the scrutinee is in WHNF the matching alternative is easily
determined, and must be treated with the specialized $\Mapsto$ judgement, which
only applies to the matching constructor.
%
Alternatives not matching the scrutinee, as mentioned in the discussion of the
$Alt0$ rule, could use resources arbitrarily as they will never be executed,
however, we uniformly treat non-matching alternatives as if the scrutinee were
not in WHNF. The rule for typing case alternatives whose scrutinee is in WHNF
is thus given by:
\[
\TypeCaseWHNF
\]
We note that it might seem unusual to specialize a rule for expressions in
WHNF, as programs scrutinizing an expression in WHNF are rarely written by a
developer. Yet, our system is designed to be suitable for optimising compilers
in which intermediate programs commonly scrutinize expressions in WHNF.
%
Foreshadowing, type preservation for the case-reduction substituting the case
binder and pattern variables in the alternative is not possible to prove
without branching on WHNF-ness, since otherwise the $\D$-substitution is not
well-defined.

\subsubsection{Splitting and tagging fragments}

Intuitively, in case alternatives whose scrutinee is not in weak head normal form,
% (and for scrutinees in WHNF which don't match the case alternative)
the proof-irrelevant resources introduced by the case expression must be fully
consumed, either via the case binder $z$, or by using all linear pattern-bound
variables (for uniformity, we also treat alternatives that do not match a
scrutinee in WHNF this way).

However, unlike with scrutinees in WHNF, the resources used by a
scrutinee not in WHNF do not necessarily match those used by each
sub-expression of the expression evaluated to WHNF.
%
Therefore, there is no direct mapping between the usage environments of the
linear pattern-bound variables and the resources used in the scrutinee.
% \todo{WAIT, there never was!
% Only when the K matches the scrutinee. Maybe we need a separate judgement for
% typing other constructors, which would be standalone and show up in this
% section too?}

We introduce \emph{tagged resources} to guarantee all linearly-bound pattern
variables are jointly used to consume all resources occurring in the
environment (in alternative to the case binder), or not at all. Given linear
resources $[\D_s]$ used to type a scrutinee, and a pattern
$K~\ov{x_\omega},\ov{y_i}$ with $i$ linear components, we assign a usage
environment $\D_i$ to each linear pattern variable where, $\D_i$ is obtained from the
scrutinee environment tagged with the constructor name and linear-variable
index $\lctag{\D_s}{K_i}$, and $\y[\D_i]$ is introduced in $\G$.
\[
\TypeAltNNotWHNF
\]
The tag consists of a constructor name $K$ and an index $i$ identifying the
position of the pattern variable among all bound variables in that pattern.
%
The key idea is that a linear resource $x$ can be split into $n$ resources at a
given constructor, where $n$ is the number of positional linear arguments of
the constructor.
%
This is given by the rule:
\[
\TypeVarSplit
\]
By assigning to each linear pattern variable a fragment of the scrutinee
resources with a tag, we guarantee that all linear pattern variables are
simultaneously used to consume all the scrutinee resources, since for any of
scrutinee resources to be used by a linear pattern-bound var be used, the
resources must be $Split$ for the fragments corresponding to that $\D$-var to
be consumed, and, consequently, the remaining fragments have to be consumed
through the other linear pattern-bound variables.
%
For instance, in $\lambda x.~\ccase{x}{z~\{K~a~b\to (a,b)\}}$, where $x$ is a
linear variable, the case alternatives are typed with $\irr{x}$ (proof
irrelevant $x$), $z$ is introduced as $\z[\irr{x}]$, and the pattern variables
are introduced as $\var[a][\lctag{\irr{x}}{K_1}]$ and
$\var[b][\lctag{\irr{x}}{K_2}]$, assuming both components of $K$ are linear.
%
We note how $Split$ can be applied both to relevant and proof irrelevant linear resources in $\D$.

% Using tags for fragments instead of fractions (e.g. $\D_s*i/n$) is necessary
% to guarantee we cannot use the same variable multiple times to consume
% multiple fractions of the resource. It also has the added benefit of allowing
% mixing of pattern variables bound at different alternatives (e.g.~$\lambda
% x~y.~\ccase{(x,y)}{(a,b)\to\ccase{(a,b)}{(z,w)\to(a,w)}}$).

%%%%% \section{Linear Core Examples}
%%%%% 
%%%%% \todo[inline]{If I have no time...}
%%%%% 
%%%%% Linear Mini-Core~\cite{cite:minicore} lists examples of Core programs where
%%%%% semantic linearity must be understood in order for them to be well-typed. In
%%%%% this section, we show those examples in Linear Core ($\lambda^\pi_\Delta$),
%%%%% briefly explaining why they are indeed well-typed.
%%%%% 
%%%%% \paragraph{Equations}
%%%%% 
%%%%% The Linear Haskell function is compiled in Linear Core as\\
%%%%% %
%%%%% \begin{minipage}{0.47\textwidth}
%%%%% \begin{code}
%%%%% data C = Red | Green | Blue
%%%%% f :: C -o C -o C
%%%%% f Red q = q
%%%%% f p Green = p
%%%%% f Blue q = q
%%%%% \end{code}
%%%%% \end{minipage}
%%%%% \begin{minipage}{0.47\textwidth}
%%%%% \[
%%%%% \begin{array}{ll}
%%%%% \lambda p{:}_1C~q{:}_1C.~\ccase{p}{p2{:}_{\{p\}}C \\
%%%%% \{Red \to q \\
%%%%% ; \_ \to \ccase{q}{q2{:}_{\{q\}}C\\
%%%%%   \{Green \to p2\\
%%%%%   ; \_ \to \ccase{p2}{p3{:}_{\{p\}}C \\
%%%%%   \{Blue \to q2\}} \}} \}}
%%%%% \end{array}
%%%%% \]
%%%%% \end{minipage}
%%%%% 
%%%%% \paragraph{Unrestricted Fields}
%%%%% 
%%%%% The following is well-typed:
%%%%% Let $\datatype{K}{K : A \lolli B \to C}$, and $f$:
%%%%% \[
%%%%% \lambda \xl.~\ccase{x}{\var[z][x]~\{ K~a~b \to (z, b) \}}
%%%%% \]
%%%%% 
%%%%% \paragraph{Wildcard}
%%%%% 
%%%%% The following is ill-typed:
%%%%% \begin{code}
%%%%% f = \x -> case x of z { _ -> True }
%%%%% \end{code}
%%%%% 
%%%%% \paragraph{Duplication}
%%%%% 
%%%%% The following is ill-typed:
%%%%% \begin{code}
%%%%% data Foo = Foo A
%%%%% f = \x -> case x of z { Foo a -> (z, a) }
%%%%% \end{code}

% }}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% {{{ Linear Core as a GHC Plugin
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Linear Core as a GHC Plugin\label{sec:discuss:implementation}}

We have implemented the Linear Core type system as a plugin for the Glasgow
Haskell Compiler.
GHC plugins allow developers to inspect and modify programs being compiled by
GHC, at different stages of compilation~\cite{10.1145/3331545.3342599}.
%
In particular, for any given Haskell module, Core plugins run for (and receive
as input) every intermediate program produced in the compilation process, i.e.
from desugaring and after each optimising transformation.
%
This implementation further substantiates our claim that Linear Core is
suitable for the intermediate language of an optimising compiler.

The GHC Linear Core plugin~\cite{cite:linear-core-plugin} implements a
typechecker for Linear Core: given a Core program, our plugin typechecks the
linearity of all expressions bound in that program, failing if a linear
resource is not used \emph{exactly once} according to semantic linearity of the
$(\lambda^\pi_\Delta)$ system.
%
Most notably, the plugin successfully validates linearity throughout the
(optimising) compilation of linearity-heavy libraries, namely
\texttt{linear-base} and \texttt{linear-smc}, except in expressions whose
linearity depends on so-called \emph{multiplicity coercions}, which are an
avenue of future work ($\S$~\ref{sec:future-work}) exploring the intersection
of Linear Core with type equality coercions.
% Furthermore, the implementation accepts all example programs from Chapter~\ref{sec:linearity-semantically}
% deemed well-typed by Linear Core, which are marked with a
% \colorbox{notyet}{\notyetcolorname} background.

% Additionally, we discuss the implementation of the Linear Core type system
% directly in the Glasgow Haskell Compiler.

% This section discusses the implementation of Linear Core as a GHC Plugin, with
% a dash of painful history in the attempt of implementing Linear Core directly
% into GHC.

The implementation of Linear Core as a typechecker does not follow directly
from the description of the type system because Linear Core is not
\emph{syntax-directed}. Specifically, the most challenging features are
splitting linear resources amongst sub-derivations and consuming fragments of
resources through pattern-bound variables.
%
We use the following techniques to tackle the \emph{non-syntax-directedness} of
Linear Core, thus making the system more suitable to implement:
%
\begin{itemize}

\item Instead of non-deterministically splitting linear resources amongst
sub-derivations, we thread input/output linear resources through each
derivation using the resource management for linear logic described
by~\cite{DBLP:journals/tcs/CervesatoHP00}.
% \todo{Escrever uma regra?}

% In a case expression whose scrutinee is not in WHNF,
\item Pattern variables bound in a case alternative, for a scrutinee not in WHNF,
are introduced as $\D$-variables with usage environment $\lctag{\irr{\D}}{K_i}$,
where $\D$ are the scrutinee resources and $K_i$ the tag of that pattern
variable. To use the resources through the pattern-bound $\D$-vars, they must
be first $Split$ into fragments.

We consume tagged fragments of a resource \emph{as needed}, i.e., when a resource,
whose usage environment has a fragmented resource with tag $K_j$, is used, we $Split$
the matching resource according to the constructor $K$ and consume the fragment
$K_j$, rather than eagerly determining which resources need to be fragmented to
do so.
%
We note that it is safe to destructively fragment the resource, i.e. removing
the whole resource and only leaving the split fragments, because resources are
only $Split$ when a fragment is needed and, consequently, if a fragment is
consumed, using the ``whole'' resource as well violates linearity.

\end{itemize}
%
Furthermore, our implementation must infer the usage environments of binders in
a recursive let group before using them to typecheck the let body. We use a
naive $O(n^2)$ algorithm (where $n$ is the number of let bindings) to determine
these usage environments, but discuss an avenue of further research regarding
this inference challenge in Section~\ref{sec:future-work}.

The usage environments of a recursive group of binders (that is not necessarily
strongly-connected) is computed in two separate passes.
%
First, we calculate a \emph{naive environment} by recording the linear
resources semantically used in the binder bodies, while counting \emph{uses}
(not syntactic occurrences) of the mutually recursive let variables.
%
Second, the binder names and corresponding \emph{naive environment} (mapping
each linear resource to $1$ and the recursive variables to the $n$ number of
times they are \emph{used}) in Cartesian pairs are given as input to
Algorithm~\ref{computeRecUsages}, which computes the actual usage environment
of each binder.
%
% TODO: I should probably use the re-computed usageEnvs instead of the naiveUsageEnvs.
% I'm pretty sure it might fail in some inputs if I keep using the naiveUsageEnvs.
\begin{algorithm}
$usageEnvs \gets naiveUsageEnvs.map(fst)$\;
\For{$(bind, U) \in naiveUsageEnvs$}{
    \For{$V \in usageEnvs$}{
        $V \gets sup(V[bind]*U\setminus\{bind\}, V\setminus\{bind\})$
    }
}
\caption{computeRecUsages\label{computeRecUsages}}
\end{algorithm}
%
Intuitively, the algorithm, for each recursive binder, iterates over the
(initially naive) usage environments and substitutes occurrences of the
recursive binders by their corresponding usage environment,
scaled up by the amount of times that recursive binder is used in the
environment being updated. The result is the least upper bound of linear
resources used by each strongly-connected group of mutually recursive bindings
(for each binder of such a group), since all occurrences of the recursive
binders in the usage environments will be substituted away by the corresponding
recursive usage environments.
%
% The high-level description is:
% \begin{itemize}
%     \item Given a list of binders and their naive environment ($(f, F), (g, G),
%     (h, H)$) in which each use of $f, g, h$ is mapped to a count $n$ in $F, G, H$
%
%     \item For each pair of a letrec-bound variable and corresponding usage
%     environment, update all bindings and their usage as described in the
%     algorithm
%
%     \item After iterating through all bound rec vars, all usage environments
%         will be free of recursive bind usages, and hence describe the ``final'' usage environment
% \end{itemize}
%
The complexity of computing a usage environment for $n$ recursive let binders
is quadratic in $n$,
% i.e. the algorithm has $O(n^2)$ complexity,
but this is not an issue since it is uncommon to have more than a handful of
binders in the same recursive let block.

The results of running the Linear Core GHC plugin on large established
libraries focused around linear types are given by
Figure~\ref{fig:core-plugin-res}.
%
We compiled the libraries \texttt{linear-smc}, a library presented in the
work~\cite{10.1145/3471874.3472980} (1500 lines), \texttt{linear-base} (4000
lines), the Haskell standard library for programming with linear types,
comprised of over 100 modules, and
\texttt{priority-sesh}~\cite{10.1145/3471874.3472979} (1400 lines), a
session-types library, using our plugin.
%
We count the number of programs accepted by our implementation, where each
top-level binding in a module counts as a program, and every such binding is
typechecked once per optimisation pass. i.e., we typecheck all intermediate
programs produced by GHC. The total amount of programs rejected by the
typechecker are given in the ``Total Rejected'' column, but we distil these
rejections into ``Unique Rejections'' by removing duplicate rejections (those
of the same program and for the same linearity-violating reason, but occurring
at different stages). We further categorize the unique rejections into ``Linear
modulo Call-by-name'', ``Linear Rejected'', ``Not Linear Rejected'', and
``Unknown Rejected''.
%
Using the plugin, \texttt{linear-base} takes 35 seconds to compile, instead of
20 seconds. We note, however, that the implementation is not performance
conscious, and types every intermediate program from scratch, instead of
maintaining linearity information throughout the pipeline.

Programs ``linear modulo call-by-name'' are a class of programs
which scrutinize a variable, but then uses the variable in the case
alternatives of the case scrutinizing it. As will be made clear in
Section~\ref{sec:reverse-binder-swap-considered-harmful}, these programs can be
understood as linear as long as applications of linear functions binding these
variables are \emph{not} reduced \emph{call-by-name}, because, in doing so,
linear resources are duplicated. In Core, these programs are not seen as
linear, and thus GHC does not not reduce them using a call-by-name evaluation strategy.
% We have a flag in our implementation to accept some of these programs...
%
``Linear Rejected'' programs include different kinds of programs which are
rejected by the Linear Core system, but are still semantically linear. An
example is a program to which a rewrite rule was applied, resulting in an
application of unrestricted function to a linear resource in the first argument
to |build|:
\begin{code}
take :: Int -> Replicator a -o [a]
take = \ (ds :: Int) (r :: Replicator b) ->
     case ds of
       1 -> build (\(c :: b -> a -> a) (n :: a) -> c (extract r) n)
       ...
\end{code}
We know the linear resource is still used linearly because |build| will
always instance the unrestricted function to $(:)$ (read \emph{Cons}), which is
linear. Additionally, these include programs scrutinizing a value of a type of
which all constructors have no linear components, but only matching on the
default alternative, programs where common-sub-expression elimination
substituted more than one occurrence of |Ur x| by |y|, in the body of
scrutinizing |y| (a linear variable), and other programs affected by rewrite rules.
%
``Not Linear Rejected'' indicate programs that we do not understand as linear,
semantically, and are simultaneously rejected by Linear Core and its
implementation. An example, where the last component of |HashMap|, |wwz|,
is linear, but is not being consumed in the case alternative matching on it:
\begin{code}
jssvi :: Ur Bool -o Set Int -o Ur (TestT IO ())
jssvi (a :: Ur Bool) (b :: Set Int)
  = case a of
    Ur ss -> case b  of
      HashMap wwx wwy wwz ->
        jump wjssAd ss
\end{code}
%
Finally, ``Unknown Rejected'' programs are those whose validity we did not
check. These include both programs accepted by Linear Core, but not by its
implementation, and programs that are simply rejected by Linear Core, but which
were not categorized.

\input{prototype/core-plugin-results}

The results indicate our mostly direct implementation of Linear Core is
successful in accepting the vast majority of the thousands of intermediate
programs produced by GHC when compiling libraries that make extensive use of
linear types.
%
The programs rejected by the Linear Core plugin, in \texttt{linear-base},
besides validating that our implementation is faithful to the
$\lambda^\pi_\Delta$ system of insofar as programs that should not be accepted
are deemed ill-typed, provide further insight into the remaining
details required to fully typecheck linearity in a mature optimising compiler.

% Talk about using our plugin on linear-base and other code bases... If I can get
% a few more case studies it would be pretty good. But then it's imperative to
% also use -dlinear-lint and make sure my plugin rejects a few of the examples

%%% \subsection{Consuming tagged resources as needed}
%%% 
%%% As discussed in Section~\ref{}, pattern-bound linear variables are
%%% put in the context with a \emph{tagged} usage environment with the resources of
%%% the scrutinee. In a \emph{tagged} usage environment environment, all resources
%%% are tagged with a constructor and an index into the many fields of the
%%% constructor.
%%% 
%%% In practice, a resource might have more than one tag. For example, in the following
%%% program, after the first pattern match, |a| and |b| have, respectively, usage
%%% environments $\{\lctag{x}{K_1}\}$ and $\{\lctag{x}{K_2}\}$:
%%% \begin{code}
%%% f x = case x of
%%%        K a b -> case a of
%%%         Pair n p -> (n,p)
%%% \end{code}
%%% However, in the following alternative, |n| has usage environment
%%% $\{\lctag{\lctag{x}{K_1}}{Pair_1}\}$ and |p| has
%%% $\{\lctag{\lctag{x}{K_1}}{Pair_2}\}$. To typecheck
%%% |(n,p)|, one has to $Split$ |x| first on |K| and then on |Pair|, in order for
%%% the usage environments to match.
%%% 
%%% In our implementation, we split resources on demand (and don't directly allow
%%% splitting linear resources), i.e. when we use a tagged resource we split the
%%% linear resource in the linear environment (if available), but never split otherwise.
%%% %
%%% Namely, starting on the innermost tag (the closest to the variable name), we
%%% substitute the linear resource for its split fragments, and then we iteratively
%%% further split those fragments if there are additional tags.
%%% %
%%% We note that it is safe to destructively split the resource (i.e. removing the
%%% original and only leaving the split fragments) because we only split resources
%%% when we need to consume a fragment, and as soon as one fragment is consumed
%%% then using the original ``whole'' variable would violate linearity.
%%% 
%%% In the example, if |n| is used, we have to use its usage environment, which in
%%% turn entails using $\lctag{\lctag{x}{K_1}}{Pair_1}$, which has two tags. In this order, we:
%%% \begin{itemize}
%%% \item Split $x$ into $\lctag{x}{K_1}$ and $\lctag{x}{K_2}$
%%% \item Split $\lctag{x}{K_1}$ and $\lctag{x}{K_2}$ into
%%%   \begin{itemize}
%%%   \item $\lctag{\lctag{x}{K_1}}{Pair_1}$ and $\lctag{\lctag{x}{K_1}}{Pair_2}$
%%%   \item Leave $\lctag{x}{K_2}$ untouched, as we only split on demand, and we aren't using a fragment of $\lctag{x}{K_2}$.
%%%   \end{itemize}
%%% \item Consume $\lctag{\lctag{x}{K_1}}{Pair_1}$, the usage environment of $n$, by removing it from the typing environment.
%%% \end{itemize}

%%%\subsection{Merging Linear Core into GHC\label{sec:merging-linear-core}}
%%%
%%%Describe the ticket for linear Core, the pending MRs, and the difficulty in
%%%even annotating the bind site across optimisations regardless of multiplicities.

% }}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% {{{ Metatheory
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Metatheory\label{sec:main:metatheory}}

The $\lambda^\pi_\D$ system is sound: well-typed programs in Linear Core do not
get \emph{stuck}. Besides type safety ($\S$~\ref{sec:type-safety-meta}), we
prove multiple optimising transformations preserve linearity
($\S$~\ref{sec:optimisations-preserve-types-meta}), and prove an auxiliary
result regarding proof irrelevant resources, stating that a case alternative
well-typed in a proof irrelevant context is also well-typed if proof irrelevant
resources are substituted by an arbitrary environment of relevant resources.
%
Additionally, we state our assumptions that outline an isomorphism between
using a linear variable $\xl$ and a $\D$-variable $\xD$ that consumes existing
resources $\D$, for any $\D$.

\subsection{Assumptions}

We use two main assumptions in our proofs, which are dual.
%
First, a program well-typed with a linear variable ($\xl$) is equivalently
well-typed if that same linear variable were instead $\D$-bound ($\xD$) with
usage environment $\D$, $\D$ were available in the linear context instead of
the linear variable.
% , and occurrences of $x$ were substituted by $\D$ in the
% usage environments of other $\D$-vars in $\G$.
%
% First, we state that a linear variable $\xl$ can be replaced by a $\D$-bound
% variable as long as the context
%
\LinearDeltaRelationLemma
%
\noindent Second, a program well-typed with resources $\D$ and $\D$-bound
variable ($\xD$) is equivalently well-typed, as long as $\D$ is consumed
through the use of $x$, if $x$ is moved to the linear context, resources $\D$
are removed from the linear context, and occurrences of $\D$ whole in usage
environments are substituted by $x$ (occurrences of fragments of $\D$ in usage
environments are unimportant since $\D$ was consumed whole by $x$, not by any
of the fragment-using $\D$-vars).

\DeltaLinearRelationLemma

% Intuitively, a linear variable

We additionally state that unrestricted resources are equivalent to $\D$-bound
variables with an empty ($\cdot$) usage environment:

\DeltaUnrestrictedRelationLemma

\subsection{Irrelevance}

As discussed above, proof irrelevant resources are resources that can only be
consumed indirectly, and are used to type case expressions whose scrutinee is
not in WHNF, essentially encoding that the scrutinee resources must be consumed
through the case binder or the linear pattern-bound variables.
%
As a case expression is evaluated, the scrutinee will eventually be in WHNF,
which must then be typed with rule $Case_{\textrm{WHNF}}$.
%
Crucially, these rules must ``work together'' in the system, in the sense that
case expressions typed using the $Case_{\textrm{Not WHNF}}$ rule must also be
well-typed after the scrutinee is evaluated to WHNF, which is then typed using
the $Case_{\textrm{WHNF}}$ rule.

% The type preservation theorem states that a well-typed expression
% remains well-typed in the presence of evaluation. Specifically, when the case
% expression whose scrutinee is evaluated to WHNF is handled in the
% preservation proof.

The \emph{Irrelevance} lemma is required to prove preservation for that
evaluation case. We need to prove that the alternatives of a case expression
typed with proof irrelevant resources are still well-typed when the proof
irrelevant resource is substituted by the scrutinee resources as it is evaluated to WHNF.
In this sense, the \emph{Irrelevance} lemma witnesses the soundness of typing a
case alternative with proof irrelevant resources in a certain context with respect to
typing the same expression with arbitrary resources (we note, however, typing
an alternative with proof irrelevant resources is not complete wrt using
arbitrary resources -- a counter example needs only to use a resource
directly).

\WHNFConvSoundness

\noindent Intuitively, the lemma holds since proof irrelevant resources can
only be used through the case binder or pattern-bound variables. If we
consistently replace the proof irrelevant resources both in the typing
environment and in the usage environments containing them, the expression
remains well-typed. We note that the proof irrelevant resources are always
unique when introduced in such a case alternative (we always take the scrutinee
environment to make irrelevant, and allow nested ``irrelevantness''), so the
case binder has the only occurrence of those resources in the $\G$ environment.
% (being somewhat akin to congruence).
%
The proof is given in Section~\ref{sec:proof:irrelevance}.

\subsection{Type safety\label{sec:type-safety-meta}}

We prove type safety of the Linear Core system via the standard type
preservation and progress results. As is customary, we make use of multiple
substitution lemmas, one for each kind of variable: unrestricted variables
$\xo$, linear variables $\xl$, and $\D$-bound variables $\xD$.

% We start with the auxiliary results, as we will make use of them in a select part of the preservation proof.

\TypePreservationTheorem
%
\noindent Type preservation states that a well-typed expression $e$ that
evaluates to $e'$ remains well-typed under the same context:
%
The proof ($\S$~\ref{sec:proof:type-preservation}) is done by structural induction on the reductions $e \longrightarrow
e'$ from the operational semantics. Most cases are straightforward and usually
appeal to one or more of the substitution lemmas described below. The most
interesting case is that of case expressions whose scrutinee can be further
evaluated -- we branch on whether the scrutinee becomes in WHNF, and invoke the
\emph{Irrelevance} lemma if so.
%
This case guarantees that the separation of rules for treating scrutinees is
consistent, in the sense that a well-typed case expression with a scrutinee not
in WHNF remains well-typed after the scrutinee is evaluated to WHNF.

\ProgressTheorem
%
\noindent Progress states that the evaluation of a well-typed term does not block:
Similarly, progress is proved by induction on typing ($\S$~\ref{sec:proof:progress}).

\subsubsection{Substitution Lemmas}

The preservation and progress theorems depend on multiple substitution lemmas,
one for each kind of variable, as is standard.
% The proofs are given in Section~\ref{sec:proof:substitution-lemmas}.
% The ... themselves diverge from their common formulation, because
% $\D$-variables refer to linearly bound variables, so substitution must take into account

The linear substitution lemma states that a well-typed expression $e$ with a
linear variable $x$ of type $\s$ remains well-typed if
occurrences of $x$ in the $e$ are replaced by an expression $e'$ of the same
type $\s$, and occurrences of $x$ in the linear context and in usage
environments of $\D$-bound variables are replaced by the linear context $\D'$
used to type $e'$:

\LinearSubstitutionLemma

\noindent Where $\G[\D'/x]$ substitutes all occurrences of $x$ in the usage
environments of $\D$-variables in $\G$ by the linear variables in $\D'$.
% ($x$ couldn't appear anywhere besides usage environments of $\D$-bound variables, since $x$ is linear).

The substitution of the resource in the usage environments is illustrated
by the following example. Consider the term $\llet{y = use~x}{y}$ where $use$ and $x$ are free variables:
if we replace occurrences of $x$ by $e'$ (where $\G;\D \vdash e : \s$), then the
``real'' usage environment of $y$ goes from $\{x\}$ to $\D$. If we don't update
the usage environment of $y$ accordingly, we'll ultimately be typing
$y{:}_{\{x\}}\vp$ with $\D$ instead of $x$, which is not valid.

The linear substitution lemma extends to case alternatives as well.
The lemma for substitution of linear variables in case alternatives is similar
to the linear substitution lemma, applied to the case alternative judgement.
% Is slightly more involved, in the sense that there are more environments in which the
% substitution $[\D/x]$ must be applied (for the same reason):
%
\LinearSubstitutionAltsLemma
%
\noindent We further require that the environment annotated in the case
alternative judgement, $\D_s$, is a subset of the environment used to type the
whole alternative $\D_s \subseteq \D$. In all occurrences of the alternative
judgement (in $Case_{\textrm{WHNF}}$ and $Case_{\textrm{Not WHNF}}$), the
environment annotating the alternative judgement is \emph{always} a subset of
the alternative environment.

The substitution lemma for unrestricted variables follows the usual
formulation, with the added restriction (common to linear type systems) that
the expression $e'$ that is going to substitute the unrestricted variable $x$
is typed on an empty linear environment:
%
\UnrestrictedSubstitutionLemma
%
\noindent Similarly, we also prove the substitution of unrestricted variables preserves types on an alternative case expression:
%
\UnrestrictedSubstitutionAltsLemma

Finally, we introduce the lemma stating that substitution of $\D$-bound
variables by expressions of the same type preserves the type of the original
expression.
%
What distinguishes this lemma from traditional substitution lemmas is that the
usage environment $\D$ of the variable $x$ being substituted by expression $e'$
must match exactly the typing environment $\D$ of $e'$ and the
environment of the original expression doesn't change with the substitution:
%
\DeltaSubstitutionLemma
%
\noindent Intuitively, if $x$ is well-typed with $\D$ in $e$, substituting $x$
by an expression $e'$ which is typed in the same environment $\D$ allows the
distribution of resources $\D,\D'$ used to type $e$ across sub-derivations to remain
unchanged. To prove the theorems, we don't need a ``stronger'' substitution of
$\D$-vars lemma (allowing arbitrary resources $\D''$ to type $e'$, as in other
substitution lemmas), as we only ever substitute $\D$-variables by expressions
whose typing environment matches the variables usage environment. However, it
is not obvious whether such a lemma is possible to prove for $\D$-variables
(e.g. let $\G;\D \vdash e :\s$ and $\G; \D' \vdash \llet{x = e'}{x}$, if we
substitute $e$ for $x$ the resources $\D'$ are no longer consumed).

The $\D$-substitution lemma on case alternatives reflects again that the typing
environment of the expression substitution the variable must match its usage
environment. We recall that $\D_s \subseteq \D,\D'$ states that the annotated
environment is always contained in the typing environment, which is true of all
occurrences of this judgement. An alternative formulation of this lemma could
instead explicitly list $\D_s$ as part of the typing environment for the same
effect:

\DeltaSubstitutionAltsLemma

The proofs for substitution lemmas of linear, unrestricted, and
$\D$-variables are available in Section~\ref{sec:proof:substitution-lemmas}.

%TODO! Substitution of proof-irrelevant linear variables preserves typing. The
%term always remains the same because $x$ cannot occur in any term, however, all
%variables that refer to $x$ in their usage environment must now refer the usage env. of the substitee (e.g. $[x] => [\D]$).
%This seems trivial to see correct, since all occurrences are in environments, so we get some equivalence similar to the one we need for the proof of Alt0.
%
%TODO: Multiplicity substitution preserves typing lemma
%
%TODO: Canonical forms lemma
%
%TODO: Corollary of $\Delta$-var subst. for $\ov{\Delta}$
%
%TODO: Constructor app typing:
%If $\Gamma, \Delta \vdash K~\ov{e}$ and $K{:}\ov{\sigma\to\pi}~T~\ov{p} \in \Gamma$ and $\hasnolinearvars{\Gamma}$
%then $\ov{\Gamma, \Delta_i \vdash e_i : \sigma_i}$

\subsection{Optimisations preserve linearity\label{sec:optimisations-preserve-types-meta}}

One of the primary goals of the Linear Core type system is being suitable for
intermediate representations of optimising compilers for lazy languages with
linear types. In light of this goal, we prove that \emph{multiple optimising
transformations} are type preserving in Linear Core, and thus preserve linearity.

The optimising transformations proved sound wrt Linear Core in this section
have been previously explained and motivated in
Section~\ref{sec:core-to-core-transformations}.
%
Transformations are described by an arbitrary well-typed expression with a certain shape, on
the left hand side (lhs) of the arrow $\Longrightarrow$, resulting in an expression on
the right hand side (rhs) that we prove to be well-typed.
%
% For our proofs, we assume the lhs to be a well-typed expression and prove the
% rhs is well-typed as well.
%
For each transformation, we describe the intuition behind the transformation
preserving linearity in our system.

\subsubsection{Inlining}

% To the best of our knowledge, there is no linear type system for which inlining
% preserves linearity\footnote{https://github.com/ghc-proposals/ghc-proposals/blob/master/proposals/0111-linear-types.rst\#id90}

\input{language-v4/proofs/optimizations/Inlining}

\subsubsection{\texorpdfstring{$\beta$}{Beta}-reduction}

\input{language-v4/proofs/optimizations/BetaReduction}

\subsubsection{Case of known constructor}

\input{language-v4/proofs/optimizations/CaseOfKnownConstructor}

\subsubsection{Let floating}

\input{language-v4/proofs/optimizations/LetFloating}

\subsubsection{\texorpdfstring{$\eta$}{Eta}-conversions}

\input{language-v4/proofs/optimizations/EtaConvs}

\subsubsection{Binder Swap}

The binder swap transformation applies to case expressions whose scrutinee is a
single variable $x$, and it substitutes occurrences of $x$ in the case
alternatives for the case binder $z$. If $x$ is a linear resource, $x$ cannot
occur in the case alternatives (as we conservatively consider variables are not
in WHNF), so the substitution preserves types vacuously. Otherwise, $x$ can be
freely substituted by $z$, since $z$ is also an unrestricted resource (it's
usage environment is empty because $x$ is unrestricted).

\input{language-v4/proofs/optimizations/BinderSwap}

\subsubsection{Reverse Binder Swap Considered Harmful\label{sec:reverse-binder-swap-considered-harmful}}

The reverse binder swap transformation substitutes occurrences of the case
binder $z$ in case alternatives by the scrutinee, when the scrutinee is a
variable $x$.

\ReverseBinderSwapTheorem

\noindent This is exactly reverse from what the binder swap transformation
does in hope of eliminating multiple uses of $x$ so as to inline it. However, by
using the scrutinee $x$ instead of the case binder, we might be able to float
out expressions from the alternative using the case binder. For example, we
might float an expensive computation involving $z$ out of the case alternative,
where $z$ is out of scope but $x$ isn't:
\[
\begin{array}{l}
\lambda x.~\lletrec{go~y = \ccase{x}{z~\{(a,b) \to \dots (expensive~z) \dots\} }}{\dots}\\
\Longrightarrow_\textrm{Reverse binder swap}\\
\lambda x.~\lletrec{go~y = \ccase{x}{z~\{(a,b) \to \dots (expensive~x) \dots\} }}{\dots}\\
\Longrightarrow_\textrm{Float out}\\
\lambda x.~\llet{t = expensive~x}{\lletrec{go~y = \ccase{x}{z~\{(a,b) \to \dots t \dots\} }}{\dots}}\\
\end{array}
\]
If $go$ is a loop, by applying the reverse binder swap we now only compute
$expensive~x$ once instead of in every loop iteration.

Despite GHC applying the reverse binder swap transformation to core programs
during Core-to-Core optimisation passes, this optimisation violates linearity
when considered as a transformation on Linear Core programs. In practice, the
optimisation preserves linearity in Core when applied as part of the GHC
transformation pipeline only due to the occurrence analyser being naive with
regard to semantic linearity. Initially, it might seem as though an expression
in which a variable $x$ occurs both in the case scrutinee and in the
alternatives is linear, for example:
\[
\G; \xl \vdash \ccase{x}{\_ \to x} : \s
\]
The reasoning is done by branching on whether $x$ refers to an expression in
WHNF or an unevaluated thunk.
\begin{itemize}
\item If $x$ refers to an unevaluated expression, then scrutinizing it results
in the expression bound by $x$ to be evaluated to WHNF. In a subsequent use of
$x$ in the alternatives, $x$ refers to the evaluated scrutinee in WHNF, which
must be consumed. Since $x$ is just another name (like the case binder) for the
scrutinee in WHNF, we may use it instead of the case binder or pattern
variables.
\item If $x$ already refers to an expression evaluated to WHNF, then
scrutinizing it in a case expression is a no-op, thus we may use it again (in
mutual exclusion with the case binder and pattern variables)
\end{itemize}
%
% This idea would be encoded by a possible rule like $$\TypeCaseVar$$
%
Even though, on its own, it makes intuitive sense that this example indeed uses
$x$ linearly, when considered as part of a complete type system, allowing this
expression to be linear makes the system unsound.

We recall that $\beta$-reduction reduces an application of a linear function
using call-by-name -- if we know the argument is used exactly once, a binding
to share the result of computing the argument is unnecessary, so we instead
substitute the argument expression for the linearly-bound variable in the
$\lambda$-body directly.

Consider the function application $$(\lambda x.~\ccase{x}{\_ \to x})~(use~y)$$
where $y$ is a free linear variable. Assuming $\lambda x.~\ccase{x}{\_ \to x}$
is a linear function by the reasoning above, $\beta$-reduction transforms the
application in $(\ccase{x}{\_ \to x})[use~y/x]$, i.e. $\ccase{use~y}{\_ \to
use~y}$. Now, $y$ is a linear variable \emph{consumed} in the scrutinee of the case
expression, yet it occurs in the body of the case alternative as well --
linearity is violated by using the linear resource $y$ twice.

Essentially, Linear Core would be unsound, and even duplicate resources, if the
above kind of expressions, where linear variable scrutinees occur in the
alternatives body, were well-typed, because of its interaction with the
call-by-name reduction of linear functions. In this sense, the reverse binder
swap is an optimisation that creates ill-typed expressions from well-typed
ones, so it is deemed an invalid optimisation that doesn't preserve types in
our system.

The reverse binder swap is not a problem in the GHC simplifier because of the
weaker notion of linearity understood by occurrence analysis. Occurrence
analysis is a static analysis pass which can be used to determine whether a
lambda application can be $\beta$-reduced call-by-name, and $\ccase{x}{\_ \to
x}$ is \emph{not} seen as using $x$ linearly by the analysis. Thus,
$\beta$-reduction is done with call-by-need on such an expression. If the above
example were reduced with call-by-need:
\[
\begin{array}{l}
(\lambda x.~\ccase{x}{\_ \to x})~(use~y)\\
\Longrightarrow_\textrm{call-by-need $\beta$-reduction}\\
\llet{x = use~y}{\ccase{y}{\_ \to y}}
\end{array}
\]
Then the computation using $y$ would be let-bound, and $y$ used as a scrutinee
variable, which is indeed an expression semantically linear in $x$.

Concluding, in being able to understand more programs as linear, our type
system allows more expressions to be considered linear for $\beta$-reduction
without a let-allocation, however, it makes reverse binder swap an invalid
transformation since its output, when considered linear, might violate
linearity when further optimised.

% (Link to ticket)
% 
% \begin{tabbing}
% (1) All optimisations preserve (semantic) linearity\\
% (2) If a function is (semantically) linear, then we can evaluate it using call-by-name and preserve linearity\\
% (3) Reverse binder swap is an optimisation\\
% (4) If reverse binder swap is applied to a case scrutinizing a linear resource in the body (`e`) of a linear function `f`, then the function is still linear by (1)\\
% (5) If we evaluate `f`, we do it call-by-name because of (2)\\
% (6) Call-by-name substitution of the linear argument in the body of a function has been reverse binder swapped doesn’t preserve linearity\\
% (7) Contradiction: By 3 and 1, `f` is linear after reverse-binder-swap. By 2, we can substitute arguments to `f` call-by-name and preserve linearity, which contradicts with 6 that says call-by-name substitution after reverse binder swap violates linearity\\
% \end{tabbing}

% Conclusion:
% Either we need to forfeit that we can always substitute call-by-name linear
% function applications, or we forfeit that reverse binder swap preserves
% linearity (instead, it preserves a weaker notion of linearity understood by the
% syntatic-occurrence-analyzer)

% \todo[inline]{Reverse-binder-swap is only well-defined in certain scenarios
% where the optimizations don't apply call-by-name beta-reduction after the
% reverse-binder-swap optimization -- otherwise we would duplicate resources.
% In this case, it is not a matter of syntatic vs semantic linearity
% }

% \todo[inline]{
% Mention, from ``Call-by-name, call-by-value, call-by-need and the linear lambda calculus'':
% The call-by-name calculus is not entirely suitable for reasoning about
% functional programs in lazy languages, because the beta rule may copy the
% argument of a function any number of times. The call-by-need calculus uses a
% diferent notion of reduction, observationally equivalent to the call-by-name
% calculus. But call-by-need, like call-by-value, guarantees that the argument
% to a function is not copied before it is reduced to a value.
% }

% It's also interesting to note that reverse-binder-swap preserves linearity under pure call-by-need but not under call-by-name, because
% (In the sense that if EVEN linear functions reduce call-by-need rather than call-by-name, then it would preserve optimisations)

% Reduce call-by-name linear function application
% \begin{code}
% f y = (\x. case x of _ -> x:1) (h y)
% ===>
% f y = (case x of _ -> x)[h y/x]
% ===>
% f y = (case h y of _ -> h y) -- consume y twice.
% \end{code}
% 
% Vs. call-by-need
% \begin{code}
%  (\y = (\x. case x of _ -> x:1) (h y)) e
% ===>
%     let y = e in
%        let x = h y in
%         case x of _ -> x
% ===>
%        let x = h e_v in
%         case x of _ -> x
% ===>
%         case x_v of _ -> x_v
% \end{code}

\subsubsection{Case of Case}

The case of case transformation applies to case expressions whose scrutinee is
another case expression, and returns the innermost case expression transformed by repeating
the outermost case expression in each alternative of the innermost case,
scrutinizing the original alternative body.

Intuitively, since the scrutinee of the outermost case is not in WHNF, no
resources from it can directly occur in the outermost alternatives. By moving
the outermost alternatives inwards with a different scrutinee, the alternatives
remain well-typed because they are typed using either the case binder or the
pattern bound variables, which, by the \emph{Irrelevance} lemma, makes it
well-typed for any scrutinee consuming arbitrary resources. The proof is given in Section~\ref{sec:proof:caseofcase}.

\CaseOfCaseTheorem

% }}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Conclusion\label{sec:discussion}}

Linear Core is an intermediate language with a type system system that
understands (semantic) linearity in the presence of laziness, suitable for
optimising compilers with these characteristics which leverage laziness and
(possibly) linearity in its transformations.

In this chapter, we compare our contributions and Linear Core to related
existing works in the literature,
% (including linearity in a lazy, evaluation models in terms of linearity, and
% the Core system).
%
consider further research
% deemed out of the scope of this work and of the Linear Core type system
(notably, so-called \emph{multiplicity coercions} to handle the
interaction between linearity and \emph{coercions}, a key feature of Core which we
left out our system), and conclude.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% {{{ Related Work
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Related Work}

In this section we discuss related work, namely, Linear
Haskell~\cite{cite:linearhaskell}, Linear Mini-Core~\cite{cite:minicore}, and
linearity-influenced optimising
transformations~\cite{cite:let-floating,peytonjones1997a,cite:linearhaskell}.

% TODO: A brief introduction to the related work section?

% \subsection{Formalization of Core}\todo{terrible paragraph name, and terrible paragraph}
% 
% As such, there exists no formal definition of Core that
% accounts for linearity. In this context, we intend to introduce a linearly typed
% System $F_C$ with multiplicity annotations and typing rules to serve
% as a basis for a linear Core. Critically, this Core linear language
% must account for call-by-need evaluation semantics and be valid in
% light of Core-to-Core optimising transformations.

% \parawith{System FC}

%System $F_C$~\cite{cite:systemfc} (Section~\ref{sec:core}) does not
%account for linearity in its original design, and, to the best of our
%knowledge, no extension to System $F_C$ with linearity and non-strict
%semantics exists.
%

% \begin{itemize}
% \item SystemFC tal como está não tem linearidade de todo
% \item Formalmente nao temos published definição de linearidade no Core
% \item Regras para sistema tipo FC com linearidade
% \item mas uma extensão tipo linear lambda calculus nao consegui exprimir as transformações do core
% \end{itemize}

% Haskell Core's foundational language was imbued with linear types, but it does
% not account for linearity with the whole of the type system
% 
% Multiplicity annotations in SystemFC?
% 
% Rules?

\subsubsection{Linear Haskell\label{sec:related-work-linear-haskell}}

% \todo[inline]{Say something about Linear Haskell's lazy operational semantics,
% but the type system not being concerned with linearity in the presence of
% laziness}

Haskell, contrary to most programming languages with linear types, has existed
for 31 years of its life \emph{without} linear types. As such, the introduction
of linear types to Haskell comes with added challenges that do not exist in
languages that were designed with linear types from the start:
%
\begin{itemize}
    \item Backwards compatibility. The addition of linear types shouldn't break
        all existing Haskell code.
    \item Code re-usability. The linearly-typed part of Haskell's ecosystem and
        its non-linearly-typed counterpart should fit in together and it must be
        possible to define functions readily usable by both sides
        simultaneously.
    \item Future-proofing. Haskell, despite being an
        industrial-strength language, is also a petri-dish for experimentation
        and innovation in the field of programming languages. Therefore, Linear
        Haskell takes care to accommodate possible future features, in
        particular, its design is forwards compatible with affine and dependent
        types.
\end{itemize}
%
Linear Haskell~\cite{cite:linearhaskell} is thus concerned with retrofitting
linear types in Haskell, taking into consideration the above design goals, but
is not concerned with extending Haskell's intermediate language(s),
which presents its own challenges.

Nonetheless, while the Linear Haskell work keeps Core unchanged, its
implementation in GHC does modify and extend Core with linearity/multiplicity
annotations. Core's type system is unable to type \emph{semantic} linearity of
programs (in contrast to \emph{syntactic} linearity), which results in Core
rejecting linear programs resulting from optimising transformations that
leverage the non-strict semantics of Core.
%
Linear Core overcomes the limitations of Core's linear type system derived from
Linear Haskell by understanding semantic linearity in the presence of laziness,
and provably accepts multiple Core-to-Core passes. Linear Core, ultimately, can
also be seen as a system that validates the programs written in Linear Haskell
and are compiled by GHC, by guaranteeing (through typing) that linear resources
are still used exactly once throughout the optimising transformations.

\subsubsection{Linear Mini-Core\label{sec:linear-mini-core}}

Linear Mini-Core~\cite{cite:minicore} is a specification of linear types in
Core as they were being implemented in GHC, and doubles as the (unpublished)
precursor to our work. Linear Mini-Core first observes the incapacity of
Core's type system to accept linear programs after transformations, and first
introduces usage environments for let-bound variables with the same goal of
Linear Core of specifying a linear type system for Core that accepts the
optimising transformations.

We draw from Linear Mini-Core's the rule for non-recursive let expressions and
how let-bound variables are annotated with a usage environment, however, our
work further explores the interaction of laziness with linearity in depth, and
diverges in rules for typing other constructs, notably, case expressions and
case alternatives. Furthermore, unlike Mini-Core, we prove type safety of our
system and that multiple optimising transformations, when applied to Linear
Core programs, preserve linearity as understood by the system.

% \subsection{OutsideIn(X)\label{related-work-gadts}}
% 
% Defines constraint-based type system parametrized over X which does not account
% for local type refinements regarding linearity.
% 
% Se for modificar o typechecker com as multiplicity coercions vou ter de falar
% disto.

% \subsection{Rust}
% 
% Rust has a core based on linear types. Describe Rust's architecture?
% How do they handle linearity plus optimisations
% They probabluy don't typecheck linearity in Core

\subsubsection{Linearity-influenced optimisations}

Core-to-Core transformations appear in many works across the research
literature~\cite{cite:let-floating,peytonjones1997a,santos1995compilation,peytonjones2002secrets,baker-finch2004constructed,maurer2017compiling,Breitner2016_1000054251,sergey_vytiniotis_jones_breitner_2017},
all designed in the context of a typed language (Core) which does not have
linear types. However,
~\cite{cite:let-floating,peytonjones1997a,cite:linearhaskell} observe that
certain optimisations (in particular, let-floating and inlining) greatly
benefit from linearity analysis and, in order to improve those transformation,
linear-type-inspired systems were created specifically for the purpose of the
transformation.

By fully supporting linear types in Core, these optimising transformations
could be informed by the language inherent linearity, and, consequently, avoid
an ad-hoc or incomplete linear-type inference pass custom-built for
optimisations. Additionally, the linearity information may potentially be used
to the benefit of optimising transformations that currently don't take any
linearity into account.

% \begin{itemize}
% \item Transfs. core to core aparecem em vários artigos, e são desenhadas no contexto de uma linguagem tipificada mas que não é linearly typed.
% \item nestes dois artigos é observado que se houvesse a capacidade de explorar linearidade podiamos fazer as coisas de forma diferente
% \item Todas estas optimizaçoes de decadas foram desenhadas sem linear types e há sitios onde linear types podiam ajudar mas não existiam na altura
% \item POdemos usar linear types multiplicitpiadads para lazy language core q definimos para nao ter de fazer sistemas lineares de proposito para optimizações
% \item Ser ad-hoc incompleto ou nao feito de todo
% \end{itemize}

% \parawith{A transformation based optimiser for Haskell}
% They discuss a cardinality analysis based on a linear type system but create (an
% ad-hoc?) one suited. Comparison in the measure of creating optimizations based
% on linearity.
% 
% \parawith{Let-Floating: Moving Bindings to Give Faster Programs\label{sec:rw:let-floating}}
% In the paper~\cite{cite:let-floating}...
% They say they are doing work on a linear type system to identify places where
% the lambda is linearly used and therefore floating-in is beneficial and
% floating-out not as productive.

% \subsection{Call-by-value, call-by-name and call-by-value...}
% 
% The work~\cite{cite:call-by-name-value-and-need-linear-lambda-calculus}...

% }}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% {{{ Future Work
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Future Work\label{sec:future-work}}

In this section we highlight some avenues of further research. Briefly,
these include \emph{multiplicity coercions}, optimisations leveraging
linearity, resource inference for usage environments, and ultimately using
Linear Core in a mature optimising compiler with lazy evaluation and linear
types -- the Glasgow Haskell Compiler. Lastly, we discuss the
% we hypothesize Linear($X$), a linear system parametrized on the evaluation strategy, and the
generalization of Linear Core to the surface Haskell language.

\parawith{Multiplicity Coercions.}
Linear Core doesn't have type equality coercions, a flagship feature of GHC
Core's type system.
%
Coercions, briefly explained in Section~\ref{sec:core},
allow the Core intermediate language to encode a panoply of Haskell source
type-level features such as GADTs, type families or newtypes.
%
In Linear Haskell, multiplicities are introduced as annotations to function
arrows which specify the linearity of the function. In practice,
multiplicities are simply types of kind |Multiplicity|, where |One| and |Many|
are the type constructors of the kind |Multiplicity|; multiplicity polymorphism
follows from type polymorphism, where multiplicity variables are
just type variables. Encoding multiplicities as types allows Haskell programs
to leverage features available for types to naturally extend to multiplicities
as well.
%
Consequently, we might define, e.g., using a GADT |SBool| and a type family
|If|, the function |dep| which is linear in the second argument if the first
argument is |STrue| and unrestricted otherwise:
%
\begin{limitation}
\begin{code}
data SBool :: Bool -> Type where
  STrue :: SBool True
  SFalse :: SBool False

type family If b t f where
  If True t _ = t
  If False _ f = f

dep :: SBool b -> Int %(If b One Many) -> Int
dep STrue x = x
dep SFalse _ = 0
\end{code}
\end{limitation}
%
This example is linear and should be accepted. However, the example is rejected
by the GHC's Core type checker. Critically, Core doesn't currently understand
so-called \emph{multiplicity coercions}. Even though after matching on |STrue|
we have access to a coercion from the function multiplicity $m$ to $1$ ($m \sim
1$), we cannot use this coercion to determine whether the usages of the linear
resources match the multiplicity.
%
Studying the interaction between coercions and multiplicities is a main avenue
of future work for Linear Core.

% In GHC, multiplicity coercions are tracked by issue $19517$\footnote{https://gitlab.haskell.org/ghc/ghc/-/issues/19517}.

% \todo[inline]{Relate to levity polymorphism and runtime rep coercions (Sam)}

\parawith{Optimisations leveraging linearity.}
We only briefly mentioned how linearity can inform optimisations to produce
more performant programs. We leave exploring optimisations unblocked by
preserving linearity in the intermediate language with Linear Core as future
work. Linearity influenced optimising transformations have been also discussed
by Linear Haskell~\cite{cite:linearhaskell} and
in~\cite{cite:let-floating,peytonjones1997a}. An obvious candidate is
\emph{inlining}, which is applied based on heuristics from information provided
by the \emph{cardinality analysis} pass that counts occurrences of bound
variables.  Linearity can be used to non-heuristically inform
the inliner~\cite{cite:linearhaskell}. Additionally, we argue that in Linear
Core accepting more programs as linear there are more chances to use linearity,
in contrast to a linear type system which does not account for lazy evaluation
and thus rejects more programs.

\parawith{Usage environment resource inference.}
In Section~\ref{sec:linearity-semantically}, we explained that the linear
resources used by a group of recursive bindings aren't obvious and must be
consistent with each other (i.e. considering the mutually-recursive calls) as
though the resources used by each binder are the solution to a set determined
by the recursive bindings group.  In Section~\ref{sec:main:linear-core}, we
further likened the challenge of determining usage environments for a recursive
group of bindings to a unification problem as that solved by the Hindley-Milner
type inference algorithm~\cite{DBLP:conf/popl/DamasM82} based on generating and solving
constraints. Even though these are useful observations, our implementation of
Linear Core uses a naive algorithm to determine the usage environments, thereby
leaving as future work the design of a principled algorithm to determine the
usage environments of recursive group of bindings.

\parawith{Linear Core in the Glasgow Haskell Compiler.}
Linear Core is suitable as the intermediate language of an optimising compiler
for a linear and lazy language such as Haskell Core, in that optimising
transformations in Linear Core preserve types \emph{and} linearity, since Linear
Core understands semantic linearity in the presence of laziness, unlike
Core's current type system under which optimisations currently violate
linearity.
%
Integrating Linear Core in the Glasgow Haskell Compiler is one of the ultimate
goals of our work. Core's current type system ignores linearity due to
its limitation in understanding semantic linearity, and our work fills this gap
and would allow Core to be linearly typed all throughout.
%
A linearly typed Core that preserves linearity throughout the optimisation
pipeline of GHC both validates the correctness of the compiler, which is
already achieved to a great extent by preserving (non-linear) types, and
informs optimisations, allowing the compiler to generate more performant programs.

Implementing Linear Core in GHC is a challenging endeavour, since we must account
for all other Core features (e.g. strict constructor fields) and more
optimisations. Despite our initiative in this
direction\footnote{https://gitlab.haskell.org/ghc/ghc/-/issues/23218}, we leave
this as future work.

\parawith{Generalizing Linear Core to Haskell.}
Linear types, despite their compile-time correctness guarantees regarding
resource management, impose a burden on programmers in being a restrictive
typing discipline (witnessed, e.g., by Linear
Constraints~\cite{cite:linearconstraints}). Linear Core eases the restrictions
of linear typing by being more flexible in understanding linearity for lazily
evaluated languages such as Haskell. In this sense, it is an avenue of future
work to apply the ideas from Linear Core to the surface Haskell language.

% \begin{itemize}
% 
% \item Linear(X), a linear type system defined by the underlying definition of
% evaluation (which in turn implies how consuming a resource is defined)
% 
% \item Implementation in Core
% 
% \item Generalization to source level language, being more permissive in the
% handling of resources imposes less burden on the programmer
% 
% \item It's harder to typecheck linearity like this in the source level because
% of the interaction with other source features, but seems feasible and an
% improvement to the usability of linear types. It allows more lazy functional
% programming idioms with linear types (also because laziness and strictness is less well defined as in Core, bc opts)
% 
% \item Beautiful inference algorithm for recursive usage environments -- insight
% that looks like inference for recursive function principle types, but haven't
% figured it out -- connection to type inference / hindley milner
% 
% \item We kind of ignore the multiplicity semiring. We should discuss
% how we don't do ring operations ... but that's kind of wrong.
% 
% % \item We know the case binder to ALWAYS be in WHNF, perhaps there could
% % be some annotation on the case binder s.t. we know nothing happens when we
% % scrutinize it as a single variable
% 
% \item Mechanizing the system and metatheory
% 
% \end{itemize}

% }}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% {{{ Conclusion
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% We give up on this section and instead describe "syntax-directedness" in the implementation chapter
%
%
% \section{Syntax Directed System}
% 
% \todo[inline]{In the other system we assume that the recursive lets are strongly connected, i.e. the expressions always}
% 
% \input{language-v4/SyntaxDirectedSystem}
% 
% \subsection{Inferring usage environments}
% 
% \todo[inline]{The difference between this and the previous section is a bit blurry}
% 
% \todo[inline]{There's one more concern: usage environments aren't readily
% available, especially in recursive lets. We must perform inference of usage
% environments before we can typecheck using them. This is how:}
% 
% \todo[inline]{Rather, we define a syntax directed type system that infers usage environments while checking...}
%

\section{Conclusion}

Optimising compilers with a typed and lazy intermediate language with linear
types (of which GHC is the prime example) leverage laziness to heavily
transform and rewrite programs into simpler forms.
%
However, these optimising transformations push the interaction between
linearity and laziness to the limits where linearity can no longer be seen
syntactically, despite being maintained semantically, in the sense that linear
resources are still used exactly once when the optimised program is run.

In this work we explored linearity in the presence of laziness by example
through the interactions of linear types with lazy (recursive) let bindings and
case expressions that evaluate their scrutinee to Weak Head Normal Form. Most
example programs were linear semantically, but not syntactically.
%
We developed a linear type system, Linear Core, for an intermediate language
akin to GHC Core, with laziness and linearity. In contrast to GHC Core's type
system, or any other linear type system (to the best of our knowledge), our
type system understands semantic linearity, and can thus correctly type a wider
range of linear programs, as those explored in the semantic linearity examples.
%
Crucially, we proved soundness of the type system, and proved multiple
optimising transformations preserve linearity, despite most violating linearity
in other linear type systems. Additionally, we implemented Linear Core as a GHC
plugin to further explore its suitability in the intermediate language of an
optimising compiler.

Concluding, Linear Core is a suitable type system for linear, lazy,
intermediate languages of optimising compilers such as GHC, as it understands
linearity in the presence of laziness s.t. optimisations preserve types and
linearity, and further unblocks optimisations influenced by linearity, e.g.
inlining and call-by-name $\beta$-reduction for applications of (semantically)
linear functions.

% }}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\bibliographystyle{plainyr}
\bibliography{references}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% {{{ Appendix
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\appendix

\section{Type Safety Proofs}

\subsection{Type Preservation\label{sec:proof:type-preservation}}

\input{language-v4/proofs/TypePreservationTheorem}

\subsection{Progress\label{sec:proof:progress}}

\input{language-v4/proofs/ProgressTheorem}

\subsection{Irrelevance\label{sec:proof:irrelevance}}

\input{language-v4/proofs/WHNFConvSoundness}

\subsection{Substitution Lemmas\label{sec:proof:substitution-lemmas}}

\input{language-v4/proofs/LinearSubstitutionLemma}

\input{language-v4/proofs/UnrestrictedSubstitutionLemma}

\input{language-v4/proofs/DeltaSubstitutionLemma}

\section{Optimisations preserve linearity}

Proofs are given inline in Chapter~\ref{sec:optimisations-preserve-types-meta},
with the exception of the proof that \emph{case-of-case} preserves types, which
is lengthier than the others.

\subsection{Case of Case\label{sec:proof:caseofcase}}

\input{language-v4/proofs/optimizations/CaseOfCase}

% }}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

% TODO: - In the letrec case: congratulations on finding an inference algorithm
% (though I will want to see the proof some day, I don't see why it works yet).
% You should clarify that, during linting, you will have a usage environment
% annotation and won't need to run the inference algorithm. This algorithm is
% only needed when you first create a letrec.

% vim: fen fdm=marker

### 2024-01-23 icfp.tex
stopped at newMArray

@@
 \usepackage{listings}
+\usepackage{todonotes}
+\usepackage[justification=centering]{subfig}
 
@@
 
+\newcommand{\incode}[1]{\lstinline{#1}}
+
 % \renewcommand{\Varid}[1]{\textcolor{Sepia}{\id{#1}}}
@@
 
+\definecolor{darkblue}{rgb}{0,0,0.5}
+\definecolor{darkgreen}{rgb}{0,0.3,0}
+\definecolor{darkpink}{rgb}{0.4,0,0.3}
+\definecolor{graygreen}{rgb}{0.3,0.5,0.3}
+\definecolor{grayblue}{rgb}{0.2,0.2,0.6}
+\definecolor{grayred}{rgb}{0.5,0.2,0.2}
+
 \lstset{
@@
   stepnumber=1,
-  numbers=left,
-  numbersep=5pt,
+  %numbers=left,
+  %numbersep=5pt,
   numberstyle=\ttfamily\tiny\color[gray]{0.3},
@@
   showspaces=false,
-  keywordstyle=\bfseries\rmfamily,
+  keywordstyle=\bfseries\rmfamily\color{darkblue}, 
   columns=flexible,
@@
 %
-Linearity in Haskell stands out from linearity in other languages for the
-following reasons, essential to our work:
-
-\begin{itemize}
+Linearity in Haskell stands out from linearity in other languages:
+%\begin{itemize}
     % \item Linear types were only added to the language roughly \emph{31 years
@@
     %     We describe Linear Haskell in detail in Section~\ref{sec:linear-haskell}.
-
-    \item Linear types permeate Haskell down to (its) Core, the
+%
+%\item
+  Linear types permeate Haskell down to (its) Core, the
         intermediate language into which Haskell is
@@
         transformations are very likely to have introduced an error
-        in the resulting program. We present Core and its formal
-        basis, System~$F_C$~\cite{cite:systemfc}, in Section~\ref{sec:core}.
+        in the resulting program.
+        %We present Core and its formal
+        %basis, System~$F_C$~\cite{cite:systemfc}, in Section~\ref{sec:core}.
         % TODO: \item values in rust are linear by default while non-linear is
         % the haskell default?
-
-    \item Both Haskell and its intermediate language Core are \emph{lazily}
+%
+        % \item
+
+        Both Haskell and its intermediate language Core are \emph{lazily}
         evaluated, i.e., expressions in Haskell are only evaluated when needed,
@@
         the resource is not used).
-
+%
         In eagerly evaluated languages, the distinction between syntactic uses of a
@@
         %
-        We review lazy and eager evaluation strategies in Section~\ref{sec:background:evaluation-strategies}.
-
-\end{itemize}
+        %We review lazy and eager evaluation strategies in Section~\ref{sec:background:evaluation-strategies}.
+
+%\end{itemize}
 %
@@
 language for the Glasgow Haskell Compiler,
-%ever since Linear Haskell was implemented,
 both in its expressiveness to completely represent a Haskell with
 linear types and in enabling more extensive compiler validations plus improved optimisations,
-% it is currently impossible to enable linearity checking in Core. The reason is
 linearity is effectively ignored in Core. The reason is not evidently clear:
@@
 We review background concepts fundamental to our work in
-Chapter~\ref{sec:background}, including linear type systems, linear types in
+Section~\ref{sec:background}, including linear type systems, linear types in
 Haskell, evaluation strategies, Core's type system, and multiple optimising
@@
 avenues for further research (highlighting so-called \emph{multiplicity
-coercions}) in Chapter~\ref{sec:discussion}, which concludes the document.
+coercions}) in Section~\ref{sec:discussion}, which concludes the document.
 
@@
 Consider the following program in a functional Haskell-like language,
-where a computation that closes the given |handle| is bound to |x| before the
-|handle| is returned:
-%
-\begin{code}
-f : Handle \lolli Handle
+where a computation that closes the given \lstinline{handle} is bound to \lstinline{x} before the
+\lstinline{handle} is returned:
+%
+\begin{lstlisting}
+f : Handle *' $\lolli$ '* Handle
 f handle = let x = close handle in handle
-\end{code}
-%
-In this seemingly innocent example, the |handle| appears to be closed before
+\end{lstlisting}
+%
+In this example, the \lstinline{handle} appears to be closed before
 returned, whereas in fact the handle will only be closed if the let bound
@@
 The example illustrates that \emph{consuming} a resource is not necessarily
-synonymous with using it syntactically, as depending on the evaluation strategy
-of the language, the computation that closes the handle might or not be
-evaluated, and if it isn't, the |handle| in that unused computation
+synonymous with its syntactic occurrences, since depending on the evaluation strategy
+of the language the computation that closes the handle might not be
+executed, and if so, the \lstinline{handle} in that unused computation
 is not consumed.
@@
 Expanding on this, consider the above example program under distinct
-evaluation strategies:
-%
-\begin{description}
-
-\item[Call-by-value] With \emph{eager} evaluation semantics, the let bound expression |close
-handle| is eagerly evaluated, and the |handle| will be closed before being
+evaluation strategies: under call-by-value, %With \emph{eager}
+% evaluation semantics,
+the let
+  bound expression \lstinline{close handle} is eagerly evaluated, and the \lstinline{handle} will be closed before being
 returned. It is clear that a linear type system should not accept such a
-program since the linear resource |handle| is duplicated -- it is used in a
+program since the linear resource \lstinline{handle} is duplicated -- it is used in a
 computation that closes it, while still being made available to the caller of
-the function.
-
-\item[Call-by-need] On the other hand, with \emph{lazy} evaluation semantics, the let
-bound expression will only be evaluated when the binding |x| is needed. We return
-the |handle| right away, and the let binding is forgotten as it cannot be
-used outside the scope of the function, so the handle is not closed by |f|.
+the function;
+%
+on the other hand, with \emph{call-by-need}, the let
+bound expression will only be evaluated when the binding \lstinline{x} is needed. We return
+the \lstinline{handle} right away, and the let binding is forgotten as it cannot be
+used outside the scope of the function, so the handle is not closed by \lstinline{f}.
 Under the lens of \emph{call-by-need} evaluation, \emph{using} a resource in a
@@
 
-\end{description}
-%
-%Rather, \emph{consuming} a resource is intrinsically defined by the evaluation model of the language.
-%
-% Under distinct evaluation strategies, the above example (and programs in general) have different semantics:
-%
-%
 Intuitively, a computation that depends on a linear resource to produce a
@@
 will not consume that resource.
-
+%
 From this observation, and exploring the connection between computation and evaluation,
@@
 %
-Indeed, under \emph{call-by-value}, syntactic occurrences of a linear resource
-directly correspond to semantically using that resource\footnote{With the minor exception
-of trivial aliases, which don't entail any computation even in
-\emph{call-by-value}. In theory, we could use in mutual exclusion any of the
-aliases to refer to a resource without loss of linearity} because \emph{all}
-expressions are eagerly evaluated -- if all computations are eagerly run, all
-linear resources required by computations are \emph{eagerly consumed}.
+% Indeed, under \emph{call-by-value}, syntactic occurrences of a linear resource
+% directly correspond to semantically using that resource\footnote{With the minor exception
+% of trivial aliases, which don't entail any computation even in
+% \emph{call-by-value}. In theory, we could use in mutual exclusion any of the
+% aliases to refer to a resource without loss of linearity} because \emph{all}
+% expressions are eagerly evaluated -- if all computations are eagerly run, all
+% linear resources required by computations are \emph{eagerly consumed}.
 
@@
 
-\subsubsection{Let bindings}
+\paragraph{Let bindings}
 We start our discussion with non-strict (non-recursive) let bindings, i.e. let bindings whose
@@
 In Core, a let binding entails the creation of a \emph{thunk} that suspends the
-evaluation of the let body (for background, see
-Section~\ref{sec:background:evaluation-strategies}). When the binding itself is evaluated, the \emph{thunk} is
+evaluation of the let body.
+% (for background, see
+% Section~\ref{sec:background:evaluation-strategies})
+When the binding itself is evaluated, the \emph{thunk} is
 \emph{forced} and the evaluation is carried out. The result overwrites the
@@
 
+    \begin{notyet}
+\begin{lstlisting}
+f1 :: (a *' $\lolli$ '* b) -> a *' $\lolli$ '* b
+f1 use x = let y = use x in y
+\end{lstlisting}
+\end{notyet}
 In a linear type system, a non-strict let binding that depends on a linear
-resource |x| doesn't consume the resource as long as the binding isn't
+resource \lstinline{x} doesn't consume the resource as long as the binding isn't
 evaluated -- the suspended computation only uses the resource if it is run. For
-this reason, we can't naively tell whether |x| is consumed just by looking at
+this reason, we can't naively tell whether \lstinline{x} is consumed just by looking at
 the let binding body. In the following example, we assign a computation that
-depends on the resource |x| to a binder, which is then returned:
-%
-\begin{notyet}
-\begin{code}
-f1 :: (a -o b) -> a -o b
-f1 use x =
-  let y = use x
-   in y
-\end{code}
-\end{notyet}
-%
-The linear resource |x| is used exactly once, since it is used exactly once in
+depends on the resource \lstinline{x} to a binder, which is then
+returned.
+
+% \begin{figure}
+%   \begin{minipage}{.33\textwidth}
+%     \begin{notyet}
+% \begin{lstlisting}
+% f1 :: (a *' $\lolli$ '* b) -> a *' $\lolli$ '* b
+% f1 use x = let y = use x in y
+% \end{lstlisting}
+% \end{notyet}
+% \caption{one}
+% \end{minipage}%
+% \begin{minipage}{.33\textwidth}
+%       \begin{notyet}
+% \begin{lstlisting}
+% f2 :: (a *' $\lolli$ '* a) -> a *' $\lolli$ '* a
+% f2 use x = let y = use x in use x
+% \end{lstlisting}
+% \end{notyet}
+% \caption{two}
+% \end{minipage}%
+% \begin{minipage}{.33\textwidth}
+% \begin{notyet}
+% \begin{lstlisting}
+% f3 :: (a *' $\lolli$ '* a) -> Bool -> a *' $\lolli$ '* a
+% f3 use bool x =   let y = use x in
+%    case bool of
+%       True -> x
+%       False -> y
+% \end{lstlisting}
+% \end{notyet}
+% \caption{Three}
+% \end{minipage}%
+% \end{figure}
+
+%
+The linear resource \lstinline{x} is used exactly once, since it is used exactly once in
 the body of the binding and the binding is used exactly once in the let body.
@@
 \begin{notyet}
-\begin{code}
-f2 :: (a -o a) -> a -o a
-f2 use x =
-  let y = use x
-  in use x
-\end{code}
+\begin{lstlisting}
+f2 :: (a *' $\lolli$ '* a) -> a *' $\lolli$ '* a
+f2 use x = let y = use x in use x
+\end{lstlisting}
 \end{notyet}
@@
 \begin{notyet}
-\begin{code}
-f3 :: (a -o a) -> Bool -> a -o a
-f3 use bool x =
-  let y = use x
-  in case bool of
+\begin{lstlisting}
+f3 :: (a *' $\lolli$ '* a) -> Bool -> a *' $\lolli$ '* a
+f3 use bool x = let y = use x in case bool of
       True -> x 
       False -> y
-\end{code}
+\end{lstlisting}
 \end{notyet}
 %
-This example essentially merges |f1| with |f2|, using |x| directly in one
-branch and using |y| in the other. Semantically, this program is linear because
-the linear resource |x| ends up being used exactly once in both case
+This example essentially merges \lstinline{f1} with \lstinline{f2}, using \lstinline{x} directly in one
+branch and using \lstinline{y} in the other. Semantically, this program is linear because
+the linear resource \lstinline{x} ends up being used exactly once in both case
 alternatives, directly or indirectly.
@@
 \begin{noway}
-\begin{code}
-f4 :: (a -o b) -> a -o (b, b)
-f4 use x =
-  let y = use x
-  in (y, y)
-\end{code}
+\begin{lstlisting}
+f4 :: (a *' $\lolli$ '* b) -> a *' $\lolli$ '* (b, b)
+f4 use x = let y = use x in (y, y)
+\end{lstlisting}
 \end{noway}
@@
 despite being effectively computed just once, because said result may still
-contain (parts of) the linear resource. The trivial example is |f4| applied to
-|id| -- the result of computing |id x| is |x|, and |x| must definitely not be
-shared! Indeed, if the result of the computation involving the linear resource
+contain (parts of) the linear resource. The trivial example is \lstinline{f4} applied to
+\incode{id} -- the result of computing \incode{id x} is \incode{x}, and \incode{x} must definitely not be
+shared. Indeed, if the result of the computation involving the linear resource
 was, e.g., an unrestricted integer, then sharing the result would not involve consuming the
@@
 
-Lastly, consider a program which defines two let bindings |z| and |y|, where
-|z| uses |y| which in turn uses the linear resource |x|:
+Lastly, consider a program which defines two let bindings \incode{z} and \incode{y}, where
+\incode{z} uses \incode{y} which in turn uses the linear resource \incode{x}:
 %
 \begin{noway}
-\begin{code}
-f5 :: (a -o a) -> a -o ()
-f5 use x =
-  let y = use x
-  let z = use y
-  in ()
-\end{code}
+\begin{lstlisting}
+f5 :: (a *' $\lolli$ '* a) -> a *' $\lolli$ '* ()
+f5 use x = let y = use x in let z = use y in ()
+\end{lstlisting}
 \end{noway}
 %
-Even though the binding |y| is used in |z|, |x| is still never consumed because
-|z| isn't evaluated in the let body, and consequently |y| isn't evaluated
-either -- never consuming |x|. We use this example to highlight that even for
+Even though the binding \incode{y} is used in \incode{z}, \incode{x} is still never consumed because
+\incode{z} isn't evaluated in the let body, and consequently \incode{y} isn't evaluated
+either -- never consuming \incode{x}. We use this example to highlight that even for
 let bound variables, the syntactic occurrence of a variable isn't enough to
-determine whether it is used. Instead, we ought to think of uses of |y| as
-implying using |x|, and therefore uses of |z| imply using |x|, however, if
-neither is used, then |x| isn't used. Since |x| is effectively discarded, this
+determine whether it is used. Instead, we ought to think of uses of \incode{y} as
+implying using \incode{x}, and therefore uses of \incode{z} imply using \incode{x}, however, if
+neither is used, then \incode{x} isn't used. Since \incode{x} is effectively discarded, this
 example also violates linearity.
@@
 linearly, either the binder occurs exactly once $y$, or the resources $\ov{x}$
-do). We'll later see how we can encode this principle of mutual exclusivity
-between let bindings and their dependencies using so called \emph{usage
+do). We discuss how to encode this principle between let bindings and their dependencies using so called \emph{usage
 environments}, in Section~\ref{sec:usage-environments}.
 
-\subsubsection{Recursive let bindings\label{sec:semantic-linearity-examples:recursive-lets}}
+\vspace{-.5cm}
+\paragraph{Recursive let bindings\label{sec:semantic-linearity-examples:recursive-lets}}
 Second, we look into recursive let bindings. For the most part,
@@
 Consider the following program, that calls a recursive let-bound function
-defined in terms of the linear resource |x| and itself:
+defined in terms of the linear resource \incode{x} and itself:
 %
 \begin{notyet}
-\begin{code}
-f6 :: Bool -> a -o a
-f6 bool x =
-  let go b
-        = case b of
+\begin{lstlisting}
+f6 :: Bool -> a *' $\lolli$ '* a
+f6 bool x = let go b = case b of
                      True -> x
@@
                    in go bool
-\end{code}
+\end{lstlisting}
 \end{notyet}
 %
-Function |f6| is semantically linear because, iff it is consumed exactly once,
-then |x| is consumed exactly once. We can see this by case analysis on |go|'s argument
-\begin{itemize}
-\item When |bool| is |True|, we'll use the resource |x|
-\item When |bool| is |False|, we recurse by calling |go| on |True|, which in turn will use the resource |x|.
-\end{itemize}
-In |go|'s body, |x| is used directly in one branch and indirectly in the
-other, by recursively calling |go| (which we know will result in using |x|
+Function \incode{f6} is semantically linear because, iff it is consumed exactly once,
+then \incode{x} is consumed exactly once. We can see this by case analysis on \incode{go}'s argument:
+%\begin{itemize}%
+%\item
+when \incode{bool} is \incode{True}, we'll use the resource \incode{x};
+%\item
+ when \incode{bool} is \incode{False}, we recurse by calling \incode{go} on \incode{True}, which in turn will use the resource \incode{x}.
+%\end{itemize}
+In \incode{go}'s body, \incode{x} is used directly in one branch and indirectly in the
+other, by recursively calling \incode{go} (which we know will result in using \incode{x}
 linearly).
@@
 %
-It so happens that |go| will terminate on any input, and will always consume
-|x|. However, termination is not a requirement for a binding to use |x| linearly,
-and we could have a similar example in which |go| might never terminate but still
-uses |x| linearly if evaluated:
-%
+
+It so happens that \incode{go} will terminate on any input, and will always consume
+\incode{x}. However, termination is not a requirement for a binding to use \incode{x} linearly,
+and we could have a similar example in which \incode{go} might never terminate but still
+uses \incode{x} linearly if evaluated:
+
 \begin{notyet}
-\begin{code}
-f7 :: Bool -> a -o a
-f7 bool x =
-  let go b
-        = case b of
+\begin{lstlisting}
+f7 :: Bool -> a *' $\lolli$ '*  a
+f7 bool x = let go b = case b of
                      True -> x
@@
                    in go bool
-\end{code}
+          \end{lstlisting}
         \end{notyet}
-%
+
 The key to linearity in the presence of non-termination is Linear Haskell's
-definition of a linear function: \emph{if a linear function application (|f u|) is
-consumed exactly once, then the argument (|u|) is consumed exactly once}.
-If |f u| doesn't terminate, it is never consumed, thus the claim holds
-vacuously; that's why |f8| typechecks:
-\begin{working}
-\begin{code}
-f8 :: a -o b
-f8 x = f8 x
-\end{code}
-\end{working}
-%
-If |go| doesn't terminate, we aren't able to compute (nor consume) the result
-of |f7|, so we do not promise anything about |x| being consumed (|f7|'s
-linearity holds trivially). If it did terminate, it would consume |x| exactly
-once (e.g. if |go| was applied to |True|).
+definition of a linear function: \emph{if a linear function application (\incode{f u}) is
+consumed exactly once, then the argument (\incode{u}) is consumed exactly once}.
+If \incode{f u} doesn't terminate, it is never consumed, thus the claim holds
+vacuously.
+% that's why \incode{f8} typechecks:
+% \begin{working}
+% \begin{lstlisting}
+% f8 :: a *' $\lolli$ '* b
+% f8 x = f8 x
+% \end{lstlisting}
+% \end{working}
+%
+If \incode{go} doesn't terminate, we aren't able to compute (nor consume) the result
+of \incode{f7}, so we do not promise anything about \incode{x} being consumed (\incode{f7}'s
+linearity holds trivially). If it did terminate, it would consume \incode{x} exactly
+once (e.g. if \incode{go} was applied to \incode{True}).
 
@@
 arising from given binding $x$, such that:
-\begin{enumerate}
-\item Occurrences of $x$ in its own body are synonymous with using all resources in $\Delta$ exactly once,
-\item And if the binding $x$ is fully evaluated, then all resources in $\Delta$ are consumed exactly once.
-\end{enumerate}
+%\begin{enumerate}
+(1) Occurrences of $x$ in its own body are synonymous with using all resources in $\Delta$ exactly once;
+(2) and if the binding $x$ is fully evaluated, then all resources in $\Delta$ are consumed exactly once.
+%\end{enumerate}
 Finding a solution to this equation is akin to finding a (principle) type for a
@@
 \begin{noway}
-\begin{code}
-f9 :: Bool -> Bool -o Bool
+\begin{lstlisting}
+f9 :: Bool -> Bool *' $\lolli$ '* Bool
 f9 bool x =
@@
   in go bool
-\end{code}
+\end{lstlisting}
 \end{noway}
@@
 \begin{notyet}
-\begin{code}
-f10 :: Bool -> a -o a
+\begin{lstlisting}
+f10 :: Bool -> a *' $\lolli$ '* a
 f10 bool x =
@@
   in go1 bool
-\end{code}
+\end{lstlisting}
 \end{notyet}
@@
 
-\subsubsection{Case expressions}
-
+\paragraph{Case expressions}
 Finally, we discuss semantic linearity for case expressions, which have been
@@
 \begin{working}
-\begin{code}
-newMArray :: Int -> (MArray a -o Unrestricted b) -o b
-\end{code}
+\begin{lstlisting}
+newMArray :: Int -> (MArray a *' $\lolli$ '* Unrestricted b) *' $\lolli$ '* b
+\end{lstlisting}
 \end{working}

### 2024-01-23 icfp.tex
.

@@
-\documentclass[acmsmall, screen, review]{acmart}
+\documentclass[acmsmall,review]{acmart}
 \PassOptionsToPackage{dvipsnames}{xcolor}
@@
    \country{Something}}
- \email{xxx\@yyy.com}
+ \email{xxx@yyy.com}
 \author{Bernardo Toninho}
@@
   \country{Portugal}}
-\email{btoninho\@fct.unl.pt}
+\email{btoninho@fct.unl.pt}
 %% TODO: AFFILIATION, EMAIL
@@
 %
-We review background concepts fundamental to our work in
-Section~\ref{sec:background}, including linear type systems, linear types in
-Haskell, evaluation strategies, Core's type system, and multiple optimising
-transformations applied by GHC in its compilation process.
+% We review background concepts fundamental to our work in
+% Section~\ref{sec:background}, including linear type systems, linear types in
+% Haskell, evaluation strategies, Core's type system, and multiple optimising
+% transformations applied by GHC in its compilation process.
 %
 We compare our contributions to related work and discuss possible
-avenues for further research (highlighting so-called \emph{multiplicity
-coercions}) in Section~\ref{sec:discussion}, which concludes the document.
+avenues for further research
+% (highlighting so-called
+% \emph{multiplicity coercions})
+in Section~\ref{sec:discussion}, which concludes the document.
 
@@
 The next example exposes the case in which the let binder is ignored in the let
-body. Here, the linear resource |x| is used in |y|'s body and in the let
-body, however, the resource is still used semantically linearly because |y| isn't used at all, thus |x| is consumed just once in the let body:
+body. Here, the linear resource \incode{x} is used in \incode{y}'s body and in the let
+body, however, the resource is still used semantically linearly
+because \incode{y} isn't used at all, thus \incode{x} is consumed just
+once in the let body: 
 %
@@
 from transformations such as inlining, which can substitute out occurrences
-of the binder (e.g. |y| is inlined in the let body).
+of the binder (e.g. \incode{y} is inlined in the let body).
 
@@
 
-\vspace{-.5cm}
+%\vspace{-.5cm}
 \paragraph{Recursive let bindings\label{sec:semantic-linearity-examples:recursive-lets}}
@@
 binding undoubtedly entails using a linear resource more than once. For
-example, if we use a linear resource |x| in one case alternative, and invoke
-the recursive call more than once, we might eventually consume |x| more than
+example, if we use a linear resource \incode{x} in one case alternative, and invoke
+the recursive call more than once, we might eventually consume \incode{x} more than
 once:
@@
 \end{noway}
-Note that if returned |x| instead of |go bool| in the let body, then, despite
-the binding using |x| more than once, we would still consume |x| exactly once,
+Note that if returned \incode{x} instead of \incode{go bool} in the let body, then, despite
+the binding using \incode{x} more than once, we would still consume \incode{x} exactly once,
 since recursive bindings are still lazy.
@@
 f10 bool x =
-  let go1 b
-        = case b of
+  let go1 b = case b of
            True -> go2 b
            False -> go1 (not b)
-      go2 b
-        = case b of
+       go2 b = case b of
            True -> x
@@
 mutually recursive bindings to determine which resources will be consumed.
-In this case, |go1| and |go2| both consume |x| exactly once if evaluated.
+In this case, \incode{go1} and \incode{go2} both consume \incode{x} exactly once if evaluated.
 We additionally note that a strongly connected group of recursive bindings
@@
 
-In short, case expressions enable us to consume resources and thus write
+In short, case expressions enable us to consume resources and so write
 functions that fully consume their linear arguments. To understand exactly how,
@@
 Haskell~\cite{cite:linearhaskell}:
-\begin{itemize}
-    \item To consume a value of atomic base type (such as~\texttt{Int} or
-        \texttt{Ptr}) exactly once, just evaluate~it.
-    \item To consume a function value exactly once, apply it to one argument,
-        and consume its result exactly once.
-    \item To consume a value of an algebraic datatype exactly once,
+%begin{itemize}
+% \item
+to consume a value of atomic base type (such as~\texttt{Int} or
+        \texttt{Ptr}) exactly once, just evaluate it;
+        % \item
+        to consume a function value exactly once, apply it to one argument,
+        and consume its result exactly once;
+        % \item
+        To consume a value of an algebraic datatype exactly once,
         pattern-match on it, and consume all its linear components exactly once.
@@
         % consumed once.
-\end{itemize}
-That is, we can consume a linear resource by fully evaluating it, through case
+%\end{itemize}
+Thus, we can consume a linear resource by fully evaluating it, through case
 expressions.
@@
 $\ov{\rho_i \to e_i}$ are the case \emph{alternatives}, composed of a pattern
-$\rho_i$ and of the corresponding expression $e_i$. Critically:
-\begin{enumerate}
-\item The case scrutinee is always evaluated to Weak Head Normal Form (WHNF).
-
-\item Evaluating to WHNF an expression that is already in WHNF is a no-op, that is,
-no computation whatsoever occurs, unlike evaluating a e.g. function application.
-
-\item The case binder is an alias to the result of evaluating the scrutinee to WHNF.
-
-\item The alternative patterns are always exhaustive, i.e. there always exists
-a pattern that matches the WHNF of a value resulting from evaluating the
-scrutinee, where a pattern is either a wildcard that matches all expressions
-($\_$), or a constructor and its linear and non-linear component binders
-($K~\ov{x}\ov{y}$, with $\ov{x}$ as linearly-bound variables and $\ov{y}$ as
-unrestricted ones).
-
-\end{enumerate}
+$\rho_i$ and of the corresponding expression $e_i$. Critically,
+%\begin{enumerate}
+%\item
+the case scrutinee is always evaluated to Weak Head Normal Form (WHNF);
+%
+% \item
+evaluating an expression that is already in WHNF is a no-op, that is,
+no computation occurs; 
+%
+the case binder is an alias to the result of evaluating the scrutinee
+to WHNF; and,
+the alternative patterns are always exhaustive.
+% i.e. there always exists
+% a pattern that matches the WHNF of a value resulting from evaluating the
+% scrutinee, where a pattern is either a wildcard that matches all expressions
+% ($\_$), or a constructor and its linear and non-linear component binders
+% ($K~\ov{x}\ov{y}$, with $\ov{x}$ as linearly-bound variables and $\ov{y}$ as
+% unrestricted ones).
+
+%\end{enumerate}
 
@@
 \begin{working}
-\begin{code}
-f11 :: a -o b -o (a -o b -o c) -> c
+\begin{lstlisting}
+f11 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
 f11 x y use = case (x,y) of { (a,b) -> use a b }
-\end{code}
+\end{lstlisting}
 \end{working}
@@
 \begin{notyet}
-\begin{code}
-f12 :: a -o b -o (a -o b -o c) -> c
+\begin{lstlisting}
+f12 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
 f12 x y use = case (x,y) of z { (a,b) -> use x y }
-\end{code}
+\end{lstlisting}
 \end{notyet}
-We justify that |f12| is linear by appealing to property 2 -- since the
+We justify that \incode{f12} is linear by appealing to the fact that the
 expression (the tuple) being scrutinized is already in WHNF, evaluating it will
@@
 \begin{noway}
-\begin{code}
-f13 :: a -o a -o (a -o a -o c) -> c
+\begin{lstlisting}
+f13 :: a *' $\lolli$ '* a *' $\lolli$ '* (a *' $\lolli$ '* a *' $\lolli$ '* c) -> c
 f13 x y use = case (x,y) of z { (a,b) -> use a x }
-\end{code}
+\end{lstlisting}
 \end{noway}
@@
 \begin{notyet}
-\begin{code}
-f14 :: a -o b -o (a -o b -o c) -> c
+\begin{lstlisting}
+f14 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
 f14 x y use = case (x,y) of z { (a,b) -> uncurry use z }
-\end{code}
+\end{lstlisting}
 \end{notyet}
@@
 \begin{notyet}
-\begin{code}
-f15 :: a -o b -o (a -o b -o (c,d)) -> (c,d)
+\begin{lstlisting}
+f15 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* (c,d)) -> (c,d)
 f15 x y use = case use x y of z { (a,b) -> z }
-\end{code}
+\end{lstlisting}
 \end{notyet}
@@
 \begin{noway}
-\begin{code}
-doubleFree :: Ptr -o (Ptr -o Result) -> Result
+\begin{lstlisting}
+doubleFree :: Ptr *' $\lolli$ '* (Ptr *' $\lolli$ '* Result) -> Result
 doubleFree x free = case free x of z { Result v -> free x }
-\end{code}
+\end{lstlisting}
 \end{noway}
 \begin{noway}
-\begin{code}
+\begin{lstlisting}
 leakPointer :: Ptr -> ()
 leakPointer x = case id x of z { _ -> () }
-\end{code}
+\end{lstlisting}
 \end{noway}
@@
 that the resources used in the scrutinee are fully consumed, or risk them being only ``almost'' consumed. Take for example
-|use| in |f15| to simply be |(,)|: it is not sufficient for |use x y| to be
-evaluated to WHNF to consume |x| and |y|. Otherwise, if all the resources were
+\incode{use} in \incode{f15} to simply be \incode{(,)}: it is not sufficient for \incode{use x y} to be
+evaluated to WHNF to consume \incode{x} and \incode{y}. Otherwise, if all the resources were
 considered to be fully consumed after the scrutinee were evaluated in a case
 expression, we could simply ignore the pattern variables, effectively
-discarding linear resources (for cases such as the |use = (,)| example). In
+discarding linear resources (for cases such as the \incode{use = (,)} example). In
 short, if the scrutinee is not in WHNF we must either consume the case binder
@@
 result of evaluating the scrutinee to be unrestricted, so we can use the case
-binder unrestrictedly and refer to it zero or more times. For example, this
-program is semantically linear:
+binder unrestrictedly and refer to it zero or more times. For example,
+the program in Figure~\ref{fig:nolin} is semantically linear.
+
+\begin{figure}[t]
+ \begin{minipage}{0.5\textwidth}
 \begin{notyet}
-\begin{code}
-f16 :: () -o ()
+\begin{lstlisting}
+f16 :: () *' $\lolli$ '* ()
 f16 x = case x of z { () -> z <> z }
-\end{code}
+\end{lstlisting}
 \end{notyet}
-A second example of an unrestricted pattern, where |K2| has no fields and |K1|
-has one linear field, seems as though it shouldn't typecheck since the resource
-$x$ must have been fully consumed to take the |K2| branch, but because the
-scrutinee is known to be |K1|, the |K2| branch is \emph{absurd}, so, in reality
+\vspace{-0.5cm}
+\caption{No Linear Fields\label{fig:nolin}}
+\end{minipage}%
+\begin{minipage}{0.5\textwidth}
+\begin{limitation}
+\begin{lstlisting}
+f :: a *' $\lolli$ '* a
+f x = case K1 x of z { K2 -> x; K1 a -> x }
+\end{lstlisting}
+\end{limitation}
+\vspace{-0.5cm}
+\caption{Absurd Branches\label{fig:absurd}}
+\end{minipage}
+\end{figure}
+
+A second example of an unrestricted pattern, where \incode{K2} has no fields and \incode{K1}
+has one linear field, is shown in Figure~\ref{fig:absurd}. Seemingly, the resource
+$x$ must have been fully consumed to take the \incode{K2} branch, but because the
+scrutinee is known to be \incode{K1}, the \incode{K2} branch is \emph{absurd}, so, in reality
 any resource could be freely used in that branch, and the example is
@@
 %
-\begin{limitation}
-\begin{code}
-f :: a -o a
-f x = case K1 x of z { K2 -> x; K1 a -> x }
-\end{code}
-\end{limitation}
+%
 %
@@
 programs undergo in an optimising compiler, many programs such as this
-naturally occur (e.g. if the definition of a function is inlined in the
+naturally occur (e.g.,~if the definition of a function is inlined in the
 scrutinee).
@@
 that resources in WHNF scrutinees aren't consumed, we are able to construct
-more contrived examples, the following two of which the first doesn't typecheck
-because the same linear field is used twice, but the second one does since it
-uses each linear field exactly once (despite pattern matching on the same
-components twice)
+more contrived examples (Figures~\ref{fig:badcase}
+and~\ref{fig:goodcase}). the example in Figure~\ref{fig:badcase} does not typecheck
+because the same linear field is used twice, but the example in
+Figure~\ref{fig:goodcase} does since it
+uses each linear field exactly once, despite pattern matching on the same
+components twice.
+
+\begin{figure}[t]
+  \begin{minipage}{0.5\textwidth}
 \begin{noway}
-\begin{code}
+\begin{lstlisting}
 f w = case w of z
-        (a,b) ->
-          case (a,b) of z'
-            (c,d) ->
-               (a,c)
-\end{code}
+        (a,b) -> case (a,b) of z'
+                        (c,d) -> (a,c)
+\end{lstlisting}
 \end{noway}
+\vspace{-0.5cm}
+\caption{Bad Case\label{fig:badcase}}
+\end{minipage}%
+\begin{minipage}{0.5\textwidth}
 \begin{notyet}
-\begin{code}
+\begin{lstlisting}
 f w = case w of z
-        (a,b) ->
-          case (a,b) of z'
-            (c,d) ->
-               (a,d)
-\end{code}
+        (a,b) -> case (a,b) of z'
+                        (c,d) -> (a,d)
+\end{lstlisting}
 \end{notyet}
-
-Before last, we consider the default case alternatives, also known as wildcards
+\vspace{-0.5cm}
+\caption{Good Case\label{fig:goodcase}}
+\end{minipage}
+\end{figure}
+
+Finally, we consider the default case alternatives, also known as wildcards
 (written $\_$), in the presence of linearity: matching against the wildcard
@@
 the case binder, as it's the only way to linearly consume the result of
-evaluating the scrutinee.
-
-Finally, we discuss the special case of a case expression scrutinizing a
-variable $x$:
-\[
+evaluating the scrutinee. We also discuss the special case of a case
+expression scrutinizing a variable $x$:
+$
 \lambda x.~\ccase{x}{\_ \to x}
-\]
+$. 
 It might seem as though the program is linear:
-\begin{itemize}
-\item If $x$ is in WHNF, then scrutinizing it is a no-op, and returning $x$
-just returns the resource intact.
-\item If $x$ is not in WHNF, then scrutinizing it evaluates it to WHNF, and
+%begin{itemize}
+% \item
+if $x$ is in WHNF, then scrutinizing it is a no-op, and returning $x$
+just returns the resource intact;
+% \item
+if $x$ is not in WHNF, then scrutinizing it evaluates it to WHNF, and
 returning $x$ returns the result of evaluating $x$ that still had to be
 consumed exactly once.
-\end{itemize}
+%\end{itemize}
 However, in practice, it depends on the evaluation strategy. If linear function
@@
 \begin{array}{l}
-(\lambda x.~\ccase{x}{\_ \to x})~(free~x)\\
-\Longrightarrow_{CBN~\beta-reduction}\\
+(\lambda x.~\ccase{x}{\_ \to x})~(free~x)
+\Longrightarrow_{CBN}
 \ccase{free~x}{\_ \to free~x}
@@
 Linear Core takes the idea of annotating lets with usage environments from the
-unpublished Linear Mini-Core document by Arnaud Spiwack et.
-all~\cite{cite:minicore}, which first tentatively tackled Core's
+unpublished Linear Mini-Core document by Spiwack et.
+al~\cite{cite:minicore}, which first tentatively tackled Core's
 linearity issues. We discuss this work in more detail in
@@
 %
-The main typing judgement reads ``expression $e$ has type $\s \to \vp$ under
-the unrestricted environment $\G$ and linear environment $\D$'':
-\[
-\G;\D \vdash e : \s \to \vp
-\]
+The main typing judgement, written $\G;\D \vdash e : \s \to \vp$,
+denotes that expression $e$ has type $\s \to \vp$ under
+the unrestricted environment $\G$ and linear environment $\D$.
+
 Occurrences of unrestricted variables from $\G$ are well-typed as long as the linear
@@
 \[
-\TypeLamElimL
+\TypeLamElimL \quad \TypeLamElimW
 \]
@@
 be unrestricted, i.e. no linear variables can be used to type them.
-\[
-\TypeLamElimW
-\]
+%
 Typing linear and unrestricted function applications separately is less general
@@
 thought they are linear functions and linear fields, because $p$ can be
-instantiated at both $\omega$ and $1$ (so types using $p$ must be well-typed at both instantiations).
+instantiated at both $\omega$ and $1$.
 \[
-\TypeMultLamIntro
+\TypeMultLamIntro \quad \TypeMultLamElim
 \]
@@
 $\s[\pi/p]$.
-\[
-\TypeMultLamElim
-\]
 The rule additionally requires that $\pi$ be \emph{well-formed} in order
@@
 
-\subsubsection{\texorpdfstring{$\D$}{Delta}-bound variables}
+\paragraph{\texorpdfstring{$\D$}{Delta}-bound variables}
 
@@
 %
-\begin{enumerate}
-\item Using $u$ is equivalent to using all the linear resources in $\Delta$
+%\begin{enumerate}
+%\item
+(1) using $u$ is equivalent to using all the linear resources in $\Delta$;
 %\item $u$ can only be used once, since the $\D$ resources will not be available afterwards
-\item Using $u$ is mutually exclusive with using the $\Delta$ resources it depends on elsehow
-\item $u$ can be safely discarded as long as the resources in $\Delta$ are consumed elsehow
-\end{enumerate}
+%\item
+(2) using $u$ is mutually exclusive with using the $\Delta$ resources it depends on elsehow;
+% \item
+(3) $u$ can be safely discarded as long as the resources in $\Delta$ are consumed elsehow.
+%\end{enumerate}
 %
 Fortunately, since linear resources must be linearly split across
-sub-derivations, (2) follows from (1): consuming the linear resources
+sub-derivations, (2) follows from (1) since consuming the linear resources
 in $\Delta$ to type $u$ makes them unavailable in the context of any other
@@
 the expression is ill-typed).
-
+%
 These observations all boil down to one typing rule for $\Delta$-bound
-variables, which fundamentally encodes (1), implying the other two bullets:
+variables, which fundamentally encodes (1), implying the other two points:
 \[
@@
 expression to progress evaluation.}. An expression in Weak Head Normal Form can
-either be:
-\begin{itemize}
-\item A lambda expression $\lambda x.~e$,
-\item or a datatype constructor application $K~\ov{e}$
-\end{itemize}
+either be %\begin{itemize}
+% \item
+a lambda expression $\lambda x.~e$ 
+or a datatype constructor application $K~\ov{e}$.
+%\end{itemize}
 In both cases, the sub-expressions $e$ or $\ov{e}$ occurring in the lambda body
@@
 well-typed (these resources will be consumed when the expression is evaluated).
-%
+
 As will be made clear in later sections, we need to devise a specialized typing
@@
 in WHNF and terms that are not in WHNF.
-%
 Following the discussion on expressions in weak head normal form, we present a
@@
 \[
-\TypeAltNNotWHNF
+  \begin{array}{cc}
+    \TypeAltNNotWHNF & \TypeVarSplit
+    \end{array}
 \]
@@
 %
-This is given by the rule:
-\[
-\TypeVarSplit
-\]
-By assigning to each linear pattern variable a fragment of the scrutinee
+This is given by the Split rule. By assigning to each linear pattern variable a fragment of the scrutinee
 resources with a tag, we guarantee that all linear pattern variables are
@@
 times they are \emph{used}) in Cartesian pairs are given as input to
-Algorithm~\ref{computeRecUsages}, which computes the actual usage environment
+% Algorithm~\ref{computeRecUsages},
+an algorithm which computes the actual usage environment
 of each binder.
@@
 % I'm pretty sure it might fail in some inputs if I keep using the naiveUsageEnvs.
-\begin{algorithm}
-$usageEnvs \gets naiveUsageEnvs.map(fst)$\;
-\For{$(bind, U) \in naiveUsageEnvs$}{
-    \For{$V \in usageEnvs$}{
-        $V \gets sup(V[bind]*U\setminus\{bind\}, V\setminus\{bind\})$
-    }
-}
-\caption{computeRecUsages\label{computeRecUsages}}
-\end{algorithm}
+% \begin{algorithm}
+% $usageEnvs \gets naiveUsageEnvs.map(fst)$\;
+% \For{$(bind, U) \in naiveUsageEnvs$}{
+%     \For{$V \in usageEnvs$}{
+%         $V \gets sup(V[bind]*U\setminus\{bind\}, V\setminus\{bind\})$
+%     }
+% }
+% \caption{computeRecUsages\label{computeRecUsages}}
+% \end{algorithm}
 %
@@
 application of unrestricted function to a linear resource in the first argument
-to |build|:
-\begin{code}
-take :: Int -> Replicator a -o [a]
+to \incode{build}:
+\begin{lstlisting}
+take :: Int -> Replicator a *' $\lolli$ '* [a]
 take = \ (ds :: Int) (r :: Replicator b) ->
@@
        ...
-\end{code}
-We know the linear resource is still used linearly because |build| will
+\end{lstlisting}
+We know the linear resource is still used linearly because \incode{build} will
 always instance the unrestricted function to $(:)$ (read \emph{Cons}), which is
@@
 default alternative, programs where common-sub-expression elimination
-substituted more than one occurrence of |Ur x| by |y|, in the body of
-scrutinizing |y| (a linear variable), and other programs affected by rewrite rules.
+substituted more than one occurrence of \incode{Ur x} by \incode{y}, in the body of
+scrutinizing \incode{y} (a linear variable), and other programs affected by rewrite rules.
 %
@@
 semantically, and are simultaneously rejected by Linear Core and its
-implementation. An example, where the last component of |HashMap|, |wwz|,
+implementation. An example, where the last component of \incode{HashMap}, \incode{wwz},
 is linear, but is not being consumed in the case alternative matching on it:
-\begin{code}
-jssvi :: Ur Bool -o Set Int -o Ur (TestT IO ())
+\begin{lstlisting}
+jssvi :: Ur Bool *' $\lolli$ '* Set Int *' $\lolli$ '* Ur (TestT IO ())
 jssvi (a :: Ur Bool) (b :: Set Int)
   = case a of
-    Ur ss -> case b  of
-      HashMap wwx wwy wwz ->
-        jump wjssAd ss
-\end{code}
+     Ur ss -> case b of { HashMap wwx wwy wwz -> jump wjssAd ss }
+\end{lstlisting}
 %
@@
 
-The \emph{Irrelevance} lemma is required to prove preservation for that
-evaluation case. We need to prove that the alternatives of a case expression
-typed with proof irrelevant resources are still well-typed when the proof
-irrelevant resource is substituted by the scrutinee resources as it is evaluated to WHNF.
-In this sense, the \emph{Irrelevance} lemma witnesses the soundness of typing a
+% The \emph{Irrelevance} lemma is required to prove preservation for that
+% evaluation case. We need to prove that the alternatives of a case expression
+% typed with proof irrelevant resources are still well-typed when the proof
+% irrelevant resource is substituted by the scrutinee resources as it is evaluated to WHNF.
+% In this sense, the \emph{Irrelevance} lemma witnesses the soundness of typing a
+% case alternative with proof irrelevant resources in a certain context with respect to
+% typing the same expression with arbitrary resources (we note, however, typing
+% an alternative with proof irrelevant resources is not complete wrt using
+% arbitrary resources -- a counter example needs only to use a resource
+% directly).
+
+\WHNFConvSoundness
+
+The \emph{Irrelevance} lemma witnesses the soundness of typing a
 case alternative with proof irrelevant resources in a certain context with respect to
@@
 
-\WHNFConvSoundness
-
-\noindent Intuitively, the lemma holds since proof irrelevant resources can
-only be used through the case binder or pattern-bound variables. If we
-consistently replace the proof irrelevant resources both in the typing
-environment and in the usage environments containing them, the expression
-remains well-typed. We note that the proof irrelevant resources are always
-unique when introduced in such a case alternative (we always take the scrutinee
-environment to make irrelevant, and allow nested ``irrelevantness''), so the
-case binder has the only occurrence of those resources in the $\G$ environment.
-% (being somewhat akin to congruence).
-%
-The proof is given in Section~\ref{sec:proof:irrelevance}.
+% \noindent Intuitively, the lemma holds since proof irrelevant resources can
+% only be used through the case binder or pattern-bound variables. If we
+% consistently replace the proof irrelevant resources both in the typing
+% environment and in the usage environments containing them, the expression
+% remains well-typed. We note that the proof irrelevant resources are always
+% unique when introduced in such a case alternative (we always take the scrutinee
+% environment to make irrelevant, and allow nested ``irrelevantness''), so the
+% case binder has the only occurrence of those resources in the $\G$ environment.
+% % (being somewhat akin to congruence).
+% %
+% The proof is given in Section~\ref{sec:proof:irrelevance}.
 
@@
 %
-\noindent Type preservation states that a well-typed expression $e$ that
-evaluates to $e'$ remains well-typed under the same context:
-%
-The proof ($\S$~\ref{sec:proof:type-preservation}) is done by structural induction on the reductions $e \longrightarrow
+\ProgressTheorem
+% \noindent Type preservation states that a well-typed expression $e$ that
+% evaluates to $e'$ remains well-typed under the same context:
+%
+Type preservation
+% ($\S$~\ref{sec:proof:type-preservation})
+follows by structural induction on the reductions $e \longrightarrow
 e'$ from the operational semantics. Most cases are straightforward and usually
@@
 %
-This case guarantees that the separation of rules for treating scrutinees is
-consistent, in the sense that a well-typed case expression with a scrutinee not
-in WHNF remains well-typed after the scrutinee is evaluated to WHNF.
-
-\ProgressTheorem
-%
-\noindent Progress states that the evaluation of a well-typed term does not block:
-Similarly, progress is proved by induction on typing ($\S$~\ref{sec:proof:progress}).
-
-\subsubsection{Substitution Lemmas}
-
-The preservation and progress theorems depend on multiple substitution lemmas,
-one for each kind of variable, as is standard.
-% The proofs are given in Section~\ref{sec:proof:substitution-lemmas}.
-% The ... themselves diverge from their common formulation, because
-% $\D$-variables refer to linearly bound variables, so substitution must take into account
-
-The linear substitution lemma states that a well-typed expression $e$ with a
-linear variable $x$ of type $\s$ remains well-typed if
-occurrences of $x$ in the $e$ are replaced by an expression $e'$ of the same
-type $\s$, and occurrences of $x$ in the linear context and in usage
-environments of $\D$-bound variables are replaced by the linear context $\D'$
-used to type $e'$:
-
-\LinearSubstitutionLemma
-
-\noindent Where $\G[\D'/x]$ substitutes all occurrences of $x$ in the usage
-environments of $\D$-variables in $\G$ by the linear variables in $\D'$.
-% ($x$ couldn't appear anywhere besides usage environments of $\D$-bound variables, since $x$ is linear).
-
-The substitution of the resource in the usage environments is illustrated
-by the following example. Consider the term $\llet{y = use~x}{y}$ where $use$ and $x$ are free variables:
-if we replace occurrences of $x$ by $e'$ (where $\G;\D \vdash e : \s$), then the
-``real'' usage environment of $y$ goes from $\{x\}$ to $\D$. If we don't update
-the usage environment of $y$ accordingly, we'll ultimately be typing
-$y{:}_{\{x\}}\vp$ with $\D$ instead of $x$, which is not valid.
-
-The linear substitution lemma extends to case alternatives as well.
-The lemma for substitution of linear variables in case alternatives is similar
-to the linear substitution lemma, applied to the case alternative judgement.
-% Is slightly more involved, in the sense that there are more environments in which the
-% substitution $[\D/x]$ must be applied (for the same reason):
-%
-\LinearSubstitutionAltsLemma
-%
-\noindent We further require that the environment annotated in the case
-alternative judgement, $\D_s$, is a subset of the environment used to type the
-whole alternative $\D_s \subseteq \D$. In all occurrences of the alternative
-judgement (in $Case_{\textrm{WHNF}}$ and $Case_{\textrm{Not WHNF}}$), the
-environment annotating the alternative judgement is \emph{always} a subset of
-the alternative environment.
-
-The substitution lemma for unrestricted variables follows the usual
-formulation, with the added restriction (common to linear type systems) that
-the expression $e'$ that is going to substitute the unrestricted variable $x$
-is typed on an empty linear environment:
-%
-\UnrestrictedSubstitutionLemma
-%
-\noindent Similarly, we also prove the substitution of unrestricted variables preserves types on an alternative case expression:
-%
-\UnrestrictedSubstitutionAltsLemma
-
-Finally, we introduce the lemma stating that substitution of $\D$-bound
-variables by expressions of the same type preserves the type of the original
-expression.
-%
-What distinguishes this lemma from traditional substitution lemmas is that the
-usage environment $\D$ of the variable $x$ being substituted by expression $e'$
-must match exactly the typing environment $\D$ of $e'$ and the
-environment of the original expression doesn't change with the substitution:
-%
-\DeltaSubstitutionLemma
-%
-\noindent Intuitively, if $x$ is well-typed with $\D$ in $e$, substituting $x$
-by an expression $e'$ which is typed in the same environment $\D$ allows the
-distribution of resources $\D,\D'$ used to type $e$ across sub-derivations to remain
-unchanged. To prove the theorems, we don't need a ``stronger'' substitution of
-$\D$-vars lemma (allowing arbitrary resources $\D''$ to type $e'$, as in other
-substitution lemmas), as we only ever substitute $\D$-variables by expressions
-whose typing environment matches the variables usage environment. However, it
-is not obvious whether such a lemma is possible to prove for $\D$-variables
-(e.g. let $\G;\D \vdash e :\s$ and $\G; \D' \vdash \llet{x = e'}{x}$, if we
-substitute $e$ for $x$ the resources $\D'$ are no longer consumed).
-
-The $\D$-substitution lemma on case alternatives reflects again that the typing
-environment of the expression substitution the variable must match its usage
-environment. We recall that $\D_s \subseteq \D,\D'$ states that the annotated
-environment is always contained in the typing environment, which is true of all
-occurrences of this judgement. An alternative formulation of this lemma could
-instead explicitly list $\D_s$ as part of the typing environment for the same
-effect:
-
-\DeltaSubstitutionAltsLemma
-
-The proofs for substitution lemmas of linear, unrestricted, and
-$\D$-variables are available in Section~\ref{sec:proof:substitution-lemmas}.
+% This case guarantees that the separation of rules for treating scrutinees is
+% consistent, in the sense that a well-typed case expression with a scrutinee not
+% in WHNF remains well-typed after the scrutinee is evaluated to WHNF.
+%
+%
+%
+%\noindent
+% Progress states that the evaluation of a well-typed term does not block:
+% Similarly, progress is proved by induction on typing ($\S$~\ref{sec:proof:progress}).
+
+% \subsubsection{Substitution Lemmas}
+
+% The preservation and progress theorems depend on multiple substitution lemmas,
+% one for each kind of variable, as is standard.
+% % The proofs are given in Section~\ref{sec:proof:substitution-lemmas}.
+% % The ... themselves diverge from their common formulation, because
+% % $\D$-variables refer to linearly bound variables, so substitution must take into account
+
+% The linear substitution lemma states that a well-typed expression $e$ with a
+% linear variable $x$ of type $\s$ remains well-typed if
+% occurrences of $x$ in the $e$ are replaced by an expression $e'$ of the same
+% type $\s$, and occurrences of $x$ in the linear context and in usage
+% environments of $\D$-bound variables are replaced by the linear context $\D'$
+% used to type $e'$:
+
+% \LinearSubstitutionLemma
+
+% \noindent Where $\G[\D'/x]$ substitutes all occurrences of $x$ in the usage
+% environments of $\D$-variables in $\G$ by the linear variables in $\D'$.
+% % ($x$ couldn't appear anywhere besides usage environments of $\D$-bound variables, since $x$ is linear).
+
+% The substitution of the resource in the usage environments is illustrated
+% by the following example. Consider the term $\llet{y = use~x}{y}$ where $use$ and $x$ are free variables:
+% if we replace occurrences of $x$ by $e'$ (where $\G;\D \vdash e : \s$), then the
+% ``real'' usage environment of $y$ goes from $\{x\}$ to $\D$. If we don't update
+% the usage environment of $y$ accordingly, we'll ultimately be typing
+% $y{:}_{\{x\}}\vp$ with $\D$ instead of $x$, which is not valid.
+
+% The linear substitution lemma extends to case alternatives as well.
+% The lemma for substitution of linear variables in case alternatives is similar
+% to the linear substitution lemma, applied to the case alternative judgement.
+% % Is slightly more involved, in the sense that there are more environments in which the
+% % substitution $[\D/x]$ must be applied (for the same reason):
+% %
+% \LinearSubstitutionAltsLemma
+% %
+% \noindent We further require that the environment annotated in the case
+% alternative judgement, $\D_s$, is a subset of the environment used to type the
+% whole alternative $\D_s \subseteq \D$. In all occurrences of the alternative
+% judgement (in $Case_{\textrm{WHNF}}$ and $Case_{\textrm{Not WHNF}}$), the
+% environment annotating the alternative judgement is \emph{always} a subset of
+% the alternative environment.
+
+% The substitution lemma for unrestricted variables follows the usual
+% formulation, with the added restriction (common to linear type systems) that
+% the expression $e'$ that is going to substitute the unrestricted variable $x$
+% is typed on an empty linear environment:
+% %
+% \UnrestrictedSubstitutionLemma
+% %
+% \noindent Similarly, we also prove the substitution of unrestricted variables preserves types on an alternative case expression:
+% %
+% \UnrestrictedSubstitutionAltsLemma
+
+% Finally, we introduce the lemma stating that substitution of $\D$-bound
+% variables by expressions of the same type preserves the type of the original
+% expression.
+% %
+% What distinguishes this lemma from traditional substitution lemmas is that the
+% usage environment $\D$ of the variable $x$ being substituted by expression $e'$
+% must match exactly the typing environment $\D$ of $e'$ and the
+% environment of the original expression doesn't change with the substitution:
+% %
+% \DeltaSubstitutionLemma
+% %
+% \noindent Intuitively, if $x$ is well-typed with $\D$ in $e$, substituting $x$
+% by an expression $e'$ which is typed in the same environment $\D$ allows the
+% distribution of resources $\D,\D'$ used to type $e$ across sub-derivations to remain
+% unchanged. To prove the theorems, we don't need a ``stronger'' substitution of
+% $\D$-vars lemma (allowing arbitrary resources $\D''$ to type $e'$, as in other
+% substitution lemmas), as we only ever substitute $\D$-variables by expressions
+% whose typing environment matches the variables usage environment. However, it
+% is not obvious whether such a lemma is possible to prove for $\D$-variables
+% (e.g. let $\G;\D \vdash e :\s$ and $\G; \D' \vdash \llet{x = e'}{x}$, if we
+% substitute $e$ for $x$ the resources $\D'$ are no longer consumed).
+
+% The $\D$-substitution lemma on case alternatives reflects again that the typing
+% environment of the expression substitution the variable must match its usage
+% environment. We recall that $\D_s \subseteq \D,\D'$ states that the annotated
+% environment is always contained in the typing environment, which is true of all
+% occurrences of this judgement. An alternative formulation of this lemma could
+% instead explicitly list $\D_s$ as part of the typing environment for the same
+% effect:
+
+% \DeltaSubstitutionAltsLemma
+
+% The proofs for substitution lemmas of linear, unrestricted, and
+% $\D$-variables are available in Section~\ref{sec:proof:substitution-lemmas}.
 
@@
 
-The optimising transformations proved sound wrt Linear Core in this section
-have been previously explained and motivated in
-Section~\ref{sec:core-to-core-transformations}.
-%
+% The optimising transformations proved sound wrt Linear Core in this section
+% have been previously explained and motivated in
+% Section~\ref{sec:core-to-core-transformations}.
+% %
 Transformations are described by an arbitrary well-typed expression with a certain shape, on
@@
 
-\subsubsection{Inlining}
-
-% To the best of our knowledge, there is no linear type system for which inlining
-% preserves linearity\footnote{https://github.com/ghc-proposals/ghc-proposals/blob/master/proposals/0111-linear-types.rst\#id90}
-
-\input{language-v4/proofs/optimizations/Inlining}
-
-\subsubsection{\texorpdfstring{$\beta$}{Beta}-reduction}
-
-\input{language-v4/proofs/optimizations/BetaReduction}
-
-\subsubsection{Case of known constructor}
-
-\input{language-v4/proofs/optimizations/CaseOfKnownConstructor}
-
-\subsubsection{Let floating}
-
-\input{language-v4/proofs/optimizations/LetFloating}
-
-\subsubsection{\texorpdfstring{$\eta$}{Eta}-conversions}
-
-\input{language-v4/proofs/optimizations/EtaConvs}
-
-\subsubsection{Binder Swap}
-
-The binder swap transformation applies to case expressions whose scrutinee is a
-single variable $x$, and it substitutes occurrences of $x$ in the case
-alternatives for the case binder $z$. If $x$ is a linear resource, $x$ cannot
-occur in the case alternatives (as we conservatively consider variables are not
-in WHNF), so the substitution preserves types vacuously. Otherwise, $x$ can be
-freely substituted by $z$, since $z$ is also an unrestricted resource (it's
-usage environment is empty because $x$ is unrestricted).
-
-\input{language-v4/proofs/optimizations/BinderSwap}
+\todo[inline]{Aqui uma lista itemizada das transformações todas. As
+  provas todas para apendice.}
+
+% \subsubsection{Inlining}
+
+% % To the best of our knowledge, there is no linear type system for which inlining
+% % preserves linearity\footnote{https://github.com/ghc-proposals/ghc-proposals/blob/master/proposals/0111-linear-types.rst\#id90}
+
+% \input{language-v4/proofs/optimizations/Inlining}
+
+% \subsubsection{\texorpdfstring{$\beta$}{Beta}-reduction}
+
+% \input{language-v4/proofs/optimizations/BetaReduction}
+
+% \subsubsection{Case of known constructor}
+
+% \input{language-v4/proofs/optimizations/CaseOfKnownConstructor}
+
+% \subsubsection{Let floating}
+
+% \input{language-v4/proofs/optimizations/LetFloating}
+
+% \subsubsection{\texorpdfstring{$\eta$}{Eta}-conversions}
+
+% \input{language-v4/proofs/optimizations/EtaConvs}
+
+% \subsubsection{Binder Swap}
+
+% The binder swap transformation applies to case expressions whose scrutinee is a
+% single variable $x$, and it substitutes occurrences of $x$ in the case
+% alternatives for the case binder $z$. If $x$ is a linear resource, $x$ cannot
+% occur in the case alternatives (as we conservatively consider variables are not
+% in WHNF), so the substitution preserves types vacuously. Otherwise, $x$ can be
+% freely substituted by $z$, since $z$ is also an unrestricted resource (it's
+% usage environment is empty because $x$ is unrestricted).
+
+% \input{language-v4/proofs/optimizations/BinderSwap}
 
@@
 \begin{array}{l}
-(\lambda x.~\ccase{x}{\_ \to x})~(use~y)\\
-\Longrightarrow_\textrm{call-by-need $\beta$-reduction}\\
+(\lambda x.~\ccase{x}{\_ \to x})~(use~y)
+\Longrightarrow_\textrm{CBNeed}
 \llet{x = use~y}{\ccase{y}{\_ \to y}}
@@
 
-\subsubsection{Case of Case}
-
-The case of case transformation applies to case expressions whose scrutinee is
-another case expression, and returns the innermost case expression transformed by repeating
-the outermost case expression in each alternative of the innermost case,
-scrutinizing the original alternative body.
-
-Intuitively, since the scrutinee of the outermost case is not in WHNF, no
-resources from it can directly occur in the outermost alternatives. By moving
-the outermost alternatives inwards with a different scrutinee, the alternatives
-remain well-typed because they are typed using either the case binder or the
-pattern bound variables, which, by the \emph{Irrelevance} lemma, makes it
-well-typed for any scrutinee consuming arbitrary resources. The proof is given in Section~\ref{sec:proof:caseofcase}.
-
-\CaseOfCaseTheorem
+% \subsubsection{Case of Case}
+
+% The case of case transformation applies to case expressions whose scrutinee is
+% another case expression, and returns the innermost case expression transformed by repeating
+% the outermost case expression in each alternative of the innermost case,
+% scrutinizing the original alternative body.
+
+% Intuitively, since the scrutinee of the outermost case is not in WHNF, no
+% resources from it can directly occur in the outermost alternatives. By moving
+% the outermost alternatives inwards with a different scrutinee, the alternatives
+% remain well-typed because they are typed using either the case binder or the
+% pattern bound variables, which, by the \emph{Irrelevance} lemma, makes it
+% well-typed for any scrutinee consuming arbitrary resources. The proof is given in Section~\ref{sec:proof:caseofcase}.
+
+% \CaseOfCaseTheorem
 
@@
 
-In this chapter, we compare our contributions and Linear Core to related
-existing works in the literature,
-% (including linearity in a lazy, evaluation models in terms of linearity, and
-% the Core system).
-%
-consider further research
-% deemed out of the scope of this work and of the Linear Core type system
-(notably, so-called \emph{multiplicity coercions} to handle the
-interaction between linearity and \emph{coercions}, a key feature of Core which we
-left out our system), and conclude.
+% In this chapter, we compare our contributions and Linear Core to related
+% existing works in the literature,
+% % (including linearity in a lazy, evaluation models in terms of linearity, and
+% % the Core system).
+% %
+% consider further research
+% % deemed out of the scope of this work and of the Linear Core type system
+% (notably, so-called \emph{multiplicity coercions} to handle the
+% interaction between linearity and \emph{coercions}, a key feature of Core which we
+% left out our system), and conclude.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Related Work}
+\subsection{Related Work}
 
@@
 
-\subsubsection{Linear Haskell\label{sec:related-work-linear-haskell}}
+\paragraph{Linear Haskell\label{sec:related-work-linear-haskell}}
 
@@
 
-\subsubsection{Linear Mini-Core\label{sec:linear-mini-core}}
+\paragraph{Linear Mini-Core\label{sec:linear-mini-core}}
 
@@
 
-\subsubsection{Linearity-influenced optimisations}
+\paragraph{Linearity-influenced optimisations}
 
@@
 transformation.
-
+%
 By fully supporting linear types in Core, these optimising transformations
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Future Work\label{sec:future-work}}
-
-In this section we highlight some avenues of further research. Briefly,
+\subsection{Future Work\label{sec:future-work}}
+
+We highlight some avenues of further research. Briefly,
 these include \emph{multiplicity coercions}, optimisations leveraging
@@
 
-\parawith{Multiplicity Coercions.}
+\paragraph{Multiplicity Coercions.}
 Linear Core doesn't have type equality coercions, a flagship feature of GHC
@@
 %
-Coercions, briefly explained in Section~\ref{sec:core},
+Coercions %briefly explained in Section~\ref{sec:core},
 allow the Core intermediate language to encode a panoply of Haskell source
@@
 arrows which specify the linearity of the function. In practice,
-multiplicities are simply types of kind |Multiplicity|, where |One| and |Many|
-are the type constructors of the kind |Multiplicity|; multiplicity polymorphism
+multiplicities are simply types of kind \incode{Multiplicity}, where \incode{One} and \incode{Many}
+are the type constructors of the kind \incode{Multiplicity}; multiplicity polymorphism
 follows from type polymorphism, where multiplicity variables are
@@
 as well.
-%
-Consequently, we might define, e.g., using a GADT |SBool| and a type family
-|If|, the function |dep| which is linear in the second argument if the first
-argument is |STrue| and unrestricted otherwise:
-%
-\begin{limitation}
-\begin{code}
-data SBool :: Bool -> Type where
-  STrue :: SBool True
-  SFalse :: SBool False
-
-type family If b t f where
-  If True t _ = t
-  If False _ f = f
-
-dep :: SBool b -> Int %(If b One Many) -> Int
-dep STrue x = x
-dep SFalse _ = 0
-\end{code}
-\end{limitation}
-%
-This example is linear and should be accepted. However, the example is rejected
-by the GHC's Core type checker. Critically, Core doesn't currently understand
-so-called \emph{multiplicity coercions}. Even though after matching on |STrue|
-we have access to a coercion from the function multiplicity $m$ to $1$ ($m \sim
-1$), we cannot use this coercion to determine whether the usages of the linear
-resources match the multiplicity.
-%
-Studying the interaction between coercions and multiplicities is a main avenue
+For instance, it is possible via the use of GADTs to define a
+function whose linearity of its second argument depends on the value
+of its first argument, internally realized through so-called
+\emph{multiplicity coercions}. Currently, Core cannot make use of such
+coercions to determine whether the usages of linear resources match
+their intended multiplicity. Studying the interaction between coercions and multiplicities is a main avenue
 of future work for Linear Core.
+%
+% Consequently, we might define, e.g., using a GADT \incode{SBool} and a type family
+% |If|, the function \incode{dep} which is linear in the second argument if the first
+% argument is \incode{STrue} and unrestricted otherwise:
+% %
+% \begin{limitation}
+% \begin{lstlisting}
+% data SBool :: Bool -> Type where
+%   STrue :: SBool True
+%   SFalse :: SBool False
+
+% type family If b t f where
+%   If True t _ = t
+%   If False _ f = f
+
+% dep :: SBool b -> Int %(If b One Many) -> Int
+% dep STrue x = x
+% dep SFalse _ = 0
+% \end{lstlisting}
+% \end{limitation}
+% %
+% This example is linear and should be accepted. However, the example is rejected
+% by the GHC's Core type checker.
+% Critically, Core doesn't currently understand
+% so-called \emph{multiplicity coercions}. Even though after matching on |STrue|
+% we have access to a coercion from the function multiplicity $m$ to $1$ ($m \sim
+% 1$), we cannot use this coercion to determine whether the usages of the linear
+% resources match the multiplicity.
+% %
+% Studying the interaction between coercions and multiplicities is a main avenue
+% of future work for Linear Core.
 
@@
 
-\parawith{Optimisations leveraging linearity.}
+\paragraph{Optimisations leveraging linearity.}
 We only briefly mentioned how linearity can inform optimisations to produce
@@
 
-\parawith{Usage environment resource inference.}
+\paragraph{Usage environment resource inference.}
 In Section~\ref{sec:linearity-semantically}, we explained that the linear
@@
 
-\parawith{Linear Core in the Glasgow Haskell Compiler.}
+\paragraph{Linear Core in the Glasgow Haskell Compiler.}
 Linear Core is suitable as the intermediate language of an optimising compiler
@@
 
-\parawith{Generalizing Linear Core to Haskell.}
+\paragraph{Generalizing Linear Core to Haskell.}
 Linear types, despite their compile-time correctness guarantees regarding
@@
 
-\section{Conclusion}
-
-Optimising compilers with a typed and lazy intermediate language with linear
-types (of which GHC is the prime example) leverage laziness to heavily
-transform and rewrite programs into simpler forms.
-%
-However, these optimising transformations push the interaction between
-linearity and laziness to the limits where linearity can no longer be seen
-syntactically, despite being maintained semantically, in the sense that linear
-resources are still used exactly once when the optimised program is run.
-
-In this work we explored linearity in the presence of laziness by example
-through the interactions of linear types with lazy (recursive) let bindings and
-case expressions that evaluate their scrutinee to Weak Head Normal Form. Most
-example programs were linear semantically, but not syntactically.
-%
-We developed a linear type system, Linear Core, for an intermediate language
-akin to GHC Core, with laziness and linearity. In contrast to GHC Core's type
-system, or any other linear type system (to the best of our knowledge), our
-type system understands semantic linearity, and can thus correctly type a wider
-range of linear programs, as those explored in the semantic linearity examples.
-%
-Crucially, we proved soundness of the type system, and proved multiple
-optimising transformations preserve linearity, despite most violating linearity
-in other linear type systems. Additionally, we implemented Linear Core as a GHC
-plugin to further explore its suitability in the intermediate language of an
-optimising compiler.
-
-Concluding, Linear Core is a suitable type system for linear, lazy,
-intermediate languages of optimising compilers such as GHC, as it understands
-linearity in the presence of laziness s.t. optimisations preserve types and
-linearity, and further unblocks optimisations influenced by linearity, e.g.
-inlining and call-by-name $\beta$-reduction for applications of (semantically)
-linear functions.
+% \section{Conclusion}
+
+% Optimising compilers with a typed and lazy intermediate language with linear
+% types (of which GHC is the prime example) leverage laziness to heavily
+% transform and rewrite programs into simpler forms.
+% %
+% However, these optimising transformations push the interaction between
+% linearity and laziness to the limits where linearity can no longer be seen
+% syntactically, despite being maintained semantically, in the sense that linear
+% resources are still used exactly once when the optimised program is run.
+
+% In this work we explored linearity in the presence of laziness by example
+% through the interactions of linear types with lazy (recursive) let bindings and
+% case expressions that evaluate their scrutinee to Weak Head Normal Form. Most
+% example programs were linear semantically, but not syntactically.
+% %
+% We developed a linear type system, Linear Core, for an intermediate language
+% akin to GHC Core, with laziness and linearity. In contrast to GHC Core's type
+% system, or any other linear type system (to the best of our knowledge), our
+% type system understands semantic linearity, and can thus correctly type a wider
+% range of linear programs, as those explored in the semantic linearity examples.
+% %
+% Crucially, we proved soundness of the type system, and proved multiple
+% optimising transformations preserve linearity, despite most violating linearity
+% in other linear type systems. Additionally, we implemented Linear Core as a GHC
+% plugin to further explore its suitability in the intermediate language of an
+% optimising compiler.
+
+% Concluding, Linear Core is a suitable type system for linear, lazy,
+% intermediate languages of optimising compilers such as GHC, as it understands
+% linearity in the presence of laziness s.t. optimisations preserve types and
+% linearity, and further unblocks optimisations influenced by linearity, e.g.
+% inlining and call-by-name $\beta$-reduction for applications of (semantically)
+% linear functions.
 

### 2024-01-24 icfp.tex
revised intro

@@
 % handle mutable state safely~\cite{}
-
-Consider the following program in which allocated memory is freed twice.
+%
+For instance, consider the following program in which allocated memory is freed twice.
 % We know this to be the dreaded double-free error which will crash the program at runtime.
@@
 %
-Linearity in Haskell stands out from linearity in other languages:
-%\begin{itemize}
-    % \item Linear types were only added to the language roughly \emph{31 years
-    %     after} Haskell's inception, unlike Rust and Idris 2 which were designed
-    %     with linearity from the start. It is an especially difficult endeavour
-    %     to add linear types to a well-established language with a large library
-    %     ecosystem and many active users, rather than to develop the language
-    %     from the ground up with linear types in mind. The successful
-    %     approach as implemented in the Glasgow Haskell Compiler (GHC), the leading
-    %     Haskell compiler, is based on Linear Haskell~\cite{cite:linearhaskell}, which describes a linear type system designed
-    %     with retaining backwards-compatibility and allowing code reuse across
-    %     linear and non-linear users of the same libraries.
-    %     We describe Linear Haskell in detail in Section~\ref{sec:linear-haskell}.
-%
-%\item
-  Linear types permeate Haskell down to (its) Core, the
-        intermediate language into which Haskell is
-        translated. Core is a minimal, explicitly typed,
-        functional language, to which multiple
-        Core-to-Core optimising transformations are
-        applied during compilation. Due to Core's minimal design, typechecking
-        Core programs is very efficient and doing so serves as a sanity check to the
+
+However, linearity in Haskell stands out from linearity in other
+languages given that linear types permeate Haskell down to (its) Core,
+the intermediate language into which Haskell is translated. Core is a
+minimal, explicitly typed, functional language, to which multiple
+Core-to-Core optimising transformations are applied during
+compilation. Due to Core's minimal design, typechecking Core programs
+is very efficient and doing so serves as a sanity check to the
 correction of the source transformations. If the resulting optimised
-        Core program fails to typecheck, the optimising
-        transformations are very likely to have introduced an error
-        in the resulting program.
-        %We present Core and its formal
-        %basis, System~$F_C$~\cite{cite:systemfc}, in Section~\ref{sec:core}.
-        % TODO: \item values in rust are linear by default while non-linear is
-        % the haskell default?
-%
-        % \item
+Core program fails to typecheck, the optimising transformations are
+very likely to have introduced an error in the resulting program.
+%
 
@@
         evaluated, i.e., expressions in Haskell are only evaluated when needed,
-        unlike C or Rust, which are \emph{eagerly} evaluated.
-        %
-        Laziness allows an optimising compiler to aggressively transform the source
-        program without changing its semantics and, indeed, the Glasgow Haskell
-        Compiler (GHC) heavily transforms Core by leveraging its laziness.
-        %
-        However, lazy evaluation interacts non-trivially with linearity.
-        Intuitively, since expressions are not necessarily evaluated, an
-        occurrence of a linear resource in an expression does not necessarily
+        (unlike Rust, which is \emph{eagerly} evaluated).
+        %
+        % Laziness allows an optimising compiler to aggressively transform the source
+        % program without changing its semantics and, indeed, the Glasgow Haskell
+        % Compiler (GHC) heavily transforms Core by leveraging its laziness.
+        %
+        As we make clear in this paper, lazy evaluation interacts non-trivially with linearity.
+        Intuitively, since expressions are not necessarily evaluated,
+        a syntactic occurrence of a linear resource in an expression does not necessarily
         entail consuming that resource (i.e., if the expression is not evaluated,
@@
         %
-        This interaction is unique to Haskell since, as far as we know, it is the
+        This interaction is unique to Haskell since, to the best of
+        our knowledge, it is the
         only language featuring both laziness and linearity.
-        %
-        %We review lazy and eager evaluation strategies in Section~\ref{sec:background:evaluation-strategies}.
-
-%\end{itemize}
-%
-
-% Linear core good Aligned with the philosophy of having a \emph{typed}
-% intermediate language, the integration of linear types in Haskell required
-% extending Core with linear types.
-%
-Much like a typed Core ensures that the translation from Haskell (dubbed
-\emph{desugaring}) and the subsequent optimising transformations applied to
-Core are correctly implemented, a \emph{linearly typed} Core guarantees that
-linear resource usage in the source language is not violated by the translation
-process and the compiler optimisation passes.
-%
-It is crucial that a program's behaviour enforced by linear types is \emph{not}
-changed by the compiler in the desugaring or optimisation stages (optimisations
-should not destroy linearity!) and a linearity aware Core type-checker is key in
-providing such guarantees -- it would be disastrous if the compiler, e.g.,
-duplicated a pointer to heap-allocated memory that was previously just used
-once and then freed in the original program.
+
+Just as Core's type system provides a degree of validation to the
+translation from Haskell (dubbed \emph{desugaring}) and the subsequent
+optimising transformations, a \emph{linearly typed} Core would guarantee that
+linear resource usage in the source language is not violated by desugaring
+and the compiler optimisation passes. Moreover, linearity information in
+Core can inform Core-to-Core optimising 
+transformations~\cite{cite:let-floating,peytonjones1997a,cite:linearhaskell},
+such as inlining and $\beta$-reduction, to produce more efficient programs.
+%
+% It is crucial that a program's behaviour enforced by linear types is \emph{not}
+% changed by the compiler in the desugaring or optimisation stages (optimisations
+% should not destroy linearity!) and a linearity aware Core type-checker is key in
+% providing such guarantees -- it would be disastrous if the compiler, e.g.,
+% duplicated a pointer to heap-allocated memory that was previously just used
+% once and then freed in the original program.
 %TODO: linearidade pode informar otimizacoes
 %
-Even more, linearity in Core can inform Core-to-Core optimising
-transformations~\cite{cite:let-floating,peytonjones1997a,cite:linearhaskell},
-including inlining and $\beta$-reduction, to produce more performant programs.
+% Even more, linearity in Core can inform Core-to-Core optimising
+% transformations~\cite{cite:let-floating,peytonjones1997a,cite:linearhaskell},
+% including inlining and $\beta$-reduction, to produce more performant programs.
 
@@
 
-In spite of a linearly typed Core ultimately being the desired intermediate
-language for the Glasgow Haskell Compiler,
-both in its expressiveness to completely represent a Haskell with
-linear types and in enabling more extensive compiler validations plus improved optimisations,
-linearity is effectively ignored in Core. The reason is not evidently clear:
-if we can typecheck linearity in the surface level Haskell, it should also be
-possible, and natural, to do so in Core.
-% why do we fail to do so in Core?
-
-The desugaring process from surface level Haskell to Core, and the subsequent
-Core-to-Core optimising transformations, eliminate, and rearrange, most of the
-syntactic constructs through which linearity checking is performed -- often
-resulting in programs completely different from the original, especially due to
-the flexibility laziness provides a compiler in the optimisations it may
-perform. % without changing the program semantics.
-%
-Crucially, since optimisations do not destroy linearity, the resulting program
-is still linear \emph{semantically}, however, the current linear type system
-fails to recognize it as linear.
-%
-% The optimisations transform programs that are linear into ones that don't
-% \emph{look} linear, but remain linear \emph{semantically}.
+Despite of the advantages of a linearly typed Core, the status quo is
+that linearity is effectively ignored in Core, despite being checked
+in the source Haskell code. The reason is that the desugaring
+and the subsequent Core-to-Core optimising transformations,
+eliminate and rearrange most of the syntactic constructs through
+which linearity checking is performed -- often resulting in programs
+that are syntactically very different from the original, especially
+due to the flexibility laziness provides a compiler in the
+optimisations it may perform.
+%
+Crucially, while optimisations aim to preserve the semantics of
+programs and so preserve a program's linear resource consumption,
+a \emph{naive} syntactic analysis of the optimised programs (as
+performed by the current Core type checker) often fails to recognize
+the program as linear.
 %
@@
 the latter results from inlining $y$ in the let body of the former. Despite the
-result no longer \emph{looking} linear (as there are now two syntactic
-occurrences of the linear resource $x$), the program \emph{is} indeed linear
-because the let-bound expression freeing $x$ is never evaluated, so $x$ is
-consumed exactly once when it is freed in the let body:
+the program no longer appearing linear (as there are now two 
+occurrences of $x$), the program \emph{is} indeed linear
+because the let-bound expression is never evaluated, so $x$ is
+consumed exactly once (when it is freed in the let body):
 % executed, $x$ will be freed exactly once:
@@
 lazy evaluation are rejected by Core's current linear type system.
-
+%
 In this work, we address this limitation (and its implications on validating
 and influencing optimising transformations) by designing a type system which
-understands semantic linearity in the presence of laziness and is suitable for
-the intermediate language of an optimising compiler.
-%
-In detail, our contributions are:
+understands \emph{semantic linearity} in the presence of laziness and is suitable for
+the intermediate language of an optimising compiler such as the
+Glasgow Haskell Compiler.
+%
+Our contributions are as follows:
 %
@@
 
-\item We explain and provide insights into \emph{semantic} linearity in
-contrast to \emph{syntactic} linearity, in Haskell, by example
+\item We introduce the concept of \emph{semantic} linearity, in
+contrast to \emph{syntactic} linearity in Haskell, by example
 ($\S$~\ref{sec:linearity-semantically}).
@@
 \item We introduce Linear Core, a type system for a linear lazy language with
-all key features from Core (except for type equality coercions), which,
-crucially, understands semantic linearity in the presence of laziness. To the
-best of our knowledge, this is the first type system to understand linearity
-semantically in the context of lazy evaluation
+the key features from Core (except for type equality coercions), which,
+accounts for semantic linearity in the presence of laziness. To the
+best of our knowledge, this is the first type system to capture linearity
+semantically in this context
 ($\S$~\ref{sec:main:linear-core}).
 
-\item We prove Linear Core to be sound (well-typed Linear Core programs do not
-get \emph{stuck}) and prove that multiple optimising transformations (which
-currently violate linearity in Core) preserve types and linearity in Linear
+\item We show the type soundness of Linear Core and prove that
+  multiple optimising transformations, which
+fail to type check in Core, are validated by Linear
 Core ($\S$~\ref{sec:main:metatheory}).
 
-\item We implement Linear Core as a GHC plugin which typechecks linearity in
+\item We implement Linear Core as a GHC plugin which checks linearity in
 all intermediate Core programs produced during the compilation process, showing
-it accepts the intermediate programs resulting from (laziness-abusing)
-transformations in linearity-heavy Haskell libraries, such as
+it accepts the programs resulting from transformations in libraries such as
 \texttt{linear-base} ($\S$~\ref{sec:discuss:implementation}).
@@
 A linear type system statically guarantees that linear resources are
-\emph{consumed} exactly once. Consequently, whether a program is well-typed
-under a linear type system intrinsically depends on the precise definition of
-\emph{consuming} a resource. Even though \emph{consuming} a resource is
-commonly regarded as synonymous with the \emph{syntactic occurrence} of the resource
-in the program, that is not always the case.
-%
-In fact, this chapter highlights the distinction between using resources
-\emph{syntactically} and \emph{semantically} as a principal limitation of
-linear type systems for non-strict languages, with examples motivated by the
-constructs available in GHC Core and how they are evaluated.
-
-% and we show how a linear type system found on
-% \emph{consuming resources semantically} rather than \emph{using resources
-% syntactically} can accept
-
+\emph{consumed} exactly once. Consequently, whether a program is
+well-typed under a linear type system intrinsically depends on the
+definition of \emph{consuming} a resource.
+In general, \emph{consuming} a resource is equated with its 
+\emph{syntactic occurrence} in the program. However, as this section
+makes clear, when considering linear type systems for non-strict
+languages, a distinction between using
+resources \emph{syntactically} and \emph{semantically} is needed.
+%
 Consider the following program in a functional Haskell-like language,
@@
 %
-In this example, the \lstinline{handle} appears to be closed before
+
+In the code above, the \lstinline{handle} appears to be closed before
 returned, whereas in fact the handle will only be closed if the let bound
-computation is effectively run (i.e. evaluated).
+computation is effectively run (i.e.,~evaluated).
 %

### 2024-01-24 icfp.tex
revised a bit of 2
Turns assumptions into lemmas, and make lemma 2 provable

@@
 
+\todo[inline]{Cut this one}
 Let bindings can also go unused if they are defined before branching on case
@@
 
+\renewcommand{\DeltaLinearRelationLemma}{
+\begin{lemma}[$\D \Rightarrow 1$]
+Any $\Delta$-variable with a fully-irrelevant usage environment can be made a linear variable as long as its irrelevant environment is dropped from the linear environment (and the unrestricted env.
+is updated accordingly).\\
+If $\G,\x[\irr{\D}]; \irr{\D},\D' \vdash e : \vp$
+then $\G[x/\irr{\D}]; \D',\xl \vdash e :\vp$.
+\end{lemma}
+}
+
+\renewcommand{\LinearDeltaRelationLemma}{
+\begin{lemma}[$1 \Rightarrow \D$]
+A linear variable can be moved to the unrestricted context as a $\D$-var with any usage environment $\D$ by introducing $\D$ in the linear resources (and the unrestricted env.
+is updated accordingly).\\
+If $\G; \D',\xl \vdash e :\vp$
+then $\G[\D/x],\xD; \D,\D' \vdash e : \vp$.
+\end{lemma}
+}
+
+\renewcommand{\DeltaUnrestrictedRelationLemma}{
+\begin{lemma}[$x{:}_\omega\sigma = x{:}_{\cdot}\sigma$]
+An unrestricted variable is equivalent to a $\D$-var with an empty usage environment.\\
+$\G,\xo; \D \vdash e : \vp$ iff $\G,\x[\cdot]; \D \vdash e : \vp$
+\end{lemma}
+}
+
 We use two main assumptions in our proofs, which are dual.

### 2024-01-24 icfp.tex
notes

--- 2024-01-24 icfp.tex
@@
 
-\title{Type-checking Linearity in GHC Core}
-\subtitle{Semantic Linearity in a Lazy Optimising Compiler}
+\title{Semantic Linearity in a Lazy Optimising Compiler}
+\subtitle{Type-checking Linearity in GHC Core}
 
@@
 \affiliation{
-   \institution{Well-Typed}
+   \institution{Well-Typed LLP}
   %\city{City}
-   \country{Something}}
+  %\country{Something}
+ }
  \email{xxx@yyy.com}
@@
 \begin{lstlisting}
-f13 :: a *' $\lolli$ '* a *' $\lolli$ '* (a *' $\lolli$ '* a *' $\lolli$ '* c) -> c
+f13 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* a *' $\lolli$ '* c) -> c
 f13 x y use = case (x,y) of z { (a,b) -> use a x }
@@
 \end{noway}
+% Remover
 \begin{noway}
@@
 
+%Remover estes
 \begin{figure}[t]
@@
 
+%Trocar ordem com a metatheory
 \section{Linear Core as a GHC Plugin\label{sec:discuss:implementation}}
@@
 
-The usage environments of a recursive group of binders (that is not necessarily
-strongly-connected) is computed in two separate passes.
-%
-First, we calculate a \emph{naive environment} by recording the linear
-resources semantically used in the binder bodies, while counting \emph{uses}
-(not syntactic occurrences) of the mutually recursive let variables.
-%
-Second, the binder names and corresponding \emph{naive environment} (mapping
-each linear resource to $1$ and the recursive variables to the $n$ number of
-times they are \emph{used}) in Cartesian pairs are given as input to
-% Algorithm~\ref{computeRecUsages},
-an algorithm which computes the actual usage environment
-of each binder.
+% The usage environments of a recursive group of binders (that is not necessarily
+% strongly-connected) is computed in two separate passes.
+% %
+% First, we calculate a \emph{naive environment} by recording the linear
+% resources semantically used in the binder bodies, while counting \emph{uses}
+% (not syntactic occurrences) of the mutually recursive let variables.
+% %
+% Second, the binder names and corresponding \emph{naive environment} (mapping
+% each linear resource to $1$ and the recursive variables to the $n$ number of
+% times they are \emph{used}) in Cartesian pairs are given as input to
+% % Algorithm~\ref{computeRecUsages},
+% an algorithm which computes the actual usage environment
+% of each binder.
 %
@@
 %
-Intuitively, the algorithm, for each recursive binder, iterates over the
-(initially naive) usage environments and substitutes occurrences of the
-recursive binders by their corresponding usage environment,
-scaled up by the amount of times that recursive binder is used in the
-environment being updated. The result is the least upper bound of linear
-resources used by each strongly-connected group of mutually recursive bindings
-(for each binder of such a group), since all occurrences of the recursive
-binders in the usage environments will be substituted away by the corresponding
-recursive usage environments.
+% Intuitively, the algorithm, for each recursive binder, iterates over the
+% (initially naive) usage environments and substitutes occurrences of the
+% recursive binders by their corresponding usage environment,
+% scaled up by the amount of times that recursive binder is used in the
+% environment being updated. The result is the least upper bound of linear
+% resources used by each strongly-connected group of mutually recursive bindings
+% (for each binder of such a group), since all occurrences of the recursive
+% binders in the usage environments will be substituted away by the corresponding
+% recursive usage environments.
 %
@@
 %
-The complexity of computing a usage environment for $n$ recursive let binders
-is quadratic in $n$,
-% i.e. the algorithm has $O(n^2)$ complexity,
-but this is not an issue since it is uncommon to have more than a handful of
-binders in the same recursive let block.
+% The complexity of computing a usage environment for $n$ recursive let binders
+% is quadratic in $n$,
+% % i.e. the algorithm has $O(n^2)$ complexity,
+% but this is not an issue since it is uncommon to have more than a handful of
+% binders in the same recursive let block.
 
@@
 example is a program to which a rewrite rule was applied, resulting in an
-application of unrestricted function to a linear resource in the first argument
-to \incode{build}:
-\begin{lstlisting}
-take :: Int -> Replicator a *' $\lolli$ '* [a]
-take = \ (ds :: Int) (r :: Replicator b) ->
-     case ds of
-       1 -> build (\(c :: b -> a -> a) (n :: a) -> c (extract r) n)
-       ...
-\end{lstlisting}
-We know the linear resource is still used linearly because \incode{build} will
-always instance the unrestricted function to $(:)$ (read \emph{Cons}), which is
-linear. Additionally, these include programs scrutinizing a value of a type of
-which all constructors have no linear components, but only matching on the
-default alternative, programs where common-sub-expression elimination
-substituted more than one occurrence of \incode{Ur x} by \incode{y}, in the body of
-scrutinizing \incode{y} (a linear variable), and other programs affected by rewrite rules.
+application of unrestricted function to a linear resource.
+% in the first argument
+% to \incode{build}:
+% \begin{lstlisting}
+% take :: Int -> Replicator a *' $\lolli$ '* [a]
+% take = \ (ds :: Int) (r :: Replicator b) ->
+%      case ds of
+%        1 -> build (\(c :: b -> a -> a) (n :: a) -> c (extract r) n)
+%        ...
+% \end{lstlisting}
+% We know the linear resource is still used linearly because \incode{build} will
+% always instance the unrestricted function to $(:)$ (read \emph{Cons}), which is
+% linear. Additionally, these include programs scrutinizing a value of a type of
+% which all constructors have no linear components, but only matching on the
+% default alternative, programs where common-sub-expression elimination
+% substituted more than one occurrence of \incode{Ur x} by \incode{y}, in the body of
+% scrutinizing \incode{y} (a linear variable), and other programs affected by rewrite rules.
 %
@@
 semantically, and are simultaneously rejected by Linear Core and its
-implementation. An example, where the last component of \incode{HashMap}, \incode{wwz},
-is linear, but is not being consumed in the case alternative matching on it:
-\begin{lstlisting}
-jssvi :: Ur Bool *' $\lolli$ '* Set Int *' $\lolli$ '* Ur (TestT IO ())
-jssvi (a :: Ur Bool) (b :: Set Int)
-  = case a of
-     Ur ss -> case b of { HashMap wwx wwy wwz -> jump wjssAd ss }
-\end{lstlisting}
+implementation. These programs were not necessarily identified as bad
+by the linter.\footnote{rever}
+% An example, where the last component of \incode{HashMap}, \incode{wwz},
+% is linear, but is not being consumed in the case alternative matching on it:
+% \begin{lstlisting}
+% jssvi :: Ur Bool *' $\lolli$ '* Set Int *' $\lolli$ '* Ur (TestT IO ())
+% jssvi (a :: Ur Bool) (b :: Set Int)
+%   = case a of
+%      Ur ss -> case b of { HashMap wwx wwy wwz -> jump wjssAd ss }
+% \end{lstlisting}
 %
@@
 
-\section{Conclusion\label{sec:discussion}}
-
-Linear Core is an intermediate language with a type system system that
-understands (semantic) linearity in the presence of laziness, suitable for
-optimising compilers with these characteristics which leverage laziness and
-(possibly) linearity in its transformations.
+\section{Related and Future Work\label{sec:discussion}}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Related Work}
-
-In this section we discuss related work, namely, Linear
-Haskell~\cite{cite:linearhaskell}, Linear Mini-Core~\cite{cite:minicore}, and
-linearity-influenced optimising
-transformations~\cite{cite:let-floating,peytonjones1997a,cite:linearhaskell}.
+%paragraph{Related Work}
+
+% In this section we discuss related work, namely, Linear
+% Haskell~\cite{cite:linearhaskell}, Linear Mini-Core~\cite{cite:minicore}, and
+% linearity-influenced optimising
+% transformations~\cite{cite:let-floating,peytonjones1997a,cite:linearhaskell}.
 
@@
 
-\paragraph{Linear Haskell\label{sec:related-work-linear-haskell}}
-
+\paragraph{Linear Haskell\label{sec:related-work-linear-haskell}.}
+%
 % \todo[inline]{Say something about Linear Haskell's lazy operational semantics,
@@
 % laziness}
-
-Haskell, contrary to most programming languages with linear types, has existed
-for 31 years of its life \emph{without} linear types. As such, the introduction
-of linear types to Haskell comes with added challenges that do not exist in
-languages that were designed with linear types from the start:
-%
-\begin{itemize}
-    \item Backwards compatibility. The addition of linear types shouldn't break
-        all existing Haskell code.
-    \item Code re-usability. The linearly-typed part of Haskell's ecosystem and
-        its non-linearly-typed counterpart should fit in together and it must be
-        possible to define functions readily usable by both sides
-        simultaneously.
-    \item Future-proofing. Haskell, despite being an
-        industrial-strength language, is also a petri-dish for experimentation
-        and innovation in the field of programming languages. Therefore, Linear
-        Haskell takes care to accommodate possible future features, in
-        particular, its design is forwards compatible with affine and dependent
-        types.
-\end{itemize}
-%
-Linear Haskell~\cite{cite:linearhaskell} is thus concerned with retrofitting
-linear types in Haskell, taking into consideration the above design goals, but
-is not concerned with extending Haskell's intermediate language(s),
+%
+% Haskell, contrary to most programming languages with linear types, has existed
+% for 31 years of its life \emph{without} linear types. As such, the introduction
+% of linear types to Haskell comes with added challenges that do not exist in
+% languages that were designed with linear types from the start:
+% %
+% \begin{itemize}
+%     \item Backwards compatibility. The addition of linear types shouldn't break
+%         all existing Haskell code.
+%     \item Code re-usability. The linearly-typed part of Haskell's ecosystem and
+%         its non-linearly-typed counterpart should fit in together and it must be
+%         possible to define functions readily usable by both sides
+%         simultaneously.
+%     \item Future-proofing. Haskell, despite being an
+%         industrial-strength language, is also a petri-dish for experimentation
+%         and innovation in the field of programming languages. Therefore, Linear
+%         Haskell takes care to accommodate possible future features, in
+%         particular, its design is forwards compatible with affine and dependent
+%         types.
+% \end{itemize}
+% %
+Linear Haskell~\cite{cite:linearhaskell} is concerned with retrofitting
+linear types in Haskell but
+is not concerned with extending GHC's intermediate language(s),
 which presents its own challenges. 
-
+%
 Nonetheless, while the Linear Haskell work keeps Core unchanged, its

### 2024-01-25 icfp.tex
passed through Section 2

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Introduction}
+\section{Introduction}\label{sec:intro}
 
@@
 %
-Expanding on this, consider the above example program under distinct
-evaluation strategies: under call-by-value, %With \emph{eager}
+Expanding on this, consider the above example program under
+under call-by-need semantics.  %With \emph{eager}
 % evaluation semantics,
-the let
-  bound expression \lstinline{close handle} is eagerly evaluated, and the \lstinline{handle} will be closed before being
-returned. It is clear that a linear type system should not accept such a
-program since the linear resource \lstinline{handle} is duplicated -- it is used in a
-computation that closes it, while still being made available to the caller of
-the function;
-%
-on the other hand, with \emph{call-by-need}, the let
+% the let
+%   bound expression \lstinline{close handle} is eagerly evaluated, and the \lstinline{handle} will be closed before being
+% returned. It is clear that a linear type system should not accept such a
+% program since the linear resource \lstinline{handle} is duplicated -- it is used in a
+% computation that closes it, while still being made available to the caller of
+% the function;
+% %
+% on the other hand, with \emph{call-by-need},
+The let
 bound expression will only be evaluated when the binding \lstinline{x} is needed. We return
 the \lstinline{handle} right away, and the let binding is forgotten as it cannot be
-used outside the scope of the function, so the handle is not closed by \lstinline{f}.
-Under the lens of \emph{call-by-need} evaluation, \emph{using} a resource in a
+used outside the scope of the function, so the handle is not closed by
+calling \lstinline{f}.
+Under the lens of \emph{call-by-need}, \emph{using} a resource in a
 let binding only results in the resource being \emph{consumed} if the binding
@@
 \emph{call-by-need} evaluation should accept the above program, unlike a linear
-type system for the same program evaluated \emph{call-by-value}.
+type system for the same program evaluated with \emph{call-by-value}, where
+the let-bound expression is always evaluated.
 
@@
 From this observation, and exploring the connection between computation and evaluation,
-% \todo{Alguma dificuldade em dizer exatamente como é que evaluation drives/is computation}
-it becomes clear that \emph{linearity} and \emph{consuming resources}, in the
-above example and for programs in general, should be defined in function of the
-language's evaluation strategy.
-%
-We turn our focus to \emph{linearity} under \emph{call-by-need}, not only
+% it becomes clear that \emph{linearity} and \emph{consuming resources}, in the
+% above example and for programs in general, should be defined in function of the
+% language's evaluation strategy.
+%
+we turn our focus to \emph{linearity} under \emph{call-by-need}, not only
 because GHC Core is \emph{call-by-need}, but also because the distinction
@@
 that optimising transformations preserve linearity, and with the goal
-of understanding linearity in a non-strict context, this section helps the
-reader build an intuition for semantic linearity through examples of Core
-programs that are semantically linear but rejected by Core's linear type
+of understanding linearity in a non-strict context, this section aims to
+build an intuition for semantic linearity through examples of Core
+programs that are semantically linear but rejected by (the current) Core linear type
 system.
@@
 In the examples, a \colorbox{working}{\workingcolorname} background highlights programs that are
-syntactically linear and are accepted by Core's naive linear type system. A
+syntactically linear and accepted by Core's naive linear type system. A
 \colorbox{notyet}{\notyetcolorname} or \colorbox{limitation}{\limitationcolorname} background mark programs that are
-semantically linear, but are not seen as linear by Core's (naive wrt laziness) linear type
+\emph{semantically} linear, but are not seen as linear by Core's (naive wrt laziness) linear type
 system. Notably, the linear type system we develop in this work accepts all
@@
 A \colorbox{noway}{\nowaycolorname} background indicates that the program
-simply isn't linear, not even semantically, i.e. the program effectively
-discards or duplicates linear resources.
+violates linearity semantically, i.e.~the program effectively
+discards or duplicates linear resources. We discuss the three main
+syntactic constructs of Core, let bindings (non-recursive and
+recursive) and case expressions.
 
@@
 
-    \begin{notyet}
-\begin{lstlisting}
-f1 :: (a *' $\lolli$ '* b) -> a *' $\lolli$ '* b
-f1 use x = let y = use x in y
-\end{lstlisting}
-\end{notyet}
-In a linear type system, a non-strict let binding that depends on a linear
-resource \lstinline{x} doesn't consume the resource as long as the binding isn't
-evaluated -- the suspended computation only uses the resource if it is run. For
-this reason, we can't naively tell whether \lstinline{x} is consumed just by looking at
-the let binding body. In the following example, we assign a computation that
-depends on the resource \lstinline{x} to a binder, which is then
-returned.
+    %\begin{notyet}
+% \begin{lstlisting}
+% f1 :: (a *' $\lolli$ '* b) -> a *' $\lolli$ '* b
+% f1 use x = let y = use x in y
+% \end{lstlisting}
+% \end{notyet}
+% In a linear type system, a non-strict let binding that depends on a linear
+% resource \lstinline{x} doesn't consume the resource as long as the binding isn't
+% evaluated -- the suspended computation only uses the resource if it is run. For
+% this reason, we can't naively tell whether \lstinline{x} is consumed just by looking at
+% the let binding body. In the following example, we assign a computation that
+% depends on the resource \lstinline{x} to a binder, which is then
+% returned.
 
@@
 
-%
-The linear resource \lstinline{x} is used exactly once, since it is used exactly once in
-the body of the binding and the binding is used exactly once in the let body.
-%
-According to Linear Haskell's core calculus~$\lambda_{\to}^{q}$~\cite{cite:linearhaskell}, let
-bound variables are annotated with a multiplicity which is multiplied (as per
-the multiplicity semiring) by the multiplicity of all variables that are free
-in the binder's body.
-%
-In short, if a let binder is linear (has
-multiplicity $1$) then the linear variables free in its body are only used
-once; if the let binder is unrestricted (has multiplicity $\omega$) then the
-resources in its body are consumed many times, meaning no linear variables can
-occur in that let binder's body.
-%
-Unfortunately, GHC's implementation of Linear
-Haskell doesn't seem to infer multiplicities for lets yet, so while the above
-program should typecheck in Linear Haskell, it is rejected by GHC.
-
-The next example exposes the case in which the let binder is ignored in the let
-body. Here, the linear resource \incode{x} is used in \incode{y}'s body and in the let
-body, however, the resource is still used semantically linearly
-because \incode{y} isn't used at all, thus \incode{x} is consumed just
-once in the let body: 
-%
-\begin{notyet}
-\begin{lstlisting}
-f2 :: (a *' $\lolli$ '* a) -> a *' $\lolli$ '* a
-f2 use x = let y = use x in use x
-\end{lstlisting}
-\end{notyet}
-%
-Programmers don't often write bindings that are completely unused, yet, an
-optimising compiler will produce intermediate programs with unused bindings\footnote{Unused bindings are then
-also dropped by the optimising compiler} 
-from transformations such as inlining, which can substitute out occurrences
-of the binder (e.g. \incode{y} is inlined in the let body).
-
-\todo[inline]{Cut this one}
+% %
+% The linear resource \lstinline{x} is used exactly once, since it is used exactly once in
+% the body of the binding and the binding is used exactly once in the let body.
+% %
+% According to Linear Haskell's core calculus~$\lambda_{\to}^{q}$~\cite{cite:linearhaskell}, let
+% bound variables are annotated with a multiplicity which is multiplied (as per
+% the multiplicity semiring) by the multiplicity of all variables that are free
+% in the binder's body.
+% %
+% In short, if a let binder is linear (has
+% multiplicity $1$) then the linear variables free in its body are only used
+% once; if the let binder is unrestricted (has multiplicity $\omega$) then the
+% resources in its body are consumed many times, meaning no linear variables can
+% occur in that let binder's body.
+% %
+% Unfortunately, GHC's implementation of Linear
+% Haskell doesn't seem to infer multiplicities for lets yet, so while the above
+% program should typecheck in Linear Haskell, it is rejected by GHC.
+
+% The next example exposes the case in which the let binder is ignored in the let
+% body. Here, the linear resource \incode{x} is used in \incode{y}'s body and in the let
+% body, however, the resource is still used semantically linearly
+% because \incode{y} isn't used at all, thus \incode{x} is consumed just
+% once in the let body: 
+% %
+% \begin{notyet}
+% \begin{lstlisting}
+% f2 :: (a *' $\lolli$ '* a) -> a *' $\lolli$ '* a
+% f2 use x = let y = use x in use x
+% \end{lstlisting}
+% \end{notyet}
+% %
+% Programmers don't often write bindings that are completely unused, yet, an
+% optimising compiler will produce intermediate programs with unused bindings\footnote{Unused bindings are then
+% also dropped by the optimising compiler} 
+% from transformations such as inlining, which can substitute out occurrences
+% of the binder (e.g. \incode{y} is inlined in the let body).
+
+% \todo[inline]{Cut this one}
+In the example from the beginning of
+Section~\ref{sec:linearity-semantically} (similar to that towards the end
+of Section~\ref{sec:intro}), we discussed a scenario where an
+unused let binding can be used to justify a function where its
+argument occurs both in the (unused) let-bound expression and in the
+body of the let.
 Let bindings can also go unused if they are defined before branching on case
@@
 be evaluated only if it occurs in that branch.
-Both optimising transformations (float-out), and programmers used to non-strict evaluation, can produce
-programs with bindings that are selectively used in the case alternatives, for
+Both optimising transformations (float-out), and programmers used to
+non-strict evaluation often produce
+programs with bindings that are selectively used in case alternatives, for
 instance:
@@
 \begin{lstlisting}
-f3 :: (a *' $\lolli$ '* a) -> Bool -> a *' $\lolli$ '* a
-f3 use bool x = let y = use x in case bool of
+f :: (a *' $\lolli$ '* a) -> Bool -> a *' $\lolli$ '* a
+f use bool x = let y = use x in case bool of
       True -> x 
@@
 %
-This example essentially merges \lstinline{f1} with \lstinline{f2}, using \lstinline{x} directly in one
-branch and using \lstinline{y} in the other. Semantically, this program is linear because
-the linear resource \lstinline{x} ends up being used exactly once in both case
-alternatives, directly or indirectly.
-
-Shifting our focus from not using a let binding to using it (more than once),
-we reiterate that a let binding creates a \emph{thunk} which is only evaluated
-once, and re-used subsequently. Despite the binder body only being evaluated
-once, and thus its resources only used once to compute a result, we can still
-only consume said result of the computation once -- perhaps surprisingly, as the
+This example uses \lstinline{x} directly in one
+branch and the let-bound \lstinline{y} in the other. Semantically, this program is linear because
+the linear resource \lstinline{x} ends up is used exactly once in both case
+alternatives, directly in the \lstinline{True} branch or via the
+let-bound \lstlinline{y} in the \lstinline{False} branch.
+
+Shifting our focus from not using a let binding to using it
+more than once.
+Despite the binder body only being evaluated
+once due to call-by-need, and thus its resources being only used once to compute a result, we can still
+only consume the result of the computation once -- perhaps surprisingly, as the
 perception so far is that ``resources are consumed during computation'' and
@@
 \begin{lstlisting}
-f4 :: (a *' $\lolli$ '* b) -> a *' $\lolli$ '* (b, b)
-f4 use x = let y = use x in (y, y)
+f1 :: (a *' $\lolli$ '* b) -> a *' $\lolli$ '* (b, b)
+f1 use x = let y = use x in (y, y)
 \end{lstlisting}
@@
 Intuitively, the result of the computation must also be used exactly once,
-despite being effectively computed just once, because said result may still
-contain (parts of) the linear resource. The trivial example is \lstinline{f4} applied to
+despite being effectively computed just once, since the result may
+still capture (parts of) the linear resource. The trivial example is \lstinline{f1} applied to
 \incode{id} -- the result of computing \incode{id x} is \incode{x}, and \incode{x} must definitely not be
@@
 \begin{lstlisting}
-f5 :: (a *' $\lolli$ '* a) -> a *' $\lolli$ '* ()
-f5 use x = let y = use x in let z = use y in ()
+f2 :: (a *' $\lolli$ '* a) -> a *' $\lolli$ '* ()
+f2 use x = let y = use x in let z = use y in ()
 \end{lstlisting}
@@
 % \parawith{Summary}
-The examples so far build an intuition for semantic linearity in the presence
-of lazy let bindings. In essence, an unused let binding doesn't consume any
+% The examples so far build an intuition for semantic linearity in the presence
+% of lazy let bindings. 
+In essence, an unused let binding doesn't consume any
 resources, and a let binding used exactly once consumes its resources exactly
@@
 \paragraph{Recursive let bindings\label{sec:semantic-linearity-examples:recursive-lets}}
-Second, we look into recursive let bindings. For the most part,
+% Second, we look into recursive let bindings.
+For the most part,
 recursive let bindings behave as non-recursive let bindings, i.e. we must use them \emph{at
@@
 bindings is that the binders bound in that group can occur, without restrictions, in
-the bodies of those same binders. The same way that, in a let body, evaluating
-a binding that uses some resource exactly once consumes that resource once,
-using the binding in its own definition also entails using that resource once.
+the bodies of those same binders. % The same way that, in a let body, evaluating
+% a binding that uses some resource exactly once consumes that resource once,
+% using the binding in its own definition also entails using that resource once.
 Consider the following program, that calls a recursive let-bound function
@@
 \begin{lstlisting}
-f6 :: Bool -> a *' $\lolli$ '* a
-f6 bool x = let go b = case b of
+f3 :: Bool -> a *' $\lolli$ '* a
+f3 bool x = let go b = case b of
                      True -> x
@@
 %
-Function \incode{f6} is semantically linear because, iff it is consumed exactly once,
+Function \incode{f3} is semantically linear because, iff it is consumed exactly once,
 then \incode{x} is consumed exactly once. We can see this by case analysis on \incode{go}'s argument:
@@
 \begin{lstlisting}
-f7 :: Bool -> a *' $\lolli$ '*  a
-f7 bool x = let go b = case b of
+f4 :: Bool -> a *' $\lolli$ '*  a
+f4 bool x = let go b = case b of
                      True -> x
@@
 If \incode{go} doesn't terminate, we aren't able to compute (nor consume) the result
-of \incode{f7}, so we do not promise anything about \incode{x} being consumed (\incode{f7}'s
+of \incode{f4}, so we do not promise anything about \incode{x} being consumed (\incode{f4}'s
 linearity holds trivially). If it did terminate, it would consume \incode{x} exactly
@@
 
-Determining the linear resources used in a recursive binding might feel
-peculiar since we need to know the linear resources used by the binder to determine the linear resources it uses.
-%
-The paradoxical definition is difficult to grasp, just how learning that a
-function can be defined in terms of itself is perplexing when one is first
-introduced to general recursion.
-%
-Informally, we \emph{assume} the binding will consume some linear resources
-exactly once, and use that assumption when reasoning about recursive calls such
-that those linear resources are used exactly once.
-
-%This high-level reasoning isn't amenable for a computer that must check
-%whether the program is linear.
-
-Generalizing, we need to find a set of linear resources ($\Delta$) that satisfies the recursive equation
-\footnote{This set of resources will basically be the least upper bound of the sets of resources used in each
-mutually recursive binding scaled by the times each binding was used}
-arising from given binding $x$, such that:
-%\begin{enumerate}
-(1) Occurrences of $x$ in its own body are synonymous with using all resources in $\Delta$ exactly once;
-(2) and if the binding $x$ is fully evaluated, then all resources in $\Delta$ are consumed exactly once.
-%\end{enumerate}
-Finding a solution to this equation is akin to finding a (principle) type for a
-recursive binding: the binding needs to be given a type such that occurrences of
-that binding in its own body typecheck using that type.
-%
-Foreshadowing, the core system we developed assumes recursive let bindings to
-be annotated with a set of resources satisfying the equations; but we also
-present an algorithm to determine this solution, and distinguish between an
-\emph{inference} and a \emph{checking} phase, where we first determine the
-linear resources used by a group of recursive bindings and only then check
-whether the binding is linear, in our implementation of checking of recursive
-lets.
+% Determining the linear resources used in a recursive binding might feel
+% peculiar since we need to know the linear resources used by the binder to determine the linear resources it uses.
+% %
+% The paradoxical definition is difficult to grasp, just how learning that a
+% function can be defined in terms of itself is perplexing when one is first
+% introduced to general recursion.
+% %
+% Informally, we \emph{assume} the recursive binding will consume some linear resources
+% exactly once, and use that assumption when reasoning about recursive calls such
+% that those linear resources are used exactly once.
+% Generalizing, we need to find a set of linear resources ($\Delta$) that satisfies the recursive equation
+% \footnote{This set of resources will basically be the least upper bound of the sets of resources used in each
+% mutually recursive binding scaled by the times each binding was used}
+% arising from given binding $x$, such that:
+% %\begin{enumerate}
+% (1) Occurrences of $x$ in its own body are synonymous with using all resources in $\Delta$ exactly once;
+% (2) and if the binding $x$ is fully evaluated, then all resources in $\Delta$ are consumed exactly once.
+% %\end{enumerate}
+% Finding a solution to this equation is akin to finding a (principle) type for a
+% recursive binding: the binding needs to be given a type such that occurrences of
+% that binding in its own body typecheck using that type.
+% %
+% Foreshadowing, the core system we developed assumes recursive let bindings to
+% be annotated with a set of resources satisfying the equations; but we also
+% present an algorithm to determine this solution, and distinguish between an
+% \emph{inference} and a \emph{checking} phase, where we first determine the
+% linear resources used by a group of recursive bindings and only then check
+% whether the binding is linear, in our implementation of checking of recursive
+% lets.
 % ROMES:TODO:!!!
@@
 
-There might not be a solution to the set of equations. In this case, the
-binding undoubtedly entails using a linear resource more than once. For
-example, if we use a linear resource \incode{x} in one case alternative, and invoke
+% There might not be a solution to the set of equations. In this case, the
+% binding undoubtedly entails using a linear resource more than once.
+
+Note that if we use a linear resource \incode{x} in one case alternative, and invoke
 the recursive call more than once, we might eventually consume \incode{x} more than
@@
 \begin{lstlisting}
-f9 :: Bool -> Bool *' $\lolli$ '* Bool
-f9 bool x =
+f5 :: Bool -> Bool *' $\lolli$ '* Bool
+f5 bool x =
   let go b
@@
 \end{noway}
-Note that if returned \incode{x} instead of \incode{go bool} in the let body, then, despite
-the binding using \incode{x} more than once, we would still consume \incode{x} exactly once,
+However, if we returned \incode{x} instead of \incode{go bool} in the let body, then, despite
+the binding using \incode{x} more than once, we would consume \incode{x} exactly once,
 since recursive bindings are still lazy.
-
-Lastly, we extend our single-binding running example to use two mutually recursive bindings that
-depend a linear resource:
-\begin{notyet}
-\begin{lstlisting}
-f10 :: Bool -> a *' $\lolli$ '* a
-f10 bool x =
-  let go1 b = case b of
-           True -> go2 b
-           False -> go1 (not b)
-       go2 b = case b of
-           True -> x
-           False -> go1 b
-  in go1 bool
-\end{lstlisting}
-\end{notyet}
-As before, we must find a solution to the set of equations defined by the
-mutually recursive bindings to determine which resources will be consumed.
-In this case, \incode{go1} and \incode{go2} both consume \incode{x} exactly once if evaluated.
-We additionally note that a strongly connected group of recursive bindings
-(i.e. all bindings are transitively reachable from any of them) will always
-consume the same set of resources -- if all bindings are potentially reachable,
-then all linear resources are too.
+%
+These ideas extend to mutually recursive bindings that depend on
+linear resources, as we discuss in Section~\ref{sec:recursivelets}.
+
+% Lastly, we extend our single-binding running example to use two mutually recursive bindings that
+% depend a linear resource:
+% \begin{notyet}
+% \begin{lstlisting}
+% f10 :: Bool -> a *' $\lolli$ '* a
+% f10 bool x =
+%   let go1 b = case b of
+%            True -> go2 b
+%            False -> go1 (not b)
+%        go2 b = case b of
+%            True -> x
+%            False -> go1 b
+%   in go1 bool
+% \end{lstlisting}
+% \end{notyet}
+% As before, we must find a solution to the set of equations defined by the
+% mutually recursive bindings to determine which resources will be consumed.
+% In this case, \incode{go1} and \incode{go2} both consume \incode{x} exactly once if evaluated.
+% We additionally note that a strongly connected group of recursive bindings
+% (i.e. all bindings are transitively reachable from any of them) will always
+% consume the same set of resources -- if all bindings are potentially reachable,
+% then all linear resources are too.
 
 % \parawith{Summary}
-Summarising, recursive let bindings behave like non-recursive let bindings in
-that if they aren't consumed, the resources they depend on aren't consumed
-either.  However, recursive let bindings are defined in terms of themselves, so
-the set of linear resources that will be consumed when the binder is evaluated
-is also defined in terms of itself (we need it to determine what resources are
-used when we recurse). We can intuitively think of this set of linear resources
-that will be consumed as a solution to a set of equations defined by a group of
-mutually recursive bindings, which we are able to reason about without an
-algorithm for simpler programs. In our work, the core type system isn't
-concerned with deriving said solution, but we present a simple algorithm for
-inferring it with our implementation.
+% Summarising, recursive let bindings behave like non-recursive let bindings in
+% that if they aren't consumed, the resources they depend on aren't consumed
+% either.  However, recursive let bindings are defined in terms of themselves, so
+% the set of linear resources that will be consumed when the binder is evaluated
+% is also defined in terms of itself (we need it to determine what resources are
+% used when we recurse). We can intuitively think of this set of linear resources
+% that will be consumed as a solution to a set of equations defined by a group of
+% mutually recursive bindings, which we are able to reason about without an
+% algorithm for simpler programs. In our work, the core type system isn't
+% concerned with deriving said solution, but we present a simple algorithm for
+% inferring it with our implementation.
 
 \paragraph{Case expressions}
-Finally, we discuss semantic linearity for case expressions, which have been
-purposefully left for last as the key ingredient that brings together the
-semantic linearity insights developed thus far, because, essentially,
+Finally, we discuss case expressions,
+which are the key ingredient that brings together the
+semantic linearity insights developed thus far given that
 \emph{case expressions drive evaluation} and semantic linearity can only be
-understood in function of how expressions are evaluated.
-
-Up until now, the example functions have always linearly transformed linear
-resources, taking into consideration how expressions will be evaluated (and thus
-consumed) to determine if resources are being used linearly. However, there
+understood in terms of how expressions are evaluated.
+
+Up until now, there
 have been no examples in which linear resources are \emph{fully consumed} in
@@
 application simply needs to be evaluated to guarantee its linear argument is
-consumed (functions that return an unrestricted value).
+consumed (e.g.,~functions that return an unrestricted value).
 %
@@
 %
-For example, the entry point to the linear array API presented in Linear
-Haskell takes such a function as its second argument:
-\begin{working}
-\begin{lstlisting}
-newMArray :: Int -> (MArray a *' $\lolli$ '* Unrestricted b) *' $\lolli$ '* b
-\end{lstlisting}
-\end{working}
+% For example, the entry point to the linear array API presented in Linear
+% Haskell takes such a function as its second argument:
+% \begin{working}
+% \begin{lstlisting}
+% newMArray :: Int -> (MArray a *' $\lolli$ '* Unrestricted b) *' $\lolli$ '* b
+% \end{lstlisting}
+% \end{working}
 %(The second argument is a function that consumes |MArray a| linearly
@@
 %body.)
-
+%
 % ROMES:TODO: Há qualquer coisa que não gosto nada! na coerência deste paragrafo
-
-In short, case expressions enable us to consume resources and so write
-functions that fully consume their linear arguments. To understand exactly how,
+%
+Case expressions enable us to consume resources and so write
+functions that fully consume their linear arguments.
+% CORTEI
+To understand exactly how,
 we turn to the definition of \emph{consuming} a resource from Linear
@@
 $\ov{\rho_i \to e_i}$ are the case \emph{alternatives}, composed of a pattern
-$\rho_i$ and of the corresponding expression $e_i$. Critically,
+$\rho_i$ and of the corresponding expression $e_i$. 
 %\begin{enumerate}
 %\item
-the case scrutinee is always evaluated to Weak Head Normal Form (WHNF);
+The case scrutinee is always evaluated to Weak Head Normal Form (WHNF) and
 %
 % \item
-evaluating an expression that is already in WHNF is a no-op, that is,
-no computation occurs; 
+% evaluating an expression that is already in WHNF is a no-op, that is,
+% no computation occurs; 
 %
 the case binder is an alias to the result of evaluating the scrutinee
-to WHNF; and,
-the alternative patterns are always exhaustive.
+to WHNF. Moreover, in Core, the alternative patterns are always exhaustive.
 % i.e. there always exists
@@
 
-To explore these case properties in the presence of linearity, we start with an
+We start with an
 example of a program that constructs a tuple from linear resources then pattern
@@
 \begin{lstlisting}
-f11 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
-f11 x y use = case (x,y) of { (a,b) -> use a b }
+f6 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
+f6 x y use = case (x,y) of { (a,b) -> use a b }
 \end{lstlisting}
@@
 What might be more surprising is that a similar program which discards the
-pattern variables and instead uses the resources in the scrutinee is also
-semantically linear, despite not being accepted by Linear Haskell:
+pattern variables and instead uses the resources in the scrutinee is 
+\emph{semantically} linear, despite not being accepted by Linear Haskell:
 \begin{notyet}
 \begin{lstlisting}
-f12 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
-f12 x y use = case (x,y) of z { (a,b) -> use x y }
+f7 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
+f7 x y use = case (x,y) of z { (a,b) -> use x y }
 \end{lstlisting}
 \end{notyet}
-We justify that \incode{f12} is linear by appealing to the fact that the
-expression (the tuple) being scrutinized is already in WHNF, evaluating it will
-not consume neither $x$ nor $y$. Even if the tuple was constructed with
+We justify that \incode{f7} is linear by appealing to the fact that 
+the tuple being scrutinized is already in WHNF and so evaluating it will
+not consume either $x$ nor $y$. Even if the tuple was constructed with
 two expressions using $x$ and $y$ respectively, no computation would happen
-since we aren't using neither $a$ nor $b$ (thereby never forcing the arguments
+since we are not using $a$ or $b$ (thereby never forcing the arguments
 of the tuple). However, if we did use $a$ in the case body, then $x$ would be unavailable:
@@
 \begin{lstlisting}
-f13 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* a *' $\lolli$ '* c) -> c
-f13 x y use = case (x,y) of z { (a,b) -> use a x }
+f8 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* a *' $\lolli$ '* c) -> c
+f8 x y use = case (x,y) of z { (a,b) -> use a x }
 \end{lstlisting}
@@
 bindings being mutually exclusive to the resources that define them.
-By forcing the pattern variable (or the let binding), we run the computations
-defined in terms of the linear variables used for that constructor argument (or
-let binder body), but otherwise, if we don't use those binders, then we don't
-run the computation thus no resources are consumed.
-
-A third option for this example is to use the case binder $z$ instead of $a,b$ or $x,y$:
+By forcing the pattern variable we run the computations
+defined in terms of the linear variables used for that constructor
+argument. Otherwise, if we do not use the pattern variables, then we
+do not run the computation and thus no resources are consumed.
+
+A third option is to use the case binder $z$ instead of $a,b$ or $x,y$:
 \begin{notyet}
 \begin{lstlisting}
-f14 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
-f14 x y use = case (x,y) of z { (a,b) -> uncurry use z }
+f9 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
+f9 x y use = case (x,y) of z { (a,b) -> uncurry use z }
 \end{lstlisting}
@@
 resource $y$, and the case binder $z$ entails using both $a$ and $b$.
-
+%
 Dually, consider the scrutinee to be an expression that's not in WHNF, s.t.
-evaluating it to WHNF will require doing computation and thus consume linear
-resources that are used in it:
+evaluating it to WHNF will require performing computation and thus consume linear
+resources:
 \begin{notyet}
 \begin{lstlisting}
-f15 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* (c,d)) -> (c,d)
-f15 x y use = case use x y of z { (a,b) -> z }
+f10 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* (c,d)) -> (c,d)
+f10 x y use = case use x y of z { (a,b) -> z }
 \end{lstlisting}
@@
 alternatives, but we \emph{must} still use either the case binder $z$ or the linear
-pattern variables $a,b$, e.g. it would be quite disastrous if any of the following typechecked:
-\begin{noway}
-\begin{lstlisting}
-doubleFree :: Ptr *' $\lolli$ '* (Ptr *' $\lolli$ '* Result) -> Result
-doubleFree x free = case free x of z { Result v -> free x }
-\end{lstlisting}
-\end{noway}
-% Remover
-\begin{noway}
-\begin{lstlisting}
-leakPointer :: Ptr -> ()
-leakPointer x = case id x of z { _ -> () }
-\end{lstlisting}
-\end{noway}
+pattern variables $a,b$. %  e.g. it would be quite disastrous if any of the following typechecked:
+% \begin{noway}
+% \begin{lstlisting}
+% doubleFree :: Ptr *' $\lolli$ '* (Ptr *' $\lolli$ '* Result) -> Result
+% doubleFree x free = case free x of z { Result v -> free x }
+% \end{lstlisting}
+% \end{noway}
+% % Remover
+% \begin{noway}
+% \begin{lstlisting}
+% leakPointer :: Ptr -> ()
+% leakPointer x = case id x of z { _ -> () }
+% \end{lstlisting}
+% \end{noway}
 %
@@
 %
-The result of evaluating the scrutinee must be consumed exactly to guarantee
-that the resources used in the scrutinee are fully consumed, or risk them being only ``almost'' consumed. Take for example
-\incode{use} in \incode{f15} to simply be \incode{(,)}: it is not sufficient for \incode{use x y} to be
+The result of evaluating the scrutinee must be consumed exactly once to guarantee
+that the resources used in the scrutinee are fully consumed. %, or
+                                %risk them being only ``almost''
+                                %consumed.
+For instance, take
+\incode{use} in \incode{f10} to be \incode{(,)}: it is not sufficient for \incode{use x y} to be
 evaluated to WHNF to consume \incode{x} and \incode{y}. Otherwise, if all the resources were
 considered to be fully consumed after the scrutinee were evaluated in a case
-expression, we could simply ignore the pattern variables, effectively
-discarding linear resources (for cases such as the \incode{use = (,)} example). In
+expression, we could simply ignore the pattern variables, potentially
+discarding linear resources. In
 short, if the scrutinee is not in WHNF we must either consume the case binder
@@
 
-However, we must also consider pattern matches on constructors without
+We now focus on pattern matching on constructors without
 any linear components. If the constructor has no linear fields, it means the
@@
 \begin{lstlisting}
-f16 :: () *' $\lolli$ '* ()
-f16 x = case x of z { () -> z <> z }
+f11 :: () *' $\lolli$ '* ()
+f11 x = case x of z { () -> z <> z }
 \end{lstlisting}
@@
 \begin{lstlisting}
-f :: a *' $\lolli$ '* a
-f x = case K1 x of z { K2 -> x; K1 a -> x }
+f12 :: a *' $\lolli$ '* a
+f12 x = case K1 x of z { K2 -> x; K1 a -> x }
 \end{lstlisting}
@@
 any resource could be freely used in that branch, and the example is
-\emph{semantically} linear (despite not being seen as so by our system):
-%
-%
-%
-This particular example has a known constructor being scrutinized which might
-seem like an unrealistic example, but we recall that during the transformations
-programs undergo in an optimising compiler, many programs such as this
-naturally occur (e.g.,~if the definition of a function is inlined in the
-scrutinee).
+\emph{semantically} linear (despite not being seen as so by our system).
+%
+%
+%
+This particular example might seem unrealistic due to the pattern
+match on a known constructor, but we note that during the
+transformations programs undergo in an optimising compiler, many
+programs such as this naturally occur (e.g.,~if the definition of a
+function is inlined in the scrutinee).
 %
@@
 
-Further exploring that each linear field must be consumed exactly once, and
-that resources in WHNF scrutinees aren't consumed, we are able to construct
-more contrived examples (Figures~\ref{fig:badcase}
-and~\ref{fig:goodcase}). the example in Figure~\ref{fig:badcase} does not typecheck
-because the same linear field is used twice, but the example in
-Figure~\ref{fig:goodcase} does since it
-uses each linear field exactly once, despite pattern matching on the same
-components twice.
-
-%Remover estes
-\begin{figure}[t]
-  \begin{minipage}{0.5\textwidth}
-\begin{noway}
-\begin{lstlisting}
-f w = case w of z
-        (a,b) -> case (a,b) of z'
-                        (c,d) -> (a,c)
-\end{lstlisting}
-\end{noway}
-\vspace{-0.5cm}
-\caption{Bad Case\label{fig:badcase}}
-\end{minipage}%
-\begin{minipage}{0.5\textwidth}
-\begin{notyet}
-\begin{lstlisting}
-f w = case w of z
-        (a,b) -> case (a,b) of z'
-                        (c,d) -> (a,d)
-\end{lstlisting}
-\end{notyet}
-\vspace{-0.5cm}
-\caption{Good Case\label{fig:goodcase}}
-\end{minipage}
-\end{figure}
+% Further exploring that each linear field must be consumed exactly once, and
+% that resources in WHNF scrutinees aren't consumed, we are able to construct
+% more contrived examples (Figures~\ref{fig:badcase}
+% and~\ref{fig:goodcase}). the example in Figure~\ref{fig:badcase} does not typecheck
+% because the same linear field is used twice, but the example in
+% Figure~\ref{fig:goodcase} does since it
+% uses each linear field exactly once, despite pattern matching on the same
+% components twice.
+
+% %Remover estes
+% \begin{figure}[t]
+%   \begin{minipage}{0.5\textwidth}
+% \begin{noway}
+% \begin{lstlisting}
+% f w = case w of z
+%         (a,b) -> case (a,b) of z'
+%                         (c,d) -> (a,c)
+% \end{lstlisting}
+% \end{noway}
+% \vspace{-0.5cm}
+% \caption{Bad Case\label{fig:badcase}}
+% \end{minipage}%
+% \begin{minipage}{0.5\textwidth}
+% \begin{notyet}
+% \begin{lstlisting}
+% f w = case w of z
+%         (a,b) -> case (a,b) of z'
+%                         (c,d) -> (a,d)
+% \end{lstlisting}
+% \end{notyet}
+% \vspace{-0.5cm}
+% \caption{Good Case\label{fig:goodcase}}
+% \end{minipage}
+% \end{figure}
 
@@
 is in WHNF, we can either use the resources from the scrutinee or the case
-binder in that alternative, if the scrutinee is not in WHNF, we \emph{must} use
+binder in that alternative. If the scrutinee is not in WHNF, we \emph{must} use
 the case binder, as it's the only way to linearly consume the result of
@@
 linear applications call-by-name, does not accept the above program as linear.
-Foreshadowing, this particular interaction between evaluation and linearity
+This interaction between evaluation and linearity
 comes up in the type preservation proof of our program, and is again explored
@@
 % \parawith{Summary}
-In summary, case expressions evaluate their scrutinees to WHNF,
-introduce a case binder, and bind pattern variables. If the scrutinee is
-already in WHNF, all resources occurring in it are still available in the case
-alternative, alongside the case binder and the pattern-bound variables. In the
-case alternative, either the resources of the scrutinee, the case binder, or
-the linearly bound pattern variables must be used exactly once, but mutually
-exclusively. For scrutinees not in WHNF, in the case alternative, either the
-case binder or the linear pattern variables need to be used, in mutual exclusion.
-If the pattern doesn't bind any linear resources, then it may be consumed
-unrestrictedly, and therefore the case binder may also be used unrestrictedly.
+% In summary, case expressions evaluate their scrutinees to WHNF,
+% introduce a case binder, and bind pattern variables. If the scrutinee is
+% already in WHNF, all resources occurring in it are still available in the case
+% alternative, alongside the case binder and the pattern-bound variables. In the
+% case alternative, either the resources of the scrutinee, the case binder, or
+% the linearly bound pattern variables must be used exactly once, but mutually
+% exclusively. For scrutinees not in WHNF, in the case alternative, either the
+% case binder or the linear pattern variables need to be used, in mutual exclusion.
+% If the pattern doesn't bind any linear resources, then it may be consumed
+% unrestrictedly, and therefore the case binder may also be used unrestrictedly.
 
@@
 
-\subsubsection{Recursive let bindings}
+\subsubsection{Recursive let bindings}\label{sec:recursivelets}
 

### 2024-01-26 icfp.tex
passed up to section 3.6.2

@@
 
-Finally, we consider the default case alternatives, also known as wildcards
+Finally, we consider default case alternatives, also known as wildcards
 (written $\_$), in the presence of linearity: matching against the wildcard
@@
 
-In this chapter, we develop a linear calculus $\lambda_\Delta^\pi$, dubbed \emph{Linear Core}, that 
-combines the linearity-in-the-arrow and multiplicity polymorphism introduced by
-Linear Haskell~\cite{cite:linearhaskell} with all the key features from GHC's Core
-language, except for type equality coercions\footnote{We explain a main avenue of
-future work, multiplicity coercions, in Section~\ref{sec:future-work}}.
-%
-Specifically, our core calculus is a linear lambda calculus with algebraic
-datatypes, case expressions, recursive let bindings, and multiplicity
-polymorphism.
-
-Linear Core makes much more precise the various insights discussed in the
+In this section, we develop a linear calculus $\lambda_\Delta^\pi$,
+dubbed \emph{Linear Core}, that combines the linearity-in-the-arrow
+and multiplicity polymorphism introduced by Linear
+Haskell~\cite{cite:linearhaskell}, featuring algebraic datatypes, case
+expressions, recursive let bindings, and multiplicity polymorphism.
+(i.e.,~all of Core's features with the exception of equality
+coercions).  We discuss multiplicity coercions as future work in
+Section~\ref{sec:future-work}.
+%
+
+Linear Core makes precise the various insights discussed in the
 previous chapter by crystallizing them together in a linear type system for which we
-prove soundness via the usual preservation and progress theorems. Crucially,
-the Linear Core type system accepts all the \emph{semantically linear} example
-programs (highlighted with \colorbox{notyet}{\notyetcolorname})
+prove soundness via the standard type preservation and progress theorems. Crucially,
+Linear Core typing accepts all the \emph{semantically linear} examples
+(highlighted with \colorbox{notyet}{\notyetcolorname})
 from Section~\ref{sec:semantic-linearity-examples}, which Core currently
@@
 %
-Besides type safety, we prove that multiple optimising Core-to-Core
-transformations preserve linearity in Linear Core. These same transformations
-don't preserve linearity under Core's current type system. As far as we know,
-we are the first to prove optimisations preserve types in a non-strict linear
-language.
+% Besides type safety, we prove that multiple optimising Core-to-Core
+% transformations preserve linearity in Linear Core. These same transformations
+% don't preserve linearity under Core's current type system. As far as we know,
+% we are the first to prove optimisations preserve types in a non-strict linear
+% language.
 
@@
 evaluated or returned, by annotating relevant binders with \emph{usage
-environments} (\S~\ref{sec:usage-environments}).
-%
-The second key idea is to have two distinct rules for
-case expressions, branching on whether the scrutinee is in Weak Head Normal
+environments} (Section~\ref{sec:usage-environments}).
+%
+The second key idea is to have two distinct rules for case
+expressions, branching on whether the scrutinee is in Weak Head Normal
 Form, and using ``proof irrelevance'' to track resources that are no
-longer available but haven't yet been fully consumed (\S~\ref{sec:lc-case-exps}). Additionally, we introduce
-tagged resources to split the resources between the pattern-bound variables as
-usage environments, encoding that pattern variables jointly ``finish
-consuming'' the scrutinee resources.
+longer available but haven't yet been fully consumed
+(Section~\ref{sec:lc-case-exps}). Additionally, we introduce tagged
+resources to split the resources between the pattern-bound variables
+as usage environments, encoding that pattern variables jointly
+``finish consuming'' the scrutinee resources.
 % 
@@
 
-We present Linear Core's syntax and type system iteratively, starting with the
-judgements and base linear calculi rules for multiplicity and term lambdas plus
-the variable rules ($\S~\ref{sec:base-calculi}$).
+We present Linear Core's syntax and type system incrementally, starting with the
+judgements and base linear calculi rules ($\S~\ref{sec:base-calculi}$).
 %
@@
 
-The complete syntax of Linear Core is given by Figure~\ref{fig:full-linear-core-syntax}.
+The complete syntax of Linear Core is given in Figure~\ref{fig:full-linear-core-syntax}.
 % Figure~\ref{fig:linear-core-types} and Figure~\ref{fig:linear-core-terms}.
@@
 %
-\[
-\SyntaxTypes
-\]
-%
-The terms are variables $x,y,z$, data constructors $K$, multiplicity
+
+%
+Terms are variables $x,y,z$, data constructors $K$, multiplicity
 abstractions $\Lambda p.~e$ and applications $e~\pi$, term abstractions
@@
 i.e. there is always at least one pattern which matches the scrutinized expression.
-\[
-\SyntaxTerms
-\]
-Linear Core takes the idea of annotating lets with usage environments from the
-unpublished Linear Mini-Core document by Spiwack et.
-al~\cite{cite:minicore}, which first tentatively tackled Core's
-linearity issues. We discuss this work in more detail in
-Section~\ref{sec:linear-mini-core}.
+
+% Linear Core takes the idea of annotating lets with usage environments from the
+% unpublished Linear Mini-Core document by Spiwack et.
+% al~\cite{cite:minicore}, which first tentatively tackled Core's
+% linearity issues. We discuss this work in more detail in
+% Section~\ref{sec:linear-mini-core}.
 
@@
 
-The (small-step) operational semantics of Linear Core are given by
+The (small-step) operational semantics of Linear Core are given in
 Figure~\ref{fig:linear-core-operational-semantics}. We use call-by-name
@@
 Specifically, function applications are reduced by the standard
-\emph{call-by-name} $\beta$-reduction, substituting the argument whole by
-occurrences of the lambda binder in its body, case expressions evaluate their
+\emph{call-by-name} $\beta$-reduction, substituting the argument by
+occurrences of the binder in its body, case expressions evaluate their
 scrutinee to WHNF, substituting the result by the case binder and constructor
@@
 
-\subsection{Typing Foundations\label{sec:base-calculi}}
+\subsection{Base Typing System\label{sec:base-calculi}}
 
@@
 %
-Otherwise, the base rules of the calculus for, multiplicity and term,
-abstraction and application are quite similar. In this section, we present
-the linear calculi's typing rules that share much in common with
-$\lambda^q_\to$, and, in the subsequent sections, the rules encoding the
-novel insights from Linear Core in typing semantic linearity, that were
-explored by example in Section~\ref{sec:linearity-semantically}.
-%
-We note, however, that we handle multiplicity polymorphism differently from
+
+The core rules of the calculus for, multiplicity and term,
+abstraction and application are similar to that of $\lambda^q_\to$ and
+mostly standard. 
+We note that we handle multiplicity polymorphism differently from
 Linear Haskell in ignoring the multiplicity semiring and, instead,
@@
 %
-The full type system is given by Figure~\ref{fig:linear-core-typing-rules},
-with auxiliary judgements given by
+The full type system is given in Figure~\ref{fig:linear-core-typing-rules},
+with auxiliary judgements given in
 Figure~\ref{fig:linear-core-other-judgements}.
@@
 
-We start with the main typing judgement. As is customary for linear type
+We start with the main typing judgement.
+As is common for linear type
 systems, we use two typing environments, an \emph{unrestricted} $\G$ and
 a \emph{linear} $\D$ environment.
-%
-Variables in $\G$ can be freely discarded (\emph{weakened}) and duplicated
-(\emph{contracted}), while resources in $\D$ must be used exactly once (hence
-can't be discarded nor duplicated). Despite not having explicit weakening and
-contraction rules in our system, they are available as admissible rules for
-$\G$ (but not for $\D$), since, equivalently (via
-\cite{91621fae-5e53-3497-8291-32b2fab5a743}), resources from $\G$ are
-duplicated for sub-derivations and may unrestrictedly exist in the variable
-rules.
+Thus, the main judgement is written $\G;\D \vdash e :
+\s$,
+denoting that expression $e$ has type $\s$ under
+the unrestricted environment $\G$ and linear environment $\D$.
+
+%
+Variables in $\G$ can be freely discarded (\emph{weakened}) and
+duplicated (\emph{contracted}), while resources in $\D$ must be used
+exactly once (hence can't be discarded nor duplicated). Despite not
+having explicit weakening and contraction rules in our system, they
+are available as admissible rules for $\G$ (but not for $\D$), since,
+equivalently \cite{91621fae-5e53-3497-8291-32b2fab5a743}, resources
+from $\G$ are duplicated for sub-derivations and may unrestrictedly
+exist in the variable rules.
 % TODO: Likely rewrite that
 %
-The main typing judgement, written $\G;\D \vdash e : \s \to \vp$,
-denotes that expression $e$ has type $\s \to \vp$ under
-the unrestricted environment $\G$ and linear environment $\D$.
 
@@
 environment is empty, while occurrences of linear variables are only well-typed
-when the variable being typed is the only resource available in the linear context.
-\[
-\begin{array}{cc}
-\TypeVarOmega & \TypeLinearVar
-\end{array}
-\]
+when the variable being typed is the only resource available in the
+linear context (rules $Var_\omega$ and $Var_1$, respectively).
+% \[
+% \begin{array}{cc}
+% \TypeVarOmega & \TypeLinearVar
+% \end{array}
+% \]
 In both cases, the linear context must contain exactly what is required to
@@
 Unrestricted functions are introduced via the function type ($\s \to_\pi \vp$) with $\pi =
-\omega$, and the $\lambda$-bound variable is introduced in $\G$:
-\[
-\begin{array}{cc}
-\TypeLamIntroL & \TypeLamIntroW
-\end{array}
-\]
-A linear function application is well-typed if there exists a disjoint split of the
-linear resources into $\D,\D'$ s.t. the function and argument, each under a
-distinct split, are both well-typed and the argument type matches the
-function's expected argument type. Conversely, unrestricted resources are
-duplicated and available whole to both sub-derivations.
-\[
-\TypeLamElimL \quad \TypeLamElimW
-\]
-An unrestricted function, unlike a linear one, consumes its argument
-unrestrictedly (zero or more times). Therefore, in an unrestricted function
-application, allowing any linear resources to occur in the argument expression
-entails consuming those resources unrestrictedly, since the
-variable binding the argument expression could be discarded or used more than
-once in the function body. Thus, argument expressions to unrestricted functions must also
-be unrestricted, i.e. no linear variables can be used to type them.
-%
+\omega$, and the $\lambda$-bound variable is introduced in $\G$ (rules
+$\lambda I_1$ and $\Lambda I$). Function application is standard, with
+linear function application splitting the linear context in two
+disjoint parts, one used to type the argument and the other for the
+function.
+Arguments of unrestricted functions must also be unrestricted, i.e. no
+linear variables can be used to type them. 
+% \[
+% \begin{array}{cc}
+% \TypeLamIntroL & \TypeLamIntroW
+% \end{array}
+% \]
+% A linear function application is well-typed if there exists a disjoint split of the
+% linear resources into $\D,\D'$ s.t. the function and argument, each under a
+% distinct split, are both well-typed and the argument type matches the
+% function's expected argument type. Conversely, unrestricted resources are
+% duplicated and available whole to both sub-derivations.
+% \[
+% \TypeLamElimL \quad \TypeLamElimW
+% \]
+% An unrestricted function, unlike a linear one, consumes its argument
+% unrestrictedly (zero or more times). Therefore, in an unrestricted function
+% application, allowing any linear resources to occur in the argument expression
+% entails consuming those resources unrestrictedly, since the
+% variable binding the argument expression could be discarded or used more than
+% once in the function body. Thus, argument expressions to unrestricted functions must also
+% be unrestricted, i.e. no linear variables can be used to type them.
+% %
+
 Typing linear and unrestricted function applications separately is less general
@@
 semiring) the multiplicities of the resources used to type the argument by $\pi$, however, our objective of typing semantic
-linearity does not benefit much from doing so, and by keeping the simple approach we can have
-the linear and unrestricted environments be separate.
+linearity does not benefit much from doing so and so we opt for a
+simpler design. 
 % \todo{We might only need multiplicities for the case-of-case definition as it
@@
 $p$ (in the unrestricted context), and construct expressions of type
-$\forall p.~\dots$, i.e. a type universally quantified over a multiplicity
+$\forall p.~\s$, i.e. a type universally quantified over a multiplicity
 variable $p$. We note that, in the body of the abstraction, function types annotated
@@
 in-scope  multiplicity variable in $\G$.
-\[
-\TypeWellFormedMult
-\]
-
-These rules conclude the foundations of our linear calculi. In subsequent
-sections we type (recursive) let bindings and case expressions,
-accounting for semantic linearity as per the insights construed in
-Section~\ref{sec:linearity-semantically}, effectively distilling them into the
-key ideas of our work -- encoded as rules.
+% \[
+% \TypeWellFormedMult
+% \]
+
+% These rules conclude the foundations of our linear calculi. In subsequent
+% sections we type (recursive) let bindings and case expressions,
+% accounting for semantic linearity as per the insights construed in
+% Section~\ref{sec:linearity-semantically}, effectively distilling them into the
+% key ideas of our work -- encoded as rules.
 
@@
 A \emph{usage environment} $\Delta$ is the means to encode the idea that lazy
-bindings don't consume the resources required by the bound expression when
+bindings do not consume the resources required by the bound expression when
 defined, but rather when the bindings themselves are fully consumed.
@@
 environment} to denote that consuming these variables equates to consuming the
-resources in the usage environment $\D$ they are annotated with, where a usage
-environment is essentially a multiset of linear resources. $\Delta$-bound
-variables are introduced by a handful of constructs, namely, (recursive) let
+resources in the annotated usage environment $\D$ (i.e.,~essentially a multiset of linear resources).
+%
+$\Delta$-bound
+variables are introduced by several of constructs: (recursive) let
 binders, case binders, and case pattern variables. In the following example, as
-per the insights into semantic linearity developed in
+per the insights developed in
 Section~\ref{sec:semantic-linearity-examples}, the resources required to
-typecheck the body of the binder $u$, $x$ and $y$, are only used if the let-var
+typecheck the body of the binder $u$, $x$ and $y$, are only used if
+the let-bound variable
 $u$ is consumed in the let-body $e$.  Accordingly, the usage environment of
@@
 Finally, we note that usage environments bear a strong resemblance to the linear
-typing environments to the right of the semicolon in the main typing judgement,
+typing environments in the main typing judgement,
 i.e. the environment with the linear resources required to type an expression.
@@
 former are used to annotate variables, while the latter used to type
-expressions. Yet, this distinction is slightly blurred after introducing how
-typing environments can be moved to usage environments, or otherwise occurs in
-rules relating the two.
+expressions.
+% Yet, this distinction is slightly blurred after introducing how
+% typing environments can be moved to usage environments, or otherwise occurs in
+% rules relating the two.
 
@@
 %\item
-(2) using $u$ is mutually exclusive with using the $\Delta$ resources it depends on elsehow;
+(2) using $u$ is mutually exclusive with using the $\Delta$ resources it depends on;
 % \item
-(3) $u$ can be safely discarded as long as the resources in $\Delta$ are consumed elsehow.
+(3) $u$ can be safely discarded as long as the resources in $\Delta$ are consumed.
 %\end{enumerate}
@@
 Fortunately, since linear resources must be linearly split across
-sub-derivations, (2) follows from (1) since consuming the linear resources
-in $\Delta$ to type $u$ makes them unavailable in the context of any other
-derivation. Therefore, expressions using these resources a second time, directly, or indirectly through the same
-(or other) usage environment, is ill-typed, as the resources are
-already allocated to the derivation of $u$. Similarly, (3) also follows from
-(1), because if the linear resources aren't consumed in the $\Delta$-var
-derivation, they must be consumed in an alternative derivation (or otherwise
-the expression is ill-typed).
-%
-These observations all boil down to one typing rule for $\Delta$-bound
-variables, which fundamentally encodes (1), implying the other two points:
+sub-derivations, (2) follows from (1) since consuming the linear
+resources in $\Delta$ to type $u$ makes them unavailable in the
+context of any other derivation. Therefore, expressions using these
+resources a second time, directly, or indirectly through the same (or
+other) usage environment, are ill-typed, as the resources are already
+allocated to the derivation of $u$. Similarly, (3) also follows from
+(1), because if the linear resources aren't consumed in the
+$\Delta$-var derivation, they must be consumed in an alternative
+derivation (or otherwise the expression is ill-typed).
+%
+These observations all lead to the typing rule for $\Delta$-bound
+variables:
+
+\vspace{-0.5cm}
 \[
@@
 that variable.
-% ROMES:TODO: Dizer alguma coisa tipo (blurring the distinction between usage envs and typing envs)?
-
-$\D$-variables are always introduced in $\Gamma$ since they can be discarded
-and duplicated, despite multiple occurrences of the same $\Delta$-variable not
-possibly being well-typed as, ultimately, it would imply non-linear usage of
-linear resources.
+%
+$\D$-variables are unrestricted 
+since they can be discarded and duplicated, despite multiple
+occurrences of the same $\Delta$-variable not being
+well-typed as, ultimately, it would imply non-linear usage of linear
+resources.
 % multiple occurrences of the same
@@
 
-\subsubsection{Lazy let bindings}
+\subsection{Lazy let bindings}
 
@@
 mutual exclusion with the linear resources required to type the expression it
-binds, and either \emph{must} be used, or we'd be discarding resources.
-
-Indeed, usage environments allow us to encode mutual exclusivity between
-alternative ways of consuming linear resources (between $\D$-vars and direct
-resource usage). Let-bound variables are the canonical example of a
-$\Delta$-bound variable, that is, let-variables bind expressions in which
+binds, and either \emph{must} be used.
+
+% Indeed, usage environments allow us to encode mutual exclusivity between
+% alternative ways of consuming linear resources (between $\D$-vars and direct
+% resource usage). 
+Let-bound variables are the canonical example of a
+$\Delta$-bound variable, that is, let-bound variables bind expressions in which
 the resources required to type them are consumed lazily rather than eagerly.
 %
-Effectively, annotating let-bound variables with a usage environment $\D$
+Annotating let-bound variables with a usage environment $\D$
 \emph{delays} the consumption of resources to when the variables themselves are
-used.
-
-Summarily, let-bindings introduce $\Delta$-variables whose usage environments
-are the linear typing environments of the bindings' bodies:
+used:
 \[
@@
 \]
-The rule for non-recursive let bindings splits the linear environment in $\D$
+The rule for (non-recursive) let bindings splits the linear environment in $\D$
 and $\D'$. $\D$ is used to type the body $e$ of the let binding $x$. Perhaps
@@
 doesn't consume resources, and the binding $x$ being $\D$-bound reflects that
-its usage entails consuming the resources $\D$ the expression $e$ it binds depends on.
+its usage entails consuming the resources $\D$ the expression it binds depends on.
 % \todo[inline]{Let bindings are hard, if they are used then we use resources. If
@@
 
-\subsubsection{Recursive let bindings}\label{sec:recursivelets}
-
-Recursive let bindings are very similar to non-recursive ones, the main
-exception being that the recursive bindings may be defined in terms of
-themselves, and we may have more than one binding. In our system, groups of
+\subsection{Recursive let bindings}\label{sec:recursivelets}
+
+% Recursive let bindings are very similar to non-recursive ones, the main
+% exception being that the recursive bindings may be defined in terms of
+% themselves, and we may have more than one binding.
+In our system, groups of
 recursive let bindings are always assumed to be strongly connected, that is,
@@
 that they all (transitively) depend on one another.
-
+%
 As before, recursive let bindings bind expressions \emph{lazily}, so they similarly
@@
 %
-However, as shown by example in Section
+However, as discussed in Section
 \ref{sec:semantic-linearity-examples:recursive-lets}, we must consider how
@@
 %, i.e. a least upper bound.
-%
+
 Extrapolating to a strongly-connected group of recursive bindings, (mutually)
@@
 The typing rule for recursive groups of bindings leverages our assumption that
-all recursive let bindings are strongly connected and exactly the observation
+all recursive let bindings are strongly connected and the observation
 that every binder in a strongly connected group of recursive bindings is typed
@@
 \]
-Unfortunately, this formulation is ill-suited for a syntax-directed system
-(from which an implementation is direct) because determining a particular $\D$
-to type and annotate all binders is difficult. We present our system and
+This formulation is not syntax-directed 
+% (from which an implementation is direct) 
+because determining a particular $\D$
+to type and annotate all binders is not obvious. We present our system and
 metatheory agnostically to the challenge of inferring this linear typing
@@
 %
-We propose a naive algorithm for inferring usage environments of recursive
-bindings in Section~\ref{sec:discuss:implementation} orthogonally to the theory
+We propose a naive $O(n^2)$ algorithm for inferring usage environments of recursive
+bindings in Section~\ref{sec:discuss:implementation}, orthogonally to the theory
 developed in this section.
 %
-The algorithm is a $O(n^2)$ traversal over the so-called \emph{naive usage
-environments} used to type each binding.
+% The algorithm is a $O(n^2)$ traversal over the so-called \emph{naive usage
+% environments} used to type each binding.
 %
@@
 traditionally achieved through the Hindley–Milner inference
-algorithm~\cite{DBLP:conf/popl/DamasM82}, there might be an opportunity to
-develop a better algorithm leveraging existing inference techniques.
-%
-Despite being a seemingly useful observation, we leave exploring a potential
-connection between inference of usage environments and type inference
-algorithms as future work.
+algorithm~\cite{DBLP:conf/popl/DamasM82}. % , there might be an opportunity to
+% develop a better algorithm leveraging existing inference techniques.
+% %
+% Despite being a seemingly useful observation, 
+We leave the exploration of this connection
+as future work.
 
@@
 %
-a case expression \emph{evaluates its scrutinee} to Weak Head Normal Form
-(WHNF)~\cite{10.5555/1096899}, \emph{then} selects the case alternative corresponding to the pattern
-matching the Weak Head Normal Form of the scrutinee\footnote{In our calculus,
-the alternatives are always exhaustive, i.e. there always exists at least one
-pattern which matches the scrutinee in its WHNF, so we're guaranteed to have an
-expression to progress evaluation.}. An expression in Weak Head Normal Form can
-either be %\begin{itemize}
+a case expression \emph{evaluates its scrutinee} to Weak Head Normal
+Form (WHNF)~\cite{10.5555/1096899}, \emph{then} selects the case
+alternative corresponding to the pattern matching the WHNF of the
+scrutinee\footnote{In our calculus, the alternatives are always
+  exhaustive, i.e. there always exists at least one pattern which
+  matches the scrutinee in its WHNF, so we're guaranteed to have an
+  expression to progress evaluation.}. An expression in WHNF can either be %\begin{itemize}
 % \item
-a lambda expression $\lambda x.~e$ 
+a $\lambda$-abstraction $\lambda x.~e$ 
 or a datatype constructor application $K~\ov{e}$.
@@
 In both cases, the sub-expressions $e$ or $\ov{e}$ occurring in the lambda body
-or as constructor arguments needn't be evaluated for the lambda or constructor
-application to be in weak head normal form (if otherwise all sub-expressions
-were fully evaluated the whole expression would also be in \emph{normal form}).
+or as constructor arguments need not be evaluated for the lambda or constructor
+application to be in WHNF. % (if otherwise all sub-expressions
+% were fully evaluated the whole expression would also be in \emph{normal form}).
 % (that is why it is called \emph{weak} head normal form).
@@
 in WHNF and terms that are not in WHNF.
-Following the discussion on expressions in weak head normal form, we present a
+Following the discussion on expressions in WHNF, we present a
 typing judgement $\G;\D \Vdash e : \s \gtrdot \ov{\D_i}$, and a rule for each
@@
 To type a constructor application $K~\ov{e_\omega e_i}$, where $e_\omega$
-are unrestricted arguments and $e_i$ the linear arguments of the
+are the unrestricted arguments and $e_i$ the linear arguments of the
 constructor, we split the resources $\D$ into a disjoint set of resources
@@
 
-Recall that the operational semantics encode the evaluation of case expressions as:
-%
-\begin{tabbing}
-$\ccase{e}{z~\{\ov{\rho_i \to e_i}\}} \longrightarrow^* \ccase{e'}{z~\{\ov{\rho_i \to e_i}\}}$\`where $e'$ is in WHNF and $e$ is not\\
-$\ccase{K~\ov{e}}{z~\{K~\ov{x} \to e_i\}} \longrightarrow e_i\ov{[e/x]}[K~\ov{e}/z]$\\
-$\ccase{e}{z~\{\_ \to e_i\}} \longrightarrow e_i[e/z]$\`where $e$ is in WHNF\\
-\end{tabbing}
-%
-When a scrutinee $K~\ov{e}$ matches a constructor pattern $K~\ov{\x[\pi]}$,
+Recall that the operational semantics encode the evaluation of case
+expressions such that 
+% %
+% \begin{tabbing}
+% $\ccase{e}{z~\{\ov{\rho_i \to e_i}\}} \longrightarrow^* \ccase{e'}{z~\{\ov{\rho_i \to e_i}\}}$\`where $e'$ is in WHNF and $e$ is not\\
+% $\ccase{K~\ov{e}}{z~\{K~\ov{x} \to e_i\}} \longrightarrow e_i\ov{[e/x]}[K~\ov{e}/z]$\\
+% $\ccase{e}{z~\{\_ \to e_i\}} \longrightarrow e_i[e/z]$\`where $e$ is in WHNF\\
+% \end{tabbing}
+% %
+when a scrutinee $K~\ov{e}$ matches a constructor pattern $K~\ov{\x[\pi]}$,
 evaluation proceeds in the case alternative corresponding to the matching
@@
 We highlight that, when evaluating a case expression, computation only
-effectively happens when a scrutinee not in WHNF is evaluated to WHNF. When the
-scrutinee is already in WHNF, evaluation continues in the alternative by
-substituting in the appropriate scrutinee expressions, but without having
-performed any computation.
+happens when a scrutinee (not in WHNF) is evaluated to
+WHNF. Otherwise, evaluation continues in the alternative by
+substituting in the appropriate scrutinee expressions, but without
+having performed any additional computation.
 % (the scrutinee was already in weak head normal form).
@@
 
-The dichotomy between evaluation (hence resource usage) of a case expression
+The dichotomy between evaluation (i.e.,~resource usage) of a case expression
 whose scrutinee is in weak head normal form, or otherwise, leads to one of our
 key insights: we must \emph{branch on weak head normal formed-ness} to
-type case expressions.
-%
-When the scrutinee is already in weak head normal form, the resources are
+accurately type case expressions.
+%
+When the scrutinee is already in WHNF, the resources are
 unused upon evaluation and thus available in the alternatives.
@@
 %
-To illustrate, consider a case expression with a scrutinee in weak head normal
-form and another whose scrutinee is not:
+To illustrate, consider the following case expressions:
 \[
@@
 
-Linear resources must be used exactly once, but there are three \emph{competing} ways
+There are three \emph{competing} ways
 to use the resources from a scrutinee in WHNF in a case alternative: directly, via
@@
 % We type a case expression whose scrutinee is in weak head normal form with
-This suggests the following rule:
+This suggests the following first attempt at a rule for cases of
+scrutinees in WHNF:
 \[
@@
 First, we assert this rule is only applicable to expressions in weak head
-normal form. Second, we use the typing judgement for expressions in WHNF
-previously introduced to determine the split of resources amongst the scrutinee
+normal form. Second, we appeal to the typing judgement for expressions in WHNF
+to determine the split of resources amongst the scrutinee
 sub-expressions. Finally, we type all case alternatives with the same context, using the
-$\vdash_{alt}$ judgement. 
-%
-Specifically:
-\begin{itemize}
-\item We introduce the case binder $z$ in the environment as a $\D$-bound
+$\vdash_{alt}$ judgement, introducing
+the case binder $z$ in the environment as a $\D$-bound
 variable whose usage environment is the linear resources used to type the
-scrutinee.
-\item We make all the resources $\ov{\D_i}$ used to type the scrutinee available in the linear typing environment.
-\item We annotate the \emph{alt} judgement with the disjoint set of linear resources $\ov{\D_i}$ used
-to typecheck the scrutinee sub-expressions.
-\item We annotate the judgement with the name of the case binder $z$ and use
-the $\Mapsto$ arrow in the judgement -- this is of most importance when typing
-the alternative itself, and will be motivated together with the alternative
-judgement below.
-\end{itemize}
+scrutinee. We also 
+make all the resources $\ov{\D_i}$ used to type the scrutinee
+available in the linear typing environment.
+
+The $\vdash_{alt}$ judgement is 
+annotated with the disjoint set of linear resources $\ov{\D_i}$ used
+to typecheck the scrutinee sub-expressions,
+the name of the case binder $z$.
+
 %
@@
 The alternative judgement $\G;\D \vdash_{alt} \rho \to e :^z_\D \s \Rightarrow
-\vp$ is used to type case alternatives, but it encompasses three
-``sub-judgements``, distinguished by the arrow that is used:
+\vp$ is used to type case alternatives, encompassing three
+``modes``, distinguished by the kind of arrow that is used:
 for alternatives of case expressions whose scrutinee is in WHNF ($\Mapsto$),
@@
 %
-Following the $Case_\textrm{WHNF}$ rule in which we use the $\Mapsto$ alternative
-judgement, the rule for typing a case alternative whose pattern is a
-constructor with $n > 0$ linear components is:
+% Following the $Case_\textrm{WHNF}$ rule in which we use the $\Mapsto$ alternative
+% judgement, t
+The rule for typing an alternative whose pattern is a
+constructor with $n > 0$ linear components (in mode $\Mapsto$) is:
 \[
@@
 whose usage environment matches the linear resources required to type the
-corresponding constructor argument in the scrutinee, which comes annotated in
+corresponding constructor argument in the scrutinee, which is annotated in
 the judgement ($\ov{\D_i}$). Unrestricted fields of the constructor are
@@
 
-Secondly, the rule for alternatives that match on the wildcard pattern:
+The rule for alternatives that match on the wildcard pattern is:
 \[
@@
 To type a wildcard alternative we simply type the expression with the main
-judgement, ignoring all annotations on the judgement; recalling that the case
+judgement, ignoring all annotations on the judgement. We recall that the case
 binder was already introduced in the environment with the appropriate usage
-environment by the case expression, rather than in the case alternative rule.
+environment by the case expression.
 
@@
 constructors without linear components.
-
-% Indeed, matching on a constructor without
-% linear arguments entails that all the scrutinee resources have been fully
-% consumed, and thus are no longer available.
-
 Taking into account that case expressions introduce the linear resources of the
@@
 The $Alt0$ rule essentially encodes this insight, and is applicable regardless
-of the WHNF-ness of the scrutinee (hence the $\Rightarrow$ arrow), as long as
+of the WHNF-ness of the scrutinee ($\Rightarrow$ mode), as long as
 the constructor pattern has no linear fields:
@@
 \]
-The rule deletes the annotated scrutinee environment $\D_s$ from two select environments:
-\begin{itemize}
-
-\item The linear typing environment, effectively deleting the resources from
-the scrutinee made available here by the case expressions (written
-$\D[\cdot/\D_s]$, a substitution of the scrutinee typing environment by the
-empty linear environment $\cdot$).
-
-\item The usage environment of the case binder $z$, written $\G[\cdot/\D_s]_z$
+The rule deletes the annotated scrutinee environment $\D_s$ from
+%two select environments:
+the linear typing context (written $\D[\cdot/\D_s]$, a substitution of
+the scrutinee typing environment by the empty linear environment $\cdot$)
+and from the usage environment of the case binder $z$, written $\G[\cdot/\D_s]_z$
 to denote replacing the usage environment of the variable $z$ in $\G$, which
@@
 
-\end{itemize}
+% \begin{itemize}
+
+% \item The linear typing environment, effectively deleting the resources from
+% the scrutinee made available by the case expressions (written
+% $\D[\cdot/\D_s]$, a substitution of the scrutinee typing environment by the
+% empty linear environment $\cdot$).
+
+% \item The usage environment of the case binder $z$, written $\G[\cdot/\D_s]_z$
+% to denote replacing the usage environment of the variable $z$ in $\G$, which
+% is necessarily $\D_s$ (since we always annotate the judgement with the
+% environment of the scrutinee), by the empty environment.
+
+% \end{itemize}
 %
@@
 pattern is unrestricted when evaluated to WHNF, implying that all linear
-resources have been consumed to produce it, and the result is something that
-can be freely discarded or duplicated.
-%
-It ensures that when we match on an unrestricted pattern we no longer need to
+resources have been consumed to produce it, and the result may 
+be freely discarded or duplicated.
+%
+This ensures that when we match on an unrestricted pattern we no longer need to
 consume the scrutinee resources. Otherwise, for example,
@@
 %
-However, let us consider two discrete situations -- pattern matches in a case
+However, let us consider two situations -- pattern matches in a case
 expression whose scrutinee is in WHNF, and matches on a case expression whose
@@
 %
-\begin{enumerate}
-\item When the scrutinee is in WHNF, it is either an unrestricted expression
+%\begin{enumerate}
+When the scrutinee is in WHNF, it is either an unrestricted expression
 against which any match will only introduce unrestricted variables, or an
-expression that depends on linear resources. The first case trivially allows
-any resource from the scrutinee in the alternatives as well. The second is
+expression that depends on linear resources. The former trivially allows
+any resource from the scrutinee in the alternatives as well. The latter is
 further divided: 
-\begin{enumerate}
-
-\item The pattern is unrestricted while the
+%\begin{enumerate}
+%
+%\item
+The pattern is unrestricted while the
 scrutinee is not, so entering this branch is impossible as long as the case
 expression is well-typed; by contradiction, the linear resources from the
-scrutinee could occur unrestrictedly in that branch, since from falsity
-anything follows (\emph{ex falso quodlibet}).
-%
-For uniformity, we type such alternatives as those for scrutinees that are not in WHNF.
-
-\item The pattern is linear and matches the scrutinee, in which case the
-$AltN_{\textrm{WHNF}}$ is applicable instead of $Alt0$.
-
-\item The pattern is linear but does not match the scrutinee and so, the same
-reasoning as (1) above applies: any resource could theoretically be used in
-such alternatives, however, for uniformity, it is also typed as though the
-scrutinee were not in WHNF.
-
-\end{enumerate}
-
-\item However, if the scrutinee is not in WHNF, the resources occurring in the
+scrutinee could occur unrestrictedly in that branch.
+%
+For uniformity, we type such alternatives as those for scrutinees that
+are not in WHNF;
+%
+The pattern is linear and matches the scrutinee, in which case the
+$AltN_{\textrm{WHNF}}$ is applicable instead of $Alt0$;
+%
+The pattern is linear but does not match the scrutinee and so any
+resource could be used in such alternatives. For uniformity,
+it is also typed as though the scrutinee were not in WHNF.
+
+If the scrutinee is not in WHNF, the resources occurring in the
 scrutinee will be consumed when evaluation occurs. Therefore, the resources
 used in the scrutinee cannot occur in the alternative body
-(e.g. $x$ cannot occur in the alternative in $\ccase{close~x}{\{K_1 \to x\}}$)
+(e.g. $x$ cannot occur in $e$ in $\ccase{close~x}{\{K_1 \to e\}}$)
 -- regardless of the pattern.
@@
 %
-We guarantee resources from a scrutinee that is not in weak head normal form
-cannot occur/directly be used in any case alternative, in our rule for typing
+We guarantee resources from a scrutinee that is not in WHNF
+cannot occur/directly be used in any case alternative in our rule for typing
 cases not in WHNF, which we introduce below.
 
-\end{enumerate}
+%\end{enumerate}
 

### 2024-01-29 icfp.tex
passed up to section 3.6.2
worked up to section 4

@@
 
-\subsubsection{Proof irrelevant resources}
+\todo[inline]{Considerar alguns exemplos aqui para ilustrar esta
+  ``análise de casos''}
+
+\subsubsection{Pattern matching on non-WHNF terms via proof irrelevant resources}
 
@@
 
-Resources used in a scrutinee that is not in weak head normal form must
-definitely not be used in the case alternatives
-% since they have been used in the evaluation of the scrutinee, as shown in
-% the example above.
+\todo[inline]{Todo este paragrafo parece-me redundante}
+Resources used in a scrutinee that is not in WHNF must
+not be used in the case alternatives.
+%
 %
@@
 \emph{fully} consume all resources occurring in the scrutinee, the scrutinee
-must be evaluated either to \emph{normal form} or s.t. all linear components of
-the scrutinee are fully evaluated, as witnessed by the $Alt0$ rule. In short,
-for a case expression whose scrutinee is not in WHNF:
-% We tackle this in due time, in the proof irrelevance section.
-\begin{itemize}
-
-\item The scrutinee resources must \emph{not} be used directly in the case alternatives;
-\item But the result of evaluating the scrutinee to WHNF must still be
+must be evaluated either to \emph{normal form} or such that all linear components of
+the scrutinee are fully evaluated, as witnessed by the $Alt0$ rule above. In short,
+for a case expression whose scrutinee is not in WHNF, the scrutinee
+resources must \emph{not} be used directly in the case alternatives,
+but the result of evaluating the scrutinee to WHNF must still be
 consumed, as all sub-expressions of the scrutinee remain unevaluated and must
-be consumed.
-\item Since the scrutinee resources cannot be consumed directly, they must be
+be consumed. Since the scrutinee resources cannot be consumed directly, they must be
 consumed indirectly through $\D$-variables, namely, either the case binder, or
 the linear pattern-bound variables introduced in the alternative.
-
-\end{itemize}
+% We tackle this in due time, in the proof irrelevance section.
+% \begin{itemize}
+
+% \item The scrutinee resources must \emph{not} be used directly in the case alternatives;
+% \item But the result of evaluating the scrutinee to WHNF must still be
+% consumed, as all sub-expressions of the scrutinee remain unevaluated and must
+% be consumed.
+% \item Since the scrutinee resources cannot be consumed directly, they must be
+% consumed indirectly through $\D$-variables, namely, either the case binder, or
+% the linear pattern-bound variables introduced in the alternative.
+
+% \end{itemize}
 
@@
 
-We introduce \emph{proof irrelevant} resources, denoted as linear resources
-within square brackets $[\D]$, to encode linear resources that cannot be
-directly used (the $Var$ rule is not applicable). Proof irrelevant resources
+We introduce \emph{proof irrelevant} resources, written as linear resources
+within square brackets $[ x{:}\s]$ and lifted to contexts as $[\D]$,
+to encode linear resources that cannot be
+directly used (i.e.,~to which the $Var$ rule is not applicable). Proof irrelevant resources
 are linear resources in all other senses, meaning they must be used
 \emph{exactly once}. However, since proof irrelevant resources cannot be
-forgotten neither used directly, they have to be consumed \emph{indirectly} --
+discarded nor used directly, they have to be consumed \emph{indirectly} --
 by $\D$-bound variables.
@@
 \]
-Note how the rule is quite similar to the one for scrutinees in WHNF, only
-diverging in that the resources in the case binder, typing environment, and
-judgement annotation, are made irrelevant.
+Note how the rule is similar to the one for scrutinees in WHNF, but
+requiring the resources in the case binder, typing environment, and
+judgement annotation to be made irrelevant.
 
@@
 only makes sense for the constructor pattern matching the scrutinee.
-Essentially, if the scrutinee is in WHNF the matching alternative is easily
-determined, and must be treated with the specialized $\Mapsto$ judgement, which
-only applies to the matching constructor.
-%
+% Essentially, if the scrutinee is in WHNF the matching alternative is easily
+% determined, and must be treated with the specialized $\Mapsto$ judgement, which
+% only applies to the matching constructor.
+% %
 Alternatives not matching the scrutinee, as mentioned in the discussion of the
@@
 however, we uniformly treat non-matching alternatives as if the scrutinee were
-not in WHNF. The rule for typing case alternatives whose scrutinee is in WHNF
-is thus given by:
+not in WHNF. Having introduced proof irrelevant resources, we can now
+present the updated and fully general $Case_\textrm{WHNF}$ rule:
 \[
@@
 We note that it might seem unusual to specialize a rule for expressions in
-WHNF, as programs scrutinizing an expression in WHNF are rarely written by a
-developer. Yet, our system is designed to be suitable for optimising compilers
+WHNF, as programs scrutinizing an expression in WHNF are rarely
+written by humans.
+Yet, our system is designed to be suitable for optimising compilers
 in which intermediate programs commonly scrutinize expressions in WHNF.
 %
-Foreshadowing, type preservation for the case-reduction substituting the case
-binder and pattern variables in the alternative is not possible to prove
-without branching on WHNF-ness, since otherwise the $\D$-substitution is not
-well-defined.
+Crucially, type preservation for the case-reduction substituting the
+case binder and pattern variables in the alternative is not possible
+to prove without branching on WHNF-ness, since otherwise the
+$\D$-substitution is not well-defined.  \todo[inline]{What is the
+  $\Delta$-substitution? Clarify}
 
@@
 
-Intuitively, in case alternatives whose scrutinee is not in weak head normal form,
-% (and for scrutinees in WHNF which don't match the case alternative)
-the proof-irrelevant resources introduced by the case expression must be fully
-consumed, either via the case binder $z$, or by using all linear pattern-bound
-variables (for uniformity, we also treat alternatives that do not match a
-scrutinee in WHNF this way).
-
-However, unlike with scrutinees in WHNF, the resources used by a
+%% NOTA -- Sempre a repetir a mesma coisa
+% Intuitively, in case alternatives whose scrutinee is not in WHNF,
+% % (and for scrutinees in WHNF which don't match the case alternative)
+% the proof-irrelevant resources introduced by the case expression must be fully
+% consumed, either via the case binder $z$, or by using all linear pattern-bound
+% variables (for uniformity, we also treat alternatives that do not match a
+% scrutinee in WHNF this way).
+
+Unlike with scrutinees in WHNF, the resources used by a
 scrutinee not in WHNF do not necessarily match those used by each
@@
 % section too?}
-
+%
 We introduce \emph{tagged resources} to guarantee all linearly-bound pattern
@@
 environment $\D_i$ to each linear pattern variable where, $\D_i$ is obtained from the
-scrutinee environment tagged with the constructor name and linear-variable
+scrutinee environment \emph{tagged} with the constructor name and linear-variable
 index $\lctag{\D_s}{K_i}$, and $\y[\D_i]$ is introduced in $\G$.
@@
 %
-This is given by the Split rule. By assigning to each linear pattern variable a fragment of the scrutinee
-resources with a tag, we guarantee that all linear pattern variables are
-simultaneously used to consume all the scrutinee resources, since for any of
-scrutinee resources to be used by a linear pattern-bound var be used, the
-resources must be $Split$ for the fragments corresponding to that $\D$-var to
-be consumed, and, consequently, the remaining fragments have to be consumed
-through the other linear pattern-bound variables.
-%
-For instance, in $\lambda x.~\ccase{x}{z~\{K~a~b\to (a,b)\}}$, where $x$ is a
-linear variable, the case alternatives are typed with $\irr{x}$ (proof
-irrelevant $x$), $z$ is introduced as $\z[\irr{x}]$, and the pattern variables
+This is given by the Split rule. By assigning to each linear pattern
+variable a fragment of the scrutinee resources with a tag, we
+guarantee that all linear pattern variables are simultaneously used to
+consume all the scrutinee resources.
+For instance, in the term $\lambda x.~\ccase{x}{z~\{K~a~b\to
+  (a,b)\}}$, where $x$ is a linear variable, the case alternative is
+typed with $\irr{x}$,
+the case binder $z$ is introduced as $\z[\irr{x}]$, and the pattern variables
 are introduced as $\var[a][\lctag{\irr{x}}{K_1}]$ and
 $\var[b][\lctag{\irr{x}}{K_2}]$, assuming both components of $K$ are linear.
+\todo[inline]{Previous explanation was very confusing, try to explain
+  what the split is doing in terms of the example!}
+% since for any of the scrutinee
+% resources to be used by a linear pattern-bound variable usage, the
+% resources must be $Split$ for the fragments corresponding to that
+% $\D$-var to be consumed, and, consequently, the remaining fragments
+% have to be consumed through the other linear pattern-bound variables.
+
 %

### 2024-01-29 icfp.tex
passed through seciton 5

@@
 
-\todo[inline]{Todo este paragrafo parece-me redundante}
-Resources used in a scrutinee that is not in WHNF must
-not be used in the case alternatives.
-%
-%
-However, it is not sufficient to evaluate the scrutinee to weak head normal
-form to \emph{fully} consume all resources used in the scrutinee, since
-sub-expressions such as constructor arguments will be left unevaluated. To
-\emph{fully} consume all resources occurring in the scrutinee, the scrutinee
-must be evaluated either to \emph{normal form} or such that all linear components of
-the scrutinee are fully evaluated, as witnessed by the $Alt0$ rule above. In short,
-for a case expression whose scrutinee is not in WHNF, the scrutinee
-resources must \emph{not} be used directly in the case alternatives,
-but the result of evaluating the scrutinee to WHNF must still be
-consumed, as all sub-expressions of the scrutinee remain unevaluated and must
-be consumed. Since the scrutinee resources cannot be consumed directly, they must be
-consumed indirectly through $\D$-variables, namely, either the case binder, or
-the linear pattern-bound variables introduced in the alternative.
+\todo[inline]{Todo este paragrafo que esta commented out parece-me redundante}
+% Resources used in a scrutinee that is not in WHNF must
+% not be used in the case alternatives.
+% %
+% %
+% However, it is not sufficient to evaluate the scrutinee to weak head normal
+% form to \emph{fully} consume all resources used in the scrutinee, since
+% sub-expressions such as constructor arguments will be left unevaluated. To
+% \emph{fully} consume all resources occurring in the scrutinee, the scrutinee
+% must be evaluated either to \emph{normal form} or such that all linear components of
+% the scrutinee are fully evaluated, as witnessed by the $Alt0$ rule above. In short,
+% for a case expression whose scrutinee is not in WHNF, the scrutinee
+% resources must \emph{not} be used directly in the case alternatives,
+% but the result of evaluating the scrutinee to WHNF must still be
+% consumed, as all sub-expressions of the scrutinee remain unevaluated and must
+% be consumed. Since the scrutinee resources cannot be consumed directly, they must be
+% consumed indirectly through $\D$-variables, namely, either the case binder, or
+% the linear pattern-bound variables introduced in the alternative.
+%ATE AQUI
 % We tackle this in due time, in the proof irrelevance section.
@@
 not in WHNF. Having introduced proof irrelevant resources, we can now
-present the updated and fully general $Case_\textrm{WHNF}$ rule:
+present the correct $Case_\textrm{WHNF}$ rule:
 \[
@@
 suitable for the intermediate language of an optimising compiler.
-
+\todo[inline]{Tem de ser anonimizado, provavelmente alterando a bib
+  entry para omitir autores e o proprio link}
 The GHC Linear Core plugin~\cite{cite:linear-core-plugin} implements a
@@
 %
-Most notably, the plugin successfully validates linearity throughout the
+The plugin successfully validates linearity throughout the
 (optimising) compilation of linearity-heavy libraries, namely
@@
 linearity depends on so-called \emph{multiplicity coercions}, which are an
-avenue of future work ($\S$~\ref{sec:future-work}) exploring the intersection
+avenue of future work (Section~\ref{sec:future-work}) exploring the intersection
 of Linear Core with type equality coercions.
@@
 
-The implementation of Linear Core as a typechecker does not follow directly
-from the description of the type system because Linear Core is not
-\emph{syntax-directed}. Specifically, the most challenging features are
-splitting linear resources amongst sub-derivations and consuming fragments of
-resources through pattern-bound variables.
-%
-We use the following techniques to tackle the \emph{non-syntax-directedness} of
-Linear Core, thus making the system more suitable to implement:
-%
-\begin{itemize}
-
-\item Instead of non-deterministically splitting linear resources amongst
+The implementation of Linear Core as a typechecker does not follow
+directly from the description of the type system because Linear Core
+is not \emph{syntax-directed}. Specifically, the most challenging
+features are splitting linear resources amongst sub-derivations and
+consuming fragments of resources through pattern-bound variables.
+Instead of non-deterministically splitting linear resources amongst
 sub-derivations, we thread input/output linear resources through each
-derivation using the resource management for linear logic described
-by~\cite{DBLP:journals/tcs/CervesatoHP00}.
-% \todo{Escrever uma regra?}
-
-% In a case expression whose scrutinee is not in WHNF,
-\item Pattern variables bound in a case alternative, for a scrutinee not in WHNF,
-are introduced as $\D$-variables with usage environment $\lctag{\irr{\D}}{K_i}$,
-where $\D$ are the scrutinee resources and $K_i$ the tag of that pattern
-variable. To use the resources through the pattern-bound $\D$-vars, they must
-be first $Split$ into fragments.
-
-We consume tagged fragments of a resource \emph{as needed}, i.e., when a resource,
-whose usage environment has a fragmented resource with tag $K_j$, is used, we $Split$
-the matching resource according to the constructor $K$ and consume the fragment
-$K_j$, rather than eagerly determining which resources need to be fragmented to
-do so.
-%
-We note that it is safe to destructively fragment the resource, i.e. removing
-the whole resource and only leaving the split fragments, because resources are
-only $Split$ when a fragment is needed and, consequently, if a fragment is
-consumed, using the ``whole'' resource as well violates linearity.
-
-\end{itemize}
-%
-Furthermore, our implementation must infer the usage environments of binders in
+derivation~\cite{DBLP:journals/tcs/CervesatoHP00}.  Pattern variables
+bound in a case alternative for a scrutinee not in WHNF, are
+introduced as $\D$-variables with usage environment
+$\lctag{\irr{\D}}{K_i}$, where $\D$ are the scrutinee resources and
+$K_i$ the tag of that pattern variable. To use the resources through
+the pattern-bound $\D$-vars, they must be first $Split$ into
+fragments. We consume tagged fragments of a resource \emph{as needed},
+i.e., when a resource, whose usage environment has a fragmented
+resource with tag $K_j$, is used, we $Split$ the matching resource
+according to the constructor $K$ and consume the fragment $K_j$,
+rather than eagerly determining which resources need to be fragmented
+to do so.
+Finally, our implementation must infer the usage environments of binders in
 a recursive let group before using them to typecheck the let body. We use a
 naive $O(n^2)$ algorithm (where $n$ is the number of let bindings) to determine
-these usage environments, but discuss an avenue of further research regarding
-this inference challenge in Section~\ref{sec:future-work}.
+these usage environments. Further research regarding
+this inference challenge is discussed in Section~\ref{sec:future-work}.
+%
+%
+%
+% We use the following techniques to tackle the \emph{non-syntax-directedness} of
+% Linear Core:
+%
+% \begin{itemize}
+
+% % \item Instead of non-deterministically splitting linear resources amongst
+% % sub-derivations, we thread input/output linear resources through each
+% % derivation using the resource management for linear logic described
+% % by~\cite{DBLP:journals/tcs/CervesatoHP00}.
+% % \todo{Escrever uma regra?}
+
+% % In a case expression whose scrutinee is not in WHNF,
+% \item Pattern variables bound in a case alternative, for a scrutinee not in WHNF,
+% are introduced as $\D$-variables with usage environment $\lctag{\irr{\D}}{K_i}$,
+% where $\D$ are the scrutinee resources and $K_i$ the tag of that pattern
+% variable. To use the resources through the pattern-bound $\D$-vars, they must
+% be first $Split$ into fragments.
+
+% We consume tagged fragments of a resource \emph{as needed}, i.e., when a resource,
+% whose usage environment has a fragmented resource with tag $K_j$, is used, we $Split$
+% the matching resource according to the constructor $K$ and consume the fragment
+% $K_j$, rather than eagerly determining which resources need to be fragmented to
+% do so.
+% %
+% We note that it is safe to destructively fragment the resource, i.e. removing
+% the whole resource and only leaving the split fragments, because resources are
+% only $Split$ when a fragment is needed and, consequently, if a fragment is
+% consumed, using the ``whole'' resource as well violates linearity.
+
+% \end{itemize}
+%
 
@@
 
-The results of running the Linear Core GHC plugin on large established
+The results of running the Linear Core GHC plugin on large 
 libraries focused around linear types are given by
@@
 %
-We compiled the libraries \texttt{linear-smc}, a library presented in the
-work~\cite{10.1145/3471874.3472980} (1500 lines), \texttt{linear-base} (4000
+We compiled the libraries \texttt{linear-smc}~\cite{10.1145/3471874.3472980} (1500 lines), \texttt{linear-base} (4000
 lines), the Haskell standard library for programming with linear types,
@@
 %
-We count the number of programs accepted by our implementation, where each
-top-level binding in a module counts as a program, and every such binding is
-typechecked once per optimisation pass. i.e., we typecheck all intermediate
-programs produced by GHC. The total amount of programs rejected by the
-typechecker are given in the ``Total Rejected'' column, but we distil these
-rejections into ``Unique Rejections'' by removing duplicate rejections (those
-of the same program and for the same linearity-violating reason, but occurring
-at different stages). We further categorize the unique rejections into ``Linear
-modulo Call-by-name'', ``Linear Rejected'', ``Not Linear Rejected'', and
-``Unknown Rejected''.
-%
-Using the plugin, \texttt{linear-base} takes 35 seconds to compile, instead of
-20 seconds. We note, however, that the implementation is not performance
-conscious, and types every intermediate program from scratch, instead of
-maintaining linearity information throughout the pipeline.
+We count the number of programs accepted by our implementation, where
+each top-level binding in a module counts as a program, and every such
+binding is typechecked once per optimisation pass. i.e., we typecheck
+all intermediate programs produced by GHC. The total amount of
+programs rejected by the typechecker are given in the ``Total
+Rejected'' column, but we distil these rejections into ``Unique
+Rejections'' by removing duplicate rejections (those of the same
+program and for the same linearity-violating reason, but occurring at
+different stages). We further categorize the unique rejections into
+``Linear modulo Call-by-name'', ``Linear Rejected'', ``Not Linear
+Rejected'', and ``Unknown Rejected''.  We note that our implementation
+is not performance conscious. For instance, \texttt{linear-base} takes
+35 seconds to compile with our plugin, instead of 20 seconds.
+%
+% Using the plugin, \texttt{linear-base} takes 35 seconds to compile, instead of
+% 20 seconds. We note, however, that the implementation is not performance
+% conscious, and types every intermediate program from scratch, instead of
+% maintaining linearity information throughout the pipeline.
 
 Programs ``linear modulo call-by-name'' are a class of programs
-which scrutinize a variable, but then uses the variable in the case
-alternatives of the case scrutinizing it. As will be made clear in
+which scrutinize a variable, but then use the variable in the case
+alternatives of the case scrutinizing it. As discussed in
 Section~\ref{sec:reverse-binder-swap-considered-harmful}, these programs can be
 understood as linear as long as applications of linear functions binding these
-variables are \emph{not} reduced \emph{call-by-name}, because, in doing so,
-linear resources are duplicated. In Core, these programs are not seen as
+variables are \emph{not} reduced \emph{call-by-name}.
+In Core, these programs are not seen as
 linear, and thus GHC does not not reduce them using a call-by-name evaluation strategy.
@@
 semantically, and are simultaneously rejected by Linear Core and its
-implementation. These programs were not necessarily identified as bad
-by the linter.\footnote{rever}
+implementation. % These programs were not necessarily identified as bad
+% by the linter.\footnote{rever}
 % An example, where the last component of \incode{HashMap}, \incode{wwz},
@@
 
+{\small
   \input{prototype/core-plugin-results}
+  }
 

### 2024-01-30 icfp.tex
up to 5.2, now at 25 pages

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Metatheory\label{sec:main:metatheory}}
-
-The $\lambda^\pi_\D$ system is sound: well-typed programs in Linear Core do not
-get \emph{stuck}. Besides type safety ($\S$~\ref{sec:type-safety-meta}), we
-prove multiple optimising transformations preserve linearity
-($\S$~\ref{sec:optimisations-preserve-types-meta}), and prove an auxiliary
-result regarding proof irrelevant resources, stating that a case alternative
-well-typed in a proof irrelevant context is also well-typed if proof irrelevant
-resources are substituted by an arbitrary environment of relevant resources.
-%
-Additionally, we state our assumptions that outline an isomorphism between
-using a linear variable $\xl$ and a $\D$-variable $\xD$ that consumes existing
-resources $\D$, for any $\D$.
-
-\subsection{Assumptions}
+\section{Metatheory of Linear Core\label{sec:main:metatheory}}
+
+The $\lambda^\pi_\D$ type system is sound: well-typed programs in Linear Core do not
+get \emph{stuck}. Besides type safety, we
+prove that multiple optimising transformations used in GHC preserve
+linearity in the sense of being well-typed in our system.
+%
+% Additionally, we state our assumptions that outline an isomorphism between
+% using a linear variable $\xl$ and a $\D$-variable $\xD$ that consumes existing
+% resources $\D$, for any $\D$.
 
 \renewcommand{\DeltaLinearRelationLemma}{
-\begin{lemma}[$\D \Rightarrow 1$]
-Any $\Delta$-variable with a fully-irrelevant usage environment can be made a linear variable as long as its irrelevant environment is dropped from the linear environment (and the unrestricted env.
+\begin{lemma}[$\Delta$-bound to Linear]
+% Any $\Delta$-variable with a fully-irrelevant usage environment can be made a linear variable as long as its irrelevant environment is dropped from the linear environment (and the unrestricted env.
 is updated accordingly).\\
@@
 \renewcommand{\LinearDeltaRelationLemma}{
-\begin{lemma}[$1 \Rightarrow \D$]
-A linear variable can be moved to the unrestricted context as a $\D$-var with any usage environment $\D$ by introducing $\D$ in the linear resources (and the unrestricted env.
-is updated accordingly).\\
+  \begin{lemma}[Linear to $\Delta$-bound]
+\\
 If $\G; \D',\xl \vdash e :\vp$
@@
 \renewcommand{\DeltaUnrestrictedRelationLemma}{
-\begin{lemma}[$x{:}_\omega\sigma = x{:}_{\cdot}\sigma$]
-An unrestricted variable is equivalent to a $\D$-var with an empty usage environment.\\
+\begin{lemma}[Unrestricted and $\Delta$-bound]
+%An unrestricted variable is equivalent to a $\D$-var with an empty usage environment.\\
 $\G,\xo; \D \vdash e : \vp$ iff $\G,\x[\cdot]; \D \vdash e : \vp$
@@
 
-We use two main assumptions in our proofs, which are dual.
-%
-First, a program well-typed with a linear variable ($\xl$) is equivalently
-well-typed if that same linear variable were instead $\D$-bound ($\xD$) with
-usage environment $\D$, $\D$ were available in the linear context instead of
-the linear variable.
-% , and occurrences of $x$ were substituted by $\D$ in the
-% usage environments of other $\D$-vars in $\G$.
-%
-% First, we state that a linear variable $\xl$ can be replaced by a $\D$-bound
-% variable as long as the context
-%
+\paragraph{$\Delta$-bound variables} Our development relies on the following characterization of $\D$-bound
+variables:
+A well-typed program with a linear variable ($\xl$) is equivalently
+well-typed if the linear variable were instead $\D$-bound ($\xD$) with
+usage environment $\D$, with $\D$ available in the linear context instead of
+the linear variable (and the unrestricted context updated accordingly).
 \LinearDeltaRelationLemma
-%
-\noindent Second, a program well-typed with resources $\D$ and $\D$-bound
-variable ($\xD$) is equivalently well-typed, as long as $\D$ is consumed
-through the use of $x$, if $x$ is moved to the linear context, resources $\D$
-are removed from the linear context, and occurrences of $\D$ whole in usage
-environments are substituted by $x$ (occurrences of fragments of $\D$ in usage
-environments are unimportant since $\D$ was consumed whole by $x$, not by any
-of the fragment-using $\D$-vars).
+
+\noindent Dually, any $\Delta$-bound variable with an irrelevant usage
+environment can be made into a linear variable as long as its
+(irrelevant) environment is dropped from the linear context.
+
+% Dually, a well-typed program with resources $\D$ and $\D$-bound
+% variable ($\xD$) is equivalently well-typed, as long as $\D$ is consumed
+% through the use of $x$, if $x$ is moved to the linear context, resources $\D$
+% are removed from the linear context, and occurrences of $\D$ whole in usage
+% environments are substituted by $x$ (occurrences of fragments of $\D$ in usage
+% environments are unimportant since $\D$ was consumed whole by $x$, not by any
+% of the fragment-using $\D$-vars).
 
@@
 
-We additionally state that unrestricted resources are equivalent to $\D$-bound
-variables with an empty ($\cdot$) usage environment:
+\noindent Finally, unrestricted resources are equivalent to $\D$-bound
+variables with an empty ($\cdot$) usage environment.
 
@@
 
-\subsection{Irrelevance}
-
-As discussed above, proof irrelevant resources are resources that can only be
-consumed indirectly, and are used to type case expressions whose scrutinee is
-not in WHNF, essentially encoding that the scrutinee resources must be consumed
-through the case binder or the linear pattern-bound variables.
-%
-As a case expression is evaluated, the scrutinee will eventually be in WHNF,
+\paragraph{Irrelevance}
+%
+As discussed, proof irrelevant resources can only be consumed
+indirectly, essentially encoding that the (non-WHNF) case scrutinee
+resources must be consumed through the case binder or the linear
+pattern-bound variables.
+%
+As a case expression is evaluated, the scrutinee will eventually
+reduce to WHNF,
 which must then be typed with rule $Case_{\textrm{WHNF}}$.
 %
-Crucially, these rules must ``work together'' in the system, in the sense that
+Crucially, the two case rules must be in harmony in the system, in the sense that
 case expressions typed using the $Case_{\textrm{Not WHNF}}$ rule must also be
-well-typed after the scrutinee is evaluated to WHNF, which is then typed using
-the $Case_{\textrm{WHNF}}$ rule.
+well-typed by the $Case_{\textrm{WHNF}}$ once the scrutinee is evaluated to WHNF.
 
@@
 The \emph{Irrelevance} lemma witnesses the soundness of typing a
-case alternative with proof irrelevant resources in a certain context with respect to
-typing the same expression with arbitrary resources (we note, however, typing
-an alternative with proof irrelevant resources is not complete wrt using
+case alternative with proof irrelevant resources with respect to
+typing the same expression with arbitrary resources. However, typing
+a case alternative with proof irrelevant resources is not complete wrt using
 arbitrary resources -- a counter example needs only to use a resource
-directly).
+directly.
 

### 2024-01-30 icfp.tex
finished section 5

@@
 patterns $\rho_1\dots \rho_n$ and corresponding right hand sides $e'_1\dots
-e'_n$. Notably, (recursive) let-bound
+e'_n$. We often omit the case binder $z$ when it is unused. Notably, (recursive) let-bound
 binders and case-bound binders are annotated
@@
 matching on the same constructor with an unrestricted field allows it to be
-used unrestrictedly. Programs are a set of declarations and a top-level
-expression.
+used unrestrictedly.
+%Programs are a set of declarations and a top-level
+%expression.
 
@@
 
-\noindent Finally, unrestricted resources are equivalent to $\D$-bound
-variables with an empty ($\cdot$) usage environment.
+\noindent Finally, unrestricted and $\D$-bound
+variables with an empty usage environment are equivalent.
 
@@
 follows by structural induction on the reductions $e \longrightarrow
-e'$ from the operational semantics. Most cases are straightforward and usually
-appeal to one or more of the substitution lemmas described below. The most
+e'$ from the operational semantics. Most cases are straightforward, appealing to one or more
+substitution lemmas (see Appendix~\ref{sec:proof:type-preservation}). The most
 interesting case is that of case expressions whose scrutinee can be further
-evaluated -- we branch on whether the scrutinee becomes in WHNF, and invoke the
-\emph{Irrelevance} lemma if so.
+evaluated -- we branch on whether the scrutinee becomes in WHNF and appeal to Lemma~\ref{lem:irrev} if so.
 %
@@
 \subsection{Optimisations preserve linearity\label{sec:optimisations-preserve-types-meta}}
-
-One of the primary goals of the Linear Core type system is being suitable for
-intermediate representations of optimising compilers for lazy languages with
-linear types. In light of this goal, we prove that \emph{multiple optimising
-transformations} are type preserving in Linear Core, and thus preserve linearity.
+\todo[inline]{Check Commuting case-let in Fig.~\ref{fig:opttrans}!}
+A main goal of the Linear Core type system is 
+to serve as a typed intermediate representation for optimising
+compilers of lazy languages with linearity. In light of this,
+we show that multiple optimising transformations are type
+preserving in Linear Core (i.e.,~preserve linearity semantically).
 
@@
 % %
-Transformations are described by an arbitrary well-typed expression with a certain shape, on
-the left hand side (lhs) of the arrow $\Longrightarrow$, resulting in an expression on
-the right hand side (rhs) that we prove to be well-typed.
-%
-% For our proofs, we assume the lhs to be a well-typed expression and prove the
-% rhs is well-typed as well.
-%
-For each transformation, we describe the intuition behind the transformation
-preserving linearity in our system.
-
-\todo[inline]{Aqui uma lista itemizada das transformações todas. As
-  provas todas para apendice.}
-
+%\{\overline{\rho\to e'}\}
+\begin{figure}[t]
+
+  {\small
+  \[
+    \begin{array}{ll}
+   (\llet{\xD = e}{e'} )  \Longrightarrow \llet{\xD =       e}{e'[e/x]} &                    \mbox{Inlining}\\
+       (\lambda \x[\pi][\s].~e)~e'  \Longrightarrow e[e'/x]  &                \mbox{$\beta$-reduction}\\
+      (\ccase{(K~\ov{e})}{\z[\D][\s]~\{\overline{\rho_j\to e_j}, K~\ov{x} \to e_i, \overline{\rho_k\to e_k}\}} )\Longrightarrow  e_i\ov{[e/x]}[K~\ov{e}/z] &
+                                  \mbox{Case of known constr.}\\
+      (\ccase{(\ccase{e_c}{\z[\D]~\{\ov{\rho_{c_i}\to e_{c_i}}\}})}{\var[w][\irr{\D,\D'}][\s']~\{\ov{\rho_i\to e_i}\}})\\
+            \Longrightarrow 
+      \ccase{e_c}{\z[\D]~\{\ov{\rho_{c_i} \to \ccase{e_{c_i}}{w~\{\ov{\rho_i\to e_i}\}}}\}}
+      & \mbox{Case-of-case}\\
+      (\lambda \y[\pi].~\llet{\xD = e}{e'})
+      \Longrightarrow
+      \llet{\xD=e}{\lambda\y[\pi].~e'} & \mbox{Commuting let-$\lambda$}\\
+      ((\llet{v = e}{b})~a) \Longrightarrow \llet{v = e}{b~a} &\mbox{Commuting let-app.}\\
+      (\ccase{(\llet{v = e}{b})}{\{\overline{\rho\to e'}\}}) \Longrightarrow
+      \llet{v =  e}{\ccase{b}{\{\overline{\rho\to e'}\}}}
+                               &
+                                 \mbox{Commuting let-case}\\
+      (\ccase{e}{\{\overline{\rho_j\to e_j}, \rho \to E[e_1], \overline{\rho_k\to e_k}\}} ) \qquad (\mbox{with $x$ fresh}) \\
+     \Longrightarrow \llet{x = e_1}{\ccase{e}{\{\overline{\rho_j\to e_j}, \rho \to E[t], \overline{\rho_k\to e_k}\}} 
+                                                                        & \mbox{Commuting case-let}\\
+      (\llet{x = (\llet{v = e}{b})}{c}) \Longrightarrow \llet{v = e}{\llet{x = b}{c}}
+                                                                      & \mbox{Commuting let-let}\\
+      f \Longrightarrow \lambda x.(f~x) & \mbox{$\eta$-expansion}\\
+      \lambda x.(f~x) \Longrightarrow f & \mbox{$\eta$-reduction}\\
+      (\ccase{x}{z~\{\ov{\rho_i\to e_i}\}}) \Longrightarrow \ccase{x}{z~\{\ov{\rho_i\to  e_i[z/x]}\}}
+                                                                      & \mbox{Binder swap}\\
+      % (\ccase{x}{z~\{\ov{\rho_i\to e_i}\}} )\Longrightarrow \ccase{x}{z~\{\ov{\rho_i\to e_i[x/z]}\}}
+      %                                                                 & \mbox{Reverse binder swap}
+    \end{array}
+  \]}
+  \caption{Optimising transformations validated by Linear Core\label{fig:opttrans} (see Appendix~\ref{app:optimisations} for details)} 
+\end{figure}
+
+We describe Core-to-Core transformations as $e_1 \Longrightarrow e_2$,
+where $e_1$ is an arbitrary \emph{well-typed} expression
+matching a certain shape (or of a certain type), which is
+transformed into expression $e_2$. Validating a transformation
+entails showing that $e_2$ is well-typed. We often annotate the arrow
+with the name of the transformation when describing chains of such
+transformations.
+
+The list of transformations we have
+validated is given in Figure~\ref{fig:opttrans}.
+%
+The behaviour of most transformations can be inferred from its name:
+Inlining substitutes a let-bound expression in the let body; $\beta$-reduction
+and case of known constructor effectively perform evaluation of the given expression;
+case-of-case commutes cases in scrutinee position inside the outer case alternatives;
+let floating and  for let-bound expressions to be floated out of $\lambda$-abstractions;
+the four commuting conversions allow let-bound expressions to be commuted with other
+Core constructs; $\eta$-expansion and reduction apply the $\eta$-laws to terms of function type;
+finally, binder swap substitutes occurrences of a scrutinee variable for the
+case binder. While many of the transformations above are not obviously optimising,
+they are chained in the compiler pipeline to expose optimisation opportunities.
+
+% Transformations are described by an arbitrary well-typed expression with a certain shape, on
+% the left hand side (lhs) of the arrow $\Longrightarrow$, resulting in an expression on
+% the right hand side (rhs) that we prove to be well-typed.
+% %
+% % For our proofs, we assume the lhs to be a well-typed expression and prove the
+% % rhs is well-typed as well.
+% %
+% For each transformation, we describe the intuition behind the transformation
+% preserving linearity in our system.
+
+% \todo[inline]{Aqui uma lista itemizada das transformações todas. As
+%   provas todas para apendice.}
+
+%%% COMENTADO AQUI
 % \subsubsection{Inlining}
@@
 
-\subsubsection{Reverse Binder Swap Considered Harmful\label{sec:reverse-binder-swap-considered-harmful}}
-
-The reverse binder swap transformation substitutes occurrences of the case
-binder $z$ in case alternatives by the scrutinee, when the scrutinee is a
+\paragraph{Reverse Binder Swap\label{sec:reverse-binder-swap-considered-harmful}}
+
+We now discuss an issue we have found with the so-called reverse binder swap
+transformation.
+%
+%
+Reverse binder swap is dual to the binder swap transformation,
+substituting occurrences of the case
+binder $z$ by the scrutinee in the special case when the scrutinee is a
 variable $x$.
-
-\ReverseBinderSwapTheorem
-
-\noindent This is exactly reverse from what the binder swap transformation
-does in hope of eliminating multiple uses of $x$ so as to inline it. However, by
+%
+% his is dual to what the binder swap transformation
+% does in hope of eliminating multiple uses of $x$ (allowing for inlining). 
+By
 using the scrutinee $x$ instead of the case binder, we might be able to float
@@
 might float an expensive computation involving $z$ out of the case alternative,
-where $z$ is out of scope but $x$ isn't:
-\[
+where $z$ is now out of scope:
+
+{\small\[
 \begin{array}{l}
-\lambda x.~\lletrec{go~y = \ccase{x}{z~\{(a,b) \to \dots (expensive~z) \dots\} }}{\dots}\\
+(\lambda x.~\lletrec{go~y = \ccase{x}{z~\{(a,b) \to \ E[expensive~z] \} }}{\dots})
 \Longrightarrow_\textrm{Reverse binder swap}\\
-\lambda x.~\lletrec{go~y = \ccase{x}{z~\{(a,b) \to \dots (expensive~x) \dots\} }}{\dots}\\
-\Longrightarrow_\textrm{Float out}\\
-\lambda x.~\llet{t = expensive~x}{\lletrec{go~y = \ccase{x}{z~\{(a,b) \to \dots t \dots\} }}{\dots}}\\
+(\lambda x.~\lletrec{go~y = \ccase{x}{z~\{(a,b) \to E[expensive~x] \} }}{\dots})
+\Longrightarrow_\textrm{Commuting case-let}\\
+\lambda x.~\llet{t = \mathit{expensive}~x}{(\lletrec{go~y = \ccase{x}{z~\{(a,b) \to E[t] \} }}{\dots})}\\
 \end{array}
-\]
-If $go$ is a loop, by applying the reverse binder swap we now only compute
-$expensive~x$ once instead of in every loop iteration.
-
-Despite GHC applying the reverse binder swap transformation to core programs
-during Core-to-Core optimisation passes, this optimisation violates linearity
+\]}
+
+\noindent If $go$ is compiled into a loop, we only compute
+$\mathit{expensive}~x$ once instead of in every loop iteration.
+
+Despite GHC applying the reverse binder swap transformation 
+during its Core-to-Core optimisation passes, this optimisation actually \emph{breaks} linearity
 when considered as a transformation on Linear Core programs. In practice, the
-optimisation preserves linearity in Core when applied as part of the GHC
-transformation pipeline only due to the occurrence analyser being naive with
+optimisation preserves linearity when applied as part of the GHC
+pipeline due to the occurrence analyser being naive with
 regard to semantic linearity. Initially, it might seem as though an expression
 in which a variable $x$ occurs both in the case scrutinee and in the
-alternatives is linear, for example:
-\[
-\G; \xl \vdash \ccase{x}{\_ \to x} : \s
-\]
-The reasoning is done by branching on whether $x$ refers to an expression in
-WHNF or an unevaluated thunk.
-\begin{itemize}
-\item If $x$ refers to an unevaluated expression, then scrutinizing it results
+alternatives is linear (e.g.,~$\ccase{x}{\_ \to x}$):
+% \[
+% \G; \xl \vdash \ccase{x}{\_ \to x} : \s
+% \]
+% The reasoning is done by branching on whether $x$ refers to an expression in
+% WHNF or an unevaluated thunk.
+% \begin{itemize}
+% \item
+  If $x$ refers to an unevaluated expression, then scrutinizing it results
 in the expression bound by $x$ to be evaluated to WHNF. In a subsequent use of
@@
 variables.
-\item If $x$ already refers to an expression evaluated to WHNF, then
+%
+On the other hand, if $x$ already refers to an expression evaluated to WHNF, then
 scrutinizing it in a case expression is a no-op, thus we may use it again (in
-mutual exclusion with the case binder and pattern variables)
-\end{itemize}
+mutual exclusion with the case binder and pattern variables).
+%\end{itemize}
 %
@@
 %
+
 Even though, on its own, it makes intuitive sense that this example indeed uses
 $x$ linearly, when considered as part of a complete type system, allowing this
-expression to be linear makes the system unsound.
-
-We recall that $\beta$-reduction reduces an application of a linear function
+expression to be linear makes the system \emph{unsound}.
+To see why this is the case, recall that $\beta$-reduction reduces an application of a linear function
 using call-by-name -- if we know the argument is used exactly once, a binding
@@
 
-Consider the function application $$(\lambda x.~\ccase{x}{\_ \to x})~(use~y)$$
+Consider the function application $(\lambda x.~\ccase{x}{\_ \to x})~(use~y)$
 where $y$ is a free linear variable. Assuming $\lambda x.~\ccase{x}{\_ \to x}$
-is a linear function by the reasoning above, $\beta$-reduction transforms the
-application in $(\ccase{x}{\_ \to x})[use~y/x]$, i.e. $\ccase{use~y}{\_ \to
+is a linear function, $\beta$-reduction transforms the
+application into $\ccase{use~y}{\_ \to
   use~y}$. Now, $y$ is a linear variable \emph{consumed} in the scrutinee of the case
-expression, yet it occurs in the body of the case alternative as well --
+that also occurs in the case alternative --
 linearity is violated by using the linear resource $y$ twice.
-
-Essentially, Linear Core would be unsound, and even duplicate resources, if the
-above kind of expressions, where linear variable scrutinees occur in the
-alternatives body, were well-typed, because of its interaction with the
-call-by-name reduction of linear functions. In this sense, the reverse binder
+%
+% Essentially, Linear Core would be unsound, and even duplicate resources, if the
+% above kind of expressions, where linear variable scrutinees occur in the
+% alternatives body, were well-typed, because of its interaction with the
+% call-by-name reduction of linear functions.
+In this sense, the reverse binder
 swap is an optimisation that creates ill-typed expressions from well-typed
-ones, so it is deemed an invalid optimisation that doesn't preserve types in
+ones, so it is deemed an \emph{invalid} optimisation that does not preserve types in
 our system.
 
-The reverse binder swap is not a problem in the GHC simplifier because of the
+Reverse binder swap is not a problem in the GHC simplifier because of the
 weaker notion of linearity understood by occurrence analysis. Occurrence
@@
 $\beta$-reduction is done with call-by-need on such an expression. If the above
-example were reduced with call-by-need:
-\[
-\begin{array}{l}
-(\lambda x.~\ccase{x}{\_ \to x})~(use~y)
-\Longrightarrow_\textrm{CBNeed}
-\llet{x = use~y}{\ccase{y}{\_ \to y}}
-\end{array}
-\]
-Then the computation using $y$ would be let-bound, and $y$ used as a scrutinee
-variable, which is indeed an expression semantically linear in $x$.
-
-Concluding, in being able to understand more programs as linear, our type
-system allows more expressions to be considered linear for $\beta$-reduction
-without a let-allocation, however, it makes reverse binder swap an invalid
-transformation since its output, when considered linear, might violate
-linearity when further optimised.
+example were reduced with call-by-need
+% \[
+% \begin{array}{l}
+% (\lambda x.~\ccase{x}{\_ \to x})~(use~y)
+% \Longrightarrow_\textrm{CBNeed}
+% \llet{x = use~y}{\ccase{y}{\_ \to y}}
+% \end{array}
+% \]
+then the computation using $y$ would be let-bound, and $y$ used as a scrutinee
+variable, which is indeed semantically linear in $x$.
+
+% Concluding, in being able to understand more programs as linear, our type
+% system allows more expressions to be considered linear for $\beta$-reduction
+% without a let-allocation, however, it makes reverse binder swap an invalid
+% transformation since its output, when considered linear, might violate
+% linearity when further optimised.
 
@@
 
-\section{Optimisations preserve linearity}
+\section{Optimisations preserve linearity}\label{app:optimisations}
 

### 2024-01-30 icfp.tex
Passed through Section 6

@@
 % %
-Linear Haskell~\cite{cite:linearhaskell} is concerned with retrofitting
-linear types in Haskell but
-is not concerned with extending GHC's intermediate language(s),
+Linear Haskell~\cite{cite:linearhaskell} augments Haskell with 
+linear types. However, the development is
+not concerned with extending GHC's intermediate language(s),
 which presents its own challenges. 
 %
-Nonetheless, while the Linear Haskell work keeps Core unchanged, its
+Nonetheless, while Linear Haskell keeps GHC Core unchanged, its
 implementation in GHC does modify and extend Core with linearity/multiplicity
 annotations. Core's type system is unable to type \emph{semantic} linearity of
-programs (in contrast to \emph{syntactic} linearity), which results in Core
+programs in the sense elaborated in our work, which results in Core
 rejecting linear programs resulting from optimising transformations that
@@
 %
-Linear Core overcomes the limitations of Core's linear type system derived from
-Linear Haskell by understanding semantic linearity in the presence of laziness,
-and provably accepts multiple Core-to-Core passes. Linear Core, ultimately, can
+Our work overcomes the limitations of Core's linear type system derived from
+Linear Haskell by understanding linearity semantically, in the presence of laziness,
+and accepting multiple Core-to-Core passes.
+Linear Core can
 also be seen as a system that validates the programs written in Linear Haskell
@@
 precursor to our work. Linear Mini-Core first observes the incapacity of
-Core's type system to accept linear programs after transformations, and first
+Core's type system to accept linear programs after transformations, and 
 introduces usage environments for let-bound variables with the same goal of
@@
 
-We draw from Linear Mini-Core's the rule for non-recursive let expressions and
-how let-bound variables are annotated with a usage environment, however, our
+We draw from Linear Mini-Core the rule for non-recursive let expressions and
+how let-bound variables are annotated with a usage environment. However, our
 work further explores the interaction of laziness with linearity in depth, and
-diverges in rules for typing other constructs, notably, case expressions and
+diverges significantly in rules for typing other constructs, notably, case expressions and
 case alternatives. Furthermore, unlike Mini-Core, we prove type safety of our
@@
 all designed in the context of a typed language (Core) which does not have
-linear types. However,
-~\cite{cite:let-floating,peytonjones1997a,cite:linearhaskell} observe that
+linear types. However, some works~\cite{cite:let-floating,peytonjones1997a,cite:linearhaskell} have observed that
 certain optimisations (in particular, let-floating and inlining) greatly
 benefit from linearity analysis and, in order to improve those transformation,
-linear-type-inspired systems were created specifically for the purpose of the
-transformation.
+special purpose linear-type-inspired systems were created and implemented.
 %
 By fully supporting linear types in Core, these optimising transformations
-could be informed by the language inherent linearity, and, consequently, avoid
-an ad-hoc or incomplete linear-type inference pass custom-built for
+could be informed by the language's intrinsic linearity, and, consequently, avoid
+an ad-hoc or incomplete linear type inference pass that is custom-built for
 optimisations. Additionally, the linearity information may potentially be used
-to the benefit of optimising transformations that currently don't take any
+to the benefit of optimising transformations that currently do not take 
 linearity into account.
@@
 
-We highlight some avenues of further research. Briefly,
+We highlight some avenues of future work. Briefly,
 these include \emph{multiplicity coercions}, optimisations leveraging
@@
 \paragraph{Multiplicity Coercions.}
-Linear Core doesn't have type equality coercions, a flagship feature of GHC
+Linear Core does not have type equality coercions, a flagship feature of GHC
 Core's type system.
@@
 informs optimisations, allowing the compiler to generate more performant programs.
-
+%
 Implementing Linear Core in GHC is a challenging endeavour, since we must account
@@
 Constraints~\cite{cite:linearconstraints}). Linear Core eases the restrictions
-of linear typing by being more flexible in understanding linearity for lazily
-evaluated languages such as Haskell. In this sense, it is an avenue of future
-work to apply the ideas from Linear Core to the surface Haskell language.
+of linear typing by being more flexible in understanding linearity for lazy
+languages. In this sense, it is an avenue of future
+work to apply the ideas from this work to the surface Haskell language.
 

### 2024-01-30 icfp.tex
fixed bibliography and compilation errors in appendices

@@
-\documentclass[acmsmall,review]{acmart}
+\documentclass[acmsmall,review,anonymous]{acmart}
 \PassOptionsToPackage{dvipsnames}{xcolor}
@@
 
-\newtheorem{theorem}{Theorem}
-\newtheorem{proposition}[theorem]{Proposition}
+%\newtheorem{theorem}{Theorem}
+%\newtheorem{proposition}[theorem]{Proposition}
 % \newtheorem{lemma}[theorem]{Lemma}%[theorem]
-\newtheorem{lemma}{Lemma}%[theorem]
-\newtheorem{sublemma}{Lemma}[lemma]
-\newtheorem{assumption}{Assumption}
+%\newtheorem{lemma}{Lemma}%[theorem]
+%\newtheorem{sublemma}{Lemma}[lemma]
+%\newtheorem{assumption}{Assumption}
 
@@
 alternatives, directly in the \lstinline{True} branch or via the
-let-bound \lstlinline{y} in the \lstinline{False} branch.
+let-bound \lstinline{y} in the \lstinline{False} branch.
 
@@
 suitable for the intermediate language of an optimising compiler.
-\todo[inline]{Tem de ser anonimizado, provavelmente alterando a bib
-  entry para omitir autores e o proprio link}
+%\todo[inline]{Tem de ser anonimizado, provavelmente alterando a bib
+  %entry para omitir autores e o proprio link}
 The GHC Linear Core plugin~\cite{cite:linear-core-plugin} implements a
@@
   \begin{lemma}[Linear to $\Delta$-bound]
-\\
 If $\G; \D',\xl \vdash e :\vp$
@@
       (\ccase{e}{\{\overline{\rho_j\to e_j}, \rho \to E[e_1], \overline{\rho_k\to e_k}\}} ) \qquad (\mbox{with $x$ fresh}) \\
-     \Longrightarrow \llet{x = e_1}{\ccase{e}{\{\overline{\rho_j\to e_j}, \rho \to E[t], \overline{\rho_k\to e_k}\}} 
+     \Longrightarrow \llet{x = e_1}{\ccase{e}{\{\overline{\rho_j\to e_j}, \rho \to E[t], \overline{\rho_k\to e_k}\}}} 
                                                                         & \mbox{Commuting case-let}\\
@@
 
-\bibliographystyle{plainyr}
+\bibliographystyle{ACM-Reference-Format}
 \bibliography{references}
@@
 
+\subsection{Substitution Lemmas\label{sec:proof:substitution-lemmas}}
+
+\input{language-v4/proofs/LinearSubstitutionLemma}
+
+\input{language-v4/proofs/UnrestrictedSubstitutionLemma}
+
+\input{language-v4/proofs/DeltaSubstitutionLemma}
+
+\subsection{Irrelevance\label{sec:proof:irrelevance}}
+
+\input{language-v4/proofs/WHNFConvSoundness}
+
 \subsection{Type Preservation\label{sec:proof:type-preservation}}
@@
 
-\subsection{Irrelevance\label{sec:proof:irrelevance}}
-
-\input{language-v4/proofs/WHNFConvSoundness}
-
-\subsection{Substitution Lemmas\label{sec:proof:substitution-lemmas}}
-
-\input{language-v4/proofs/LinearSubstitutionLemma}
-
-\input{language-v4/proofs/UnrestrictedSubstitutionLemma}
-
-\input{language-v4/proofs/DeltaSubstitutionLemma}
-
 \section{Optimisations preserve linearity}\label{app:optimisations}
 
-Proofs are given inline in Chapter~\ref{sec:optimisations-preserve-types-meta},
-with the exception of the proof that \emph{case-of-case} preserves types, which
-is lengthier than the others.
+\subsection{Inlining}
+\input{language-v4/proofs/optimizations/Inlining}
+
+\subsection{\texorpdfstring{$\beta$}{Beta}-reduction}
+
+\input{language-v4/proofs/optimizations/BetaReduction}
+
+\subsection{Case of known constructor}
+
+\input{language-v4/proofs/optimizations/CaseOfKnownConstructor}
 
@@
 
+The case of case transformation applies to case expressions whose scrutinee is
+another case expression, and returns the innermost case expression transformed by repeating
+the outermost case expression in each alternative of the innermost case,
+scrutinizing the original alternative body.
+
+Intuitively, since the scrutinee of the outermost case is not in WHNF, no
+resources from it can directly occur in the outermost alternatives. By moving
+the outermost alternatives inwards with a different scrutinee, the alternatives
+remain well-typed because they are typed using either the case binder or the
+pattern bound variables, which, by the \emph{Irrelevance} lemma, makes it
+well-typed for any scrutinee consuming arbitrary resources.
+
 \input{language-v4/proofs/optimizations/CaseOfCase}
+
+\subsection{Commuting let}
+
+\input{language-v4/proofs/optimizations/LetFloating}
+
+\subsection{\texorpdfstring{$\eta$}{Eta}-conversions}
+
+\input{language-v4/proofs/optimizations/EtaConvs}
+
+\subsection{Binder Swap}
+
+The binder swap transformation applies to case expressions whose scrutinee is a
+single variable $x$, and it substitutes occurrences of $x$ in the case
+alternatives for the case binder $z$. If $x$ is a linear resource, $x$ cannot
+occur in the case alternatives (as we conservatively consider variables are not
+in WHNF), so the substitution preserves types vacuously. Otherwise, $x$ can be
+freely substituted by $z$, since $z$ is also an unrestricted resource (it's
+usage environment is empty because $x$ is unrestricted).
+
+\input{language-v4/proofs/optimizations/BinderSwap}
 

### 2024-02-07 icfp.tex
RM Edit Section 1

@@
 In programming languages with a linear type system, not using certain resources
-or using them twice is flagged as a type error. Linear types can thus be used
-to, for instance, statically guarantee that socket descriptors are closed or
-heap-allocated memory is freed, exactly once (leaks and double-frees become
-type errors), or guarantee channel-based communication protocols are
-deadlock-free~\cite{10.1007/978-3-642-15375-4_16},
+or using them twice is flagged as a type error. For instance, linear types can
+statically enforce correct usage of file or socket descriptors, or
+that heap-allocated memory is freed exactly once (leaks and double-frees become
+type errors), or even guarantee deadlock freedom in channel-based communication
+protocols~\cite{10.1007/978-3-642-15375-4_16},
 % implement safe
@@
 %
-For instance, consider the following program in which allocated memory is freed twice.
+The following program with a runtime error (allocated memory is freed twice) is
+accepted by a C-like type system, however, under the lens of a linear type
+system, if $p$ is deemed to be a linear resource created by the call to
+\texttt{malloc}, we would instead reject it:
 % We know this to be the dreaded double-free error which will crash the program at runtime.
-Regardless of the \emph{double-free} error, a C-like type system will
-accept such a program:
 % \centering
@@
 \end{code}
-Under the lens of a linear type system, if $p$ is deemed to be a
-linear resource created by the call to \texttt{malloc}, it
-must be used \emph{exactly once}.  However, the program above uses $p$ twice,
-in the two calls to \texttt{free} and so fails to type-check.
-%
-In this sense, linear typing
-effectively ensures the program does not compile with a double-free error.
 
@@
 ownership types build on linear types to guarantee memory safety
-without garbage collection or reference counting, and, more recently,
+without garbage collection, and, more recently,
 Haskell~\cite{cite:linearhaskell}, a pure, functional, and
@@
 However, linearity in Haskell stands out from linearity in other
-languages given that linear types permeate Haskell down to (its) Core,
-the intermediate language into which Haskell is translated. Core is a
-minimal, explicitly typed, functional language, to which multiple
-Core-to-Core optimising transformations are applied during
-compilation. Due to Core's minimal design, typechecking Core programs
-is very efficient and doing so serves as a sanity check to the
-correction of the source transformations. If the resulting optimised
-Core program fails to typecheck, the optimising transformations are
-very likely to have introduced an error in the resulting program.
-%
-
-Both Haskell and its intermediate language Core are \emph{lazily}
-        evaluated, i.e., expressions in Haskell are only evaluated when needed,
-        (unlike Rust, which is \emph{eagerly} evaluated).
+languages because linear types permeate Haskell down to (its) Core, the
+intermediate language into which Haskell is translated by the Glasgow Haskell Compiler (GHC).
+%
+Core is a minimal, explicitly typed, functional language to which multiple
+Core-to-Core optimising transformations are applied during compilation.
+%
+Typechecking Core serves as a sanity check to the correction of source
+transformations, since unsound ones will likely introduce ill-typed
+expressions, making them easy to spot.
+%
+% If a Core program fails to typecheck, the
+% optimising transformations are very likely to have introduced an error in the
+% resulting program.
+%
+
+Furthermore, both Haskell and its intermediate language Core are \emph{lazily}
+        evaluated -- expressions in Haskell are only evaluated when needed
+        (unlike, e.g., Rust, which is \emph{eagerly} evaluated).
         %
@@
 
-Despite of the advantages of a linearly typed Core, the status quo is
+Despite the advantages of a linearly typed Core, the status quo is
 that linearity is effectively ignored in Core, despite being checked
 in the source Haskell code. The reason is that the desugaring
-and the subsequent Core-to-Core optimising transformations,
+and the subsequent Core-to-Core optimising transformations
 eliminate and rearrange most of the syntactic constructs through
@@
 Crucially, while optimisations aim to preserve the semantics of
-programs and so preserve a program's linear resource consumption,
+programs and therefore a program's linear resource consumption,
 a \emph{naive} syntactic analysis of the optimised programs (as
@@
 For instance, let $x$ be a linear resource in the two following programs, where
-the latter results from inlining $y$ in the let body of the former. Despite the
-the program no longer appearing linear (as there are now two 
+the latter results from inlining $y$ in the let body of the former. Even though the
+the program no longer appears linear (as there are now two 
 occurrences of $x$), the program \emph{is} indeed linear
@@
 %
-We compare our contributions to related work and discuss possible
-avenues for further research
+% We compare our contributions to related work and discuss possible
+% avenues for further research
 % (highlighting so-called
 % \emph{multiplicity coercions})
-in Section~\ref{sec:discussion}, which concludes the document.
+% in Section~\ref{sec:discussion}, which concludes the document.
 

### 2024-02-11 icfp.tex
minor changes

--- 2024-01-30 icfp.tex
@@
 \renewcommand{\DeltaLinearRelationLemma}{
-\begin{lemma}[$\Delta$-bound to Linear]
+\begin{lemma}[$\Delta$-bound to Linear]\label{lem:deltaone}
 % Any $\Delta$-variable with a fully-irrelevant usage environment can be made a linear variable as long as its irrelevant environment is dropped from the linear environment (and the unrestricted env.
@@
 \renewcommand{\LinearDeltaRelationLemma}{
-  \begin{lemma}[Linear to $\Delta$-bound]
+  \begin{lemma}[Linear to $\Delta$-bound]\label{lem:onedelta}
 If $\G; \D',\xl \vdash e :\vp$
@@
 \renewcommand{\DeltaUnrestrictedRelationLemma}{
-\begin{lemma}[Unrestricted and $\Delta$-bound]
+\begin{lemma}[Unrestricted and $\Delta$-bound]\label{lem:undelta}
 %An unrestricted variable is equivalent to a $\D$-var with an empty usage environment.\\

### 2024-02-11 language-v4/proofs/WHNFConvSoundness.tex
Restate theorems and lemmas
minor changes

--- 2023-09-12 language-v4/proofs/WHNFConvSoundness.tex
@@
-\WHNFConvSoundness
+\irrelevancelemma*
 
@@
 $\irr{\irr{\D}}\neq\irr{\D}$)\\
-    (4) $\G; \z[1], \D' \vdash e : \vp$\` by $\D \Rightarrow 1$ lemma (2,3)\\
-    (5) $\G, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash e : \vp$\` by $1 \Rightarrow \D$ lemma (4)\\
-    (6) $\G, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash_{alt} \_ \to e :^z_{\ov{\D_i}} \s \Rightarrow \vp$\` by $Alt\_$\\
+    (4) $\G[z/\irr{\D}]; \z[1], \D' \vdash e : \vp$\` by Lemma~\ref{lem:deltaone} (2,3)\\
+    (5) $\G[\ov{\D_i}/\irr{\D}], \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash e : \vp$\` by Lemma~\ref{lem:onedelta} (4)\\
+    (6) $\G, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash e : \vp$\` by (3), resources in $\G$ with $\irr{\D}$ (parts) in\\\` the usage environment cannot be used in $e$\\
+    (7) $\G, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash_{alt} \_ \to e :^z_{\ov{\D_i}} \s \Rightarrow \vp$\` by $Alt\_$\\
 \end{tabbing}
@@
 \item[Case:] $AltN_{\textrm{Not WHNF}}$
+    We prove the theorem for constructing both an $AltN_{\textrm{Not WHNF}}$
+        and an $AltN_{\textrm{WHNF}}$ from a proof-irrelevant $AltN_{\textrm{Not
+        WHNF}}$, to prove the statement holds for any $\Rightarrow$ kind for
+        $AltN$, rather than just $\Rrightarrow$ or $\Mapsto$.
 \begin{tabbing}
-    (1) $\G, \z[\irr{\D}]; \irr{\D}, \D' \vdash_{alt} K~\ov{\xo},\ov{y_i{:}_{1}\s_i}^n\to e :^z_{\irr{\D}} \s \Rightarrow \vp$\\
+    (1) $\G, \z[\irr{\D}]; \irr{\D}, \D' \vdash_{alt} K~\ov{\xo},\ov{y_i{:}_{1}\s_i}^n\to e :^z_{\irr{\D}} \s \Rrightarrow \vp$\\
     (2) $\G,\z[\irr{\D}],\ov{\xo},\ov{y_j{:}_{\D_j}\s_j}^n; \irr{\D}, \D' \vdash e : \vp$\`by inv. on $AltN_{\textrm{Not WHNF}}$\\
@@
     Subcase $\irr{\D}$ is consumed through $z$\\
-    (4) $\G,\ov{\xo},\ov{y_j{:}_{\D_j}\s_j}^n; \z[1], \D' \vdash e : \vp$\`by $\D \Rightarrow 1$ lemma (2,subcase)\\
-    (5) $\G,\ov{\xo},\ov{y_j{:}_{\D_i}\s_j}^n; \z[1], \D' \vdash e : \vp$\`by $\ov{y}$ does not consume resources,\\\` and Weaken? (HELP move $\D_j$ to $\D_i$)\\
-    (6) $\G,\ov{\xo},\ov{y_j{:}_{\D_i}\s_j}^n, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash e : \vp$\`by $1 \Rightarrow \D$ lemma (5)\\
+    (4) $\G,\ov{\xo},\ov{y_j{:}_{\D_j[z/\irr{\D}]}\s_j}^n; \z[1], \D' \vdash e : \vp$\`by Lemma~\ref{lem:deltaone} (2,subcase)\\\`and vars in $\G$ with $\irr{\D}$ cannot be used by subcase\\
+    Subcase constructing $AltN_{\textrm{WHNF}}$ ($\Mapsto$)\\
+    (5) $\G,\ov{\xo},\ov{y_j{:}_{\D_i}\s_j}^n; \z[1], \D' \vdash e : \vp$\`by $\ov{y}$ is not used, and $Weaken$\\
+    (6) $\G,\ov{\xo},\ov{y_j{:}_{\D_i}\s_j}^n, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash e : \vp$\`by Lemma~\ref{lem:onedelta} (5)\\\` and $\G$ vars do not mention $z$ by (4) \\
     (7) $\G, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash_{alt} K~\ov{\xo},\ov{y_i{:}_{1}\s_i}^n \to e :^z_{\ov{\D_i}} \s \Mapsto \vp$\`by $AltN_{\textrm{WHNF}}$ (6)\\
+    Subcase constructing $AltN_{\textrm{Not WHNF}}$ ($\Rrightarrow$)\\
+    (5) $\G,\ov{\xo},\ov{y_j{:}_{\lctag{\ov{\D_i}}{K_j}}\s_j}^n; \z[1], \D' \vdash e : \vp$\`by $\ov{y}$ does not consume resources\\
+    (6) $\G,\ov{\xo},\ov{y_j{:}_{\lctag{\ov{\D_i}}{K_j}}\s_j}^n, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash e : \vp$\`by $1 \Rightarrow \D$ lemma (5)\\\` and $\G$ vars do not mention $z$ by (4) \\
+    (7) $\G, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash_{alt} K~\ov{\xo},\ov{y_i{:}_{1}\s_i}^n \to e :^z_{\ov{\D_i}} \s \Rrightarrow \vp$\`by $AltN_{\textrm{Not WHNF}}$ (6)\\
     Subcase $\irr{\D}$ is (fully) consumed by $\ov{y}$ (after splitting)\\
-    (4) $\G,\z[\irr{\D}],\ov{\xo};\ov{y_j{:}_{1}\s_j}^n, \D' \vdash e : \vp$\`by $\D \Rightarrow 1$ lemma (2, subcase)\\
-    (5) $\G,\z[\ov{\D_i}],\ov{\xo};\ov{y_j{:}_{1}\s_j}^n, \D' \vdash e : \vp$\`by Weaken and $z$ does not consume resources (HELP SAME KIND)\\
-    (6) $\G,\z[\ov{\D_i}],\ov{\xo},\ov{y_j{:}_{\D_i}\s_j}^n; \ov{\D_i}, \D' \vdash e : \vp$\`by $1 \Rightarrow \D$ lemma (5)\\
+    (4) $\G,\z[\ov{y}],\ov{\xo};\ov{y_j{:}_{1}\s_j}^n, \D' \vdash e : \vp$\`by Lemma~\ref{lem:deltaone} (2, subcase)\\\` and vars in $\G$ with $\irr{\D}$ cannot be used by subcase\\
+    (5) $\G,\z[\ov{\D_i}],\ov{\xo};\ov{y_j{:}_{1}\s_j}^n, \D' \vdash e : \vp$\`by $Weaken$ and $z$ does not consume resources\\
+    Subcase constructing $AltN_{\textrm{WHNF}}$ ($\Mapsto$)\\
+    (6) $\G,\z[\ov{\D_i}],\ov{\xo},\ov{y_j{:}_{\D_i}\s_j}^n; \ov{\D_i}, \D' \vdash e : \vp$\`by Lemma~\ref{lem:onedelta} (5)\\\`and vars in $\G$ do not mention $\ov{y}$ (4,5)\\
     (7) $\G,\z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash K~\ov{\xo},\ov{y_i{:}_{1}\s_i}^n\to e :^z_{\ov{\D_i}} \s \Mapsto \vp$\`by $AltN_{\textrm{WHNF}}$\\
+    Subcase constructing $AltN_{\textrm{Not WHNF}}$ ($\Rrightarrow$)\\
+    (6) $\G,\z[\ov{\D_i}],\ov{\xo},\ov{y_j{:}_{\lctag{\ov{\D_i}}{K_j}}\s_j}^n; \ov{\D_i}, \D' \vdash e : \vp$\`by $1 \Rightarrow \D$ lemma (5)\\\`and vars in $\G$ do not mention $\ov{y}$ (4,5)\\
+    (7) $\G,\z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash K~\ov{\xo},\ov{y_i{:}_{1}\s_i}^n\to e :^z_{\ov{\D_i}} \s \Rrightarrow \vp$\`by $AltN_{\textrm{Not WHNF}}$\\
 \end{tabbing}

### 2024-02-12 icfp.tex
fixed proofs in appendix

--- 2024-02-07 icfp.tex
@@
 syntactic occurrence of a resource does not necessarily entail
-\emph{using} that resource when the program is evaluated.
-%
-Semantic linearity is especially necessary in optimising compilers for
-languages combining linearity and laziness: optimisations leverage laziness to
-heavily rewrite the source program, pushing the interaction of linearity and
-laziness to its limit, regardless of the original program typing linearity
-conservatively.
+\emph{using} that resource when the program is executed.
+Semantic linearity becomes especially vital when considering the
+internal language of an optimising compiler, where optimisations
+heavily rewrite the source program in ways that can often break
+syntactic linearity but that aim to preserve the program's semantics.
+% %
+% Semantic linearity is especially necessary in optimising compilers for
+% languages combining linearity and laziness: optimisations leverage laziness to
+% heavily rewrite the source program, pushing the interaction of linearity and
+% laziness to its limit, regardless of the original program typing linearity
+% conservatively.
 %
@@
 compiler plugin to validate the system against linearity-heavy libraries,
-including \texttt{linear-base}, in the heart of the compiler.
+including \texttt{linear-base}.%^, in the heart of the compiler.
 \end{abstract}
@@
 
-\item We implement Linear Core as a GHC plugin which checks linearity in
+\item We discuss our implementation of Linear Core as a GHC
+  plugin\footnote{We plan to submit our implementation as a companion
+    artifact to this paper.} which checks linearity in
 all intermediate Core programs produced during the compilation process, showing
@@
 \end{itemize}
+
+\noindent Additional proofs and definitions are given in Appendices~\ref{app:proofs}
+and~\ref{app:optimisations}.
+
 %
@@
 %
-Crucially, type preservation for the case-reduction substituting the
-case binder and pattern variables in the alternative is not possible
-to prove without branching on WHNF-ness, since otherwise the
-$\D$-substitution is not well-defined.  \todo[inline]{What is the
-  $\Delta$-substitution? Clarify}
+% Crucially, type preservation for the case-reduction substituting the
+% case binder and pattern variables in the alternative is not possible
+% to prove without branching on WHNF-ness, since otherwise the
+% $\D$-substitution is not well-defined.  \todo[inline]{What is the
+%   $\Delta$-substitution? Clarify}
 
@@
 \renewcommand{\DeltaLinearRelationLemma}{
-\begin{lemma}[$\Delta$-bound to Linear]
+\begin{restatable}[$\Delta$-bound to Linear]{lemma}{deltaone}\label{lem:deltaone}
 % Any $\Delta$-variable with a fully-irrelevant usage environment can be made a linear variable as long as its irrelevant environment is dropped from the linear environment (and the unrestricted env.
@@
 then $\G[x/\irr{\D}]; \D',\xl \vdash e :\vp$.
-\end{lemma}
+\end{restatable}
 }
@@
 \renewcommand{\LinearDeltaRelationLemma}{
-  \begin{lemma}[Linear to $\Delta$-bound]
+  \begin{restatable}[Linear to $\Delta$-bound]{lemma}{onedelta}\label{lem:onedelta}
 If $\G; \D',\xl \vdash e :\vp$
-then $\G[\D/x],\xD; \D,\D' \vdash e : \vp$.
-\end{lemma}
+then $\G[\D/x],\xD; \D,\D' \vdash e : \vp$, with fresh $\D$.
+\end{restatable}
 }
@@
 \renewcommand{\DeltaUnrestrictedRelationLemma}{
-\begin{lemma}[Unrestricted and $\Delta$-bound]
+\begin{restatable}[Unrestricted and $\Delta$-bound]{lemma}{undelta}\label{lem:undelta}
 %An unrestricted variable is equivalent to a $\D$-var with an empty usage environment.\\
 $\G,\xo; \D \vdash e : \vp$ iff $\G,\x[\cdot]; \D \vdash e : \vp$
-\end{lemma}
+\end{restatable}
 }
@@
 well-typed if the linear variable were instead $\D$-bound ($\xD$) with
-usage environment $\D$, with $\D$ available in the linear context instead of
+usage environment $\D$, with $\D$
+
+available in the linear context instead of
 the linear variable (and the unrestricted context updated accordingly).
@@
 
-\section{Type Safety Proofs}
+\section{Type Safety Proofs}\label{app:proofs}
+
+\subsection{Auxiliary Lemmas}
+The following lemmas presuppose the variable names across the typing
+contexts are distinct.
+
+  \begin{lemma}[Linear to $\Delta$-bound]~\label{lem:onedelta-app}
+
+    \begin{enumerate}
+
+      \item If $\G; \D',\xl \vdash e :\vp$
+        then $\G[\D/x],\xD; \D,\D' \vdash e : \vp$ (with $\Delta$ fresh).
+      \item If $\G ; \D' , \xl \vdash_{alt} \rho \rightarrow e
+        :^z_{\Delta_s} \sigma \Rightarrow \rho$ then
+        $\G[\D/x],\xD;\D,\D' \vdash_{alt} \rho \rightarrow e
+        :^z_{\Delta_s} \sigma \Rightarrow \rho$ (with $\Delta$ fresh).
+
+      \end{enumerate}
+\end{lemma}
+\begin{proof}
+Straightforward induction on the structure of the given derivation,
+reconstructing the original derivation and applying rule
+$\mathit{Var}_\Delta$ where rule $\mathit{Var}_1$ was applied on
+$x$ in the original derivation.
+\end{proof}
+\onedelta*
+
+\begin{proof}
+By Lemma~\ref{lem:onedelta-app}(1).
+\end{proof}
+
+\begin{lemma}[$\Delta$-bound to Linear]~\label{lem:deltaone-app}
+
+  \begin{enumerate}
+ \item If $\G,\x[\irr{\D}]; \irr{\D},\D' \vdash e : \vp$
+   then $\G[x/\irr{\D}]; \D',\xl \vdash e :\vp$.
+ \item If $\G ,\x[\irr{\D}]; \irr{\D}, \D' \vdash_{alt} \rho \rightarrow e
+   :^z_{\Delta_s} \sigma \Rightarrow \rho$ then
+   $\G[x/\irr{\D}];\D',\xl \vdash_{alt} \rho \rightarrow e
+   :^z_{\Delta_s} \sigma \Rightarrow \rho$.
+  \end{enumerate}
+\end{lemma}
+\begin{proof}
+Straightforward induction on the structure of the given derivation,
+reconstructing the original derivation and applying rule
+$\mathit{Var}_1$ where rule $\mathit{Var}_\D$ was applied on
+$x$ in the original derivation.
+\end{proof}
+
+\deltaone*
+\begin{proof}
+By Lemma~\ref{lem:deltaone-app}(1).
+\end{proof}
+
+\begin{lemma}[Unrestricted and $\Delta$-bound]~\label{lem:undelta-app}
+  \begin{enumerate}
+    \item $\G,\xo; \D \vdash e : \vp$ iff $\G,\x[\cdot]; \D \vdash e :
+      \vp$
+    \item $\G,\xo;\D \vdash_{alt} \rho \rightarrow e
+      :^z_{\Delta_s} \sigma \Rightarrow \rho$ iff
+      $\G,\x[\cdot];\D \vdash_{alt} \rho \rightarrow e
+   :^z_{\Delta_s} \sigma \Rightarrow \rho$
+  \end{enumerate}
+\end{lemma}
+\begin{proof}
+Straightforward induction on the structure of the given derivation,
+noting that rules $\mathit{Var}_\Delta$ and $\mathit{Var}_\omega$
+coincide when $\Delta$ is empty.
+\end{proof}
+
+\undelta*
+
+\begin{proof}
+By Lemma~\ref{lem:undelta-app}.
+\end{proof}
+
+\subsection{Irrelevance\label{sec:proof:irrelevance}}
+
+\input{language-v4/proofs/WHNFConvSoundness}
 
@@
 \input{language-v4/proofs/DeltaSubstitutionLemma}
-
-\subsection{Irrelevance\label{sec:proof:irrelevance}}
-
-\input{language-v4/proofs/WHNFConvSoundness}
 

### 2024-02-12 language-v4/proofs/WHNFConvSoundness.tex
fixed proofs in appendix

@@
     (5) $\G,\ov{\xo},\ov{y_j{:}_{\lctag{\ov{\D_i}}{K_j}}\s_j}^n; \z[1], \D' \vdash e : \vp$\`by $\ov{y}$ does not consume resources\\
-    (6) $\G,\ov{\xo},\ov{y_j{:}_{\lctag{\ov{\D_i}}{K_j}}\s_j}^n, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash e : \vp$\`by $1 \Rightarrow \D$ lemma (5)\\\` and $\G$ vars do not mention $z$ by (4) \\
+    (6) $\G,\ov{\xo},\ov{y_j{:}_{\lctag{\ov{\D_i}}{K_j}}\s_j}^n, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash e : \vp$\`by Lemma~\ref{lem:onedelta} (5)\\\` and $\G$ vars do not mention $z$ by (4) \\
     (7) $\G, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash_{alt} K~\ov{\xo},\ov{y_i{:}_{1}\s_i}^n \to e :^z_{\ov{\D_i}} \s \Rrightarrow \vp$\`by $AltN_{\textrm{Not WHNF}}$ (6)\\

### 2024-02-12 icfp.tex
proof of commuting case-let

@@
                                  \mbox{Commuting let-case}\\
-      (\ccase{e}{\{\overline{\rho_j\to e_j}, \rho \to E[e_1], \overline{\rho_k\to e_k}\}} ) \qquad (\mbox{with $x$ fresh}) \\
-     \Longrightarrow \llet{x = e_1}{\ccase{e}{\{\overline{\rho_j\to e_j}, \rho \to E[t], \overline{\rho_k\to e_k}\}}} 
+      (\ccase{e}{\{\overline{\rho_j\to e_j}, \rho \to E[e_1],
+      \overline{\rho_k\to e_k}\}} ) \qquad (\mbox{$x$ fresh}) \\
+     \Longrightarrow \llet{x = e_1}{\ccase{e}{\{\overline{\rho_j\to e_j}, \rho \to E[x], \overline{\rho_k\to e_k}\}}} 
                                                                         & \mbox{Commuting case-let}\\
@@
   \]}
-  \caption{Optimising transformations validated by Linear Core\label{fig:opttrans} (see Appendix~\ref{app:optimisations} for details)} 
+  \caption{Optimising transformations validated by Linear
+    Core\label{fig:opttrans} (for commuting case-let we
+      require $e_1$ not to capture the case binder, variables bound by the
+      pattern $\rho$ and the context $E[{-}]$ -- see Appendix~\ref{app:optimisations} for
+    details).} 
 \end{figure}

### 2024-02-12 icfp.tex
removed todo on case-let

@@
 \subsection{Optimisations preserve linearity\label{sec:optimisations-preserve-types-meta}}
-\todo[inline]{Check Commuting case-let in Fig.~\ref{fig:opttrans}!}
+%\todo[inline]{Check Commuting case-let in Fig.~\ref{fig:opttrans}!}
 A main goal of the Linear Core type system is 

### 2024-02-12 icfp.tex
typos

@@
 If $\G; \D',\xl \vdash e :\vp$
-then $\G[\D/x],\xD; \D,\D' \vdash e : \vp$, with fresh $\D$.
+then $\G[\D/x],\xD; \D,\D' \vdash e : \vp$ ($\Delta$ fresh).
 \end{restatable}
@@
 well-typed if the linear variable were instead $\D$-bound ($\xD$) with
-usage environment $\D$, with $\D$
-
-available in the linear context instead of
-the linear variable (and the unrestricted context updated accordingly).
+usage environment $\D$, with $\D$ available in the linear context (and the unrestricted context updated accordingly).
+
 \LinearDeltaRelationLemma

### 2024-02-12 icfp.tex
typos

@@
 
-\paragraph{$\Delta$-bound variables} Our development relies on the following characterization of $\D$-bound
-variables:
+\paragraph{$\Delta$-bound variables} % Our development relies on a characterization of $\D$-bound
+% variables
 A well-typed program with a linear variable ($\xl$) is equivalently

### 2024-02-12 icfp.tex
RM Edits Sec 2. prelude

@@
 %
-Consider the following program in a functional Haskell-like language,
-where a computation that closes the given \lstinline{handle} is bound to \lstinline{x} before the
-\lstinline{handle} is returned:
+For instance, in the following program in a Haskell-like language there are two
+syntactic occurrences of \emph{handle}, and a let-bound computation that closes
+this handle is bound to \lstinline{x} before the \lstinline{handle} is
+returned. However, if the program is evaluated lazily, the handle is \emph{not}
+closed because the computation that closes the handle is never evaluated (since
+\lstinline{x} goes unused).
 %
@@
 %
-
-In the code above, the \lstinline{handle} appears to be closed before
-returned, whereas in fact the handle will only be closed if the let bound
-computation is effectively run (i.e.,~evaluated).
-%
-The example illustrates that \emph{consuming} a resource is not necessarily
-synonymous with its syntactic occurrences, since depending on the evaluation strategy
-of the language the computation that closes the handle might not be
-executed, and if so, the \lstinline{handle} in that unused computation
-is not consumed.
-%
-Expanding on this, consider the above example program under
-under call-by-need semantics.  %With \emph{eager}
-% evaluation semantics,
+% Intuitively, a computation that depends on a linear resource to produce a
+% result consumes that resource iff the result is effectively computed; in
+% contrast, a computation that depends on a linear resource, but is never run,
+% will not consume that resource.
+%
+% Even though there are two occurrences of \lstinline{handle} in the program, and
+% it looks as though the \lstinline{handle} may be closed before returned, this function may
+% actually only \emph{consume} used once -- it depends on the evaluation strategy of the
+% language -- the \lstinline{handle} is closed iff the let bound computation is
+% evaluated.
+%
+%
+% Intuitively, \emph{consuming} a resource is not necessarily synonymous with its
+% syntactic occurrence. In the example, \emph{non-strict} evaluation means the
+% handle is never closed by the function.
+%
+%
+% Depending on the evaluation strategy of the language, a computation that closes
+% a handle might not be evaluated, and thus is not consumed.
+% With \emph{eager} evaluation semantics,
 % the let
@@
 % on the other hand, with \emph{call-by-need},
-The let
-bound expression will only be evaluated when the binding \lstinline{x} is needed. We return
-the \lstinline{handle} right away, and the let binding is forgotten as it cannot be
-used outside the scope of the function, so the handle is not closed by
-calling \lstinline{f}.
-Under the lens of \emph{call-by-need}, \emph{using} a resource in a
-let binding only results in the resource being \emph{consumed} if the binding
-itself is \emph{consumed}. We argue that a linear type system under
-\emph{call-by-need} evaluation should accept the above program, unlike a linear
-type system for the same program evaluated with \emph{call-by-value}, where
-the let-bound expression is always evaluated.
+%
+% Reconsider the \lstinline{handle} example program under
+% \emph{call-by-need} evaluation semantics:
+% The let bound expression is only evaluated if the binding \lstinline{x} is
+% needed. The function returns the \lstinline{handle} right away, and forgets the
+% let bound \lstinline{x} as its scope is exited, such that \lstinline{x} will
+% never be needed -- consequently, the \lstinline{handle} is never closed by the
+% function...
+%
+%
+%
 
 Intuitively, a computation that depends on a linear resource to produce a
-result consumes that resource iff the result is effectively computed; in
+result consumes that resource iff the result is effectively consumed; in
 contrast, a computation that depends on a linear resource, but is never run,
 will not consume that resource.
+Under the lens of \emph{non-strict} evaluation, this means that a syntactic
+occurrence of a resource in a let binding only entails consuming that resource
+iff the binding itself is consumed.
+%
+We argue that a linear type system for a language with non-strict evaluation
+semantics should accept the above program, unlike a linear type system for the
+same program if it were to be evaluated \emph{eagerly}.
 %
@@
 build an intuition for semantic linearity through examples of Core
-programs that are semantically linear but rejected by (the current) Core linear type
+programs that are semantically linear but are rejected by (the current) Core linear type
 system.
@@
 \colorbox{notyet}{\notyetcolorname} or \colorbox{limitation}{\limitationcolorname} background mark programs that are
-\emph{semantically} linear, but are not seen as linear by Core's (naive wrt laziness) linear type
+\emph{semantically} linear, but are not seen as linear by Core's (naive wrt lazy evaluation) linear type
 system. Notably, the linear type system we develop in this work accepts all

### 2024-02-12 icfp.tex
RM: Edits Sec 2. lets

@@
 f :: (a *' $\lolli$ '* a) -> Bool -> a *' $\lolli$ '* a
-f use bool x = let y = use x in case bool of
+f use bool x = let y = use x in
+  case bool of
     True -> x 
@@
 %
-This example uses \lstinline{x} directly in one
-branch and the let-bound \lstinline{y} in the other. Semantically, this program is linear because
-the linear resource \lstinline{x} ends up is used exactly once in both case
-alternatives, directly in the \lstinline{True} branch or via the
+This example uses \lstinline{x} directly in one branch and the let-bound
+\lstinline{y} in the other. Semantically, as opposed to syntactically, this
+program is linear (i.e. linear resources are used \emph{exactly once}) because
+the linear resource \lstinline{x} is used exactly once in both case
+alternatives, either directly in the \lstinline{True} branch, or \emph{via} the
 let-bound \lstinline{y} in the \lstinline{False} branch.
 
-Shifting our focus from not using a let binding to using it
-more than once.
-Despite the binder body only being evaluated
-once due to call-by-need, and thus its resources being only used once to compute a result, we can still
-only consume the result of the computation once -- perhaps surprisingly, as the
+% Shifting our focus from not using a let binding to using it
+% more than once.
+On the other hand, let us consider a let-binding \lstinline{y} that is used
+twice, and that binds an expression that depends on a linear resource
+\lstinline{x}.
+Despite the binder body only being evaluated once due to call-by-need, and thus
+its resources being only used once to compute a result, we can still only
+consume the result of the computation once -- perhaps surprisingly, as the
 perception so far is that ``resources are consumed during computation'' and
@@
 shared. Indeed, if the result of the computation involving the linear resource
-was, e.g., an unrestricted integer, then sharing the result would not involve consuming the
-resource more than once.
+was insteaed, e.g., an unrestricted integer, then sharing the result would not
+involve consuming the resource more than once.
 % ROMES:TODO:!!!!
@@
 %
-Concretely, the result of evaluating a let binder body using linear resources, if computed, must be
-consumed exactly once, or, otherwise, we risk discarding or duplicating said resources.
+The result of evaluating a let-binder's body which depends on linear resources,
+if computed, must be consumed exactly once, or, otherwise, we risk discarding
+or duplicating (the parts of) the resources that were captured in the result.
 
@@
 either -- never consuming \incode{x}. We use this example to highlight that even for
-let bound variables, the syntactic occurrence of a variable isn't enough to
+let-bound variables, the syntactic occurrence of a variable isn't enough to
 determine whether it is used. Instead, we ought to think of uses of \incode{y} as

### 2024-02-12 icfp.tex
RM edits recursive lets and cases
Missing: case of variable

@@
 In \incode{go}'s body, \incode{x} is used directly in one branch and indirectly in the
-other, by recursively calling \incode{go} (which we know will result in using \incode{x}
-linearly).
+other (by recursing, which we know will result in using \incode{x} linearly).
 % ROMES:TODO:!!
@@
 semantic linearity insights developed thus far given that
-\emph{case expressions drive evaluation} and semantic linearity can only be
+\emph{case expressions drive evaluation}, and semantic linearity can only be
 understood in terms of how expressions are evaluated.
@@
         % \item
-        To consume a value of an algebraic datatype exactly once,
+        to consume a value of an algebraic datatype exactly once,
         pattern-match on it, and consume all its linear components exactly once.
@@
 f6 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
-f6 x y use = case (x,y) of { (a,b) -> use a b }
+f6 x y use = case (x,y) of z { (a,b) -> use a b }
 \end{lstlisting}
@@
 Unlike when the scrutinee was in WHNF, we can no longer use $x,y$ in the case
-alternatives, but we \emph{must} still use either the case binder $z$ or the linear
+alternatives, but we must still use either the case binder $z$ or the linear
 pattern variables $a,b$. %  e.g. it would be quite disastrous if any of the following typechecked:
@@
 the program in Figure~\ref{fig:nolin} is semantically linear.
-
+%
 \begin{figure}[t]
@@
 \end{figure}
-
+%
 A second example of an unrestricted pattern, where \incode{K2} has no fields and \incode{K1}
@@
 \emph{semantically} linear (despite not being seen as so by our system).
-%
-%
-%
-This particular example might seem unrealistic due to the pattern
-match on a known constructor, but we note that during the
+
+Many of these examples might seem unrealistic due to the pattern
+match on a known constructor, but we note that, during the
 transformations programs undergo in an optimising compiler, many
-programs such as this naturally occur (e.g.,~if the definition of a
+such programs naturally occur (e.g.,~if the definition of a
 function is inlined in the scrutinee).

### 2024-02-12 icfp.tex
RM edits end Sec. 2

@@
 
-Finally, we consider default case alternatives, also known as wildcards
-(written $\_$), in the presence of linearity: matching against the wildcard
-doesn't provide new information, so linearity is seen as before but without
-fully consuming the scrutinee linear resources (in non-linear patterns) nor
-binding new linear resources (in linear patterns). In short, if the scrutinee
-is in WHNF, we can either use the resources from the scrutinee or the case
-binder in that alternative. If the scrutinee is not in WHNF, we \emph{must} use
-the case binder, as it's the only way to linearly consume the result of
-evaluating the scrutinee. We also discuss the special case of a case
-expression scrutinizing a variable $x$:
+As for \emph{default} case alternatives, also known as \emph{wildcards}
+(written $\_$): matching a wildcard doesn't provide new linearity information,
+unlike matching against a pattern with or without linear components, however,
+the scrutinee will still be evaluated to WHNF.
+%
+% so linearity is seen as before but without
+% fully consuming the scrutinee linear resources (in non-linear patterns) nor
+% binding new linear resources (in linear patterns).
+%
+Briefly, when matching a wildcard, if the scrutinee is already in WHNF, we can
+either use the resources from the scrutinee directly, or the case binder in
+that alternative. If the scrutinee is not in WHNF, we \emph{must} use the case
+binder, as it's the only way to linearly consume the result of evaluating the
+scrutinee.
+
+Finally, we discuss case expressions which scrutinize a
+single variable $x$, e.g.:
 $
@@
 $. 
-It might seem as though the program is linear:
+It might seem as though this function is linear:
 %begin{itemize}
@@
 % \item
-if $x$ is not in WHNF, then scrutinizing it evaluates it to WHNF, and
-returning $x$ returns the result of evaluating $x$ that still had to be
-consumed exactly once.
+if $x$ is not in WHNF, then scrutinizing it evaluates it to WHNF, and returning
+$x$ returns the result of evaluating $x$ that still had to be consumed exactly
+once.
 %\end{itemize}
-However, in practice, it depends on the evaluation strategy. If linear function
-applications are $\beta$-reduced \emph{call-by-name} (a common practice, as
-linear functions use their argument exactly once), and the above function is
-considered linear, then an application might duplicate linear resources during
-evaluation. For example:
+However, in practice, this program's linearity can only be determined by
+further branching down on the non-strict evaluation strategy. If linear
+function applications, in general, are $\beta$-reduced \emph{call-by-name} (a desireable
+practice, as linear functions are known to use their argument exactly once)
+% and thus call-by-name reduction will definitely not duplicate computation
+while the above function is considered linear, then such an application might
+duplicate linear resources during evaluation. For instance:
 \[
@@
 \]
-Therefore, the type system we present in the next section, which evaluates
-linear applications call-by-name, does not accept the above program as linear.
-This interaction between evaluation and linearity
-comes up in the type preservation proof of our program, and is again explored
-with the reverse binder swap transformation in
+On the other hand, if linear function applications are reduced
+\emph{call-by-need}, the above example would instead be semantically linear
+because the result of freeing $x$ would be bound to some $y$ first.
+%
+The type system we present in the next section, which evaluates linear
+applications call-by-name, does not accept the above function scrutinizing a
+variable as linear. 
+This interaction between evaluation and linearity comes up in the type
+preservation proof of our program, and is again explored with the reverse
+binder swap transformation in
 Section~\ref{sec:optimisations-preserve-types-meta}.

### 2024-02-12 icfp.tex
anonymization

@@
 optimisations. Despite our initiative in this
-direction\footnote{https://gitlab.haskell.org/ghc/ghc/-/issues/23218}, we leave
+direction\footnote{[Link removed due to anonymization]}%\footnote{https://gitlab.haskell.org/ghc/ghc/-/issues/23218}
+, we leave
 this as future work.

### 2024-02-13 icfp.tex
RM edits until 3.3

@@
 contrast to \emph{syntactic} linearity in Haskell, by example
-($\S$~\ref{sec:linearity-semantically}).
+(Section~\ref{sec:linearity-semantically}).
 
@@
 semantically in this context
-($\S$~\ref{sec:main:linear-core}).
+(Section~\ref{sec:main:linear-core}).
 
@@
 fail to type check in Core, are validated by Linear
-Core ($\S$~\ref{sec:main:metatheory}).
+Core (Section~\ref{sec:main:metatheory}).
 
@@
 it accepts the programs resulting from transformations in libraries such as
-\texttt{linear-base} ($\S$~\ref{sec:discuss:implementation}).
+\texttt{linear-base} (Section~\ref{sec:discuss:implementation}).
 
@@
 and multiplicity polymorphism introduced by Linear
-Haskell~\cite{cite:linearhaskell}, featuring algebraic datatypes, case
-expressions, recursive let bindings, and multiplicity polymorphism.
-(i.e.,~all of Core's features with the exception of equality
-coercions).  We discuss multiplicity coercions as future work in
+Haskell~\cite{cite:linearhaskell} with all the key features from GHC's Core
+language (algebraic datatypes, case expressions, and recursive lazy let bindings)
+except for equality coercions, which we discuss as future work in
 Section~\ref{sec:future-work}.
 %
-
+%
 Linear Core makes precise the various insights discussed in the
-previous chapter by crystallizing them together in a linear type system for which we
+previous section by crystallizing them together in a linear type system for which we
 prove soundness via the standard type preservation and progress theorems. Crucially,
@@
 %
-% Besides type safety, we prove that multiple optimising Core-to-Core
-% transformations preserve linearity in Linear Core. These same transformations
-% don't preserve linearity under Core's current type system. As far as we know,
-% we are the first to prove optimisations preserve types in a non-strict linear
-% language.
+Besides type safety, we prove that multiple optimising Core-to-Core
+transformations preserve linearity in Linear Core, while violating linearity
+under Core's current type system.
+%
+% As far as we know, we are the first to prove optimisations preserve types in
+% a non-strict linear language.
 
@@
 longer available but haven't yet been fully consumed
-(Section~\ref{sec:lc-case-exps}). Additionally, we introduce tagged
-resources to split the resources between the pattern-bound variables
-as usage environments, encoding that pattern variables jointly
-``finish consuming'' the scrutinee resources.
+(Section~\ref{sec:lc-case-exps}). Additionally, we introduce tags to split
+resources between the pattern-bound variables, %^as usage environments
+encoding that all linear components of a pattern must be used.
 % 
@@
 We present Linear Core's syntax and type system incrementally, starting with the
-judgements and base linear calculi rules ($\S~\ref{sec:base-calculi}$).
+judgements and base linear calculi rules (Section~\ref{sec:base-calculi}).
 %
 Then, usage environments, the rule for $\Delta$-bound variables, and rules for
-(recursive) let bindings ($\S~\ref{sec:usage-environments}$).
+(recursive) let bindings (Section~\ref{sec:usage-environments}).
 %
@@
 along with the key insights to do so, namely branching on WHNF-ness of
-scrutinee, proof irrelevant resources, and tagged variables
-($\S~\ref{sec:lc-case-exps}$).
+scrutinee, proof irrelevant resources, and tagging (Section~\ref{sec:lc-case-exps}).
 
@@
 Datatype declarations $\datatype{T~\overline{p}}{\overline{K:\overline{\sigma
-\to_\pi}~T~\overline{p} }}$ involve the name of the type being declared $T$
+\to_\pi}~T~\overline{p} }}$ have the name of the type being declared $T$
 parametrized over multiplicity variables $\ov{p}$, and a set of the data
@@
 % mutable heap to store \emph{thunks} and the values they are overwritten with.
-Furthermore, linear function applications, even in a \emph{call-by-need} system, are
-usually reduced \emph{call-by-name} as the function argument is guaranteed to
+Furthermore, linear function applications, even in a \emph{call-by-need} system, can
+usually be reduced \emph{call-by-name} as the function argument is guaranteed to
 be used exactly once (thus avoiding unnecessarily allocating memory on the heap
@@
 %
-
+%
 The core rules of the calculus for, multiplicity and term,
@@
 % %
-
+%
 Typing linear and unrestricted function applications separately is less general
@@
 semiring) the multiplicities of the resources used to type the argument by $\pi$, however, our objective of typing semantic
-linearity does not benefit much from doing so and so we opt for a
+linearity does not benefit much from doing so and therefore we opt for a
 simpler design. 
@@
 $\forall p.~\s$, i.e. a type universally quantified over a multiplicity
-variable $p$. We note that, in the body of the abstraction, function types annotated
-with a $p$ variable and datatype fields with multiplicity $p$ are typed as
-thought they are linear functions and linear fields, because $p$ can be
-instantiated at both $\omega$ and $1$.
+variable $p$. Again, conservatively, in the body of the abstraction, function
+types annotated with a $p$ variable and datatype fields with multiplicity $p$
+are typed as thought they are linear functions and linear fields, because $p$
+can be instantiated at both $\omega$ and $1$.
 \[
@@
 bindings do not consume the resources required by the bound expression when
-defined, but rather when the bindings themselves are fully consumed.
+defined, but rather when the bindings themselves are consumed.
 Specifically, we annotate so-called $\Delta$-bound variables with a \emph{usage

### 2024-02-13 icfp.tex
RM edits 3.3

@@
 Specifically, we annotate so-called $\Delta$-bound variables with a \emph{usage
-environment} to denote that consuming these variables equates to consuming the
-resources in the annotated usage environment $\D$ (i.e.,~essentially a multiset of linear resources).
-%
-$\Delta$-bound
-variables are introduced by several of constructs: (recursive) let
-binders, case binders, and case pattern variables. In the following example, as
-per the insights developed in
-Section~\ref{sec:semantic-linearity-examples}, the resources required to
-typecheck the body of the binder $u$, $x$ and $y$, are only used if
-the let-bound variable
-$u$ is consumed in the let-body $e$.  Accordingly, the usage environment of
-the let-bound $u$ is $\{x,y\}$:
+environment} to denote that consuming these $\D$-variables equates to consuming the
+resources in the annotated usage environment $\D$ (which is essentially a
+multiset of linear resources).
+%
+$\Delta$-bound variables are introduced by several constructs as (recursive)
+let binders, case binders, and case pattern variables.
+%
+For example, the let-binder $u$ in the following program is annotated with a
+usage environment $\{x,y\}$, the resources that are used in its body. Thus, in
+the expression $e$, consuming $u$ will equate to consuming the linear resources
+$x$ and $y$:
+% , as per the insights developed in
+% Section~\ref{sec:semantic-linearity-examples},
 \[
@@
 
-Finally, we note that usage environments bear a strong resemblance to the linear
-typing environments in the main typing judgement,
-i.e. the environment with the linear resources required to type an expression.
-%
-In fact, usage environments and linear typing contexts differ only in that the
-former are used to annotate variables, while the latter used to type
-expressions.
+% Finally, we note that usage environments bear a strong resemblance to the linear
+% typing environments in the main typing judgement,
+% i.e. the environment with the linear resources required to type an expression.
+%
+% In fact, usage environments and linear typing contexts differ only in that the
+% former used to annotate variables, while the latter used to type
+% expressions.
+%
 % Yet, this distinction is slightly blurred after introducing how
@@
 resources in $\Delta$ to type $u$ makes them unavailable in the
-context of any other derivation. Therefore, expressions using these
-resources a second time, directly, or indirectly through the same (or
-other) usage environment, are ill-typed, as the resources are already
-allocated to the derivation of $u$. Similarly, (3) also follows from
-(1), because if the linear resources aren't consumed in the
-$\Delta$-var derivation, they must be consumed in an alternative
-derivation (or otherwise the expression is ill-typed).
+context of any other derivation.
+% Therefore, expressions using these
+% resources a second time, directly, or indirectly through the same (or
+% other) usage environment, are ill-typed, as the resources are already
+% allocated to the derivation of $u$.
+Similarly, (3) also follows from (1), because if the linear resources aren't
+consumed in the $\Delta$-var derivation, they must be consumed in an
+alternative derivation (or otherwise the expression is ill-typed).
 %

### 2024-02-13 icfp.tex
RM edits 3.4

@@
 Let-bound variables are the canonical example of a
-$\Delta$-bound variable, that is, let-bound variables bind expressions in which
-the resources required to type them are consumed lazily rather than eagerly.
+$\Delta$-bound variable.
+% , that is, let-bound variables bind expressions in which
+% the resources required to type them are consumed lazily rather than eagerly.
 %
@@
 environment to type the let body $e'$, alongside the unrestricted $x$ binding
-annotated with the usage environment $\D$. Ultimately, the resources being
-available in $e'$ reflects the fact that typing a lazily bound expression
-doesn't consume resources, and the binding $x$ being $\D$-bound reflects that
-its usage entails consuming the resources $\D$ the expression it binds depends on.
+annotated with the usage environment $\D$. In essence, the resources of $e$ being
+available in $e'$ reflects how typing the body of a lazy binder
+doesn't consume resources, and the binding $x$ being an unrestricted
+$\D$-variable ensures that using it will consume the resources $\D$ the
+expression it binds depends on.
 % \todo[inline]{Let bindings are hard, if they are used then we use resources. If

### 2024-02-15 icfp.tex
RM edits Sec. 3.5

@@
 % themselves, and we may have more than one binding.
-In our system, groups of
-recursive let bindings are always assumed to be strongly connected, that is,
-all the bindings in a recursive let group are mutually recursive in the sense
-that they all (transitively) depend on one another.
-%
-As before, recursive let bindings bind expressions \emph{lazily}, so they similarly
-introduce a $\D$-variable for each binding, and resources required to type the
+%
+Recursive let bindings also bind expressions \emph{lazily}, so, similarly, we
+introduce a $\D$-variable for each binding. Resources required to type the
 let-bindings are still available in the body of the let, to later be consumed
-via $\D$-variables or directly, if the let-bindings are unused.
-%
-However, as discussed in Section
-\ref{sec:semantic-linearity-examples:recursive-lets}, we must consider how
-recursive uses of a binder in its own definition entails consuming all resources otherwise
-required to type the binder's body.
-%, i.e. a least upper bound.
-
-Extrapolating to a strongly-connected group of recursive bindings, (mutually)
-recursive uses of binders entail consuming all resources
-required to type those binders. By definition, those binders in turn
-recursively use binders that used them, and thus all the resources otherwise
-required to type them.
-%
-Ultimately, all binders in a strongly-connected group of mutually recursive let
+via $\D$-variables, or, directly, if the let-bindings are unused.
+%
+Additionally, as seen in Section
+\ref{sec:semantic-linearity-examples:recursive-lets}, recursive uses of a
+binder in its own definition entail consuming all resources otherwise required
+to type the binder's body.
+
+As for groups of recursive let bindings, we recall that, in our
+system, all recursive let groups are strongly-connected -- all bindings in a
+recursive let group are mutually recursive in the sense that they all
+(transitively) depend on one another.
+%
+% In light of it, (mutually) recursive uses of binders entail consuming all the
+% resources required to type all of the binders.
+%
+By definition of strongly-connected, any binder in the mutually recursive group
+may end up using any binder from the group. Accordingly, occurrences of any
+such binder must be typed conservatively, using up the sum of resources needed
+to type each of the binders,
+%
+i.e., all binders in a strongly-connected group of mutually recursive let
 bindings have to be typed with the same linear resources (which are the least
@@
 
-The typing rule for recursive groups of bindings leverages our assumption that
-all recursive let bindings are strongly connected and the observation
-that every binder in a strongly connected group of recursive bindings is typed
-with the same linear context. Consequently, all bindings of the recursive group
-are introduced as $\D$-vars with the same $\D$ environment -- using any one of
-the bindings in a recursive group entails consuming all resources required to
-type that same group, which is also why we can use the same linear resources to
-type each binder:
+The typing rule for recursive groups of bindings leverages both the assumption
+that all recursive let bindings are strongly connected and the observation that
+every binder in a strongly connected group of recursive bindings is typed with
+the same linear context. Consequently, all bindings of the recursive group are
+introduced as $\D$-vars with the same $\D$ environment -- using any one of the
+bindings in a recursive group entails consuming all resources required to type
+that same group, which is also why we can use the same linear resources to type
+each binder:
 \[
@@
 %
-Inference of usage environments for recursive binding groups bears some
-resemblance to the inference of principle types for recursive bindings
-traditionally achieved through the Hindley–Milner inference
+However, we note that inference of usage environments for recursive binding
+groups bears some resemblance to the inference of principle types for recursive
+bindings traditionally achieved through the Hindley–Milner inference
 algorithm~\cite{DBLP:conf/popl/DamasM82}. % , there might be an opportunity to

### 2024-02-15 icfp.tex
RM edits until 3.6

@@
 a case expression \emph{evaluates its scrutinee} to Weak Head Normal
-Form (WHNF)~\cite{10.5555/1096899}, \emph{then} selects the case
+Form (WHNF)~\cite{10.5555/1096899}, and \emph{then} selects the case
 alternative corresponding to the pattern matching the WHNF of the
@@
 Accordingly, these sub-expressions might still depend on linear resources to be
-well-typed (these resources will be consumed when the expression is evaluated).
-
-As will be made clear in later sections, we need to devise a specialized typing
+well-typed (these resources will be consumed when the expressions are evaluated).
+
+As will be made clear in this section, we need to devise a specialized typing
 judgement for scrutinees that is able to distinguish between terms
@@
 the main typing judgement and trivially ``outputs'' the whole $\D$ environment,
-as there is always only a single sub-expression in lambdas (the lambda body).
-
-Recall that the operational semantics encode the evaluation of case
-expressions such that 
+as there is always only a single sub-expression in lambda abstractions (the body).
+
+% Precisamos disto?
 % %
@@
 % %
-when a scrutinee $K~\ov{e}$ matches a constructor pattern $K~\ov{\x[\pi]}$,
-evaluation proceeds in the case alternative corresponding to the matching
-pattern, with occurrences of $\ov{x}$ being substituted by $\ov{e}$, and
-occurrences of the case binder $z$ substituted by the whole scrutinee
-$K~\ov{e}$. Constructors and lambda expressions otherwise match the wildcard
-pattern whose alternative body is evaluated only substituting the case binder by the
-scrutinee.
+% Recall how the operational semantics encode the evaluation of case
+% expressions such that,
+% when a scrutinee $K~\ov{e}$ matches a constructor pattern $K~\ov{\x[\pi]}$,
+% evaluation proceeds in the case alternative corresponding to the matching
+% pattern, with occurrences of $\ov{x}$ being substituted by $\ov{e}$ and
+% occurrences of the case binder $z$ substituted by the whole scrutinee
+% $K~\ov{e}$. Constructors and lambda expressions may otherwise match match the wildcard
+% pattern, resuming evaluation on the alternative body only substituting the case
+% binder by the scrutinee.
 
@@
 available in the linear typing environment.
-
+%
 The $\vdash_{alt}$ judgement is 
@@
 to typecheck the scrutinee sub-expressions,
-the name of the case binder $z$.
+and the name of the case binder $z$.
 
@@
 %
-This ensures that when we match on an unrestricted pattern we no longer need to
-consume the scrutinee resources. Otherwise, for example,
-$\ccase{K_1~x}{\{K_2 \to K_2,K_1~y \to K_1~y\}}$ would not be well-typed since
-the resource $x$ is not consumed in the first branch.
+Reactively deleting the resources that typed the scrutinee from the linear
+environment ensures that when we match on an unrestricted pattern there are no
+remaining linear resources to consume. Otherwise, e.g., $\ccase{K_1~x}{\{K_2 \to
+K_2,K_1~y \to K_1~y\}}$ would not be well-typed since the resource $x$ would
+be left \emph{un}-consumed in the first alternative, even though we know by
+matching on $K_1$ that it has already been consumed.
 %
@@
 unrestricted expression, the case binder too may be used unrestrictedly, which
-we allow by making its usage environment empty.
-
-It might seem as though deleting the resources from the environment in this
-rule is necessary to guarantee a resource is not used after it is consumed.
-%
-However, let us consider two situations -- pattern matches in a case
-expression whose scrutinee is in WHNF, and matches on a case expression whose
-scrutinee is \emph{not} in WHNF:
-%
-%\begin{enumerate}
-When the scrutinee is in WHNF, it is either an unrestricted expression
-against which any match will only introduce unrestricted variables, or an
-expression that depends on linear resources. The former trivially allows
-any resource from the scrutinee in the alternatives as well. The latter is
-further divided: 
-%\begin{enumerate}
-%
-%\item
-The pattern is unrestricted while the
-scrutinee is not, so entering this branch is impossible as long as the case
-expression is well-typed; by contradiction, the linear resources from the
-scrutinee could occur unrestrictedly in that branch.
-%
-For uniformity, we type such alternatives as those for scrutinees that
-are not in WHNF;
-%
-The pattern is linear and matches the scrutinee, in which case the
-$AltN_{\textrm{WHNF}}$ is applicable instead of $Alt0$;
-%
-The pattern is linear but does not match the scrutinee and so any
-resource could be used in such alternatives. For uniformity,
-it is also typed as though the scrutinee were not in WHNF.
-
-If the scrutinee is not in WHNF, the resources occurring in the
-scrutinee will be consumed when evaluation occurs. Therefore, the resources
-used in the scrutinee cannot occur in the alternative body
-(e.g. $x$ cannot occur in $e$ in $\ccase{close~x}{\{K_1 \to e\}}$)
--- regardless of the pattern.
-%
-% In fact, the resources from a scrutinee that is not in weak head normal form
-% cannot occur in any of the alternatives, even ones matching on constructors
-% with linear components, as the resources may have been consumed when evaluating
-% the expression to weak head normal form.
-%
-We guarantee resources from a scrutinee that is not in WHNF
-cannot occur/directly be used in any case alternative in our rule for typing
-cases not in WHNF, which we introduce below.
+we allow by emptying its usage environment.
+
+%It might seem as though deleting the resources from the environment in $\textsc{Alt}_0$
+%is necessary to guarantee a resource is not used after it is consumed.
+%%
+%However, let us consider two situations -- pattern matches in a case
+%expression whose scrutinee is in WHNF, and matches on a case expression whose
+%scrutinee is \emph{not} in WHNF:
+%%
+%%\begin{enumerate}
+%When the scrutinee is in WHNF, it is either an unrestricted expression
+%against which any match will only introduce unrestricted variables, or an
+%expression that depends on linear resources. The former trivially allows
+%any resource from the scrutinee in the alternatives as well. The latter is
+%further divided: 
+%%\begin{enumerate}
+%%
+%%\item
+%The pattern is unrestricted while the
+%scrutinee is not, so entering this branch is impossible as long as the case
+%expression is well-typed; by contradiction, the linear resources from the
+%scrutinee could occur unrestrictedly in that branch.
+%%
+%For uniformity, we type such alternatives as those for scrutinees that
+%are not in WHNF;
+%%
+%The pattern is linear and matches the scrutinee, in which case the
+%$AltN_{\textrm{WHNF}}$ is applicable instead of $Alt0$;
+%%
+%The pattern is linear but does not match the scrutinee and so any
+%resource could be used in such alternatives. For uniformity,
+%it is also typed as though the scrutinee were not in WHNF.
+
+%If the scrutinee is not in WHNF, the resources occurring in the
+%scrutinee will be consumed when evaluation occurs. Therefore, the resources
+%used in the scrutinee cannot occur in the alternative body
+%(e.g. $x$ cannot occur in $e$ in $\ccase{close~x}{\{K_1 \to e\}}$)
+%-- regardless of the pattern.
+%%
+%% In fact, the resources from a scrutinee that is not in weak head normal form
+%% cannot occur in any of the alternatives, even ones matching on constructors
+%% with linear components, as the resources may have been consumed when evaluating
+%% the expression to weak head normal form.
+%%
+%We guarantee resources from a scrutinee that is not in WHNF
+%cannot occur/directly be used in any case alternative in our rule for typing
+%cases not in WHNF, which we introduce below.
 
@@
 
-% \todo[inline]{The trick here is to separate the case rules into two separate
-% rules, one that fires when the scrutinee is in WHNF, the other when it isn't.}
-
-\todo[inline]{Considerar alguns exemplos aqui para ilustrar esta
-  ``análise de casos''}
-
 \subsubsection{Pattern matching on non-WHNF terms via proof irrelevant resources}
 
-% \todo[inline]{O que é que eu vou fazer aqui? O que falta é conseguir tratar uma
-% case expression de forma flexivel o suficiente. Explorar o scrutinee estar em
-% WHNF ou não, agora é preciso usar essas ideias em typing rules.  Só preciso de
-% dizer que para tipificar o caso em que não está em WHNF de forma rigorosa num
-% sistema de tipos: Manter resources in scope pq é preciso consumir; Mas não
-% podem ser usadas diretamente}
-
-\todo[inline]{Todo este paragrafo que esta commented out parece-me redundante}
-% Resources used in a scrutinee that is not in WHNF must
-% not be used in the case alternatives.
-% %
-% %
-% However, it is not sufficient to evaluate the scrutinee to weak head normal
-% form to \emph{fully} consume all resources used in the scrutinee, since
-% sub-expressions such as constructor arguments will be left unevaluated. To
-% \emph{fully} consume all resources occurring in the scrutinee, the scrutinee
-% must be evaluated either to \emph{normal form} or such that all linear components of
-% the scrutinee are fully evaluated, as witnessed by the $Alt0$ rule above. In short,
-% for a case expression whose scrutinee is not in WHNF, the scrutinee
-% resources must \emph{not} be used directly in the case alternatives,
-% but the result of evaluating the scrutinee to WHNF must still be
-% consumed, as all sub-expressions of the scrutinee remain unevaluated and must
-% be consumed. Since the scrutinee resources cannot be consumed directly, they must be
-% consumed indirectly through $\D$-variables, namely, either the case binder, or
-% the linear pattern-bound variables introduced in the alternative.
-%ATE AQUI
+Resources used to type a scrutinee that is not in WHNF \emph{must not} be used in
+the case alternatives.
+%
+However, evaluating the scrutinee to WHNF is not sufficient to
+consume all resources used by the scrutinee, since sub-expressions such
+as constructor arguments will be left unevaluated and may still contain
+(``parts of'') the linear resources. To \emph{fully} consume all
+resources occurring in the scrutinee, all linear components of the
+constructor must also be fully evaluated, as witnessed by the
+$\textsc{Alt}_0$ rule above.
+%
+In essence, to type cases scrutinizing expressions not in WHNF, we need to
+forbid the usage of the scrutinee resources while tracking that we necessarily
+``finish consuming'' them, either via the case binder, or by linearly using all
+linear pattern-bound variables introduced in the alternative.
+
 % We tackle this in due time, in the proof irrelevance section.
@@
 to encode linear resources that cannot be
-directly used (i.e.,~to which the $Var$ rule is not applicable). Proof irrelevant resources
+directly used (i.e.,~to which the $\textsc{Var}$ rule is not applicable). Proof irrelevant resources
 are linear resources in all other senses, meaning they must be used
@@
 discarded nor used directly, they have to be consumed \emph{indirectly} --
-by $\D$-bound variables.
-
-To type a case expression whose scrutinee is in weak head normal form, we
+by $\D$-bound variables, namely, the case binder, or the linear pattern-bound variables.
+
+Hence, to type a case expression whose scrutinee is in weak head normal form, we
 type the scrutinee with linear resources $\D$ and type the case
@@
 % %
-Alternatives not matching the scrutinee, as mentioned in the discussion of the
-$Alt0$ rule, could use resources arbitrarily as they will never be executed,
-however, we uniformly treat non-matching alternatives as if the scrutinee were
-not in WHNF. Having introduced proof irrelevant resources, we can now
-present the correct $Case_\textrm{WHNF}$ rule:
+Alternatives not matching the scrutinee could use resources arbitrarily as they
+will never be executed, however, we uniformly treat non-matching alternatives
+as if the scrutinee were not in WHNF. Having introduced proof irrelevant
+resources, we can now present the correct $Case_\textrm{WHNF}$ rule:
 \[

### 2024-02-17 icfp.tex
RM edits 3.6.3

@@
 
-Unlike with scrutinees in WHNF, the resources used by a
-scrutinee not in WHNF do not necessarily match those used by each
-sub-expression of the expression evaluated to WHNF.
-%
-Therefore, there is no direct mapping between the usage environments of the
-linear pattern-bound variables and the resources used in the scrutinee.
+As opposed to scrutinees in WHNF, where the resources used to type a scrutinee
+can be cleanly divided amongst the various sub-expressions of e.g. the
+constructor and henceforth each pattern variable, there is no direct mapping
+between the resources typing a scrutinee \emph{not in WHNF} and the usage
+environments of pattern variables in any alternative.
+
+%
+% That is, there is no direct mapping between the usage environments of the
+% linear pattern-bound variables and the resources used in the scrutinee.
 % \todo{WAIT, there never was!
@@
 %
-We introduce \emph{tagged resources} to guarantee all linearly-bound pattern
-variables are jointly used to consume all resources occurring in the
-environment (in alternative to the case binder), or not at all. Given linear
-resources $[\D_s]$ used to type a scrutinee, and a pattern
-$K~\ov{x_\omega},\ov{y_i}$ with $i$ linear components, we assign a usage
-environment $\D_i$ to each linear pattern variable where, $\D_i$ is obtained from the
-scrutinee environment \emph{tagged} with the constructor name and linear-variable
-index $\lctag{\D_s}{K_i}$, and $\y[\D_i]$ is introduced in $\G$.
+We introduce \emph{tagged resources} as a means to enforce all pattern-bound
+variables for a scrutinee not in WHNF are either \emph{jointly} used to consume
+all resources occurring in the environment, or not at all (instead, the case
+binder may be used). Given linear resources $[\D_s]$ used to type a scrutinee,
+and a pattern $K~\ov{x_\omega},\ov{y_i}$ with $i$ linear components, we assign
+a usage environment $\D_i$ to each linear pattern variable where $\D_i$ is
+obtained from the scrutinee environment \emph{tagged} with the constructor name
+and linear-variable index $\lctag{\irr{\D_s}\!}{K_i}$. $y_i{:}_{\D_i}\s$ is
+introduced in $\G$, just like other $\D$-variables.
 \[
@@
 \]
-The tag consists of a constructor name $K$ and an index $i$ identifying the
-position of the pattern variable among all bound variables in that pattern.
-%
-The key idea is that a linear resource $x$ can be split into $n$ resources at a
-given constructor, where $n$ is the number of positional linear arguments of
-the constructor.
-%
-This is given by the Split rule. By assigning to each linear pattern
-variable a fragment of the scrutinee resources with a tag, we
-guarantee that all linear pattern variables are simultaneously used to
-consume all the scrutinee resources.
-For instance, in the term $\lambda x.~\ccase{x}{z~\{K~a~b\to
+%
+% The tag consists of a constructor name $K$ and an index $i$ identifying the
+% position of the pattern variable among all bound variables in that pattern.
+%
+Having tagged the resources in the usage environment of each pattern variable
+differently, we need now only to express that (1) the pattern $\D$-variables
+can be used (i.e., the tagged resources need be available in the linear
+environment in order for the $\textsc{Var}_\D$ rule to be applicable) and (2)
+if we consume one of the pattern variables with a non-empty usage environment,
+then \emph{all} pattern variables for linear components must also be consumed.
+
+The key idea to guarantee these is to provide a $\textsc{Split}$ rule that
+allows a linear resource $x{:}_1\s$ to be split into $n$ resources at a given
+constructor $K$, where $n$ is the number of linear components of the
+constructor, and each resource resulting from the split is \emph{tagged} with
+$K$ and the positional index of a linear component (i.e. $x{:}_1\s$ can be
+split into $\ov{\lctag{x{:}_1\s}{K_i}}^n$).
+%
+By assigning to each pattern variable a \emph{fragment} of the scrutinee
+resources (with a tag), we require the scrutinee resources to be
+$\textsc{Split}$ in order to use pattern variables at all.  Moreover, the
+remaining tagged resources cannot be used directly, yet need still be consumed
+exactly once. Thus, the choice to consume the tagged resources via usage
+environments of the other pattern variables is forced, ensuring that no
+variable for a constructor's linear component can go unused (the tagged environments are disjoint).
+
+For instance, in the term $\lambda x.~\ccase{f~x}{z~\{K~a~b\to
   (a,b)\}}$, where $x$ is a linear variable, the case alternative is
@@
 $\var[b][\lctag{\irr{x}}{K_2}]$, assuming both components of $K$ are linear.
-\todo[inline]{Previous explanation was very confusing, try to explain
-  what the split is doing in terms of the example!}
+%
+The occurrences of $a$ and $b$ are well-typed because we can first
+$\textsc{Split}$ $[x{:}_1\s]$ into $\lctag{[x{:}_1\s]}{K_1}, \lctag{[x{:}_1\s]}{K_2}$,
+%
 % since for any of the scrutinee
@@
 % have to be consumed through the other linear pattern-bound variables.
-
-%
-We note how $Split$ can be applied both to relevant and proof irrelevant linear resources in $\D$.
+%
+noting that $Split$ can be applied both to relevant and proof irrelevant linear
+resources in $\D$.
 

### 2024-02-17 icfp.tex
RM edit section on GHC plugin

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% {{{ Metatheory
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Metatheory of Linear Core\label{sec:main:metatheory}}
+
+The $\lambda^\pi_\D$ type system is sound: well-typed programs in Linear Core do not
+get \emph{stuck}. Besides type safety, we
+prove that multiple optimising transformations used in GHC preserve
+linearity in the sense of being well-typed in our system.
+%
+% Additionally, we state our assumptions that outline an isomorphism between
+% using a linear variable $\xl$ and a $\D$-variable $\xD$ that consumes existing
+% resources $\D$, for any $\D$.
+
+\renewcommand{\DeltaLinearRelationLemma}{
+\begin{restatable}[$\Delta$-bound to Linear]{lemma}{deltaone}\label{lem:deltaone}
+% Any $\Delta$-variable with a fully-irrelevant usage environment can be made a linear variable as long as its irrelevant environment is dropped from the linear environment (and the unrestricted env.
+is updated accordingly).\\
+If $\G,\x[\irr{\D}]; \irr{\D},\D' \vdash e : \vp$ 
+then $\G[x/\irr{\D}]; \D',\xl \vdash e :\vp$.
+\end{restatable}
+}
+
+\renewcommand{\LinearDeltaRelationLemma}{
+  \begin{restatable}[Linear to $\Delta$-bound]{lemma}{onedelta}\label{lem:onedelta}
+If $\G; \D',\xl \vdash e :\vp$
+then $\G[\D/x],\xD; \D,\D' \vdash e : \vp$ ($\Delta$ fresh).
+\end{restatable}
+}
+
+\renewcommand{\DeltaUnrestrictedRelationLemma}{
+\begin{restatable}[Unrestricted and $\Delta$-bound]{lemma}{undelta}\label{lem:undelta}
+%An unrestricted variable is equivalent to a $\D$-var with an empty usage environment.\\
+$\G,\xo; \D \vdash e : \vp$ iff $\G,\x[\cdot]; \D \vdash e : \vp$
+\end{restatable}
+}
+
+\paragraph{$\Delta$-bound variables} % Our development relies on a characterization of $\D$-bound
+% variables
+A well-typed program with a linear variable ($\xl$) is equivalently
+well-typed if the linear variable were instead $\D$-bound ($\xD$) with
+usage environment $\D$, with $\D$ available in the linear context (and the unrestricted context updated accordingly).
+
+\LinearDeltaRelationLemma
+
+\noindent Dually, any $\Delta$-bound variable with an irrelevant usage
+environment can be made into a linear variable as long as its
+(irrelevant) environment is dropped from the linear context.
+
+% Dually, a well-typed program with resources $\D$ and $\D$-bound
+% variable ($\xD$) is equivalently well-typed, as long as $\D$ is consumed
+% through the use of $x$, if $x$ is moved to the linear context, resources $\D$
+% are removed from the linear context, and occurrences of $\D$ whole in usage
+% environments are substituted by $x$ (occurrences of fragments of $\D$ in usage
+% environments are unimportant since $\D$ was consumed whole by $x$, not by any
+% of the fragment-using $\D$-vars).
+
+\DeltaLinearRelationLemma
+
+% Intuitively, a linear variable
+
+\noindent Finally, unrestricted and $\D$-bound
+variables with an empty usage environment are equivalent.
+
+\DeltaUnrestrictedRelationLemma
+
+\paragraph{Irrelevance}
+%
+As discussed, proof irrelevant resources can only be consumed
+indirectly, essentially encoding that the (non-WHNF) case scrutinee
+resources must be consumed through the case binder or the linear
+pattern-bound variables.
+%
+As a case expression is evaluated, the scrutinee will eventually
+reduce to WHNF,
+which must then be typed with rule $Case_{\textrm{WHNF}}$.
+%
+Crucially, the two case rules must be in harmony in the system, in the sense that
+case expressions typed using the $Case_{\textrm{Not WHNF}}$ rule must also be
+well-typed by the $Case_{\textrm{WHNF}}$ once the scrutinee is evaluated to WHNF.
+
+% The type preservation theorem states that a well-typed expression
+% remains well-typed in the presence of evaluation. Specifically, when the case
+% expression whose scrutinee is evaluated to WHNF is handled in the
+% preservation proof.
+
+% The \emph{Irrelevance} lemma is required to prove preservation for that
+% evaluation case. We need to prove that the alternatives of a case expression
+% typed with proof irrelevant resources are still well-typed when the proof
+% irrelevant resource is substituted by the scrutinee resources as it is evaluated to WHNF.
+% In this sense, the \emph{Irrelevance} lemma witnesses the soundness of typing a
+% case alternative with proof irrelevant resources in a certain context with respect to
+% typing the same expression with arbitrary resources (we note, however, typing
+% an alternative with proof irrelevant resources is not complete wrt using
+% arbitrary resources -- a counter example needs only to use a resource
+% directly).
+
+\WHNFConvSoundness
+
+The \emph{Irrelevance} lemma witnesses the soundness of typing a
+case alternative with proof irrelevant resources with respect to
+typing the same expression with arbitrary resources. However, typing
+a case alternative with proof irrelevant resources is not complete wrt using
+arbitrary resources -- a counter example needs only to use a resource
+directly.
+
+% \noindent Intuitively, the lemma holds since proof irrelevant resources can
+% only be used through the case binder or pattern-bound variables. If we
+% consistently replace the proof irrelevant resources both in the typing
+% environment and in the usage environments containing them, the expression
+% remains well-typed. We note that the proof irrelevant resources are always
+% unique when introduced in such a case alternative (we always take the scrutinee
+% environment to make irrelevant, and allow nested ``irrelevantness''), so the
+% case binder has the only occurrence of those resources in the $\G$ environment.
+% % (being somewhat akin to congruence).
+% %
+% The proof is given in Section~\ref{sec:proof:irrelevance}.
+
+\subsection{Type safety\label{sec:type-safety-meta}}
+
+We prove type safety of the Linear Core system via the standard type
+preservation and progress results. As is customary, we make use of multiple
+substitution lemmas, one for each kind of variable: unrestricted variables
+$\xo$, linear variables $\xl$, and $\D$-bound variables $\xD$.
+
+% We start with the auxiliary results, as we will make use of them in a select part of the preservation proof.
+
+\TypePreservationTheorem
+%
+\ProgressTheorem
+% \noindent Type preservation states that a well-typed expression $e$ that
+% evaluates to $e'$ remains well-typed under the same context:
+%
+Type preservation
+% ($\S$~\ref{sec:proof:type-preservation})
+follows by structural induction on the reductions $e \longrightarrow
+e'$ from the operational semantics. Most cases are straightforward, appealing to one or more
+substitution lemmas (see Appendix~\ref{sec:proof:type-preservation}). The most
+interesting case is that of case expressions whose scrutinee can be further
+evaluated -- we branch on whether the scrutinee becomes in WHNF and appeal to Lemma~\ref{lem:irrev} if so.
+%
+% This case guarantees that the separation of rules for treating scrutinees is
+% consistent, in the sense that a well-typed case expression with a scrutinee not
+% in WHNF remains well-typed after the scrutinee is evaluated to WHNF.
+%
+%
+%
+%\noindent
+% Progress states that the evaluation of a well-typed term does not block:
+% Similarly, progress is proved by induction on typing ($\S$~\ref{sec:proof:progress}).
+
+% \subsubsection{Substitution Lemmas}
+
+% The preservation and progress theorems depend on multiple substitution lemmas,
+% one for each kind of variable, as is standard.
+% % The proofs are given in Section~\ref{sec:proof:substitution-lemmas}.
+% % The ... themselves diverge from their common formulation, because
+% % $\D$-variables refer to linearly bound variables, so substitution must take into account
+
+% The linear substitution lemma states that a well-typed expression $e$ with a
+% linear variable $x$ of type $\s$ remains well-typed if
+% occurrences of $x$ in the $e$ are replaced by an expression $e'$ of the same
+% type $\s$, and occurrences of $x$ in the linear context and in usage
+% environments of $\D$-bound variables are replaced by the linear context $\D'$
+% used to type $e'$:
+
+% \LinearSubstitutionLemma
+
+% \noindent Where $\G[\D'/x]$ substitutes all occurrences of $x$ in the usage
+% environments of $\D$-variables in $\G$ by the linear variables in $\D'$.
+% % ($x$ couldn't appear anywhere besides usage environments of $\D$-bound variables, since $x$ is linear).
+
+% The substitution of the resource in the usage environments is illustrated
+% by the following example. Consider the term $\llet{y = use~x}{y}$ where $use$ and $x$ are free variables:
+% if we replace occurrences of $x$ by $e'$ (where $\G;\D \vdash e : \s$), then the
+% ``real'' usage environment of $y$ goes from $\{x\}$ to $\D$. If we don't update
+% the usage environment of $y$ accordingly, we'll ultimately be typing
+% $y{:}_{\{x\}}\vp$ with $\D$ instead of $x$, which is not valid.
+
+% The linear substitution lemma extends to case alternatives as well.
+% The lemma for substitution of linear variables in case alternatives is similar
+% to the linear substitution lemma, applied to the case alternative judgement.
+% % Is slightly more involved, in the sense that there are more environments in which the
+% % substitution $[\D/x]$ must be applied (for the same reason):
+% %
+% \LinearSubstitutionAltsLemma
+% %
+% \noindent We further require that the environment annotated in the case
+% alternative judgement, $\D_s$, is a subset of the environment used to type the
+% whole alternative $\D_s \subseteq \D$. In all occurrences of the alternative
+% judgement (in $Case_{\textrm{WHNF}}$ and $Case_{\textrm{Not WHNF}}$), the
+% environment annotating the alternative judgement is \emph{always} a subset of
+% the alternative environment.
+
+% The substitution lemma for unrestricted variables follows the usual
+% formulation, with the added restriction (common to linear type systems) that
+% the expression $e'$ that is going to substitute the unrestricted variable $x$
+% is typed on an empty linear environment:
+% %
+% \UnrestrictedSubstitutionLemma
+% %
+% \noindent Similarly, we also prove the substitution of unrestricted variables preserves types on an alternative case expression:
+% %
+% \UnrestrictedSubstitutionAltsLemma
+
+% Finally, we introduce the lemma stating that substitution of $\D$-bound
+% variables by expressions of the same type preserves the type of the original
+% expression.
+% %
+% What distinguishes this lemma from traditional substitution lemmas is that the
+% usage environment $\D$ of the variable $x$ being substituted by expression $e'$
+% must match exactly the typing environment $\D$ of $e'$ and the
+% environment of the original expression doesn't change with the substitution:
+% %
+% \DeltaSubstitutionLemma
+% %
+% \noindent Intuitively, if $x$ is well-typed with $\D$ in $e$, substituting $x$
+% by an expression $e'$ which is typed in the same environment $\D$ allows the
+% distribution of resources $\D,\D'$ used to type $e$ across sub-derivations to remain
+% unchanged. To prove the theorems, we don't need a ``stronger'' substitution of
+% $\D$-vars lemma (allowing arbitrary resources $\D''$ to type $e'$, as in other
+% substitution lemmas), as we only ever substitute $\D$-variables by expressions
+% whose typing environment matches the variables usage environment. However, it
+% is not obvious whether such a lemma is possible to prove for $\D$-variables
+% (e.g. let $\G;\D \vdash e :\s$ and $\G; \D' \vdash \llet{x = e'}{x}$, if we
+% substitute $e$ for $x$ the resources $\D'$ are no longer consumed).
+
+% The $\D$-substitution lemma on case alternatives reflects again that the typing
+% environment of the expression substitution the variable must match its usage
+% environment. We recall that $\D_s \subseteq \D,\D'$ states that the annotated
+% environment is always contained in the typing environment, which is true of all
+% occurrences of this judgement. An alternative formulation of this lemma could
+% instead explicitly list $\D_s$ as part of the typing environment for the same
+% effect:
+
+% \DeltaSubstitutionAltsLemma
+
+% The proofs for substitution lemmas of linear, unrestricted, and
+% $\D$-variables are available in Section~\ref{sec:proof:substitution-lemmas}.
+
+%TODO! Substitution of proof-irrelevant linear variables preserves typing. The
+%term always remains the same because $x$ cannot occur in any term, however, all
+%variables that refer to $x$ in their usage environment must now refer the usage env. of the substitee (e.g. $[x] => [\D]$).
+%This seems trivial to see correct, since all occurrences are in environments, so we get some equivalence similar to the one we need for the proof of Alt0.
+%
+%TODO: Multiplicity substitution preserves typing lemma
+%
+%TODO: Canonical forms lemma
+%
+%TODO: Corollary of $\Delta$-var subst. for $\ov{\Delta}$
+%
+%TODO: Constructor app typing:
+%If $\Gamma, \Delta \vdash K~\ov{e}$ and $K{:}\ov{\sigma\to\pi}~T~\ov{p} \in \Gamma$ and $\hasnolinearvars{\Gamma}$
+%then $\ov{\Gamma, \Delta_i \vdash e_i : \sigma_i}$
+
+\subsection{Optimisations preserve linearity\label{sec:optimisations-preserve-types-meta}}
+%\todo[inline]{Check Commuting case-let in Fig.~\ref{fig:opttrans}!}
+A main goal of the Linear Core type system is 
+to serve as a typed intermediate representation for optimising
+compilers of lazy languages with linearity. In light of this,
+we show that multiple optimising transformations are type
+preserving in Linear Core (i.e.,~preserve linearity semantically).
+
+% The optimising transformations proved sound wrt Linear Core in this section
+% have been previously explained and motivated in
+% Section~\ref{sec:core-to-core-transformations}.
+% %
+%\{\overline{\rho\to e'}\}
+\begin{figure}[t]
+
+  {\small
+  \[
+    \begin{array}{ll}
+   (\llet{\xD = e}{e'} )  \Longrightarrow \llet{\xD =       e}{e'[e/x]} &                    \mbox{Inlining}\\
+       (\lambda \x[\pi][\s].~e)~e'  \Longrightarrow e[e'/x]  &                \mbox{$\beta$-reduction}\\
+      (\ccase{(K~\ov{e})}{\z[\D][\s]~\{\overline{\rho_j\to e_j}, K~\ov{x} \to e_i, \overline{\rho_k\to e_k}\}} )\Longrightarrow  e_i\ov{[e/x]}[K~\ov{e}/z] &
+                                  \mbox{Case of known constr.}\\
+      (\ccase{(\ccase{e_c}{\z[\D]~\{\ov{\rho_{c_i}\to e_{c_i}}\}})}{\var[w][\irr{\D,\D'}][\s']~\{\ov{\rho_i\to e_i}\}})\\
+            \Longrightarrow 
+            %{{{
+      \ccase{e_c}{\z[\D]~\{\ov{\rho_{c_i} \to \ccase{e_{c_i}}{w~\{\ov{\rho_i\to e_i}\}}}\}}
+      & \mbox{Case-of-case}\\
+      (\lambda \y[\pi].~\llet{\xD = e}{e'})
+      \Longrightarrow
+      \llet{\xD=e}{\lambda\y[\pi].~e'} & \mbox{Commuting let-$\lambda$}\\
+      ((\llet{v = e}{b})~a) \Longrightarrow \llet{v = e}{b~a} &\mbox{Commuting let-app.}\\
+      (\ccase{(\llet{v = e}{b})}{\{\overline{\rho\to e'}\}}) \Longrightarrow
+      %{{{
+      \llet{v =  e}{\ccase{b}{\{\overline{\rho\to e'}\}}}
+                               &
+                                 \mbox{Commuting let-case}\\
+      (\ccase{e}{\{\overline{\rho_j\to e_j}, \rho \to E[e_1],
+      \overline{\rho_k\to e_k}\}} ) \qquad (\mbox{$x$ fresh}) \\
+      %{{{
+     \Longrightarrow \llet{x = e_1}{\ccase{e}{\{\overline{\rho_j\to e_j}, \rho \to E[x], \overline{\rho_k\to e_k}\}}} 
+                                                                        & \mbox{Commuting case-let}\\
+      (\llet{x = (\llet{v = e}{b})}{c}) \Longrightarrow \llet{v = e}{\llet{x = b}{c}}
+                                                                      & \mbox{Commuting let-let}\\
+      f \Longrightarrow \lambda x.(f~x) & \mbox{$\eta$-expansion}\\
+      \lambda x.(f~x) \Longrightarrow f & \mbox{$\eta$-reduction}\\
+      (\ccase{x}{z~\{\ov{\rho_i\to e_i}\}}) \Longrightarrow \ccase{x}{z~\{\ov{\rho_i\to  e_i[z/x]}\}}
+                                                                      & \mbox{Binder swap}\\
+      % (\ccase{x}{z~\{\ov{\rho_i\to e_i}\}} )\Longrightarrow \ccase{x}{z~\{\ov{\rho_i\to e_i[x/z]}\}}
+      %                                                                 & \mbox{Reverse binder swap}
+    \end{array}
+  \]}
+  \caption{Optimising transformations validated by Linear
+    Core\label{fig:opttrans} (for commuting case-let we
+      require $e_1$ not to capture the case binder, variables bound by the
+      pattern $\rho$ and the context $E[{-}]$ -- see Appendix~\ref{app:optimisations} for
+    details).} 
+\end{figure}
+
+We describe Core-to-Core transformations as $e_1 \Longrightarrow e_2$,
+where $e_1$ is an arbitrary \emph{well-typed} expression
+matching a certain shape (or of a certain type), which is
+transformed into expression $e_2$. Validating a transformation
+entails showing that $e_2$ is well-typed. We often annotate the arrow
+with the name of the transformation when describing chains of such
+transformations.
+
+The list of transformations we have
+validated is given in Figure~\ref{fig:opttrans}.
+%
+The behaviour of most transformations can be inferred from its name:
+Inlining substitutes a let-bound expression in the let body; $\beta$-reduction
+and case of known constructor effectively perform evaluation of the given expression;
+case-of-case commutes cases in scrutinee position inside the outer case alternatives;
+let floating and  for let-bound expressions to be floated out of $\lambda$-abstractions;
+the four commuting conversions allow let-bound expressions to be commuted with other
+Core constructs; $\eta$-expansion and reduction apply the $\eta$-laws to terms of function type;
+finally, binder swap substitutes occurrences of a scrutinee variable for the
+case binder. While many of the transformations above are not obviously optimising,
+they are chained in the compiler pipeline to expose optimisation opportunities.
+
+% Transformations are described by an arbitrary well-typed expression with a certain shape, on
+% the left hand side (lhs) of the arrow $\Longrightarrow$, resulting in an expression on
+% the right hand side (rhs) that we prove to be well-typed.
+% %
+% % For our proofs, we assume the lhs to be a well-typed expression and prove the
+% % rhs is well-typed as well.
+% %
+% For each transformation, we describe the intuition behind the transformation
+% preserving linearity in our system.
+
+% \todo[inline]{Aqui uma lista itemizada das transformações todas. As
+%   provas todas para apendice.}
+
+%%% COMENTADO AQUI
+% \subsubsection{Inlining}
+
+% % To the best of our knowledge, there is no linear type system for which inlining
+% % preserves linearity\footnote{https://github.com/ghc-proposals/ghc-proposals/blob/master/proposals/0111-linear-types.rst\#id90}
+
+% \input{language-v4/proofs/optimizations/Inlining}
+
+% \subsubsection{\texorpdfstring{$\beta$}{Beta}-reduction}
+
+% \input{language-v4/proofs/optimizations/BetaReduction}
+
+% \subsubsection{Case of known constructor}
+
+% \input{language-v4/proofs/optimizations/CaseOfKnownConstructor}
+
+% \subsubsection{Let floating}
+
+% \input{language-v4/proofs/optimizations/LetFloating}
+
+% \subsubsection{\texorpdfstring{$\eta$}{Eta}-conversions}
+
+% \input{language-v4/proofs/optimizations/EtaConvs}
+
+% \subsubsection{Binder Swap}
+
+% The binder swap transformation applies to case expressions whose scrutinee is a
+% single variable $x$, and it substitutes occurrences of $x$ in the case
+% alternatives for the case binder $z$. If $x$ is a linear resource, $x$ cannot
+% occur in the case alternatives (as we conservatively consider variables are not
+% in WHNF), so the substitution preserves types vacuously. Otherwise, $x$ can be
+% freely substituted by $z$, since $z$ is also an unrestricted resource (it's
+% usage environment is empty because $x$ is unrestricted).
+
+% \input{language-v4/proofs/optimizations/BinderSwap}
+
+\paragraph{Reverse Binder Swap\label{sec:reverse-binder-swap-considered-harmful}}
+
+We now discuss an issue we have found with the so-called reverse binder swap
+transformation.
+%
+%
+Reverse binder swap is dual to the binder swap transformation,
+substituting occurrences of the case
+binder $z$ by the scrutinee in the special case when the scrutinee is a
+variable $x$.
+%
+% his is dual to what the binder swap transformation
+% does in hope of eliminating multiple uses of $x$ (allowing for inlining). 
+By
+using the scrutinee $x$ instead of the case binder, we might be able to float
+out expressions from the alternative using the case binder. For example, we
+might float an expensive computation involving $z$ out of the case alternative,
+where $z$ is now out of scope:
+
+{\small\[
+\begin{array}{l}
+(\lambda x.~\lletrec{go~y = \ccase{x}{z~\{(a,b) \to \ E[expensive~z] \} }}{\dots})
+\Longrightarrow_\textrm{Reverse binder swap}\\
+(\lambda x.~\lletrec{go~y = \ccase{x}{z~\{(a,b) \to E[expensive~x] \} }}{\dots})
+\Longrightarrow_\textrm{Commuting case-let}\\
+\lambda x.~\llet{t = \mathit{expensive}~x}{(\lletrec{go~y = \ccase{x}{z~\{(a,b) \to E[t] \} }}{\dots})}\\
+\end{array}
+\]}
+
+\noindent If $go$ is compiled into a loop, we only compute
+$\mathit{expensive}~x$ once instead of in every loop iteration.
+
+Despite GHC applying the reverse binder swap transformation 
+during its Core-to-Core optimisation passes, this optimisation actually \emph{breaks} linearity
+when considered as a transformation on Linear Core programs. In practice, the
+optimisation preserves linearity when applied as part of the GHC
+pipeline due to the occurrence analyser being naive with
+regard to semantic linearity. Initially, it might seem as though an expression
+in which a variable $x$ occurs both in the case scrutinee and in the
+alternatives is linear (e.g.,~$\ccase{x}{\_ \to x}$):
+% \[
+% \G; \xl \vdash \ccase{x}{\_ \to x} : \s
+% \]
+% The reasoning is done by branching on whether $x$ refers to an expression in
+% WHNF or an unevaluated thunk.
+% \begin{itemize}
+% \item
+  If $x$ refers to an unevaluated expression, then scrutinizing it results
+in the expression bound by $x$ to be evaluated to WHNF. In a subsequent use of
+$x$ in the alternatives, $x$ refers to the evaluated scrutinee in WHNF, which
+must be consumed. Since $x$ is just another name (like the case binder) for the
+scrutinee in WHNF, we may use it instead of the case binder or pattern
+variables.
+%
+On the other hand, if $x$ already refers to an expression evaluated to WHNF, then
+scrutinizing it in a case expression is a no-op, thus we may use it again (in
+mutual exclusion with the case binder and pattern variables).
+%\end{itemize}
+%
+% This idea would be encoded by a possible rule like $$\TypeCaseVar$$
+%
+
+Even though, on its own, it makes intuitive sense that this example indeed uses
+$x$ linearly, when considered as part of a complete type system, allowing this
+expression to be linear makes the system \emph{unsound}.
+To see why this is the case, recall that $\beta$-reduction reduces an application of a linear function
+using call-by-name -- if we know the argument is used exactly once, a binding
+to share the result of computing the argument is unnecessary, so we instead
+substitute the argument expression for the linearly-bound variable in the
+$\lambda$-body directly.
+
+Consider the function application $(\lambda x.~\ccase{x}{\_ \to x})~(use~y)$
+where $y$ is a free linear variable. Assuming $\lambda x.~\ccase{x}{\_ \to x}$
+is a linear function, $\beta$-reduction transforms the
+application into $\ccase{use~y}{\_ \to
+  use~y}$. Now, $y$ is a linear variable \emph{consumed} in the scrutinee of the case
+that also occurs in the case alternative --
+linearity is violated by using the linear resource $y$ twice.
+%
+% Essentially, Linear Core would be unsound, and even duplicate resources, if the
+% above kind of expressions, where linear variable scrutinees occur in the
+% alternatives body, were well-typed, because of its interaction with the
+% call-by-name reduction of linear functions.
+In this sense, the reverse binder
+swap is an optimisation that creates ill-typed expressions from well-typed
+ones, so it is deemed an \emph{invalid} optimisation that does not preserve types in
+our system.
+
+Reverse binder swap is not a problem in the GHC simplifier because of the
+weaker notion of linearity understood by occurrence analysis. Occurrence
+analysis is a static analysis pass which can be used to determine whether a
+lambda application can be $\beta$-reduced call-by-name, and $\ccase{x}{\_ \to
+x}$ is \emph{not} seen as using $x$ linearly by the analysis. Thus,
+$\beta$-reduction is done with call-by-need on such an expression. If the above
+example were reduced with call-by-need
+% \[
+% \begin{array}{l}
+% (\lambda x.~\ccase{x}{\_ \to x})~(use~y)
+% \Longrightarrow_\textrm{CBNeed}
+% \llet{x = use~y}{\ccase{y}{\_ \to y}}
+% \end{array}
+% \]
+then the computation using $y$ would be let-bound, and $y$ used as a scrutinee
+variable, which is indeed semantically linear in $x$.
+
+% Concluding, in being able to understand more programs as linear, our type
+% system allows more expressions to be considered linear for $\beta$-reduction
+% without a let-allocation, however, it makes reverse binder swap an invalid
+% transformation since its output, when considered linear, might violate
+% linearity when further optimised.
+
+% (Link to ticket)
+% 
+% \begin{tabbing}
+% (1) All optimisations preserve (semantic) linearity\\
+% (2) If a function is (semantically) linear, then we can evaluate it using call-by-name and preserve linearity\\
+% (3) Reverse binder swap is an optimisation\\
+% (4) If reverse binder swap is applied to a case scrutinizing a linear resource in the body (`e`) of a linear function `f`, then the function is still linear by (1)\\
+% (5) If we evaluate `f`, we do it call-by-name because of (2)\\
+% (6) Call-by-name substitution of the linear argument in the body of a function has been reverse binder swapped doesn’t preserve linearity\\
+% (7) Contradiction: By 3 and 1, `f` is linear after reverse-binder-swap.
+By 2, we can substitute arguments to `f` call-by-name and preserve linearity, which contradicts with 6 that says call-by-name substitution after reverse binder swap violates linearity\\
+% \end{tabbing}
+
+% Conclusion:
+% Either we need to forfeit that we can always substitute call-by-name linear
+% function applications, or we forfeit that reverse binder swap preserves
+% linearity (instead, it preserves a weaker notion of linearity understood by the
+% syntatic-occurrence-analyzer)
+
+% \todo[inline]{Reverse-binder-swap is only well-defined in certain scenarios
+% where the optimizations don't apply call-by-name beta-reduction after the
+% reverse-binder-swap optimization -- otherwise we would duplicate resources.
+% In this case, it is not a matter of syntatic vs semantic linearity
+% }
+
+% \todo[inline]{
+% Mention, from ``Call-by-name, call-by-value, call-by-need and the linear lambda calculus'':
+% The call-by-name calculus is not entirely suitable for reasoning about
+% functional programs in lazy languages, because the beta rule may copy the
+% argument of a function any number of times. The call-by-need calculus uses a
+% diferent notion of reduction, observationally equivalent to the call-by-name
+% calculus. But call-by-need, like call-by-value, guarantees that the argument
+% to a function is not copied before it is reduced to a value.
+% }
+
+% It's also interesting to note that reverse-binder-swap preserves linearity under pure call-by-need but not under call-by-name, because
+% (In the sense that if EVEN linear functions reduce call-by-need rather than call-by-name, then it would preserve optimisations)
+
+% Reduce call-by-name linear function application
+% \begin{code}
+% f y = (\x. case x of _ -> x:1) (h y)
+% ===>
+% f y = (case x of _ -> x)[h y/x]
+% ===>
+% f y = (case h y of _ -> h y) -- consume y twice.
+% \end{code}
+% 
+% Vs. call-by-need
+% \begin{code}
+%  (\y = (\x. case x of _ -> x:1) (h y)) e
+% ===>
+%     let y = e in
+%        let x = h y in
+%         case x of _ -> x
+% ===>
+%        let x = h e_v in
+%         case x of _ -> x
+% ===>
+%         case x_v of _ -> x_v
+% \end{code}
+
+% \subsubsection{Case of Case}
+
+% The case of case transformation applies to case expressions whose scrutinee is
+% another case expression, and returns the innermost case expression transformed by repeating
+% the outermost case expression in each alternative of the innermost case,
+% scrutinizing the original alternative body.
+
+% Intuitively, since the scrutinee of the outermost case is not in WHNF, no
+% resources from it can directly occur in the outermost alternatives. By moving
+% the outermost alternatives inwards with a different scrutinee, the alternatives
+% remain well-typed because they are typed using either the case binder or the
+% pattern bound variables, which, by the \emph{Irrelevance} lemma, makes it
+% well-typed for any scrutinee consuming arbitrary resources. The proof is given in Section~\ref{sec:proof:caseofcase}.
+
+% \CaseOfCaseTheorem
+
+% }}}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % {{{ Linear Core as a GHC Plugin
@@
 
-%Trocar ordem com a metatheory
 \section{Linear Core as a GHC Plugin\label{sec:discuss:implementation}}
@@
 %
-The plugin successfully validates linearity throughout the
-(optimising) compilation of linearity-heavy libraries, namely
-\texttt{linear-base} and \texttt{linear-smc}, except in expressions whose
-linearity depends on so-called \emph{multiplicity coercions}, which are an
-avenue of future work (Section~\ref{sec:future-work}) exploring the intersection
-of Linear Core with type equality coercions.
 % Furthermore, the implementation accepts all example programs from Chapter~\ref{sec:linearity-semantically}
@@
 sub-derivations, we thread input/output linear resources through each
-derivation~\cite{DBLP:journals/tcs/CervesatoHP00}.  Pattern variables
-bound in a case alternative for a scrutinee not in WHNF, are
-introduced as $\D$-variables with usage environment
-$\lctag{\irr{\D}}{K_i}$, where $\D$ are the scrutinee resources and
-$K_i$ the tag of that pattern variable. To use the resources through
-the pattern-bound $\D$-vars, they must be first $Split$ into
-fragments. We consume tagged fragments of a resource \emph{as needed},
-i.e., when a resource, whose usage environment has a fragmented
-resource with tag $K_j$, is used, we $Split$ the matching resource
-according to the constructor $K$ and consume the fragment $K_j$,
-rather than eagerly determining which resources need to be fragmented
-to do so.
+derivation~\cite{DBLP:journals/tcs/CervesatoHP00}.
+%
+Pattern variables bound in a case alternative, for a scrutinee not in WHNF, are
+introduced as $\D$-variables with usage environment $\lctag{\irr{\D}}{K_i}$,
+where $\D$ are the scrutinee resources and $K_i$ the tag of that pattern
+variable. To use these pattern variables, the scrutinee resources must be first
+$Split$ into fragments. Instead of guessing which resources need to be split,
+the implementation consumes tagged fragments of a resource \emph{as needed},
+i.e., only when consuming a resource whose usage environment has a resource
+tagged $K_j$ do we $Split$ that resource on $K$ and consume the $K_j$ fragment.
+%
 Finally, our implementation must infer the usage environments of binders in
@@
 
+% TOdo: INCLUDE THIS??
+% The plugin successfully validates linearity throughout the
+% (optimising) compilation of linearity-heavy libraries, namely
+% \texttt{linear-base} and \texttt{linear-smc}, except in expressions whose
+% linearity depends on so-called \emph{multiplicity coercions}, which are an
+% avenue of future work (Section~\ref{sec:future-work}) exploring the intersection
+% of Linear Core with type equality coercions.
+
 The results of running the Linear Core GHC plugin on large 
@@
 %
-We compiled the libraries \texttt{linear-smc}~\cite{10.1145/3471874.3472980} (1500 lines), \texttt{linear-base} (4000
-lines), the Haskell standard library for programming with linear types,
-comprised of over 100 modules, and
+%
+We compiled the Haskell standard library for programming with linear types,
+\texttt{linear-base} (4000 lines, comprising over 100 modules), the
+\texttt{linear-smc}~\cite{10.1145/3471874.3472980} (1500 lines) library, and
 \texttt{priority-sesh}~\cite{10.1145/3471874.3472979} (1400 lines), a
@@
 binding is typechecked once per optimisation pass. i.e., we typecheck
-all intermediate programs produced by GHC. The total amount of
-programs rejected by the typechecker are given in the ``Total
-Rejected'' column, but we distil these rejections into ``Unique
-Rejections'' by removing duplicate rejections (those of the same
-program and for the same linearity-violating reason, but occurring at
-different stages). We further categorize the unique rejections into
-``Linear modulo Call-by-name'', ``Linear Rejected'', ``Not Linear
-Rejected'', and ``Unknown Rejected''.  We note that our implementation
-is not performance conscious. For instance, \texttt{linear-base} takes
-35 seconds to compile with our plugin, instead of 20 seconds.
-%
-% Using the plugin, \texttt{linear-base} takes 35 seconds to compile, instead of
-% 20 seconds. We note, however, that the implementation is not performance
-% conscious, and types every intermediate program from scratch, instead of
-% maintaining linearity information throughout the pipeline.
-
-Programs ``linear modulo call-by-name'' are a class of programs
-which scrutinize a variable, but then use the variable in the case
-alternatives of the case scrutinizing it. As discussed in
-Section~\ref{sec:reverse-binder-swap-considered-harmful}, these programs can be
-understood as linear as long as applications of linear functions binding these
-variables are \emph{not} reduced \emph{call-by-name}.
-In Core, these programs are not seen as
-linear, and thus GHC does not not reduce them using a call-by-name evaluation strategy.
-% We have a flag in our implementation to accept some of these programs...
-%
-``Linear Rejected'' programs include different kinds of programs which are
-rejected by the Linear Core system, but are still semantically linear. An
-example is a program to which a rewrite rule was applied, resulting in an
-application of unrestricted function to a linear resource.
-% in the first argument
-% to \incode{build}:
-% \begin{lstlisting}
-% take :: Int -> Replicator a *' $\lolli$ '* [a]
-% take = \ (ds :: Int) (r :: Replicator b) ->
-%      case ds of
-%        1 -> build (\(c :: b -> a -> a) (n :: a) -> c (extract r) n)
-%        ...
-% \end{lstlisting}
-% We know the linear resource is still used linearly because \incode{build} will
-% always instance the unrestricted function to $(:)$ (read \emph{Cons}), which is
-% linear. Additionally, these include programs scrutinizing a value of a type of
-% which all constructors have no linear components, but only matching on the
-% default alternative, programs where common-sub-expression elimination
-% substituted more than one occurrence of \incode{Ur x} by \incode{y}, in the body of
-% scrutinizing \incode{y} (a linear variable), and other programs affected by rewrite rules.
-%
-``Not Linear Rejected'' indicate programs that we do not understand as linear,
-semantically, and are simultaneously rejected by Linear Core and its
-implementation. % These programs were not necessarily identified as bad
-% by the linter.\footnote{rever}
-% An example, where the last component of \incode{HashMap}, \incode{wwz},
-% is linear, but is not being consumed in the case alternative matching on it:
-% \begin{lstlisting}
-% jssvi :: Ur Bool *' $\lolli$ '* Set Int *' $\lolli$ '* Ur (TestT IO ())
-% jssvi (a :: Ur Bool) (b :: Set Int)
-%   = case a of
-%      Ur ss -> case b of { HashMap wwx wwy wwz -> jump wjssAd ss }
-% \end{lstlisting}
-%
-Finally, ``Unknown Rejected'' programs are those whose validity we did not
-check. These include both programs accepted by Linear Core, but not by its
-implementation, and programs that are simply rejected by Linear Core, but which
-were not categorized.
+all intermediate programs produced by GHC. The total amount of programs
+rejected by the typechecker are counted dually, also accross every optimisation
+pass, without halting compilation when a program is rejected. We note that our
+implementation is not performance conscious, and types every intermediate
+program from scratch, instead of preserving information throughout the
+pipeline.
+
+%``Total Rejected'' column, but we distil these rejections into ``Unique
+%Rejections'' by removing duplicate rejections (those of the same
+%program and for the same linearity-violating reason, but occurring at
+%different stages). We further categorize the unique rejections into
+%``Linear modulo Call-by-name'', ``Linear Rejected'', ``Not Linear
+%Rejected'', and ``Unknown Rejected''.  We note that our implementation
+%is not performance conscious. For instance, \texttt{linear-base} takes
+%35 seconds to compile with our plugin, instead of 20 seconds.
+%%
+%% Using the plugin, \texttt{linear-base} takes 35 seconds to compile, instead of
+%% 20 seconds. We note, however, that the implementation is not performance
+%% conscious, and types every intermediate program from scratch, instead of
+%% maintaining linearity information throughout the pipeline.
+
+%Programs ``linear modulo call-by-name'' are a class of programs
+%which scrutinize a variable, but then use the variable in the case
+%alternatives of the case scrutinizing it. As discussed in
+%Section~\ref{sec:reverse-binder-swap-considered-harmful}, these programs can be
+%understood as linear as long as applications of linear functions binding these
+%variables are \emph{not} reduced \emph{call-by-name}.
+%In Core, these programs are not seen as
+%linear, and thus GHC does not not reduce them using a call-by-name evaluation strategy.
+%% We have a flag in our implementation to accept some of these programs...
+%%
+%``Linear Rejected'' programs include different kinds of programs which are
+%rejected by the Linear Core system, but are still semantically linear. An
+%example is a program to which a rewrite rule was applied, resulting in an
+%application of unrestricted function to a linear resource.
+%% in the first argument
+%% to \incode{build}:
+%% \begin{lstlisting}
+%% take :: Int -> Replicator a *' $\lolli$ '* [a]
+%% take = \ (ds :: Int) (r :: Replicator b) ->
+%%      case ds of
+%%        1 -> build (\(c :: b -> a -> a) (n :: a) -> c (extract r) n)
+%%        ...
+%% \end{lstlisting}
+%% We know the linear resource is still used linearly because \incode{build} will
+%% always instance the unrestricted function to $(:)$ (read \emph{Cons}), which is
+%% linear. Additionally, these include programs scrutinizing a value of a type of
+%% which all constructors have no linear components, but only matching on the
+%% default alternative, programs where common-sub-expression elimination
+%% substituted more than one occurrence of \incode{Ur x} by \incode{y}, in the body of
+%% scrutinizing \incode{y} (a linear variable), and other programs affected by rewrite rules.
+%%
+%``Not Linear Rejected'' indicate programs that we do not understand as linear,
+%semantically, and are simultaneously rejected by Linear Core and its
+%implementation. % These programs were not necessarily identified as bad
+%% by the linter.\footnote{rever}
+%% An example, where the last component of \incode{HashMap}, \incode{wwz},
+%% is linear, but is not being consumed in the case alternative matching on it:
+%% \begin{lstlisting}
+%% jssvi :: Ur Bool *' $\lolli$ '* Set Int *' $\lolli$ '* Ur (TestT IO ())
+%% jssvi (a :: Ur Bool) (b :: Set Int)
+%%   = case a of
+%%      Ur ss -> case b of { HashMap wwx wwy wwz -> jump wjssAd ss }
+%% \end{lstlisting}
+%%
+%Finally, ``Unknown Rejected'' programs are those whose validity we did not
+%check. These include both programs accepted by Linear Core, but not by its
+%implementation, and programs that are simply rejected by Linear Core, but which
+%were not categorized.
 
 {\small
-  \input{prototype/core-plugin-results}
+  \input{prototype/core-plugin-results-2}
   }
-
+%
 The results indicate our mostly direct implementation of Linear Core is
@@
 linear types.
-%
-The programs rejected by the Linear Core plugin, in \texttt{linear-base},
+
+However, Linear Core does not accept every single program produced by the
+optimisation pipeline of GHC. Our manual analysis of rejected programs revealed
+excellent results. Programs considered ill-typed by the Linear Core
+implementation include:
+%
+\begin{itemize}
+    \item Programs we expected GHC to produce, but know are not seen as linear by $\lambda^\pi_\Delta$.
+        Namely, programs which scrutinize a variable, but then use the variable in the
+        case alternatives of the case scrutinizing it. As discussed in
+        Section~\ref{sec:reverse-binder-swap-considered-harmful}, these programs can be
+        understood as linear as long as applications of linear functions binding these
+        variables are reduced \emph{call-by-need}, which is not the case in our system.
+
+    \item Programs rejected due to the lack of ``linearity coercions'' in
+        $\lambda^\pi_\Delta$. So-called \emph{multiplicity coercions} extending
+        Linear Core could make these programs be understood as linear.
+
+    \item Programs which \emph{actually violate linearity}, in that the term
+        violates the linear type it is annotated with. These are the programs
+        we ultimately want to be able to reject by having a linearly typed Core
+        validate intermediate programs. We found both a program which outright
+        discarded a linear resource, and that rewrite rules implementing stream
+        fusion~\cite{10.1145/165180.165214,10.1145/1291151.1291199,10.1007/978-3-642-24276-2_2}, and
+        others too, could produce linearly-invalid programs (in one such case,
+        the type of \lstinline{build} was not general enough to accommodate linearity).
+\end{itemize}
+%
+Overall, the programs rejected by the Linear Core plugin,
+%, in \texttt{linear-base},
 besides validating that our implementation is faithful to the
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% {{{ Metatheory
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Metatheory of Linear Core\label{sec:main:metatheory}}
-
-The $\lambda^\pi_\D$ type system is sound: well-typed programs in Linear Core do not
-get \emph{stuck}. Besides type safety, we
-prove that multiple optimising transformations used in GHC preserve
-linearity in the sense of being well-typed in our system.
-%
-% Additionally, we state our assumptions that outline an isomorphism between
-% using a linear variable $\xl$ and a $\D$-variable $\xD$ that consumes existing
-% resources $\D$, for any $\D$.
-
-\renewcommand{\DeltaLinearRelationLemma}{
-\begin{restatable}[$\Delta$-bound to Linear]{lemma}{deltaone}\label{lem:deltaone}
-% Any $\Delta$-variable with a fully-irrelevant usage environment can be made a linear variable as long as its irrelevant environment is dropped from the linear environment (and the unrestricted env.
-is updated accordingly).\\
-If $\G,\x[\irr{\D}]; \irr{\D},\D' \vdash e : \vp$ 
-then $\G[x/\irr{\D}]; \D',\xl \vdash e :\vp$.
-\end{restatable}
-}
-
-\renewcommand{\LinearDeltaRelationLemma}{
-  \begin{restatable}[Linear to $\Delta$-bound]{lemma}{onedelta}\label{lem:onedelta}
-If $\G; \D',\xl \vdash e :\vp$
-then $\G[\D/x],\xD; \D,\D' \vdash e : \vp$ ($\Delta$ fresh).
-\end{restatable}
-}
-
-\renewcommand{\DeltaUnrestrictedRelationLemma}{
-\begin{restatable}[Unrestricted and $\Delta$-bound]{lemma}{undelta}\label{lem:undelta}
-%An unrestricted variable is equivalent to a $\D$-var with an empty usage environment.\\
-$\G,\xo; \D \vdash e : \vp$ iff $\G,\x[\cdot]; \D \vdash e : \vp$
-\end{restatable}
-}
-
-\paragraph{$\Delta$-bound variables} % Our development relies on a characterization of $\D$-bound
-% variables
-A well-typed program with a linear variable ($\xl$) is equivalently
-well-typed if the linear variable were instead $\D$-bound ($\xD$) with
-usage environment $\D$, with $\D$ available in the linear context (and the unrestricted context updated accordingly).
-
-\LinearDeltaRelationLemma
-
-\noindent Dually, any $\Delta$-bound variable with an irrelevant usage
-environment can be made into a linear variable as long as its
-(irrelevant) environment is dropped from the linear context.
-
-% Dually, a well-typed program with resources $\D$ and $\D$-bound
-% variable ($\xD$) is equivalently well-typed, as long as $\D$ is consumed
-% through the use of $x$, if $x$ is moved to the linear context, resources $\D$
-% are removed from the linear context, and occurrences of $\D$ whole in usage
-% environments are substituted by $x$ (occurrences of fragments of $\D$ in usage
-% environments are unimportant since $\D$ was consumed whole by $x$, not by any
-% of the fragment-using $\D$-vars).
-
-\DeltaLinearRelationLemma
-
-% Intuitively, a linear variable
-
-\noindent Finally, unrestricted and $\D$-bound
-variables with an empty usage environment are equivalent.
-
-\DeltaUnrestrictedRelationLemma
-
-\paragraph{Irrelevance}
-%
-As discussed, proof irrelevant resources can only be consumed
-indirectly, essentially encoding that the (non-WHNF) case scrutinee
-resources must be consumed through the case binder or the linear
-pattern-bound variables.
-%
-As a case expression is evaluated, the scrutinee will eventually
-reduce to WHNF,
-which must then be typed with rule $Case_{\textrm{WHNF}}$.
-%
-Crucially, the two case rules must be in harmony in the system, in the sense that
-case expressions typed using the $Case_{\textrm{Not WHNF}}$ rule must also be
-well-typed by the $Case_{\textrm{WHNF}}$ once the scrutinee is evaluated to WHNF.
-
-% The type preservation theorem states that a well-typed expression
-% remains well-typed in the presence of evaluation. Specifically, when the case
-% expression whose scrutinee is evaluated to WHNF is handled in the
-% preservation proof.
-
-% The \emph{Irrelevance} lemma is required to prove preservation for that
-% evaluation case. We need to prove that the alternatives of a case expression
-% typed with proof irrelevant resources are still well-typed when the proof
-% irrelevant resource is substituted by the scrutinee resources as it is evaluated to WHNF.
-% In this sense, the \emph{Irrelevance} lemma witnesses the soundness of typing a
-% case alternative with proof irrelevant resources in a certain context with respect to
-% typing the same expression with arbitrary resources (we note, however, typing
-% an alternative with proof irrelevant resources is not complete wrt using
-% arbitrary resources -- a counter example needs only to use a resource
-% directly).
-
-\WHNFConvSoundness
-
-The \emph{Irrelevance} lemma witnesses the soundness of typing a
-case alternative with proof irrelevant resources with respect to
-typing the same expression with arbitrary resources. However, typing
-a case alternative with proof irrelevant resources is not complete wrt using
-arbitrary resources -- a counter example needs only to use a resource
-directly.
-
-% \noindent Intuitively, the lemma holds since proof irrelevant resources can
-% only be used through the case binder or pattern-bound variables. If we
-% consistently replace the proof irrelevant resources both in the typing
-% environment and in the usage environments containing them, the expression
-% remains well-typed. We note that the proof irrelevant resources are always
-% unique when introduced in such a case alternative (we always take the scrutinee
-% environment to make irrelevant, and allow nested ``irrelevantness''), so the
-% case binder has the only occurrence of those resources in the $\G$ environment.
-% % (being somewhat akin to congruence).
-% %
-% The proof is given in Section~\ref{sec:proof:irrelevance}.
-
-\subsection{Type safety\label{sec:type-safety-meta}}
-
-We prove type safety of the Linear Core system via the standard type
-preservation and progress results. As is customary, we make use of multiple
-substitution lemmas, one for each kind of variable: unrestricted variables
-$\xo$, linear variables $\xl$, and $\D$-bound variables $\xD$.
-
-% We start with the auxiliary results, as we will make use of them in a select part of the preservation proof.
-
-\TypePreservationTheorem
-%
-\ProgressTheorem
-% \noindent Type preservation states that a well-typed expression $e$ that
-% evaluates to $e'$ remains well-typed under the same context:
-%
-Type preservation
-% ($\S$~\ref{sec:proof:type-preservation})
-follows by structural induction on the reductions $e \longrightarrow
-e'$ from the operational semantics. Most cases are straightforward, appealing to one or more
-substitution lemmas (see Appendix~\ref{sec:proof:type-preservation}). The most
-interesting case is that of case expressions whose scrutinee can be further
-evaluated -- we branch on whether the scrutinee becomes in WHNF and appeal to Lemma~\ref{lem:irrev} if so.
-%
-% This case guarantees that the separation of rules for treating scrutinees is
-% consistent, in the sense that a well-typed case expression with a scrutinee not
-% in WHNF remains well-typed after the scrutinee is evaluated to WHNF.
-%
-%
-%
-%\noindent
-% Progress states that the evaluation of a well-typed term does not block:
-% Similarly, progress is proved by induction on typing ($\S$~\ref{sec:proof:progress}).
-
-% \subsubsection{Substitution Lemmas}
-
-% The preservation and progress theorems depend on multiple substitution lemmas,
-% one for each kind of variable, as is standard.
-% % The proofs are given in Section~\ref{sec:proof:substitution-lemmas}.
-% % The ... themselves diverge from their common formulation, because
-% % $\D$-variables refer to linearly bound variables, so substitution must take into account
-
-% The linear substitution lemma states that a well-typed expression $e$ with a
-% linear variable $x$ of type $\s$ remains well-typed if
-% occurrences of $x$ in the $e$ are replaced by an expression $e'$ of the same
-% type $\s$, and occurrences of $x$ in the linear context and in usage
-% environments of $\D$-bound variables are replaced by the linear context $\D'$
-% used to type $e'$:
-
-% \LinearSubstitutionLemma
-
-% \noindent Where $\G[\D'/x]$ substitutes all occurrences of $x$ in the usage
-% environments of $\D$-variables in $\G$ by the linear variables in $\D'$.
-% % ($x$ couldn't appear anywhere besides usage environments of $\D$-bound variables, since $x$ is linear).
-
-% The substitution of the resource in the usage environments is illustrated
-% by the following example. Consider the term $\llet{y = use~x}{y}$ where $use$ and $x$ are free variables:
-% if we replace occurrences of $x$ by $e'$ (where $\G;\D \vdash e : \s$), then the
-% ``real'' usage environment of $y$ goes from $\{x\}$ to $\D$. If we don't update
-% the usage environment of $y$ accordingly, we'll ultimately be typing
-% $y{:}_{\{x\}}\vp$ with $\D$ instead of $x$, which is not valid.
-
-% The linear substitution lemma extends to case alternatives as well.
-% The lemma for substitution of linear variables in case alternatives is similar
-% to the linear substitution lemma, applied to the case alternative judgement.
-% % Is slightly more involved, in the sense that there are more environments in which the
-% % substitution $[\D/x]$ must be applied (for the same reason):
-% %
-% \LinearSubstitutionAltsLemma
-% %
-% \noindent We further require that the environment annotated in the case
-% alternative judgement, $\D_s$, is a subset of the environment used to type the
-% whole alternative $\D_s \subseteq \D$. In all occurrences of the alternative
-% judgement (in $Case_{\textrm{WHNF}}$ and $Case_{\textrm{Not WHNF}}$), the
-% environment annotating the alternative judgement is \emph{always} a subset of
-% the alternative environment.
-
-% The substitution lemma for unrestricted variables follows the usual
-% formulation, with the added restriction (common to linear type systems) that
-% the expression $e'$ that is going to substitute the unrestricted variable $x$
-% is typed on an empty linear environment:
-% %
-% \UnrestrictedSubstitutionLemma
-% %
-% \noindent Similarly, we also prove the substitution of unrestricted variables preserves types on an alternative case expression:
-% %
-% \UnrestrictedSubstitutionAltsLemma
-
-% Finally, we introduce the lemma stating that substitution of $\D$-bound
-% variables by expressions of the same type preserves the type of the original
-% expression.
-% %
-% What distinguishes this lemma from traditional substitution lemmas is that the
-% usage environment $\D$ of the variable $x$ being substituted by expression $e'$
-% must match exactly the typing environment $\D$ of $e'$ and the
-% environment of the original expression doesn't change with the substitution:
-% %
-% \DeltaSubstitutionLemma
-% %
-% \noindent Intuitively, if $x$ is well-typed with $\D$ in $e$, substituting $x$
-% by an expression $e'$ which is typed in the same environment $\D$ allows the
-% distribution of resources $\D,\D'$ used to type $e$ across sub-derivations to remain
-% unchanged. To prove the theorems, we don't need a ``stronger'' substitution of
-% $\D$-vars lemma (allowing arbitrary resources $\D''$ to type $e'$, as in other
-% substitution lemmas), as we only ever substitute $\D$-variables by expressions
-% whose typing environment matches the variables usage environment. However, it
-% is not obvious whether such a lemma is possible to prove for $\D$-variables
-% (e.g. let $\G;\D \vdash e :\s$ and $\G; \D' \vdash \llet{x = e'}{x}$, if we
-% substitute $e$ for $x$ the resources $\D'$ are no longer consumed).
-
-% The $\D$-substitution lemma on case alternatives reflects again that the typing
-% environment of the expression substitution the variable must match its usage
-% environment. We recall that $\D_s \subseteq \D,\D'$ states that the annotated
-% environment is always contained in the typing environment, which is true of all
-% occurrences of this judgement. An alternative formulation of this lemma could
-% instead explicitly list $\D_s$ as part of the typing environment for the same
-% effect:
-
-% \DeltaSubstitutionAltsLemma
-
-% The proofs for substitution lemmas of linear, unrestricted, and
-% $\D$-variables are available in Section~\ref{sec:proof:substitution-lemmas}.
-
-%TODO! Substitution of proof-irrelevant linear variables preserves typing. The
-%term always remains the same because $x$ cannot occur in any term, however, all
-%variables that refer to $x$ in their usage environment must now refer the usage env. of the substitee (e.g. $[x] => [\D]$).
-%This seems trivial to see correct, since all occurrences are in environments, so we get some equivalence similar to the one we need for the proof of Alt0.
-%
-%TODO: Multiplicity substitution preserves typing lemma
-%
-%TODO: Canonical forms lemma
-%
-%TODO: Corollary of $\Delta$-var subst. for $\ov{\Delta}$
-%
-%TODO: Constructor app typing:
-%If $\Gamma, \Delta \vdash K~\ov{e}$ and $K{:}\ov{\sigma\to\pi}~T~\ov{p} \in \Gamma$ and $\hasnolinearvars{\Gamma}$
-%then $\ov{\Gamma, \Delta_i \vdash e_i : \sigma_i}$
-
-\subsection{Optimisations preserve linearity\label{sec:optimisations-preserve-types-meta}}
-%\todo[inline]{Check Commuting case-let in Fig.~\ref{fig:opttrans}!}
-A main goal of the Linear Core type system is 
-to serve as a typed intermediate representation for optimising
-compilers of lazy languages with linearity. In light of this,
-we show that multiple optimising transformations are type
-preserving in Linear Core (i.e.,~preserve linearity semantically).
-
-% The optimising transformations proved sound wrt Linear Core in this section
-% have been previously explained and motivated in
-% Section~\ref{sec:core-to-core-transformations}.
-% %
-%\{\overline{\rho\to e'}\}
-\begin{figure}[t]
-
-  {\small
-  \[
-    \begin{array}{ll}
-   (\llet{\xD = e}{e'} )  \Longrightarrow \llet{\xD =       e}{e'[e/x]} &                    \mbox{Inlining}\\
-       (\lambda \x[\pi][\s].~e)~e'  \Longrightarrow e[e'/x]  &                \mbox{$\beta$-reduction}\\
-      (\ccase{(K~\ov{e})}{\z[\D][\s]~\{\overline{\rho_j\to e_j}, K~\ov{x} \to e_i, \overline{\rho_k\to e_k}\}} )\Longrightarrow  e_i\ov{[e/x]}[K~\ov{e}/z] &
-                                  \mbox{Case of known constr.}\\
-      (\ccase{(\ccase{e_c}{\z[\D]~\{\ov{\rho_{c_i}\to e_{c_i}}\}})}{\var[w][\irr{\D,\D'}][\s']~\{\ov{\rho_i\to e_i}\}})\\
-            \Longrightarrow 
-      \ccase{e_c}{\z[\D]~\{\ov{\rho_{c_i} \to \ccase{e_{c_i}}{w~\{\ov{\rho_i\to e_i}\}}}\}}
-      & \mbox{Case-of-case}\\
-      (\lambda \y[\pi].~\llet{\xD = e}{e'})
-      \Longrightarrow
-      \llet{\xD=e}{\lambda\y[\pi].~e'} & \mbox{Commuting let-$\lambda$}\\
-      ((\llet{v = e}{b})~a) \Longrightarrow \llet{v = e}{b~a} &\mbox{Commuting let-app.}\\
-      (\ccase{(\llet{v = e}{b})}{\{\overline{\rho\to e'}\}}) \Longrightarrow
-      \llet{v =  e}{\ccase{b}{\{\overline{\rho\to e'}\}}}
-                               &
-                                 \mbox{Commuting let-case}\\
-      (\ccase{e}{\{\overline{\rho_j\to e_j}, \rho \to E[e_1],
-      \overline{\rho_k\to e_k}\}} ) \qquad (\mbox{$x$ fresh}) \\
-     \Longrightarrow \llet{x = e_1}{\ccase{e}{\{\overline{\rho_j\to e_j}, \rho \to E[x], \overline{\rho_k\to e_k}\}}} 
-                                                                        & \mbox{Commuting case-let}\\
-      (\llet{x = (\llet{v = e}{b})}{c}) \Longrightarrow \llet{v = e}{\llet{x = b}{c}}
-                                                                      & \mbox{Commuting let-let}\\
-      f \Longrightarrow \lambda x.(f~x) & \mbox{$\eta$-expansion}\\
-      \lambda x.(f~x) \Longrightarrow f & \mbox{$\eta$-reduction}\\
-      (\ccase{x}{z~\{\ov{\rho_i\to e_i}\}}) \Longrightarrow \ccase{x}{z~\{\ov{\rho_i\to  e_i[z/x]}\}}
-                                                                      & \mbox{Binder swap}\\
-      % (\ccase{x}{z~\{\ov{\rho_i\to e_i}\}} )\Longrightarrow \ccase{x}{z~\{\ov{\rho_i\to e_i[x/z]}\}}
-      %                                                                 & \mbox{Reverse binder swap}
-    \end{array}
-  \]}
-  \caption{Optimising transformations validated by Linear
-    Core\label{fig:opttrans} (for commuting case-let we
-      require $e_1$ not to capture the case binder, variables bound by the
-      pattern $\rho$ and the context $E[{-}]$ -- see Appendix~\ref{app:optimisations} for
-    details).} 
-\end{figure}
-
-We describe Core-to-Core transformations as $e_1 \Longrightarrow e_2$,
-where $e_1$ is an arbitrary \emph{well-typed} expression
-matching a certain shape (or of a certain type), which is
-transformed into expression $e_2$. Validating a transformation
-entails showing that $e_2$ is well-typed. We often annotate the arrow
-with the name of the transformation when describing chains of such
-transformations.
-
-The list of transformations we have
-validated is given in Figure~\ref{fig:opttrans}.
-%
-The behaviour of most transformations can be inferred from its name:
-Inlining substitutes a let-bound expression in the let body; $\beta$-reduction
-and case of known constructor effectively perform evaluation of the given expression;
-case-of-case commutes cases in scrutinee position inside the outer case alternatives;
-let floating and  for let-bound expressions to be floated out of $\lambda$-abstractions;
-the four commuting conversions allow let-bound expressions to be commuted with other
-Core constructs; $\eta$-expansion and reduction apply the $\eta$-laws to terms of function type;
-finally, binder swap substitutes occurrences of a scrutinee variable for the
-case binder. While many of the transformations above are not obviously optimising,
-they are chained in the compiler pipeline to expose optimisation opportunities.
-
-% Transformations are described by an arbitrary well-typed expression with a certain shape, on
-% the left hand side (lhs) of the arrow $\Longrightarrow$, resulting in an expression on
-% the right hand side (rhs) that we prove to be well-typed.
-% %
-% % For our proofs, we assume the lhs to be a well-typed expression and prove the
-% % rhs is well-typed as well.
-% %
-% For each transformation, we describe the intuition behind the transformation
-% preserving linearity in our system.
-
-% \todo[inline]{Aqui uma lista itemizada das transformações todas. As
-%   provas todas para apendice.}
-
-%%% COMENTADO AQUI
-% \subsubsection{Inlining}
-
-% % To the best of our knowledge, there is no linear type system for which inlining
-% % preserves linearity\footnote{https://github.com/ghc-proposals/ghc-proposals/blob/master/proposals/0111-linear-types.rst\#id90}
-
-% \input{language-v4/proofs/optimizations/Inlining}
-
-% \subsubsection{\texorpdfstring{$\beta$}{Beta}-reduction}
-
-% \input{language-v4/proofs/optimizations/BetaReduction}
-
-% \subsubsection{Case of known constructor}
-
-% \input{language-v4/proofs/optimizations/CaseOfKnownConstructor}
-
-% \subsubsection{Let floating}
-
-% \input{language-v4/proofs/optimizations/LetFloating}
-
-% \subsubsection{\texorpdfstring{$\eta$}{Eta}-conversions}
-
-% \input{language-v4/proofs/optimizations/EtaConvs}
-
-% \subsubsection{Binder Swap}
-
-% The binder swap transformation applies to case expressions whose scrutinee is a
-% single variable $x$, and it substitutes occurrences of $x$ in the case
-% alternatives for the case binder $z$. If $x$ is a linear resource, $x$ cannot
-% occur in the case alternatives (as we conservatively consider variables are not
-% in WHNF), so the substitution preserves types vacuously. Otherwise, $x$ can be
-% freely substituted by $z$, since $z$ is also an unrestricted resource (it's
-% usage environment is empty because $x$ is unrestricted).
-
-% \input{language-v4/proofs/optimizations/BinderSwap}
-
-\paragraph{Reverse Binder Swap\label{sec:reverse-binder-swap-considered-harmful}}
-
-We now discuss an issue we have found with the so-called reverse binder swap
-transformation.
-%
-%
-Reverse binder swap is dual to the binder swap transformation,
-substituting occurrences of the case
-binder $z$ by the scrutinee in the special case when the scrutinee is a
-variable $x$.
-%
-% his is dual to what the binder swap transformation
-% does in hope of eliminating multiple uses of $x$ (allowing for inlining). 
-By
-using the scrutinee $x$ instead of the case binder, we might be able to float
-out expressions from the alternative using the case binder. For example, we
-might float an expensive computation involving $z$ out of the case alternative,
-where $z$ is now out of scope:
-
-{\small\[
-\begin{array}{l}
-(\lambda x.~\lletrec{go~y = \ccase{x}{z~\{(a,b) \to \ E[expensive~z] \} }}{\dots})
-\Longrightarrow_\textrm{Reverse binder swap}\\
-(\lambda x.~\lletrec{go~y = \ccase{x}{z~\{(a,b) \to E[expensive~x] \} }}{\dots})
-\Longrightarrow_\textrm{Commuting case-let}\\
-\lambda x.~\llet{t = \mathit{expensive}~x}{(\lletrec{go~y = \ccase{x}{z~\{(a,b) \to E[t] \} }}{\dots})}\\
-\end{array}
-\]}
-
-\noindent If $go$ is compiled into a loop, we only compute
-$\mathit{expensive}~x$ once instead of in every loop iteration.
-
-Despite GHC applying the reverse binder swap transformation 
-during its Core-to-Core optimisation passes, this optimisation actually \emph{breaks} linearity
-when considered as a transformation on Linear Core programs. In practice, the
-optimisation preserves linearity when applied as part of the GHC
-pipeline due to the occurrence analyser being naive with
-regard to semantic linearity. Initially, it might seem as though an expression
-in which a variable $x$ occurs both in the case scrutinee and in the
-alternatives is linear (e.g.,~$\ccase{x}{\_ \to x}$):
-% \[
-% \G; \xl \vdash \ccase{x}{\_ \to x} : \s
-% \]
-% The reasoning is done by branching on whether $x$ refers to an expression in
-% WHNF or an unevaluated thunk.
-% \begin{itemize}
-% \item
-  If $x$ refers to an unevaluated expression, then scrutinizing it results
-in the expression bound by $x$ to be evaluated to WHNF. In a subsequent use of
-$x$ in the alternatives, $x$ refers to the evaluated scrutinee in WHNF, which
-must be consumed. Since $x$ is just another name (like the case binder) for the
-scrutinee in WHNF, we may use it instead of the case binder or pattern
-variables.
-%
-On the other hand, if $x$ already refers to an expression evaluated to WHNF, then
-scrutinizing it in a case expression is a no-op, thus we may use it again (in
-mutual exclusion with the case binder and pattern variables).
-%\end{itemize}
-%
-% This idea would be encoded by a possible rule like $$\TypeCaseVar$$
-%
-
-Even though, on its own, it makes intuitive sense that this example indeed uses
-$x$ linearly, when considered as part of a complete type system, allowing this
-expression to be linear makes the system \emph{unsound}.
-To see why this is the case, recall that $\beta$-reduction reduces an application of a linear function
-using call-by-name -- if we know the argument is used exactly once, a binding
-to share the result of computing the argument is unnecessary, so we instead
-substitute the argument expression for the linearly-bound variable in the
-$\lambda$-body directly.
-
-Consider the function application $(\lambda x.~\ccase{x}{\_ \to x})~(use~y)$
-where $y$ is a free linear variable. Assuming $\lambda x.~\ccase{x}{\_ \to x}$
-is a linear function, $\beta$-reduction transforms the
-application into $\ccase{use~y}{\_ \to
-  use~y}$. Now, $y$ is a linear variable \emph{consumed} in the scrutinee of the case
-that also occurs in the case alternative --
-linearity is violated by using the linear resource $y$ twice.
-%
-% Essentially, Linear Core would be unsound, and even duplicate resources, if the
-% above kind of expressions, where linear variable scrutinees occur in the
-% alternatives body, were well-typed, because of its interaction with the
-% call-by-name reduction of linear functions.
-In this sense, the reverse binder
-swap is an optimisation that creates ill-typed expressions from well-typed
-ones, so it is deemed an \emph{invalid} optimisation that does not preserve types in
-our system.
-
-Reverse binder swap is not a problem in the GHC simplifier because of the
-weaker notion of linearity understood by occurrence analysis. Occurrence
-analysis is a static analysis pass which can be used to determine whether a
-lambda application can be $\beta$-reduced call-by-name, and $\ccase{x}{\_ \to
-x}$ is \emph{not} seen as using $x$ linearly by the analysis. Thus,
-$\beta$-reduction is done with call-by-need on such an expression. If the above
-example were reduced with call-by-need
-% \[
-% \begin{array}{l}
-% (\lambda x.~\ccase{x}{\_ \to x})~(use~y)
-% \Longrightarrow_\textrm{CBNeed}
-% \llet{x = use~y}{\ccase{y}{\_ \to y}}
-% \end{array}
-% \]
-then the computation using $y$ would be let-bound, and $y$ used as a scrutinee
-variable, which is indeed semantically linear in $x$.
-
-% Concluding, in being able to understand more programs as linear, our type
-% system allows more expressions to be considered linear for $\beta$-reduction
-% without a let-allocation, however, it makes reverse binder swap an invalid
-% transformation since its output, when considered linear, might violate
-% linearity when further optimised.
-
-% (Link to ticket)
-% 
-% \begin{tabbing}
-% (1) All optimisations preserve (semantic) linearity\\
-% (2) If a function is (semantically) linear, then we can evaluate it using call-by-name and preserve linearity\\
-% (3) Reverse binder swap is an optimisation\\
-% (4) If reverse binder swap is applied to a case scrutinizing a linear resource in the body (`e`) of a linear function `f`, then the function is still linear by (1)\\
-% (5) If we evaluate `f`, we do it call-by-name because of (2)\\
-% (6) Call-by-name substitution of the linear argument in the body of a function has been reverse binder swapped doesn’t preserve linearity\\
-% (7) Contradiction: By 3 and 1, `f` is linear after reverse-binder-swap.
-By 2, we can substitute arguments to `f` call-by-name and preserve linearity, which contradicts with 6 that says call-by-name substitution after reverse binder swap violates linearity\\
-% \end{tabbing}
-
-% Conclusion:
-% Either we need to forfeit that we can always substitute call-by-name linear
-% function applications, or we forfeit that reverse binder swap preserves
-% linearity (instead, it preserves a weaker notion of linearity understood by the
-% syntatic-occurrence-analyzer)
-
-% \todo[inline]{Reverse-binder-swap is only well-defined in certain scenarios
-% where the optimizations don't apply call-by-name beta-reduction after the
-% reverse-binder-swap optimization -- otherwise we would duplicate resources.
-% In this case, it is not a matter of syntatic vs semantic linearity
-% }
-
-% \todo[inline]{
-% Mention, from ``Call-by-name, call-by-value, call-by-need and the linear lambda calculus'':
-% The call-by-name calculus is not entirely suitable for reasoning about
-% functional programs in lazy languages, because the beta rule may copy the
-% argument of a function any number of times. The call-by-need calculus uses a
-% diferent notion of reduction, observationally equivalent to the call-by-name
-% calculus. But call-by-need, like call-by-value, guarantees that the argument
-% to a function is not copied before it is reduced to a value.
-% }
-
-% It's also interesting to note that reverse-binder-swap preserves linearity under pure call-by-need but not under call-by-name, because
-% (In the sense that if EVEN linear functions reduce call-by-need rather than call-by-name, then it would preserve optimisations)
-
-% Reduce call-by-name linear function application
-% \begin{code}
-% f y = (\x. case x of _ -> x:1) (h y)
-% ===>
-% f y = (case x of _ -> x)[h y/x]
-% ===>
-% f y = (case h y of _ -> h y) -- consume y twice.
-% \end{code}
-% 
-% Vs. call-by-need
-% \begin{code}
-%  (\y = (\x. case x of _ -> x:1) (h y)) e
-% ===>
-%     let y = e in
-%        let x = h y in
-%         case x of _ -> x
-% ===>
-%        let x = h e_v in
-%         case x of _ -> x
-% ===>
-%         case x_v of _ -> x_v
-% \end{code}
-
-% \subsubsection{Case of Case}
-
-% The case of case transformation applies to case expressions whose scrutinee is
-% another case expression, and returns the innermost case expression transformed by repeating
-% the outermost case expression in each alternative of the innermost case,
-% scrutinizing the original alternative body.
-
-% Intuitively, since the scrutinee of the outermost case is not in WHNF, no
-% resources from it can directly occur in the outermost alternatives. By moving
-% the outermost alternatives inwards with a different scrutinee, the alternatives
-% remain well-typed because they are typed using either the case binder or the
-% pattern bound variables, which, by the \emph{Irrelevance} lemma, makes it
-% well-typed for any scrutinee consuming arbitrary resources. The proof is given in Section~\ref{sec:proof:caseofcase}.
-
-% \CaseOfCaseTheorem
-
-% }}}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2024-02-18 icfp.tex
RM edits metatheory section
And, especially, completely rewrites the reverse binder swap section.

@@
 we show that multiple optimising transformations are type
-preserving in Linear Core (i.e.,~preserve linearity semantically).
+preserving in Linear Core. %(i.e.,~preserve linearity semantically).
 
@@
 
-We now discuss an issue we have found with the so-called reverse binder swap
-transformation.
-%
-%
-Reverse binder swap is dual to the binder swap transformation,
-substituting occurrences of the case
-binder $z$ by the scrutinee in the special case when the scrutinee is a
-variable $x$.
+Reverse binder swap is dual to the binder swap transformation, substituting
+occurrences of the case binder $z$ by the scrutinee in the special case when
+the scrutinee is a variable $x$.
 %
@@
 % does in hope of eliminating multiple uses of $x$ (allowing for inlining). 
-By
-using the scrutinee $x$ instead of the case binder, we might be able to float
-out expressions from the alternative using the case binder. For example, we
-might float an expensive computation involving $z$ out of the case alternative,
-where $z$ is now out of scope:
+By using the scrutinee $x$ instead of the case binder, we might be able to
+float out expressions from the alternative using the case binder. For example,
+we might float an expensive computation involving $z$ out of the case
+alternative, where $z$ is now out of scope:
 
@@
 
-Despite GHC applying the reverse binder swap transformation 
-during its Core-to-Core optimisation passes, this optimisation actually \emph{breaks} linearity
-when considered as a transformation on Linear Core programs. In practice, the
-optimisation preserves linearity when applied as part of the GHC
-pipeline due to the occurrence analyser being naive with
-regard to semantic linearity. Initially, it might seem as though an expression
-in which a variable $x$ occurs both in the case scrutinee and in the
-alternatives is linear (e.g.,~$\ccase{x}{\_ \to x}$):
+Even though GHC applies the reverse binder swap transformation in the
+Core-to-Core passes, \emph{this optimisation violates linearity} when considered
+as a transformation \emph{on Linear Core programs} because the result (second program
+above) is ill-typed -- the resource $x$ of the scrutinee not in WHNF is used
+again in an alternative, despite it no longer being available according to our system.
+
+An expression in which a variable $x$ occurs both in the case scrutinee and in
+the alternatives (e.g.,~$\ccase{x}{\_ \to x}$) could be considered linear if
+considered in isolation:
 % \[
@@
 % \item
-  If $x$ refers to an unevaluated expression, then scrutinizing it results
-in the expression bound by $x$ to be evaluated to WHNF. In a subsequent use of
+If $x$ refers to an unevaluated expression, then scrutinizing will force the expression
+bound by $x$ to be evaluated to WHNF. In a subsequent use of
 $x$ in the alternatives, $x$ refers to the evaluated scrutinee in WHNF, which
@@
 mutual exclusion with the case binder and pattern variables).
-%\end{itemize}
-%
-% This idea would be encoded by a possible rule like $$\TypeCaseVar$$
-%
-
-Even though, on its own, it makes intuitive sense that this example indeed uses
-$x$ linearly, when considered as part of a complete type system, allowing this
-expression to be linear makes the system \emph{unsound}.
-To see why this is the case, recall that $\beta$-reduction reduces an application of a linear function
-using call-by-name -- if we know the argument is used exactly once, a binding
-to share the result of computing the argument is unnecessary, so we instead
-substitute the argument expression for the linearly-bound variable in the
-$\lambda$-body directly.
-
+
+However, allowing this expression to be linear makes $\lambda^\pi_\Delta$ unsound!
 Consider the function application $(\lambda x.~\ccase{x}{\_ \to x})~(use~y)$
-where $y$ is a free linear variable. Assuming $\lambda x.~\ccase{x}{\_ \to x}$
-is a linear function, $\beta$-reduction transforms the
-application into $\ccase{use~y}{\_ \to
-  use~y}$. Now, $y$ is a linear variable \emph{consumed} in the scrutinee of the case
-that also occurs in the case alternative --
-linearity is violated by using the linear resource $y$ twice.
-%
-% Essentially, Linear Core would be unsound, and even duplicate resources, if the
-% above kind of expressions, where linear variable scrutinees occur in the
-% alternatives body, were well-typed, because of its interaction with the
-% call-by-name reduction of linear functions.
-In this sense, the reverse binder
-swap is an optimisation that creates ill-typed expressions from well-typed
-ones, so it is deemed an \emph{invalid} optimisation that does not preserve types in
-our system.
-
-Reverse binder swap is not a problem in the GHC simplifier because of the
-weaker notion of linearity understood by occurrence analysis. Occurrence
-analysis is a static analysis pass which can be used to determine whether a
-lambda application can be $\beta$-reduced call-by-name, and $\ccase{x}{\_ \to
-x}$ is \emph{not} seen as using $x$ linearly by the analysis. Thus,
-$\beta$-reduction is done with call-by-need on such an expression. If the above
-example were reduced with call-by-need
-% \[
-% \begin{array}{l}
-% (\lambda x.~\ccase{x}{\_ \to x})~(use~y)
-% \Longrightarrow_\textrm{CBNeed}
-% \llet{x = use~y}{\ccase{y}{\_ \to y}}
-% \end{array}
-% \]
-then the computation using $y$ would be let-bound, and $y$ used as a scrutinee
-variable, which is indeed semantically linear in $x$.
+where $y$ is a free linear variable, and assume $\lambda x.~\ccase{x}{\_ \to x}$
+to be a linear function. According to $\lambda^\pi_\Delta$'s call-by-name
+$\beta$-reduction, we have:
+\[
+\begin{array}{l}
+(\lambda x.~\ccase{x}{\_ \to x})~(use~y)
+\Longrightarrow_\textrm{CBName}
+\ccase{use~y}{\_ \to use~y}
+\end{array}
+\]
+After the reduction, $y$ is a linear variable \emph{consumed} in the scrutinee
+of the case that also occurs in the case alternative -- the resource
+$y$ is being effectively duplicated.
+%
+If the application were instead reduced with call-by-need
+\[
+\begin{array}{l}
+(\lambda x.~\ccase{x}{\_ \to x})~(use~y)
+\Longrightarrow_\textrm{CBNeed}
+\llet{x = use~y}{\ccase{x}{\_ \to x}}
+\end{array}
+\]
+then the computation using $y$ would be let-bound, and $y$ would be used as a
+scrutinee variable, which preserves semantic linearity.
+
+For an optimising compiler such as GHC, the reverse binder swap transformation
+violating linearity in Linear Core is relevant because optimising
+compilers do not necessarily allocate a binding/use call-by-need for
+every function application -- our result shows that failing to do so
+when calling a function such as the one above is \emph{unsound}.
+%
+In an optimising compiler, a binding may be deemed unnecessary if the function
+argument is inferred to be used exactly once. Linearity can play a
+(non-heuristic) role in this inference, for a linear function uses its
+argument exactly once by definition.
+%
+% Since the compiler may leverage this information (we
+% briefly discuss \emph{linearity-influenced optimisations} in
+% Section~\ref{sec:discussion}).
+%
+However, it is incompatible to both type expressions like $(\lambda
+x.~\ccase{x}{\_ \to x})$ as linear functions and consider
+call-by-name $\beta$-reduction of linear functions unconditionally
+safe (without risk of duplicating resources nor work).
+%
+Furthermore, considering such an use of $x$ to be linear also makes inlining of
+linear let bindings no longer unconditionally safe.
+
+% $\beta$-reduction reduces an application of a linear function using
+% call-by-name -- If we know the argument is used exactly once, a binding to
+% share the result of computing the argument is unnecessary, so we instead
+% substitute the argument expression for the linearly-bound variable in the
+% $\lambda$-body directly.
+%%
+%% Essentially, Linear Core would be unsound, and even duplicate resources, if the
+%% above kind of expressions, where linear variable scrutinees occur in the
+%% alternatives body, were well-typed, because of its interaction with the
+%% call-by-name reduction of linear functions.
+%In this sense, the reverse binder
+%swap is an optimisation that creates ill-typed expressions from well-typed
+%ones, so it is deemed an \emph{invalid} optimisation that does not preserve types in
+%our system.
+
+%% Reverse binder swap is not a problem in the GHC simplifier because of the
+%% weaker notion of linearity understood by occurrence analysis. Occurrence
+%% analysis is a static analysis pass which can be used to determine whether a
+%% lambda application can be $\beta$-reduced call-by-name, and $\ccase{x}{\_ \to
+%% x}$ is \emph{not} seen as using $x$ linearly by the analysis. Thus,
+%% $\beta$-reduction is done with call-by-need on such an expression.
+%
+% In practice, the optimisation preserves linearity when applied as part of the
+% GHC pipeline due to the occurrence analyser being naive with regard to semantic
+% linearity.
 
@@
         discarded a linear resource, and that rewrite rules implementing stream
-        fusion~\cite{10.1145/165180.165214,10.1145/1291151.1291199,10.1007/978-3-642-24276-2_2}, and
+        fusion~\cite{10.1145/165180.165214,10.1145/1291151.1291199}, and
         others too, could produce linearly-invalid programs (in one such case,

### 2024-02-18 icfp.tex
RM edits related & future work

@@
 %
-We propose a naive $O(n^2)$ algorithm for inferring usage environments of recursive
-bindings in Section~\ref{sec:discuss:implementation}, orthogonally to the theory
-developed in this section.
+Our implementation uses a naive $O(n^2)$ algorithm for inferring usage
+environments of recursive bindings.
+% , which is not relevant to the theory developed in this section.
 %
@@
 %
-However, we note that inference of usage environments for recursive binding
+Moreover, we note that inference of usage environments for recursive binding
 groups bears some resemblance to the inference of principle types for recursive
@@
 % %
-Linear Haskell~\cite{cite:linearhaskell} augments Haskell with 
-linear types. However, the development is
-not concerned with extending GHC's intermediate language(s),
-which presents its own challenges. 
-%
-Nonetheless, while Linear Haskell keeps GHC Core unchanged, its
-implementation in GHC does modify and extend Core with linearity/multiplicity
-annotations. Core's type system is unable to type \emph{semantic} linearity of
-programs in the sense elaborated in our work, which results in Core
-rejecting linear programs resulting from optimising transformations that
-leverage the non-strict semantics of Core.
+Linear Haskell~\cite{cite:linearhaskell} augments the Haskell surface language
+with linear types, but it is not concerned with extending GHC's intermediate
+language(s), which presents its own challenges. 
+%
+However, the implementation Linear Haskell in GHC does modify and extend Core
+with linearity/multiplicity annotations. Core's type system is unable to type
+\emph{semantic} linearity of programs in the sense elaborated in our work,
+which results in Core rejecting most linear programs resulting from optimising
+transformations that leverage the non-strict semantics of Core.
 %
 Our work overcomes the limitations of Core's linear type system derived from
-Linear Haskell by understanding linearity semantically, in the presence of laziness,
-and accepting multiple Core-to-Core passes.
-Linear Core can
-also be seen as a system that validates the programs written in Linear Haskell
-and are compiled by GHC, by guaranteeing (through typing) that linear resources
-are still used exactly once throughout the optimising transformations.
+Linear Haskell by understanding linearity semantically in the presence of
+laziness, and by proving linearity-preserving multiple Core-to-Core
+optimisations GHC.
+Linear Core can also be seen as a system that validates the programs written in
+Linear Haskell and are compiled by GHC, by guaranteeing (through typing) that
+linear resources are still used exactly once throughout the optimising
+transformations.
 
@@
 %
-By fully supporting linear types in Core, these optimising transformations
-could be informed by the language's intrinsic linearity, and, consequently, avoid
-an ad-hoc or incomplete linear type inference pass that is custom-built for
-optimisations. Additionally, the linearity information may potentially be used
-to the benefit of optimising transformations that currently do not take 
-linearity into account.
+Preserving linear types in Core throughout the compilation pipeline allows the
+optimiser to leverage non-heuristic linearity information where, previously, it
+would resort solely to ad-hoc or incomplete custom-built linearity inference
+passes (naturally, these passes are still necessary to optimise programs not
+using explicit linearity or multiplicity annotations). Linearity may
+potentially be used to the benefit of other optimising transformations that
+currently do not take it into account.
 
@@
 allow the Core intermediate language to encode a panoply of Haskell source
-type-level features such as GADTs, type families or newtypes.
+type-level features such as GADTs, type families, or newtypes.
 %
@@
 variables.  Linearity can be used to non-heuristically inform
-the inliner~\cite{cite:linearhaskell}. Additionally, we argue that in Linear
-Core accepting more programs as linear there are more chances to use linearity,
-in contrast to a linear type system which does not account for lazy evaluation
-and thus rejects more programs.
-
-\paragraph{Usage environment resource inference.}
-In Section~\ref{sec:linearity-semantically}, we explained that the linear
-resources used by a group of recursive bindings aren't obvious and must be
-consistent with each other (i.e. considering the mutually-recursive calls) as
-though the resources used by each binder are the solution to a set determined
-by the recursive bindings group.  In Section~\ref{sec:main:linear-core}, we
-further likened the challenge of determining usage environments for a recursive
-group of bindings to a unification problem as that solved by the Hindley-Milner
-type inference algorithm~\cite{DBLP:conf/popl/DamasM82} based on generating and solving
-constraints. Even though these are useful observations, our implementation of
-Linear Core uses a naive algorithm to determine the usage environments, thereby
-leaving as future work the design of a principled algorithm to determine the
-usage environments of recursive group of bindings.
+the inliner~\cite{cite:linearhaskell}.
+% Additionally, we argue that in Linear Core accepting more programs as linear
+% there are more chances to use linearity, in contrast to a linear type system
+% which does not account for lazy evaluation and thus rejects more programs.
+
+% \paragraph{Usage environment resource inference.}
+% In Section~\ref{sec:linearity-semantically}, we discussed how there is not an obvious way to determine the set of linear
+% resources used by a group of (mutually) recursive bindings, .
+% though the resources used by each binder are the solution to a set determined
+% by the recursive bindings group.  In Section~\ref{sec:main:linear-core}, we
+% further likened the challenge of determining usage environments for a recursive
+% group of bindings to a unification problem as that solved by the Hindley-Milner
+% type inference algorithm~\cite{DBLP:conf/popl/DamasM82} based on generating and solving
+% constraints. Even though these are useful observations, our implementation of
+% Linear Core uses a naive algorithm to determine the usage environments, thereby
+% leaving as future work the design of a principled algorithm to determine the
+% usage environments of recursive group of bindings.
 
@@
 %
-Implementing Linear Core in GHC is a challenging endeavour, since we must account
-for all other Core features (e.g. strict constructor fields) and more
-optimisations. Despite our initiative in this
-direction\footnote{[Link removed due to anonymization]}%\footnote{https://gitlab.haskell.org/ghc/ghc/-/issues/23218}
-, we leave
-this as future work.
+Implementing Linear Core in GHC is a challenging endeavour, since we must
+account for all other Core features (e.g. strict constructor fields), propagate
+new or modified data throughout the entire pipeline, and accept more
+optimisations. Despite our initiative in this direction\footnote{[Link removed
+due to
+anonymization]}%\footnote{https://gitlab.haskell.org/ghc/ghc/-/issues/23218} ,
+we leave this as future work.
 

### 2024-02-18 icfp.tex
.

--- 2024-02-17 icfp.tex
@@
 shared. Indeed, if the result of the computation involving the linear resource
-was insteaed, e.g., an unrestricted integer, then sharing the result would not
+was instead, e.g., an unrestricted integer, then sharing the result would not
 involve consuming the resource more than once.
@@
 
-As opposed to scrutinees in WHNF, where the resources used to type a scrutinee
-can be cleanly divided amongst the various sub-expressions of e.g. the
-constructor and henceforth each pattern variable, there is no direct mapping
-between the resources typing a scrutinee \emph{not in WHNF} and the usage
-environments of pattern variables in any alternative.
-
-%
-% That is, there is no direct mapping between the usage environments of the
-% linear pattern-bound variables and the resources used in the scrutinee.
+Unlike with scrutinees in WHNF, the resources used by a
+scrutinee not in WHNF do not necessarily match those used by each
+sub-expression of the expression evaluated to WHNF.
+%
+Therefore, there is no direct mapping between the usage environments of the
+linear pattern-bound variables and the resources used in the scrutinee.
 % \todo{WAIT, there never was!
@@
 %
-We introduce \emph{tagged resources} as a means to enforce all pattern-bound
-variables for a scrutinee not in WHNF are either \emph{jointly} used to consume
-all resources occurring in the environment, or not at all (instead, the case
-binder may be used). Given linear resources $[\D_s]$ used to type a scrutinee,
-and a pattern $K~\ov{x_\omega},\ov{y_i}$ with $i$ linear components, we assign
-a usage environment $\D_i$ to each linear pattern variable where $\D_i$ is
-obtained from the scrutinee environment \emph{tagged} with the constructor name
-and linear-variable index $\lctag{\irr{\D_s}\!}{K_i}$. $y_i{:}_{\D_i}\s$ is
-introduced in $\G$, just like other $\D$-variables.
+We introduce \emph{tagged resources} to guarantee all linearly-bound pattern
+variables are jointly used to consume all resources occurring in the
+environment (in alternative to the case binder), or not at all. Given linear
+resources $[\D_s]$ used to type a scrutinee, and a pattern
+$K~\ov{x_\omega},\ov{y_i}$ with $i$ linear components, we assign a usage
+environment $\D_i$ to each linear pattern variable where, $\D_i$ is obtained from the
+scrutinee environment \emph{tagged} with the constructor name and linear-variable
+index $\lctag{\D_s}{K_i}$, and $\y[\D_i]$ is introduced in $\G$.
 \[
@@
 \]
-%
-% The tag consists of a constructor name $K$ and an index $i$ identifying the
-% position of the pattern variable among all bound variables in that pattern.
-%
-Having tagged the resources in the usage environment of each pattern variable
-differently, we need now only to express that (1) the pattern $\D$-variables
-can be used (i.e., the tagged resources need be available in the linear
-environment in order for the $\textsc{Var}_\D$ rule to be applicable) and (2)
-if we consume one of the pattern variables with a non-empty usage environment,
-then \emph{all} pattern variables for linear components must also be consumed.
-
-The key idea to guarantee these is to provide a $\textsc{Split}$ rule that
-allows a linear resource $x{:}_1\s$ to be split into $n$ resources at a given
-constructor $K$, where $n$ is the number of linear components of the
-constructor, and each resource resulting from the split is \emph{tagged} with
-$K$ and the positional index of a linear component (i.e. $x{:}_1\s$ can be
-split into $\ov{\lctag{x{:}_1\s}{K_i}}^n$).
-%
-By assigning to each pattern variable a \emph{fragment} of the scrutinee
-resources (with a tag), we require the scrutinee resources to be
-$\textsc{Split}$ in order to use pattern variables at all.  Moreover, the
-remaining tagged resources cannot be used directly, yet need still be consumed
-exactly once. Thus, the choice to consume the tagged resources via usage
-environments of the other pattern variables is forced, ensuring that no
-variable for a constructor's linear component can go unused (the tagged environments are disjoint).
-
-For instance, in the term $\lambda x.~\ccase{f~x}{z~\{K~a~b\to
+The tag consists of a constructor name $K$ and an index $i$ identifying the
+position of the pattern variable among all bound variables in that pattern.
+%
+The key idea is that a linear resource $x$ can be split into $n$ resources at a
+given constructor, where $n$ is the number of positional linear arguments of
+the constructor.
+%
+This is given by the Split rule. By assigning to each linear pattern
+variable a fragment of the scrutinee resources with a tag, we
+guarantee that all linear pattern variables are simultaneously used to
+consume all the scrutinee resources.
+For instance, in the term $\lambda x.~\ccase{x}{z~\{K~a~b\to
   (a,b)\}}$, where $x$ is a linear variable, the case alternative is
@@
 $\var[b][\lctag{\irr{x}}{K_2}]$, assuming both components of $K$ are linear.
-%
-The occurrences of $a$ and $b$ are well-typed because we can first
-$\textsc{Split}$ $[x{:}_1\s]$ into $\lctag{[x{:}_1\s]}{K_1}, \lctag{[x{:}_1\s]}{K_2}$,
-%
+\todo[inline]{Previous explanation was very confusing, try to explain
+  what the split is doing in terms of the example!}
 % since for any of the scrutinee
@@
 % have to be consumed through the other linear pattern-bound variables.
-%
-noting that $Split$ can be applied both to relevant and proof irrelevant linear
-resources in $\D$.
+
+%
+We note how $Split$ can be applied both to relevant and proof irrelevant linear resources in $\D$.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% {{{ Metatheory
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Metatheory of Linear Core\label{sec:main:metatheory}}
-
-The $\lambda^\pi_\D$ type system is sound: well-typed programs in Linear Core do not
-get \emph{stuck}. Besides type safety, we
-prove that multiple optimising transformations used in GHC preserve
-linearity in the sense of being well-typed in our system.
-%
-% Additionally, we state our assumptions that outline an isomorphism between
-% using a linear variable $\xl$ and a $\D$-variable $\xD$ that consumes existing
-% resources $\D$, for any $\D$.
-
-\renewcommand{\DeltaLinearRelationLemma}{
-\begin{restatable}[$\Delta$-bound to Linear]{lemma}{deltaone}\label{lem:deltaone}
-% Any $\Delta$-variable with a fully-irrelevant usage environment can be made a linear variable as long as its irrelevant environment is dropped from the linear environment (and the unrestricted env.
-is updated accordingly).\\
-If $\G,\x[\irr{\D}]; \irr{\D},\D' \vdash e : \vp$ 
-then $\G[x/\irr{\D}]; \D',\xl \vdash e :\vp$.
-\end{restatable}
-}
-
-\renewcommand{\LinearDeltaRelationLemma}{
-  \begin{restatable}[Linear to $\Delta$-bound]{lemma}{onedelta}\label{lem:onedelta}
-If $\G; \D',\xl \vdash e :\vp$
-then $\G[\D/x],\xD; \D,\D' \vdash e : \vp$ ($\Delta$ fresh).
-\end{restatable}
-}
-
-\renewcommand{\DeltaUnrestrictedRelationLemma}{
-\begin{restatable}[Unrestricted and $\Delta$-bound]{lemma}{undelta}\label{lem:undelta}
-%An unrestricted variable is equivalent to a $\D$-var with an empty usage environment.\\
-$\G,\xo; \D \vdash e : \vp$ iff $\G,\x[\cdot]; \D \vdash e : \vp$
-\end{restatable}
-}
-
-\paragraph{$\Delta$-bound variables} % Our development relies on a characterization of $\D$-bound
-% variables
-A well-typed program with a linear variable ($\xl$) is equivalently
-well-typed if the linear variable were instead $\D$-bound ($\xD$) with
-usage environment $\D$, with $\D$ available in the linear context (and the unrestricted context updated accordingly).
-
-\LinearDeltaRelationLemma
-
-\noindent Dually, any $\Delta$-bound variable with an irrelevant usage
-environment can be made into a linear variable as long as its
-(irrelevant) environment is dropped from the linear context.
-
-% Dually, a well-typed program with resources $\D$ and $\D$-bound
-% variable ($\xD$) is equivalently well-typed, as long as $\D$ is consumed
-% through the use of $x$, if $x$ is moved to the linear context, resources $\D$
-% are removed from the linear context, and occurrences of $\D$ whole in usage
-% environments are substituted by $x$ (occurrences of fragments of $\D$ in usage
-% environments are unimportant since $\D$ was consumed whole by $x$, not by any
-% of the fragment-using $\D$-vars).
-
-\DeltaLinearRelationLemma
-
-% Intuitively, a linear variable
-
-\noindent Finally, unrestricted and $\D$-bound
-variables with an empty usage environment are equivalent.
-
-\DeltaUnrestrictedRelationLemma
-
-\paragraph{Irrelevance}
-%
-As discussed, proof irrelevant resources can only be consumed
-indirectly, essentially encoding that the (non-WHNF) case scrutinee
-resources must be consumed through the case binder or the linear
-pattern-bound variables.
-%
-As a case expression is evaluated, the scrutinee will eventually
-reduce to WHNF,
-which must then be typed with rule $Case_{\textrm{WHNF}}$.
-%
-Crucially, the two case rules must be in harmony in the system, in the sense that
-case expressions typed using the $Case_{\textrm{Not WHNF}}$ rule must also be
-well-typed by the $Case_{\textrm{WHNF}}$ once the scrutinee is evaluated to WHNF.
-
-% The type preservation theorem states that a well-typed expression
-% remains well-typed in the presence of evaluation. Specifically, when the case
-% expression whose scrutinee is evaluated to WHNF is handled in the
-% preservation proof.
-
-% The \emph{Irrelevance} lemma is required to prove preservation for that
-% evaluation case. We need to prove that the alternatives of a case expression
-% typed with proof irrelevant resources are still well-typed when the proof
-% irrelevant resource is substituted by the scrutinee resources as it is evaluated to WHNF.
-% In this sense, the \emph{Irrelevance} lemma witnesses the soundness of typing a
-% case alternative with proof irrelevant resources in a certain context with respect to
-% typing the same expression with arbitrary resources (we note, however, typing
-% an alternative with proof irrelevant resources is not complete wrt using
-% arbitrary resources -- a counter example needs only to use a resource
-% directly).
-
-\WHNFConvSoundness
-
-The \emph{Irrelevance} lemma witnesses the soundness of typing a
-case alternative with proof irrelevant resources with respect to
-typing the same expression with arbitrary resources. However, typing
-a case alternative with proof irrelevant resources is not complete wrt using
-arbitrary resources -- a counter example needs only to use a resource
-directly.
-
-% \noindent Intuitively, the lemma holds since proof irrelevant resources can
-% only be used through the case binder or pattern-bound variables. If we
-% consistently replace the proof irrelevant resources both in the typing
-% environment and in the usage environments containing them, the expression
-% remains well-typed. We note that the proof irrelevant resources are always
-% unique when introduced in such a case alternative (we always take the scrutinee
-% environment to make irrelevant, and allow nested ``irrelevantness''), so the
-% case binder has the only occurrence of those resources in the $\G$ environment.
-% % (being somewhat akin to congruence).
-% %
-% The proof is given in Section~\ref{sec:proof:irrelevance}.
-
-\subsection{Type safety\label{sec:type-safety-meta}}
-
-We prove type safety of the Linear Core system via the standard type
-preservation and progress results. As is customary, we make use of multiple
-substitution lemmas, one for each kind of variable: unrestricted variables
-$\xo$, linear variables $\xl$, and $\D$-bound variables $\xD$.
-
-% We start with the auxiliary results, as we will make use of them in a select part of the preservation proof.
-
-\TypePreservationTheorem
-%
-\ProgressTheorem
-% \noindent Type preservation states that a well-typed expression $e$ that
-% evaluates to $e'$ remains well-typed under the same context:
-%
-Type preservation
-% ($\S$~\ref{sec:proof:type-preservation})
-follows by structural induction on the reductions $e \longrightarrow
-e'$ from the operational semantics. Most cases are straightforward, appealing to one or more
-substitution lemmas (see Appendix~\ref{sec:proof:type-preservation}). The most
-interesting case is that of case expressions whose scrutinee can be further
-evaluated -- we branch on whether the scrutinee becomes in WHNF and appeal to Lemma~\ref{lem:irrev} if so.
-%
-% This case guarantees that the separation of rules for treating scrutinees is
-% consistent, in the sense that a well-typed case expression with a scrutinee not
-% in WHNF remains well-typed after the scrutinee is evaluated to WHNF.
-%
-%
-%
-%\noindent
-% Progress states that the evaluation of a well-typed term does not block:
-% Similarly, progress is proved by induction on typing ($\S$~\ref{sec:proof:progress}).
-
-% \subsubsection{Substitution Lemmas}
-
-% The preservation and progress theorems depend on multiple substitution lemmas,
-% one for each kind of variable, as is standard.
-% % The proofs are given in Section~\ref{sec:proof:substitution-lemmas}.
-% % The ... themselves diverge from their common formulation, because
-% % $\D$-variables refer to linearly bound variables, so substitution must take into account
-
-% The linear substitution lemma states that a well-typed expression $e$ with a
-% linear variable $x$ of type $\s$ remains well-typed if
-% occurrences of $x$ in the $e$ are replaced by an expression $e'$ of the same
-% type $\s$, and occurrences of $x$ in the linear context and in usage
-% environments of $\D$-bound variables are replaced by the linear context $\D'$
-% used to type $e'$:
-
-% \LinearSubstitutionLemma
-
-% \noindent Where $\G[\D'/x]$ substitutes all occurrences of $x$ in the usage
-% environments of $\D$-variables in $\G$ by the linear variables in $\D'$.
-% % ($x$ couldn't appear anywhere besides usage environments of $\D$-bound variables, since $x$ is linear).
-
-% The substitution of the resource in the usage environments is illustrated
-% by the following example. Consider the term $\llet{y = use~x}{y}$ where $use$ and $x$ are free variables:
-% if we replace occurrences of $x$ by $e'$ (where $\G;\D \vdash e : \s$), then the
-% ``real'' usage environment of $y$ goes from $\{x\}$ to $\D$. If we don't update
-% the usage environment of $y$ accordingly, we'll ultimately be typing
-% $y{:}_{\{x\}}\vp$ with $\D$ instead of $x$, which is not valid.
-
-% The linear substitution lemma extends to case alternatives as well.
-% The lemma for substitution of linear variables in case alternatives is similar
-% to the linear substitution lemma, applied to the case alternative judgement.
-% % Is slightly more involved, in the sense that there are more environments in which the
-% % substitution $[\D/x]$ must be applied (for the same reason):
-% %
-% \LinearSubstitutionAltsLemma
-% %
-% \noindent We further require that the environment annotated in the case
-% alternative judgement, $\D_s$, is a subset of the environment used to type the
-% whole alternative $\D_s \subseteq \D$. In all occurrences of the alternative
-% judgement (in $Case_{\textrm{WHNF}}$ and $Case_{\textrm{Not WHNF}}$), the
-% environment annotating the alternative judgement is \emph{always} a subset of
-% the alternative environment.
-
-% The substitution lemma for unrestricted variables follows the usual
-% formulation, with the added restriction (common to linear type systems) that
-% the expression $e'$ that is going to substitute the unrestricted variable $x$
-% is typed on an empty linear environment:
-% %
-% \UnrestrictedSubstitutionLemma
-% %
-% \noindent Similarly, we also prove the substitution of unrestricted variables preserves types on an alternative case expression:
-% %
-% \UnrestrictedSubstitutionAltsLemma
-
-% Finally, we introduce the lemma stating that substitution of $\D$-bound
-% variables by expressions of the same type preserves the type of the original
-% expression.
-% %
-% What distinguishes this lemma from traditional substitution lemmas is that the
-% usage environment $\D$ of the variable $x$ being substituted by expression $e'$
-% must match exactly the typing environment $\D$ of $e'$ and the
-% environment of the original expression doesn't change with the substitution:
-% %
-% \DeltaSubstitutionLemma
-% %
-% \noindent Intuitively, if $x$ is well-typed with $\D$ in $e$, substituting $x$
-% by an expression $e'$ which is typed in the same environment $\D$ allows the
-% distribution of resources $\D,\D'$ used to type $e$ across sub-derivations to remain
-% unchanged. To prove the theorems, we don't need a ``stronger'' substitution of
-% $\D$-vars lemma (allowing arbitrary resources $\D''$ to type $e'$, as in other
-% substitution lemmas), as we only ever substitute $\D$-variables by expressions
-% whose typing environment matches the variables usage environment. However, it
-% is not obvious whether such a lemma is possible to prove for $\D$-variables
-% (e.g. let $\G;\D \vdash e :\s$ and $\G; \D' \vdash \llet{x = e'}{x}$, if we
-% substitute $e$ for $x$ the resources $\D'$ are no longer consumed).
-
-% The $\D$-substitution lemma on case alternatives reflects again that the typing
-% environment of the expression substitution the variable must match its usage
-% environment. We recall that $\D_s \subseteq \D,\D'$ states that the annotated
-% environment is always contained in the typing environment, which is true of all
-% occurrences of this judgement. An alternative formulation of this lemma could
-% instead explicitly list $\D_s$ as part of the typing environment for the same
-% effect:
-
-% \DeltaSubstitutionAltsLemma
-
-% The proofs for substitution lemmas of linear, unrestricted, and
-% $\D$-variables are available in Section~\ref{sec:proof:substitution-lemmas}.
-
-%TODO! Substitution of proof-irrelevant linear variables preserves typing. The
-%term always remains the same because $x$ cannot occur in any term, however, all
-%variables that refer to $x$ in their usage environment must now refer the usage env. of the substitee (e.g. $[x] => [\D]$).
-%This seems trivial to see correct, since all occurrences are in environments, so we get some equivalence similar to the one we need for the proof of Alt0.
-%
-%TODO: Multiplicity substitution preserves typing lemma
-%
-%TODO: Canonical forms lemma
-%
-%TODO: Corollary of $\Delta$-var subst. for $\ov{\Delta}$
-%
-%TODO: Constructor app typing:
-%If $\Gamma, \Delta \vdash K~\ov{e}$ and $K{:}\ov{\sigma\to\pi}~T~\ov{p} \in \Gamma$ and $\hasnolinearvars{\Gamma}$
-%then $\ov{\Gamma, \Delta_i \vdash e_i : \sigma_i}$
-
-\subsection{Optimisations preserve linearity\label{sec:optimisations-preserve-types-meta}}
-%\todo[inline]{Check Commuting case-let in Fig.~\ref{fig:opttrans}!}
-A main goal of the Linear Core type system is 
-to serve as a typed intermediate representation for optimising
-compilers of lazy languages with linearity. In light of this,
-we show that multiple optimising transformations are type
-preserving in Linear Core (i.e.,~preserve linearity semantically).
-
-% The optimising transformations proved sound wrt Linear Core in this section
-% have been previously explained and motivated in
-% Section~\ref{sec:core-to-core-transformations}.
-% %
-%\{\overline{\rho\to e'}\}
-\begin{figure}[t]
-
-  {\small
-  \[
-    \begin{array}{ll}
-   (\llet{\xD = e}{e'} )  \Longrightarrow \llet{\xD =       e}{e'[e/x]} &                    \mbox{Inlining}\\
-       (\lambda \x[\pi][\s].~e)~e'  \Longrightarrow e[e'/x]  &                \mbox{$\beta$-reduction}\\
-      (\ccase{(K~\ov{e})}{\z[\D][\s]~\{\overline{\rho_j\to e_j}, K~\ov{x} \to e_i, \overline{\rho_k\to e_k}\}} )\Longrightarrow  e_i\ov{[e/x]}[K~\ov{e}/z] &
-                                  \mbox{Case of known constr.}\\
-      (\ccase{(\ccase{e_c}{\z[\D]~\{\ov{\rho_{c_i}\to e_{c_i}}\}})}{\var[w][\irr{\D,\D'}][\s']~\{\ov{\rho_i\to e_i}\}})\\
-            \Longrightarrow 
-            %{{{
-      \ccase{e_c}{\z[\D]~\{\ov{\rho_{c_i} \to \ccase{e_{c_i}}{w~\{\ov{\rho_i\to e_i}\}}}\}}
-      & \mbox{Case-of-case}\\
-      (\lambda \y[\pi].~\llet{\xD = e}{e'})
-      \Longrightarrow
-      \llet{\xD=e}{\lambda\y[\pi].~e'} & \mbox{Commuting let-$\lambda$}\\
-      ((\llet{v = e}{b})~a) \Longrightarrow \llet{v = e}{b~a} &\mbox{Commuting let-app.}\\
-      (\ccase{(\llet{v = e}{b})}{\{\overline{\rho\to e'}\}}) \Longrightarrow
-      %{{{
-      \llet{v =  e}{\ccase{b}{\{\overline{\rho\to e'}\}}}
-                               &
-                                 \mbox{Commuting let-case}\\
-      (\ccase{e}{\{\overline{\rho_j\to e_j}, \rho \to E[e_1],
-      \overline{\rho_k\to e_k}\}} ) \qquad (\mbox{$x$ fresh}) \\
-      %{{{
-     \Longrightarrow \llet{x = e_1}{\ccase{e}{\{\overline{\rho_j\to e_j}, \rho \to E[x], \overline{\rho_k\to e_k}\}}} 
-                                                                        & \mbox{Commuting case-let}\\
-      (\llet{x = (\llet{v = e}{b})}{c}) \Longrightarrow \llet{v = e}{\llet{x = b}{c}}
-                                                                      & \mbox{Commuting let-let}\\
-      f \Longrightarrow \lambda x.(f~x) & \mbox{$\eta$-expansion}\\
-      \lambda x.(f~x) \Longrightarrow f & \mbox{$\eta$-reduction}\\
-      (\ccase{x}{z~\{\ov{\rho_i\to e_i}\}}) \Longrightarrow \ccase{x}{z~\{\ov{\rho_i\to  e_i[z/x]}\}}
-                                                                      & \mbox{Binder swap}\\
-      % (\ccase{x}{z~\{\ov{\rho_i\to e_i}\}} )\Longrightarrow \ccase{x}{z~\{\ov{\rho_i\to e_i[x/z]}\}}
-      %                                                                 & \mbox{Reverse binder swap}
-    \end{array}
-  \]}
-  \caption{Optimising transformations validated by Linear
-    Core\label{fig:opttrans} (for commuting case-let we
-      require $e_1$ not to capture the case binder, variables bound by the
-      pattern $\rho$ and the context $E[{-}]$ -- see Appendix~\ref{app:optimisations} for
-    details).} 
-\end{figure}
-
-We describe Core-to-Core transformations as $e_1 \Longrightarrow e_2$,
-where $e_1$ is an arbitrary \emph{well-typed} expression
-matching a certain shape (or of a certain type), which is
-transformed into expression $e_2$. Validating a transformation
-entails showing that $e_2$ is well-typed. We often annotate the arrow
-with the name of the transformation when describing chains of such
-transformations.
-
-The list of transformations we have
-validated is given in Figure~\ref{fig:opttrans}.
-%
-The behaviour of most transformations can be inferred from its name:
-Inlining substitutes a let-bound expression in the let body; $\beta$-reduction
-and case of known constructor effectively perform evaluation of the given expression;
-case-of-case commutes cases in scrutinee position inside the outer case alternatives;
-let floating and  for let-bound expressions to be floated out of $\lambda$-abstractions;
-the four commuting conversions allow let-bound expressions to be commuted with other
-Core constructs; $\eta$-expansion and reduction apply the $\eta$-laws to terms of function type;
-finally, binder swap substitutes occurrences of a scrutinee variable for the
-case binder. While many of the transformations above are not obviously optimising,
-they are chained in the compiler pipeline to expose optimisation opportunities.
-
-% Transformations are described by an arbitrary well-typed expression with a certain shape, on
-% the left hand side (lhs) of the arrow $\Longrightarrow$, resulting in an expression on
-% the right hand side (rhs) that we prove to be well-typed.
-% %
-% % For our proofs, we assume the lhs to be a well-typed expression and prove the
-% % rhs is well-typed as well.
-% %
-% For each transformation, we describe the intuition behind the transformation
-% preserving linearity in our system.
-
-% \todo[inline]{Aqui uma lista itemizada das transformações todas. As
-%   provas todas para apendice.}
-
-%%% COMENTADO AQUI
-% \subsubsection{Inlining}
-
-% % To the best of our knowledge, there is no linear type system for which inlining
-% % preserves linearity\footnote{https://github.com/ghc-proposals/ghc-proposals/blob/master/proposals/0111-linear-types.rst\#id90}
-
-% \input{language-v4/proofs/optimizations/Inlining}
-
-% \subsubsection{\texorpdfstring{$\beta$}{Beta}-reduction}
-
-% \input{language-v4/proofs/optimizations/BetaReduction}
-
-% \subsubsection{Case of known constructor}
-
-% \input{language-v4/proofs/optimizations/CaseOfKnownConstructor}
-
-% \subsubsection{Let floating}
-
-% \input{language-v4/proofs/optimizations/LetFloating}
-
-% \subsubsection{\texorpdfstring{$\eta$}{Eta}-conversions}
-
-% \input{language-v4/proofs/optimizations/EtaConvs}
-
-% \subsubsection{Binder Swap}
-
-% The binder swap transformation applies to case expressions whose scrutinee is a
-% single variable $x$, and it substitutes occurrences of $x$ in the case
-% alternatives for the case binder $z$. If $x$ is a linear resource, $x$ cannot
-% occur in the case alternatives (as we conservatively consider variables are not
-% in WHNF), so the substitution preserves types vacuously. Otherwise, $x$ can be
-% freely substituted by $z$, since $z$ is also an unrestricted resource (it's
-% usage environment is empty because $x$ is unrestricted).
-
-% \input{language-v4/proofs/optimizations/BinderSwap}
-
-\paragraph{Reverse Binder Swap\label{sec:reverse-binder-swap-considered-harmful}}
-
-We now discuss an issue we have found with the so-called reverse binder swap
-transformation.
-%
-%
-Reverse binder swap is dual to the binder swap transformation,
-substituting occurrences of the case
-binder $z$ by the scrutinee in the special case when the scrutinee is a
-variable $x$.
-%
-% his is dual to what the binder swap transformation
-% does in hope of eliminating multiple uses of $x$ (allowing for inlining). 
-By
-using the scrutinee $x$ instead of the case binder, we might be able to float
-out expressions from the alternative using the case binder. For example, we
-might float an expensive computation involving $z$ out of the case alternative,
-where $z$ is now out of scope:
-
-{\small\[
-\begin{array}{l}
-(\lambda x.~\lletrec{go~y = \ccase{x}{z~\{(a,b) \to \ E[expensive~z] \} }}{\dots})
-\Longrightarrow_\textrm{Reverse binder swap}\\
-(\lambda x.~\lletrec{go~y = \ccase{x}{z~\{(a,b) \to E[expensive~x] \} }}{\dots})
-\Longrightarrow_\textrm{Commuting case-let}\\
-\lambda x.~\llet{t = \mathit{expensive}~x}{(\lletrec{go~y = \ccase{x}{z~\{(a,b) \to E[t] \} }}{\dots})}\\
-\end{array}
-\]}
-
-\noindent If $go$ is compiled into a loop, we only compute
-$\mathit{expensive}~x$ once instead of in every loop iteration.
-
-Despite GHC applying the reverse binder swap transformation 
-during its Core-to-Core optimisation passes, this optimisation actually \emph{breaks} linearity
-when considered as a transformation on Linear Core programs. In practice, the
-optimisation preserves linearity when applied as part of the GHC
-pipeline due to the occurrence analyser being naive with
-regard to semantic linearity. Initially, it might seem as though an expression
-in which a variable $x$ occurs both in the case scrutinee and in the
-alternatives is linear (e.g.,~$\ccase{x}{\_ \to x}$):
-% \[
-% \G; \xl \vdash \ccase{x}{\_ \to x} : \s
-% \]
-% The reasoning is done by branching on whether $x$ refers to an expression in
-% WHNF or an unevaluated thunk.
-% \begin{itemize}
-% \item
-  If $x$ refers to an unevaluated expression, then scrutinizing it results
-in the expression bound by $x$ to be evaluated to WHNF. In a subsequent use of
-$x$ in the alternatives, $x$ refers to the evaluated scrutinee in WHNF, which
-must be consumed. Since $x$ is just another name (like the case binder) for the
-scrutinee in WHNF, we may use it instead of the case binder or pattern
-variables.
-%
-On the other hand, if $x$ already refers to an expression evaluated to WHNF, then
-scrutinizing it in a case expression is a no-op, thus we may use it again (in
-mutual exclusion with the case binder and pattern variables).
-%\end{itemize}
-%
-% This idea would be encoded by a possible rule like $$\TypeCaseVar$$
-%
-
-Even though, on its own, it makes intuitive sense that this example indeed uses
-$x$ linearly, when considered as part of a complete type system, allowing this
-expression to be linear makes the system \emph{unsound}.
-To see why this is the case, recall that $\beta$-reduction reduces an application of a linear function
-using call-by-name -- if we know the argument is used exactly once, a binding
-to share the result of computing the argument is unnecessary, so we instead
-substitute the argument expression for the linearly-bound variable in the
-$\lambda$-body directly.
-
-Consider the function application $(\lambda x.~\ccase{x}{\_ \to x})~(use~y)$
-where $y$ is a free linear variable. Assuming $\lambda x.~\ccase{x}{\_ \to x}$
-is a linear function, $\beta$-reduction transforms the
-application into $\ccase{use~y}{\_ \to
-  use~y}$. Now, $y$ is a linear variable \emph{consumed} in the scrutinee of the case
-that also occurs in the case alternative --
-linearity is violated by using the linear resource $y$ twice.
-%
-% Essentially, Linear Core would be unsound, and even duplicate resources, if the
-% above kind of expressions, where linear variable scrutinees occur in the
-% alternatives body, were well-typed, because of its interaction with the
-% call-by-name reduction of linear functions.
-In this sense, the reverse binder
-swap is an optimisation that creates ill-typed expressions from well-typed
-ones, so it is deemed an \emph{invalid} optimisation that does not preserve types in
-our system.
-
-Reverse binder swap is not a problem in the GHC simplifier because of the
-weaker notion of linearity understood by occurrence analysis. Occurrence
-analysis is a static analysis pass which can be used to determine whether a
-lambda application can be $\beta$-reduced call-by-name, and $\ccase{x}{\_ \to
-x}$ is \emph{not} seen as using $x$ linearly by the analysis. Thus,
-$\beta$-reduction is done with call-by-need on such an expression. If the above
-example were reduced with call-by-need
-% \[
-% \begin{array}{l}
-% (\lambda x.~\ccase{x}{\_ \to x})~(use~y)
-% \Longrightarrow_\textrm{CBNeed}
-% \llet{x = use~y}{\ccase{y}{\_ \to y}}
-% \end{array}
-% \]
-then the computation using $y$ would be let-bound, and $y$ used as a scrutinee
-variable, which is indeed semantically linear in $x$.
-
-% Concluding, in being able to understand more programs as linear, our type
-% system allows more expressions to be considered linear for $\beta$-reduction
-% without a let-allocation, however, it makes reverse binder swap an invalid
-% transformation since its output, when considered linear, might violate
-% linearity when further optimised.
-
-% (Link to ticket)
-% 
-% \begin{tabbing}
-% (1) All optimisations preserve (semantic) linearity\\
-% (2) If a function is (semantically) linear, then we can evaluate it using call-by-name and preserve linearity\\
-% (3) Reverse binder swap is an optimisation\\
-% (4) If reverse binder swap is applied to a case scrutinizing a linear resource in the body (`e`) of a linear function `f`, then the function is still linear by (1)\\
-% (5) If we evaluate `f`, we do it call-by-name because of (2)\\
-% (6) Call-by-name substitution of the linear argument in the body of a function has been reverse binder swapped doesn’t preserve linearity\\
-% (7) Contradiction: By 3 and 1, `f` is linear after reverse-binder-swap.
-By 2, we can substitute arguments to `f` call-by-name and preserve linearity, which contradicts with 6 that says call-by-name substitution after reverse binder swap violates linearity\\
-% \end{tabbing}
-
-% Conclusion:
-% Either we need to forfeit that we can always substitute call-by-name linear
-% function applications, or we forfeit that reverse binder swap preserves
-% linearity (instead, it preserves a weaker notion of linearity understood by the
-% syntatic-occurrence-analyzer)
-
-% \todo[inline]{Reverse-binder-swap is only well-defined in certain scenarios
-% where the optimizations don't apply call-by-name beta-reduction after the
-% reverse-binder-swap optimization -- otherwise we would duplicate resources.
-% In this case, it is not a matter of syntatic vs semantic linearity
-% }
-
-% \todo[inline]{
-% Mention, from ``Call-by-name, call-by-value, call-by-need and the linear lambda calculus'':
-% The call-by-name calculus is not entirely suitable for reasoning about
-% functional programs in lazy languages, because the beta rule may copy the
-% argument of a function any number of times. The call-by-need calculus uses a
-% diferent notion of reduction, observationally equivalent to the call-by-name
-% calculus. But call-by-need, like call-by-value, guarantees that the argument
-% to a function is not copied before it is reduced to a value.
-% }
-
-% It's also interesting to note that reverse-binder-swap preserves linearity under pure call-by-need but not under call-by-name, because
-% (In the sense that if EVEN linear functions reduce call-by-need rather than call-by-name, then it would preserve optimisations)
-
-% Reduce call-by-name linear function application
-% \begin{code}
-% f y = (\x. case x of _ -> x:1) (h y)
-% ===>
-% f y = (case x of _ -> x)[h y/x]
-% ===>
-% f y = (case h y of _ -> h y) -- consume y twice.
-% \end{code}
-% 
-% Vs. call-by-need
-% \begin{code}
-%  (\y = (\x. case x of _ -> x:1) (h y)) e
-% ===>
-%     let y = e in
-%        let x = h y in
-%         case x of _ -> x
-% ===>
-%        let x = h e_v in
-%         case x of _ -> x
-% ===>
-%         case x_v of _ -> x_v
-% \end{code}
-
-% \subsubsection{Case of Case}
-
-% The case of case transformation applies to case expressions whose scrutinee is
-% another case expression, and returns the innermost case expression transformed by repeating
-% the outermost case expression in each alternative of the innermost case,
-% scrutinizing the original alternative body.
-
-% Intuitively, since the scrutinee of the outermost case is not in WHNF, no
-% resources from it can directly occur in the outermost alternatives. By moving
-% the outermost alternatives inwards with a different scrutinee, the alternatives
-% remain well-typed because they are typed using either the case binder or the
-% pattern bound variables, which, by the \emph{Irrelevance} lemma, makes it
-% well-typed for any scrutinee consuming arbitrary resources. The proof is given in Section~\ref{sec:proof:caseofcase}.
-
-% \CaseOfCaseTheorem
-
-% }}}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % {{{ Linear Core as a GHC Plugin
@@
 %
+The plugin successfully validates linearity throughout the
+(optimising) compilation of linearity-heavy libraries, namely
+\texttt{linear-base} and \texttt{linear-smc}, except in expressions whose
+linearity depends on so-called \emph{multiplicity coercions}, which are an
+avenue of future work (Section~\ref{sec:future-work}) exploring the intersection
+of Linear Core with type equality coercions.
 % Furthermore, the implementation accepts all example programs from Chapter~\ref{sec:linearity-semantically}
@@
 sub-derivations, we thread input/output linear resources through each
-derivation~\cite{DBLP:journals/tcs/CervesatoHP00}.
-%
-Pattern variables bound in a case alternative, for a scrutinee not in WHNF, are
-introduced as $\D$-variables with usage environment $\lctag{\irr{\D}}{K_i}$,
-where $\D$ are the scrutinee resources and $K_i$ the tag of that pattern
-variable. To use these pattern variables, the scrutinee resources must be first
-$Split$ into fragments. Instead of guessing which resources need to be split,
-the implementation consumes tagged fragments of a resource \emph{as needed},
-i.e., only when consuming a resource whose usage environment has a resource
-tagged $K_j$ do we $Split$ that resource on $K$ and consume the $K_j$ fragment.
-%
+derivation~\cite{DBLP:journals/tcs/CervesatoHP00}.  Pattern variables
+bound in a case alternative for a scrutinee not in WHNF, are
+introduced as $\D$-variables with usage environment
+$\lctag{\irr{\D}}{K_i}$, where $\D$ are the scrutinee resources and
+$K_i$ the tag of that pattern variable. To use the resources through
+the pattern-bound $\D$-vars, they must be first $Split$ into
+fragments. We consume tagged fragments of a resource \emph{as needed},
+i.e., when a resource, whose usage environment has a fragmented
+resource with tag $K_j$, is used, we $Split$ the matching resource
+according to the constructor $K$ and consume the fragment $K_j$,
+rather than eagerly determining which resources need to be fragmented
+to do so.
 Finally, our implementation must infer the usage environments of binders in
@@
 
-% TOdo: INCLUDE THIS??
-% The plugin successfully validates linearity throughout the
-% (optimising) compilation of linearity-heavy libraries, namely
-% \texttt{linear-base} and \texttt{linear-smc}, except in expressions whose
-% linearity depends on so-called \emph{multiplicity coercions}, which are an
-% avenue of future work (Section~\ref{sec:future-work}) exploring the intersection
-% of Linear Core with type equality coercions.
-
 The results of running the Linear Core GHC plugin on large 
@@
 %
-%
-We compiled the Haskell standard library for programming with linear types,
-\texttt{linear-base} (4000 lines, comprising over 100 modules), the
-\texttt{linear-smc}~\cite{10.1145/3471874.3472980} (1500 lines) library, and
+We compiled the libraries \texttt{linear-smc}~\cite{10.1145/3471874.3472980} (1500 lines), \texttt{linear-base} (4000
+lines), the Haskell standard library for programming with linear types,
+comprised of over 100 modules, and
 \texttt{priority-sesh}~\cite{10.1145/3471874.3472979} (1400 lines), a
@@
 binding is typechecked once per optimisation pass. i.e., we typecheck
-all intermediate programs produced by GHC. The total amount of programs
-rejected by the typechecker are counted dually, also accross every optimisation
-pass, without halting compilation when a program is rejected. We note that our
-implementation is not performance conscious, and types every intermediate
-program from scratch, instead of preserving information throughout the
-pipeline.
-
-%``Total Rejected'' column, but we distil these rejections into ``Unique
-%Rejections'' by removing duplicate rejections (those of the same
-%program and for the same linearity-violating reason, but occurring at
-%different stages). We further categorize the unique rejections into
-%``Linear modulo Call-by-name'', ``Linear Rejected'', ``Not Linear
-%Rejected'', and ``Unknown Rejected''.  We note that our implementation
-%is not performance conscious. For instance, \texttt{linear-base} takes
-%35 seconds to compile with our plugin, instead of 20 seconds.
-%%
-%% Using the plugin, \texttt{linear-base} takes 35 seconds to compile, instead of
-%% 20 seconds. We note, however, that the implementation is not performance
-%% conscious, and types every intermediate program from scratch, instead of
-%% maintaining linearity information throughout the pipeline.
-
-%Programs ``linear modulo call-by-name'' are a class of programs
-%which scrutinize a variable, but then use the variable in the case
-%alternatives of the case scrutinizing it. As discussed in
-%Section~\ref{sec:reverse-binder-swap-considered-harmful}, these programs can be
-%understood as linear as long as applications of linear functions binding these
-%variables are \emph{not} reduced \emph{call-by-name}.
-%In Core, these programs are not seen as
-%linear, and thus GHC does not not reduce them using a call-by-name evaluation strategy.
-%% We have a flag in our implementation to accept some of these programs...
-%%
-%``Linear Rejected'' programs include different kinds of programs which are
-%rejected by the Linear Core system, but are still semantically linear. An
-%example is a program to which a rewrite rule was applied, resulting in an
-%application of unrestricted function to a linear resource.
-%% in the first argument
-%% to \incode{build}:
-%% \begin{lstlisting}
-%% take :: Int -> Replicator a *' $\lolli$ '* [a]
-%% take = \ (ds :: Int) (r :: Replicator b) ->
-%%      case ds of
-%%        1 -> build (\(c :: b -> a -> a) (n :: a) -> c (extract r) n)
-%%        ...
-%% \end{lstlisting}
-%% We know the linear resource is still used linearly because \incode{build} will
-%% always instance the unrestricted function to $(:)$ (read \emph{Cons}), which is
-%% linear. Additionally, these include programs scrutinizing a value of a type of
-%% which all constructors have no linear components, but only matching on the
-%% default alternative, programs where common-sub-expression elimination
-%% substituted more than one occurrence of \incode{Ur x} by \incode{y}, in the body of
-%% scrutinizing \incode{y} (a linear variable), and other programs affected by rewrite rules.
-%%
-%``Not Linear Rejected'' indicate programs that we do not understand as linear,
-%semantically, and are simultaneously rejected by Linear Core and its
-%implementation. % These programs were not necessarily identified as bad
-%% by the linter.\footnote{rever}
-%% An example, where the last component of \incode{HashMap}, \incode{wwz},
-%% is linear, but is not being consumed in the case alternative matching on it:
-%% \begin{lstlisting}
-%% jssvi :: Ur Bool *' $\lolli$ '* Set Int *' $\lolli$ '* Ur (TestT IO ())
-%% jssvi (a :: Ur Bool) (b :: Set Int)
-%%   = case a of
-%%      Ur ss -> case b of { HashMap wwx wwy wwz -> jump wjssAd ss }
-%% \end{lstlisting}
-%%
-%Finally, ``Unknown Rejected'' programs are those whose validity we did not
-%check. These include both programs accepted by Linear Core, but not by its
-%implementation, and programs that are simply rejected by Linear Core, but which
-%were not categorized.
+all intermediate programs produced by GHC. The total amount of
+programs rejected by the typechecker are given in the ``Total
+Rejected'' column, but we distil these rejections into ``Unique
+Rejections'' by removing duplicate rejections (those of the same
+program and for the same linearity-violating reason, but occurring at
+different stages). We further categorize the unique rejections into
+``Linear modulo Call-by-name'', ``Linear Rejected'', ``Not Linear
+Rejected'', and ``Unknown Rejected''.  We note that our implementation
+is not performance conscious. For instance, \texttt{linear-base} takes
+35 seconds to compile with our plugin, instead of 20 seconds.
+%
+% Using the plugin, \texttt{linear-base} takes 35 seconds to compile, instead of
+% 20 seconds. We note, however, that the implementation is not performance
+% conscious, and types every intermediate program from scratch, instead of
+% maintaining linearity information throughout the pipeline.
+
+Programs ``linear modulo call-by-name'' are a class of programs
+which scrutinize a variable, but then use the variable in the case
+alternatives of the case scrutinizing it. As discussed in
+Section~\ref{sec:reverse-binder-swap-considered-harmful}, these programs can be
+understood as linear as long as applications of linear functions binding these
+variables are \emph{not} reduced \emph{call-by-name}.
+In Core, these programs are not seen as
+linear, and thus GHC does not not reduce them using a call-by-name evaluation strategy.
+% We have a flag in our implementation to accept some of these programs...
+%
+``Linear Rejected'' programs include different kinds of programs which are
+rejected by the Linear Core system, but are still semantically linear. An
+example is a program to which a rewrite rule was applied, resulting in an
+application of unrestricted function to a linear resource.
+% in the first argument
+% to \incode{build}:
+% \begin{lstlisting}
+% take :: Int -> Replicator a *' $\lolli$ '* [a]
+% take = \ (ds :: Int) (r :: Replicator b) ->
+%      case ds of
+%        1 -> build (\(c :: b -> a -> a) (n :: a) -> c (extract r) n)
+%        ...
+% \end{lstlisting}
+% We know the linear resource is still used linearly because \incode{build} will
+% always instance the unrestricted function to $(:)$ (read \emph{Cons}), which is
+% linear. Additionally, these include programs scrutinizing a value of a type of
+% which all constructors have no linear components, but only matching on the
+% default alternative, programs where common-sub-expression elimination
+% substituted more than one occurrence of \incode{Ur x} by \incode{y}, in the body of
+% scrutinizing \incode{y} (a linear variable), and other programs affected by rewrite rules.
+%
+``Not Linear Rejected'' indicate programs that we do not understand as linear,
+semantically, and are simultaneously rejected by Linear Core and its
+implementation. % These programs were not necessarily identified as bad
+% by the linter.\footnote{rever}
+% An example, where the last component of \incode{HashMap}, \incode{wwz},
+% is linear, but is not being consumed in the case alternative matching on it:
+% \begin{lstlisting}
+% jssvi :: Ur Bool *' $\lolli$ '* Set Int *' $\lolli$ '* Ur (TestT IO ())
+% jssvi (a :: Ur Bool) (b :: Set Int)
+%   = case a of
+%      Ur ss -> case b of { HashMap wwx wwy wwz -> jump wjssAd ss }
+% \end{lstlisting}
+%
+Finally, ``Unknown Rejected'' programs are those whose validity we did not
+check. These include both programs accepted by Linear Core, but not by its
+implementation, and programs that are simply rejected by Linear Core, but which
+were not categorized.
 
 {\small
-  \input{prototype/core-plugin-results-2}
+  \input{prototype/core-plugin-results}
   }
-%
+
 The results indicate our mostly direct implementation of Linear Core is
@@
 linear types.
-
-However, Linear Core does not accept every single program produced by the
-optimisation pipeline of GHC. Our manual analysis of rejected programs revealed
-excellent results. Programs considered ill-typed by the Linear Core
-implementation include:
-%
-\begin{itemize}
-    \item Programs we expected GHC to produce, but know are not seen as linear by $\lambda^\pi_\Delta$.
-        Namely, programs which scrutinize a variable, but then use the variable in the
-        case alternatives of the case scrutinizing it. As discussed in
-        Section~\ref{sec:reverse-binder-swap-considered-harmful}, these programs can be
-        understood as linear as long as applications of linear functions binding these
-        variables are reduced \emph{call-by-need}, which is not the case in our system.
-
-    \item Programs rejected due to the lack of ``linearity coercions'' in
-        $\lambda^\pi_\Delta$. So-called \emph{multiplicity coercions} extending
-        Linear Core could make these programs be understood as linear.
-
-    \item Programs which \emph{actually violate linearity}, in that the term
-        violates the linear type it is annotated with. These are the programs
-        we ultimately want to be able to reject by having a linearly typed Core
-        validate intermediate programs. We found both a program which outright
-        discarded a linear resource, and that rewrite rules implementing stream
-        fusion~\cite{10.1145/165180.165214,10.1145/1291151.1291199,10.1007/978-3-642-24276-2_2}, and
-        others too, could produce linearly-invalid programs (in one such case,
-        the type of \lstinline{build} was not general enough to accommodate linearity).
-\end{itemize}
-%
-Overall, the programs rejected by the Linear Core plugin,
-%, in \texttt{linear-base},
+%
+The programs rejected by the Linear Core plugin, in \texttt{linear-base},
 besides validating that our implementation is faithful to the
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% {{{ Metatheory
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Metatheory of Linear Core\label{sec:main:metatheory}}
+
+The $\lambda^\pi_\D$ type system is sound: well-typed programs in Linear Core do not
+get \emph{stuck}. Besides type safety, we
+prove that multiple optimising transformations used in GHC preserve
+linearity in the sense of being well-typed in our system.
+%
+% Additionally, we state our assumptions that outline an isomorphism between
+% using a linear variable $\xl$ and a $\D$-variable $\xD$ that consumes existing
+% resources $\D$, for any $\D$.
+
+\renewcommand{\DeltaLinearRelationLemma}{
+\begin{restatable}[$\Delta$-bound to Linear]{lemma}{deltaone}\label{lem:deltaone}
+% Any $\Delta$-variable with a fully-irrelevant usage environment can be made a linear variable as long as its irrelevant environment is dropped from the linear environment (and the unrestricted env.
+is updated accordingly).\\
+If $\G,\x[\irr{\D}]; \irr{\D},\D' \vdash e : \vp$ 
+then $\G[x/\irr{\D}]; \D',\xl \vdash e :\vp$.
+\end{restatable}
+}
+
+\renewcommand{\LinearDeltaRelationLemma}{
+  \begin{restatable}[Linear to $\Delta$-bound]{lemma}{onedelta}\label{lem:onedelta}
+If $\G; \D',\xl \vdash e :\vp$
+then $\G[\D/x],\xD; \D,\D' \vdash e : \vp$ ($\Delta$ fresh).
+\end{restatable}
+}
+
+\renewcommand{\DeltaUnrestrictedRelationLemma}{
+\begin{restatable}[Unrestricted and $\Delta$-bound]{lemma}{undelta}\label{lem:undelta}
+%An unrestricted variable is equivalent to a $\D$-var with an empty usage environment.\\
+$\G,\xo; \D \vdash e : \vp$ iff $\G,\x[\cdot]; \D \vdash e : \vp$
+\end{restatable}
+}
+
+\paragraph{$\Delta$-bound variables} % Our development relies on a characterization of $\D$-bound
+% variables
+A well-typed program with a linear variable ($\xl$) is equivalently
+well-typed if the linear variable were instead $\D$-bound ($\xD$) with
+usage environment $\D$, with $\D$ available in the linear context (and the unrestricted context updated accordingly).
+
+\LinearDeltaRelationLemma
+
+\noindent Dually, any $\Delta$-bound variable with an irrelevant usage
+environment can be made into a linear variable as long as its
+(irrelevant) environment is dropped from the linear context.
+
+% Dually, a well-typed program with resources $\D$ and $\D$-bound
+% variable ($\xD$) is equivalently well-typed, as long as $\D$ is consumed
+% through the use of $x$, if $x$ is moved to the linear context, resources $\D$
+% are removed from the linear context, and occurrences of $\D$ whole in usage
+% environments are substituted by $x$ (occurrences of fragments of $\D$ in usage
+% environments are unimportant since $\D$ was consumed whole by $x$, not by any
+% of the fragment-using $\D$-vars).
+
+\DeltaLinearRelationLemma
+
+% Intuitively, a linear variable
+
+\noindent Finally, unrestricted and $\D$-bound
+variables with an empty usage environment are equivalent.
+
+\DeltaUnrestrictedRelationLemma
+
+\paragraph{Irrelevance}
+%
+As discussed, proof irrelevant resources can only be consumed
+indirectly, essentially encoding that the (non-WHNF) case scrutinee
+resources must be consumed through the case binder or the linear
+pattern-bound variables.
+%
+As a case expression is evaluated, the scrutinee will eventually
+reduce to WHNF,
+which must then be typed with rule $Case_{\textrm{WHNF}}$.
+%
+Crucially, the two case rules must be in harmony in the system, in the sense that
+case expressions typed using the $Case_{\textrm{Not WHNF}}$ rule must also be
+well-typed by the $Case_{\textrm{WHNF}}$ once the scrutinee is evaluated to WHNF.
+
+% The type preservation theorem states that a well-typed expression
+% remains well-typed in the presence of evaluation. Specifically, when the case
+% expression whose scrutinee is evaluated to WHNF is handled in the
+% preservation proof.
+
+% The \emph{Irrelevance} lemma is required to prove preservation for that
+% evaluation case. We need to prove that the alternatives of a case expression
+% typed with proof irrelevant resources are still well-typed when the proof
+% irrelevant resource is substituted by the scrutinee resources as it is evaluated to WHNF.
+% In this sense, the \emph{Irrelevance} lemma witnesses the soundness of typing a
+% case alternative with proof irrelevant resources in a certain context with respect to
+% typing the same expression with arbitrary resources (we note, however, typing
+% an alternative with proof irrelevant resources is not complete wrt using
+% arbitrary resources -- a counter example needs only to use a resource
+% directly).
+
+\WHNFConvSoundness
+
+The \emph{Irrelevance} lemma witnesses the soundness of typing a
+case alternative with proof irrelevant resources with respect to
+typing the same expression with arbitrary resources. However, typing
+a case alternative with proof irrelevant resources is not complete wrt using
+arbitrary resources -- a counter example needs only to use a resource
+directly.
+
+% \noindent Intuitively, the lemma holds since proof irrelevant resources can
+% only be used through the case binder or pattern-bound variables. If we
+% consistently replace the proof irrelevant resources both in the typing
+% environment and in the usage environments containing them, the expression
+% remains well-typed. We note that the proof irrelevant resources are always
+% unique when introduced in such a case alternative (we always take the scrutinee
+% environment to make irrelevant, and allow nested ``irrelevantness''), so the
+% case binder has the only occurrence of those resources in the $\G$ environment.
+% % (being somewhat akin to congruence).
+% %
+% The proof is given in Section~\ref{sec:proof:irrelevance}.
+
+\subsection{Type safety\label{sec:type-safety-meta}}
+
+We prove type safety of the Linear Core system via the standard type
+preservation and progress results. As is customary, we make use of multiple
+substitution lemmas, one for each kind of variable: unrestricted variables
+$\xo$, linear variables $\xl$, and $\D$-bound variables $\xD$.
+
+% We start with the auxiliary results, as we will make use of them in a select part of the preservation proof.
+
+\TypePreservationTheorem
+%
+\ProgressTheorem
+% \noindent Type preservation states that a well-typed expression $e$ that
+% evaluates to $e'$ remains well-typed under the same context:
+%
+Type preservation
+% ($\S$~\ref{sec:proof:type-preservation})
+follows by structural induction on the reductions $e \longrightarrow
+e'$ from the operational semantics. Most cases are straightforward, appealing to one or more
+substitution lemmas (see Appendix~\ref{sec:proof:type-preservation}). The most
+interesting case is that of case expressions whose scrutinee can be further
+evaluated -- we branch on whether the scrutinee becomes in WHNF and appeal to Lemma~\ref{lem:irrev} if so.
+%
+% This case guarantees that the separation of rules for treating scrutinees is
+% consistent, in the sense that a well-typed case expression with a scrutinee not
+% in WHNF remains well-typed after the scrutinee is evaluated to WHNF.
+%
+%
+%
+%\noindent
+% Progress states that the evaluation of a well-typed term does not block:
+% Similarly, progress is proved by induction on typing ($\S$~\ref{sec:proof:progress}).
+
+% \subsubsection{Substitution Lemmas}
+
+% The preservation and progress theorems depend on multiple substitution lemmas,
+% one for each kind of variable, as is standard.
+% % The proofs are given in Section~\ref{sec:proof:substitution-lemmas}.
+% % The ... themselves diverge from their common formulation, because
+% % $\D$-variables refer to linearly bound variables, so substitution must take into account
+
+% The linear substitution lemma states that a well-typed expression $e$ with a
+% linear variable $x$ of type $\s$ remains well-typed if
+% occurrences of $x$ in the $e$ are replaced by an expression $e'$ of the same
+% type $\s$, and occurrences of $x$ in the linear context and in usage
+% environments of $\D$-bound variables are replaced by the linear context $\D'$
+% used to type $e'$:
+
+% \LinearSubstitutionLemma
+
+% \noindent Where $\G[\D'/x]$ substitutes all occurrences of $x$ in the usage
+% environments of $\D$-variables in $\G$ by the linear variables in $\D'$.
+% % ($x$ couldn't appear anywhere besides usage environments of $\D$-bound variables, since $x$ is linear).
+
+% The substitution of the resource in the usage environments is illustrated
+% by the following example. Consider the term $\llet{y = use~x}{y}$ where $use$ and $x$ are free variables:
+% if we replace occurrences of $x$ by $e'$ (where $\G;\D \vdash e : \s$), then the
+% ``real'' usage environment of $y$ goes from $\{x\}$ to $\D$. If we don't update
+% the usage environment of $y$ accordingly, we'll ultimately be typing
+% $y{:}_{\{x\}}\vp$ with $\D$ instead of $x$, which is not valid.
+
+% The linear substitution lemma extends to case alternatives as well.
+% The lemma for substitution of linear variables in case alternatives is similar
+% to the linear substitution lemma, applied to the case alternative judgement.
+% % Is slightly more involved, in the sense that there are more environments in which the
+% % substitution $[\D/x]$ must be applied (for the same reason):
+% %
+% \LinearSubstitutionAltsLemma
+% %
+% \noindent We further require that the environment annotated in the case
+% alternative judgement, $\D_s$, is a subset of the environment used to type the
+% whole alternative $\D_s \subseteq \D$. In all occurrences of the alternative
+% judgement (in $Case_{\textrm{WHNF}}$ and $Case_{\textrm{Not WHNF}}$), the
+% environment annotating the alternative judgement is \emph{always} a subset of
+% the alternative environment.
+
+% The substitution lemma for unrestricted variables follows the usual
+% formulation, with the added restriction (common to linear type systems) that
+% the expression $e'$ that is going to substitute the unrestricted variable $x$
+% is typed on an empty linear environment:
+% %
+% \UnrestrictedSubstitutionLemma
+% %
+% \noindent Similarly, we also prove the substitution of unrestricted variables preserves types on an alternative case expression:
+% %
+% \UnrestrictedSubstitutionAltsLemma
+
+% Finally, we introduce the lemma stating that substitution of $\D$-bound
+% variables by expressions of the same type preserves the type of the original
+% expression.
+% %
+% What distinguishes this lemma from traditional substitution lemmas is that the
+% usage environment $\D$ of the variable $x$ being substituted by expression $e'$
+% must match exactly the typing environment $\D$ of $e'$ and the
+% environment of the original expression doesn't change with the substitution:
+% %
+% \DeltaSubstitutionLemma
+% %
+% \noindent Intuitively, if $x$ is well-typed with $\D$ in $e$, substituting $x$
+% by an expression $e'$ which is typed in the same environment $\D$ allows the
+% distribution of resources $\D,\D'$ used to type $e$ across sub-derivations to remain
+% unchanged. To prove the theorems, we don't need a ``stronger'' substitution of
+% $\D$-vars lemma (allowing arbitrary resources $\D''$ to type $e'$, as in other
+% substitution lemmas), as we only ever substitute $\D$-variables by expressions
+% whose typing environment matches the variables usage environment. However, it
+% is not obvious whether such a lemma is possible to prove for $\D$-variables
+% (e.g. let $\G;\D \vdash e :\s$ and $\G; \D' \vdash \llet{x = e'}{x}$, if we
+% substitute $e$ for $x$ the resources $\D'$ are no longer consumed).
+
+% The $\D$-substitution lemma on case alternatives reflects again that the typing
+% environment of the expression substitution the variable must match its usage
+% environment. We recall that $\D_s \subseteq \D,\D'$ states that the annotated
+% environment is always contained in the typing environment, which is true of all
+% occurrences of this judgement. An alternative formulation of this lemma could
+% instead explicitly list $\D_s$ as part of the typing environment for the same
+% effect:
+
+% \DeltaSubstitutionAltsLemma
+
+% The proofs for substitution lemmas of linear, unrestricted, and
+% $\D$-variables are available in Section~\ref{sec:proof:substitution-lemmas}.
+
+%TODO! Substitution of proof-irrelevant linear variables preserves typing. The
+%term always remains the same because $x$ cannot occur in any term, however, all
+%variables that refer to $x$ in their usage environment must now refer the usage env. of the substitee (e.g. $[x] => [\D]$).
+%This seems trivial to see correct, since all occurrences are in environments, so we get some equivalence similar to the one we need for the proof of Alt0.
+%
+%TODO: Multiplicity substitution preserves typing lemma
+%
+%TODO: Canonical forms lemma
+%
+%TODO: Corollary of $\Delta$-var subst. for $\ov{\Delta}$
+%
+%TODO: Constructor app typing:
+%If $\Gamma, \Delta \vdash K~\ov{e}$ and $K{:}\ov{\sigma\to\pi}~T~\ov{p} \in \Gamma$ and $\hasnolinearvars{\Gamma}$
+%then $\ov{\Gamma, \Delta_i \vdash e_i : \sigma_i}$
+
+\subsection{Optimisations preserve linearity\label{sec:optimisations-preserve-types-meta}}
+%\todo[inline]{Check Commuting case-let in Fig.~\ref{fig:opttrans}!}
+A main goal of the Linear Core type system is 
+to serve as a typed intermediate representation for optimising
+compilers of lazy languages with linearity. In light of this,
+we show that multiple optimising transformations are type
+preserving in Linear Core (i.e.,~preserve linearity semantically).
+
+% The optimising transformations proved sound wrt Linear Core in this section
+% have been previously explained and motivated in
+% Section~\ref{sec:core-to-core-transformations}.
+% %
+%\{\overline{\rho\to e'}\}
+\begin{figure}[t]
+
+  {\small
+  \[
+    \begin{array}{ll}
+   (\llet{\xD = e}{e'} )  \Longrightarrow \llet{\xD =       e}{e'[e/x]} &                    \mbox{Inlining}\\
+       (\lambda \x[\pi][\s].~e)~e'  \Longrightarrow e[e'/x]  &                \mbox{$\beta$-reduction}\\
+      (\ccase{(K~\ov{e})}{\z[\D][\s]~\{\overline{\rho_j\to e_j}, K~\ov{x} \to e_i, \overline{\rho_k\to e_k}\}} )\Longrightarrow  e_i\ov{[e/x]}[K~\ov{e}/z] &
+                                  \mbox{Case of known constr.}\\
+      (\ccase{(\ccase{e_c}{\z[\D]~\{\ov{\rho_{c_i}\to e_{c_i}}\}})}{\var[w][\irr{\D,\D'}][\s']~\{\ov{\rho_i\to e_i}\}})\\
+            \Longrightarrow 
+      \ccase{e_c}{\z[\D]~\{\ov{\rho_{c_i} \to \ccase{e_{c_i}}{w~\{\ov{\rho_i\to e_i}\}}}\}}
+      & \mbox{Case-of-case}\\
+      (\lambda \y[\pi].~\llet{\xD = e}{e'})
+      \Longrightarrow
+      \llet{\xD=e}{\lambda\y[\pi].~e'} & \mbox{Commuting let-$\lambda$}\\
+      ((\llet{v = e}{b})~a) \Longrightarrow \llet{v = e}{b~a} &\mbox{Commuting let-app.}\\
+      (\ccase{(\llet{v = e}{b})}{\{\overline{\rho\to e'}\}}) \Longrightarrow
+      \llet{v =  e}{\ccase{b}{\{\overline{\rho\to e'}\}}}
+                               &
+                                 \mbox{Commuting let-case}\\
+      (\ccase{e}{\{\overline{\rho_j\to e_j}, \rho \to E[e_1],
+      \overline{\rho_k\to e_k}\}} ) \qquad (\mbox{$x$ fresh}) \\
+     \Longrightarrow \llet{x = e_1}{\ccase{e}{\{\overline{\rho_j\to e_j}, \rho \to E[x], \overline{\rho_k\to e_k}\}}} 
+                                                                        & \mbox{Commuting case-let}\\
+      (\llet{x = (\llet{v = e}{b})}{c}) \Longrightarrow \llet{v = e}{\llet{x = b}{c}}
+                                                                      & \mbox{Commuting let-let}\\
+      f \Longrightarrow \lambda x.(f~x) & \mbox{$\eta$-expansion}\\
+      \lambda x.(f~x) \Longrightarrow f & \mbox{$\eta$-reduction}\\
+      (\ccase{x}{z~\{\ov{\rho_i\to e_i}\}}) \Longrightarrow \ccase{x}{z~\{\ov{\rho_i\to  e_i[z/x]}\}}
+                                                                      & \mbox{Binder swap}\\
+      % (\ccase{x}{z~\{\ov{\rho_i\to e_i}\}} )\Longrightarrow \ccase{x}{z~\{\ov{\rho_i\to e_i[x/z]}\}}
+      %                                                                 & \mbox{Reverse binder swap}
+    \end{array}
+  \]}
+  \caption{Optimising transformations validated by Linear
+    Core\label{fig:opttrans} (for commuting case-let we
+      require $e_1$ not to capture the case binder, variables bound by the
+      pattern $\rho$ and the context $E[{-}]$ -- see Appendix~\ref{app:optimisations} for
+    details).} 
+\end{figure}
+
+We describe Core-to-Core transformations as $e_1 \Longrightarrow e_2$,
+where $e_1$ is an arbitrary \emph{well-typed} expression
+matching a certain shape (or of a certain type), which is
+transformed into expression $e_2$. Validating a transformation
+entails showing that $e_2$ is well-typed. We often annotate the arrow
+with the name of the transformation when describing chains of such
+transformations.
+
+The list of transformations we have
+validated is given in Figure~\ref{fig:opttrans}.
+%
+The behaviour of most transformations can be inferred from its name:
+Inlining substitutes a let-bound expression in the let body; $\beta$-reduction
+and case of known constructor effectively perform evaluation of the given expression;
+case-of-case commutes cases in scrutinee position inside the outer case alternatives;
+let floating and  for let-bound expressions to be floated out of $\lambda$-abstractions;
+the four commuting conversions allow let-bound expressions to be commuted with other
+Core constructs; $\eta$-expansion and reduction apply the $\eta$-laws to terms of function type;
+finally, binder swap substitutes occurrences of a scrutinee variable for the
+case binder. While many of the transformations above are not obviously optimising,
+they are chained in the compiler pipeline to expose optimisation opportunities.
+
+% Transformations are described by an arbitrary well-typed expression with a certain shape, on
+% the left hand side (lhs) of the arrow $\Longrightarrow$, resulting in an expression on
+% the right hand side (rhs) that we prove to be well-typed.
+% %
+% % For our proofs, we assume the lhs to be a well-typed expression and prove the
+% % rhs is well-typed as well.
+% %
+% For each transformation, we describe the intuition behind the transformation
+% preserving linearity in our system.
+
+% \todo[inline]{Aqui uma lista itemizada das transformações todas. As
+%   provas todas para apendice.}
+
+%%% COMENTADO AQUI
+% \subsubsection{Inlining}
+
+% % To the best of our knowledge, there is no linear type system for which inlining
+% % preserves linearity\footnote{https://github.com/ghc-proposals/ghc-proposals/blob/master/proposals/0111-linear-types.rst\#id90}
+
+% \input{language-v4/proofs/optimizations/Inlining}
+
+% \subsubsection{\texorpdfstring{$\beta$}{Beta}-reduction}
+
+% \input{language-v4/proofs/optimizations/BetaReduction}
+
+% \subsubsection{Case of known constructor}
+
+% \input{language-v4/proofs/optimizations/CaseOfKnownConstructor}
+
+% \subsubsection{Let floating}
+
+% \input{language-v4/proofs/optimizations/LetFloating}
+
+% \subsubsection{\texorpdfstring{$\eta$}{Eta}-conversions}
+
+% \input{language-v4/proofs/optimizations/EtaConvs}
+
+% \subsubsection{Binder Swap}
+
+% The binder swap transformation applies to case expressions whose scrutinee is a
+% single variable $x$, and it substitutes occurrences of $x$ in the case
+% alternatives for the case binder $z$. If $x$ is a linear resource, $x$ cannot
+% occur in the case alternatives (as we conservatively consider variables are not
+% in WHNF), so the substitution preserves types vacuously. Otherwise, $x$ can be
+% freely substituted by $z$, since $z$ is also an unrestricted resource (it's
+% usage environment is empty because $x$ is unrestricted).
+
+% \input{language-v4/proofs/optimizations/BinderSwap}
+
+\paragraph{Reverse Binder Swap\label{sec:reverse-binder-swap-considered-harmful}}
+
+We now discuss an issue we have found with the so-called reverse binder swap
+transformation.
+%
+%
+Reverse binder swap is dual to the binder swap transformation,
+substituting occurrences of the case
+binder $z$ by the scrutinee in the special case when the scrutinee is a
+variable $x$.
+%
+% his is dual to what the binder swap transformation
+% does in hope of eliminating multiple uses of $x$ (allowing for inlining). 
+By
+using the scrutinee $x$ instead of the case binder, we might be able to float
+out expressions from the alternative using the case binder. For example, we
+might float an expensive computation involving $z$ out of the case alternative,
+where $z$ is now out of scope:
+
+{\small\[
+\begin{array}{l}
+(\lambda x.~\lletrec{go~y = \ccase{x}{z~\{(a,b) \to \ E[expensive~z] \} }}{\dots})
+\Longrightarrow_\textrm{Reverse binder swap}\\
+(\lambda x.~\lletrec{go~y = \ccase{x}{z~\{(a,b) \to E[expensive~x] \} }}{\dots})
+\Longrightarrow_\textrm{Commuting case-let}\\
+\lambda x.~\llet{t = \mathit{expensive}~x}{(\lletrec{go~y = \ccase{x}{z~\{(a,b) \to E[t] \} }}{\dots})}\\
+\end{array}
+\]}
+
+\noindent If $go$ is compiled into a loop, we only compute
+$\mathit{expensive}~x$ once instead of in every loop iteration.
+
+Despite GHC applying the reverse binder swap transformation 
+during its Core-to-Core optimisation passes, this optimisation actually \emph{breaks} linearity
+when considered as a transformation on Linear Core programs. In practice, the
+optimisation preserves linearity when applied as part of the GHC
+pipeline due to the occurrence analyser being naive with
+regard to semantic linearity. Initially, it might seem as though an expression
+in which a variable $x$ occurs both in the case scrutinee and in the
+alternatives is linear (e.g.,~$\ccase{x}{\_ \to x}$):
+% \[
+% \G; \xl \vdash \ccase{x}{\_ \to x} : \s
+% \]
+% The reasoning is done by branching on whether $x$ refers to an expression in
+% WHNF or an unevaluated thunk.
+% \begin{itemize}
+% \item
+  If $x$ refers to an unevaluated expression, then scrutinizing it results
+in the expression bound by $x$ to be evaluated to WHNF. In a subsequent use of
+$x$ in the alternatives, $x$ refers to the evaluated scrutinee in WHNF, which
+must be consumed. Since $x$ is just another name (like the case binder) for the
+scrutinee in WHNF, we may use it instead of the case binder or pattern
+variables.
+%
+On the other hand, if $x$ already refers to an expression evaluated to WHNF, then
+scrutinizing it in a case expression is a no-op, thus we may use it again (in
+mutual exclusion with the case binder and pattern variables).
+%\end{itemize}
+%
+% This idea would be encoded by a possible rule like $$\TypeCaseVar$$
+%
+
+Even though, on its own, it makes intuitive sense that this example indeed uses
+$x$ linearly, when considered as part of a complete type system, allowing this
+expression to be linear makes the system \emph{unsound}.
+To see why this is the case, recall that $\beta$-reduction reduces an application of a linear function
+using call-by-name -- if we know the argument is used exactly once, a binding
+to share the result of computing the argument is unnecessary, so we instead
+substitute the argument expression for the linearly-bound variable in the
+$\lambda$-body directly.
+
+Consider the function application $(\lambda x.~\ccase{x}{\_ \to x})~(use~y)$
+where $y$ is a free linear variable. Assuming $\lambda x.~\ccase{x}{\_ \to x}$
+is a linear function, $\beta$-reduction transforms the
+application into $\ccase{use~y}{\_ \to
+  use~y}$. Now, $y$ is a linear variable \emph{consumed} in the scrutinee of the case
+that also occurs in the case alternative --
+linearity is violated by using the linear resource $y$ twice.
+%
+% Essentially, Linear Core would be unsound, and even duplicate resources, if the
+% above kind of expressions, where linear variable scrutinees occur in the
+% alternatives body, were well-typed, because of its interaction with the
+% call-by-name reduction of linear functions.
+In this sense, the reverse binder
+swap is an optimisation that creates ill-typed expressions from well-typed
+ones, so it is deemed an \emph{invalid} optimisation that does not preserve types in
+our system.
+
+Reverse binder swap is not a problem in the GHC simplifier because of the
+weaker notion of linearity understood by occurrence analysis. Occurrence
+analysis is a static analysis pass which can be used to determine whether a
+lambda application can be $\beta$-reduced call-by-name, and $\ccase{x}{\_ \to
+x}$ is \emph{not} seen as using $x$ linearly by the analysis. Thus,
+$\beta$-reduction is done with call-by-need on such an expression. If the above
+example were reduced with call-by-need
+% \[
+% \begin{array}{l}
+% (\lambda x.~\ccase{x}{\_ \to x})~(use~y)
+% \Longrightarrow_\textrm{CBNeed}
+% \llet{x = use~y}{\ccase{y}{\_ \to y}}
+% \end{array}
+% \]
+then the computation using $y$ would be let-bound, and $y$ used as a scrutinee
+variable, which is indeed semantically linear in $x$.
+
+% Concluding, in being able to understand more programs as linear, our type
+% system allows more expressions to be considered linear for $\beta$-reduction
+% without a let-allocation, however, it makes reverse binder swap an invalid
+% transformation since its output, when considered linear, might violate
+% linearity when further optimised.
+
+% (Link to ticket)
+% 
+% \begin{tabbing}
+% (1) All optimisations preserve (semantic) linearity\\
+% (2) If a function is (semantically) linear, then we can evaluate it using call-by-name and preserve linearity\\
+% (3) Reverse binder swap is an optimisation\\
+% (4) If reverse binder swap is applied to a case scrutinizing a linear resource in the body (`e`) of a linear function `f`, then the function is still linear by (1)\\
+% (5) If we evaluate `f`, we do it call-by-name because of (2)\\
+% (6) Call-by-name substitution of the linear argument in the body of a function has been reverse binder swapped doesn’t preserve linearity\\
+% (7) Contradiction: By 3 and 1, `f` is linear after reverse-binder-swap.
+By 2, we can substitute arguments to `f` call-by-name and preserve linearity, which contradicts with 6 that says call-by-name substitution after reverse binder swap violates linearity\\
+% \end{tabbing}
+
+% Conclusion:
+% Either we need to forfeit that we can always substitute call-by-name linear
+% function applications, or we forfeit that reverse binder swap preserves
+% linearity (instead, it preserves a weaker notion of linearity understood by the
+% syntatic-occurrence-analyzer)
+
+% \todo[inline]{Reverse-binder-swap is only well-defined in certain scenarios
+% where the optimizations don't apply call-by-name beta-reduction after the
+% reverse-binder-swap optimization -- otherwise we would duplicate resources.
+% In this case, it is not a matter of syntatic vs semantic linearity
+% }
+
+% \todo[inline]{
+% Mention, from ``Call-by-name, call-by-value, call-by-need and the linear lambda calculus'':
+% The call-by-name calculus is not entirely suitable for reasoning about
+% functional programs in lazy languages, because the beta rule may copy the
+% argument of a function any number of times. The call-by-need calculus uses a
+% diferent notion of reduction, observationally equivalent to the call-by-name
+% calculus. But call-by-need, like call-by-value, guarantees that the argument
+% to a function is not copied before it is reduced to a value.
+% }
+
+% It's also interesting to note that reverse-binder-swap preserves linearity under pure call-by-need but not under call-by-name, because
+% (In the sense that if EVEN linear functions reduce call-by-need rather than call-by-name, then it would preserve optimisations)
+
+% Reduce call-by-name linear function application
+% \begin{code}
+% f y = (\x. case x of _ -> x:1) (h y)
+% ===>
+% f y = (case x of _ -> x)[h y/x]
+% ===>
+% f y = (case h y of _ -> h y) -- consume y twice.
+% \end{code}
+% 
+% Vs. call-by-need
+% \begin{code}
+%  (\y = (\x. case x of _ -> x:1) (h y)) e
+% ===>
+%     let y = e in
+%        let x = h y in
+%         case x of _ -> x
+% ===>
+%        let x = h e_v in
+%         case x of _ -> x
+% ===>
+%         case x_v of _ -> x_v
+% \end{code}
+
+% \subsubsection{Case of Case}
+
+% The case of case transformation applies to case expressions whose scrutinee is
+% another case expression, and returns the innermost case expression transformed by repeating
+% the outermost case expression in each alternative of the innermost case,
+% scrutinizing the original alternative body.
+
+% Intuitively, since the scrutinee of the outermost case is not in WHNF, no
+% resources from it can directly occur in the outermost alternatives. By moving
+% the outermost alternatives inwards with a different scrutinee, the alternatives
+% remain well-typed because they are typed using either the case binder or the
+% pattern bound variables, which, by the \emph{Irrelevance} lemma, makes it
+% well-typed for any scrutinee consuming arbitrary resources. The proof is given in Section~\ref{sec:proof:caseofcase}.
+
+% \CaseOfCaseTheorem
+
+% }}}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2024-02-18 prototype/core-plugin-results-2.tex

\begin{figure}[h]
\centering{
\begin{tabular}{|l|c|c|c|}
    \toprule
    \thead{Library} & \thead{Total\\Accepted} & \thead{Total\\Rejected} & \thead{Accept Rate}\\
    \midrule
    linear-smc & 19438 & 4 & 99.98\% \\
    \makecell{priority-sesh} & 6781 & 19 & 99.72\% \\
    linear-base & 112311 & 538 & 99.52\% \\
    \bottomrule
\end{tabular}
}
\caption{Linear Core Plugin on Linear Libraries}
\label{fig:core-plugin-res}
\end{figure}

### 2024-02-19 icfp.tex
typos and minor stuff

@@
 since we are not using $a$ or $b$ (thereby never forcing the arguments
-of the tuple). However, if we did use $a$ in the case body, then $x$ would be unavailable:
+of the tuple). However, if we did use $a$ in the case body, then $x$
+would be unavailable (and so \incode{f8} below cannot be accepted):
 \begin{noway}
@@
 \end{notyet}
-Again, $z$ is mutually exclusive with $a,b$ and with $x,y$, but at least one of
+Again, $z$ is mutually exclusive with $(a,b)$ and with $(x,y)$, but at least one of
 the three must occur to ensure the linear resources are consumed. In this
@@
 For instance, take
-\incode{use} in \incode{f10} to be \incode{(,)}: it is not sufficient for \incode{use x y} to be
+\incode{use} in \incode{f10} to be the pairing function \incode{(,)}: it is not sufficient for \incode{use x y} to be
 evaluated to WHNF to consume \incode{x} and \incode{y}. Otherwise, if all the resources were
@@
 % \item
-if $x$ is in WHNF, then scrutinizing it is a no-op, and returning $x$
+if the argument instantiating $x$ is in WHNF, then scrutinizing it is a no-op, and returning $x$
 just returns the resource intact;
 % \item
-if $x$ is not in WHNF, then scrutinizing it evaluates it to WHNF, and returning
+if the argument instantiating$x$ is not in WHNF, then scrutinizing it evaluates it to WHNF, and returning
 $x$ returns the result of evaluating $x$ that still had to be consumed exactly
@@
 However, in practice, this program's linearity can only be determined by
-further branching down on the non-strict evaluation strategy. If linear
-function applications, in general, are $\beta$-reduced \emph{call-by-name} (a desireable
+further analysis on the non-strict evaluation strategy. If linear
+function applications, in general, are $\beta$-reduced \emph{call-by-name} (a desirable
 practice, as linear functions are known to use their argument exactly once)
@@
 %
-The type system we present in the next section, which evaluates linear
+The system we present in the next section, which evaluates linear
 applications call-by-name, does not accept the above function scrutinizing a
 variable as linear. 
-This interaction between evaluation and linearity comes up in the type
-preservation proof of our program, and is again explored with the reverse
-binder swap transformation in
+This interaction between evaluation and linearity arises in the type
+preservation proof of our system, and is again explored with the reverse
+binder swap transformation discussed in
 Section~\ref{sec:optimisations-preserve-types-meta}.
@@
 consumed linearly (ultimately, in order to consume the scrutinee linearly).
-Additionally, the set of patterns in a case expression is guaranteed to be exhaustive,
+Additionally, the set of patterns in a case expression is assumed to be exhaustive,
 i.e. there is always at least one pattern which matches the scrutinized expression.
@@
 The rule reads that an occurrence of a $\Delta$-bound variable is well-typed if
-the linear environment is exactly the resources in the usage environment of
+the linear environment is made up exactly of the resources in the usage environment of
 that variable.
@@
 alternative corresponding to the pattern matching the WHNF of the
-scrutinee\footnote{In our calculus, the alternatives are always
+scrutinee\footnote{In our calculus, the alternatives are assumed to be
   exhaustive, i.e. there always exists at least one pattern which
@@
 As will be made clear in this section, we need to devise a specialized typing
-judgement for scrutinees that is able to distinguish between terms
+for scrutinees that distinguishes between terms
 in WHNF and terms that are not in WHNF.
@@
 whose scrutinee is in weak head normal form, or otherwise, leads to one of our
-key insights: we must \emph{branch on weak head normal formed-ness} to
+key insights: we must \emph{branch} on weak head normal formed-ness to
 accurately type case expressions.
@@
 \vp$ is used to type case alternatives, encompassing three
-``modes``, distinguished by the kind of arrow that is used:
+``modes'', distinguished by the kind of arrow that is used:
 for alternatives of case expressions whose scrutinee is in WHNF ($\Mapsto$),
@@
 
-Hence, to type a case expression whose scrutinee is in weak head normal form, we
+Hence, to type a case expression whose scrutinee is in WHNF, we
 type the scrutinee with linear resources $\D$ and type the case
@@
 
-The key idea to guarantee these is to provide a $\textsc{Split}$ rule that
+The key idea to guarantee these is to provide a $\textit{Split}$ rule that
 allows a linear resource $x{:}_1\s$ to be split into $n$ resources at a given
@@
 case-of-case commutes cases in scrutinee position inside the outer case alternatives;
-let floating and  for let-bound expressions to be floated out of $\lambda$-abstractions;
-the four commuting conversions allow let-bound expressions to be commuted with other
+%let floating and  for let-bound expressions to be floated out of $\lambda$-abstractions;
+the five commuting conversions allow let-bound expressions to be commuted with other
 Core constructs; $\eta$-expansion and reduction apply the $\eta$-laws to terms of function type;
@@
 naive $O(n^2)$ algorithm (where $n$ is the number of let bindings) to determine
-these usage environments. Further research regarding
-this inference challenge is discussed in Section~\ref{sec:future-work}.
+these usage environments. %Further research regarding
+%this inference challenge is discussed in Section~\ref{sec:future-work}.
 %
@@
 each top-level binding in a module counts as a program, and every such
-binding is typechecked once per optimisation pass. i.e., we typecheck
-all intermediate programs produced by GHC. The total amount of programs
-rejected by the typechecker are counted dually, also accross every optimisation
+binding is typechecked once per optimisation pass (i.e., we typecheck
+all intermediate programs produced by GHC). The total amount of programs
+rejected by the typechecker are counted dually, also across every optimisation
 pass, without halting compilation when a program is rejected. We note that our
@@
         $\lambda^\pi_\Delta$. So-called \emph{multiplicity coercions} extending
-        Linear Core could make these programs be understood as linear.
+        Linear Core could make these programs be understood as linear
+        (discussed in Section~\ref{sec:future-work}).
 
@@
         discarded a linear resource, and that rewrite rules implementing stream
-        fusion~\cite{10.1145/165180.165214,10.1145/1291151.1291199}, and
-        others too, could produce linearly-invalid programs (in one such case,
+        fusion~\cite{10.1145/165180.165214,10.1145/1291151.1291199}
+        (among others), could produce linearly-invalid programs (in one such case,
         the type of \lstinline{build} was not general enough to accommodate linearity).

### 2024-02-19 icfp.tex
typo

@@
 % \item
-if the argument instantiating$x$ is not in WHNF, then scrutinizing it evaluates it to WHNF, and returning
+if the argument instantiating $x$ is not in WHNF, then scrutinizing it evaluates it to WHNF, and returning
 $x$ returns the result of evaluating $x$ that still had to be consumed exactly

### 2024-02-25 icfp.tex
taken into account feedback from Adam Gundry

@@
 
-Lastly, consider a program which defines two let bindings \incode{z} and \incode{y}, where
+Lastly, consider a (ill-typed) program which defines two let bindings \incode{z} and \incode{y}, where
 \incode{z} uses \incode{y} which in turn uses the linear resource \incode{x}:
@@
 the recursive call more than once, we might eventually consume \incode{x} more than
-once:
+once (and so \incode{f5} below is ill-typed):
 %
@@
 
-A third option is to use the case binder $z$ instead of $a,b$ or $x,y$:
+A third (semantically well-typed, but rejected by Linear Haskell) option is to use the case binder $z$ instead of $a,b$ or $x,y$:
 \begin{notyet}
@@
 evaluating it to WHNF will require performing computation and thus consume linear
-resources:
+resources (\incode{f10} below is well-typed in our system but rejected
+by Linear Haskell):
 \begin{notyet}
@@
 ``modes'', distinguished by the kind of arrow that is used:
-for alternatives of case expressions whose scrutinee is in WHNF ($\Mapsto$),
-for case expressions in which the scrutinee is not in WHNF ($\Rrightarrow$),
+for alternatives of case expressions whose scrutinee is in WHNF ($\Rightarrow_\mathsf{WHNF}$),
+for case expressions in which the scrutinee is not in WHNF ($\Rightarrow_\mathsf{NWHNF}$),
 and for alternatives agnostic to the WHNF-ness of the scrutinee
@@
 The rule for typing an alternative whose pattern is a
-constructor with $n > 0$ linear components (in mode $\Mapsto$) is:
+constructor with $n > 0$ linear components (in mode $\Rightarrow_{\mathsf{NWHNF}}$) is:
 \[
@@
 
-Hence, to type a case expression whose scrutinee is in WHNF, we
+Hence, to type a case expression whose scrutinee is not in WHNF, we
 type the scrutinee with linear resources $\D$ and type the case
@@
 
+In the following we write $\Mapsto$ to stand for
+$\Rightarrow_\mathsf{WHNF}$ and $\Rrightarrow$ for $\Rightarrow_\mathsf{NWHNF}$.
+
 \subsection{Auxiliary Lemmas}

### 2024-02-28 icfp.tex
typo in S6
fixed usage env in 3.3

@@
 For example, the let-binder $u$ in the following program is annotated with a
-usage environment $\{x,y\}$, the resources that are used in its body. Thus, in
+usage environment $\{x{:}_1\s,y{:}_1\s\}$, the resources that are used in its body. Thus, in
 the expression $e$, consuming $u$ will equate to consuming the linear resources
@@
 \[
-f = \lambda \xl.~\lambda \y[1].~\llet{u = (x,y)}{e}
+f = \lambda \xl.~\lambda \y[1].~\llet{u_{\{x:_1\s,y:_1\s\}} = (x,y)}{e}
 \]
@@
 %
-However, the implementation Linear Haskell in GHC does modify and extend Core
+However, the implementation of Linear Haskell in GHC does modify and extend Core
 with linearity/multiplicity annotations. Core's type system is unable to type

### 2024-02-28 icfp.tex
fixed 3 arrows

@@
 environment, annotating the judgement with the proof irrelevant resources,
-and using the $\Rrightarrow$ judgement:
+and using the $\Rightarrow_{\textsf{NWHNF}}$ judgement:
 \[
@@
 Finally, we recall the tentative $Case_\textrm{WHNF}$ rule presented before and
-highlight its flaw: the $\G;\D \vdash_{alt} \rho \to e :^z_{\D_s} \s \Mapsto
+highlight its flaw: the $\G;\D \vdash_{alt} \rho \to e :^z_{\D_s} \s \Rightarrow_{\textsf{WHNF}}
 \vp$ judgement is only well-defined for patterns $\rho$ matching the WHNF form

### 2025-01-10 icfp.tex
RM edit S1

@@
-\documentclass[acmsmall,review,anonymous]{acmart}
+\documentclass[acmsmall,review,anonymous,screen]{acmart}
 \PassOptionsToPackage{dvipsnames}{xcolor}
@@
 
+\begin{CCSXML}
+<ccs2012>
+   <concept>
+       <concept_id>10003752.10003790.10003801</concept_id>
+       <concept_desc>Theory of computation~Linear logic</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10003752.10003790.10011740</concept_id>
+       <concept_desc>Theory of computation~Type theory</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10011007.10011006.10011008.10011009.10011012</concept_id>
+       <concept_desc>Software and its engineering~Functional languages</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10011007.10011006.10011041</concept_id>
+       <concept_desc>Software and its engineering~Compilers</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+ </ccs2012>
+\end{CCSXML}
+
+\ccsdesc[500]{Theory of computation~Linear logic}
+\ccsdesc[500]{Theory of computation~Type theory}
+\ccsdesc[500]{Software and its engineering~Functional languages}
+\ccsdesc[500]{Software and its engineering~Compilers}
+
+\keywords{Linear Types, Lazy Evaluation, Haskell, GHC Core}
+
 \begin{document}
@@
 %
+
+\todo{REWRITE}
 We present Linear Core, the first type system that understands semantic
@@
 to existing type systems by enforcing that certain \emph{resources} (e.g.~a
-file handle) must be used \emph{exactly once}.
+file handle) are used \emph{exactly once}.
 %
@@
 The following program with a runtime error (allocated memory is freed twice) is
-accepted by a C-like type system, however, under the lens of a linear type
-system, if $p$ is deemed to be a linear resource created by the call to
-\texttt{malloc}, we would instead reject it:
+accepted by a C-like type system. Conversely, under the lens of a linear type
+system where $p$ is deemed a linear resource created by the call to
+\texttt{malloc}, the program is rejected:
 % We know this to be the dreaded double-free error which will crash the program at runtime.
@@
 
-Furthermore, both Haskell and its intermediate language Core are \emph{lazily}
-        evaluated -- expressions in Haskell are only evaluated when needed
-        (unlike, e.g., Rust, which is \emph{eagerly} evaluated).
-        %
-        % Laziness allows an optimising compiler to aggressively transform the source
-        % program without changing its semantics and, indeed, the Glasgow Haskell
-        % Compiler (GHC) heavily transforms Core by leveraging its laziness.
-        %
-        As we make clear in this paper, lazy evaluation interacts non-trivially with linearity.
-        Intuitively, since expressions are not necessarily evaluated,
-        a syntactic occurrence of a linear resource in an expression does not necessarily
-        entail consuming that resource (i.e., if the expression is not evaluated,
-        the resource is not used).
-%
-        In eagerly evaluated languages, the distinction between syntactic uses of a
-        resource and the actual use of linear resources at runtime does not exist --
-        an occurrence of a variable in the program always entails consuming it.
-        %
-        This interaction is unique to Haskell since, to the best of
-        our knowledge, it is the
-        only language featuring both laziness and linearity.
+Both Haskell and its intermediate language Core are lazily
+        evaluated.
+        % -- expressions in Haskell are only evaluated when needed.
+        As we make clear in this paper, \emph{lazy evaluation interacts non-trivially with linearity}.
+        Intuitively, a syntactic occurrence of a linear resource does not
+        necessarily entail consuming that resource because expressions are not
+        necessarily evaluated (i.e., if the expression is not evaluated, the
+        resource is not used).
+        %
+        Under eager evaluation, a syntactic occurrence of a resource always implies its use at runtime.
+        %
+        % Under eager/call-by-value evaluation, the distinction between syntactic uses of a
+        % resource and the actual use of linear resources at runtime does not exist --
+        % an occurrence of a variable in the program always entails consuming it.
+        %
+
+        % Include?
+        % This interaction is unique to Haskell since, to the best of
+        % our knowledge, it is the
+        % only language featuring both laziness and linearity.
 
@@
 translation from Haskell (dubbed \emph{desugaring}) and the subsequent
-optimising transformations, a \emph{linearly typed} Core would guarantee that
+optimising transformations, a \emph{linearly typed Core} would guarantee that
 linear resource usage in the source language is not violated by desugaring
@@
 For instance, let $x$ be a linear resource in the two following programs, where
-the latter results from inlining $y$ in the let body of the former. Even though the
+the latter results from inlining $y$ in the let body of the former. Even though
 the program no longer appears linear (as there are now two 
-occurrences of $x$), the program \emph{is} indeed linear
+occurrences of $x$), the program is indeed linear
 because the let-bound expression is never evaluated, so $x$ is
@@
 The Core optimising transformations expose a fundamental limitation of Core's
-linear type system: it does not account for the call-by-need evaluation model
-of Core, and thus a whole class of programs that are linear under the lens of
+linear type system: it does not account for the lazy evaluation model
+of Core and, thus, a whole class of programs that are linear under the lens of
 lazy evaluation are rejected by Core's current linear type system.
 %
-In this work, we address this limitation (and its implications on validating
-and influencing optimising transformations) by designing a type system which
-understands \emph{semantic linearity} in the presence of laziness and is suitable for
-the intermediate language of an optimising compiler such as the
-Glasgow Haskell Compiler.
+In this work, we address this limitation by exploring how syntactic linearity
+breaks down in the presence of laziness and designing a type system which,
+derived from our observations, accepts the vast majority of valid non-strict
+programs and optimising transformations previously rejected by traditional
+linear typing disciplines. Additionally, we show our system
+%
+% and its implications on validating and influencing optimising transformations
+% by designing a type system which analyses how syntactic linearity breaks down
+% in the presence of laziness
+is suitable for the intermediate language of an
+optimising compiler by implementing it as a Glasgow Haskell Compiler plugin.
 %

### 2025-01-12 icfp.tex
Touches Section 2

@@
 \emph{syntactic occurrence} in the program. However, as this section
-makes clear, when considering linear type systems for non-strict
+makes clear, in linear type systems for non-strict
 languages, a distinction between using
@@
 %
-For instance, in the following program in a Haskell-like language there are two
-syntactic occurrences of \emph{handle}, and a let-bound computation that closes
-this handle is bound to \lstinline{x} before the \lstinline{handle} is
-returned. However, if the program is evaluated lazily, the handle is \emph{not}
-closed because the computation that closes the handle is never evaluated (since
-\lstinline{x} goes unused).
+For instance, in the following Haskell-like program, there are two syntactic
+occurrences of \lstinline{handle}: one is in a let-bound computation that
+closes this handle, the other is returned. If the program is evaluated lazily,
+the handle will not be closed because the computation that closes the handle is
+never evaluated (\lstinline{x} goes unused).
 %
@@
 will not consume that resource.
-Under the lens of \emph{non-strict} evaluation, this means that a syntactic
+Under the lens of non-strict evaluation, this means that a syntactic
 occurrence of a resource in a let binding only entails consuming that resource
@@
 semantics should accept the above program, unlike a linear type system for the
-same program if it were to be evaluated \emph{eagerly}.
+same program if it were to be evaluated eagerly.
 %
@@
 %
-we turn our focus to \emph{linearity} under \emph{call-by-need}, not only
-because GHC Core is \emph{call-by-need}, but also because the distinction
-between semantically and syntactically consuming a resource is only exposed
-under \emph{non-strict} semantics.
+we turn our focus to linearity under non-strict evaluation, both since GHC Core is
+non-strict and because the distinction between semantically and syntactically
+consuming a resource is only exposed under non-strict semantics.
 %
@@
 
-Aligned with our original motivation of typechecking linearity in GHC Core such
+Aligned with our motivation of typing linearity in GHC Core such
 that optimising transformations preserve linearity, and with the goal
@@
 build an intuition for semantic linearity through examples of Core
-programs that are semantically linear but are rejected by (the current) Core linear type
-system.
+programs that are semantically linear but are rejected by (the current) Core
+linear type system.
 % BETTER?:
@@
 \colorbox{notyet}{\notyetcolorname} or \colorbox{limitation}{\limitationcolorname} background mark programs that are
-\emph{semantically} linear, but are not seen as linear by Core's (naive wrt lazy evaluation) linear type
+semantically linear, but are not seen as linear by Core's (naive wrt lazy evaluation) linear type
 system. Notably, the linear type system we develop in this work accepts all
@@
 
+\todo{Should we have a section starting with simple function application?}
+
 \paragraph{Let bindings}
-We start our discussion with non-strict (non-recursive) let bindings, i.e. let bindings whose
-body is evaluated only when the binding is needed, rather than when declared.
-In Core, a let binding entails the creation of a \emph{thunk} that suspends the
-evaluation of the let body.
-% (for background, see
-% Section~\ref{sec:background:evaluation-strategies})
-When the binding itself is evaluated, the \emph{thunk} is
-\emph{forced} and the evaluation is carried out. The result overwrites the
-\emph{thunk} -- the let binding now points to the result of the evaluation.
+% We start our discussion with non-strict (non-recursive) let bindings, i.e. let bindings whose
+% body is evaluated only when the binding is needed, rather than when declared.
+%
+% In Core, let bindings entail creating \emph{thunks} that suspends the
+% evaluation of the let body.
+% % (for background, see
+% % Section~\ref{sec:background:evaluation-strategies})
+% When the binding itself is evaluated, the \emph{thunk} is
+% \emph{forced} and the evaluation is carried out. The result overwrites the
+% \emph{thunk} -- the let binding now points to the result of the evaluation.
+%
 % A \emph{thunk} is \emph{forced} (and the suspended computation is evaluated) when
 % the binding itself is evaluated.
-
+%
 %\begin{notyet}
@@
 % returned.
-
+%
 % \begin{figure}
@@
 % \end{figure}
-
+%
+%
+%
+%
 % %
@@
 
-% \todo[inline]{Cut this one}
-In the example from the beginning of
-Section~\ref{sec:linearity-semantically} (similar to that towards the end
-of Section~\ref{sec:intro}), we discussed a scenario where an
-unused let binding can be used to justify a function where its
-argument occurs both in the (unused) let-bound expression and in the
-body of the let.
-Let bindings can also go unused if they are defined before branching on case
-alternatives. At runtime, depending on the branch taken, the let binding will
-be evaluated only if it occurs in that branch.
-Both optimising transformations (float-out), and programmers used to
-non-strict evaluation often produce
-programs with bindings that are selectively used in case alternatives, for
-instance:
+The example at the start of Section~\ref{sec:linearity-semantically}
+% (similar to towards the end of Section~\ref{sec:intro}),
+trivially showed how linearity can be more subtle in a non-strict language with
+a simple unused let binding.
+%
+Less trivially, let-bound expressions that capture linears variables
+may be evaluated depending on the case-expression branch taken at runtime.
+Both optimising transformations (float-out) and programmers used to
+non-strict evaluation are keen to write programs with bindings that are selectively
+used in case alternatives, such as:
 %
@@
 %
-This example uses \lstinline{x} directly in one branch and the let-bound
+This example returns \lstinline{x} in one branch and the let-bound
 \lstinline{y} in the other. Semantically, as opposed to syntactically, this
-program is linear (i.e. linear resources are used \emph{exactly once}) because
-the linear resource \lstinline{x} is used exactly once in both case
-alternatives, either directly in the \lstinline{True} branch, or \emph{via} the
+program is linear % (i.e. linear resources are used exactly once)
+because the linear resource \lstinline{x} is used exactly once regardless of which branch is
+taken: either directly in the \lstinline{True} branch, or \emph{via} the
 let-bound \lstinline{y} in the \lstinline{False} branch.
 
-% Shifting our focus from not using a let binding to using it
-% more than once.
-On the other hand, let us consider a let-binding \lstinline{y} that is used
-twice, and that binds an expression that depends on a linear resource
-\lstinline{x}.
-Despite the binder body only being evaluated once due to call-by-need, and thus
-its resources being only used once to compute a result, we can still only
-consume the result of the computation once -- perhaps surprisingly, as the
-perception so far is that ``resources are consumed during computation'' and
-multiple uses of the same let binder share the result that was computed only
-once. Illustratively, the following program must \emph{not} typecheck:
+That said, a let-bound \lstinline{y} that captures a linear resource
+\lstinline{x} must still be used exactly once, regardless of whether evaluation
+inlines the let (ala call-by-name) or creates a thunk (ala call-by-need):
+%
+% \lstinline{y} must still be used exactly once.  When \lstinline{y} is inlined
+% in the body, this is trivial to see since multiple occurrences of will be 
 %
@@
 %
-Intuitively, the result of the computation must also be used exactly once,
-despite being effectively computed just once, since the result may
-still capture (parts of) the linear resource. The trivial example is \lstinline{f1} applied to
-\incode{id} -- the result of computing \incode{id x} is \incode{x}, and \incode{x} must definitely not be
-shared. Indeed, if the result of the computation involving the linear resource
-was instead, e.g., an unrestricted integer, then sharing the result would not
-involve consuming the resource more than once.
-% ROMES:TODO:!!!!
-%\todo{this kind of hints that maybe somehow in the
-%alternative scrutinizing a "trivial" atom we could make the let binding
-%unrestricted after we were sure it was evaluated once, but that isn't easy
-%without erasing too many things before, we kind of tried this once}
-%
-The result of evaluating a let-binder's body which depends on linear resources,
-if computed, must be consumed exactly once, or, otherwise, we risk discarding
-or duplicating (the parts of) the resources that were captured in the result.
+Linearity is trivially violated if \lstinline{y} is inlined.
+%
+On the other hand, if the computation using the resource is suspended under a
+thunk, it will be evaluated only once when the binding is first forced and
+re-used subsequently. Despite the computation (here, \lstinline{use x}) only
+being evaluated once, the program must still be rejected.
+%
+Intuitively, linearity may be violated if the computation leaks
+the resource or parts of it. For instance, if \lstinline{use = id}, the program
+must clearly be rejected.
+%
+% TODO:NEW: Include?
+% If the computation fully consumed the linear resource to produce an
+% unrestricted result which was forced, the binding could be shared...
+% OLD Indeed, if the result of the computation involving the linear resource
+% was instead, e.g., an unrestricted integer, then sharing the result would not
+% involve consuming the resource more than once.
+%
+%
+% Redundant?
+% In short, the result of evaluating a let-binder's body which depends on linear resources,
+% if computed, must be consumed exactly once, or, otherwise, we risk discarding
+% or duplicating (the parts of) the resources that were captured in the result.
 

### 2025-01-20 icfp_submission_2024.tex
historical copy

--- 2024-02-28 icfp.tex
@@
 % \date{ }
+
+\begin{CCSXML}
+<ccs2012>
+   <concept>
+       <concept_id>10003752.10003790.10003801</concept_id>
+       <concept_desc>Theory of computation~Linear logic</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10003752.10003790.10011740</concept_id>
+       <concept_desc>Theory of computation~Type theory</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10011007.10011006.10011008.10011009.10011012</concept_id>
+       <concept_desc>Software and its engineering~Functional languages</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10011007.10011006.10011041</concept_id>
+       <concept_desc>Software and its engineering~Compilers</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+ </ccs2012>
+\end{CCSXML}
+
+\ccsdesc[500]{Theory of computation~Linear logic}
+\ccsdesc[500]{Theory of computation~Type theory}
+\ccsdesc[500]{Software and its engineering~Functional languages}
+\ccsdesc[500]{Software and its engineering~Compilers}
+
+\keywords{Linear Types, Lazy Evaluation, Haskell, GHC Core}
 

### 2025-01-21 icfp.tex
tweak abstract

@@
 \begin{abstract}
-Linear type systems guarantee linear resources are used \emph{exactly once}.
+Linear type systems guarantee linear resources are used exactly once.
 Traditionally, using a resource is synonymous with its \emph{syntactic}
-occurrence in the program, however, under the lens of \emph{lazy} evaluation,
+occurrence in the program, however, under the lens of non-strict evaluation,
 linearity can be further understood \emph{semantically}, where a
 syntactic occurrence of a resource does not necessarily entail
-\emph{using} that resource when the program is executed.
+using that resource when the program is executed.
 Semantic linearity becomes especially vital when considering the
@@
 
-\todo{REWRITE}
 We present Linear Core, the first type system that understands semantic
-linearity in the presence of laziness, suitable for the Core intermediate
-language of the Glasgow Haskell Compiler. We prove Linear Core is both type
-safe and that multiple optimising transformations preserve linearity in Linear
-Core while failing to do so in Core. We have implemented Linear Core as a
-compiler plugin to validate the system against linearity-heavy libraries,
-including \texttt{linear-base}.%^, in the heart of the compiler.
+linearity in the presence of non-strict evaluation, suitable for the Core
+intermediate language of the Glasgow Haskell Compiler. We prove Linear Core
+guarantees linear resource usage via a type safety argument and prove
+multiple optimising transformations preserve linearity in Linear Core while
+failing to do so in Core. We have implemented Linear Core as a compiler
+plugin to validate the system against linearity-heavy libraries, including
+\texttt{linear-base}.%^, in the heart of the compiler.
 \end{abstract}

### 2025-01-21 icfp.tex
1 and 2:header

@@
 intermediate language of the Glasgow Haskell Compiler. We prove Linear Core
-guarantees linear resource usage via a type safety argument and prove
+guarantees linear resource usage via a type safety and bisimilarity argument and prove
 multiple optimising transformations preserve linearity in Linear Core while
@@
 Linear type systems~\cite{cite:linear-logic,cite:barberdill} add expressiveness
-to existing type systems by enforcing that certain \emph{resources} (e.g.~a
+to existing type systems by enforcing that certain resources (e.g.~a
 file handle) are used \emph{exactly once}.
@@
         % -- expressions in Haskell are only evaluated when needed.
-        As we make clear in this paper, \emph{lazy evaluation interacts non-trivially with linearity}.
+        As we make clear in this paper, \emph{non-strict evaluation interacts non-trivially with linearity}.
         Intuitively, a syntactic occurrence of a linear resource does not
@@
 translation from Haskell (dubbed \emph{desugaring}) and the subsequent
-optimising transformations, a \emph{linearly typed Core} would guarantee that
+optimising transformations, a linearly typed Core would guarantee that
 linear resource usage in the source language is not violated by desugaring
@@
 in the source Haskell code. The reason is that the desugaring
-and the subsequent Core-to-Core optimising transformations
+and subsequent Core-to-Core optimising transformations
 eliminate and rearrange most of the syntactic constructs through
@@
 %
-For instance, let $x$ be a linear resource in the two following programs, where
-the latter results from inlining $y$ in the let body of the former. Even though
-the program no longer appears linear (as there are now two 
-occurrences of $x$), the program is indeed linear
-because the let-bound expression is never evaluated, so $x$ is
-consumed exactly once (when it is freed in the let body):
+As a trivial example, let $x$ be a linear resource in the next two expressions,
+where the latter results from inlining $y$ without eliminating the binder. Even
+though the second expression no longer appears linear (as there are now two occurrences
+of $x$), it is indeed linear because the let-bound expression is never
+evaluated, thus $x$ is still consumed exactly once (when it is freed in the let body):
 % executed, $x$ will be freed exactly once:
@@
 The Core optimising transformations expose a fundamental limitation of Core's
-linear type system: it does not account for the lazy evaluation model
+linear type system: it does not account for the non-strict evaluation model
 of Core and, thus, a whole class of programs that are linear under the lens of
-lazy evaluation are rejected by Core's current linear type system.
+non-strict evaluation are rejected by Core's current linear type system.
 %
 In this work, we address this limitation by exploring how syntactic linearity
-breaks down in the presence of laziness and designing a type system which,
+breaks down in the presence of non-strictness and designing a type system which,
 derived from our observations, accepts the vast majority of valid non-strict
@@
 
-\item We introduce Linear Core, a type system for a linear lazy language with
-the key features from Core (except for type equality coercions), which,
+\item We introduce Linear Core, a non-strict linear language with
+the key features from Core (except for type equality coercions), whose type system
 accounts for semantic linearity in the presence of laziness. To the
@@
 
-\item We show the type soundness of Linear Core and prove that
-  multiple optimising transformations, which
-fail to type check in Core, are validated by Linear
-Core (Section~\ref{sec:main:metatheory}).
+\item We prove our type system is sound, guarantees linear resource usage, and
+    prove that multiple optimising transformations, which are rejected in Core,
+        are validated by Linear Core (Section~\ref{sec:main:metatheory}).
 
@@
 A linear type system statically guarantees that linear resources are
-\emph{consumed} exactly once. Consequently, whether a program is
+consumed exactly once. Consequently, whether a program is
 well-typed under a linear type system intrinsically depends on the
-definition of \emph{consuming} a resource.
-In general, \emph{consuming} a resource is equated with its 
+definition of consuming a resource.
+In general, consuming a resource is equated with its 
 \emph{syntactic occurrence} in the program. However, as this section
@@
 
+%Under the lens of non-strict evaluation, a syntactic
+%occurrence of a resource in an expression only entails consuming that resource
+%iff the expression itself is fully consumed.
+%
 Intuitively, a computation that depends on a linear resource to produce a
-result consumes that resource iff the result is effectively consumed; in
+result consumes that resource iff the result is fully consumed; in
 contrast, a computation that depends on a linear resource, but is never run,
 will not consume that resource.
-Under the lens of non-strict evaluation, this means that a syntactic
-occurrence of a resource in a let binding only entails consuming that resource
-iff the binding itself is consumed.
 %
@@
 we turn our focus to linearity under non-strict evaluation, both since GHC Core is
-non-strict and because the distinction between semantically and syntactically
+lazy and because the distinction between semantically and syntactically
 consuming a resource is only exposed under non-strict semantics.

### 2025-01-21 icfp.tex
More 2.1

@@
 
-Aligned with our motivation of typing linearity in GHC Core such
-that optimising transformations preserve linearity, and with the goal
-of understanding linearity in a non-strict context, this section aims to
-build an intuition for semantic linearity through examples of Core
-programs that are semantically linear but are rejected by (the current) Core
-linear type system.
+This section aims to build an intuition for semantic linearity through examples
+of Core programs that can be understood as linear but are rejected by Core's
+linear type system (or, in fact, by all linear type systems we are aware of).
+%
+% Needed?
+%Aligned with our motivation of typing linearity in GHC Core such
+%that optimising transformations preserve linearity, and with the goal
+%of understanding linearity in a non-strict context,
+%
 % BETTER?:
@@
 %
-In the examples, a \colorbox{working}{\workingcolorname} background highlights programs that are
-syntactically linear and accepted by Core's naive linear type system. A
-\colorbox{notyet}{\notyetcolorname} or \colorbox{limitation}{\limitationcolorname} background mark programs that are
-semantically linear, but are not seen as linear by Core's (naive wrt lazy evaluation) linear type
-system. Notably, the linear type system we develop in this work accepts all
+In the examples, a \colorbox{working}{\workingcolorname} background highlights
+programs that are syntactically linear and accepted by Core's linear type
+system.
+%
+A \colorbox{notyet}{\notyetcolorname} or
+\colorbox{limitation}{\limitationcolorname} background mark programs that are
+semantically linear, but not seen as linear by Core's type system.
+%
+Notably, the linear type system we develop in this work accepts all
 \colorbox{notyet}{\notyetcolorname} programs.
+%
 % (while the few \colorbox{limitation}{\limitationcolorname} programs are not accepted).
@@
 
-\todo{Should we have a section starting with simple function application?}
+%\todo{Should we have a subsection starting with simple function application?}
 
@@
 %
-Less trivially, let-bound expressions that capture linears variables
-may be evaluated depending on the case-expression branch taken at runtime.
+Less trivially, a let-bound expression that captures a linear variable
+could be conditionally needed at runtime, depending on the branch taken.
 Both optimising transformations (float-out) and programmers used to
@@
 On the other hand, if the computation using the resource is suspended under a
-thunk, it will be evaluated only once when the binding is first forced and
+thunk, it will only be evaluated once, when the binding is first forced, and
 re-used subsequently. Despite the computation (here, \lstinline{use x}) only
@@
 Even though the binding \incode{y} is used in \incode{z}, \incode{x} is still never consumed because
-\incode{z} isn't evaluated in the let body, and consequently \incode{y} isn't evaluated
+\incode{z} isn't evaluated in the let body, and, consequently, \incode{y} isn't evaluated
 either -- never consuming \incode{x}. We use this example to highlight that even for
@@
 once. Let binders that depend on linear resources must be used \emph{at most
-once} -- let bound variables are \emph{affine} in the let body.
-%
-Moreover, if the let binding ($y$) isn't used in the let body, then the
-resources it depends on ($\ov{x}$) must still be used -- the binding $y$ is
-mutually exclusive with the resources $\ov{x}$ (for the resources to be used
-linearly, either the binder occurs exactly once $y$, or the resources $\ov{x}$
-do). We discuss how to encode this principle between let bindings and their dependencies using so called \emph{usage
+once}, mutually exclusively with the linear resources themselves
+(in a sense, let-vars are affine in the let-body).
+%
+%Moreover, if the let binding \incode{y} isn't used in the let body, then the
+%resources it depends on ($\ov{x}$) must still be used -- the binding $y$ is
+%mutually exclusive with the resources $\ov{x}$ (for the resources to be used
+%linearly, either the binder occurs exactly once $y$, or the resources $\ov{x}$
+%do).
+%
+We discuss how to encode this principle between let bindings and their dependencies using so called \emph{usage
 environments}, in Section~\ref{sec:usage-environments}.

### 2025-01-22 icfp.tex
Touches S2.1 Rec Let

@@
 %
-Even though the binding \incode{y} is used in \incode{z}, \incode{x} is still never consumed because
-\incode{z} isn't evaluated in the let body, and, consequently, \incode{y} isn't evaluated
-either -- never consuming \incode{x}. We use this example to highlight that even for
-let-bound variables, the syntactic occurrence of a variable isn't enough to
-determine whether it is used. Instead, we ought to think of uses of \incode{y} as
-implying using \incode{x}, and therefore uses of \incode{z} imply using \incode{x}, however, if
-neither is used, then \incode{x} isn't used. Since \incode{x} is effectively discarded, this
-example also violates linearity.
+Even though the binding \incode{y} is used in \incode{z}, \incode{x} is never
+consumed because \incode{z} isn't evaluated in the let body, and, thus,
+\incode{y} isn't evaluated either. We use this example to highlight that even
+for let-bound variables, the syntactic occurrence of a variable isn't enough to
+determine whether it is used. Instead, we ought to think of uses of \incode{y}
+as implying using \incode{x}, and therefore uses of \incode{z} imply using
+\incode{x}, however, if neither is used, then \incode{x} isn't used. Since
+\incode{x} is effectively discarded, this example also violates linearity.
 
@@
 once. Let binders that depend on linear resources must be used \emph{at most
-once}, mutually exclusively with the linear resources themselves
+once} in mutual exclusion with the linear resources themselves
 (in a sense, let-vars are affine in the let-body).
@@
 \paragraph{Recursive let bindings\label{sec:semantic-linearity-examples:recursive-lets}}
-% Second, we look into recursive let bindings.
-For the most part,
-recursive let bindings behave as non-recursive let bindings, i.e. we must use them \emph{at
-most once} because, when evaluated, the linear resources used in the binders
-bodies are consumed. The defining property of a group of mutually recursive let
-bindings is that the binders bound in that group can occur, without restrictions, in
-the bodies of those same binders. % The same way that, in a let body, evaluating
+
+%Recursive lets behave similarly to non-recursive let bindings in the sense we
+%must use them \emph{at most once} because, when evaluated, the linear resources
+%used in the binders bodies are consumed.
+%
+
+Mutually recursive let-bindings, i.e. bindings with definitions that
+recursively depend on the bindings they define, behave similarly to
+non-recursive lets as far as their usage in the let continuation is concerned.
+%recursive lets must still be used at most once, in mutual exclusion with the
+%linear resources they capture.
+%
+The challenge lies in understanding linearity in the definitions of the
+recursive binders and which resources exactly are consumed when one of the
+binders in the recursive group is used.
+
+% The same way that, in a let body, evaluating
 % a binding that uses some resource exactly once consumes that resource once,

### 2025-01-24 icfp_submission_2024.tex
started instrumented semantics

@@
 
+\section{Instrumented Operational Semantics}
+
+In this section we develop a lazy big-step operational semantics for $\lambda_\Delta^\pi$ in the
+style of Launchbury~\cite{10.1145/158511.158618}. The semantics is
+equipped with an environment which assigns values to variables and is
+mutated in order to express lazy evaluation. Terms are explicit about
+their sharing (in the sense of shared reductions in lazy evaluation),
+and so we perform a translation step that makes all sharing explicit
+through let-binders.
+Below we write $\llet{x{:}_{\Delta}\sigma = e,\dots,y{:}_{\Delta'}=e''
+}{e'}$ for the iterated let-binding of variables $x , \dots , y$.       
+\[
+  \begin{array}{lclr}
+    x^* & \triangleq & x\\
+
+    (\Lambda p.~e)^* & \triangleq & \Lambda p.~(e)^*\\
+    (\lambda x:_\pi\sigma .e)^* & \triangleq & \lambda x:_\pi\sigma . (e)^*\\
+    (e~\pi)^* & \triangleq & (e)^*~\pi\\
+    (e~x)^* & \triangleq & (e)^*~x\\
+    (e~e')^* & \triangleq & \llet{x{:}_{\Delta}\sigma =
+                            (e')^*}{(e)^*~x}     \\ &&\qquad \mbox{with
+                            $\Gamma;\Delta' \vdash e :
+                            \sigma\rightarrow_\pi \varphi$ and
+                                                   $\Gamma;\Delta \vdash e' : \sigma$}\\
+    K~e_1~\dots~e_n & \triangleq & \llet{x_1:_{\Delta_1} \sigma_1 =
+                                   (e_1)^*,\dots,x_n:_{\Delta_n}
+                                   \sigma_n =
+                                   (e_n)^*}{K~x_1~\dots~x_n} \\ &&
+                                                               \qquad\mbox{with
+                                                               $\Gamma;\Delta_1\vdash
+                                                               e_1 :
+                                                               \sigma_1
+                                                               \dots \Gamma;\Delta_n\vdash
+                                                               e_n :
+                                                               \sigma_n$}\\                                              
+(\ccase{e}{z{:}_{\Delta}\sigma~\{\overline{\rho\to e'}\}})^* &
+                                                               \triangleq
+                                &
+                                  \ccase{(e)^*}{z{:}_{\Delta}\sigma~\{\overline{\rho\to (e')^*}\}}\\
+    (\llet{x{:}_{\Delta}\sigma = e}{e'}      )^* & \triangleq & \llet{x{:}_{\Delta}\sigma  = (e)^*}{(e')^*}    \\
+    (\lletrec{\overline{x{:}_{\Delta}\sigma = e}}{e'}  )^* & \triangleq &\lletrec{\overline{x{:}_{\Delta}\sigma = (e)^*}}{(e')^*}  
+      \end{array}
+\]
+
+The big-step operational semantics is defined via the relation
+$\Gamma : e \Downarrow \Delta : v$ where $e$ is an expression, $v$ a
+value, $\Gamma$ and $\Delta$ are environments with bindings of the
+form $x:_\omega\sigma = e$ which assigns the expression $e$ to the
+variable $x$ of the given type.
+
+\[
+  \begin{array}{c}
+    \infer[]
+    {\,}
+    {\Gamma : \Lambda p.e \Downarrow \Gamma : \Lambda p.e}
+    \qquad
+    \infer[]
+    {\Gamma : e\Downarrow \Delta : \Lambda p.e' \quad \Delta :
+    e'[\pi/p]\Downarrow \Theta : v}
+    {\Gamma : e~\pi\Downarrow \Theta : v }\\[1.5em]
+    \infer[]
+    {\,}
+    {\Gamma : \lambda x:_\pi \sigma . e \Downarrow \Gamma : \lambda x:_\pi \sigma . e }
+    \qquad
+    \infer[]
+    {\Gamma : e \Downarrow \Delta : \lambda x:_\pi \sigma . e' \quad
+    \Delta : e'[x/y] \Downarrow \Theta : v }
+    {\Gamma : e~x \Downarrow \Theta : v}\\[1.5em]
+    \infer[]
+    {\Gamma : e \Downarrow \Delta : v}
+    {(\Gamma , x:_\omega \sigma = e) : x \Downarrow (\Delta,x:_\omega
+    \sigma = v) : v}
+    \qquad
+    \infer[]
+    {(\Gamma,x:_\omega \sigma = e) : e' \Downarrow \Delta : v}
+    {\Gamma : \llet{x{:}_{\Delta}\sigma = e}{e'}\Downarrow \Delta : v
+    }\\[1.5em]
+    \infer[]
+    {\,}
+    {\Gamma : K~x_1~\dots~x_n \Downarrow \Gamma : K~x_1~\dots~x_n}
+    \qquad
+    \infer[]
+    {\Gamma : e \Downarrow \Delta : K~x_1~\dots~x_n \quad
+    \Delta : e'[x_i/y_i] \Downarrow \Theta : v }
+    {\Gamma : \ccase{e}{z{:}_{\Delta}\sigma~\{\overline{\dots,K~y_1~\dots~y_n\to e'}\}}
+    \Downarrow \Theta : v}\\[1.5em]
+    \infer[]
+    {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
+    % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
+    % \Downarrow \Theta_n : v_n \quad
+    (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e' \Downarrow \Theta
+    : v
+    }
+    {\Gamma : \lletrec{\overline{x_i{:}_{\Delta}\sigma_i = e_i}}{e'}
+    \Downarrow \Theta : v}
+    \end{array}
+\]
+
+We follow an approach
+similar to that of \cite{cite:linearhaskell}.
+
 % }}}

### 2025-01-24 icfp_submission_2024.tex
added linearity-aware semantics

@@
     \infer[]
-    {\Gamma : e \Downarrow \Delta : \lambda x:_\pi \sigma . e' \quad
+    {\Gamma : e \Downarrow \Delta : \lambda y:_\pi \sigma . e' \quad
     \Delta : e'[x/y] \Downarrow \Theta : v }
@@
     \infer[]
-    {(\Gamma,x:_\omega \sigma = e) : e' \Downarrow \Delta : v}
-    {\Gamma : \llet{x{:}_{\Delta}\sigma = e}{e'}\Downarrow \Delta : v
+    {(\Gamma,x:_\omega \sigma = e) : e' \Downarrow \Theta : v}
+    {\Gamma : \llet{x{:}_{\Delta}\sigma = e}{e'}\Downarrow \Theta : v
     }\\[1.5em]
@@
     {\Gamma : e \Downarrow \Delta : K~x_1~\dots~x_n \quad
-    \Delta : e'[x_i/y_i] \Downarrow \Theta : v }
-    {\Gamma : \ccase{e}{z{:}_{\Delta}\sigma~\{\overline{\dots,K~y_1~\dots~y_n\to e'}\}}
+    \Delta : e'[x_i/y_i][K~x_1~\dots~x_n/z] \Downarrow \Theta : v }
+    {\Gamma :
+    \ccase{e}{z{:}_{\Delta'}\sigma~\{{\dots,K~\ov{y_i}, {\dots} \to e'}\}}
     \Downarrow \Theta : v}\\[1.5em]
+    \infer[]
+    {\Gamma : e \Downarrow \Delta : K~x_1~\dots~x_n \quad
+    \Delta : e'[K~x_1~\dots~x_n/z] \Downarrow \Theta : v }
+    {\Gamma :
+    \ccase{e}{z{:}_{\Delta'}\sigma~\{\dots , \_\to
+    e'}\} \Downarrow \Theta : v}
+    \qquad
     \infer[]
@@
 We follow an approach
-similar to that of \cite{cite:linearhaskell}.
+similar to that of \cite{cite:linearhaskell}, defining a linearity
+aware operational semantics in which linear variables are erased from
+the runtime environment, ensuring that any term violating linearity
+will result in a stuck state. The semantics is type-aware, defined by
+the judgment $\Gamma ; \Delta \vdash (\Theta \mid e) \Downarrow (\Theta'
+\mid v) : \sigma$ where $\Gamma$ and $\Delta$ are typing judgments
+($\Gamma$ tracks unrestricted variables, $\Delta$ tracks linear
+assumptions);
+$\Theta$ and $\Theta'$ are runtime environments, consisting of bindings of the form
+$x :_\pi \sigma = e$; $e$ is the expression being evaluated and $v$
+its resulting value, both of type $\sigma$.
+
+\[
+  \begin{array}{c}
+    \infer[]
+    {\,}
+    {\Gamma;\Delta \vdash (\Theta \mid \Lambda p.e) \Downarrow (\Theta \mid
+    \Lambda p.e) : \forall p . \sigma }
+    \\[1.5em]
+    \infer[]
+    {\Gamma ; \Delta \vdash (\Theta \mid  e) \Downarrow (\Theta' \mid
+    \Lambda p.e') : \forall p . \sigma \quad \Gamma ;\Delta \vdash
+    (\Theta' \mid e'[\pi/p]) \Downarrow (\Theta'' : v) : \sigma}
+    {\Gamma ; \Delta \vdash (\Theta \mid e~\pi) \Downarrow (\Theta''
+    \mid v) : \sigma  }\\[1.5em]
+    \infer[]
+    {\,}
+    {\Gamma ;\Delta \vdash (\Theta \mid \lambda x:_\pi \sigma . e)
+    \Downarrow (\Theta \mid \lambda x:_\pi\sigma .e) : \sigma
+    \rightarrow_\pi \tau }\\[1.5em]
+    \infer[]
+    {\Gamma;\Delta \vdash  (\Theta \mid e) \Downarrow (\Theta' \mid
+    \lambda y:_\pi \sigma . e') : \sigma \rightarrow_\pi \tau
+    \quad
+    \Gamma ; \Delta \vdash (\Theta' \mid e'[x/y]) \Downarrow (\Theta''
+    : v) : \tau }
+    {\Gamma ; \Delta\vdash (\Theta \mid  e~x ) \Downarrow (\Theta'
+    \mid v) : \tau}\\[1.5em]
+
+       \infer[]
+    {\Gamma , x:_{\Delta'} \sigma ; \Delta  \vdash (\Theta \mid e)
+    \Downarrow (\Theta' \mid v) : \sigma}
+    {\Gamma ; \Delta \vdash (\Theta , x:_{\Delta'} \sigma = e \mid x)
+    \Downarrow (\Theta' , x:_{\Delta'} \sigma = v \mid v) : \sigma}
+    \\[1.5em]
+           \infer[]
+    {\Gamma ;\Delta  \vdash (\Theta \mid e)
+    \Downarrow (\Theta' \mid v) : \sigma}
+    {\Gamma ; \Delta \vdash (\Theta , x:_1 \sigma = e \mid x)
+    \Downarrow (\Theta' \mid v) : \sigma}
+    \\[1.5em]
+
+    \infer[]
+    {\Gamma ;\Delta \vdash (\Theta,x:_{\Delta'} \sigma = e \mid e')
+    \Downarrow (\Theta' \mid v) : \tau}
+    {\Gamma ; \Delta \vdash (\Theta \mid \llet{x{:}_{\Delta}\sigma =
+    e}{e'})\Downarrow (\Theta' \mid v) : \tau
+    }\\[1.5em]
+
+    \infer[]
+    {\,}
+    {\Gamma ; \Delta \vdash (\Theta \mid K~x_1~\dots~x_n) \Downarrow
+    (\Theta \mid K~x_1~\dots~x_n) : T}
+    \\[1.5em]
+
+    \infer[]
+    {\Gamma;\Delta \vdash (\Theta \mid e) \Downarrow (\Theta' :
+    K~x_1~\dots~x_n) : T \quad
+    \Gamma ;\Delta \vdash (\Theta' \mid e'[x_i/y_i][K~x_1~\dots~x_n/z]
+    ) \Downarrow (\Theta'' : v) :\tau }
+    {\Gamma ;\Delta \vdash (\Theta \mid
+    \ccase{e}{z{:}_{\Delta'}\sigma~\{{\dots,K~\ov{y_i}, {\dots} \to e'}\}})
+    \Downarrow (\Theta'' \mid  v) : \tau}\\[1.5em]
+
+    \infer[]
+    {\Gamma ;\Delta \vdash (\Theta \mid e) \Downarrow (\Theta' \mid
+    K~x_1~\dots~x_n) : T \quad
+    \Gamma ;\Delta \vdash (\Theta' \mid  e'[K~x_1~\dots~x_n/z])
+    \Downarrow (\Theta'' \mid v) : \tau }
+    {\Gamma ; \Delta \vdash (\Theta \mid 
+    \ccase{e}{z{:}_{\Delta'}\sigma~\{\dots , \_\to
+    e'}\}) \Downarrow (\Theta'' \mid v) : \tau}
+    \\[1.5em]
+    \infer[]
+    {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
+    % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
+    % \Downarrow \Theta_n : v_n \quad
+    (\Gamma ; \Delta \vdash (\Theta, \overline{x_i:_\omega \sigma_i =
+    e_i} \mid e' )\Downarrow (\Theta'
+    \mid v) : \tau
+    }
+    {\Gamma ;\Delta \vdash (\Theta \mid \lletrec{\overline{x_i{:}_{\Delta}\sigma_i = e_i}}{e'})
+    \Downarrow (\Theta' \mid v) : \tau}
+
+   \end{array}
+\]
 

### 2025-01-24 icfp.tex
Move new section to live .tex file

@@
 
+\section{Instrumented Operational Semantics}
+
+In this section we develop a lazy big-step operational semantics for $\lambda_\Delta^\pi$ in the
+style of Launchbury~\cite{10.1145/158511.158618}. The semantics is
+equipped with an environment which assigns values to variables and is
+mutated in order to express lazy evaluation. Terms are explicit about
+their sharing (in the sense of shared reductions in lazy evaluation),
+and so we perform a translation step that makes all sharing explicit
+through let-binders.
+Below we write $\llet{x{:}_{\Delta}\sigma = e,\dots,y{:}_{\Delta'}=e''
+}{e'}$ for the iterated let-binding of variables $x , \dots , y$.       
+\[
+  \begin{array}{lclr}
+    x^* & \triangleq & x\\
+
+    (\Lambda p.~e)^* & \triangleq & \Lambda p.~(e)^*\\
+    (\lambda x:_\pi\sigma .e)^* & \triangleq & \lambda x:_\pi\sigma . (e)^*\\
+    (e~\pi)^* & \triangleq & (e)^*~\pi\\
+    (e~x)^* & \triangleq & (e)^*~x\\
+    (e~e')^* & \triangleq & \llet{x{:}_{\Delta}\sigma =
+                            (e')^*}{(e)^*~x}     \\ &&\qquad \mbox{with
+                            $\Gamma;\Delta' \vdash e :
+                            \sigma\rightarrow_\pi \varphi$ and
+                                                   $\Gamma;\Delta \vdash e' : \sigma$}\\
+    K~e_1~\dots~e_n & \triangleq & \llet{x_1:_{\Delta_1} \sigma_1 =
+                                   (e_1)^*,\dots,x_n:_{\Delta_n}
+                                   \sigma_n =
+                                   (e_n)^*}{K~x_1~\dots~x_n} \\ &&
+                                                               \qquad\mbox{with
+                                                               $\Gamma;\Delta_1\vdash
+                                                               e_1 :
+                                                               \sigma_1
+                                                               \dots \Gamma;\Delta_n\vdash
+                                                               e_n :
+                                                               \sigma_n$}\\                                              
+(\ccase{e}{z{:}_{\Delta}\sigma~\{\overline{\rho\to e'}\}})^* &
+                                                               \triangleq
+                                &
+                                  \ccase{(e)^*}{z{:}_{\Delta}\sigma~\{\overline{\rho\to (e')^*}\}}\\
+    (\llet{x{:}_{\Delta}\sigma = e}{e'}      )^* & \triangleq & \llet{x{:}_{\Delta}\sigma  = (e)^*}{(e')^*}    \\
+    (\lletrec{\overline{x{:}_{\Delta}\sigma = e}}{e'}  )^* & \triangleq &\lletrec{\overline{x{:}_{\Delta}\sigma = (e)^*}}{(e')^*}  
+      \end{array}
+\]
+
+The big-step operational semantics is defined via the relation
+$\Gamma : e \Downarrow \Delta : v$ where $e$ is an expression, $v$ a
+value, $\Gamma$ and $\Delta$ are environments with bindings of the
+form $x:_\omega\sigma = e$ which assigns the expression $e$ to the
+variable $x$ of the given type.
+
+\[
+  \begin{array}{c}
+    \infer[]
+    {\,}
+    {\Gamma : \Lambda p.e \Downarrow \Gamma : \Lambda p.e}
+    \qquad
+    \infer[]
+    {\Gamma : e\Downarrow \Delta : \Lambda p.e' \quad \Delta :
+    e'[\pi/p]\Downarrow \Theta : v}
+    {\Gamma : e~\pi\Downarrow \Theta : v }\\[1.5em]
+    \infer[]
+    {\,}
+    {\Gamma : \lambda x:_\pi \sigma . e \Downarrow \Gamma : \lambda x:_\pi \sigma . e }
+    \qquad
+    \infer[]
+    {\Gamma : e \Downarrow \Delta : \lambda y:_\pi \sigma . e' \quad
+    \Delta : e'[x/y] \Downarrow \Theta : v }
+    {\Gamma : e~x \Downarrow \Theta : v}\\[1.5em]
+    \infer[]
+    {\Gamma : e \Downarrow \Delta : v}
+    {(\Gamma , x:_\omega \sigma = e) : x \Downarrow (\Delta,x:_\omega
+    \sigma = v) : v}
+    \qquad
+    \infer[]
+    {(\Gamma,x:_\omega \sigma = e) : e' \Downarrow \Theta : v}
+    {\Gamma : \llet{x{:}_{\Delta}\sigma = e}{e'}\Downarrow \Theta : v
+    }\\[1.5em]
+    \infer[]
+    {\,}
+    {\Gamma : K~x_1~\dots~x_n \Downarrow \Gamma : K~x_1~\dots~x_n}
+    \qquad
+    \infer[]
+    {\Gamma : e \Downarrow \Delta : K~x_1~\dots~x_n \quad
+    \Delta : e'[x_i/y_i][K~x_1~\dots~x_n/z] \Downarrow \Theta : v }
+    {\Gamma :
+    \ccase{e}{z{:}_{\Delta'}\sigma~\{{\dots,K~\ov{y_i}, {\dots} \to e'}\}}
+    \Downarrow \Theta : v}\\[1.5em]
+    \infer[]
+    {\Gamma : e \Downarrow \Delta : K~x_1~\dots~x_n \quad
+    \Delta : e'[K~x_1~\dots~x_n/z] \Downarrow \Theta : v }
+    {\Gamma :
+    \ccase{e}{z{:}_{\Delta'}\sigma~\{\dots , \_\to
+    e'}\} \Downarrow \Theta : v}
+    \qquad
+    \infer[]
+    {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
+    % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
+    % \Downarrow \Theta_n : v_n \quad
+    (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e' \Downarrow \Theta
+    : v
+    }
+    {\Gamma : \lletrec{\overline{x_i{:}_{\Delta}\sigma_i = e_i}}{e'}
+    \Downarrow \Theta : v}
+    \end{array}
+\]
+
+We follow an approach
+similar to that of \cite{cite:linearhaskell}, defining a linearity
+aware operational semantics in which linear variables are erased from
+the runtime environment, ensuring that any term violating linearity
+will result in a stuck state. The semantics is type-aware, defined by
+the judgment $\Gamma ; \Delta \vdash (\Theta \mid e) \Downarrow (\Theta'
+\mid v) : \sigma$ where $\Gamma$ and $\Delta$ are typing judgments
+($\Gamma$ tracks unrestricted variables, $\Delta$ tracks linear
+assumptions);
+$\Theta$ and $\Theta'$ are runtime environments, consisting of bindings of the form
+$x :_\pi \sigma = e$; $e$ is the expression being evaluated and $v$
+its resulting value, both of type $\sigma$.
+
+\[
+  \begin{array}{c}
+    \infer[]
+    {\,}
+    {\Gamma;\Delta \vdash (\Theta \mid \Lambda p.e) \Downarrow (\Theta \mid
+    \Lambda p.e) : \forall p . \sigma }
+    \\[1.5em]
+    \infer[]
+    {\Gamma ; \Delta \vdash (\Theta \mid  e) \Downarrow (\Theta' \mid
+    \Lambda p.e') : \forall p . \sigma \quad \Gamma ;\Delta \vdash
+    (\Theta' \mid e'[\pi/p]) \Downarrow (\Theta'' : v) : \sigma}
+    {\Gamma ; \Delta \vdash (\Theta \mid e~\pi) \Downarrow (\Theta''
+    \mid v) : \sigma  }\\[1.5em]
+    \infer[]
+    {\,}
+    {\Gamma ;\Delta \vdash (\Theta \mid \lambda x:_\pi \sigma . e)
+    \Downarrow (\Theta \mid \lambda x:_\pi\sigma .e) : \sigma
+    \rightarrow_\pi \tau }\\[1.5em]
+    \infer[]
+    {\Gamma;\Delta \vdash  (\Theta \mid e) \Downarrow (\Theta' \mid
+    \lambda y:_\pi \sigma . e') : \sigma \rightarrow_\pi \tau
+    \quad
+    \Gamma ; \Delta \vdash (\Theta' \mid e'[x/y]) \Downarrow (\Theta''
+    : v) : \tau }
+    {\Gamma ; \Delta\vdash (\Theta \mid  e~x ) \Downarrow (\Theta'
+    \mid v) : \tau}\\[1.5em]
+
+       \infer[]
+    {\Gamma , x:_{\Delta'} \sigma ; \Delta  \vdash (\Theta \mid e)
+    \Downarrow (\Theta' \mid v) : \sigma}
+    {\Gamma ; \Delta \vdash (\Theta , x:_{\Delta'} \sigma = e \mid x)
+    \Downarrow (\Theta' , x:_{\Delta'} \sigma = v \mid v) : \sigma}
+    \\[1.5em]
+           \infer[]
+    {\Gamma ;\Delta  \vdash (\Theta \mid e)
+    \Downarrow (\Theta' \mid v) : \sigma}
+    {\Gamma ; \Delta \vdash (\Theta , x:_1 \sigma = e \mid x)
+    \Downarrow (\Theta' \mid v) : \sigma}
+    \\[1.5em]
+
+    \infer[]
+    {\Gamma ;\Delta \vdash (\Theta,x:_{\Delta'} \sigma = e \mid e')
+    \Downarrow (\Theta' \mid v) : \tau}
+    {\Gamma ; \Delta \vdash (\Theta \mid \llet{x{:}_{\Delta}\sigma =
+    e}{e'})\Downarrow (\Theta' \mid v) : \tau
+    }\\[1.5em]
+
+    \infer[]
+    {\,}
+    {\Gamma ; \Delta \vdash (\Theta \mid K~x_1~\dots~x_n) \Downarrow
+    (\Theta \mid K~x_1~\dots~x_n) : T}
+    \\[1.5em]
+
+    \infer[]
+    {\Gamma;\Delta \vdash (\Theta \mid e) \Downarrow (\Theta' :
+    K~x_1~\dots~x_n) : T \quad
+    \Gamma ;\Delta \vdash (\Theta' \mid e'[x_i/y_i][K~x_1~\dots~x_n/z]
+    ) \Downarrow (\Theta'' : v) :\tau }
+    {\Gamma ;\Delta \vdash (\Theta \mid
+    \ccase{e}{z{:}_{\Delta'}\sigma~\{{\dots,K~\ov{y_i}, {\dots} \to e'}\}})
+    \Downarrow (\Theta'' \mid  v) : \tau}\\[1.5em]
+
+    \infer[]
+    {\Gamma ;\Delta \vdash (\Theta \mid e) \Downarrow (\Theta' \mid
+    K~x_1~\dots~x_n) : T \quad
+    \Gamma ;\Delta \vdash (\Theta' \mid  e'[K~x_1~\dots~x_n/z])
+    \Downarrow (\Theta'' \mid v) : \tau }
+    {\Gamma ; \Delta \vdash (\Theta \mid 
+    \ccase{e}{z{:}_{\Delta'}\sigma~\{\dots , \_\to
+    e'}\}) \Downarrow (\Theta'' \mid v) : \tau}
+    \\[1.5em]
+    \infer[]
+    {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
+    % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
+    % \Downarrow \Theta_n : v_n \quad
+    (\Gamma ; \Delta \vdash (\Theta, \overline{x_i:_\omega \sigma_i =
+    e_i} \mid e' )\Downarrow (\Theta'
+    \mid v) : \tau
+    }
+    {\Gamma ;\Delta \vdash (\Theta \mid \lletrec{\overline{x_i{:}_{\Delta}\sigma_i = e_i}}{e'})
+    \Downarrow (\Theta' \mid v) : \tau}
+
+   \end{array}
+\]
+
 % }}}

### 2025-01-27 icfp.tex
RM new semantics fixes
tweaked semantics

@@
 through let-binders.
-Below we write $\llet{x{:}_{\Delta}\sigma = e,\dots,y{:}_{\Delta'}=e''
-}{e'}$ for the iterated let-binding of variables $x , \dots , y$.       
+Below we write $\llet{\ov{x_i{:}_{\Delta_i}\sigma = e_i}}{e'}$ for the iterated
+let-binding of variables $\ov{x_i}$.
 \[
@@
                             $\Gamma;\Delta' \vdash e :
-                            \sigma\rightarrow_\pi \varphi$ and
+                            \sigma\rightarrow_\pi \varphi$} \\ &&\qquad \mbox{and
                                        $\Gamma;\Delta \vdash e' : \sigma$}\\
-    K~e_1~\dots~e_n & \triangleq & \llet{x_1:_{\Delta_1} \sigma_1 =
-                                   (e_1)^*,\dots,x_n:_{\Delta_n}
-                                   \sigma_n =
-                                   (e_n)^*}{K~x_1~\dots~x_n} \\ &&
+      K~\ov{e_i} & \triangleq & \llet{\ov{x_i:_{\Delta_i} \sigma_i =
+                                   (e_i)^*}}{K~\ov{x_i}} \\ &&
                                        \qquad\mbox{with
-                                                               $\Gamma;\Delta_1\vdash
-                                                               e_1 :
-                                                               \sigma_1
-                                                               \dots \Gamma;\Delta_n\vdash
-                                                               e_n :
-                                                               \sigma_n$}\\                                              
+                                       $\ov{\Gamma;\Delta_i\vdash
+                                       e_i :
+                                       \sigma_i}$}\\                                              
 (\ccase{e}{z{:}_{\Delta}\sigma~\{\overline{\rho\to e'}\}})^* &
@@
     {\,}
-    {\Gamma : K~x_1~\dots~x_n \Downarrow \Gamma : K~x_1~\dots~x_n}
+    {\Gamma : K~\ov{x_i} \Downarrow \Gamma : K~\ov{x_i}}
     \qquad
     \infer[]
-    {\Gamma : e \Downarrow \Delta : K~x_1~\dots~x_n \quad
-    \Delta : e'[x_i/y_i][K~x_1~\dots~x_n/z] \Downarrow \Theta : v }
+    {\Gamma : e \Downarrow \Delta : K~\ov{x_i} \quad
+    \Delta : e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] \Downarrow \Theta : v }
     {\Gamma :
@@
     \infer[]
-    {\Gamma : e \Downarrow \Delta : K~x_1~\dots~x_n \quad
-    \Delta : e'[K~x_1~\dots~x_n/z] \Downarrow \Theta : v }
+    {\Gamma : e \Downarrow \Delta : K~\ov{x_i} \quad
+    \Delta : e'[K~\ov{x_i}/z] \Downarrow \Theta : v }
     {\Gamma :
@@
     \Lambda p.e') : \forall p . \sigma \quad \Gamma ;\Delta \vdash
-    (\Theta' \mid e'[\pi/p]) \Downarrow (\Theta'' : v) : \sigma}
+    (\Theta' \mid e'[\pi/p]) \Downarrow (\Theta'' \mid v) : \sigma}
     {\Gamma ; \Delta \vdash (\Theta \mid e~\pi) \Downarrow (\Theta''
@@
     \quad
-    \Gamma ; \Delta \vdash (\Theta' \mid e'[x/y]) \Downarrow (\Theta''
-    : v) : \tau }
-    {\Gamma ; \Delta\vdash (\Theta \mid  e~x ) \Downarrow (\Theta'
+    \Gamma ; \Delta \vdash (\Theta' , y:_{\pi} \sigma = x \mid e') \Downarrow (\Theta''
+    \mid v) : \tau }
+    {\Gamma ; \Delta\vdash (\Theta \mid  e~x ) \Downarrow (\Theta''
     \mid v) : \tau}\\[1.5em]
@@
     \Downarrow (\Theta' \mid v) : \tau}
-    {\Gamma ; \Delta \vdash (\Theta \mid \llet{x{:}_{\Delta}\sigma =
+    {\Gamma ; \Delta \vdash (\Theta \mid \llet{x{:}_{\Delta'}\sigma =
     e}{e'})\Downarrow (\Theta' \mid v) : \tau
@@
     {\,}
-    {\Gamma ; \Delta \vdash (\Theta \mid K~x_1~\dots~x_n) \Downarrow
-    (\Theta \mid K~x_1~\dots~x_n) : T}
+    {\Gamma ; \Delta \vdash (\Theta \mid K~\ov{x_i}) \Downarrow
+    (\Theta \mid K~\ov{x_i}) : T}
     \\[1.5em]
@@
     {\Gamma;\Delta \vdash (\Theta \mid e) \Downarrow (\Theta' :
-    K~x_1~\dots~x_n) : T \quad
-    \Gamma ;\Delta \vdash (\Theta' \mid e'[x_i/y_i][K~x_1~\dots~x_n/z]
-    ) \Downarrow (\Theta'' : v) :\tau }
+    K~\ov{x_i}) : T \quad
+    \Gamma ;\Delta \vdash (\Theta' \mid e'\ov{[x_i/y_i]}[K~\ov{x_i}/z]
+    ) \Downarrow (\Theta'' \mid v) :\tau }
     {\Gamma ;\Delta \vdash (\Theta \mid
@@
     {\Gamma ;\Delta \vdash (\Theta \mid e) \Downarrow (\Theta' \mid
-    K~x_1~\dots~x_n) : T \quad
-    \Gamma ;\Delta \vdash (\Theta' \mid  e'[K~x_1~\dots~x_n/z])
+    K~\ov{x_i}) : T \quad
+    \Gamma ;\Delta \vdash (\Theta' \mid  e'[K~\ov{x_i}/z])
     \Downarrow (\Theta'' \mid v) : \tau }
@@
 
+ \begin{definition}[Well-typed State]
+   We write $\vdash \Theta \mid e : \tau$ to denote:
+
+   \[
+\vdash \llet{\Theta}{e} : \tau
+   \]
+
+ \end{definition}
+
+\begin{theorem}[Type Preservation]
+  Let $(\Theta \mid e) \Downarrow (\Theta' \mid v) : \tau$.
+  If $\vdash \Theta \mid e : \tau$ then $\vdash \Theta' \mid v : \tau$. 
+ \end{theorem}
+
 % }}}

### 2025-01-29 icfp.tex
comments on the proof
added well-typed state and env expansion

@@
     {\Gamma :
-    \ccase{e}{z{:}_{\Delta'}\sigma~\{{\dots,K~\ov{y_i}, {\dots} \to e'}\}}
+    \ccase{e}{z{:}_{\Delta'}\sigma~\{{\dots,K~\ov{y_i} \to e', {\dots}}\}}
     \Downarrow \Theta : v}\\[1.5em]
@@
 the judgment $\Gamma ; \Delta \vdash (\Theta \mid e) \Downarrow (\Theta'
-\mid v) : \sigma$ where $\Gamma$ and $\Delta$ are typing judgments
+\mid v) : \sigma , \Sigma$ where $\Gamma$ and $\Delta$ are typing judgments
 ($\Gamma$ tracks unrestricted variables, $\Delta$ tracks linear
@@
 $x :_\pi \sigma = e$; $e$ is the expression being evaluated and $v$
-its resulting value, both of type $\sigma$.
+its resulting value, both of type $\sigma$; $\Sigma$ is a list of assignments of the form
+$e :_\pi \tau$ or $x :_\Delta \tau$ which are expressions in which variables in $\Theta$ are used.
 
@@
     {\Gamma;\Delta \vdash (\Theta \mid \Lambda p.e) \Downarrow (\Theta \mid
-    \Lambda p.e) : \forall p . \sigma }
+    \Lambda p.e) : \forall p . \sigma , \Sigma }
     \\[1.5em]
@@
     {\Gamma ; \Delta \vdash (\Theta \mid  e) \Downarrow (\Theta' \mid
-    \Lambda p.e') : \forall p . \sigma \quad \Gamma ;\Delta \vdash
-    (\Theta' \mid e'[\pi/p]) \Downarrow (\Theta'' \mid v) : \sigma}
+    \Lambda p.e') : \forall p . \sigma , \Sigma \quad \Gamma ;\Delta \vdash
+    (\Theta' \mid e'[\pi/p]) \Downarrow (\Theta'' \mid v) : \sigma, \Sigma}
     {\Gamma ; \Delta \vdash (\Theta \mid e~\pi) \Downarrow (\Theta''
-    \mid v) : \sigma  }\\[1.5em]
+    \mid v) : \sigma , \Sigma  }\\[1.5em]
     \infer[]
@@
     \Downarrow (\Theta \mid \lambda x:_\pi\sigma .e) : \sigma
-    \rightarrow_\pi \tau }\\[1.5em]
+    \rightarrow_\pi \tau,\Sigma }\\[1.5em]
     \infer[]
     {\Gamma;\Delta \vdash  (\Theta \mid e) \Downarrow (\Theta' \mid
-    \lambda y:_\pi \sigma . e') : \sigma \rightarrow_\pi \tau
+    \lambda y:_\pi \sigma . e') : \sigma \rightarrow_\pi \tau , x:\sigma, \Sigma
     \quad
     \Gamma ; \Delta \vdash (\Theta' , y:_{\pi} \sigma = x \mid e') \Downarrow (\Theta''
-    \mid v) : \tau }
+    \mid v) : \tau , \Sigma  }
     {\Gamma ; \Delta\vdash (\Theta \mid  e~x ) \Downarrow (\Theta''
-    \mid v) : \tau}\\[1.5em]
+    \mid v) : \tau , \Sigma }\\[1.5em]
 
@@
     {\Gamma , x:_{\Delta'} \sigma ; \Delta  \vdash (\Theta \mid e)
-    \Downarrow (\Theta' \mid v) : \sigma}
+    \Downarrow (\Theta' \mid v) : \sigma , \Sigma}
     {\Gamma ; \Delta \vdash (\Theta , x:_{\Delta'} \sigma = e \mid x)
-    \Downarrow (\Theta' , x:_{\Delta'} \sigma = v \mid v) : \sigma}
+    \Downarrow (\Theta' , x:_{\Delta'} \sigma = v \mid v) : \sigma, \Sigma}
     \\[1.5em]
@@
     {\Gamma ;\Delta  \vdash (\Theta \mid e)
-    \Downarrow (\Theta' \mid v) : \sigma}
+    \Downarrow (\Theta' \mid v) : \sigma,\Sigma}
     {\Gamma ; \Delta \vdash (\Theta , x:_1 \sigma = e \mid x)
-    \Downarrow (\Theta' \mid v) : \sigma}
+    \Downarrow (\Theta' \mid v) : \sigma,\Sigma}
     \\[1.5em]
@@
     {\Gamma ;\Delta \vdash (\Theta,x:_{\Delta'} \sigma = e \mid e')
-    \Downarrow (\Theta' \mid v) : \tau}
+    \Downarrow (\Theta' \mid v) : \tau,\Sigma}
     {\Gamma ; \Delta \vdash (\Theta \mid \llet{x{:}_{\Delta'}\sigma =
-    e}{e'})\Downarrow (\Theta' \mid v) : \tau
+    e}{e'})\Downarrow (\Theta' \mid v) : \tau , \Sigma
     }\\[1.5em]
@@
     {\Gamma ; \Delta \vdash (\Theta \mid K~\ov{x_i}) \Downarrow
-    (\Theta \mid K~\ov{x_i}) : T}
+    (\Theta \mid K~\ov{x_i}) : T , \Sigma}
     \\[1.5em]
@@
     \infer[]
-    {\Gamma;\Delta \vdash (\Theta \mid e) \Downarrow (\Theta' :
-    K~\ov{x_i}) : T \quad
+    { K : \ov{ \sigma_i
+    \rightarrow_\omega \sigma_j \lolli} T
+    \\
+    \Gamma, \ov{x_i :_\omega \sigma_i} , z:_{\Delta'} T ;\Delta , \ov{x_j :_1 \sigma_j} \vdash (\Theta \mid e) \Downarrow (\Theta' :
+    K~\ov{x_i}) : T , e' : \tau ,\Sigma \\
     \Gamma ;\Delta \vdash (\Theta' \mid e'\ov{[x_i/y_i]}[K~\ov{x_i}/z]
-    ) \Downarrow (\Theta'' \mid v) :\tau }
+    ) \Downarrow (\Theta'' \mid v) :\tau , \Sigma }
     {\Gamma ;\Delta \vdash (\Theta \mid
-    \ccase{e}{z{:}_{\Delta'}\sigma~\{{\dots,K~\ov{y_i}, {\dots} \to e'}\}})
-    \Downarrow (\Theta'' \mid  v) : \tau}\\[1.5em]
+    \ccase{e}{z{:}_{\Delta'}T~\{{\dots,K~\ov{y_i} \to e', {\dots} }\}})
+    \Downarrow (\Theta'' \mid  v) : \tau , \Sigma}\\[1.5em]
 
     \infer[]
-    {\Gamma ;\Delta \vdash (\Theta \mid e) \Downarrow (\Theta' \mid
-    K~\ov{x_i}) : T \quad
+    {K : \ov{ \sigma_i
+    \rightarrow_\omega \sigma_j \lolli} T\\
+    \Gamma, z:_{\Delta'} T ;\Delta \vdash (\Theta \mid e) \Downarrow (\Theta' \mid
+    K~\ov{x_i}) : T , e' , \Sigma \quad
     \Gamma ;\Delta \vdash (\Theta' \mid  e'[K~\ov{x_i}/z])
-    \Downarrow (\Theta'' \mid v) : \tau }
+    \Downarrow (\Theta'' \mid v) : \tau,\Sigma }
     {\Gamma ; \Delta \vdash (\Theta \mid 
-    \ccase{e}{z{:}_{\Delta'}\sigma~\{\dots , \_\to
+    \ccase{e}{z{:}_{\Delta'}T~\{\dots , \_\to
     e'}\}) \Downarrow (\Theta'' \mid v) : \tau}
     \\[1.5em]
-    \infer[]
-    {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
-    % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
-    % \Downarrow \Theta_n : v_n \quad
-    (\Gamma ; \Delta \vdash (\Theta, \overline{x_i:_\omega \sigma_i =
-    e_i} \mid e' )\Downarrow (\Theta'
-    \mid v) : \tau
-    }
-    {\Gamma ;\Delta \vdash (\Theta \mid \lletrec{\overline{x_i{:}_{\Delta}\sigma_i = e_i}}{e'})
-    \Downarrow (\Theta' \mid v) : \tau}
+    % \infer[]
+    % {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
+    % % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
+    % % \Downarrow \Theta_n : v_n \quad
+    % (\Gamma ; \Delta \vdash (\Theta, \overline{x_i:_\omega \sigma_i =
+    % e_i} \mid e' )\Downarrow (\Theta'
+    % \mid v) : \tau
+    % }
+    % {\Gamma ;\Delta \vdash (\Theta \mid \lletrec{\overline{x_i{:}_{\Delta}\sigma_i = e_i}}{e'})
+    % \Downarrow (\Theta' \mid v) : \tau}
 
@@
 
+ We write $\Theta {\uparrow^1}$ for the linear bindings in evaluation environment $\Theta$ and $\Theta {\uparrow^\omega}$ for all other bindings.
+
+ \begin{definition}[Environment Expansion]
+   We rely on an expansion of a runtime environment $\Theta$ for the evaluation of a term $e$ into a (typed) term, written $[\Theta \mid e]$, defined inductively as follows:
+   \[
+     \begin{array}{lcl}
+       \left[ \cdot \mid e \right] & \triangleq & e\\
+       \left[ \Theta , x:_1\sigma = e' \mid e \right] & \triangleq  & (\lambda x_1:\sigma . [\Theta \mid e])~e'\\
+       \left[\Theta , x:_\Delta \sigma = e' \mid e \right] & \triangleq & \llet{x:_\Delta \sigma = e'}{[\Theta \mid e]}
+    \end{array}
+   \]
+ \end{definition}
+
  \begin{definition}[Well-typed State]
-   We write $\vdash \Theta \mid e : \tau$ to denote:
+   We write $\G;\D \vdash \Theta \mid e : \tau,\Sigma$ to denote the following typing judgment:
 
    \[
-\vdash \llet{\Theta}{e} : \tau
+\G ; \D \vdash [ \Theta \mid (e,\mathit{terms}(\Sigma))] : \tau \otimes \bigotimes(\Sigma)
    \]
-
+   where $\mathit{terms}(e_1 : \sigma_1 , \dots , e_n : \sigma_n)$ stands for the tuple $(e_1 ,( \dots , (e_n , ())))$ and
+   $\bigotimes(e_1 : \sigma_1 , \dots , e_n : \sigma_n)$ for $\sigma_1 \otimes ( \dots (\sigma_n \otimes ()))$.
  \end{definition}
 
+ \begin{lemma}
+   Let $\G ; \D , \ov{x_i :_{\Delta_i} \tau_i} ; \ov{x_j :_1 : \tau_j} \vdash  e : \sigma$, we have that
+   $\G ; \D , \ov{\Delta_i} \vdash [ \ov{x_i :_{\Delta_i} \tau_i = e_i} , \ov{x_j :_1 : \tau_j = e_j} \mid e] : \sigma$
+   \end{lemma}
+
 \begin{theorem}[Type Preservation]
-  Let $(\Theta \mid e) \Downarrow (\Theta' \mid v) : \tau$.
-  If $\vdash \Theta \mid e : \tau$ then $\vdash \Theta' \mid v : \tau$. 
+  Let $\G ; \D \vdash (\Theta \mid e) \Downarrow (\Theta' \mid v) : \tau , \Sigma$.
+  If $\G ; \D \vdash \Theta \mid e : \tau , \Sigma$ then $\G ; \D \vdash \Theta' \mid v : \tau , \Sigma$
+
 \end{theorem}
+
+% G ; D |- (Theta | e x) \Downarrow (Theta'' | v) : tau, Sigma
+
+% QUEREMOS PROVAR:
+% Se
+% G ; D |- (Theta | e x) : tau , Sigma
+% entao
+% G ; D |- (Theta'' | v) : tau, Sigma
+
+% por i.h.
+% Se G;D |- (Theta | e) : sigma -> tau, x:tau , Sigma
+% entao
+% G ; D |- (Theta' | \y:sigma.e') : sigma->tau, x:tau,Sigma
+
+% por i.h.
+% Se G ; D |- (Theta' , y:sigma = x | e') : tau, Sigma
+% entao
+% G;D |- (Theta'' | v) : tau , Sigma
+
+% Gamma ; Delta |- [Theta | (e x , terms(Sigma))] : (tau , types(Sigma))
+% Gamma ; Delta |- [Theta | (e , terms(x:tau,Sigma))] : (tau->sigma , types(x:tau,Sigma))
+
+% Daqui Gamma ; Delta |- [Theta' | (\y:_1 sigma.e' , terms(x:tau,Sigma))] : (sigma->tau, types(x:tau,Sigma))
+% tirar Gamma ; Delta |- [Theta',y:sigma = x | (e' , terms(Sigma))] : (sigma->tau, types(Sigma))
+
+% [Theta' , x_1 : tau = y | e] = (\x_1:tau . [Theta' | e] ) y
+% [ Theta' , x_Delta:tau = e' | e] = let x_Delta = e' in [Theta' | e]
+% [ . | (e,e')] = (e,e')
+
+% Gamma ; Delta |- [Theta | (e , terms(Sigma))] : (tau , types(Sigma))
+
+ % Gamma ; Delta |- Theta | e  : tau, Sigma
 

### 2025-02-19 icfp.tex
environment expansion keeps env outerwards
fix useful lemma
drop unnecessary \x from typing env in x=e|x
fix typo
bad lemma update
downarrow^1_ fix Lemma C3
Fix Case reduction auxiliary Sigma things
Tweaks and fixes to operational semantic
No, the e' in the Sigma is not substituted
There are two cases:

In the "e is in WHNF" we don't need that induction hypothesis to deduce
the result

In the "e is not in WHNF", the resources introduced in e' by the case
are "irrelevant", so we can move the \D bound variables to linear and
wipe that from the environment to reconstruct (e, e') and apply the
ind

@@
 
-\subsection{Language Syntax and Operational Semantics}
+\subsection{Language Syntax}
 
@@
 
-The (small-step) operational semantics of Linear Core are given in
-Figure~\ref{fig:linear-core-operational-semantics}. We use call-by-name
-evaluation for Linear Core as it captures the non-strict semantics in which
-our type system understands linearity, while being simpler to reason about than
-call-by-need operational semantics.
-% which is traditionally modelled with a
-% mutable heap to store \emph{thunks} and the values they are overwritten with.
-Furthermore, linear function applications, even in a \emph{call-by-need} system, can
-usually be reduced \emph{call-by-name} as the function argument is guaranteed to
-be used exactly once (thus avoiding unnecessarily allocating memory on the heap
-for a redundant \emph{thunk}).
-%
-Specifically, function applications are reduced by the standard
-\emph{call-by-name} $\beta$-reduction, substituting the argument by
-occurrences of the binder in its body, case expressions evaluate their
-scrutinee to WHNF, substituting the result by the case binder and constructor
-arguments for pattern-bound bound variables matching on that same constructor.
-
-\input{language-v4/OperationalSemantics}
+% The (small-step) operational semantics of Linear Core are given in
+% Figure~\ref{fig:linear-core-operational-semantics}. We use call-by-name
+% evaluation for Linear Core as it captures the non-strict semantics in which
+% our type system understands linearity, while being simpler to reason about than
+% call-by-need operational semantics.
+% % which is traditionally modelled with a
+% % mutable heap to store \emph{thunks} and the values they are overwritten with.
+% Furthermore, linear function applications, even in a \emph{call-by-need} system, can
+% usually be reduced \emph{call-by-name} as the function argument is guaranteed to
+% be used exactly once (thus avoiding unnecessarily allocating memory on the heap
+% for a redundant \emph{thunk}).
+% %
+% Specifically, function applications are reduced by the standard
+% \emph{call-by-name} $\beta$-reduction, substituting the argument by
+% occurrences of the binder in its body, case expressions evaluate their
+% scrutinee to WHNF, substituting the result by the case binder and constructor
+% arguments for pattern-bound bound variables matching on that same constructor.
+
+% \input{language-v4/OperationalSemantics}
 
@@
 
+We begin by presenting the operational semantics of $\lambda^\pi_\D$,
+consisting of two lazy big-step semantics in the
+style of Launchbury~\cite{10.1145/158511.158618}.
+
+\subsection{Operational Semantics}
+
+The first semantics, presented in Figure~\ref{fig:opsem}, captures
+lazy evaluation but ignores linearity information. The second
+semantics is instrumented with type information and is thus
+\emph{linearity aware}, resulting in a stuck state when a linear
+variable is used more than once. We show that for well-typed terms,
+the two semantics coincide.
+
+Both semantics are
+equipped with an environment which assigns values to variables and is
+mutated in order to express lazy evaluation. Terms are explicit about
+their sharing (in the sense of shared reductions in lazy evaluation),
+and so we perform a translation step that makes all sharing explicit
+through let-binders (see Appendix~\ref{app:opsem}).
+We write $\llet{\ov{x_i{:}_{\Delta_i}\sigma = e_i}}{e'}$ for the iterated
+let-binding of variables $\ov{x_i}$.
+
+\begin{figure}
+
+  {\small
+    \[
+  \begin{array}{c}
+    \infer[]
+    {\,}
+    {\Theta : \Lambda p.e \Downarrow \Theta : \Lambda p.e}
+    \qquad
+    \infer[]
+    {\Theta : e\Downarrow \Theta' : \Lambda p.e' \quad \Theta' :
+    e'[\pi/p]\Downarrow \Theta'' : v}
+    {\Theta : e~\pi\Downarrow \Theta'' : v }\\[1.5em]
+    \infer[]
+    {\,}
+    {\Theta : \lambda x:_\pi \sigma . e \Downarrow \Theta : \lambda x:_\pi \sigma . e }
+    \qquad
+    \infer[]
+    {\Theta : e \Downarrow \Theta' : \lambda y:_\pi \sigma . e' \quad
+    \Theta' : e'[x/y] \Downarrow \Theta'' : v }
+    {\Gamma : e~x \Downarrow \Theta'' : v}\\[1.5em]
+    \infer[]
+    {\Theta : e \Downarrow \Theta' : v}
+    {(\Theta , x:_\omega \sigma = e) : x \Downarrow (\Theta',x:_\omega
+    \sigma = v) : v}
+    \qquad
+    \infer[]
+    {(\Theta,x:_\omega \sigma = e) : e' \Downarrow \Theta' : v}
+    {\Theta : \llet{x{:}_{\Delta}\sigma = e}{e'}\Downarrow \Theta' : v
+    }\\[1.5em]
+    \infer[]
+    {\,}
+    {\Theta : K~\ov{x_i} \Downarrow \Theta : K~\ov{x_i}}
+    \qquad
+    \infer[]
+    {\Theta : e \Downarrow \Theta' : K~\ov{x_i} \quad
+    \Theta' : e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] \Downarrow \Theta'' : v }
+    {\Theta :
+    \ccase{e}{z{:}_{\Delta'}\sigma~\{{\dots,K~\ov{y_i} \to e', {\dots}}\}}
+    \Downarrow \Theta'' : v}\\[1.5em]
+    \infer[]
+    {\Theta : e \Downarrow \Theta' : K~\ov{x_i} \quad
+    \Theta' : e'[K~\ov{x_i}/z] \Downarrow \Theta'' : v }
+    {\Theta :
+    \ccase{e}{z{:}_{\Delta'}\sigma~\{\dots , \_\to
+    e'}\} \Downarrow \Theta'' : v}
+    \qquad
+    \infer[]
+    {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
+    % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
+    % \Downarrow \Theta_n : v_n \quad
+    (\Theta, \overline{x_i:_\omega \sigma_i = e_i}) : e' \Downarrow \Theta'
+    : v
+    }
+    {\Theta : \lletrec{\overline{x_i{:}_{\Delta}\sigma_i = e_i}}{e'}
+    \Downarrow \Theta' : v}
+    \end{array}
+\]}
+    \caption{Operational Semantics of $\lambda^\pi_\D$\label{fig:opsem}}
+  \end{figure}
+
+  The lazy big-step operational semantics is defined via the relation
+$\Theta : e \Downarrow \Theta' : v$ where $e$ is an expression, $v$ a
+value, $\Theta$ and $\Theta'$ are environments with bindings of the
+form $x:_\omega\sigma = e$ which assigns the expression $e$ to the
+variable $x$ of the given type.
+
+We follow an approach
+similar to that of \cite{cite:linearhaskell}, defining a
+linearity-aware operational semantics in which linear variables are
+erased from 
+the runtime environment, ensuring that any term violating linearity
+will result in a stuck state. The semantics is also type-aware (for
+the purposes of proving type safety), defined by
+the judgment $\Gamma ; \Delta \vdash (\Theta \mid e) \Downarrow (\Theta'
+\mid v) : \sigma , \Sigma$ where $\Gamma$ and $\Delta$ are typing judgments
+($\Gamma$ tracks unrestricted variables, $\Delta$ tracks linear
+assumptions);
+$\Theta$ and $\Theta'$ are runtime environments, consisting of bindings of the form
+$x :_\pi \sigma = e$; $e$ is the expression being evaluated and $v$
+its resulting value, both of type $\sigma$; $\Sigma$ is a list of assignments of the form
+$e :_\pi \tau$ or $x :_\Delta \tau$ which are expressions in which
+variables in $\Theta$ are used.
+
+\begin{figure}
+\[
+  \begin{array}{c}
+    \infer[]
+    {\,}
+    {\Gamma;\Delta \vdash (\Theta \mid \Lambda p.e) \Downarrow (\Theta \mid
+    \Lambda p.e) : \forall p . \sigma , \Sigma }
+    \\[1.5em]
+    \infer[]
+    {\Gamma ; \Delta \vdash (\Theta \mid  e) \Downarrow (\Theta' \mid
+    \Lambda p.e') : \forall p . \sigma , \Sigma \quad \Gamma ;\Delta \vdash
+      (\Theta' \mid e'[\pi/p]) \Downarrow (\Theta'' \mid v) : \sigma[\pi/p], \Sigma}
+    {\Gamma ; \Delta \vdash (\Theta \mid e~\pi) \Downarrow (\Theta''
+      \mid v) : \sigma[\pi/p] , \Sigma  }\\[1.5em]
+    \infer[]
+    {\,}
+    {\Gamma ;\Delta \vdash (\Theta \mid \lambda x:_\pi \sigma . e)
+    \Downarrow (\Theta \mid \lambda x:_\pi\sigma .e) : \sigma
+    \rightarrow_\pi \tau,\Sigma }\\[1.5em]
+    \infer[]
+    {\Gamma;\Delta \vdash  (\Theta \mid e) \Downarrow (\Theta' \mid
+    \lambda y:_\pi \sigma . e') : \sigma \rightarrow_\pi \tau , x:\sigma, \Sigma
+    \quad
+    \Gamma ; \Delta \vdash (\Theta' , y:_{\pi} \sigma = x \mid e') \Downarrow (\Theta''
+    \mid v) : \tau , \Sigma  }
+    {\Gamma ; \Delta\vdash (\Theta \mid  e~x ) \Downarrow (\Theta''
+    \mid v) : \tau , \Sigma }\\[1.5em]
+
+       \infer[]
+    {\Gamma ; \Delta  \vdash (\Theta \mid e)
+    \Downarrow (\Theta' \mid v) : \sigma , \Sigma}
+    {\Gamma ; \Delta \vdash (\Theta , x:_{\Delta'} \sigma = e \mid x)
+      \Downarrow (\Theta' , x:_{\Delta''} \sigma = v \mid v) : \sigma, \Sigma}
+    \\[1.5em]
+           \infer[]
+    {\Gamma ;\Delta  \vdash (\Theta \mid e)
+    \Downarrow (\Theta' \mid v) : \sigma,\Sigma}
+    {\Gamma ; \Delta \vdash (\Theta , x:_1 \sigma = e \mid x)
+    \Downarrow (\Theta' \mid v) : \sigma,\Sigma}
+    \\[1.5em]
+
+    \infer[]
+    {\Gamma ;\Delta \vdash (\Theta,x:_{\Delta'} \sigma = e \mid e')
+    \Downarrow (\Theta' \mid v) : \tau,\Sigma}
+    {\Gamma ; \Delta \vdash (\Theta \mid \llet{x{:}_{\Delta'}\sigma =
+    e}{e'})\Downarrow (\Theta' \mid v) : \tau , \Sigma
+    }\\[1.5em]
+
+    \infer[]
+    {\,}
+    {\Gamma ; \Delta \vdash (\Theta \mid K~\ov{x_i}) \Downarrow
+    (\Theta \mid K~\ov{x_i}) : T , \Sigma}
+    \\[1.5em]
+
+    \infer[]
+    { K : \ov{ \sigma_i
+    \rightarrow_\omega \sigma_j \lolli} T
+    \\
+    \Gamma, \ov{y_i :_\omega \sigma_i} ;\Delta , \ov{y_j :_1 \sigma_j} \vdash (\Theta \mid e) \Downarrow (\Theta' \mid
+      K~\ov{x_i}) : T , e' : \tau ,\Sigma \\
+    \Gamma ;\Delta \vdash (\Theta' \mid e'\ov{[x_i/y_{i,j}]}[K~\ov{x_i}/z]
+    ) \Downarrow (\Theta'' \mid v) :\tau , \Sigma }
+    {\Gamma ;\Delta \vdash (\Theta \mid
+    \ccase{e}{z{:}_{\Delta'}T~\{{\dots,K~\ov{y_{i,j}} \to e', {\dots} }\}})
+    \Downarrow (\Theta'' \mid  v) : \tau , \Sigma}\\[1.5em]
+
+    \infer[]
+    {K : \ov{ \sigma_i
+    \rightarrow_\omega \sigma_j \lolli} T\\
+    \Gamma ; \Delta, z{:}_1\s \vdash (\Theta \mid e) \Downarrow (\Theta' \mid
+    K~\ov{x_i}) : T , e', \Sigma \quad
+    \Gamma ;\Delta \vdash (\Theta' \mid  e'[K~\ov{x_i}/z])
+    \Downarrow (\Theta'' \mid v) : \tau,\Sigma }
+    {\Gamma ; \Delta \vdash (\Theta \mid 
+    \ccase{e}{z{:}_{\Delta'}T~\{\dots , \_\to
+    e'}\}) \Downarrow (\Theta'' \mid v) : \tau}
+    \\[1.5em]
+    % \infer[]
+    % {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
+    % % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
+    % % \Downarrow \Theta_n : v_n \quad
+    % (\Gamma ; \Delta \vdash (\Theta, \overline{x_i:_\omega \sigma_i =
+    % e_i} \mid e' )\Downarrow (\Theta'
+    % \mid v) : \tau
+    % }
+    % {\Gamma ;\Delta \vdash (\Theta \mid \lletrec{\overline{x_i{:}_{\Delta}\sigma_i = e_i}}{e'})
+    % \Downarrow (\Theta' \mid v) : \tau}
+
+   \end{array}
+ \]
+ \caption{Linearity-aware Operational Semantics~\label{fig:linopsem}}
+
+\end{figure}
+
+\subsection{Results}
+
 The $\lambda^\pi_\D$ type system is sound: well-typed programs in Linear Core do not
@@
 
-\section{Instrumented Operational Semantics}
+\section{Instrumented Operational Semantics}\label{app:opsem}
 
@@
     \Lambda p.e') : \forall p . \sigma , \Sigma \quad \Gamma ;\Delta \vdash
-    (\Theta' \mid e'[\pi/p]) \Downarrow (\Theta'' \mid v) : \sigma, \Sigma}
+      (\Theta' \mid e'[\pi/p]) \Downarrow (\Theta'' \mid v) : \sigma[\pi/p], \Sigma}
     {\Gamma ; \Delta \vdash (\Theta \mid e~\pi) \Downarrow (\Theta''
-    \mid v) : \sigma , \Sigma  }\\[1.5em]
+      \mid v) : \sigma[\pi/p] , \Sigma  }\\[1.5em]
     \infer[]
@@
        \infer[]
-    {\Gamma , x:_{\Delta'} \sigma ; \Delta  \vdash (\Theta \mid e)
+    {\Gamma ; \Delta  \vdash (\Theta \mid e)
     \Downarrow (\Theta' \mid v) : \sigma , \Sigma}
     {\Gamma ; \Delta \vdash (\Theta , x:_{\Delta'} \sigma = e \mid x)
-    \Downarrow (\Theta' , x:_{\Delta'} \sigma = v \mid v) : \sigma, \Sigma}
+      \Downarrow (\Theta' , x:_{\Delta''} \sigma = v \mid v) : \sigma, \Sigma}
     \\[1.5em]
@@
     \\
-    \Gamma, \ov{x_i :_\omega \sigma_i} , z:_{\Delta'} T ;\Delta , \ov{x_j :_1 \sigma_j} \vdash (\Theta \mid e) \Downarrow (\Theta' :
+    \Gamma, \ov{y_i :_\omega \sigma_i} ;\Delta , \ov{y_j :_1 \sigma_j} \vdash (\Theta \mid e) \Downarrow (\Theta' \mid
       K~\ov{x_i}) : T , e' : \tau ,\Sigma \\
-    \Gamma ;\Delta \vdash (\Theta' \mid e'\ov{[x_i/y_i]}[K~\ov{x_i}/z]
+    \Gamma ;\Delta \vdash (\Theta' \mid e'\ov{[x_i/y_{i,j}]}[K~\ov{x_i}/z]
     ) \Downarrow (\Theta'' \mid v) :\tau , \Sigma }
     {\Gamma ;\Delta \vdash (\Theta \mid
-    \ccase{e}{z{:}_{\Delta'}T~\{{\dots,K~\ov{y_i} \to e', {\dots} }\}})
+    \ccase{e}{z{:}_{\Delta'}T~\{{\dots,K~\ov{y_{i,j}} \to e', {\dots} }\}})
     \Downarrow (\Theta'' \mid  v) : \tau , \Sigma}\\[1.5em]
@@
     \rightarrow_\omega \sigma_j \lolli} T\\
-    \Gamma, z:_{\Delta'} T ;\Delta \vdash (\Theta \mid e) \Downarrow (\Theta' \mid
-    K~\ov{x_i}) : T , e' , \Sigma \quad
+    \Gamma ; \Delta, z{:}_1\s \vdash (\Theta \mid e) \Downarrow (\Theta' \mid
+    K~\ov{x_i}) : T , e', \Sigma \quad
     \Gamma ;\Delta \vdash (\Theta' \mid  e'[K~\ov{x_i}/z])
@@
        \left[ \cdot \mid e \right] & \triangleq & e\\
-       \left[ \Theta , x:_1\sigma = e' \mid e \right] & \triangleq  & (\lambda x_1:\sigma . [\Theta \mid e])~e'\\
-       \left[\Theta , x:_\Delta \sigma = e' \mid e \right] & \triangleq & \llet{x:_\Delta \sigma = e'}{[\Theta \mid e]}
+       \left[ \Theta , x:_1\sigma = e' \mid e \right] & \triangleq  & [\Theta \mid (\lambda x_1:\sigma . e)~e']\\
+       \left[\Theta , x:_\Delta \sigma = e' \mid e \right] & \triangleq & [\Theta \mid \llet{x:_\Delta \sigma = e'}{e}]
     \end{array}
@@
  \begin{lemma}
-   Let $\G ; \D , \ov{x_i :_{\Delta_i} \tau_i} ; \ov{x_j :_1 : \tau_j} \vdash  e : \sigma$, we have that
-   $\G ; \D , \ov{\Delta_i} \vdash [ \ov{x_i :_{\Delta_i} \tau_i = e_i} , \ov{x_j :_1 : \tau_j = e_j} \mid e] : \sigma$
+   Iff $\G ; \D \vdash [ \Theta \mid e] : \sigma$
+     then $\G , \Theta\uparrow^{\omega} ; \D , \Theta\uparrow^1 \vdash  (e, \Theta\downarrow^1_e) : \sigma \otimes \bigotimes(\Theta\downarrow^1_\sigma)$
    \end{lemma}

### 2025-02-19 icfp.tex
added Progress and partial derivations

@@
 consisting of two lazy big-step semantics in the
-style of Launchbury~\cite{10.1145/158511.158618}.
+style of Launchbury~\cite{10.1145/158511.158618}, one of which is
+instrumented with linearity information, producing stuck states when
+linearity is violated. We then show the
+type safety  of $\lambda^\pi_\D$ via the standard progress  and type
+preservation results (Section~\ref{sec:typesafe}), which in our setting entails that programs
+preserve linearity. Finally, we show that multiple optimising
+transformations such as those performed by GHC on (Linear) Core are
+type preserving
+(Section~\ref{sec:optimisations-preserve-types-meta}), and thus
+linearity preserving.
 
@@
 The first semantics, presented in Figure~\ref{fig:opsem}, captures
-lazy evaluation but ignores linearity information. The second
-semantics is instrumented with type information and is thus
+standard lazy evaluation but ignores linearity information. The second
+semantics is significantly instrumented with type information and is thus
 \emph{linearity aware}, resulting in a stuck state when a linear
-variable is used more than once. We show that for well-typed terms,
-the two semantics coincide.
+variable is used more than once. The former is a more direct model of
+an implementation of lazy evaluation, however it is not so convenient
+to reason about formally. The latter semantics is augmented with the additional
+information necessary to strengthen the induction hypothesis to make our
+type safety proofs go through.
+%
+We show that for well-typed terms, the two semantics coincide.
 
@@
 the runtime environment, ensuring that any term violating linearity
-will result in a stuck state. The semantics is also type-aware (for
-the purposes of proving type safety), defined by
+will result in a stuck state. The semantics is also type-aware and
+carries with it sufficient data to reconstruct typing derivations (for
+the purposes of proving type safety). The semantics is defined by
 the judgment $\Gamma ; \Delta \vdash (\Theta \mid e) \Downarrow (\Theta'
 \mid v) : \sigma , \Sigma$ where $\Gamma$ and $\Delta$ are typing judgments
-($\Gamma$ tracks unrestricted variables, $\Delta$ tracks linear
+($\Gamma$ tracks unrestricted and $\Delta$-bound variables, $\Delta$ tracks linear
 assumptions);
@@
 $e :_\pi \tau$ or $x :_\Delta \tau$ which are expressions in which
-variables in $\Theta$ are used.
+bindings in $\Theta$ are used. This additional data is needed to
+inductively show that the evaluation state is well-typed (and
+linearity preserving).
 
@@
 
-\subsection{Results}
+\subsection{Type Safety}\label{sec:typesafe}
 
@@
 
-\subsection{Type safety\label{sec:type-safety-meta}}
+%\subsection{Type safety\label{sec:type-safety-meta}}
 
@@
 
+ Following~\cite{Gunter1993APA,cite:linearhaskell}, we also consider
+ partial derivations of our semantics in order to be able to state a
+ progress result. Given our set of rules defining judgments of the
+ form $a \Downarrow b$ with ordered premises, we define a total
+ derivation of $a \Downarrow b$ as a tree in the standard way. A
+ partial derivation $a \Downarrow?$ is either a root labelled with $a
+ \Downarrow?$ or an application of a rule matching $a$ on the left of
+ the evaluation arrow, where exactly one of the premises $a'
+ \Downarrow?$ has a partial  derivation, all the premises to the left
+ of $a' \Downarrow?$ have a total derivation, and the premises to the
+ right of $a'\Downarrow?$ are not known yet.
+ %
+
+ By construction, in a partial derivation, there is only one premise
+ $b\Downarrow?$ with no sub-derivation. We call $b$ the head of such a
+ partial derivation.
+We write $a \Downarrow^* b$ for the relation which holds when $b$ is
+the head of some partial derivation with root $a \Downarrow?$.  We
+write $a \Downarrow^* b$  the partial evaluation relation.
+
  We write $\Theta {\uparrow^1}$ for the linear bindings in evaluation environment $\Theta$ and $\Theta {\uparrow^\omega}$ for all other bindings.
@@
 \begin{theorem}[Type Preservation]
-  Let $\G ; \D \vdash (\Theta \mid e) \Downarrow (\Theta' \mid v) : \tau , \Sigma$.
+  Let $\G ; \D \vdash (\Theta \mid e) \Downarrow (\Theta' \mid v) :
+  \tau , \Sigma$ or $\G ; \D \vdash (\Theta \mid e) \Downarrow^* (\Theta' \mid v) :
+  \tau , \Sigma$.
   If $\G ; \D \vdash \Theta \mid e : \tau , \Sigma$ then $\G ; \D \vdash \Theta' \mid v : \tau , \Sigma$
@@
 \end{theorem}
+
+\begin{theorem}[Progress]
+Let $\G ; \D \vdash \Theta \mid e : \tau,\Sigma$. Any partial derivation $\G; \D \vdash (\Theta \mid e)
+\Downarrow?,\Sigma$ can be extended. 
+\end{theorem}
+
+\begin{proof}
+ Essentially all cases are standard. The only potential issue is the
+ (linear) variable rule.
+It suffices to show that whenever a variable is evaluated it is
+present in the evaluation environment. This situation is impossible
+since the state
+$\G ; \D \vdash \Theta \mid x : \tau , \Sigma$ where $x$ is absent
+from $\Theta$ is not well-typed and so cannot be the head of a partial derivation.
+\end{proof}
 

### 2025-02-21 icfp.tex
added bisimulation between two op sems, type preservation and progress for natural semantics through bisimulation

@@
     \mid v) : \tau , \Sigma }\\[1.5em]
+           \infer[]
+    {\Gamma;\Delta \vdash  (\Theta \mid e) \Downarrow (\Theta' \mid
+    \lambda y:_\omega \sigma . e') : \sigma \rightarrow_\omega \tau , x:\sigma, \Sigma
+    \quad
+            \Gamma ; \Delta \vdash (\Theta' \mid e'[x/y]) \Downarrow (\Theta''
+    \mid v) : \tau , \Sigma  }
+    {\Gamma ; \Delta\vdash (\Theta \mid  e~x ) \Downarrow (\Theta''
+    \mid v) : \tau , \Sigma }\\[1.5em]
 
@@
     {\Gamma;\Delta \vdash  (\Theta \mid e) \Downarrow (\Theta' \mid
-    \lambda y:_\pi \sigma . e') : \sigma \rightarrow_\pi \tau , x:\sigma, \Sigma
+    \lambda y:_1 \sigma . e') : \sigma \rightarrow_1 \tau , x:\sigma, \Sigma
     \quad
-    \Gamma ; \Delta \vdash (\Theta' , y:_{\pi} \sigma = x \mid e') \Downarrow (\Theta''
+    \Gamma ; \Delta \vdash (\Theta' , y:_{1} \sigma = x \mid e') \Downarrow (\Theta''
     \mid v) : \tau , \Sigma  }
@@
     \mid v) : \tau , \Sigma }\\[1.5em]
-
+        \infer[]
+    {\Gamma;\Delta \vdash  (\Theta \mid e) \Downarrow (\Theta' \mid
+    \lambda y:_\omega \sigma . e') : \sigma \rightarrow_\omega \tau , x:\sigma, \Sigma
+    \quad
+            \Gamma ; \Delta \vdash (\Theta' \mid e'[x/y]) \Downarrow (\Theta''
+    \mid v) : \tau , \Sigma  }
+    {\Gamma ; \Delta\vdash (\Theta \mid  e~x ) \Downarrow (\Theta''
+    \mid v) : \tau , \Sigma }\\[1.5em]
        \infer[]
@@
 
-\begin{theorem}[Type Preservation]
+\begin{theorem}[Type Preservation]\label{thm:typepres}
   Let $\G ; \D \vdash (\Theta \mid e) \Downarrow (\Theta' \mid v) :
@@
 
-\begin{theorem}[Progress]
+\begin{theorem}[Progress]\label{thm:prog}
 Let $\G ; \D \vdash \Theta \mid e : \tau,\Sigma$. Any partial derivation $\G; \D \vdash (\Theta \mid e)
@@
 $\G ; \D \vdash \Theta \mid x : \tau , \Sigma$ where $x$ is absent
-from $\Theta$ is not well-typed and so cannot be the head of a partial derivation.
+from $\Theta$ is not well-typed and so cannot be the head of a partial
+derivation. 
+\end{proof}
+
+We now relate our two operational semantics through a bisimulation,
+allowing us to transfer type safety of the instrumented semantics to
+the lazy big-step semantics.
+
+\begin{definition}[Bisimulation]
+We define the relation $\gamma(\Theta : e)(\G ; \D \vdash \Theta' \mid
+e' : \tau,\Sigma)$ between a well-typed state and an ordinary state of
+the lazy big-step semantics if $e = e'$ and $\Theta' \leq \Theta\uparrow^\omega$
+\end{definition}
+
+\begin{lemma}~\label{lem:naturaltoinstr}
+  \begin{itemize}
+    \item For all $\gamma(\Theta : e)(\G ; \D \vdash \Theta' \mid
+e : \tau,\Sigma)$ such that $\Theta : e \Downarrow \Theta'' : v$
+there exists a well-typed state $\G ; \D \vdash (\Theta''' \mid v) :
+\tau,\Sigma$ such that $\G ; \D \vdash (\Theta' \mid e) \Downarrow
+(\Theta''' \mid v) : \tau,\Sigma$ and $\gamma(\Theta'' : v)(\G ; \D
+\vdash \Theta''' \mid v : \tau,\Sigma)$
+\item For all $\gamma(\Theta : e)(\G ; \D \vdash \Theta' \mid
+e : \tau,\Sigma)$ such that $\Theta : e \Downarrow^* \Theta'' : v$
+there exists a well-typed state $\G ; \D \vdash (\Theta''' \mid v) :
+\tau,\Sigma$ such that $\G ; \D \vdash (\Theta' \mid e) \Downarrow^*
+(\Theta''' \mid v) : \tau,\Sigma$ and $\gamma(\Theta'' : v)(\G ; \D
+\vdash \Theta''' \mid v : \tau,\Sigma)$
+\end{itemize}
+\end{lemma}
+\begin{proof}
+By induction on the given operational semantics rules.
+  \end{proof}
+
+\begin{lemma}~\label{lem:instrtonatural}
+  \begin{itemize}
+\item  For all $\gamma(\Theta : e)(\G ; \D \vdash \Theta' \mid
+e : \tau,\Sigma)$ such that $\G ; \D \vdash (\Theta' \mid e) \Downarrow
+(\Theta'' \mid v) : \tau,\Sigma$, there exists a state $\Theta''' : v$
+such that $\Theta : e \Downarrow \Theta''' : v$ and
+$\gamma(\Theta''' : v)(\G ; \D \vdash (\Theta'' \mid v) :
+\tau,\Sigma)$
+\item For all $\gamma(\Theta : e)(\G ; \D \vdash \Theta' \mid
+e : \tau,\Sigma)$ such that $\G ; \D \vdash (\Theta' \mid e) \Downarrow^*
+(\Theta'' \mid v) : \tau,\Sigma$, there exists a state $\Theta''' : v$
+such that $\Theta : e \Downarrow^* \Theta''' : v$ and
+$\gamma(\Theta''' : v)(\G ; \D \vdash (\Theta'' \mid v) :
+\tau,\Sigma)$
+   \end{itemize}
+ \end{lemma}
+\begin{proof}
+By induction on the given operational semantics rules.
+  \end{proof}
+
+  We now derive type safety for the lazy natural semantics.
+
+  \begin{theorem}[Type Preservation]
+For any well-typed $[\Theta \mid e]$, if $\Theta : e \Downarrow
+\Theta' : e'$ or $\Theta : e \Downarrow^* \Theta' e'$ then we have
+that $[\Theta \mid e']$ is well-typed.
+\end{theorem}
+\begin{proof}
+By Lemma~\ref{lem:naturaltoinstr} and Theorem~\ref{thm:typepres}.
+\end{proof}
+
+\begin{theorem}[Progress]
+For any partial derivation of $\Theta : e \Downarrow?$, for $[\Theta |
+e]$ well-typed, the derivation can be extended.
+\end{theorem}
+\begin{proof}
+Follows from Lemma~\ref{lem:instrtonatural} and Theorem~\ref{thm:prog}.
  \end{proof}

### 2025-02-22 icfp.tex
2.1: recursive let bindings agian

@@
 %
-we turn our focus to linearity under non-strict evaluation, both since GHC Core is
+we turn our focus to linearity under \todo{either non strict simpler sentence or say focus on lazy. decide}non-strict evaluation, both since GHC Core is
 lazy and because the distinction between semantically and syntactically
@@
 violates linearity semantically, i.e.~the program effectively
-discards or duplicates linear resources. We discuss the three main
-syntactic constructs of Core, let bindings (non-recursive and
-recursive) and case expressions.
+discards or duplicates linear resources.
+
+% We discuss recursive and recursive let bindings, and case expressions.
 
@@
 %
-Linearity is trivially violated if \lstinline{y} is inlined.
-%
-On the other hand, if the computation using the resource is suspended under a
-thunk, it will only be evaluated once, when the binding is first forced, and
-re-used subsequently. Despite the computation (here, \lstinline{use x}) only
-being evaluated once, the program must still be rejected.
-%
-Intuitively, linearity may be violated if the computation leaks
-the resource or parts of it. For instance, if \lstinline{use = id}, the program
-must clearly be rejected.
+If \lstinline{y} is inlined, linearity is trivially violated.
+%
+On the other hand, if \lstinline{y} becomes a thunk, \lstinline{use x} will
+only be evaluated once regardless of how many times \lstinline{y} is used.
+Even so, this program must be rejected under call-by-need.
+%
+Intuitively, linearity is violated if the computation leaks the resource or
+parts of it. For instance, if \lstinline{use = id}, the program must clearly be
+rejected.
 %
@@
 % of lazy let bindings. 
-In essence, an unused let binding doesn't consume any
-resources, and a let binding used exactly once consumes its resources exactly
-once. Let binders that depend on linear resources must be used \emph{at most
-once} in mutual exclusion with the linear resources themselves
-(in a sense, let-vars are affine in the let-body).
+In essence, an unused let binding doesn't consume any resources, and a let
+binding used exactly once consumes the resources it captures exactly once. Let
+binders that depend on linear resources must be used \emph{at most once} in
+mutual exclusion with the linear resources themselves (in a sense, let-vars are
+affine in the let-body).
 %
@@
 
-Mutually recursive let-bindings, i.e. bindings with definitions that
-recursively depend on the bindings they define, behave similarly to
-non-recursive lets as far as their usage in the let continuation is concerned.
+Recursive let-bindings
+%, i.e. bindings with definitions that recursively depend on the bindings they define,
+behave similarly to non-recursive lets as far as their usage in the let
+continuation is concerned.
 %recursive lets must still be used at most once, in mutual exclusion with the
@@
 binders in the recursive group is used.
-
+%
 % The same way that, in a let body, evaluating
@@
 % using the binding in its own definition also entails using that resource once.
-Consider the following program, that calls a recursive let-bound function
+Consider the following program, which calls a recursive let-bound function
 defined in terms of the linear resource \incode{x} and itself:
@@
 It so happens that \incode{go} will terminate on any input, and will always consume
-\incode{x}. However, termination is not a requirement for a binding to use \incode{x} linearly,
+\incode{x}. However, termination is not a requirement for a binding to use a resource linearly,
 and we could have a similar example in which \incode{go} might never terminate but still
@@
 
+That said, the linear type system must statically reject any recursive group that could
+lead to non-linear resource usage.
+%
 % Determining the linear resources used in a recursive binding might feel
-% peculiar since we need to know the linear resources used by the binder to determine the linear resources it uses.
+% peculiar since we need to know the linear resources used by the binder to
+% determine the linear resources it uses.
 % %
@@
 %determining how many times each thing gets used}
-
+%
 % There might not be a solution to the set of equations. In this case, the
 % binding undoubtedly entails using a linear resource more than once.
-
-Note that if we use a linear resource \incode{x} in one case alternative, and invoke
-the recursive call more than once, we might eventually consume \incode{x} more than
-once (and so \incode{f5} below is ill-typed):
+%
+For instance, if the linear resource \incode{x} is used in one case alternative and
+recursion used more than once in the other, \incode{x} may end up being consumed
+more than once:
+% (thus, \incode{f5} below is ill-typed):
 %
@@
 \end{noway}
-However, if we returned \incode{x} instead of \incode{go bool} in the let body, then, despite
-the binding using \incode{x} more than once, we would consume \incode{x} exactly once,
-since recursive bindings are still lazy.
-%
-These ideas extend to mutually recursive bindings that depend on
-linear resources, as we discuss in Section~\ref{sec:recursivelets}.
+%However, if we returned \incode{x} instead of \incode{go bool} in the let body, then, despite
+%the binding using \incode{x} more than once, we would consume \incode{x} exactly once,
+%since recursive bindings are still lazy.
+%
+
+%Intuitively, to statically determine whether a group of recursive 
+%
+%Informally, we \emph{assume} the recursive binding will consume some linear resources
+%exactly once, and use that assumption when reasoning about recursive calls such
+%that those linear resources are used exactly once.
+
+To determine whether the usage of a binding from a recursive group is linearly
+valid, we must infer the superset of resources that may end up being consumed
+in a single use of any of the bindings. For a strongly connected recursive
+group, this superset is the same for all bindings by definition.
+%
+Intuitively, this inference amounts to finding a set of linear resources
+($\Delta$) that satisfies the set of equations derived from the bindings
+$\ov{x_i}$ in the group such that
+%\begin{enumerate}
+occurrences of $x_i$ in its own body are synonymous with using all resources in $\Delta$ exactly once.
+%(2) and if the binding $x$ is fully evaluated, then all resources in $\Delta$ are consumed exactly once.
+%\end{enumerate}
+%Finding a solution to this equation is akin to finding a (principle) type for a
+%recursive binding such that occurrences of itself in its definition are also well-typed.
+%
+It follows from the solution being the same for all bindings that
+at most a single variable from the recursive group can be used in the continuation.
+%
+Recursive let bindings in our system are discussed in Section~\ref{sec:recursivelets}.
 

### 2025-02-22 icfp.tex
2.1: finish pass + acknowledgements

@@
 \paragraph{Case expressions}
-Finally, we discuss case expressions,
-which are the key ingredient that brings together the
-semantic linearity insights developed thus far given that
-\emph{case expressions drive evaluation}, and semantic linearity can only be
-understood in terms of how expressions are evaluated.
-
-Up until now, there
-have been no examples in which linear resources are \emph{fully consumed} in
-the bodies of linear functions. In other words, all example functions so far
-return a value that has to be itself consumed exactly once to ensure the linear
-argument is, in turn, consumed exactly once -- as opposed to functions whose
-application simply needs to be evaluated to guarantee its linear argument is
-consumed (e.g.,~functions that return an unrestricted value).
+Above, we have observed more programs can be accepted as linear by delaying
+resource consumption in (recursive) let bindings until they are evaluated.
+On the other hand, \emph{case expressions drive evaluation} -- and, therefore,
+do consume resources.
+
+%and semantic linearity can only be understood in terms of how expressions are
+%evaluated.
+
+%Up until now, there have been no examples in which linear resources are
+%\emph{fully consumed} in the bodies of linear functions.
+
+%%%%%%%%%%%%%%% KEEEP ????? %%%%%%%
+%%%%%%%%%All example functions so far returned a value that has to be itself consumed
+%%%%%%%%%exactly once to ensure the linear argument is, in turn, consumed exactly once
+%%%%%%%%%-- as opposed to functions whose application simply needs to be evaluated to
+%%%%%%%%%guarantee its linear argument is consumed (e.g.,~functions that return an
+%%%%%%%%%unrestricted value).
+
 %
@@
 %
-Case expressions enable us to consume resources and so write
-functions that fully consume their linear arguments.
+%Case expressions enable us to consume resources and so write
+%functions that fully consume their linear arguments.
 % CORTEI
@@
 %\end{itemize}
-Thus, we can consume a linear resource by fully evaluating it, through case
-expressions.
+Thus, we can consume a linear resource by fully evaluating it -- and evaluation
+happens in case expressions.
 
@@
 
-In Core, case expressions are of the form $\ccase{e_s}{z~\{\ov{\rho_i \to e_i}\}}$,
+In Core, cases are of the form $\ccase{e_s}{z~\{\ov{\rho_i \to e_i}\}}$,
 where $e_s$ is the case \emph{scrutinee}, $z$ is the case \emph{binder}, and
@@
 of the tuple). However, if we did use $a$ in the case body, then $x$
-would be unavailable (and so \incode{f8} below cannot be accepted):
+would be unavailable (as opposed to \incode{y}) and \incode{f8} below must be rejected:
 \begin{noway}
@@
 resource $y$, and the case binder $z$ entails using both $a$ and $b$.
-%
+
 Dually, consider the scrutinee to be an expression that's not in WHNF, s.t.
@@
 
-We now focus on pattern matching on constructors without
-any linear components. If the constructor has no linear fields, it means the
-result can be consumed unrestrictedly and, therefore, all linear resources used
-in the computation have been fully consumed.
-%
-Consequently, in a branch of a constructor without linear fields we know the
-result of evaluating the scrutinee to be unrestricted, so we can use the case
-binder unrestrictedly and refer to it zero or more times. For example,
-the program in Figure~\ref{fig:nolin} is semantically linear.
-%
-\begin{figure}[t]
+On the other hand, if we match on a constructor without any linear components,
+all linear resources used in the scrutinee must have been fully consumed in
+that branch since the result of evaluating it can be consumed unrestrictedly
+(by definition, a constructor application to unrestricted arguments is also
+unrestricted).
+%
+Hence, in the branch for a constructor without linear fields, the case
+binder can also be used unrestrictedly. For example, the program in
+Figure~\ref{fig:nolin} is semantically linear.
+%
+\begin{figure}[h]
  \begin{minipage}{0.5\textwidth}
@@
 %
-A second example of an unrestricted pattern, where \incode{K2} has no fields and \incode{K1}
-has one linear field, is shown in Figure~\ref{fig:absurd}. Seemingly, the resource
-$x$ must have been fully consumed to take the \incode{K2} branch, but because the
-scrutinee is known to be \incode{K1}, the \incode{K2} branch is \emph{absurd}, so, in reality
-any resource could be freely used in that branch, and the example is
-\emph{semantically} linear (despite not being seen as so by our system).
-
-Many of these examples might seem unrealistic due to the pattern
-match on a known constructor, but we note that, during the
-transformations programs undergo in an optimising compiler, many
-such programs naturally occur (e.g.,~if the definition of a
-function is inlined in the scrutinee).
+A second example of an unrestricted pattern, where \incode{K2} has no fields
+and \incode{K1} has one linear field, is shown in Figure~\ref{fig:absurd}.
+The use of \incode{x} in the \incode{K2} branch is valid despite \incode{x}
+also occurring in the scrutinee because this branch is never taken. In fact,
+any arbitrary resource could be used in absurd branches and be understood as
+linear (despite not being seen as such by our system).
+
+Many of these examples might seem unrealistic due to the pattern match on a
+known constructor, but we note that in between the transformations programs
+undergo in an optimising compiler, many such programs naturally occur (e.g.,~if
+the definition of a function is inlined in the scrutinee).
 %
@@
 (written $\_$): matching a wildcard doesn't provide new linearity information,
-unlike matching against a pattern with or without linear components, however,
-the scrutinee will still be evaluated to WHNF.
+unlike a pattern with linear or unrestricted components, however, the
+scrutinee is still evaluated to WHNF.
 %
@@
 %
-Briefly, when matching a wildcard, if the scrutinee is already in WHNF, we can
-either use the resources from the scrutinee directly, or the case binder in
-that alternative. If the scrutinee is not in WHNF, we \emph{must} use the case
-binder, as it's the only way to linearly consume the result of evaluating the
-scrutinee.
-
-Finally, we discuss case expressions which scrutinize a
+Briefly, when matching a wildcard, if the scrutinee is already in WHNF,
+programs that either use the resources from the scrutinee directly, or the case
+binder in that alternative, can be seen as linear. If the scrutinee is not in
+WHNF, we \emph{must} use the case binder, as it's the only way to linearly
+consume the result of evaluating the scrutinee.
+
+Finally, we discuss the special case of scrutinizing a
 single variable $x$, e.g.:
@@
 further analysis on the non-strict evaluation strategy. If linear
-function applications, in general, are $\beta$-reduced \emph{call-by-name} (a desirable
-practice, as linear functions are known to use their argument exactly once)
+function applications are $\beta$-reduced \emph{call-by-name}
+(often desirable, as linear functions are known to use their argument exactly
+once)
 % and thus call-by-name reduction will definitely not duplicate computation
-while the above function is considered linear, then such an application might
+while the example function is considered linear, then such an application might
 duplicate linear resources during evaluation. For instance:
@@
 %
-The system we present in the next section, which evaluates linear
-applications call-by-name, does not accept the above function scrutinizing a
-variable as linear. 
-This interaction between evaluation and linearity arises in the type
-preservation proof of our system, and is again explored with the reverse
-binder swap transformation discussed in
-Section~\ref{sec:optimisations-preserve-types-meta}.
+The system we present in the next section does accept this function
+scrutinizing a variable as linear, as our operational semantics are
+call-by-need.
+This interaction between evaluation and linearity is considered specifically in the
+reverse binder swap transformation discussed in
+Section~\ref{sec:optimisations-preserve-types-meta}.\todo{Figure out precisely reverse binder swap proof and reevaluate}
 
@@
 
+\section*{Acknowledgements}
+
+We thank Arnaud Spiwack for insightful discussions and Well-Typed for funding
+this work.\todo{Do these have to be anonymized?}
+
 % \begin{itemize}

### 2025-02-23 icfp.tex
Pass over till end of 3.1

@@
 Haskell~\cite{cite:linearhaskell} with all the key features from GHC's Core
-language (algebraic datatypes, case expressions, and recursive lazy let bindings)
+language (algebraic datatypes, case expressions, and recursive lazy let bindings),
 except for equality coercions, which we discuss as future work in
@@
 %
-%
+
 Linear Core makes precise the various insights discussed in the
-previous section by crystallizing them together in a linear type system for which we
-prove soundness via the standard type preservation and progress theorems. Crucially,
-Linear Core typing accepts all the \emph{semantically linear} examples
+previous section in a linear type system that we prove is sound and guarantees
+linear resource usage at runtime via a standard soundness argument on a
+call-by-need big-step operational semantics that becomes stuck when linear
+variables in the runtime heap are used more than once.
+\todo{O que digo aqui sobre a semantica launchbury style?}
+Crucially,
+Linear Core typing accepts all the semantically linear examples
 (highlighted with \colorbox{notyet}{\notyetcolorname})
@@
 %
-Besides type safety, we prove that multiple optimising Core-to-Core
-transformations preserve linearity in Linear Core, while violating linearity
-under Core's current type system.
-%
-% As far as we know, we are the first to prove optimisations preserve types in
-% a non-strict linear language.
-
-The first key idea for typing linearity semantically is to delay \emph{consuming a
-resource} to when a computation that depends on that resource is effectively
-evaluated or returned, by annotating relevant binders with \emph{usage
-environments} (Section~\ref{sec:usage-environments}).
+Additionally, we prove that multiple optimising Core-to-Core transformations
+preserve linearity in Linear Core, while violating it under Core's
+current type system.
+%
+%To the best of our knowledge, we are the first to prove optimisations preserve
+%types in a non-strict linear language.
+% 
+Despite the focus on GHC Core, the fundamental ideas for
+understanding linearity in our call-by-need calculus can be readily applied to
+other non-strict languages.
+
+Linear Core's first key idea is to delay consuming
+resources until a computation that depends on those resources is effectively
+evaluated or returned. Delayed consumption is tracked by annotating relevant
+binders with \emph{usage environments} (Section~\ref{sec:usage-environments}).
 %
@@
 expressions, branching on whether the scrutinee is in Weak Head Normal
-Form, and using ``proof irrelevance'' to track resources that are no
+Form or not, and using ``proof irrelevance'' to track resources that are no
 longer available but haven't yet been fully consumed
 (Section~\ref{sec:lc-case-exps}). Additionally, we introduce tags to split
-resources between the pattern-bound variables, %^as usage environments
-encoding that all linear components of a pattern must be used.
-% 
-We also note that despite the focus on GHC Core, the fundamental ideas for
-understanding linearity in a call-by-need calculus can be readily applied to
-other non-strict languages.
+resources amongst pattern-bound variables and %^as usage environments
+encode that all linear components of a pattern must be used together.
 
@@
 %
-Finally, we introduce the rules to typecheck case expressions and alternatives,
+Finally, we introduce the rules to type case expressions and case alternatives,
 along with the key insights to do so, namely branching on WHNF-ness of
@@
 $\lletrec{\ov{\xD = e}}{e'}$, where the overline denotes a set of distinct
-bindings $x_1{:}_{\D_1}\s_1\dots x_n{:}_{\D_n}\s_n$ and associated expressions
+bindings $x_1{:}_{\D}\s_1\dots x_n{:}_{\D}\s_n$ and associated expressions
 $e_1\dots e_n$, and case expressions $\ccase{e}{\zD~\{\ov{\rho \to e'}\}}$,
@@
 with a so-called \emph{usage environment} $\Delta$ -- a fundamental construct
-for type-checking semantic linearity in the presence of laziness we present in
+for typing semantic linearity in the presence of laziness that we present in
 Section~\ref{sec:usage-environments}.
 %
-Case patterns $\rho$ can be either the \emph{default} or \emph{wildcard}
+Case patterns $\rho$ can be either the \emph{default/wildcard}
 pattern $\_$, which matches any expression, or a constructor $K$ and a set of
 variables that bind its arguments, where each field of the constructor has an
-associated multiplicity denoting whether the pattern-bound variables must
-consumed linearly (ultimately, in order to consume the scrutinee linearly).
+associated multiplicity denoting which pattern-bound variables must be
+consumed linearly or unrestrictedly.
 Additionally, the set of patterns in a case expression is assumed to be exhaustive,

### 2025-02-23 icfp.tex
done 3.2

@@
 %
-The core rules of the calculus for, multiplicity and term,
-abstraction and application are similar to that of $\lambda^q_\to$ and
-mostly standard. 
+The core rules of the calculus for abstraction and application are similar to
+that of $\lambda^q_\to$ and mostly standard. 
 We note that we handle multiplicity polymorphism differently from
@@
 
-We start with the main typing judgement.
-As is common for linear type
-systems, we use two typing environments, an \emph{unrestricted} $\G$ and
-a \emph{linear} $\D$ environment.
-Thus, the main judgement is written $\G;\D \vdash e :
-\s$,
-denoting that expression $e$ has type $\s$ under
-the unrestricted environment $\G$ and linear environment $\D$.
-
-%
-Variables in $\G$ can be freely discarded (\emph{weakened}) and
-duplicated (\emph{contracted}), while resources in $\D$ must be used
-exactly once (hence can't be discarded nor duplicated). Despite not
-having explicit weakening and contraction rules in our system, they
-are available as admissible rules for $\G$ (but not for $\D$), since,
-equivalently \cite{91621fae-5e53-3497-8291-32b2fab5a743}, resources
-from $\G$ are duplicated for sub-derivations and may unrestrictedly
-exist in the variable rules.
-% TODO: Likely rewrite that
+%We start with the main typing judgement.
+%As is common for linear type systems, we use two typing environments, an
+%unrestricted $\G$ and a linear $\D$ environment.
+The main judgement is written $\G;\D \vdash e : \s$ to denote that expression
+$e$ has type $\s$ under the unrestricted environment $\G$ and linear
+environment $\D$.
+%
+%
+Variables in $\G$ can be freely discarded (\emph{weakened}) and duplicated
+(\emph{contracted}), while resources in $\D$ must be used exactly once. Despite
+not having explicit weakening and contraction rules in our system, they are
+available as admissible rules for $\G$ (but not for $\D$), since, equivalently
+\cite{91621fae-5e53-3497-8291-32b2fab5a743}, resources from $\G$ are duplicated
+for sub-derivations and may unrestrictedly exist in the variable rules.
+% TODO: Likely rewrite that?
 %
@@
 linear function application splitting the linear context in two
-disjoint parts, one used to type the argument and the other for the
+disjoint parts, one used to type the argument and the other the
 function.
@@
 
-Multiplicity abstractions ($\Lambda p.~e$) introduce a multiplicity variable
-$p$ (in the unrestricted context), and construct expressions of type
-$\forall p.~\s$, i.e. a type universally quantified over a multiplicity
-variable $p$. Again, conservatively, in the body of the abstraction, function
-types annotated with a $p$ variable and datatype fields with multiplicity $p$
-are typed as thought they are linear functions and linear fields, because $p$
-can be instantiated at both $\omega$ and $1$.
-\[
-\TypeMultLamIntro \quad \TypeMultLamElim
-\]
-A multiplicity application instantiates a multiplicity-polymorphic type
+Multiplicity abstractions ($\Lambda I$) introduce a multiplicity variable $p$,
+and construct expressions of type $\forall p.~\s$, i.e. a type universally
+quantified over a multiplicity variable $p$. Again, conservatively, in the body
+of the abstraction, function types annotated with a $p$ variable and datatype
+fields with multiplicity $p$ are typed as thought they are linear functions and
+linear fields, because $p$ can be instantiated at both $\omega$ and $1$.
+%\[
+%%\TypeMultLamIntro \quad \TypeMultLamElim
+%\]
+A multiplicity application ($\Lambda E$) instantiates a multiplicity-polymorphic type
 $\forall p.~\s$ at a particular (argument) multiplicity $\pi$, resulting in an

### 2025-02-23 icfp.tex
3.3 done

@@
 %
-$\Delta$-bound variables are introduced by several constructs as (recursive)
-let binders, case binders, and case pattern variables.
+$\Delta$-bound variables are introduced in the unrestricted environment by
+several constructs such as (recursive) let binders, case binders, and case
+pattern variables.
 %
 For example, the let-binder $u$ in the following program is annotated with a
-usage environment $\{x{:}_1\s,y{:}_1\s\}$, the resources that are used in its body. Thus, in
-the expression $e$, consuming $u$ will equate to consuming the linear resources
-$x$ and $y$:
+usage environment $\{x{:}_1\s,y{:}_1\s\}$, the resources that are used in its
+body. Thus, in the expression $e$, consuming $u$ will equate to consuming the
+linear resources $x$ and $y$:
 % , as per the insights developed in
@@
 Dually, using the linear resources directly means they are no longer available
-to consume through the usage environment of the $\Delta$-bound variable.
+to consume through the usage environment of the $\Delta$-bound variable. The
+$\Delta$-bound variable can be left unused since it is unrestricted, regardless
+of the variables it is annotated with.
 
@@
 %
-$\D$-variables are unrestricted 
-since they can be discarded and duplicated, despite multiple
-occurrences of the same $\Delta$-variable not being
-well-typed as, ultimately, it would imply non-linear usage of linear
-resources.
+$\D$-vars are unrestricted and can be discarded and duplicated, despite
+multiple occurrences of the same $\Delta$-var in an expression not possibly being well-typed
+since that would imply non-linear usage of linear resources.
 % multiple occurrences of the same

### 2025-02-23 icfp.tex
3.4 done

@@
 In Section~\ref{sec:semantic-linearity-examples}, we discussed how linear
-resources used in let-bound expressions are only consumed when the same let-bound
-expressions are fully evaluated, i.e. linear resources required by let-bound
-expressions are consumed lazily.
-%
-Moreover, resources from a let-bound expression cannot be used \emph{together}
-with the variable binding, since said resources would end up being consumed
-more than once, violating (semantic) linearity -- the binder has to be used in
-mutual exclusion with the linear resources required to type the expression it
-binds, and either \emph{must} be used.
+resources used in let-bound expressions are only consumed when the respective
+binders are evaluated in the let-continuation.
+%, i.e. linear resources required by let-bound expressions are consumed lazily.
+%
+Moreover, resources captured by a let-bound expression cannot be used
+simultaneously with the let-binder in the continuation, since said
+resources would end up being consumed more than once and violate (semantic)
+linearity. The binder has to be used in mutual exclusion with the linear
+resources captured by the expression it binds, and either the binding or the
+resources \emph{must} be used.
 
@@
 %
-Annotating let-bound variables with a usage environment $\D$
-\emph{delays} the consumption of resources to when the variables themselves are
-used:
+Annotating let-bound variables with a usage environment $\D$ delays the
+consumption of resources to when the variables themselves are used and
+guarantees the desired mutual exclusivity property between the binder and the resources:
 \[
@@
 $\D$-variable ensures that using it will consume the resources $\D$ the
-expression it binds depends on.
+expression captures, \emph{if} it is used.
 % \todo[inline]{Let bindings are hard, if they are used then we use resources. If

### 2025-02-23 icfp.tex
3.5 and start 3.6

@@
 %
-Recursive let bindings also bind expressions \emph{lazily}, so, similarly, we
+Recursive let bindings also bind expressions lazily, so, similarly, we
 introduce a $\D$-variable for each binding. Resources required to type the
@@
 %
-Additionally, as seen in Section
-\ref{sec:semantic-linearity-examples:recursive-lets}, recursive uses of a
-binder in its own definition entail consuming all resources otherwise required
-to type the binder's body.
-
-As for groups of recursive let bindings, we recall that, in our
-system, all recursive let groups are strongly-connected -- all bindings in a
-recursive let group are mutually recursive in the sense that they all
-(transitively) depend on one another.
+%Additionally, as seen in Section
+%\ref{sec:semantic-linearity-examples:recursive-lets}, recursive uses of a
+%binder in its own definition entail consuming all resources otherwise required
+%to type the binder's body.
+
+%We assume all mutually recursive groups of bindings are strongly-connected.
+%As for groups of recursive let bindings, we recall that, in our
+%system, all recursive let groups are strongly-connected -- all bindings in a
+%recursive let group are mutually recursive in the sense that they all
+%(transitively) depend on one another.
 %
@@
 %
-By definition of strongly-connected, any binder in the mutually recursive group
-may end up using any binder from the group. Accordingly, occurrences of any
-such binder must be typed conservatively, using up the sum of resources needed
-to type each of the binders,
-%
-i.e., all binders in a strongly-connected group of mutually recursive let
-bindings have to be typed with the same linear resources (which are the least
-upper bound of resources needed to type the bodies of all binders, accounting
-for uses of mutually-recursive binders).
+%By definition, any binder in such a group may end up recursively using
+%any other binder from the group and thus use up any resource occurring.
+%
+%Accordingly, occurrences of any such binder
+%must be typed conservatively, using up the sum of resources needed to type each
+%of the binders,
+%%
+%i.e.,
+%Accordingly, all the binders in a strongly-connected group of mutually recursive let
+%bindings have to be typed with the same linear resources (which are the least
+%upper bound of resources needed to type the bodies of all binders, accounting
+%for uses of mutually-recursive binders).
 %
@@
 The typing rule for recursive groups of bindings leverages both the assumption
-that all recursive let bindings are strongly connected and the observation that
-every binder in a strongly connected group of recursive bindings is typed with
-the same linear context. Consequently, all bindings of the recursive group are
-introduced as $\D$-vars with the same $\D$ environment -- using any one of the
-bindings in a recursive group entails consuming all resources required to type
-that same group, which is also why we can use the same linear resources to type
-each binder:
+that all mutually recursive groups are strongly connected and the corollary
+that every binder in such a group must be typed with the same linear context,
+as observed in \ref{sec:semantic-linearity-examples:recursive-lets}.
+Consequently, all bindings of the recursive group are introduced as $\D$-vars
+which all share the same $\D$ environment -- using any one of the bindings in a
+recursive group entails consuming all resources required to type that same
+group -- so, at most a single binder from the group can be used in the let-continuation.
+%\todo{should we say how this means at most once of the binders from the group can be used?} OK
+%, which is also why we can use the same linear resources to type each binder:
 \[
@@
 
-Case expressions \emph{drive evaluation} --
+%Case expressions \emph{drive evaluation} --
 % thus they are key to realize a type system that understands linearity in the presence of lazy evaluation.
@@
 %
-a case expression \emph{evaluates its scrutinee} to Weak Head Normal
-Form (WHNF)~\cite{10.5555/1096899}, and \emph{then} selects the case
-alternative corresponding to the pattern matching the WHNF of the
-scrutinee\footnote{In our calculus, the alternatives are assumed to be
+A case expressions drive evaluation by evaluating their scrutinee to Weak Head
+Normal Form (WHNF)~\cite{10.5555/1096899}. Then, the case
+alternative whose pattern matches the result of evaluating the
+scrutinee is taken\footnote{In our calculus, the alternatives are assumed to be
   exhaustive, i.e. there always exists at least one pattern which
   matches the scrutinee in its WHNF, so we're guaranteed to have an
-  expression to progress evaluation.}. An expression in WHNF can either be %\begin{itemize}
+  expression to progress evaluation.}.
+An expression in WHNF can either be %\begin{itemize}
 % \item
-a $\lambda$-abstraction $\lambda x.~e$ 
-or a datatype constructor application $K~\ov{e}$.
+a $\lambda$-abstraction $(\lambda x.~e)$ 
+or a datatype constructor application $(K~\ov{e})$.
 %\end{itemize}
@@
 Accordingly, these sub-expressions might still depend on linear resources to be
-well-typed (these resources will be consumed when the expressions are evaluated).
+well-typed (such resources will only be consumed when the sub-expressions themselves are evaluated).
 

### 2025-02-23 icfp.tex
Up to 3.6.2

@@
 %
-A case expressions drive evaluation by evaluating their scrutinee to Weak Head
+A case expression drives evaluation by evaluating its scrutinee to Weak Head
 Normal Form (WHNF)~\cite{10.5555/1096899}. Then, the case
@@
 Therefore, resources used to type a scrutinee not in
-WHNF will be consumed when the case is evaluated, making said resources unavailable in the case
+WHNF will be consumed when the case is evaluated and no longer be directly available in the case
 alternatives. Conversely, when the scrutinee is already in WHNF, linear
-resources required to type the scrutinee are still available in the alternatives.
-The linear resources used by an expression in WHNF are exactly those which
-occur to the right of $\gtrdot$ in the WHNF judgement shown above
-(corresponding to the resources required to typecheck the lambda body or the
-constructor arguments).
+resources required to type the scrutinee are still directly available in the
+alternatives.
+%
+%%%%???
+%%%%The linear resources used to type an expression in WHNF are exactly those which
+%%%%occur to the right of $\gtrdot$ in the WHNF judgement shown above
+%%%%(corresponding to the resources required to typecheck the lambda body or the
+%%%%constructor arguments).
 
@@
 
-The dichotomy between evaluation (i.e.,~resource usage) of a case expression
-whose scrutinee is in weak head normal form, or otherwise, leads to one of our
-key insights: we must \emph{branch} on weak head normal formed-ness to
-accurately type case expressions.
+The dichotomy between evaluation (and resource usage) of a case expression
+whose scrutinee is in weak head normal form and one whose scrutinee is not
+leads to one of the key insights of $\lambda^\pi_\Delta$: we must \emph{branch}
+on weak head normal formed-ness to accurately type case expressions.
 %
@@
 \begin{array}{ccc}
-(1)~\lambda x.~\ccase{K~x}{\_ \to x} &  & (2)~\lambda x.~\ccase{free~x}{\_ \to x}
+    (1)~\lambda x.~\ccase{K~x}{z~\{\_ \to x\}} &  & (2)~\lambda x.~\ccase{free~x}{z~\{\_ \to x\}}
 \end{array}
 \]
-The first function uses $x$ linearly, while the second does not.
+The first function uses $x$ linearly, but the second does not.
 %
@@
 
-There are three \emph{competing} ways
+There are three competing ways
 to use the resources from a scrutinee in WHNF in a case alternative: directly, via
-the case binder, or by using \emph{all} the pattern-bound variables.
+the case binder, or by using pattern-bound variables.
 %
@@
 ways of consuming resources -- it follows that case binders and pattern-bound
-variables are another instance of $\D$-bound variables. Intuitively, resources
-in a scrutinee that is already in WHNF are only properly consumed when all
-(linear) fields of the pattern are used, satisfying the definition of
-consuming resources given in Linear Haskell.
+variables are yet another instance of $\D$-bound variables.
+%
+%Resources in a scrutinee that is already in WHNF are only consumed when all
+%(linear) fields of the pattern are used, satisfying the definition of
+%consuming resources given in Linear Haskell.
 %
 % We type a case expression whose scrutinee is in weak head normal form with
-This suggests the following first attempt at a rule for cases of
-scrutinees in WHNF:
+This suggests the following first rule for cases of scrutinees already in WHNF:
 \[
@@
 \]
+
 First, we assert this rule is only applicable to expressions in weak head
@@
 $\vdash_{alt}$ judgement, introducing
-the case binder $z$ in the environment as a $\D$-bound
-variable whose usage environment is the linear resources used to type the
-scrutinee. We also 
-make all the resources $\ov{\D_i}$ used to type the scrutinee
-available in the linear typing environment.
-%
-The $\vdash_{alt}$ judgement is 
-annotated with the disjoint set of linear resources $\ov{\D_i}$ used
-to typecheck the scrutinee sub-expressions,
-and the name of the case binder $z$.
-
-%
-Despite the key intuitions of typing a case expressions whose scrutinee is in
-WHNF being conveyed by this rule, our treatment of such a case expression is
-slightly more involved, as will be presented after discussing cases of
-scrutinees not in WHNF.
+the case binder $z$ in the unrestricted environment as a $\D$-bound
+variable whose usage environment is the linear resources $\ov{\D_i}$ used to type the
+scrutinee. Those same resources $\ov{\D_i}$ are again made available in the
+linear typing environment of the alternative, similarly to how the resources
+used to type a $Let$ binder are still available in the continuation.
+%
+%The $\vdash_{alt}$ judgement is 
+%annotated with the disjoint set of linear resources $\ov{\D_i}$ used
+%to type the scrutinee sub-expressions,
+%and the name of the case binder $z$.
+
+%
+Despite the main idea for typing ``WHNF case expressions'' being conveyed by
+this rule, in our system it is slightly more involved, as will be seen after
+discussing cases of scrutinees not in WHNF.
 
@@
 \vp$ is used to type case alternatives, encompassing three
-``modes'', distinguished by the kind of arrow that is used:
-for alternatives of case expressions whose scrutinee is in WHNF ($\Rightarrow_\mathsf{WHNF}$),
-for case expressions in which the scrutinee is not in WHNF ($\Rightarrow_\mathsf{NWHNF}$),
+``modes'' distinguished by the kind of arrow that is used:
+for alternatives of case expressions whose scrutinee is in WHNF
+($\Rightarrow_\mathsf{WHNF}$), not in WHNF ($\Rightarrow_\mathsf{NWHNF}$),
 and for alternatives agnostic to the WHNF-ness of the scrutinee
@@
 % judgement, t
-The rule for typing an alternative whose pattern is a
-constructor with $n > 0$ linear components (in mode $\Rightarrow_{\mathsf{NWHNF}}$) is:
+
+First, an alternative whose pattern is a constructor with $n > 0$ linear components
+under the $\Rightarrow_{\mathsf{NWHNF}}$ mode is typed as:
 \[
@@
 \]
-The rule states that, for such a pattern matching a scrutinee already in
+The rule states that, for such a match on a scrutinee already in
 WHNF, we introduce the linear components of the pattern as $\D$-bound variables
-whose usage environment matches the linear resources required to type the
-corresponding constructor argument in the scrutinee, which is annotated in
-the judgement ($\ov{\D_i}$). Unrestricted fields of the constructor are
-introduced as unrestricted variables. We note that the typing environment $\D$
-always contains the resources $\ov{\D_i}$ in uses of the alternative
-judgement.
-
-The rule for alternatives that match on the wildcard pattern is:
+with usage environment matching the linear resources required to type the
+corresponding constructor argument in the scrutinee. The resources required to
+type each constructor sub-expression come annotated in the judgement (as
+${:}_{\ov{\D_i}}$). Unrestricted fields of the constructor are introduced as
+unrestricted variables. We note that the typing environment $\D$ always
+contains the resources $\ov{\D_i}$ in uses of the alternative judgement.
+
+Second, the rule for alternatives that match on the wildcard pattern, be the
+original scrutinee in WHNF or not (note the use of $\Rightarrow$, which
+is applicable both under $\Rightarrow_{WHNF}$ and $\Rightarrow_{NWHNF}$):
 \[
@@
 To type a wildcard alternative we simply type the expression with the main
-judgement, ignoring all annotations on the judgement. We recall that the case
-binder was already introduced in the environment with the appropriate usage
-environment by the case expression.
-
-Finally, consider an alternative matching on a case constructor without any
-linear components. According to the definition of consuming a resource from
-Linear Haskell, the linear resources of a scrutinee matching such a pattern are
-fully consumed in the body of the corresponding alternative, since the
-scrutinee must have been evaluated to a form that does not have any linear
-components.
-%
-This definition agrees with the intuition we have developed by example in the
-previous section, and with the typing rule we devised for alternatives matching
-constructors without linear components.
-Taking into account that case expressions introduce the linear resources of the
-scrutinee in the typing environment of all alternatives, and in the usage
-environment of the case binder, we must reactively update the typing
-environments after matching on such a pattern.
+judgement, ignoring all annotations on the judgement. The case binder will
+have already been introduced in the environment with the appropriate usage
+environment by the relevant (WHNF or not) case expression rule.
+
+Lastly, consider an alternative matching on a data constructor without any
+linear components. The linear resources used to type a scrutinee matching such
+a pattern are fully consumed during evaluation: the resulting unrestricted
+constructor application can only be well-typed under an empty linear
+environment.
+%
+Consequently, the resources to type the scrutinee which are carried over to the
+case-alternative environments by the $Case_{WHNF}$ and $Case_{NWHNF}$ rules
+(see the latter below), must be reactively destroyed from branches where the
+pattern is unrestricted. In failing to do this, we'd allow the fully-consumed
+resources to be used again in the branch!
+% Really, in the WHNF case it is not really necessary to remove the things
+% since they must already be empty be definition.
+%
+%This definition agrees with the intuition we have developed by example in the
+%previous section, and with the typing rule we devised for alternatives matching
+%constructors without linear components.
+%
+%Taking into account that case expressions introduce the linear resources of the
+%scrutinee in the typing environment of all alternatives, and in the usage
+%environment of the case binder, we must reactively update the typing
+%environments after matching on such a pattern.
 %
@@
 the scrutinee typing environment by the empty linear environment $\cdot$)
-and from the usage environment of the case binder $z$, written $\G[\cdot/\D_s]_z$
-to denote replacing the usage environment of the variable $z$ in $\G$, which
-is necessarily $\D_s$ (since we always annotate the judgement with the
-environment of the scrutinee), by the empty environment.
+and from the usage environment of the case binder $z$ (written $\G[\cdot/\D_s]_z$
+to similarly delete $\D_s$ from the usage environment of the $\D$-var $z$ in $\G$).
+Given that the case-binder $z$ is always annotated with $\D_s$ (essentially,
+the linear environment typing the scrutinee), this latter substitution makes
+the case-binder unrestricted in branches with unrestricted patterns.
+%-- agreeing with our observation in Section \label{sec:semantic-linearity-examples}.
 
@@
 %
-The rule faithfully encodes the notion that an expression matching such a
-pattern is unrestricted when evaluated to WHNF, implying that all linear
-resources have been consumed to produce it, and the result may 
-be freely discarded or duplicated.
-%
-Reactively deleting the resources that typed the scrutinee from the linear
-environment ensures that when we match on an unrestricted pattern there are no
-remaining linear resources to consume. Otherwise, e.g., $\ccase{K_1~x}{\{K_2 \to
-K_2,K_1~y \to K_1~y\}}$ would not be well-typed since the resource $x$ would
-be left \emph{un}-consumed in the first alternative, even though we know by
-matching on $K_1$ that it has already been consumed.
-%
-Furthermore, since the case binder in such an alternative refers to the
-unrestricted expression, the case binder too may be used unrestrictedly, which
-we allow by emptying its usage environment.
+This rule observes that any linear resources required to type an expression
+that reduces to an unrestricted constructor application are fully consumed
+through evaluation and the resulting unrestricted expression can be freely
+discarded or duplicated (e.g. via the case binder).
+%
+%Reactively deleting the resources that typed the scrutinee from the linear
+%environment ensures that when we match on an unrestricted pattern there are no
+%remaining linear resources to consume. Otherwise, e.g., $\ccase{K_1~x}{\{K_2 \to
+%K_2,K_1~y \to K_1~y\}}$ would not be well-typed since the resource $x$ would
+%be left \emph{un}-consumed in the first alternative, even though we know by
+%matching on $K_1$ that it has already been consumed.
+%
+%Furthermore, since the case binder in such an alternative refers to the
+%unrestricted expression, the case binder too may be used unrestrictedly, which
+%we allow by emptying its usage environment.
 
@@
 
-We thank Arnaud Spiwack for insightful discussions and Well-Typed for funding
-this work.\todo{Do these have to be anonymized?}
+We thank Arnaud Spiwack for valuable and insightful discussions.\todo{Do these have to be anonymized?}
 

### 2025-02-24 icfp.tex
Start 3.6.2

@@
 
-\subsubsection{Pattern matching on non-WHNF terms via proof irrelevant resources}
-
-Resources used to type a scrutinee that is not in WHNF \emph{must not} be used in
-the case alternatives.
-%
-However, evaluating the scrutinee to WHNF is not sufficient to
+\subsubsection{Proof irrelevant resources}
+
+When the scrutinee is not in WHNF, it is not known statically which of the
+scrutinee resources end up being consumed during evaluation and which would
+still be available in the branch continuation.
+%
+Therefore, cases with non-WHNF scrutinees must by typed such that no resource
+from the scrutinee can be used directly in the branch again, or risk
+consuming the same resource twice.
+%
+However, simply evaluating the scrutinee to WHNF does not guarantee all
+resources are consumed (unless the result matches an unrestricted pattern).
+In the case where the matching pattern has some linear components
+
+is not sufficient to
 consume all resources used by the scrutinee, since sub-expressions such
@@
 are linear resources in all other senses, meaning they must be used
-\emph{exactly once}. However, since proof irrelevant resources cannot be
-discarded nor used directly, they have to be consumed \emph{indirectly} --
-by $\D$-bound variables, namely, the case binder, or the linear pattern-bound variables.
+exactly once. However, since proof irrelevant resources cannot be
+discarded nor used directly, they have to be consumed \emph{indirectly} via
+$\D$-bound variables, namely, the case binder, or the linear pattern-bound
+variables.
 

### 2025-02-24 icfp.tex
improve

@@
 intermediate language of the Glasgow Haskell Compiler. We prove Linear Core
-guarantees linear resource usage via a type safety and bisimilarity argument and prove
-multiple optimising transformations preserve linearity in Linear Core while
-failing to do so in Core. We have implemented Linear Core as a compiler
-plugin to validate the system against linearity-heavy libraries, including
+is sound and guarantees linear resource usage and prove multiple optimising
+transformations preserve linearity in Linear Core while failing to do so in
+Core. We have implemented Linear Core as a compiler plugin to validate the
+system against linearity-heavy libraries, including
 \texttt{linear-base}.%^, in the heart of the compiler.
@@
 consume the result of evaluating the scrutinee.
-
-Finally, we discuss the special case of scrutinizing a
-single variable $x$, e.g.:
-$
-\lambda x.~\ccase{x}{\_ \to x}
-$. 
-It might seem as though this function is linear:
-%begin{itemize}
-% \item
-if the argument instantiating $x$ is in WHNF, then scrutinizing it is a no-op, and returning $x$
-just returns the resource intact;
-% \item
-if the argument instantiating $x$ is not in WHNF, then scrutinizing it evaluates it to WHNF, and returning
-$x$ returns the result of evaluating $x$ that still had to be consumed exactly
-once.
-%\end{itemize}
-However, in practice, this program's linearity can only be determined by
-further analysis on the non-strict evaluation strategy. If linear
-function applications are $\beta$-reduced \emph{call-by-name}
-(often desirable, as linear functions are known to use their argument exactly
-once)
-% and thus call-by-name reduction will definitely not duplicate computation
-while the example function is considered linear, then such an application might
-duplicate linear resources during evaluation. For instance:
-\[
-\begin{array}{l}
-(\lambda x.~\ccase{x}{\_ \to x})~(free~x)
-\Longrightarrow_{CBN}
-\ccase{free~x}{\_ \to free~x}
-\end{array}
-\]
-On the other hand, if linear function applications are reduced
-\emph{call-by-need}, the above example would instead be semantically linear
-because the result of freeing $x$ would be bound to some $y$ first.
-%
-The system we present in the next section does accept this function
-scrutinizing a variable as linear, as our operational semantics are
-call-by-need.
-This interaction between evaluation and linearity is considered specifically in the
-reverse binder swap transformation discussed in
-Section~\ref{sec:optimisations-preserve-types-meta}.\todo{Figure out precisely reverse binder swap proof and reevaluate}
 
@@
 
+\todo{UPDATE}
+Finally, we discuss the special case of scrutinizing a
+single variable $x$, e.g.:
+$
+\lambda x.~\ccase{x}{\_ \to x}
+$. 
+It might seem as though this function is linear:
+%begin{itemize}
+% \item
+if the argument instantiating $x$ is in WHNF, then scrutinizing it is a no-op, and returning $x$
+just returns the resource intact;
+% \item
+if the argument instantiating $x$ is not in WHNF, then scrutinizing it evaluates it to WHNF, and returning
+$x$ returns the result of evaluating $x$ that still had to be consumed exactly
+once.
+%\end{itemize}
+However, in practice, this program's linearity can only be determined by
+further analysis on the non-strict evaluation strategy. If linear
+function applications are $\beta$-reduced \emph{call-by-name}
+(often desirable, as linear functions are known to use their argument exactly
+once)
+% and thus call-by-name reduction will definitely not duplicate computation
+while the example function is considered linear, then such an application might
+duplicate linear resources during evaluation. For instance:
+\[
+\begin{array}{l}
+(\lambda x.~\ccase{x}{\_ \to x})~(free~x)
+\Longrightarrow_{CBN}
+\ccase{free~x}{\_ \to free~x}
+\end{array}
+\]
+On the other hand, if linear function applications are reduced
+\emph{call-by-need}, the above example would instead be semantically linear
+because the result of freeing $x$ would be bound to some $y$ first.
+%
+The system we present in the next section does accept this function
+scrutinizing a variable as linear, as our operational semantics are
+call-by-need.
+This interaction between evaluation and linearity is considered specifically in the
+reverse binder swap transformation discussed in
+Section~\ref{sec:optimisations-preserve-types-meta}.\todo{Figure out precisely reverse binder swap proof and reevaluate}
+
 Reverse binder swap is dual to the binder swap transformation, substituting
@@
        \infer[]
-    {\Gamma ; \Delta  \vdash (\Theta \mid e)
-    \Downarrow (\Theta' \mid v) : \sigma , \Sigma}
+    {\Gamma ; \Delta  \vdash (\Theta, x:_{\Delta'} \sigma = e \mid e)
+    \Downarrow (\Theta' \mid v) : \sigma , \Sigma \\
+      \D'' = \D' \setminus (\Theta \setminus \Theta')
+      }
     {\Gamma ; \Delta \vdash (\Theta , x:_{\Delta'} \sigma = e \mid x)
@@
     \\[1.5em]
-    % \infer[]
-    % {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
-    % % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
-    % % \Downarrow \Theta_n : v_n \quad
-    % (\Gamma ; \Delta \vdash (\Theta, \overline{x_i:_\omega \sigma_i =
-    % e_i} \mid e' )\Downarrow (\Theta'
-    % \mid v) : \tau
-    % }
-    % {\Gamma ;\Delta \vdash (\Theta \mid \lletrec{\overline{x_i{:}_{\Delta}\sigma_i = e_i}}{e'})
-    % \Downarrow (\Theta' \mid v) : \tau}
+
+     \infer[]
+     {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
+     % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
+     % \Downarrow \Theta_n : v_n \quad
+     (\Gamma ; \Delta \vdash (\Theta, \overline{x_i:_\Delta' \sigma_i =
+     e_i} \mid e' )\Downarrow (\Theta'
+     \mid v) : \tau
+     }
+     {\Gamma ;\Delta \vdash (\Theta \mid \lletrec{\overline{x_i{:}_{\Delta'}\sigma_i = e_i}}{e'})
+     \Downarrow (\Theta' \mid v) : \tau}
 

### 2025-02-24 icfp.tex
.

--- 2025-02-24 icfp.tex
@@
        \infer[]
-    {\Gamma ; \Delta  \vdash (\Theta \mid e)
+    {\Gamma ; \Delta  \vdash (\Theta,  \mid e)
     \Downarrow (\Theta' \mid v) : \sigma , \Sigma}
@@
     \\[1.5em]
-    % \infer[]
-    % {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
-    % % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
-    % % \Downarrow \Theta_n : v_n \quad
-    % (\Gamma ; \Delta \vdash (\Theta, \overline{x_i:_\omega \sigma_i =
-    % e_i} \mid e' )\Downarrow (\Theta'
-    % \mid v) : \tau
-    % }
-    % {\Gamma ;\Delta \vdash (\Theta \mid \lletrec{\overline{x_i{:}_{\Delta}\sigma_i = e_i}}{e'})
-    % \Downarrow (\Theta' \mid v) : \tau}
+    \infer[]
+    {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
+    % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
+    % \Downarrow \Theta_n : v_n \quad
+    (\Gamma ; \Delta \vdash (\Theta, \overline{x_i:_\Delta \sigma_i =
+    e_i} \mid e' )\Downarrow (\Theta'
+    \mid v) : \tau
+    }
+    {\Gamma ;\Delta \vdash (\Theta \mid \lletrec{\overline{x_i{:}_{\Delta}\sigma_i = e_i}}{e'})
+    \Downarrow (\Theta' \mid v) : \tau}
 
@@
 allowing us to transfer type safety of the instrumented semantics to
-the lazy big-step semantics.
+the lazy big-step semantics. The relation essentially allows the
+execution environment to treat some unrestricted bindings
+of the natural semantics as linear.
 
@@
 e' : \tau,\Sigma)$ between a well-typed state and an ordinary state of
-the lazy big-step semantics if $e = e'$ and $\Theta' \leq \Theta\uparrow^\omega$
+the lazy big-step semantics if $e = e'$ and $\Theta'{\uparrow^\omega}
+\leq \Theta$
 \end{definition}
@@
 For any well-typed $[\Theta \mid e]$, if $\Theta : e \Downarrow
-\Theta' : e'$ or $\Theta : e \Downarrow^* \Theta' e'$ then we have
+\Theta' : e'$ or $\Theta : e \Downarrow^* \Theta' : e'$ then we have
 that $[\Theta \mid e']$ is well-typed.

### 2025-02-24 icfp.tex

--- 2025-02-24 icfp.tex
@@
 We begin by presenting the operational semantics of $\lambda^\pi_\D$,
-consisting of two lazy big-step semantics in the
-style of Launchbury~\cite{10.1145/158511.158618}, one of which is
-instrumented with linearity information, producing stuck states when
-linearity is violated. We then show the
-type safety  of $\lambda^\pi_\D$ via the standard progress  and type
-preservation results (Section~\ref{sec:typesafe}), which in our setting entails that programs
-preserve linearity. Finally, we show that multiple optimising
+consisting of a lazy natural semantics in the
+style of Launchbury~\cite{10.1145/158511.158618}.
+This semantics conveys the expected behavior of an implementation of
+lazy evaluation and is agnostic to any linearity) information.
+Thus, such a semantics is ill-suited for meta-reasoning about
+linearity and so we develop an \emph{instrumented} semantics that is
+enriched with sufficient information to allow us to establish type
+safety, showing that linearity is preserved. We show that the natural
+and instrumented semantics are bisimilar for well-typed terms and so
+derive type safety in the natural semantics
+(Section~\ref{sec:typesafe}). Finally, we show that multiple optimising
 transformations such as those performed by GHC on (Linear) Core are
@@
 
-The first semantics, presented in Figure~\ref{fig:opsem}, captures
-standard lazy evaluation but ignores linearity information. The second
-semantics is significantly instrumented with type information and is thus
-\emph{linearity aware}, resulting in a stuck state when a linear
-variable is used more than once. The former is a more direct model of
-an implementation of lazy evaluation, however it is not so convenient
-to reason about formally. The latter semantics is augmented with the additional
-information necessary to strengthen the induction hypothesis to make our
-type safety proofs go through.
-%
-We show that for well-typed terms, the two semantics coincide.
-
-Both semantics are
-equipped with an environment which assigns values to variables and is
-mutated in order to express lazy evaluation. Terms are explicit about
+The Launchbury-style natural semantics, presented in Figure~\ref{fig:opsem}, captures
+standard lazy evaluation while ignoring linearity information.
+% The second
+% semantics is significantly instrumented with type information and is thus
+% \emph{linearity aware}, resulting in a stuck state when a linear
+% variable is used more than once. The former is a more direct model of
+% an implementation of lazy evaluation, however it is not so convenient
+% to reason about formally. The latter semantics is augmented with the additional
+% information necessary to strengthen the induction hypothesis to make our
+% type safety proofs go through.
+% %
+% We show that for well-typed terms, the two semantics coincide. 
+%
+%Both semantics are
+The natural semantics is equipped with an environment at assigns values to variables and is
+mutated in order to express evaluation. Terms are explicit about
 their sharing (in the sense of shared reductions in lazy evaluation),
@@
 \]}
-    \caption{Operational Semantics of $\lambda^\pi_\D$\label{fig:opsem}}
+    \caption{Natural Semantics of $\lambda^\pi_\D$\label{fig:opsem}}
   \end{figure}
 
-  The lazy big-step operational semantics is defined via the relation
+  The natural semantics is defined via the relation
 $\Theta : e \Downarrow \Theta' : v$ where $e$ is an expression, $v$ a
@@
 form $x:_\omega\sigma = e$ which assigns the expression $e$ to the
-variable $x$ of the given type.
-
+variable $x$ of the given type. The rules are standard, augmenting the
+runtime environment with let-bound expressions which are evaluated
+(only once) if the let-bound variable is forced.
+
+% The second
+% semantics is significantly instrumented with type information and is thus
+% \emph{linearity aware}, resulting in a stuck state when a linear
+% variable is used more than once. The former is a more direct model of
+% an implementation of lazy evaluation, however it is not so convenient
+% to reason about formally. The latter semantics is augmented with the additional
+% information necessary to strengthen the induction hypothesis to make our
+% type safety proofs go through.
+% %
+% We show that for well-typed terms, the two semantics coincide.
+
+While the natural semantics is a mostly direct model of an implementation
+of lazy evaluation, it is inconvenient for formal reasoning about
+well-typed linear terms.
+%
 We follow an approach
 similar to that of \cite{cite:linearhaskell}, defining a
-linearity-aware operational semantics in which linear variables are
+an \emph{instrumented} operational semantics in which linear variables are
 erased from 
-the runtime environment, ensuring that any term violating linearity
+the runtime environment once they are forced, ensuring that any term violating linearity
 will result in a stuck state. The semantics is also type-aware and
-carries with it sufficient data to reconstruct typing derivations (for
-the purposes of proving type safety). The semantics is defined by
-the judgment $\Gamma ; \Delta \vdash (\Theta \mid e) \Downarrow (\Theta'
-\mid v) : \sigma , \Sigma$ where $\Gamma$ and $\Delta$ are typing judgments
+carries with it sufficient data to reconstruct typing derivations for
+the purposes of showing type safety.
+
+The semantics is presented in
+Figure~\ref{fig:linopsem}, defining the judgment $\Gamma ; \Delta \vdash (\Theta \mid e) \Downarrow (\Theta'
+\mid v) : \sigma , \Sigma$ where $\Gamma$ and $\Delta$ are typing
+contexts for expressions in $\Sigma$
 ($\Gamma$ tracks unrestricted and $\Delta$-bound variables, $\Delta$ tracks linear
@@
 $e :_\pi \tau$ or $x :_\Delta \tau$ which are expressions in which
-bindings in $\Theta$ are used. This additional data is needed to
-inductively show that the evaluation state is well-typed (and
-linearity preserving).
+bindings in $\Theta$ may also be used. This additional data is needed to
+inductively show that the overall evaluation state is well-typed (and
+linearity preserving). The rules mostly mirror those of the natural
+semantics, with the exception of the linear variable rule which erases
+the binding from the execution environment. Both semantics define
+evaluation relations of the form $a \Downarrow b$, where $a$ and $b$
+are evaluation states. We also rely on a notion
+of partial derivations and partial evaluation (see
+Appendix~\ref{app:opsem}), written $a\Downarrow^* b$ which allows us
+to reason about progress.
 
@@
        \infer[]
-    {\Gamma ; \Delta  \vdash (\Theta \mid e)
+    {\Gamma ; \Delta  \vdash (\Theta  , x:_{\Delta'} \sigma = e\mid e)
     \Downarrow (\Theta' \mid v) : \sigma , \Sigma}
@@
     \\[1.5em]
-    % \infer[]
-    % {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
-    % % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
-    % % \Downarrow \Theta_n : v_n \quad
-    % (\Gamma ; \Delta \vdash (\Theta, \overline{x_i:_\omega \sigma_i =
-    % e_i} \mid e' )\Downarrow (\Theta'
-    % \mid v) : \tau
-    % }
-    % {\Gamma ;\Delta \vdash (\Theta \mid \lletrec{\overline{x_i{:}_{\Delta}\sigma_i = e_i}}{e'})
-    % \Downarrow (\Theta' \mid v) : \tau}
+    \infer[]
+    {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
+    % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
+    % \Downarrow \Theta_n : v_n \quad
+    \Gamma ; \Delta \vdash (\Theta, \overline{x_i:_\omega \sigma_i =
+    e_i} \mid e' )\Downarrow (\Theta'
+    \mid v) : \tau
+    }
+    {\Gamma ;\Delta \vdash (\Theta \mid \lletrec{\overline{x_i{:}_{\Delta}\sigma_i = e_i}}{e'})
+    \Downarrow (\Theta' \mid v) : \tau}
 
@@
  \]
- \caption{Linearity-aware Operational Semantics~\label{fig:linopsem}}
+ \caption{Instrumented Operational Semantics~\label{fig:linopsem}}
 
@@
 The $\lambda^\pi_\D$ type system is sound: well-typed programs in Linear Core do not
-get \emph{stuck}. Besides type safety, we
-prove that multiple optimising transformations used in GHC preserve
-linearity in the sense of being well-typed in our system.
+get \emph{stuck} and their evaluation preserves linearity.
 %
@@
 substitution lemmas, one for each kind of variable: unrestricted variables
-$\xo$, linear variables $\xl$, and $\D$-bound variables $\xD$.
+$\xo$, linear variables $\xl$, and $\D$-bound variables $\xD$. The
+development is reported in Appendices~\ref{app:proofs}
+and~\ref{app:opsem}, where type safety is established directly for the
+instrumented semantics and ported to the natural semantics via a
+bisimulation argument.
+
+  \begin{theorem}[Type Preservation]
+    For any well-typed evaluation state $a$, if $a \Downarrow
+    b$ or $a \Downarrow^* b$ then $b$ is well-typed.
+\end{theorem}
+\begin{theorem}[Progress]
+Let $a$ be a well-typed evaluation state. For any partial evaluation $a \Downarrow^* b$ 
+the evaluation can be extended.
+\end{theorem}
 
@@
 
-\TypePreservationTheorem
-%
-\ProgressTheorem
+% \TypePreservationTheorem
+% %
+% \ProgressTheorem
 % \noindent Type preservation states that a well-typed expression $e$ that
@@
 %
-Type preservation
-% ($\S$~\ref{sec:proof:type-preservation})
-follows by structural induction on the reductions $e \longrightarrow
-e'$ from the operational semantics. Most cases are straightforward, appealing to one or more
-substitution lemmas (see Appendix~\ref{sec:proof:type-preservation}). The most
-interesting case is that of case expressions whose scrutinee can be further
-evaluated -- we branch on whether the scrutinee becomes in WHNF and appeal to Lemma~\ref{lem:irrev} if so.
+% Type preservation
+% % ($\S$~\ref{sec:proof:type-preservation})
+% follows by structural induction on the reductions $e \longrightarrow
+% e'$ from the operational semantics. Most cases are straightforward, appealing to one or more
+% substitution lemmas (see Appendix~\ref{sec:proof:type-preservation}). The most
+% interesting case is that of case expressions whose scrutinee can be further
+% evaluated -- we branch on whether the scrutinee becomes in WHNF and appeal to Lemma~\ref{lem:irrev} if so.
 %
@@
 
-\input{language-v4/proofs/TypePreservationTheorem}
+%\input{language-v4/proofs/TypePreservationTheorem}
 
@@
 
-\input{language-v4/proofs/ProgressTheorem}
+%\input{language-v4/proofs/ProgressTheorem}
 
@@
 The big-step operational semantics is defined via the relation
-$\Gamma : e \Downarrow \Delta : v$ where $e$ is an expression, $v$ a
-value, $\Gamma$ and $\Delta$ are environments with bindings of the
+$\Theta : e \Downarrow \Theta' : v$ where $e$ is an expression, $v$ a
+value, $\Theta$ and $\Theta'$ are environments with bindings of the
 form $x:_\omega\sigma = e$ which assigns the expression $e$ to the
@@
     {\,}
-    {\Gamma : \Lambda p.e \Downarrow \Gamma : \Lambda p.e}
+    {\Theta : \Lambda p.e \Downarrow \Theta : \Lambda p.e}
     \qquad
     \infer[]
-    {\Gamma : e\Downarrow \Delta : \Lambda p.e' \quad \Delta :
-    e'[\pi/p]\Downarrow \Theta : v}
-    {\Gamma : e~\pi\Downarrow \Theta : v }\\[1.5em]
+    {\Theta  : e\Downarrow \Theta' : \Lambda p.e' \quad \Theta' :
+    e'[\pi/p]\Downarrow \Theta'' : v}
+    {\Theta : e~\pi\Downarrow \Theta'' : v }\\[1.5em]
     \infer[]
     {\,}
-    {\Gamma : \lambda x:_\pi \sigma . e \Downarrow \Gamma : \lambda x:_\pi \sigma . e }
+    {\Theta : \lambda x:_\pi \sigma . e \Downarrow \Theta : \lambda x:_\pi \sigma . e }
     \qquad
     \infer[]
-    {\Gamma : e \Downarrow \Delta : \lambda y:_\pi \sigma . e' \quad
-    \Delta : e'[x/y] \Downarrow \Theta : v }
-    {\Gamma : e~x \Downarrow \Theta : v}\\[1.5em]
+    {\Theta : e \Downarrow \Theta' : \lambda y:_\pi \sigma . e' \quad
+    \Theta' : e'[x/y] \Downarrow \Theta'' : v }
+    {\Theta : e~x \Downarrow \Theta'' : v}\\[1.5em]
     \infer[]
-    {\Gamma : e \Downarrow \Delta : v}
-    {(\Gamma , x:_\omega \sigma = e) : x \Downarrow (\Delta,x:_\omega
+    {(\Theta , x:_\omega \sigma = e): e \Downarrow \Theta' : v}
+    {(\Theta , x:_\omega \sigma = e) : x \Downarrow (\Theta',x:_\omega
     \sigma = v) : v}
@@
     \infer[]
-    {(\Gamma,x:_\omega \sigma = e) : e' \Downarrow \Theta : v}
-    {\Gamma : \llet{x{:}_{\Delta}\sigma = e}{e'}\Downarrow \Theta : v
+    {(\Theta,x:_\omega \sigma = e) : e' \Downarrow \Theta' : v}
+    {\Theta : \llet{x{:}_{\Delta}\sigma = e}{e'}\Downarrow \Theta' : v
     }\\[1.5em]
@@
     {\,}
-    {\Gamma : K~\ov{x_i} \Downarrow \Gamma : K~\ov{x_i}}
+    {\Theta : K~\ov{x_i} \Downarrow \Theta : K~\ov{x_i}}
     \qquad
     \infer[]
-    {\Gamma : e \Downarrow \Delta : K~\ov{x_i} \quad
-    \Delta : e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] \Downarrow \Theta : v }
-    {\Gamma :
+    {\Theta : e \Downarrow \Theta' : K~\ov{x_i} \quad
+    \Theta' : e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] \Downarrow \Theta'' : v }
+    {\Theta :
     \ccase{e}{z{:}_{\Delta'}\sigma~\{{\dots,K~\ov{y_i} \to e', {\dots}}\}}
-    \Downarrow \Theta : v}\\[1.5em]
+    \Downarrow \Theta'' : v}\\[1.5em]
     \infer[]
-    {\Gamma : e \Downarrow \Delta : K~\ov{x_i} \quad
-    \Delta : e'[K~\ov{x_i}/z] \Downarrow \Theta : v }
-    {\Gamma :
+    {\Theta : e \Downarrow \Theta' : K~\ov{x_i} \quad
+    \Theta' : e'[K~\ov{x_i}/z] \Downarrow \Theta'' : v }
+    {\Theta :
     \ccase{e}{z{:}_{\Delta'}\sigma~\{\dots , \_\to
-    e'}\} \Downarrow \Theta : v}
+    e'}\} \Downarrow \Theta'' : v}
     \qquad
@@
     % \Downarrow \Theta_n : v_n \quad
-    (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e' \Downarrow \Theta
+    (\Theta, \overline{x_i:_\omega \sigma_i = e_i}) : e' \Downarrow \Theta'
     : v
     }
-    {\Gamma : \lletrec{\overline{x_i{:}_{\Delta}\sigma_i = e_i}}{e'}
-    \Downarrow \Theta : v}
+    {\Theta : \lletrec{\overline{x_i{:}_{\Delta}\sigma_i = e_i}}{e'}
+    \Downarrow \Theta' : v}
     \end{array}
@@
 
-We now relate our two operational semantics through a bisimulation,
+We relate our two operational semantics through a bisimulation,
 allowing us to transfer type safety of the instrumented semantics to
-the lazy big-step semantics.
+the lazy big-step semantics. The relation essentially allows the
+execution environment to treat some unrestricted bindings
+of the natural semantics as linear.
 

### 2025-02-24 icfp.tex
minor updates

@@
 call-by-need big-step operational semantics that becomes stuck when linear
-variables in the runtime heap are used more than once.
-\todo{O que digo aqui sobre a semantica launchbury style?}
+variables in the runtime heap are used more than once. We show this
+semantics to be bisimilar to a Launchbury-style~\cite{10.1145/158511.158618} natural semantics.
+%\todo{O que digo aqui sobre a semantica launchbury style?}
 Crucially,
@@
 
-\section*{Acknowledgements}
-
-We thank Arnaud Spiwack for valuable and insightful discussions.\todo{Do these have to be anonymized?}
+% \section*{Acknowledgements}
+
+% We thank Arnaud Spiwack for valuable and insightful discussions.\todo{Do these have to be anonymized?}
 

### 2025-02-25 icfp.tex
3.6.2 done

@@
 
-When the scrutinee is not in WHNF, it is not known statically which of the
-scrutinee resources end up being consumed during evaluation and which would
-still be available in the branch continuation.
-%
-Therefore, cases with non-WHNF scrutinees must by typed such that no resource
+When the scrutinee is not in WHNF, it is not statically known which resources
+will have been consumed after evaluating said scrutinee, or which
+are transferred into the branch continuation, e.g. as in $\ccase{f\ x}{z\ \{K\
+a \rightarrow e\}}$ where $f = K$, but opaque.
+\begin{enumerate}
+
+    \item Consequently, cases with non-WHNF scrutinees must by typed such that no resource
     from the scrutinee can be used directly in the branch again, or risk
-consuming the same resource twice.
-%
-However, simply evaluating the scrutinee to WHNF does not guarantee all
-resources are consumed (unless the result matches an unrestricted pattern).
-In the case where the matching pattern has some linear components
-
-is not sufficient to
-consume all resources used by the scrutinee, since sub-expressions such
-as constructor arguments will be left unevaluated and may still contain
-(``parts of'') the linear resources. To \emph{fully} consume all
-resources occurring in the scrutinee, all linear components of the
-constructor must also be fully evaluated, as witnessed by the
-$\textsc{Alt}_0$ rule above.
-%
-In essence, to type cases scrutinizing expressions not in WHNF, we need to
-forbid the usage of the scrutinee resources while tracking that we necessarily
-``finish consuming'' them, either via the case binder, or by linearly using all
-linear pattern-bound variables introduced in the alternative.
+    consuming the same resource twice. In the example, $x$ must not directly occur
+    in $e$ since $f$ \emph{could} have consumed it.
+
+    \item On the other hand, since evaluation to WHNF does not guarantee all
+    resources do end up being consumed (unless the result matches an
+    unrestricted pattern), we must ensure any which transfer from the
+    scrutinee to the alternative are ultimately consumed. (In the example,
+    $x$ is not consumed by evaluating the scrutinee).
+
+    Considering (1), we must resort to a more uniform method of guaranteeing we
+    ``finish consuming`` the scrutinee. And there are only two ways of
+    uniformly referring to the remaining linear resources in the newly-evaluated scrutinee:
+
+        \begin{itemize}
+            \item The case binder, which refers to the whole result of evaluating the scrutinee
+            \item The linear components of a constructor pattern, which refer to expressions that may contain linear resources.
+        \end{itemize}
+
+%To guarantee all post-evaluation unconsumed scrutinee resources are consumed in
+%the alternative, without statically knowing
+%Statically we can't know which resources are ``transferred'', but  guarantee
+%
+\end{enumerate}
+
+%is not sufficient to
+%consume all resources used by the scrutinee, since sub-expressions such
+%as constructor arguments will be left unevaluated and may still contain
+%(``parts of'') the linear resources. To \emph{fully} consume all
+%resources occurring in the scrutinee, all linear components of the
+%constructor must also be fully evaluated, as witnessed by the
+%$\textsc{Alt}_0$ rule above.
+%%
+%In essence, to type cases scrutinizing expressions not in WHNF, we need to
+%forbid the usage of the scrutinee resources while tracking that we necessarily
+%``finish consuming'' them, either via the case binder, or by linearly using all
+%linear pattern-bound variables introduced in the alternative.
 
@@
 discarded nor used directly, they have to be consumed \emph{indirectly} via
-$\D$-bound variables, namely, the case binder, or the linear pattern-bound
+$\D$-bound variables, namely, the case binder, or, in mutual exclusion, the linear pattern-bound
 variables.
@@
 having the same proof irrelevant linear context $[\D]$ in the typing
-environment, annotating the judgement with the proof irrelevant resources,
-and using the $\Rightarrow_{\textsf{NWHNF}}$ judgement:
+environment, and annotating the judgement with the proof irrelevant resources
+for use in the $\Rightarrow_{\textsf{NWHNF}}$ judgement:
 \[
@@
 Note how the rule is similar to the one for scrutinees in WHNF, but
-requiring the resources in the case binder, typing environment, and
-judgement annotation to be made irrelevant.
+requires the resources in the case binder, typing environment, and
+judgement annotation to be made ``irrelevant''.
 
@@
 as if the scrutinee were not in WHNF. Having introduced proof irrelevant
-resources, we can now present the correct $Case_\textrm{WHNF}$ rule:
+resources, we can now present the final $Case_\textrm{WHNF}$ rule:
 \[
@@
 due to
-anonymization]}%\footnote{https://gitlab.haskell.org/ghc/ghc/-/issues/23218} ,
+anonymization]} %\footnote{https://gitlab.haskell.org/ghc/ghc/-/issues/23218} ,
 we leave this as future work.

### 2025-02-25 icfp.tex
3.6.3; and 3 whole is done on this pass

@@
 As opposed to scrutinees in WHNF, where the resources used to type a scrutinee
-can be cleanly divided amongst the various sub-expressions of e.g. the
-constructor and henceforth each pattern variable, there is no direct mapping
-between the resources typing a scrutinee \emph{not in WHNF} and the usage
+can be cleanly divided amongst the various sub-expressions of a constructor
+application and, henceforth, each pattern variable, there is no direct mapping
+between the resources typing a scrutinee not in WHNF and the usage
 environments of pattern variables in any alternative.
@@
 obtained from the scrutinee environment \emph{tagged} with the constructor name
-and linear-variable index $\lctag{\irr{\D_s}\!}{K_i}$. $y_i{:}_{\D_i}\s$ is
+and linear-variable index $\lctag{\irr{\D_s}\!}{K_i}$. Then, $y_i{:}_{\D_i}\s$ is
 introduced in $\G$, just like other $\D$-variables.
@@
 %
-Having tagged the resources in the usage environment of each pattern variable
-differently, we need now only to express that (1) the pattern $\D$-variables
-can be used (i.e., the tagged resources need be available in the linear
+Having uniquely tagged the resources in the usage environment of each pattern
+variable, we need now only to express that (1) the pattern-bound $\D$-variables can
+be used (i.e., the tagged resources need be available in the linear
 environment in order for the $\textsc{Var}_\D$ rule to be applicable) and (2)
-if we consume one of the pattern variables with a non-empty usage environment,
-then \emph{all} pattern variables for linear components must also be consumed.
-
-The key idea to guarantee these is to provide a $\textit{Split}$ rule that
+if a $K$-pattern-bound $\D$-var is consumed, all remaining $K$-$\D$-vars,
+standing for the remaining linear components of the same pattern, must also be
+consumed.
+
+To satisfy these two properties, we introduce a $\textit{Split}$ rule that
 allows a linear resource $x{:}_1\s$ to be split into $n$ resources at a given
@@
 remaining tagged resources cannot be used directly, yet need still be consumed
-exactly once. Thus, the choice to consume the tagged resources via usage
+exactly once. Thus, the choice to consume the tagged resources via the usage
 environments of the other pattern variables is forced, ensuring that no

### 2025-02-25 icfp.tex
big pass, stopped at 3.6.1

@@
 
-Linear type systems~\cite{cite:linear-logic,cite:barberdill} add expressiveness
-to existing type systems by enforcing that certain resources (e.g.~a
+Linear types~\cite{cite:linear-logic,cite:barberdill} increase the expressiveness
+of type systems by allowing them to enforce that certain resources (e.g.~a
 file handle) are used \emph{exactly once}.
@@
 statically enforce correct usage of file or socket descriptors, or
-that heap-allocated memory is freed exactly once (leaks and double-frees become
+that heap-allocated memory is freed exactly once (i.e.,~leaks and double-frees become
 type errors), or even guarantee deadlock freedom in channel-based communication
@@
 %
-Typechecking Core serves as a sanity check to the correction of source
-transformations, since unsound ones will likely introduce ill-typed
-expressions, making them easy to spot.
+Typechecking Core serves as a sanity check to the correction of such
+transformations, since unsound ones are likely to introduce ill-typed
+expressions. %, making them easy to spot.
 %
@@
         Intuitively, a syntactic occurrence of a linear resource does not
-        necessarily entail consuming that resource because expressions are not
+        necessarily entail consuming that resource since expressions are not
         necessarily evaluated (i.e., if the expression is not evaluated, the
@@
         %
-        Under eager evaluation, a syntactic occurrence of a resource always implies its use at runtime.
+        Under eager evaluation, a syntactic occurrence of a resource
+        along a given control-flow path always implies its use at runtime.
         %
@@
 Despite the advantages of a linearly typed Core, the status quo is
-that linearity is effectively ignored in Core, despite being checked
+that linearity is effectively ignored in Core, being only checked
 in the source Haskell code. The reason is that the desugaring
@@
 that are syntactically very different from the original, especially
-due to the flexibility laziness provides a compiler in the
-optimisations it may perform.
-%
-Crucially, while optimisations aim to preserve the semantics of
+due to the flexibility provided by laziness wrt the
+optimisations a compiler may perform.
+
+While compiler optimisations aim to preserve the semantics of
 programs and therefore a program's linear resource consumption,
@@
 of $x$), it is indeed linear because the let-bound expression is never
-evaluated, thus $x$ is still consumed exactly once (when it is freed in the let body):
+evaluated and $x$ is still consumed exactly once (when it is freed in the let body):
 % executed, $x$ will be freed exactly once:
@@
 \item We introduce Linear Core, a non-strict linear language with
-the key features from Core (except for type equality coercions), whose type system
+the key features of Core (except for type equality coercions), whose type system
 accounts for semantic linearity in the presence of laziness. To the
@@
 
-\noindent Additional proofs and definitions are given in Appendices~\ref{app:proofs}
-and~\ref{app:optimisations}.
+\noindent Additional proofs and definitions are given in
+Appendices~\ref{app:proofs},~\ref{app:optimisations} and~\ref{app:opsem}.
 
@@
 In general, consuming a resource is equated with its 
-\emph{syntactic occurrence} in the program. However, as this section
+\emph{syntactic occurrence} along a control-flow path in the program. However, as this section
 makes clear, in linear type systems for non-strict
@@
 %
-From this observation, and exploring the connection between computation and evaluation,
+Given this observation, %and exploring the connection between computation and evaluation,
 % it becomes clear that \emph{linearity} and \emph{consuming resources}, in the
@@
 %
-we turn our focus to linearity under \todo{either non strict simpler sentence or say focus on lazy. decide}non-strict evaluation, both since GHC Core is
-lazy and because the distinction between semantically and syntactically
-consuming a resource is only exposed under non-strict semantics.
-%
+we turn our focus to linearity under lazy evaluation since the distinction between semantically and syntactically
+consuming a resource is only exposed under non-strict semantics (and since GHC Core is
+lazy).
+%.
+
 % Indeed, under \emph{call-by-value}, syntactic occurrences of a linear resource
@@
 
-This section aims to build an intuition for semantic linearity through examples
-of Core programs that can be understood as linear but are rejected by Core's
-linear type system (or, in fact, by all linear type systems we are aware of).
+This section aims to develop an intuition for semantic linearity through simple
+Core programs that can be understood as linear but are rejected by Core's
+linear type system (or, in fact, by all linear type systems we are
+aware of prior to this work).
 %
@@
 % (similar to towards the end of Section~\ref{sec:intro}),
-trivially showed how linearity can be more subtle in a non-strict language with
-a simple unused let binding.
-%
-Less trivially, a let-bound expression that captures a linear variable
-could be conditionally needed at runtime, depending on the branch taken.
+highlighted the subtlety of linearity in a non-strict language with
+a single unused let binding.
+%
+%Less trivially,
+In general, a let-bound expression that captures a linear variable
+can be conditionally needed at runtime, depending on the branch taken.
 Both optimising transformations (float-out) and programmers used to
@@
 \lstinline{x} must still be used exactly once, regardless of whether evaluation
-inlines the let (ala call-by-name) or creates a thunk (ala call-by-need):
+inlines the let (\emph{à la} call-by-name) or creates a thunk (\emph{à
+  la} call-by-need):
 %
@@
 as implying using \incode{x}, and therefore uses of \incode{z} imply using
-\incode{x}, however, if neither is used, then \incode{x} isn't used. Since
+\incode{x}, however, if neither are used, then \incode{x} isn't used. Since
 \incode{x} is effectively discarded, this example also violates linearity.
@@
 binders that depend on linear resources must be used \emph{at most once} in
-mutual exclusion with the linear resources themselves (in a sense, let-vars are
-affine in the let-body).
+mutual exclusion with the linear resources themselves (in a sense,
+let-bound variables are affine in the let body).
 %
@@
 The challenge lies in understanding linearity in the definitions of the
-recursive binders and which resources exactly are consumed when one of the
+recursive binders and determining which resources are consumed when one of the
 binders in the recursive group is used.
@@
 %
-If \incode{go} doesn't terminate, we aren't able to compute (nor consume) the result
+If \incode{go} doesn't terminate, we aren't able to compute (or consume) the result
 of \incode{f4}, so we do not promise anything about \incode{x} being consumed (\incode{f4}'s
@@
 
-That said, the linear type system must statically reject any recursive group that could
+Naturally, a linear type system must statically reject any recursive group that could
 lead to non-linear resource usage.
@@
 For instance, if the linear resource \incode{x} is used in one case alternative and
-recursion used more than once in the other, \incode{x} may end up being consumed
+recursion is used more than once in another, \incode{x} may end up being consumed
 more than once:
@@
 
-To determine whether the usage of a binding from a recursive group is linearly
-valid, we must infer the superset of resources that may end up being consumed
+To determine whether the usage of a binding from a recursive group is
+linear,
+we must infer the superset of resources that may end up being consumed
 in a single use of any of the bindings. For a strongly connected recursive
@@
 %
-Recursive let bindings in our system are discussed in Section~\ref{sec:recursivelets}.
+The treatment of recursive let bindings in our system is discussed in Section~\ref{sec:recursivelets}.
 
@@
 \paragraph{Case expressions}
-Above, we have observed more programs can be accepted as linear by delaying
+We have observed that more programs can be accepted as linear by delaying
 resource consumption in (recursive) let bindings until they are evaluated.
@@
 the tuple being scrutinized is already in WHNF and so evaluating it will
-not consume either $x$ nor $y$. Even if the tuple was constructed with
+not consume either $x$ or $y$. Even if the tuple was constructed with
 two expressions using $x$ and $y$ respectively, no computation would happen
@@
 
-Many of these examples might seem unrealistic due to the pattern match on a
-known constructor, but we note that in between the transformations programs
+While many of these examples may seem artificial due to the pattern match on a
+known constructor, we note that in between the transformations programs
 undergo in an optimising compiler, many such programs naturally occur (e.g.,~if
@@
 As for \emph{default} case alternatives, also known as \emph{wildcards}
-(written $\_$): matching a wildcard doesn't provide new linearity information,
-unlike a pattern with linear or unrestricted components, however, the
+(written $\_$): matching a wildcard doesn't provide any linearity information,
+% unlike a pattern with linear or unrestricted components,
+%
+but the
 scrutinee is still evaluated to WHNF.
@@
 %
-Briefly, when matching a wildcard, if the scrutinee is already in WHNF,
+When matching a wildcard, if the scrutinee is already in WHNF,
 programs that either use the resources from the scrutinee directly, or the case
@@
 and multiplicity polymorphism introduced by Linear
-Haskell~\cite{cite:linearhaskell} with all the key features from GHC's Core
+Haskell~\cite{cite:linearhaskell} with all the key features of GHC's Core
 language (algebraic datatypes, case expressions, and recursive lazy let bindings),
@@
 previous section in a linear type system that we prove is sound and guarantees
-linear resource usage at runtime via a standard soundness argument on a
+linear resource usage at runtime.
+This result is obtained (see Section~\ref{sec:main:metatheory}) via a soundness argument on a
 call-by-need big-step operational semantics that becomes stuck when linear
 variables in the runtime heap are used more than once. We show this
-semantics to be bisimilar to a Launchbury-style~\cite{10.1145/158511.158618} natural semantics.
+semantics to be bisimilar to a
+Launchbury-style~\cite{10.1145/158511.158618} natural semantics. 
 %\todo{O que digo aqui sobre a semantica launchbury style?}
@@
 Form or not, and using ``proof irrelevance'' to track resources that are no
-longer available but haven't yet been fully consumed
+longer available but have yet to be fully consumed
 (Section~\ref{sec:lc-case-exps}). Additionally, we introduce tags to split
 resources amongst pattern-bound variables and %^as usage environments
-encode that all linear components of a pattern must be used together.
+encode that all linear components of a pattern must be used in tandem.
 
@@
 We note that we handle multiplicity polymorphism differently from
-Linear Haskell in ignoring the multiplicity semiring and, instead,
-conservatively treating all multiplicity polymorphic functions as linear.
+Linear Haskell by ignoring the multiplicity semiring and
+conservatively treating all multiplicity polymorphic functions as
+linear, for the sake of simplicity.
 %
@@
 %
-Typing linear and unrestricted function applications separately is less general
-than typing applications of functions of any multiplicity $\pi$ by scaling (per the multiplicity
-semiring) the multiplicities of the resources used to type the argument by $\pi$, however, our objective of typing semantic
-linearity does not benefit much from doing so and therefore we opt for a
-simpler design. 
+Typing linear and unrestricted function application separately is less
+general than typing applications of functions of any multiplicity
+$\pi$ by scaling (per the multiplicity semiring) the multiplicities of
+the resources used to type the argument by $\pi$, however, our
+objective of typing semantic linearity does not benefit much from
+doing so and therefore we opt for a simpler design.
 % \todo{We might only need multiplicities for the case-of-case definition as it
@@
 of the abstraction, function types annotated with a $p$ variable and datatype
-fields with multiplicity $p$ are typed as thought they are linear functions and
+fields with multiplicity $p$ are typed as though they are linear functions and
 linear fields, because $p$ can be instantiated at both $\omega$ and $1$.
@@
 
-A \emph{usage environment} $\Delta$ is the means to encode the idea that lazy
-bindings do not consume the resources required by the bound expression when
-defined, but rather when the bindings themselves are consumed.
+A \emph{usage environment} $\Delta$ encodes the idea that lazy
+bindings do not consume resources upon definition, but rather when the bindings themselves are consumed.
 Specifically, we annotate so-called $\Delta$-bound variables with a \emph{usage
@@
 For example, the let-binder $u$ in the following program is annotated with a
-usage environment $\{x{:}_1\s,y{:}_1\s\}$, the resources that are used in its
+usage environment $\{x{:}_1\s,y{:}_1\s\}$ tracking the resources that are used in its
 body. Thus, in the expression $e$, consuming $u$ will equate to consuming the
@@
 \]
-The rule reads that an occurrence of a $\Delta$-bound variable is well-typed if
+The rule states that an occurrence of a $\Delta$-bound variable is well-typed if
 the linear environment is made up exactly of the resources in the usage environment of
@@
 resources used in let-bound expressions are only consumed when the respective
-binders are evaluated in the let-continuation.
+binders are evaluated in the let body.
 %, i.e. linear resources required by let-bound expressions are consumed lazily.
@@
 Moreover, resources captured by a let-bound expression cannot be used
-simultaneously with the let-binder in the continuation, since said
+simultaneously with the let-binder in its body, since said
 resources would end up being consumed more than once and violate (semantic)
 linearity. The binder has to be used in mutual exclusion with the linear
-resources captured by the expression it binds, and either the binding or the
+resources captured by the expression it binds -- either the binding or the
 resources \emph{must} be used.
@@
 consumption of resources to when the variables themselves are used and
-guarantees the desired mutual exclusivity property between the binder and the resources:
+guarantees the desired mutual exclusion property between the binder and the resources:
 \[
@@
 %
-Recursive let bindings also bind expressions lazily, so, similarly, we
-introduce a $\D$-variable for each binding. Resources required to type the
+Recursive let bindings also bind expressions lazily and so
+introduce a $\D$-bound variable for each binding. Resources required to type the
 let-bindings are still available in the body of the let, to later be consumed
-via $\D$-variables, or, directly, if the let-bindings are unused.
+via $\D$-bound variables, or, directly, if the let-bindings are unused.
 %
@@
 that every binder in such a group must be typed with the same linear context,
-as observed in \ref{sec:semantic-linearity-examples:recursive-lets}.
-Consequently, all bindings of the recursive group are introduced as $\D$-vars
-which all share the same $\D$ environment -- using any one of the bindings in a
+as observed in Section~\ref{sec:semantic-linearity-examples:recursive-lets}.
+Consequently, all bindings of the recursive group are introduced as
+$\D$-bound variables
+sharing the same usage environment -- using any one of the bindings in a
 recursive group entails consuming all resources required to type that same
-group -- so, at most a single binder from the group can be used in the let-continuation.
+group -- so, at most a single binder from the group can be used in the
+let body.
 %\todo{should we say how this means at most once of the binders from the group can be used?} OK
@@
 Accordingly, these sub-expressions might still depend on linear resources to be
-well-typed (such resources will only be consumed when the sub-expressions themselves are evaluated).
+well-typed and such resources will only be consumed when the sub-expressions themselves are evaluated.
 
@@
 the main typing judgement and trivially ``outputs'' the whole $\D$ environment,
-as there is always only a single sub-expression in lambda abstractions (the body).
+as there is always only a single sub-expression in lambda abstractions.
 
@@
 We highlight that, when evaluating a case expression, computation only
-happens when a scrutinee (not in WHNF) is evaluated to
+happens when a scrutinee that is not in WHNF is evaluated to
 WHNF. Otherwise, evaluation continues in the alternative by
@@
 Therefore, resources used to type a scrutinee not in
-WHNF will be consumed when the case is evaluated and no longer be directly available in the case
+WHNF will be consumed when the case is evaluated and will no longer be directly available in the case
 alternatives. Conversely, when the scrutinee is already in WHNF, linear

### 2025-02-25 icfp.tex
big pass, up to 4

@@
 Alternatives may also use the case binder or pattern variables, referring to, respectively,
-the whole scrutinee (and all resources used to type the scrutinee) or
-constructor arguments (and the resources to type each argument).
+the whole scrutinee and all its used resources; or
+constructor arguments and the resources used to type them.
 % Using the case binder entails using all the resources required by the scrutinee, and using a pattern
@@
 %
-Recall how $\D$-variables can encode mutual exclusivity between alternative
+Recall how $\D$-bound variables encode mutual exclusion between alternative
 ways of consuming resources -- it follows that case binders and pattern-bound
@@
 %and the name of the case binder $z$.
-
-%
-Despite the main idea for typing ``WHNF case expressions'' being conveyed by
-this rule, in our system it is slightly more involved, as will be seen after
+%
+%
+Although the main idea for typing ``WHNF case expressions'' is conveyed by
+this rule, the full rule of our system is slightly more involved, as will be seen after
 discussing cases of scrutinees not in WHNF.
@@
 \vp$ is used to type case alternatives, encompassing three
-``modes'' distinguished by the kind of arrow that is used:
+``modes'' that are distinguished by the kind of arrow that is used:
 for alternatives of case expressions whose scrutinee is in WHNF
@@
 
-Second, the rule for alternatives that match on the wildcard pattern, be the
-original scrutinee in WHNF or not (note the use of $\Rightarrow$, which
+Second, the rule for alternatives that match on the wildcard pattern,
+regardless of the WHNF status of the scrutinee (note the use of $\Rightarrow$, which
 is applicable both under $\Rightarrow_{WHNF}$ and $\Rightarrow_{NWHNF}$):
@@
 To type a wildcard alternative we simply type the expression with the main
-judgement, ignoring all annotations on the judgement. The case binder will
+judgement, ignoring all annotations. The case binder will
 have already been introduced in the environment with the appropriate usage
-environment by the relevant (WHNF or not) case expression rule.
+environment by the relevant %(WHNF or not)
+case expression rule.
 
@@
 (see the latter below), must be reactively destroyed from branches where the
-pattern is unrestricted. In failing to do this, we'd allow the fully-consumed
-resources to be used again in the branch!
+pattern is unrestricted. In failing to do this, we would allow the fully-consumed
+resources to be used again in the branch.
 % Really, in the WHNF case it is not really necessary to remove the things
@@
 The $Alt0$ rule essentially encodes this insight, and is applicable regardless
-of the WHNF-ness of the scrutinee ($\Rightarrow$ mode), as long as
+of the WHNF status of the scrutinee ($\Rightarrow$ mode), as long as
 the constructor pattern has no linear fields:
@@
 
-    \item Consequently, cases with non-WHNF scrutinees must by typed such that no resource
-    from the scrutinee can be used directly in the branch again, or risk
+    \item Consequently, cases with non-WHNF scrutinees must be typed such that no resource
+    from the scrutinee can be used directly in the branch again, or we
+    risk
     consuming the same resource twice. In the example, $x$ must not directly occur
@@
     \item On the other hand, since evaluation to WHNF does not guarantee all
-    resources do end up being consumed (unless the result matches an
-    unrestricted pattern), we must ensure any which transfer from the
-    scrutinee to the alternative are ultimately consumed. (In the example,
+    resources are consumed (unless the result matches an
+    unrestricted pattern), we must ensure that any resources that transfer from the
+    scrutinee to the alternative are ultimately consumed (in the example,
     $x$ is not consumed by evaluating the scrutinee).
@@
     Considering (1), we must resort to a more uniform method of guaranteeing we
-    ``finish consuming`` the scrutinee. And there are only two ways of
+    ``finish consuming'' the scrutinee. There are only two ways of
     uniformly referring to the remaining linear resources in the newly-evaluated scrutinee:
@@
         \begin{itemize}
-            \item The case binder, which refers to the whole result of evaluating the scrutinee
+            \item The case binder, which refers to the whole result of evaluating the scrutinee;
             \item The linear components of a constructor pattern, which refer to expressions that may contain linear resources.
@@
 Hence, to type a case expression whose scrutinee is not in WHNF, we
-type the scrutinee with linear resources $\D$ and type the case
+type the scrutinee with linear resources $\D$ and the case
 alternatives by introducing the case binder with a usage environment $[\D]$,
@@
 Alternatives not matching the scrutinee could use resources arbitrarily as they
-will never be executed, however, we uniformly treat non-matching alternatives
+will never be executed. We uniformly treat non-matching alternatives
 as if the scrutinee were not in WHNF. Having introduced proof irrelevant
-resources, we can now present the final $Case_\textrm{WHNF}$ rule:
+resources, we can now present the full $Case_\textrm{WHNF}$ rule:
 \[
@@
 %
-We introduce \emph{tagged resources} as a means to enforce all pattern-bound
+We introduce \emph{tagged resources} as a means to enforce that all pattern-bound
 variables for a scrutinee not in WHNF are either \emph{jointly} used to consume
@@
 Having uniquely tagged the resources in the usage environment of each pattern
-variable, we need now only to express that (1) the pattern-bound $\D$-variables can
+variable, we need only to express that (1) the pattern-bound $\D$-variables can
 be used (i.e., the tagged resources need be available in the linear
 environment in order for the $\textsc{Var}_\D$ rule to be applicable) and (2)
-if a $K$-pattern-bound $\D$-var is consumed, all remaining $K$-$\D$-vars,
+if a $K$-pattern-bound $\D$-variable is consumed, all remaining $K$-$\D$-variables,
 standing for the remaining linear components of the same pattern, must also be
@@
 
-To satisfy these two properties, we introduce a $\textit{Split}$ rule that
+To satisfy these two constraints, we introduce a $\textit{Split}$ rule that
 allows a linear resource $x{:}_1\s$ to be split into $n$ resources at a given
@@
 $\textsc{Split}$ in order to use pattern variables at all.  Moreover, the
-remaining tagged resources cannot be used directly, yet need still be consumed
+remaining tagged resources cannot be used directly, yet need to still be consumed
 exactly once. Thus, the choice to consume the tagged resources via the usage

### 2025-02-25 icfp.tex
big pass, up to 5

@@
 are transferred into the branch continuation, e.g. as in $\ccase{f\ x}{z\ \{K\
-a \rightarrow e\}}$ where $f = K$, but opaque.
+a \rightarrow e\}}$ where $f$ is defined as $f = K$:
 \begin{enumerate}
 
-    \item Consequently, cases with non-WHNF scrutinees must be typed such that no resource
+    \item Cases with non-WHNF scrutinees must be typed such that no resource
     from the scrutinee can be used directly in the branch again, or we
@@
 
-We introduce \emph{proof irrelevant} resources, written as linear resources
-within square brackets $[ x{:}\s]$ and lifted to contexts as $[\D]$,
-to encode linear resources that cannot be
-directly used (i.e.,~to which the $\textsc{Var}$ rule is not applicable). Proof irrelevant resources
-are linear resources in all other senses, meaning they must be used
+To encode  linear resources that cannot be
+directly used (i.e.,~to which the $\textsc{Var}$ rule is not
+applicable) we introduce \emph{proof irrelevant} resources, written as linear resources
+within square brackets $[ x{:}\s]$ and lifted to contexts as $[\D]$.
+Proof irrelevant resources
+are linear resources in every sense, meaning they must be used
 exactly once. However, since proof irrelevant resources cannot be
-discarded nor used directly, they have to be consumed \emph{indirectly} via
+discarded or used directly, they have to be consumed \emph{indirectly} via
 $\D$-bound variables, namely, the case binder, or, in mutual exclusion, the linear pattern-bound
@@
 
-We begin by presenting the operational semantics of $\lambda^\pi_\D$,
+We develop the metatheory of Linear Core by first 
+presenting the operational semantics of $\lambda^\pi_\D$,
 consisting of a lazy natural semantics in the
@@
 This semantics conveys the expected behavior of an implementation of
-lazy evaluation and is agnostic to any linearity) information.
-Thus, such a semantics is ill-suited for meta-reasoning about
-linearity and so we develop an \emph{instrumented} semantics that is
+lazy evaluation and is agnostic to any linearity information.
+Such a semantics is ill-suited for reasoning about
+linearity and so we develop an \emph{instrumented} linearity-aware semantics that is
 enriched with sufficient information to allow us to establish type
@@
 %Both semantics are
-The natural semantics is equipped with an environment at assigns values to variables and is
+The natural semantics is equipped with an evaluation environment that
+maps variables to expressions and is
 mutated in order to express evaluation. Terms are explicit about
@@
     \Theta' : e'[x/y] \Downarrow \Theta'' : v }
-    {\Gamma : e~x \Downarrow \Theta'' : v}\\[1.5em]
+    {\Theta : e~x \Downarrow \Theta'' : v}\\[1.5em]
     \infer[]
-    {\Theta : e \Downarrow \Theta' : v}
+    {(\Theta,  x:_\omega \sigma = e) : e \Downarrow \Theta' : v}
     {(\Theta , x:_\omega \sigma = e) : x \Downarrow (\Theta',x:_\omega
@@
 $\Theta : e \Downarrow \Theta' : v$ where $e$ is an expression, $v$ a
-value, $\Theta$ and $\Theta'$ are environments with bindings of the
+value, $\Theta$ and $\Theta'$ are evaluation environments with bindings of the
 form $x:_\omega\sigma = e$ which assigns the expression $e$ to the
@@
 assumptions);
-$\Theta$ and $\Theta'$ are runtime environments, consisting of bindings of the form
+$\Theta$ and $\Theta'$ are evaluation environments, consisting of bindings of the form
 $x :_\pi \sigma = e$; $e$ is the expression being evaluated and $v$
@@
 of partial derivations and partial evaluation (see
-Appendix~\ref{app:opsem}), written $a\Downarrow^* b$ which allows us
-to reason about progress.
+Appendix~\ref{app:opsem}), written $a\Downarrow^* b$, which allows us
+to state and reason about a progress property.
 
@@
 
-The $\lambda^\pi_\D$ type system is sound: well-typed programs in Linear Core do not
-get \emph{stuck} and their evaluation preserves linearity.
+Our results rely on a series of lemmas relating linear, unrestricted and
+$\Delta$-bound variables, as well as a lemma characterizing irrelevant
+resources.
 %
@@
 %
-As discussed, proof irrelevant resources can only be consumed
+Proof irrelevant resources can only be consumed
 indirectly, essentially encoding that the (non-WHNF) case scrutinee
@@
 %\subsection{Type safety\label{sec:type-safety-meta}}
-
+% The $\lambda^\pi_\D$ type system is sound: well-typed programs in Linear Core do not
+% get \emph{stuck} and their evaluation preserves linearity.
 We prove type safety of the Linear Core system via the standard type

### 2025-02-25 icfp.tex
big pass, up to 6

@@
 %
-This implementation further substantiates our claim that Linear Core is
+Our implementation further substantiates our claim that Linear Core is
 suitable for the intermediate language of an optimising compiler.
@@
 %
-Pattern variables bound in a case alternative, for a scrutinee not in WHNF, are
+For a scrutinee not in WHNF, pattern variables in a case alternative are
 introduced as $\D$-variables with usage environment $\lctag{\irr{\D}}{K_i}$,
@@
 The results of running the Linear Core GHC plugin on large 
-libraries focused around linear types are given by
+libraries focused around linear types are reported in
 Figure~\ref{fig:core-plugin-res}.
@@
 %
-The results indicate our mostly direct implementation of Linear Core is
+The results indicate that our mostly direct implementation of Linear Core is
 successful in accepting the vast majority of the thousands of intermediate
@@
 linear types.
-
+%
 However, Linear Core does not accept every single program produced by the

### 2025-02-26 icfp.tex
reverse binder swap muito bonito; sec 4 pass

@@
 are transferred into the branch continuation, e.g. as in $\ccase{f\ x}{z\ \{K\
-a \rightarrow e\}}$ where $f$ is defined as $f = K$:
+a \rightarrow e\}}$:
 \begin{enumerate}
@@
     unrestricted pattern), we must ensure that any resources that transfer from the
-    scrutinee to the alternative are ultimately consumed (in the example,
-    $x$ is not consumed by evaluating the scrutinee).
+    scrutinee to the alternative are ultimately consumed (in the example, $x$
+    is not consumed by evaluating the scrutinee for $f = K$).
 
@@
 enriched with sufficient information to allow us to establish type
-safety, showing that linearity is preserved. We show that the natural
+safety, thus showing that linearity is preserved. We show that the natural
 and instrumented semantics are bisimilar for well-typed terms and so
@@
 (Section~\ref{sec:typesafe}). Finally, we show that multiple optimising
-transformations such as those performed by GHC on (Linear) Core are
+transformations such as those performed by GHC on Linear Core are
 type preserving
@@
 maps variables to expressions and is
-mutated in order to express evaluation. Terms are explicit about
+mutated in order to express evaluation\todo{aqui não seria sharing?}. Terms are explicit about
 their sharing (in the sense of shared reductions in lazy evaluation),
@@
 $\Theta : e \Downarrow \Theta' : v$ where $e$ is an expression, $v$ a
-value, $\Theta$ and $\Theta'$ are evaluation environments with bindings of the
-form $x:_\omega\sigma = e$ which assigns the expression $e$ to the
-variable $x$ of the given type. The rules are standard, augmenting the
-runtime environment with let-bound expressions which are evaluated
-(only once) if the let-bound variable is forced.
+value, and $\Theta$ and $\Theta'$ are evaluation environments with bindings of
+the form $x:_\omega\sigma = e$ which assigns the expression $e$ to the variable
+$x$ of the given type $\s$. The rules are standard, augmenting the runtime
+environment with let-bound expressions which are evaluated (only once) if the
+corresponding let-bound variable is forced.
 
@@
 We follow an approach
-similar to that of \cite{cite:linearhaskell}, defining a
+similar to that of \cite{cite:linearhaskell}, defining
 an \emph{instrumented} operational semantics in which linear variables are
@@
 
-The semantics is presented in
+The instrumented semantics is presented in
 Figure~\ref{fig:linopsem}, defining the judgment $\Gamma ; \Delta \vdash (\Theta \mid e) \Downarrow (\Theta'
@@
 $\Theta$ and $\Theta'$ are evaluation environments, consisting of bindings of the form
-$x :_\pi \sigma = e$; $e$ is the expression being evaluated and $v$
+$x :_\pi \sigma = e$ or $x :_\Delta \sigma = e$; $e$ is the expression being evaluated and $v$
 its resulting value, both of type $\sigma$; $\Sigma$ is a list of assignments of the form
-$e :_\pi \tau$ or $x :_\Delta \tau$ which are expressions in which
+$e :_\pi \tau$ or $x :_\Delta \tau$\todo{isto não me fez sentido} which are expressions in which
 bindings in $\Theta$ may also be used. This additional data is needed to
@@
 Appendix~\ref{app:opsem}), written $a\Downarrow^* b$, which allows us
-to state and reason about a progress property.
+to state and reason about a progress property.\todo{Na semantica O omega e 1 ainda estão separados, é assim?}
 
@@
 
-\todo{UPDATE}
-Finally, we discuss the special case of scrutinizing a
-single variable $x$, e.g.:
-$
-\lambda x.~\ccase{x}{\_ \to x}
-$. 
-It might seem as though this function is linear:
-%begin{itemize}
-% \item
-if the argument instantiating $x$ is in WHNF, then scrutinizing it is a no-op, and returning $x$
-just returns the resource intact;
-% \item
-if the argument instantiating $x$ is not in WHNF, then scrutinizing it evaluates it to WHNF, and returning
-$x$ returns the result of evaluating $x$ that still had to be consumed exactly
-once.
-%\end{itemize}
-However, in practice, this program's linearity can only be determined by
-further analysis on the non-strict evaluation strategy. If linear
-function applications are $\beta$-reduced \emph{call-by-name}
-(often desirable, as linear functions are known to use their argument exactly
-once)
-% and thus call-by-name reduction will definitely not duplicate computation
-while the example function is considered linear, then such an application might
-duplicate linear resources during evaluation. For instance:
-\[
-\begin{array}{l}
-(\lambda x.~\ccase{x}{\_ \to x})~(free~x)
-\Longrightarrow_{CBN}
-\ccase{free~x}{\_ \to free~x}
-\end{array}
-\]
-On the other hand, if linear function applications are reduced
-\emph{call-by-need}, the above example would instead be semantically linear
-because the result of freeing $x$ would be bound to some $y$ first.
-%
-The system we present in the next section does accept this function
-scrutinizing a variable as linear, as our operational semantics are
-call-by-need.
-This interaction between evaluation and linearity is considered specifically in the
-reverse binder swap transformation discussed in
-Section~\ref{sec:optimisations-preserve-types-meta}.\todo{Figure out precisely reverse binder swap proof and reevaluate}
+%Finally, we discuss the special case of scrutinizing a
+%single variable $x$, e.g.:
+%$
+%\lambda x.~\ccase{x}{\_ \to x}
+%$. 
+%It might seem as though this function is linear:
+%%begin{itemize}
+%% \item
+%if the argument instantiating $x$ is in WHNF, then scrutinizing it is a no-op, and returning $x$
+%just returns the resource intact;
+%% \item
+%if the argument instantiating $x$ is not in WHNF, then scrutinizing it evaluates it to WHNF, and returning
+%$x$ returns the result of evaluating $x$ that still had to be consumed exactly
+%once.
+%%\end{itemize}
+%However, in practice, this program's linearity can only be determined by
+%further analysis on the non-strict evaluation strategy. If linear
+%function applications are $\beta$-reduced \emph{call-by-name}
+%(often desirable, as linear functions are known to use their argument exactly
+%once)
+%% and thus call-by-name reduction will definitely not duplicate computation
+%while the example function is considered linear, then such an application might
+%duplicate linear resources during evaluation. For instance:
+%\[
+%\begin{array}{l}
+%(\lambda x.~\ccase{x}{\_ \to x})~(free~x)
+%\Longrightarrow_{CBN}
+%\ccase{free~x}{\_ \to free~x}
+%\end{array}
+%\]
+%On the other hand, if linear function applications are reduced
+%\emph{call-by-need}, the above example would instead be semantically linear
+%because the result of freeing $x$ would be bound to some $y$ first.
+%%
+%The system we present in the next section does accept this function
+%scrutinizing a variable as linear, as our operational semantics are
+%call-by-need.
+%This interaction between evaluation and linearity is considered specifically in the
+%reverse binder swap transformation discussed in
+%Section~\ref{sec:optimisations-preserve-types-meta}.\todo{Figure out precisely reverse binder swap proof and reevaluate}
+%
 
@@
 By using the scrutinee $x$ instead of the case binder, we might be able to
-float out expressions from the alternative using the case binder. For example,
-we might float an expensive computation involving $z$ out of the case
-alternative, where $z$ is now out of scope:
-
-{\small\[
-\begin{array}{l}
-(\lambda x.~\lletrec{go~y = \ccase{x}{z~\{(a,b) \to \ E[expensive~z] \} }}{\dots})
-\Longrightarrow_\textrm{Reverse binder swap}\\
-(\lambda x.~\lletrec{go~y = \ccase{x}{z~\{(a,b) \to E[expensive~x] \} }}{\dots})
-\Longrightarrow_\textrm{Commuting case-let}\\
-\lambda x.~\llet{t = \mathit{expensive}~x}{(\lletrec{go~y = \ccase{x}{z~\{(a,b) \to E[t] \} }}{\dots})}\\
-\end{array}
-\]}
-
-\noindent If $go$ is compiled into a loop, we only compute
-$\mathit{expensive}~x$ once instead of in every loop iteration.
-
+float out expressions from the alternative using the case binder:
+\[
+    \ccase{x}{z~\{\rho \rightarrow e\}} \Rightarrow \ccase{x}{z~\{\rho \rightarrow e[x/z]\}}
+\]
+
+%For example, we might float an expensive computation involving $z$ out of the
+%case alternative, where $z$ is now out of scope:
+%
+%{\small\[
+%\begin{array}{l}
+%(\lambda x.~\lletrec{go~y = \ccase{x}{z~\{(a,b) \to \ E[expensive~z] \} }}{\dots})
+%\Longrightarrow_\textrm{Reverse binder swap}\\
+%(\lambda x.~\lletrec{go~y = \ccase{x}{z~\{(a,b) \to E[expensive~x] \} }}{\dots})
+%\Longrightarrow_\textrm{Commuting case-let}\\
+%\lambda x.~\llet{t = \mathit{expensive}~x}{(\lletrec{go~y = \ccase{x}{z~\{(a,b) \to E[t] \} }}{\dots})}\\
+%\end{array}
+%\]}
+%
+%\noindent If $go$ is compiled into a loop, we only compute
+%$\mathit{expensive}~x$ once instead of in every loop iteration.
+%
 Even though GHC applies the reverse binder swap transformation in the
-Core-to-Core passes, \emph{this optimisation violates linearity} when considered
-as a transformation \emph{on Linear Core programs} because the result (second program
-above) is ill-typed -- the resource $x$ of the scrutinee not in WHNF is used
-again in an alternative, despite it no longer being available according to our system.
-
-An expression in which a variable $x$ occurs both in the case scrutinee and in
-the alternatives (e.g.,~$\ccase{x}{\_ \to x}$) could be considered linear if
-considered in isolation:
+Core-to-Core passes, this optimisation violates linearity in Linear Core
+because the transformed program is rejected -- the scrutinee $x$, a single
+variable, is not considered to be in WHNF by our system, thus occurrences of
+$x$ in any alternative are ill-typed in it. This is a conservative choice in $\lambda^\pi_\Delta$ for
+typing the subtle case where the scrutinee is a single variable:
+
+Consider $\ccase{x}{\_ \to x}$. Perhaps surprisingly, this expression can be
+considered semantically linear in a call-by-need calculus but violates
+linearity in a call-by-name one:
+%
+%An expression in which a variable $x$ occurs both in the case scrutinee and in
+%the alternatives (e.g.,~$\ccase{x}{\_ \to x}$) could be considered linear if
+%considered in isolation:
 % \[
@@
 % \item
-If $x$ refers to an unevaluated expression, then scrutinizing will force the expression
-bound by $x$ to be evaluated to WHNF. In a subsequent use of
-$x$ in the alternatives, $x$ refers to the evaluated scrutinee in WHNF, which
-must be consumed. Since $x$ is just another name (like the case binder) for the
-scrutinee in WHNF, we may use it instead of the case binder or pattern
-variables.
-%
-On the other hand, if $x$ already refers to an expression evaluated to WHNF, then
-scrutinizing it in a case expression is a no-op, thus we may use it again (in
-mutual exclusion with the case binder and pattern variables).
-
-However, allowing this expression to be linear makes $\lambda^\pi_\Delta$ unsound!
-Consider the function application $(\lambda x.~\ccase{x}{\_ \to x})~(use~y)$
-where $y$ is a free linear variable, and assume $\lambda x.~\ccase{x}{\_ \to x}$
-to be a linear function. According to $\lambda^\pi_\Delta$'s call-by-name
-$\beta$-reduction, we have:
+Under call-by-need, if $x$ refers to an unevaluated expression, scrutinizing it forces the expression
+bound by $x$ to WHNF. In a subsequent use of
+$x$ in an alternative, $x$ refers to the already evaluated scrutinee in WHNF.
+Since $x$ is just another name (like the case binder) for the scrutinee in
+WHNF, its use is also valid in the alternatives.
+%
+Dually, if $x$ refers to an expression already in WHNF, no evaluation takes
+place and $x$ is still just another name for the scrutinee in WHNF.
+
+However, under call-by-name, evaluation of $x$ is not shared:
+%However, allowing this expression to be linear makes $\lambda^\pi_\Delta$ unsound!
+%Consider the function application $(\lambda x.~\ccase{x}{\_ \to x})~(use~y)$
+%where $y$ is a free linear variable, and assume $\lambda x.~\ccase{x}{\_ \to x}$
+%to be a linear function. According to $\lambda^\pi_\Delta$'s call-by-name
+%$\beta$-reduction, we have:
 \[
@@
 \]
-After the reduction, $y$ is a linear variable \emph{consumed} in the scrutinee
-of the case that also occurs in the case alternative -- the resource
-$y$ is being effectively duplicated.
-%
-If the application were instead reduced with call-by-need
+After the reduction, $y$ is a linear variable consumed both in the scrutinee
+and in the case alternative -- $y$ has been duplicated!
+%
+If the application is instead reduced call-by-need, $y$ is not duplicated:
 \[
@@
 \]
-then the computation using $y$ would be let-bound, and $y$ would be used as a
-scrutinee variable, which preserves semantic linearity.
-
-For an optimising compiler such as GHC, the reverse binder swap transformation
-violating linearity in Linear Core is relevant because optimising
-compilers do not necessarily allocate a binding/use call-by-need for
-every function application -- our result shows that failing to do so
-when calling a function such as the one above is \emph{unsound}.
-%
-In an optimising compiler, a binding may be deemed unnecessary if the function
-argument is inferred to be used exactly once. Linearity can play a
-(non-heuristic) role in this inference, for a linear function uses its
-argument exactly once by definition.
+%then the computation using $y$ would be let-bound, and $y$ would be used as a
+%scrutinee variable, which preserves semantic linearity.
+
+In an optimising compiler such as GHC, there is no definitive evaluation
+strategy. It is mostly up to heuristics to determine what to inline
+\emph{à la} call-by-name and what to create let-bindings for \emph{à
+la} call-by-need.
+%
+In this flexible evaluation setting, we highlight the reverse binder swap is
+only a linearity-preserving transformation as long as a linear variable with
+more than one occurrence is never substituted for an expression with free
+linear variables.
+%
+%This is unfortunately at odds with
+%Moreover, allowing the reverse binder swap also challenges the 
+%
+%It additionally challenges the non-heuristic property that linear variables can
+%be unconditionally inlined.......
+
+%the aggressive inlining passes blur evaluation
+%there is no definitive evaluation
+%strategy in the sense functional compilers use both when agressively inline
+%expressions \emph{à la} call-by-name.
+%call-by-need for
+%every function application -- our result shows that failing to do so
+%when calling a function such as the one above is \emph{unsound}.
+%
+%In an optimising compiler, a binding may be deemed unnecessary if the function
+%argument is inferred to be used exactly once. Linearity can play a
+%(non-heuristic) role in this inference, for a linear function uses its
+%argument exactly once by definition.
 %
@@
 %
-However, it is incompatible to both type expressions like $(\lambda
-x.~\ccase{x}{\_ \to x})$ as linear functions and consider
-call-by-name $\beta$-reduction of linear functions unconditionally
-safe (without risk of duplicating resources nor work).
-%
-Furthermore, considering such an use of $x$ to be linear also makes inlining of
-linear let bindings no longer unconditionally safe.
+%However, it is incompatible to both type expressions like $(\lambda
+%x.~\ccase{x}{\_ \to x})$ as linear functions and consider
+%call-by-name $\beta$-reduction of linear functions unconditionally
+%safe (without risk of duplicating resources nor work).
+%
+%Furthermore, considering such an use of $x$ to be linear also makes inlining of
+%linear let bindings no longer unconditionally safe.
 

### 2025-02-26 icfp.tex
pass 5

@@
     \item Programs we expected GHC to produce, but know are not seen as linear by $\lambda^\pi_\Delta$.
-        Namely, programs which scrutinize a variable, but then use the variable in the
-        case alternatives of the case scrutinizing it. As discussed in
-        Section~\ref{sec:reverse-binder-swap-considered-harmful}, these programs can be
-        understood as linear as long as applications of linear functions binding these
-        variables are reduced \emph{call-by-need}, which is not the case in our system.
+        Namely, programs resulting from the reverse binder swap which
+        scrutinize a variable and then use it again in the case alternatives.
+        As discussed in Section~\ref{sec:reverse-binder-swap-considered-harmful}, these
+        programs could be understood as linear as long as all applications of
+        linear functions are reduced \emph{call-by-need}, but our system
+        conservatively rejects it in light of the flexible evaluation strategy
+        used by optimizing compilers.
 
@@
 
-    \item Programs which \emph{actually violate linearity}, in that the term
+    \item Programs which actually violate linearity, in that the term
         violates the linear type it is annotated with. These are the programs
-        we ultimately want to be able to reject by having a linearly typed Core
+        we ultimately want to spot and reject by having a linearly typed Core
         validate intermediate programs. We found both a program which outright
-        discarded a linear resource, and that rewrite rules implementing stream
+        discarded a linear resource and that rewrite rules implementing stream
         fusion~\cite{10.1145/165180.165214,10.1145/1291151.1291199}
-        (among others), could produce linearly-invalid programs (in one such case,
+        (among other rules) could produce linearly-invalid programs (in one such case,
         the type of \lstinline{build} was not general enough to accommodate linearity).
@@
 besides validating that our implementation is faithful to the
-$\lambda^\pi_\Delta$ system of insofar as programs that should not be accepted
+$\lambda^\pi_\Delta$ system insofar as programs that should not be accepted
 are deemed ill-typed, provide further insight into the remaining

### 2025-02-26 icfp.tex
minor pass 6

@@
 with linearity/multiplicity annotations. Core's type system is unable to type
-\emph{semantic} linearity of programs in the sense elaborated in our work,
+semantic linearity of programs in the sense elaborated in our work,
 which results in Core rejecting most linear programs resulting from optimising
@@
 linear resources are still used exactly once throughout the optimising
-transformations.
+transformations.\todo{add part about operational semantics and proof being similar to them}
 
@@
 diverges significantly in rules for typing other constructs, notably, case expressions and
-case alternatives. Furthermore, unlike Mini-Core, we prove type safety of our
-system and that multiple optimising transformations, when applied to Linear
-Core programs, preserve linearity as understood by the system.
+case alternatives. Furthermore, unlike Mini-Core, we prove Linear Core is type
+safe, guarantees linear resource usage, and that multiple optimising
+transformations, when applied to Linear Core programs, preserve linearity.
 

### 2025-02-26 icfp.tex
appendix shuffle

@@
 maps variables to expressions and is
-mutated in order to express evaluation\todo{aqui não seria sharing?}. Terms are explicit about
+mutated in order to express shared evaluation. Terms are explicit about
 their sharing (in the sense of shared reductions in lazy evaluation),
@@
 its resulting value, both of type $\sigma$; $\Sigma$ is a list of assignments of the form
-$e :_\pi \tau$ or $x :_\Delta \tau$\todo{isto não me fez sentido} which are expressions in which
+$e : \tau$ which are expressions in which
 bindings in $\Theta$ may also be used. This additional data is needed to
@@
     {\Gamma ; \Delta  \vdash (\Theta  , x:_{\Delta'} \sigma = e\mid e)
-    \Downarrow (\Theta' \mid v) : \sigma , \Sigma}
+    \Downarrow (\Theta' \mid v) : \sigma , \Sigma \and \D'' = \D' \setminus (\Theta \setminus \Theta')}
     {\Gamma ; \Delta \vdash (\Theta , x:_{\Delta'} \sigma = e \mid x)
-      \Downarrow (\Theta' , x:_{\Delta''} \sigma = v \mid v) : \sigma, \Sigma}
+      \Downarrow (\Theta' , x:_{\Delta''} \sigma = v \mid v) : \sigma,
+    \Sigma }
     \\[1.5em]
@@
     % \Downarrow \Theta_n : v_n \quad
-    \Gamma ; \Delta \vdash (\Theta, \overline{x_i:_\omega \sigma_i =
+    \Gamma ; \Delta \vdash (\Theta, \overline{x_i:_{\Delta'} \sigma_i =
     e_i} \mid e' )\Downarrow (\Theta'
@@
     }
-    {\Gamma ;\Delta \vdash (\Theta \mid \lletrec{\overline{x_i{:}_{\Delta}\sigma_i = e_i}}{e'})
+    {\Gamma ;\Delta \vdash (\Theta \mid \lletrec{\overline{x_i{:}_{\Delta'}\sigma_i = e_i}}{e'})
     \Downarrow (\Theta' \mid v) : \tau}
@@
 $x$ in any alternative are ill-typed in it. This is a conservative choice in $\lambda^\pi_\Delta$ for
-typing the subtle case where the scrutinee is a single variable:
+typing the subtle case where the scrutinee is a single variable.
 
 Consider $\ccase{x}{\_ \to x}$. Perhaps surprisingly, this expression can be
-considered semantically linear in a call-by-need calculus but violates
-linearity in a call-by-name one:
+considered semantically linear under call-by-need but violates
+linearity under call-by-name:
 %
@@
 
-\section{Type Safety Proofs}\label{app:proofs}
-
-In the following we write $\Mapsto$ to stand for
-$\Rightarrow_\mathsf{WHNF}$ and $\Rrightarrow$ for $\Rightarrow_\mathsf{NWHNF}$.
-
-\subsection{Auxiliary Lemmas}
-The following lemmas presuppose the variable names across the typing
-contexts are distinct.
-
-  \begin{lemma}[Linear to $\Delta$-bound]~\label{lem:onedelta-app}
-
-    \begin{enumerate}
-
-      \item If $\G; \D',\xl \vdash e :\vp$
-        then $\G[\D/x],\xD; \D,\D' \vdash e : \vp$ (with $\Delta$ fresh).
-      \item If $\G ; \D' , \xl \vdash_{alt} \rho \rightarrow e
-        :^z_{\Delta_s} \sigma \Rightarrow \rho$ then
-        $\G[\D/x],\xD;\D,\D' \vdash_{alt} \rho \rightarrow e
-        :^z_{\Delta_s} \sigma \Rightarrow \rho$ (with $\Delta$ fresh).
-
-      \end{enumerate}
-\end{lemma}
-\begin{proof}
-Straightforward induction on the structure of the given derivation,
-reconstructing the original derivation and applying rule
-$\mathit{Var}_\Delta$ where rule $\mathit{Var}_1$ was applied on
-$x$ in the original derivation.
-\end{proof}
-\onedelta*
-
-\begin{proof}
-By Lemma~\ref{lem:onedelta-app}(1).
-\end{proof}
-
-\begin{lemma}[$\Delta$-bound to Linear]~\label{lem:deltaone-app}
-
-  \begin{enumerate}
- \item If $\G,\x[\irr{\D}]; \irr{\D},\D' \vdash e : \vp$
-   then $\G[x/\irr{\D}]; \D',\xl \vdash e :\vp$.
- \item If $\G ,\x[\irr{\D}]; \irr{\D}, \D' \vdash_{alt} \rho \rightarrow e
-   :^z_{\Delta_s} \sigma \Rightarrow \rho$ then
-   $\G[x/\irr{\D}];\D',\xl \vdash_{alt} \rho \rightarrow e
-   :^z_{\Delta_s} \sigma \Rightarrow \rho$.
-  \end{enumerate}
-\end{lemma}
-\begin{proof}
-Straightforward induction on the structure of the given derivation,
-reconstructing the original derivation and applying rule
-$\mathit{Var}_1$ where rule $\mathit{Var}_\D$ was applied on
-$x$ in the original derivation.
-\end{proof}
-
-\deltaone*
-\begin{proof}
-By Lemma~\ref{lem:deltaone-app}(1).
-\end{proof}
-
-\begin{lemma}[Unrestricted and $\Delta$-bound]~\label{lem:undelta-app}
-  \begin{enumerate}
-    \item $\G,\xo; \D \vdash e : \vp$ iff $\G,\x[\cdot]; \D \vdash e :
-      \vp$
-    \item $\G,\xo;\D \vdash_{alt} \rho \rightarrow e
-      :^z_{\Delta_s} \sigma \Rightarrow \rho$ iff
-      $\G,\x[\cdot];\D \vdash_{alt} \rho \rightarrow e
-   :^z_{\Delta_s} \sigma \Rightarrow \rho$
-  \end{enumerate}
-\end{lemma}
-\begin{proof}
-Straightforward induction on the structure of the given derivation,
-noting that rules $\mathit{Var}_\Delta$ and $\mathit{Var}_\omega$
-coincide when $\Delta$ is empty.
-\end{proof}
-
-\undelta*
-
-\begin{proof}
-By Lemma~\ref{lem:undelta-app}.
-\end{proof}
-
-\subsection{Irrelevance\label{sec:proof:irrelevance}}
-
-\input{language-v4/proofs/WHNFConvSoundness}
-
-\subsection{Substitution Lemmas\label{sec:proof:substitution-lemmas}}
-
-\input{language-v4/proofs/LinearSubstitutionLemma}
-
-\input{language-v4/proofs/UnrestrictedSubstitutionLemma}
-
-\input{language-v4/proofs/DeltaSubstitutionLemma}
-
-\subsection{Type Preservation\label{sec:proof:type-preservation}}
-
-%\input{language-v4/proofs/TypePreservationTheorem}
-
-\subsection{Progress\label{sec:proof:progress}}
-
-%\input{language-v4/proofs/ProgressTheorem}
-
-\section{Optimisations preserve linearity}\label{app:optimisations}
-
-\subsection{Inlining}
-\input{language-v4/proofs/optimizations/Inlining}
-
-\subsection{\texorpdfstring{$\beta$}{Beta}-reduction}
-
-\input{language-v4/proofs/optimizations/BetaReduction}
-
-\subsection{Case of known constructor}
-
-\input{language-v4/proofs/optimizations/CaseOfKnownConstructor}
-
-\subsection{Case of Case\label{sec:proof:caseofcase}}
-
-The case of case transformation applies to case expressions whose scrutinee is
-another case expression, and returns the innermost case expression transformed by repeating
-the outermost case expression in each alternative of the innermost case,
-scrutinizing the original alternative body.
-
-Intuitively, since the scrutinee of the outermost case is not in WHNF, no
-resources from it can directly occur in the outermost alternatives. By moving
-the outermost alternatives inwards with a different scrutinee, the alternatives
-remain well-typed because they are typed using either the case binder or the
-pattern bound variables, which, by the \emph{Irrelevance} lemma, makes it
-well-typed for any scrutinee consuming arbitrary resources.
-
-\input{language-v4/proofs/optimizations/CaseOfCase}
-
-\subsection{Commuting let}
-
-\input{language-v4/proofs/optimizations/LetFloating}
-
-\subsection{\texorpdfstring{$\eta$}{Eta}-conversions}
-
-\input{language-v4/proofs/optimizations/EtaConvs}
-
-\subsection{Binder Swap}
-
-The binder swap transformation applies to case expressions whose scrutinee is a
-single variable $x$, and it substitutes occurrences of $x$ in the case
-alternatives for the case binder $z$. If $x$ is a linear resource, $x$ cannot
-occur in the case alternatives (as we conservatively consider variables are not
-in WHNF), so the substitution preserves types vacuously. Otherwise, $x$ can be
-freely substituted by $z$, since $z$ is also an unrestricted resource (it's
-usage environment is empty because $x$ is unrestricted).
-
-\input{language-v4/proofs/optimizations/BinderSwap}
-
-\section{Instrumented Operational Semantics}\label{app:opsem}
-
-In this section we develop a lazy big-step operational semantics for $\lambda_\Delta^\pi$ in the
-style of Launchbury~\cite{10.1145/158511.158618}. The semantics is
-equipped with an environment which assigns values to variables and is
+\section{Natural and Instrumented Operational Semantics}\label{app:opsem}
+
+We define a lazy big-step operational semantics for $\lambda_\Delta^\pi$ in the
+style of Launchbury~\cite{10.1145/158511.158618} natural semantics. The semantics is
+equipped with an environment which assigns expressions to variables and is
 mutated in order to express lazy evaluation. Terms are explicit about
 their sharing (in the sense of shared reductions in lazy evaluation),
-and so we perform a translation step that makes all sharing explicit
-through let-binders.
-Below we write $\llet{\ov{x_i{:}_{\Delta_i}\sigma = e_i}}{e'}$ for the iterated
+and so we perform a translation that makes all sharing explicit
+through let-binders. This translation is presented in Figure~\ref{fig:explicitsharing}.
+We write $\llet{\ov{x_i{:}_{\Delta_i}\sigma = e_i}}{e'}$ for the iterated
 let-binding of variables $\ov{x_i}$.
+\begin{figure}
 \[
@@
     \]
-
-The big-step operational semantics is defined via the relation
+    \caption{Explicit Sharing Translation\label{fig:explicitsharing}}
+  \end{figure}
+
+The natural semantics (Figure~\ref{fig:natsemapp}) is defined via the relation
 $\Theta : e \Downarrow \Theta' : v$ where $e$ is an expression, $v$ a
@@
 
+\begin{figure}
 \[
@@
   \]
-
-We follow an approach
-similar to that of \cite{cite:linearhaskell}, defining a linearity
-aware operational semantics in which linear variables are erased from
+  \caption{Natural Semantics for $\lambda_\Delta^\pi$\label{fig:natsemapp}}
+\end{figure}
+
+Our instrumented, linearity-aware semantics (akin to that
+of~\cite{cite:linearhaskell})
+is defined in Figure~\ref{fig:instrsemapp}.
+In this semantics, linear variables are erased from
 the runtime environment, ensuring that any term violating linearity
-will result in a stuck state. The semantics is type-aware, defined by
+produces a stuck state. The semantics is defined by
 the judgment $\Gamma ; \Delta \vdash (\Theta \mid e) \Downarrow (\Theta'
@@
 ($\Gamma$ tracks unrestricted variables, $\Delta$ tracks linear
-assumptions);
+assumptions) for variables occurring in $\Sigma$;
 $\Theta$ and $\Theta'$ are runtime environments, consisting of bindings of the form
-$x :_\pi \sigma = e$; $e$ is the expression being evaluated and $v$
+$x :_\pi \sigma = e$ or $x :_\Delta \sigma = e$ ; $e$ is the expression being evaluated and $v$
 its resulting value, both of type $\sigma$; $\Sigma$ is a list of assignments of the form
-$e :_\pi \tau$ or $x :_\Delta \tau$ which are expressions in which variables in $\Theta$ are used.
-
+$e : \tau$, which are expressions in which variables in $\Theta$ are
+used.
+
+\begin{figure}
 \[
@@
      % \Downarrow \Theta_n : v_n \quad
-     (\Gamma ; \Delta \vdash (\Theta, \overline{x_i:_\Delta' \sigma_i =
+     (\Gamma ; \Delta \vdash (\Theta, \overline{x_i:_{\Delta'} \sigma_i =
      e_i} \mid e' )\Downarrow (\Theta'
@@
  \]
+ \caption{Instrumented Semantics for $\lambda_\Delta^\pi$\label{fig:instrsemapp}}
+ \end{figure}
 
@@
 
- We write $\Theta {\uparrow^1}$ for the linear bindings in evaluation environment $\Theta$ and $\Theta {\uparrow^\omega}$ for all other bindings.
+Below, we write $\Theta {\uparrow^1}$ for the linear bindings in evaluation environment $\Theta$ and $\Theta {\uparrow^\omega}$ for all other bindings.
 
@@
 
+We note that for the treatment of recursive let groups, environment
+expansion needs to happen for the entire group of let-bound
+variables. This is easily achieved by extending the runtime
+environment with information about recursive let groupings, which we
+omit for the sake of presentation.
+
  \begin{definition}[Well-typed State]
@@
 
-\begin{theorem}[Type Preservation]\label{thm:typepres}
-  Let $\G ; \D \vdash (\Theta \mid e) \Downarrow (\Theta' \mid v) :
-  \tau , \Sigma$ or $\G ; \D \vdash (\Theta \mid e) \Downarrow^* (\Theta' \mid v) :
-  \tau , \Sigma$.
-  If $\G ; \D \vdash \Theta \mid e : \tau , \Sigma$ then $\G ; \D \vdash \Theta' \mid v : \tau , \Sigma$
-
-\end{theorem}
-
-\begin{theorem}[Progress]\label{thm:prog}
-Let $\G ; \D \vdash \Theta \mid e : \tau,\Sigma$. Any partial derivation $\G; \D \vdash (\Theta \mid e)
-\Downarrow?,\Sigma$ can be extended. 
-\end{theorem}
-
-\begin{proof}
- Essentially all cases are standard. The only potential issue is the
- (linear) variable rule.
-It suffices to show that whenever a variable is evaluated it is
-present in the evaluation environment. This situation is impossible
-since the state
-$\G ; \D \vdash \Theta \mid x : \tau , \Sigma$ where $x$ is absent
-from $\Theta$ is not well-typed and so cannot be the head of a partial
-derivation. 
-\end{proof}
+\subsection{Bisimulation}
 
@@
 the lazy big-step semantics if $e = e'$ and $\Theta'{\uparrow^\omega}
-\leq \Theta$
+\leq \Theta$.
 \end{definition}
@@
 
-  We now derive type safety for the lazy natural semantics.
+\section{Type Safety Proofs}\label{app:proofs}
+
+In the following we write $\Mapsto$ to stand for
+$\Rightarrow_\mathsf{WHNF}$ and $\Rrightarrow$ for $\Rightarrow_\mathsf{NWHNF}$.
+
+\subsection{Auxiliary Lemmas}
+The following lemmas presuppose the variable names across the typing
+contexts are distinct.
+
+  \begin{lemma}[Linear to $\Delta$-bound]~\label{lem:onedelta-app}
+
+    \begin{enumerate}
+
+      \item If $\G; \D',\xl \vdash e :\vp$
+        then $\G[\D/x],\xD; \D,\D' \vdash e : \vp$ (with $\Delta$ fresh).
+      \item If $\G ; \D' , \xl \vdash_{alt} \rho \rightarrow e
+        :^z_{\Delta_s} \sigma \Rightarrow \rho$ then
+        $\G[\D/x],\xD;\D,\D' \vdash_{alt} \rho \rightarrow e
+        :^z_{\Delta_s} \sigma \Rightarrow \rho$ (with $\Delta$ fresh).
+
+      \end{enumerate}
+\end{lemma}
+\begin{proof}
+Straightforward induction on the structure of the given derivation,
+reconstructing the original derivation and applying rule
+$\mathit{Var}_\Delta$ where rule $\mathit{Var}_1$ was applied on
+$x$ in the original derivation.
+\end{proof}
+\onedelta*
+
+\begin{proof}
+By Lemma~\ref{lem:onedelta-app}(1).
+\end{proof}
+
+\begin{lemma}[$\Delta$-bound to Linear]~\label{lem:deltaone-app}
+
+  \begin{enumerate}
+ \item If $\G,\x[\irr{\D}]; \irr{\D},\D' \vdash e : \vp$
+   then $\G[x/\irr{\D}]; \D',\xl \vdash e :\vp$.
+ \item If $\G ,\x[\irr{\D}]; \irr{\D}, \D' \vdash_{alt} \rho \rightarrow e
+   :^z_{\Delta_s} \sigma \Rightarrow \rho$ then
+   $\G[x/\irr{\D}];\D',\xl \vdash_{alt} \rho \rightarrow e
+   :^z_{\Delta_s} \sigma \Rightarrow \rho$.
+  \end{enumerate}
+\end{lemma}
+\begin{proof}
+Straightforward induction on the structure of the given derivation,
+reconstructing the original derivation and applying rule
+$\mathit{Var}_1$ where rule $\mathit{Var}_\D$ was applied on
+$x$ in the original derivation.
+\end{proof}
+
+\deltaone*
+\begin{proof}
+By Lemma~\ref{lem:deltaone-app}(1).
+\end{proof}
+
+\begin{lemma}[Unrestricted and $\Delta$-bound]~\label{lem:undelta-app}
+  \begin{enumerate}
+    \item $\G,\xo; \D \vdash e : \vp$ iff $\G,\x[\cdot]; \D \vdash e :
+      \vp$
+    \item $\G,\xo;\D \vdash_{alt} \rho \rightarrow e
+      :^z_{\Delta_s} \sigma \Rightarrow \rho$ iff
+      $\G,\x[\cdot];\D \vdash_{alt} \rho \rightarrow e
+   :^z_{\Delta_s} \sigma \Rightarrow \rho$
+  \end{enumerate}
+\end{lemma}
+\begin{proof}
+Straightforward induction on the structure of the given derivation,
+noting that rules $\mathit{Var}_\Delta$ and $\mathit{Var}_\omega$
+coincide when $\Delta$ is empty.
+\end{proof}
+
+\undelta*
+
+\begin{proof}
+By Lemma~\ref{lem:undelta-app}.
+\end{proof}
+
+\subsection{Irrelevance\label{sec:proof:irrelevance}}
+
+\input{language-v4/proofs/WHNFConvSoundness}
+
+\subsection{Substitution Lemmas\label{sec:proof:substitution-lemmas}}
+
+\input{language-v4/proofs/LinearSubstitutionLemma}
+
+\input{language-v4/proofs/UnrestrictedSubstitutionLemma}
+
+\input{language-v4/proofs/DeltaSubstitutionLemma}
+
+\subsection{Type Preservation\label{sec:proof:type-preservation}}
+
+% \input{language-v4/proofs/TypePreservationTheorem}
+\begin{theorem}[Type Preservation]\label{thm:typepres}
+  Let $\G ; \D \vdash (\Theta \mid e) \Downarrow (\Theta' \mid v) :
+  \tau , \Sigma$ or $\G ; \D \vdash (\Theta \mid e) \Downarrow^* (\Theta' \mid v) :
+  \tau , \Sigma$.
+  If $\G ; \D \vdash \Theta \mid e : \tau , \Sigma$ then $\G ; \D \vdash \Theta' \mid v : \tau , \Sigma$
+
+\end{theorem}
+
+We also derive type preservation for the lazy natural semantics.
 
@@
 
+\subsection{Progress\label{sec:proof:progress}}
+
+%\input{language-v4/proofs/ProgressTheorem}
+
+\begin{theorem}[Progress]\label{thm:prog}
+Let $\G ; \D \vdash \Theta \mid e : \tau,\Sigma$. Any partial derivation $\G; \D \vdash (\Theta \mid e)
+\Downarrow?,\Sigma$ can be extended. 
+\end{theorem}
+
+\begin{proof}
+ Essentially all cases are standard. The only potential issue is the
+ (linear) variable rule.
+It suffices to show that whenever a variable is evaluated it is
+present in the evaluation environment. This situation is impossible
+since the state
+$\G ; \D \vdash \Theta \mid x : \tau , \Sigma$ where $x$ is absent
+from $\Theta$ is not well-typed and so cannot be the head of a partial
+derivation. 
+\end{proof}
+
 \begin{theorem}[Progress]
@@
 
+\section{Optimisations preserve linearity}\label{app:optimisations}
+
+\subsection{Inlining}
+\input{language-v4/proofs/optimizations/Inlining}
+
+\subsection{\texorpdfstring{$\beta$}{Beta}-reduction}
+
+\input{language-v4/proofs/optimizations/BetaReduction}
+
+\subsection{Case of known constructor}
+
+\input{language-v4/proofs/optimizations/CaseOfKnownConstructor}
+
+\subsection{Case of Case\label{sec:proof:caseofcase}}
+
+The case of case transformation applies to case expressions whose scrutinee is
+another case expression, and returns the innermost case expression transformed by repeating
+the outermost case expression in each alternative of the innermost case,
+scrutinizing the original alternative body.
+
+Intuitively, since the scrutinee of the outermost case is not in WHNF, no
+resources from it can directly occur in the outermost alternatives. By moving
+the outermost alternatives inwards with a different scrutinee, the alternatives
+remain well-typed because they are typed using either the case binder or the
+pattern bound variables, which, by the \emph{Irrelevance} lemma, makes it
+well-typed for any scrutinee consuming arbitrary resources.
+
+\input{language-v4/proofs/optimizations/CaseOfCase}
+
+\subsection{Commuting let}
+
+\input{language-v4/proofs/optimizations/LetFloating}
+
+\subsection{\texorpdfstring{$\eta$}{Eta}-conversions}
+
+\input{language-v4/proofs/optimizations/EtaConvs}
+
+\subsection{Binder Swap}
+
+The binder swap transformation applies to case expressions whose scrutinee is a
+single variable $x$, and it substitutes occurrences of $x$ in the case
+alternatives for the case binder $z$. If $x$ is a linear resource, $x$ cannot
+occur in the case alternatives (as we conservatively consider variables are not
+in WHNF), so the substitution preserves types vacuously. Otherwise, $x$ can be
+freely substituted by $z$, since $z$ is also an unrestricted resource (it's
+usage environment is empty because $x$ is unrestricted).
+
+\input{language-v4/proofs/optimizations/BinderSwap}
+
 % G ; D |- (Theta | e x) \Downarrow (Theta'' | v) : tau, Sigma

### 2025-02-26 icfp.tex
author
Lemma C3 proof typeset

@@
 \author{Rodrigo Mesquita}
-\affiliation{
-  \department{Departamento de Inform\'{a}tica}
-  \institution{FCT-NOVA, Universidade Nova de Lisboa}
+%\affiliation{
+  %\department{Departamento de Inform\'{a}tica}
+  %\institution{FCT-NOVA, Universidade Nova de Lisboa}
   %\city{City}
-  \country{Portugal}}
+  %\country{Portugal}}
 \affiliation{
@@
  }
- \email{xxx@yyy.com}
+ \email{rodrigo@well-typed.com}
 \author{Bernardo Toninho}
@@
 
-Below, we write $\Theta {\uparrow^1}$ for the linear bindings in evaluation environment $\Theta$ and $\Theta {\uparrow^\omega}$ for all other bindings.
+Below, we write $\Theta {\uparrow^1}$ for the linear bindings in evaluation
+environment $\Theta$ and $\Theta {\uparrow^\omega}$ for all other bindings.
+$\Theta\downarrow^1_e$ refers to the expressions bound to linear variables in
+the runtime environment and $\Theta\downarrow^1_\sigma$ refers to the types of
+those expressions.
 
@@
        \left[ \cdot \mid e \right] & \triangleq & e\\
-       \left[ \Theta , x:_1\sigma = e' \mid e \right] & \triangleq  & [\Theta \mid (\lambda x_1:\sigma . e)~e']\\
+       \left[ \Theta , \xl = e' \mid e \right] & \triangleq  & [\Theta \mid (\lambda \xl . e)~e']\\
        \left[\Theta , x:_\Delta \sigma = e' \mid e \right] & \triangleq & [\Theta \mid \llet{x:_\Delta \sigma = e'}{e}]
@@
    \end{lemma}
+\begin{proof}
+By structural induction on the environment expansion definition.
+\begin{description}
+%\item[Case:] $\left[\cdot \mid e\right] \triangleq e$ Trivial
+\item[Case:] $\left[\Theta, \xl = y \mid e\right] \triangleq \left[\Theta \mid (\lambda \xl. e)~y\right]$
+\begin{tabbing}
+    (1) $\G; \D \vdash \left[ \Theta, \xl = y \mid e \right] : \phi$\\
+    (2) $\G; \D \vdash \left[ \Theta \mid (\lambda \xl. e)~y \right] : \phi$\` by def.\\
+    (3) $\G, \Theta\uparrow^\omega; \D, \Theta\uparrow^1 \vdash ((\lambda \xl. e)~y, \Theta\downarrow^1_e) : \phi \otimes \bigotimes(\Theta\downarrow^1_\sigma)$\` by i.h.\\
+    (4) $Let \D_1,\D_2,\D_3 = \D, \Theta\uparrow^1 $\\
+    (5) $\G, \Theta\uparrow^\omega; \D_3 \vdash \Theta\downarrow^1_e : \bigotimes(\Theta\downarrow^1_\sigma)$\\
+    (6) $\G, \Theta\uparrow^\omega; \D_2 \vdash y : \s$\` by inv\\
+    (7) $\G, \Theta\uparrow^\omega; \D_1 \vdash (\lambda \xl. e) : \s \to \phi $\` by inv\\
+    (8) $\G, \Theta\uparrow^\omega; \D_1, \xl \vdash e : \phi $\` by inv\\
+    (9) $\G, \Theta\uparrow^\omega; \D_1, \D_2, \xl \vdash (e, y) : \phi \otimes \sigma$\` by pair\\
+    (10) $\G, \Theta\uparrow^\omega; \D, \xl \vdash (e, y, \Theta\downarrow^1_e) : \phi \otimes \sigma \otimes \bigotimes(\Theta\downarrow^1_\sigma)$\` by pair\\
+\end{tabbing}
+\item[Case:] $\left[\Theta, \xD = e' \mid e\right] \triangleq \left[\Theta \mid \llet{\xD = e'}{e}\right]$
+\begin{tabbing}
+    (1) $\G; \D' \vdash \left[ \Theta, \xD = y \mid e \right] : \phi$\\
+    (2) $\G; \D' \vdash \left[ \Theta \mid \llet{\xD = e'}{e} \right] : \phi$\` by def.\\
+    (3) $\G, \Theta\uparrow^\omega; \D', \Theta\uparrow^1 \vdash (\llet{\xD = e'}{e}, \Theta\downarrow^1_e) : \phi \otimes \bigotimes(\Theta\downarrow^1_\sigma)$\` by i.h.\\
+    (4) $Let \D,\D_2,\D_3 = \D', \Theta\uparrow^1 $\\
+    (5) $\G, \Theta\uparrow^\omega; \D_3 \vdash \Theta\downarrow^1_e : \bigotimes(\Theta\downarrow^1_\sigma)$\\
+    (6) $\G, \Theta\uparrow^\omega; \D \vdash e' : \s$\` by inv on let\\
+    (7) $\G, \Theta\uparrow^\omega, \xD; \D,\D_2 \vdash e : \phi $\` by inv on let\\
+    (8) $\G, \Theta\uparrow^\omega, \xD; \D,\D_2,\D_3 \vdash (e, \Theta\downarrow^1_e) : \phi \otimes \bigotimes(\Theta\downarrow^1_\sigma)$\` by pair\\
+\end{tabbing}
+\end{description}
+\end{proof}
 

### 2025-02-27 icfp.tex
start on preservation typesetting

@@
 
- \begin{lemma}
+ \begin{lemma}~\label{lem:unwrapenv}
    Iff $\G ; \D \vdash [ \Theta \mid e] : \sigma$
@@
     (3) $\G, \Theta\uparrow^\omega; \D, \Theta\uparrow^1 \vdash ((\lambda \xl. e)~y, \Theta\downarrow^1_e) : \phi \otimes \bigotimes(\Theta\downarrow^1_\sigma)$\` by i.h.\\
-    (4) $Let \D_1,\D_2,\D_3 = \D, \Theta\uparrow^1 $\\
+    (4) $Let~\D_1,\D_2,\D_3 = \D, \Theta\uparrow^1 $\\
     (5) $\G, \Theta\uparrow^\omega; \D_3 \vdash \Theta\downarrow^1_e : \bigotimes(\Theta\downarrow^1_\sigma)$\\
@@
     (3) $\G, \Theta\uparrow^\omega; \D', \Theta\uparrow^1 \vdash (\llet{\xD = e'}{e}, \Theta\downarrow^1_e) : \phi \otimes \bigotimes(\Theta\downarrow^1_\sigma)$\` by i.h.\\
-    (4) $Let \D,\D_2,\D_3 = \D', \Theta\uparrow^1 $\\
+    (4) $Let~\D,\D_2,\D_3 = \D', \Theta\uparrow^1 $\\
     (5) $\G, \Theta\uparrow^\omega; \D_3 \vdash \Theta\downarrow^1_e : \bigotimes(\Theta\downarrow^1_\sigma)$\\
@@
 
+\begin{proof}
+By structural induction on the instrumented operational semantics reductions.
+\begin{description}
+    \item[Case:] $\Gamma; \Delta \vdash (\Theta \mid \ccase{e}{z{:}_{\Delta'}\s~\{K~\ov{y_i}\to e'}\}) \Downarrow (\Theta'' \mid v) : \tau$
+\begin{tabbing}
+    (1) $\Gamma; \Delta \vdash \Theta \mid \ccase{e}{z{:}_{\Delta_a}\s~\{K~\ov{y_i}\to e'}\} : \tau, \Sigma$\\
+    (2) $\G, \Theta\uparrow^\omega; \D, \Theta\uparrow^1 \vdash (\ccase{e}{z{:}_{\Delta_a}\s~\{K~\ov{y_i}\to e'}\}, \Theta\downarrow^1_e, \Sigma) : \tau \otimes \bigotimes(\Theta\downarrow^1_\sigma) \otimes \bigotimes(\Sigma)$\`by \ref{lem:unwrapenv}\\
+    (3) $Let~\G' = \G, \Theta\uparrow^\omega; \D_a,\D_b,\D_2,\D_3 = \D,\Theta\uparrow^1$\`s.t. $(\D_2,\D_3)$ types $(\Theta\downarrow^1_e, \Sigma)$\\
+    (4) $\G'; \D_a, \D_b \vdash \ccase{e}{z{:}_{\D_a}\s\{K~\ov{y_i}\to e'\}} : \tau$\\
+    Subcase $e$ is in WHNF (thus $= K~\ov{x_i}$)\\
+    (5) $\G'; \D_a \Vdash K~\ov{x_i} : \s \gtrdot \ov{\D_i}$\` by subcase\\
+    (6) $\G', z{:}_{\D_a}\s; \D_a, \D_b \vdash_{alt} K~\ov{y_i} \to e' :^z_{\D_i} \s \Rightarrow \tau$\` by inv\\
+    Subsubcase $\vdash_{alt}$ $Alt_{NWHNF}$\\
+    (7) $\G', z{:}_{\D_a}\s, \ov{y_i{:}_\omega\s_i}, \ov{y_j{:}_{\Delta_j}\s_j}; \D_a, \D_b \vdash e' : \tau$\` by inv\\
+    (8) $\Theta' = \Theta$\` in this subcase because $(\Theta \mid K~\ov{x_i})\Downarrow(\Theta \mid K~\ov{x_i})$\\
+    (9) $\G'; \ov{\D_i} \vdash K~\ov{x_i} : \s$\` by inv (5)\\
+    (10) $\ov{\G'; \cdot \vdash x_i : \s_i}$\` by inv (9) for unrestricted $x$s\\
+    (11) $\ov{\G'; \D_j \vdash x_j : \s_j}$\` by inv (9) for linear fields $x$s\\
+    (12) $\G', z{:}_{\D_a}\s; \D_a, \D_b \vdash e'\ov{[x_{i,j}/y_{i,j}] : \tau}$\` by $\D$-subst and $\omega$-subst lemmas (7,10,11)\\
+    (13) $\G'; \D_a, \D_b \vdash e'\ov{[x_{i,j}/y_{i,j}]}[K~\ov{x_{i,j}}/z] : \tau$\` by $\D$-subst lemma and $\ov{\D_i} = \D_a$ (9,12)\\
+    (14) $\G; \D \vdash \Theta \mid e'\ov{[x_{i,j}/y_{i,j}]}[K~\ov{x_{i,j}}/z] : \tau, \Sigma$\` by reconstructing\\
+    (15) $\G; \D \vdash \Theta'' \mid v : \tau, \Sigma$\` by induction hypothesis\\
+    Subsubcase $\vdash_{alt}$ $Alt_{0}$\\
+    (7) $(\G', z{:}_{\D_a}\s)[\cdot/\D_a]_z, \ov{y_i{:}_\omega\sigma_i}; (\D_a, \D_b)[\cdot / \D_a] \vdash e' : \tau$\` by inv subcase\\
+    (8) $\G', z{:}_{\cdot}, \ov{y_i{:}_\omega\sigma_i}; \D_b \vdash e' : \tau$\` by $z$ cannot occur in $\G'$\\
+    (9) Since $\G';\D_a \Vdash K~\ov{x_i}$, in this subcase $\D_a = \cdot$\\
+    (10) $\G', \z{:}_\cdot\sigma, \ov{y{:}_\omega\sigma}; \D_b \vdash e' : \tau$\` by (9)\\
+    (11) $\G', \z{:}_\cdot\sigma; \D_b \vdash e'\ov{[x_i/y_i]} : \tau$\` by $\omega$-subst (10)\\
+    (12) $\G'; \D_b \vdash e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] : \tau$\` by $\D$-subst\\
+    (13) $\G; \D \vdash \Theta \mid e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] : \tau, \Sigma$\` by reconstructing\\
+    (14) $\G; \D \vdash \Theta'' \mid v : \tau, \Sigma$\`by induction hypothesis\\
+    Subcase $e$ is not in WHNF\\
+    (5) $\G'; \D_a \vdash e : \s$\` by inv on case\\
+    (6) $\G', z{:}_{\irr{\D_a}}\s; \irr{\D_a}, \D_b \vdash_{alt} K~\ov{y_i} \to e' :^z_{\irr{\D_a}} \s $\` by inv on case\\
+\end{tabbing}
+\end{description}
+\end{proof}
+
 We also derive type preservation for the lazy natural semantics.

### 2025-02-27 icfp.tex
Fix in instrumented semantics
Type Preservation for Case K~xs

@@
     \\
-    \Gamma, \ov{y_i :_\omega \sigma_i} ;\Delta , \ov{y_j :_1 \sigma_j} \vdash (\Theta \mid e) \Downarrow (\Theta' \mid
+    \Gamma, z{:}_{\ov{y_j}}T, \ov{y_i :_\omega \sigma_i} ;\Delta , \ov{y_j :_1 \sigma_j} \vdash (\Theta \mid e) \Downarrow (\Theta' \mid
       K~\ov{x_i}) : T , e' : \tau ,\Sigma \\
@@
     (1) $\Gamma; \Delta \vdash \Theta \mid \ccase{e}{z{:}_{\Delta_a}\s~\{K~\ov{y_i}\to e'}\} : \tau, \Sigma$\\
-    (2) $\G, \Theta\uparrow^\omega; \D, \Theta\uparrow^1 \vdash (\ccase{e}{z{:}_{\Delta_a}\s~\{K~\ov{y_i}\to e'}\}, \Theta\downarrow^1_e, \Sigma) : \tau \otimes \bigotimes(\Theta\downarrow^1_\sigma) \otimes \bigotimes(\Sigma)$\`by \ref{lem:unwrapenv}\\
+    (2) $\G, \Theta{\uparrow}^\omega; \D, \Theta{\uparrow}^1 \vdash (\ccase{e}{z{:}_{\Delta_a}\s~\{K~\ov{y_i}\to e'}\}, \Theta{\downarrow}^1_e, \Sigma) : \tau \otimes \bigotimes(\Theta{\downarrow}^1_\sigma) \otimes \bigotimes(\Sigma)$\`by \ref{lem:unwrapenv}\\
     (3) $Let~\G' = \G, \Theta\uparrow^\omega; \D_a,\D_b,\D_2,\D_3 = \D,\Theta\uparrow^1$\`s.t. $(\D_2,\D_3)$ types $(\Theta\downarrow^1_e, \Sigma)$\\
@@
     (6) $\G', z{:}_{\D_a}\s; \D_a, \D_b \vdash_{alt} K~\ov{y_i} \to e' :^z_{\D_i} \s \Rightarrow \tau$\` by inv\\
-    Subsubcase $\vdash_{alt}$ $Alt_{NWHNF}$\\
+    Subsubcase $\vdash_{alt}$ $Alt_{WHNF}$\\
     (7) $\G', z{:}_{\D_a}\s, \ov{y_i{:}_\omega\s_i}, \ov{y_j{:}_{\Delta_j}\s_j}; \D_a, \D_b \vdash e' : \tau$\` by inv\\
@@
     (5) $\G'; \D_a \vdash e : \s$\` by inv on case\\
-    (6) $\G', z{:}_{\irr{\D_a}}\s; \irr{\D_a}, \D_b \vdash_{alt} K~\ov{y_i} \to e' :^z_{\irr{\D_a}} \s $\` by inv on case\\
+    (6) $\G', z{:}_{\irr{\D_a}}\s; \irr{\D_a}, \D_b \vdash_{alt} K~\ov{y_{i,j}} \to e' :^z_{\irr{\D_a}} \s $\` by inv on case\\
+    Subsubcase $\vdash_{alt}$ $Alt_{NWHNF}$\\
+    (7) $\G', z{:}_{\irr{\D_a}}\s, \ov{y_i{:}_\omega\s_i}, \ov{y_j{:}_{\Delta_j}\s_j}; \irr{\D_a}, \D_b \vdash e' : \tau$\` for \ov{\D_j = \lctag{\irr{\D_a}}{K_j}}, by inv.\\
+    (8) $\G', z{:}_{\ov{y_j{:}_1\s_j}}\s, \ov{y_i{:}_\omega\s_i}; \ov{y_j{:}_1\s_j}, \D_b \vdash e' : \tau$\` by $\D$-bound to linear\\
+    (9) $\G', z{:}_{\ov{y_j{:}_1\s_j}}\s, \ov{y_i{:}_\omega\s_i}; \D_a, \D_b, \ov{y_j{:}_1\s_j} \vdash (e, e') : \sigma \otimes \tau$\` by pair (5,8)\\
+    (10) $\G, z{:}_{\ov{y_j{:}_1\s_j}}\s, \ov{y_i{:}_\omega\s_i}; \D, \ov{y_j{:}_1\s_j} \vdash \Theta \mid e : \s, e':\tau, \Sigma$\`by reconstructing\\
+    (11) $\G, z{:}_{\ov{y_j{:}_1\s_j}}\s, \ov{y_i{:}_\omega\s_i}; \D, \ov{y_j{:}_1\s_j} \vdash \Theta' \mid K~\ov{x_i} : \s, e':\tau, \Sigma$\`by induction hypothesis\\
+    (12) $\G, \Theta'{\uparrow}^\omega, \ov{y_i{:}_\omega\s_i}; \D, \ov{y_j{:}_1\s_j}, \Theta'{\uparrow}^1 \vdash (K~\ov{x_{i,j}}, \Theta'{\downarrow}^1_e, e', \Sigma) : \s \bigotimes(\Theta'\downarrow^1_\sigma) \otimes \tau \bigotimes(\Sigma)$\`by \ref{lem:unwrapenv}\\
+    (13) $Let~\G''=\G,\Theta'{\uparrow}^\omega~and~\D_c\D_d\D_4\D_5=\D,\Theta'{\uparrow}^1$\` where $(\D_4, \D_5)$ types $(\Theta'{\downarrow^1_e}, \Sigma)$\\
+    (13a) $\G''; \D_c \vdash K~\ov{x_{i,j}} : \s$\\
+    (13b) $\G'', z{:}_{\ov{y_j{:}_1\s_j}}\s, \ov{y_i{:}_\omega\s_i}; \D_d, \ov{y_j{:}_1\s_j} \vdash e' : \tau$\\
+    (14) $\ov{\G''; \cdot \vdash x_i : \s_i}$\\
+    (15) $\ov{\G''; \D_j \vdash x_j : \s_j}$\` for $\ov{D_j}=\D_c$, by inv\\
+    (16) $\G'', z{:}_{\D_c}\s; \D_c, \D_d \vdash e'\ov{[x_{i,j}/y_{i,j}]} : \tau$\`by $1$-substs and $\omega$-substs lemmas\\
+    (17) $\G''; \D_c, \D_d \vdash e'\ov{[x_{i,j}/y_{i,j}]}[K~\ov{x_{i,j}}/z] : \tau$\`by $\D$-subst lemma (13a,16)\\
+    (18) $\G; \D \vdash \Theta' \mid e'\ov{[x_{i,j}/y_{i,j}]}[K~\ov{x_{i,j}}/z] : \tau, \Sigma$\`by reconstructing\\
+    (19) $\G; \D \vdash \Theta'' \mid v : \tau, \Sigma$\`by induction hypothesis\\
+    Subsubcase $\vdash_{alt}$ $Alt_0$\\
+    (7) $\G', z{:}_\cdot\s, \ov{y_i{:}_\omega\s_i}; \D_b \vdash e' : \tau$\` by inv\\
+    (8) $\G', z{:}_\cdot\s, \ov{y_i{:}_\omega\s_i}; \D_a, \D_b \vdash (e, e') : \sigma \otimes \tau$\` by pair (5,7)\\
+    (9) $\G, z{:}_\cdot\s, \ov{y_i{:}_\omega\s_i}; \D \vdash \Theta \mid e : \sigma, e' : \tau, \Sigma$\` by reconstructing\\
+    (10) $\G, z{:}_\cdot\s, \ov{y_i{:}_\omega\s_i}; \D \vdash \Theta \mid e : \sigma, e' : \tau, \Sigma$\` by reconstructing\\
+    (11) $\G, z{:}_\cdot\s, \ov{y_i{:}_\omega\s_i}; \D \vdash \Theta' \mid K~\ov{x_i} : \sigma, e' : \tau, \Sigma$\` by induction hypothesis\\
+    (12) $Let~\G'' = \G,\Theta'{\uparrow}^\omega~and~\D_d,\D_4,\D_5 = \D, \Theta'{\uparrow}^1$\` where $(\D_4, \D_5)$ types $(\Theta'{\downarrow^1_e}, \Sigma)$\\
+    (12a) $\G''; \cdot \vdash K~\ov{x_i} : \s$\`since $K$ has no linear fields in this subcase\\
+    (12b) $\G'', z{:}_{\cdot}\s, \ov{y_i{:}_\omega\s_i}; \D_d \vdash e' : \tau$\\
+    (13) $\ov{\G''; \cdot \vdash x_{i} : \s_i}$\`by inv (12a)\\
+    (14) $\G'', z{:}_{\cdot}\s; \D_d \vdash e'\ov{[x_i/y_i]} : \tau$\` by $\omega$-subst lemma\\
+    (15) $\G''; \D_d \vdash e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] : \tau$\` by $\D$-subst lemma\\
+    (16) $\G; \D \vdash \Theta' \mid e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] : \tau, \Sigma$\` by reconstructing\\
+    (17) $\G; \D \vdash \Theta'' \mid v : \tau, \Sigma$\` by induction hypothesis\\
 \end{tabbing}

### 2025-02-27 icfp.tex
preservation case ALT_
Finish type preservation proof typesetting

@@
 \begin{description}
+    \item[Case:] $\Gamma; \Delta \vdash (\Theta, \xl = y \mid x) \Downarrow (\Theta' \mid v) : \s, \Sigma$
+\begin{tabbing}
+    (1) $\G; \D \vdash \Theta, \xl = y \mid x : \s, \Sigma$\\
+    (2) $\G, \Theta{\uparrow}^\omega; \D, \Theta{\uparrow}^1, \xl \vdash (x, y, \Theta{\downarrow}^1_e, \Sigma) : \s \otimes \s \bigotimes(\Theta{\downarrow^1_e}) \bigotimes(\Sigma)$\`by \ref{lem:unwrapenv}\\
+    (3) $Let~\G';\D_1,\D_2,\D_3 = \G, \Theta{\uparrow}^\omega; \D, \Theta{\uparrow}^1$\` s.t. $(\D_2,\D_3)$ types $(\Theta{\downarrow}^1_e,\Sigma)$\\
+    (4) $\G'; \xl \vdash x : \s$\\
+    (5) $\G'; \D_1 \vdash y : \s$\\
+    (6) $\G'; \D_1, \D_2, \D_3 \vdash (y, \Theta{\downarrow}^1_e, \Sigma) : \s \bigotimes(\Theta{\downarrow^1_e}) \bigotimes(\Sigma)$
+    (7) $\G; \D \vdash \Theta \mid y : \s, \Sigma$\`by reconstructing\\
+    (8) $\G; \D \vdash \Theta' \mid v : \s, \Sigma$\` by induction hypothesis\\
+\end{tabbing}
+    \item[Case:] $\Gamma; \Delta \vdash (\Theta, x{:}_{\D'}\s = e \mid x) \Downarrow (\Theta', x{:}_{\D_a}\s = v \mid v) : \s, \Sigma$
+\begin{tabbing}
+    (1) $\Gamma; \Delta \vdash (\Theta, x{:}_{\D'}\s = e \mid x) : \s, \Sigma$\\
+    (2) $\G, \Theta{\uparrow}^\omega; \D, \Theta{\uparrow}^1 \vdash \llet{x{:}_{\D'} = e}{(x, \Theta{\downarrow}^1_e, \Sigma)} : \s \bigotimes(\Theta{\downarrow}^1_\s) \bigotimes(\Sigma)$\`by \ref{lem:unwrapenv}\\
+    (3) $Let~\D',\D'' = \D,\Theta{\uparrow}^1$\\
+    (4) $\G, \Theta{\uparrow}^\omega; \D' \vdash e : \s$\`by inv on let (2)\\
+    (5) $\G, \Theta{\uparrow}^\omega; \D', \D'' \vdash (x, \Theta{\downarrow}^1_e, \Sigma) : \s \bigotimes(\Theta{\downarrow}^1_\s) \bigotimes(\Sigma)$\`by inv on let (2)\\
+    (6) $\G, \Theta{\uparrow}^\omega; \D' \vdash x : \s$\\
+    (7) $\G, \Theta{\uparrow}^\omega; \D'' \vdash (\Theta{\downarrow}^1_e, \Sigma) : \bigotimes(\Theta{\downarrow}^1_\s) \bigotimes(\Sigma)$\\
+    (8) $\G, \Theta{\uparrow}^\omega; \D', \D'' \vdash (e, \Theta{\downarrow}^1_e, \Sigma) : \tau \bigotimes(\Theta{\downarrow}^1_\s) \bigotimes(\Sigma)$\` by pair (4,7)\\
+    (9) $\G; \D \vdash \Theta \mid e : \tau,\Sigma$\` by reconstructing\\
+    (10) $\G; \D \vdash \Theta \mid v : \tau,\Sigma$\` by induction hypothesis\\
+    (11) $Let~\G';\D_a,\D_b = \G, \Theta'{\uparrow}^\omega, \D, \Theta'{\uparrow}^1$\\
+    (11a) $\G'; \D_a \vdash v : \s$
+    (11b) $\G'; \D_b \vdash (\Theta'{\downarrow}^1_e, \Sigma) : \bigotimes(\Theta'{\downarrow}^1_\s) \bigotimes(\Sigma)$\\
+    (12) $\G'; \D_a \vdash \llet{x{:}_{\D_a}\s = v}{v} : \s$\`by let (11a,11a)\\
+    (13) $\G; \D \vdash \Theta' \mid \llet{x{:}_{\D_a}\s = v}{v} : \s, \Sigma$\`by reconstructing\\
+    (14) $\G; \D \vdash \Theta', x{:}_{\D_a}\s = v \mid v : \s, \Sigma$\`by env. expansion\\
+\end{tabbing}
+Note: It's slightly more elaborate to accommodate recursion. The runtime
+environment bindings must be tagged with the recursive group such that the
+environment expansion can decide whether to unwrap to a letrec group or
+not. Unlike the case proved here, when the $\D$-binding belongs to a let group,
+the whole group is unwrapped and we can invert the assumption to get an
+$e$ typed with the $x$s from the group.\\
 \item[Case:] $\Gamma; \Delta \vdash (\Theta \mid \ccase{e}{z{:}_{\Delta'}\s~\{K~\ov{y_i}\to e'}\}) \Downarrow (\Theta'' \mid v) : \tau$
@@
 \end{tabbing}
+\item[Case:] $\Gamma; \Delta \vdash (\Theta \mid \ccase{e}{z{:}_{\Delta'}\s~\{\_ \to e'}\}) \Downarrow (\Theta'' \mid v) : \tau$
+\begin{tabbing}
+    (1) $\Gamma; \Delta \vdash \Theta \mid \ccase{e}{z{:}_{\Delta_a}\s~\{\_ \to e'}\} : \tau, \Sigma$\\
+    (2) $\G, \Theta{\uparrow}^\omega; \D, \Theta{\uparrow}^1 \vdash (\ccase{e}{z{:}_{\Delta_a}\s~\{\_ \to e'}\}, \Theta{\downarrow}^1_e, \Sigma) : \tau \otimes \bigotimes(\Theta{\downarrow}^1_\sigma) \otimes \bigotimes(\Sigma)$\`by \ref{lem:unwrapenv}\\
+    (3) $Let~\G' = \G, \Theta\uparrow^\omega; \D_a,\D_b,\D_2,\D_3 = \D,\Theta\uparrow^1$\`s.t. $(\D_2,\D_3)$ types $(\Theta\downarrow^1_e, \Sigma)$\\
+    (4) $\G'; \D_a, \D_b \vdash \ccase{e}{z{:}_{\D_a}\s\{\_ \to e'\}} : \tau$\\
+    (5) $\G'; \D_a \vdash e : \sigma$\\
+    Subcase $e$ is in WHNF\\
+    (-) Similar to the same subcase in the case-with-pattern-vars reduction,\\
+        except only $Alt_{\_}$ is relevant which is far simpler\\
+    Subcase $e$ is not in WHNF\\
+    (6) $\G', z{:}_{\irr{\D_a}}\s; \irr{\D_a}, \D_b \vdash e' : \tau$\\
+    (7) $\G'; \D_b, z{:}_1\s \vdash e' : \tau$\` by $\D$-bound to linear\\
+    (8) $\G'; \D_a, \D_b, z{:}_1\s \vdash (e, e') : \sigma \otimes \tau$\` by pair\\
+    (9) $\G; \D, z{:}_1\s \vdash \Theta \mid e : \sigma, e' : \tau, \Sigma$\` by reconstructing\\
+    (10) $\G; \D, z{:}_1\s \vdash \Theta' \mid K~\ov{x_i} : \sigma, e' : \tau, \Sigma$\` by induction hypothesis\\
+    (11) $Let~\G';\D_c,\D_d,\D_4,\D_5 = \G, \Theta'{\uparrow}^\omega; \Delta, \Theta'{\uparrow}^1$\`s.t. $(\D_4,\D_5)$ types $(\Theta'{\downarrow}^1_e, \Sigma)$\\
+    (12) $\G';\D_c \vdash K~\ov{x_i} : \s$\\
+    (13) $\G';\D_d,z{:}_1\s \vdash e' : \tau$\\
+    (14) $\G';\D_c, \D_d \vdash e'[K~\ov{x_i}/z] : \tau$\` by $1$-subst lemma\\
+    (15) $\G;\D \vdash \Theta' \mid e'[K~\ov{x_i}/z] : \tau, \Sigma$\` by reconstructing\\
+    (16) $\G;\D \vdash \Theta'' \mid v : \tau, \Sigma$\` by induction hypothesis\\
+\end{tabbing}
 \end{description}

### 2025-02-27 icfp.tex
erased trivial cases of subst lemmas

--- 2025-02-27 icfp.tex
@@
 \author{Rodrigo Mesquita}
-%\affiliation{
-  %\department{Departamento de Inform\'{a}tica}
-  %\institution{FCT-NOVA, Universidade Nova de Lisboa}
+\affiliation{
+  \department{Departamento de Inform\'{a}tica}
+  \institution{FCT-NOVA, Universidade Nova de Lisboa}
   %\city{City}
-  %\country{Portugal}}
+  \country{Portugal}}
 \affiliation{
@@
  }
- \email{rodrigo@well-typed.com}
+ \email{xxx@yyy.com}
 \author{Bernardo Toninho}
@@
 Appendix~\ref{app:opsem}), written $a\Downarrow^* b$, which allows us
-to state and reason about a progress property.\todo{Na semantica O omega e 1 ainda estão separados, é assim?}
+to state and reason about a progress property.
 
@@
     \item Programs rejected due to the lack of ``linearity coercions'' in
-        $\lambda^\pi_\Delta$. So-called \emph{multiplicity coercions} extending
-        Linear Core could make these programs be understood as linear
+        $\lambda^\pi_\Delta$. Extending Linear Core with \emph{multiplicity coercions} 
+        could make these programs be understood as linear
         (discussed in Section~\ref{sec:future-work}).
@@
 Linear Haskell by understanding linearity semantically in the presence of
-laziness, and by proving linearity-preserving multiple Core-to-Core
-optimisations GHC.
+laziness, and by showing that multiple Core-to-Core
+optimisations employed by GHC are linearity preserving, using a
+linearity-aware semantics that is essentially identical to that of~\cite{cite:linearhaskell}.
 Linear Core can also be seen as a system that validates the programs written in
@@
 linear resources are still used exactly once throughout the optimising
-transformations.\todo{add part about operational semantics and proof being similar to them}
+transformations.
 
@@
 
-Below, we write $\Theta {\uparrow^1}$ for the linear bindings in evaluation
-environment $\Theta$ and $\Theta {\uparrow^\omega}$ for all other bindings.
-$\Theta\downarrow^1_e$ refers to the expressions bound to linear variables in
-the runtime environment and $\Theta\downarrow^1_\sigma$ refers to the types of
-those expressions.
+Below, we write $\Theta {\uparrow^1}$ for the linear bindings in evaluation environment $\Theta$ and $\Theta {\uparrow^\omega}$ for all other bindings.
 
@@
        \left[ \cdot \mid e \right] & \triangleq & e\\
-       \left[ \Theta , \xl = e' \mid e \right] & \triangleq  & [\Theta \mid (\lambda \xl . e)~e']\\
+       \left[ \Theta , x:_1\sigma = e' \mid e \right] & \triangleq  & [\Theta \mid (\lambda x_1:\sigma . e)~e']\\
        \left[\Theta , x:_\Delta \sigma = e' \mid e \right] & \triangleq & [\Theta \mid \llet{x:_\Delta \sigma = e'}{e}]
@@
 
- \begin{lemma}~\label{lem:unwrapenv}
+ \begin{lemma}
    Iff $\G ; \D \vdash [ \Theta \mid e] : \sigma$
@@
    \end{lemma}
-\begin{proof}
-By structural induction on the environment expansion definition.
-\begin{description}
-%\item[Case:] $\left[\cdot \mid e\right] \triangleq e$ Trivial
-\item[Case:] $\left[\Theta, \xl = y \mid e\right] \triangleq \left[\Theta \mid (\lambda \xl. e)~y\right]$
-\begin{tabbing}
-    (1) $\G; \D \vdash \left[ \Theta, \xl = y \mid e \right] : \phi$\\
-    (2) $\G; \D \vdash \left[ \Theta \mid (\lambda \xl. e)~y \right] : \phi$\` by def.\\
-    (3) $\G, \Theta\uparrow^\omega; \D, \Theta\uparrow^1 \vdash ((\lambda \xl. e)~y, \Theta\downarrow^1_e) : \phi \otimes \bigotimes(\Theta\downarrow^1_\sigma)$\` by i.h.\\
-    (4) $Let~\D_1,\D_2,\D_3 = \D, \Theta\uparrow^1 $\\
-    (5) $\G, \Theta\uparrow^\omega; \D_3 \vdash \Theta\downarrow^1_e : \bigotimes(\Theta\downarrow^1_\sigma)$\\
-    (6) $\G, \Theta\uparrow^\omega; \D_2 \vdash y : \s$\` by inv\\
-    (7) $\G, \Theta\uparrow^\omega; \D_1 \vdash (\lambda \xl. e) : \s \to \phi $\` by inv\\
-    (8) $\G, \Theta\uparrow^\omega; \D_1, \xl \vdash e : \phi $\` by inv\\
-    (9) $\G, \Theta\uparrow^\omega; \D_1, \D_2, \xl \vdash (e, y) : \phi \otimes \sigma$\` by pair\\
-    (10) $\G, \Theta\uparrow^\omega; \D, \xl \vdash (e, y, \Theta\downarrow^1_e) : \phi \otimes \sigma \otimes \bigotimes(\Theta\downarrow^1_\sigma)$\` by pair\\
-\end{tabbing}
-\item[Case:] $\left[\Theta, \xD = e' \mid e\right] \triangleq \left[\Theta \mid \llet{\xD = e'}{e}\right]$
-\begin{tabbing}
-    (1) $\G; \D' \vdash \left[ \Theta, \xD = y \mid e \right] : \phi$\\
-    (2) $\G; \D' \vdash \left[ \Theta \mid \llet{\xD = e'}{e} \right] : \phi$\` by def.\\
-    (3) $\G, \Theta\uparrow^\omega; \D', \Theta\uparrow^1 \vdash (\llet{\xD = e'}{e}, \Theta\downarrow^1_e) : \phi \otimes \bigotimes(\Theta\downarrow^1_\sigma)$\` by i.h.\\
-    (4) $Let~\D,\D_2,\D_3 = \D', \Theta\uparrow^1 $\\
-    (5) $\G, \Theta\uparrow^\omega; \D_3 \vdash \Theta\downarrow^1_e : \bigotimes(\Theta\downarrow^1_\sigma)$\\
-    (6) $\G, \Theta\uparrow^\omega; \D \vdash e' : \s$\` by inv on let\\
-    (7) $\G, \Theta\uparrow^\omega, \xD; \D,\D_2 \vdash e : \phi $\` by inv on let\\
-    (8) $\G, \Theta\uparrow^\omega, \xD; \D,\D_2,\D_3 \vdash (e, \Theta\downarrow^1_e) : \phi \otimes \bigotimes(\Theta\downarrow^1_\sigma)$\` by pair\\
-\end{tabbing}
-\end{description}
-\end{proof}
 
@@
 
-\begin{proof}
-By structural induction on the instrumented operational semantics reductions.
-\begin{description}
-    \item[Case:] $\Gamma; \Delta \vdash (\Theta \mid \ccase{e}{z{:}_{\Delta'}\s~\{K~\ov{y_i}\to e'}\}) \Downarrow (\Theta'' \mid v) : \tau$
-\begin{tabbing}
-    (1) $\Gamma; \Delta \vdash \Theta \mid \ccase{e}{z{:}_{\Delta_a}\s~\{K~\ov{y_i}\to e'}\} : \tau, \Sigma$\\
-    (2) $\G, \Theta\uparrow^\omega; \D, \Theta\uparrow^1 \vdash (\ccase{e}{z{:}_{\Delta_a}\s~\{K~\ov{y_i}\to e'}\}, \Theta\downarrow^1_e, \Sigma) : \tau \otimes \bigotimes(\Theta\downarrow^1_\sigma) \otimes \bigotimes(\Sigma)$\`by \ref{lem:unwrapenv}\\
-    (3) $Let~\G' = \G, \Theta\uparrow^\omega; \D_a,\D_b,\D_2,\D_3 = \D,\Theta\uparrow^1$\`s.t. $(\D_2,\D_3)$ types $(\Theta\downarrow^1_e, \Sigma)$\\
-    (4) $\G'; \D_a, \D_b \vdash \ccase{e}{z{:}_{\D_a}\s\{K~\ov{y_i}\to e'\}} : \tau$\\
-    Subcase $e$ is in WHNF (thus $= K~\ov{x_i}$)\\
-    (5) $\G'; \D_a \Vdash K~\ov{x_i} : \s \gtrdot \ov{\D_i}$\` by subcase\\
-    (6) $\G', z{:}_{\D_a}\s; \D_a, \D_b \vdash_{alt} K~\ov{y_i} \to e' :^z_{\D_i} \s \Rightarrow \tau$\` by inv\\
-    Subsubcase $\vdash_{alt}$ $Alt_{NWHNF}$\\
-    (7) $\G', z{:}_{\D_a}\s, \ov{y_i{:}_\omega\s_i}, \ov{y_j{:}_{\Delta_j}\s_j}; \D_a, \D_b \vdash e' : \tau$\` by inv\\
-    (8) $\Theta' = \Theta$\` in this subcase because $(\Theta \mid K~\ov{x_i})\Downarrow(\Theta \mid K~\ov{x_i})$\\
-    (9) $\G'; \ov{\D_i} \vdash K~\ov{x_i} : \s$\` by inv (5)\\
-    (10) $\ov{\G'; \cdot \vdash x_i : \s_i}$\` by inv (9) for unrestricted $x$s\\
-    (11) $\ov{\G'; \D_j \vdash x_j : \s_j}$\` by inv (9) for linear fields $x$s\\
-    (12) $\G', z{:}_{\D_a}\s; \D_a, \D_b \vdash e'\ov{[x_{i,j}/y_{i,j}] : \tau}$\` by $\D$-subst and $\omega$-subst lemmas (7,10,11)\\
-    (13) $\G'; \D_a, \D_b \vdash e'\ov{[x_{i,j}/y_{i,j}]}[K~\ov{x_{i,j}}/z] : \tau$\` by $\D$-subst lemma and $\ov{\D_i} = \D_a$ (9,12)\\
-    (14) $\G; \D \vdash \Theta \mid e'\ov{[x_{i,j}/y_{i,j}]}[K~\ov{x_{i,j}}/z] : \tau, \Sigma$\` by reconstructing\\
-    (15) $\G; \D \vdash \Theta'' \mid v : \tau, \Sigma$\` by induction hypothesis\\
-    Subsubcase $\vdash_{alt}$ $Alt_{0}$\\
-    (7) $(\G', z{:}_{\D_a}\s)[\cdot/\D_a]_z, \ov{y_i{:}_\omega\sigma_i}; (\D_a, \D_b)[\cdot / \D_a] \vdash e' : \tau$\` by inv subcase\\
-    (8) $\G', z{:}_{\cdot}, \ov{y_i{:}_\omega\sigma_i}; \D_b \vdash e' : \tau$\` by $z$ cannot occur in $\G'$\\
-    (9) Since $\G';\D_a \Vdash K~\ov{x_i}$, in this subcase $\D_a = \cdot$\\
-    (10) $\G', \z{:}_\cdot\sigma, \ov{y{:}_\omega\sigma}; \D_b \vdash e' : \tau$\` by (9)\\
-    (11) $\G', \z{:}_\cdot\sigma; \D_b \vdash e'\ov{[x_i/y_i]} : \tau$\` by $\omega$-subst (10)\\
-    (12) $\G'; \D_b \vdash e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] : \tau$\` by $\D$-subst\\
-    (13) $\G; \D \vdash \Theta \mid e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] : \tau, \Sigma$\` by reconstructing\\
-    (14) $\G; \D \vdash \Theta'' \mid v : \tau, \Sigma$\`by induction hypothesis\\
-    Subcase $e$ is not in WHNF\\
-    (5) $\G'; \D_a \vdash e : \s$\` by inv on case\\
-    (6) $\G', z{:}_{\irr{\D_a}}\s; \irr{\D_a}, \D_b \vdash_{alt} K~\ov{y_i} \to e' :^z_{\irr{\D_a}} \s $\` by inv on case\\
-\end{tabbing}
-\end{description}
-\end{proof}
-
 We also derive type preservation for the lazy natural semantics.

### 2025-02-27 icfp.tex
touched up type pres proof

@@
  \begin{lemma}~\label{lem:unwrapenv}
-   Iff $\G ; \D \vdash [ \Theta \mid e] : \sigma$
-     then $\G , \Theta\uparrow^{\omega} ; \D , \Theta\uparrow^1 \vdash  (e, \Theta\downarrow^1_e) : \sigma \otimes \bigotimes(\Theta\downarrow^1_\sigma)$
+   $\G ; \D \vdash [ \Theta \mid e] : \sigma$
+     iff $\G , \Theta\uparrow^{\omega} ; \D , \Theta\uparrow^1 \vdash  (e, \Theta\downarrow^1_e) : \sigma \otimes \bigotimes(\Theta\downarrow^1_\sigma)$
    \end{lemma}
 \begin{proof}
-By structural induction on the environment expansion definition.
+By structural induction on the environment expansion definition in the
+forward direction and by a straightforward induction on $\Theta$ in the
+converse direction. We show only the forward direction.
 \begin{description}
@@
 \begin{proof}
-By structural induction on the instrumented operational semantics reductions.
+By structural induction on the instrumented operational semantics. We
+omit many trivial cases.
 \begin{description}
@@
     (5) $\G'; \D_1 \vdash y : \s$\\
-    (6) $\G'; \D_1, \D_2, \D_3 \vdash (y, \Theta{\downarrow}^1_e, \Sigma) : \s \bigotimes(\Theta{\downarrow^1_e}) \bigotimes(\Sigma)$
-    (7) $\G; \D \vdash \Theta \mid y : \s, \Sigma$\`by reconstructing\\
+    (6) $\G'; \D_1, \D_2, \D_3 \vdash (y, \Theta{\downarrow}^1_e, \Sigma) : \s \bigotimes(\Theta{\downarrow^1_e}) \bigotimes(\Sigma)$\\
+    (7) $\G; \D \vdash \Theta \mid y : \s, \Sigma$\`by well-typed state\\
     (8) $\G; \D \vdash \Theta' \mid v : \s, \Sigma$\` by induction hypothesis\\
@@
     (8) $\G, \Theta{\uparrow}^\omega; \D', \D'' \vdash (e, \Theta{\downarrow}^1_e, \Sigma) : \tau \bigotimes(\Theta{\downarrow}^1_\s) \bigotimes(\Sigma)$\` by pair (4,7)\\
-    (9) $\G; \D \vdash \Theta \mid e : \tau,\Sigma$\` by reconstructing\\
+    (9) $\G; \D \vdash \Theta \mid e : \tau,\Sigma$\` by well-typed state\\
     (10) $\G; \D \vdash \Theta \mid v : \tau,\Sigma$\` by induction hypothesis\\
@@
     (12) $\G'; \D_a \vdash \llet{x{:}_{\D_a}\s = v}{v} : \s$\`by let (11a,11a)\\
-    (13) $\G; \D \vdash \Theta' \mid \llet{x{:}_{\D_a}\s = v}{v} : \s, \Sigma$\`by reconstructing\\
+    (13) $\G; \D \vdash \Theta' \mid \llet{x{:}_{\D_a}\s = v}{v} : \s, \Sigma$\`by well-typed state\\
     (14) $\G; \D \vdash \Theta', x{:}_{\D_a}\s = v \mid v : \s, \Sigma$\`by env. expansion\\
@@
     (13) $\G'; \D_a, \D_b \vdash e'\ov{[x_{i,j}/y_{i,j}]}[K~\ov{x_{i,j}}/z] : \tau$\` by $\D$-subst lemma and $\ov{\D_i} = \D_a$ (9,12)\\
-    (14) $\G; \D \vdash \Theta \mid e'\ov{[x_{i,j}/y_{i,j}]}[K~\ov{x_{i,j}}/z] : \tau, \Sigma$\` by reconstructing\\
+    (14) $\G; \D \vdash \Theta \mid e'\ov{[x_{i,j}/y_{i,j}]}[K~\ov{x_{i,j}}/z] : \tau, \Sigma$\` by well-typed state\\
     (15) $\G; \D \vdash \Theta'' \mid v : \tau, \Sigma$\` by induction hypothesis\\
@@
     (12) $\G'; \D_b \vdash e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] : \tau$\` by $\D$-subst\\
-    (13) $\G; \D \vdash \Theta \mid e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] : \tau, \Sigma$\` by reconstructing\\
+    (13) $\G; \D \vdash \Theta \mid e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] : \tau, \Sigma$\` by well-typed state\\
     (14) $\G; \D \vdash \Theta'' \mid v : \tau, \Sigma$\`by induction hypothesis\\
@@
     (9) $\G', z{:}_{\ov{y_j{:}_1\s_j}}\s, \ov{y_i{:}_\omega\s_i}; \D_a, \D_b, \ov{y_j{:}_1\s_j} \vdash (e, e') : \sigma \otimes \tau$\` by pair (5,8)\\
-    (10) $\G, z{:}_{\ov{y_j{:}_1\s_j}}\s, \ov{y_i{:}_\omega\s_i}; \D, \ov{y_j{:}_1\s_j} \vdash \Theta \mid e : \s, e':\tau, \Sigma$\`by reconstructing\\
+    (10) $\G, z{:}_{\ov{y_j{:}_1\s_j}}\s, \ov{y_i{:}_\omega\s_i}; \D, \ov{y_j{:}_1\s_j} \vdash \Theta \mid e : \s, e':\tau, \Sigma$\`by well-typed state\\
     (11) $\G, z{:}_{\ov{y_j{:}_1\s_j}}\s, \ov{y_i{:}_\omega\s_i}; \D, \ov{y_j{:}_1\s_j} \vdash \Theta' \mid K~\ov{x_i} : \s, e':\tau, \Sigma$\`by induction hypothesis\\
@@
     (17) $\G''; \D_c, \D_d \vdash e'\ov{[x_{i,j}/y_{i,j}]}[K~\ov{x_{i,j}}/z] : \tau$\`by $\D$-subst lemma (13a,16)\\
-    (18) $\G; \D \vdash \Theta' \mid e'\ov{[x_{i,j}/y_{i,j}]}[K~\ov{x_{i,j}}/z] : \tau, \Sigma$\`by reconstructing\\
+    (18) $\G; \D \vdash \Theta' \mid e'\ov{[x_{i,j}/y_{i,j}]}[K~\ov{x_{i,j}}/z] : \tau, \Sigma$\`by well-typed state\\
     (19) $\G; \D \vdash \Theta'' \mid v : \tau, \Sigma$\`by induction hypothesis\\
@@
     (8) $\G', z{:}_\cdot\s, \ov{y_i{:}_\omega\s_i}; \D_a, \D_b \vdash (e, e') : \sigma \otimes \tau$\` by pair (5,7)\\
-    (9) $\G, z{:}_\cdot\s, \ov{y_i{:}_\omega\s_i}; \D \vdash \Theta \mid e : \sigma, e' : \tau, \Sigma$\` by reconstructing\\
-    (10) $\G, z{:}_\cdot\s, \ov{y_i{:}_\omega\s_i}; \D \vdash \Theta \mid e : \sigma, e' : \tau, \Sigma$\` by reconstructing\\
-    (11) $\G, z{:}_\cdot\s, \ov{y_i{:}_\omega\s_i}; \D \vdash \Theta' \mid K~\ov{x_i} : \sigma, e' : \tau, \Sigma$\` by induction hypothesis\\
-    (12) $Let~\G'' = \G,\Theta'{\uparrow}^\omega~and~\D_d,\D_4,\D_5 = \D, \Theta'{\uparrow}^1$\` where $(\D_4, \D_5)$ types $(\Theta'{\downarrow^1_e}, \Sigma)$\\
-    (12a) $\G''; \cdot \vdash K~\ov{x_i} : \s$\`since $K$ has no linear fields in this subcase\\
-    (12b) $\G'', z{:}_{\cdot}\s, \ov{y_i{:}_\omega\s_i}; \D_d \vdash e' : \tau$\\
-    (13) $\ov{\G''; \cdot \vdash x_{i} : \s_i}$\`by inv (12a)\\
-    (14) $\G'', z{:}_{\cdot}\s; \D_d \vdash e'\ov{[x_i/y_i]} : \tau$\` by $\omega$-subst lemma\\
-    (15) $\G''; \D_d \vdash e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] : \tau$\` by $\D$-subst lemma\\
-    (16) $\G; \D \vdash \Theta' \mid e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] : \tau, \Sigma$\` by reconstructing\\
-    (17) $\G; \D \vdash \Theta'' \mid v : \tau, \Sigma$\` by induction hypothesis\\
+    (9) $\G, z{:}_\cdot\s, \ov{y_i{:}_\omega\s_i}; \D \vdash \Theta \mid e : \sigma, e' : \tau, \Sigma$\` by well-typed state\\
+    % (10) $\G, z{:}_\cdot\s, \ov{y_i{:}_\omega\s_i}; \D \vdash \Theta \mid e : \sigma, e' : \tau, \Sigma$\` by reconstructing\\
+    (10) $\G, z{:}_\cdot\s, \ov{y_i{:}_\omega\s_i}; \D \vdash \Theta' \mid K~\ov{x_i} : \sigma, e' : \tau, \Sigma$\` by induction hypothesis\\
+    (11) $Let~\G'' = \G,\Theta'{\uparrow}^\omega~and~\D_d,\D_4,\D_5 = \D, \Theta'{\uparrow}^1$\` where $(\D_4, \D_5)$ types $(\Theta'{\downarrow^1_e}, \Sigma)$\\
+    (11a) $\G''; \cdot \vdash K~\ov{x_i} : \s$\`since $K$ has no linear fields in this subcase\\
+    (11b) $\G'', z{:}_{\cdot}\s, \ov{y_i{:}_\omega\s_i}; \D_d \vdash e' : \tau$\\
+    (12) $\ov{\G''; \cdot \vdash x_{i} : \s_i}$\`by inv (11a)\\
+    (13) $\G'', z{:}_{\cdot}\s; \D_d \vdash e'\ov{[x_i/y_i]} : \tau$\` by $\omega$-subst lemma\\
+    (14) $\G''; \D_d \vdash e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] : \tau$\` by $\D$-subst lemma\\
+    (15) $\G; \D \vdash \Theta' \mid e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] :
+    \tau, \Sigma$\` by well-typed state\\
+    (16) $\G; \D \vdash \Theta'' \mid v : \tau, \Sigma$\` by induction hypothesis\\
 \end{tabbing}
@@
     (8) $\G'; \D_a, \D_b, z{:}_1\s \vdash (e, e') : \sigma \otimes \tau$\` by pair\\
-    (9) $\G; \D, z{:}_1\s \vdash \Theta \mid e : \sigma, e' : \tau, \Sigma$\` by reconstructing\\
+    (9) $\G; \D, z{:}_1\s \vdash \Theta \mid e : \sigma, e' : \tau,
+    \Sigma$\` by well-typed state\\
     (10) $\G; \D, z{:}_1\s \vdash \Theta' \mid K~\ov{x_i} : \sigma, e' : \tau, \Sigma$\` by induction hypothesis\\
@@
     (14) $\G';\D_c, \D_d \vdash e'[K~\ov{x_i}/z] : \tau$\` by $1$-subst lemma\\
-    (15) $\G;\D \vdash \Theta' \mid e'[K~\ov{x_i}/z] : \tau, \Sigma$\` by reconstructing\\
+    (15) $\G;\D \vdash \Theta' \mid e'[K~\ov{x_i}/z] : \tau, \Sigma$\` by well-typed state\\
     (16) $\G;\D \vdash \Theta'' \mid v : \tau, \Sigma$\` by induction hypothesis\\

### 2025-07-09 icfp_submission_2025.tex
save submission 2025 as separate file

--- 2025-02-27 icfp.tex
@@
 intermediate language of the Glasgow Haskell Compiler. We prove Linear Core
-is sound and guarantees linear resource usage and prove multiple optimising
+is sound and guarantees linear resource usage, and prove multiple optimising
 transformations preserve linearity in Linear Core while failing to do so in
@@
         Intuitively, a syntactic occurrence of a linear resource does not
-        necessarily entail consuming that resource since expressions are not
-        necessarily evaluated (i.e., if the expression is not evaluated, the
+        necessarily entail consuming that resource because expressions are not
+        necessarily evaluated (if the expression is not evaluated, the
         resource is not used).
         %
-        Under eager evaluation, a syntactic occurrence of a resource
+        In contrast, under eager evaluation, a syntactic occurrence of a resource
         along a given control-flow path always implies its use at runtime.

### 2025-07-09 icfp.tex
revert

@@
 %
-Core is a minimal, explicitly typed, functional language to which multiple
+Core is a minimal, explicitly typed, lazy functional language to which multiple
 Core-to-Core optimising transformations are applied during compilation.
@@
         % -- expressions in Haskell are only evaluated when needed.
-        As we make clear in this paper, \emph{non-strict evaluation interacts non-trivially with linearity}.
+        As we make clear in this paper, non-strict evaluation interacts non-trivially with linearity.
+        We give a precise definition of what it 
         Intuitively, a syntactic occurrence of a linear resource does not
-        necessarily entail consuming that resource since expressions are not
-        necessarily evaluated (i.e., if the expression is not evaluated, the
+        necessarily entail consuming that resource because expressions are not
+        necessarily evaluated (if the expression is not evaluated, the
         resource is not used).
         %
-        Under eager evaluation, a syntactic occurrence of a resource
+        In contrast, under eager evaluation, a syntactic occurrence of a resource
         along a given control-flow path always implies its use at runtime.

### 2025-07-09 icfp.tex
Update part about semantic linearity intuition and precise semantics in intro

@@
 
-Both Haskell and its intermediate language Core are lazily
-        evaluated.
-        % -- expressions in Haskell are only evaluated when needed.
-        As we make clear in this paper, non-strict evaluation interacts non-trivially with linearity.
-        We give a precise definition of what it 
-        Intuitively, a syntactic occurrence of a linear resource does not
-        necessarily entail consuming that resource because expressions are not
-        necessarily evaluated (if the expression is not evaluated, the
-        resource is not used).
-        %
-        In contrast, under eager evaluation, a syntactic occurrence of a resource
-        along a given control-flow path always implies its use at runtime.
+% Both Haskell and its intermediate language Core are lazily
+%         evaluated.
+%         % -- expressions in Haskell are only evaluated when needed.
+%         As we make clear in this paper, non-strict evaluation interacts non-trivially with linearity.
+%         We give a precise definition of what it 
+%         Intuitively, a syntactic occurrence of a linear resource does not
+%         necessarily entail consuming that resource because expressions are not
+%         necessarily evaluated (if the expression is not evaluated, the
+%         resource is not used).
+%         %
+%         In contrast, under eager evaluation, a syntactic occurrence of a resource
+%         along a given control-flow path always implies its use at runtime.
         %
@@
 In this work, we address this limitation by exploring how syntactic linearity
-breaks down in the presence of non-strictness and designing a type system which,
-derived from our observations, accepts the vast majority of valid non-strict
-programs and optimising transformations previously rejected by traditional
-linear typing disciplines. Additionally, we show our system
+breaks down in the presence of non-strictness and designing a linear type
+system which accepts the vast majority of programs and optimising
+transformations that, while correct under non-strict evaluation, are rejected
+by traditional linear typing disciplines.
+%
+%
+Additionally, we show our system
 %
@@
 (Section~\ref{sec:linearity-semantically}).
+%
+A precise definition of \emph{semantic} linearity is given by our instrumented
+operational semantics, under which a program using a linear resource
+non-linearly gets stuck (Section~\ref{sec:main:metatheory}).
 
@@
 best of our knowledge, this is the first type system to capture linearity
-semantically in this context
-(Section~\ref{sec:main:linear-core}).
+semantically in this context (Section~\ref{sec:main:linear-core}).
 

### 2025-07-10 icfp.tex
Clarify case expression WHNF judgment

@@
 or as constructor arguments need not be evaluated for the lambda or constructor
-application to be in WHNF. % (if otherwise all sub-expressions
+application to be in WHNF % (if otherwise all sub-expressions
 % were fully evaluated the whole expression would also be in \emph{normal form}).
@@
 %
-Accordingly, these sub-expressions might still depend on linear resources to be
-well-typed and such resources will only be consumed when the sub-expressions themselves are evaluated.
-
-As will be made clear in this section, we need to devise a specialized typing
-for scrutinees that distinguishes between terms
-in WHNF and terms that are not in WHNF.
-Following the discussion on expressions in WHNF, we present a
-typing judgement $\G;\D \Vdash e : \s \gtrdot \ov{\D_i}$, and a rule for each
-of the forms given above:
+(and they may capture linear resources).
+
+A key detail is that when the scrutinee is already in WHNF, evaluation continues
+in the alternative simply by substituting the pattern variables by the
+scrutinee sub-expressions. No resource from the scrutinee can be consumed since
+no computation happens at all.
+% (the scrutinee was already in weak head normal form).
+%In short, no computation happens if the scrutinee is already in WHNF.
+%
+% In light of this, our type system will need to distinguish between terms in WHNF and terms
+% that are not in WHNF.
+%
+% TODO While it may seem ad-hoc to type an expression differently
+%
+We introduce a typing judgement $\G;\D \Vdash e : \s \gtrdot \ov{\D_i}$ to
+extract additional information from the static structure of terms in WHNF:
 \[
@@
 
-We highlight that, when evaluating a case expression, computation only
-happens when a scrutinee that is not in WHNF is evaluated to
-WHNF. Otherwise, evaluation continues in the alternative by
-substituting in the appropriate scrutinee expressions, but without
-having performed any additional computation.
-% (the scrutinee was already in weak head normal form).
-%In short, no computation happens if the scrutinee is already in WHNF.
-%
-In terms of linearity, resources are consumed only if evaluation happens.
-Therefore, resources used to type a scrutinee not in
-WHNF will be consumed when the case is evaluated and will no longer be directly available in the case
-alternatives. Conversely, when the scrutinee is already in WHNF, linear
-resources required to type the scrutinee are still directly available in the
-alternatives.
-%
 %%%%???

### 2025-07-10 icfp.tex
add reference to fractional permissions
Fix typos

@@
 \vp$ is used to type case alternatives, encompassing three
-``modes'' that are distinguished by the kind of arrow that is used:
+modes that are distinguished by the kind of arrow that is used:
 for alternatives of case expressions whose scrutinee is in WHNF
@@
 First, an alternative whose pattern is a constructor with $n > 0$ linear components
-under the $\Rightarrow_{\mathsf{NWHNF}}$ mode is typed as:
+under the $\Rightarrow_{\mathsf{WHNF}}$ mode is typed as:
 \[
@@
 
-To satisfy these two constraints, we introduce a $\textit{Split}$ rule that
+To satisfy these two constraints, we introduce a $\textsc{Split}$ rule that
 allows a linear resource $x{:}_1\s$ to be split into $n$ resources at a given
@@
 variable for a constructor's linear component can go unused (the tagged environments are disjoint).
+%
+Tagged resources are inspired by fractional permissions in separation logic~\cite{10.5555/1760267.1760273}.
 
@@
 %
-noting that $Split$ can be applied both to relevant and proof irrelevant linear
+noting that $\textsc{Split}$ can be applied both to relevant and proof irrelevant linear
 resources in $\D$.
@@
 variable. To use these pattern variables, the scrutinee resources must be first
-$Split$ into fragments. Instead of guessing which resources need to be split,
+$\textsc{Split}$ into fragments. Instead of guessing which resources need to be split,
 the implementation consumes tagged fragments of a resource \emph{as needed},
 i.e., only when consuming a resource whose usage environment has a resource
-tagged $K_j$ do we $Split$ that resource on $K$ and consume the $K_j$ fragment.
+tagged $K_j$ do we $\textsc{Split}$ that resource on $K$ and consume the $K_j$ fragment.
 %

### 2025-07-11 popl26.tex
made popl26 file

--- 2025-07-10 icfp.tex
@@
 
-To determine whether the usage of a binding from a recursive group is
-linear,
-we must infer the superset of resources that may end up being consumed
-in a single use of any of the bindings. For a strongly connected recursive
-group, this superset is the same for all bindings by definition.
-%
-Intuitively, this inference amounts to finding a set of linear resources
-($\Delta$) that satisfies the set of equations derived from the bindings
-$\ov{x_i}$ in the group such that
-%\begin{enumerate}
-occurrences of $x_i$ in its own body are synonymous with using all resources in $\Delta$ exactly once.
+To determine whether the usage of a binding from a recursive group is linear,
+we must infer the superset of resources that may end up being consumed in a
+single use of any of the bindings. For a strongly connected recursive group,
+this superset is the same for all bindings by definition.
+%
+% Intuitively, this inference amounts to finding the linear resources
+% ($\Delta$) which satisfy set of equations derived from the bindings
+% $\ov{x_i}$ in the group such that
+% %\begin{enumerate}
+% occurrences of $x_i$ in its own body are synonymous with using all resources in $\Delta$ exactly once.
 %(2) and if the binding $x$ is fully evaluated, then all resources in $\Delta$ are consumed exactly once.
@@
 %
-It follows from the solution being the same for all bindings that
-at most a single variable from the recursive group can be used in the continuation.
+It follows from the inferred resources used being the same for all bindings
+that at most a single variable from the recursive group can be used in the
+continuation.
 %

### 2025-07-11 popl26.tex
related work and such

@@
 \usepackage[justification=centering]{subfig}
+
+%% Bibliography style
+\bibliographystyle{ACM-Reference-Format}
+%% Citation style
+%% Note: author/year citations are required for papers published as an
+%% issue of PACMPL.
+\citestyle{acmauthoryear}   %% For author/year citations
 
@@
 \begin{abstract}
-Linear type systems guarantee linear resources are used exactly once.
+Linear type systems can be used to guarantee (linear) resources are used exactly once.
 Traditionally, using a resource is synonymous with its \emph{syntactic}
@@
 Linear types~\cite{cite:linear-logic,cite:barberdill} increase the expressiveness
-of type systems by allowing them to enforce that certain resources (e.g.~a
-file handle) are used \emph{exactly once}.
+of type systems by allowing them to enforce desirable usage patterns
+for certain resources (e.g.~a
+file handle). Typically, linear types can be used to enforce resources are used \emph{exactly once}.
 %
@@
 \begin{code}
-let p = malloc(4); in free(p); free(p);
+let p = malloc(4) in free(p); free(p);
 \end{code}
@@
 %
-Consequently, few general purpose programming languages have linear
+Consequently, few general purpose programming languages have linear or substructural
 type systems. Among them are Idris~2~\cite{brady:LIPIcs.ECOOP.2021.9},
@@
 
-However, linearity in Haskell stands out from linearity in other
+Linearity in Haskell stands out from linearity in other
 languages because linear types permeate Haskell down to (its) Core, the
@@
 %
-Core is a minimal, explicitly typed, lazy functional language to which multiple
-Core-to-Core optimising transformations are applied during compilation.
+Core is a minimal, explicitly typed, lazy functional language with
+both linear and unrestricted types, to which multiple
+Core-to-Core optimising transformations are applied during
+compilation. Notably Core does not have a (higher-order or otherwise)
+store.
 %
@@
 
-Just as Core's type system provides a degree of validation to the
+Just as Core's type system \cite{10.1145/1190315.1190324} provides a degree of validation to the
 translation from Haskell (dubbed \emph{desugaring}) and the subsequent
@@
 
-Despite the advantages of a linearly typed Core, the status quo is
+Despite the advantages of adding linear types to Core, the status quo is
 that linearity is effectively ignored in Core, being only checked
@@
 While compiler optimisations aim to preserve the semantics of
-programs and therefore a program's linear resource consumption,
+programs and therefore a program's resource consumption,
 a \emph{naive} syntactic analysis of the optimised programs (as
@@
 As a trivial example, let $x$ be a linear resource in the next two expressions,
-where the latter results from inlining $y$ without eliminating the binder. Even
+where the latter results from inlining $y$ without eliminating the
+binder (a common optimising transformation in GHC). Even
 though the second expression no longer appears linear (as there are now two occurrences
 of $x$), it is indeed linear because the let-bound expression is never
-evaluated and $x$ is still consumed exactly once (when it is freed in the let body):
+evaluated under \emph{lazy evaluation} and $x$ is still consumed exactly once (when it is freed in the let body):
 % executed, $x$ will be freed exactly once:
@@
 The Core optimising transformations expose a fundamental limitation of Core's
-linear type system: it does not account for the non-strict evaluation model
+type system: it does not account for the non-strict evaluation model
 of Core and, thus, a whole class of programs that are linear under the lens of
@@
 In this work, we address this limitation by exploring how syntactic linearity
-breaks down in the presence of non-strictness and designing a linear type
-system which accepts the vast majority of programs and optimising
+breaks down in the presence of non-strictness and design a linear type
+system for Core which accepts the vast majority of programs and optimising
 transformations that, while correct under non-strict evaluation, are rejected
@@
 
-A linear type system statically guarantees that linear resources are
+A linear type system typically guarantees that linear resources are
 consumed exactly once. Consequently, whether a program is
-well-typed under a linear type system intrinsically depends on the
+well-typed under this lens intrinsically depends on the
 definition of consuming a resource.
@@
 %
-For instance, in the following Haskell-like program, there are two syntactic
+For instance, in the following Haskell-like program defining a linear
+function \lstinline{f} (i.e., a function that should use its argument
+exactly once), there are two syntactic
 occurrences of \lstinline{handle}: one is in a let-bound computation that
@@
 the handle will not be closed because the computation that closes the handle is
-never evaluated (\lstinline{x} goes unused).
+never evaluated (the non-linear \lstinline{x} goes unused).
 %
@@
 semantics should accept the above program, unlike a linear type system for the
-same program if it were to be evaluated eagerly.
+same program if it were to be evaluated eagerly, where the let-bound
+expression would always be evaluated.
 %
@@
 non-strict evaluation are keen to write programs with bindings that are selectively
-used in case alternatives, such as:
+used in case alternatives, such as (we write $\lolli$ for linear
+functions and \lstinline{->} for non-linear functions):
 %
@@
 because the linear resource \lstinline{x} is used exactly once regardless of which branch is
-taken: either directly in the \lstinline{True} branch, or \emph{via} the
-let-bound \lstinline{y} in the \lstinline{False} branch.
+taken: either directly in the \lstinline{True} branch, or \emph{via}
+the use of the let-bound \lstinline{y} in the \lstinline{False} branch.
 
@@
 
-In this section, we develop a linear calculus $\lambda_\Delta^\pi$,
+In this section, we develop a pure linear calculus $\lambda_\Delta^\pi$,
 dubbed \emph{Linear Core}, that combines the linearity-in-the-arrow
@@
 
+Note that since the operational semantics gets stuck if a linear
+variable is used more than once, progress ensures that linear
+functions consume their arguments at most once if their result is
+consumed exactly once. Type preservation ensures that evaluation of a
+closed term returns an evaluation environment with no linear variables.
+
   \begin{theorem}[Type Preservation]
@@
   \end{theorem}
+
 \begin{theorem}[Progress]
@@
 
+\paragraph{Linearity-aware Semantics}
+Several other works explore linearity aware semantics,
+dating back to \cite{Chirimar_Gunter_Riecke_1996} and
+\cite{TURNER1999231}.
+The former uses a computational interpretation of linear logic to give
+an account of reference counting. The latter present a heap-based
+operational interpretation of linear logic and explore memory usage
+properties under call-by-name and call-by-need, showing both satisfy
+different properties under this lens. Their semantics is similar to
+ours, but does not account for pattern matching.
+More recently, \cite{DBLP:journals/pacmpl/ChoudhuryEEW21} and
+\cite{cite:linearhaskell} both make use of usage-aware
+semantics. Choudhury et al. do so in the context of graded dependent
+type theory, using the usage-aware heap-based operational semantics to
+show correct accounting of resource usage in their setting, as we do
+in ours, but they do not model
+call-by-need. \citet{cite:linearhaskell} provide a similar semantics
+to ours, but present only a naive linear typing system that does not
+leverage non-strict evaluation.
+
 \paragraph{Linear Mini-Core\label{sec:linear-mini-core}}
@@
 
-\bibliographystyle{ACM-Reference-Format}
 \bibliography{references}

### 2025-10-07 popl26-accepted-before-rebuttal.tex
Copy popl26 to popl26-accepted-before-rebuttal.tex

\documentclass[acmsmall,review,anonymous,screen]{acmart}
\PassOptionsToPackage{dvipsnames}{xcolor}
\usepackage{boldline}
\usepackage{xcolor}
\usepackage{xargs}
\usepackage{framed}
\usepackage[framemethod=default]{mdframed}
\usepackage{mathpartir}
\usepackage{mathtools}
\usepackage{soul}
\usepackage{cmll} % ROMES:TODO: not allowed?
\usepackage{amsthm}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{cleveref}
\usepackage{makecell}
\usepackage{fancyvrb}
\usepackage{thmtools, thm-restate}
\usepackage{listings}
\usepackage{todonotes}
\usepackage[justification=centering]{subfig}

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
\citestyle{acmauthoryear}   %% For author/year citations

%\newtheorem{theorem}{Theorem}
%\newtheorem{proposition}[theorem]{Proposition}
% \newtheorem{lemma}[theorem]{Lemma}%[theorem]
%\newtheorem{lemma}{Lemma}%[theorem]
%\newtheorem{sublemma}{Lemma}[lemma]
%\newtheorem{assumption}{Assumption}

%%%%%%%%%%%%%%  Color-related things   %%%%%%%%%%%%%%

% Work around lhs2tex#82
\let\Bbbk\undefined

%include polycode.fmt
%format ⊸ = "\lolli"

% don't, for now
%%%%subst keyword a = "\textcolor{BlueViolet}{\textbf{" a "}}"

\newcommand{\myFor}[2]{\For{$#1$}{$#2$}}
\newcommand{\id}[1]{\textsf{\textsl{#1}}}

\newcommand{\incode}[1]{\lstinline{#1}}

% \renewcommand{\Varid}[1]{\textcolor{Sepia}{\id{#1}}}
% \renewcommand{\Conid}[1]{\textcolor{OliveGreen}{\id{#1}}}

%%%%%%%%%%%%  End of Color-related things   %%%%%%%%%%%%

% It might make sense to add pretty formating of individual things
% like "forall", cf.
% https://github.com/goldfirere/thesis/blob/master/tex/rae.fmt

% colorboxes, from rae's thesis as well
\definecolor{working}{rgb}{0.9,1,0.9}
\newmdenv[hidealllines=true,backgroundcolor=working,innerleftmargin=0pt,innerrightmargin=0pt,innertopmargin=-3pt,innerbottommargin=-3pt,skipabove=3pt,skipbelow=3pt]{working}
\newcommand{\workingcolorname}{light green}

\definecolor{notyet}{rgb}{1,1,0.85}
\newmdenv[hidealllines=true,backgroundcolor=notyet,innerleftmargin=0pt,innerrightmargin=0pt,innertopmargin=-3pt,innerbottommargin=-3pt,skipabove=3pt,skipbelow=3pt]{notyet}
\newcommand{\notyetcolorname}{light yellow}

\definecolor{noway}{rgb}{1,0.9,0.9}
\newmdenv[hidealllines=true,backgroundcolor=noway,innerleftmargin=0pt,innerrightmargin=0pt,innertopmargin=-3pt,innerbottommargin=-3pt,skipabove=3pt,skipbelow=3pt]{noway}
\newcommand{\nowaycolorname}{light red}

\definecolor{limitation}{rgb}{1.0, 0.875, 0.75}
\newmdenv[hidealllines=true,backgroundcolor=limitation,innerleftmargin=0pt,innerrightmargin=0pt,innertopmargin=-3pt,innerbottommargin=-3pt,skipabove=3pt,skipbelow=3pt]{limitation}
\newcommand{\limitationcolorname}{light orange}

\DefineVerbatimEnvironment{code}{Verbatim}{fontsize=\small}
\DefineVerbatimEnvironment{example}{Verbatim}{fontsize=\small}

\newcommand{\parawith}[1]{\paragraph{\emph{#1}}}
\newcommand{\lolli}{\multimap}
\newcommand{\tensor}{\otimes}
\newcommand{\one}{\mathbf{1}}
\newcommand{\bang}{{!}}

\newcommand{\llet}[2]{\mathsf{let}~#1~\mathsf{in}~#2}
\newcommand{\lletrec}[2]{\mathsf{letrec}~#1~\mathsf{in}~#2}
\newcommand{\ccase}[2]{\mathsf{case}~#1~\mathsf{of}~#2}

\definecolor{darkblue}{rgb}{0,0,0.5}
\definecolor{darkgreen}{rgb}{0,0.3,0}
\definecolor{darkpink}{rgb}{0.4,0,0.3}
\definecolor{graygreen}{rgb}{0.3,0.5,0.3}
\definecolor{grayblue}{rgb}{0.2,0.2,0.6}
\definecolor{grayred}{rgb}{0.5,0.2,0.2}

\lstset{
  frame=none,
  xleftmargin=2pt,
  stepnumber=1,
  %numbers=left,
  %numbersep=5pt,
  numberstyle=\ttfamily\tiny\color[gray]{0.3},
  belowcaptionskip=\bigskipamount,
  captionpos=b,
  escapeinside={*'}{'*},
  language=haskell,
  tabsize=2,
  emphstyle={\bf},
  commentstyle=\it,
  stringstyle=\mdseries\rmfamily,
  showspaces=false,
  keywordstyle=\bfseries\rmfamily\color{darkblue}, 
  columns=flexible,
  basicstyle=\small\sffamily,
  showstringspaces=false,
  morecomment=[l]\%,
}

\input{proof}
\input{language-v2/proof}
\input{language-v3/proof}
\input{language-v4/proof}
\input{language-v4/Syntax}
\input{language-v4/TypingRules}
\input{language-v4/TheoremsAndLemmas}

\title{Semantic Linearity in a Lazy Optimising Compiler}
\subtitle{Type-checking Linearity in GHC Core}

\author{Rodrigo Mesquita}
%\affiliation{
  %\department{Departamento de Inform\'{a}tica}
  %\institution{FCT-NOVA, Universidade Nova de Lisboa}
  %\city{City}
  %\country{Portugal}}
\affiliation{
   \institution{Well-Typed LLP}
  %\city{City}
  %\country{Something}
 }
 \email{rodrigo@well-typed.com}
\author{Bernardo Toninho}
\orcid{0000-0002-0746-7514}
\affiliation{
  \department{Departamento de Inform\'{a}tica}
  \institution{NOVA-LINCS, FCT-NOVA, Universidade Nova de Lisboa}
  %\city{City}
  \country{Portugal}}
\email{btoninho@fct.unl.pt}
%% TODO: AFFILIATION, EMAIL
%% TODO: CCS CONCEPTS and KEYWORDS
%% TODO: ACM RIGHTS notice...
% CHECK: https://www.acm.org/binaries/content/assets/publications/taps/latex-best_practices-06-may-2020.pdf

% \date{ }

\begin{CCSXML}
<ccs2012>
   <concept>
       <concept_id>10003752.10003790.10003801</concept_id>
       <concept_desc>Theory of computation~Linear logic</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
   <concept>
       <concept_id>10003752.10003790.10011740</concept_id>
       <concept_desc>Theory of computation~Type theory</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
   <concept>
       <concept_id>10011007.10011006.10011008.10011009.10011012</concept_id>
       <concept_desc>Software and its engineering~Functional languages</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
   <concept>
       <concept_id>10011007.10011006.10011041</concept_id>
       <concept_desc>Software and its engineering~Compilers</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
 </ccs2012>
\end{CCSXML}

\ccsdesc[500]{Theory of computation~Linear logic}
\ccsdesc[500]{Theory of computation~Type theory}
\ccsdesc[500]{Software and its engineering~Functional languages}
\ccsdesc[500]{Software and its engineering~Compilers}

\keywords{Linear Types, Lazy Evaluation, Haskell, GHC Core}

\begin{document}

\begin{abstract}
Linear type systems can be used to guarantee (linear) resources are used exactly once.
Traditionally, using a resource is synonymous with its \emph{syntactic}
occurrence in the program, however, under the lens of non-strict evaluation,
linearity can be further understood \emph{semantically}, where a
syntactic occurrence of a resource does not necessarily entail
using that resource when the program is executed.
Semantic linearity becomes especially vital when considering the
internal language of an optimising compiler, where optimisations
heavily rewrite the source program in ways that can often break
syntactic linearity but that aim to preserve the program's semantics.
% %
% Semantic linearity is especially necessary in optimising compilers for
% languages combining linearity and laziness: optimisations leverage laziness to
% heavily rewrite the source program, pushing the interaction of linearity and
% laziness to its limit, regardless of the original program typing linearity
% conservatively.
%

We present Linear Core, the first type system that understands semantic
linearity in the presence of non-strict evaluation, suitable for the Core
intermediate language of the Glasgow Haskell Compiler. We prove Linear Core
is sound and guarantees linear resource usage, and prove multiple optimising
transformations preserve linearity in Linear Core while failing to do so in
Core. We have implemented Linear Core as a compiler plugin to validate the
system against linearity-heavy libraries, including
\texttt{linear-base}.%^, in the heart of the compiler.
\end{abstract}

\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% {{{ Introduction
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}\label{sec:intro}

% Statically safe programming languages provide compile time correctness
% guarantees by having the compiler rule out certain classes of errors
% or invalid programs. Moreover, static typing
% allows programmers to state and enforce (compile-time) invariants
% relevant to their problem domain.
% In this sense, type safety entails that all
% possible executions of a type-correct program cannot exhibit behaviors
% deemed ``wrong'' by the type system design. This idea is captured in
% the motto ``well-typed programs do not go wrong''.%~\cite{}.

Linear types~\cite{cite:linear-logic,cite:barberdill} increase the expressiveness
of type systems by allowing them to enforce desirable usage patterns
for certain resources (e.g.~a
file handle). Typically, linear types can be used to enforce resources are used \emph{exactly once}.
%
In programming languages with a linear type system, not using certain resources
or using them twice is flagged as a type error. For instance, linear types can
statically enforce correct usage of file or socket descriptors, or
that heap-allocated memory is freed exactly once (i.e.,~leaks and double-frees become
type errors), or even guarantee deadlock freedom in channel-based communication
protocols~\cite{10.1007/978-3-642-15375-4_16},
% implement safe
% high-performance language interoperability~\cite{}, 
%guarantee that quantum entangled states are not duplicated~\cite{}
among other correctness properties~\cite{10.1145/3373718.3394765,10.1145/3527313,cite:linearhaskell}.
% handle mutable state safely~\cite{}
%
The following program with a runtime error (allocated memory is freed twice) is
accepted by a C-like type system. Conversely, under the lens of a linear type
system where $p$ is deemed a linear resource created by the call to
\texttt{malloc}, the program is rejected:
% We know this to be the dreaded double-free error which will crash the program at runtime.
% \centering
\begin{code}
let p = malloc(4) in free(p); free(p);
\end{code}

Despite their promise and extensive presence in research
literature~\cite{Wadler1990LinearTC,CERVESATO2000133,10.1093/logcom/2.3.297},
an effective design combining linear and non-linear typing is both
challenging and necessary to bring the advantages of linear typing to
mainstream languages.
%
Consequently, few general purpose programming languages have linear or substructural
type systems. Among them are Idris~2~\cite{brady:LIPIcs.ECOOP.2021.9},
% a linearly and dependently typed language based on Quantitative Type Theory,
Rust~\cite{10.1145/2692956.2663188}, a language whose
ownership types build on linear types to guarantee memory safety
without garbage collection, and, more recently,
Haskell~\cite{cite:linearhaskell}, a pure, functional, and
\emph{lazy} general purpose programming language.
%
% Besides Haskell's supporting linear
% types according to the said paper, Idris 2\cite{} supports linear types in a
% dependently typed setting, Clean\cite{} has uniqueness types which are closely
% related to linear types, and Rust\cite{} has ownership types which build from
% linear types. 
%

Linearity in Haskell stands out from linearity in other
languages because linear types permeate Haskell down to (its) Core, the
intermediate language into which Haskell is translated by the Glasgow Haskell Compiler (GHC).
%
Core is a minimal, explicitly typed, lazy functional language with
both linear and unrestricted types, to which multiple
Core-to-Core optimising transformations are applied during
compilation. Notably Core does not have a (higher-order or otherwise)
store.
%
Typechecking Core serves as a sanity check to the correction of such
transformations, since unsound ones are likely to introduce ill-typed
expressions. %, making them easy to spot.
%
% If a Core program fails to typecheck, the
% optimising transformations are very likely to have introduced an error in the
% resulting program.
%

% Both Haskell and its intermediate language Core are lazily
%         evaluated.
%         % -- expressions in Haskell are only evaluated when needed.
%         As we make clear in this paper, non-strict evaluation interacts non-trivially with linearity.
%         We give a precise definition of what it 
%         Intuitively, a syntactic occurrence of a linear resource does not
%         necessarily entail consuming that resource because expressions are not
%         necessarily evaluated (if the expression is not evaluated, the
%         resource is not used).
%         %
%         In contrast, under eager evaluation, a syntactic occurrence of a resource
%         along a given control-flow path always implies its use at runtime.
        %
        % Under eager/call-by-value evaluation, the distinction between syntactic uses of a
        % resource and the actual use of linear resources at runtime does not exist --
        % an occurrence of a variable in the program always entails consuming it.
        %

        % Include?
        % This interaction is unique to Haskell since, to the best of
        % our knowledge, it is the
        % only language featuring both laziness and linearity.

Just as Core's type system \cite{10.1145/1190315.1190324} provides a degree of validation to the
translation from Haskell (dubbed \emph{desugaring}) and the subsequent
optimising transformations, a linearly typed Core would guarantee that
linear resource usage in the source language is not violated by desugaring
and the compiler optimisation passes. Moreover, linearity information in
Core can inform Core-to-Core optimising 
transformations~\cite{cite:let-floating,peytonjones1997a,cite:linearhaskell},
such as inlining and $\beta$-reduction, to produce more efficient programs.
%
% It is crucial that a program's behaviour enforced by linear types is \emph{not}
% changed by the compiler in the desugaring or optimisation stages (optimisations
% should not destroy linearity!) and a linearity aware Core type-checker is key in
% providing such guarantees -- it would be disastrous if the compiler, e.g.,
% duplicated a pointer to heap-allocated memory that was previously just used
% once and then freed in the original program.
%TODO: linearidade pode informar otimizacoes
%
% Even more, linearity in Core can inform Core-to-Core optimising
% transformations~\cite{cite:let-floating,peytonjones1997a,cite:linearhaskell},
% including inlining and $\beta$-reduction, to produce more performant programs.

% Linear core actually not so good
% Additionally, while not yet a reality, linearity in Core could be used to inform
% certain program optimisations, i.e. having linear types in Core could be used to
% further optimise certain programs and, therefore, benefit the runtime
% performance characteristics of our programs. For example, Linear Haskell\cite{}
% describes as future work an improvement to the inlining optimisation: Inlining
% is a centerpiece program optimisation primarily because of the subsequent
% optimising opportunities unlocked by inlining. However, it relies on a heuristic
% process known as \emph{cardinality analysis} to discover safe inlining
% opportunities. Unfortunately, heuristics can be volatile and fail in identifying
% crucial inlining opportunities resulting in programs that fall short of their
% performance potential. On the contrary, the linearity information could be
% integrated in the analysis and used as a (non-heuristic) cardinality
% \emph{declaration}.

% While the current version of Core is linearity-aware (i.e.,~Core has so-called
% multiplicity annotations in variable binders), its type system does not (fully)
% validate the linearity constraints in the desugared program and essentially
% fails to type-check programs resulting from several optimising transformations
% that are necessary to produce efficient object code. The reason for this latter
% point is not evidently clear:
% %
% if we can typecheck linearity in the surface level Haskell why do we fail to do
% so in Core?
%

Despite the advantages of adding linear types to Core, the status quo is
that linearity is effectively ignored in Core, being only checked
in the source Haskell code. The reason is that the desugaring
and subsequent Core-to-Core optimising transformations
eliminate and rearrange most of the syntactic constructs through
which linearity checking is performed -- often resulting in programs
that are syntactically very different from the original, especially
due to the flexibility provided by laziness wrt the
optimisations a compiler may perform.

While compiler optimisations aim to preserve the semantics of
programs and therefore a program's resource consumption,
a \emph{naive} syntactic analysis of the optimised programs (as
performed by the current Core type checker) often fails to recognize
the program as linear.
%
As a trivial example, let $x$ be a linear resource in the next two expressions,
where the latter results from inlining $y$ without eliminating the
binder (a common optimising transformation in GHC). Even
though the second expression no longer appears linear (as there are now two occurrences
of $x$), it is indeed linear because the let-bound expression is never
evaluated under \emph{lazy evaluation} and $x$ is still consumed exactly once (when it is freed in the let body):
% executed, $x$ will be freed exactly once:
%
% , \emph{is} linear, as the let bound expression freeing $x$
% is never evaluated (because is not needed) -- thus $x$ is only consumed once
% when freed in the let body:
\[
\begin{array}{ccc}
\llet{y = \textsf{free}~x}{y} & \Longrightarrow_{Inlining} & \llet{y = \textsf{free}~x}{\textsf{free}~x}
\end{array}
\]

% \begin{boxedminipage}
% \begin{code}
% let y = free x
% in free x
% \end{code}
% \end{boxedminipage}

The Core optimising transformations expose a fundamental limitation of Core's
type system: it does not account for the non-strict evaluation model
of Core and, thus, a whole class of programs that are linear under the lens of
non-strict evaluation are rejected by Core's current linear type system.
%
In this work, we address this limitation by exploring how syntactic linearity
breaks down in the presence of non-strictness and design a linear type
system for Core which accepts the vast majority of programs and optimising
transformations that, while correct under non-strict evaluation, are rejected
by traditional linear typing disciplines.
%
%
Additionally, we show our system
%
% and its implications on validating and influencing optimising transformations
% by designing a type system which analyses how syntactic linearity breaks down
% in the presence of laziness
is suitable for the intermediate language of an
optimising compiler by implementing it as a Glasgow Haskell Compiler plugin.
%
Our contributions are as follows:
%
\begin{itemize}

\item We introduce the concept of \emph{semantic} linearity, in
contrast to \emph{syntactic} linearity in Haskell, by example
(Section~\ref{sec:linearity-semantically}).
%
A precise definition of \emph{semantic} linearity is given by our instrumented
operational semantics, under which a program using a linear resource
non-linearly gets stuck (Section~\ref{sec:main:metatheory}).

\item We introduce Linear Core, a non-strict linear language with
the key features of Core (except for type equality coercions), whose type system
accounts for semantic linearity in the presence of laziness. To the
best of our knowledge, this is the first type system to capture linearity
semantically in this context (Section~\ref{sec:main:linear-core}).

\item We prove our type system is sound, guarantees linear resource usage, and
    prove that multiple optimising transformations, which are rejected in Core,
        are validated by Linear Core (Section~\ref{sec:main:metatheory}).

\item We discuss our implementation of Linear Core as a GHC
  plugin\footnote{We plan to submit our implementation as a companion
    artifact to this paper.} which checks linearity in
all intermediate Core programs produced during the compilation process, showing
it accepts the programs resulting from transformations in libraries such as
\texttt{linear-base} (Section~\ref{sec:discuss:implementation}).

\end{itemize}

\noindent Additional proofs and definitions are given in
Appendices~\ref{app:proofs},~\ref{app:optimisations} and~\ref{app:opsem}.

%
% We review background concepts fundamental to our work in
% Section~\ref{sec:background}, including linear type systems, linear types in
% Haskell, evaluation strategies, Core's type system, and multiple optimising
% transformations applied by GHC in its compilation process.
%
% We compare our contributions to related work and discuss possible
% avenues for further research
% (highlighting so-called
% \emph{multiplicity coercions})
% in Section~\ref{sec:discussion}, which concludes the document.

% In fact, we are not aware of any linear type system which
% understands linearity in the presence of laziness.

% \todo[inline]{In reality, the Core optimising transformations only expose a
% more fundamental issue in the existing linear types in Haskell -- its mismatch
% with the evaluation model. In call-by-need languages a mention of a variable
% might not entail using that variable - it depends on it being required or not.
% We're the first to do so as far as we know}

% Concluding, by extending Core / System $F_C$ with linearity and multiplicity
% annotations such that we can desugar all of Linear Haskell and validate it
% accross transformations taking into consideration Core's call-by-need
% semantics, we can validate the surface level linear type's implementation, we
% can guarantee optimising transformations do not destroy linearity, and we might
% be able to inform optimising transformations with linearity.

% Consider the following recursive let
% definition, where $x$ is a linear variable that must be used exactly once, would
% not typecheck in a source Haskell program since the typechecker cannot tell that
% $x$ is used linearly, but this program might occur naturally in Core from
% transformations on a program that did succesfully typecheck:
% \begin{code}
% letrec f = \case
%         True -> f False
%         False -> x
% in f True
% \end{code}

% Alternatively, one might imagine Haskell being desugared into an intermediate
% representation to which multiple different optimising transformations are
% applied but on which no integrity checks are done

% Despite \emph{want} a linearly-typed core
% because a linearly-typed Core ensures that desugaring
% Haskell and optimising transformations don't destroy linearity.

% In theory, because the Core language must also know about linearity, we should

% A remedy is to use the multiplicity annotations of λq → as cardinality declarations. Formalising
% and implementing the integration of multiplicities in the cardinality analysis is left as future work.

% In theory, we should typecheck \emph{linearity} in \textbf{Core} just the same
% as the typechecking verification we had with the existing type annotations prior
% to the addition of linear types, that is, our Core program with linearity
% annotations should be typechecked after the optimising transformations...

% }}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% {{{ Linearity, Semantically
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Linearity, Semantically\label{sec:linearity-semantically}}

A linear type system typically guarantees that linear resources are
consumed exactly once. Consequently, whether a program is
well-typed under this lens intrinsically depends on the
definition of consuming a resource.
In general, consuming a resource is equated with its 
\emph{syntactic occurrence} along a control-flow path in the program. However, as this section
makes clear, in linear type systems for non-strict
languages, a distinction between using
resources \emph{syntactically} and \emph{semantically} is needed.
%
For instance, in the following Haskell-like program defining a linear
function \lstinline{f} (i.e., a function that should use its argument
exactly once), there are two syntactic
occurrences of \lstinline{handle}: one is in a let-bound computation that
closes this handle, the other is returned. If the program is evaluated lazily,
the handle will not be closed because the computation that closes the handle is
never evaluated (the non-linear \lstinline{x} goes unused).
%
\begin{lstlisting}
f : Handle *' $\lolli$ '* Handle
f handle = let x = close handle in handle
\end{lstlisting}
%
% Intuitively, a computation that depends on a linear resource to produce a
% result consumes that resource iff the result is effectively computed; in
% contrast, a computation that depends on a linear resource, but is never run,
% will not consume that resource.
%
% Even though there are two occurrences of \lstinline{handle} in the program, and
% it looks as though the \lstinline{handle} may be closed before returned, this function may
% actually only \emph{consume} used once -- it depends on the evaluation strategy of the
% language -- the \lstinline{handle} is closed iff the let bound computation is
% evaluated.
%
%
% Intuitively, \emph{consuming} a resource is not necessarily synonymous with its
% syntactic occurrence. In the example, \emph{non-strict} evaluation means the
% handle is never closed by the function.
%
%
% Depending on the evaluation strategy of the language, a computation that closes
% a handle might not be evaluated, and thus is not consumed.
% With \emph{eager} evaluation semantics,
% the let
%   bound expression \lstinline{close handle} is eagerly evaluated, and the \lstinline{handle} will be closed before being
% returned. It is clear that a linear type system should not accept such a
% program since the linear resource \lstinline{handle} is duplicated -- it is used in a
% computation that closes it, while still being made available to the caller of
% the function;
% %
% on the other hand, with \emph{call-by-need},
%
% Reconsider the \lstinline{handle} example program under
% \emph{call-by-need} evaluation semantics:
% The let bound expression is only evaluated if the binding \lstinline{x} is
% needed. The function returns the \lstinline{handle} right away, and forgets the
% let bound \lstinline{x} as its scope is exited, such that \lstinline{x} will
% never be needed -- consequently, the \lstinline{handle} is never closed by the
% function...
%
%
%

%Under the lens of non-strict evaluation, a syntactic
%occurrence of a resource in an expression only entails consuming that resource
%iff the expression itself is fully consumed.
%
Intuitively, a computation that depends on a linear resource to produce a
result consumes that resource iff the result is fully consumed; in
contrast, a computation that depends on a linear resource, but is never run,
will not consume that resource.
%
We argue that a linear type system for a language with non-strict evaluation
semantics should accept the above program, unlike a linear type system for the
same program if it were to be evaluated eagerly, where the let-bound
expression would always be evaluated.
%
Given this observation, %and exploring the connection between computation and evaluation,
% it becomes clear that \emph{linearity} and \emph{consuming resources}, in the
% above example and for programs in general, should be defined in function of the
% language's evaluation strategy.
%
we turn our focus to linearity under lazy evaluation since the distinction between semantically and syntactically
consuming a resource is only exposed under non-strict semantics (and since GHC Core is
lazy).
%.

% Indeed, under \emph{call-by-value}, syntactic occurrences of a linear resource
% directly correspond to semantically using that resource\footnote{With the minor exception
% of trivial aliases, which don't entail any computation even in
% \emph{call-by-value}. In theory, we could use in mutual exclusion any of the
% aliases to refer to a resource without loss of linearity} because \emph{all}
% expressions are eagerly evaluated -- if all computations are eagerly run, all
% linear resources required by computations are \emph{eagerly consumed}.

% ROMES:IMPORTANT:TODO:
% \subsection{Reductions / Function applications}
% 
% \todo[inline]{unrestricted call-by-name with resources can duplicate the resources, as if it were unsound?}
% We reduce function applications in two distinct ways, call by name (for linear
% functions) and call by need (we've now introduced linear lets, so we can look
% at this now)
% 
% Foreshadow to issues with opt (reverse binder swap)?
% 
% Or maybe just drop this section altogether

\subsection{Semantic Linearity by Example\label{sec:semantic-linearity-examples}}

This section aims to develop an intuition for semantic linearity through simple
Core programs that can be understood as linear but are rejected by Core's
linear type system (or, in fact, by all linear type systems we are
aware of prior to this work).
%
% Needed?
%Aligned with our motivation of typing linearity in GHC Core such
%that optimising transformations preserve linearity, and with the goal
%of understanding linearity in a non-strict context,
%
% BETTER?:
%Aligned with our motivation of typechecking linearity in GHC Core such
%that optimising transformations preserve linearity, and in order to explain
%linearity in a non-strict context, we illustrate semantic linearity
%through examples of Core programs that are semantically linear but rejected by
%Core's linear type system -- building intuition towards semantic linearity in a
%lazy language.
%
In the examples, a \colorbox{working}{\workingcolorname} background highlights
programs that are syntactically linear and accepted by Core's linear type
system.
%
A \colorbox{notyet}{\notyetcolorname} or
\colorbox{limitation}{\limitationcolorname} background mark programs that are
semantically linear, but not seen as linear by Core's type system.
%
Notably, the linear type system we develop in this work accepts all
\colorbox{notyet}{\notyetcolorname} programs.
%
% (while the few \colorbox{limitation}{\limitationcolorname} programs are not accepted).
A \colorbox{noway}{\nowaycolorname} background indicates that the program
violates linearity semantically, i.e.~the program effectively
discards or duplicates linear resources.

% We discuss recursive and recursive let bindings, and case expressions.

%\todo{Should we have a subsection starting with simple function application?}

\paragraph{Let bindings}
% We start our discussion with non-strict (non-recursive) let bindings, i.e. let bindings whose
% body is evaluated only when the binding is needed, rather than when declared.
%
% In Core, let bindings entail creating \emph{thunks} that suspends the
% evaluation of the let body.
% % (for background, see
% % Section~\ref{sec:background:evaluation-strategies})
% When the binding itself is evaluated, the \emph{thunk} is
% \emph{forced} and the evaluation is carried out. The result overwrites the
% \emph{thunk} -- the let binding now points to the result of the evaluation.
%
% A \emph{thunk} is \emph{forced} (and the suspended computation is evaluated) when
% the binding itself is evaluated.
%
%\begin{notyet}
% \begin{lstlisting}
% f1 :: (a *' $\lolli$ '* b) -> a *' $\lolli$ '* b
% f1 use x = let y = use x in y
% \end{lstlisting}
% \end{notyet}
% In a linear type system, a non-strict let binding that depends on a linear
% resource \lstinline{x} doesn't consume the resource as long as the binding isn't
% evaluated -- the suspended computation only uses the resource if it is run. For
% this reason, we can't naively tell whether \lstinline{x} is consumed just by looking at
% the let binding body. In the following example, we assign a computation that
% depends on the resource \lstinline{x} to a binder, which is then
% returned.
%
% \begin{figure}
%   \begin{minipage}{.33\textwidth}
%     \begin{notyet}
% \begin{lstlisting}
% f1 :: (a *' $\lolli$ '* b) -> a *' $\lolli$ '* b
% f1 use x = let y = use x in y
% \end{lstlisting}
% \end{notyet}
% \caption{one}
% \end{minipage}%
% \begin{minipage}{.33\textwidth}
%       \begin{notyet}
% \begin{lstlisting}
% f2 :: (a *' $\lolli$ '* a) -> a *' $\lolli$ '* a
% f2 use x = let y = use x in use x
% \end{lstlisting}
% \end{notyet}
% \caption{two}
% \end{minipage}%
% \begin{minipage}{.33\textwidth}
% \begin{notyet}
% \begin{lstlisting}
% f3 :: (a *' $\lolli$ '* a) -> Bool -> a *' $\lolli$ '* a
% f3 use bool x =   let y = use x in
%    case bool of
%       True -> x
%       False -> y
% \end{lstlisting}
% \end{notyet}
% \caption{Three}
% \end{minipage}%
% \end{figure}
%
%
%
%
% %
% The linear resource \lstinline{x} is used exactly once, since it is used exactly once in
% the body of the binding and the binding is used exactly once in the let body.
% %
% According to Linear Haskell's core calculus~$\lambda_{\to}^{q}$~\cite{cite:linearhaskell}, let
% bound variables are annotated with a multiplicity which is multiplied (as per
% the multiplicity semiring) by the multiplicity of all variables that are free
% in the binder's body.
% %
% In short, if a let binder is linear (has
% multiplicity $1$) then the linear variables free in its body are only used
% once; if the let binder is unrestricted (has multiplicity $\omega$) then the
% resources in its body are consumed many times, meaning no linear variables can
% occur in that let binder's body.
% %
% Unfortunately, GHC's implementation of Linear
% Haskell doesn't seem to infer multiplicities for lets yet, so while the above
% program should typecheck in Linear Haskell, it is rejected by GHC.

% The next example exposes the case in which the let binder is ignored in the let
% body. Here, the linear resource \incode{x} is used in \incode{y}'s body and in the let
% body, however, the resource is still used semantically linearly
% because \incode{y} isn't used at all, thus \incode{x} is consumed just
% once in the let body: 
% %
% \begin{notyet}
% \begin{lstlisting}
% f2 :: (a *' $\lolli$ '* a) -> a *' $\lolli$ '* a
% f2 use x = let y = use x in use x
% \end{lstlisting}
% \end{notyet}
% %
% Programmers don't often write bindings that are completely unused, yet, an
% optimising compiler will produce intermediate programs with unused bindings\footnote{Unused bindings are then
% also dropped by the optimising compiler} 
% from transformations such as inlining, which can substitute out occurrences
% of the binder (e.g. \incode{y} is inlined in the let body).

The example at the start of Section~\ref{sec:linearity-semantically}
% (similar to towards the end of Section~\ref{sec:intro}),
highlighted the subtlety of linearity in a non-strict language with
a single unused let binding.
%
%Less trivially,
In general, a let-bound expression that captures a linear variable
can be conditionally needed at runtime, depending on the branch taken.
Both optimising transformations (float-out) and programmers used to
non-strict evaluation are keen to write programs with bindings that are selectively
used in case alternatives, such as (we write $\lolli$ for linear
functions and \lstinline{->} for non-linear functions):
%
\begin{notyet}
\begin{lstlisting}
f :: (a *' $\lolli$ '* a) -> Bool -> a *' $\lolli$ '* a
f use bool x = let y = use x in
  case bool of
    True -> x 
    False -> y
\end{lstlisting}
\end{notyet}
%
This example returns \lstinline{x} in one branch and the let-bound
\lstinline{y} in the other. Semantically, as opposed to syntactically, this
program is linear % (i.e. linear resources are used exactly once)
because the linear resource \lstinline{x} is used exactly once regardless of which branch is
taken: either directly in the \lstinline{True} branch, or \emph{via}
the use of the let-bound \lstinline{y} in the \lstinline{False} branch.

That said, a let-bound \lstinline{y} that captures a linear resource
\lstinline{x} must still be used exactly once, regardless of whether evaluation
inlines the let (\emph{à la} call-by-name) or creates a thunk (\emph{à
  la} call-by-need):
%
% \lstinline{y} must still be used exactly once.  When \lstinline{y} is inlined
% in the body, this is trivial to see since multiple occurrences of will be 
%
\begin{noway}
\begin{lstlisting}
f1 :: (a *' $\lolli$ '* b) -> a *' $\lolli$ '* (b, b)
f1 use x = let y = use x in (y, y)
\end{lstlisting}
\end{noway}
%
If \lstinline{y} is inlined, linearity is trivially violated.
%
On the other hand, if \lstinline{y} becomes a thunk, \lstinline{use x} will
only be evaluated once regardless of how many times \lstinline{y} is used.
Even so, this program must be rejected under call-by-need.
%
Intuitively, linearity is violated if the computation leaks the resource or
parts of it. For instance, if \lstinline{use = id}, the program must clearly be
rejected.
%
% TODO:NEW: Include?
% If the computation fully consumed the linear resource to produce an
% unrestricted result which was forced, the binding could be shared...
% OLD Indeed, if the result of the computation involving the linear resource
% was instead, e.g., an unrestricted integer, then sharing the result would not
% involve consuming the resource more than once.
%
%
% Redundant?
% In short, the result of evaluating a let-binder's body which depends on linear resources,
% if computed, must be consumed exactly once, or, otherwise, we risk discarding
% or duplicating (the parts of) the resources that were captured in the result.

Lastly, consider a (ill-typed) program which defines two let bindings \incode{z} and \incode{y}, where
\incode{z} uses \incode{y} which in turn uses the linear resource \incode{x}:
%
\begin{noway}
\begin{lstlisting}
f2 :: (a *' $\lolli$ '* a) -> a *' $\lolli$ '* ()
f2 use x = let y = use x in let z = use y in ()
\end{lstlisting}
\end{noway}
%
Even though the binding \incode{y} is used in \incode{z}, \incode{x} is never
consumed because \incode{z} isn't evaluated in the let body, and, thus,
\incode{y} isn't evaluated either. We use this example to highlight that even
for let-bound variables, the syntactic occurrence of a variable isn't enough to
determine whether it is used. Instead, we ought to think of uses of \incode{y}
as implying using \incode{x}, and therefore uses of \incode{z} imply using
\incode{x}, however, if neither are used, then \incode{x} isn't used. Since
\incode{x} is effectively discarded, this example also violates linearity.

% \parawith{Summary}
% The examples so far build an intuition for semantic linearity in the presence
% of lazy let bindings. 
In essence, an unused let binding doesn't consume any resources, and a let
binding used exactly once consumes the resources it captures exactly once. Let
binders that depend on linear resources must be used \emph{at most once} in
mutual exclusion with the linear resources themselves (in a sense,
let-bound variables are affine in the let body).
%
%Moreover, if the let binding \incode{y} isn't used in the let body, then the
%resources it depends on ($\ov{x}$) must still be used -- the binding $y$ is
%mutually exclusive with the resources $\ov{x}$ (for the resources to be used
%linearly, either the binder occurs exactly once $y$, or the resources $\ov{x}$
%do).
%
We discuss how to encode this principle between let bindings and their dependencies using so called \emph{usage
environments}, in Section~\ref{sec:usage-environments}.

%\vspace{-.5cm}
\paragraph{Recursive let bindings\label{sec:semantic-linearity-examples:recursive-lets}}

%Recursive lets behave similarly to non-recursive let bindings in the sense we
%must use them \emph{at most once} because, when evaluated, the linear resources
%used in the binders bodies are consumed.
%

Recursive let-bindings
%, i.e. bindings with definitions that recursively depend on the bindings they define,
behave similarly to non-recursive lets as far as their usage in the let
continuation is concerned.
%recursive lets must still be used at most once, in mutual exclusion with the
%linear resources they capture.
%
The challenge lies in understanding linearity in the definitions of the
recursive binders and determining which resources are consumed when one of the
binders in the recursive group is used.
%
% The same way that, in a let body, evaluating
% a binding that uses some resource exactly once consumes that resource once,
% using the binding in its own definition also entails using that resource once.
Consider the following program, which calls a recursive let-bound function
defined in terms of the linear resource \incode{x} and itself:
%
\begin{notyet}
\begin{lstlisting}
f3 :: Bool -> a *' $\lolli$ '* a
f3 bool x = let go b = case b of
                     True -> x
                     False -> go (not b)
                   in go bool
\end{lstlisting}
\end{notyet}
%
Function \incode{f3} is semantically linear because, iff it is consumed exactly once,
then \incode{x} is consumed exactly once. We can see this by case analysis on \incode{go}'s argument:
%\begin{itemize}%
%\item
when \incode{bool} is \incode{True}, we'll use the resource \incode{x};
%\item
 when \incode{bool} is \incode{False}, we recurse by calling \incode{go} on \incode{True}, which in turn will use the resource \incode{x}.
%\end{itemize}
In \incode{go}'s body, \incode{x} is used directly in one branch and indirectly in the
other (by recursing, which we know will result in using \incode{x} linearly).
% ROMES:TODO:!!
%\todo{this bit is quite hard to explain. It is some sort of cyclic
%argument -- we kind of assume go uses x linearly s.t. when go itself is used
%then we're using x linearly. recursion...}
%

It so happens that \incode{go} will terminate on any input, and will always consume
\incode{x}. However, termination is not a requirement for a binding to use a resource linearly,
and we could have a similar example in which \incode{go} might never terminate but still
uses \incode{x} linearly if evaluated:

\begin{notyet}
\begin{lstlisting}
f4 :: Bool -> a *' $\lolli$ '*  a
f4 bool x = let go b = case b of
                     True -> x
                     False -> go b
                   in go bool
          \end{lstlisting}
        \end{notyet}

The key to linearity in the presence of non-termination is Linear Haskell's
definition of a linear function: \emph{if a linear function application (\incode{f u}) is
consumed exactly once, then the argument (\incode{u}) is consumed exactly once}.
If \incode{f u} doesn't terminate, it is never consumed, thus the claim holds
vacuously.
% that's why \incode{f8} typechecks:
% \begin{working}
% \begin{lstlisting}
% f8 :: a *' $\lolli$ '* b
% f8 x = f8 x
% \end{lstlisting}
% \end{working}
%
If \incode{go} doesn't terminate, we aren't able to compute (or consume) the result
of \incode{f4}, so we do not promise anything about \incode{x} being consumed (\incode{f4}'s
linearity holds trivially). If it did terminate, it would consume \incode{x} exactly
once (e.g. if \incode{go} was applied to \incode{True}).

Naturally, a linear type system must statically reject any recursive group that could
lead to non-linear resource usage.
%
% Determining the linear resources used in a recursive binding might feel
% peculiar since we need to know the linear resources used by the binder to
% determine the linear resources it uses.
% %
% The paradoxical definition is difficult to grasp, just how learning that a
% function can be defined in terms of itself is perplexing when one is first
% introduced to general recursion.
% %
% Informally, we \emph{assume} the recursive binding will consume some linear resources
% exactly once, and use that assumption when reasoning about recursive calls such
% that those linear resources are used exactly once.
% Generalizing, we need to find a set of linear resources ($\Delta$) that satisfies the recursive equation
% \footnote{This set of resources will basically be the least upper bound of the sets of resources used in each
% mutually recursive binding scaled by the times each binding was used}
% arising from given binding $x$, such that:
% %\begin{enumerate}
% (1) Occurrences of $x$ in its own body are synonymous with using all resources in $\Delta$ exactly once;
% (2) and if the binding $x$ is fully evaluated, then all resources in $\Delta$ are consumed exactly once.
% %\end{enumerate}
% Finding a solution to this equation is akin to finding a (principle) type for a
% recursive binding: the binding needs to be given a type such that occurrences of
% that binding in its own body typecheck using that type.
% %
% Foreshadowing, the core system we developed assumes recursive let bindings to
% be annotated with a set of resources satisfying the equations; but we also
% present an algorithm to determine this solution, and distinguish between an
% \emph{inference} and a \emph{checking} phase, where we first determine the
% linear resources used by a group of recursive bindings and only then check
% whether the binding is linear, in our implementation of checking of recursive
% lets.
% ROMES:TODO:!!!
%\todo{We might not need the inferrence phase, somehow. Anyhow it seems like
%our inferrence is not really about determining a solution but more about
%determining how many times each thing gets used}
%
% There might not be a solution to the set of equations. In this case, the
% binding undoubtedly entails using a linear resource more than once.
%
For instance, if the linear resource \incode{x} is used in one case alternative and
recursion is used more than once in another, \incode{x} may end up being consumed
more than once:
% (thus, \incode{f5} below is ill-typed):
%
\begin{noway}
\begin{lstlisting}
f5 :: Bool -> Bool *' $\lolli$ '* Bool
f5 bool x =
  let go b
        = case b of
           True -> x
           False -> go (not b) && go True
  in go bool
\end{lstlisting}
\end{noway}
%However, if we returned \incode{x} instead of \incode{go bool} in the let body, then, despite
%the binding using \incode{x} more than once, we would consume \incode{x} exactly once,
%since recursive bindings are still lazy.
%

%Intuitively, to statically determine whether a group of recursive 
%
%Informally, we \emph{assume} the recursive binding will consume some linear resources
%exactly once, and use that assumption when reasoning about recursive calls such
%that those linear resources are used exactly once.

To determine whether the usage of a binding from a recursive group is linear,
we must infer the superset of resources that may end up being consumed in a
single use of any of the bindings. For a strongly connected recursive group,
this superset is the same for all bindings by definition.
%
% Intuitively, this inference amounts to finding the linear resources
% ($\Delta$) which satisfy set of equations derived from the bindings
% $\ov{x_i}$ in the group such that
% %\begin{enumerate}
% occurrences of $x_i$ in its own body are synonymous with using all resources in $\Delta$ exactly once.
%(2) and if the binding $x$ is fully evaluated, then all resources in $\Delta$ are consumed exactly once.
%\end{enumerate}
%Finding a solution to this equation is akin to finding a (principle) type for a
%recursive binding such that occurrences of itself in its definition are also well-typed.
%
It follows from the inferred resources used being the same for all bindings
that at most a single variable from the recursive group can be used in the
continuation.
%
The treatment of recursive let bindings in our system is discussed in Section~\ref{sec:recursivelets}.

% Lastly, we extend our single-binding running example to use two mutually recursive bindings that
% depend a linear resource:
% \begin{notyet}
% \begin{lstlisting}
% f10 :: Bool -> a *' $\lolli$ '* a
% f10 bool x =
%   let go1 b = case b of
%            True -> go2 b
%            False -> go1 (not b)
%        go2 b = case b of
%            True -> x
%            False -> go1 b
%   in go1 bool
% \end{lstlisting}
% \end{notyet}
% As before, we must find a solution to the set of equations defined by the
% mutually recursive bindings to determine which resources will be consumed.
% In this case, \incode{go1} and \incode{go2} both consume \incode{x} exactly once if evaluated.
% We additionally note that a strongly connected group of recursive bindings
% (i.e. all bindings are transitively reachable from any of them) will always
% consume the same set of resources -- if all bindings are potentially reachable,
% then all linear resources are too.

% \parawith{Summary}
% Summarising, recursive let bindings behave like non-recursive let bindings in
% that if they aren't consumed, the resources they depend on aren't consumed
% either.  However, recursive let bindings are defined in terms of themselves, so
% the set of linear resources that will be consumed when the binder is evaluated
% is also defined in terms of itself (we need it to determine what resources are
% used when we recurse). We can intuitively think of this set of linear resources
% that will be consumed as a solution to a set of equations defined by a group of
% mutually recursive bindings, which we are able to reason about without an
% algorithm for simpler programs. In our work, the core type system isn't
% concerned with deriving said solution, but we present a simple algorithm for
% inferring it with our implementation.

\paragraph{Case expressions}
We have observed that more programs can be accepted as linear by delaying
resource consumption in (recursive) let bindings until they are evaluated.
On the other hand, \emph{case expressions drive evaluation} -- and, therefore,
do consume resources.

%and semantic linearity can only be understood in terms of how expressions are
%evaluated.

%Up until now, there have been no examples in which linear resources are
%\emph{fully consumed} in the bodies of linear functions.

%%%%%%%%%%%%%%% KEEEP ????? %%%%%%%
%%%%%%%%%All example functions so far returned a value that has to be itself consumed
%%%%%%%%%exactly once to ensure the linear argument is, in turn, consumed exactly once
%%%%%%%%%-- as opposed to functions whose application simply needs to be evaluated to
%%%%%%%%%guarantee its linear argument is consumed (e.g.,~functions that return an
%%%%%%%%%unrestricted value).

%
%The latter are of particular relevance because linearly-typed abstractions
%usually require such a function to provide newly-created linear resources to.
%
% For example, the entry point to the linear array API presented in Linear
% Haskell takes such a function as its second argument:
% \begin{working}
% \begin{lstlisting}
% newMArray :: Int -> (MArray a *' $\lolli$ '* Unrestricted b) *' $\lolli$ '* b
% \end{lstlisting}
% \end{working}
%(The second argument is a function that consumes |MArray a| linearly
%and returns an unrestricted result -- we don't need to consume said result
%exactly once to guarantee that |MArray a| is used linearly in the function
%body.)
%
% ROMES:TODO: Há qualquer coisa que não gosto nada! na coerência deste paragrafo
%
%Case expressions enable us to consume resources and so write
%functions that fully consume their linear arguments.
% CORTEI
To understand exactly how,
we turn to the definition of \emph{consuming} a resource from Linear
Haskell~\cite{cite:linearhaskell}:
%begin{itemize}
% \item
to consume a value of atomic base type (such as~\texttt{Int} or
        \texttt{Ptr}) exactly once, just evaluate it;
        % \item
        to consume a function value exactly once, apply it to one argument,
        and consume its result exactly once;
        % \item
        to consume a value of an algebraic datatype exactly once,
        pattern-match on it, and consume all its linear components exactly once.
        % For example, a linear pair (equivalent to $\tensor$) is consumed exactly
        % once if pattern-matched on \emph{and} both the first and second element are
        % consumed once.
%\end{itemize}
Thus, we can consume a linear resource by fully evaluating it -- and evaluation
happens in case expressions.

%%% In Section~\ref{sec:generalizing-evaluation-consuming} we generalize the idea
%%% that consuming a resource is deeply tied to evaluation.  Here, we continue
%%% building intuition for semantic linearity, first reviewing how case expressions
%%% evaluate expressions, and then exploring how they consume resources, by way of example.

In Core, cases are of the form $\ccase{e_s}{z~\{\ov{\rho_i \to e_i}\}}$,
where $e_s$ is the case \emph{scrutinee}, $z$ is the case \emph{binder}, and
$\ov{\rho_i \to e_i}$ are the case \emph{alternatives}, composed of a pattern
$\rho_i$ and of the corresponding expression $e_i$. 
%\begin{enumerate}
%\item
The case scrutinee is always evaluated to Weak Head Normal Form (WHNF) and
%
% \item
% evaluating an expression that is already in WHNF is a no-op, that is,
% no computation occurs; 
%
the case binder is an alias to the result of evaluating the scrutinee
to WHNF. Moreover, in Core, the alternative patterns are always exhaustive.
% i.e. there always exists
% a pattern that matches the WHNF of a value resulting from evaluating the
% scrutinee, where a pattern is either a wildcard that matches all expressions
% ($\_$), or a constructor and its linear and non-linear component binders
% ($K~\ov{x}\ov{y}$, with $\ov{x}$ as linearly-bound variables and $\ov{y}$ as
% unrestricted ones).

%\end{enumerate}

% ROMES:TODO: We should talk about WHNF in the background, not here.
% \parawith{WHNF} An expression is in Weak Head Normal Form when

We start with an
example of a program that constructs a tuple from linear resources then pattern
matches on it, then uses both linearly-bound variables from the tuple pattern
match. This is well-typed in Linear Haskell:
\begin{working}
\begin{lstlisting}
f6 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
f6 x y use = case (x,y) of z { (a,b) -> use a b }
\end{lstlisting}
\end{working}
What might be more surprising is that a similar program which discards the
pattern variables and instead uses the resources in the scrutinee is 
\emph{semantically} linear, despite not being accepted by Linear Haskell:
\begin{notyet}
\begin{lstlisting}
f7 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
f7 x y use = case (x,y) of z { (a,b) -> use x y }
\end{lstlisting}
\end{notyet}
We justify that \incode{f7} is linear by appealing to the fact that 
the tuple being scrutinized is already in WHNF and so evaluating it will
not consume either $x$ or $y$. Even if the tuple was constructed with
two expressions using $x$ and $y$ respectively, no computation would happen
since we are not using $a$ or $b$ (thereby never forcing the arguments
of the tuple). However, if we did use $a$ in the case body, then $x$
would be unavailable (as opposed to \incode{y}) and \incode{f8} below must be rejected:
\begin{noway}
\begin{lstlisting}
f8 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* a *' $\lolli$ '* c) -> c
f8 x y use = case (x,y) of z { (a,b) -> use a x }
\end{lstlisting}
\end{noway}
This idea that $x$ and $a$ are mutually exclusive is the same behind let
bindings being mutually exclusive to the resources that define them.
By forcing the pattern variable we run the computations
defined in terms of the linear variables used for that constructor
argument. Otherwise, if we do not use the pattern variables, then we
do not run the computation and thus no resources are consumed.

A third (semantically well-typed, but rejected by Linear Haskell) option is to use the case binder $z$ instead of $a,b$ or $x,y$:
\begin{notyet}
\begin{lstlisting}
f9 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
f9 x y use = case (x,y) of z { (a,b) -> uncurry use z }
\end{lstlisting}
\end{notyet}
Again, $z$ is mutually exclusive with $(a,b)$ and with $(x,y)$, but at least one of
the three must occur to ensure the linear resources are consumed. In this
example, we can think that using $a$ entails using the resource $x$, $b$ the
resource $y$, and the case binder $z$ entails using both $a$ and $b$.

Dually, consider the scrutinee to be an expression that's not in WHNF, s.t.
evaluating it to WHNF will require performing computation and thus consume linear
resources (\incode{f10} below is well-typed in our system but rejected
by Linear Haskell):
\begin{notyet}
\begin{lstlisting}
f10 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* (c,d)) -> (c,d)
f10 x y use = case use x y of z { (a,b) -> z }
\end{lstlisting}
\end{notyet}
Unlike when the scrutinee was in WHNF, we can no longer use $x,y$ in the case
alternatives, but we must still use either the case binder $z$ or the linear
pattern variables $a,b$. %  e.g. it would be quite disastrous if any of the following typechecked:
% \begin{noway}
% \begin{lstlisting}
% doubleFree :: Ptr *' $\lolli$ '* (Ptr *' $\lolli$ '* Result) -> Result
% doubleFree x free = case free x of z { Result v -> free x }
% \end{lstlisting}
% \end{noway}
% % Remover
% \begin{noway}
% \begin{lstlisting}
% leakPointer :: Ptr -> ()
% leakPointer x = case id x of z { _ -> () }
% \end{lstlisting}
% \end{noway}
%
%whereas this is fine:
%\begin{notyet}
%\begin{code}
%f16 :: Ptr -o (Ptr -o (Value,Int)) -> (Value,Int)
%f16 x free = case K (free x) of z { K y -> free x }
%\end{code}
%\end{notyet}
%
The result of evaluating the scrutinee must be consumed exactly once to guarantee
that the resources used in the scrutinee are fully consumed. %, or
                                %risk them being only ``almost''
                                %consumed.
For instance, take
\incode{use} in \incode{f10} to be the pairing function \incode{(,)}: it is not sufficient for \incode{use x y} to be
evaluated to WHNF to consume \incode{x} and \incode{y}. Otherwise, if all the resources were
considered to be fully consumed after the scrutinee were evaluated in a case
expression, we could simply ignore the pattern variables, potentially
discarding linear resources. In
short, if the scrutinee is not in WHNF we must either consume the case binder
or the linear components of the pattern.

On the other hand, if we match on a constructor without any linear components,
all linear resources used in the scrutinee must have been fully consumed in
that branch since the result of evaluating it can be consumed unrestrictedly
(by definition, a constructor application to unrestricted arguments is also
unrestricted).
%
Hence, in the branch for a constructor without linear fields, the case
binder can also be used unrestrictedly. For example, the program in
Figure~\ref{fig:nolin} is semantically linear.
%
\begin{figure}[h]
 \begin{minipage}{0.5\textwidth}
\begin{notyet}
\begin{lstlisting}
f11 :: () *' $\lolli$ '* ()
f11 x = case x of z { () -> z <> z }
\end{lstlisting}
\end{notyet}
\vspace{-0.5cm}
\caption{No Linear Fields\label{fig:nolin}}
\end{minipage}%
\begin{minipage}{0.5\textwidth}
\begin{limitation}
\begin{lstlisting}
f12 :: a *' $\lolli$ '* a
f12 x = case K1 x of z { K2 -> x; K1 a -> x }
\end{lstlisting}
\end{limitation}
\vspace{-0.5cm}
\caption{Absurd Branches\label{fig:absurd}}
\end{minipage}
\end{figure}
%
A second example of an unrestricted pattern, where \incode{K2} has no fields
and \incode{K1} has one linear field, is shown in Figure~\ref{fig:absurd}.
The use of \incode{x} in the \incode{K2} branch is valid despite \incode{x}
also occurring in the scrutinee because this branch is never taken. In fact,
any arbitrary resource could be used in absurd branches and be understood as
linear (despite not being seen as such by our system).

While many of these examples may seem artificial due to the pattern match on a
known constructor, we note that in between the transformations programs
undergo in an optimising compiler, many such programs naturally occur (e.g.,~if
the definition of a function is inlined in the scrutinee).
%
% However, the scrutinee before evaluation likely isn't in WHNF, thus scrutinee
% resources cannot directly occur in the case alternatives, since they might be
% potentially consumed when evaluating the scrutinee, the program leading up to
% this one wouldn't have been linear (even semantically).

% Further exploring that each linear field must be consumed exactly once, and
% that resources in WHNF scrutinees aren't consumed, we are able to construct
% more contrived examples (Figures~\ref{fig:badcase}
% and~\ref{fig:goodcase}). the example in Figure~\ref{fig:badcase} does not typecheck
% because the same linear field is used twice, but the example in
% Figure~\ref{fig:goodcase} does since it
% uses each linear field exactly once, despite pattern matching on the same
% components twice.

% %Remover estes
% \begin{figure}[t]
%   \begin{minipage}{0.5\textwidth}
% \begin{noway}
% \begin{lstlisting}
% f w = case w of z
%         (a,b) -> case (a,b) of z'
%                         (c,d) -> (a,c)
% \end{lstlisting}
% \end{noway}
% \vspace{-0.5cm}
% \caption{Bad Case\label{fig:badcase}}
% \end{minipage}%
% \begin{minipage}{0.5\textwidth}
% \begin{notyet}
% \begin{lstlisting}
% f w = case w of z
%         (a,b) -> case (a,b) of z'
%                         (c,d) -> (a,d)
% \end{lstlisting}
% \end{notyet}
% \vspace{-0.5cm}
% \caption{Good Case\label{fig:goodcase}}
% \end{minipage}
% \end{figure}

As for \emph{default} case alternatives, also known as \emph{wildcards}
(written $\_$): matching a wildcard doesn't provide any linearity information,
% unlike a pattern with linear or unrestricted components,
%
but the
scrutinee is still evaluated to WHNF.
%
% so linearity is seen as before but without
% fully consuming the scrutinee linear resources (in non-linear patterns) nor
% binding new linear resources (in linear patterns).
%
When matching a wildcard, if the scrutinee is already in WHNF,
programs that either use the resources from the scrutinee directly, or the case
binder in that alternative, can be seen as linear. If the scrutinee is not in
WHNF, we \emph{must} use the case binder, as it's the only way to linearly
consume the result of evaluating the scrutinee.

% \parawith{Summary}
% In summary, case expressions evaluate their scrutinees to WHNF,
% introduce a case binder, and bind pattern variables. If the scrutinee is
% already in WHNF, all resources occurring in it are still available in the case
% alternative, alongside the case binder and the pattern-bound variables. In the
% case alternative, either the resources of the scrutinee, the case binder, or
% the linearly bound pattern variables must be used exactly once, but mutually
% exclusively. For scrutinees not in WHNF, in the case alternative, either the
% case binder or the linear pattern variables need to be used, in mutual exclusion.
% If the pattern doesn't bind any linear resources, then it may be consumed
% unrestrictedly, and therefore the case binder may also be used unrestrictedly.

% This is fine, because |x| in the case alternative is known to be K...
% Variables really are weird case...
% ROMES:TODO: Do something about this?
% \begin{noway}
% \begin{code}
% f :: a -o (a -o K) -o a
% f x use = case x of z { K -> x }
% \end{code}
% \end{noway}

%We used to call NOT IN WHNF "negative" hah
%\begin{code}
%-- This is NOT OK since (g x y) is negative (eliminates g)
%f x y = case g x y of z
%          K1 a b -> (x,y)
%          K2 w   -> (x,y)
%\end{code}

%% The harder harder things

%\todo[inline]{Should we discuss this? It would be fine, but we're not able to see this because of call-by-name substitution}
%\begin{code}
%f x = case x of z { _ -> x }
%\end{code}

%\todo[inline]{the harder ones regarding reverse binder swap. these give some
%intuition, but this is an unsound optimisation in some contexts}
%\begin{code}
%f y = let x = use y
%       in case x of z { K a b -> expensive x; K2 w -> w }
%\end{code}
%
%\begin{code}
%f y = let x = use y
%      let t = expensive x
%       in case x of z { K a b -> t; K2 w -> w }
%\end{code}
%
%\begin{code}
%f x = case x of z { K a b -> expensive x; K2 w -> w }
%
%
%f y = let x = use y
%       in case x of z { K a b -> expensive x; K2 w -> w }
%
%
%f y = let x = use y
%      let t = expensive x
%       in case x of z { K a b -> t; K2 w -> w }
%\end{code}

% ROMES: We don't have time to do this I think
%%%\subsection{Generalizing linearity in function of evaluation\label{sec:generalizing-evaluation-consuming}}
%%%
%%%\todo[inline]{Deixar isto para último, dificil generalizar, se necessario cortar}
%%%
%%%Indeed, as hinted towards in the previous section, there's a deep connection
%%%between \emph{evaluation} and \emph{consuming resources}.
%%%
%%%Definition X.Y: A linear resource is consumed when it is either fully evaluated
%%%(NF) to a value, or when it is returned s.t. an application of that function
%%%being fully evaluated would fully evaluate the resource to a value. Or something like that.
%%%%
%%%Note how this generalizes Linear Haskell's definition of consuming a resource: $\dots$
%%%
%%%\begin{itemize}
%%%\item We could almost say that eventually everything all linear
%%%resources must be evaluated to NF to be consumed, or returned by a function
%%%s.t. a continuation of that function has to evaluate the result to NF., or something.
%%%
%%%\item Discuss our own generalized (call-by-value, call-by-name, etc)
%%%definition of consuming resources by evaluation. Something like, if an
%%%expression is fully evaluated, all linear resources that expression depends on
%%%to compute a result are consumed, or something...
%%%
%%%\item How does this relate to strictness? Reference the section of Linear
%%%Haskell about linearity and strictness, and basically revisit what they say.
%%%
%%%\end{itemize}

% }}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% {{{ Linear Core
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{A Type System for Semantic Linearity in Core\label{sec:main:linear-core}}

In this section, we develop a pure linear calculus $\lambda_\Delta^\pi$,
dubbed \emph{Linear Core}, that combines the linearity-in-the-arrow
and multiplicity polymorphism introduced by Linear
Haskell~\cite{cite:linearhaskell} with all the key features of GHC's Core
language (algebraic datatypes, case expressions, and recursive lazy let bindings),
except for equality coercions, which we discuss as future work in
Section~\ref{sec:future-work}. 
%

Linear Core makes precise the various insights discussed in the
previous section in a linear type system that we prove is sound and guarantees
linear resource usage at runtime.
This result is obtained (see Section~\ref{sec:main:metatheory}) via a soundness argument on a
call-by-need big-step operational semantics that becomes stuck when linear
variables in the runtime heap are used more than once. We show this
semantics to be bisimilar to a
Launchbury-style~\cite{10.1145/158511.158618} natural semantics. 
%\todo{O que digo aqui sobre a semantica launchbury style?}
Crucially,
Linear Core typing accepts all the semantically linear examples
(highlighted with \colorbox{notyet}{\notyetcolorname})
from Section~\ref{sec:semantic-linearity-examples}, which Core currently
rejects.
%
Additionally, we prove that multiple optimising Core-to-Core transformations
preserve linearity in Linear Core, while violating it under Core's
current type system.
%
%To the best of our knowledge, we are the first to prove optimisations preserve
%types in a non-strict linear language.
% 
Despite the focus on GHC Core, the fundamental ideas for
understanding linearity in our call-by-need calculus can be readily applied to
other non-strict languages.

Linear Core's first key idea is to delay consuming
resources until a computation that depends on those resources is effectively
evaluated or returned. Delayed consumption is tracked by annotating relevant
binders with \emph{usage environments} (Section~\ref{sec:usage-environments}).
%
The second key idea is to have two distinct rules for case
expressions, branching on whether the scrutinee is in Weak Head Normal
Form or not, and using ``proof irrelevance'' to track resources that are no
longer available but have yet to be fully consumed
(Section~\ref{sec:lc-case-exps}). Additionally, we introduce tags to split
resources amongst pattern-bound variables and %^as usage environments
encode that all linear components of a pattern must be used in tandem.

We present Linear Core's syntax and type system incrementally, starting with the
judgements and base linear calculi rules (Section~\ref{sec:base-calculi}).
%
Then, usage environments, the rule for $\Delta$-bound variables, and rules for
(recursive) let bindings (Section~\ref{sec:usage-environments}).
%
Finally, we introduce the rules to type case expressions and case alternatives,
along with the key insights to do so, namely branching on WHNF-ness of
scrutinee, proof irrelevant resources, and tagging (Section~\ref{sec:lc-case-exps}).

% We start by introducing the Core-like language, $\dots$ usage environments as a
% way to encode choice between the way a resource is used $\dots$\todo{$\dots$}
%

%\todo[inline]{Explicar algumas das ideias fundamentais, e apresentar as regras
%iterativamente. Podemos começar com as triviais e avançar para os dois pontos
%mais difíceis : Lets e Cases}

\subsection{Language Syntax}

The complete syntax of Linear Core is given in Figure~\ref{fig:full-linear-core-syntax}.
% Figure~\ref{fig:linear-core-types} and Figure~\ref{fig:linear-core-terms}.
%
The types of Linear Core are algebraic datatypes, function types, and
multiplicity schemes to support multiplicity polymorphism: datatypes
($T~\ov{p}$) are parametrised by multiplicities, function types
($\vp\to_\pi\s$) are also annotated with a multiplicity, which can be $1$,
$\omega$ (read \emph{many}), or a multiplicity variable $p$ introduced by a
multiplicity universal scheme ($\forall p.~\vp$).
%

%
Terms are variables $x,y,z$, data constructors $K$, multiplicity
abstractions $\Lambda p.~e$ and applications $e~\pi$, term abstractions
$\lambda \x[\pi].~e$ and applications $e~e'$, where lambda binders are
annotated with a multiplicity $\pi$ and a type $\s$. Then, there are
non-recursive let bindings $\llet{\xD = e}{e'}$, recursive let bindings
$\lletrec{\ov{\xD = e}}{e'}$, where the overline denotes a set of distinct
bindings $x_1{:}_{\D}\s_1\dots x_n{:}_{\D}\s_n$ and associated expressions
$e_1\dots e_n$, and case expressions $\ccase{e}{\zD~\{\ov{\rho \to e'}\}}$,
where $z$ is the case binder and the overline denotes a set of distinct
patterns $\rho_1\dots \rho_n$ and corresponding right hand sides $e'_1\dots
e'_n$. We often omit the case binder $z$ when it is unused. Notably, (recursive) let-bound
binders and case-bound binders are annotated
with a so-called \emph{usage environment} $\Delta$ -- a fundamental construct
for typing semantic linearity in the presence of laziness that we present in
Section~\ref{sec:usage-environments}.
%
Case patterns $\rho$ can be either the \emph{default/wildcard}
pattern $\_$, which matches any expression, or a constructor $K$ and a set of
variables that bind its arguments, where each field of the constructor has an
associated multiplicity denoting which pattern-bound variables must be
consumed linearly or unrestrictedly.
Additionally, the set of patterns in a case expression is assumed to be exhaustive,
i.e. there is always at least one pattern which matches the scrutinized expression.

% Linear Core takes the idea of annotating lets with usage environments from the
% unpublished Linear Mini-Core document by Spiwack et.
% al~\cite{cite:minicore}, which first tentatively tackled Core's
% linearity issues. We discuss this work in more detail in
% Section~\ref{sec:linear-mini-core}.

Datatype declarations $\datatype{T~\overline{p}}{\overline{K:\overline{\sigma
\to_\pi}~T~\overline{p} }}$ have the name of the type being declared $T$
parametrized over multiplicity variables $\ov{p}$, and a set of the data
constructors $K$ with signatures indicating the type and multiplicity of the
constructor arguments. Note that a linear resource is used many times when a
constructor with an unrestricted field is applied to it, since, dually, pattern
matching on the same constructor with an unrestricted field allows it to be
used unrestrictedly.
%Programs are a set of declarations and a top-level
%expression.

\SyntaxFull

%Linear Core is similar to Core without coercions, and with multiplicity
%abstractions instead of type abstractions.
%The key differences are that lets, recursive lets, case binders and
%pattern-bound variables
%the product of combining Linear Haskell's arrow linearity and
%multiplicity polymorphism with Core's key features, except for type equality
%coercions --
%a linear lambda calculus with algebraic datatypes, case
%expressions, recursive let bindings, and multiplicity abstractions. The

% The (small-step) operational semantics of Linear Core are given in
% Figure~\ref{fig:linear-core-operational-semantics}. We use call-by-name
% evaluation for Linear Core as it captures the non-strict semantics in which
% our type system understands linearity, while being simpler to reason about than
% call-by-need operational semantics.
% % which is traditionally modelled with a
% % mutable heap to store \emph{thunks} and the values they are overwritten with.
% Furthermore, linear function applications, even in a \emph{call-by-need} system, can
% usually be reduced \emph{call-by-name} as the function argument is guaranteed to
% be used exactly once (thus avoiding unnecessarily allocating memory on the heap
% for a redundant \emph{thunk}).
% %
% Specifically, function applications are reduced by the standard
% \emph{call-by-name} $\beta$-reduction, substituting the argument by
% occurrences of the binder in its body, case expressions evaluate their
% scrutinee to WHNF, substituting the result by the case binder and constructor
% arguments for pattern-bound bound variables matching on that same constructor.

% \input{language-v4/OperationalSemantics}

\subsection{Base Typing System\label{sec:base-calculi}}

Linear Core ($\lambda^\pi_\Delta$) is a linear lambda calculus akin to Linear
Haskell's $\lambda^q_\to$ in that both have multiplicity polymorphism,
(recursive) let bindings, case expressions, and algebraic data types.
$\lambda^\pi_\Delta$ diverges from $\lambda^q_\to$ primarily when typing lets,
case expressions, and alternatives (in its purpose to type semantic linearity).
%, and secondarily in only treating multiplicity polymorphism superficially.
%
% \todo{And our treatment of multiplicity polymorphism completely ignores
% algebraic treatment of multiplicities with semiring operations!!! Would it be
% sufficient to add a rule for application of variable multiplicity functions?}
%
%
The core rules of the calculus for abstraction and application are similar to
that of $\lambda^q_\to$ and mostly standard. 
We note that we handle multiplicity polymorphism differently from
Linear Haskell by ignoring the multiplicity semiring and
conservatively treating all multiplicity polymorphic functions as
linear, for the sake of simplicity.
%
The full type system is given in Figure~\ref{fig:linear-core-typing-rules},
with auxiliary judgements given in
Figure~\ref{fig:linear-core-other-judgements}.

\TypingRules
\TypingRulesOther

%We start with the main typing judgement.
%As is common for linear type systems, we use two typing environments, an
%unrestricted $\G$ and a linear $\D$ environment.
The main judgement is written $\G;\D \vdash e : \s$ to denote that expression
$e$ has type $\s$ under the unrestricted environment $\G$ and linear
environment $\D$.
%
%
Variables in $\G$ can be freely discarded (\emph{weakened}) and duplicated
(\emph{contracted}), while resources in $\D$ must be used exactly once. Despite
not having explicit weakening and contraction rules in our system, they are
available as admissible rules for $\G$ (but not for $\D$), since, equivalently
\cite{91621fae-5e53-3497-8291-32b2fab5a743}, resources from $\G$ are duplicated
for sub-derivations and may unrestrictedly exist in the variable rules.
% TODO: Likely rewrite that?
%

Occurrences of unrestricted variables from $\G$ are well-typed as long as the linear
environment is empty, while occurrences of linear variables are only well-typed
when the variable being typed is the only resource available in the
linear context (rules $Var_\omega$ and $Var_1$, respectively).
% \[
% \begin{array}{cc}
% \TypeVarOmega & \TypeLinearVar
% \end{array}
% \]
In both cases, the linear context must contain exactly what is required to
prove the proposition, whereas the unrestricted context may contain arbitrary
variables.
%
Variables in contexts are annotated with their type and multiplicity, so
$\x[\pi]$ is a variable named $x$ of type $\s$ and multiplicity $\pi$.

Linear functions are introduced via the function type ($\s \to_\pi \vp$) with
$\pi = 1$, i.e. a function of type $\s \to_1 \vp$ (or, equivalently, $\s \lolli \vp$)
introduces a linear resource of type $\s$ in the linear environment $\D$ to then type an expression of type $\varphi$.
%
Unrestricted functions are introduced via the function type ($\s \to_\pi \vp$) with $\pi =
\omega$, and the $\lambda$-bound variable is introduced in $\G$ (rules
$\lambda I_1$ and $\Lambda I$). Function application is standard, with
linear function application splitting the linear context in two
disjoint parts, one used to type the argument and the other the
function.
Arguments of unrestricted functions must also be unrestricted, i.e. no
linear variables can be used to type them. 
% \[
% \begin{array}{cc}
% \TypeLamIntroL & \TypeLamIntroW
% \end{array}
% \]
% A linear function application is well-typed if there exists a disjoint split of the
% linear resources into $\D,\D'$ s.t. the function and argument, each under a
% distinct split, are both well-typed and the argument type matches the
% function's expected argument type. Conversely, unrestricted resources are
% duplicated and available whole to both sub-derivations.
% \[
% \TypeLamElimL \quad \TypeLamElimW
% \]
% An unrestricted function, unlike a linear one, consumes its argument
% unrestrictedly (zero or more times). Therefore, in an unrestricted function
% application, allowing any linear resources to occur in the argument expression
% entails consuming those resources unrestrictedly, since the
% variable binding the argument expression could be discarded or used more than
% once in the function body. Thus, argument expressions to unrestricted functions must also
% be unrestricted, i.e. no linear variables can be used to type them.
% %
%
Typing linear and unrestricted function application separately is less
general than typing applications of functions of any multiplicity
$\pi$ by scaling (per the multiplicity semiring) the multiplicities of
the resources used to type the argument by $\pi$, however, our
objective of typing semantic linearity does not benefit much from
doing so and therefore we opt for a simpler design.
% \todo{We might only need multiplicities for the case-of-case definition as it
% exists in ghc?, even then couldn't we do semiring scaling without variables?
% Also, dislike sentence}

Multiplicity abstractions ($\Lambda I$) introduce a multiplicity variable $p$,
and construct expressions of type $\forall p.~\s$, i.e. a type universally
quantified over a multiplicity variable $p$. Again, conservatively, in the body
of the abstraction, function types annotated with a $p$ variable and datatype
fields with multiplicity $p$ are typed as though they are linear functions and
linear fields, because $p$ can be instantiated at both $\omega$ and $1$.
%\[
%%\TypeMultLamIntro \quad \TypeMultLamElim
%\]
A multiplicity application ($\Lambda E$) instantiates a multiplicity-polymorphic type
$\forall p.~\s$ at a particular (argument) multiplicity $\pi$, resulting in an
expression of type $\s$ where occurrences of $p$ are substituted by $\pi$, i.e.
$\s[\pi/p]$.
The rule additionally requires that $\pi$ be \emph{well-formed} in order
for the expression to be well-typed, using the judgement $\G \vdash_{mult}
\pi$, where well-formedness is given by $\pi$ either being $1$, $\omega$, or an
in-scope  multiplicity variable in $\G$.
% \[
% \TypeWellFormedMult
% \]

% These rules conclude the foundations of our linear calculi. In subsequent
% sections we type (recursive) let bindings and case expressions,
% accounting for semantic linearity as per the insights construed in
% Section~\ref{sec:linearity-semantically}, effectively distilling them into the
% key ideas of our work -- encoded as rules.

\subsection{Usage environments\label{sec:usage-environments}}

A \emph{usage environment} $\Delta$ encodes the idea that lazy
bindings do not consume resources upon definition, but rather when the bindings themselves are consumed.
Specifically, we annotate so-called $\Delta$-bound variables with a \emph{usage
environment} to denote that consuming these $\D$-variables equates to consuming the
resources in the annotated usage environment $\D$ (which is essentially a
multiset of linear resources).
%
$\Delta$-bound variables are introduced in the unrestricted environment by
several constructs such as (recursive) let binders, case binders, and case
pattern variables.
%
For example, the let-binder $u$ in the following program is annotated with a
usage environment $\{x{:}_1\s,y{:}_1\s\}$ tracking the resources that are used in its
body. Thus, in the expression $e$, consuming $u$ will equate to consuming the
linear resources $x$ and $y$:
% , as per the insights developed in
% Section~\ref{sec:semantic-linearity-examples},
\[
f = \lambda \xl.~\lambda \y[1].~\llet{u_{\{x:_1\s,y:_1\s\}} = (x,y)}{e}
\]
Furthermore, usage environments guarantee that using a $\Delta$-bound variable
is mutually exclusive with directly using the resources it is annotated with --
using the $\Delta$-bound variable consumes all linear resources listed in its
usage environment, meaning they are no longer available for direct usage.
Dually, using the linear resources directly means they are no longer available
to consume through the usage environment of the $\Delta$-bound variable. The
$\Delta$-bound variable can be left unused since it is unrestricted, regardless
of the variables it is annotated with.

% Finally, we note that usage environments bear a strong resemblance to the linear
% typing environments in the main typing judgement,
% i.e. the environment with the linear resources required to type an expression.
%
% In fact, usage environments and linear typing contexts differ only in that the
% former used to annotate variables, while the latter used to type
% expressions.
%
% Yet, this distinction is slightly blurred after introducing how
% typing environments can be moved to usage environments, or otherwise occurs in
% rules relating the two.

\paragraph{\texorpdfstring{$\D$}{Delta}-bound variables}

A $\Delta$-bound variable $u$ is a variable annotated with a usage environment $\Delta$. Crucially, for any $\Delta$-bound variable $u$:
%
%\begin{enumerate}
%\item
(1) using $u$ is equivalent to using all the linear resources in $\Delta$;
%\item $u$ can only be used once, since the $\D$ resources will not be available afterwards
%\item
(2) using $u$ is mutually exclusive with using the $\Delta$ resources it depends on;
% \item
(3) $u$ can be safely discarded as long as the resources in $\Delta$ are consumed.
%\end{enumerate}
%
Fortunately, since linear resources must be linearly split across
sub-derivations, (2) follows from (1) since consuming the linear
resources in $\Delta$ to type $u$ makes them unavailable in the
context of any other derivation.
% Therefore, expressions using these
% resources a second time, directly, or indirectly through the same (or
% other) usage environment, are ill-typed, as the resources are already
% allocated to the derivation of $u$.
Similarly, (3) also follows from (1), because if the linear resources aren't
consumed in the $\Delta$-var derivation, they must be consumed in an
alternative derivation (or otherwise the expression is ill-typed).
%
These observations all lead to the typing rule for $\Delta$-bound
variables:

\vspace{-0.5cm}
\[
\TypeVarDelta
\]
The rule states that an occurrence of a $\Delta$-bound variable is well-typed if
the linear environment is made up exactly of the resources in the usage environment of
that variable.
%
$\D$-vars are unrestricted and can be discarded and duplicated, despite
multiple occurrences of the same $\Delta$-var in an expression not possibly being well-typed
since that would imply non-linear usage of linear resources.
% multiple occurrences of the same
% $\D$-variable imply non-linear usage of resources that must be used linearly.

\subsection{Lazy let bindings}

In Section~\ref{sec:semantic-linearity-examples}, we discussed how linear
resources used in let-bound expressions are only consumed when the respective
binders are evaluated in the let body.
%, i.e. linear resources required by let-bound expressions are consumed lazily.
%
Moreover, resources captured by a let-bound expression cannot be used
simultaneously with the let-binder in its body, since said
resources would end up being consumed more than once and violate (semantic)
linearity. The binder has to be used in mutual exclusion with the linear
resources captured by the expression it binds -- either the binding or the
resources \emph{must} be used.

% Indeed, usage environments allow us to encode mutual exclusivity between
% alternative ways of consuming linear resources (between $\D$-vars and direct
% resource usage). 
Let-bound variables are the canonical example of a
$\Delta$-bound variable.
% , that is, let-bound variables bind expressions in which
% the resources required to type them are consumed lazily rather than eagerly.
%
Annotating let-bound variables with a usage environment $\D$ delays the
consumption of resources to when the variables themselves are used and
guarantees the desired mutual exclusion property between the binder and the resources:
\[
\TypeLet
\]
The rule for (non-recursive) let bindings splits the linear environment in $\D$
and $\D'$. $\D$ is used to type the body $e$ of the let binding $x$. Perhaps
surprisingly, the resources $\D$ used to type $e$ are still available in the
environment to type the let body $e'$, alongside the unrestricted $x$ binding
annotated with the usage environment $\D$. In essence, the resources of $e$ being
available in $e'$ reflects how typing the body of a lazy binder
doesn't consume resources, and the binding $x$ being an unrestricted
$\D$-variable ensures that using it will consume the resources $\D$ the
expression captures, \emph{if} it is used.
% \todo[inline]{Let bindings are hard, if they are used then we use resources. If
% they don't get used then we use no resources! In practice, resources that show
% up in the body of the let must be used, be it by using the let binder, or by
% using them directly. This makes the let binder and the resources in its body
% mutual exclusive.}

% \todo[inline]{Explain the idea of suspended computation, and how resources will
% be consumed to some extent when we force the computation -- also foreshadowing
% that evaluation to WHNF doesn't necessarily consume all resources}

\subsection{Recursive let bindings}\label{sec:recursivelets}

% Recursive let bindings are very similar to non-recursive ones, the main
% exception being that the recursive bindings may be defined in terms of
% themselves, and we may have more than one binding.
%
Recursive let bindings also bind expressions lazily and so
introduce a $\D$-bound variable for each binding. Resources required to type the
let-bindings are still available in the body of the let, to later be consumed
via $\D$-bound variables, or, directly, if the let-bindings are unused.
%
%Additionally, as seen in Section
%\ref{sec:semantic-linearity-examples:recursive-lets}, recursive uses of a
%binder in its own definition entail consuming all resources otherwise required
%to type the binder's body.

%We assume all mutually recursive groups of bindings are strongly-connected.
%As for groups of recursive let bindings, we recall that, in our
%system, all recursive let groups are strongly-connected -- all bindings in a
%recursive let group are mutually recursive in the sense that they all
%(transitively) depend on one another.
%
% In light of it, (mutually) recursive uses of binders entail consuming all the
% resources required to type all of the binders.
%
%By definition, any binder in such a group may end up recursively using
%any other binder from the group and thus use up any resource occurring.
%
%Accordingly, occurrences of any such binder
%must be typed conservatively, using up the sum of resources needed to type each
%of the binders,
%%
%i.e.,
%Accordingly, all the binders in a strongly-connected group of mutually recursive let
%bindings have to be typed with the same linear resources (which are the least
%upper bound of resources needed to type the bodies of all binders, accounting
%for uses of mutually-recursive binders).
%
% Therefore, we conclude the least upper bound of resources required to type a
% mutually recursive group of let bindings to be the same for any such group that
% is strongly-connected.

The typing rule for recursive groups of bindings leverages both the assumption
that all mutually recursive groups are strongly connected and the corollary
that every binder in such a group must be typed with the same linear context,
as observed in Section~\ref{sec:semantic-linearity-examples:recursive-lets}.
Consequently, all bindings of the recursive group are introduced as
$\D$-bound variables
sharing the same usage environment -- using any one of the bindings in a
recursive group entails consuming all resources required to type that same
group -- so, at most a single binder from the group can be used in the
let body.
%\todo{should we say how this means at most once of the binders from the group can be used?} OK
%, which is also why we can use the same linear resources to type each binder:
\[
\TypeLetRec
\]
This formulation is not syntax-directed 
% (from which an implementation is direct) 
because determining a particular $\D$
to type and annotate all binders is not obvious. We present our system and
metatheory agnostically to the challenge of inferring this linear typing
environment by assuming recursive let expressions are annotated with the
correct typing environment.

In practice, determining this typing environment $\D$ amounts to finding a
least upper bound of the resources needed to type each mutually-recursive
binding that (transitively) uses all binders in the recursive group.
%
Our implementation uses a naive $O(n^2)$ algorithm for inferring usage
environments of recursive bindings.
% , which is not relevant to the theory developed in this section.
%
% The algorithm is a $O(n^2)$ traversal over the so-called \emph{naive usage
% environments} used to type each binding.
%
Moreover, we note that inference of usage environments for recursive binding
groups bears some resemblance to the inference of principle types for recursive
bindings traditionally achieved through the Hindley–Milner inference
algorithm~\cite{DBLP:conf/popl/DamasM82}. % , there might be an opportunity to
% develop a better algorithm leveraging existing inference techniques.
% %
% Despite being a seemingly useful observation, 
We leave the exploration of this connection
as future work.

% We present a naive algorithm for inferring usage environments of recursive bindings in
% Section~\ref{sec:impl:recursive-alg} and leave exploring this potential
% connection as future work.

\subsection{Case Expressions\label{sec:lc-case-exps}}

%Case expressions \emph{drive evaluation} --
% thus they are key to realize a type system that understands linearity in the presence of lazy evaluation.
%
%However, the evaluation of case expressions is considerably nuanced,
%
%In lazy let bindings, computations  can be a case
%expression can effectively consume resources rather than just 
%
A case expression drives evaluation by evaluating its scrutinee to Weak Head
Normal Form (WHNF)~\cite{10.5555/1096899}. Then, the case
alternative whose pattern matches the result of evaluating the
scrutinee is taken\footnote{In our calculus, the alternatives are assumed to be
  exhaustive, i.e. there always exists at least one pattern which
  matches the scrutinee in its WHNF, so we're guaranteed to have an
  expression to progress evaluation.}.
An expression in WHNF can either be %\begin{itemize}
% \item
a $\lambda$-abstraction $(\lambda x.~e)$ 
or a datatype constructor application $(K~\ov{e})$.
%\end{itemize}
In both cases, the sub-expressions $e$ or $\ov{e}$ occurring in the lambda body
or as constructor arguments need not be evaluated for the lambda or constructor
application to be in WHNF % (if otherwise all sub-expressions
% were fully evaluated the whole expression would also be in \emph{normal form}).
% (that is why it is called \emph{weak} head normal form).
%
(and they may capture linear resources).

A key detail is that when the scrutinee is already in WHNF, evaluation continues
in the alternative simply by substituting the pattern variables by the
scrutinee sub-expressions. No resource from the scrutinee can be consumed since
no computation happens at all.
% (the scrutinee was already in weak head normal form).
%In short, no computation happens if the scrutinee is already in WHNF.
%
% In light of this, our type system will need to distinguish between terms in WHNF and terms
% that are not in WHNF.
%
% TODO While it may seem ad-hoc to type an expression differently
%
We introduce a typing judgement $\G;\D \Vdash e : \s \gtrdot \ov{\D_i}$ to
extract additional information from the static structure of terms in WHNF:
\[
    \TypeWHNFCons
\qquad
    \TypeWHNFLam
\]
This judgement differs from the main typing judgement in that (1) it only
applies to expressions in weak head normal form, and (2) it ``outputs'' (to the right of $\gtrdot$) a
disjoint set of linear environments ($\ov{\D_i}$), where each environment corresponds to the
linear resources used by a sub-expression of the WHNF expression.
%
To type a constructor application $K~\ov{e_\omega e_i}$, where $e_\omega$
are the unrestricted arguments and $e_i$ the linear arguments of the
constructor, we split the resources $\D$ into a disjoint set of resources
$\ov{\D_i}$ required to type each linear argument individually and return exactly
that split of the resources; the unrestricted $e_\omega$ expressions must be
typed on an empty linear environment. A lambda expression is typed with
the main typing judgement and trivially ``outputs'' the whole $\D$ environment,
as there is always only a single sub-expression in lambda abstractions.

% Precisamos disto?
% %
% \begin{tabbing}
% $\ccase{e}{z~\{\ov{\rho_i \to e_i}\}} \longrightarrow^* \ccase{e'}{z~\{\ov{\rho_i \to e_i}\}}$\`where $e'$ is in WHNF and $e$ is not\\
% $\ccase{K~\ov{e}}{z~\{K~\ov{x} \to e_i\}} \longrightarrow e_i\ov{[e/x]}[K~\ov{e}/z]$\\
% $\ccase{e}{z~\{\_ \to e_i\}} \longrightarrow e_i[e/z]$\`where $e$ is in WHNF\\
% \end{tabbing}
% %
% Recall how the operational semantics encode the evaluation of case
% expressions such that,
% when a scrutinee $K~\ov{e}$ matches a constructor pattern $K~\ov{\x[\pi]}$,
% evaluation proceeds in the case alternative corresponding to the matching
% pattern, with occurrences of $\ov{x}$ being substituted by $\ov{e}$ and
% occurrences of the case binder $z$ substituted by the whole scrutinee
% $K~\ov{e}$. Constructors and lambda expressions may otherwise match match the wildcard
% pattern, resuming evaluation on the alternative body only substituting the case
% binder by the scrutinee.

%%%%???
%%%%The linear resources used to type an expression in WHNF are exactly those which
%%%%occur to the right of $\gtrdot$ in the WHNF judgement shown above
%%%%(corresponding to the resources required to typecheck the lambda body or the
%%%%constructor arguments).

%Recalling that patterns are either a wildcard that matches anything or data
%constructors parametrised by variables to bind the constructor arguments in the
%case alternative
%
%Let's consider how resources are consumed when the case exp ...
%
%Patterns data constructors with variables to bind the constructor arguments parametrised
%
%\begin{itemize}
%\item The wildcard pattern matches against anything, and is the only pattern that matches against a lambda function
%\item We can match against the data constructor and 
%\end{itemize}
%
%Considering patterns are either data constructors and variables bounding
%A function (lambda expression) can only be matched against the wildcard pattern $\_$, but 
%
%Evaluation continues in the selected branch by
%substituting, first, the pattern variables by the (possibly unevaluated)
%expressions used as arguments to the

% \todo[inline]{Case expressions are the means by which we do evaluation and
% pattern matching -- when things are scrutinized, we evaluate them (if they
% aren't evaluated -- tag is 0), and then compare the result against multiple
% alternatives}

% \todo[inline]{Item 2.
%     Pattern matching an expression that is evaluated will not consume all
%     the resources that define that computation -- because of laziness, we only
%     evaluate things to WHNF. To fully consume a value, we need to consume all
%     the linear components of that pattern.
% }

\subsubsection{Branching on WHNF-ness}

The dichotomy between evaluation (and resource usage) of a case expression
whose scrutinee is in weak head normal form and one whose scrutinee is not
leads to one of the key insights of $\lambda^\pi_\Delta$: we must \emph{branch}
on weak head normal formed-ness to accurately type case expressions.
%
When the scrutinee is already in WHNF, the resources are
unused upon evaluation and thus available in the alternatives.
%
When it is not, resources will be consumed and cannot be used in the
alternative.
%
To illustrate, consider the following case expressions:
\[
\begin{array}{ccc}
    (1)~\lambda x.~\ccase{K~x}{z~\{\_ \to x\}} &  & (2)~\lambda x.~\ccase{free~x}{z~\{\_ \to x\}}
\end{array}
\]
The first function uses $x$ linearly, but the second does not.
%
Alternatives may also use the case binder or pattern variables, referring to, respectively,
the whole scrutinee and all its used resources; or
constructor arguments and the resources used to type them.
% Using the case binder entails using all the resources required by the scrutinee, and using a pattern
% variable implies using the resources of the corresponding constructor argument.

There are three competing ways
to use the resources from a scrutinee in WHNF in a case alternative: directly, via
the case binder, or by using pattern-bound variables.
%
Recall how $\D$-bound variables encode mutual exclusion between alternative
ways of consuming resources -- it follows that case binders and pattern-bound
variables are yet another instance of $\D$-bound variables.
%
%Resources in a scrutinee that is already in WHNF are only consumed when all
%(linear) fields of the pattern are used, satisfying the definition of
%consuming resources given in Linear Haskell.
%
% We type a case expression whose scrutinee is in weak head normal form with
This suggests the following first rule for cases of scrutinees already in WHNF:
\[
\TypeCaseWHNFIntermediate
\]

First, we assert this rule is only applicable to expressions in weak head
normal form. Second, we appeal to the typing judgement for expressions in WHNF
to determine the split of resources amongst the scrutinee
sub-expressions. Finally, we type all case alternatives with the same context, using the
$\vdash_{alt}$ judgement, introducing
the case binder $z$ in the unrestricted environment as a $\D$-bound
variable whose usage environment is the linear resources $\ov{\D_i}$ used to type the
scrutinee. Those same resources $\ov{\D_i}$ are again made available in the
linear typing environment of the alternative, similarly to how the resources
used to type a $Let$ binder are still available in the continuation.
%
%The $\vdash_{alt}$ judgement is 
%annotated with the disjoint set of linear resources $\ov{\D_i}$ used
%to type the scrutinee sub-expressions,
%and the name of the case binder $z$.
%
%
Although the main idea for typing ``WHNF case expressions'' is conveyed by
this rule, the full rule of our system is slightly more involved, as will be seen after
discussing cases of scrutinees not in WHNF.

The alternative judgement $\G;\D \vdash_{alt} \rho \to e :^z_\D \s \Rightarrow
\vp$ is used to type case alternatives, encompassing three
modes that are distinguished by the kind of arrow that is used:
for alternatives of case expressions whose scrutinee is in WHNF
($\Rightarrow_\mathsf{WHNF}$), not in WHNF ($\Rightarrow_\mathsf{NWHNF}$),
and for alternatives agnostic to the WHNF-ness of the scrutinee
($\Rightarrow$), with $\Rightarrow$ also generalizing the other two.
%
% Following the $Case_\textrm{WHNF}$ rule in which we use the $\Mapsto$ alternative
% judgement, t

First, an alternative whose pattern is a constructor with $n > 0$ linear components
under the $\Rightarrow_{\mathsf{WHNF}}$ mode is typed as:
\[
\TypeAltNWHNF
\]
The rule states that, for such a match on a scrutinee already in
WHNF, we introduce the linear components of the pattern as $\D$-bound variables
with usage environment matching the linear resources required to type the
corresponding constructor argument in the scrutinee. The resources required to
type each constructor sub-expression come annotated in the judgement (as
${:}_{\ov{\D_i}}$). Unrestricted fields of the constructor are introduced as
unrestricted variables. We note that the typing environment $\D$ always
contains the resources $\ov{\D_i}$ in uses of the alternative judgement.

Second, the rule for alternatives that match on the wildcard pattern,
regardless of the WHNF status of the scrutinee (note the use of $\Rightarrow$, which
is applicable both under $\Rightarrow_{WHNF}$ and $\Rightarrow_{NWHNF}$):
\[
\TypeAltWild
\]
To type a wildcard alternative we simply type the expression with the main
judgement, ignoring all annotations. The case binder will
have already been introduced in the environment with the appropriate usage
environment by the relevant %(WHNF or not)
case expression rule.

Lastly, consider an alternative matching on a data constructor without any
linear components. The linear resources used to type a scrutinee matching such
a pattern are fully consumed during evaluation: the resulting unrestricted
constructor application can only be well-typed under an empty linear
environment.
%
Consequently, the resources to type the scrutinee which are carried over to the
case-alternative environments by the $Case_{WHNF}$ and $Case_{NWHNF}$ rules
(see the latter below), must be reactively destroyed from branches where the
pattern is unrestricted. In failing to do this, we would allow the fully-consumed
resources to be used again in the branch.
% Really, in the WHNF case it is not really necessary to remove the things
% since they must already be empty be definition.
%
%This definition agrees with the intuition we have developed by example in the
%previous section, and with the typing rule we devised for alternatives matching
%constructors without linear components.
%
%Taking into account that case expressions introduce the linear resources of the
%scrutinee in the typing environment of all alternatives, and in the usage
%environment of the case binder, we must reactively update the typing
%environments after matching on such a pattern.
%
The $Alt0$ rule essentially encodes this insight, and is applicable regardless
of the WHNF status of the scrutinee ($\Rightarrow$ mode), as long as
the constructor pattern has no linear fields:
%
\[
\TypeAltZero
\]
The rule deletes the annotated scrutinee environment $\D_s$ from
%two select environments:
the linear typing context (written $\D[\cdot/\D_s]$, a substitution of
the scrutinee typing environment by the empty linear environment $\cdot$)
and from the usage environment of the case binder $z$ (written $\G[\cdot/\D_s]_z$
to similarly delete $\D_s$ from the usage environment of the $\D$-var $z$ in $\G$).
Given that the case-binder $z$ is always annotated with $\D_s$ (essentially,
the linear environment typing the scrutinee), this latter substitution makes
the case-binder unrestricted in branches with unrestricted patterns.
%-- agreeing with our observation in Section \label{sec:semantic-linearity-examples}.

% \begin{itemize}

% \item The linear typing environment, effectively deleting the resources from
% the scrutinee made available by the case expressions (written
% $\D[\cdot/\D_s]$, a substitution of the scrutinee typing environment by the
% empty linear environment $\cdot$).

% \item The usage environment of the case binder $z$, written $\G[\cdot/\D_s]_z$
% to denote replacing the usage environment of the variable $z$ in $\G$, which
% is necessarily $\D_s$ (since we always annotate the judgement with the
% environment of the scrutinee), by the empty environment.

% \end{itemize}
%
This rule observes that any linear resources required to type an expression
that reduces to an unrestricted constructor application are fully consumed
through evaluation and the resulting unrestricted expression can be freely
discarded or duplicated (e.g. via the case binder).
%
%Reactively deleting the resources that typed the scrutinee from the linear
%environment ensures that when we match on an unrestricted pattern there are no
%remaining linear resources to consume. Otherwise, e.g., $\ccase{K_1~x}{\{K_2 \to
%K_2,K_1~y \to K_1~y\}}$ would not be well-typed since the resource $x$ would
%be left \emph{un}-consumed in the first alternative, even though we know by
%matching on $K_1$ that it has already been consumed.
%
%Furthermore, since the case binder in such an alternative refers to the
%unrestricted expression, the case binder too may be used unrestrictedly, which
%we allow by emptying its usage environment.

%It might seem as though deleting the resources from the environment in $\textsc{Alt}_0$
%is necessary to guarantee a resource is not used after it is consumed.
%%
%However, let us consider two situations -- pattern matches in a case
%expression whose scrutinee is in WHNF, and matches on a case expression whose
%scrutinee is \emph{not} in WHNF:
%%
%%\begin{enumerate}
%When the scrutinee is in WHNF, it is either an unrestricted expression
%against which any match will only introduce unrestricted variables, or an
%expression that depends on linear resources. The former trivially allows
%any resource from the scrutinee in the alternatives as well. The latter is
%further divided: 
%%\begin{enumerate}
%%
%%\item
%The pattern is unrestricted while the
%scrutinee is not, so entering this branch is impossible as long as the case
%expression is well-typed; by contradiction, the linear resources from the
%scrutinee could occur unrestrictedly in that branch.
%%
%For uniformity, we type such alternatives as those for scrutinees that
%are not in WHNF;
%%
%The pattern is linear and matches the scrutinee, in which case the
%$AltN_{\textrm{WHNF}}$ is applicable instead of $Alt0$;
%%
%The pattern is linear but does not match the scrutinee and so any
%resource could be used in such alternatives. For uniformity,
%it is also typed as though the scrutinee were not in WHNF.

%If the scrutinee is not in WHNF, the resources occurring in the
%scrutinee will be consumed when evaluation occurs. Therefore, the resources
%used in the scrutinee cannot occur in the alternative body
%(e.g. $x$ cannot occur in $e$ in $\ccase{close~x}{\{K_1 \to e\}}$)
%-- regardless of the pattern.
%%
%% In fact, the resources from a scrutinee that is not in weak head normal form
%% cannot occur in any of the alternatives, even ones matching on constructors
%% with linear components, as the resources may have been consumed when evaluating
%% the expression to weak head normal form.
%%
%We guarantee resources from a scrutinee that is not in WHNF
%cannot occur/directly be used in any case alternative in our rule for typing
%cases not in WHNF, which we introduce below.

%\end{enumerate}

\subsubsection{Proof irrelevant resources}

When the scrutinee is not in WHNF, it is not statically known which resources
will have been consumed after evaluating said scrutinee, or which
are transferred into the branch continuation, e.g. as in $\ccase{f\ x}{z\ \{K\
a \rightarrow e\}}$:
\begin{enumerate}

    \item Cases with non-WHNF scrutinees must be typed such that no resource
    from the scrutinee can be used directly in the branch again, or we
    risk
    consuming the same resource twice. In the example, $x$ must not directly occur
    in $e$ since $f$ \emph{could} have consumed it.

    \item On the other hand, since evaluation to WHNF does not guarantee all
    resources are consumed (unless the result matches an
    unrestricted pattern), we must ensure that any resources that transfer from the
    scrutinee to the alternative are ultimately consumed (in the example, $x$
    is not consumed by evaluating the scrutinee for $f = K$).

    Considering (1), we must resort to a more uniform method of guaranteeing we
    ``finish consuming'' the scrutinee. There are only two ways of
    uniformly referring to the remaining linear resources in the newly-evaluated scrutinee:

        \begin{itemize}
            \item The case binder, which refers to the whole result of evaluating the scrutinee;
            \item The linear components of a constructor pattern, which refer to expressions that may contain linear resources.
        \end{itemize}

%To guarantee all post-evaluation unconsumed scrutinee resources are consumed in
%the alternative, without statically knowing
%Statically we can't know which resources are ``transferred'', but  guarantee
%
\end{enumerate}

%is not sufficient to
%consume all resources used by the scrutinee, since sub-expressions such
%as constructor arguments will be left unevaluated and may still contain
%(``parts of'') the linear resources. To \emph{fully} consume all
%resources occurring in the scrutinee, all linear components of the
%constructor must also be fully evaluated, as witnessed by the
%$\textsc{Alt}_0$ rule above.
%%
%In essence, to type cases scrutinizing expressions not in WHNF, we need to
%forbid the usage of the scrutinee resources while tracking that we necessarily
%``finish consuming'' them, either via the case binder, or by linearly using all
%linear pattern-bound variables introduced in the alternative.

% We tackle this in due time, in the proof irrelevance section.
% \begin{itemize}

% \item The scrutinee resources must \emph{not} be used directly in the case alternatives;
% \item But the result of evaluating the scrutinee to WHNF must still be
% consumed, as all sub-expressions of the scrutinee remain unevaluated and must
% be consumed.
% \item Since the scrutinee resources cannot be consumed directly, they must be
% consumed indirectly through $\D$-variables, namely, either the case binder, or
% the linear pattern-bound variables introduced in the alternative.

% \end{itemize}

% In alternatives of a case where the scrutinee is not in
% WHNF, we must also consume the result of evaluating the scrutinee to WHNF, but the
% scrutinee resources must definitely not be available for consumption. In
% practice, the result of evaluating the scrutinee must be consumed by using
% either the case binder or all the linear components of a constructor pattern,
% except for patterns matching an unrestricted pattern, which are handled with
% the $Alt0$ rule. For WHNF scrutinees, we encode mutual exclusivity between
% consuming resources directly, with the case binder, or through linear pattern
% variables, by introducing the latter two as $\D$-bound variables.  In essence,
% for the counterpart not-WHNF scrutinees, either the case binder or linear
% pattern-bound variables \emph{must} still be used to guarantee the evaluation
% result is consumed (thus their usage environment cannot be empty), but the
% scrutinee resources cannot be used directly.

To encode  linear resources that cannot be
directly used (i.e.,~to which the $\textsc{Var}$ rule is not
applicable) we introduce \emph{proof irrelevant} resources, written as linear resources
within square brackets $[ x{:}\s]$ and lifted to contexts as $[\D]$.
Proof irrelevant resources
are linear resources in every sense, meaning they must be used
exactly once. However, since proof irrelevant resources cannot be
discarded or used directly, they have to be consumed \emph{indirectly} via
$\D$-bound variables, namely, the case binder, or, in mutual exclusion, the linear pattern-bound
variables.

Hence, to type a case expression whose scrutinee is not in WHNF, we
type the scrutinee with linear resources $\D$ and the case
alternatives by introducing the case binder with a usage environment $[\D]$,
having the same proof irrelevant linear context $[\D]$ in the typing
environment, and annotating the judgement with the proof irrelevant resources
for use in the $\Rightarrow_{\textsf{NWHNF}}$ judgement:
\[
\TypeCaseNotWHNF
\]
Note how the rule is similar to the one for scrutinees in WHNF, but
requires the resources in the case binder, typing environment, and
judgement annotation to be made ``irrelevant''.

% In practice, we can't know which resources are consumed by evaluating a given
% expression. The resources become in a limbo state -- they cannot be used
% directly because they might have been consumed, but they mustn't be considered
% as consumed, because they might not have been.  We say these resources enter a
% proof irrelevant state. They must still be tracked as though they weren't
% consumed, but they cannot be used directly to construct the program. How can we
% ensure these proof irrelevant resource variables are fully consumed? With usage
% environments -- for the case binder and for the pattern variables, and
% otherwise propagate

% \todo[inline]{The case binder and pattern variables will consume the scrutinee
% resources, be those irrelevant or relevant resources}

Finally, we recall the tentative $Case_\textrm{WHNF}$ rule presented before and
highlight its flaw: the $\G;\D \vdash_{alt} \rho \to e :^z_{\D_s} \s \Rightarrow_{\textsf{WHNF}}
\vp$ judgement is only well-defined for patterns $\rho$ matching the WHNF form
of the scrutinee, as the distribution of resources per constructor components
only makes sense for the constructor pattern matching the scrutinee.
% Essentially, if the scrutinee is in WHNF the matching alternative is easily
% determined, and must be treated with the specialized $\Mapsto$ judgement, which
% only applies to the matching constructor.
% %
Alternatives not matching the scrutinee could use resources arbitrarily as they
will never be executed. We uniformly treat non-matching alternatives
as if the scrutinee were not in WHNF. Having introduced proof irrelevant
resources, we can now present the full $Case_\textrm{WHNF}$ rule:
\[
\TypeCaseWHNF
\]
We note that it might seem unusual to specialize a rule for expressions in
WHNF, as programs scrutinizing an expression in WHNF are rarely
written by humans.
Yet, our system is designed to be suitable for optimising compilers
in which intermediate programs commonly scrutinize expressions in WHNF.
%
% Crucially, type preservation for the case-reduction substituting the
% case binder and pattern variables in the alternative is not possible
% to prove without branching on WHNF-ness, since otherwise the
% $\D$-substitution is not well-defined.  \todo[inline]{What is the
%   $\Delta$-substitution? Clarify}

\subsubsection{Splitting and tagging fragments}

%% NOTA -- Sempre a repetir a mesma coisa
% Intuitively, in case alternatives whose scrutinee is not in WHNF,
% % (and for scrutinees in WHNF which don't match the case alternative)
% the proof-irrelevant resources introduced by the case expression must be fully
% consumed, either via the case binder $z$, or by using all linear pattern-bound
% variables (for uniformity, we also treat alternatives that do not match a
% scrutinee in WHNF this way).

As opposed to scrutinees in WHNF, where the resources used to type a scrutinee
can be cleanly divided amongst the various sub-expressions of a constructor
application and, henceforth, each pattern variable, there is no direct mapping
between the resources typing a scrutinee not in WHNF and the usage
environments of pattern variables in any alternative.

%
% That is, there is no direct mapping between the usage environments of the
% linear pattern-bound variables and the resources used in the scrutinee.
% \todo{WAIT, there never was!
% Only when the K matches the scrutinee. Maybe we need a separate judgement for
% typing other constructors, which would be standalone and show up in this
% section too?}
%
We introduce \emph{tagged resources} as a means to enforce that all pattern-bound
variables for a scrutinee not in WHNF are either \emph{jointly} used to consume
all resources occurring in the environment, or not at all (instead, the case
binder may be used). Given linear resources $[\D_s]$ used to type a scrutinee,
and a pattern $K~\ov{x_\omega},\ov{y_i}$ with $i$ linear components, we assign
a usage environment $\D_i$ to each linear pattern variable where $\D_i$ is
obtained from the scrutinee environment \emph{tagged} with the constructor name
and linear-variable index $\lctag{\irr{\D_s}\!}{K_i}$. Then, $y_i{:}_{\D_i}\s$ is
introduced in $\G$, just like other $\D$-variables.
\[
  \begin{array}{cc}
    \TypeAltNNotWHNF & \TypeVarSplit
    \end{array}
\]
%
% The tag consists of a constructor name $K$ and an index $i$ identifying the
% position of the pattern variable among all bound variables in that pattern.
%
Having uniquely tagged the resources in the usage environment of each pattern
variable, we need only to express that (1) the pattern-bound $\D$-variables can
be used (i.e., the tagged resources need be available in the linear
environment in order for the $\textsc{Var}_\D$ rule to be applicable) and (2)
if a $K$-pattern-bound $\D$-variable is consumed, all remaining $K$-$\D$-variables,
standing for the remaining linear components of the same pattern, must also be
consumed.

To satisfy these two constraints, we introduce a $\textsc{Split}$ rule that
allows a linear resource $x{:}_1\s$ to be split into $n$ resources at a given
constructor $K$, where $n$ is the number of linear components of the
constructor, and each resource resulting from the split is \emph{tagged} with
$K$ and the positional index of a linear component (i.e. $x{:}_1\s$ can be
split into $\ov{\lctag{x{:}_1\s}{K_i}}^n$).
%
By assigning to each pattern variable a \emph{fragment} of the scrutinee
resources (with a tag), we require the scrutinee resources to be
$\textsc{Split}$ in order to use pattern variables at all.  Moreover, the
remaining tagged resources cannot be used directly, yet need to still be consumed
exactly once. Thus, the choice to consume the tagged resources via the usage
environments of the other pattern variables is forced, ensuring that no
variable for a constructor's linear component can go unused (the tagged environments are disjoint).
%
Tagged resources are inspired by fractional permissions in separation logic~\cite{10.5555/1760267.1760273}.

For instance, in the term $\lambda x.~\ccase{f~x}{z~\{K~a~b\to
  (a,b)\}}$, where $x$ is a linear variable, the case alternative is
typed with $\irr{x}$,
the case binder $z$ is introduced as $\z[\irr{x}]$, and the pattern variables
are introduced as $\var[a][\lctag{\irr{x}}{K_1}]$ and
$\var[b][\lctag{\irr{x}}{K_2}]$, assuming both components of $K$ are linear.
%
The occurrences of $a$ and $b$ are well-typed because we can first
$\textsc{Split}$ $[x{:}_1\s]$ into $\lctag{[x{:}_1\s]}{K_1}, \lctag{[x{:}_1\s]}{K_2}$,
%
% since for any of the scrutinee
% resources to be used by a linear pattern-bound variable usage, the
% resources must be $Split$ for the fragments corresponding to that
% $\D$-var to be consumed, and, consequently, the remaining fragments
% have to be consumed through the other linear pattern-bound variables.
%
noting that $\textsc{Split}$ can be applied both to relevant and proof irrelevant linear
resources in $\D$.

% Using tags for fragments instead of fractions (e.g. $\D_s*i/n$) is necessary
% to guarantee we cannot use the same variable multiple times to consume
% multiple fractions of the resource. It also has the added benefit of allowing
% mixing of pattern variables bound at different alternatives (e.g.~$\lambda
% x~y.~\ccase{(x,y)}{(a,b)\to\ccase{(a,b)}{(z,w)\to(a,w)}}$).

%%%%% \section{Linear Core Examples}
%%%%% 
%%%%% \todo[inline]{If I have no time...}
%%%%% 
%%%%% Linear Mini-Core~\cite{cite:minicore} lists examples of Core programs where
%%%%% semantic linearity must be understood in order for them to be well-typed. In
%%%%% this section, we show those examples in Linear Core ($\lambda^\pi_\Delta$),
%%%%% briefly explaining why they are indeed well-typed.
%%%%% 
%%%%% \paragraph{Equations}
%%%%% 
%%%%% The Linear Haskell function is compiled in Linear Core as\\
%%%%% %
%%%%% \begin{minipage}{0.47\textwidth}
%%%%% \begin{code}
%%%%% data C = Red | Green | Blue
%%%%% f :: C -o C -o C
%%%%% f Red q = q
%%%%% f p Green = p
%%%%% f Blue q = q
%%%%% \end{code}
%%%%% \end{minipage}
%%%%% \begin{minipage}{0.47\textwidth}
%%%%% \[
%%%%% \begin{array}{ll}
%%%%% \lambda p{:}_1C~q{:}_1C.~\ccase{p}{p2{:}_{\{p\}}C \\
%%%%% \{Red \to q \\
%%%%% ; \_ \to \ccase{q}{q2{:}_{\{q\}}C\\
%%%%%   \{Green \to p2\\
%%%%%   ; \_ \to \ccase{p2}{p3{:}_{\{p\}}C \\
%%%%%   \{Blue \to q2\}} \}} \}}
%%%%% \end{array}
%%%%% \]
%%%%% \end{minipage}
%%%%% 
%%%%% \paragraph{Unrestricted Fields}
%%%%% 
%%%%% The following is well-typed:
%%%%% Let $\datatype{K}{K : A \lolli B \to C}$, and $f$:
%%%%% \[
%%%%% \lambda \xl.~\ccase{x}{\var[z][x]~\{ K~a~b \to (z, b) \}}
%%%%% \]
%%%%% 
%%%%% \paragraph{Wildcard}
%%%%% 
%%%%% The following is ill-typed:
%%%%% \begin{code}
%%%%% f = \x -> case x of z { _ -> True }
%%%%% \end{code}
%%%%% 
%%%%% \paragraph{Duplication}
%%%%% 
%%%%% The following is ill-typed:
%%%%% \begin{code}
%%%%% data Foo = Foo A
%%%%% f = \x -> case x of z { Foo a -> (z, a) }
%%%%% \end{code}

% }}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% {{{ Metatheory
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Metatheory of Linear Core\label{sec:main:metatheory}}

We develop the metatheory of Linear Core by first 
presenting the operational semantics of $\lambda^\pi_\D$,
consisting of a lazy natural semantics in the
style of Launchbury~\cite{10.1145/158511.158618}.
This semantics conveys the expected behavior of an implementation of
lazy evaluation and is agnostic to any linearity information.
Such a semantics is ill-suited for reasoning about
linearity and so we develop an \emph{instrumented} linearity-aware semantics that is
enriched with sufficient information to allow us to establish type
safety, thus showing that linearity is preserved. We show that the natural
and instrumented semantics are bisimilar for well-typed terms and so
derive type safety in the natural semantics
(Section~\ref{sec:typesafe}). Finally, we show that multiple optimising
transformations such as those performed by GHC on Linear Core are
type preserving
(Section~\ref{sec:optimisations-preserve-types-meta}), and thus
linearity preserving.

\subsection{Operational Semantics}

The Launchbury-style natural semantics, presented in Figure~\ref{fig:opsem}, captures
standard lazy evaluation while ignoring linearity information.
% The second
% semantics is significantly instrumented with type information and is thus
% \emph{linearity aware}, resulting in a stuck state when a linear
% variable is used more than once. The former is a more direct model of
% an implementation of lazy evaluation, however it is not so convenient
% to reason about formally. The latter semantics is augmented with the additional
% information necessary to strengthen the induction hypothesis to make our
% type safety proofs go through.
% %
% We show that for well-typed terms, the two semantics coincide. 
%
%Both semantics are
The natural semantics is equipped with an evaluation environment that
maps variables to expressions and is
mutated in order to express shared evaluation. Terms are explicit about
their sharing (in the sense of shared reductions in lazy evaluation),
and so we perform a translation step that makes all sharing explicit
through let-binders (see Appendix~\ref{app:opsem}).
We write $\llet{\ov{x_i{:}_{\Delta_i}\sigma = e_i}}{e'}$ for the iterated
let-binding of variables $\ov{x_i}$.

\begin{figure}

  {\small
    \[
  \begin{array}{c}
    \infer[]
    {\,}
    {\Theta : \Lambda p.e \Downarrow \Theta : \Lambda p.e}
    \qquad
    \infer[]
    {\Theta : e\Downarrow \Theta' : \Lambda p.e' \quad \Theta' :
    e'[\pi/p]\Downarrow \Theta'' : v}
    {\Theta : e~\pi\Downarrow \Theta'' : v }\\[1.5em]
    \infer[]
    {\,}
    {\Theta : \lambda x:_\pi \sigma . e \Downarrow \Theta : \lambda x:_\pi \sigma . e }
    \qquad
    \infer[]
    {\Theta : e \Downarrow \Theta' : \lambda y:_\pi \sigma . e' \quad
    \Theta' : e'[x/y] \Downarrow \Theta'' : v }
    {\Theta : e~x \Downarrow \Theta'' : v}\\[1.5em]
    \infer[]
    {(\Theta,  x:_\omega \sigma = e) : e \Downarrow \Theta' : v}
    {(\Theta , x:_\omega \sigma = e) : x \Downarrow (\Theta',x:_\omega
    \sigma = v) : v}
    \qquad
    \infer[]
    {(\Theta,x:_\omega \sigma = e) : e' \Downarrow \Theta' : v}
    {\Theta : \llet{x{:}_{\Delta}\sigma = e}{e'}\Downarrow \Theta' : v
    }\\[1.5em]
    \infer[]
    {\,}
    {\Theta : K~\ov{x_i} \Downarrow \Theta : K~\ov{x_i}}
    \qquad
    \infer[]
    {\Theta : e \Downarrow \Theta' : K~\ov{x_i} \quad
    \Theta' : e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] \Downarrow \Theta'' : v }
    {\Theta :
    \ccase{e}{z{:}_{\Delta'}\sigma~\{{\dots,K~\ov{y_i} \to e', {\dots}}\}}
    \Downarrow \Theta'' : v}\\[1.5em]
    \infer[]
    {\Theta : e \Downarrow \Theta' : K~\ov{x_i} \quad
    \Theta' : e'[K~\ov{x_i}/z] \Downarrow \Theta'' : v }
    {\Theta :
    \ccase{e}{z{:}_{\Delta'}\sigma~\{\dots , \_\to
    e'}\} \Downarrow \Theta'' : v}
    \qquad
    \infer[]
    {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
    % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
    % \Downarrow \Theta_n : v_n \quad
    (\Theta, \overline{x_i:_\omega \sigma_i = e_i}) : e' \Downarrow \Theta'
    : v
    }
    {\Theta : \lletrec{\overline{x_i{:}_{\Delta}\sigma_i = e_i}}{e'}
    \Downarrow \Theta' : v}
    \end{array}
\]}
    \caption{Natural Semantics of $\lambda^\pi_\D$\label{fig:opsem}}
  \end{figure}

  The natural semantics is defined via the relation
$\Theta : e \Downarrow \Theta' : v$ where $e$ is an expression, $v$ a
value, and $\Theta$ and $\Theta'$ are evaluation environments with bindings of
the form $x:_\omega\sigma = e$ which assigns the expression $e$ to the variable
$x$ of the given type $\s$. The rules are standard, augmenting the runtime
environment with let-bound expressions which are evaluated (only once) if the
corresponding let-bound variable is forced.

% The second
% semantics is significantly instrumented with type information and is thus
% \emph{linearity aware}, resulting in a stuck state when a linear
% variable is used more than once. The former is a more direct model of
% an implementation of lazy evaluation, however it is not so convenient
% to reason about formally. The latter semantics is augmented with the additional
% information necessary to strengthen the induction hypothesis to make our
% type safety proofs go through.
% %
% We show that for well-typed terms, the two semantics coincide.

While the natural semantics is a mostly direct model of an implementation
of lazy evaluation, it is inconvenient for formal reasoning about
well-typed linear terms.
%
We follow an approach
similar to that of \cite{cite:linearhaskell}, defining
an \emph{instrumented} operational semantics in which linear variables are
erased from 
the runtime environment once they are forced, ensuring that any term violating linearity
will result in a stuck state. The semantics is also type-aware and
carries with it sufficient data to reconstruct typing derivations for
the purposes of showing type safety.

The instrumented semantics is presented in
Figure~\ref{fig:linopsem}, defining the judgment $\Gamma ; \Delta \vdash (\Theta \mid e) \Downarrow (\Theta'
\mid v) : \sigma , \Sigma$ where $\Gamma$ and $\Delta$ are typing
contexts for expressions in $\Sigma$
($\Gamma$ tracks unrestricted and $\Delta$-bound variables, $\Delta$ tracks linear
assumptions);
$\Theta$ and $\Theta'$ are evaluation environments, consisting of bindings of the form
$x :_\pi \sigma = e$ or $x :_\Delta \sigma = e$; $e$ is the expression being evaluated and $v$
its resulting value, both of type $\sigma$; $\Sigma$ is a list of assignments of the form
$e : \tau$ which are expressions in which
bindings in $\Theta$ may also be used. This additional data is needed to
inductively show that the overall evaluation state is well-typed (and
linearity preserving). The rules mostly mirror those of the natural
semantics, with the exception of the linear variable rule which erases
the binding from the execution environment. Both semantics define
evaluation relations of the form $a \Downarrow b$, where $a$ and $b$
are evaluation states. We also rely on a notion
of partial derivations and partial evaluation (see
Appendix~\ref{app:opsem}), written $a\Downarrow^* b$, which allows us
to state and reason about a progress property.

\begin{figure}
\[
  \begin{array}{c}
    \infer[]
    {\,}
    {\Gamma;\Delta \vdash (\Theta \mid \Lambda p.e) \Downarrow (\Theta \mid
    \Lambda p.e) : \forall p . \sigma , \Sigma }
    \\[1.5em]
    \infer[]
    {\Gamma ; \Delta \vdash (\Theta \mid  e) \Downarrow (\Theta' \mid
    \Lambda p.e') : \forall p . \sigma , \Sigma \quad \Gamma ;\Delta \vdash
      (\Theta' \mid e'[\pi/p]) \Downarrow (\Theta'' \mid v) : \sigma[\pi/p], \Sigma}
    {\Gamma ; \Delta \vdash (\Theta \mid e~\pi) \Downarrow (\Theta''
      \mid v) : \sigma[\pi/p] , \Sigma  }\\[1.5em]
    \infer[]
    {\,}
    {\Gamma ;\Delta \vdash (\Theta \mid \lambda x:_\pi \sigma . e)
    \Downarrow (\Theta \mid \lambda x:_\pi\sigma .e) : \sigma
    \rightarrow_\pi \tau,\Sigma }\\[1.5em]
    \infer[]
    {\Gamma;\Delta \vdash  (\Theta \mid e) \Downarrow (\Theta' \mid
    \lambda y:_1 \sigma . e') : \sigma \rightarrow_1 \tau , x:\sigma, \Sigma
    \quad
    \Gamma ; \Delta \vdash (\Theta' , y:_{1} \sigma = x \mid e') \Downarrow (\Theta''
    \mid v) : \tau , \Sigma  }
    {\Gamma ; \Delta\vdash (\Theta \mid  e~x ) \Downarrow (\Theta''
    \mid v) : \tau , \Sigma }\\[1.5em]
           \infer[]
    {\Gamma;\Delta \vdash  (\Theta \mid e) \Downarrow (\Theta' \mid
    \lambda y:_\omega \sigma . e') : \sigma \rightarrow_\omega \tau , x:\sigma, \Sigma
    \quad
            \Gamma ; \Delta \vdash (\Theta' \mid e'[x/y]) \Downarrow (\Theta''
    \mid v) : \tau , \Sigma  }
    {\Gamma ; \Delta\vdash (\Theta \mid  e~x ) \Downarrow (\Theta''
    \mid v) : \tau , \Sigma }\\[1.5em]

       \infer[]
    {\Gamma ; \Delta  \vdash (\Theta  , x:_{\Delta'} \sigma = e\mid e)
    \Downarrow (\Theta' \mid v) : \sigma , \Sigma \and \D'' = \D' \setminus (\Theta \setminus \Theta')}
    {\Gamma ; \Delta \vdash (\Theta , x:_{\Delta'} \sigma = e \mid x)
      \Downarrow (\Theta' , x:_{\Delta''} \sigma = v \mid v) : \sigma,
    \Sigma }
    \\[1.5em]
           \infer[]
    {\Gamma ;\Delta  \vdash (\Theta \mid e)
    \Downarrow (\Theta' \mid v) : \sigma,\Sigma}
    {\Gamma ; \Delta \vdash (\Theta , x:_1 \sigma = e \mid x)
    \Downarrow (\Theta' \mid v) : \sigma,\Sigma}
    \\[1.5em]

    \infer[]
    {\Gamma ;\Delta \vdash (\Theta,x:_{\Delta'} \sigma = e \mid e')
    \Downarrow (\Theta' \mid v) : \tau,\Sigma}
    {\Gamma ; \Delta \vdash (\Theta \mid \llet{x{:}_{\Delta'}\sigma =
    e}{e'})\Downarrow (\Theta' \mid v) : \tau , \Sigma
    }\\[1.5em]

    \infer[]
    {\,}
    {\Gamma ; \Delta \vdash (\Theta \mid K~\ov{x_i}) \Downarrow
    (\Theta \mid K~\ov{x_i}) : T , \Sigma}
    \\[1.5em]

    \infer[]
    { K : \ov{ \sigma_i
    \rightarrow_\omega \sigma_j \lolli} T
    \\
    \Gamma, z{:}_{\ov{y_j}}T, \ov{y_i :_\omega \sigma_i} ;\Delta , \ov{y_j :_1 \sigma_j} \vdash (\Theta \mid e) \Downarrow (\Theta' \mid
      K~\ov{x_i}) : T , e' : \tau ,\Sigma \\
    \Gamma ;\Delta \vdash (\Theta' \mid e'\ov{[x_i/y_{i,j}]}[K~\ov{x_i}/z]
    ) \Downarrow (\Theta'' \mid v) :\tau , \Sigma }
    {\Gamma ;\Delta \vdash (\Theta \mid
    \ccase{e}{z{:}_{\Delta'}T~\{{\dots,K~\ov{y_{i,j}} \to e', {\dots} }\}})
    \Downarrow (\Theta'' \mid  v) : \tau , \Sigma}\\[1.5em]

    \infer[]
    {K : \ov{ \sigma_i
    \rightarrow_\omega \sigma_j \lolli} T\\
    \Gamma ; \Delta, z{:}_1\s \vdash (\Theta \mid e) \Downarrow (\Theta' \mid
    K~\ov{x_i}) : T , e', \Sigma \quad
    \Gamma ;\Delta \vdash (\Theta' \mid  e'[K~\ov{x_i}/z])
    \Downarrow (\Theta'' \mid v) : \tau,\Sigma }
    {\Gamma ; \Delta \vdash (\Theta \mid 
    \ccase{e}{z{:}_{\Delta'}T~\{\dots , \_\to
    e'}\}) \Downarrow (\Theta'' \mid v) : \tau}
    \\[1.5em]
    \infer[]
    {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
    % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
    % \Downarrow \Theta_n : v_n \quad
    \Gamma ; \Delta \vdash (\Theta, \overline{x_i:_{\Delta'} \sigma_i =
    e_i} \mid e' )\Downarrow (\Theta'
    \mid v) : \tau
    }
    {\Gamma ;\Delta \vdash (\Theta \mid \lletrec{\overline{x_i{:}_{\Delta'}\sigma_i = e_i}}{e'})
    \Downarrow (\Theta' \mid v) : \tau}

   \end{array}
 \]
 \caption{Instrumented Operational Semantics~\label{fig:linopsem}}

\end{figure}

\subsection{Type Safety}\label{sec:typesafe}

Our results rely on a series of lemmas relating linear, unrestricted and
$\Delta$-bound variables, as well as a lemma characterizing irrelevant
resources.
%
% Additionally, we state our assumptions that outline an isomorphism between
% using a linear variable $\xl$ and a $\D$-variable $\xD$ that consumes existing
% resources $\D$, for any $\D$.

\renewcommand{\DeltaLinearRelationLemma}{
\begin{restatable}[$\Delta$-bound to Linear]{lemma}{deltaone}\label{lem:deltaone}
% Any $\Delta$-variable with a fully-irrelevant usage environment can be made a linear variable as long as its irrelevant environment is dropped from the linear environment (and the unrestricted env. is updated accordingly).\\
If $\G,\x[\irr{\D}]; \irr{\D},\D' \vdash e : \vp$ 
then $\G[x/\irr{\D}]; \D',\xl \vdash e :\vp$.
\end{restatable}
}

\renewcommand{\LinearDeltaRelationLemma}{
  \begin{restatable}[Linear to $\Delta$-bound]{lemma}{onedelta}\label{lem:onedelta}
If $\G; \D',\xl \vdash e :\vp$
then $\G[\D/x],\xD; \D,\D' \vdash e : \vp$ ($\Delta$ fresh).
\end{restatable}
}

\renewcommand{\DeltaUnrestrictedRelationLemma}{
\begin{restatable}[Unrestricted and $\Delta$-bound]{lemma}{undelta}\label{lem:undelta}
%An unrestricted variable is equivalent to a $\D$-var with an empty usage environment.\\
$\G,\xo; \D \vdash e : \vp$ iff $\G,\x[\cdot]; \D \vdash e : \vp$
\end{restatable}
}

\paragraph{$\Delta$-bound variables} % Our development relies on a characterization of $\D$-bound
% variables
A well-typed program with a linear variable ($\xl$) is equivalently
well-typed if the linear variable were instead $\D$-bound ($\xD$) with
usage environment $\D$, with $\D$ available in the linear context (and the unrestricted context updated accordingly).

\LinearDeltaRelationLemma

\noindent Dually, any $\Delta$-bound variable with an irrelevant usage
environment can be made into a linear variable as long as its
(irrelevant) environment is dropped from the linear context.

% Dually, a well-typed program with resources $\D$ and $\D$-bound
% variable ($\xD$) is equivalently well-typed, as long as $\D$ is consumed
% through the use of $x$, if $x$ is moved to the linear context, resources $\D$
% are removed from the linear context, and occurrences of $\D$ whole in usage
% environments are substituted by $x$ (occurrences of fragments of $\D$ in usage
% environments are unimportant since $\D$ was consumed whole by $x$, not by any
% of the fragment-using $\D$-vars).

\DeltaLinearRelationLemma

% Intuitively, a linear variable

\noindent Finally, unrestricted and $\D$-bound
variables with an empty usage environment are equivalent.

\DeltaUnrestrictedRelationLemma

\paragraph{Irrelevance}
%
Proof irrelevant resources can only be consumed
indirectly, essentially encoding that the (non-WHNF) case scrutinee
resources must be consumed through the case binder or the linear
pattern-bound variables.
%
As a case expression is evaluated, the scrutinee will eventually
reduce to WHNF,
which must then be typed with rule $Case_{\textrm{WHNF}}$.
%
Crucially, the two case rules must be in harmony in the system, in the sense that
case expressions typed using the $Case_{\textrm{Not WHNF}}$ rule must also be
well-typed by the $Case_{\textrm{WHNF}}$ once the scrutinee is evaluated to WHNF.

% The type preservation theorem states that a well-typed expression
% remains well-typed in the presence of evaluation. Specifically, when the case
% expression whose scrutinee is evaluated to WHNF is handled in the
% preservation proof.

% The \emph{Irrelevance} lemma is required to prove preservation for that
% evaluation case. We need to prove that the alternatives of a case expression
% typed with proof irrelevant resources are still well-typed when the proof
% irrelevant resource is substituted by the scrutinee resources as it is evaluated to WHNF.
% In this sense, the \emph{Irrelevance} lemma witnesses the soundness of typing a
% case alternative with proof irrelevant resources in a certain context with respect to
% typing the same expression with arbitrary resources (we note, however, typing
% an alternative with proof irrelevant resources is not complete wrt using
% arbitrary resources -- a counter example needs only to use a resource
% directly).

\WHNFConvSoundness

The \emph{Irrelevance} lemma witnesses the soundness of typing a
case alternative with proof irrelevant resources with respect to
typing the same expression with arbitrary resources. However, typing
a case alternative with proof irrelevant resources is not complete wrt using
arbitrary resources -- a counter example needs only to use a resource
directly.

% \noindent Intuitively, the lemma holds since proof irrelevant resources can
% only be used through the case binder or pattern-bound variables. If we
% consistently replace the proof irrelevant resources both in the typing
% environment and in the usage environments containing them, the expression
% remains well-typed. We note that the proof irrelevant resources are always
% unique when introduced in such a case alternative (we always take the scrutinee
% environment to make irrelevant, and allow nested ``irrelevantness''), so the
% case binder has the only occurrence of those resources in the $\G$ environment.
% % (being somewhat akin to congruence).
% %
% The proof is given in Section~\ref{sec:proof:irrelevance}.

%\subsection{Type safety\label{sec:type-safety-meta}}
% The $\lambda^\pi_\D$ type system is sound: well-typed programs in Linear Core do not
% get \emph{stuck} and their evaluation preserves linearity.
We prove type safety of the Linear Core system via the standard type
preservation and progress results. As is customary, we make use of multiple
substitution lemmas, one for each kind of variable: unrestricted variables
$\xo$, linear variables $\xl$, and $\D$-bound variables $\xD$. The
development is reported in Appendices~\ref{app:proofs}
and~\ref{app:opsem}, where type safety is established directly for the
instrumented semantics and ported to the natural semantics via a
bisimulation argument.

Note that since the operational semantics gets stuck if a linear
variable is used more than once, progress ensures that linear
functions consume their arguments at most once if their result is
consumed exactly once. Type preservation ensures that evaluation of a
closed term returns an evaluation environment with no linear variables.

  \begin{theorem}[Type Preservation]
    For any well-typed evaluation state $a$, if $a \Downarrow
    b$ or $a \Downarrow^* b$ then $b$ is well-typed.
  \end{theorem}

\begin{theorem}[Progress]
Let $a$ be a well-typed evaluation state. For any partial evaluation $a \Downarrow^* b$ 
the evaluation can be extended.
\end{theorem}

% We start with the auxiliary results, as we will make use of them in a select part of the preservation proof.

% \TypePreservationTheorem
% %
% \ProgressTheorem
% \noindent Type preservation states that a well-typed expression $e$ that
% evaluates to $e'$ remains well-typed under the same context:
%
% Type preservation
% % ($\S$~\ref{sec:proof:type-preservation})
% follows by structural induction on the reductions $e \longrightarrow
% e'$ from the operational semantics. Most cases are straightforward, appealing to one or more
% substitution lemmas (see Appendix~\ref{sec:proof:type-preservation}). The most
% interesting case is that of case expressions whose scrutinee can be further
% evaluated -- we branch on whether the scrutinee becomes in WHNF and appeal to Lemma~\ref{lem:irrev} if so.
%
% This case guarantees that the separation of rules for treating scrutinees is
% consistent, in the sense that a well-typed case expression with a scrutinee not
% in WHNF remains well-typed after the scrutinee is evaluated to WHNF.
%
%
%
%\noindent
% Progress states that the evaluation of a well-typed term does not block:
% Similarly, progress is proved by induction on typing ($\S$~\ref{sec:proof:progress}).

% \subsubsection{Substitution Lemmas}

% The preservation and progress theorems depend on multiple substitution lemmas,
% one for each kind of variable, as is standard.
% % The proofs are given in Section~\ref{sec:proof:substitution-lemmas}.
% % The ... themselves diverge from their common formulation, because
% % $\D$-variables refer to linearly bound variables, so substitution must take into account

% The linear substitution lemma states that a well-typed expression $e$ with a
% linear variable $x$ of type $\s$ remains well-typed if
% occurrences of $x$ in the $e$ are replaced by an expression $e'$ of the same
% type $\s$, and occurrences of $x$ in the linear context and in usage
% environments of $\D$-bound variables are replaced by the linear context $\D'$
% used to type $e'$:

% \LinearSubstitutionLemma

% \noindent Where $\G[\D'/x]$ substitutes all occurrences of $x$ in the usage
% environments of $\D$-variables in $\G$ by the linear variables in $\D'$.
% % ($x$ couldn't appear anywhere besides usage environments of $\D$-bound variables, since $x$ is linear).

% The substitution of the resource in the usage environments is illustrated
% by the following example. Consider the term $\llet{y = use~x}{y}$ where $use$ and $x$ are free variables:
% if we replace occurrences of $x$ by $e'$ (where $\G;\D \vdash e : \s$), then the
% ``real'' usage environment of $y$ goes from $\{x\}$ to $\D$. If we don't update
% the usage environment of $y$ accordingly, we'll ultimately be typing
% $y{:}_{\{x\}}\vp$ with $\D$ instead of $x$, which is not valid.

% The linear substitution lemma extends to case alternatives as well.
% The lemma for substitution of linear variables in case alternatives is similar
% to the linear substitution lemma, applied to the case alternative judgement.
% % Is slightly more involved, in the sense that there are more environments in which the
% % substitution $[\D/x]$ must be applied (for the same reason):
% %
% \LinearSubstitutionAltsLemma
% %
% \noindent We further require that the environment annotated in the case
% alternative judgement, $\D_s$, is a subset of the environment used to type the
% whole alternative $\D_s \subseteq \D$. In all occurrences of the alternative
% judgement (in $Case_{\textrm{WHNF}}$ and $Case_{\textrm{Not WHNF}}$), the
% environment annotating the alternative judgement is \emph{always} a subset of
% the alternative environment.

% The substitution lemma for unrestricted variables follows the usual
% formulation, with the added restriction (common to linear type systems) that
% the expression $e'$ that is going to substitute the unrestricted variable $x$
% is typed on an empty linear environment:
% %
% \UnrestrictedSubstitutionLemma
% %
% \noindent Similarly, we also prove the substitution of unrestricted variables preserves types on an alternative case expression:
% %
% \UnrestrictedSubstitutionAltsLemma

% Finally, we introduce the lemma stating that substitution of $\D$-bound
% variables by expressions of the same type preserves the type of the original
% expression.
% %
% What distinguishes this lemma from traditional substitution lemmas is that the
% usage environment $\D$ of the variable $x$ being substituted by expression $e'$
% must match exactly the typing environment $\D$ of $e'$ and the
% environment of the original expression doesn't change with the substitution:
% %
% \DeltaSubstitutionLemma
% %
% \noindent Intuitively, if $x$ is well-typed with $\D$ in $e$, substituting $x$
% by an expression $e'$ which is typed in the same environment $\D$ allows the
% distribution of resources $\D,\D'$ used to type $e$ across sub-derivations to remain
% unchanged. To prove the theorems, we don't need a ``stronger'' substitution of
% $\D$-vars lemma (allowing arbitrary resources $\D''$ to type $e'$, as in other
% substitution lemmas), as we only ever substitute $\D$-variables by expressions
% whose typing environment matches the variables usage environment. However, it
% is not obvious whether such a lemma is possible to prove for $\D$-variables
% (e.g. let $\G;\D \vdash e :\s$ and $\G; \D' \vdash \llet{x = e'}{x}$, if we
% substitute $e$ for $x$ the resources $\D'$ are no longer consumed).

% The $\D$-substitution lemma on case alternatives reflects again that the typing
% environment of the expression substitution the variable must match its usage
% environment. We recall that $\D_s \subseteq \D,\D'$ states that the annotated
% environment is always contained in the typing environment, which is true of all
% occurrences of this judgement. An alternative formulation of this lemma could
% instead explicitly list $\D_s$ as part of the typing environment for the same
% effect:

% \DeltaSubstitutionAltsLemma

% The proofs for substitution lemmas of linear, unrestricted, and
% $\D$-variables are available in Section~\ref{sec:proof:substitution-lemmas}.

%TODO! Substitution of proof-irrelevant linear variables preserves typing. The
%term always remains the same because $x$ cannot occur in any term, however, all
%variables that refer to $x$ in their usage environment must now refer the usage env. of the substitee (e.g. $[x] => [\D]$).
%This seems trivial to see correct, since all occurrences are in environments, so we get some equivalence similar to the one we need for the proof of Alt0.
%
%TODO: Multiplicity substitution preserves typing lemma
%
%TODO: Canonical forms lemma
%
%TODO: Corollary of $\Delta$-var subst. for $\ov{\Delta}$
%
%TODO: Constructor app typing:
%If $\Gamma, \Delta \vdash K~\ov{e}$ and $K{:}\ov{\sigma\to\pi}~T~\ov{p} \in \Gamma$ and $\hasnolinearvars{\Gamma}$
%then $\ov{\Gamma, \Delta_i \vdash e_i : \sigma_i}$

\subsection{Optimisations preserve linearity\label{sec:optimisations-preserve-types-meta}}
%\todo[inline]{Check Commuting case-let in Fig.~\ref{fig:opttrans}!}
A main goal of the Linear Core type system is 
to serve as a typed intermediate representation for optimising
compilers of lazy languages with linearity. In light of this,
we show that multiple optimising transformations are type
preserving in Linear Core. %(i.e.,~preserve linearity semantically).

% The optimising transformations proved sound wrt Linear Core in this section
% have been previously explained and motivated in
% Section~\ref{sec:core-to-core-transformations}.
% %
%\{\overline{\rho\to e'}\}
\begin{figure}[t]

  {\small
  \[
    \begin{array}{ll}
   (\llet{\xD = e}{e'} )  \Longrightarrow \llet{\xD =       e}{e'[e/x]} &                    \mbox{Inlining}\\
       (\lambda \x[\pi][\s].~e)~e'  \Longrightarrow e[e'/x]  &                \mbox{$\beta$-reduction}\\
      (\ccase{(K~\ov{e})}{\z[\D][\s]~\{\overline{\rho_j\to e_j}, K~\ov{x} \to e_i, \overline{\rho_k\to e_k}\}} )\Longrightarrow  e_i\ov{[e/x]}[K~\ov{e}/z] &
                                  \mbox{Case of known constr.}\\
      (\ccase{(\ccase{e_c}{\z[\D]~\{\ov{\rho_{c_i}\to e_{c_i}}\}})}{\var[w][\irr{\D,\D'}][\s']~\{\ov{\rho_i\to e_i}\}})\\
            \Longrightarrow 
            %{{{
      \ccase{e_c}{\z[\D]~\{\ov{\rho_{c_i} \to \ccase{e_{c_i}}{w~\{\ov{\rho_i\to e_i}\}}}\}}
      & \mbox{Case-of-case}\\
      (\lambda \y[\pi].~\llet{\xD = e}{e'})
      \Longrightarrow
      \llet{\xD=e}{\lambda\y[\pi].~e'} & \mbox{Commuting let-$\lambda$}\\
      ((\llet{v = e}{b})~a) \Longrightarrow \llet{v = e}{b~a} &\mbox{Commuting let-app.}\\
      (\ccase{(\llet{v = e}{b})}{\{\overline{\rho\to e'}\}}) \Longrightarrow
      %{{{
      \llet{v =  e}{\ccase{b}{\{\overline{\rho\to e'}\}}}
                               &
                                 \mbox{Commuting let-case}\\
      (\ccase{e}{\{\overline{\rho_j\to e_j}, \rho \to E[e_1],
      \overline{\rho_k\to e_k}\}} ) \qquad (\mbox{$x$ fresh}) \\
      %{{{
     \Longrightarrow \llet{x = e_1}{\ccase{e}{\{\overline{\rho_j\to e_j}, \rho \to E[x], \overline{\rho_k\to e_k}\}}} 
                                                                        & \mbox{Commuting case-let}\\
      (\llet{x = (\llet{v = e}{b})}{c}) \Longrightarrow \llet{v = e}{\llet{x = b}{c}}
                                                                      & \mbox{Commuting let-let}\\
      f \Longrightarrow \lambda x.(f~x) & \mbox{$\eta$-expansion}\\
      \lambda x.(f~x) \Longrightarrow f & \mbox{$\eta$-reduction}\\
      (\ccase{x}{z~\{\ov{\rho_i\to e_i}\}}) \Longrightarrow \ccase{x}{z~\{\ov{\rho_i\to  e_i[z/x]}\}}
                                                                      & \mbox{Binder swap}\\
      % (\ccase{x}{z~\{\ov{\rho_i\to e_i}\}} )\Longrightarrow \ccase{x}{z~\{\ov{\rho_i\to e_i[x/z]}\}}
      %                                                                 & \mbox{Reverse binder swap}
    \end{array}
  \]}
  \caption{Optimising transformations validated by Linear
    Core\label{fig:opttrans} (for commuting case-let we
      require $e_1$ not to capture the case binder, variables bound by the
      pattern $\rho$ and the context $E[{-}]$ -- see Appendix~\ref{app:optimisations} for
    details).} 
\end{figure}

We describe Core-to-Core transformations as $e_1 \Longrightarrow e_2$,
where $e_1$ is an arbitrary \emph{well-typed} expression
matching a certain shape (or of a certain type), which is
transformed into expression $e_2$. Validating a transformation
entails showing that $e_2$ is well-typed. We often annotate the arrow
with the name of the transformation when describing chains of such
transformations.

The list of transformations we have
validated is given in Figure~\ref{fig:opttrans}.
%
The behaviour of most transformations can be inferred from its name:
Inlining substitutes a let-bound expression in the let body; $\beta$-reduction
and case of known constructor effectively perform evaluation of the given expression;
case-of-case commutes cases in scrutinee position inside the outer case alternatives;
%let floating and  for let-bound expressions to be floated out of $\lambda$-abstractions;
the five commuting conversions allow let-bound expressions to be commuted with other
Core constructs; $\eta$-expansion and reduction apply the $\eta$-laws to terms of function type;
finally, binder swap substitutes occurrences of a scrutinee variable for the
case binder. While many of the transformations above are not obviously optimising,
they are chained in the compiler pipeline to expose optimisation opportunities.

% Transformations are described by an arbitrary well-typed expression with a certain shape, on
% the left hand side (lhs) of the arrow $\Longrightarrow$, resulting in an expression on
% the right hand side (rhs) that we prove to be well-typed.
% %
% % For our proofs, we assume the lhs to be a well-typed expression and prove the
% % rhs is well-typed as well.
% %
% For each transformation, we describe the intuition behind the transformation
% preserving linearity in our system.

% \todo[inline]{Aqui uma lista itemizada das transformações todas. As
%   provas todas para apendice.}

%%% COMENTADO AQUI
% \subsubsection{Inlining}

% % To the best of our knowledge, there is no linear type system for which inlining
% % preserves linearity\footnote{https://github.com/ghc-proposals/ghc-proposals/blob/master/proposals/0111-linear-types.rst\#id90}

% \input{language-v4/proofs/optimizations/Inlining}

% \subsubsection{\texorpdfstring{$\beta$}{Beta}-reduction}

% \input{language-v4/proofs/optimizations/BetaReduction}

% \subsubsection{Case of known constructor}

% \input{language-v4/proofs/optimizations/CaseOfKnownConstructor}

% \subsubsection{Let floating}

% \input{language-v4/proofs/optimizations/LetFloating}

% \subsubsection{\texorpdfstring{$\eta$}{Eta}-conversions}

% \input{language-v4/proofs/optimizations/EtaConvs}

% \subsubsection{Binder Swap}

% The binder swap transformation applies to case expressions whose scrutinee is a
% single variable $x$, and it substitutes occurrences of $x$ in the case
% alternatives for the case binder $z$. If $x$ is a linear resource, $x$ cannot
% occur in the case alternatives (as we conservatively consider variables are not
% in WHNF), so the substitution preserves types vacuously. Otherwise, $x$ can be
% freely substituted by $z$, since $z$ is also an unrestricted resource (it's
% usage environment is empty because $x$ is unrestricted).

% \input{language-v4/proofs/optimizations/BinderSwap}

\paragraph{Reverse Binder Swap\label{sec:reverse-binder-swap-considered-harmful}}

%Finally, we discuss the special case of scrutinizing a
%single variable $x$, e.g.:
%$
%\lambda x.~\ccase{x}{\_ \to x}
%$. 
%It might seem as though this function is linear:
%%begin{itemize}
%% \item
%if the argument instantiating $x$ is in WHNF, then scrutinizing it is a no-op, and returning $x$
%just returns the resource intact;
%% \item
%if the argument instantiating $x$ is not in WHNF, then scrutinizing it evaluates it to WHNF, and returning
%$x$ returns the result of evaluating $x$ that still had to be consumed exactly
%once.
%%\end{itemize}
%However, in practice, this program's linearity can only be determined by
%further analysis on the non-strict evaluation strategy. If linear
%function applications are $\beta$-reduced \emph{call-by-name}
%(often desirable, as linear functions are known to use their argument exactly
%once)
%% and thus call-by-name reduction will definitely not duplicate computation
%while the example function is considered linear, then such an application might
%duplicate linear resources during evaluation. For instance:
%\[
%\begin{array}{l}
%(\lambda x.~\ccase{x}{\_ \to x})~(free~x)
%\Longrightarrow_{CBN}
%\ccase{free~x}{\_ \to free~x}
%\end{array}
%\]
%On the other hand, if linear function applications are reduced
%\emph{call-by-need}, the above example would instead be semantically linear
%because the result of freeing $x$ would be bound to some $y$ first.
%%
%The system we present in the next section does accept this function
%scrutinizing a variable as linear, as our operational semantics are
%call-by-need.
%This interaction between evaluation and linearity is considered specifically in the
%reverse binder swap transformation discussed in
%Section~\ref{sec:optimisations-preserve-types-meta}.\todo{Figure out precisely reverse binder swap proof and reevaluate}
%

Reverse binder swap is dual to the binder swap transformation, substituting
occurrences of the case binder $z$ by the scrutinee in the special case when
the scrutinee is a variable $x$.
%
% his is dual to what the binder swap transformation
% does in hope of eliminating multiple uses of $x$ (allowing for inlining). 
By using the scrutinee $x$ instead of the case binder, we might be able to
float out expressions from the alternative using the case binder:
\[
    \ccase{x}{z~\{\rho \rightarrow e\}} \Rightarrow \ccase{x}{z~\{\rho \rightarrow e[x/z]\}}
\]

%For example, we might float an expensive computation involving $z$ out of the
%case alternative, where $z$ is now out of scope:
%
%{\small\[
%\begin{array}{l}
%(\lambda x.~\lletrec{go~y = \ccase{x}{z~\{(a,b) \to \ E[expensive~z] \} }}{\dots})
%\Longrightarrow_\textrm{Reverse binder swap}\\
%(\lambda x.~\lletrec{go~y = \ccase{x}{z~\{(a,b) \to E[expensive~x] \} }}{\dots})
%\Longrightarrow_\textrm{Commuting case-let}\\
%\lambda x.~\llet{t = \mathit{expensive}~x}{(\lletrec{go~y = \ccase{x}{z~\{(a,b) \to E[t] \} }}{\dots})}\\
%\end{array}
%\]}
%
%\noindent If $go$ is compiled into a loop, we only compute
%$\mathit{expensive}~x$ once instead of in every loop iteration.
%
Even though GHC applies the reverse binder swap transformation in the
Core-to-Core passes, this optimisation violates linearity in Linear Core
because the transformed program is rejected -- the scrutinee $x$, a single
variable, is not considered to be in WHNF by our system, thus occurrences of
$x$ in any alternative are ill-typed in it. This is a conservative choice in $\lambda^\pi_\Delta$ for
typing the subtle case where the scrutinee is a single variable.

Consider $\ccase{x}{\_ \to x}$. Perhaps surprisingly, this expression can be
considered semantically linear under call-by-need but violates
linearity under call-by-name:
%
%An expression in which a variable $x$ occurs both in the case scrutinee and in
%the alternatives (e.g.,~$\ccase{x}{\_ \to x}$) could be considered linear if
%considered in isolation:
% \[
% \G; \xl \vdash \ccase{x}{\_ \to x} : \s
% \]
% The reasoning is done by branching on whether $x$ refers to an expression in
% WHNF or an unevaluated thunk.
% \begin{itemize}
% \item
Under call-by-need, if $x$ refers to an unevaluated expression, scrutinizing it forces the expression
bound by $x$ to WHNF. In a subsequent use of
$x$ in an alternative, $x$ refers to the already evaluated scrutinee in WHNF.
Since $x$ is just another name (like the case binder) for the scrutinee in
WHNF, its use is also valid in the alternatives.
%
Dually, if $x$ refers to an expression already in WHNF, no evaluation takes
place and $x$ is still just another name for the scrutinee in WHNF.

However, under call-by-name, evaluation of $x$ is not shared:
%However, allowing this expression to be linear makes $\lambda^\pi_\Delta$ unsound!
%Consider the function application $(\lambda x.~\ccase{x}{\_ \to x})~(use~y)$
%where $y$ is a free linear variable, and assume $\lambda x.~\ccase{x}{\_ \to x}$
%to be a linear function. According to $\lambda^\pi_\Delta$'s call-by-name
%$\beta$-reduction, we have:
\[
\begin{array}{l}
(\lambda x.~\ccase{x}{\_ \to x})~(use~y)
\Longrightarrow_\textrm{CBName}
\ccase{use~y}{\_ \to use~y}
\end{array}
\]
After the reduction, $y$ is a linear variable consumed both in the scrutinee
and in the case alternative -- $y$ has been duplicated!
%
If the application is instead reduced call-by-need, $y$ is not duplicated:
\[
\begin{array}{l}
(\lambda x.~\ccase{x}{\_ \to x})~(use~y)
\Longrightarrow_\textrm{CBNeed}
\llet{x = use~y}{\ccase{x}{\_ \to x}}
\end{array}
\]
%then the computation using $y$ would be let-bound, and $y$ would be used as a
%scrutinee variable, which preserves semantic linearity.

In an optimising compiler such as GHC, there is no definitive evaluation
strategy. It is mostly up to heuristics to determine what to inline
\emph{à la} call-by-name and what to create let-bindings for \emph{à
la} call-by-need.
%
In this flexible evaluation setting, we highlight the reverse binder swap is
only a linearity-preserving transformation as long as a linear variable with
more than one occurrence is never substituted for an expression with free
linear variables.
%
%This is unfortunately at odds with
%Moreover, allowing the reverse binder swap also challenges the 
%
%It additionally challenges the non-heuristic property that linear variables can
%be unconditionally inlined.......

%the aggressive inlining passes blur evaluation
%there is no definitive evaluation
%strategy in the sense functional compilers use both when agressively inline
%expressions \emph{à la} call-by-name.
%call-by-need for
%every function application -- our result shows that failing to do so
%when calling a function such as the one above is \emph{unsound}.
%
%In an optimising compiler, a binding may be deemed unnecessary if the function
%argument is inferred to be used exactly once. Linearity can play a
%(non-heuristic) role in this inference, for a linear function uses its
%argument exactly once by definition.
%
% Since the compiler may leverage this information (we
% briefly discuss \emph{linearity-influenced optimisations} in
% Section~\ref{sec:discussion}).
%
%However, it is incompatible to both type expressions like $(\lambda
%x.~\ccase{x}{\_ \to x})$ as linear functions and consider
%call-by-name $\beta$-reduction of linear functions unconditionally
%safe (without risk of duplicating resources nor work).
%
%Furthermore, considering such an use of $x$ to be linear also makes inlining of
%linear let bindings no longer unconditionally safe.

% $\beta$-reduction reduces an application of a linear function using
% call-by-name -- If we know the argument is used exactly once, a binding to
% share the result of computing the argument is unnecessary, so we instead
% substitute the argument expression for the linearly-bound variable in the
% $\lambda$-body directly.
%%
%% Essentially, Linear Core would be unsound, and even duplicate resources, if the
%% above kind of expressions, where linear variable scrutinees occur in the
%% alternatives body, were well-typed, because of its interaction with the
%% call-by-name reduction of linear functions.
%In this sense, the reverse binder
%swap is an optimisation that creates ill-typed expressions from well-typed
%ones, so it is deemed an \emph{invalid} optimisation that does not preserve types in
%our system.

%% Reverse binder swap is not a problem in the GHC simplifier because of the
%% weaker notion of linearity understood by occurrence analysis. Occurrence
%% analysis is a static analysis pass which can be used to determine whether a
%% lambda application can be $\beta$-reduced call-by-name, and $\ccase{x}{\_ \to
%% x}$ is \emph{not} seen as using $x$ linearly by the analysis. Thus,
%% $\beta$-reduction is done with call-by-need on such an expression.
%
% In practice, the optimisation preserves linearity when applied as part of the
% GHC pipeline due to the occurrence analyser being naive with regard to semantic
% linearity.

% Concluding, in being able to understand more programs as linear, our type
% system allows more expressions to be considered linear for $\beta$-reduction
% without a let-allocation, however, it makes reverse binder swap an invalid
% transformation since its output, when considered linear, might violate
% linearity when further optimised.

% (Link to ticket)
% 
% \begin{tabbing}
% (1) All optimisations preserve (semantic) linearity\\
% (2) If a function is (semantically) linear, then we can evaluate it using call-by-name and preserve linearity\\
% (3) Reverse binder swap is an optimisation\\
% (4) If reverse binder swap is applied to a case scrutinizing a linear resource in the body (`e`) of a linear function `f`, then the function is still linear by (1)\\
% (5) If we evaluate `f`, we do it call-by-name because of (2)\\
% (6) Call-by-name substitution of the linear argument in the body of a function has been reverse binder swapped doesn’t preserve linearity\\
% (7) Contradiction: By 3 and 1, `f` is linear after reverse-binder-swap. By 2, we can substitute arguments to `f` call-by-name and preserve linearity, which contradicts with 6 that says call-by-name substitution after reverse binder swap violates linearity\\
% \end{tabbing}

% Conclusion:
% Either we need to forfeit that we can always substitute call-by-name linear
% function applications, or we forfeit that reverse binder swap preserves
% linearity (instead, it preserves a weaker notion of linearity understood by the
% syntatic-occurrence-analyzer)

% \todo[inline]{Reverse-binder-swap is only well-defined in certain scenarios
% where the optimizations don't apply call-by-name beta-reduction after the
% reverse-binder-swap optimization -- otherwise we would duplicate resources.
% In this case, it is not a matter of syntatic vs semantic linearity
% }

% \todo[inline]{
% Mention, from ``Call-by-name, call-by-value, call-by-need and the linear lambda calculus'':
% The call-by-name calculus is not entirely suitable for reasoning about
% functional programs in lazy languages, because the beta rule may copy the
% argument of a function any number of times. The call-by-need calculus uses a
% diferent notion of reduction, observationally equivalent to the call-by-name
% calculus. But call-by-need, like call-by-value, guarantees that the argument
% to a function is not copied before it is reduced to a value.
% }

% It's also interesting to note that reverse-binder-swap preserves linearity under pure call-by-need but not under call-by-name, because
% (In the sense that if EVEN linear functions reduce call-by-need rather than call-by-name, then it would preserve optimisations)

% Reduce call-by-name linear function application
% \begin{code}
% f y = (\x. case x of _ -> x:1) (h y)
% ===>
% f y = (case x of _ -> x)[h y/x]
% ===>
% f y = (case h y of _ -> h y) -- consume y twice.
% \end{code}
% 
% Vs. call-by-need
% \begin{code}
%  (\y = (\x. case x of _ -> x:1) (h y)) e
% ===>
%     let y = e in
%        let x = h y in
%         case x of _ -> x
% ===>
%        let x = h e_v in
%         case x of _ -> x
% ===>
%         case x_v of _ -> x_v
% \end{code}

% \subsubsection{Case of Case}

% The case of case transformation applies to case expressions whose scrutinee is
% another case expression, and returns the innermost case expression transformed by repeating
% the outermost case expression in each alternative of the innermost case,
% scrutinizing the original alternative body.

% Intuitively, since the scrutinee of the outermost case is not in WHNF, no
% resources from it can directly occur in the outermost alternatives. By moving
% the outermost alternatives inwards with a different scrutinee, the alternatives
% remain well-typed because they are typed using either the case binder or the
% pattern bound variables, which, by the \emph{Irrelevance} lemma, makes it
% well-typed for any scrutinee consuming arbitrary resources. The proof is given in Section~\ref{sec:proof:caseofcase}.

% \CaseOfCaseTheorem

% }}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% {{{ Linear Core as a GHC Plugin
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Linear Core as a GHC Plugin\label{sec:discuss:implementation}}

We have implemented the Linear Core type system as a plugin for the Glasgow
Haskell Compiler.
GHC plugins allow developers to inspect and modify programs being compiled by
GHC, at different stages of compilation~\cite{10.1145/3331545.3342599}.
%
In particular, for any given Haskell module, Core plugins run for (and receive
as input) every intermediate program produced in the compilation process, i.e.
from desugaring and after each optimising transformation.
%
Our implementation further substantiates our claim that Linear Core is
suitable for the intermediate language of an optimising compiler.
%\todo[inline]{Tem de ser anonimizado, provavelmente alterando a bib
  %entry para omitir autores e o proprio link}
The GHC Linear Core plugin~\cite{cite:linear-core-plugin} implements a
typechecker for Linear Core: given a Core program, our plugin typechecks the
linearity of all expressions bound in that program, failing if a linear
resource is not used \emph{exactly once} according to semantic linearity of the
$(\lambda^\pi_\Delta)$ system.
%
% Furthermore, the implementation accepts all example programs from Chapter~\ref{sec:linearity-semantically}
% deemed well-typed by Linear Core, which are marked with a
% \colorbox{notyet}{\notyetcolorname} background.

% Additionally, we discuss the implementation of the Linear Core type system
% directly in the Glasgow Haskell Compiler.

% This section discusses the implementation of Linear Core as a GHC Plugin, with
% a dash of painful history in the attempt of implementing Linear Core directly
% into GHC.

The implementation of Linear Core as a typechecker does not follow
directly from the description of the type system because Linear Core
is not \emph{syntax-directed}. Specifically, the most challenging
features are splitting linear resources amongst sub-derivations and
consuming fragments of resources through pattern-bound variables.
Instead of non-deterministically splitting linear resources amongst
sub-derivations, we thread input/output linear resources through each
derivation~\cite{DBLP:journals/tcs/CervesatoHP00}.
%
For a scrutinee not in WHNF, pattern variables in a case alternative are
introduced as $\D$-variables with usage environment $\lctag{\irr{\D}}{K_i}$,
where $\D$ are the scrutinee resources and $K_i$ the tag of that pattern
variable. To use these pattern variables, the scrutinee resources must be first
$\textsc{Split}$ into fragments. Instead of guessing which resources need to be split,
the implementation consumes tagged fragments of a resource \emph{as needed},
i.e., only when consuming a resource whose usage environment has a resource
tagged $K_j$ do we $\textsc{Split}$ that resource on $K$ and consume the $K_j$ fragment.
%
Finally, our implementation must infer the usage environments of binders in
a recursive let group before using them to typecheck the let body. We use a
naive $O(n^2)$ algorithm (where $n$ is the number of let bindings) to determine
these usage environments. %Further research regarding
%this inference challenge is discussed in Section~\ref{sec:future-work}.
%
%
%
% We use the following techniques to tackle the \emph{non-syntax-directedness} of
% Linear Core:
%
% \begin{itemize}

% % \item Instead of non-deterministically splitting linear resources amongst
% % sub-derivations, we thread input/output linear resources through each
% % derivation using the resource management for linear logic described
% % by~\cite{DBLP:journals/tcs/CervesatoHP00}.
% % \todo{Escrever uma regra?}

% % In a case expression whose scrutinee is not in WHNF,
% \item Pattern variables bound in a case alternative, for a scrutinee not in WHNF,
% are introduced as $\D$-variables with usage environment $\lctag{\irr{\D}}{K_i}$,
% where $\D$ are the scrutinee resources and $K_i$ the tag of that pattern
% variable. To use the resources through the pattern-bound $\D$-vars, they must
% be first $Split$ into fragments.

% We consume tagged fragments of a resource \emph{as needed}, i.e., when a resource,
% whose usage environment has a fragmented resource with tag $K_j$, is used, we $Split$
% the matching resource according to the constructor $K$ and consume the fragment
% $K_j$, rather than eagerly determining which resources need to be fragmented to
% do so.
% %
% We note that it is safe to destructively fragment the resource, i.e. removing
% the whole resource and only leaving the split fragments, because resources are
% only $Split$ when a fragment is needed and, consequently, if a fragment is
% consumed, using the ``whole'' resource as well violates linearity.

% \end{itemize}
%

% The usage environments of a recursive group of binders (that is not necessarily
% strongly-connected) is computed in two separate passes.
% %
% First, we calculate a \emph{naive environment} by recording the linear
% resources semantically used in the binder bodies, while counting \emph{uses}
% (not syntactic occurrences) of the mutually recursive let variables.
% %
% Second, the binder names and corresponding \emph{naive environment} (mapping
% each linear resource to $1$ and the recursive variables to the $n$ number of
% times they are \emph{used}) in Cartesian pairs are given as input to
% % Algorithm~\ref{computeRecUsages},
% an algorithm which computes the actual usage environment
% of each binder.
%
% TODO: I should probably use the re-computed usageEnvs instead of the naiveUsageEnvs.
% I'm pretty sure it might fail in some inputs if I keep using the naiveUsageEnvs.
% \begin{algorithm}
% $usageEnvs \gets naiveUsageEnvs.map(fst)$\;
% \For{$(bind, U) \in naiveUsageEnvs$}{
%     \For{$V \in usageEnvs$}{
%         $V \gets sup(V[bind]*U\setminus\{bind\}, V\setminus\{bind\})$
%     }
% }
% \caption{computeRecUsages\label{computeRecUsages}}
% \end{algorithm}
%
% Intuitively, the algorithm, for each recursive binder, iterates over the
% (initially naive) usage environments and substitutes occurrences of the
% recursive binders by their corresponding usage environment,
% scaled up by the amount of times that recursive binder is used in the
% environment being updated. The result is the least upper bound of linear
% resources used by each strongly-connected group of mutually recursive bindings
% (for each binder of such a group), since all occurrences of the recursive
% binders in the usage environments will be substituted away by the corresponding
% recursive usage environments.
%
% The high-level description is:
% \begin{itemize}
%     \item Given a list of binders and their naive environment ($(f, F), (g, G),
%     (h, H)$) in which each use of $f, g, h$ is mapped to a count $n$ in $F, G, H$
%
%     \item For each pair of a letrec-bound variable and corresponding usage
%     environment, update all bindings and their usage as described in the
%     algorithm
%
%     \item After iterating through all bound rec vars, all usage environments
%         will be free of recursive bind usages, and hence describe the ``final'' usage environment
% \end{itemize}
%
% The complexity of computing a usage environment for $n$ recursive let binders
% is quadratic in $n$,
% % i.e. the algorithm has $O(n^2)$ complexity,
% but this is not an issue since it is uncommon to have more than a handful of
% binders in the same recursive let block.

% TOdo: INCLUDE THIS??
% The plugin successfully validates linearity throughout the
% (optimising) compilation of linearity-heavy libraries, namely
% \texttt{linear-base} and \texttt{linear-smc}, except in expressions whose
% linearity depends on so-called \emph{multiplicity coercions}, which are an
% avenue of future work (Section~\ref{sec:future-work}) exploring the intersection
% of Linear Core with type equality coercions.

The results of running the Linear Core GHC plugin on large 
libraries focused around linear types are reported in
Figure~\ref{fig:core-plugin-res}.
%
%
We compiled the Haskell standard library for programming with linear types,
\texttt{linear-base} (4000 lines, comprising over 100 modules), the
\texttt{linear-smc}~\cite{10.1145/3471874.3472980} (1500 lines) library, and
\texttt{priority-sesh}~\cite{10.1145/3471874.3472979} (1400 lines), a
session-types library, using our plugin.
%
We count the number of programs accepted by our implementation, where
each top-level binding in a module counts as a program, and every such
binding is typechecked once per optimisation pass (i.e., we typecheck
all intermediate programs produced by GHC). The total amount of programs
rejected by the typechecker are counted dually, also across every optimisation
pass, without halting compilation when a program is rejected. We note that our
implementation is not performance conscious, and types every intermediate
program from scratch, instead of preserving information throughout the
pipeline.

%``Total Rejected'' column, but we distil these rejections into ``Unique
%Rejections'' by removing duplicate rejections (those of the same
%program and for the same linearity-violating reason, but occurring at
%different stages). We further categorize the unique rejections into
%``Linear modulo Call-by-name'', ``Linear Rejected'', ``Not Linear
%Rejected'', and ``Unknown Rejected''.  We note that our implementation
%is not performance conscious. For instance, \texttt{linear-base} takes
%35 seconds to compile with our plugin, instead of 20 seconds.
%%
%% Using the plugin, \texttt{linear-base} takes 35 seconds to compile, instead of
%% 20 seconds. We note, however, that the implementation is not performance
%% conscious, and types every intermediate program from scratch, instead of
%% maintaining linearity information throughout the pipeline.

%Programs ``linear modulo call-by-name'' are a class of programs
%which scrutinize a variable, but then use the variable in the case
%alternatives of the case scrutinizing it. As discussed in
%Section~\ref{sec:reverse-binder-swap-considered-harmful}, these programs can be
%understood as linear as long as applications of linear functions binding these
%variables are \emph{not} reduced \emph{call-by-name}.
%In Core, these programs are not seen as
%linear, and thus GHC does not not reduce them using a call-by-name evaluation strategy.
%% We have a flag in our implementation to accept some of these programs...
%%
%``Linear Rejected'' programs include different kinds of programs which are
%rejected by the Linear Core system, but are still semantically linear. An
%example is a program to which a rewrite rule was applied, resulting in an
%application of unrestricted function to a linear resource.
%% in the first argument
%% to \incode{build}:
%% \begin{lstlisting}
%% take :: Int -> Replicator a *' $\lolli$ '* [a]
%% take = \ (ds :: Int) (r :: Replicator b) ->
%%      case ds of
%%        1 -> build (\(c :: b -> a -> a) (n :: a) -> c (extract r) n)
%%        ...
%% \end{lstlisting}
%% We know the linear resource is still used linearly because \incode{build} will
%% always instance the unrestricted function to $(:)$ (read \emph{Cons}), which is
%% linear. Additionally, these include programs scrutinizing a value of a type of
%% which all constructors have no linear components, but only matching on the
%% default alternative, programs where common-sub-expression elimination
%% substituted more than one occurrence of \incode{Ur x} by \incode{y}, in the body of
%% scrutinizing \incode{y} (a linear variable), and other programs affected by rewrite rules.
%%
%``Not Linear Rejected'' indicate programs that we do not understand as linear,
%semantically, and are simultaneously rejected by Linear Core and its
%implementation. % These programs were not necessarily identified as bad
%% by the linter.\footnote{rever}
%% An example, where the last component of \incode{HashMap}, \incode{wwz},
%% is linear, but is not being consumed in the case alternative matching on it:
%% \begin{lstlisting}
%% jssvi :: Ur Bool *' $\lolli$ '* Set Int *' $\lolli$ '* Ur (TestT IO ())
%% jssvi (a :: Ur Bool) (b :: Set Int)
%%   = case a of
%%      Ur ss -> case b of { HashMap wwx wwy wwz -> jump wjssAd ss }
%% \end{lstlisting}
%%
%Finally, ``Unknown Rejected'' programs are those whose validity we did not
%check. These include both programs accepted by Linear Core, but not by its
%implementation, and programs that are simply rejected by Linear Core, but which
%were not categorized.

{\small
  \input{prototype/core-plugin-results-2}
  }
%
The results indicate that our mostly direct implementation of Linear Core is
successful in accepting the vast majority of the thousands of intermediate
programs produced by GHC when compiling libraries that make extensive use of
linear types.
%
However, Linear Core does not accept every single program produced by the
optimisation pipeline of GHC. Our manual analysis of rejected programs revealed
excellent results. Programs considered ill-typed by the Linear Core
implementation include:
%
\begin{itemize}
    \item Programs we expected GHC to produce, but know are not seen as linear by $\lambda^\pi_\Delta$.
        Namely, programs resulting from the reverse binder swap which
        scrutinize a variable and then use it again in the case alternatives.
        As discussed in Section~\ref{sec:reverse-binder-swap-considered-harmful}, these
        programs could be understood as linear as long as all applications of
        linear functions are reduced \emph{call-by-need}, but our system
        conservatively rejects it in light of the flexible evaluation strategy
        used by optimizing compilers.

    \item Programs rejected due to the lack of ``linearity coercions'' in
        $\lambda^\pi_\Delta$. Extending Linear Core with \emph{multiplicity coercions} 
        could make these programs be understood as linear
        (discussed in Section~\ref{sec:future-work}).

    \item Programs which actually violate linearity, in that the term
        violates the linear type it is annotated with. These are the programs
        we ultimately want to spot and reject by having a linearly typed Core
        validate intermediate programs. We found both a program which outright
        discarded a linear resource and that rewrite rules implementing stream
        fusion~\cite{10.1145/165180.165214,10.1145/1291151.1291199}
        (among other rules) could produce linearly-invalid programs (in one such case,
        the type of \lstinline{build} was not general enough to accommodate linearity).
\end{itemize}
%
Overall, the programs rejected by the Linear Core plugin,
%, in \texttt{linear-base},
besides validating that our implementation is faithful to the
$\lambda^\pi_\Delta$ system insofar as programs that should not be accepted
are deemed ill-typed, provide further insight into the remaining
details required to fully typecheck linearity in a mature optimising compiler.

% Talk about using our plugin on linear-base and other code bases... If I can get
% a few more case studies it would be pretty good. But then it's imperative to
% also use -dlinear-lint and make sure my plugin rejects a few of the examples

%%% \subsection{Consuming tagged resources as needed}
%%% 
%%% As discussed in Section~\ref{}, pattern-bound linear variables are
%%% put in the context with a \emph{tagged} usage environment with the resources of
%%% the scrutinee. In a \emph{tagged} usage environment environment, all resources
%%% are tagged with a constructor and an index into the many fields of the
%%% constructor.
%%% 
%%% In practice, a resource might have more than one tag. For example, in the following
%%% program, after the first pattern match, |a| and |b| have, respectively, usage
%%% environments $\{\lctag{x}{K_1}\}$ and $\{\lctag{x}{K_2}\}$:
%%% \begin{code}
%%% f x = case x of
%%%        K a b -> case a of
%%%         Pair n p -> (n,p)
%%% \end{code}
%%% However, in the following alternative, |n| has usage environment
%%% $\{\lctag{\lctag{x}{K_1}}{Pair_1}\}$ and |p| has
%%% $\{\lctag{\lctag{x}{K_1}}{Pair_2}\}$. To typecheck
%%% |(n,p)|, one has to $Split$ |x| first on |K| and then on |Pair|, in order for
%%% the usage environments to match.
%%% 
%%% In our implementation, we split resources on demand (and don't directly allow
%%% splitting linear resources), i.e. when we use a tagged resource we split the
%%% linear resource in the linear environment (if available), but never split otherwise.
%%% %
%%% Namely, starting on the innermost tag (the closest to the variable name), we
%%% substitute the linear resource for its split fragments, and then we iteratively
%%% further split those fragments if there are additional tags.
%%% %
%%% We note that it is safe to destructively split the resource (i.e. removing the
%%% original and only leaving the split fragments) because we only split resources
%%% when we need to consume a fragment, and as soon as one fragment is consumed
%%% then using the original ``whole'' variable would violate linearity.
%%% 
%%% In the example, if |n| is used, we have to use its usage environment, which in
%%% turn entails using $\lctag{\lctag{x}{K_1}}{Pair_1}$, which has two tags. In this order, we:
%%% \begin{itemize}
%%% \item Split $x$ into $\lctag{x}{K_1}$ and $\lctag{x}{K_2}$
%%% \item Split $\lctag{x}{K_1}$ and $\lctag{x}{K_2}$ into
%%%   \begin{itemize}
%%%   \item $\lctag{\lctag{x}{K_1}}{Pair_1}$ and $\lctag{\lctag{x}{K_1}}{Pair_2}$
%%%   \item Leave $\lctag{x}{K_2}$ untouched, as we only split on demand, and we aren't using a fragment of $\lctag{x}{K_2}$.
%%%   \end{itemize}
%%% \item Consume $\lctag{\lctag{x}{K_1}}{Pair_1}$, the usage environment of $n$, by removing it from the typing environment.
%%% \end{itemize}

%%%\subsection{Merging Linear Core into GHC\label{sec:merging-linear-core}}
%%%
%%%Describe the ticket for linear Core, the pending MRs, and the difficulty in
%%%even annotating the bind site across optimisations regardless of multiplicities.

% }}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Related and Future Work\label{sec:discussion}}

% In this chapter, we compare our contributions and Linear Core to related
% existing works in the literature,
% % (including linearity in a lazy, evaluation models in terms of linearity, and
% % the Core system).
% %
% consider further research
% % deemed out of the scope of this work and of the Linear Core type system
% (notably, so-called \emph{multiplicity coercions} to handle the
% interaction between linearity and \emph{coercions}, a key feature of Core which we
% left out our system), and conclude.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% {{{ Related Work
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%paragraph{Related Work}

% In this section we discuss related work, namely, Linear
% Haskell~\cite{cite:linearhaskell}, Linear Mini-Core~\cite{cite:minicore}, and
% linearity-influenced optimising
% transformations~\cite{cite:let-floating,peytonjones1997a,cite:linearhaskell}.

% TODO: A brief introduction to the related work section?

% \subsection{Formalization of Core}\todo{terrible paragraph name, and terrible paragraph}
% 
% As such, there exists no formal definition of Core that
% accounts for linearity. In this context, we intend to introduce a linearly typed
% System $F_C$ with multiplicity annotations and typing rules to serve
% as a basis for a linear Core. Critically, this Core linear language
% must account for call-by-need evaluation semantics and be valid in
% light of Core-to-Core optimising transformations.

% \parawith{System FC}

%System $F_C$~\cite{cite:systemfc} (Section~\ref{sec:core}) does not
%account for linearity in its original design, and, to the best of our
%knowledge, no extension to System $F_C$ with linearity and non-strict
%semantics exists.
%

% \begin{itemize}
% \item SystemFC tal como está não tem linearidade de todo
% \item Formalmente nao temos published definição de linearidade no Core
% \item Regras para sistema tipo FC com linearidade
% \item mas uma extensão tipo linear lambda calculus nao consegui exprimir as transformações do core
% \end{itemize}

% Haskell Core's foundational language was imbued with linear types, but it does
% not account for linearity with the whole of the type system
% 
% Multiplicity annotations in SystemFC?
% 
% Rules?

\paragraph{Linear Haskell\label{sec:related-work-linear-haskell}.}
%
% \todo[inline]{Say something about Linear Haskell's lazy operational semantics,
% but the type system not being concerned with linearity in the presence of
% laziness}
%
% Haskell, contrary to most programming languages with linear types, has existed
% for 31 years of its life \emph{without} linear types. As such, the introduction
% of linear types to Haskell comes with added challenges that do not exist in
% languages that were designed with linear types from the start:
% %
% \begin{itemize}
%     \item Backwards compatibility. The addition of linear types shouldn't break
%         all existing Haskell code.
%     \item Code re-usability. The linearly-typed part of Haskell's ecosystem and
%         its non-linearly-typed counterpart should fit in together and it must be
%         possible to define functions readily usable by both sides
%         simultaneously.
%     \item Future-proofing. Haskell, despite being an
%         industrial-strength language, is also a petri-dish for experimentation
%         and innovation in the field of programming languages. Therefore, Linear
%         Haskell takes care to accommodate possible future features, in
%         particular, its design is forwards compatible with affine and dependent
%         types.
% \end{itemize}
% %
Linear Haskell~\cite{cite:linearhaskell} augments the Haskell surface language
with linear types, but it is not concerned with extending GHC's intermediate
language(s), which presents its own challenges. 
%
However, the implementation of Linear Haskell in GHC does modify and extend Core
with linearity/multiplicity annotations. Core's type system is unable to type
semantic linearity of programs in the sense elaborated in our work,
which results in Core rejecting most linear programs resulting from optimising
transformations that leverage the non-strict semantics of Core.
%
Our work overcomes the limitations of Core's linear type system derived from
Linear Haskell by understanding linearity semantically in the presence of
laziness, and by showing that multiple Core-to-Core
optimisations employed by GHC are linearity preserving, using a
linearity-aware semantics that is essentially identical to that of~\cite{cite:linearhaskell}.
Linear Core can also be seen as a system that validates the programs written in
Linear Haskell and are compiled by GHC, by guaranteeing (through typing) that
linear resources are still used exactly once throughout the optimising
transformations.

\paragraph{Linearity-aware Semantics}
Several other works explore linearity aware semantics,
dating back to \cite{Chirimar_Gunter_Riecke_1996} and
\cite{TURNER1999231}.
The former uses a computational interpretation of linear logic to give
an account of reference counting. The latter present a heap-based
operational interpretation of linear logic and explore memory usage
properties under call-by-name and call-by-need, showing both satisfy
different properties under this lens. Their semantics is similar to
ours, but does not account for pattern matching.
More recently, \cite{DBLP:journals/pacmpl/ChoudhuryEEW21,DBLP:conf/esop/MarshallVO22} and
\cite{cite:linearhaskell} all make use of usage-aware
semantics. \citet{DBLP:journals/pacmpl/ChoudhuryEEW21} do so in the context of graded dependent
type theory, using the usage-aware heap-based operational semantics to
show correct accounting of resource usage in their setting, as we do
in ours, but they do not model
call-by-need, only call-by-name. The work of \cite{DBLP:conf/esop/MarshallVO22} is
applies the ideas of \cite{DBLP:journals/pacmpl/ChoudhuryEEW21}
to a non-graded (also call-by-name) setting. \citet{cite:linearhaskell} provide a similar semantics
to ours, but present only a naive linear typing system that does not
leverage non-strict evaluation.

\paragraph{Linear Mini-Core\label{sec:linear-mini-core}}

Linear Mini-Core~\cite{cite:minicore} is a specification of linear types in
Core as they were being implemented in GHC, and doubles as the (unpublished)
precursor to our work. Linear Mini-Core first observes the incapacity of
Core's type system to accept linear programs after transformations, and 
introduces usage environments for let-bound variables with the same goal of
Linear Core of specifying a linear type system for Core that accepts the
optimising transformations.

We draw from Linear Mini-Core the rule for non-recursive let expressions and
how let-bound variables are annotated with a usage environment. However, our
work further explores the interaction of laziness with linearity in depth, and
diverges significantly in rules for typing other constructs, notably, case expressions and
case alternatives. Furthermore, unlike Mini-Core, we prove Linear Core is type
safe, guarantees linear resource usage, and that multiple optimising
transformations, when applied to Linear Core programs, preserve linearity.

% \subsection{OutsideIn(X)\label{related-work-gadts}}
% 
% Defines constraint-based type system parametrized over X which does not account
% for local type refinements regarding linearity.
% 
% Se for modificar o typechecker com as multiplicity coercions vou ter de falar
% disto.

% \subsection{Rust}
% 
% Rust has a core based on linear types. Describe Rust's architecture?
% How do they handle linearity plus optimisations
% They probabluy don't typecheck linearity in Core

\paragraph{Linearity-influenced optimisations}

Core-to-Core transformations appear in many works across the research
literature~\cite{cite:let-floating,peytonjones1997a,santos1995compilation,peytonjones2002secrets,baker-finch2004constructed,maurer2017compiling,Breitner2016_1000054251,sergey_vytiniotis_jones_breitner_2017},
all designed in the context of a typed language (Core) which does not have
linear types. However, some works~\cite{cite:let-floating,peytonjones1997a,cite:linearhaskell} have observed that
certain optimisations (in particular, let-floating and inlining) greatly
benefit from linearity analysis and, in order to improve those transformation,
special purpose linear-type-inspired systems were created and implemented.
%
Preserving linear types in Core throughout the compilation pipeline allows the
optimiser to leverage non-heuristic linearity information where, previously, it
would resort solely to ad-hoc or incomplete custom-built linearity inference
passes (naturally, these passes are still necessary to optimise programs not
using explicit linearity or multiplicity annotations). Linearity may
potentially be used to the benefit of other optimising transformations that
currently do not take it into account.

% \begin{itemize}
% \item Transfs. core to core aparecem em vários artigos, e são desenhadas no contexto de uma linguagem tipificada mas que não é linearly typed.
% \item nestes dois artigos é observado que se houvesse a capacidade de explorar linearidade podiamos fazer as coisas de forma diferente
% \item Todas estas optimizaçoes de decadas foram desenhadas sem linear types e há sitios onde linear types podiam ajudar mas não existiam na altura
% \item POdemos usar linear types multiplicitpiadads para lazy language core q definimos para nao ter de fazer sistemas lineares de proposito para optimizações
% \item Ser ad-hoc incompleto ou nao feito de todo
% \end{itemize}

% \parawith{A transformation based optimiser for Haskell}
% They discuss a cardinality analysis based on a linear type system but create (an
% ad-hoc?) one suited. Comparison in the measure of creating optimizations based
% on linearity.
% 
% \parawith{Let-Floating: Moving Bindings to Give Faster Programs\label{sec:rw:let-floating}}
% In the paper~\cite{cite:let-floating}...
% They say they are doing work on a linear type system to identify places where
% the lambda is linearly used and therefore floating-in is beneficial and
% floating-out not as productive.

% \subsection{Call-by-value, call-by-name and call-by-value...}
% 
% The work~\cite{cite:call-by-name-value-and-need-linear-lambda-calculus}...

% }}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% {{{ Future Work
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Future Work\label{sec:future-work}}

We highlight some avenues of future work. Briefly,
these include \emph{multiplicity coercions}, optimisations leveraging
linearity, resource inference for usage environments, and ultimately using
Linear Core in a mature optimising compiler with lazy evaluation and linear
types -- the Glasgow Haskell Compiler. Lastly, we discuss the
% we hypothesize Linear($X$), a linear system parametrized on the evaluation strategy, and the
generalization of Linear Core to the surface Haskell language.

\paragraph{Multiplicity Coercions.}
Linear Core does not have type equality coercions, a flagship feature of GHC
Core's type system.
%
Coercions %briefly explained in Section~\ref{sec:core},
allow the Core intermediate language to encode a panoply of Haskell source
type-level features such as GADTs, type families, or newtypes.
%
In Linear Haskell, multiplicities are introduced as annotations to function
arrows which specify the linearity of the function. In practice,
multiplicities are simply types of kind \incode{Multiplicity}, where \incode{One} and \incode{Many}
are the type constructors of the kind \incode{Multiplicity}; multiplicity polymorphism
follows from type polymorphism, where multiplicity variables are
just type variables. Encoding multiplicities as types allows Haskell programs
to leverage features available for types to naturally extend to multiplicities
as well.
For instance, it is possible via the use of GADTs to define a
function whose linearity of its second argument depends on the value
of its first argument, internally realized through so-called
\emph{multiplicity coercions}. Currently, Core cannot make use of such
coercions to determine whether the usages of linear resources match
their intended multiplicity. Studying the interaction between coercions and multiplicities is a main avenue
of future work for Linear Core.
%
% Consequently, we might define, e.g., using a GADT \incode{SBool} and a type family
% |If|, the function \incode{dep} which is linear in the second argument if the first
% argument is \incode{STrue} and unrestricted otherwise:
% %
% \begin{limitation}
% \begin{lstlisting}
% data SBool :: Bool -> Type where
%   STrue :: SBool True
%   SFalse :: SBool False

% type family If b t f where
%   If True t _ = t
%   If False _ f = f

% dep :: SBool b -> Int %(If b One Many) -> Int
% dep STrue x = x
% dep SFalse _ = 0
% \end{lstlisting}
% \end{limitation}
% %
% This example is linear and should be accepted. However, the example is rejected
% by the GHC's Core type checker.
% Critically, Core doesn't currently understand
% so-called \emph{multiplicity coercions}. Even though after matching on |STrue|
% we have access to a coercion from the function multiplicity $m$ to $1$ ($m \sim
% 1$), we cannot use this coercion to determine whether the usages of the linear
% resources match the multiplicity.
% %
% Studying the interaction between coercions and multiplicities is a main avenue
% of future work for Linear Core.

% In GHC, multiplicity coercions are tracked by issue $19517$\footnote{https://gitlab.haskell.org/ghc/ghc/-/issues/19517}.

% \todo[inline]{Relate to levity polymorphism and runtime rep coercions (Sam)}

\paragraph{Optimisations leveraging linearity.}
We only briefly mentioned how linearity can inform optimisations to produce
more performant programs. We leave exploring optimisations unblocked by
preserving linearity in the intermediate language with Linear Core as future
work. Linearity influenced optimising transformations have been also discussed
by Linear Haskell~\cite{cite:linearhaskell} and
in~\cite{cite:let-floating,peytonjones1997a}. An obvious candidate is
\emph{inlining}, which is applied based on heuristics from information provided
by the \emph{cardinality analysis} pass that counts occurrences of bound
variables.  Linearity can be used to non-heuristically inform
the inliner~\cite{cite:linearhaskell}.
% Additionally, we argue that in Linear Core accepting more programs as linear
% there are more chances to use linearity, in contrast to a linear type system
% which does not account for lazy evaluation and thus rejects more programs.

% \paragraph{Usage environment resource inference.}
% In Section~\ref{sec:linearity-semantically}, we discussed how there is not an obvious way to determine the set of linear
% resources used by a group of (mutually) recursive bindings, .
% though the resources used by each binder are the solution to a set determined
% by the recursive bindings group.  In Section~\ref{sec:main:linear-core}, we
% further likened the challenge of determining usage environments for a recursive
% group of bindings to a unification problem as that solved by the Hindley-Milner
% type inference algorithm~\cite{DBLP:conf/popl/DamasM82} based on generating and solving
% constraints. Even though these are useful observations, our implementation of
% Linear Core uses a naive algorithm to determine the usage environments, thereby
% leaving as future work the design of a principled algorithm to determine the
% usage environments of recursive group of bindings.

\paragraph{Linear Core in the Glasgow Haskell Compiler.}
Linear Core is suitable as the intermediate language of an optimising compiler
for a linear and lazy language such as Haskell Core, in that optimising
transformations in Linear Core preserve types \emph{and} linearity, since Linear
Core understands semantic linearity in the presence of laziness, unlike
Core's current type system under which optimisations currently violate
linearity.
%
Integrating Linear Core in the Glasgow Haskell Compiler is one of the ultimate
goals of our work. Core's current type system ignores linearity due to
its limitation in understanding semantic linearity, and our work fills this gap
and would allow Core to be linearly typed all throughout.
%
A linearly typed Core that preserves linearity throughout the optimisation
pipeline of GHC both validates the correctness of the compiler, which is
already achieved to a great extent by preserving (non-linear) types, and
informs optimisations, allowing the compiler to generate more performant programs.
%
Implementing Linear Core in GHC is a challenging endeavour, since we must
account for all other Core features (e.g. strict constructor fields), propagate
new or modified data throughout the entire pipeline, and accept more
optimisations. Despite our initiative in this direction\footnote{[Link removed
due to
anonymization]} %\footnote{https://gitlab.haskell.org/ghc/ghc/-/issues/23218} ,
we leave this as future work.

\paragraph{Generalizing Linear Core to Haskell.}
Linear types, despite their compile-time correctness guarantees regarding
resource management, impose a burden on programmers in being a restrictive
typing discipline (witnessed, e.g., by Linear
Constraints~\cite{cite:linearconstraints}). Linear Core eases the restrictions
of linear typing by being more flexible in understanding linearity for lazy
languages. In this sense, it is an avenue of future
work to apply the ideas from this work to the surface Haskell language.

% \section*{Acknowledgements}

% We thank Arnaud Spiwack for valuable and insightful discussions.\todo{Do these have to be anonymized?}

% \begin{itemize}
% 
% \item Linear(X), a linear type system defined by the underlying definition of
% evaluation (which in turn implies how consuming a resource is defined)
% 
% \item Implementation in Core
% 
% \item Generalization to source level language, being more permissive in the
% handling of resources imposes less burden on the programmer
% 
% \item It's harder to typecheck linearity like this in the source level because
% of the interaction with other source features, but seems feasible and an
% improvement to the usability of linear types. It allows more lazy functional
% programming idioms with linear types (also because laziness and strictness is less well defined as in Core, bc opts)
% 
% \item Beautiful inference algorithm for recursive usage environments -- insight
% that looks like inference for recursive function principle types, but haven't
% figured it out -- connection to type inference / hindley milner
% 
% \item We kind of ignore the multiplicity semiring. We should discuss
% how we don't do ring operations ... but that's kind of wrong.
% 
% % \item We know the case binder to ALWAYS be in WHNF, perhaps there could
% % be some annotation on the case binder s.t. we know nothing happens when we
% % scrutinize it as a single variable
% 
% \item Mechanizing the system and metatheory
% 
% \end{itemize}

% }}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% {{{ Conclusion
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% We give up on this section and instead describe "syntax-directedness" in the implementation chapter
%
%
% \section{Syntax Directed System}
% 
% \todo[inline]{In the other system we assume that the recursive lets are strongly connected, i.e. the expressions always}
% 
% \input{language-v4/SyntaxDirectedSystem}
% 
% \subsection{Inferring usage environments}
% 
% \todo[inline]{The difference between this and the previous section is a bit blurry}
% 
% \todo[inline]{There's one more concern: usage environments aren't readily
% available, especially in recursive lets. We must perform inference of usage
% environments before we can typecheck using them. This is how:}
% 
% \todo[inline]{Rather, we define a syntax directed type system that infers usage environments while checking...}
%

% \section{Conclusion}

% Optimising compilers with a typed and lazy intermediate language with linear
% types (of which GHC is the prime example) leverage laziness to heavily
% transform and rewrite programs into simpler forms.
% %
% However, these optimising transformations push the interaction between
% linearity and laziness to the limits where linearity can no longer be seen
% syntactically, despite being maintained semantically, in the sense that linear
% resources are still used exactly once when the optimised program is run.

% In this work we explored linearity in the presence of laziness by example
% through the interactions of linear types with lazy (recursive) let bindings and
% case expressions that evaluate their scrutinee to Weak Head Normal Form. Most
% example programs were linear semantically, but not syntactically.
% %
% We developed a linear type system, Linear Core, for an intermediate language
% akin to GHC Core, with laziness and linearity. In contrast to GHC Core's type
% system, or any other linear type system (to the best of our knowledge), our
% type system understands semantic linearity, and can thus correctly type a wider
% range of linear programs, as those explored in the semantic linearity examples.
% %
% Crucially, we proved soundness of the type system, and proved multiple
% optimising transformations preserve linearity, despite most violating linearity
% in other linear type systems. Additionally, we implemented Linear Core as a GHC
% plugin to further explore its suitability in the intermediate language of an
% optimising compiler.

% Concluding, Linear Core is a suitable type system for linear, lazy,
% intermediate languages of optimising compilers such as GHC, as it understands
% linearity in the presence of laziness s.t. optimisations preserve types and
% linearity, and further unblocks optimisations influenced by linearity, e.g.
% inlining and call-by-name $\beta$-reduction for applications of (semantically)
% linear functions.

% }}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\bibliography{references}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% {{{ Appendix
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\appendix

\section{Natural and Instrumented Operational Semantics}\label{app:opsem}

We define a lazy big-step operational semantics for $\lambda_\Delta^\pi$ in the
style of Launchbury~\cite{10.1145/158511.158618} natural semantics. The semantics is
equipped with an environment which assigns expressions to variables and is
mutated in order to express lazy evaluation. Terms are explicit about
their sharing (in the sense of shared reductions in lazy evaluation),
and so we perform a translation that makes all sharing explicit
through let-binders. This translation is presented in Figure~\ref{fig:explicitsharing}.
We write $\llet{\ov{x_i{:}_{\Delta_i}\sigma = e_i}}{e'}$ for the iterated
let-binding of variables $\ov{x_i}$.
\begin{figure}
\[
  \begin{array}{lclr}
    x^* & \triangleq & x\\

    (\Lambda p.~e)^* & \triangleq & \Lambda p.~(e)^*\\
    (\lambda x:_\pi\sigma .e)^* & \triangleq & \lambda x:_\pi\sigma . (e)^*\\
    (e~\pi)^* & \triangleq & (e)^*~\pi\\
    (e~x)^* & \triangleq & (e)^*~x\\
    (e~e')^* & \triangleq & \llet{x{:}_{\Delta}\sigma =
                            (e')^*}{(e)^*~x}     \\ &&\qquad \mbox{with
                            $\Gamma;\Delta' \vdash e :
                            \sigma\rightarrow_\pi \varphi$} \\ &&\qquad \mbox{and
                                       $\Gamma;\Delta \vdash e' : \sigma$}\\
      K~\ov{e_i} & \triangleq & \llet{\ov{x_i:_{\Delta_i} \sigma_i =
                                   (e_i)^*}}{K~\ov{x_i}} \\ &&
                                       \qquad\mbox{with
                                       $\ov{\Gamma;\Delta_i\vdash
                                       e_i :
                                       \sigma_i}$}\\                                              
(\ccase{e}{z{:}_{\Delta}\sigma~\{\overline{\rho\to e'}\}})^* &
                                                               \triangleq
                                &
                                  \ccase{(e)^*}{z{:}_{\Delta}\sigma~\{\overline{\rho\to (e')^*}\}}\\
    (\llet{x{:}_{\Delta}\sigma = e}{e'}      )^* & \triangleq & \llet{x{:}_{\Delta}\sigma  = (e)^*}{(e')^*}    \\
    (\lletrec{\overline{x{:}_{\Delta}\sigma = e}}{e'}  )^* & \triangleq &\lletrec{\overline{x{:}_{\Delta}\sigma = (e)^*}}{(e')^*}  
      \end{array}
    \]
    \caption{Explicit Sharing Translation\label{fig:explicitsharing}}
  \end{figure}

The natural semantics (Figure~\ref{fig:natsemapp}) is defined via the relation
$\Theta : e \Downarrow \Theta' : v$ where $e$ is an expression, $v$ a
value, $\Theta$ and $\Theta'$ are environments with bindings of the
form $x:_\omega\sigma = e$ which assigns the expression $e$ to the
variable $x$ of the given type.

\begin{figure}
\[
  \begin{array}{c}
    \infer[]
    {\,}
    {\Theta : \Lambda p.e \Downarrow \Theta : \Lambda p.e}
    \qquad
    \infer[]
    {\Theta  : e\Downarrow \Theta' : \Lambda p.e' \quad \Theta' :
    e'[\pi/p]\Downarrow \Theta'' : v}
    {\Theta : e~\pi\Downarrow \Theta'' : v }\\[1.5em]
    \infer[]
    {\,}
    {\Theta : \lambda x:_\pi \sigma . e \Downarrow \Theta : \lambda x:_\pi \sigma . e }
    \qquad
    \infer[]
    {\Theta : e \Downarrow \Theta' : \lambda y:_\pi \sigma . e' \quad
    \Theta' : e'[x/y] \Downarrow \Theta'' : v }
    {\Theta : e~x \Downarrow \Theta'' : v}\\[1.5em]
    \infer[]
    {(\Theta , x:_\omega \sigma = e): e \Downarrow \Theta' : v}
    {(\Theta , x:_\omega \sigma = e) : x \Downarrow (\Theta',x:_\omega
    \sigma = v) : v}
    \qquad
    \infer[]
    {(\Theta,x:_\omega \sigma = e) : e' \Downarrow \Theta' : v}
    {\Theta : \llet{x{:}_{\Delta}\sigma = e}{e'}\Downarrow \Theta' : v
    }\\[1.5em]
    \infer[]
    {\,}
    {\Theta : K~\ov{x_i} \Downarrow \Theta : K~\ov{x_i}}
    \qquad
    \infer[]
    {\Theta : e \Downarrow \Theta' : K~\ov{x_i} \quad
    \Theta' : e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] \Downarrow \Theta'' : v }
    {\Theta :
    \ccase{e}{z{:}_{\Delta'}\sigma~\{{\dots,K~\ov{y_i} \to e', {\dots}}\}}
    \Downarrow \Theta'' : v}\\[1.5em]
    \infer[]
    {\Theta : e \Downarrow \Theta' : K~\ov{x_i} \quad
    \Theta' : e'[K~\ov{x_i}/z] \Downarrow \Theta'' : v }
    {\Theta :
    \ccase{e}{z{:}_{\Delta'}\sigma~\{\dots , \_\to
    e'}\} \Downarrow \Theta'' : v}
    \qquad
    \infer[]
    {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
    % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
    % \Downarrow \Theta_n : v_n \quad
    (\Theta, \overline{x_i:_\omega \sigma_i = e_i}) : e' \Downarrow \Theta'
    : v
    }
    {\Theta : \lletrec{\overline{x_i{:}_{\Delta}\sigma_i = e_i}}{e'}
    \Downarrow \Theta' : v}
    \end{array}
  \]
  \caption{Natural Semantics for $\lambda_\Delta^\pi$\label{fig:natsemapp}}
\end{figure}

Our instrumented, linearity-aware semantics (akin to that
of~\cite{cite:linearhaskell})
is defined in Figure~\ref{fig:instrsemapp}.
In this semantics, linear variables are erased from
the runtime environment, ensuring that any term violating linearity
produces a stuck state. The semantics is defined by
the judgment $\Gamma ; \Delta \vdash (\Theta \mid e) \Downarrow (\Theta'
\mid v) : \sigma , \Sigma$ where $\Gamma$ and $\Delta$ are typing judgments
($\Gamma$ tracks unrestricted variables, $\Delta$ tracks linear
assumptions) for variables occurring in $\Sigma$;
$\Theta$ and $\Theta'$ are runtime environments, consisting of bindings of the form
$x :_\pi \sigma = e$ or $x :_\Delta \sigma = e$ ; $e$ is the expression being evaluated and $v$
its resulting value, both of type $\sigma$; $\Sigma$ is a list of assignments of the form
$e : \tau$, which are expressions in which variables in $\Theta$ are
used.

\begin{figure}
\[
  \begin{array}{c}
    \infer[]
    {\,}
    {\Gamma;\Delta \vdash (\Theta \mid \Lambda p.e) \Downarrow (\Theta \mid
    \Lambda p.e) : \forall p . \sigma , \Sigma }
    \\[1.5em]
    \infer[]
    {\Gamma ; \Delta \vdash (\Theta \mid  e) \Downarrow (\Theta' \mid
    \Lambda p.e') : \forall p . \sigma , \Sigma \quad \Gamma ;\Delta \vdash
      (\Theta' \mid e'[\pi/p]) \Downarrow (\Theta'' \mid v) : \sigma[\pi/p], \Sigma}
    {\Gamma ; \Delta \vdash (\Theta \mid e~\pi) \Downarrow (\Theta''
      \mid v) : \sigma[\pi/p] , \Sigma  }\\[1.5em]
    \infer[]
    {\,}
    {\Gamma ;\Delta \vdash (\Theta \mid \lambda x:_\pi \sigma . e)
    \Downarrow (\Theta \mid \lambda x:_\pi\sigma .e) : \sigma
    \rightarrow_\pi \tau,\Sigma }\\[1.5em]
    \infer[]
    {\Gamma;\Delta \vdash  (\Theta \mid e) \Downarrow (\Theta' \mid
    \lambda y:_1 \sigma . e') : \sigma \rightarrow_1 \tau , x:\sigma, \Sigma
    \quad
    \Gamma ; \Delta \vdash (\Theta' , y:_{1} \sigma = x \mid e') \Downarrow (\Theta''
    \mid v) : \tau , \Sigma  }
    {\Gamma ; \Delta\vdash (\Theta \mid  e~x ) \Downarrow (\Theta''
    \mid v) : \tau , \Sigma }\\[1.5em]
        \infer[]
    {\Gamma;\Delta \vdash  (\Theta \mid e) \Downarrow (\Theta' \mid
    \lambda y:_\omega \sigma . e') : \sigma \rightarrow_\omega \tau , x:\sigma, \Sigma
    \quad
            \Gamma ; \Delta \vdash (\Theta' \mid e'[x/y]) \Downarrow (\Theta''
    \mid v) : \tau , \Sigma  }
    {\Gamma ; \Delta\vdash (\Theta \mid  e~x ) \Downarrow (\Theta''
    \mid v) : \tau , \Sigma }\\[1.5em]
       \infer[]
    {\Gamma ; \Delta  \vdash (\Theta, x:_{\Delta'} \sigma = e \mid e)
    \Downarrow (\Theta' \mid v) : \sigma , \Sigma \\
      \D'' = \D' \setminus (\Theta \setminus \Theta')
      }
    {\Gamma ; \Delta \vdash (\Theta , x:_{\Delta'} \sigma = e \mid x)
      \Downarrow (\Theta' , x:_{\Delta''} \sigma = v \mid v) : \sigma, \Sigma}
    \\[1.5em]
           \infer[]
    {\Gamma ;\Delta  \vdash (\Theta \mid e)
    \Downarrow (\Theta' \mid v) : \sigma,\Sigma}
    {\Gamma ; \Delta \vdash (\Theta , x:_1 \sigma = e \mid x)
    \Downarrow (\Theta' \mid v) : \sigma,\Sigma}
    \\[1.5em]

    \infer[]
    {\Gamma ;\Delta \vdash (\Theta,x:_{\Delta'} \sigma = e \mid e')
    \Downarrow (\Theta' \mid v) : \tau,\Sigma}
    {\Gamma ; \Delta \vdash (\Theta \mid \llet{x{:}_{\Delta'}\sigma =
    e}{e'})\Downarrow (\Theta' \mid v) : \tau , \Sigma
    }\\[1.5em]

    \infer[]
    {\,}
    {\Gamma ; \Delta \vdash (\Theta \mid K~\ov{x_i}) \Downarrow
    (\Theta \mid K~\ov{x_i}) : T , \Sigma}
    \\[1.5em]

    \infer[]
    { K : \ov{ \sigma_i
    \rightarrow_\omega \sigma_j \lolli} T
    \\
    \Gamma, z{:}_{\ov{y_j}}T, \ov{y_i :_\omega \sigma_i} ;\Delta , \ov{y_j :_1 \sigma_j} \vdash (\Theta \mid e) \Downarrow (\Theta' \mid
      K~\ov{x_i}) : T , e' : \tau ,\Sigma \\
    \Gamma ;\Delta \vdash (\Theta' \mid e'\ov{[x_i/y_{i,j}]}[K~\ov{x_i}/z]
    ) \Downarrow (\Theta'' \mid v) :\tau , \Sigma }
    {\Gamma ;\Delta \vdash (\Theta \mid
    \ccase{e}{z{:}_{\Delta'}T~\{{\dots,K~\ov{y_{i,j}} \to e', {\dots} }\}})
    \Downarrow (\Theta'' \mid  v) : \tau , \Sigma}\\[1.5em]

    \infer[]
    {K : \ov{ \sigma_i
    \rightarrow_\omega \sigma_j \lolli} T\\
    \Gamma ; \Delta, z{:}_1\s \vdash (\Theta \mid e) \Downarrow (\Theta' \mid
    K~\ov{x_i}) : T , e', \Sigma \quad
    \Gamma ;\Delta \vdash (\Theta' \mid  e'[K~\ov{x_i}/z])
    \Downarrow (\Theta'' \mid v) : \tau,\Sigma }
    {\Gamma ; \Delta \vdash (\Theta \mid 
    \ccase{e}{z{:}_{\Delta'}T~\{\dots , \_\to
    e'}\}) \Downarrow (\Theta'' \mid v) : \tau}
    \\[1.5em]

     \infer[]
     {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
     % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
     % \Downarrow \Theta_n : v_n \quad
     (\Gamma ; \Delta \vdash (\Theta, \overline{x_i:_{\Delta'} \sigma_i =
     e_i} \mid e' )\Downarrow (\Theta'
     \mid v) : \tau
     }
     {\Gamma ;\Delta \vdash (\Theta \mid \lletrec{\overline{x_i{:}_{\Delta'}\sigma_i = e_i}}{e'})
     \Downarrow (\Theta' \mid v) : \tau}

   \end{array}
 \]
 \caption{Instrumented Semantics for $\lambda_\Delta^\pi$\label{fig:instrsemapp}}
 \end{figure}

 Following~\cite{Gunter1993APA,cite:linearhaskell}, we also consider
 partial derivations of our semantics in order to be able to state a
 progress result. Given our set of rules defining judgments of the
 form $a \Downarrow b$ with ordered premises, we define a total
 derivation of $a \Downarrow b$ as a tree in the standard way. A
 partial derivation $a \Downarrow?$ is either a root labelled with $a
 \Downarrow?$ or an application of a rule matching $a$ on the left of
 the evaluation arrow, where exactly one of the premises $a'
 \Downarrow?$ has a partial  derivation, all the premises to the left
 of $a' \Downarrow?$ have a total derivation, and the premises to the
 right of $a'\Downarrow?$ are not known yet.
 %

 By construction, in a partial derivation, there is only one premise
 $b\Downarrow?$ with no sub-derivation. We call $b$ the head of such a
 partial derivation.
We write $a \Downarrow^* b$ for the relation which holds when $b$ is
the head of some partial derivation with root $a \Downarrow?$.  We
write $a \Downarrow^* b$  the partial evaluation relation.

Below, we write $\Theta {\uparrow^1}$ for the linear bindings in evaluation
environment $\Theta$ and $\Theta {\uparrow^\omega}$ for all other bindings.
$\Theta\downarrow^1_e$ refers to the expressions bound to linear variables in
the runtime environment and $\Theta\downarrow^1_\sigma$ refers to the types of
those expressions.

 \begin{definition}[Environment Expansion]
   We rely on an expansion of a runtime environment $\Theta$ for the evaluation of a term $e$ into a (typed) term, written $[\Theta \mid e]$, defined inductively as follows:
   \[
     \begin{array}{lcl}
       \left[ \cdot \mid e \right] & \triangleq & e\\
       \left[ \Theta , \xl = e' \mid e \right] & \triangleq  & [\Theta \mid (\lambda \xl . e)~e']\\
       \left[\Theta , x:_\Delta \sigma = e' \mid e \right] & \triangleq & [\Theta \mid \llet{x:_\Delta \sigma = e'}{e}]
    \end{array}
   \]
 \end{definition}

We note that for the treatment of recursive let groups, environment
expansion needs to happen for the entire group of let-bound
variables. This is easily achieved by extending the runtime
environment with information about recursive let groupings, which we
omit for the sake of presentation.

 \begin{definition}[Well-typed State]
   We write $\G;\D \vdash \Theta \mid e : \tau,\Sigma$ to denote the following typing judgment:

   \[
\G ; \D \vdash [ \Theta \mid (e,\mathit{terms}(\Sigma))] : \tau \otimes \bigotimes(\Sigma)
   \]
   where $\mathit{terms}(e_1 : \sigma_1 , \dots , e_n : \sigma_n)$ stands for the tuple $(e_1 ,( \dots , (e_n , ())))$ and
   $\bigotimes(e_1 : \sigma_1 , \dots , e_n : \sigma_n)$ for $\sigma_1 \otimes ( \dots (\sigma_n \otimes ()))$.
 \end{definition}

 \begin{lemma}~\label{lem:unwrapenv}
   $\G ; \D \vdash [ \Theta \mid e] : \sigma$
     iff $\G , \Theta\uparrow^{\omega} ; \D , \Theta\uparrow^1 \vdash  (e, \Theta\downarrow^1_e) : \sigma \otimes \bigotimes(\Theta\downarrow^1_\sigma)$
   \end{lemma}
\begin{proof}
By structural induction on the environment expansion definition in the
forward direction and by a straightforward induction on $\Theta$ in the
converse direction. We show only the forward direction.
\begin{description}
%\item[Case:] $\left[\cdot \mid e\right] \triangleq e$ Trivial
\item[Case:] $\left[\Theta, \xl = y \mid e\right] \triangleq \left[\Theta \mid (\lambda \xl. e)~y\right]$
\begin{tabbing}
    (1) $\G; \D \vdash \left[ \Theta, \xl = y \mid e \right] : \phi$\\
    (2) $\G; \D \vdash \left[ \Theta \mid (\lambda \xl. e)~y \right] : \phi$\` by def.\\
    (3) $\G, \Theta\uparrow^\omega; \D, \Theta\uparrow^1 \vdash ((\lambda \xl. e)~y, \Theta\downarrow^1_e) : \phi \otimes \bigotimes(\Theta\downarrow^1_\sigma)$\` by i.h.\\
    (4) $Let~\D_1,\D_2,\D_3 = \D, \Theta\uparrow^1 $\\
    (5) $\G, \Theta\uparrow^\omega; \D_3 \vdash \Theta\downarrow^1_e : \bigotimes(\Theta\downarrow^1_\sigma)$\\
    (6) $\G, \Theta\uparrow^\omega; \D_2 \vdash y : \s$\` by inv\\
    (7) $\G, \Theta\uparrow^\omega; \D_1 \vdash (\lambda \xl. e) : \s \to \phi $\` by inv\\
    (8) $\G, \Theta\uparrow^\omega; \D_1, \xl \vdash e : \phi $\` by inv\\
    (9) $\G, \Theta\uparrow^\omega; \D_1, \D_2, \xl \vdash (e, y) : \phi \otimes \sigma$\` by pair\\
    (10) $\G, \Theta\uparrow^\omega; \D, \xl \vdash (e, y, \Theta\downarrow^1_e) : \phi \otimes \sigma \otimes \bigotimes(\Theta\downarrow^1_\sigma)$\` by pair\\
\end{tabbing}
\item[Case:] $\left[\Theta, \xD = e' \mid e\right] \triangleq \left[\Theta \mid \llet{\xD = e'}{e}\right]$
\begin{tabbing}
    (1) $\G; \D' \vdash \left[ \Theta, \xD = y \mid e \right] : \phi$\\
    (2) $\G; \D' \vdash \left[ \Theta \mid \llet{\xD = e'}{e} \right] : \phi$\` by def.\\
    (3) $\G, \Theta\uparrow^\omega; \D', \Theta\uparrow^1 \vdash (\llet{\xD = e'}{e}, \Theta\downarrow^1_e) : \phi \otimes \bigotimes(\Theta\downarrow^1_\sigma)$\` by i.h.\\
    (4) $Let~\D,\D_2,\D_3 = \D', \Theta\uparrow^1 $\\
    (5) $\G, \Theta\uparrow^\omega; \D_3 \vdash \Theta\downarrow^1_e : \bigotimes(\Theta\downarrow^1_\sigma)$\\
    (6) $\G, \Theta\uparrow^\omega; \D \vdash e' : \s$\` by inv on let\\
    (7) $\G, \Theta\uparrow^\omega, \xD; \D,\D_2 \vdash e : \phi $\` by inv on let\\
    (8) $\G, \Theta\uparrow^\omega, \xD; \D,\D_2,\D_3 \vdash (e, \Theta\downarrow^1_e) : \phi \otimes \bigotimes(\Theta\downarrow^1_\sigma)$\` by pair\\
\end{tabbing}
\end{description}
\end{proof}

\subsection{Bisimulation}

We relate our two operational semantics through a bisimulation,
allowing us to transfer type safety of the instrumented semantics to
the lazy big-step semantics. The relation essentially allows the
execution environment to treat some unrestricted bindings
of the natural semantics as linear.

\begin{definition}[Bisimulation]
We define the relation $\gamma(\Theta : e)(\G ; \D \vdash \Theta' \mid
e' : \tau,\Sigma)$ between a well-typed state and an ordinary state of
the lazy big-step semantics if $e = e'$ and $\Theta'{\uparrow^\omega}
\leq \Theta$.
\end{definition}

\begin{lemma}~\label{lem:naturaltoinstr}
  \begin{itemize}
    \item For all $\gamma(\Theta : e)(\G ; \D \vdash \Theta' \mid
e : \tau,\Sigma)$ such that $\Theta : e \Downarrow \Theta'' : v$
there exists a well-typed state $\G ; \D \vdash (\Theta''' \mid v) :
\tau,\Sigma$ such that $\G ; \D \vdash (\Theta' \mid e) \Downarrow
(\Theta''' \mid v) : \tau,\Sigma$ and $\gamma(\Theta'' : v)(\G ; \D
\vdash \Theta''' \mid v : \tau,\Sigma)$
\item For all $\gamma(\Theta : e)(\G ; \D \vdash \Theta' \mid
e : \tau,\Sigma)$ such that $\Theta : e \Downarrow^* \Theta'' : v$
there exists a well-typed state $\G ; \D \vdash (\Theta''' \mid v) :
\tau,\Sigma$ such that $\G ; \D \vdash (\Theta' \mid e) \Downarrow^*
(\Theta''' \mid v) : \tau,\Sigma$ and $\gamma(\Theta'' : v)(\G ; \D
\vdash \Theta''' \mid v : \tau,\Sigma)$
\end{itemize}
\end{lemma}
\begin{proof}
By induction on the given operational semantics rules.
  \end{proof}

\begin{lemma}~\label{lem:instrtonatural}
  \begin{itemize}
\item  For all $\gamma(\Theta : e)(\G ; \D \vdash \Theta' \mid
e : \tau,\Sigma)$ such that $\G ; \D \vdash (\Theta' \mid e) \Downarrow
(\Theta'' \mid v) : \tau,\Sigma$, there exists a state $\Theta''' : v$
such that $\Theta : e \Downarrow \Theta''' : v$ and
$\gamma(\Theta''' : v)(\G ; \D \vdash (\Theta'' \mid v) :
\tau,\Sigma)$
\item For all $\gamma(\Theta : e)(\G ; \D \vdash \Theta' \mid
e : \tau,\Sigma)$ such that $\G ; \D \vdash (\Theta' \mid e) \Downarrow^*
(\Theta'' \mid v) : \tau,\Sigma$, there exists a state $\Theta''' : v$
such that $\Theta : e \Downarrow^* \Theta''' : v$ and
$\gamma(\Theta''' : v)(\G ; \D \vdash (\Theta'' \mid v) :
\tau,\Sigma)$
   \end{itemize}
 \end{lemma}
\begin{proof}
By induction on the given operational semantics rules.
  \end{proof}

\section{Type Safety Proofs}\label{app:proofs}

In the following we write $\Mapsto$ to stand for
$\Rightarrow_\mathsf{WHNF}$ and $\Rrightarrow$ for $\Rightarrow_\mathsf{NWHNF}$.

\subsection{Auxiliary Lemmas}
The following lemmas presuppose the variable names across the typing
contexts are distinct.

  \begin{lemma}[Linear to $\Delta$-bound]~\label{lem:onedelta-app}

    \begin{enumerate}

      \item If $\G; \D',\xl \vdash e :\vp$
        then $\G[\D/x],\xD; \D,\D' \vdash e : \vp$ (with $\Delta$ fresh).
      \item If $\G ; \D' , \xl \vdash_{alt} \rho \rightarrow e
        :^z_{\Delta_s} \sigma \Rightarrow \rho$ then
        $\G[\D/x],\xD;\D,\D' \vdash_{alt} \rho \rightarrow e
        :^z_{\Delta_s} \sigma \Rightarrow \rho$ (with $\Delta$ fresh).

      \end{enumerate}
\end{lemma}
\begin{proof}
Straightforward induction on the structure of the given derivation,
reconstructing the original derivation and applying rule
$\mathit{Var}_\Delta$ where rule $\mathit{Var}_1$ was applied on
$x$ in the original derivation.
\end{proof}
\onedelta*

\begin{proof}
By Lemma~\ref{lem:onedelta-app}(1).
\end{proof}

\begin{lemma}[$\Delta$-bound to Linear]~\label{lem:deltaone-app}

  \begin{enumerate}
 \item If $\G,\x[\irr{\D}]; \irr{\D},\D' \vdash e : \vp$
   then $\G[x/\irr{\D}]; \D',\xl \vdash e :\vp$.
 \item If $\G ,\x[\irr{\D}]; \irr{\D}, \D' \vdash_{alt} \rho \rightarrow e
   :^z_{\Delta_s} \sigma \Rightarrow \rho$ then
   $\G[x/\irr{\D}];\D',\xl \vdash_{alt} \rho \rightarrow e
   :^z_{\Delta_s} \sigma \Rightarrow \rho$.
  \end{enumerate}
\end{lemma}
\begin{proof}
Straightforward induction on the structure of the given derivation,
reconstructing the original derivation and applying rule
$\mathit{Var}_1$ where rule $\mathit{Var}_\D$ was applied on
$x$ in the original derivation.
\end{proof}

\deltaone*
\begin{proof}
By Lemma~\ref{lem:deltaone-app}(1).
\end{proof}

\begin{lemma}[Unrestricted and $\Delta$-bound]~\label{lem:undelta-app}
  \begin{enumerate}
    \item $\G,\xo; \D \vdash e : \vp$ iff $\G,\x[\cdot]; \D \vdash e :
      \vp$
    \item $\G,\xo;\D \vdash_{alt} \rho \rightarrow e
      :^z_{\Delta_s} \sigma \Rightarrow \rho$ iff
      $\G,\x[\cdot];\D \vdash_{alt} \rho \rightarrow e
   :^z_{\Delta_s} \sigma \Rightarrow \rho$
  \end{enumerate}
\end{lemma}
\begin{proof}
Straightforward induction on the structure of the given derivation,
noting that rules $\mathit{Var}_\Delta$ and $\mathit{Var}_\omega$
coincide when $\Delta$ is empty.
\end{proof}

\undelta*

\begin{proof}
By Lemma~\ref{lem:undelta-app}.
\end{proof}

\subsection{Irrelevance\label{sec:proof:irrelevance}}

\input{language-v4/proofs/WHNFConvSoundness}

\subsection{Substitution Lemmas\label{sec:proof:substitution-lemmas}}

\input{language-v4/proofs/LinearSubstitutionLemma}

\input{language-v4/proofs/UnrestrictedSubstitutionLemma}

\input{language-v4/proofs/DeltaSubstitutionLemma}

\subsection{Type Preservation\label{sec:proof:type-preservation}}

% \input{language-v4/proofs/TypePreservationTheorem}
\begin{theorem}[Type Preservation]\label{thm:typepres}
  Let $\G ; \D \vdash (\Theta \mid e) \Downarrow (\Theta' \mid v) :
  \tau , \Sigma$ or $\G ; \D \vdash (\Theta \mid e) \Downarrow^* (\Theta' \mid v) :
  \tau , \Sigma$.
  If $\G ; \D \vdash \Theta \mid e : \tau , \Sigma$ then $\G ; \D \vdash \Theta' \mid v : \tau , \Sigma$

\end{theorem}

\begin{proof}
By structural induction on the instrumented operational semantics. We
omit many trivial cases.
\begin{description}
    \item[Case:] $\Gamma; \Delta \vdash (\Theta, \xl = y \mid x) \Downarrow (\Theta' \mid v) : \s, \Sigma$
\begin{tabbing}
    (1) $\G; \D \vdash \Theta, \xl = y \mid x : \s, \Sigma$\\
    (2) $\G, \Theta{\uparrow}^\omega; \D, \Theta{\uparrow}^1, \xl \vdash (x, y, \Theta{\downarrow}^1_e, \Sigma) : \s \otimes \s \bigotimes(\Theta{\downarrow^1_e}) \bigotimes(\Sigma)$\`by \ref{lem:unwrapenv}\\
    (3) $Let~\G';\D_1,\D_2,\D_3 = \G, \Theta{\uparrow}^\omega; \D, \Theta{\uparrow}^1$\` s.t. $(\D_2,\D_3)$ types $(\Theta{\downarrow}^1_e,\Sigma)$\\
    (4) $\G'; \xl \vdash x : \s$\\
    (5) $\G'; \D_1 \vdash y : \s$\\
    (6) $\G'; \D_1, \D_2, \D_3 \vdash (y, \Theta{\downarrow}^1_e, \Sigma) : \s \bigotimes(\Theta{\downarrow^1_e}) \bigotimes(\Sigma)$\\
    (7) $\G; \D \vdash \Theta \mid y : \s, \Sigma$\`by well-typed state\\
    (8) $\G; \D \vdash \Theta' \mid v : \s, \Sigma$\` by induction hypothesis\\
\end{tabbing}
    \item[Case:] $\Gamma; \Delta \vdash (\Theta, x{:}_{\D'}\s = e \mid x) \Downarrow (\Theta', x{:}_{\D_a}\s = v \mid v) : \s, \Sigma$
\begin{tabbing}
    (1) $\Gamma; \Delta \vdash (\Theta, x{:}_{\D'}\s = e \mid x) : \s, \Sigma$\\
    (2) $\G, \Theta{\uparrow}^\omega; \D, \Theta{\uparrow}^1 \vdash \llet{x{:}_{\D'} = e}{(x, \Theta{\downarrow}^1_e, \Sigma)} : \s \bigotimes(\Theta{\downarrow}^1_\s) \bigotimes(\Sigma)$\`by \ref{lem:unwrapenv}\\
    (3) $Let~\D',\D'' = \D,\Theta{\uparrow}^1$\\
    (4) $\G, \Theta{\uparrow}^\omega; \D' \vdash e : \s$\`by inv on let (2)\\
    (5) $\G, \Theta{\uparrow}^\omega; \D', \D'' \vdash (x, \Theta{\downarrow}^1_e, \Sigma) : \s \bigotimes(\Theta{\downarrow}^1_\s) \bigotimes(\Sigma)$\`by inv on let (2)\\
    (6) $\G, \Theta{\uparrow}^\omega; \D' \vdash x : \s$\\
    (7) $\G, \Theta{\uparrow}^\omega; \D'' \vdash (\Theta{\downarrow}^1_e, \Sigma) : \bigotimes(\Theta{\downarrow}^1_\s) \bigotimes(\Sigma)$\\
    (8) $\G, \Theta{\uparrow}^\omega; \D', \D'' \vdash (e, \Theta{\downarrow}^1_e, \Sigma) : \tau \bigotimes(\Theta{\downarrow}^1_\s) \bigotimes(\Sigma)$\` by pair (4,7)\\
    (9) $\G; \D \vdash \Theta \mid e : \tau,\Sigma$\` by well-typed state\\
    (10) $\G; \D \vdash \Theta \mid v : \tau,\Sigma$\` by induction hypothesis\\
    (11) $Let~\G';\D_a,\D_b = \G, \Theta'{\uparrow}^\omega, \D, \Theta'{\uparrow}^1$\\
    (11a) $\G'; \D_a \vdash v : \s$
    (11b) $\G'; \D_b \vdash (\Theta'{\downarrow}^1_e, \Sigma) : \bigotimes(\Theta'{\downarrow}^1_\s) \bigotimes(\Sigma)$\\
    (12) $\G'; \D_a \vdash \llet{x{:}_{\D_a}\s = v}{v} : \s$\`by let (11a,11a)\\
    (13) $\G; \D \vdash \Theta' \mid \llet{x{:}_{\D_a}\s = v}{v} : \s, \Sigma$\`by well-typed state\\
    (14) $\G; \D \vdash \Theta', x{:}_{\D_a}\s = v \mid v : \s, \Sigma$\`by env. expansion\\
\end{tabbing}
Note: It's slightly more elaborate to accommodate recursion. The runtime
environment bindings must be tagged with the recursive group such that the
environment expansion can decide whether to unwrap to a letrec group or
not. Unlike the case proved here, when the $\D$-binding belongs to a let group,
the whole group is unwrapped and we can invert the assumption to get an
$e$ typed with the $x$s from the group.\\
\item[Case:] $\Gamma; \Delta \vdash (\Theta \mid \ccase{e}{z{:}_{\Delta'}\s~\{K~\ov{y_i}\to e'}\}) \Downarrow (\Theta'' \mid v) : \tau$
\begin{tabbing}
    (1) $\Gamma; \Delta \vdash \Theta \mid \ccase{e}{z{:}_{\Delta_a}\s~\{K~\ov{y_i}\to e'}\} : \tau, \Sigma$\\
    (2) $\G, \Theta{\uparrow}^\omega; \D, \Theta{\uparrow}^1 \vdash (\ccase{e}{z{:}_{\Delta_a}\s~\{K~\ov{y_i}\to e'}\}, \Theta{\downarrow}^1_e, \Sigma) : \tau \otimes \bigotimes(\Theta{\downarrow}^1_\sigma) \otimes \bigotimes(\Sigma)$\`by \ref{lem:unwrapenv}\\
    (3) $Let~\G' = \G, \Theta\uparrow^\omega; \D_a,\D_b,\D_2,\D_3 = \D,\Theta\uparrow^1$\`s.t. $(\D_2,\D_3)$ types $(\Theta\downarrow^1_e, \Sigma)$\\
    (4) $\G'; \D_a, \D_b \vdash \ccase{e}{z{:}_{\D_a}\s\{K~\ov{y_i}\to e'\}} : \tau$\\
    Subcase $e$ is in WHNF (thus $= K~\ov{x_i}$)\\
    (5) $\G'; \D_a \Vdash K~\ov{x_i} : \s \gtrdot \ov{\D_i}$\` by subcase\\
    (6) $\G', z{:}_{\D_a}\s; \D_a, \D_b \vdash_{alt} K~\ov{y_i} \to e' :^z_{\D_i} \s \Rightarrow \tau$\` by inv\\
    Subsubcase $\vdash_{alt}$ $Alt_{WHNF}$\\
    (7) $\G', z{:}_{\D_a}\s, \ov{y_i{:}_\omega\s_i}, \ov{y_j{:}_{\Delta_j}\s_j}; \D_a, \D_b \vdash e' : \tau$\` by inv\\
    (8) $\Theta' = \Theta$\` in this subcase because $(\Theta \mid K~\ov{x_i})\Downarrow(\Theta \mid K~\ov{x_i})$\\
    (9) $\G'; \ov{\D_i} \vdash K~\ov{x_i} : \s$\` by inv (5)\\
    (10) $\ov{\G'; \cdot \vdash x_i : \s_i}$\` by inv (9) for unrestricted $x$s\\
    (11) $\ov{\G'; \D_j \vdash x_j : \s_j}$\` by inv (9) for linear fields $x$s\\
    (12) $\G', z{:}_{\D_a}\s; \D_a, \D_b \vdash e'\ov{[x_{i,j}/y_{i,j}] : \tau}$\` by $\D$-subst and $\omega$-subst lemmas (7,10,11)\\
    (13) $\G'; \D_a, \D_b \vdash e'\ov{[x_{i,j}/y_{i,j}]}[K~\ov{x_{i,j}}/z] : \tau$\` by $\D$-subst lemma and $\ov{\D_i} = \D_a$ (9,12)\\
    (14) $\G; \D \vdash \Theta \mid e'\ov{[x_{i,j}/y_{i,j}]}[K~\ov{x_{i,j}}/z] : \tau, \Sigma$\` by well-typed state\\
    (15) $\G; \D \vdash \Theta'' \mid v : \tau, \Sigma$\` by induction hypothesis\\
    Subsubcase $\vdash_{alt}$ $Alt_{0}$\\
    (7) $(\G', z{:}_{\D_a}\s)[\cdot/\D_a]_z, \ov{y_i{:}_\omega\sigma_i}; (\D_a, \D_b)[\cdot / \D_a] \vdash e' : \tau$\` by inv subcase\\
    (8) $\G', z{:}_{\cdot}, \ov{y_i{:}_\omega\sigma_i}; \D_b \vdash e' : \tau$\` by $z$ cannot occur in $\G'$\\
    (9) Since $\G';\D_a \Vdash K~\ov{x_i}$, in this subcase $\D_a = \cdot$\\
    (10) $\G', \z{:}_\cdot\sigma, \ov{y{:}_\omega\sigma}; \D_b \vdash e' : \tau$\` by (9)\\
    (11) $\G', \z{:}_\cdot\sigma; \D_b \vdash e'\ov{[x_i/y_i]} : \tau$\` by $\omega$-subst (10)\\
    (12) $\G'; \D_b \vdash e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] : \tau$\` by $\D$-subst\\
    (13) $\G; \D \vdash \Theta \mid e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] : \tau, \Sigma$\` by well-typed state\\
    (14) $\G; \D \vdash \Theta'' \mid v : \tau, \Sigma$\`by induction hypothesis\\
    Subcase $e$ is not in WHNF\\
    (5) $\G'; \D_a \vdash e : \s$\` by inv on case\\
    (6) $\G', z{:}_{\irr{\D_a}}\s; \irr{\D_a}, \D_b \vdash_{alt} K~\ov{y_{i,j}} \to e' :^z_{\irr{\D_a}} \s $\` by inv on case\\
    Subsubcase $\vdash_{alt}$ $Alt_{NWHNF}$\\
    (7) $\G', z{:}_{\irr{\D_a}}\s, \ov{y_i{:}_\omega\s_i}, \ov{y_j{:}_{\Delta_j}\s_j}; \irr{\D_a}, \D_b \vdash e' : \tau$\` for \ov{\D_j = \lctag{\irr{\D_a}}{K_j}}, by inv.\\
    (8) $\G', z{:}_{\ov{y_j{:}_1\s_j}}\s, \ov{y_i{:}_\omega\s_i}; \ov{y_j{:}_1\s_j}, \D_b \vdash e' : \tau$\` by $\D$-bound to linear\\
    (9) $\G', z{:}_{\ov{y_j{:}_1\s_j}}\s, \ov{y_i{:}_\omega\s_i}; \D_a, \D_b, \ov{y_j{:}_1\s_j} \vdash (e, e') : \sigma \otimes \tau$\` by pair (5,8)\\
    (10) $\G, z{:}_{\ov{y_j{:}_1\s_j}}\s, \ov{y_i{:}_\omega\s_i}; \D, \ov{y_j{:}_1\s_j} \vdash \Theta \mid e : \s, e':\tau, \Sigma$\`by well-typed state\\
    (11) $\G, z{:}_{\ov{y_j{:}_1\s_j}}\s, \ov{y_i{:}_\omega\s_i}; \D, \ov{y_j{:}_1\s_j} \vdash \Theta' \mid K~\ov{x_i} : \s, e':\tau, \Sigma$\`by induction hypothesis\\
    (12) $\G, \Theta'{\uparrow}^\omega, \ov{y_i{:}_\omega\s_i}; \D, \ov{y_j{:}_1\s_j}, \Theta'{\uparrow}^1 \vdash (K~\ov{x_{i,j}}, \Theta'{\downarrow}^1_e, e', \Sigma) : \s \bigotimes(\Theta'\downarrow^1_\sigma) \otimes \tau \bigotimes(\Sigma)$\`by \ref{lem:unwrapenv}\\
    (13) $Let~\G''=\G,\Theta'{\uparrow}^\omega~and~\D_c\D_d\D_4\D_5=\D,\Theta'{\uparrow}^1$\` where $(\D_4, \D_5)$ types $(\Theta'{\downarrow^1_e}, \Sigma)$\\
    (13a) $\G''; \D_c \vdash K~\ov{x_{i,j}} : \s$\\
    (13b) $\G'', z{:}_{\ov{y_j{:}_1\s_j}}\s, \ov{y_i{:}_\omega\s_i}; \D_d, \ov{y_j{:}_1\s_j} \vdash e' : \tau$\\
    (14) $\ov{\G''; \cdot \vdash x_i : \s_i}$\\
    (15) $\ov{\G''; \D_j \vdash x_j : \s_j}$\` for $\ov{D_j}=\D_c$, by inv\\
    (16) $\G'', z{:}_{\D_c}\s; \D_c, \D_d \vdash e'\ov{[x_{i,j}/y_{i,j}]} : \tau$\`by $1$-substs and $\omega$-substs lemmas\\
    (17) $\G''; \D_c, \D_d \vdash e'\ov{[x_{i,j}/y_{i,j}]}[K~\ov{x_{i,j}}/z] : \tau$\`by $\D$-subst lemma (13a,16)\\
    (18) $\G; \D \vdash \Theta' \mid e'\ov{[x_{i,j}/y_{i,j}]}[K~\ov{x_{i,j}}/z] : \tau, \Sigma$\`by well-typed state\\
    (19) $\G; \D \vdash \Theta'' \mid v : \tau, \Sigma$\`by induction hypothesis\\
    Subsubcase $\vdash_{alt}$ $Alt_0$\\
    (7) $\G', z{:}_\cdot\s, \ov{y_i{:}_\omega\s_i}; \D_b \vdash e' : \tau$\` by inv\\
    (8) $\G', z{:}_\cdot\s, \ov{y_i{:}_\omega\s_i}; \D_a, \D_b \vdash (e, e') : \sigma \otimes \tau$\` by pair (5,7)\\
    (9) $\G, z{:}_\cdot\s, \ov{y_i{:}_\omega\s_i}; \D \vdash \Theta \mid e : \sigma, e' : \tau, \Sigma$\` by well-typed state\\
    % (10) $\G, z{:}_\cdot\s, \ov{y_i{:}_\omega\s_i}; \D \vdash \Theta \mid e : \sigma, e' : \tau, \Sigma$\` by reconstructing\\
    (10) $\G, z{:}_\cdot\s, \ov{y_i{:}_\omega\s_i}; \D \vdash \Theta' \mid K~\ov{x_i} : \sigma, e' : \tau, \Sigma$\` by induction hypothesis\\
    (11) $Let~\G'' = \G,\Theta'{\uparrow}^\omega~and~\D_d,\D_4,\D_5 = \D, \Theta'{\uparrow}^1$\` where $(\D_4, \D_5)$ types $(\Theta'{\downarrow^1_e}, \Sigma)$\\
    (11a) $\G''; \cdot \vdash K~\ov{x_i} : \s$\`since $K$ has no linear fields in this subcase\\
    (11b) $\G'', z{:}_{\cdot}\s, \ov{y_i{:}_\omega\s_i}; \D_d \vdash e' : \tau$\\
    (12) $\ov{\G''; \cdot \vdash x_{i} : \s_i}$\`by inv (11a)\\
    (13) $\G'', z{:}_{\cdot}\s; \D_d \vdash e'\ov{[x_i/y_i]} : \tau$\` by $\omega$-subst lemma\\
    (14) $\G''; \D_d \vdash e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] : \tau$\` by $\D$-subst lemma\\
    (15) $\G; \D \vdash \Theta' \mid e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] :
    \tau, \Sigma$\` by well-typed state\\
    (16) $\G; \D \vdash \Theta'' \mid v : \tau, \Sigma$\` by induction hypothesis\\
\end{tabbing}
\item[Case:] $\Gamma; \Delta \vdash (\Theta \mid \ccase{e}{z{:}_{\Delta'}\s~\{\_ \to e'}\}) \Downarrow (\Theta'' \mid v) : \tau$
\begin{tabbing}
    (1) $\Gamma; \Delta \vdash \Theta \mid \ccase{e}{z{:}_{\Delta_a}\s~\{\_ \to e'}\} : \tau, \Sigma$\\
    (2) $\G, \Theta{\uparrow}^\omega; \D, \Theta{\uparrow}^1 \vdash (\ccase{e}{z{:}_{\Delta_a}\s~\{\_ \to e'}\}, \Theta{\downarrow}^1_e, \Sigma) : \tau \otimes \bigotimes(\Theta{\downarrow}^1_\sigma) \otimes \bigotimes(\Sigma)$\`by \ref{lem:unwrapenv}\\
    (3) $Let~\G' = \G, \Theta\uparrow^\omega; \D_a,\D_b,\D_2,\D_3 = \D,\Theta\uparrow^1$\`s.t. $(\D_2,\D_3)$ types $(\Theta\downarrow^1_e, \Sigma)$\\
    (4) $\G'; \D_a, \D_b \vdash \ccase{e}{z{:}_{\D_a}\s\{\_ \to e'\}} : \tau$\\
    (5) $\G'; \D_a \vdash e : \sigma$\\
    Subcase $e$ is in WHNF\\
    (-) Similar to the same subcase in the case-with-pattern-vars reduction,\\
        except only $Alt_{\_}$ is relevant which is far simpler\\
    Subcase $e$ is not in WHNF\\
    (6) $\G', z{:}_{\irr{\D_a}}\s; \irr{\D_a}, \D_b \vdash e' : \tau$\\
    (7) $\G'; \D_b, z{:}_1\s \vdash e' : \tau$\` by $\D$-bound to linear\\
    (8) $\G'; \D_a, \D_b, z{:}_1\s \vdash (e, e') : \sigma \otimes \tau$\` by pair\\
    (9) $\G; \D, z{:}_1\s \vdash \Theta \mid e : \sigma, e' : \tau,
    \Sigma$\` by well-typed state\\
    (10) $\G; \D, z{:}_1\s \vdash \Theta' \mid K~\ov{x_i} : \sigma, e' : \tau, \Sigma$\` by induction hypothesis\\
    (11) $Let~\G';\D_c,\D_d,\D_4,\D_5 = \G, \Theta'{\uparrow}^\omega; \Delta, \Theta'{\uparrow}^1$\`s.t. $(\D_4,\D_5)$ types $(\Theta'{\downarrow}^1_e, \Sigma)$\\
    (12) $\G';\D_c \vdash K~\ov{x_i} : \s$\\
    (13) $\G';\D_d,z{:}_1\s \vdash e' : \tau$\\
    (14) $\G';\D_c, \D_d \vdash e'[K~\ov{x_i}/z] : \tau$\` by $1$-subst lemma\\
    (15) $\G;\D \vdash \Theta' \mid e'[K~\ov{x_i}/z] : \tau, \Sigma$\` by well-typed state\\
    (16) $\G;\D \vdash \Theta'' \mid v : \tau, \Sigma$\` by induction hypothesis\\
\end{tabbing}
\end{description}
\end{proof}

We also derive type preservation for the lazy natural semantics.

  \begin{theorem}[Type Preservation]
For any well-typed $[\Theta \mid e]$, if $\Theta : e \Downarrow
\Theta' : e'$ or $\Theta : e \Downarrow^* \Theta' : e'$ then we have
that $[\Theta \mid e']$ is well-typed.
\end{theorem}
\begin{proof}
By Lemma~\ref{lem:naturaltoinstr} and Theorem~\ref{thm:typepres}.
\end{proof}

\subsection{Progress\label{sec:proof:progress}}

%\input{language-v4/proofs/ProgressTheorem}

\begin{theorem}[Progress]\label{thm:prog}
Let $\G ; \D \vdash \Theta \mid e : \tau,\Sigma$. Any partial derivation $\G; \D \vdash (\Theta \mid e)
\Downarrow?,\Sigma$ can be extended. 
\end{theorem}

\begin{proof}
 Essentially all cases are standard. The only potential issue is the
 (linear) variable rule.
It suffices to show that whenever a variable is evaluated it is
present in the evaluation environment. This situation is impossible
since the state
$\G ; \D \vdash \Theta \mid x : \tau , \Sigma$ where $x$ is absent
from $\Theta$ is not well-typed and so cannot be the head of a partial
derivation. 
\end{proof}

\begin{theorem}[Progress]
For any partial derivation of $\Theta : e \Downarrow?$, for $[\Theta |
e]$ well-typed, the derivation can be extended.
\end{theorem}
\begin{proof}
Follows from Lemma~\ref{lem:instrtonatural} and Theorem~\ref{thm:prog}.
 \end{proof}

\section{Optimisations preserve linearity}\label{app:optimisations}

\subsection{Inlining}
\input{language-v4/proofs/optimizations/Inlining}

\subsection{\texorpdfstring{$\beta$}{Beta}-reduction}

\input{language-v4/proofs/optimizations/BetaReduction}

\subsection{Case of known constructor}

\input{language-v4/proofs/optimizations/CaseOfKnownConstructor}

\subsection{Case of Case\label{sec:proof:caseofcase}}

The case of case transformation applies to case expressions whose scrutinee is
another case expression, and returns the innermost case expression transformed by repeating
the outermost case expression in each alternative of the innermost case,
scrutinizing the original alternative body.

Intuitively, since the scrutinee of the outermost case is not in WHNF, no
resources from it can directly occur in the outermost alternatives. By moving
the outermost alternatives inwards with a different scrutinee, the alternatives
remain well-typed because they are typed using either the case binder or the
pattern bound variables, which, by the \emph{Irrelevance} lemma, makes it
well-typed for any scrutinee consuming arbitrary resources.

\input{language-v4/proofs/optimizations/CaseOfCase}

\subsection{Commuting let}

\input{language-v4/proofs/optimizations/LetFloating}

\subsection{\texorpdfstring{$\eta$}{Eta}-conversions}

\input{language-v4/proofs/optimizations/EtaConvs}

\subsection{Binder Swap}

The binder swap transformation applies to case expressions whose scrutinee is a
single variable $x$, and it substitutes occurrences of $x$ in the case
alternatives for the case binder $z$. If $x$ is a linear resource, $x$ cannot
occur in the case alternatives (as we conservatively consider variables are not
in WHNF), so the substitution preserves types vacuously. Otherwise, $x$ can be
freely substituted by $z$, since $z$ is also an unrestricted resource (it's
usage environment is empty because $x$ is unrestricted).

\input{language-v4/proofs/optimizations/BinderSwap}

% G ; D |- (Theta | e x) \Downarrow (Theta'' | v) : tau, Sigma

% QUEREMOS PROVAR:
% Se
% G ; D |- (Theta | e x) : tau , Sigma
% entao
% G ; D |- (Theta'' | v) : tau, Sigma

% por i.h.
% Se G;D |- (Theta | e) : sigma -> tau, x:tau , Sigma
% entao
% G ; D |- (Theta' | \y:sigma.e') : sigma->tau, x:tau,Sigma

% por i.h.
% Se G ; D |- (Theta' , y:sigma = x | e') : tau, Sigma
% entao
% G;D |- (Theta'' | v) : tau , Sigma

% Gamma ; Delta |- [Theta | (e x , terms(Sigma))] : (tau , types(Sigma))
% Gamma ; Delta |- [Theta | (e , terms(x:tau,Sigma))] : (tau->sigma , types(x:tau,Sigma))

% Daqui Gamma ; Delta |- [Theta' | (\y:_1 sigma.e' , terms(x:tau,Sigma))] : (sigma->tau, types(x:tau,Sigma))
% tirar Gamma ; Delta |- [Theta',y:sigma = x | (e' , terms(Sigma))] : (sigma->tau, types(Sigma))

% [Theta' , x_1 : tau = y | e] = (\x_1:tau . [Theta' | e] ) y
% [ Theta' , x_Delta:tau = e' | e] = let x_Delta = e' in [Theta' | e]
% [ . | (e,e')] = (e,e')

% Gamma ; Delta |- [Theta | (e , terms(Sigma))] : (tau , types(Sigma))

 % Gamma ; Delta |- Theta | e  : tau, Sigma

% }}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

% TODO: - In the letrec case: congratulations on finding an inference algorithm
% (though I will want to see the proof some day, I don't see why it works yet).
% You should clarify that, during linting, you will have a usage environment
% annotation and won't need to run the inference algorithm. This algorithm is
% only needed when you first create a letrec.

% vim: fen fdm=marker

### 2025-10-08 popl26.tex
anonymous; affiliation

@@
-\documentclass[acmsmall,review,anonymous,screen]{acmart}
+\documentclass[acmsmall,review,screen]{acmart}
 \PassOptionsToPackage{dvipsnames}{xcolor}
@@
    \institution{Well-Typed LLP}
-  %\city{City}
-  %\country{Something}
+   % \city{City}
+   \country{United Kingdom}
  }
@@
 \email{btoninho@fct.unl.pt}
-%% TODO: AFFILIATION, EMAIL
-%% TODO: CCS CONCEPTS and KEYWORDS
-%% TODO: ACM RIGHTS notice...
 % CHECK: https://www.acm.org/binaries/content/assets/publications/taps/latex-best_practices-06-may-2020.pdf
@@
 ours, but does not account for pattern matching.
-More recently, \cite{DBLP:journals/pacmpl/ChoudhuryEEW21} and
-\cite{cite:linearhaskell} both make use of usage-aware
-semantics. Choudhury et al. do so in the context of graded dependent
+More recently, \cite{DBLP:journals/pacmpl/ChoudhuryEEW21,DBLP:conf/esop/MarshallVO22} and
+\cite{cite:linearhaskell} all make use of usage-aware
+semantics. \citet{DBLP:journals/pacmpl/ChoudhuryEEW21} do so in the context of graded dependent
 type theory, using the usage-aware heap-based operational semantics to
@@
 in ours, but they do not model
-call-by-need. \citet{cite:linearhaskell} provide a similar semantics
+call-by-need, only call-by-name. The work of \cite{DBLP:conf/esop/MarshallVO22} is
+applies the ideas of \cite{DBLP:journals/pacmpl/ChoudhuryEEW21}
+to a non-graded (also call-by-name) setting. \citet{cite:linearhaskell} provide a similar semantics
 to ours, but present only a naive linear typing system that does not

### 2025-10-13 popl26.tex
started minor points up to 3.3

@@
 }
+
+ \newcommand{\ROUNDTWO}[1]{{\color{red}#1}}
+%\newcommand{\ROUNDTWO}[1]{#1}
 
@@
 %
-Additionally, we show our system
+Additionally, we \ROUNDTWO{that} show our system
 %
@@
 
-\item We prove our type system is sound, guarantees linear resource usage, and
+\item We prove that our type system is sound, guarantees linear resource usage, and
     prove that multiple optimising transformations, which are rejected in Core,
@@
 To determine whether the usage of a binding from a recursive group is linear,
-we must infer the superset of resources that may end up being consumed in a
+we must \ROUNDTWO{check} the superset of resources that may end up being consumed in a
 single use of any of the bindings. For a strongly connected recursive group,
@@
 %
-It follows from the inferred resources used being the same for all bindings
+It follows from the resources used being the same for all bindings
 that at most a single variable from the recursive group can be used in the
@@
 Linear Core makes precise the various insights discussed in the
-previous section in a linear type system that we prove is sound and guarantees
+previous section in a linear type system that we prove \ROUNDTWO{to be} sound and guarantee
 linear resource usage at runtime.
@@
 call-by-need big-step operational semantics that becomes stuck when linear
-variables in the runtime heap are used more than once. We show this
+variables in the runtime heap are used more than once, \ROUNDTWO{a
+  technical approach that is natural in the literature \cite{}}. We show this
 semantics to be bisimilar to a
@@
 expressions, branching on whether the scrutinee is in Weak Head Normal
-Form or not, and using ``proof irrelevance'' to track resources that are no
+Form or not, and using  \ROUNDTWO{irrelevance annotations} to track resources that are no
 longer available but have yet to be fully consumed
@@
 We note that we handle multiplicity polymorphism differently from
-Linear Haskell by ignoring the multiplicity semiring and
-conservatively treating all multiplicity polymorphic functions as
-linear, for the sake of simplicity.
+Linear Haskell by \ROUNDTWO{effectively} ignoring the multiplicity semiring and
+treating all multiplicity polymorphic \ROUNDTWO{variables}
+\ROUNDTWO{parametrically}, for the sake of simplicity. \ROUNDTWO{Such
+  variables cannot be duplicated in our system.}
 %
@@
 Occurrences of unrestricted variables from $\G$ are well-typed as long as the linear
-environment is empty, while occurrences of linear variables are only well-typed
+environment is empty, while occurrences of linear \ROUNDTWO{(or multiplicity polymorphic)} variables
+are only well-typed
 when the variable being typed is the only resource available in the
-linear context (rules $Var_\omega$ and $Var_1$, respectively).
+linear context (rules $Var_\omega$, $Var_1$ \ROUNDTWO{and $Var_p$}, respectively).
 % \[
@@
 % \]
-In both cases, the linear context must contain exactly what is required to
-prove the proposition, whereas the unrestricted context may contain arbitrary
+In \ROUNDTWO{the $Var_1$ and $Var_p$ cases}, the linear context must contain exactly \ROUNDTWO{the
+  appropriate variable binding}, whereas the unrestricted context may contain arbitrary
 variables.
@@
 and construct expressions of type $\forall p.~\s$, i.e. a type universally
-quantified over a multiplicity variable $p$. Again, conservatively, in the body
-of the abstraction, function types annotated with a $p$ variable and datatype
-fields with multiplicity $p$ are typed as though they are linear functions and
-linear fields, because $p$ can be instantiated at both $\omega$ and $1$.
+quantified over a multiplicity variable $p$. \ROUNDTWO{In the body
+of the abstraction, function types and datatype fields annotated with the multiplicity
+variable $p$ are typed parametrically in their multiplicity, since} $p$ can be instantiated at both $\omega$ and $1$.
 %\[
@@
 A \emph{usage environment} $\Delta$ encodes the idea that lazy
-bindings do not consume resources upon definition, but rather when the bindings themselves are consumed.
-Specifically, we annotate so-called $\Delta$-bound variables with a \emph{usage
-environment} to denote that consuming these $\D$-variables equates to consuming the
-resources in the annotated usage environment $\D$ (which is essentially a
-multiset of linear resources).
-%
-$\Delta$-bound variables are introduced in the unrestricted environment by
+\ROUNDTWO{variable} bindings do not consume resources upon definition, but rather when the bindings themselves are consumed.
+Specifically, we annotate so-called $\Delta$-bound variables with \ROUNDTWO{their} \emph{usage
+environment} \ROUNDTWO{$\Delta$} to denote that consuming
+\ROUNDTWO{such variables must be equated with} consuming the
+resources in the annotated \ROUNDTWO{typing} environment.
+%
+\ROUNDTWO{Such} variables are introduced in the unrestricted environment by
 several constructs such as (recursive) let binders, case binders, and case

### 2025-10-13 popl26.tex
fig 3 tweak

@@
 \ROUNDTWO{Such} variables are introduced in the unrestricted environment by
-several constructs such as (recursive) let binders, case binders, and case
+\ROUNDTWO{multiple} constructs such as (recursive) let binders, case binders, and case
 pattern variables.
 %
-For example, the let-binder $u$ in the following program is annotated with a
+For example, the let-bound variable $u$ in the following program is annotated with a
 usage environment $\{x{:}_1\s,y{:}_1\s\}$ tracking the resources that are used in its

### 2025-10-13 popl26.tex
fig 4

@@
 
-\title{Semantic Linearity in a Lazy Optimising Compiler}
-\subtitle{Type-checking Linearity in GHC Core}
+%\title{Semantic linearity in a lazy optimizing compiler]
+\title{Effectively Checking Semantic Linearity in a Lazy Functional Language}
+%\subtitle{Type-checking Linearity in GHC Core}
 
@@
 \affiliation{
-  \department{Departamento de Inform\'{a}tica}
-  \institution{NOVA-LINCS, FCT-NOVA, Universidade Nova de Lisboa}
+  \department{Departamento de Engenharia Inform\'{a}tica}
+  \institution{Instituto Superior Técnico, Universidade de Lisboa and INESC-ID}
   %\city{City}
   \country{Portugal}}
-\email{btoninho@fct.unl.pt}
+\email{bernardo.toninho@tecnico.ulisboa.pt}
 % CHECK: https://www.acm.org/binaries/content/assets/publications/taps/latex-best_practices-06-may-2020.pdf
@@
 \[
-f = \lambda \xl.~\lambda \y[1].~\llet{u_{\{x:_1\s,y:_1\s\}} = (x,y)}{e}
+f = \lambda \x:_1\tau.~\lambda \y[1].~\llet{u_{\{x:_1\tau,y:_1\s\}} = (x,y)}{e}
 \]
@@
 $\Delta$-bound variable can be left unused since it is unrestricted, regardless
-of the variables it is annotated with.
+of the variables it is annotated with. \ROUNDTWO{In the example above,
+  our type system ensures that \emph{either} $x$ and $y$ are consumed
+  or $u$ is consumed in $e$, since $u$ has usage environment $\{x:_1\tau,y:_1\s\}$.}
 
@@
 
-A $\Delta$-bound variable $u$ is a variable annotated with a usage environment $\Delta$. Crucially, for any $\Delta$-bound variable $u$:
+A $\Delta$-bound variable $u$ is a variable annotated with
+\ROUNDTWO{some} usage environment \ROUNDTWO{$\Delta'$}.
+\ROUNDTWO{For conciseness, we use the meta-variable $\Delta$ to refer to the usage
+  environment of a $\Delta$-bound variable.}
+Crucially, for any $\Delta$-bound variable $u$:
 %
@@
 Similarly, (3) also follows from (1), because if the linear resources aren't
-consumed in the $\Delta$-var derivation, they must be consumed in an
-alternative derivation (or otherwise the expression is ill-typed).
+consumed \ROUNDTWO{by a use of $u$}, they must be consumed in 
+some other derivation (or otherwise the \ROUNDTWO{overall} expression is ill-typed).
 %

### 2025-10-13 popl26.tex
.

@@
 %unrestricted $\G$ and a linear $\D$ environment.
-The main judgement is written $\G;\D \vdash e : \s$ to denote that expression
-$e$ has type $\s$ under the unrestricted environment $\G$ and linear
+The main judgement is written $\G;\D \vdash e : \tau$ to denote that expression
+$e$ has type $\tau$ under the unrestricted environment $\G$ and linear
 environment $\D$.
@@
 %
-These observations all lead to the typing rule for $\Delta$-bound
+These observations lead to the typing rule for $\Delta$-bound
 variables:
@@
 %
-$\D$-vars are unrestricted and can be discarded and duplicated, despite
-multiple occurrences of the same $\Delta$-var in an expression not possibly being well-typed
-since that would imply non-linear usage of linear resources.
+\ROUNDTWO{$\Delta$-bound variables} are unrestricted and can be discarded and duplicated, despite
+multiple occurrences of the \ROUNDTWO{variable} in an expression not
+being well-typed
+\ROUNDTWO{due to} non-linear usage of linear resources.
 % multiple occurrences of the same

### 2025-10-14 popl26.tex
popl: Update abstract
And title and orcid

@@
 %\title{Semantic linearity in a lazy optimizing compiler]
-\title{Effectively Checking Semantic Linearity in a Lazy Functional Language}
+\title{Lazy Linearity for a Core Functional Language}
 %\subtitle{Type-checking Linearity in GHC Core}
@@
 \author{Rodrigo Mesquita}
+\orcid{0009-0005-1467-115X}
 %\affiliation{
@@
 \begin{abstract}
-Linear type systems can be used to guarantee (linear) resources are used exactly once.
-Traditionally, using a resource is synonymous with its \emph{syntactic}
-occurrence in the program, however, under the lens of non-strict evaluation,
-linearity can be further understood \emph{semantically}, where a
-syntactic occurrence of a resource does not necessarily entail
+Traditionally, \ROUNDTWO{in linearly typed languages, consuming} a linear resource is synonymous
+with its syntactic occurrence in the program. However, under the lens of
+non-strict evaluation, linearity can be further understood semantically,
+where a syntactic occurrence of a resource does not necessarily entail
 using that resource when the program is executed.
-Semantic linearity becomes especially vital when considering the
-internal language of an optimising compiler, where optimisations
-heavily rewrite the source program in ways that can often break
-syntactic linearity but that aim to preserve the program's semantics.
+%
+\ROUNDTWO{While this distinction is largely unexplored, it becomes unavoidable
+    for} Haskell's optimising compiler, which heavily rewrites the source
+    program in ways that break syntactic linearity but preserve the program's
+    semantics.
+% \ROUNDTWO{This distinction becomes vital in the Haskell optimising compiler,} where
+% optimisations heavily rewrite the source program in ways that break syntactic
+% linearity but preserve the program's semantics
+%
+% This distinction is 
+% \ROUNDTWO{This distinction is critical for the Haskell optimising compiler, 
+%
+% when considering the internal language of an optimising compiler, where
+% optimisations heavily rewrite the source program in ways that break syntactic
+% linearity but preserve the program's semantics.
+%
+%
 % %
@@
 %
-
-We present Linear Core, the first type system that understands semantic
-linearity in the presence of non-strict evaluation, suitable for the Core
-intermediate language of the Glasgow Haskell Compiler. We prove Linear Core
-is sound and guarantees linear resource usage, and prove multiple optimising
-transformations preserve linearity in Linear Core while failing to do so in
-Core. We have implemented Linear Core as a compiler plugin to validate the
-system against linearity-heavy libraries, including
+%
+We introduce Linear Core, \ROUNDTWO{a novel system which accepts the lazy
+semantics of linearity statically} and
+% \ROUNDTWO{Linear Core explores the interaction between laziness and linearity}
+is suitable for the Core intermediate language of the
+Glasgow Haskell Compiler. We prove \ROUNDTWO{that} Linear Core is sound,
+guarantees linear resource usage, and \ROUNDTWO{that} multiple
+optimising transformations preserve linearity in Linear Core while failing
+to do so in Core. We have implemented Linear Core as a compiler plugin to
+validate the system against linearity-heavy libraries, including
 \texttt{linear-base}.%^, in the heart of the compiler.

### 2025-10-14 popl26.tex
Up to P2 L72

@@
 
-Linear types~\cite{cite:linear-logic,cite:barberdill} increase the expressiveness
-of type systems by allowing them to enforce desirable usage patterns
-for certain resources (e.g.~a
-file handle). Typically, linear types can be used to enforce resources are used \emph{exactly once}.
-%
-In programming languages with a linear type system, not using certain resources
-or using them twice is flagged as a type error. For instance, linear types can
-statically enforce correct usage of file or socket descriptors, or
-that heap-allocated memory is freed exactly once (i.e.,~leaks and double-frees become
-type errors), or even guarantee deadlock freedom in channel-based communication
-protocols~\cite{10.1007/978-3-642-15375-4_16},
+Linear types~\cite{cite:linear-logic,cite:barberdill} increase the
+expressiveness of type systems by \ROUNDTWO{guaranteeing that certain
+resources} are used \emph{exactly once}.
+%
+In programming languages with a linear type system, \ROUNDTWO{discarding so-called linear} resources,
+or using them twice, is flagged as a type error. Linear types can \ROUNDTWO{be used to}, for instance,
+statically enforce correct usage of file descriptors, that heap-allocated
+memory is freed exactly once \ROUNDTWO{s.t.}~leaks and double-frees become type errors,
+or guarantee deadlock freedom in channel-based communication protocols~\cite{10.1007/978-3-642-15375-4_16},
 % implement safe
@@
 % a linearly and dependently typed language based on Quantitative Type Theory,
-Rust~\cite{10.1145/2692956.2663188}, a language whose
+Rust~\cite{10.1145/2692956.2663188}, whose
 ownership types build on linear types to guarantee memory safety
@@
 Haskell~\cite{cite:linearhaskell}, a pure, functional, and
-\emph{lazy} general purpose programming language.
+\emph{lazy} general purpose language.
 %
@@
 
-Linearity in Haskell stands out from linearity in other
-languages because linear types permeate Haskell down to (its) Core, the
-intermediate language into which Haskell is translated by the Glasgow Haskell Compiler (GHC).
-%
-Core is a minimal, explicitly typed, lazy functional language with
-both linear and unrestricted types, to which multiple
-Core-to-Core optimising transformations are applied during
-compilation. Notably Core does not have a (higher-order or otherwise)
-store.
-%
-Typechecking Core serves as a sanity check to the correction of such
-transformations, since unsound ones are likely to introduce ill-typed
-expressions. %, making them easy to spot.
+Linearity in Haskell stands out from linearity in other languages because
+linear types permeate Haskell down to Core, the intermediate language into
+which Haskell is translated by the Glasgow Haskell Compiler (GHC).
+%
+Core is a minimal, explicitly typed, \ROUNDTWO{pure} lazy functional language with both
+linear and unrestricted types, to which multiple Core-to-Core optimising
+transformations are applied during compilation.
+%
+Notably, Core \ROUNDTWO{is typechecked after each transformation}. This serves
+as a sanity check to the correction of \ROUNDTWO{the implementation} of
+\ROUNDTWO{such} transformations, since unsound ones are likely to introduce
+ill-typed expressions. %, making them easy to spot.
 %
@@
 % including inlining and $\beta$-reduction, to produce more performant programs.
-
+%
 % Linear core actually not so good
@@
 % \emph{declaration}.
-
+%
 % While the current version of Core is linearity-aware (i.e.,~Core has so-called
@@
 %
-
-Despite the advantages of adding linear types to Core, the status quo is
+%
+However, despite the advantages of having linear types in Core, the status quo is
 that linearity is effectively ignored in Core, being only checked
-in the source Haskell code. The reason is that the desugaring
-and subsequent Core-to-Core optimising transformations
-eliminate and rearrange most of the syntactic constructs through
-which linearity checking is performed -- often resulting in programs
-that are syntactically very different from the original, especially
-due to the flexibility provided by laziness wrt the
+in the source Haskell code. The reason is that, after desugaring, the
+Core-to-Core optimising transformations eliminate and rearrange most of the
+syntactic constructs through which linearity checking is performed -- often
+resulting in programs that are syntactically very different from the original,
+especially due to the flexibility provided by laziness with regard to the
 optimisations a compiler may perform.
@@
 %
-Additionally, we \ROUNDTWO{that} show our system
+Additionally, we show \ROUNDTWO{that} our system
 %

### 2025-10-14 popl26.tex
.

@@
 %
-\ROUNDTWO{While this distinction is largely unexplored, it becomes unavoidable
-    for} Haskell's optimising compiler, which heavily rewrites the source
+\ROUNDTWO{While this distinction has been largely unexplored, it turns
+  out to be inescapable in} Haskell's optimising compiler, which heavily rewrites the source
     program in ways that break syntactic linearity but preserve the program's
@@
 % \ROUNDTWO{Linear Core explores the interaction between laziness and linearity}
-is suitable for the Core intermediate language of the
+is suitable for \ROUNDTWO{lazy languages such as } the Core intermediate language of the
 Glasgow Haskell Compiler. We prove \ROUNDTWO{that} Linear Core is sound,
-guarantees linear resource usage, and \ROUNDTWO{that} multiple
+\ROUNDTWO{guaranteeing} linear resource usage, and \ROUNDTWO{that} multiple
 optimising transformations preserve linearity in Linear Core while failing

### 2025-10-15 popl26.tex
Up to start of section 2

@@
 %
-%
-However, despite the advantages of having linear types in Core, the status quo is
+\ROUNDTWO{However, despite the advantages of having linear types in Core, the status quo is}
 that linearity is effectively ignored in Core, being only checked
-in the source Haskell code. The reason is that, after desugaring, the
+in the source Haskell code. The reason is that, \ROUNDTWO{after} desugaring, the
 Core-to-Core optimising transformations eliminate and rearrange most of the
@@
 programs and therefore a program's resource consumption,
-a \emph{naive} syntactic analysis of the optimised programs (as
-performed by the current Core type checker) often fails to recognize
-the program as linear.
+a naive syntactic analysis of the optimised programs as
+performed by the current Core type checker often fails to recognize
+\ROUNDTWO{resulting} programs as linear.
 %
@@
 where the latter results from inlining $y$ without eliminating the
-binder (a common optimising transformation in GHC). Even
+binder -- a common optimising transformation in GHC. Even
 though the second expression no longer appears linear (as there are now two occurrences
 of $x$), it is indeed linear because the let-bound expression is never
-evaluated under \emph{lazy evaluation} and $x$ is still consumed exactly once (when it is freed in the let body):
+evaluated under \ROUNDTWO{non-strict} evaluation and $x$ is still consumed exactly once (when it is freed in the let body):
 % executed, $x$ will be freed exactly once:
@@
 The Core optimising transformations expose a fundamental limitation of Core's
-type system: it does not account for the non-strict evaluation model
+\ROUNDTWO{linear} type system: it does not account for the non-strict evaluation model
 of Core and, thus, a whole class of programs that are linear under the lens of
-non-strict evaluation are rejected by Core's current linear type system.
+non-strict evaluation are rejected by Core's current \ROUNDTWO{checker}.
 %
@@
 
-\item We introduce the concept of \emph{semantic} linearity, in
-contrast to \emph{syntactic} linearity in Haskell, by example
-(Section~\ref{sec:linearity-semantically}).
-%
-A precise definition of \emph{semantic} linearity is given by our instrumented
-operational semantics, under which a program using a linear resource
-non-linearly gets stuck (Section~\ref{sec:main:metatheory}).
-
-\item We introduce Linear Core, a non-strict linear language with
-the key features of Core (except for type equality coercions), whose type system
-accounts for semantic linearity in the presence of laziness. To the
-best of our knowledge, this is the first type system to capture linearity
-semantically in this context (Section~\ref{sec:main:linear-core}).
+\item We \ROUNDTWO{illustrate} the \ROUNDTWO{lazy semantics} of linearity, in
+contrast to the \ROUNDTWO{traditional} syntactic \ROUNDTWO{understanding
+of} linearity, in Haskell, by example (Section~\ref{sec:linearity-semantically}).
+%
+A precise \ROUNDTWO{semantic definition of} linearity, \ROUNDTWO{closely
+following that of Linear Haskell~\cite{cite:linearhaskell}}, is given by
+our instrumented operational semantics, under which a program using a
+linear resource non-linearly gets stuck
+(Section~\ref{sec:main:metatheory}).
+
+\item We introduce Linear Core, a \ROUNDTWO{lazy} linear language with
+the key features of Core (except for type equality coercions), whose type
+system \ROUNDTWO{accepts the lazy} semantics of linearity. To the best of our
+knowledge, this is the first type system \ROUNDTWO{to accept programs
+which depend on non-strict evaluation to be understood as linear}
+(Section~\ref{sec:main:linear-core}).
 
@@
 
-\noindent Additional proofs and definitions are given in
+\noindent \ROUNDTWO{While GHC's Core language is our main focus, we
+note that the ideas explored in this work can potentially be applied to any
+language with non-strict features, even if the language itself is strict.}
+
+Additional proofs and definitions are given in \ROUNDTWO{TODO}
 Appendices~\ref{app:proofs},~\ref{app:optimisations} and~\ref{app:opsem}.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Linearity, Semantically\label{sec:linearity-semantically}}
-
-A linear type system typically guarantees that linear resources are
-consumed exactly once. Consequently, whether a program is
-well-typed under this lens intrinsically depends on the
-definition of consuming a resource.
-In general, consuming a resource is equated with its 
-\emph{syntactic occurrence} along a control-flow path in the program. However, as this section
-makes clear, in linear type systems for non-strict
-languages, a distinction between using
-resources \emph{syntactically} and \emph{semantically} is needed.
+\section{Linearity, Lazily\label{sec:linearity-semantically}}
+
+Linear type systems guarantee that linear resources are consumed exactly once.
+\ROUNDTWO{Thus}, whether a program is \ROUNDTWO{accepted as linear} intrinsically depends on
+the definition of consuming a resource.
+%
+\ROUNDTWO{Traditionally}, consuming a resource \ROUNDTWO{has been} equated with its 
+syntactic occurrence along a control-flow path in the program. However, as this
+section \ROUNDTWO{aims to} make clear, in linear type systems for non-strict languages, a
+distinction between using resources \emph{syntactically} and \emph{semantically} is needed.
 %
@@
 
-\subsection{Semantic Linearity by Example\label{sec:semantic-linearity-examples}}
+\subsection{Lazy Linearity by Example\label{sec:semantic-linearity-examples}}
 

### 2025-10-15 popl26.tex
Até subsection 2.Let Bindings

@@
 Just as Core's type system \cite{10.1145/1190315.1190324} provides a degree of validation to the
-translation from Haskell (dubbed \emph{desugaring}) and the subsequent
+translation from Haskell, dubbed \emph{desugaring}, and the subsequent
 optimising transformations, a linearly typed Core would guarantee that
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Linearity, Lazily\label{sec:linearity-semantically}}
+\section{Linearity, Semantically\label{sec:linearity-semantically}}
 
@@
 syntactic occurrence along a control-flow path in the program. However, as this
-section \ROUNDTWO{aims to} make clear, in linear type systems for non-strict languages, a
-distinction between using resources \emph{syntactically} and \emph{semantically} is needed.
-%
-For instance, in the following Haskell-like program defining a linear
-function \lstinline{f} (i.e., a function that should use its argument
-exactly once), there are two syntactic
+section \ROUNDTWO{aims to} make clear, \ROUNDTWO{in a non-strict context, a
+syntactic definition of linearity is much too conservative}.
+%
+For instance, in the following Haskell-like program, \lstinline{f} is a linear
+function (a function that must use its argument
+exactly once), but there are two syntactic
 occurrences of \lstinline{handle}: one is in a let-bound computation that
-closes this handle, the other is returned. If the program is evaluated lazily,
-the handle will not be closed because the computation that closes the handle is
-never evaluated (the non-linear \lstinline{x} goes unused).
+closes this handle, the other \ROUNDTWO{is in the first branch}. If the program is
+evaluated lazily, the handle will, \ROUNDTWO{dynamically, either be closed in
+the second branch, or returned untouched in the first}, since the computation
+which closes the handle is \ROUNDTWO{not needed in the first branch}.
 %
@@
 f : Handle *' $\lolli$ '* Handle
-f handle = let x = close handle in handle
+f handle =
+    let closed = close handle in
+    if <cond>
+        then handle
+        else closed
 \end{lstlisting}
@@
 %
-Intuitively, a computation that depends on a linear resource to produce a
+Intuitively, a computation that depends on a linear resource to produce some
 result consumes that resource iff the result is fully consumed; in
@@
 we turn our focus to linearity under lazy evaluation since the distinction between semantically and syntactically
-consuming a resource is only exposed under non-strict semantics (and since GHC Core is
-lazy).
+consuming a resource is only exposed under non-strict semantics, and because GHC Core is
+lazy.
 %.
@@
 
-\subsection{Lazy Linearity by Example\label{sec:semantic-linearity-examples}}
+\subsection{Semantic Linearity by Example\label{sec:semantic-linearity-examples}}
 
@@
 non-strict evaluation are keen to write programs with bindings that are selectively
-used in case alternatives, such as (we write $\lolli$ for linear
-functions and \lstinline{->} for non-linear functions):
+used in case alternatives, such as in the following program. Note that we write
+$\lolli$ for linear functions and \lstinline{->} for non-linear functions:
 %

### 2025-10-15 popl26.tex
Almost finish Let Bindings S2

@@
 
-Lastly, consider a (ill-typed) program which defines two let bindings \incode{z} and \incode{y}, where
-\incode{z} uses \incode{y} which in turn uses the linear resource \incode{x}:
+Lastly, consider an ill-typed program which defines two let bindings \incode{z}
+and \incode{y}, where \incode{z} uses \incode{y} which in turn uses the linear
+resource \incode{x}:
 %
@@
 %
-Even though the binding \incode{y} is used in \incode{z}, \incode{x} is never
-consumed because \incode{z} isn't evaluated in the let body, and, thus,
-\incode{y} isn't evaluated either. We use this example to highlight that even
-for let-bound variables, the syntactic occurrence of a variable isn't enough to
-determine whether it is used. Instead, we ought to think of uses of \incode{y}
-as implying using \incode{x}, and therefore uses of \incode{z} imply using
-\incode{x}, however, if neither are used, then \incode{x} isn't used. Since
-\incode{x} is effectively discarded, this example also violates linearity.
+Even though \incode{y} occurs in \incode{z}, \incode{x} is never consumed
+\ROUNDTWO{because \incode{z} is discarded}. \ROUNDTWO{This example highlights
+that naively checking for syntactic occurrences of let-bound variables such as
+\incode{y} is also inadequate to track resource usage}.
+%
+\ROUNDTWO{A better mental model is that} using \incode{y} implies
+using \incode{x}, and thus using \incode{z} implies using \incode{x}
+\ROUNDTWO{too}. If neither \incode{z} nor \incode{y} are used, then \incode{x}
+is \ROUNDTWO{effectively discarded}.
 

### 2025-10-16 popl26.tex
revised case rule typesetting

@@
 
-Just as Core's type system \cite{10.1145/1190315.1190324} provides a degree of validation to the
+Just as Core's type system \cite{10.1145/1190315.1190324} provides a
+degree of validation to the 
 translation from Haskell, dubbed \emph{desugaring}, and the subsequent
@@
 bindings $x_1{:}_{\D}\s_1\dots x_n{:}_{\D}\s_n$ and associated expressions
-$e_1\dots e_n$, and case expressions $\ccase{e}{\zD~\{\ov{\rho \to e'}\}}$,
+$e_1\dots e_n$, and case expressions $\ccase{e}{\zD~\{\ov{\rho \Rightarrow e'}\}}$,
 where $z$ is the case binder and the overline denotes a set of distinct
@@
 
-\vspace{-0.5cm}
+%\vspace{-0.5cm}
 \[
@@
 the linear environment is made up exactly of the resources in the usage environment of
-that variable.
+that variable, \ROUNDTWO{denoted by the $\Delta = \Delta'$ premise of
+  the rule.}
 %
@@
 Moreover, resources captured by a let-bound expression cannot be used
-simultaneously with the let-binder in its body, since said
-resources would end up being consumed more than once and violate (semantic)
-linearity. The binder has to be used in mutual exclusion with the linear
-resources captured by the expression it binds -- either the binding or the
+simultaneously with the let-binder in its body, since \ROUNDTWO{those}
+resources would end up being consumed more than once and violate
+\ROUNDTWO{the semantics of}
+linearity. % The binder has to be used in mutual exclusion with the linear
+% resources captured by the expression it binds --
+Either the binding or the
 resources \emph{must} be used.
@@
 %
-Annotating let-bound variables with a usage environment $\D$ delays the
+Annotating let-bound variables with \ROUNDTWO{their} usage environment $\D$ delays the
 consumption of resources to when the variables themselves are used and

### 2025-10-16 popl26.tex
up to recursive let

@@
 and $\D'$. $\D$ is used to type the body $e$ of the let binding $x$. Perhaps
-surprisingly, the resources $\D$ used to type $e$ are still available in the
-environment to type the let body $e'$, alongside the unrestricted $x$ binding
-annotated with the usage environment $\D$. In essence, the resources of $e$ being
-available in $e'$ reflects how typing the body of a lazy binder
-doesn't consume resources, and the binding $x$ being an unrestricted
-$\D$-variable ensures that using it will consume the resources $\D$ the
-expression captures, \emph{if} it is used.
+surprisingly, the resources $\D$ are still available to type the let body $e'$, alongside the unrestricted $x$ binding
+annotated with the usage environment $\D$. The resources of $e$ being
+available in $e'$ reflects how the body of a lazy binder
+doesn't consume resources outright, and $x$ being an
+\ROUNDTWO{unrestricted $\Delta$-bound variable}
+ensures that using $x$ will consume the resources $\D$ that $e$
+captures.
 % \todo[inline]{Let bindings are hard, if they are used then we use resources. If
@@
 Recursive let bindings also bind expressions lazily and so
-introduce a $\D$-bound variable for each binding. Resources required to type the
-let-bindings are still available in the body of the let, to later be consumed
-via $\D$-bound variables, or, directly, if the let-bindings are unused.
+introduce a $\D$-bound variable for each binding. \ROUNDTWO{The} resources required to type the
+let-bindings are still available in the \ROUNDTWO{let body}. \ROUNDTWO{These
+resources may} be consumed
+via \ROUNDTWO{the corresponding} $\D$-bound variables, or directly if the let-bindings are unused.
 %
@@
 % (from which an implementation is direct) 
-because determining a particular $\D$
-to type and annotate all binders is not obvious. We present our system and
-metatheory agnostically to the challenge of inferring this linear typing
-environment by assuming recursive let expressions are annotated with the
-correct typing environment.
+since \ROUNDTWO{it does not describe how to annotate binders with} a
+particular \ROUNDTWO{usage environment} $\D$.
+\ROUNDTWO{Our system simply assumes the annotations are explicitly
+  provided in the source and checks the correctness of the overall
+  recursive let.}
 

### 2025-10-16 popl26.tex
Up to start S2.1.3

@@
 %
-\ROUNDTWO{However, despite the advantages of having linear types in Core, the status quo is}
+\ROUNDTWO{However}, despite the advantages of \ROUNDTWO{having} linear types in Core, the status quo is
 that linearity is effectively ignored in Core, being only checked
@@
 following that of Linear Haskell~\cite{cite:linearhaskell}}, is given by
-our instrumented operational semantics, under which a program using a
+our instrumented operational semantics under which a program using a
 linear resource non-linearly gets stuck
@@
 evaluated lazily, the handle will, \ROUNDTWO{dynamically, either be closed in
-the second branch, or returned untouched in the first}, since the computation
+the second branch, or be returned untouched in the first}, since the computation
 which closes the handle is \ROUNDTWO{not needed in the first branch}.
@@
 %
-we turn our focus to linearity under lazy evaluation since the distinction between semantically and syntactically
-consuming a resource is only exposed under non-strict semantics, and because GHC Core is
-lazy.
+we turn our focus to linearity under lazy evaluation
+\ROUNDTWO{specifically} since the distinction between semantically and
+syntactically consuming a resource is only exposed under non-strict semantics,
+and because GHC Core is lazy.
 %.
@@
 To understand exactly how,
-we turn to the definition of \emph{consuming} a resource from Linear
+we turn to the definition of consuming a resource from Linear
 Haskell~\cite{cite:linearhaskell}:
@@
 %\end{itemize}
-Thus, we can consume a linear resource by fully evaluating it -- and evaluation
-happens in case expressions.
+Thus, we can consume a linear resource by fully evaluating it, and evaluation
+happens with case expressions.
 

### 2025-10-16 popl26.tex
WIP case examples

@@
 % unrestricted ones).
-
+%
 %\end{enumerate}
-
+%
 % ROMES:TODO: We should talk about WHNF in the background, not here.
 % \parawith{WHNF} An expression is in Weak Head Normal Form when 
-
-We start with an
-example of a program that constructs a tuple from linear resources then pattern
-matches on it, then uses both linearly-bound variables from the tuple pattern
-match. This is well-typed in Linear Haskell:
+%
+\ROUNDTWO{The first} example program constructs a tuple from linear resources,
+matches on it, then uses both linearly-bound variables from the tuple pattern.
+This is well-typed in Linear Haskell:
 \begin{working}
@@
 \end{working}
-What might be more surprising is that a similar program which discards the
-pattern variables and instead uses the resources in the scrutinee is 
-\emph{semantically} linear, despite not being accepted by Linear Haskell:
+\ROUNDTWO{Perhaps surprisingly}, a similar program which discards the pattern
+variables, and instead uses the resources \ROUNDTWO{from} the scrutinee
+\ROUNDTWO{again}, is still linear, despite not being accepted by Linear Haskell:
 \begin{notyet}
@@
 \end{notyet}
-We justify that \incode{f7} is linear by appealing to the fact that 
-the tuple being scrutinized is already in WHNF and so evaluating it will
-not consume either $x$ or $y$. Even if the tuple was constructed with
-two expressions using $x$ and $y$ respectively, no computation would happen
-since we are not using $a$ or $b$ (thereby never forcing the arguments
-of the tuple). However, if we did use $a$ in the case body, then $x$
-would be unavailable (as opposed to \incode{y}) and \incode{f8} below must be rejected:
+\incode{f7} \ROUNDTWO{can be seen as} linear by \ROUNDTWO{realizing} that
+the tuple being scrutinized is already in WHNF, so evaluating it again does
+not consume either $x$ \ROUNDTWO{nor} $y$. Even if the tuple \ROUNDTWO{paired}
+two \ROUNDTWO{arbitrary} expressions which \ROUNDTWO{captured} $x$ and $y$,
+\ROUNDTWO{because $a$ and $b$ are not forced in the continuation, $x$ and $y$ would remain unused.}
+\ROUNDTWO{On the other hand, as seen in the next example}, if $a$ were used in 
+the case body, then $x$ \ROUNDTWO{is consumed via $a$ and must not be used again.}
+Thus, \incode{f8} must be rejected:
 \begin{noway}
@@
 \end{noway}
-This idea that $x$ and $a$ are mutually exclusive is the same behind let
-bindings being mutually exclusive to the resources that define them.
-By forcing the pattern variable we run the computations
-defined in terms of the linear variables used for that constructor
-argument. Otherwise, if we do not use the pattern variables, then we
-do not run the computation and thus no resources are consumed.
+This idea that $x$ and $a$ are mutually exclusive is \ROUNDTWO{akin to how} let
+bindings \ROUNDTWO{are} mutually exclusive \ROUNDTWO{with the set} of resources
+\ROUNDTWO{they capture, in the let-continuation}.
+\ROUNDTWO{Forcing a pattern variable evaluates the corresponding expression to
+which the constructor was applied, and all linear variables captured by that
+expression may }.
+Otherwise, if we do not use the
+pattern variables, then we do not run the computation and thus no resources are
+consumed.
 

### 2025-10-17 popl26.tex
.

@@
 the key features of Core (except for type equality coercions), whose type
-system \ROUNDTWO{accepts the lazy} semantics of linearity. To the best of our
+system \ROUNDTWO{is designed to reconcile with the lazy} semantics of
+linearity. \ROUNDTWO{Our system achieves this via a novel combination
+  of features: explicit tracking linear dependencies of variables
+  through \emph{usage environments}
+  (Section~\ref{sec:usage-environments}), codifying that lazy variable
+  bindings do not consume resources upon definition, but rather when
+  the variables are consumed;
+  a distinct treatment of case
+  scrutinees in weak-head normal form (Section~\ref{sec:onwhnf}),
+  identifying that linear resources captured by case expressions in
+  WHNF can be safely used in branches since the case will not further
+  evaluate its scrutinee; and a notion of irrelevant and tagged resources
+  (Section~\ref{sec:irrev}), which allow for fine-grained control over
+  variable usage. Namely, linear resources occurring in non-WHNF case
+  scrutinees can no longer be used in the alternative, while the
+  pattern variables (or the case binder) must necessarily be used. } 
+%
+To the best of our
 knowledge, this is the first type system \ROUNDTWO{to accept programs
@@
 
-Additional proofs and definitions are given in \ROUNDTWO{TODO}
-Appendices~\ref{app:proofs},~\ref{app:optimisations} and~\ref{app:opsem}.
+Additional proofs and definitions are given in \ROUNDTWO{the extended
+  version of the paper~\cite{techreport}.}
+%Appendices~\ref{app:proofs},~\ref{app:optimisations} and~\ref{app:opsem}.
 
@@
 section \ROUNDTWO{aims to} make clear, \ROUNDTWO{in a non-strict context, a
-syntactic definition of linearity is much too conservative}.
+syntactic definition of linearity is overly conservative}.
 %
@@
 occurrences of \lstinline{handle}: one is in a let-bound computation that
-closes this handle, the other \ROUNDTWO{is in the first branch}. If the program is
-evaluated lazily, the handle will, \ROUNDTWO{dynamically, either be closed in
-the second branch, or be returned untouched in the first}, since the computation
-which closes the handle is \ROUNDTWO{not needed in the first branch}.
+closes this handle, the other \ROUNDTWO{is in the then branch of a
+  conditional}. If the program is 
+evaluated lazily, the handle will, \ROUNDTWO{dynamically, either be
+  used exactly once (i.e,~closed) in
+the else branch, or be used exactly once (i.e.,~returned) in the then branch}, since the 
+\ROUNDTWO{let-bound variable is not needed in that branch and
+so the expression which closes the handle is never evaluated}.
 %
@@
 
-\subsubsection{Branching on WHNF-ness}
+\subsubsection{Branching on WHNF-ness}\label{sec:onwhnf}
 
@@
 
-\subsubsection{Proof irrelevant resources}
+\subsubsection{Proof irrelevant resources}\label{sec:irrev}
 

### 2025-10-17 popl26.tex
more tweaks
More on cases

@@
 \ROUNDTWO{they capture, in the let-continuation}.
-\ROUNDTWO{Forcing a pattern variable evaluates the corresponding expression to
-which the constructor was applied, and all linear variables captured by that
-expression may }.
-Otherwise, if we do not use the
-pattern variables, then we do not run the computation and thus no resources are
-consumed.
-
+\ROUNDTWO{Forcing a pattern variable evaluates the expression for the
+corresponding constructor field, and all linear variables captured by that
+expression may end up being used}.
+% If we do not use the pattern variables, then we do not run the computation and
+% thus no resources are consumed.
+%
 A third (semantically well-typed, but rejected by Linear Haskell) option is to use the case binder $z$ instead of $a,b$ or $x,y$:
@@
 \end{notyet}
-Again, $z$ is mutually exclusive with $(a,b)$ and with $(x,y)$, but at least one of
-the three must occur to ensure the linear resources are consumed. In this
-example, we can think that using $a$ entails using the resource $x$, $b$ the
-resource $y$, and the case binder $z$ entails using both $a$ and $b$.
+Again, $z$ is mutually exclusive with $(a,b)$ and with $(x,y)$, but at least
+one of the three must \ROUNDTWO{be used} to ensure the
+linear resources are consumed. Essentially, in this example, using $a$ entails
+using the resource $x$, $b$ the resource $y$, and the case binder $z$ entails
+using both $a$ and $b$.
 
@@
 
-\subsubsection{Branching on WHNF-ness}\label{sec:onwhnf}
+\subsubsection{\ROUNDTWO{Discriminating by} WHNF-ness}\label{sec:onwhnf}
 

### 2025-10-17 popl26.tex
reconcile with -> accept

@@
 the key features of Core (except for type equality coercions), whose type
-system \ROUNDTWO{is designed to reconcile with the lazy} semantics of
+system \ROUNDTWO{is designed to accept the lazy} semantics of
 linearity. \ROUNDTWO{Our system achieves this via a novel combination

### 2025-10-17 popl26.tex
End S2

@@
 %
-A third (semantically well-typed, but rejected by Linear Haskell) option is to use the case binder $z$ instead of $a,b$ or $x,y$:
+A third option, semantically \ROUNDTWO{linear}, but rejected by Linear Haskell, is to use
+the case binder $z$ instead of $a,b$ or $x,y$:
 \begin{notyet}
@@
 
-Dually, consider the scrutinee to be an expression that's not in WHNF, s.t.
-evaluating it to WHNF will require performing computation and thus consume linear
-resources (\incode{f10} below is well-typed in our system but rejected
-by Linear Haskell):
+\ROUNDTWO{Conversely}, if the scrutinee is an expression that's not in WHNF,
+evaluation to WHNF \ROUNDTWO{may indeed} consume \ROUNDTWO{some or all of the}
+linear resources \ROUNDTWO{captured by said scrutinee}. \ROUNDTWO{Consider the following
+non-WHNF scrutinee example},
+\incode{f10}, \ROUNDTWO{which is accepted} by our system:
 \begin{notyet}
@@
 \end{notyet}
-Unlike when the scrutinee was in WHNF, we can no longer use $x,y$ in the case
-alternatives, but we must still use either the case binder $z$ or the linear
-pattern variables $a,b$. %  e.g. it would be quite disastrous if any of the following typechecked:
+\ROUNDTWO{In contrast to} a scrutinee in WHNF, \ROUNDTWO{the resources captured
+by a non-WHNF scrutinee} (here, $x,y$), \ROUNDTWO{conservatively, can no longer
+be used} in the case alternatives, \ROUNDTWO{since reducing \incode{use x y} to
+WHNF may consume (parts of) $x$ and/or $y$}. \ROUNDTWO{Consequently}, either the case
+binder $z$ or the linear pattern variables $a,b$ must be used exactly
+once,
+%  e.g. it would be quite disastrous if any of the following typechecked:
 % \begin{noway}
@@
 %
-The result of evaluating the scrutinee must be consumed exactly once to guarantee
-that the resources used in the scrutinee are fully consumed. %, or
+since the linear resources in the scrutinee are \ROUNDTWO{only} fully consumed
+if the result of evaluating the scrutinee \ROUNDTWO{is also} fully consumed . %, or
                                 %risk them being only ``almost''
@@
 For instance, take
-\incode{use} in \incode{f10} to be the pairing function \incode{(,)}: it is not sufficient for \incode{use x y} to be
-evaluated to WHNF to consume \incode{x} and \incode{y}. Otherwise, if all the resources were
-considered to be fully consumed after the scrutinee were evaluated in a case
-expression, we could simply ignore the pattern variables, potentially
-discarding linear resources. In
-short, if the scrutinee is not in WHNF we must either consume the case binder
-or the linear components of the pattern.
-
-On the other hand, if we match on a constructor without any linear components,
-all linear resources used in the scrutinee must have been fully consumed in
-that branch since the result of evaluating it can be consumed unrestrictedly
-(by definition, a constructor application to unrestricted arguments is also
-unrestricted).
-%
-Hence, in the branch for a constructor without linear fields, the case
+\incode{use} in \incode{f10} to be the pairing function \incode{(,)}: it is not
+sufficient for \incode{use x y} to be evaluated to WHNF to consume \incode{x}
+and \incode{y}.
+%
+If all the resources were considered to be fully consumed after
+\ROUNDTWO{evaluating the} scrutinee to \ROUNDTWO{WHNF}, the pattern variables
+could simply be ignored and thus linear resources potentially discarded.
+In \ROUNDTWO{practice}, if the scrutinee is not in WHNF then \ROUNDTWO{either}
+the case binder or the linear components of the pattern \ROUNDTWO{must be consumed}.
+
+\ROUNDTWO{Alternatively}, if a constructor without any linear components \ROUNDTWO{is matched},
+all linear resources used in the scrutinee must \ROUNDTWO{already} have been
+fully consumed in that branch since the result of evaluating it can be consumed
+unrestrictedly (by definition, a constructor application to unrestricted
+arguments is also unrestricted).
+%
+Hence, in the \ROUNDTWO{case-}branch for a constructor without linear fields, the case
 binder can also be used unrestrictedly. For example, the program in

### 2025-10-17 popl26.tex
section 3

--- 2025-10-17 popl26.tex
@@
 Finally, we introduce the rules to type case expressions and case alternatives,
-along with the key insights to do so, namely branching on WHNF-ness of
-scrutinee, proof irrelevant resources, and tagging (Section~\ref{sec:lc-case-exps}).
+along with the key insights to do so, namely \ROUNDTWO{discriminating by} the WHNF-ness of
+scrutinee, irrelevant resources, and tagging (Section~\ref{sec:lc-case-exps}).
 
@@
 whose scrutinee is in weak head normal form and one whose scrutinee is not
-leads to one of the key insights of $\lambda^\pi_\Delta$: we must \emph{branch}
-on weak head normal formed-ness to accurately type case expressions.
+leads to one of the key insights of $\lambda^\pi_\Delta$:
+\ROUNDTWO{accurate typing of case expressions depends on whether the scrutinee is in WHNF.}
+%
 %
@@
 constructor arguments and the resources used to type them.
-% Using the case binder entails using all the resources required by the scrutinee, and using a pattern
-% variable implies using the resources of the corresponding constructor argument.
+\ROUNDTWO{The reader may wonder about the significance of case expressions
+  where the scrutinee is in WHNF. We note that in an
+  optimizing compiler (such as GHC) many intermediate programs that
+  arise due to program transformations commonly scrutinize expressions
+  in WHNF.}
 
@@
 sub-expressions. Finally, we type all case alternatives with the same context, using the
-$\vdash_{alt}$ judgement, introducing
+introducing
 the case binder $z$ in the unrestricted environment as a $\D$-bound
@@
 
-The alternative judgement $\G;\D \vdash_{alt} \rho \to e :^z_\D \s \Rightarrow
-\vp$ is used to type case alternatives, encompassing three
+The alternative judgement \ROUNDTWO{$\G;\D \vdash \rho \Rightarrow e :^z_\D 
+\s$} is used to type case alternatives, encompassing three
 modes that are distinguished by the kind of arrow that is used:
@@
 the scrutinee typing environment by the empty linear environment $\cdot$)
-and from the usage environment of the case binder $z$ (written $\G[\cdot/\D_s]_z$
-to similarly delete $\D_s$ from the usage environment of the $\D$-var $z$ in $\G$).
+and from the usage environment of the case binder $z$, written $\G[\cdot/\D_s]_z$
+to delete $\D_s$ from the usage environment of the \ROUNDTWO{$\D$-bound variable} $z$ in $\G$.
 Given that the case-binder $z$ is always annotated with $\D_s$ (essentially,
@@
 
-\subsubsection{Proof irrelevant resources}\label{sec:irrev}
-
-When the scrutinee is not in WHNF, it is not statically known which resources
-will have been consumed after evaluating said scrutinee, or which
-are transferred into the branch continuation, e.g. as in $\ccase{f\ x}{z\ \{K\
+\subsubsection{Irrelevant resources}\label{sec:irrev}
+
+When the scrutinee is not in WHNF
+\ROUNDTWO{we cannot statically determine which resources are consumed
+  upon scrutinee evaluation or which are transferred into the branch continuation}
+e.g. as in $\ccase{f\ x}{z\ \{K\
 a \rightarrow e\}}$:
@@
     is not consumed by evaluating the scrutinee for $f = K$).
-
-    Considering (1), we must resort to a more uniform method of guaranteeing we
-    ``finish consuming'' the scrutinee. There are only two ways of
-    uniformly referring to the remaining linear resources in the newly-evaluated scrutinee:
-
-        \begin{itemize}
-            \item The case binder, which refers to the whole result of evaluating the scrutinee;
-            \item The linear components of a constructor pattern, which refer to expressions that may contain linear resources.
-        \end{itemize}
 
@@
 
+\ROUNDTWO{
+    Considering (1), we must resort to a more uniform method of guaranteeing we
+    ``finish consuming'' the scrutinee. There are only two ways of
+    uniformly referring to the remaining linear resources in the newly-evaluated scrutinee:
+
+        \begin{itemize}
+            \item The case binder, which refers to the whole result of evaluating the scrutinee;
+            \item The linear components of a constructor pattern, which refer to expressions that may contain linear resources.
+        \end{itemize}}
+
 To encode  linear resources that cannot be
 directly used (i.e.,~to which the $\textsc{Var}$ rule is not
-applicable) we introduce \emph{proof irrelevant} resources, written as linear resources
-within square brackets $[ x{:}\s]$ and lifted to contexts as $[\D]$.
-Proof irrelevant resources
+applicable) we introduce \emph{irrelevant} resources, written as linear resources
+within square brackets $[ x{:}\s]$ and \ROUNDTWO{pointwise} lifted to contexts
+\ROUNDTWO{and usage environments} as $[\D]$.
+Irrelevant resources
 are linear resources in every sense, meaning they must be used
-exactly once. However, since proof irrelevant resources cannot be
+exactly once. However, since irrelevant resources cannot be
 discarded or used directly, they have to be consumed \emph{indirectly} via
@@
 type the scrutinee with linear resources $\D$ and the case
-alternatives by introducing the case binder with a usage environment $[\D]$,
-having the same proof irrelevant linear context $[\D]$ in the typing
-environment, and annotating the judgement with the proof irrelevant resources
+alternatives by introducing the case binder with an \ROUNDTWO{irrelevant} usage environment $[\D]$,
+having the same irrelevant $[\D]$ in the typing
+environment, and annotating the judgement with the irrelevant resources
 for use in the $\Rightarrow_{\textsf{NWHNF}}$ judgement:
@@
 Finally, we recall the tentative $Case_\textrm{WHNF}$ rule presented before and
-highlight its flaw: the $\G;\D \vdash_{alt} \rho \to e :^z_{\D_s} \s \Rightarrow_{\textsf{WHNF}}
-\vp$ judgement is only well-defined for patterns $\rho$ matching the WHNF form
+highlight its flaw: the $\G;\D \vdash \rho \Rightarrow_{\textsf{WHNF}}
+e :^z_{\D_s} \s$ judgement is only well-defined for patterns $\rho$
+matching the WHNF form 
 of the scrutinee, as the distribution of resources per constructor components
@@
 will never be executed. We uniformly treat non-matching alternatives
-as if the scrutinee were not in WHNF. Having introduced proof irrelevant
+as if the scrutinee were not in WHNF. Having introduced irrelevant
 resources, we can now present the full $Case_\textrm{WHNF}$ rule:
@@
 \]
-We note that it might seem unusual to specialize a rule for expressions in
+\ROUNDTWO{where the matching pattern is typed as before, with the
+  case-binder $z$ is annotated with the appropriate usage environment;
+  and the non-matching patterns are if the scrutinee were notin WHNF.}
+We again note that it might seem unusual to specialize a rule for expressions in
 WHNF, as programs scrutinizing an expression in WHNF are rarely
@@
 
-As opposed to scrutinees in WHNF, where the resources used to type a scrutinee
+As opposed to scrutinees in WHNF, where resources 
 can be cleanly divided amongst the various sub-expressions of a constructor
@@
 all resources occurring in the environment, or not at all (instead, the case
-binder may be used). Given linear resources $[\D_s]$ used to type a scrutinee,
+binder may be used). Given \ROUNDTWO{irrelevant} resources $[\D_s]$ used to type a scrutinee,
 and a pattern $K~\ov{x_\omega},\ov{y_i}$ with $i$ linear components, we assign
@@
 and linear-variable index $\lctag{\irr{\D_s}\!}{K_i}$. Then, $y_i{:}_{\D_i}\s$ is
-introduced in $\G$, just like other $\D$-variables.
+introduced in $\G$, just like other $\D$-bound variables.
 \[
@@
 Having uniquely tagged the resources in the usage environment of each pattern
-variable, we need only to express that (1) the pattern-bound $\D$-variables can
+variable, we need only express that (1) the pattern $\D$-bound variables can
 be used (i.e., the tagged resources need be available in the linear
 environment in order for the $\textsc{Var}_\D$ rule to be applicable) and (2)
-if a $K$-pattern-bound $\D$-variable is consumed, all remaining $K$-$\D$-variables,
+if a $K$-pattern $\D$-bound variable is consumed, all remaining
+$K$-$\D$-bound variables,
 standing for the remaining linear components of the same pattern, must also be
@@
 are evaluation states. We also rely on a notion
-of partial derivations and partial evaluation (see
-Appendix~\ref{app:opsem}), written $a\Downarrow^* b$, which allows us
+of partial derivations and partial evaluation (see~\cite{techreport}), written $a\Downarrow^* b$, which allows us
 to state and reason about a progress property.
@@
 well-typed if the linear variable were instead $\D$-bound ($\xD$) with
-usage environment $\D$, with $\D$ available in the linear context (and the unrestricted context updated accordingly).
+usage environment $\D$, with $\D$ available in the linear context and
+\ROUNDTWO{with any occurrences of $x$ in usage environments in $\G$
+  expanded into $x$'s usage environment $\Delta$, with the abuse of notation $\Gamma[\Delta/x]$.
 

### 2025-10-17 popl26.tex
Make key ideas more obvious in S3

@@
 % 
-Despite the focus on GHC Core, the fundamental ideas for
-understanding linearity in our call-by-need calculus can be readily applied to
-other non-strict languages.
-
-Linear Core's first key idea is to delay consuming
-resources until a computation that depends on those resources is effectively
-evaluated or returned. Delayed consumption is tracked by annotating relevant
-binders with \emph{usage environments} (Section~\ref{sec:usage-environments}).
-%
-The second key idea is to have two distinct rules for case
-expressions, branching on whether the scrutinee is in Weak Head Normal
-Form or not, and using  \ROUNDTWO{irrelevance annotations} to track resources that are no
-longer available but have yet to be fully consumed
-(Section~\ref{sec:lc-case-exps}). Additionally, we introduce tags to split
-resources amongst pattern-bound variables and %^as usage environments
-encode that all linear components of a pattern must be used in tandem.
+Despite the focus on GHC Core, the fundamental ideas for \ROUNDTWO{typing} linearity in
+our call-by-need calculus may be applied to other languages \ROUNDTWO{which
+combine non-strict features or semantics with linearity}.
+
+\ROUNDTWO{The key ideas that make Linear Core work are:}
+
+\ROUNDTWO{
+\begin{enumerate}
+    \item Usage environments (Section~\ref{sec:usage-environments})
+    \item Distinct treatment of case scrutinees in WHNF (Section~\ref{sec:onwhnf})
+    \item Irrelevant and tagged resources (Sections~\ref{sec:irrev}~and~\ref{sec:tagging})
+\end{enumerate}
+
+\noindent (1) Encodes the idea that lazy bindings do not consume resources upon
+definition, but rather when the bound variables themselves are consumed.
+Consuming a variable with some usage environment $\Delta$ equates to consuming
+the variables contained in $\Delta$.
+%
+(2) Case expressions in WHNF may capture ambient linear resources, but these
+resources can be safely used in branches since the case will not further
+evaluate its scrutinee. On the other hand, scrutinizing a non-WHNF expression
+must be treated more conservatively.
+%
+(3) Irrelevant resources make existing bindings unusable while forcing the use
+of others. Namely, linear resources occurring in non-WHNF case scrutinees can
+no longer be used in the alternatives, while the pattern variables or the case
+binder must necessarily be used. Tagged resources follow the observation that
+some variables must be used jointly or not at all.
+}
+%
+% Linear Core's first key idea is to delay consuming
+% resources until a computation that depends on those resources is effectively
+% evaluated or returned. Delayed consumption is tracked by annotating relevant
+% binders with \emph{usage environments} (Section~\ref{sec:usage-environments}).
+%
+% The second key idea is to have two distinct rules for case
+% expressions, \ROUNDTWO{discriminating} on whether the scrutinee is in Weak Head Normal
+% Form or not, and using  \ROUNDTWO{irrelevance annotations} to track resources that are no
+% longer available but have yet to be fully consumed
+% (Section~\ref{sec:lc-case-exps}). Additionally, we introduce tags to split
+% resources amongst pattern-bound variables and %^as usage environments
+% encode that all linear components of a pattern must be used in tandem.
 
@@
 
-\subsubsection{Splitting and tagging fragments}
+\subsubsection{Splitting and tagging fragments}\label{sec:tagging}
 

### 2025-10-17 popl26.tex
Simplify list of key ideas in S1

@@
   through \emph{usage environments}
-  (Section~\ref{sec:usage-environments}), codifying that lazy variable
-  bindings do not consume resources upon definition, but rather when
-  the variables are consumed;
+  (Section~\ref{sec:usage-environments});
   a distinct treatment of case
-  scrutinees in weak-head normal form (Section~\ref{sec:onwhnf}),
-  identifying that linear resources captured by case expressions in
-  WHNF can be safely used in branches since the case will not further
-  evaluate its scrutinee; and a notion of irrelevant and tagged resources
+  scrutinees in weak-head normal form (Section~\ref{sec:onwhnf});
+  and a notion of irrelevant and tagged resources
   (Section~\ref{sec:irrev}), which allow for fine-grained control over
-  variable usage. Namely, linear resources occurring in non-WHNF case
-  scrutinees can no longer be used in the alternative, while the
-  pattern variables (or the case binder) must necessarily be used. } 
+  variable usage.} 
 %
@@
 of others. Namely, linear resources occurring in non-WHNF case scrutinees can
-no longer be used in the alternatives, while the pattern variables or the case
-binder must necessarily be used. Tagged resources follow the observation that
+no longer be used in the alternatives, while the pattern variables (or the case
+binder) must necessarily be used. Tagged resources follow the observation that
 some variables must be used jointly or not at all.
@@
 along with the key insights to do so, namely \ROUNDTWO{discriminating by} the WHNF-ness of
-scrutinee, irrelevant resources, and tagging (Section~\ref{sec:lc-case-exps}).
+the scrutinee, irrelevant resources, and tagging (Section~\ref{sec:lc-case-exps}).
 
@@
 \ROUNDTWO{with any occurrences of $x$ in usage environments in $\G$
-  expanded into $x$'s usage environment $\Delta$, with the abuse of notation $\Gamma[\Delta/x]$.
+  expanded into $x$'s usage environment $\Delta$, with the abuse of notation $\Gamma[\Delta/x]$}.
 

### 2025-10-17 popl26.tex
Tweak to introduction contribs

@@
 linearity. \ROUNDTWO{Our system achieves this via a novel combination
-  of features: explicit tracking linear dependencies of variables
-  through \emph{usage environments}
-  (Section~\ref{sec:usage-environments});
+  of features: explicit tracking of captured linear resources through
+    so-called usage environments (Section~\ref{sec:usage-environments});
   a distinct treatment of case

### 2025-10-17 popl26.tex
emph

@@
   of features: explicit tracking of captured linear resources through
-    so-called usage environments (Section~\ref{sec:usage-environments});
+        \emph{usage environments} (Section~\ref{sec:usage-environments});
   a distinct treatment of case

### 2025-10-17 popl26.tex
4

@@
 linearity. \ROUNDTWO{Our system achieves this via a novel combination
-  of features: explicit tracking of captured linear resources through
-        \emph{usage environments} (Section~\ref{sec:usage-environments});
+  of features: explicit tracking linear dependencies of variables
+  through \emph{usage environments}
+  (Section~\ref{sec:usage-environments}), codifying that lazy variable
+  bindings do not consume resources upon definition, but rather when
+  the variables are consumed;
   a distinct treatment of case
-  scrutinees in weak-head normal form (Section~\ref{sec:onwhnf});
-  and a notion of irrelevant and tagged resources
+  scrutinees in weak-head normal form (Section~\ref{sec:onwhnf}),
+  identifying that linear resources captured by case expressions in
+  WHNF can be safely used in branches since the case will not further
+  evaluate its scrutinee; and a notion of irrelevant and tagged resources
   (Section~\ref{sec:irrev}), which allow for fine-grained control over
-  variable usage.} 
+  variable usage. Namely, linear resources occurring in non-WHNF case
+  scrutinees can no longer be used in the alternative, while the
+  pattern variables (or the case binder) must necessarily be used. } 
 %
@@
 % 
-Despite the focus on GHC Core, the fundamental ideas for \ROUNDTWO{typing} linearity in
-our call-by-need calculus may be applied to other languages \ROUNDTWO{which
-combine non-strict features or semantics with linearity}.
-
-\ROUNDTWO{The key ideas that make Linear Core work are:}
-
-\ROUNDTWO{
-\begin{enumerate}
-    \item Usage environments (Section~\ref{sec:usage-environments})
-    \item Distinct treatment of case scrutinees in WHNF (Section~\ref{sec:onwhnf})
-    \item Irrelevant and tagged resources (Sections~\ref{sec:irrev}~and~\ref{sec:tagging})
-\end{enumerate}
-
-\noindent (1) Encodes the idea that lazy bindings do not consume resources upon
-definition, but rather when the bound variables themselves are consumed.
-Consuming a variable with some usage environment $\Delta$ equates to consuming
-the variables contained in $\Delta$.
-%
-(2) Case expressions in WHNF may capture ambient linear resources, but these
-resources can be safely used in branches since the case will not further
-evaluate its scrutinee. On the other hand, scrutinizing a non-WHNF expression
-must be treated more conservatively.
-%
-(3) Irrelevant resources make existing bindings unusable while forcing the use
-of others. Namely, linear resources occurring in non-WHNF case scrutinees can
-no longer be used in the alternatives, while the pattern variables (or the case
-binder) must necessarily be used. Tagged resources follow the observation that
-some variables must be used jointly or not at all.
-}
-%
-% Linear Core's first key idea is to delay consuming
-% resources until a computation that depends on those resources is effectively
-% evaluated or returned. Delayed consumption is tracked by annotating relevant
-% binders with \emph{usage environments} (Section~\ref{sec:usage-environments}).
-%
-% The second key idea is to have two distinct rules for case
-% expressions, \ROUNDTWO{discriminating} on whether the scrutinee is in Weak Head Normal
-% Form or not, and using  \ROUNDTWO{irrelevance annotations} to track resources that are no
-% longer available but have yet to be fully consumed
-% (Section~\ref{sec:lc-case-exps}). Additionally, we introduce tags to split
-% resources amongst pattern-bound variables and %^as usage environments
-% encode that all linear components of a pattern must be used in tandem.
+Despite the focus on GHC Core, the fundamental ideas for
+understanding linearity in our call-by-need calculus can be readily applied to
+other non-strict languages.
+
+Linear Core's first key idea is to delay consuming
+resources until a computation that depends on those resources is effectively
+evaluated or returned. Delayed consumption is tracked by annotating relevant
+binders with \emph{usage environments} (Section~\ref{sec:usage-environments}).
+%
+The second key idea is to have two distinct rules for case
+expressions, branching on whether the scrutinee is in Weak Head Normal
+Form or not, and using  \ROUNDTWO{irrelevance annotations} to track resources that are no
+longer available but have yet to be fully consumed
+(Section~\ref{sec:lc-case-exps}). Additionally, we introduce tags to split
+resources amongst pattern-bound variables and %^as usage environments
+encode that all linear components of a pattern must be used in tandem.
 
@@
 along with the key insights to do so, namely \ROUNDTWO{discriminating by} the WHNF-ness of
-the scrutinee, irrelevant resources, and tagging (Section~\ref{sec:lc-case-exps}).
+scrutinee, irrelevant resources, and tagging (Section~\ref{sec:lc-case-exps}).
 
@@
 
-\subsubsection{Splitting and tagging fragments}\label{sec:tagging}
+\subsubsection{Splitting and tagging fragments}
 
@@
 is updated accordingly).\\
-If $\G,\x[\irr{\D}]; \irr{\D},\D' \vdash e : \vp$ 
-then $\G[x/\irr{\D}]; \D',\xl \vdash e :\vp$.
+If $\G,\x[\irr{\D}]; \irr{\D},\D' \vdash e : \s$ 
+then $\G[x/\irr{\D}]; \D',\xl \vdash e :\s$.
 \end{restatable}
@@
   \begin{restatable}[Linear to $\Delta$-bound]{lemma}{onedelta}\label{lem:onedelta}
-If $\G; \D',\xl \vdash e :\vp$
-then $\G[\D/x],\xD; \D,\D' \vdash e : \vp$ ($\Delta$ fresh).
+If $\G; \D',\xl \vdash e :\s$
+then $\G[\D/x],\xD; \D,\D' \vdash e : \s$ ($\Delta$ fresh).
 \end{restatable}
@@
 %An unrestricted variable is equivalent to a $\D$-var with an empty usage environment.\\
-$\G,\xo; \D \vdash e : \vp$ iff $\G,\x[\cdot]; \D \vdash e : \vp$
+$\G,\xo; \D \vdash e : \s$ iff $\G,\x[\ROUNDTWO{\emptyset}]; \D \vdash e : \s$
 \end{restatable}
@@
 \ROUNDTWO{with any occurrences of $x$ in usage environments in $\G$
-  expanded into $x$'s usage environment $\Delta$, with the abuse of notation $\Gamma[\Delta/x]$}.
+  expanded into $x$'s usage environment $\Delta$, with the abuse of notation $\Gamma[\Delta/x]$.}
 
@@
 environment can be made into a linear variable as long as its
-(irrelevant) environment is dropped from the linear context.
+(irrelevant) environment is dropped from the linear context
+\ROUNDTWO{and all the occurrences of the set of variables $[\D]$ in
+  usage environments in $\Gamma$ are collapsed to $x$, written $\Gamma[x/[\Delta]]$}.
 
@@
 %
-Proof irrelevant resources can only be consumed
-indirectly, essentially encoding that the (non-WHNF) case scrutinee
+Irrelevant resources \ROUNDTWO{(Section~\ref{sec:irrev})} can only be consumed
+indirectly, essentially encoding that the non-WHNF case scrutinee
 resources must be consumed through the case binder or the linear
@@
 The \emph{Irrelevance} lemma witnesses the soundness of typing a
-case alternative with proof irrelevant resources with respect to
+case alternative with irrelevant resources with respect to
 typing the same expression with arbitrary resources. However, typing
-a case alternative with proof irrelevant resources is not complete wrt using
+a case alternative with irrelevant resources is not complete wrt using
 arbitrary resources -- a counter example needs only to use a resource
@@
 $\xo$, linear variables $\xl$, and $\D$-bound variables $\xD$. The
-development is reported in Appendices~\ref{app:proofs}
-and~\ref{app:opsem}, where type safety is established directly for the
+development is reported in \cite{techreport}, where type safety is established directly for the
 instrumented semantics and ported to the natural semantics via a
@@
     \begin{array}{ll}
-   (\llet{\xD = e}{e'} )  \Longrightarrow \llet{\xD =       e}{e'[e/x]} &                    \mbox{Inlining}\\
-       (\lambda \x[\pi][\s].~e)~e'  \Longrightarrow e[e'/x]  &                \mbox{$\beta$-reduction}\\
+   (\llet{\xD = e}{e'} )  \Longrightarrow \llet{\xD =       e}{e'[e/x]} &                    \mbox{Inlining}\\[1ex]
+       (\lambda \x[\pi][\s].~e)~e'  \Longrightarrow e[e'/x]  &                \mbox{$\beta$-reduction}\\[1ex]
       (\ccase{(K~\ov{e})}{\z[\D][\s]~\{\overline{\rho_j\to e_j}, K~\ov{x} \to e_i, \overline{\rho_k\to e_k}\}} )\Longrightarrow  e_i\ov{[e/x]}[K~\ov{e}/z] &
-                                  \mbox{Case of known constr.}\\
+                                  \mbox{Case of known constr.}\\[1.5ex]
       (\ccase{(\ccase{e_c}{\z[\D]~\{\ov{\rho_{c_i}\to e_{c_i}}\}})}{\var[w][\irr{\D,\D'}][\s']~\{\ov{\rho_i\to e_i}\}})\\
@@
       \ccase{e_c}{\z[\D]~\{\ov{\rho_{c_i} \to \ccase{e_{c_i}}{w~\{\ov{\rho_i\to e_i}\}}}\}}
-      & \mbox{Case-of-case}\\
+      & \mbox{Case-of-case}\\[1.5ex]
       (\lambda \y[\pi].~\llet{\xD = e}{e'})
       \Longrightarrow
-      \llet{\xD=e}{\lambda\y[\pi].~e'} & \mbox{Commuting let-$\lambda$}\\
-      ((\llet{v = e}{b})~a) \Longrightarrow \llet{v = e}{b~a} &\mbox{Commuting let-app.}\\
+      \llet{\xD=e}{\lambda\y[\pi].~e'} & \mbox{Commuting let-$\lambda$}\\[1ex]
+      ((\llet{v = e}{b})~a) \Longrightarrow \llet{v = e}{b~a} &\mbox{Commuting let-app.}\\[1ex]
       (\ccase{(\llet{v = e}{b})}{\{\overline{\rho\to e'}\}}) \Longrightarrow
@@
                                &
-                                 \mbox{Commuting let-case}\\
+                                 \mbox{Commuting let-case}\\[1ex]
       (\ccase{e}{\{\overline{\rho_j\to e_j}, \rho \to E[e_1],
-      \overline{\rho_k\to e_k}\}} ) \qquad (\mbox{$x$ fresh}) \\
+      \overline{\rho_k\to e_k}\}} ) \qquad (\mbox{$x$ fresh}) \\[1.5ex]
       %{{{
      \Longrightarrow \llet{x = e_1}{\ccase{e}{\{\overline{\rho_j\to e_j}, \rho \to E[x], \overline{\rho_k\to e_k}\}}} 
-                                                                        & \mbox{Commuting case-let}\\
+                                                                        & \mbox{Commuting case-let}\\[1.5ex]
       (\llet{x = (\llet{v = e}{b})}{c}) \Longrightarrow \llet{v = e}{\llet{x = b}{c}}
-                                                                      & \mbox{Commuting let-let}\\
-      f \Longrightarrow \lambda x.(f~x) & \mbox{$\eta$-expansion}\\
-      \lambda x.(f~x) \Longrightarrow f & \mbox{$\eta$-reduction}\\
+                                                                      & \mbox{Commuting let-let}\\[1ex]
+      f \Longrightarrow \lambda x.(f~x) & \mbox{$\eta$-expansion}\\[1ex]
+      \lambda x.(f~x) \Longrightarrow f & \mbox{$\eta$-reduction}\\[1ex]
       (\ccase{x}{z~\{\ov{\rho_i\to e_i}\}}) \Longrightarrow \ccase{x}{z~\{\ov{\rho_i\to  e_i[z/x]}\}}
-                                                                      & \mbox{Binder swap}\\
+                                                                      & \mbox{Binder swap}\\[1ex]
       % (\ccase{x}{z~\{\ov{\rho_i\to e_i}\}} )\Longrightarrow \ccase{x}{z~\{\ov{\rho_i\to e_i[x/z]}\}}
@@
       require $e_1$ not to capture the case binder, variables bound by the
-      pattern $\rho$ and the context $E[{-}]$ -- see Appendix~\ref{app:optimisations} for
+      pattern $\rho$ and the context $E[{-}]$ -- see \cite{techreport} for
     details).} 

### 2025-10-17 popl26.tex

@@
 linearity. \ROUNDTWO{Our system achieves this via a novel combination
-  of features: explicit tracking linear dependencies of variables
-  through \emph{usage environments}
-  (Section~\ref{sec:usage-environments}), codifying that lazy variable
-  bindings do not consume resources upon definition, but rather when
-  the variables are consumed;
+  of features: explicit tracking of captured linear resources through
+        \emph{usage environments} (Section~\ref{sec:usage-environments});
   a distinct treatment of case
-  scrutinees in weak-head normal form (Section~\ref{sec:onwhnf}),
-  identifying that linear resources captured by case expressions in
-  WHNF can be safely used in branches since the case will not further
-  evaluate its scrutinee; and a notion of irrelevant and tagged resources
+  scrutinees in weak-head normal form (Section~\ref{sec:onwhnf});
+  and a notion of irrelevant and tagged resources
   (Section~\ref{sec:irrev}), which allow for fine-grained control over
-  variable usage. Namely, linear resources occurring in non-WHNF case
-  scrutinees can no longer be used in the alternative, while the
-  pattern variables (or the case binder) must necessarily be used. } 
+  variable usage.} 
 %
@@
 % 
-Despite the focus on GHC Core, the fundamental ideas for
-understanding linearity in our call-by-need calculus can be readily applied to
-other non-strict languages.
-
-Linear Core's first key idea is to delay consuming
-resources until a computation that depends on those resources is effectively
-evaluated or returned. Delayed consumption is tracked by annotating relevant
-binders with \emph{usage environments} (Section~\ref{sec:usage-environments}).
-%
-The second key idea is to have two distinct rules for case
-expressions, branching on whether the scrutinee is in Weak Head Normal
-Form or not, and using  \ROUNDTWO{irrelevance annotations} to track resources that are no
-longer available but have yet to be fully consumed
-(Section~\ref{sec:lc-case-exps}). Additionally, we introduce tags to split
-resources amongst pattern-bound variables and %^as usage environments
-encode that all linear components of a pattern must be used in tandem.
+Despite the focus on GHC Core, the fundamental ideas for \ROUNDTWO{typing} linearity in
+our call-by-need calculus may be applied to other languages \ROUNDTWO{which
+combine non-strict features or semantics with linearity}.
+
+\ROUNDTWO{The key ideas that make Linear Core work are:}
+
+\ROUNDTWO{
+\begin{enumerate}
+    \item Usage environments (Section~\ref{sec:usage-environments})
+    \item Distinct treatment of case scrutinees in WHNF (Section~\ref{sec:onwhnf})
+    \item Irrelevant and tagged resources (Sections~\ref{sec:irrev}~and~\ref{sec:tagging})
+\end{enumerate}
+
+\noindent (1) Encodes the idea that lazy bindings do not consume resources upon
+definition, but rather when the bound variables themselves are consumed.
+Consuming a variable with some usage environment $\Delta$ equates to consuming
+the variables contained in $\Delta$.
+%
+(2) Case expressions in WHNF may capture ambient linear resources, but these
+resources can be safely used in branches since the case will not further
+evaluate its scrutinee. On the other hand, scrutinizing a non-WHNF expression
+must be treated more conservatively.
+%
+(3) Irrelevant resources make existing bindings unusable while forcing the use
+of others. Namely, linear resources occurring in non-WHNF case scrutinees can
+no longer be used in the alternatives, while the pattern variables (or the case
+binder) must necessarily be used. Tagged resources follow the observation that
+some variables must be used jointly or not at all.
+}
+%
+% Linear Core's first key idea is to delay consuming
+% resources until a computation that depends on those resources is effectively
+% evaluated or returned. Delayed consumption is tracked by annotating relevant
+% binders with \emph{usage environments} (Section~\ref{sec:usage-environments}).
+%
+% The second key idea is to have two distinct rules for case
+% expressions, \ROUNDTWO{discriminating} on whether the scrutinee is in Weak Head Normal
+% Form or not, and using  \ROUNDTWO{irrelevance annotations} to track resources that are no
+% longer available but have yet to be fully consumed
+% (Section~\ref{sec:lc-case-exps}). Additionally, we introduce tags to split
+% resources amongst pattern-bound variables and %^as usage environments
+% encode that all linear components of a pattern must be used in tandem.
 
@@
 \ROUNDTWO{with any occurrences of $x$ in usage environments in $\G$
-  expanded into $x$'s usage environment $\Delta$, with the abuse of notation $\Gamma[\Delta/x]$.}
+  expanded into $x$'s usage environment $\Delta$, with the abuse of notation $\Gamma[\Delta/x]$}.
 

### 2025-10-17 popl26.tex
Link added no more anonymization

@@
 new or modified data throughout the entire pipeline, and accept more
-optimisations. Despite our initiative in this direction\footnote{[Link removed
-due to
-anonymization]} %\footnote{https://gitlab.haskell.org/ghc/ghc/-/issues/23218} ,
+optimisations. Despite our initiative in this
+direction~\footnote{\url{https://gitlab.haskell.org/ghc/ghc/-/merge_requests/10310}},
+% \footnote{[Link removed due to anonymization]}
 we leave this as future work.

### 2025-10-18 popl26.tex
related work and metatheory review

--- 2025-10-17 popl26.tex
@@
 without garbage collection, and, more recently,
-Haskell~\cite{cite:linearhaskell}, a pure, functional, and
+Haskell~\cite{10.1145/3158093}, a pure, functional, and
 \emph{lazy} general purpose language.
@@
 Core can inform Core-to-Core optimising 
-transformations~\cite{cite:let-floating,peytonjones1997a,cite:linearhaskell},
+transformations~\cite{cite:let-floating,peytonjones1997a,10.1145/3158093},
 such as inlining and $\beta$-reduction, to produce more efficient programs.
@@
 we turn to the definition of consuming a resource from Linear
-Haskell~\cite{cite:linearhaskell}:
+Haskell~\cite{10.1145/3158093}:
 %begin{itemize}
@@
 and multiplicity polymorphism introduced by Linear
-Haskell~\cite{cite:linearhaskell} with all the key features of GHC's Core
+Haskell~\cite{10.1145/3158093} with all the key features of GHC's Core
 language (algebraic datatypes, case expressions, and recursive lazy let bindings),
@@
 variables in the runtime heap are used more than once, \ROUNDTWO{a
-  technical approach that is natural in the literature \cite{}}. We show this
+  technical approach that is natural in the literature~\cite{10.5555/1076265,10.1145/3471874.3472980}}. We show this
 semantics to be bisimilar to a
@@
 We follow an approach
-similar to that of \cite{cite:linearhaskell}, defining
+similar to that of \cite{cite:linearhaskell,10.1145/3158093}, defining
 an \emph{instrumented} operational semantics in which linear variables are
@@
 
-The \emph{Irrelevance} lemma witnesses the soundness of typing a
-case alternative with irrelevant resources with respect to
-typing the same expression with arbitrary resources. However, typing
+The \emph{Irrelevance} lemma \ROUNDTWO{shows that if a case
+  alternative is well-typed with irrelevant resources, then it is
+  well-typed with arbitrary resources.}
+%
+However, typing
 a case alternative with irrelevant resources is not complete wrt using
@@
 
-Note that since the operational semantics gets stuck if a linear
-variable is used more than once, progress ensures that linear
+\ROUNDTWO{We adopt the standard semantic technique of dealing with badly-formed
+terms, such as trying to add two values that are not numbers, by
+having such terms be stuck. In our setting, linear variables used more
+than once result in stuck states. Thus,} progress ensures that linear
 functions consume their arguments at most once if their result is
-consumed exactly once. Type preservation ensures that evaluation of a
+consumed exactly once.
+%
+%
+Type preservation ensures that evaluation of a
 closed term returns an evaluation environment with no linear variables.
@@
 case binder. While many of the transformations above are not obviously optimising,
-they are chained in the compiler pipeline to expose optimisation opportunities.
+they are chained in the compiler pipeline to expose optimisation
+opportunities.
 
@@
 linearity of all expressions bound in that program, failing if a linear
-resource is not used \emph{exactly once} according to semantic linearity of the
-$(\lambda^\pi_\Delta)$ system.
+resource is not used \emph{exactly once} according to \ROUNDTWO{our typing discipline}.
 %
@@
 directly from the description of the type system because Linear Core
-is not \emph{syntax-directed}. Specifically, the most challenging
-features are splitting linear resources amongst sub-derivations and
+is not \emph{syntax-directed}, \ROUNDTWO{given that the goal of our
+  design is to be as simple to reason with as possible.} Specifically,
+the \ROUNDTWO{features of Linear Core that are not syntax-directed are
+  essentially the}
+splitting of linear resources amongst sub-derivations and
 consuming fragments of resources through pattern-bound variables.
-Instead of non-deterministically splitting linear resources amongst
-sub-derivations, we thread input/output linear resources through each
+\ROUNDTWO{As standard in substructural typing systems, our rules
+  non-deterministically split linear resources as needed.}
+\ROUNDTWO{In our implementation,} we thread input/output linear resources through each
 derivation~\cite{DBLP:journals/tcs/CervesatoHP00}.
-%
+
 For a scrutinee not in WHNF, pattern variables in a case alternative are
@@
         programs could be understood as linear as long as all applications of
-        linear functions are reduced \emph{call-by-need}, but our system
-        conservatively rejects it in light of the flexible evaluation strategy
-        used by optimizing compilers.
+        linear functions are reduced \emph{call-by-need}.
 
@@
 % %
-Linear Haskell~\cite{cite:linearhaskell} augments the Haskell surface language
+Linear Haskell~\cite{10.1145/3158093} augments the Haskell surface language
 with linear types, but it is not concerned with extending GHC's intermediate
@@
 \paragraph{Linearity-aware Semantics}
-Several other works explore linearity aware semantics,
+Several other work explore linearity aware semantics,
 dating back to \cite{Chirimar_Gunter_Riecke_1996} and
@@
 More recently, \cite{DBLP:journals/pacmpl/ChoudhuryEEW21,DBLP:conf/esop/MarshallVO22} and
-\cite{cite:linearhaskell} all make use of usage-aware
+\cite{cite:linearhaskell,10.1145/3158093} all make use of usage-aware
 semantics. \citet{DBLP:journals/pacmpl/ChoudhuryEEW21} do so in the context of graded dependent
@@
 applies the ideas of \cite{DBLP:journals/pacmpl/ChoudhuryEEW21}
-to a non-graded (also call-by-name) setting. \citet{cite:linearhaskell} provide a similar semantics
+to a non-graded call-by-name setting. \citet{10.1145/3158093} provide a similar semantics
 to ours, but present only a naive linear typing system that does not
 leverage non-strict evaluation.
-
-\paragraph{Linear Mini-Core\label{sec:linear-mini-core}}
+\ROUNDTWO{We also note the non-strict language
+  Clean~\cite{10.1007/3-540-18317-5_20} that employs uniqueness types
+  to ensure resources are used at most once.}
+
+\paragraph{Inspirations for Linear Core\label{sec:linear-mini-core}}
 
@@
 optimising transformations.
-
+%
 We draw from Linear Mini-Core the rule for non-recursive let expressions and
@@
 transformations, when applied to Linear Core programs, preserve linearity.
+\ROUNDTWO{As previously noted in the paper, our irrelevant resources
+  are similar to 0-multiplicity
+  values from QTT~\cite{brady:LIPIcs.ECOOP.2021.9} and our tagged resources are
+  related to fractional permissions from separation
+  logic~\cite{10.5555/1760267.1760273}.}
 
@@
 
-\paragraph{Linearity-influenced optimisations}
-
-Core-to-Core transformations appear in many works across the research
-literature~\cite{cite:let-floating,peytonjones1997a,santos1995compilation,peytonjones2002secrets,baker-finch2004constructed,maurer2017compiling,Breitner2016_1000054251,sergey_vytiniotis_jones_breitner_2017},
+\paragraph{Linearity-directed optimisations}
+
+Core-to-Core transformations appear in multiple papers~\cite{cite:let-floating,peytonjones1997a,santos1995compilation,peytonjones2002secrets,baker-finch2004constructed,maurer2017compiling,Breitner2016_1000054251,sergey_vytiniotis_jones_breitner_2017},
 all designed in the context of a typed language (Core) which does not have
-linear types. However, some works~\cite{cite:let-floating,peytonjones1997a,cite:linearhaskell} have observed that
-certain optimisations (in particular, let-floating and inlining) greatly
-benefit from linearity analysis and, in order to improve those transformation,
+linear types. However, some authors~\cite{cite:let-floating,peytonjones1997a,10.1145/3158093} have observed that
+optimisations (in particular, let-floating and inlining) can greatly
+benefit from linearity analysis and, in order to improve those transformations,
 special purpose linear-type-inspired systems were created and implemented.
@@
 passes (naturally, these passes are still necessary to optimise programs not
-using explicit linearity or multiplicity annotations). Linearity may
-potentially be used to the benefit of other optimising transformations that
-currently do not take it into account.
+using explicit linearity or multiplicity annotations). \ROUNDTWO{Linearity may
+potentially be used to the benefit of other transformations.
+An obvious candidate is
+\emph{inlining}, which is applied based on heuristics from information provided
+by the \emph{cardinality analysis} pass that counts occurrences of bound
+variables.  Linearity can be used to non-heuristically inform
+the inliner~\cite{10.1145/3158093}.}
 
@@
 just type variables. Encoding multiplicities as types allows Haskell programs
-to leverage features available for types to naturally extend to multiplicities
-as well.
+to leverage \ROUNDTWO{type-level features when handling multiplicities}.
 For instance, it is possible via the use of GADTs to define a
@@
 
-\paragraph{Optimisations leveraging linearity.}
-We only briefly mentioned how linearity can inform optimisations to produce
-more performant programs. We leave exploring optimisations unblocked by
-preserving linearity in the intermediate language with Linear Core as future
-work. Linearity influenced optimising transformations have been also discussed
-by Linear Haskell~\cite{cite:linearhaskell} and
-in~\cite{cite:let-floating,peytonjones1997a}. An obvious candidate is
-\emph{inlining}, which is applied based on heuristics from information provided
-by the \emph{cardinality analysis} pass that counts occurrences of bound
-variables.  Linearity can be used to non-heuristically inform
-the inliner~\cite{cite:linearhaskell}.
+% \paragraph{Optimisations leveraging linearity.}
+% We only briefly mentioned how linearity can inform optimisations to produce
+% more performant programs. We leave exploring optimisations unblocked by
+% preserving linearity in the intermediate language with Linear Core as future
+% work. Linearity influenced optimising transformations have been also discussed
+% by Linear Haskell~\cite{cite:linearhaskell} and
+% in~\cite{cite:let-floating,peytonjones1997a}. 
 % Additionally, we argue that in Linear Core accepting more programs as linear
@@
 \paragraph{Linear Core in the Glasgow Haskell Compiler.}
-Linear Core is suitable as the intermediate language of an optimising compiler
-for a linear and lazy language such as Haskell Core, in that optimising
-transformations in Linear Core preserve types \emph{and} linearity, since Linear
-Core understands semantic linearity in the presence of laziness, unlike
-Core's current type system under which optimisations currently violate
-linearity.
+% Linear Core is suitable as the intermediate language of an optimising compiler
+% for a linear and lazy language such as Haskell Core, in that optimising
+% transformations in Linear Core preserve types \emph{and} linearity, since Linear
+% Core understands semantic linearity in the presence of laziness, unlike
+% Core's current type system under which optimisations currently violate
+% linearity.
 %
@@
 %
-A linearly typed Core that preserves linearity throughout the optimisation
-pipeline of GHC both validates the correctness of the compiler, which is
-already achieved to a great extent by preserving (non-linear) types, and
-informs optimisations, allowing the compiler to generate more performant programs.
+% A linearly typed Core that preserves linearity throughout the optimisation
+% pipeline of GHC both validates the correctness of the compiler, which is
+% already achieved to a great extent by preserving (non-linear) types, and
+% informs optimisations, allowing the compiler to generate more performant programs.
 %
@@
 new or modified data throughout the entire pipeline, and accept more
-optimisations. Despite our initiative in this direction\footnote{[Link removed
-due to
-anonymization]} %\footnote{https://gitlab.haskell.org/ghc/ghc/-/issues/23218} ,
+optimisations. Despite our initiative in this direction\footnote{https://gitlab.haskell.org/ghc/ghc/-/issues/23218},
 we leave this as future work.
@@
 Our instrumented, linearity-aware semantics (akin to that
-of~\cite{cite:linearhaskell})
+of~\cite{cite:linearhaskell,10.1145/3158093})
 is defined in Figure~\ref{fig:instrsemapp}.

### 2025-10-18 language-v4/proofs/WHNFConvSoundness.tex
appendix clean up

@@
 \begin{tabbing}
-    (1) $\G,\z[\irr{\D}];\irr{\D}, \D' \vdash_{alt} \_ \to e :^z_{\irr{\D}} \s \Rightarrow \vp$\\
-    (2) $\G, \z[\irr{\D}];\irr{\D},\D' \vdash e : \vp$\\
+    (1) $\G,\z[\irr{\D}];\irr{\D}, \D' \vdash \_ \ROUNDTWO{\Rightarrow} e :^z_{\irr{\D}} \s $\\
+    (2) $\G, \z[\irr{\D}];\irr{\D},\D' \vdash e : \s$\\
     (3) $\irr{\D}$ is used through $z$\`since $\irr{\D}$ can't otherwise be used\\\`and is introduced in this alternative uniquely \\\`(since we have multi-tier proof irrelevance, i.e.
 $\irr{\irr{\D}}\neq\irr{\D}$)\\
-    (4) $\G[z/\irr{\D}]; \z[1], \D' \vdash e : \vp$\` by Lemma~\ref{lem:deltaone} (2,3)\\
-    (5) $\G[\ov{\D_i}/\irr{\D}], \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash e : \vp$\` by Lemma~\ref{lem:onedelta} (4)\\
-    (6) $\G, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash e : \vp$\` by (3), resources in $\G$ with $\irr{\D}$ (parts) in\\\` the usage environment cannot be used in $e$\\
-    (7) $\G, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash_{alt} \_ \to e :^z_{\ov{\D_i}} \s \Rightarrow \vp$\` by $Alt\_$\\
+    (4) $\G[z/\irr{\D}]; \z[1], \D' \vdash e : \s$\` by Lemma~\ref{lem:deltaone} (2,3)\\
+    (5) $\G[\ov{\D_i}/\irr{\D}], \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash e : \s$\` by Lemma~\ref{lem:onedelta} (4)\\
+    (6) $\G, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash e : \s$\` by (3), resources in $\G$ with $\irr{\D}$ (parts) in\\\` the usage environment cannot be used in $e$\\
+    (7) $\G, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash \_ \,\ROUNDTWO{\Rightarrow}\, e :^z_{\ov{\D_i}} \s $\` by $Alt\_$\\
 \end{tabbing}
@@
 \begin{tabbing}
-    (1) $\G, \z[\irr{\D}]; \irr{\D}, \D' \vdash_{alt} K~\ov{\xo}\to e :^z_{\irr{\D}} \s \Rightarrow \vp$\\
-    (2) $\G, \z[\cdot], \ov{\xo}; \D' \vdash e : \vp$\` by inv. on $Alt0$ and def. of empty subst.\\
-    (3) $\G, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash_{alt} K~\ov{\xo}\to e :^z_{\ov{\D_i}} \s \Rightarrow \vp$\` by $Alt0$ (2)\\
+    (1) $\G, \z[\irr{\D}]; \irr{\D}, \D' \vdash K~\ov{\xo}\ROUNDTWO{\Rightarrow} e :^z_{\irr{\D}} \s $\\
+    (2) $\G, \z[\cdot], \ov{\xo}; \D' \vdash e : \s$\` by inv. on $Alt0$ and def. of empty subst.\\
+    (3) $\G, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash K~\ov{\xo}\ROUNDTWO{\Rightarrow} e :^z_{\ov{\D_i}} \s $\` by $Alt0$ (2)\\
 \end{tabbing}
@@
 \begin{tabbing}
-    (1) $\G, \z[\irr{\D}]; \irr{\D}, \D' \vdash_{alt} K~\ov{\xo},\ov{y_i{:}_{1}\s_i}^n\to e :^z_{\irr{\D}} \s \Rrightarrow \vp$\\
-    (2) $\G,\z[\irr{\D}],\ov{\xo},\ov{y_j{:}_{\D_j}\s_j}^n; \irr{\D}, \D' \vdash e : \vp$\`by inv. on $AltN_{\textrm{Not WHNF}}$\\
+    (1) $\G, \z[\irr{\D}]; \irr{\D}, \D' \vdash K~\ov{\xo},\ov{y_i{:}_{1}\s_i}^n\ROUNDTWO{\Rightarrow} e :^z_{\irr{\D}} \s $\\
+    (2) $\G,\z[\irr{\D}],\ov{\xo},\ov{y_j{:}_{\D_j}\s_j}^n; \irr{\D}, \D' \vdash e : \s$\`by inv. on $AltN_{\textrm{Not WHNF}}$\\
     (3) $\ov{\D_j} = \ov{\lctag{\irr{\D}}{K_j}}$\`by inv. on $AltN_{\textrm{Not WHNF}}$\\
     Subcase $\irr{\D}$ is consumed through $z$\\
-    (4) $\G,\ov{\xo},\ov{y_j{:}_{\D_j[z/\irr{\D}]}\s_j}^n; \z[1], \D' \vdash e : \vp$\`by Lemma~\ref{lem:deltaone} (2,subcase)\\\`and vars in $\G$ with $\irr{\D}$ cannot be used by subcase\\
+    (4) $\G,\ov{\xo},\ov{y_j{:}_{\D_j[z/\irr{\D}]}\s_j}^n; \z[1], \D' \vdash e : \s$\`by Lemma~\ref{lem:deltaone} (2,subcase)\\\`and vars in $\G$ with $\irr{\D}$ cannot be used by subcase\\
     Subcase constructing $AltN_{\textrm{WHNF}}$ ($\Mapsto$)\\
-    (5) $\G,\ov{\xo},\ov{y_j{:}_{\D_i}\s_j}^n; \z[1], \D' \vdash e : \vp$\`by $\ov{y}$ is not used, and $Weaken$\\
-    (6) $\G,\ov{\xo},\ov{y_j{:}_{\D_i}\s_j}^n, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash e : \vp$\`by Lemma~\ref{lem:onedelta} (5)\\\` and $\G$ vars do not mention $z$ by (4) \\
-    (7) $\G, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash_{alt} K~\ov{\xo},\ov{y_i{:}_{1}\s_i}^n \to e :^z_{\ov{\D_i}} \s \Mapsto \vp$\`by $AltN_{\textrm{WHNF}}$ (6)\\
+    (5) $\G,\ov{\xo},\ov{y_j{:}_{\D_i}\s_j}^n; \z[1], \D' \vdash e : \s$\`by $\ov{y}$ is not used, and $Weaken$\\
+    (6) $\G,\ov{\xo},\ov{y_j{:}_{\D_i}\s_j}^n, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash e : \s$\`by Lemma~\ref{lem:onedelta} (5)\\\` and $\G$ vars do not mention $z$ by (4) \\
+    (7) $\G, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash K~\ov{\xo},\ov{y_i{:}_{1}\s_i}^n \ROUNDTWO{\Mapsto} e :^z_{\ov{\D_i}} \s $\`by $AltN_{\textrm{WHNF}}$ (6)\\
     Subcase constructing $AltN_{\textrm{Not WHNF}}$ ($\Rrightarrow$)\\
-    (5) $\G,\ov{\xo},\ov{y_j{:}_{\lctag{\ov{\D_i}}{K_j}}\s_j}^n; \z[1], \D' \vdash e : \vp$\`by $\ov{y}$ does not consume resources\\
-    (6) $\G,\ov{\xo},\ov{y_j{:}_{\lctag{\ov{\D_i}}{K_j}}\s_j}^n, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash e : \vp$\`by Lemma~\ref{lem:onedelta} (5)\\\` and $\G$ vars do not mention $z$ by (4) \\
-    (7) $\G, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash_{alt} K~\ov{\xo},\ov{y_i{:}_{1}\s_i}^n \to e :^z_{\ov{\D_i}} \s \Rrightarrow \vp$\`by $AltN_{\textrm{Not WHNF}}$ (6)\\
+    (5) $\G,\ov{\xo},\ov{y_j{:}_{\lctag{\ov{\D_i}}{K_j}}\s_j}^n; \z[1], \D' \vdash e : \s$\`by $\ov{y}$ does not consume resources\\
+    (6) $\G,\ov{\xo},\ov{y_j{:}_{\lctag{\ov{\D_i}}{K_j}}\s_j}^n, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash e : \s$\`by Lemma~\ref{lem:onedelta} (5)\\\` and $\G$ vars do not mention $z$ by (4) \\
+    (7) $\G, \z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash K~\ov{\xo},\ov{y_i{:}_{1}\s_i}^n \ROUNDTWO{\Rightarrow} e :^z_{\ov{\D_i}} \s $\`by $AltN_{\textrm{Not WHNF}}$ (6)\\
     Subcase $\irr{\D}$ is (fully) consumed by $\ov{y}$ (after splitting)\\
-    (4) $\G,\z[\ov{y}],\ov{\xo};\ov{y_j{:}_{1}\s_j}^n, \D' \vdash e : \vp$\`by Lemma~\ref{lem:deltaone} (2, subcase)\\\` and vars in $\G$ with $\irr{\D}$ cannot be used by subcase\\
-    (5) $\G,\z[\ov{\D_i}],\ov{\xo};\ov{y_j{:}_{1}\s_j}^n, \D' \vdash e : \vp$\`by $Weaken$ and $z$ does not consume resources\\
+    (4) $\G,\z[\ov{y}],\ov{\xo};\ov{y_j{:}_{1}\s_j}^n, \D' \vdash e : \s$\`by Lemma~\ref{lem:deltaone} (2, subcase)\\\` and vars in $\G$ with $\irr{\D}$ cannot be used by subcase\\
+    (5) $\G,\z[\ov{\D_i}],\ov{\xo};\ov{y_j{:}_{1}\s_j}^n, \D' \vdash e : \s$\`by $Weaken$ and $z$ does not consume resources\\
     Subcase constructing $AltN_{\textrm{WHNF}}$ ($\Mapsto$)\\
-    (6) $\G,\z[\ov{\D_i}],\ov{\xo},\ov{y_j{:}_{\D_i}\s_j}^n; \ov{\D_i}, \D' \vdash e : \vp$\`by Lemma~\ref{lem:onedelta} (5)\\\`and vars in $\G$ do not mention $\ov{y}$ (4,5)\\
-    (7) $\G,\z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash K~\ov{\xo},\ov{y_i{:}_{1}\s_i}^n\to e :^z_{\ov{\D_i}} \s \Mapsto \vp$\`by $AltN_{\textrm{WHNF}}$\\
+    (6) $\G,\z[\ov{\D_i}],\ov{\xo},\ov{y_j{:}_{\D_i}\s_j}^n; \ov{\D_i}, \D' \vdash e : \s$\`by Lemma~\ref{lem:onedelta} (5)\\\`and vars in $\G$ do not mention $\ov{y}$ (4,5)\\
+    (7) $\G,\z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash K~\ov{\xo},\ov{y_i{:}_{1}\s_i}^n\ROUNDTWO{\Mapsto}e :^z_{\ov{\D_i}} \s$\`by $AltN_{\textrm{WHNF}}$\\
     Subcase constructing $AltN_{\textrm{Not WHNF}}$ ($\Rrightarrow$)\\
-    (6) $\G,\z[\ov{\D_i}],\ov{\xo},\ov{y_j{:}_{\lctag{\ov{\D_i}}{K_j}}\s_j}^n; \ov{\D_i}, \D' \vdash e : \vp$\`by $1 \Rightarrow \D$ lemma (5)\\\`and vars in $\G$ do not mention $\ov{y}$ (4,5)\\
-    (7) $\G,\z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash K~\ov{\xo},\ov{y_i{:}_{1}\s_i}^n\to e :^z_{\ov{\D_i}} \s \Rrightarrow \vp$\`by $AltN_{\textrm{Not WHNF}}$\\
+    (6) $\G,\z[\ov{\D_i}],\ov{\xo},\ov{y_j{:}_{\lctag{\ov{\D_i}}{K_j}}\s_j}^n; \ov{\D_i}, \D' \vdash e : \s$\`by $1 \Rightarrow \D$ lemma (5)\\\`and vars in $\G$ do not mention $\ov{y}$ (4,5)\\
+    (7) $\G,\z[\ov{\D_i}]; \ov{\D_i}, \D' \vdash K~\ov{\xo},\ov{y_i{:}_{1}\s_i}^n\ROUNDTWO{\Rrightarrow} e :^z_{\ov{\D_i}} \s $\`by $AltN_{\textrm{Not WHNF}}$\\
 \end{tabbing}

### 2025-10-18 popl26.tex
appendix clean up

@@
 
-\item We introduce Linear Core, a \ROUNDTWO{lazy} linear language with
+\item We introduce Linear Core (Section~\ref{sec:main:linear-core}), a \ROUNDTWO{lazy} linear language with
 the key features of Core (except for type equality coercions), whose type
@@
 knowledge, this is the first type system \ROUNDTWO{to accept programs
-which depend on non-strict evaluation to be understood as linear}
-(Section~\ref{sec:main:linear-core}).
+which depend on non-strict evaluation to be understood as linear}.
 
@@
 \item We discuss our implementation of Linear Core as a GHC
-  plugin\footnote{We plan to submit our implementation as a companion
-    artifact to this paper.} which checks linearity in
+  plugin which checks linearity in
 all intermediate Core programs produced during the compilation process, showing
@@
 
+\ROUNDTWO{The reader may wonder whether the complexities of our
+  system hinted above are a better alternative to simply specializing
+  the optimisation passes to produce code that passes a more naive
+  linearity check. This avenue was already explored during the
+  implementation of Linear
+  Haskell in
+  GHC\footnote{https://gitlab.haskell.org/ghc/ghc/-/issues/22123},
+ exhibiting the prohibitive cost of too often restricting
+ optimisations and producing less efficient code. Our work adheres to
+ the principle that the type system bends to the optimisations, not
+ the other way around.}
 \noindent \ROUNDTWO{While GHC's Core language is our main focus, we
@@
 binder) must necessarily be used. Tagged resources follow the observation that
-some variables must be used jointly or not at all.
-}
+some variables must be used jointly or not at all.}
+
 %
@@
 the scrutinee, irrelevant resources, and tagging (Section~\ref{sec:lc-case-exps}).
-
-% We start by introducing the Core-like language, $\dots$ usage environments as a
-% way to encode choice between the way a resource is used $\dots$\todo{$\dots$}
-%
-
-%\todo[inline]{Explicar algumas das ideias fundamentais, e apresentar as regras
-%iterativamente. Podemos começar com as triviais e avançar para os dois pontos
-%mais difíceis : Lets e Cases}
 
@@
                             $\Gamma;\Delta' \vdash e :
-                            \sigma\rightarrow_\pi \varphi$} \\ &&\qquad \mbox{and
+                            \sigma\rightarrow_\pi \tau$} \\ &&\qquad \mbox{and
                                        $\Gamma;\Delta \vdash e' : \sigma$}\\
@@
                                        \sigma_i}$}\\                                              
-(\ccase{e}{z{:}_{\Delta}\sigma~\{\overline{\rho\to e'}\}})^* &
+(\ccase{e}{z{:}_{\Delta}\sigma~\{\overline{\rho\ROUNDTWO{\Rightarrow} e'}\}})^* &
                                                                \triangleq
                                 &
-                                  \ccase{(e)^*}{z{:}_{\Delta}\sigma~\{\overline{\rho\to (e')^*}\}}\\
+                                  \ccase{(e)^*}{z{:}_{\Delta}\sigma~\{\overline{\rho\ROUNDTWO{\Rightarrow} (e')^*}\}}\\
     (\llet{x{:}_{\Delta}\sigma = e}{e'}      )^* & \triangleq & \llet{x{:}_{\Delta}\sigma  = (e)^*}{(e')^*}    \\
@@
     {\Gamma ;\Delta \vdash (\Theta \mid
-    \ccase{e}{z{:}_{\Delta'}T~\{{\dots,K~\ov{y_{i,j}} \to e', {\dots} }\}})
+    \ccase{e}{z{:}_{\Delta'}T~\{{\dots,K~\ov{y_{i,j}} \,\ROUNDTWO{\Rightarrow}\, e', {\dots} }\}})
     \Downarrow (\Theta'' \mid  v) : \tau , \Sigma}\\[1.5em]
@@
     {\Gamma ; \Delta \vdash (\Theta \mid 
-    \ccase{e}{z{:}_{\Delta'}T~\{\dots , \_\to
+    \ccase{e}{z{:}_{\Delta'}T~\{\dots , \_\,\ROUNDTWO{\Rightarrow}\,
     e'}\}) \Downarrow (\Theta'' \mid v) : \tau}
@@
 $e$ typed with the $x$s from the group.\\
-\item[Case:] $\Gamma; \Delta \vdash (\Theta \mid \ccase{e}{z{:}_{\Delta'}\s~\{K~\ov{y_i}\to e'}\}) \Downarrow (\Theta'' \mid v) : \tau$
+\item[Case:] $\Gamma; \Delta \vdash (\Theta \mid \ccase{e}{z{:}_{\Delta'}\s~\{K~\ov{y_i}\Rightarrow e'}\}) \Downarrow (\Theta'' \mid v) : \tau$
 \begin{tabbing}
-    (1) $\Gamma; \Delta \vdash \Theta \mid \ccase{e}{z{:}_{\Delta_a}\s~\{K~\ov{y_i}\to e'}\} : \tau, \Sigma$\\
-    (2) $\G, \Theta{\uparrow}^\omega; \D, \Theta{\uparrow}^1 \vdash (\ccase{e}{z{:}_{\Delta_a}\s~\{K~\ov{y_i}\to e'}\}, \Theta{\downarrow}^1_e, \Sigma) : \tau \otimes \bigotimes(\Theta{\downarrow}^1_\sigma) \otimes \bigotimes(\Sigma)$\`by \ref{lem:unwrapenv}\\
+    (1) $\Gamma; \Delta \vdash \Theta \mid \ccase{e}{z{:}_{\Delta_a}\s~\{K~\ov{y_i}\Rightarrow e'}\} : \tau, \Sigma$\\
+    (2) $\G, \Theta{\uparrow}^\omega; \D, \Theta{\uparrow}^1 \vdash (\ccase{e}{z{:}_{\Delta_a}\s~\{K~\ov{y_i}\Rightarrow e'}\}, \Theta{\downarrow}^1_e, \Sigma) : \tau \otimes \bigotimes(\Theta{\downarrow}^1_\sigma) \otimes \bigotimes(\Sigma)$\`by \ref{lem:unwrapenv}\\
     (3) $Let~\G' = \G, \Theta\uparrow^\omega; \D_a,\D_b,\D_2,\D_3 = \D,\Theta\uparrow^1$\`s.t. $(\D_2,\D_3)$ types $(\Theta\downarrow^1_e, \Sigma)$\\
-    (4) $\G'; \D_a, \D_b \vdash \ccase{e}{z{:}_{\D_a}\s\{K~\ov{y_i}\to e'\}} : \tau$\\
+    (4) $\G'; \D_a, \D_b \vdash \ccase{e}{z{:}_{\D_a}\s\{K~\ov{y_i}\Rightarrow e'\}} : \tau$\\
     Subcase $e$ is in WHNF (thus $= K~\ov{x_i}$)\\
@@
     (6) $\G', z{:}_{\D_a}\s; \D_a, \D_b \vdash_{alt} K~\ov{y_i} \to e' :^z_{\D_i} \s \Rightarrow \tau$\` by inv\\
-    Subsubcase $\vdash_{alt}$ $Alt_{WHNF}$\\
+    Subsubcase $\vdash$ $Alt_{WHNF}$\\
     (7) $\G', z{:}_{\D_a}\s, \ov{y_i{:}_\omega\s_i}, \ov{y_j{:}_{\Delta_j}\s_j}; \D_a, \D_b \vdash e' : \tau$\` by inv\\
@@
     (15) $\G; \D \vdash \Theta'' \mid v : \tau, \Sigma$\` by induction hypothesis\\
-    Subsubcase $\vdash_{alt}$ $Alt_{0}$\\
+    Subsubcase $\vdash$ $Alt_{0}$\\
     (7) $(\G', z{:}_{\D_a}\s)[\cdot/\D_a]_z, \ov{y_i{:}_\omega\sigma_i}; (\D_a, \D_b)[\cdot / \D_a] \vdash e' : \tau$\` by inv subcase\\
@@
     (6) $\G', z{:}_{\irr{\D_a}}\s; \irr{\D_a}, \D_b \vdash_{alt} K~\ov{y_{i,j}} \to e' :^z_{\irr{\D_a}} \s $\` by inv on case\\
-    Subsubcase $\vdash_{alt}$ $Alt_{NWHNF}$\\
+    Subsubcase $\vdash$ $Alt_{NWHNF}$\\
     (7) $\G', z{:}_{\irr{\D_a}}\s, \ov{y_i{:}_\omega\s_i}, \ov{y_j{:}_{\Delta_j}\s_j}; \irr{\D_a}, \D_b \vdash e' : \tau$\` for \ov{\D_j = \lctag{\irr{\D_a}}{K_j}}, by inv.\\
@@
     (19) $\G; \D \vdash \Theta'' \mid v : \tau, \Sigma$\`by induction hypothesis\\
-    Subsubcase $\vdash_{alt}$ $Alt_0$\\
+    Subsubcase $\vdash$ $Alt_0$\\
     (7) $\G', z{:}_\cdot\s, \ov{y_i{:}_\omega\s_i}; \D_b \vdash e' : \tau$\` by inv\\
@@
 \end{tabbing}
-\item[Case:] $\Gamma; \Delta \vdash (\Theta \mid \ccase{e}{z{:}_{\Delta'}\s~\{\_ \to e'}\}) \Downarrow (\Theta'' \mid v) : \tau$
+\item[Case:] $\Gamma; \Delta \vdash (\Theta \mid \ccase{e}{z{:}_{\Delta'}\s~\{\_ \Rightarrow e'}\}) \Downarrow (\Theta'' \mid v) : \tau$
 \begin{tabbing}
-    (1) $\Gamma; \Delta \vdash \Theta \mid \ccase{e}{z{:}_{\Delta_a}\s~\{\_ \to e'}\} : \tau, \Sigma$\\
-    (2) $\G, \Theta{\uparrow}^\omega; \D, \Theta{\uparrow}^1 \vdash (\ccase{e}{z{:}_{\Delta_a}\s~\{\_ \to e'}\}, \Theta{\downarrow}^1_e, \Sigma) : \tau \otimes \bigotimes(\Theta{\downarrow}^1_\sigma) \otimes \bigotimes(\Sigma)$\`by \ref{lem:unwrapenv}\\
+    (1) $\Gamma; \Delta \vdash \Theta \mid \ccase{e}{z{:}_{\Delta_a}\s~\{\_ \Rightarrow e'}\} : \tau, \Sigma$\\
+    (2) $\G, \Theta{\uparrow}^\omega; \D, \Theta{\uparrow}^1 \vdash (\ccase{e}{z{:}_{\Delta_a}\s~\{\_ \Rightarrow e'}\}, \Theta{\downarrow}^1_e, \Sigma) : \tau \otimes \bigotimes(\Theta{\downarrow}^1_\sigma) \otimes \bigotimes(\Sigma)$\`by \ref{lem:unwrapenv}\\
     (3) $Let~\G' = \G, \Theta\uparrow^\omega; \D_a,\D_b,\D_2,\D_3 = \D,\Theta\uparrow^1$\`s.t. $(\D_2,\D_3)$ types $(\Theta\downarrow^1_e, \Sigma)$\\
-    (4) $\G'; \D_a, \D_b \vdash \ccase{e}{z{:}_{\D_a}\s\{\_ \to e'\}} : \tau$\\
+    (4) $\G'; \D_a, \D_b \vdash \ccase{e}{z{:}_{\D_a}\s\{\_ \Rightarrow e'\}} : \tau$\\
     (5) $\G'; \D_a \vdash e : \sigma$\\

### 2025-10-18 popl26.tex
exceptions

@@
 
- \newcommand{\ROUNDTWO}[1]{{\color{red}#1}}
+ \newcommand{\ROUNDTWO}[1]{{\color{purple}#1}}
 %\newcommand{\ROUNDTWO}[1]{#1}
@@
     {\Theta :
-    \ccase{e}{z{:}_{\Delta'}\sigma~\{{\dots,K~\ov{y_i} \to e', {\dots}}\}}
+    \ccase{e}{z{:}_{\Delta'}\sigma~\{{\dots,K~\ov{y_i} \Rightarrow e', {\dots}}\}}
     \Downarrow \Theta'' : v}\\[1.5em]
@@
     {\Theta :
-    \ccase{e}{z{:}_{\Delta'}\sigma~\{\dots , \_\to
+    \ccase{e}{z{:}_{\Delta'}\sigma~\{\dots , \_\Rightarrow
     e'}\} \Downarrow \Theta'' : v}
@@
     {\Gamma ;\Delta \vdash (\Theta \mid
-    \ccase{e}{z{:}_{\Delta'}T~\{{\dots,K~\ov{y_{i,j}} \to e', {\dots} }\}})
+    \ccase{e}{z{:}_{\Delta'}T~\{{\dots,K~\ov{y_{i,j}} \Rightarrow e', {\dots} }\}})
     \Downarrow (\Theta'' \mid  v) : \tau , \Sigma}\\[1.5em]
@@
     {\Gamma ; \Delta \vdash (\Theta \mid 
-    \ccase{e}{z{:}_{\Delta'}T~\{\dots , \_\to
+    \ccase{e}{z{:}_{\Delta'}T~\{\dots , \_\Rightarrow
     e'}\}) \Downarrow (\Theta'' \mid v) : \tau}
@@
 transformations.
+\ROUNDTWO{We note that both Linear Haskell and our work has no
+  special treatment of exceptions. This is also the case in GHC Core,
+  where exceptions have no special status and so there is
+  no natural way of dealing with the interaction between exceptions and
+  linearity in Core itself without a major overhaul of the exception
+  mechanisms of the language. }
 

### 2025-10-18 popl26.tex
explanation of casewhnf

@@
 \]
-\ROUNDTWO{where the matching pattern is typed as before, with the
-  case-binder $z$ is annotated with the appropriate usage environment;
-  and the non-matching patterns are if the scrutinee were notin WHNF.}
+\ROUNDTWO{where we appeal to the pattern judgment for WHNF scrutinees
+  for pattern $\rho_j$, allowing the $\ov{\Delta_i}$ resources to be
+  consumed either through the case-binder or outright; and to the
+  non-WHNF judgment for the remaining branches, where the
+  $\Delta$ resources may only be consumed via the case-binder.}
 We again note that it might seem unusual to specialize a rule for expressions in

### 2025-10-18 popl26.tex
target

@@
  the other way around.}
-\noindent \ROUNDTWO{While GHC's Core language is our main focus, we
+
+ \ROUNDTWO{While GHC's Core language is our main focus, we
 note that the ideas explored in this work can potentially be applied to any
-language with non-strict features, even if the language itself is strict.}
+language with non-strict features, even if the language itself is
+strict. We assume some familiarity with Haskell and linear types for
+functional languages, but try to explain our work and motivations
+without assuming much knowledge of GHC and Core. }
 

### 2025-10-18 popl26.tex
color clarification
super and subscripts

@@
 \colorbox{limitation}{\limitationcolorname} background mark programs that are
-semantically linear, but not seen as linear by Core's type system.
-%
-Notably, the linear type system we develop in this work accepts all
-\colorbox{notyet}{\notyetcolorname} programs.
+semantically linear, but not seen as linear by Core's type system. \ROUNDTWO{Our
+system accepts all
+\colorbox{notyet}{\notyetcolorname} programs, but not the
+\colorbox{limitation}{\limitationcolorname} examples. }
+%
 %
@@
 (recursive) let bindings, case expressions, and algebraic data types.
+\ROUNDTWO{We use the $\pi$ superscript to highlight multiplicity
+  polymorphism and the $\Delta$ subscript to emphasize usage
+  environments, a key feature of our work.}
 $\lambda^\pi_\Delta$ diverges from $\lambda^q_\to$ primarily when typing lets,

### 2025-10-18 popl26.tex

@@
   implementation of Linear
-  Haskell in
-  GHC\footnote{https://gitlab.haskell.org/ghc/ghc/-/issues/22123},
+  Haskell in GHC,
  exhibiting the prohibitive cost of too often restricting
@@
  the principle that the type system bends to the optimisations, not
- the other way around.}
+ the other way around\footnote{\url{https://gitlab.haskell.org/ghc/ghc/-/blob/c85c845dc5ad539bf28f1b8c5c1dbb349e3f3d25/compiler/GHC/Core/Lint.hs\#L3225-3235}}.}
 
@@
 new or modified data throughout the entire pipeline, and accept more
-optimisations. Despite our initiative in this direction\footnote{https://gitlab.haskell.org/ghc/ghc/-/issues/23218},
+optimisations. Despite our initiative in this direction\footnote{\url{https://gitlab.haskell.org/ghc/ghc/-/merge_requests/10310}},
 we leave this as future work.

### 2025-10-18 popl26.tex
tweak

@@
 (recursive) let bindings, case expressions, and algebraic data types.
-\ROUNDTWO{We use the $\pi$ superscript to highlight multiplicity
-  polymorphism and the $\Delta$ subscript to emphasize usage
+\ROUNDTWO{We use the $\pi$ superscript to stand for multiplicity
+  and the $\Delta$ subscript to emphasize usage
   environments, a key feature of our work.}

### 2025-10-18 popl26.tex
delete redudnatnd

@@
 $\lambda^\pi_\Delta$ diverges from $\lambda^q_\to$ primarily when typing lets,
-case expressions, and alternatives (in its purpose to type semantic linearity).
+case expressions, and alternatives.
 %, and secondarily in only treating multiplicity polymorphism superficially.

### 2025-10-18 popl26.tex
25 pages

--- 2025-10-18 popl26.tex
@@
 \begin{figure}
+  {\small
 \[
@@
    \end{array}
- \]
+ \]}
  \caption{Instrumented Operational Semantics~\label{fig:linopsem}}
@@
 besides validating that our implementation is faithful to the
-$\lambda^\pi_\Delta$ system insofar as programs that should not be accepted
-are deemed ill-typed, provide further insight into the remaining
+$\lambda^\pi_\Delta$ system, provide further insight into the remaining
 details required to fully typecheck linearity in a mature optimising compiler.
@@
 of linear typing by being more flexible in understanding linearity for lazy
-languages. In this sense, it is an avenue of future
-work to apply the ideas from this work to the surface Haskell language.
+languages. It is future
+work to apply our ideas to the surface Haskell language.
 

### 2025-10-20 popl26.tex
added coverpage for supplement
Move PRINCIPLE from S1 to S3
Delete 1 backref from related work

@@
 \end{itemize}
-
-\ROUNDTWO{The reader may wonder whether the complexities of our
-  system hinted above are a better alternative to simply specializing
-  the optimisation passes to produce code that passes a more naive
-  linearity check. This avenue was already explored during the
-  implementation of Linear
-  Haskell in GHC,
- exhibiting the prohibitive cost of too often restricting
- optimisations and producing less efficient code. Our work adheres to
- the principle that the type system bends to the optimisations, not
- the other way around\footnote{\url{https://gitlab.haskell.org/ghc/ghc/-/blob/c85c845dc5ad539bf28f1b8c5c1dbb349e3f3d25/compiler/GHC/Core/Lint.hs\#L3225-3235}}.}
 
@@
 some variables must be used jointly or not at all.}
+
+\ROUNDTWO{The reader may wonder whether the complexities of our
+  system hinted above are a better alternative to simply specializing
+  the optimisation passes to produce code that passes a more naive
+  linearity check. This avenue was already explored during the
+  implementation of Linear
+  Haskell in GHC,
+ exhibiting the prohibitive cost of too often restricting
+ optimisations and producing less efficient code. Our work adheres to
+ the principle that the type system bends to the optimisations, not
+ the other way around\footnote{\url{https://gitlab.haskell.org/ghc/ghc/-/blob/c85c845dc5ad539bf28f1b8c5c1dbb349e3f3d25/compiler/GHC/Core/Lint.hs\#L3225-3235}}.}
 
@@
 transformations, when applied to Linear Core programs, preserve linearity.
-\ROUNDTWO{As previously noted in the paper, our irrelevant resources
-  are similar to 0-multiplicity
-  values from QTT~\cite{brady:LIPIcs.ECOOP.2021.9} and our tagged resources are
-  related to fractional permissions from separation
+\ROUNDTWO{Our irrelevant resources
+  have similarities with 0-multiplicity
+  values from QTT~\cite{brady:LIPIcs.ECOOP.2021.9} and, as previously noted,
+  our tagged resources are related to fractional permissions from separation
   logic~\cite{10.5555/1760267.1760273}.}
@@
 \appendix
+
+\newpage
+
+~\vspace{8cm}
+\begin{center}
+  {\huge
+    Supplemental Materials for \\ Lazy Linearity for a Core Functional
+    Language 
+    }
+\end{center}
+
+\newpage
 

### 2025-10-20 popl26.tex
revised and fit to 25'

@@
   have similarities with 0-multiplicity
-  values from QTT~\cite{brady:LIPIcs.ECOOP.2021.9} and, as previously noted,
+  values from QTT~\cite{brady:LIPIcs.ECOOP.2021.9} which are
+  disallowed from being used directly in executable code and, as previously noted,
   our tagged resources are related to fractional permissions from separation
@@
 \paragraph{Linearity-directed optimisations}
-
+%
 Core-to-Core transformations appear in multiple papers~\cite{cite:let-floating,peytonjones1997a,santos1995compilation,peytonjones2002secrets,baker-finch2004constructed,maurer2017compiling,Breitner2016_1000054251,sergey_vytiniotis_jones_breitner_2017},
@@
 these include \emph{multiplicity coercions}, optimisations leveraging
-linearity, resource inference for usage environments, and ultimately using
-Linear Core in a mature optimising compiler with lazy evaluation and linear
-types -- the Glasgow Haskell Compiler. Lastly, we discuss the
-% we hypothesize Linear($X$), a linear system parametrized on the evaluation strategy, and the
-generalization of Linear Core to the surface Haskell language.
+linearity, %resource inference for usage environments,
+and incorporating Linear Core in GHC Haskell. % Lastly, we discuss the
+% % we hypothesize Linear($X$), a linear system parametrized on the evaluation strategy, and the
+% generalization of Linear Core to the surface Haskell language.
 

### 2025-11-13 popl26.tex
25 pages

@@
 
- \newcommand{\ROUNDTWO}[1]{{\color{purple}#1}}
-%\newcommand{\ROUNDTWO}[1]{#1}
+% \newcommand{\ROUNDTWO}[1]{{\color{purple}#1}}
+\newcommand{\ROUNDTWO}[1]{#1}
 
@@
 and so we perform a translation step that makes all sharing explicit
-through let-binders (see Appendix~\ref{app:opsem}).
+through let-binders (see~\cite{techreport}).
 We write $\llet{\ov{x_i{:}_{\Delta_i}\sigma = e_i}}{e'}$ for the iterated
@@
 
-  {\small
+  {\scriptsize
   \[
@@
 %implementation, and programs that are simply rejected by Linear Core, but which
-%were not categorized.
-
+%were not categorized
+
+\vspace{-.1cm}
 {\small
@@
 %
-\begin{itemize}
-    \item Programs we expected GHC to produce, but know are not seen as linear by $\lambda^\pi_\Delta$.
+%\begin{itemize}
+% \item
+Programs we expected GHC to produce, but know are not seen as linear by $\lambda^\pi_\Delta$.
         Namely, programs resulting from the reverse binder swap which
@@
         programs could be understood as linear as long as all applications of
-        linear functions are reduced \emph{call-by-need}.
-
-    \item Programs rejected due to the lack of ``linearity coercions'' in
+        linear functions are reduced \emph{call-by-need};
+programs rejected due to the lack of ``linearity coercions'' in
         $\lambda^\pi_\Delta$. Extending Linear Core with \emph{multiplicity coercions} 
         could make these programs be understood as linear
-        (discussed in Section~\ref{sec:future-work}).
-
-    \item Programs which actually violate linearity, in that the term
+        (discussed in Section~\ref{sec:future-work});
+programs which actually violate linearity, in that the term
         violates the linear type it is annotated with. These are the programs
@@
         the type of \lstinline{build} was not general enough to accommodate linearity).
-\end{itemize}
-%
+%\end{itemize}
+
 Overall, the programs rejected by the Linear Core plugin,
@@
 
+\begin{acks}
+
+  This
+  work was supported by national funds through Fundação para a Ciência e a Tecnologia, I.P. (FCT) under projects UID/50021/2025 and UID/PRR/50021/2025.
+
+\end{acks}
+
 % \section*{Acknowledgements}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\appendix
-
-\newpage
-
-~\vspace{8cm}
-\begin{center}
-  {\huge
-    Supplemental Materials for \\ Lazy Linearity for a Core Functional
-    Language 
-    }
-\end{center}
-
-\newpage
-
-\section{Natural and Instrumented Operational Semantics}\label{app:opsem}
-
-We define a lazy big-step operational semantics for $\lambda_\Delta^\pi$ in the
-style of Launchbury~\cite{10.1145/158511.158618} natural semantics. The semantics is
-equipped with an environment which assigns expressions to variables and is
-mutated in order to express lazy evaluation. Terms are explicit about
-their sharing (in the sense of shared reductions in lazy evaluation),
-and so we perform a translation that makes all sharing explicit
-through let-binders. This translation is presented in Figure~\ref{fig:explicitsharing}.
-We write $\llet{\ov{x_i{:}_{\Delta_i}\sigma = e_i}}{e'}$ for the iterated
-let-binding of variables $\ov{x_i}$.
-\begin{figure}
-\[
-  \begin{array}{lclr}
-    x^* & \triangleq & x\\
-
-    (\Lambda p.~e)^* & \triangleq & \Lambda p.~(e)^*\\
-    (\lambda x:_\pi\sigma .e)^* & \triangleq & \lambda x:_\pi\sigma . (e)^*\\
-    (e~\pi)^* & \triangleq & (e)^*~\pi\\
-    (e~x)^* & \triangleq & (e)^*~x\\
-    (e~e')^* & \triangleq & \llet{x{:}_{\Delta}\sigma =
-                            (e')^*}{(e)^*~x}     \\ &&\qquad \mbox{with
-                            $\Gamma;\Delta' \vdash e :
-                            \sigma\rightarrow_\pi \tau$} \\ &&\qquad \mbox{and
-                                       $\Gamma;\Delta \vdash e' : \sigma$}\\
-      K~\ov{e_i} & \triangleq & \llet{\ov{x_i:_{\Delta_i} \sigma_i =
-                                   (e_i)^*}}{K~\ov{x_i}} \\ &&
-                                       \qquad\mbox{with
-                                       $\ov{\Gamma;\Delta_i\vdash
-                                       e_i :
-                                       \sigma_i}$}\\                                              
-(\ccase{e}{z{:}_{\Delta}\sigma~\{\overline{\rho\ROUNDTWO{\Rightarrow} e'}\}})^* &
-                                                               \triangleq
-                                &
-                                  \ccase{(e)^*}{z{:}_{\Delta}\sigma~\{\overline{\rho\ROUNDTWO{\Rightarrow} (e')^*}\}}\\
-    (\llet{x{:}_{\Delta}\sigma = e}{e'}      )^* & \triangleq & \llet{x{:}_{\Delta}\sigma  = (e)^*}{(e')^*}    \\
-    (\lletrec{\overline{x{:}_{\Delta}\sigma = e}}{e'}  )^* & \triangleq &\lletrec{\overline{x{:}_{\Delta}\sigma = (e)^*}}{(e')^*}  
-      \end{array}
-    \]
-    \caption{Explicit Sharing Translation\label{fig:explicitsharing}}
-  \end{figure}
-
-The natural semantics (Figure~\ref{fig:natsemapp}) is defined via the relation
-$\Theta : e \Downarrow \Theta' : v$ where $e$ is an expression, $v$ a
-value, $\Theta$ and $\Theta'$ are environments with bindings of the
-form $x:_\omega\sigma = e$ which assigns the expression $e$ to the
-variable $x$ of the given type.
-
-\begin{figure}
-\[
-  \begin{array}{c}
-    \infer[]
-    {\,}
-    {\Theta : \Lambda p.e \Downarrow \Theta : \Lambda p.e}
-    \qquad
-    \infer[]
-    {\Theta  : e\Downarrow \Theta' : \Lambda p.e' \quad \Theta' :
-    e'[\pi/p]\Downarrow \Theta'' : v}
-    {\Theta : e~\pi\Downarrow \Theta'' : v }\\[1.5em]
-    \infer[]
-    {\,}
-    {\Theta : \lambda x:_\pi \sigma . e \Downarrow \Theta : \lambda x:_\pi \sigma . e }
-    \qquad
-    \infer[]
-    {\Theta : e \Downarrow \Theta' : \lambda y:_\pi \sigma . e' \quad
-    \Theta' : e'[x/y] \Downarrow \Theta'' : v }
-    {\Theta : e~x \Downarrow \Theta'' : v}\\[1.5em]
-    \infer[]
-    {(\Theta , x:_\omega \sigma = e): e \Downarrow \Theta' : v}
-    {(\Theta , x:_\omega \sigma = e) : x \Downarrow (\Theta',x:_\omega
-    \sigma = v) : v}
-    \qquad
-    \infer[]
-    {(\Theta,x:_\omega \sigma = e) : e' \Downarrow \Theta' : v}
-    {\Theta : \llet{x{:}_{\Delta}\sigma = e}{e'}\Downarrow \Theta' : v
-    }\\[1.5em]
-    \infer[]
-    {\,}
-    {\Theta : K~\ov{x_i} \Downarrow \Theta : K~\ov{x_i}}
-    \qquad
-    \infer[]
-    {\Theta : e \Downarrow \Theta' : K~\ov{x_i} \quad
-    \Theta' : e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] \Downarrow \Theta'' : v }
-    {\Theta :
-    \ccase{e}{z{:}_{\Delta'}\sigma~\{{\dots,K~\ov{y_i} \to e', {\dots}}\}}
-    \Downarrow \Theta'' : v}\\[1.5em]
-    \infer[]
-    {\Theta : e \Downarrow \Theta' : K~\ov{x_i} \quad
-    \Theta' : e'[K~\ov{x_i}/z] \Downarrow \Theta'' : v }
-    {\Theta :
-    \ccase{e}{z{:}_{\Delta'}\sigma~\{\dots , \_\to
-    e'}\} \Downarrow \Theta'' : v}
-    \qquad
-    \infer[]
-    {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
-    % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
-    % \Downarrow \Theta_n : v_n \quad
-    (\Theta, \overline{x_i:_\omega \sigma_i = e_i}) : e' \Downarrow \Theta'
-    : v
-    }
-    {\Theta : \lletrec{\overline{x_i{:}_{\Delta}\sigma_i = e_i}}{e'}
-    \Downarrow \Theta' : v}
-    \end{array}
-  \]
-  \caption{Natural Semantics for $\lambda_\Delta^\pi$\label{fig:natsemapp}}
-\end{figure}
-
-Our instrumented, linearity-aware semantics (akin to that
-of~\cite{cite:linearhaskell,10.1145/3158093})
-is defined in Figure~\ref{fig:instrsemapp}.
-In this semantics, linear variables are erased from
-the runtime environment, ensuring that any term violating linearity
-produces a stuck state. The semantics is defined by
-the judgment $\Gamma ; \Delta \vdash (\Theta \mid e) \Downarrow (\Theta'
-\mid v) : \sigma , \Sigma$ where $\Gamma$ and $\Delta$ are typing judgments
-($\Gamma$ tracks unrestricted variables, $\Delta$ tracks linear
-assumptions) for variables occurring in $\Sigma$;
-$\Theta$ and $\Theta'$ are runtime environments, consisting of bindings of the form
-$x :_\pi \sigma = e$ or $x :_\Delta \sigma = e$ ; $e$ is the expression being evaluated and $v$
-its resulting value, both of type $\sigma$; $\Sigma$ is a list of assignments of the form
-$e : \tau$, which are expressions in which variables in $\Theta$ are
-used.
-
-\begin{figure}
-\[
-  \begin{array}{c}
-    \infer[]
-    {\,}
-    {\Gamma;\Delta \vdash (\Theta \mid \Lambda p.e) \Downarrow (\Theta \mid
-    \Lambda p.e) : \forall p . \sigma , \Sigma }
-    \\[1.5em]
-    \infer[]
-    {\Gamma ; \Delta \vdash (\Theta \mid  e) \Downarrow (\Theta' \mid
-    \Lambda p.e') : \forall p . \sigma , \Sigma \quad \Gamma ;\Delta \vdash
-      (\Theta' \mid e'[\pi/p]) \Downarrow (\Theta'' \mid v) : \sigma[\pi/p], \Sigma}
-    {\Gamma ; \Delta \vdash (\Theta \mid e~\pi) \Downarrow (\Theta''
-      \mid v) : \sigma[\pi/p] , \Sigma  }\\[1.5em]
-    \infer[]
-    {\,}
-    {\Gamma ;\Delta \vdash (\Theta \mid \lambda x:_\pi \sigma . e)
-    \Downarrow (\Theta \mid \lambda x:_\pi\sigma .e) : \sigma
-    \rightarrow_\pi \tau,\Sigma }\\[1.5em]
-    \infer[]
-    {\Gamma;\Delta \vdash  (\Theta \mid e) \Downarrow (\Theta' \mid
-    \lambda y:_1 \sigma . e') : \sigma \rightarrow_1 \tau , x:\sigma, \Sigma
-    \quad
-    \Gamma ; \Delta \vdash (\Theta' , y:_{1} \sigma = x \mid e') \Downarrow (\Theta''
-    \mid v) : \tau , \Sigma  }
-    {\Gamma ; \Delta\vdash (\Theta \mid  e~x ) \Downarrow (\Theta''
-    \mid v) : \tau , \Sigma }\\[1.5em]
-        \infer[]
-    {\Gamma;\Delta \vdash  (\Theta \mid e) \Downarrow (\Theta' \mid
-    \lambda y:_\omega \sigma . e') : \sigma \rightarrow_\omega \tau , x:\sigma, \Sigma
-    \quad
-            \Gamma ; \Delta \vdash (\Theta' \mid e'[x/y]) \Downarrow (\Theta''
-    \mid v) : \tau , \Sigma  }
-    {\Gamma ; \Delta\vdash (\Theta \mid  e~x ) \Downarrow (\Theta''
-    \mid v) : \tau , \Sigma }\\[1.5em]
-       \infer[]
-    {\Gamma ; \Delta  \vdash (\Theta, x:_{\Delta'} \sigma = e \mid e)
-    \Downarrow (\Theta' \mid v) : \sigma , \Sigma \\
-      \D'' = \D' \setminus (\Theta \setminus \Theta')
-      }
-    {\Gamma ; \Delta \vdash (\Theta , x:_{\Delta'} \sigma = e \mid x)
-      \Downarrow (\Theta' , x:_{\Delta''} \sigma = v \mid v) : \sigma, \Sigma}
-    \\[1.5em]
-           \infer[]
-    {\Gamma ;\Delta  \vdash (\Theta \mid e)
-    \Downarrow (\Theta' \mid v) : \sigma,\Sigma}
-    {\Gamma ; \Delta \vdash (\Theta , x:_1 \sigma = e \mid x)
-    \Downarrow (\Theta' \mid v) : \sigma,\Sigma}
-    \\[1.5em]
-
-    \infer[]
-    {\Gamma ;\Delta \vdash (\Theta,x:_{\Delta'} \sigma = e \mid e')
-    \Downarrow (\Theta' \mid v) : \tau,\Sigma}
-    {\Gamma ; \Delta \vdash (\Theta \mid \llet{x{:}_{\Delta'}\sigma =
-    e}{e'})\Downarrow (\Theta' \mid v) : \tau , \Sigma
-    }\\[1.5em]
-
-    \infer[]
-    {\,}
-    {\Gamma ; \Delta \vdash (\Theta \mid K~\ov{x_i}) \Downarrow
-    (\Theta \mid K~\ov{x_i}) : T , \Sigma}
-    \\[1.5em]
-
-    \infer[]
-    { K : \ov{ \sigma_i
-    \rightarrow_\omega \sigma_j \lolli} T
-    \\
-    \Gamma, z{:}_{\ov{y_j}}T, \ov{y_i :_\omega \sigma_i} ;\Delta , \ov{y_j :_1 \sigma_j} \vdash (\Theta \mid e) \Downarrow (\Theta' \mid
-      K~\ov{x_i}) : T , e' : \tau ,\Sigma \\
-    \Gamma ;\Delta \vdash (\Theta' \mid e'\ov{[x_i/y_{i,j}]}[K~\ov{x_i}/z]
-    ) \Downarrow (\Theta'' \mid v) :\tau , \Sigma }
-    {\Gamma ;\Delta \vdash (\Theta \mid
-    \ccase{e}{z{:}_{\Delta'}T~\{{\dots,K~\ov{y_{i,j}} \,\ROUNDTWO{\Rightarrow}\, e', {\dots} }\}})
-    \Downarrow (\Theta'' \mid  v) : \tau , \Sigma}\\[1.5em]
-
-    \infer[]
-    {K : \ov{ \sigma_i
-    \rightarrow_\omega \sigma_j \lolli} T\\
-    \Gamma ; \Delta, z{:}_1\s \vdash (\Theta \mid e) \Downarrow (\Theta' \mid
-    K~\ov{x_i}) : T , e', \Sigma \quad
-    \Gamma ;\Delta \vdash (\Theta' \mid  e'[K~\ov{x_i}/z])
-    \Downarrow (\Theta'' \mid v) : \tau,\Sigma }
-    {\Gamma ; \Delta \vdash (\Theta \mid 
-    \ccase{e}{z{:}_{\Delta'}T~\{\dots , \_\,\ROUNDTWO{\Rightarrow}\,
-    e'}\}) \Downarrow (\Theta'' \mid v) : \tau}
-    \\[1.5em]
-
-     \infer[]
-     {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
-     % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
-     % \Downarrow \Theta_n : v_n \quad
-     (\Gamma ; \Delta \vdash (\Theta, \overline{x_i:_{\Delta'} \sigma_i =
-     e_i} \mid e' )\Downarrow (\Theta'
-     \mid v) : \tau
-     }
-     {\Gamma ;\Delta \vdash (\Theta \mid \lletrec{\overline{x_i{:}_{\Delta'}\sigma_i = e_i}}{e'})
-     \Downarrow (\Theta' \mid v) : \tau}
-
-   \end{array}
- \]
- \caption{Instrumented Semantics for $\lambda_\Delta^\pi$\label{fig:instrsemapp}}
- \end{figure}
-
- Following~\cite{Gunter1993APA,cite:linearhaskell}, we also consider
- partial derivations of our semantics in order to be able to state a
- progress result. Given our set of rules defining judgments of the
- form $a \Downarrow b$ with ordered premises, we define a total
- derivation of $a \Downarrow b$ as a tree in the standard way. A
- partial derivation $a \Downarrow?$ is either a root labelled with $a
- \Downarrow?$ or an application of a rule matching $a$ on the left of
- the evaluation arrow, where exactly one of the premises $a'
- \Downarrow?$ has a partial  derivation, all the premises to the left
- of $a' \Downarrow?$ have a total derivation, and the premises to the
- right of $a'\Downarrow?$ are not known yet.
- %
-
- By construction, in a partial derivation, there is only one premise
- $b\Downarrow?$ with no sub-derivation. We call $b$ the head of such a
- partial derivation.
-We write $a \Downarrow^* b$ for the relation which holds when $b$ is
-the head of some partial derivation with root $a \Downarrow?$.  We
-write $a \Downarrow^* b$  the partial evaluation relation.
-
-Below, we write $\Theta {\uparrow^1}$ for the linear bindings in evaluation
-environment $\Theta$ and $\Theta {\uparrow^\omega}$ for all other bindings.
-$\Theta\downarrow^1_e$ refers to the expressions bound to linear variables in
-the runtime environment and $\Theta\downarrow^1_\sigma$ refers to the types of
-those expressions.
-
- \begin{definition}[Environment Expansion]
-   We rely on an expansion of a runtime environment $\Theta$ for the evaluation of a term $e$ into a (typed) term, written $[\Theta \mid e]$, defined inductively as follows:
-   \[
-     \begin{array}{lcl}
-       \left[ \cdot \mid e \right] & \triangleq & e\\
-       \left[ \Theta , \xl = e' \mid e \right] & \triangleq  & [\Theta \mid (\lambda \xl . e)~e']\\
-       \left[\Theta , x:_\Delta \sigma = e' \mid e \right] & \triangleq & [\Theta \mid \llet{x:_\Delta \sigma = e'}{e}]
-    \end{array}
-   \]
- \end{definition}
-
-We note that for the treatment of recursive let groups, environment
-expansion needs to happen for the entire group of let-bound
-variables. This is easily achieved by extending the runtime
-environment with information about recursive let groupings, which we
-omit for the sake of presentation.
-
- \begin{definition}[Well-typed State]
-   We write $\G;\D \vdash \Theta \mid e : \tau,\Sigma$ to denote the following typing judgment:
-
-   \[
-\G ; \D \vdash [ \Theta \mid (e,\mathit{terms}(\Sigma))] : \tau \otimes \bigotimes(\Sigma)
-   \]
-   where $\mathit{terms}(e_1 : \sigma_1 , \dots , e_n : \sigma_n)$ stands for the tuple $(e_1 ,( \dots , (e_n , ())))$ and
-   $\bigotimes(e_1 : \sigma_1 , \dots , e_n : \sigma_n)$ for $\sigma_1 \otimes ( \dots (\sigma_n \otimes ()))$.
- \end{definition}
-
- \begin{lemma}~\label{lem:unwrapenv}
-   $\G ; \D \vdash [ \Theta \mid e] : \sigma$
-     iff $\G , \Theta\uparrow^{\omega} ; \D , \Theta\uparrow^1 \vdash  (e, \Theta\downarrow^1_e) : \sigma \otimes \bigotimes(\Theta\downarrow^1_\sigma)$
-   \end{lemma}
-\begin{proof}
-By structural induction on the environment expansion definition in the
-forward direction and by a straightforward induction on $\Theta$ in the
-converse direction. We show only the forward direction.
-\begin{description}
-%\item[Case:] $\left[\cdot \mid e\right] \triangleq e$ Trivial
-\item[Case:] $\left[\Theta, \xl = y \mid e\right] \triangleq \left[\Theta \mid (\lambda \xl. e)~y\right]$
-\begin{tabbing}
-    (1) $\G; \D \vdash \left[ \Theta, \xl = y \mid e \right] : \phi$\\
-    (2) $\G; \D \vdash \left[ \Theta \mid (\lambda \xl. e)~y \right] : \phi$\` by def.\\
-    (3) $\G, \Theta\uparrow^\omega; \D, \Theta\uparrow^1 \vdash ((\lambda \xl. e)~y, \Theta\downarrow^1_e) : \phi \otimes \bigotimes(\Theta\downarrow^1_\sigma)$\` by i.h.\\
-    (4) $Let~\D_1,\D_2,\D_3 = \D, \Theta\uparrow^1 $\\
-    (5) $\G, \Theta\uparrow^\omega; \D_3 \vdash \Theta\downarrow^1_e : \bigotimes(\Theta\downarrow^1_\sigma)$\\
-    (6) $\G, \Theta\uparrow^\omega; \D_2 \vdash y : \s$\` by inv\\
-    (7) $\G, \Theta\uparrow^\omega; \D_1 \vdash (\lambda \xl. e) : \s \to \phi $\` by inv\\
-    (8) $\G, \Theta\uparrow^\omega; \D_1, \xl \vdash e : \phi $\` by inv\\
-    (9) $\G, \Theta\uparrow^\omega; \D_1, \D_2, \xl \vdash (e, y) : \phi \otimes \sigma$\` by pair\\
-    (10) $\G, \Theta\uparrow^\omega; \D, \xl \vdash (e, y, \Theta\downarrow^1_e) : \phi \otimes \sigma \otimes \bigotimes(\Theta\downarrow^1_\sigma)$\` by pair\\
-\end{tabbing}
-\item[Case:] $\left[\Theta, \xD = e' \mid e\right] \triangleq \left[\Theta \mid \llet{\xD = e'}{e}\right]$
-\begin{tabbing}
-    (1) $\G; \D' \vdash \left[ \Theta, \xD = y \mid e \right] : \phi$\\
-    (2) $\G; \D' \vdash \left[ \Theta \mid \llet{\xD = e'}{e} \right] : \phi$\` by def.\\
-    (3) $\G, \Theta\uparrow^\omega; \D', \Theta\uparrow^1 \vdash (\llet{\xD = e'}{e}, \Theta\downarrow^1_e) : \phi \otimes \bigotimes(\Theta\downarrow^1_\sigma)$\` by i.h.\\
-    (4) $Let~\D,\D_2,\D_3 = \D', \Theta\uparrow^1 $\\
-    (5) $\G, \Theta\uparrow^\omega; \D_3 \vdash \Theta\downarrow^1_e : \bigotimes(\Theta\downarrow^1_\sigma)$\\
-    (6) $\G, \Theta\uparrow^\omega; \D \vdash e' : \s$\` by inv on let\\
-    (7) $\G, \Theta\uparrow^\omega, \xD; \D,\D_2 \vdash e : \phi $\` by inv on let\\
-    (8) $\G, \Theta\uparrow^\omega, \xD; \D,\D_2,\D_3 \vdash (e, \Theta\downarrow^1_e) : \phi \otimes \bigotimes(\Theta\downarrow^1_\sigma)$\` by pair\\
-\end{tabbing}
-\end{description}
-\end{proof}
-
-\subsection{Bisimulation}
-
-We relate our two operational semantics through a bisimulation,
-allowing us to transfer type safety of the instrumented semantics to
-the lazy big-step semantics. The relation essentially allows the
-execution environment to treat some unrestricted bindings
-of the natural semantics as linear.
-
-\begin{definition}[Bisimulation]
-We define the relation $\gamma(\Theta : e)(\G ; \D \vdash \Theta' \mid
-e' : \tau,\Sigma)$ between a well-typed state and an ordinary state of
-the lazy big-step semantics if $e = e'$ and $\Theta'{\uparrow^\omega}
-\leq \Theta$.
-\end{definition}
-
-\begin{lemma}~\label{lem:naturaltoinstr}
-  \begin{itemize}
-    \item For all $\gamma(\Theta : e)(\G ; \D \vdash \Theta' \mid
-e : \tau,\Sigma)$ such that $\Theta : e \Downarrow \Theta'' : v$
-there exists a well-typed state $\G ; \D \vdash (\Theta''' \mid v) :
-\tau,\Sigma$ such that $\G ; \D \vdash (\Theta' \mid e) \Downarrow
-(\Theta''' \mid v) : \tau,\Sigma$ and $\gamma(\Theta'' : v)(\G ; \D
-\vdash \Theta''' \mid v : \tau,\Sigma)$
-\item For all $\gamma(\Theta : e)(\G ; \D \vdash \Theta' \mid
-e : \tau,\Sigma)$ such that $\Theta : e \Downarrow^* \Theta'' : v$
-there exists a well-typed state $\G ; \D \vdash (\Theta''' \mid v) :
-\tau,\Sigma$ such that $\G ; \D \vdash (\Theta' \mid e) \Downarrow^*
-(\Theta''' \mid v) : \tau,\Sigma$ and $\gamma(\Theta'' : v)(\G ; \D
-\vdash \Theta''' \mid v : \tau,\Sigma)$
-\end{itemize}
-\end{lemma}
-\begin{proof}
-By induction on the given operational semantics rules.
-  \end{proof}
-
-\begin{lemma}~\label{lem:instrtonatural}
-  \begin{itemize}
-\item  For all $\gamma(\Theta : e)(\G ; \D \vdash \Theta' \mid
-e : \tau,\Sigma)$ such that $\G ; \D \vdash (\Theta' \mid e) \Downarrow
-(\Theta'' \mid v) : \tau,\Sigma$, there exists a state $\Theta''' : v$
-such that $\Theta : e \Downarrow \Theta''' : v$ and
-$\gamma(\Theta''' : v)(\G ; \D \vdash (\Theta'' \mid v) :
-\tau,\Sigma)$
-\item For all $\gamma(\Theta : e)(\G ; \D \vdash \Theta' \mid
-e : \tau,\Sigma)$ such that $\G ; \D \vdash (\Theta' \mid e) \Downarrow^*
-(\Theta'' \mid v) : \tau,\Sigma$, there exists a state $\Theta''' : v$
-such that $\Theta : e \Downarrow^* \Theta''' : v$ and
-$\gamma(\Theta''' : v)(\G ; \D \vdash (\Theta'' \mid v) :
-\tau,\Sigma)$
-   \end{itemize}
- \end{lemma}
-\begin{proof}
-By induction on the given operational semantics rules.
-  \end{proof}
-
-\section{Type Safety Proofs}\label{app:proofs}
-
-In the following we write $\Mapsto$ to stand for
-$\Rightarrow_\mathsf{WHNF}$ and $\Rrightarrow$ for $\Rightarrow_\mathsf{NWHNF}$.
-
-\subsection{Auxiliary Lemmas}
-The following lemmas presuppose the variable names across the typing
-contexts are distinct.
-
-  \begin{lemma}[Linear to $\Delta$-bound]~\label{lem:onedelta-app}
-
-    \begin{enumerate}
-
-      \item If $\G; \D',\xl \vdash e :\vp$
-        then $\G[\D/x],\xD; \D,\D' \vdash e : \vp$ (with $\Delta$ fresh).
-      \item If $\G ; \D' , \xl \vdash_{alt} \rho \rightarrow e
-        :^z_{\Delta_s} \sigma \Rightarrow \rho$ then
-        $\G[\D/x],\xD;\D,\D' \vdash_{alt} \rho \rightarrow e
-        :^z_{\Delta_s} \sigma \Rightarrow \rho$ (with $\Delta$ fresh).
-
-      \end{enumerate}
-\end{lemma}
-\begin{proof}
-Straightforward induction on the structure of the given derivation,
-reconstructing the original derivation and applying rule
-$\mathit{Var}_\Delta$ where rule $\mathit{Var}_1$ was applied on
-$x$ in the original derivation.
-\end{proof}
-\onedelta*
-
-\begin{proof}
-By Lemma~\ref{lem:onedelta-app}(1).
-\end{proof}
-
-\begin{lemma}[$\Delta$-bound to Linear]~\label{lem:deltaone-app}
-
-  \begin{enumerate}
- \item If $\G,\x[\irr{\D}]; \irr{\D},\D' \vdash e : \vp$
-   then $\G[x/\irr{\D}]; \D',\xl \vdash e :\vp$.
- \item If $\G ,\x[\irr{\D}]; \irr{\D}, \D' \vdash_{alt} \rho \rightarrow e
-   :^z_{\Delta_s} \sigma \Rightarrow \rho$ then
-   $\G[x/\irr{\D}];\D',\xl \vdash_{alt} \rho \rightarrow e
-   :^z_{\Delta_s} \sigma \Rightarrow \rho$.
-  \end{enumerate}
-\end{lemma}
-\begin{proof}
-Straightforward induction on the structure of the given derivation,
-reconstructing the original derivation and applying rule
-$\mathit{Var}_1$ where rule $\mathit{Var}_\D$ was applied on
-$x$ in the original derivation.
-\end{proof}
-
-\deltaone*
-\begin{proof}
-By Lemma~\ref{lem:deltaone-app}(1).
-\end{proof}
-
-\begin{lemma}[Unrestricted and $\Delta$-bound]~\label{lem:undelta-app}
-  \begin{enumerate}
-    \item $\G,\xo; \D \vdash e : \vp$ iff $\G,\x[\cdot]; \D \vdash e :
-      \vp$
-    \item $\G,\xo;\D \vdash_{alt} \rho \rightarrow e
-      :^z_{\Delta_s} \sigma \Rightarrow \rho$ iff
-      $\G,\x[\cdot];\D \vdash_{alt} \rho \rightarrow e
-   :^z_{\Delta_s} \sigma \Rightarrow \rho$
-  \end{enumerate}
-\end{lemma}
-\begin{proof}
-Straightforward induction on the structure of the given derivation,
-noting that rules $\mathit{Var}_\Delta$ and $\mathit{Var}_\omega$
-coincide when $\Delta$ is empty.
-\end{proof}
-
-\undelta*
-
-\begin{proof}
-By Lemma~\ref{lem:undelta-app}.
-\end{proof}
-
-\subsection{Irrelevance\label{sec:proof:irrelevance}}
-
-\input{language-v4/proofs/WHNFConvSoundness}
-
-\subsection{Substitution Lemmas\label{sec:proof:substitution-lemmas}}
-
-\input{language-v4/proofs/LinearSubstitutionLemma}
-
-\input{language-v4/proofs/UnrestrictedSubstitutionLemma}
-
-\input{language-v4/proofs/DeltaSubstitutionLemma}
-
-\subsection{Type Preservation\label{sec:proof:type-preservation}}
-
-% \input{language-v4/proofs/TypePreservationTheorem}
-\begin{theorem}[Type Preservation]\label{thm:typepres}
-  Let $\G ; \D \vdash (\Theta \mid e) \Downarrow (\Theta' \mid v) :
-  \tau , \Sigma$ or $\G ; \D \vdash (\Theta \mid e) \Downarrow^* (\Theta' \mid v) :
-  \tau , \Sigma$.
-  If $\G ; \D \vdash \Theta \mid e : \tau , \Sigma$ then $\G ; \D \vdash \Theta' \mid v : \tau , \Sigma$
-
-\end{theorem}
-
-\begin{proof}
-By structural induction on the instrumented operational semantics. We
-omit many trivial cases.
-\begin{description}
-    \item[Case:] $\Gamma; \Delta \vdash (\Theta, \xl = y \mid x) \Downarrow (\Theta' \mid v) : \s, \Sigma$
-\begin{tabbing}
-    (1) $\G; \D \vdash \Theta, \xl = y \mid x : \s, \Sigma$\\
-    (2) $\G, \Theta{\uparrow}^\omega; \D, \Theta{\uparrow}^1, \xl \vdash (x, y, \Theta{\downarrow}^1_e, \Sigma) : \s \otimes \s \bigotimes(\Theta{\downarrow^1_e}) \bigotimes(\Sigma)$\`by \ref{lem:unwrapenv}\\
-    (3) $Let~\G';\D_1,\D_2,\D_3 = \G, \Theta{\uparrow}^\omega; \D, \Theta{\uparrow}^1$\` s.t. $(\D_2,\D_3)$ types $(\Theta{\downarrow}^1_e,\Sigma)$\\
-    (4) $\G'; \xl \vdash x : \s$\\
-    (5) $\G'; \D_1 \vdash y : \s$\\
-    (6) $\G'; \D_1, \D_2, \D_3 \vdash (y, \Theta{\downarrow}^1_e, \Sigma) : \s \bigotimes(\Theta{\downarrow^1_e}) \bigotimes(\Sigma)$\\
-    (7) $\G; \D \vdash \Theta \mid y : \s, \Sigma$\`by well-typed state\\
-    (8) $\G; \D \vdash \Theta' \mid v : \s, \Sigma$\` by induction hypothesis\\
-\end{tabbing}
-    \item[Case:] $\Gamma; \Delta \vdash (\Theta, x{:}_{\D'}\s = e \mid x) \Downarrow (\Theta', x{:}_{\D_a}\s = v \mid v) : \s, \Sigma$
-\begin{tabbing}
-    (1) $\Gamma; \Delta \vdash (\Theta, x{:}_{\D'}\s = e \mid x) : \s, \Sigma$\\
-    (2) $\G, \Theta{\uparrow}^\omega; \D, \Theta{\uparrow}^1 \vdash \llet{x{:}_{\D'} = e}{(x, \Theta{\downarrow}^1_e, \Sigma)} : \s \bigotimes(\Theta{\downarrow}^1_\s) \bigotimes(\Sigma)$\`by \ref{lem:unwrapenv}\\
-    (3) $Let~\D',\D'' = \D,\Theta{\uparrow}^1$\\
-    (4) $\G, \Theta{\uparrow}^\omega; \D' \vdash e : \s$\`by inv on let (2)\\
-    (5) $\G, \Theta{\uparrow}^\omega; \D', \D'' \vdash (x, \Theta{\downarrow}^1_e, \Sigma) : \s \bigotimes(\Theta{\downarrow}^1_\s) \bigotimes(\Sigma)$\`by inv on let (2)\\
-    (6) $\G, \Theta{\uparrow}^\omega; \D' \vdash x : \s$\\
-    (7) $\G, \Theta{\uparrow}^\omega; \D'' \vdash (\Theta{\downarrow}^1_e, \Sigma) : \bigotimes(\Theta{\downarrow}^1_\s) \bigotimes(\Sigma)$\\
-    (8) $\G, \Theta{\uparrow}^\omega; \D', \D'' \vdash (e, \Theta{\downarrow}^1_e, \Sigma) : \tau \bigotimes(\Theta{\downarrow}^1_\s) \bigotimes(\Sigma)$\` by pair (4,7)\\
-    (9) $\G; \D \vdash \Theta \mid e : \tau,\Sigma$\` by well-typed state\\
-    (10) $\G; \D \vdash \Theta \mid v : \tau,\Sigma$\` by induction hypothesis\\
-    (11) $Let~\G';\D_a,\D_b = \G, \Theta'{\uparrow}^\omega, \D, \Theta'{\uparrow}^1$\\
-    (11a) $\G'; \D_a \vdash v : \s$
-    (11b) $\G'; \D_b \vdash (\Theta'{\downarrow}^1_e, \Sigma) : \bigotimes(\Theta'{\downarrow}^1_\s) \bigotimes(\Sigma)$\\
-    (12) $\G'; \D_a \vdash \llet{x{:}_{\D_a}\s = v}{v} : \s$\`by let (11a,11a)\\
-    (13) $\G; \D \vdash \Theta' \mid \llet{x{:}_{\D_a}\s = v}{v} : \s, \Sigma$\`by well-typed state\\
-    (14) $\G; \D \vdash \Theta', x{:}_{\D_a}\s = v \mid v : \s, \Sigma$\`by env. expansion\\
-\end{tabbing}
-Note: It's slightly more elaborate to accommodate recursion. The runtime
-environment bindings must be tagged with the recursive group such that the
-environment expansion can decide whether to unwrap to a letrec group or
-not. Unlike the case proved here, when the $\D$-binding belongs to a let group,
-the whole group is unwrapped and we can invert the assumption to get an
-$e$ typed with the $x$s from the group.\\
-\item[Case:] $\Gamma; \Delta \vdash (\Theta \mid \ccase{e}{z{:}_{\Delta'}\s~\{K~\ov{y_i}\Rightarrow e'}\}) \Downarrow (\Theta'' \mid v) : \tau$
-\begin{tabbing}
-    (1) $\Gamma; \Delta \vdash \Theta \mid \ccase{e}{z{:}_{\Delta_a}\s~\{K~\ov{y_i}\Rightarrow e'}\} : \tau, \Sigma$\\
-    (2) $\G, \Theta{\uparrow}^\omega; \D, \Theta{\uparrow}^1 \vdash (\ccase{e}{z{:}_{\Delta_a}\s~\{K~\ov{y_i}\Rightarrow e'}\}, \Theta{\downarrow}^1_e, \Sigma) : \tau \otimes \bigotimes(\Theta{\downarrow}^1_\sigma) \otimes \bigotimes(\Sigma)$\`by \ref{lem:unwrapenv}\\
-    (3) $Let~\G' = \G, \Theta\uparrow^\omega; \D_a,\D_b,\D_2,\D_3 = \D,\Theta\uparrow^1$\`s.t. $(\D_2,\D_3)$ types $(\Theta\downarrow^1_e, \Sigma)$\\
-    (4) $\G'; \D_a, \D_b \vdash \ccase{e}{z{:}_{\D_a}\s\{K~\ov{y_i}\Rightarrow e'\}} : \tau$\\
-    Subcase $e$ is in WHNF (thus $= K~\ov{x_i}$)\\
-    (5) $\G'; \D_a \Vdash K~\ov{x_i} : \s \gtrdot \ov{\D_i}$\` by subcase\\
-    (6) $\G', z{:}_{\D_a}\s; \D_a, \D_b \vdash_{alt} K~\ov{y_i} \to e' :^z_{\D_i} \s \Rightarrow \tau$\` by inv\\
-    Subsubcase $\vdash$ $Alt_{WHNF}$\\
-    (7) $\G', z{:}_{\D_a}\s, \ov{y_i{:}_\omega\s_i}, \ov{y_j{:}_{\Delta_j}\s_j}; \D_a, \D_b \vdash e' : \tau$\` by inv\\
-    (8) $\Theta' = \Theta$\` in this subcase because $(\Theta \mid K~\ov{x_i})\Downarrow(\Theta \mid K~\ov{x_i})$\\
-    (9) $\G'; \ov{\D_i} \vdash K~\ov{x_i} : \s$\` by inv (5)\\
-    (10) $\ov{\G'; \cdot \vdash x_i : \s_i}$\` by inv (9) for unrestricted $x$s\\
-    (11) $\ov{\G'; \D_j \vdash x_j : \s_j}$\` by inv (9) for linear fields $x$s\\
-    (12) $\G', z{:}_{\D_a}\s; \D_a, \D_b \vdash e'\ov{[x_{i,j}/y_{i,j}] : \tau}$\` by $\D$-subst and $\omega$-subst lemmas (7,10,11)\\
-    (13) $\G'; \D_a, \D_b \vdash e'\ov{[x_{i,j}/y_{i,j}]}[K~\ov{x_{i,j}}/z] : \tau$\` by $\D$-subst lemma and $\ov{\D_i} = \D_a$ (9,12)\\
-    (14) $\G; \D \vdash \Theta \mid e'\ov{[x_{i,j}/y_{i,j}]}[K~\ov{x_{i,j}}/z] : \tau, \Sigma$\` by well-typed state\\
-    (15) $\G; \D \vdash \Theta'' \mid v : \tau, \Sigma$\` by induction hypothesis\\
-    Subsubcase $\vdash$ $Alt_{0}$\\
-    (7) $(\G', z{:}_{\D_a}\s)[\cdot/\D_a]_z, \ov{y_i{:}_\omega\sigma_i}; (\D_a, \D_b)[\cdot / \D_a] \vdash e' : \tau$\` by inv subcase\\
-    (8) $\G', z{:}_{\cdot}, \ov{y_i{:}_\omega\sigma_i}; \D_b \vdash e' : \tau$\` by $z$ cannot occur in $\G'$\\
-    (9) Since $\G';\D_a \Vdash K~\ov{x_i}$, in this subcase $\D_a = \cdot$\\
-    (10) $\G', \z{:}_\cdot\sigma, \ov{y{:}_\omega\sigma}; \D_b \vdash e' : \tau$\` by (9)\\
-    (11) $\G', \z{:}_\cdot\sigma; \D_b \vdash e'\ov{[x_i/y_i]} : \tau$\` by $\omega$-subst (10)\\
-    (12) $\G'; \D_b \vdash e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] : \tau$\` by $\D$-subst\\
-    (13) $\G; \D \vdash \Theta \mid e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] : \tau, \Sigma$\` by well-typed state\\
-    (14) $\G; \D \vdash \Theta'' \mid v : \tau, \Sigma$\`by induction hypothesis\\
-    Subcase $e$ is not in WHNF\\
-    (5) $\G'; \D_a \vdash e : \s$\` by inv on case\\
-    (6) $\G', z{:}_{\irr{\D_a}}\s; \irr{\D_a}, \D_b \vdash_{alt} K~\ov{y_{i,j}} \to e' :^z_{\irr{\D_a}} \s $\` by inv on case\\
-    Subsubcase $\vdash$ $Alt_{NWHNF}$\\
-    (7) $\G', z{:}_{\irr{\D_a}}\s, \ov{y_i{:}_\omega\s_i}, \ov{y_j{:}_{\Delta_j}\s_j}; \irr{\D_a}, \D_b \vdash e' : \tau$\` for \ov{\D_j = \lctag{\irr{\D_a}}{K_j}}, by inv.\\
-    (8) $\G', z{:}_{\ov{y_j{:}_1\s_j}}\s, \ov{y_i{:}_\omega\s_i}; \ov{y_j{:}_1\s_j}, \D_b \vdash e' : \tau$\` by $\D$-bound to linear\\
-    (9) $\G', z{:}_{\ov{y_j{:}_1\s_j}}\s, \ov{y_i{:}_\omega\s_i}; \D_a, \D_b, \ov{y_j{:}_1\s_j} \vdash (e, e') : \sigma \otimes \tau$\` by pair (5,8)\\
-    (10) $\G, z{:}_{\ov{y_j{:}_1\s_j}}\s, \ov{y_i{:}_\omega\s_i}; \D, \ov{y_j{:}_1\s_j} \vdash \Theta \mid e : \s, e':\tau, \Sigma$\`by well-typed state\\
-    (11) $\G, z{:}_{\ov{y_j{:}_1\s_j}}\s, \ov{y_i{:}_\omega\s_i}; \D, \ov{y_j{:}_1\s_j} \vdash \Theta' \mid K~\ov{x_i} : \s, e':\tau, \Sigma$\`by induction hypothesis\\
-    (12) $\G, \Theta'{\uparrow}^\omega, \ov{y_i{:}_\omega\s_i}; \D, \ov{y_j{:}_1\s_j}, \Theta'{\uparrow}^1 \vdash (K~\ov{x_{i,j}}, \Theta'{\downarrow}^1_e, e', \Sigma) : \s \bigotimes(\Theta'\downarrow^1_\sigma) \otimes \tau \bigotimes(\Sigma)$\`by \ref{lem:unwrapenv}\\
-    (13) $Let~\G''=\G,\Theta'{\uparrow}^\omega~and~\D_c\D_d\D_4\D_5=\D,\Theta'{\uparrow}^1$\` where $(\D_4, \D_5)$ types $(\Theta'{\downarrow^1_e}, \Sigma)$\\
-    (13a) $\G''; \D_c \vdash K~\ov{x_{i,j}} : \s$\\
-    (13b) $\G'', z{:}_{\ov{y_j{:}_1\s_j}}\s, \ov{y_i{:}_\omega\s_i}; \D_d, \ov{y_j{:}_1\s_j} \vdash e' : \tau$\\
-    (14) $\ov{\G''; \cdot \vdash x_i : \s_i}$\\
-    (15) $\ov{\G''; \D_j \vdash x_j : \s_j}$\` for $\ov{D_j}=\D_c$, by inv\\
-    (16) $\G'', z{:}_{\D_c}\s; \D_c, \D_d \vdash e'\ov{[x_{i,j}/y_{i,j}]} : \tau$\`by $1$-substs and $\omega$-substs lemmas\\
-    (17) $\G''; \D_c, \D_d \vdash e'\ov{[x_{i,j}/y_{i,j}]}[K~\ov{x_{i,j}}/z] : \tau$\`by $\D$-subst lemma (13a,16)\\
-    (18) $\G; \D \vdash \Theta' \mid e'\ov{[x_{i,j}/y_{i,j}]}[K~\ov{x_{i,j}}/z] : \tau, \Sigma$\`by well-typed state\\
-    (19) $\G; \D \vdash \Theta'' \mid v : \tau, \Sigma$\`by induction hypothesis\\
-    Subsubcase $\vdash$ $Alt_0$\\
-    (7) $\G', z{:}_\cdot\s, \ov{y_i{:}_\omega\s_i}; \D_b \vdash e' : \tau$\` by inv\\
-    (8) $\G', z{:}_\cdot\s, \ov{y_i{:}_\omega\s_i}; \D_a, \D_b \vdash (e, e') : \sigma \otimes \tau$\` by pair (5,7)\\
-    (9) $\G, z{:}_\cdot\s, \ov{y_i{:}_\omega\s_i}; \D \vdash \Theta \mid e : \sigma, e' : \tau, \Sigma$\` by well-typed state\\
-    % (10) $\G, z{:}_\cdot\s, \ov{y_i{:}_\omega\s_i}; \D \vdash \Theta \mid e : \sigma, e' : \tau, \Sigma$\` by reconstructing\\
-    (10) $\G, z{:}_\cdot\s, \ov{y_i{:}_\omega\s_i}; \D \vdash \Theta' \mid K~\ov{x_i} : \sigma, e' : \tau, \Sigma$\` by induction hypothesis\\
-    (11) $Let~\G'' = \G,\Theta'{\uparrow}^\omega~and~\D_d,\D_4,\D_5 = \D, \Theta'{\uparrow}^1$\` where $(\D_4, \D_5)$ types $(\Theta'{\downarrow^1_e}, \Sigma)$\\
-    (11a) $\G''; \cdot \vdash K~\ov{x_i} : \s$\`since $K$ has no linear fields in this subcase\\
-    (11b) $\G'', z{:}_{\cdot}\s, \ov{y_i{:}_\omega\s_i}; \D_d \vdash e' : \tau$\\
-    (12) $\ov{\G''; \cdot \vdash x_{i} : \s_i}$\`by inv (11a)\\
-    (13) $\G'', z{:}_{\cdot}\s; \D_d \vdash e'\ov{[x_i/y_i]} : \tau$\` by $\omega$-subst lemma\\
-    (14) $\G''; \D_d \vdash e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] : \tau$\` by $\D$-subst lemma\\
-    (15) $\G; \D \vdash \Theta' \mid e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] :
-    \tau, \Sigma$\` by well-typed state\\
-    (16) $\G; \D \vdash \Theta'' \mid v : \tau, \Sigma$\` by induction hypothesis\\
-\end{tabbing}
-\item[Case:] $\Gamma; \Delta \vdash (\Theta \mid \ccase{e}{z{:}_{\Delta'}\s~\{\_ \Rightarrow e'}\}) \Downarrow (\Theta'' \mid v) : \tau$
-\begin{tabbing}
-    (1) $\Gamma; \Delta \vdash \Theta \mid \ccase{e}{z{:}_{\Delta_a}\s~\{\_ \Rightarrow e'}\} : \tau, \Sigma$\\
-    (2) $\G, \Theta{\uparrow}^\omega; \D, \Theta{\uparrow}^1 \vdash (\ccase{e}{z{:}_{\Delta_a}\s~\{\_ \Rightarrow e'}\}, \Theta{\downarrow}^1_e, \Sigma) : \tau \otimes \bigotimes(\Theta{\downarrow}^1_\sigma) \otimes \bigotimes(\Sigma)$\`by \ref{lem:unwrapenv}\\
-    (3) $Let~\G' = \G, \Theta\uparrow^\omega; \D_a,\D_b,\D_2,\D_3 = \D,\Theta\uparrow^1$\`s.t. $(\D_2,\D_3)$ types $(\Theta\downarrow^1_e, \Sigma)$\\
-    (4) $\G'; \D_a, \D_b \vdash \ccase{e}{z{:}_{\D_a}\s\{\_ \Rightarrow e'\}} : \tau$\\
-    (5) $\G'; \D_a \vdash e : \sigma$\\
-    Subcase $e$ is in WHNF\\
-    (-) Similar to the same subcase in the case-with-pattern-vars reduction,\\
-        except only $Alt_{\_}$ is relevant which is far simpler\\
-    Subcase $e$ is not in WHNF\\
-    (6) $\G', z{:}_{\irr{\D_a}}\s; \irr{\D_a}, \D_b \vdash e' : \tau$\\
-    (7) $\G'; \D_b, z{:}_1\s \vdash e' : \tau$\` by $\D$-bound to linear\\
-    (8) $\G'; \D_a, \D_b, z{:}_1\s \vdash (e, e') : \sigma \otimes \tau$\` by pair\\
-    (9) $\G; \D, z{:}_1\s \vdash \Theta \mid e : \sigma, e' : \tau,
-    \Sigma$\` by well-typed state\\
-    (10) $\G; \D, z{:}_1\s \vdash \Theta' \mid K~\ov{x_i} : \sigma, e' : \tau, \Sigma$\` by induction hypothesis\\
-    (11) $Let~\G';\D_c,\D_d,\D_4,\D_5 = \G, \Theta'{\uparrow}^\omega; \Delta, \Theta'{\uparrow}^1$\`s.t. $(\D_4,\D_5)$ types $(\Theta'{\downarrow}^1_e, \Sigma)$\\
-    (12) $\G';\D_c \vdash K~\ov{x_i} : \s$\\
-    (13) $\G';\D_d,z{:}_1\s \vdash e' : \tau$\\
-    (14) $\G';\D_c, \D_d \vdash e'[K~\ov{x_i}/z] : \tau$\` by $1$-subst lemma\\
-    (15) $\G;\D \vdash \Theta' \mid e'[K~\ov{x_i}/z] : \tau, \Sigma$\` by well-typed state\\
-    (16) $\G;\D \vdash \Theta'' \mid v : \tau, \Sigma$\` by induction hypothesis\\
-\end{tabbing}
-\end{description}
-\end{proof}
-
-We also derive type preservation for the lazy natural semantics.
-
-  \begin{theorem}[Type Preservation]
-For any well-typed $[\Theta \mid e]$, if $\Theta : e \Downarrow
-\Theta' : e'$ or $\Theta : e \Downarrow^* \Theta' : e'$ then we have
-that $[\Theta \mid e']$ is well-typed.
-\end{theorem}
-\begin{proof}
-By Lemma~\ref{lem:naturaltoinstr} and Theorem~\ref{thm:typepres}.
-\end{proof}
-
-\subsection{Progress\label{sec:proof:progress}}
-
-%\input{language-v4/proofs/ProgressTheorem}
-
-\begin{theorem}[Progress]\label{thm:prog}
-Let $\G ; \D \vdash \Theta \mid e : \tau,\Sigma$. Any partial derivation $\G; \D \vdash (\Theta \mid e)
-\Downarrow?,\Sigma$ can be extended. 
-\end{theorem}
-
-\begin{proof}
- Essentially all cases are standard. The only potential issue is the
- (linear) variable rule.
-It suffices to show that whenever a variable is evaluated it is
-present in the evaluation environment. This situation is impossible
-since the state
-$\G ; \D \vdash \Theta \mid x : \tau , \Sigma$ where $x$ is absent
-from $\Theta$ is not well-typed and so cannot be the head of a partial
-derivation. 
-\end{proof}
-
-\begin{theorem}[Progress]
-For any partial derivation of $\Theta : e \Downarrow?$, for $[\Theta |
-e]$ well-typed, the derivation can be extended.
-\end{theorem}
-\begin{proof}
-Follows from Lemma~\ref{lem:instrtonatural} and Theorem~\ref{thm:prog}.
- \end{proof}
-
-\section{Optimisations preserve linearity}\label{app:optimisations}
-
-\subsection{Inlining}
-\input{language-v4/proofs/optimizations/Inlining}
-
-\subsection{\texorpdfstring{$\beta$}{Beta}-reduction}
-
-\input{language-v4/proofs/optimizations/BetaReduction}
-
-\subsection{Case of known constructor}
-
-\input{language-v4/proofs/optimizations/CaseOfKnownConstructor}
-
-\subsection{Case of Case\label{sec:proof:caseofcase}}
-
-The case of case transformation applies to case expressions whose scrutinee is
-another case expression, and returns the innermost case expression transformed by repeating
-the outermost case expression in each alternative of the innermost case,
-scrutinizing the original alternative body.
-
-Intuitively, since the scrutinee of the outermost case is not in WHNF, no
-resources from it can directly occur in the outermost alternatives. By moving
-the outermost alternatives inwards with a different scrutinee, the alternatives
-remain well-typed because they are typed using either the case binder or the
-pattern bound variables, which, by the \emph{Irrelevance} lemma, makes it
-well-typed for any scrutinee consuming arbitrary resources.
-
-\input{language-v4/proofs/optimizations/CaseOfCase}
-
-\subsection{Commuting let}
-
-\input{language-v4/proofs/optimizations/LetFloating}
-
-\subsection{\texorpdfstring{$\eta$}{Eta}-conversions}
-
-\input{language-v4/proofs/optimizations/EtaConvs}
-
-\subsection{Binder Swap}
-
-The binder swap transformation applies to case expressions whose scrutinee is a
-single variable $x$, and it substitutes occurrences of $x$ in the case
-alternatives for the case binder $z$. If $x$ is a linear resource, $x$ cannot
-occur in the case alternatives (as we conservatively consider variables are not
-in WHNF), so the substitution preserves types vacuously. Otherwise, $x$ can be
-freely substituted by $z$, since $z$ is also an unrestricted resource (it's
-usage environment is empty because $x$ is unrestricted).
-
-\input{language-v4/proofs/optimizations/BinderSwap}
-
-% G ; D |- (Theta | e x) \Downarrow (Theta'' | v) : tau, Sigma
-
-% QUEREMOS PROVAR:
-% Se
-% G ; D |- (Theta | e x) : tau , Sigma
-% entao
-% G ; D |- (Theta'' | v) : tau, Sigma
-
-% por i.h.
-% Se G;D |- (Theta | e) : sigma -> tau, x:tau , Sigma
-% entao
-% G ; D |- (Theta' | \y:sigma.e') : sigma->tau, x:tau,Sigma
-
-% por i.h.
-% Se G ; D |- (Theta' , y:sigma = x | e') : tau, Sigma
-% entao
-% G;D |- (Theta'' | v) : tau , Sigma
-
-% Gamma ; Delta |- [Theta | (e x , terms(Sigma))] : (tau , types(Sigma))
-% Gamma ; Delta |- [Theta | (e , terms(x:tau,Sigma))] : (tau->sigma , types(x:tau,Sigma))
-
-% Daqui Gamma ; Delta |- [Theta' | (\y:_1 sigma.e' , terms(x:tau,Sigma))] : (sigma->tau, types(x:tau,Sigma))
-% tirar Gamma ; Delta |- [Theta',y:sigma = x | (e' , terms(Sigma))] : (sigma->tau, types(Sigma))
-
-% [Theta' , x_1 : tau = y | e] = (\x_1:tau . [Theta' | e] ) y
-% [ Theta' , x_Delta:tau = e' | e] = let x_Delta = e' in [Theta' | e]
-% [ . | (e,e')] = (e,e')
-
-% Gamma ; Delta |- [Theta | (e , terms(Sigma))] : (tau , types(Sigma))
-
- % Gamma ; Delta |- Theta | e  : tau, Sigma
-
-% }}}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% \appendix
+
+% \newpage
+
+% ~\vspace{8cm}
+% \begin{center}
+%   {\huge
+%     Supplemental Materials for \\ Lazy Linearity for a Core Functional
+%     Language 
+%     }
+% \end{center}
+
+% \newpage
+
+% \section{Natural and Instrumented Operational Semantics}\label{app:opsem}
+
+% We define a lazy big-step operational semantics for $\lambda_\Delta^\pi$ in the
+% style of Launchbury~\cite{10.1145/158511.158618} natural semantics. The semantics is
+% equipped with an environment which assigns expressions to variables and is
+% mutated in order to express lazy evaluation. Terms are explicit about
+% their sharing (in the sense of shared reductions in lazy evaluation),
+% and so we perform a translation that makes all sharing explicit
+% through let-binders. This translation is presented in Figure~\ref{fig:explicitsharing}.
+% We write $\llet{\ov{x_i{:}_{\Delta_i}\sigma = e_i}}{e'}$ for the iterated
+% let-binding of variables $\ov{x_i}$.
+% \begin{figure}
+% \[
+%   \begin{array}{lclr}
+%     x^* & \triangleq & x\\
+
+%     (\Lambda p.~e)^* & \triangleq & \Lambda p.~(e)^*\\
+%     (\lambda x:_\pi\sigma .e)^* & \triangleq & \lambda x:_\pi\sigma . (e)^*\\
+%     (e~\pi)^* & \triangleq & (e)^*~\pi\\
+%     (e~x)^* & \triangleq & (e)^*~x\\
+%     (e~e')^* & \triangleq & \llet{x{:}_{\Delta}\sigma =
+%                             (e')^*}{(e)^*~x}     \\ &&\qquad \mbox{with
+%                             $\Gamma;\Delta' \vdash e :
+%                             \sigma\rightarrow_\pi \tau$} \\ &&\qquad \mbox{and
+%                                        $\Gamma;\Delta \vdash e' : \sigma$}\\
+%       K~\ov{e_i} & \triangleq & \llet{\ov{x_i:_{\Delta_i} \sigma_i =
+%                                    (e_i)^*}}{K~\ov{x_i}} \\ &&
+%                                        \qquad\mbox{with
+%                                        $\ov{\Gamma;\Delta_i\vdash
+%                                        e_i :
+%                                        \sigma_i}$}\\                                              
+% (\ccase{e}{z{:}_{\Delta}\sigma~\{\overline{\rho\ROUNDTWO{\Rightarrow} e'}\}})^* &
+%                                                                \triangleq
+%                                 &
+%                                   \ccase{(e)^*}{z{:}_{\Delta}\sigma~\{\overline{\rho\ROUNDTWO{\Rightarrow} (e')^*}\}}\\
+%     (\llet{x{:}_{\Delta}\sigma = e}{e'}      )^* & \triangleq & \llet{x{:}_{\Delta}\sigma  = (e)^*}{(e')^*}    \\
+%     (\lletrec{\overline{x{:}_{\Delta}\sigma = e}}{e'}  )^* & \triangleq &\lletrec{\overline{x{:}_{\Delta}\sigma = (e)^*}}{(e')^*}  
+%       \end{array}
+%     \]
+%     \caption{Explicit Sharing Translation\label{fig:explicitsharing}}
+%   \end{figure}
+
+% The natural semantics (Figure~\ref{fig:natsemapp}) is defined via the relation
+% $\Theta : e \Downarrow \Theta' : v$ where $e$ is an expression, $v$ a
+% value, $\Theta$ and $\Theta'$ are environments with bindings of the
+% form $x:_\omega\sigma = e$ which assigns the expression $e$ to the
+% variable $x$ of the given type.
+
+% \begin{figure}
+% \[
+%   \begin{array}{c}
+%     \infer[]
+%     {\,}
+%     {\Theta : \Lambda p.e \Downarrow \Theta : \Lambda p.e}
+%     \qquad
+%     \infer[]
+%     {\Theta  : e\Downarrow \Theta' : \Lambda p.e' \quad \Theta' :
+%     e'[\pi/p]\Downarrow \Theta'' : v}
+%     {\Theta : e~\pi\Downarrow \Theta'' : v }\\[1.5em]
+%     \infer[]
+%     {\,}
+%     {\Theta : \lambda x:_\pi \sigma . e \Downarrow \Theta : \lambda x:_\pi \sigma . e }
+%     \qquad
+%     \infer[]
+%     {\Theta : e \Downarrow \Theta' : \lambda y:_\pi \sigma . e' \quad
+%     \Theta' : e'[x/y] \Downarrow \Theta'' : v }
+%     {\Theta : e~x \Downarrow \Theta'' : v}\\[1.5em]
+%     \infer[]
+%     {(\Theta , x:_\omega \sigma = e): e \Downarrow \Theta' : v}
+%     {(\Theta , x:_\omega \sigma = e) : x \Downarrow (\Theta',x:_\omega
+%     \sigma = v) : v}
+%     \qquad
+%     \infer[]
+%     {(\Theta,x:_\omega \sigma = e) : e' \Downarrow \Theta' : v}
+%     {\Theta : \llet{x{:}_{\Delta}\sigma = e}{e'}\Downarrow \Theta' : v
+%     }\\[1.5em]
+%     \infer[]
+%     {\,}
+%     {\Theta : K~\ov{x_i} \Downarrow \Theta : K~\ov{x_i}}
+%     \qquad
+%     \infer[]
+%     {\Theta : e \Downarrow \Theta' : K~\ov{x_i} \quad
+%     \Theta' : e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] \Downarrow \Theta'' : v }
+%     {\Theta :
+%     \ccase{e}{z{:}_{\Delta'}\sigma~\{{\dots,K~\ov{y_i} \to e', {\dots}}\}}
+%     \Downarrow \Theta'' : v}\\[1.5em]
+%     \infer[]
+%     {\Theta : e \Downarrow \Theta' : K~\ov{x_i} \quad
+%     \Theta' : e'[K~\ov{x_i}/z] \Downarrow \Theta'' : v }
+%     {\Theta :
+%     \ccase{e}{z{:}_{\Delta'}\sigma~\{\dots , \_\to
+%     e'}\} \Downarrow \Theta'' : v}
+%     \qquad
+%     \infer[]
+%     {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
+%     % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
+%     % \Downarrow \Theta_n : v_n \quad
+%     (\Theta, \overline{x_i:_\omega \sigma_i = e_i}) : e' \Downarrow \Theta'
+%     : v
+%     }
+%     {\Theta : \lletrec{\overline{x_i{:}_{\Delta}\sigma_i = e_i}}{e'}
+%     \Downarrow \Theta' : v}
+%     \end{array}
+%   \]
+%   \caption{Natural Semantics for $\lambda_\Delta^\pi$\label{fig:natsemapp}}
+% \end{figure}
+
+% Our instrumented, linearity-aware semantics (akin to that
+% of~\cite{cite:linearhaskell,10.1145/3158093})
+% is defined in Figure~\ref{fig:instrsemapp}.
+% In this semantics, linear variables are erased from
+% the runtime environment, ensuring that any term violating linearity
+% produces a stuck state. The semantics is defined by
+% the judgment $\Gamma ; \Delta \vdash (\Theta \mid e) \Downarrow (\Theta'
+% \mid v) : \sigma , \Sigma$ where $\Gamma$ and $\Delta$ are typing judgments
+% ($\Gamma$ tracks unrestricted variables, $\Delta$ tracks linear
+% assumptions) for variables occurring in $\Sigma$;
+% $\Theta$ and $\Theta'$ are runtime environments, consisting of bindings of the form
+% $x :_\pi \sigma = e$ or $x :_\Delta \sigma = e$ ; $e$ is the expression being evaluated and $v$
+% its resulting value, both of type $\sigma$; $\Sigma$ is a list of assignments of the form
+% $e : \tau$, which are expressions in which variables in $\Theta$ are
+% used.
+
+% \begin{figure}
+% \[
+%   \begin{array}{c}
+%     \infer[]
+%     {\,}
+%     {\Gamma;\Delta \vdash (\Theta \mid \Lambda p.e) \Downarrow (\Theta \mid
+%     \Lambda p.e) : \forall p . \sigma , \Sigma }
+%     \\[1.5em]
+%     \infer[]
+%     {\Gamma ; \Delta \vdash (\Theta \mid  e) \Downarrow (\Theta' \mid
+%     \Lambda p.e') : \forall p . \sigma , \Sigma \quad \Gamma ;\Delta \vdash
+%       (\Theta' \mid e'[\pi/p]) \Downarrow (\Theta'' \mid v) : \sigma[\pi/p], \Sigma}
+%     {\Gamma ; \Delta \vdash (\Theta \mid e~\pi) \Downarrow (\Theta''
+%       \mid v) : \sigma[\pi/p] , \Sigma  }\\[1.5em]
+%     \infer[]
+%     {\,}
+%     {\Gamma ;\Delta \vdash (\Theta \mid \lambda x:_\pi \sigma . e)
+%     \Downarrow (\Theta \mid \lambda x:_\pi\sigma .e) : \sigma
+%     \rightarrow_\pi \tau,\Sigma }\\[1.5em]
+%     \infer[]
+%     {\Gamma;\Delta \vdash  (\Theta \mid e) \Downarrow (\Theta' \mid
+%     \lambda y:_1 \sigma . e') : \sigma \rightarrow_1 \tau , x:\sigma, \Sigma
+%     \quad
+%     \Gamma ; \Delta \vdash (\Theta' , y:_{1} \sigma = x \mid e') \Downarrow (\Theta''
+%     \mid v) : \tau , \Sigma  }
+%     {\Gamma ; \Delta\vdash (\Theta \mid  e~x ) \Downarrow (\Theta''
+%     \mid v) : \tau , \Sigma }\\[1.5em]
+%         \infer[]
+%     {\Gamma;\Delta \vdash  (\Theta \mid e) \Downarrow (\Theta' \mid
+%     \lambda y:_\omega \sigma . e') : \sigma \rightarrow_\omega \tau , x:\sigma, \Sigma
+%     \quad
+%             \Gamma ; \Delta \vdash (\Theta' \mid e'[x/y]) \Downarrow (\Theta''
+%     \mid v) : \tau , \Sigma  }
+%     {\Gamma ; \Delta\vdash (\Theta \mid  e~x ) \Downarrow (\Theta''
+%     \mid v) : \tau , \Sigma }\\[1.5em]
+%        \infer[]
+%     {\Gamma ; \Delta  \vdash (\Theta, x:_{\Delta'} \sigma = e \mid e)
+%     \Downarrow (\Theta' \mid v) : \sigma , \Sigma \\
+%       \D'' = \D' \setminus (\Theta \setminus \Theta')
+%       }
+%     {\Gamma ; \Delta \vdash (\Theta , x:_{\Delta'} \sigma = e \mid x)
+%       \Downarrow (\Theta' , x:_{\Delta''} \sigma = v \mid v) : \sigma, \Sigma}
+%     \\[1.5em]
+%            \infer[]
+%     {\Gamma ;\Delta  \vdash (\Theta \mid e)
+%     \Downarrow (\Theta' \mid v) : \sigma,\Sigma}
+%     {\Gamma ; \Delta \vdash (\Theta , x:_1 \sigma = e \mid x)
+%     \Downarrow (\Theta' \mid v) : \sigma,\Sigma}
+%     \\[1.5em]
+
+%     \infer[]
+%     {\Gamma ;\Delta \vdash (\Theta,x:_{\Delta'} \sigma = e \mid e')
+%     \Downarrow (\Theta' \mid v) : \tau,\Sigma}
+%     {\Gamma ; \Delta \vdash (\Theta \mid \llet{x{:}_{\Delta'}\sigma =
+%     e}{e'})\Downarrow (\Theta' \mid v) : \tau , \Sigma
+%     }\\[1.5em]
+
+%     \infer[]
+%     {\,}
+%     {\Gamma ; \Delta \vdash (\Theta \mid K~\ov{x_i}) \Downarrow
+%     (\Theta \mid K~\ov{x_i}) : T , \Sigma}
+%     \\[1.5em]
+
+%     \infer[]
+%     { K : \ov{ \sigma_i
+%     \rightarrow_\omega \sigma_j \lolli} T
+%     \\
+%     \Gamma, z{:}_{\ov{y_j}}T, \ov{y_i :_\omega \sigma_i} ;\Delta , \ov{y_j :_1 \sigma_j} \vdash (\Theta \mid e) \Downarrow (\Theta' \mid
+%       K~\ov{x_i}) : T , e' : \tau ,\Sigma \\
+%     \Gamma ;\Delta \vdash (\Theta' \mid e'\ov{[x_i/y_{i,j}]}[K~\ov{x_i}/z]
+%     ) \Downarrow (\Theta'' \mid v) :\tau , \Sigma }
+%     {\Gamma ;\Delta \vdash (\Theta \mid
+%     \ccase{e}{z{:}_{\Delta'}T~\{{\dots,K~\ov{y_{i,j}} \,\ROUNDTWO{\Rightarrow}\, e', {\dots} }\}})
+%     \Downarrow (\Theta'' \mid  v) : \tau , \Sigma}\\[1.5em]
+
+%     \infer[]
+%     {K : \ov{ \sigma_i
+%     \rightarrow_\omega \sigma_j \lolli} T\\
+%     \Gamma ; \Delta, z{:}_1\s \vdash (\Theta \mid e) \Downarrow (\Theta' \mid
+%     K~\ov{x_i}) : T , e', \Sigma \quad
+%     \Gamma ;\Delta \vdash (\Theta' \mid  e'[K~\ov{x_i}/z])
+%     \Downarrow (\Theta'' \mid v) : \tau,\Sigma }
+%     {\Gamma ; \Delta \vdash (\Theta \mid 
+%     \ccase{e}{z{:}_{\Delta'}T~\{\dots , \_\,\ROUNDTWO{\Rightarrow}\,
+%     e'}\}) \Downarrow (\Theta'' \mid v) : \tau}
+%     \\[1.5em]
+
+%      \infer[]
+%      {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
+%      % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
+%      % \Downarrow \Theta_n : v_n \quad
+%      (\Gamma ; \Delta \vdash (\Theta, \overline{x_i:_{\Delta'} \sigma_i =
+%      e_i} \mid e' )\Downarrow (\Theta'
+%      \mid v) : \tau
+%      }
+%      {\Gamma ;\Delta \vdash (\Theta \mid \lletrec{\overline{x_i{:}_{\Delta'}\sigma_i = e_i}}{e'})
+%      \Downarrow (\Theta' \mid v) : \tau}
+
+%    \end{array}
+%  \]
+%  \caption{Instrumented Semantics for $\lambda_\Delta^\pi$\label{fig:instrsemapp}}
+%  \end{figure}
+
+%  Following~\cite{Gunter1993APA,cite:linearhaskell}, we also consider
+%  partial derivations of our semantics in order to be able to state a
+%  progress result. Given our set of rules defining judgments of the
+%  form $a \Downarrow b$ with ordered premises, we define a total
+%  derivation of $a \Downarrow b$ as a tree in the standard way. A
+%  partial derivation $a \Downarrow?$ is either a root labelled with $a
+%  \Downarrow?$ or an application of a rule matching $a$ on the left of
+%  the evaluation arrow, where exactly one of the premises $a'
+%  \Downarrow?$ has a partial  derivation, all the premises to the left
+%  of $a' \Downarrow?$ have a total derivation, and the premises to the
+%  right of $a'\Downarrow?$ are not known yet.
+%  %
+
+%  By construction, in a partial derivation, there is only one premise
+%  $b\Downarrow?$ with no sub-derivation. We call $b$ the head of such a
+%  partial derivation.
+% We write $a \Downarrow^* b$ for the relation which holds when $b$ is
+% the head of some partial derivation with root $a \Downarrow?$.  We
+% write $a \Downarrow^* b$  the partial evaluation relation.
+
+% Below, we write $\Theta {\uparrow^1}$ for the linear bindings in evaluation
+% environment $\Theta$ and $\Theta {\uparrow^\omega}$ for all other bindings.
+% $\Theta\downarrow^1_e$ refers to the expressions bound to linear variables in
+% the runtime environment and $\Theta\downarrow^1_\sigma$ refers to the types of
+% those expressions.
+
+%  \begin{definition}[Environment Expansion]
+%    We rely on an expansion of a runtime environment $\Theta$ for the evaluation of a term $e$ into a (typed) term, written $[\Theta \mid e]$, defined inductively as follows:
+%    \[
+%      \begin{array}{lcl}
+%        \left[ \cdot \mid e \right] & \triangleq & e\\
+%        \left[ \Theta , \xl = e' \mid e \right] & \triangleq  & [\Theta \mid (\lambda \xl . e)~e']\\
+%        \left[\Theta , x:_\Delta \sigma = e' \mid e \right] & \triangleq & [\Theta \mid \llet{x:_\Delta \sigma = e'}{e}]
+%     \end{array}
+%    \]
+%  \end{definition}
+
+% We note that for the treatment of recursive let groups, environment
+% expansion needs to happen for the entire group of let-bound
+% variables. This is easily achieved by extending the runtime
+% environment with information about recursive let groupings, which we
+% omit for the sake of presentation.
+
+%  \begin{definition}[Well-typed State]
+%    We write $\G;\D \vdash \Theta \mid e : \tau,\Sigma$ to denote the following typing judgment:
+
+%    \[
+% \G ; \D \vdash [ \Theta \mid (e,\mathit{terms}(\Sigma))] : \tau \otimes \bigotimes(\Sigma)
+%    \]
+%    where $\mathit{terms}(e_1 : \sigma_1 , \dots , e_n : \sigma_n)$ stands for the tuple $(e_1 ,( \dots , (e_n , ())))$ and
+%    $\bigotimes(e_1 : \sigma_1 , \dots , e_n : \sigma_n)$ for $\sigma_1 \otimes ( \dots (\sigma_n \otimes ()))$.
+%  \end{definition}
+
+%  \begin{lemma}~\label{lem:unwrapenv}
+%    $\G ; \D \vdash [ \Theta \mid e] : \sigma$
+%      iff $\G , \Theta\uparrow^{\omega} ; \D , \Theta\uparrow^1 \vdash  (e, \Theta\downarrow^1_e) : \sigma \otimes \bigotimes(\Theta\downarrow^1_\sigma)$
+%    \end{lemma}
+% \begin{proof}
+% By structural induction on the environment expansion definition in the
+% forward direction and by a straightforward induction on $\Theta$ in the
+% converse direction. We show only the forward direction.
+% \begin{description}
+% %\item[Case:] $\left[\cdot \mid e\right] \triangleq e$ Trivial
+% \item[Case:] $\left[\Theta, \xl = y \mid e\right] \triangleq \left[\Theta \mid (\lambda \xl. e)~y\right]$
+% \begin{tabbing}
+%     (1) $\G; \D \vdash \left[ \Theta, \xl = y \mid e \right] : \phi$\\
+%     (2) $\G; \D \vdash \left[ \Theta \mid (\lambda \xl. e)~y \right] : \phi$\` by def.\\
+%     (3) $\G, \Theta\uparrow^\omega; \D, \Theta\uparrow^1 \vdash ((\lambda \xl. e)~y, \Theta\downarrow^1_e) : \phi \otimes \bigotimes(\Theta\downarrow^1_\sigma)$\` by i.h.\\
+%     (4) $Let~\D_1,\D_2,\D_3 = \D, \Theta\uparrow^1 $\\
+%     (5) $\G, \Theta\uparrow^\omega; \D_3 \vdash \Theta\downarrow^1_e : \bigotimes(\Theta\downarrow^1_\sigma)$\\
+%     (6) $\G, \Theta\uparrow^\omega; \D_2 \vdash y : \s$\` by inv\\
+%     (7) $\G, \Theta\uparrow^\omega; \D_1 \vdash (\lambda \xl. e) : \s \to \phi $\` by inv\\
+%     (8) $\G, \Theta\uparrow^\omega; \D_1, \xl \vdash e : \phi $\` by inv\\
+%     (9) $\G, \Theta\uparrow^\omega; \D_1, \D_2, \xl \vdash (e, y) : \phi \otimes \sigma$\` by pair\\
+%     (10) $\G, \Theta\uparrow^\omega; \D, \xl \vdash (e, y, \Theta\downarrow^1_e) : \phi \otimes \sigma \otimes \bigotimes(\Theta\downarrow^1_\sigma)$\` by pair\\
+% \end{tabbing}
+% \item[Case:] $\left[\Theta, \xD = e' \mid e\right] \triangleq \left[\Theta \mid \llet{\xD = e'}{e}\right]$
+% \begin{tabbing}
+%     (1) $\G; \D' \vdash \left[ \Theta, \xD = y \mid e \right] : \phi$\\
+%     (2) $\G; \D' \vdash \left[ \Theta \mid \llet{\xD = e'}{e} \right] : \phi$\` by def.\\
+%     (3) $\G, \Theta\uparrow^\omega; \D', \Theta\uparrow^1 \vdash (\llet{\xD = e'}{e}, \Theta\downarrow^1_e) : \phi \otimes \bigotimes(\Theta\downarrow^1_\sigma)$\` by i.h.\\
+%     (4) $Let~\D,\D_2,\D_3 = \D', \Theta\uparrow^1 $\\
+%     (5) $\G, \Theta\uparrow^\omega; \D_3 \vdash \Theta\downarrow^1_e : \bigotimes(\Theta\downarrow^1_\sigma)$\\
+%     (6) $\G, \Theta\uparrow^\omega; \D \vdash e' : \s$\` by inv on let\\
+%     (7) $\G, \Theta\uparrow^\omega, \xD; \D,\D_2 \vdash e : \phi $\` by inv on let\\
+%     (8) $\G, \Theta\uparrow^\omega, \xD; \D,\D_2,\D_3 \vdash (e, \Theta\downarrow^1_e) : \phi \otimes \bigotimes(\Theta\downarrow^1_\sigma)$\` by pair\\
+% \end{tabbing}
+% \end{description}
+% \end{proof}
+
+% \subsection{Bisimulation}
+
+% We relate our two operational semantics through a bisimulation,
+% allowing us to transfer type safety of the instrumented semantics to
+% the lazy big-step semantics. The relation essentially allows the
+% execution environment to treat some unrestricted bindings
+% of the natural semantics as linear.
+
+% \begin{definition}[Bisimulation]
+% We define the relation $\gamma(\Theta : e)(\G ; \D \vdash \Theta' \mid
+% e' : \tau,\Sigma)$ between a well-typed state and an ordinary state of
+% the lazy big-step semantics if $e = e'$ and $\Theta'{\uparrow^\omega}
+% \leq \Theta$.
+% \end{definition}
+
+% \begin{lemma}~\label{lem:naturaltoinstr}
+%   \begin{itemize}
+%     \item For all $\gamma(\Theta : e)(\G ; \D \vdash \Theta' \mid
+% e : \tau,\Sigma)$ such that $\Theta : e \Downarrow \Theta'' : v$
+% there exists a well-typed state $\G ; \D \vdash (\Theta''' \mid v) :
+% \tau,\Sigma$ such that $\G ; \D \vdash (\Theta' \mid e) \Downarrow
+% (\Theta''' \mid v) : \tau,\Sigma$ and $\gamma(\Theta'' : v)(\G ; \D
+% \vdash \Theta''' \mid v : \tau,\Sigma)$
+% \item For all $\gamma(\Theta : e)(\G ; \D \vdash \Theta' \mid
+% e : \tau,\Sigma)$ such that $\Theta : e \Downarrow^* \Theta'' : v$
+% there exists a well-typed state $\G ; \D \vdash (\Theta''' \mid v) :
+% \tau,\Sigma$ such that $\G ; \D \vdash (\Theta' \mid e) \Downarrow^*
+% (\Theta''' \mid v) : \tau,\Sigma$ and $\gamma(\Theta'' : v)(\G ; \D
+% \vdash \Theta''' \mid v : \tau,\Sigma)$
+% \end{itemize}
+% \end{lemma}
+% \begin{proof}
+% By induction on the given operational semantics rules.
+%   \end{proof}
+
+% \begin{lemma}~\label{lem:instrtonatural}
+%   \begin{itemize}
+% \item  For all $\gamma(\Theta : e)(\G ; \D \vdash \Theta' \mid
+% e : \tau,\Sigma)$ such that $\G ; \D \vdash (\Theta' \mid e) \Downarrow
+% (\Theta'' \mid v) : \tau,\Sigma$, there exists a state $\Theta''' : v$
+% such that $\Theta : e \Downarrow \Theta''' : v$ and
+% $\gamma(\Theta''' : v)(\G ; \D \vdash (\Theta'' \mid v) :
+% \tau,\Sigma)$
+% \item For all $\gamma(\Theta : e)(\G ; \D \vdash \Theta' \mid
+% e : \tau,\Sigma)$ such that $\G ; \D \vdash (\Theta' \mid e) \Downarrow^*
+% (\Theta'' \mid v) : \tau,\Sigma$, there exists a state $\Theta''' : v$
+% such that $\Theta : e \Downarrow^* \Theta''' : v$ and
+% $\gamma(\Theta''' : v)(\G ; \D \vdash (\Theta'' \mid v) :
+% \tau,\Sigma)$
+%    \end{itemize}
+%  \end{lemma}
+% \begin{proof}
+% By induction on the given operational semantics rules.
+%   \end{proof}
+
+% \section{Type Safety Proofs}\label{app:proofs}
+
+% In the following we write $\Mapsto$ to stand for
+% $\Rightarrow_\mathsf{WHNF}$ and $\Rrightarrow$ for $\Rightarrow_\mathsf{NWHNF}$.
+
+% \subsection{Auxiliary Lemmas}
+% The following lemmas presuppose the variable names across the typing
+% contexts are distinct.
+
+%   \begin{lemma}[Linear to $\Delta$-bound]~\label{lem:onedelta-app}
+
+%     \begin{enumerate}
+
+%       \item If $\G; \D',\xl \vdash e :\vp$
+%         then $\G[\D/x],\xD; \D,\D' \vdash e : \vp$ (with $\Delta$ fresh).
+%       \item If $\G ; \D' , \xl \vdash_{alt} \rho \rightarrow e
+%         :^z_{\Delta_s} \sigma \Rightarrow \rho$ then
+%         $\G[\D/x],\xD;\D,\D' \vdash_{alt} \rho \rightarrow e
+%         :^z_{\Delta_s} \sigma \Rightarrow \rho$ (with $\Delta$ fresh).
+
+%       \end{enumerate}
+% \end{lemma}
+% \begin{proof}
+% Straightforward induction on the structure of the given derivation,
+% reconstructing the original derivation and applying rule
+% $\mathit{Var}_\Delta$ where rule $\mathit{Var}_1$ was applied on
+% $x$ in the original derivation.
+% \end{proof}
+% \onedelta*
+
+% \begin{proof}
+% By Lemma~\ref{lem:onedelta-app}(1).
+% \end{proof}
+
+% \begin{lemma}[$\Delta$-bound to Linear]~\label{lem:deltaone-app}
+
+%   \begin{enumerate}
+%  \item If $\G,\x[\irr{\D}]; \irr{\D},\D' \vdash e : \vp$
+%    then $\G[x/\irr{\D}]; \D',\xl \vdash e :\vp$.
+%  \item If $\G ,\x[\irr{\D}]; \irr{\D}, \D' \vdash_{alt} \rho \rightarrow e
+%    :^z_{\Delta_s} \sigma \Rightarrow \rho$ then
+%    $\G[x/\irr{\D}];\D',\xl \vdash_{alt} \rho \rightarrow e
+%    :^z_{\Delta_s} \sigma \Rightarrow \rho$.
+%   \end{enumerate}
+% \end{lemma}
+% \begin{proof}
+% Straightforward induction on the structure of the given derivation,
+% reconstructing the original derivation and applying rule
+% $\mathit{Var}_1$ where rule $\mathit{Var}_\D$ was applied on
+% $x$ in the original derivation.
+% \end{proof}
+
+% \deltaone*
+% \begin{proof}
+% By Lemma~\ref{lem:deltaone-app}(1).
+% \end{proof}
+
+% \begin{lemma}[Unrestricted and $\Delta$-bound]~\label{lem:undelta-app}
+%   \begin{enumerate}
+%     \item $\G,\xo; \D \vdash e : \vp$ iff $\G,\x[\cdot]; \D \vdash e :
+%       \vp$
+%     \item $\G,\xo;\D \vdash_{alt} \rho \rightarrow e
+%       :^z_{\Delta_s} \sigma \Rightarrow \rho$ iff
+%       $\G,\x[\cdot];\D \vdash_{alt} \rho \rightarrow e
+%    :^z_{\Delta_s} \sigma \Rightarrow \rho$
+%   \end{enumerate}
+% \end{lemma}
+% \begin{proof}
+% Straightforward induction on the structure of the given derivation,
+% noting that rules $\mathit{Var}_\Delta$ and $\mathit{Var}_\omega$
+% coincide when $\Delta$ is empty.
+% \end{proof}
+
+% \undelta*
+
+% \begin{proof}
+% By Lemma~\ref{lem:undelta-app}.
+% \end{proof}
+
+% \subsection{Irrelevance\label{sec:proof:irrelevance}}
+
+% \input{language-v4/proofs/WHNFConvSoundness}
+
+% \subsection{Substitution Lemmas\label{sec:proof:substitution-lemmas}}
+
+% \input{language-v4/proofs/LinearSubstitutionLemma}
+
+% \input{language-v4/proofs/UnrestrictedSubstitutionLemma}
+
+% \input{language-v4/proofs/DeltaSubstitutionLemma}
+
+% \subsection{Type Preservation\label{sec:proof:type-preservation}}
+
+% % \input{language-v4/proofs/TypePreservationTheorem}
+% \begin{theorem}[Type Preservation]\label{thm:typepres}
+%   Let $\G ; \D \vdash (\Theta \mid e) \Downarrow (\Theta' \mid v) :
+%   \tau , \Sigma$ or $\G ; \D \vdash (\Theta \mid e) \Downarrow^* (\Theta' \mid v) :
+%   \tau , \Sigma$.
+%   If $\G ; \D \vdash \Theta \mid e : \tau , \Sigma$ then $\G ; \D \vdash \Theta' \mid v : \tau , \Sigma$
+
+% \end{theorem}
+
+% \begin{proof}
+% By structural induction on the instrumented operational semantics. We
+% omit many trivial cases.
+% \begin{description}
+%     \item[Case:] $\Gamma; \Delta \vdash (\Theta, \xl = y \mid x) \Downarrow (\Theta' \mid v) : \s, \Sigma$
+% \begin{tabbing}
+%     (1) $\G; \D \vdash \Theta, \xl = y \mid x : \s, \Sigma$\\
+%     (2) $\G, \Theta{\uparrow}^\omega; \D, \Theta{\uparrow}^1, \xl \vdash (x, y, \Theta{\downarrow}^1_e, \Sigma) : \s \otimes \s \bigotimes(\Theta{\downarrow^1_e}) \bigotimes(\Sigma)$\`by \ref{lem:unwrapenv}\\
+%     (3) $Let~\G';\D_1,\D_2,\D_3 = \G, \Theta{\uparrow}^\omega; \D, \Theta{\uparrow}^1$\` s.t. $(\D_2,\D_3)$ types $(\Theta{\downarrow}^1_e,\Sigma)$\\
+%     (4) $\G'; \xl \vdash x : \s$\\
+%     (5) $\G'; \D_1 \vdash y : \s$\\
+%     (6) $\G'; \D_1, \D_2, \D_3 \vdash (y, \Theta{\downarrow}^1_e, \Sigma) : \s \bigotimes(\Theta{\downarrow^1_e}) \bigotimes(\Sigma)$\\
+%     (7) $\G; \D \vdash \Theta \mid y : \s, \Sigma$\`by well-typed state\\
+%     (8) $\G; \D \vdash \Theta' \mid v : \s, \Sigma$\` by induction hypothesis\\
+% \end{tabbing}
+%     \item[Case:] $\Gamma; \Delta \vdash (\Theta, x{:}_{\D'}\s = e \mid x) \Downarrow (\Theta', x{:}_{\D_a}\s = v \mid v) : \s, \Sigma$
+% \begin{tabbing}
+%     (1) $\Gamma; \Delta \vdash (\Theta, x{:}_{\D'}\s = e \mid x) : \s, \Sigma$\\
+%     (2) $\G, \Theta{\uparrow}^\omega; \D, \Theta{\uparrow}^1 \vdash \llet{x{:}_{\D'} = e}{(x, \Theta{\downarrow}^1_e, \Sigma)} : \s \bigotimes(\Theta{\downarrow}^1_\s) \bigotimes(\Sigma)$\`by \ref{lem:unwrapenv}\\
+%     (3) $Let~\D',\D'' = \D,\Theta{\uparrow}^1$\\
+%     (4) $\G, \Theta{\uparrow}^\omega; \D' \vdash e : \s$\`by inv on let (2)\\
+%     (5) $\G, \Theta{\uparrow}^\omega; \D', \D'' \vdash (x, \Theta{\downarrow}^1_e, \Sigma) : \s \bigotimes(\Theta{\downarrow}^1_\s) \bigotimes(\Sigma)$\`by inv on let (2)\\
+%     (6) $\G, \Theta{\uparrow}^\omega; \D' \vdash x : \s$\\
+%     (7) $\G, \Theta{\uparrow}^\omega; \D'' \vdash (\Theta{\downarrow}^1_e, \Sigma) : \bigotimes(\Theta{\downarrow}^1_\s) \bigotimes(\Sigma)$\\
+%     (8) $\G, \Theta{\uparrow}^\omega; \D', \D'' \vdash (e, \Theta{\downarrow}^1_e, \Sigma) : \tau \bigotimes(\Theta{\downarrow}^1_\s) \bigotimes(\Sigma)$\` by pair (4,7)\\
+%     (9) $\G; \D \vdash \Theta \mid e : \tau,\Sigma$\` by well-typed state\\
+%     (10) $\G; \D \vdash \Theta \mid v : \tau,\Sigma$\` by induction hypothesis\\
+%     (11) $Let~\G';\D_a,\D_b = \G, \Theta'{\uparrow}^\omega, \D, \Theta'{\uparrow}^1$\\
+%     (11a) $\G'; \D_a \vdash v : \s$
+%     (11b) $\G'; \D_b \vdash (\Theta'{\downarrow}^1_e, \Sigma) : \bigotimes(\Theta'{\downarrow}^1_\s) \bigotimes(\Sigma)$\\
+%     (12) $\G'; \D_a \vdash \llet{x{:}_{\D_a}\s = v}{v} : \s$\`by let (11a,11a)\\
+%     (13) $\G; \D \vdash \Theta' \mid \llet{x{:}_{\D_a}\s = v}{v} : \s, \Sigma$\`by well-typed state\\
+%     (14) $\G; \D \vdash \Theta', x{:}_{\D_a}\s = v \mid v : \s, \Sigma$\`by env. expansion\\
+% \end{tabbing}
+% Note: It's slightly more elaborate to accommodate recursion. The runtime
+% environment bindings must be tagged with the recursive group such that the
+% environment expansion can decide whether to unwrap to a letrec group or
+% not. Unlike the case proved here, when the $\D$-binding belongs to a let group,
+% the whole group is unwrapped and we can invert the assumption to get an
+% $e$ typed with the $x$s from the group.\\
+% \item[Case:] $\Gamma; \Delta \vdash (\Theta \mid \ccase{e}{z{:}_{\Delta'}\s~\{K~\ov{y_i}\Rightarrow e'}\}) \Downarrow (\Theta'' \mid v) : \tau$
+% \begin{tabbing}
+%     (1) $\Gamma; \Delta \vdash \Theta \mid \ccase{e}{z{:}_{\Delta_a}\s~\{K~\ov{y_i}\Rightarrow e'}\} : \tau, \Sigma$\\
+%     (2) $\G, \Theta{\uparrow}^\omega; \D, \Theta{\uparrow}^1 \vdash (\ccase{e}{z{:}_{\Delta_a}\s~\{K~\ov{y_i}\Rightarrow e'}\}, \Theta{\downarrow}^1_e, \Sigma) : \tau \otimes \bigotimes(\Theta{\downarrow}^1_\sigma) \otimes \bigotimes(\Sigma)$\`by \ref{lem:unwrapenv}\\
+%     (3) $Let~\G' = \G, \Theta\uparrow^\omega; \D_a,\D_b,\D_2,\D_3 = \D,\Theta\uparrow^1$\`s.t. $(\D_2,\D_3)$ types $(\Theta\downarrow^1_e, \Sigma)$\\
+%     (4) $\G'; \D_a, \D_b \vdash \ccase{e}{z{:}_{\D_a}\s\{K~\ov{y_i}\Rightarrow e'\}} : \tau$\\
+%     Subcase $e$ is in WHNF (thus $= K~\ov{x_i}$)\\
+%     (5) $\G'; \D_a \Vdash K~\ov{x_i} : \s \gtrdot \ov{\D_i}$\` by subcase\\
+%     (6) $\G', z{:}_{\D_a}\s; \D_a, \D_b \vdash_{alt} K~\ov{y_i} \to e' :^z_{\D_i} \s \Rightarrow \tau$\` by inv\\
+%     Subsubcase $\vdash$ $Alt_{WHNF}$\\
+%     (7) $\G', z{:}_{\D_a}\s, \ov{y_i{:}_\omega\s_i}, \ov{y_j{:}_{\Delta_j}\s_j}; \D_a, \D_b \vdash e' : \tau$\` by inv\\
+%     (8) $\Theta' = \Theta$\` in this subcase because $(\Theta \mid K~\ov{x_i})\Downarrow(\Theta \mid K~\ov{x_i})$\\
+%     (9) $\G'; \ov{\D_i} \vdash K~\ov{x_i} : \s$\` by inv (5)\\
+%     (10) $\ov{\G'; \cdot \vdash x_i : \s_i}$\` by inv (9) for unrestricted $x$s\\
+%     (11) $\ov{\G'; \D_j \vdash x_j : \s_j}$\` by inv (9) for linear fields $x$s\\
+%     (12) $\G', z{:}_{\D_a}\s; \D_a, \D_b \vdash e'\ov{[x_{i,j}/y_{i,j}] : \tau}$\` by $\D$-subst and $\omega$-subst lemmas (7,10,11)\\
+%     (13) $\G'; \D_a, \D_b \vdash e'\ov{[x_{i,j}/y_{i,j}]}[K~\ov{x_{i,j}}/z] : \tau$\` by $\D$-subst lemma and $\ov{\D_i} = \D_a$ (9,12)\\
+%     (14) $\G; \D \vdash \Theta \mid e'\ov{[x_{i,j}/y_{i,j}]}[K~\ov{x_{i,j}}/z] : \tau, \Sigma$\` by well-typed state\\
+%     (15) $\G; \D \vdash \Theta'' \mid v : \tau, \Sigma$\` by induction hypothesis\\
+%     Subsubcase $\vdash$ $Alt_{0}$\\
+%     (7) $(\G', z{:}_{\D_a}\s)[\cdot/\D_a]_z, \ov{y_i{:}_\omega\sigma_i}; (\D_a, \D_b)[\cdot / \D_a] \vdash e' : \tau$\` by inv subcase\\
+%     (8) $\G', z{:}_{\cdot}, \ov{y_i{:}_\omega\sigma_i}; \D_b \vdash e' : \tau$\` by $z$ cannot occur in $\G'$\\
+%     (9) Since $\G';\D_a \Vdash K~\ov{x_i}$, in this subcase $\D_a = \cdot$\\
+%     (10) $\G', \z{:}_\cdot\sigma, \ov{y{:}_\omega\sigma}; \D_b \vdash e' : \tau$\` by (9)\\
+%     (11) $\G', \z{:}_\cdot\sigma; \D_b \vdash e'\ov{[x_i/y_i]} : \tau$\` by $\omega$-subst (10)\\
+%     (12) $\G'; \D_b \vdash e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] : \tau$\` by $\D$-subst\\
+%     (13) $\G; \D \vdash \Theta \mid e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] : \tau, \Sigma$\` by well-typed state\\
+%     (14) $\G; \D \vdash \Theta'' \mid v : \tau, \Sigma$\`by induction hypothesis\\
+%     Subcase $e$ is not in WHNF\\
+%     (5) $\G'; \D_a \vdash e : \s$\` by inv on case\\
+%     (6) $\G', z{:}_{\irr{\D_a}}\s; \irr{\D_a}, \D_b \vdash_{alt} K~\ov{y_{i,j}} \to e' :^z_{\irr{\D_a}} \s $\` by inv on case\\
+%     Subsubcase $\vdash$ $Alt_{NWHNF}$\\
+%     (7) $\G', z{:}_{\irr{\D_a}}\s, \ov{y_i{:}_\omega\s_i}, \ov{y_j{:}_{\Delta_j}\s_j}; \irr{\D_a}, \D_b \vdash e' : \tau$\` for \ov{\D_j = \lctag{\irr{\D_a}}{K_j}}, by inv.\\
+%     (8) $\G', z{:}_{\ov{y_j{:}_1\s_j}}\s, \ov{y_i{:}_\omega\s_i}; \ov{y_j{:}_1\s_j}, \D_b \vdash e' : \tau$\` by $\D$-bound to linear\\
+%     (9) $\G', z{:}_{\ov{y_j{:}_1\s_j}}\s, \ov{y_i{:}_\omega\s_i}; \D_a, \D_b, \ov{y_j{:}_1\s_j} \vdash (e, e') : \sigma \otimes \tau$\` by pair (5,8)\\
+%     (10) $\G, z{:}_{\ov{y_j{:}_1\s_j}}\s, \ov{y_i{:}_\omega\s_i}; \D, \ov{y_j{:}_1\s_j} \vdash \Theta \mid e : \s, e':\tau, \Sigma$\`by well-typed state\\
+%     (11) $\G, z{:}_{\ov{y_j{:}_1\s_j}}\s, \ov{y_i{:}_\omega\s_i}; \D, \ov{y_j{:}_1\s_j} \vdash \Theta' \mid K~\ov{x_i} : \s, e':\tau, \Sigma$\`by induction hypothesis\\
+%     (12) $\G, \Theta'{\uparrow}^\omega, \ov{y_i{:}_\omega\s_i}; \D, \ov{y_j{:}_1\s_j}, \Theta'{\uparrow}^1 \vdash (K~\ov{x_{i,j}}, \Theta'{\downarrow}^1_e, e', \Sigma) : \s \bigotimes(\Theta'\downarrow^1_\sigma) \otimes \tau \bigotimes(\Sigma)$\`by \ref{lem:unwrapenv}\\
+%     (13) $Let~\G''=\G,\Theta'{\uparrow}^\omega~and~\D_c\D_d\D_4\D_5=\D,\Theta'{\uparrow}^1$\` where $(\D_4, \D_5)$ types $(\Theta'{\downarrow^1_e}, \Sigma)$\\
+%     (13a) $\G''; \D_c \vdash K~\ov{x_{i,j}} : \s$\\
+%     (13b) $\G'', z{:}_{\ov{y_j{:}_1\s_j}}\s, \ov{y_i{:}_\omega\s_i}; \D_d, \ov{y_j{:}_1\s_j} \vdash e' : \tau$\\
+%     (14) $\ov{\G''; \cdot \vdash x_i : \s_i}$\\
+%     (15) $\ov{\G''; \D_j \vdash x_j : \s_j}$\` for $\ov{D_j}=\D_c$, by inv\\
+%     (16) $\G'', z{:}_{\D_c}\s; \D_c, \D_d \vdash e'\ov{[x_{i,j}/y_{i,j}]} : \tau$\`by $1$-substs and $\omega$-substs lemmas\\
+%     (17) $\G''; \D_c, \D_d \vdash e'\ov{[x_{i,j}/y_{i,j}]}[K~\ov{x_{i,j}}/z] : \tau$\`by $\D$-subst lemma (13a,16)\\
+%     (18) $\G; \D \vdash \Theta' \mid e'\ov{[x_{i,j}/y_{i,j}]}[K~\ov{x_{i,j}}/z] : \tau, \Sigma$\`by well-typed state\\
+%     (19) $\G; \D \vdash \Theta'' \mid v : \tau, \Sigma$\`by induction hypothesis\\
+%     Subsubcase $\vdash$ $Alt_0$\\
+%     (7) $\G', z{:}_\cdot\s, \ov{y_i{:}_\omega\s_i}; \D_b \vdash e' : \tau$\` by inv\\
+%     (8) $\G', z{:}_\cdot\s, \ov{y_i{:}_\omega\s_i}; \D_a, \D_b \vdash (e, e') : \sigma \otimes \tau$\` by pair (5,7)\\
+%     (9) $\G, z{:}_\cdot\s, \ov{y_i{:}_\omega\s_i}; \D \vdash \Theta \mid e : \sigma, e' : \tau, \Sigma$\` by well-typed state\\
+%     % (10) $\G, z{:}_\cdot\s, \ov{y_i{:}_\omega\s_i}; \D \vdash \Theta \mid e : \sigma, e' : \tau, \Sigma$\` by reconstructing\\
+%     (10) $\G, z{:}_\cdot\s, \ov{y_i{:}_\omega\s_i}; \D \vdash \Theta' \mid K~\ov{x_i} : \sigma, e' : \tau, \Sigma$\` by induction hypothesis\\
+%     (11) $Let~\G'' = \G,\Theta'{\uparrow}^\omega~and~\D_d,\D_4,\D_5 = \D, \Theta'{\uparrow}^1$\` where $(\D_4, \D_5)$ types $(\Theta'{\downarrow^1_e}, \Sigma)$\\
+%     (11a) $\G''; \cdot \vdash K~\ov{x_i} : \s$\`since $K$ has no linear fields in this subcase\\
+%     (11b) $\G'', z{:}_{\cdot}\s, \ov{y_i{:}_\omega\s_i}; \D_d \vdash e' : \tau$\\
+%     (12) $\ov{\G''; \cdot \vdash x_{i} : \s_i}$\`by inv (11a)\\
+%     (13) $\G'', z{:}_{\cdot}\s; \D_d \vdash e'\ov{[x_i/y_i]} : \tau$\` by $\omega$-subst lemma\\
+%     (14) $\G''; \D_d \vdash e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] : \tau$\` by $\D$-subst lemma\\
+%     (15) $\G; \D \vdash \Theta' \mid e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] :
+%     \tau, \Sigma$\` by well-typed state\\
+%     (16) $\G; \D \vdash \Theta'' \mid v : \tau, \Sigma$\` by induction hypothesis\\
+% \end{tabbing}
+% \item[Case:] $\Gamma; \Delta \vdash (\Theta \mid \ccase{e}{z{:}_{\Delta'}\s~\{\_ \Rightarrow e'}\}) \Downarrow (\Theta'' \mid v) : \tau$
+% \begin{tabbing}
+%     (1) $\Gamma; \Delta \vdash \Theta \mid \ccase{e}{z{:}_{\Delta_a}\s~\{\_ \Rightarrow e'}\} : \tau, \Sigma$\\
+%     (2) $\G, \Theta{\uparrow}^\omega; \D, \Theta{\uparrow}^1 \vdash (\ccase{e}{z{:}_{\Delta_a}\s~\{\_ \Rightarrow e'}\}, \Theta{\downarrow}^1_e, \Sigma) : \tau \otimes \bigotimes(\Theta{\downarrow}^1_\sigma) \otimes \bigotimes(\Sigma)$\`by \ref{lem:unwrapenv}\\
+%     (3) $Let~\G' = \G, \Theta\uparrow^\omega; \D_a,\D_b,\D_2,\D_3 = \D,\Theta\uparrow^1$\`s.t. $(\D_2,\D_3)$ types $(\Theta\downarrow^1_e, \Sigma)$\\
+%     (4) $\G'; \D_a, \D_b \vdash \ccase{e}{z{:}_{\D_a}\s\{\_ \Rightarrow e'\}} : \tau$\\
+%     (5) $\G'; \D_a \vdash e : \sigma$\\
+%     Subcase $e$ is in WHNF\\
+%     (-) Similar to the same subcase in the case-with-pattern-vars reduction,\\
+%         except only $Alt_{\_}$ is relevant which is far simpler\\
+%     Subcase $e$ is not in WHNF\\
+%     (6) $\G', z{:}_{\irr{\D_a}}\s; \irr{\D_a}, \D_b \vdash e' : \tau$\\
+%     (7) $\G'; \D_b, z{:}_1\s \vdash e' : \tau$\` by $\D$-bound to linear\\
+%     (8) $\G'; \D_a, \D_b, z{:}_1\s \vdash (e, e') : \sigma \otimes \tau$\` by pair\\
+%     (9) $\G; \D, z{:}_1\s \vdash \Theta \mid e : \sigma, e' : \tau,
+%     \Sigma$\` by well-typed state\\
+%     (10) $\G; \D, z{:}_1\s \vdash \Theta' \mid K~\ov{x_i} : \sigma, e' : \tau, \Sigma$\` by induction hypothesis\\
+%     (11) $Let~\G';\D_c,\D_d,\D_4,\D_5 = \G, \Theta'{\uparrow}^\omega; \Delta, \Theta'{\uparrow}^1$\`s.t. $(\D_4,\D_5)$ types $(\Theta'{\downarrow}^1_e, \Sigma)$\\
+%     (12) $\G';\D_c \vdash K~\ov{x_i} : \s$\\
+%     (13) $\G';\D_d,z{:}_1\s \vdash e' : \tau$\\
+%     (14) $\G';\D_c, \D_d \vdash e'[K~\ov{x_i}/z] : \tau$\` by $1$-subst lemma\\
+%     (15) $\G;\D \vdash \Theta' \mid e'[K~\ov{x_i}/z] : \tau, \Sigma$\` by well-typed state\\
+%     (16) $\G;\D \vdash \Theta'' \mid v : \tau, \Sigma$\` by induction hypothesis\\
+% \end{tabbing}
+% \end{description}
+% \end{proof}
+
+% We also derive type preservation for the lazy natural semantics.
+
+%   \begin{theorem}[Type Preservation]
+% For any well-typed $[\Theta \mid e]$, if $\Theta : e \Downarrow
+% \Theta' : e'$ or $\Theta : e \Downarrow^* \Theta' : e'$ then we have
+% that $[\Theta \mid e']$ is well-typed.
+% \end{theorem}
+% \begin{proof}
+% By Lemma~\ref{lem:naturaltoinstr} and Theorem~\ref{thm:typepres}.
+% \end{proof}
+
+% \subsection{Progress\label{sec:proof:progress}}
+
+% %\input{language-v4/proofs/ProgressTheorem}
+
+% \begin{theorem}[Progress]\label{thm:prog}
+% Let $\G ; \D \vdash \Theta \mid e : \tau,\Sigma$. Any partial derivation $\G; \D \vdash (\Theta \mid e)
+% \Downarrow?,\Sigma$ can be extended. 
+% \end{theorem}
+
+% \begin{proof}
+%  Essentially all cases are standard. The only potential issue is the
+%  (linear) variable rule.
+% It suffices to show that whenever a variable is evaluated it is
+% present in the evaluation environment. This situation is impossible
+% since the state
+% $\G ; \D \vdash \Theta \mid x : \tau , \Sigma$ where $x$ is absent
+% from $\Theta$ is not well-typed and so cannot be the head of a partial
+% derivation. 
+% \end{proof}
+
+% \begin{theorem}[Progress]
+% For any partial derivation of $\Theta : e \Downarrow?$, for $[\Theta |
+% e]$ well-typed, the derivation can be extended.
+% \end{theorem}
+% \begin{proof}
+% Follows from Lemma~\ref{lem:instrtonatural} and Theorem~\ref{thm:prog}.
+%  \end{proof}
+
+% \section{Optimisations preserve linearity}\label{app:optimisations}
+
+% \subsection{Inlining}
+% \input{language-v4/proofs/optimizations/Inlining}
+
+% \subsection{\texorpdfstring{$\beta$}{Beta}-reduction}
+
+% \input{language-v4/proofs/optimizations/BetaReduction}
+
+% \subsection{Case of known constructor}
+
+% \input{language-v4/proofs/optimizations/CaseOfKnownConstructor}
+
+% \subsection{Case of Case\label{sec:proof:caseofcase}}
+
+% The case of case transformation applies to case expressions whose scrutinee is
+% another case expression, and returns the innermost case expression transformed by repeating
+% the outermost case expression in each alternative of the innermost case,
+% scrutinizing the original alternative body.
+
+% Intuitively, since the scrutinee of the outermost case is not in WHNF, no
+% resources from it can directly occur in the outermost alternatives. By moving
+% the outermost alternatives inwards with a different scrutinee, the alternatives
+% remain well-typed because they are typed using either the case binder or the
+% pattern bound variables, which, by the \emph{Irrelevance} lemma, makes it
+% well-typed for any scrutinee consuming arbitrary resources.
+
+% \input{language-v4/proofs/optimizations/CaseOfCase}
+
+% \subsection{Commuting let}
+
+% \input{language-v4/proofs/optimizations/LetFloating}
+
+% \subsection{\texorpdfstring{$\eta$}{Eta}-conversions}
+
+% \input{language-v4/proofs/optimizations/EtaConvs}
+
+% \subsection{Binder Swap}
+
+% The binder swap transformation applies to case expressions whose scrutinee is a
+% single variable $x$, and it substitutes occurrences of $x$ in the case
+% alternatives for the case binder $z$. If $x$ is a linear resource, $x$ cannot
+% occur in the case alternatives (as we conservatively consider variables are not
+% in WHNF), so the substitution preserves types vacuously. Otherwise, $x$ can be
+% freely substituted by $z$, since $z$ is also an unrestricted resource (it's
+% usage environment is empty because $x$ is unrestricted).
+
+% \input{language-v4/proofs/optimizations/BinderSwap}
+
+% % G ; D |- (Theta | e x) \Downarrow (Theta'' | v) : tau, Sigma
+
+% % QUEREMOS PROVAR:
+% % Se
+% % G ; D |- (Theta | e x) : tau , Sigma
+% % entao
+% % G ; D |- (Theta'' | v) : tau, Sigma
+
+% % por i.h.
+% % Se G;D |- (Theta | e) : sigma -> tau, x:tau , Sigma
+% % entao
+% % G ; D |- (Theta' | \y:sigma.e') : sigma->tau, x:tau,Sigma
+
+% % por i.h.
+% % Se G ; D |- (Theta' , y:sigma = x | e') : tau, Sigma
+% % entao
+% % G;D |- (Theta'' | v) : tau , Sigma
+
+% % Gamma ; Delta |- [Theta | (e x , terms(Sigma))] : (tau , types(Sigma))
+% % Gamma ; Delta |- [Theta | (e , terms(x:tau,Sigma))] : (tau->sigma , types(x:tau,Sigma))
+
+% % Daqui Gamma ; Delta |- [Theta' | (\y:_1 sigma.e' , terms(x:tau,Sigma))] : (sigma->tau, types(x:tau,Sigma))
+% % tirar Gamma ; Delta |- [Theta',y:sigma = x | (e' , terms(Sigma))] : (sigma->tau, types(Sigma))
+
+% % [Theta' , x_1 : tau = y | e] = (\x_1:tau . [Theta' | e] ) y
+% % [ Theta' , x_Delta:tau = e' | e] = let x_Delta = e' in [Theta' | e]
+% % [ . | (e,e')] = (e,e')
+
+% % Gamma ; Delta |- [Theta | (e , terms(Sigma))] : (tau , types(Sigma))
+
+%  % Gamma ; Delta |- Theta | e  : tau, Sigma
+
+% % }}}
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2025-11-25 popl26.tex
added author package things
.
Put things before and after \begin{document} as requested
Don't use scriptsize, and revert back to itemize

@@
-\documentclass[acmsmall,review,screen]{acmart}
+\documentclass[acmsmall]{acmart}
 \PassOptionsToPackage{dvipsnames}{xcolor}
@@
 \input{language-v4/TheoremsAndLemmas}
-
-%\title{Semantic linearity in a lazy optimizing compiler]
-\title{Lazy Linearity for a Core Functional Language}
-%\subtitle{Type-checking Linearity in GHC Core}
-
-\author{Rodrigo Mesquita}
-\orcid{0009-0005-1467-115X}
-%\affiliation{
-  %\department{Departamento de Inform\'{a}tica}
-  %\institution{FCT-NOVA, Universidade Nova de Lisboa}
-  %\city{City}
-  %\country{Portugal}}
-\affiliation{
-   \institution{Well-Typed LLP}
-   % \city{City}
-   \country{United Kingdom}
- }
- \email{rodrigo@well-typed.com}
-\author{Bernardo Toninho}
-\orcid{0000-0002-0746-7514}
-\affiliation{
-  \department{Departamento de Engenharia Inform\'{a}tica}
-  \institution{Instituto Superior Técnico, Universidade de Lisboa and INESC-ID}
-  %\city{City}
-  \country{Portugal}}
-\email{bernardo.toninho@tecnico.ulisboa.pt}
-% CHECK: https://www.acm.org/binaries/content/assets/publications/taps/latex-best_practices-06-may-2020.pdf
-
-% \date{ }
 
@@
 
+\setcopyright{cc}
+\setcctype{by}
+\acmDOI{10.1145/3776711}
+\acmYear{2026}
+\acmJournal{PACMPL}
+\acmVolume{10}
+\acmNumber{POPL}
+\acmArticle{69}
+\acmMonth{1}
+\received{2025-07-10}
+\received[accepted]{2025-11-06}
+
 \begin{document}
+
+%\title{Semantic linearity in a lazy optimizing compiler]
+\title{Lazy Linearity for a Core Functional Language}
+%\subtitle{Type-checking Linearity in GHC Core}
+
+\author{Rodrigo Mesquita}
+\orcid{0009-0005-1467-115X}
+\affiliation{%
+  \institution{Well-Typed LLP}
+  \city{London}
+  \country{United Kingdom}
+}
+\email{rodrigo@well-typed.com}
+
+\author{Bernardo Toninho}
+\orcid{0000-0002-0746-7514}
+\affiliation{%
+  \institution{Instituto Superior Técnico, University of Lisbon}
+  \city{Lisbon}
+  \country{Portugal}
+}
+\affiliation{%
+  \institution{INESC-ID}
+  \city{Lisbon}
+  \country{Portugal}
+}
+\email{bernardo.toninho@tecnico.ulisboa.pt}
 
@@
 
-  {\scriptsize
+  {\small
   \[
@@
 excellent results. Programs considered ill-typed by the Linear Core
-implementation include:
-%
-%\begin{itemize}
-% \item
-Programs we expected GHC to produce, but know are not seen as linear by $\lambda^\pi_\Delta$.
+implementation include programs that:
+%
+\begin{itemize}
+\item we expected GHC to produce, but know are not seen as linear by $\lambda^\pi_\Delta$.
         Namely, programs resulting from the reverse binder swap which
@@
         programs could be understood as linear as long as all applications of
-        linear functions are reduced \emph{call-by-need};
-programs rejected due to the lack of ``linearity coercions'' in
+        linear functions are reduced \emph{call-by-need}
+\item were rejected due to the lack of ``linearity coercions'' in
         $\lambda^\pi_\Delta$. Extending Linear Core with \emph{multiplicity coercions} 
@@
         (discussed in Section~\ref{sec:future-work});
-programs which actually violate linearity, in that the term
+\item actually violate linearity, in that the term
         violates the linear type it is annotated with. These are the programs
@@
         the type of \lstinline{build} was not general enough to accommodate linearity).
-%\end{itemize}
+\end{itemize}
 

### 2025-11-25 popl26.tex
Improve plugin section with AE reviews

@@
 
-We have implemented the Linear Core type system as a plugin for the Glasgow
-Haskell Compiler.
+We implemented Linear Core as a plugin for the Glasgow Haskell Compiler.
 GHC plugins allow developers to inspect and modify programs being compiled by
@@
 %
-In particular, for any given Haskell module, Core plugins run for (and receive
-as input) every intermediate program produced in the compilation process, i.e.
-from desugaring and after each optimising transformation.
-%
-Our implementation further substantiates our claim that Linear Core is
-suitable for the intermediate language of an optimising compiler.
+% In particular, for any given Haskell module, Core plugins run for (and receive
+% as input) every intermediate program produced in the compilation process, i.e.
+% from desugaring and after each optimising transformation.
+%
+The Linear Core GHC plugin~\cite{cite:linear-core-plugin} is a typechecker. For
+every single intermediate Core program produced by GHC, our plugin checks the
+linearity of all expressions, failing if a linear resource is not used
+\emph{exactly once} according to \ROUNDTWO{our typing discipline}.
+%
+The plugin implementation of Linear Core provides us with strong confidence
+that the Linear Core type system is suitable for the intermediate language of
+an optimizing compiler, since it is successful in accepting the vast majority
+of real-world linear programs produced by the GHC optimizer, and even
+identified some which violate linearity.
+% Our implementation further substantiates our claim that Linear Core is
+% suitable for the intermediate language of an optimising compiler.
 %\todo[inline]{Tem de ser anonimizado, provavelmente alterando a bib
   %entry para omitir autores e o proprio link}
-The GHC Linear Core plugin~\cite{cite:linear-core-plugin} implements a
-typechecker for Linear Core: given a Core program, our plugin typechecks the
-linearity of all expressions bound in that program, failing if a linear
-resource is not used \emph{exactly once} according to \ROUNDTWO{our typing discipline}.
 %
@@
 
-The implementation of Linear Core as a typechecker does not follow
-directly from the description of the type system because Linear Core
-is not \emph{syntax-directed}, \ROUNDTWO{given that the goal of our
-  design is to be as simple to reason with as possible.} Specifically,
-the \ROUNDTWO{features of Linear Core that are not syntax-directed are
-  essentially the}
+The implementation of a Linear Core checker does not follow directly from the
+type system because Linear Core is not syntax-directed \ROUNDTWO{(the design
+aims to be as simple to reason with as possible).}
+Specifically, the \ROUNDTWO{features of Linear Core that are not
+syntax-directed are essentially the}
 splitting of linear resources amongst sub-derivations and
@@
 derivation~\cite{DBLP:journals/tcs/CervesatoHP00}.
-
-For a scrutinee not in WHNF, pattern variables in a case alternative are
-introduced as $\D$-variables with usage environment $\lctag{\irr{\D}}{K_i}$,
-where $\D$ are the scrutinee resources and $K_i$ the tag of that pattern
-variable. To use these pattern variables, the scrutinee resources must be first
-$\textsc{Split}$ into fragments. Instead of guessing which resources need to be split,
-the implementation consumes tagged fragments of a resource \emph{as needed},
-i.e., only when consuming a resource whose usage environment has a resource
-tagged $K_j$ do we $\textsc{Split}$ that resource on $K$ and consume the $K_j$ fragment.
+%
+To use pattern variables with usage environment $\lctag{\irr{\D}}{K_i}$, the
+scrutinee resources must first be $\textsc{Split}$ into fragments. Instead of
+guessing which resources need to be split,
+the implementation consumes tagged fragments of a resource as needed,
+i.e., only when consuming a resource tagged $K_j$ do we $\textsc{Split}$ that
+resource on $K$ and consume the $K_j$ fragment.
 %
@@
 naive $O(n^2)$ algorithm (where $n$ is the number of let bindings) to determine
-these usage environments. %Further research regarding
+these usage environments. We leave the study of the soundness and completeness
+of this implementation strategy to future work.
+%Further research regarding
 %this inference challenge is discussed in Section~\ref{sec:future-work}.
@@
 
-The results of running the Linear Core GHC plugin on large 
-libraries focused around linear types are reported in
-Figure~\ref{fig:core-plugin-res}.
+The results of running the Linear Core GHC plugin on real-world libraries with
+linear types are reported in Figure~\ref{fig:core-plugin-res}.
 %
@@
 We compiled the Haskell standard library for programming with linear types,
-\texttt{linear-base} (4000 lines, comprising over 100 modules), the
-\texttt{linear-smc}~\cite{10.1145/3471874.3472980} (1500 lines) library, and
-\texttt{priority-sesh}~\cite{10.1145/3471874.3472979} (1400 lines), a
-session-types library, using our plugin.
+\texttt{linear-base} (4000 lines, comprising over 100 modules),
+\texttt{linear-smc}~\cite{10.1145/3471874.3472980} (1500 lines),
+\texttt{priority-sesh}~\cite{10.1145/3471874.3472979} (1400 lines),
+\texttt{text-builder-linear}~\cite{cite:text-builder-linear} (1800 lines), and
+\texttt{linear-generics}~\cite{cite:linear-generics} (2000 lines), using our
+plugin.
 %
@@
 each top-level binding in a module counts as a program, and every such
-binding is typechecked once per optimisation pass (i.e., we typecheck
-all intermediate programs produced by GHC). The total amount of programs
-rejected by the typechecker are counted dually, also across every optimisation
-pass, without halting compilation when a program is rejected. We note that our
-implementation is not performance conscious, and types every intermediate
-program from scratch, instead of preserving information throughout the
-pipeline.
+binding is checked once per optimisation pass, i.e. we check
+all intermediate programs produced by GHC. The total of programs
+rejected is counted in the same way, without ever halting compilation when a
+program is rejected. We note that our implementation is not performance
+conscious and types every intermediate program from scratch.
+% , instead of preserving information throughout the pipeline.
 
@@
 
-\vspace{-.1cm}
 {\small
@@
         programs could be understood as linear as long as all applications of
-        linear functions are reduced \emph{call-by-need}
+        linear functions are reduced call-by-need
 \item were rejected due to the lack of ``linearity coercions'' in
-        $\lambda^\pi_\Delta$. Extending Linear Core with \emph{multiplicity coercions} 
-        could make these programs be understood as linear
-        (discussed in Section~\ref{sec:future-work});
-\item actually violate linearity, in that the term
-        violates the linear type it is annotated with. These are the programs
+        $\lambda^\pi_\Delta$. Extending Linear Core with these coercions would
+        make these programs be accepted (as discussed in Section~\ref{sec:future-work});
+\item actually violate linearity. These are the programs
         we ultimately want to spot and reject by having a linearly typed Core
@@
 \end{itemize}
-
-Overall, the programs rejected by the Linear Core plugin,
+%
 %, in \texttt{linear-base},
-besides validating that our implementation is faithful to the
-$\lambda^\pi_\Delta$ system, provide further insight into the remaining
-details required to fully typecheck linearity in a mature optimising compiler.
+Besides validating that our implementation is faithful to the
+$\lambda^\pi_\Delta$ system, these programs provide insight as to what more is
+needed to fully understand linearity in a mature optimising compiler.
 

### 2025-11-25 prototype/core-plugin-results-2.tex
started instrumented semantics
Improve plugin section with AE reviews

@@
 \centering{
-\begin{tabular}{|l|c|c|c|}
-    \toprule
-    \thead{Library} & \thead{Total\\Accepted} & \thead{Total\\Rejected} & \thead{Accept Rate}\\
-    \midrule
-    linear-smc & 19438 & 4 & 99.98\% \\
-    \makecell{priority-sesh} & 6781 & 19 & 99.72\% \\
-    linear-base & 112311 & 538 & 99.52\% \\
-    \bottomrule
+\begin{tabular}{|l|c|c|c|c|}
+    %\toprule
+    \thead{Library} & \thead{Total\\Accepted} & \thead{Total\\Rejected} & \thead{Unique\\Rejected} & \thead{Accept Rate}\\
+    %\midrule
+    linear-base & 48018 & 735 & 51 & 98.50\% \\
+    linear-smc & 18759 & 5 & 3 & 99.97\% \\
+    priority-sesh & 760 & 95 & 6 & 88.89\% \\
+    text-builder-linear & 4858 & 34 & 6 & 99.30\% \\
+    linear-generics & 63332 & 0 & 0 & 100.00\% \\
+    %\bottomrule
 \end{tabular}
 }
-\caption{Linear Core Plugin on Linear Libraries}
+\caption{Linear Core Plugin on Linear Libraries with GHC 9.10.3}
 \label{fig:core-plugin-res}

### 2025-11-25 popl26.tex
CR: Remove frames and improve and use even spacing on type system figure with newfound space
Fix colorbox spacing; kill 1 orphan

@@
 \usepackage{xargs}
-\usepackage{framed}
+\usepackage{framed} % can't use!
 \usepackage[framemethod=default]{mdframed}
@@
 %
+\begingroup
+\setlength{\fboxsep}{0pt}% remove the weird padding in colorbox until the end of this group
 In the examples, a \colorbox{working}{\workingcolorname} background highlights
@@
 \colorbox{limitation}{\limitationcolorname} background mark programs that are
-semantically linear, but not seen as linear by Core's type system. \ROUNDTWO{Our
-system accepts all
-\colorbox{notyet}{\notyetcolorname} programs, but not the
-\colorbox{limitation}{\limitationcolorname} examples. }
+semantically linear, but not seen as linear by Core's type system.
+\ROUNDTWO{Our system accepts all \colorbox{notyet}{\notyetcolorname} programs,
+but not the \colorbox{limitation}{\limitationcolorname} examples.}
 %
@@
 discards or duplicates linear resources.
+\endgroup% end group
 
@@
 and multiplicity polymorphism introduced by Linear
-Haskell~\cite{10.1145/3158093} with all the key features of GHC's Core
-language (algebraic datatypes, case expressions, and recursive lazy let bindings),
-except for equality coercions, which we discuss as future work in
-Section~\ref{sec:future-work}. 
+Haskell~\cite{10.1145/3158093} with all the key features of GHC Core: algebraic
+datatypes, case expressions, and recursive lazy let bindings; except for equality
+coercions, which we discuss as future work in Section~\ref{sec:future-work}. 
 %

### 2025-11-25 popl26.tex
More acknowledgments

@@
 
-  This
-  work was supported by national funds through Fundação para a Ciência e a Tecnologia, I.P. (FCT) under projects UID/50021/2025 and UID/PRR/50021/2025.
+We would like to thank Arnaud Spiwack for many insightful discussions on
+linearity in Core and the reviewers for their feedback, which improved the
+paper significantly. This work was supported by Well-Typed LLP and by
+national funds through Fundação para a Ciência e a Tecnologia, I.P. (FCT)
+under projects UID/50021/2025 and UID/PRR/50021/2025.
 

### 2025-11-25 popl26.tex
Camera Ready

@@
 \ROUNDTWO{In our implementation,} we thread input/output linear resources through each
-derivation~\cite{DBLP:journals/tcs/CervesatoHP00}.
+derivation~\cite{CERVESATO2000133}.
 %
@@
 \texttt{priority-sesh}~\cite{10.1145/3471874.3472979} (1400 lines),
-\texttt{text-builder-linear}~\cite{cite:text-builder-linear} (1800 lines), and
+\texttt{text-builder-linear} \cite{cite:text-builder-linear} (1800 lines), and
 \texttt{linear-generics}~\cite{cite:linear-generics} (2000 lines), using our
@@
 each top-level binding in a module counts as a program, and every such
-binding is checked once per optimisation pass, i.e. we check
-all intermediate programs produced by GHC. The total of programs
-rejected is counted in the same way, without ever halting compilation when a
-program is rejected. We note that our implementation is not performance
+binding is checked once per optimisation pass, i.e.~we check
+all intermediate programs produced by GHC. The total of rejected programs
+is counted in the same way, without halting compilation. We note that our implementation is not performance
 conscious and types every intermediate program from scratch.
